### 2015-12-30 code/Rmd/MSSM_Array.Rmd
Preliminary commit of RNASeq and microarray curation from MSBB

---
title: "Differential expression analysis of MSBB microarray data"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./MSSM_Array.Rmd",
                                 parentId ="syn5570248",
                                 entityName = 'MSBB Microarray')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE}
## It is assumed your working directory is where this file

### Clear R console screen output
cat("\014")

## Load required libraries
library(synapseClient)
library(ggplot2)
library(reshape2)
library(data.table)
library(dplyr)
library(stringr)
library(limma)
library(knitr)
library(Biobase)
library(GEOquery)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

## Needs the dev branch
library(rGithubClient)

synapseLogin()

# source utility files from ../R/lib folder
file.sources = list.files('../R/lib',pattern="*.R", full.names=T)
tmp = sapply(file.sources,source,.GlobalEnv)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn3721177';
activityName = 'Differential expression analysis';
activityDescription = 'Differential expression analysis based on image features';

thisFileName <- 'DiffExpAnal_MSSM_MA.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Takeda_contract", ref="branch", refName='master')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('Rmd/',thisFileName))
```
### Image Analysis
1. Image features used:
  * FreeSurfer.thickness..median  
  * mean.curvature..median         
  * Volume
  * travel.depth..median  
  * FreeSurfer.convexity..sulc...median
```{r load.data}
adni.mem.1_5T.3Y <- read.csv('./adni.mem_ranked_1.5T_3Yr_takeda_4_6_15.csv')
label.name.conversion <- read.csv('./label_names_conversion.csv')

adni.mem.1_5T.3Y <- dplyr::mutate(adni.mem.1_5T.3Y, 
                                  sig.feature.all = a.ks.pvals <= 0.05 & a.cor.pvals <= 0.05,
                                  sig.feature.aks = a.ks.pvals <= 0.05,
                                  feature = sapply(str_split(adni.mem.1_5T.3Y$var.nms,'\\.'),function(x){paste(x[-(length(x))],collapse='.')}),
                                  region = sapply(str_split(adni.mem.1_5T.3Y$var.nms,'\\.'),function(x){paste(x[(length(x))],collapse='.')}))
adni.mem.1_5T.3Y <- merge(adni.mem.1_5T.3Y,label.name.conversion, by.x = 'region', by.y = 'Label')
sig.table.all <- xtabs(sig.feature.all ~ feature + region,data = adni.mem.1_5T.3Y)
colnames(sig.table.all) <- label.name.conversion[match(label.name.conversion$Label,colnames(sig.table.all)),1]

unique.features <- unique(sapply(str_split(rownames(sig.table.all),'\\.'),function(x){x <- x[x!=""];l <- length(x);ifelse(l>1,paste(x[-(l)],collapse='.'),x)}))

region.ranking <- colSums(sig.table.all[rownames(sig.table.all) %in% c('FreeSurfer.thickness..median','mean.curvature..median','Volume','travel.depth..median','FreeSurfer.convexity..sulc...median'),])
region.ranking <- sort(region.ranking,decreasing = T)
region.ranking <- region.ranking[region.ranking!=0]

unique.regions.correlated <- unique(sapply(str_split(names(region.ranking),'-'),function(x){x[[length(x)]]}))

sig.table.aks <- xtabs(sig.feature.aks ~ feature + Label.name, data = adni.mem.1_5T.3Y)
region.ranking <- colSums(sig.table.aks[rownames(sig.table.aks) %in% c('FreeSurfer.thickness..median','mean.curvature..median','Volume','travel.depth..median','FreeSurfer.convexity..sulc...median'),])
region.ranking <- sort(region.ranking,decreasing = T)
region.ranking <- region.ranking[region.ranking!=0]
unique.regions <- unique(sapply(str_split(names(region.ranking),'-'),function(x){x[[length(x)]]}))

```
2. Significantly changed regions are: `r unique.regions`
3. Significantly changed regions that are also correlated with ADNI.MEM are: `r unique.regions.correlated`
4. Following brain regions were analysed for differential expression between case and control

| Brain Region                        | Synapse ID |
| ----------------------------------- | ---------- |
| middle temporal                     | syn3191101 |
| superior temporal                   | syn3191103 |
| inferior temporal                   | syn3191099 |
| precuneus superior parietal lobule  | syn3191119 |
| caudal anterior cingulate           | syn3191107 |
| frontal pole                        | syn3191095 |
| para hippocampal gyrus              | syn3191109 |

```{r differential.expression}
# Function to calculate differnetial expression
differential.expression <- function(EXPR_ID, COV_ID){

  # Download expression data
  EXPR_OBJ <- synGet(EXP_ID)
  EXPR <- read.table(EXPR_OBJ@filePath,header=T,sep='\t',check.names = F)
  EXPR <- EXPR[EXPR$Gene.Symbol!='',]
  rownames(EXPR) <- EXPR$ID
  ANNOTATIONS <- dplyr::select(EXPR,ID,GB_ACC,Gene.Symbol,ENTREZ_GENE_ID)
  EXPR <- dplyr::select(EXPR,-c(ID,GB_ACC,Gene.Symbol,ENTREZ_GENE_ID))

  Collapsed.obj <- collapseRows(EXPR,ANNOTATIONS$Gene.Symbol,rownames(EXPR))
  EXPR <- Collapsed.obj$datETcollapsed

  # Download covariates data
  COV_OBJ <- synGet(COV_ID)
  COV <- fread(COV_OBJ@filePath,sep='\t',header=T,data.table=F)
  rownames(COV) <- COV$BrainBank
  COV <- dplyr::select(COV,-(BrainBank))
  COV <- dplyr::mutate(COV,Dx = ifelse(CDR<1,'Control','AD'))

  # Get covariates classification
  FactorCovariates <- c('Sex','Race','CDR','Braak','NP1','Dx')
  ContCovariates <- c('PMI','pH','Age','PLQ_Mn','NPrSum','NTrSum')

  # Match expression to covariates data
  ind <- match(colnames(EXPR),rownames(COV))
  EXPR <- EXPR[,!is.na(ind)]
  COV <- COV[ind[!is.na(ind)],]

  # Convert covariates to factors and numeric variables accordingly
  COV[,FactorCovariates] <- lapply(COV[,FactorCovariates], factor)  
  COV[,ContCovariates] <- lapply(COV[,ContCovariates], as.numeric)

  # Find PC of gene expression and significant covariates that are highly correlated with PCs
  preAdjustedSigCovars <- runPCAandPlotCorrelations(EXPR, COV,'expression', isKeyPlot=TRUE)

  # Plot significant data
  print(preAdjustedSigCovars$PC_res[[2]]$plotData)
  print(paste('Significant covariates are:',paste(preAdjustedSigCovars$significantCovars,collapse=',')))

  # Get design matrix
  DESIGN <- getDesignMatrix(COV[,'Dx',drop=F], Intercept = F)
  print(paste('Fitting linear model with the following coefficients:',paste(colnames(DESIGN$design),collapse=',')))

#   # Calculate correlation between donors
#   correlation <- duplicateCorrelation(EXPR, design=DESIGN$design, block=ANNOTATIONS$Gene.Symbol)
#   print(paste('Consensus correlation is', correlation$consensus.correlation))
#   
  # Fit linear model
  FIT <- lmFit(EXPR, design = DESIGN$design)

  # Calculate residuals matrix
  RESIDUAL <- residuals.MArrayLM(FIT,EXPR)

  # Find correlation between residuals and covariates 
  postAdjustedSigCovars <- runPCAandPlotCorrelations(RESIDUAL, COV,'residual expression', isKeyPlot=TRUE)

  if (length(postAdjustedSigCovars$significantCovars) != 0)
    print(paste('Following coefficients have to be included in the model:',paste(postAdjustedSigCovars$significantCovars,collapse=',')))

  # Make contrast to check differential expression
  CONT.NAMES <- expand.grid(colnames(DESIGN$design)[2],colnames(DESIGN$design)[1])
  CONT.NAMES <- apply(CONT.NAMES,1,paste,collapse ='-')

  CONT <- makeContrasts(contrasts=CONT.NAMES,levels=colnames(FIT$coefficients))

  # Refit contrasts
  FIT.CONTRAST <- contrasts.fit(FIT,CONT)

  # Estimate moderated t-statistics
  FIT.CONTRAST <- eBayes(FIT.CONTRAST)

  # Obtain all the differential expession combinations
  DEXP <- list()
  DEXP$logFC <- data.frame(row.names = rownames(EXPR))
  DEXP$adj.P.Val <- data.frame(row.names = rownames(EXPR))

  for (i in colnames(CONT)){
    tmp <- topTable(FIT.CONTRAST, coef=i, number=dim(EXPR)[1])    
    DEXP$logFC[,i] <- tmp[rownames(DEXP$logFC),'logFC']
    DEXP$adj.P.Val[,i] <- tmp[rownames(DEXP$adj.P.Val),'adj.P.Val']
  }

  # Get number of significantly differentialy expressed genes with adj.P.Val <= 0.05 and abs(logFC) >= 0
  DEXP$SIG.EXP <- DEXP$adj.P.Val<=0.05 #& abs(DEXP$logFC) >= 0   
  DEXP$NUM.SIG.EXP <- colSums(DEXP$SIG.EXP)

  return(list(FIT = FIT, FIT.CONTRAST = FIT.CONTRAST, DEXP = DEXP, ANNOTATIONS = ANNOTATIONS))
}
```
### Middle Temporal Analysis
```{r middle.temporal, cache=TRUE, fig.height=10, fig.width=15}
# SynIDs of expression and covariates matrix
EXP_ID <- 'syn3191101'
COV_ID <- 'syn3205399'  
ALL_USED_IDs <- unique(c(EXP_ID,COV_ID))

# Obtain differential expession
DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)

# Print results
tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
kable(tmp)

# Get fold change and pvalues
FC <- DIFF.EXP$DEXP$logFC
colnames(FC) <- paste('MSSM.MA.MiddleTemporal',colnames(FC),sep='.')
FC <- rownameToFirstColumn(FC,'hgnc_symbol')

PVAL <- DIFF.EXP$DEXP$adj.P.Val
colnames(PVAL) <- paste('MSSM.MA.MiddleTemporal',colnames(PVAL),sep='.')
PVAL <- rownameToFirstColumn(PVAL,'hgnc_symbol')
```
### Superior Temporal Analysis 
```{r superior.temporal, cache=TRUE, fig.height=10, fig.width=15}
# SynIDs of expression and covariates matrix
EXP_ID <- 'syn3191103'
ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))

# Obtain differential expession
DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)

# Print results
tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
kable(tmp)

# Get fold change and pvalues
tmp <- DIFF.EXP$DEXP$logFC
colnames(tmp) <- paste('MSSM.MA.SuperiorTemporal',colnames(tmp),sep='.')
FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')

tmp <- DIFF.EXP$DEXP$adj.P.Val
colnames(tmp) <- paste('MSSM.MA.SuperiorTemporal',colnames(tmp),sep='.')
PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
```
### Inferior Temporal Analysis 
```{r inferior.temporal, cache=TRUE, fig.height=10, fig.width=15}
# SynIDs of expression and covariates matrix
EXP_ID <- 'syn3191099'
ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))

# Obtain differential expession
DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)

# Print results
tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
kable(tmp)

# Get fold change and pvalues
tmp <- DIFF.EXP$DEXP$logFC
colnames(tmp) <- paste('MSSM.MA.InferiorTemporal',colnames(tmp),sep='.')
FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')

tmp <- DIFF.EXP$DEXP$adj.P.Val
colnames(tmp) <- paste('MSSM.MA.InferiorTemporal',colnames(tmp),sep='.')
PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
```
### Precuneus Superior Parietal Lobule Analysis
```{r superior.parietal, cache=TRUE, fig.height=10, fig.width=15}
# SynIDs of expression and covariates matrix
EXP_ID <- 'syn3191119'
ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))

# Obtain differential expession
DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)

# Print results
tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
kable(tmp)

# Get fold change and pvalues
tmp <- DIFF.EXP$DEXP$logFC
colnames(tmp) <- paste('MSSM.MA.PrecuneusSuperiorParietal',colnames(tmp),sep='.')
FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')

tmp <- DIFF.EXP$DEXP$adj.P.Val
colnames(tmp) <- paste('MSSM.MA.PrecuneusSuperiorParietal',colnames(tmp),sep='.')
PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
```
### Caudal Anterior Cingulate Analysis
```{r anterior.cingulate, cache=TRUE, fig.height=10, fig.width=15}
# SynIDs of expression and covariates matrix
EXP_ID <- 'syn3191107'
ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))

# Obtain differential expession
DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)

# Print results
tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
kable(tmp)

# Get fold change and pvalues
tmp <- DIFF.EXP$DEXP$logFC
colnames(tmp) <- paste('MSSM.MA.CaudalAnteriorCingulate',colnames(tmp),sep='.')
FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')

tmp <- DIFF.EXP$DEXP$adj.P.Val
colnames(tmp) <- paste('MSSM.MA.CaudalAnteriorCingulate',colnames(tmp),sep='.')
PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
```
### Hippocampal Analysis
Hippocampal region shows shape changes but not correltated with ADNIMEM
```{r hippocampal, cache=TRUE, fig.height=10, fig.width=15}
# Print significant features
print('Significantly changed image features of hippocampal region are:')
tmp <- filter(adni.mem.1_5T.3Y, a.ks.pvals <= 0.05 & Label.name %in% grep('hippocampal',adni.mem.1_5T.3Y$Label.name,value=T))
kable(tmp)

# SynIDs of expression and covariates matrix
EXP_ID <- 'syn3191109'
ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))

# Obtain differential expession
DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)

# Print results
tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
kable(tmp)

# Get fold change and pvalues
tmp <- DIFF.EXP$DEXP$logFC
colnames(tmp) <- paste('MSSM.MA.Hippocampal',colnames(tmp),sep='.')
FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')

tmp <- DIFF.EXP$DEXP$adj.P.Val
colnames(tmp) <- paste('MSSM.MA.Hippocampal',colnames(tmp),sep='.')
PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
```
### Frontal Pole Analysis
Frontal pole analysis was performed (since this is one of the regions for which RNASeq is available from MSSM)
```{r frontal.pole, cache=TRUE, fig.height=10, fig.width=15}
# SynIDs of expression and covariates matrix
EXP_ID <- 'syn3191107'
ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))

# Obtain differential expession
DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)

# Print results
tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
kable(tmp)

# Get fold change and pvalues
tmp <- DIFF.EXP$DEXP$logFC
colnames(tmp) <- paste('MSSM.MA.FrontalPole',colnames(tmp),sep='.')
FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')

tmp <- DIFF.EXP$DEXP$adj.P.Val
colnames(tmp) <- paste('MSSM.MA.FrontalPole',colnames(tmp),sep='.')
PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE}
# Code
CODE <- File(paste0('./',thisFileName), name = "Differential Expression Analysis MSSM MA", parentId = parentId)
CODE <- synStore(CODE, used = ALL_USED_IDs, activityName = activityName, executed = thisFile, activityDescription = activityDescription)

# Fold change
FC <- FC[, unique(colnames(FC))]
write.table(FC, file='./FoldChange_MSSM_MA.tsv', sep='\t', row.names=F, quote=F)
FC_OBJ <- File('./FoldChange_MSSM_MA.tsv',name = 'Differential Expression Analysis MSSM MA logFC',parentId = parentId)
FC_OBJ <- synStore(FC_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=CODE$properties$id, activityDescription = activityDescription)

# Pvalue
PVAL <- PVAL[, unique(colnames(PVAL))]
write.table(PVAL, file='./AdjustedPval_MSSM_MA.tsv', sep='\t', row.names=F, quote=F)
PVAL_OBJ <- File('./AdjustedPval_MSSM_MA.tsv',name = 'Differential Expression Analysis MSSM MA adjPval',parentId = parentId)
PVAL_OBJ <- synStore(PVAL_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=CODE$properties$id, activityDescription = activityDescription)
```
|  *Results*           |  *SynapseID*                |
|  -------             |   ---------                 |
|  Log fold change     |  `r FC_OBJ$properties$id`       |
|  Adjusted pvalues    |  `r PVAL_OBJ$properties$id`     |

### 2015-12-30 code/Rmd/MSSM_RNASeq.Rmd
Preliminary commit of RNASeq and microarray curation from MSBB

---
title: "Covariates analysis of MSBB RNASeq data"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq.Rmd",
                                 parentId ="syn5570248",
                                 entityName = 'MSBB RNASeq')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

### Clear R console screen output
cat("\014")

## Load required libraries
library(synapseClient)
library(rGithubClient) ## Needs the dev branch
library(ggplot2)
library(reshape2)
library(data.table)
library(plyr)
library(dplyr)
library(tidyr)
library(stringr)
library(knitr)
library(limma)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

synapseLogin()

# source utility files from ../R/lib folder
file.sources = list.files('../R/lib',pattern="*.R", full.names=T)
tmp = sapply(file.sources,source,.GlobalEnv)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570248';
activityName = 'Differential expression analysis';
activityDescription = 'Differential expression analysis of MSBB RNASeq samples';

thisFileName <- 'MSSM_RNASeq.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('Rmd/',thisFileName))
```
Data were obtained from AMP-AD portal of MSSM, and following brain regions were analysed for differential expression between case and control:
  | Region | Name                      | 
  | :----- | ------------------------- | 
  | BM_10  | Frontal Pole (FP)         |
  | BM_22  | Superior Temporal Gyrus   |
  | BM_36  | Parahipocampal Gyrus (PHG)|

* BM 10: frontopolar prefrontal cortex, rostrolateral prefrontal cortex, or anterior prefrontal cortex, 
* BM 22: In humans it corresponds approximately to the lateral and caudal two thirds of the superior temporal gyrus, 
* BM 36: Ectorhinal area 36 is a subdivision of the cytoarchitecturally defined temporal region of cerebral cortex.

Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_ID <- 'syn3205353'
ALL_USED_IDs <- COUNT_ID
COUNT_OBJ <- synGet(COUNT_ID)
COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F, row.names = 1)

# Get metadata
METADATA_ID <- 'syn3205337'
METADATA_OBJ <- synGet(METADATA_ID)
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA <- read.table(METADATA_OBJ@filePath,sep='\t',header=T, row.names=1)
METADATA <- cbind(METADATA, data.frame(Dx = ifelse(METADATA$CDR<1,'Control','AD')))
```
### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

Remove genes that have less than 1 cpm counts in at least 50% of samples.
```{r cpmnormalisation, cache=TRUE}
ind <- intersect(rownames(METADATA), colnames(COUNT))
METADATA <- METADATA[ind,]
COUNT <- COUNT[, ind]
PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT,MIN_GENE_CPM=1, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
```
### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering, cache=TRUE}
FactorCovariates <- c('TISSUE','Donor','RACE','CDR','CERJ','SEX','NP.1','bbscore','Dx')
ContCovariates <- c('PMI','AOD','PlaqueMean')

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- rownames(METADATA)

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)

COVARIATES <- dplyr::mutate(COVARIATES, TISSUE.Dx = paste(TISSUE,Dx,sep='.'))
COVARIATES$TISSUE.Dx <- factor(COVARIATES$TISSUE.Dx)
```
Covariate correlation
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 1)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.3, row.width=0.15)
```
### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T)

# Initial design matrix with covariates under consideration
DM = getDesignMatrix(COVARIATES, Intercept = F)
```
### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
DESIGN = getDesignMat2Fact(DM$design,FactorCovariates)
rownames(DESIGN) = rownames(METADATA)
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, DESIGN,'NULL design(voom-normalized)', isKeyPlot=TRUE)

# Find significant covariates
adjustCovars = designMatVarsToCovars(getCovariatesMap(DM),preAdjustedSigCovars$significantCovars)
```
Significant covariates to adjust at FDR 0.1 are `r adjustCovars`
```{r preAdjustedSigCovars.NULL, fig.width=15, fig.height=8}
preAdjustedSigCovars[["PC_res"]][[1]]$plotData
```
### Normalisation (custom/manual design)
Re-perform normalisation with custom design matrix (manually with "TISSUE.Dx","SEX","RACE","PMI" as fixed effects and Donor as random effects)
```{r reperform.normalisation, cache=TRUE}
# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,c("TISSUE.Dx","SEX","RACE","PMI"),drop=F],Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

# Estimate correlation of random effects
VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)
correlation <- duplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION$E, design=DM1$design, block=COVARIATES$Donor)

# Re-calculate voom weights with correlation of random effects
VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F,
                                     block=COVARIATES$Donor, correlation = correlation$cor)

# Re-calculate linear fit using new weights and new design
VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION$E, design=DM1$design, weights = VOOM.ADJUSTED.GENE_EXPRESSION$weights,
                          block=COVARIATES$Donor, correlation = correlation$cor)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

# Add variable of interest back to the residuals
RESIDUAL.GENE_EXPRESSION =

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
rownames(DESIGN) <- rownames(METADATA)
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION,
                                              dplyr::select(DESIGN,-starts_with('Donor')),
                                              'all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

```{r residualSigCovars.manual, fig.width=15, fig.height=8}
residualSigCovars[["PC_res"]][[1]]$plotData
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE}
# Code
CODE <- File(paste0('./',thisFileName), name = "Differential Expression Analysis MSSM RNASeq", parentId = parentId)
CODE <- synStore(CODE, used = ALL_USED_IDs, activityName = activityName, executed = thisFile, activityDescription = activityDescription)

# Fold change
FC <- FC[, unique(colnames(FC))]
write.table(FC, file='./FoldChange_MSSM_RNASeq.tsv', sep='\t', row.names=F, quote=F)
FC_OBJ <- File('./FoldChange_MSSM_RNASeq.tsv',name = 'Differential Expression Analysis MSSM RNASeq logFC',parentId = parentId)
FC_OBJ <- synStore(FC_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=CODE$properties$id, activityDescription = activityDescription)

# Pvalue
PVAL <- PVAL[, unique(colnames(PVAL))]
write.table(PVAL, file='./AdjustedPval_MSSM_RNASeq.tsv', sep='\t', row.names=F, quote=F)
PVAL_OBJ <- File('./AdjustedPval_MSSM_RNASeq.tsv',name = 'Differential Expression Analysis MSSM RNASeq adjPval',parentId = parentId)
PVAL_OBJ <- synStore(PVAL_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=CODE$properties$id, activityDescription = activityDescription)
```
|  *Results*           |  *SynapseID*                    |
  |  -------             |   ---------                     |
  |  Log fold change     |  `r FC_OBJ$properties$id`       |
  |  Adjusted pvalues    |  `r PVAL_OBJ$properties$id`     |

### 2015-12-31 code/Rmd/MSSM_Array.Rmd
Major modifications in the MSSM microarray Rmd file

@@
 ---
-title: "Differential expression analysis of MSBB microarray data"
+title: "Covariate analysis of MSBB microarray data"
 author: "Thanneer Perumal"
@@
 # Synapse parameters
-parentId = 'syn3721177';
-activityName = 'Differential expression analysis';
-activityDescription = 'Differential expression analysis based on image features';
+parentId = 'syn5570248';
+activityName = 'Covariate analysis';
+activityDescription = 'Covariate analysis of MSBB microarrays';
 
-thisFileName <- 'DiffExpAnal_MSSM_MA.Rmd'
+thisFileName <- 'MSSM_Array.Rmd'
 
 # Github link
-thisRepo <- getRepo(repository = "th1vairam/Takeda_contract", ref="branch", refName='master')
-thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('Rmd/',thisFileName))
+thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
+thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
 ```
-### Image Analysis
-1. Image features used:
-  * FreeSurfer.thickness..median  
-  * mean.curvature..median         
-  * Volume
-  * travel.depth..median  
-  * FreeSurfer.convexity..sulc...median
-```{r load.data}
-adni.mem.1_5T.3Y <- read.csv('./adni.mem_ranked_1.5T_3Yr_takeda_4_6_15.csv')
-label.name.conversion <- read.csv('./label_names_conversion.csv')
-
-adni.mem.1_5T.3Y <- dplyr::mutate(adni.mem.1_5T.3Y, 
-                                  sig.feature.all = a.ks.pvals <= 0.05 & a.cor.pvals <= 0.05,
-                                  sig.feature.aks = a.ks.pvals <= 0.05,
-                                  feature = sapply(str_split(adni.mem.1_5T.3Y$var.nms,'\\.'),function(x){paste(x[-(length(x))],collapse='.')}),
-                                  region = sapply(str_split(adni.mem.1_5T.3Y$var.nms,'\\.'),function(x){paste(x[(length(x))],collapse='.')}))
-adni.mem.1_5T.3Y <- merge(adni.mem.1_5T.3Y,label.name.conversion, by.x = 'region', by.y = 'Label')
-sig.table.all <- xtabs(sig.feature.all ~ feature + region,data = adni.mem.1_5T.3Y)
-colnames(sig.table.all) <- label.name.conversion[match(label.name.conversion$Label,colnames(sig.table.all)),1]
-
-unique.features <- unique(sapply(str_split(rownames(sig.table.all),'\\.'),function(x){x <- x[x!=""];l <- length(x);ifelse(l>1,paste(x[-(l)],collapse='.'),x)}))
-
-region.ranking <- colSums(sig.table.all[rownames(sig.table.all) %in% c('FreeSurfer.thickness..median','mean.curvature..median','Volume','travel.depth..median','FreeSurfer.convexity..sulc...median'),])
-region.ranking <- sort(region.ranking,decreasing = T)
-region.ranking <- region.ranking[region.ranking!=0]
-
-unique.regions.correlated <- unique(sapply(str_split(names(region.ranking),'-'),function(x){x[[length(x)]]}))
-
-sig.table.aks <- xtabs(sig.feature.aks ~ feature + Label.name, data = adni.mem.1_5T.3Y)
-region.ranking <- colSums(sig.table.aks[rownames(sig.table.aks) %in% c('FreeSurfer.thickness..median','mean.curvature..median','Volume','travel.depth..median','FreeSurfer.convexity..sulc...median'),])
-region.ranking <- sort(region.ranking,decreasing = T)
-region.ranking <- region.ranking[region.ranking!=0]
-unique.regions <- unique(sapply(str_split(names(region.ranking),'-'),function(x){x[[length(x)]]}))
-
-```
-2. Significantly changed regions are: `r unique.regions`
-3. Significantly changed regions that are also correlated with ADNI.MEM are: `r unique.regions.correlated`
-4. Following brain regions were analysed for differential expression between case and control
+Following brain regions were analysed for differential expression between case and control
 
@@
 
-```{r differential.expression}
-# Function to calculate differnetial expression
-differential.expression <- function(EXPR_ID, COV_ID){
+```{r covariate analysis}
+# Function to do covariates adjustments
+covariatesAdjustments <- function(exprId, covId){
+  # Download expression data
+  expr <- read.table(synGet(exprId)@filePath, header=T, sep='\t', check.names = F)
+  expr <- expr[expr$Gene.Symbol!='',]
+  rownames(expr) <- expr$ID
+  ANNOTATIONS <- dplyr::select(expr,ID,GB_ACC,Gene.Symbol,ENTREZ_GENE_ID)
+  expr <- dplyr::select(expr,-c(ID,GB_ACC,Gene.Symbol,ENTREZ_GENE_ID))
+  rownames(expr) = ANNOTATIONS$ID
 
-  # Download expression data
-  EXPR_OBJ <- synGet(EXP_ID)
-  EXPR <- read.table(EXPR_OBJ@filePath,header=T,sep='\t',check.names = F)
-  EXPR <- EXPR[EXPR$Gene.Symbol!='',]
-  rownames(EXPR) <- EXPR$ID
-  ANNOTATIONS <- dplyr::select(EXPR,ID,GB_ACC,Gene.Symbol,ENTREZ_GENE_ID)
-  EXPR <- dplyr::select(EXPR,-c(ID,GB_ACC,Gene.Symbol,ENTREZ_GENE_ID))
-
-  Collapsed.obj <- collapseRows(EXPR,ANNOTATIONS$Gene.Symbol,rownames(EXPR))
-  EXPR <- Collapsed.obj$datETcollapsed
+  Collapsed.obj <- collapseRows(expr,ANNOTATIONS$Gene.Symbol,ANNOTATIONS$ID)
+  expr <- Collapsed.obj$datETcollapsed
 
   # Download covariates data
-  COV_OBJ <- synGet(COV_ID)
-  COV <- fread(COV_OBJ@filePath,sep='\t',header=T,data.table=F)
-  rownames(COV) <- COV$BrainBank
-  COV <- dplyr::select(COV,-(BrainBank))
-  COV <- dplyr::mutate(COV,Dx = ifelse(CDR<1,'Control','AD'))
+  covar <- fread(synGet(covId)@filePath,sep='\t',header=T,data.table=F)
+  BrainBank <- covar$BrainBank
+  covar <- dplyr::select(covar,-(BrainBank))
+  covar <- dplyr::mutate(covar,Dx = ifelse(CDR<1,'Control','AD'))
+  rownames(covar) <- BrainBank
 
@@
   # Match expression to covariates data
-  ind <- match(colnames(EXPR),rownames(COV))
-  EXPR <- EXPR[,!is.na(ind)]
-  COV <- COV[ind[!is.na(ind)],]
+  ind <- match(colnames(expr),rownames(covar))
+  expr <- expr[,!is.na(ind)]
+  covar <- covar[ind[!is.na(ind)],]
 
   # Convert covariates to factors and numeric variables accordingly
-  COV[,FactorCovariates] <- lapply(COV[,FactorCovariates], factor)  
-  COV[,ContCovariates] <- lapply(COV[,ContCovariates], as.numeric)
+  covar$Age = gsub('\\+','',covar$Age)
+  covar[,FactorCovariates] <- lapply(covar[,FactorCovariates], as.factor)
+  covar[,ContCovariates] <- lapply(covar[,ContCovariates], as.numeric)
 
   # Find PC of gene expression and significant covariates that are highly correlated with PCs
-  preAdjustedSigCovars <- runPCAandPlotCorrelations(EXPR, COV,'expression', isKeyPlot=TRUE)
+  covar.tmp = covar
+  covar.tmp[,FactorCovariates] <- lapply(covar.tmp[,FactorCovariates], as.numeric)
+  preAdjustedSigCovars <- runPCAandPlotCorrelations(expr, covar.tmp, 'expression', isKeyPlot=TRUE)
 
@@
   # Get design matrix
-  DESIGN <- getDesignMatrix(COV[,'Dx',drop=F], Intercept = F)
+  DESIGN <- getDesignMatrix(covar[,'Dx',drop=F], Intercept = F)
   print(paste('Fitting linear model with the following coefficients:',paste(colnames(DESIGN$design),collapse=',')))
 
-#   # Calculate correlation between donors
-#   correlation <- duplicateCorrelation(EXPR, design=DESIGN$design, block=ANNOTATIONS$Gene.Symbol)
-#   print(paste('Consensus correlation is', correlation$consensus.correlation))
-#   
   # Fit linear model
-  FIT <- lmFit(EXPR, design = DESIGN$design)
+  fit <- lmFit(expr, design = DESIGN$design)
 
   # Calculate residuals matrix
-  RESIDUAL <- residuals.MArrayLM(FIT,EXPR)
+  residualFit <- residuals.MArrayLM(fit, expr)
+  residualFit = residualFit + fit$coefficients[,c('DxAD','DxControl')] %*% t(DESIGN$design[,c('DxAD','DxControl')])
 
   # Find correlation between residuals and covariates 
-  postAdjustedSigCovars <- runPCAandPlotCorrelations(RESIDUAL, COV,'residual expression', isKeyPlot=TRUE)
+  postAdjustedSigCovars <- runPCAandPlotCorrelations(residualFit, covar.tmp,'residual expression', isKeyPlot=TRUE)
 
@@
 
-  # Make contrast to check differential expression
-  CONT.NAMES <- expand.grid(colnames(DESIGN$design)[2],colnames(DESIGN$design)[1])
-  CONT.NAMES <- apply(CONT.NAMES,1,paste,collapse ='-')
-
-  CONT <- makeContrasts(contrasts=CONT.NAMES,levels=colnames(FIT$coefficients))
-
-  # Refit contrasts
-  FIT.CONTRAST <- contrasts.fit(FIT,CONT)
-
-  # Estimate moderated t-statistics
-  FIT.CONTRAST <- eBayes(FIT.CONTRAST)
-
-  # Obtain all the differential expession combinations
-  DEXP <- list()
-  DEXP$logFC <- data.frame(row.names = rownames(EXPR))
-  DEXP$adj.P.Val <- data.frame(row.names = rownames(EXPR))
-
-  for (i in colnames(CONT)){
-    tmp <- topTable(FIT.CONTRAST, coef=i, number=dim(EXPR)[1])    
-    DEXP$logFC[,i] <- tmp[rownames(DEXP$logFC),'logFC']
-    DEXP$adj.P.Val[,i] <- tmp[rownames(DEXP$adj.P.Val),'adj.P.Val']
-  }
-
-  # Get number of significantly differentialy expressed genes with adj.P.Val <= 0.05 and abs(logFC) >= 0
-  DEXP$SIG.EXP <- DEXP$adj.P.Val<=0.05 #& abs(DEXP$logFC) >= 0   
-  DEXP$NUM.SIG.EXP <- colSums(DEXP$SIG.EXP)
-
-  return(list(FIT = FIT, FIT.CONTRAST = FIT.CONTRAST, DEXP = DEXP, ANNOTATIONS = ANNOTATIONS))
+  return(list(EXPR = expr,
+              COVAR = covar,
+              FIT = fit, 
+              RESIDUAL = residualFit, 
+              DESIGN = DESIGN, 
+              ANNOTATIONS = ANNOTATIONS))
 }
@@
 # SynIDs of expression and covariates matrix
-EXP_ID <- 'syn3191101'
+EXP_ID <- c('Middle Temporal' = 'syn3191101',
+            'Superior Temporal' = 'syn3191103',
+            'Inferior Temporal' = 'syn3191099',
+            'Superior Parietal Lobule' = 'syn3191119',
+            'Caudal Anterior Cingulate' = 'syn3191107',
+            'Parahippocampal Gyrus' = 'syn3191109',
+            'Frontal Pole' = 'syn3191107')
 COV_ID <- 'syn3205399'
-ALL_USED_IDs <- unique(c(EXP_ID,COV_ID))
 
-# Obtain differential expession
-DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)
-
-# Print results
-tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
-colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
-kable(tmp)
-
-# Get fold change and pvalues
-FC <- DIFF.EXP$DEXP$logFC
-colnames(FC) <- paste('MSSM.MA.MiddleTemporal',colnames(FC),sep='.')
-FC <- rownameToFirstColumn(FC,'hgnc_symbol')
-
-PVAL <- DIFF.EXP$DEXP$adj.P.Val
-colnames(PVAL) <- paste('MSSM.MA.MiddleTemporal',colnames(PVAL),sep='.')
-PVAL <- rownameToFirstColumn(PVAL,'hgnc_symbol')
-```
-### Superior Temporal Analysis 
-```{r superior.temporal, cache=TRUE, fig.height=10, fig.width=15}
-# SynIDs of expression and covariates matrix
-EXP_ID <- 'syn3191103'
-ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))
-
-# Obtain differential expession
-DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)
-
-# Print results
-tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
-colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
-kable(tmp)
-
-# Get fold change and pvalues
-tmp <- DIFF.EXP$DEXP$logFC
-colnames(tmp) <- paste('MSSM.MA.SuperiorTemporal',colnames(tmp),sep='.')
-FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
-
-tmp <- DIFF.EXP$DEXP$adj.P.Val
-colnames(tmp) <- paste('MSSM.MA.SuperiorTemporal',colnames(tmp),sep='.')
-PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
-```
-### Inferior Temporal Analysis 
-```{r inferior.temporal, cache=TRUE, fig.height=10, fig.width=15}
-# SynIDs of expression and covariates matrix
-EXP_ID <- 'syn3191099'
-ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))
-
-# Obtain differential expession
-DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)
-
-# Print results
-tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
-colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
-kable(tmp)
-
-# Get fold change and pvalues
-tmp <- DIFF.EXP$DEXP$logFC
-colnames(tmp) <- paste('MSSM.MA.InferiorTemporal',colnames(tmp),sep='.')
-FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
-
-tmp <- DIFF.EXP$DEXP$adj.P.Val
-colnames(tmp) <- paste('MSSM.MA.InferiorTemporal',colnames(tmp),sep='.')
-PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
-```
-### Precuneus Superior Parietal Lobule Analysis
-```{r superior.parietal, cache=TRUE, fig.height=10, fig.width=15}
-# SynIDs of expression and covariates matrix
-EXP_ID <- 'syn3191119'
-ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))
-
-# Obtain differential expession
-DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)
-
-# Print results
-tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
-colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
-kable(tmp)
-
-# Get fold change and pvalues
-tmp <- DIFF.EXP$DEXP$logFC
-colnames(tmp) <- paste('MSSM.MA.PrecuneusSuperiorParietal',colnames(tmp),sep='.')
-FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
-
-tmp <- DIFF.EXP$DEXP$adj.P.Val
-colnames(tmp) <- paste('MSSM.MA.PrecuneusSuperiorParietal',colnames(tmp),sep='.')
-PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
-```
-### Caudal Anterior Cingulate Analysis
-```{r anterior.cingulate, cache=TRUE, fig.height=10, fig.width=15}
-# SynIDs of expression and covariates matrix
-EXP_ID <- 'syn3191107'
-ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))
-
-# Obtain differential expession
-DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)
-
-# Print results
-tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
-colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
-kable(tmp)
-
-# Get fold change and pvalues
-tmp <- DIFF.EXP$DEXP$logFC
-colnames(tmp) <- paste('MSSM.MA.CaudalAnteriorCingulate',colnames(tmp),sep='.')
-FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
-
-tmp <- DIFF.EXP$DEXP$adj.P.Val
-colnames(tmp) <- paste('MSSM.MA.CaudalAnteriorCingulate',colnames(tmp),sep='.')
-PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
-```
-### Hippocampal Analysis
-Hippocampal region shows shape changes but not correltated with ADNIMEM
-```{r hippocampal, cache=TRUE, fig.height=10, fig.width=15}
-# Print significant features
-print('Significantly changed image features of hippocampal region are:')
-tmp <- filter(adni.mem.1_5T.3Y, a.ks.pvals <= 0.05 & Label.name %in% grep('hippocampal',adni.mem.1_5T.3Y$Label.name,value=T))
-kable(tmp)
-
-# SynIDs of expression and covariates matrix
-EXP_ID <- 'syn3191109'
-ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))
-
-# Obtain differential expession
-DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)
-
-# Print results
-tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
-colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
-kable(tmp)
-
-# Get fold change and pvalues
-tmp <- DIFF.EXP$DEXP$logFC
-colnames(tmp) <- paste('MSSM.MA.Hippocampal',colnames(tmp),sep='.')
-FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
-
-tmp <- DIFF.EXP$DEXP$adj.P.Val
-colnames(tmp) <- paste('MSSM.MA.Hippocampal',colnames(tmp),sep='.')
-PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
-```
-### Frontal Pole Analysis
-Frontal pole analysis was performed (since this is one of the regions for which RNASeq is available from MSSM)
-```{r frontal.pole, cache=TRUE, fig.height=10, fig.width=15}
-# SynIDs of expression and covariates matrix
-EXP_ID <- 'syn3191107'
-ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))
-
-# Obtain differential expession
-DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)
-
-# Print results
-tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
-colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
-kable(tmp)
-
-# Get fold change and pvalues
-tmp <- DIFF.EXP$DEXP$logFC
-colnames(tmp) <- paste('MSSM.MA.FrontalPole',colnames(tmp),sep='.')
-FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
-
-tmp <- DIFF.EXP$DEXP$adj.P.Val
-colnames(tmp) <- paste('MSSM.MA.FrontalPole',colnames(tmp),sep='.')
-PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
+# Obtain adjusted expession
+adjustedExpression <- lapply(EXP_ID, covariatesAdjustments, COV_ID)
 ```
@@
 # Code
-CODE <- File(paste0('./',thisFileName), name = "Differential Expression Analysis MSSM MA", parentId = parentId)
-CODE <- synStore(CODE, used = ALL_USED_IDs, activityName = activityName, executed = thisFile, activityDescription = activityDescription)
+CODE <- Folder(name = "MSBB Microarray", parentId ="syn5570248")
+CODE <- synStore(CODE)
 
-# Fold change
-FC <- FC[, unique(colnames(FC))]
-write.table(FC, file='./FoldChange_MSSM_MA.tsv', sep='\t', row.names=F, quote=F)
-FC_OBJ <- File('./FoldChange_MSSM_MA.tsv',name = 'Differential Expression Analysis MSSM MA logFC',parentId = parentId)
-FC_OBJ <- synStore(FC_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=CODE$properties$id, activityDescription = activityDescription)
+ALL.OBJ = mapply(function(exp, regionName, parentId, used, executed, activityName, activityDescription){
+  # Expression
+  expr = rownameToFirstColumn(exp$EXPR, 'hgnc_symbols')
+  write.table(expr, file='expression.tsv', sep = '\t', row.names=F, quote=F)
+  obj = File('expression.tsv', 
+             name = paste('Expression',regionName), 
+             parentId = parentId)
+  obj = synStore(obj,
+             used = used, 
+             activityName = activityName, 
+             executed = executed, 
+             activityDescription = activityDescription)
 
-# Pvalue
-PVAL <- PVAL[, unique(colnames(PVAL))]
-write.table(PVAL, file='./AdjustedPval_MSSM_MA.tsv', sep='\t', row.names=F, quote=F)
-PVAL_OBJ <- File('./AdjustedPval_MSSM_MA.tsv',name = 'Differential Expression Analysis MSSM MA adjPval',parentId = parentId)
-PVAL_OBJ <- synStore(PVAL_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=CODE$properties$id, activityDescription = activityDescription)
+  # Covariates
+  expr = rownameToFirstColumn(exp$COVAR, 'SampleId')
+  write.table(expr, file='covariates.tsv', sep = '\t', row.names=F, quote=F)
+  obj = File('covariates.tsv', 
+             name = paste('Covariates',regionName), 
+             parentId = parentId)
+  obj = synStore(obj,
+             used = used, 
+             activityName = activityName, 
+             executed = executed, 
+             activityDescription = activityDescription)
+},adjustedExpression, names(adjustedExpression), 
+MoreArgs = list(CODE$properties$id, ALL_USED_IDs, thisFile, activityName, activityDescription), 
+SIMPLIFY = F)       
 ```
-|  *Results*           |  *SynapseID*                |
-|  -------             |   ---------                 |
-|  Log fold change     |  `r FC_OBJ$properties$id`       |
-|  Adjusted pvalues    |  `r PVAL_OBJ$properties$id`     |

### 2015-12-31 code/Rmd/MSSM_Array.Rmd
Minor fix

@@
 ```
-### Middle Temporal Analysis
+### Extract adjust covariates
 ```{r middle.temporal, cache=TRUE, fig.height=10, fig.width=15}

### 2015-12-31 code/Rmd/MSSM_RNASeq.Rmd
Major modifications in MSBB RNASeq analysis

@@
 thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
-thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('Rmd/',thisFileName))
+thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
 ```
@@
 METADATA <- METADATA[ind,]
-COUNT <- COUNT[, ind]
+COUNT <- COUNT[1:1000, ind]
 PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT,MIN_GENE_CPM=1, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
@@
 
-# Add variable of interest back to the residuals
-RESIDUAL.GENE_EXPRESSION =
-
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
@@
 
+```{r varsToAddBack}
+# Add variable of interest back to the residuals
+varsToAddIn = c("TISSUE.DxBM_10.AD", "TISSUE.DxBM_10.Control", "TISSUE.DxBM_22.AD",
+                "TISSUE.DxBM_22.Control", "TISSUE.DxBM_36.AD","TISSUE.DxBM_36.Control")
+RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
+  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+```
+
 ### Store files in synapse
@@
 # Code
-CODE <- File(paste0('./',thisFileName), name = "Differential Expression Analysis MSSM RNASeq", parentId = parentId)
-CODE <- synStore(CODE, used = ALL_USED_IDs, activityName = activityName, executed = thisFile, activityDescription = activityDescription)
+CODE <- Folder(name = "MSBB RNASeq", parentId = parentId)
+CODE <- synStore(CODE)
 
-# Fold change
-FC <- FC[, unique(colnames(FC))]
-write.table(FC, file='./FoldChange_MSSM_RNASeq.tsv', sep='\t', row.names=F, quote=F)
-FC_OBJ <- File('./FoldChange_MSSM_RNASeq.tsv',name = 'Differential Expression Analysis MSSM RNASeq logFC',parentId = parentId)
-FC_OBJ <- synStore(FC_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=CODE$properties$id, activityDescription = activityDescription)
+# Expression
+expr = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'EnsembleIDs')
+write.table(expr, file='expression.tsv', sep = '\t', row.names=F, quote=F)
+obj = File('expression.tsv', 
+           name = 'Voom Adjusted Weighted Residual Expression', 
+           parentId = parentId)
+obj = synStore(obj,
+               used = ALL_USED_IDs, 
+               activityName = activityName, 
+               executed = thisFile, 
+               activityDescription = activityDescription)
 
-# Pvalue
-PVAL <- PVAL[, unique(colnames(PVAL))]
-write.table(PVAL, file='./AdjustedPval_MSSM_RNASeq.tsv', sep='\t', row.names=F, quote=F)
-PVAL_OBJ <- File('./AdjustedPval_MSSM_RNASeq.tsv',name = 'Differential Expression Analysis MSSM RNASeq adjPval',parentId = parentId)
-PVAL_OBJ <- synStore(PVAL_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=CODE$properties$id, activityDescription = activityDescription)
+# Covariates
+covar = rownameToFirstColumn(COVARIATES, 'SampleId')
+write.table(expr, file='covariates.tsv', sep = '\t', row.names=F, quote=F)
+obj = File('covariates.tsv', 
+           name = 'Covariates', 
+           parentId = parentId)
+obj = synStore(obj,
+               used = ALL_USED_IDs, 
+               activityName = activityName, 
+               executed = thisFile, 
+               activityDescription = activityDescription)
 ```
-|  *Results*           |  *SynapseID*                    |
-  |  -------             |   ---------                     |
-  |  Log fold change     |  `r FC_OBJ$properties$id`       |
-  |  Adjusted pvalues    |  `r PVAL_OBJ$properties$id`     |

### 2015-12-31 code/Rmd/MSSM_RNASeq.Rmd
Minor fix

@@
   tidy = FALSE,
-  cache = TRUE)
+  cache = FALSE)
 ```
@@
 parentId = 'syn5570248';
-activityName = 'Differential expression analysis';
-activityDescription = 'Differential expression analysis of MSBB RNASeq samples';
+activityName = 'Covariate analysis';
+activityDescription = 'Covariate analysis of MSBB RNASeq samples';
 
@@
 METADATA <- METADATA[ind,]
-COUNT <- COUNT[1:1000, ind]
+COUNT <- COUNT[, ind]
 PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT,MIN_GENE_CPM=1, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
@@
            name = 'Voom Adjusted Weighted Residual Expression', 
-           parentId = parentId)
+           parentId = CODE$properties$id)
 obj = synStore(obj,
@@
 covar = rownameToFirstColumn(COVARIATES, 'SampleId')
-write.table(expr, file='covariates.tsv', sep = '\t', row.names=F, quote=F)
+write.table(covar, file='covariates.tsv', sep = '\t', row.names=F, quote=F)
 obj = File('covariates.tsv', 
            name = 'Covariates', 
-           parentId = parentId)
+           parentId = CODE$properties$id)
 obj = synStore(obj,

### 2016-01-04 code/Rmd/MSSM_RNASeq.Rmd
Minor changes in MSSB RNASeq covariates code

@@
 COVARIATES$TISSUE.Dx <- factor(COVARIATES$TISSUE.Dx)
+rownames(COVARIATES) <- rownames(METADATA)
 ```
@@
 ```
-
+### Residual calculation
+Calculate weighted residuals and add back "TISSUE.Dx" to the residuals
 ```{r varsToAddBack}
@@
 ### Store files in synapse
-```{r synapse.store, include=FALSE, eval=TRUE}
+```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
 # Code
@@
 obj = File('expression.tsv', 
-           name = 'Voom Adjusted Weighted Residual Expression', 
+           name = 'Voom Adjusted Weighted Residual Expression (Tissue.Dx added)', 
            parentId = CODE$properties$id)

### 2016-01-04 code/Rmd/MSSM_Array.Rmd
Minor modification in MSSB Array covariates adjustments

--- 2015-12-31 code/Rmd/MSSM_Array.Rmd
@@
 
-```{r covariate analysis}
+```{r fxns}
 # Function to do covariates adjustments
-covariatesAdjustments <- function(exprId, covId){
+covariatesAdjustments <- function(exprId, regionName, covId){
+  inline_expr(paste0('Working on ',regionName), "md")
+
   # Download expression data
@@
 
+  # Collapse probe ids to hgnc symbols
   Collapsed.obj <- collapseRows(expr,ANNOTATIONS$Gene.Symbol,ANNOTATIONS$ID)
@@
   print(preAdjustedSigCovars$PC_res[[2]]$plotData)
-  print(paste('Significant covariates are:',paste(preAdjustedSigCovars$significantCovars,collapse=',')))
+  inline_expr(paste('Significant covariates are:',paste(preAdjustedSigCovars$significantCovars,collapse=',')), "md")
 
@@
   DESIGN <- getDesignMatrix(covar[,'Dx',drop=F], Intercept = F)
-  print(paste('Fitting linear model with the following coefficients:',paste(colnames(DESIGN$design),collapse=',')))
+  inline_expr(paste('Fitting linear model with the following coefficients:',paste(colnames(DESIGN$design),collapse=',')), "md")
 
@@
   if (length(postAdjustedSigCovars$significantCovars) != 0)
-    print(paste('Following coefficients have to be included in the model:',paste(postAdjustedSigCovars$significantCovars,collapse=',')))
+    inline_expr(paste('Following coefficients have to be included in the model:',paste(postAdjustedSigCovars$significantCovars,collapse=',')), "md")
 
@@
 ```
-### Middle Temporal Analysis
-```{r middle.temporal, cache=TRUE, fig.height=10, fig.width=15}
+### Extract expression data and adjust covariates 
+```{r covariate.analysis, cache=TRUE, fig.height=10, fig.width=15}
 # SynIDs of expression and covariates matrix
@@
 # Obtain adjusted expession
-adjustedExpression <- lapply(EXP_ID, covariatesAdjustments, COV_ID)
+adjustedExpression <- mapply(covariatesAdjustments, EXP_ID, names(EXP_ID), MoreArgs= list(COV_ID), SIMPLIFY=T)
 ```

### 2016-01-07 code/Rmd/MSSM_Array.Rmd
Minor modifications in MSSM Array normalisation

@@
 covariatesAdjustments <- function(exprId, regionName, covId){
-  inline_expr(paste0('Working on ',regionName), "md")
+  writeLines(paste0('Working on ',regionName))
 
@@
   print(preAdjustedSigCovars$PC_res[[2]]$plotData)
-  inline_expr(paste('Significant covariates are:',paste(preAdjustedSigCovars$significantCovars,collapse=',')), "md")
+  writeLines(paste('Significant covariates are:',paste(preAdjustedSigCovars$significantCovars,collapse=',')))
 
@@
   DESIGN <- getDesignMatrix(covar[,'Dx',drop=F], Intercept = F)
-  inline_expr(paste('Fitting linear model with the following coefficients:',paste(colnames(DESIGN$design),collapse=',')), "md")
+  writeLines(paste('Fitting linear model with the following coefficients:',paste(colnames(DESIGN$design),collapse=',')))
 
@@
   if (length(postAdjustedSigCovars$significantCovars) != 0)
-    inline_expr(paste('Following coefficients have to be included in the model:',paste(postAdjustedSigCovars$significantCovars,collapse=',')), "md")
+    writeLines(paste('Following coefficients have to be included in the model:',paste(postAdjustedSigCovars$significantCovars,collapse=',')))
 
@@
 ### Extract expression data and adjust covariates 
-```{r covariate.analysis, cache=TRUE, fig.height=10, fig.width=15}
+```{r covariate.analysis, cache=TRUE, fig.height=10, fig.width=15, echo=FALSE, results='asis'}
 # SynIDs of expression and covariates matrix
@@
 # Obtain adjusted expession
-adjustedExpression <- mapply(covariatesAdjustments, EXP_ID, names(EXP_ID), MoreArgs= list(COV_ID), SIMPLIFY=T)
+adjustedExpression <- mapply(covariatesAdjustments, EXP_ID, names(EXP_ID), MoreArgs= list(COV_ID), SIMPLIFY=F)
 ```

### 2016-01-07 code/Rmd/ROSMAP_RNASeq.Rmd
Added markdown to process RNASeq files from ROSMAP cohorts from AMPAD knowledge portal

---
title: "Covariate analysis of ROSMAP RNASeq"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq.Rmd",
                                 parentId ="syn5570291",
                                 entityName = 'ROSMAP RNASeq')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(synapseClient)
library(ggplot2)
library(reshape2)
library(data.table)
library(dplyr)
library(stringr)
library(limma)
library(knitr)
library(Biobase)
library(GEOquery)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)

## Needs the dev branch
library(rGithubClient)

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

synapseLogin()

# source utility files from ../R/lib folder
file.sources = list.files('../R/lib',pattern="*.R", full.names=T)
tmp = sapply(file.sources,source,.GlobalEnv)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570291';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of ROSMAP RNASeq data';

thisFileName <- 'ROSMAP_RNASeq.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_ID <- 'syn3505720';
ALL_USED_IDs <- COUNT_ID
COUNT_OBJ <- synGet(COUNT_ID)
COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F)

# COUNT_ID1 <- 'syn3505720'; #'syn3505732'
# ALL_USED_IDs <- COUNT_ID1
# COUNT_OBJ1 <- synGet(COUNT_ID1)
# COUNT1 <- read.table(COUNT_OBJ1@filePath,header=T,sep='\t',check.names = F)
# 
# # Download expression data
# COUNT_ID2 <- 'syn3505724'
# ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_ID2
# COUNT_OBJ2 <- synGet(COUNT_ID2)
# COUNT2 <- read.table(COUNT_OBJ2@filePath,header=T,sep='\t',check.names = F)

# Get clinical metadata
METADATA.CLINICAL_ID <- 'syn3191087'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.CLINICAL_ID
METADATA.CLINICAL_OBJ <- synGet(METADATA.CLINICAL_ID)
METADATA.CLINICAL <- read.table(METADATA.CLINICAL_OBJ@filePath,sep=',',header=T)

# Get technical covariates
# METADATA.TECH <- read.csv('./ROSMAP_RNA_Seq_RIN_Sample_READS.csv')
# METADATA.TECH <- METADATA.TECH[!duplicated(METADATA.TECH$Projid),]
METADATA.TECH_ID <- 'syn4300313'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.TECH_ID
METADATA.TECH_OBJ <- synGet(METADATA.TECH_ID)
METADATA.TECH <- read.table(METADATA.TECH_OBJ@filePath,sep='\t',header=T)

# Match technical and clinical covariates
METADATA <- dplyr::inner_join(METADATA.CLINICAL,METADATA.TECH)
```

### Data preprocessing
```{r preprocess.data,cache=TRUE, echo=TRUE}
# Remove samples with no cogdx, RIN and PMI scores
METADATA <- METADATA %>%
  mutate(Sampleid = paste(Sampleid, Batch, sep = '_')) %>%
  filter(Sampleid %in% colnames(COUNT)) %>%
  filter(!is.na(cogdx)) %>%
  filter(!is.na(RINcontinuous)) %>%
  filter(!is.na(pmi))  
```

```{r preprocess.data1,cache=TRUE}
indToRetain = intersect(METADATA$Sampleid, colnames(EXPR))

EXPR <- COUNT; #merge(COUNT1,COUNT2,by = 'gene_id',all=T)
rownames(EXPR) <- EXPR$gene_id
EXPR <- EXPR[, colnames(EXPR) %in% indToRetain]

# Match covariates to expression data
rownames(METADATA) = METADATA$Sampleid
METADATA = METADATA[indToRetain,]
```
Dorsolateral prefrontal cortex of `r dim(EXPR)[2]` subjects from the ROS and MAP cohorts were obtained from AMP-AD portal

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering, cache=TRUE}
#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv"
FactorCovariates <- c("Batch","msex","apoe_genotype", "race", "spanish", "cogdx")
ContCovariates <- c("RINcontinuous","age_first_ad_dx","age_death","age_at_visit_max","pmi","educ")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES <- data.frame(lapply(COVARIATES,function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'\\+','')})}))
rownames(COVARIATES) <- METADATA$Sampleid

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Covariate correlation
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.1)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.3, row.width=0.15)
```

### Gene Filtering
Retain genes with atleast 50% of the samples have >= 0 FPKM counts
```{r gene.filtering, cache=TRUE}
# Remove genes
EXPR <- EXPR[(rowSums(EXPR>0)/dim(EXPR)[2] >= 0.5),]

# To avoid taking log of zero add 1e-4 to the FPKM counts
EXPR <- log2(EXPR+1e-3)
```
`r dim(EXPR)[1]` number of genes and `r dim(EXPR)[2]` number of samples are considered for the analysis

### Significant Covariates
Correlation between pca of normalised mRNA expression and covariates
```{r preAdjusted.covariates, cache=TRUE, results='asis', echo=FALSE}
# Find correlation between PC's of gene expression with covariates
DM = getDesignMatrix(COVARIATES, Intercept = F)
DESIGN = getDesignMat2Fact(DM$design,FactorCovariates)
rownames(DESIGN) = rownames(COVARIATES)
preAdjustedSigCovars = runPCAandPlotCorrelations(EXPR, DESIGN,'non-normalized', isKeyPlot=TRUE)

# Find significant covariates
adjustCovars = designMatVarsToCovars(getCovariatesMap(DM),preAdjustedSigCovars$significantCovars)
```
Significant covariates to adjust at FDR 0.1 are `r adjustCovars`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=10, fig.height=6}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (custom design)
Fitting a linear model with cogdx, Batch, RINcontinuous, and pmi 
```{r iterative.normalisation, cache=TRUE, results='asis', echo=FALSE}
postAdjustCovars <- c('cogdx','Batch',"RINcontinuous","pmi")

writeLines(paste('Using following covariates in the final model',paste(postAdjustCovars,collapse=','),'as fixed effects',sep=' '))

# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

# Re-calculate linear fit using new weights and new design
FIT = lmFit(EXPR, design=DM1$design)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,EXPR)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
rownames(DESIGN) <- rownames(COVARIATES)
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION,
                                              DESIGN,
                                              'all adjusted design',
                                              isKeyPlot=TRUE)
```

```{r residualSigCovars.manual.ALL, fig.width=15, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```

```{r addVarsBack}
varsToAddBackIn <- c("cogdx1",  "cogdx2", "cogdx3", "cogdx4", "cogdx5", "cogdx6")
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + (FIT$coefficients[,varsToAddBackIn] %*% t(data.matrix(DESIGN[,varsToAddBackIn])))
RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'EnsemblGeneID') %>%
  tidyr::separate(col = EnsemblGeneID, into = c('ensembl_gene_id','position'), sep = '\\.')
```

```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = data.frame(cogdx = 1:6, No.OfSamples = summary(COVARIATES$cogdx))
kable(tmp)
```

Interpretation of cogdx scores

1.  NCI, No cognitive impairment (No impaired domains)
2.  MCI, Mild cognitive impairment (One impaired domain) and NO other cause of CI
3.  MCI, Mild cognitive impairment (One impaired domain) AND another cause of CI
4.  AD, Alzheimer's disease and NO other cause of CI (NINCDS PROB AD)
5.  AD, Alzheimer's disease AND another cause of CI (NINCDS POSS AD)
6.  Other dementia. Other primary cause of dementia

```{r get.gene.symbols, cache=TRUE}
# Define biomart object
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = RESIDUAL.GENE_EXPRESSION$ensembl_gene_id,
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(RESIDUAL.GENE_EXPRESSION, Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE}
# Code
CODE <- Folder(name = "Differential Expression Analysis ROSMAP RNASeq (Based on cogdx)", parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'Sampleid_batch')
write.table(COVARIATES, file = 'covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store covariates
write.table(RESIDUAL.GENE_EXPRESSION, file = 'expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('expression.tsv', name = 'FPKM Normalised Adjusted Residual Expression (cogdx added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXPRP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
```
|  *Results*            |  *SynapseID*                    |
|  -------              |   ---------                     |
|  Covariates           |  `r COV_OBJ$properties$id`      |
|  Residual Expression  |  `r EXP_OBJ$properties$id`      |

### 2016-01-08 code/Rmd/ROSMAP_RNASeq.Rmd
Minor bug fix

@@
                                  parentId ="syn5570291",
-                                 entityName = 'ROSMAP RNASeq')
+                                 entityName = 'ROSMAP RNASeq (Based on cogdx)')
 ```
@@
 ### Store files in synapse
-```{r synapse.store, include=FALSE, eval=TRUE}
+```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
 # Code
-CODE <- Folder(name = "Differential Expression Analysis ROSMAP RNASeq (Based on cogdx)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP RNASeq (Based on cogdx)", parentId = parentId)
 CODE <- synStore(CODE)
@@
 EXP_OBJ = File('expression.tsv', name = 'FPKM Normalised Adjusted Residual Expression (cogdx added)', parentId = CODE$properties$id)
-EXP_OBJ = synStore(EXPRP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)

### 2016-01-11 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
Added markdown to compute covariates for mayo cerebellum data

---
title: "Covariate analysis of UFL-Mayo-ISB RNASeq"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./Mayo_Cerebellum_RNASeq.Rmd",
                                 parentId ="syn5570325",
                                 entityName = 'UFL_Mayo_ISB Cerebellum RNASeq')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(synapseClient)
library(ggplot2)
library(reshape2)
library(data.table)
library(dplyr)
library(stringr)
library(limma)
library(knitr)
library(Biobase)
library(GEOquery)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)

## Needs the dev branch
library(rGithubClient)

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

synapseLogin()

# source utility files from ../R/lib folder
file.sources = list.files('../R/lib',pattern="*.R", full.names=T)
tmp = sapply(file.sources,source,.GlobalEnv)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570325';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of UFL_Mayo_ISB cerebellum RNASeq data';

thisFileName <- 'Mayo_Cerebellum_RNASeq.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_ID <- 'syn5201012';
ALL_USED_IDs <- COUNT_ID
COUNT_OBJ <- synGet(COUNT_ID)
COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F)

# Get clinical metadata
METADATA_ID <- 'syn5223705'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA_OBJ <- synGet(METADATA_ID)
METADATA <- read.table(METADATA_OBJ@filePath,sep=',',header=T, stringsAsFactors=F)
```

### Data preprocessing
```{r preprocess.data,cache=TRUE, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA <- METADATA %>%
  dplyr::filter(!(SampleID %in% c("742_CER")))
```

```{r preprocess.data1,cache=TRUE}
indToRetain = intersect(METADATA$SampleID, colnames(COUNT))

EXPR <- COUNT;
EXPR <- EXPR[, colnames(EXPR) %in% indToRetain]

# Match covariates to expression data
rownames(METADATA) = METADATA$SampleID
METADATA = METADATA[rownames(METADATA) %in% indToRetain,]

EXPR = EXPR[,indToRetain]
COUNT = COUNT[rownames(EXPR),colnames(EXPR)]
METADATA = METADATA[indToRetain,]
```
Dorsolateral prefrontal cortex of `r dim(EXPR)[2]` subjects from the Mayo cohorts were obtained from AMP-AD portal

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering, cache=TRUE}
FactorCovariates <- c("Source", "Diagnosis",  "Sex",  "ApoE", "Flowcell")
ContCovariates <- c("RIN", "AgeAtDeath")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$SampleID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Covariate correlation
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.1)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.3, row.width=0.15)
```

### Gene Filtering
Retain genes with atleast 50% of the samples have >= 1 gene counts
```{r gene.filtering, results='asis', echo=FALSE}
# Remove genes
EXPR <- getGeneFilteredGeneExprMatrix(EXPR, 
                                      MIN_GENE_CPM=1,
                                      MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5,
                                      verbose=T)
```

### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=NULL, plot=T)

# Initial design matrix with covariates under consideration
DM = getDesignMatrix(COVARIATES, Intercept = F)
```
Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, METADATA, by="SampleID")

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(Diagnosis), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") 
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, results='asis', echo=FALSE}
# Find correlation between PC's of gene expression with covariates
DESIGN = getDesignMat2Fact(DM$design,FactorCovariates)
colnames(DESIGN) = substr(colnames(DESIGN),1,50)
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, DESIGN,
                                                 'NULL design(voom-normalized)',
                                                 isKeyPlot=TRUE)

# Find significant covariates
adjustCovars = designMatVarsToCovars(getCovariatesMap(DM),
                                     preAdjustedSigCovars$significantCovars)
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
```{r iterative.adjusted.voom.normalisation, cache=TRUE, results='asis', echo=FALSE}
postAdjustCovars = c('Diagnosis')

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars

significantCovars = designMatVarsToCovars(getCovariatesMap(DM), 
                                          residualSigCovars$significantCovars)

covariatesEffects = designMat2CovEffects(getCovariatesMap(DM),
                                         residualSigCovars$Effects.significantCovars)

covariatesEffects = covariatesEffects[setdiff(significantCovars, postAdjustCovars)]

postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:', 
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=DM1$design, plot=F)

  # Estimate voom weights
  FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION$E, 
              design=DM1$design, 
              weights = VOOM.ADJUSTED.GENE_EXPRESSION$weights)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), postAdjustCovars)

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                DESIGN,
                                                'all adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  significantCovars = designMatVarsToCovars(getCovariatesMap(DM), residualSigCovars$significantCovars)
  covariatesEffects = designMat2CovEffects(getCovariatesMap(DM),residualSigCovars$Effects.significantCovars)
  covariatesEffects = covariatesEffects[setdiff(significantCovars,postAdjustCovars)]

  postAdjustCovars = c(postAdjustCovars,names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
modelStr <-paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=','), 'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`

```{r adjusted.voom.normalisation, results='asis', echo=FALSE}
# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F], Intercept = F)
tmp = linColumnFinder(DM1$design)
writeLines('Degenerate variables are:')
writeLines(tmp$relations)
DM1$design = DM1$design[,tmp$indepCols]

# Recalculate weights with new design
VOOM.ADJUSTED.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, 
                                     design=DM1$design, plot=T)

# Linear model with new weights and design
VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION$E, 
                          design=DM1$design, 
                          weights = VOOM.ADJUSTED.GENE_EXPRESSION$weights)  
```
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r calculate.residuals, include=FALSE}
# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                              VOOM.ADJUSTED.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              DESIGN,
                                              'residual matrix of all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
```{r addVarsBack}
varsToAddBackIn <- c("DiagnosisAD", "DiagnosisControl", "DiagnosisPSP", "DiagnosisPathologic Aging")
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddBackIn] %*% t(data.matrix(DM1$design[,varsToAddBackIn]))
RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'ensembl_gene_id') 
```

```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = as.data.frame(summary(COVARIATES$Diagnosis))
colnames(tmp) = 'No. of samples'
kable(tmp)
```

```{r get.gene.symbols, cache=TRUE}
# Define biomart object
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = RESIDUAL.GENE_EXPRESSION$ensembl_gene_id,
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(RESIDUAL.GENE_EXPRESSION, Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB Cerebellum RNASeq', parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store covariates
write.table(RESIDUAL.GENE_EXPRESSION, file = 'expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('expression.tsv', name = 'Voom Adjusted Residual Expression (Diagnosis added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
```
|  *Results*            |  *SynapseID*                    |
|  -------              |   ---------                     |
|  Covariates           |  `r COV_OBJ$properties$id`      |
|  Residual Expression  |  `r EXP_OBJ$properties$id`      |

### 2016-01-11 code/Rmd/Mayo_TemporalCortex_RNASeq.Rmd
Added markdown to compute covariates for Mayo temporal cortex

--- 2016-01-11 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
@@
 ---
-title: "Covariate analysis of UFL-Mayo-ISB RNASeq"
+title: "Covariate analysis of UFL-Mayo-ISB RNASeq temporal cortex"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./Mayo_Cerebellum_RNASeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./Mayo_TemporalCortex_RNASeq.Rmd",
                                  parentId ="syn5570325",
-                                 entityName = 'UFL_Mayo_ISB Cerebellum RNASeq')
+                                 entityName = 'UFL_Mayo_ISB Temporal Cortex RNASeq')
 ```
@@
 activityName = 'Covariate analysis';
-activityDescription = 'Covariate analysis of UFL_Mayo_ISB cerebellum RNASeq data';
+activityDescription = 'Covariate analysis of UFL_Mayo_ISB temporal cortex RNASeq data';
 
-thisFileName <- 'Mayo_Cerebellum_RNASeq.Rmd'
+thisFileName <- 'Mayo_TemporalCortex_RNASeq.Rmd'
 
@@
 # Download expression data
-COUNT_ID <- 'syn5201012';
+COUNT_ID <- 'syn4650257';
 ALL_USED_IDs <- COUNT_ID
 COUNT_OBJ <- synGet(COUNT_ID)
-COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F)
+COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep=' ',check.names = F)
 
 # Get clinical metadata
-METADATA_ID <- 'syn5223705'
+METADATA_ID <- 'syn3817650'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
@@
 ```{r preprocess.data1,cache=TRUE}
-indToRetain = intersect(METADATA$SampleID, colnames(COUNT))
+indToRetain = intersect(METADATA$ID, colnames(COUNT))
 
@@
 # Match covariates to expression data
-rownames(METADATA) = METADATA$SampleID
+rownames(METADATA) = METADATA$ID
 METADATA = METADATA[rownames(METADATA) %in% indToRetain,]
@@
 ```{r covariates.clustering, cache=TRUE}
-FactorCovariates <- c("Source", "Diagnosis",  "Sex",  "ApoE", "Flowcell")
+FactorCovariates <- c("Source", "Diagnosis",  "FLOWCELL")
 ContCovariates <- c("RIN", "AgeAtDeath")
@@
 COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
-rownames(COVARIATES) <- METADATA$SampleID
+rownames(COVARIATES) <- METADATA$ID
 
@@
 # Plot first 2 PCs
-plotdata <- data.frame(SampleID=rownames(PC$rotation), 
+plotdata <- data.frame(ID=rownames(PC$rotation), 
                        PC1=PC$rotation[,1], 
@@
 
-plotdata <- left_join(plotdata, METADATA, by="SampleID")
+plotdata <- left_join(plotdata, METADATA, by="ID")
 
@@
 # Code
-CODE <- Folder(name = 'UFL_Mayo_ISB Cerebellum RNASeq', parentId = parentId)
+CODE <- Folder(name = 'UFL_Mayo_ISB Temporal Cortex RNASeq', parentId = parentId)
 CODE <- synStore(CODE)
@@
 # Store covariates
-COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
+COVARIATES = rownameToFirstColumn(COVARIATES, 'ID')
 write.table(COVARIATES, file = 'covariates.tsv', sep = '\t', row.names=F, quote=F)

### 2016-01-11 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
Minor modifications

@@
 METADATA <- METADATA %>%
-  dplyr::filter(!(SampleID %in% c("742_CER")))
+  dplyr::filter(!(SampleID %in% c("742_CER", "11386_CER","11311_CER","1923_CER","1147_CER")))
 ```
-
+Following five samples were considered outliers and removed 742_CER, 11386_CER, 11311_CER, 1923_CER, 1147_CER
 ```{r preprocess.data1,cache=TRUE}
@@
 p <- p + theme_bw() + theme(legend.position="top") 
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
 p
+
+# Tree based clustering
+sampleTrees = flashClust::hclust(dist(t(VOOM.GENE_EXPRESSION$E)), method = "average")
+tree = WGCNA::cutreeStatic(sampleTrees, cutHeight = 200, minSize=3)
+
+WGCNA::plotDendroAndColors(sampleTrees, 
+                    labels2colors(cbind(tree,COVARIATES)),
+                    groupLabels = c('Cluster',colnames(COVARIATES)))
+collectGarbage()
 ```
+
 ### Significant Covariates

### 2016-01-11 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
Minor modifications

@@
 ---
-title: "Covariate analysis of UFL-Mayo-ISB RNASeq"
+title: "Covariate analysis of UFL-Mayo-ISB RNASeq from cerebellum"
 author: "Thanneer Perumal"
@@
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
-  VOOM.ADJUSTED.FIT$coefficients[,varsToAddBackIn] %*% t(data.matrix(DM1$design[,varsToAddBackIn]))
+  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(VOOM.ADJUSTED.FIT$coefficients[,varsToAddBackIn]))
 RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'ensembl_gene_id') 

### 2016-01-22 code/Rmd/MSSM_RNASeq.Rmd
Added outlier analysis to the MSSM RNASeq workflow

@@
 ```
+### Outlier analysis
+Clustering of initial normalised data (with NULL design)
+```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
+# Find principal components of expression to plot
+PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T)
+
+# Plot first 2 PCs
+plotdata <- data.frame(ID=rownames(PC$rotation), 
+                       PC1=PC$rotation[,1], 
+                       PC2=PC$rotation[,2])
+
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'ID'))
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(color=factor(TISSUE.Dx), shape=as.factor(CDR), size=as.numeric(PMI)))
+p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
+# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
+p
+
+COVARIATES.tmp = data.matrix(COVARIATES)
+COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
+
+tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
+cols = WGCNA::labels2colors(COVARIATES.tmp);
+tmp = gsub('\\..*','',cols)
+WGCNA::plotDendroAndColors(tree, 
+                           colors = tmp, 
+                           dendroLabels = FALSE, 
+                           abHeight = 0.80, 
+                           main = "Sample dendrogram",
+                           groupLabels = colnames(COVARIATES.tmp))
+```
+
 ### Significant Covariates

### 2016-01-22 code/Rmd/ROSMAP_RNASeq.Rmd

@@
   filter(!is.na(RINcontinuous)) %>%
-  filter(!is.na(pmi))  
+  filter(!is.na(pmi)) %>%
+  filter(!(Sampleid %in% c('380_120503_1')))
 ```
-
+Sample id 380\_120503\_1 is considered as outlier and removed
 ```{r preprocess.data1,cache=TRUE}
-indToRetain = intersect(METADATA$Sampleid, colnames(EXPR))
+indToRetain = intersect(METADATA$Sampleid, colnames(COUNT))
 
@@
 `r dim(EXPR)[1]` number of genes and `r dim(EXPR)[2]` number of samples are considered for the analysis
+
+### Outliers detection
+```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
+# Find principal components of expression to plot
+PC <- prcomp(EXPR, scale.=T)
+
+# Plot first 2 PCs
+plotdata <- data.frame(Sampleid=rownames(PC$rotation), 
+                       PC1=PC$rotation[,1], 
+                       PC2=PC$rotation[,2])
+
+plotdata <- left_join(plotdata, METADATA)
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(color=factor(Batch), shape=as.factor(cogdx), size=as.numeric(age_death)))
+p <- p + theme_bw() + theme(legend.position="top") 
+p <- p + geom_text(aes(label= Sampleid), size=4, hjust=1)
+p
+
+COVARIATES.tmp = data.matrix(COVARIATES)
+COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
+
+tree = hclust(as.dist(t(EXPR)))
+plotDendroAndColors(tree, 
+                    colors = labels2colors(COVARIATES.tmp), 
+                    dendroLabels = FALSE, 
+                    abHeight = 0.80, 
+                    main = "Sample dendrogram",
+                    groupLabels = colnames(COVARIATES.tmp))
+```
 
@@
 varsToAddBackIn <- c("cogdx1",  "cogdx2", "cogdx3", "cogdx4", "cogdx5", "cogdx6")
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + (FIT$coefficients[,varsToAddBackIn] %*% t(data.matrix(DESIGN[,varsToAddBackIn])))
+RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + t(data.matrix(DESIGN[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
 RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'EnsemblGeneID') %>%

### 2016-01-25 code/Rmd/ROSMAP_RNASeq.Rmd
Minor modifications

@@
 p <- p + theme_bw() + theme(legend.position="top") 
-p <- p + geom_text(aes(label= Sampleid), size=4, hjust=1)
+# p <- p + geom_text(aes(label= Sampleid), size=4, hjust=1)
 p
@@
 ```
-
+Adding cogdx back to the residuals
 ```{r addVarsBack}
@@
   tidyr::separate(col = EnsemblGeneID, into = c('ensembl_gene_id','position'), sep = '\\.')
+```
+
+### Clustering residual gene expression
+```{r decompse.normalise.data1, fig.height=8, fig.width=8, results='asis'}
+# Find principal components of expression to plot
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1:2)], scale.=T)
+
+# Plot first 2 PCs
+plotdata <- data.frame(Sampleid=rownames(PC$rotation), 
+                       PC1=PC$rotation[,1], 
+                       PC2=PC$rotation[,2])
+
+plotdata <- left_join(plotdata, METADATA)
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(color=factor(Batch), shape=as.factor(cogdx), size=as.numeric(age_death)))
+p <- p + theme_bw() + theme(legend.position="top") 
+# p <- p + geom_text(aes(label= Sampleid), size=4, hjust=1)
+p
+
+COVARIATES.tmp = data.matrix(COVARIATES)
+COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
+
+tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION[,-(1:2)])))
+plotDendroAndColors(tree, 
+                    colors = labels2colors(COVARIATES.tmp), 
+                    dendroLabels = FALSE, 
+                    abHeight = 0.80, 
+                    main = "Sample dendrogram",
+                    groupLabels = colnames(COVARIATES.tmp))
 ```

### 2016-01-25 code/Rmd/MSSM_RNASeq.Rmd
Minor modifications

@@
 
+### Clustering residual data
+```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
+# Find principal components of expression to plot
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T)
+
+# Plot first 2 PCs
+plotdata <- data.frame(ID=rownames(PC$rotation), 
+                       PC1=PC$rotation[,1], 
+                       PC2=PC$rotation[,2])
+
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'ID'))
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(color=factor(TISSUE.Dx), shape=as.factor(CDR), size=as.numeric(PMI)))
+p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
+# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
+p
+
+COVARIATES.tmp = data.matrix(COVARIATES)
+COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
+
+tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
+cols = WGCNA::labels2colors(COVARIATES.tmp);
+tmp = gsub('\\..*','',cols)
+WGCNA::plotDendroAndColors(tree, 
+                           colors = tmp, 
+                           dendroLabels = FALSE, 
+                           abHeight = 0.80, 
+                           main = "Sample dendrogram",
+                           groupLabels = colnames(COVARIATES.tmp))
+```
+
 ### Store files in synapse

### 2016-01-25 code/Rmd/MSSM_RNASeq.Rmd
minor changes

@@
                            groupLabels = colnames(COVARIATES.tmp))
+dev.off()
 ```
@@
 ```{r preAdjustedSigCovars.NULL, fig.width=15, fig.height=8}
-preAdjustedSigCovars[["PC_res"]][[1]]$plotData
+preAdjustedSigCovars[["PC_res"]][[2]]$plotData
 ```
@@
 ```{r residualSigCovars.manual, fig.width=15, fig.height=8}
-residualSigCovars[["PC_res"]][[1]]$plotData
+residualSigCovars[["PC_res"]][[2]]$plotData
 ```
@@
 
+```{r summary, results='asis', echo=FALSE}
+writeLines('Number of samples in each category are')
+tmp = as.data.frame(summary(COVARIATES$TISSUE.Dx))
+colnames(tmp) = 'No. of samples'
+kable(tmp)
+```
+
 ### Store files in synapse
@@
            parentId = CODE$properties$id)
-obj = synStore(obj,
+expr_obj = synStore(obj,
                used = ALL_USED_IDs, 
@@
            parentId = CODE$properties$id)
-obj = synStore(obj,
+covar_obj = synStore(obj,
                used = ALL_USED_IDs, 
@@
 ```
+|  *Results*            |  *SynapseID*                    |
+|  -------              |   ---------                     |
+|  Covariates           |  `r covar_obj$properties$id`    |
+|  Residual Expression  |  `r expr_obj$properties$id`     |

### 2016-01-25 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
Minor changes

@@
 METADATA <- METADATA %>%
-  dplyr::filter(!(SampleID %in% c("742_CER", "11386_CER","11311_CER","1923_CER","1147_CER")))
+  dplyr::filter(!(SampleID %in% c("742_CER", "11386_CER","11311_CER","1923_CER","11477_CER")))
 ```
-Following five samples were considered outliers and removed 742_CER, 11386_CER, 11311_CER, 1923_CER, 1147_CER
-```{r preprocess.data1,cache=TRUE}
+Following five samples were considered outliers and removed 742\_CER, 11386\_CER, 11311\_CER, 1923\_CER, 11477\_CER
+```{r preprocess.data1}
 indToRetain = intersect(METADATA$SampleID, colnames(COUNT))
@@
 p <- p + theme_bw() + theme(legend.position="top") 
-p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
 p
@@
 collectGarbage()
+dev.off()
 ```
@@
 Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
-```{r calculate.residuals, include=FALSE}
+```{r calculate.residuals, include=FALSE, results = 'asis'}
 # Residuals after normalisation
@@
 ```
+Add "Diagnosis"" back to the residuals
 ```{r addVarsBack}
@@
 
+### Clustering residual gene expression
+```{r decompse.normalise.data1, fig.height=8, fig.width=8, results='asis'}
+# Find principal components of expression to plot
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1)], scale.=T)
+
+# Plot first 2 PCs
+plotdata <- data.frame(SampleID=rownames(PC$rotation), 
+                       PC1=PC$rotation[,1], 
+                       PC2=PC$rotation[,2])
+
+plotdata <- left_join(plotdata, METADATA, by="SampleID")
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(color=factor(Diagnosis), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
+p <- p + theme_bw() + theme(legend.position="top") 
+# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+p
+
+# Tree based clustering
+sampleTrees = flashClust::hclust(dist(t(RESIDUAL.GENE_EXPRESSION[,-(1)])), method = "average")
+tree = WGCNA::cutreeStatic(sampleTrees, cutHeight = 200, minSize=3)
+
+WGCNA::plotDendroAndColors(sampleTrees, 
+                           labels2colors(cbind(tree,COVARIATES)),
+                           groupLabels = c('Cluster',colnames(COVARIATES)))
+collectGarbage()
+dev.off()
+```
+
 ```{r summary, results='asis', echo=FALSE}
 writeLines('Number of samples in each category are')
-tmp = as.data.frame(summary(COVARIATES$Diagnosis))
-colnames(tmp) = 'No. of samples'
+tmp = data.frame(summary(COVARIATES$Diagnosis))
+colnames(tmp) = "No. of samples"
 kable(tmp)

### 2016-01-25 code/Rmd/Mayo_TemporalCortex_RNASeq.Rmd
Minor modifications

@@
 
-```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
+```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
 ## It is assumed your working directory is where this file
@@
 METADATA <- METADATA %>%
-  dplyr::filter(!(SampleID %in% c("742_CER")))
+  dplyr::filter(!(ID %in% c("1923_TCX")))
 ```
-
+Following five samples were considered outliers and removed 1923\_TCX
 ```{r preprocess.data1,cache=TRUE}
@@
 p <- p + theme_bw() + theme(legend.position="top") 
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+p <- p + geom_text(aes(label= ID), size=4, hjust=0)
 p
+
+# Tree based clustering
+sampleTrees = flashClust::hclust(dist(t(VOOM.GENE_EXPRESSION$E)), method = "average")
+tree = WGCNA::cutreeStatic(sampleTrees, cutHeight = 200, minSize=3)
+
+WGCNA::plotDendroAndColors(sampleTrees, 
+                    labels2colors(cbind(tree,COVARIATES)),
+                    groupLabels = c('Cluster',colnames(COVARIATES)))
+collectGarbage()
+dev.off()
 ```
@@
 ```
+Adding "Diagnosis" back to the residuals
 ```{r addVarsBack}
@@
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
-  VOOM.ADJUSTED.FIT$coefficients[,varsToAddBackIn] %*% t(data.matrix(DM1$design[,varsToAddBackIn]))
+  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(VOOM.ADJUSTED.FIT$coefficients[,varsToAddBackIn]))
 RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'ensembl_gene_id') 
+```
+
+### Clustering residual gene expression
+```{r decompse.normalise.data1, fig.height=8, fig.width=8, results='asis'}
+# Find principal components of expression to plot
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1)], scale.=T)
+
+# Plot first 2 PCs
+plotdata <- data.frame(ID=rownames(PC$rotation), 
+                       PC1=PC$rotation[,1], 
+                       PC2=PC$rotation[,2])
+
+plotdata <- left_join(plotdata, METADATA, by="ID")
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(color=factor(Diagnosis), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
+p <- p + theme_bw() + theme(legend.position="top") 
+# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+p
+
+# Tree based clustering
+sampleTrees = flashClust::hclust(dist(t(RESIDUAL.GENE_EXPRESSION[,-(1)])), method = "average")
+
+plotDendroAndColors(sampleTrees, 
+                    colors = labels2colors(COVARIATES), 
+                    abHeight = 0.80, 
+                    main = "Sample dendrogram",
+                    groupLabels = colnames(COVARIATES))
+collectGarbage()
+dev.off()
 ```
@@
 writeLines('Number of samples in each category are')
-tmp = as.data.frame(summary(COVARIATES$Diagnosis))
-colnames(tmp) = 'No. of samples'
+tmp = data.frame(summary(COVARIATES$Diagnosis))
+colnames(tmp) = "No. of samples"
 kable(tmp)

### 2016-01-25 code/Rmd/Mayo_TemporalCortex_Array.Rmd
Added markdown to curate Mayo temporal cortex array

---
title: "Covariate analysis of UFL-Mayo-ISB temporal cortex arrays"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./Mayo_TemporalCortex_Array.Rmd",
                                 parentId ="syn5570325",
                                 entityName = 'UFL_Mayo_ISB Temporal Cortex Microarray')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(synapseClient)
library(ggplot2)
library(reshape2)
library(data.table)
library(dplyr)
library(stringr)
library(limma)
library(knitr)
library(Biobase)
library(GEOquery)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)

## Needs the dev branch
library(rGithubClient)

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

synapseLogin()

# source utility files from ../R/lib folder
file.sources = list.files('../R/lib',pattern="*.R", full.names=T)
tmp = sapply(file.sources,source,.GlobalEnv)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570325';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of UFL_Mayo_ISB temporal cortex array data';

thisFileName <- 'Mayo_TemporalCortex_Array.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
EXPR_ID <- 'syn3617054';
ALL_USED_IDs <- EXPR_ID
EXPR_OBJ <- synGet(EXPR_ID)
EXPR <- fread(EXPR_OBJ@filePath,sep=',',data.table=F)

# Get clinical metadata
METADATA_ID <- 'syn3617056'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA_OBJ <- synGet(METADATA_ID)
METADATA <- read.table(METADATA_OBJ@filePath,sep=',',header=T, stringsAsFactors=F)
```

### Data preprocessing
```{r preprocess.data,cache=TRUE}
rownames(EXPR) <- EXPR$IID
EXPR <- data.matrix(EXPR[,-(1:2)]) %>% t

# Fix metadata
rownames(METADATA) <-  METADATA$IID
METADATA = METADATA[colnames(EXPR),]

METADATA = dplyr::mutate(METADATA,
                         plate2 = 2*plate2,
                         plate3 = 3*plate3,
                         plate4 = 4*plate4,
                         plate5 = 5*plate5,
                         plate = plate2+plate3+plate4+plate5)
METADATA$plate[METADATA$plate == 0] = 1
```
Temporal cortex of `r dim(EXPR)[2]` subjects from the Mayo cohorts were obtained from [AMP-AD portal](`r EXPR_ID`)

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering, cache=TRUE}
FactorCovariates <- c("Dxn", "Sex",  "E4dose", "plate")
ContCovariates <- c("Age", "RIN", "RINsqAdj")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$IID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Covariate correlation
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.1)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.3, row.width=0.15)
```

### Outlier analysis
Clustering of raw data
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(EXPR, scale.=T)

# Plot first 2 PCs
plotdata <- data.frame(IID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, METADATA, by="IID")

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(Dxn), shape=as.factor(plate), size=as.numeric(Age)))
p <- p + theme_bw() + theme(legend.position="top") 
# p <- p + geom_text(aes(label= IID), size=4, hjust=0)
p

COVARIATES.tmp = data.matrix(COVARIATES)
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(EXPR)))
plotDendroAndColors(tree, 
                    colors = labels2colors(COVARIATES.tmp), 
                    dendroLabels = FALSE,
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES.tmp))
```
### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, results='asis', echo=FALSE}
# Find correlation between PC's of gene expression with covariates
DM = getDesignMatrix(COVARIATES, Intercept=F)
DESIGN = getDesignMat2Fact(DM$design,FactorCovariates)
colnames(DESIGN) = substr(colnames(DESIGN),1,50)
preAdjustedSigCovars = runPCAandPlotCorrelations(EXPR, DESIGN,
                                                 'NULL design',
                                                 isKeyPlot=TRUE)

# Find significant covariates
adjustCovars = designMatVarsToCovars(getCovariatesMap(DM),
                                     preAdjustedSigCovars$significantCovars)
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
```{r iterative.adjusted.voom.normalisation, cache=TRUE, results='asis', echo=FALSE}
postAdjustCovars = c('Dxn')

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars

significantCovars = designMatVarsToCovars(getCovariatesMap(DM), 
                                          residualSigCovars$significantCovars)

covariatesEffects = designMat2CovEffects(getCovariatesMap(DM),
                                         residualSigCovars$Effects.significantCovars)

covariatesEffects = covariatesEffects[setdiff(significantCovars, postAdjustCovars)]

postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:', 
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Fit linear model
  FIT = lmFit(EXPR, 
              design=DM1$design)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,EXPR)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), postAdjustCovars)

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                DESIGN,
                                                'all adjusted design',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  significantCovars = designMatVarsToCovars(getCovariatesMap(DM), residualSigCovars$significantCovars)
  covariatesEffects = designMat2CovEffects(getCovariatesMap(DM),residualSigCovars$Effects.significantCovars)
  covariatesEffects = covariatesEffects[setdiff(significantCovars,postAdjustCovars)]

  postAdjustCovars = c(postAdjustCovars,names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
modelStr <-paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=','), 'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`

```{r adjusted.expression, results='asis', echo=FALSE}
# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F], Intercept = F)
tmp = linColumnFinder(DM1$design)

writeLines('Degenerate variables are:')
writeLines(tmp$relations)
DM1$design = DM1$design[,tmp$indepCols]

# Linear model with new design
ADJUSTED.FIT = lmFit(EXPR, design=DM1$design)  
```
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r calculate.residuals, include=FALSE}
# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT,
                                              EXPR)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              DESIGN,
                                              'residual matrix of all adjusted design',
                                              isKeyPlot=TRUE)
```
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
```{r addVarsBack}
varsToAddBackIn <- c("Dxn0", "Dxn1")
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  ADJUSTED.FIT$coefficients[,varsToAddBackIn] %*% t(data.matrix(DM1$design[,varsToAddBackIn]))
RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'illumina_humanht_12_v4') 
```

```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = as.data.frame(summary(COVARIATES$Dxn))
colnames(tmp) = 'No. of samples'
kable(tmp)
```
Where 0 is Non-Ad and 1 is AD
```{r get.gene.symbols, cache=TRUE}
# Define biomart object
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")

# Query biomart
Illumina2HGNC <- getBM(attributes = c("illumina_humanht_12_v4", "hgnc_symbol"),
                       filters = "illumina_humanht_12_v4", 
                       values = RESIDUAL.GENE_EXPRESSION$illumina_humanht_12_v4,
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(Illumina2HGNC, RESIDUAL.GENE_EXPRESSION) %>%
  dplyr::select(illumina_humanht_12_v4, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB Temporal Cortex Microarray', parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'ID')
write.table(COVARIATES, file = 'covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store expression values
write.table(RESIDUAL.GENE_EXPRESSION, file = 'expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('expression.tsv', name = 'Voom Adjusted Residual Expression (Dxn added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
```
|  *Results*            |  *SynapseID*                    |
|  -------              |   ---------                     |
|  Covariates           |  `r COV_OBJ$properties$id`      |
|  Residual Expression  |  `r EXP_OBJ$properties$id`      |

### 2016-01-25 code/Rmd/Mayo_Cerebellum_Array.Rmd
Added markdown for processing cerebellum microarray data from mayo-ufl-isb

--- 2016-01-25 code/Rmd/Mayo_TemporalCortex_Array.Rmd
@@
 ---
-title: "Covariate analysis of UFL-Mayo-ISB temporal cortex arrays"
+title: "Covariate analysis of UFL-Mayo-ISB cerebellum arrays"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./Mayo_TemporalCortex_Array.Rmd",
+knit2synapse::knitToFolderEntity(file = "./Mayo_Cerebellum_Array.Rmd",
                                  parentId ="syn5570325",
-                                 entityName = 'UFL_Mayo_ISB Temporal Cortex Microarray')
+                                 entityName = 'UFL_Mayo_ISB Cerebellum Microarray')
 ```
@@
   tidy = FALSE,
-  cache = FALSE)
+  cache = TRUE)
 ```
@@
 activityName = 'Covariate analysis';
-activityDescription = 'Covariate analysis of UFL_Mayo_ISB temporal cortex array data';
+activityDescription = 'Covariate analysis of UFL_Mayo_ISB Cerebellum array data';
 
-thisFileName <- 'Mayo_TemporalCortex_Array.Rmd'
+thisFileName <- 'Mayo_Cerebellum_Array.Rmd'
 
@@
 # Download expression data
-EXPR_ID <- 'syn3617054';
+EXPR_ID <- 'syn3256501';
 ALL_USED_IDs <- EXPR_ID
@@
 # Get clinical metadata
-METADATA_ID <- 'syn3617056'
+METADATA_ID <- 'syn3256502'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
@@
 METADATA = dplyr::mutate(METADATA,
+                         plate0 = 6*plate0,
+                         plate1 = 1*plate1,
                          plate2 = 2*plate2,
@@
                          plate4 = 4*plate4,
-                         plate5 = 5*plate5,
-                         plate = plate2+plate3+plate4+plate5)
-METADATA$plate[METADATA$plate == 0] = 1
+                         plate = plate0+plate1+plate2+plate3+plate4)
+METADATA$plate[METADATA$plate == 0] = 5
+METADATA$plate[METADATA$plate == 6] = 0
 ```
-Temporal cortex of `r dim(EXPR)[2]` subjects from the Mayo cohorts were obtained from [AMP-AD portal](`r EXPR_ID`)
+Cerebellum of `r dim(EXPR)[2]` subjects from the Mayo cohorts were obtained from [AMP-AD portal](`r EXPR_ID`)
 
@@
                     colors = labels2colors(COVARIATES.tmp), 
-                    dendroLabels = FALSE,
-                    abHeight = 0.80, 
+#                     dendroLabels = FALSE,
+                    abHeight = 10, 
                     main = "Sample dendrogram",
@@
 # Code
-CODE <- Folder(name = 'UFL_Mayo_ISB Temporal Cortex Microarray', parentId = parentId)
+CODE <- Folder(name = 'UFL_Mayo_ISB Cerebellum Microarray', parentId = parentId)
 CODE <- synStore(CODE)

### 2016-01-26 code/Rmd/Mayo_TemporalCortex_Array.Rmd
Added differential expression to Mayo temporal cortex data

@@
 ```
+### Differential expression analysis
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diffExp}
+# Get design matrix
+DM = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F], Intercept = F)
+
+# Linear model with new design
+FIT = lmFit(EXPR, design=DM$design)
+
+# Fit contrast
+contrast = makeContrasts(contrasts="Dxn0-Dxn1", levels = colnames(FIT$coefficients))
+FIT = contrasts.fit(FIT, contrasts=contrast)
+FIT = eBayes(FIT)
+
+# Get differnetial expression
+DE = topTable(FIT, coef=1, number = dim(EXPR)[1])
+DE = rownameToFirstColumn(DE, 'illumina_humanht_12_v4') %>%
+  left_join(Illumina2HGNC)
+p <- ggplot(DE, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
+p
+
+tmp = DE %>% dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(hgnc_symbol) %>%
+  unlist %>% unique
+kable(data.frame(Comparison = "Control_vs_AD",
+           nDiffExp = length(tmp),
+           DiffExpGenes = paste(tmp, collapse = ',')))  
+```
 ### Store files in synapse
@@
                    executed = thisFile, activityDescription = activityDescription)
+
+# Store differential expression results
+write.table(DE, file = 'diffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('diffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
+DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                   executed = thisFile, activityDescription = activityDescription)
 ```
@@
 |  Residual Expression  |  `r EXP_OBJ$properties$id`      |
+|  Differential Expression  |  `r DEXP_OBJ$properties$id`      |

### 2016-01-26 code/Rmd/MSSM_RNASeq.Rmd
Added differential expression analysis to MSSM RNASeq markdown

@@
 ```
+```{r get.gene.symbols, cache=TRUE}
+# Define biomart object
+mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
 
+# Query biomart
+Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
+                       filters = "ensembl_gene_id", values = rownames(RESIDUAL.GENE_EXPRESSION),
+                       mart = mart)
+RESIDUAL.GENE_EXPRESSION <- left_join(rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id"),
+                                      Ensemble2HGNC) %>%
+  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+```
 ```{r summary, results='asis', echo=FALSE}
@@
 colnames(tmp) = 'No. of samples'
+kable(tmp)
+```
+### Differential expression analysis
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diffExp}
+# Fit contrast
+contrast = makeContrasts(contrasts=c("TISSUE.DxBM_10.Control-TISSUE.DxBM_10.AD",
+                                     "TISSUE.DxBM_22.Control-TISSUE.DxBM_22.AD",
+                                     "TISSUE.DxBM_36.Control-TISSUE.DxBM_36.AD"), 
+                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
+FIT = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
+FIT = eBayes(FIT)
+
+# Get differnetial expression
+DE = list()
+p = list()
+for (i in 1:3){
+  DE[[i]] = topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
+  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id") %>%
+    left_join(Ensemble2HGNC)
+  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
+  p[[i]] <- p[[i]] + ggtitle(gsub('TISSUE.Dx','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
+  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red')
+}
+multiplot(plotlist=p, cols = 3)
+names(DE) = gsub('TISSUE.Dx','',colnames(contrast))
+
+DE = DE %>% ldply(.id = 'Comparison')
+
+tmp = DE %>%
+  dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(Comparison, hgnc_symbol) %>%
+  group_by(Comparison) %>%
+  summarise(nDiffExp = length(unique(hgnc_symbol)), 
+            DiffExpGenes = paste(unique(hgnc_symbol), collapse = ','))
+
 kable(tmp)
@@
                activityDescription = activityDescription)
+
+# Store differential expression results
+write.table(DE, file = 'diffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('diffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
+DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                   executed = thisFile, activityDescription = activityDescription)
 ```
@@
 |  Residual Expression  |  `r expr_obj$properties$id`     |
+|  Differential Expression  |  `r DEXP_OBJ$properties$id`      |

### 2016-01-26 code/Rmd/MSSM_RNASeq.Rmd
Minor modifications

@@
   group_by(Comparison) %>%
-  summarise(nDiffExp = length(unique(hgnc_symbol)), 
+  dplyr::summarise(nDiffExp = length(unique(hgnc_symbol)), 
             DiffExpGenes = paste(unique(hgnc_symbol), collapse = ','))
@@
 write.table(DE, file = 'diffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('diffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
-DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+dexp_obj = File('diffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
+dexp_obj = synStore(dexp_obj, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
@@
 |  Residual Expression  |  `r expr_obj$properties$id`     |
-|  Differential Expression  |  `r DEXP_OBJ$properties$id`      |
+|  Differential Expression  |  `r dexp_obj$properties$id`      |

### 2016-01-26 code/Rmd/ROSMAP_RNASeq.Rmd
Added differential expression to the ROSMAP markdown

@@
 library(biomaRt)
+library(ggplot2)
 
@@
 ```{r iterative.normalisation, cache=TRUE, results='asis', echo=FALSE}
-postAdjustCovars <- c('cogdx','Batch',"RINcontinuous","pmi")
+postAdjustCovars <- c('cogdx', 'Batch', 'RINcontinuous', 'pmi')
 
@@
 
-```{r get.gene.symbols, cache=TRUE}
+```{r get.gene.symbols, cache=TRUE, eval=F}
 # Define biomart object
@@
   dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+```
+### Differential expression analysis
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diffExp}
+# Fit contrast
+contrast = makeContrasts(contrasts=c("cogdx1-cogdx2",
+                                     "cogdx1-cogdx4",
+                                     "cogdx2-cogdx4"), 
+                         levels = colnames(FIT$coefficients))
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = list()
+p = list()
+for (i in 1:3){
+  DE[[i]] = topTable(FIT.CONTR, coef=i, number = dim(EXPR)[1])
+  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id") 
+  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
+  p[[i]] <- p[[i]] + ggtitle(colnames(contrast)[i])# + xlim(c(-1.5,1.5)) + ylim(c(0,6))
+  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red')
+}
+multiplot(plotlist=p, cols = 3)
+names(DE) = colnames(contrast)
+
+DE = DE %>% plyr::ldply(.id = 'Comparison')
+
+tmp = DE %>%
+  dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(Comparison) %>%
+  group_by(Comparison) %>%
+  dplyr::summarise(nDiffExp = length(unique(hgnc_symbol)), 
+            DiffExpGenes = paste(unique(hgnc_symbol), collapse = ','))
+
+kable(tmp)
 ```

### 2016-01-26 code/Rmd/ROSMAP_RNASeq.Rmd
Minor modifications

@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp}
+```{r diffExp, fig.height=8, fig.width=16}
 # Fit contrast

### 2016-01-26 code/Rmd/ROSMAP_RNASeq.Rmd
Minor modifications

@@
 Significant covariates to adjust at FDR 0.1 are `r adjustCovars`
-```{r preAdjustedSigCovars.NULL.ALL, fig.width=10, fig.height=6}
+```{r preAdjustedSigCovars.NULL.ALL, fig.width=15, fig.height=8}
 preAdjustedSigCovars[["PC_res"]][[2]]$plotData

### 2016-01-27 code/Rmd/MSSM_RNASeq.Rmd
Minor modifications

@@
 Significant covariates to adjust at FDR 0.1 are `r adjustCovars`
-```{r preAdjustedSigCovars.NULL, fig.width=15, fig.height=8}
+```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
 preAdjustedSigCovars[["PC_res"]][[2]]$plotData
@@
 
-```{r residualSigCovars.manual, fig.width=15, fig.height=8}
+```{r residualSigCovars.manual, fig.width=20, fig.height=12}
 residualSigCovars[["PC_res"]][[2]]$plotData
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp}
+```{r diffExp, fig.height=10, fig.width=18}
 # Fit contrast

### 2016-01-27 code/Rmd/ROSMAP_RNASeq.Rmd
Minor modifications

--- 2016-01-26 code/Rmd/ROSMAP_RNASeq.Rmd
@@
                    executed = thisFile, activityDescription = activityDescription)
+
 ```

### 2016-01-27 code/Rmd/ROSMAP_RNASeq.Rmd

--- 2016-01-26 code/Rmd/ROSMAP_RNASeq.Rmd
@@
 
-```{r get.gene.symbols, cache=TRUE, eval=F}
+```{r get.gene.symbols, cache=TRUE, eval=T}
 # Define biomart object
@@
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(Comparison) %>%
+  dplyr::select(Comparison, hgnc_symbol) %>%
   group_by(Comparison) %>%

### 2016-01-27 code/Rmd/Mayo_TemporalCortex_RNASeq.Rmd
Added differential expression to Mayo temporal cortex data

@@
 METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
+METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
 METADATA <- METADATA %>%
@@
 Clustering of initial normalised data (with NULL design)
-```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
+```{r decompse.normalise.data, fig.height=9, fig.width=16, results='asis'}
 # Find principal components of expression to plot
@@
 p <- p + theme_bw() + theme(legend.position="top") 
-p <- p + geom_text(aes(label= ID), size=4, hjust=0)
+# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
 p
@@
 ```{r addVarsBack}
-varsToAddBackIn <- c("DiagnosisAD", "DiagnosisControl", "DiagnosisPSP", "DiagnosisPathologic Aging")
+varsToAddBackIn <- c("DiagnosisAD", "DiagnosisControl", "DiagnosisPSP", "DiagnosisPathologic_Aging")
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
@@
 ### Clustering residual gene expression
-```{r decompse.normalise.data1, fig.height=8, fig.width=8, results='asis'}
+```{r decompse.normalise.data1, fig.height=10, fig.width=16, results='asis'}
 # Find principal components of expression to plot
@@
 ```
+### Differential expression analysis
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diffExp}
+# Fit contrast
+contrast = makeContrasts(contrasts=c("DiagnosisControl-DiagnosisAD",
+                                     "DiagnosisControl-DiagnosisPSP",
+                                     "DiagnosisControl-DiagnosisPathologic_Aging",
+                                     "DiagnosisPathologic_Aging-DiagnosisPSP",
+                                     "DiagnosisPathologic_Aging-DiagnosisAD",
+                                     "DiagnosisPSP-DiagnosisAD"), 
+                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
+FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = list()
+p = list()
+for (i in 1:6){
+  DE[[i]] = topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
+  DE[[i]] = rownameToFirstColumn(DE[[i]], 'ensembl_gene_id') %>%
+    left_join(Ensemble2HGNC)
+  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
+  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red') 
+  p[[i]] <- p[[i]] + ggtitle(gsub('Diagnosis','',colnames(contrast)[i]))
+}
+multiplot(plotlist=p, cols = 3)
+names(DE) = gsub('Diagnosis','',colnames(contrast))
+
+DE = ldply(DE, .id = 'Comparison')
+tmp = DE %>% 
+  dplyr::filter(adj.P.Val <= 0.05) %>%
+  group_by(Comparison) %>%
+  dplyr::summarize(nDiffExpGenes = length(unique(hgnc_symbol)),
+                   DiffExpGenes = paste(unique(hgnc_symbol), collapse = ','))
+kable(tmp)
+```
+
 ### Store files in synapse

### 2016-01-27 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
Added differential expression to the Mayo RNASeq markdown

@@
 ### Data preprocessing
-```{r preprocess.data,cache=TRUE, echo=TRUE}
+```{r preprocess.data, echo=TRUE}
 # Fix metadata
 METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
+METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
 METADATA <- METADATA %>%
@@
 Determine relationship between covariates. 
-```{r covariates.clustering, cache=TRUE}
+```{r covariates.clustering}
 FactorCovariates <- c("Source", "Diagnosis",  "Sex",  "ApoE", "Flowcell")
@@
 Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
-```{r iterative.adjusted.voom.normalisation, cache=TRUE, results='asis', echo=FALSE}
+```{r iterative.adjusted.voom.normalisation, results='asis', echo=FALSE}
 postAdjustCovars = c('Diagnosis')
@@
 ```
-
 ```{r get.gene.symbols, cache=TRUE}
@@
   dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+```
+### Differential expression analysis
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diffExp}
+# Fit contrast
+contrast = makeContrasts(contrasts=c("DiagnosisControl-DiagnosisAD",
+                                     "DiagnosisControl-DiagnosisPSP",
+                                     "DiagnosisControl-DiagnosisPathologic_Aging",
+                                     "DiagnosisPathologic_Aging-DiagnosisPSP",
+                                     "DiagnosisPathologic_Aging-DiagnosisAD",
+                                     "DiagnosisPSP-DiagnosisAD"), 
+                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
+FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = list()
+p = list()
+for (i in 1:6){
+  DE[[i]] = topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
+  DE[[i]] = rownameToFirstColumn(DE[[i]], 'ensembl_gene_id') %>%
+    left_join(Ensemble2HGNC)
+  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
+  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red') 
+  p[[i]] <- p[[i]] + ggtitle(gsub('Diagnosis','',colnames(contrast)[i]))
+}
+multiplot(plotlist=p, cols = 3)
+names(DE) = gsub('Diagnosis','',colnames(contrast))
+
+DE = ldply(DE, .id = 'Comparison')
+tmp = DE %>% 
+  dplyr::filter(adj.P.Val <= 0.05) %>%
+  group_by(Comparison) %>%
+  dplyr::summarize(nDiffExpGenes = length(unique(hgnc_symbol)),
+                   DiffExpGenes = paste(unique(hgnc_symbol), collapse = ','))
+kable(tmp)
 ```

### 2016-01-27 code/Rmd/MetaAnalysis.Rmd
Added markdown to perform meta diff expression analysis on AMPAD cohorts

---
title: "Meta analysis of differential expression results from AMP-AD (Control vs AD)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)
synapseLogin()

knit2synapse::knitToFolderEntity(file = "./MetaAnalysis.Rmd", 
                                 parentId ="syn5569102",
                                 entityName = 'MetaAnalysis of Control vs AD Data')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file is

# Clear R console screen output
cat("\014")

# Load required libraries
library(synapseClient)
library(RColorBrewer)
library(ggplot2)
library(data.table)
library(plyr)
library(dplyr)
library(psych)
library(tidyr)

library(limma)
library(sva)
library(RankProd)

library(knitr)
library(rGithubClient) ## Needs the dev branch

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

synapseLogin()

# source utility files from ../R/lib folder
file.sources = list.files('../R/lib',pattern="*.R", full.names = T)
tmp = sapply(file.sources,source,.GlobalEnv)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r fxns}
# Set ggplot theme
my.theme = theme(title = element_text(size = 24), 
                 axis.text = element_text(size = 20), 
                 axis.text.x = element_text(angle = 90, hjust=1, vjust = 0.5))

# Utility function to download files from synapse
download.file <- function(id){ tmp = fread(synGet(id)@filePath, data.table=F, header=T);  }

# Utility function for combining p-values using Fishers inverse chi-square method
fishersPvalue <- function(pvalue){
  pvalue = -2*log(pvalue, base = exp(1))
  S <- 0
  N <- dim(pvalue)[2]

  correlation <- WGCNA::bicor(pvalue,use = 'pairwise.complete.obs');

  for (i in 1:(N-1))
    for (j in (i+1):N){
      if (correlation[i,j] > 0){
        S <- S + correlation[i,j]*(3.25+0.75*correlation[i,j])
      } else if (correlation[i,j] <= 0){
        S <- S + correlation[i,j]*(3.27+0.71*correlation[i,j])
      }
    }

  combinedPvalue.chi <- rowSums(pvalue)*(4*N)/(4*N+2*S)
  combinedPvalue <- sapply(combinedPvalue.chi, function(x,N){
    pchisq(x, df = 2*N, lower.tail = FALSE)
  },N)
  names(combinedPvalue) = rownames(pvalue)

  return(combinedPvalue)
}
```

### Download differential expression data
```{r download.data}
# Get all differential expression files and their synIds
DEXP_ID = c(ROSMAP.PFC = 'syn5608845',
            MSSB = 'syn5609009',
            MAYO.CER = 'syn5609850',
            MAYO.TC = 'syn5609813',
            MAYO.CER.AR = 'syn5607457',
            MAYO.TC.AR = 'syn5607475')

# Download differential expression data
DEXP = lapply(DEXP_ID, download.file)

# Extract subset of differential expression data
DEXP$ROSMAP.PFC = DEXP$ROSMAP.PFC %>% 
  dplyr::filter(Comparison == "cogdx1-cogdx4") %>%
  dplyr::select(hgnc_symbol, P.Value)

DEXP$MSSB.FP = DEXP$MSSB %>% 
  dplyr::filter(Comparison == "BM_10.Control-BM_10.AD") %>%
  dplyr::select(hgnc_symbol, P.Value)

DEXP$MSSB.STG = DEXP$MSSB %>% 
  dplyr::filter(Comparison == "BM_22.Control-BM_22.AD") %>%
  dplyr::select(hgnc_symbol, P.Value)

DEXP$MSSB.PHG = DEXP$MSSB %>% 
  dplyr::filter(Comparison == "BM_36.Control-BM_36.AD") %>%
  dplyr::select(hgnc_symbol, P.Value)

DEXP$MSSB = NULL

DEXP$MAYO.TC = DEXP$MAYO.TC %>% 
  dplyr::select(hgnc_symbol, P.Value)

DEXP$MAYO.CER.AR = DEXP$MAYO.CER.AR %>% 
  dplyr::select(hgnc_symbol, P.Value)

DEXP$MAYO.TC.AR = DEXP$MAYO.TC.AR %>% 
  dplyr::select(hgnc_symbol, P.Value)

GNames = DEXP[[1]]$hgnc_symbol
for (i in 2: length(DEXP)){
  GNames = intersect(GNames, DEXP[[i]]$hgnc_symbol)
}
GNames = GNames[!is.na(GNames)]
GNames = GNames[GNames != '']

DEXP = lapply(DEXP, function(dexp, GNames){
  dexp = dexp %>%
    dplyr::filter(hgnc_symbol %in% GNames) %>%
    group_by(hgnc_symbol) %>%
    dplyr::summarise(P.Value = min(P.Value)) %>%
    data.frame
  rownames(dexp) = dexp$hgnc_symbol
  dexp = dexp[GNames,]
  }, GNames)

DEXP = mapply(function(dexp, dataName){
  colnames(dexp) = c('hgnc_symbol', dataName)
  return(dexp)
}, DEXP, names(DEXP), SIMPLIFY = F)

DEXP = join_all(DEXP, by = "hgnc_symbol")
```
Overall `r dim(DEXP)[1]` genes in `r dim(DEXP)[2]-1` studies are analysed
### Differential expression analysis
```{r diffExp, results = 'asis'}
Adj.P.Value = apply(DEXP[,-(1)], 2, p.adjust, method = "fdr")
rownames(Adj.P.Value) = DEXP$hgnc_symbol

writeLines('Total number of genes that are differentially expressed at a FDR <= 0.05 are')
tmp = (data.frame(colSums(Adj.P.Value <= 0.05)))
colnames(tmp) = 'No. of Genes'
kable(tmp)

writeLines('Genes that are differentially expressed at a FDR <= 0.05 in 7/8 datasets')
rownames(Adj.P.Value)[which(rowSums(Adj.P.Value<=0.05)==7)]

writeLines('Genes that are differentially expressed at a FDR <= 0.05 in 6/8 datasets')
rownames(Adj.P.Value)[which(rowSums(Adj.P.Value<=0.05)==6)]
```

### Minimum Pvalue meta analysis
```{r minP}
Pval = DEXP

Pmin = data.frame(hgnc_symbols = Pval[,1], Pmin = rowMin(data.matrix(Pval[,-(1)])))
Pmin$Pmin = p.adjust(Pmin$Pmin, method = 'fdr')

tmp = dplyr::filter(Pmin, Pmin <= 0.05) 
tmp = data.frame(No.of.Genes = dim(tmp)[1], 
                 GeneList = tmp %>%
                   dplyr::select(hgnc_symbols) %>%
                   unlist %>% paste(collapse = ','))
writeLines("Genes that are atleast differentially expressed at an FDR <= 0.05 in one study are")
kable(tmp)
```

### Fisher-Brown's meta analysis
```{r fisher.brown}
Pfb = data.frame(hgnc_symbols = Pval[,1], 
                 Pfb = p.adjust(fishersPvalue(Pval[,-(1)]), method = 'fdr'))

Pfb$Pfb = p.adjust(Pfb$Pfb, method = 'fdr')

tmp = dplyr::filter(Pfb, Pfb <= 0.05) 
tmp = data.frame(No.of.Genes = dim(tmp)[1], 
                 GeneList = tmp %>%
                   dplyr::select(hgnc_symbols) %>%
                   unlist %>% paste(collapse = ','))

writeLines("Genes that are differentially expressed in the meta analysis at an FDR <= 0.05 are")
kable(tmp)
```

### Store results in synapse
```{r synapse.store, include=FALSE, cache=FALSE}
thisFileName <- 'MetaAnalysis.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='metaAnal')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))

# Store consolidated data in synapse
CODE = Folder(name = 'MetaAnalysis of Control vs AD Data', parentId = "syn5569102")
CODE = synStore(CODE)

# Store pvalues
tmp = join_all(list(Pmin,Pfb))
write.table(tmp, file = './Pval.tsv', sep = '\t', row.names=F, quote=F)
FILE_OBJ = File("./Pval.tsv", name = 'Meta Differential Expression Analysis FDR', parentId = CODE$properties$id)
FILE_OBJ = synStore(FILE_OBJ, 
                    used = as.character(c(EXP_ID, COV_ID)),
                    executed = thisFile,
                    activityName = 'Differential expression analysis of control-AD')

# Store differentially expressed genes
write.table(rownameToFirstColumn(Adj.P.Value, 'hgnc_symbol'), file = './DiffGenes.tsv', sep = '\n', row.names=F, quote=F, col.names = F)
FILE_OBJ = File("./DiffGenes.tsv", name = 'Differentially expressed genes FDR (individual studies)', parentId = CODE$properties$id)
FILE_OBJ = synStore(FILE_OBJ, 
                    used = as.character(c(EXP_ID, COV_ID)),
                    executed = thisFile,
                    activityName = 'Differential expression analysis of control-AD')
```

### 2016-01-27 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
Minor modifications

--- 2016-01-25 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
@@
 METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
+METADATA$Diagnosis = gsub("Pathologic Aging", "Pathologic_Aging", METADATA$Diagnosis)
 METADATA <- METADATA %>%
@@
 ```
+### Differential expression analysis
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diffExp}
+# Fit contrast
+contrast = makeContrasts(contrasts=c("DiagnosisControl-DiagnosisAD",
+                                     "DiagnosisControl-DiagnosisPathologic_Aging",
+                                     "DiagnosisControl-DiagnosisPSP",
+                                     "DiagnosisPathologic_Aging-DiagnosisAD",
+                                     "DiagnosisPathologic_Aging-DiagnosisAD",
+                                     "DiagnosisPSP-DiagnosisAD"),
+                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
+FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = topTable(FIT.CONTR, coef=1, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
+DE = rownameToFirstColumn(DE, 'ensembl_gene_id') %>%
+  left_join(Ensemble2HGNC)
+p <- ggplot(DE, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
+p <- p + geom_hline(yintercept = -log10(0.05))
+p
+
+tmp = DE %>% dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(hgnc_symbol) %>%
+  unlist %>% unique
+kable(data.frame(Comparison = "Control_vs_AD",
+           nDiffExp = length(tmp),
+           DiffExpGenes = paste(tmp, collapse = ',')))  
+```
+
 ### Store files in synapse

### 2016-01-27 code/Rmd/MetaAnalysis.Rmd
Minor modifications

@@
 library(tidyr)
+library(Matrix)
 
@@
 # source utility files from ../R/lib folder
-file.sources = list.files('../R/lib',pattern="*.R", full.names = T)
-tmp = sapply(file.sources,source,.GlobalEnv)
+source('../R/lib/rownameToFirstColumn.R')
+# file.sources = list.files('../R/lib/rownfile.sources',pattern="row*.R", full.names = T)
+# tmp = sapply(file.sources,source,.GlobalEnv)
 
@@
 writeLines('Genes that are differentially expressed at a FDR <= 0.05 in 7/8 datasets')
-rownames(Adj.P.Value)[which(rowSums(Adj.P.Value<=0.05)==7)]
+paste(rownames(Adj.P.Value)[which(rowSums(Adj.P.Value<=0.05)==7)], collapse = ',')
 
 writeLines('Genes that are differentially expressed at a FDR <= 0.05 in 6/8 datasets')
-rownames(Adj.P.Value)[which(rowSums(Adj.P.Value<=0.05)==6)]
+paste(rownames(Adj.P.Value)[which(rowSums(Adj.P.Value<=0.05)==6)], collapse = ',')
 ```
@@
 FILE_OBJ = synStore(FILE_OBJ, 
-                    used = as.character(c(EXP_ID, COV_ID)),
+                    used = as.character(DEXP_ID),
                     executed = thisFile,

### 2016-01-27 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
Minor modifications

--- 2016-01-25 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
@@
 ```
-Following five samples were considered outliers and removed 742\_CER, 11386\_CER, 11311\_CER, 1923\_CER, 11477\_CER
+Following five samples were considered outliers and removed 742\_CER, 11386\_CER, 11311\_CER, 1923\_CER, 11477\_CER, based on the eucledian distance from other samples and/or seperated on the first 2 PC's
 ```{r preprocess.data1}
@@
 # Define biomart object
-mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "www.ensembl.org")
 
@@
 |  Residual Expression  |  `r EXP_OBJ$properties$id`      |
+
+### Source Code
+[This File](`r thisFile`)

### 2016-01-27 code/Rmd/Mayo_TemporalCortex_RNASeq.Rmd

@@
 
-```{r synapse.parameters, include=FALSE, cache=TRUE}
+```{r synapse.parameters, include=FALSE}
 # Synapse parameters
@@
 ### Data preprocessing
-```{r preprocess.data,cache=TRUE, echo=TRUE}
+```{r preprocess.data,cache=FALSE, echo=TRUE}
 # Fix metadata
@@
 Following five samples were considered outliers and removed 1923\_TCX
-```{r preprocess.data1,cache=TRUE}
+```{r preprocess.data1}
 indToRetain = intersect(METADATA$ID, colnames(COUNT))
@@
 Determine relationship between covariates. 
-```{r covariates.clustering, cache=TRUE}
+```{r covariates.clustering}
 FactorCovariates <- c("Source", "Diagnosis",  "FLOWCELL")
@@
 
-```{r get.gene.symbols, cache=TRUE}
+```{r get.gene.symbols}
 # Define biomart object

### 2016-01-28 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd

--- 2016-01-27 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
@@
 |  Differential Expression  |  `r DEXP_OBJ$properties$id`      |
+
+### Source Code
+[This File](`r thisFile`)

### 2016-01-29 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
Minor changes

@@
 Clustering of initial normalised data (with NULL design)
-```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
+```{r decompse.normalise.data1, fig.height=8, fig.width=8, results='asis'}
 # Find principal components of expression to plot
@@
 p
-
+```
+```{r decompse.normalise.data2, fig.height=10, fig.width=25, results='asis'}
 # Tree based clustering
@@
 ### Clustering residual gene expression
-```{r decompse.normalise.data1, fig.height=8, fig.width=8, results='asis'}
+```{r decompse.normalise.data3, fig.height=8, fig.width=8, results='asis'}
 # Find principal components of expression to plot
@@
 p
-
+```
+```{r decompse.normalise.data4, fig.height=8, fig.width=25, results='asis'}
 # Tree based clustering
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp}
+```{r diffExp, fig.height=10, fig.width=16}
 # Fit contrast

### 2016-01-29 code/Rmd/Mayo_Cerebellum_Array.Rmd

@@
                    executed = thisFile, activityDescription = activityDescription)
+
+# Store differential expression results
+write.table(DE, file = 'diffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('diffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
+DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                   executed = thisFile, activityDescription = activityDescription)
+
 ```

### 2016-01-29 code/Rmd/Mayo_TemporalCortex_RNASeq.Rmd

@@
 ```
-Following five samples were considered outliers and removed 1923\_TCX
+Following samples were considered outliers and removed based on PCA and/or eucledian distance: 1923\_TCX
 ```{r preprocess.data1}
@@
 Initial normalisation usign voom (with NULL design)
+
 ```{r initial.voom.normalisation}
@@
 ```
+
 Clustering of initial normalised data (with NULL design)
-```{r decompse.normalise.data, fig.height=9, fig.width=16, results='asis'}
+```{r decompse.normalise.data1, fig.height=10, fig.width=10, results='asis'}
 # Find principal components of expression to plot
@@
 p
-
+```
+```{r decompse.normalise.data2, fig.height=9, fig.width=25, results='asis'}
 # Tree based clustering
@@
 VOOM.ADJUSTED.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, 
-                                     design=DM1$design, plot=T)
+                                     design=DM1$design, plot=F)
 
@@
 ### Clustering residual gene expression
-```{r decompse.normalise.data1, fig.height=10, fig.width=16, results='asis'}
+```{r decompse.normalise.data3, fig.height=10, fig.width=9, results='asis'}
 # Find principal components of expression to plot
@@
 p
-
+```
+```{r decompse.normalise.data4, fig.height=10, fig.width=25, results='asis'}
 # Tree based clustering
@@
 |  Differential Expression  |  `r DEXP_OBJ$properties$id`      |
+
+### Source R Code
+[This File](`r thisFile`)

### 2016-01-29 code/Rmd/ROSMAP_RNASeq.Rmd

@@
 library(ggplot2)
+library(WGCNA)
 
@@
 tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION[,-(1:2)])))
-plotDendroAndColors(tree, 
+WGCNA::plotDendroAndColors(tree, 
                     colors = labels2colors(COVARIATES.tmp), 
@@
 
-DE = DE %>% plyr::ldply(.id = 'Comparison')
+DE = DE %>% plyr::ldply(.id = 'Comparison') %>%
+  tidyr::separate(ensembl_gene_id, into=c("ensemble_id","position")) %>%
+  dplyr::rename(ensembl_gene_id = ensemble_id) %>%
+  left_join(Ensemble2HGNC)
 

### 2016-01-30 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
Minor modifications

--- 2016-01-28 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
@@
 author: "Thanneer Perumal"
-date: "`r date()`"
-output: html_document
+date: '`r date()`'
+output:
+  html_document: default
+  pdf_document:
+    keep_tex: yes
+    latex_engine: xelatex
 ---
@@
 library(data.table)
+library(plyr)
 library(dplyr)
@@
 
-```{r synapse.parameters, include=FALSE, cache=TRUE}
+```{r synapse_parameters, include=FALSE, cache=TRUE}
 # Synapse parameters
@@
 #### Obtain count matrix and metadata from synapse.
-```{r download.data, cache=TRUE}
+```{r download_data, cache=TRUE}
 # Download expression data
@@
 ### Data preprocessing
-```{r preprocess.data, echo=TRUE}
+```{r preprocess_data, echo=TRUE}
 # Fix metadata
@@
 Following five samples were considered outliers and removed 742\_CER, 11386\_CER, 11311\_CER, 1923\_CER, 11477\_CER, based on the eucledian distance from other samples and/or seperated on the first 2 PC's
-```{r preprocess.data1}
+```{r preprocess_data1}
 indToRetain = intersect(METADATA$SampleID, colnames(COUNT))
@@
 Determine relationship between covariates. 
-```{r covariates.clustering}
+```{r covariates_clustering}
 FactorCovariates <- c("Source", "Diagnosis",  "Sex",  "ApoE", "Flowcell")
@@
 Covariate correlation
-```{r covariates.correlation, fig.width=10, fig.height=10}
+```{r covariates_correlation, fig.width=10, fig.height=10}
 COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.1)
@@
 Retain genes with atleast 50% of the samples have >= 1 gene counts
-```{r gene.filtering, results='asis', echo=FALSE}
+```{r gene_filtering, results='asis', echo=FALSE}
 # Remove genes
@@
 Initial normalisation usign voom (with NULL design)
-```{r initial.voom.normalisation}
+```{r initial_voom_normalisation}
 # Initial normalisation of gene expression
@@
 Clustering of initial normalised data (with NULL design)
-```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
+```{r decompse_normalise_data, fig.height=8, fig.width=8, results='asis'}
 # Find principal components of expression to plot
@@
 Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
-```{r preAdjusted.covariates, results='asis', echo=FALSE}
+```{r preAdjusted_covariates, results='asis', echo=FALSE}
 # Find correlation between PC's of gene expression with covariates
@@
 Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
-```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
+```{r preAdjustedSigCovars_NULL_ALL, fig.width=25, fig.height=12}
 preAdjustedSigCovars[["PC_res"]][[2]]$plotData
@@
 Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
-```{r iterative.adjusted.voom.normalisation, results='asis', echo=FALSE}
+```{r iterative_adjusted_voom_normalisation, results='asis', echo=FALSE}
 postAdjustCovars = c('Diagnosis')
@@
 
-```{r adjusted.voom.normalisation, results='asis', echo=FALSE}
+```{r adjusted_voom_normalisation, results='asis', echo=FALSE}
 # Post adjusted design matrix
@@
 Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
-```{r calculate.residuals, include=FALSE, results = 'asis'}
+```{r calculate_residuals, include=FALSE, results = 'asis'}
 # Residuals after normalisation
@@
 ```
-```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
+```{r residualSigCovars_ALL, fig.width=25, fig.height=12}
 residualSigCovars[["PC_res"]][[2]]$plotData
@@
 ### Clustering residual gene expression
-```{r decompse.normalise.data1, fig.height=8, fig.width=8, results='asis'}
+```{r decompse_normalise_data1, fig.height=8, fig.width=8, results='asis'}
 # Find principal components of expression to plot
@@
 ```
-```{r get.gene.symbols, cache=TRUE}
+```{r get_gene_symbols, cache=TRUE}
 # Define biomart object
@@
 ### Store files in synapse
-```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
+```{r synapse_store, include=FALSE, eval=TRUE, cache=FALSE}
 # Code

### 2016-01-30 code/Rmd/Mayo_TemporalCortex_RNASeq.Rmd
Minor modifications

--- 2016-01-27 code/Rmd/Mayo_TemporalCortex_RNASeq.Rmd
@@
 knit2synapse::knitToFolderEntity(file = "./Mayo_TemporalCortex_RNASeq.Rmd",
-                                 parentId ="syn5570325",
+                                 parentId ='syn5612983',#"syn5570325",
                                  entityName = 'UFL_Mayo_ISB Temporal Cortex RNASeq')
@@
 library(biomaRt)
+library(fpc)
 
@@
 # Synapse parameters
-parentId = 'syn5570325';
+parentId = 'syn5612983'#'syn5570325';
 activityName = 'Covariate analysis';
@@
 
+# Tree based clustering
+sampleTrees = flashClust::hclust(dist(t(RESIDUAL.GENE_EXPRESSION[,-(1)])), method = "average")
+
+# Cluster expression
+clust = cutreeStatic(sampleTrees, cutHeight = 100, minSize = 10)
+
 # Plot first 2 PCs
@@
                        PC1=PC$rotation[,1], 
-                       PC2=PC$rotation[,2])
+                       PC2=PC$rotation[,2],
+                       clust = clust)
 
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=factor(Diagnosis), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
+p <- p + geom_point(aes(color=factor(clust), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
 p <- p + theme_bw() + theme(legend.position="top") 
@@
 
-# Tree based clustering
-sampleTrees = flashClust::hclust(dist(t(RESIDUAL.GENE_EXPRESSION[,-(1)])), method = "average")
+names(clust) = sampleTrees$labels
+clust = clust[rownames(COVARIATES)]
 
 plotDendroAndColors(sampleTrees, 
-                    colors = labels2colors(COVARIATES), 
+                    colors = labels2colors(cbind(data.frame(clust = clust), COVARIATES)), 
                     abHeight = 0.80, 
@@
 dev.off()
+
+writeLines(paste('Samples that deviate from others are', paste(sampleTrees$labels[clust==0], collapse = ',')))
 ```

### 2016-01-30 code/Rmd/Mayo_TemporalCortex_RNASeq.Rmd

--- 2016-01-29 code/Rmd/Mayo_TemporalCortex_RNASeq.Rmd
@@
 p
+
+names(clust) = sampleTrees$labels
+clust = clust[rownames(COVARIATES)]
 ```
@@
 # Tree based clustering
-sampleTrees = flashClust::hclust(dist(t(RESIDUAL.GENE_EXPRESSION[,-(1)])), method = "average")
-
 plotDendroAndColors(sampleTrees, 
-                    colors = labels2colors(COVARIATES), 
+                    colors = labels2colors(cbind(data.frame(clust = clust), COVARIATES)), 
                     abHeight = 0.80, 
@@
 
-DE = ldply(DE, .id = 'Comparison')
+DE = plyr::ldply(DE, .id = 'Comparison')
 tmp = DE %>% 
@@
   group_by(Comparison) %>%
-  dplyr::summarize(nDiffExpGenes = length(unique(hgnc_symbol)),
-                   DiffExpGenes = paste(unique(hgnc_symbol), collapse = ','))
+  dplyr::summarize(nDiffExpGenes = length(unique(hgnc_symbol)))
 kable(tmp)

### 2016-02-02 code/Rmd/Mayo_TemporalCortex_RNASeq.Rmd
Minor modifications

@@
   cache = TRUE)
+dev.off()
 ```
@@
 sampleTrees = flashClust::hclust(dist(t(VOOM.GENE_EXPRESSION$E)), method = "average")
-tree = WGCNA::cutreeStatic(sampleTrees, cutHeight = 200, minSize=3)
+tree = WGCNA::cutreeStatic(sampleTrees, cutHeight = 135, minSize=10)
 
@@
 dev.off()
+
+tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[tree == 0], collapse = ','))
+tmp = base::gsub('_','\\\\_',tmp)
+writeLines(tmp)
+tmp = paste('Samples in cluster 2 are', paste(sampleTrees$labels[tree == 2], collapse = ','))
+tmp = base::gsub('_','\\\\_',tmp)
+writeLines(tmp)
 ```
@@
 
-writeLines(paste('Samples that deviate from others are', paste(sampleTrees$labels[clust==0], collapse = ',')))
+tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[tree == 0], collapse = ','))
+tmp = base::gsub('_','\\\\_',tmp)
+writeLines(tmp)
 ```

### 2016-02-02 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd

@@
 date: '`r date()`'
-output:
-  html_document: default
-  pdf_document:
-    keep_tex: yes
-    latex_engine: xelatex
+output: "html_document"
 ---
@@
 Clustering of initial normalised data (with NULL design)
-```{r decompse_normalise_data, fig.height=8, fig.width=8, results='asis'}
+```{r decompse.normalise.data1, fig.height=8, fig.width=8, results='asis'}
 # Find principal components of expression to plot
@@
 ### Clustering residual gene expression
-```{r decompse_normalise_data1, fig.height=8, fig.width=8, results='asis'}
+```{r decompse.normalise.data3, fig.height=8, fig.width=8, results='asis'}
 # Find principal components of expression to plot

### 2016-02-02 code/Rmd/Mayo_TemporalCortex_RNASeq.Rmd

--- 2016-01-30 code/Rmd/Mayo_TemporalCortex_RNASeq.Rmd
@@
 knit2synapse::knitToFolderEntity(file = "./Mayo_TemporalCortex_RNASeq.Rmd",
-                                 parentId ='syn5612983',#"syn5570325",
+                                 parentId = "syn5570325",
                                  entityName = 'UFL_Mayo_ISB Temporal Cortex RNASeq')
@@
 # Synapse parameters
-parentId = 'syn5612983'#'syn5570325';
+parentId = 'syn5570325';
 activityName = 'Covariate analysis';

### 2016-02-09 code/Rmd/ROSMAP_RNASeq.Rmd
Minor bug fix in ROSMAP RNASeq processing

@@
 METADATA = METADATA[indToRetain,]
+
+# Arrange rows of metadata in the same order as columns of expression
+METADATA = METADATA[colnames(EXPR),]
 ```

### 2016-02-09 code/Rmd/ROSMAP_RNASeq.Rmd
Minor modifications

@@
 ### Outliers detection
-```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
+```{r decompse.normalise.data1, fig.height=8, fig.width=8, results='asis'}
 # Find principal components of expression to plot
@@
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
-
+```
+```{r decompse.normalise.data2, fig.height=8, fig.width=16, results='asis'}
 tree = hclust(as.dist(t(EXPR)))
@@
                     groupLabels = colnames(COVARIATES.tmp))
+dev.off()
 ```
@@
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
-
+```
+```{r decompse.normalise.data4, fig.height=8, fig.width=16, results='asis'}
 tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION[,-(1:2)])))

### 2016-02-28 code/Rmd/synNetGen.Rmd
Added markdown for generating synthetic networks

---
title: "Synthetic network and data generation using GeneNetWeaver(GNW)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./synNetGen.Rmd",
                                 parentId ="syn5569099",
                                 entityName = 'Synthetic Networks')
```
#### Clone GNW repo from github
```{r getGNW, eval=FALSE}
wd = getwd()
setwd('../../../')

# Clone GNW repo
system('git clone https://github.com/tschaffter/gnw.git')

setwd(wd)
```
#### Initialise libraries
```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(synapseClient)
library(ggplot2)
library(reshape2)
library(data.table)
library(dplyr)
library(stringr)
library(limma)
library(knitr)
library(Biobase)
library(GEOquery)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(ggplot2)
library(WGCNA)

## Needs the dev branch
library(rGithubClient)

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

synapseLogin()

# source utility files from ../R/lib folder
file.sources = list.files('../R/lib',pattern="*.R", full.names=T)
tmp = sapply(file.sources,source,.GlobalEnv)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5569099';
activityName = 'Synthetic network generation';
activityDescription = 'Synthetic network data from ecoli using GNW';

thisFileName <- 'synNetGen.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='synNet')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))

# Create folder to store results
FOLD = Folder(name = 'Synthetic Networks', parentId = parentId)
FOLD = synStore(FOLD)
```
#### Extract network
```{r extractNet}
# Extract settings text file
con = file('../../../gnw//sandbox/settings.txt')
settings.txt = readLines(con)
close(con)

netSize = c(a = 1e1, b = 1e2, c = 1e3, d = 1e4)
for (net in names(netSize))){
  # Number of minimum regulators
  minReg = max(5, round(netSize[[net]]*0.5))

  # Create output directory
  outputDir = paste0('SubNetSize',net)
  system(paste0('mkdir ', outputDir))

  # Write settings file to the output directory
  con = file(paste0(outputDir,'/settings.txt'))
  writeLines(settings.txt, con)
  close(con)

  # Extract 10 subnetworks of size n from ecoli data
  system(paste('java','-jar','../../../gnw/gnw-3.1.2b.jar',
                 '--extract','-c',paste0(outputDir,'/settings.txt'), '--input-net',
                 '../../../gnw/sandbox/ecoli_transcriptional_network_regulonDB_6_7.tsv',
                 '--random-seed','--greedy-selection','--num-subnets=10',
                 paste0('--subnet-size=',netSize[[net]]),'--output-net-format=4',
                 paste0('--output-path=',outputDir),
                 paste0('--min-num-regulators=',minReg)))

  # Convert generated subnetworks to tsv data
  FNames = dir(outputDir, pattern='*.xml',full.names=T)
  for (fnames in FNames){
    system(paste('java','-jar','../../../gnw/gnw-3.1.2b.jar',
                 '--transform','-c',paste0(outputDir,'/settings.txt'), '--input-net',
                 fnames, '--output-net-format=0',
                 paste0('--output-path=',outputDir)))
  }

  # Change settings for simulations  
  tmp = settings.txt
  tmp[grep('outputDirectory = ', tmp)] = paste('outputDirectory =',outputDir)
  tmp[grep('ssDualKnockouts',tmp)] = "ssDualKnockouts = 0"
  tmp[grep('tsDREAM4TimeSeries',tmp)] = "tsDREAM4TimeSeries = 0"

  con = file(paste0(outputDir,'/settings.txt'))
  writeLines(tmp, con)
  close(con)

  # Simulate from extracted networks
  for (fnames in FNames){
    system(paste('java','-jar','../../../gnw/gnw-3.1.2b.jar',
                 '--simulate','-c',paste0(outputDir,'/settings.txt'), 
                 '--input-net', fnames, '--output-path',outputDir))

    # Read all data into one file
    allFiles = dir(pattern = 'ecoli')
    system(paste('mv',allFiles[grep('.xml',allFiles)],outputDir))
    system(paste('mv',allFiles[grep('goldstandard.tsv',allFiles)],outputDir))

    fileName = tools::file_path_sans_ext(fnames) %>% str_replace(paste0(outputDir,'/'),'')

    data = list()

    extractData <- function(pat,allfiles, file.name){
      read.table(allfiles[grep(paste0(file.name,pat),allfiles)], header=T, sep = '\t', quote='')
    }

    allPatterns1 = c('_wildtype.tsv','_knockdowns.tsv', '_multifactorial.tsv', '_multifactorial_perturbations.tsv', '_knockouts.tsv')
    data[['1']] = lapply(allPatterns1, extractData, allFiles, fileName) %>%
      do.call(rbind,.)

    allPatterns2 = c('_nonoise_wildtype.tsv','_nonoise_knockdowns.tsv', '_nonoise_multifactorial.tsv', '_nonoise_knockouts.tsv')
    data[['2']] = lapply(allPatterns2, extractData, allFiles, fileName) %>%
      do.call(rbind, .)

    allPatterns3 = c('_noexpnoise_wildtype.tsv','_noexpnoise_knockdowns.tsv', '_noexpnoise_multifactorial.tsv', '_noexpnoise_knockouts.tsv')
    data[['3']] = lapply(allPatterns3, extractData, allFiles, fileName) %>%
      do.call(rbind, .)

    # Store data in synaspe
    fileNum = str_split(str_split(fnames, '-')[[1]][2],'\\.')[[1]][1]
    SUB.FOLD = Folder(name = paste0(outputDir, '_', fileNum), parentId = FOLD$properties$id)
    SUB.FOLD = synStore(SUB.FOLD)

    objs = lapply(names(data), function(x, Data, subFolder){
      tmp = data.matrix(Data[[x]]) %>% t %>% rownameToFirstColumn('GeneNames')
      colnames(tmp)[-1] = paste0('EXP',colnames(tmp)[-1])

      # Store data in synapse
      write.table(tmp, file = paste0('Expression_',x,'.tsv'), sep = '\t', row.names=F, quote=F)
      OBJ = File(paste0('Expression_',x,'.tsv'), 
                 name = paste('Expression',x), 
                 parentId = subFolder$properties$id)
      OBJ = synStore(OBJ, activityName = activityName,
                     executed = thisFile, activityDescription = activityDescription)
    }, data, SUB.FOLD)
    system('rm ecoli*')
  }
}
```

### 2016-02-28 code/Rmd/synNetGen.Rmd
Minor changes

@@
 library(synapseClient)
-library(ggplot2)
-library(reshape2)
 library(data.table)
@@
 library(stringr)
-library(limma)
 library(knitr)
-library(Biobase)
-library(GEOquery)
-library(RColorBrewer)
-library(gplots)
-library(WGCNA)
-library(psych)
-library(edgeR)
-library(biomaRt)
-library(ggplot2)
-library(WGCNA)
 
 ## Needs the dev branch
-library(rGithubClient)
+library(githubr)
 

### 2016-02-29 code/Rmd/synNetGen.Rmd
Added special ACL privileges to gold standard

@@
     }, data, SUB.FOLD)
+
+    OBJ = File(paste0(tools::file_path_sans_ext(fnames), '.tsv'),
+               name = 'Gold Standard', parentId = SUB.FOLD$properties$id)
+    OBJ = synStore(OBJ, activityName = activityName, executed = thisFile, activityDescription = activityDescription)
+
+    # Set private acl to the gold standard file
+    resAcc = ResourceAccess(principalId = 3319864, 
+                            accessType = c('CREATE', 'READ', 'UPDATE', 'DELETE', 'CHANGE_PERMISSIONS', 'DOWNLOAD', 'UPLOAD', 'PARTICIPATE', 'SUBMIT', 'READ_PRIVATE_SUBMISSION', 'UPDATE_SUBMISSION', 'DELETE_SUBMISSION', 'TEAM_MEMBERSHIP_UPDATE', 'SEND_MESSAGE', 'CHANGE_SETTINGS'))
+    acl = AccessControlList(id = OBJ$properties$id, resourceAccess = ResourceAccessList(resAcc))
+    tmp = synCreateEntityACL(acl)
+
     system('rm ecoli*')

### 2016-02-29 code/R/synNetGen.R
Added R code for generating synthetic networks

@@ -1,0 +1,142 @@
+## It is assumed your working directory is where this file
+
+## Load required libraries
+library(synapseClient)
+library(data.table)
+library(dplyr)
+library(stringr)
+library(knitr)
+
+## Needs the dev branch
+library(githubr)
+
+synapseLogin()
+
+# source utility files from ../R/lib folder
+source('./lib/rownameToFirstColumn.R')
+
+# Synapse parameters
+parentId = 'syn5569099';
+activityName = 'Synthetic network generation';
+activityDescription = 'Synthetic network data from ecoli using GNW';
+
+thisFileName <- 'synNetGen.R'
+
+# Github link
+thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='synNet')
+thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/R/',thisFileName))
+
+# Create folder to store results
+FOLD = Folder(name = 'Synthetic Networks', parentId = parentId)
+FOLD = synStore(FOLD)
+
+#### Extract network
+# Extract settings text file
+con = file('../../../gnw//sandbox/settings.txt')
+settings.txt = readLines(con)
+close(con)
+
+netSize = c(a = 1e1, b = 1e2, c = 1e3, d = 1e4)
+for (net in names(netSize)){
+  # Number of minimum regulators
+  minReg = max(5, round(netSize[[net]]*0.5))
+  
+  # Create output directory
+  outputDir = paste0('SubNetSize',net)
+  system(paste0('mkdir ', outputDir))
+
+  # Write settings file to the output directory
+  con = file(paste0(outputDir,'/settings.txt'))
+  writeLines(settings.txt, con)
+  close(con)
+
+  # Extract 10 subnetworks of size n from ecoli data
+  system(paste('java','-jar','../../../gnw/gnw-3.1.2b.jar',
+                 '--extract','-c',paste0(outputDir,'/settings.txt'), '--input-net',
+                 '../../../gnw/sandbox/ecoli_transcriptional_network_regulonDB_6_7.tsv',
+                 '--random-seed','--greedy-selection','--num-subnets=10',
+                 paste0('--subnet-size=',netSize[[net]]),'--output-net-format=4',
+                 paste0('--output-path=',outputDir),
[…]

### 2016-04-19 code/Rmd/MetaAnalysis.Rmd
Minor modifications

@@
 # Utility function to download files from synapse
-download.file <- function(id){ tmp = fread(synGet(id)@filePath, data.table=F, header=T);  }
+download.file <- function(id){ tmp = fread(synGet(as.character(id))@filePath, data.table=F, header=T);  }
 
@@
   dplyr::filter(Comparison == "cogdx1-cogdx4") %>%
-  dplyr::select(hgnc_symbol, P.Value)
+  dplyr::select(hgnc_symbol, P.Value, logFC)
 
@@
   dplyr::filter(Comparison == "BM_10.Control-BM_10.AD") %>%
-  dplyr::select(hgnc_symbol, P.Value)
+  dplyr::select(hgnc_symbol, P.Value, logFC)
 
@@
   dplyr::filter(Comparison == "BM_22.Control-BM_22.AD") %>%
-  dplyr::select(hgnc_symbol, P.Value)
+  dplyr::select(hgnc_symbol, P.Value, logFC)
 
@@
   dplyr::filter(Comparison == "BM_36.Control-BM_36.AD") %>%
-  dplyr::select(hgnc_symbol, P.Value)
+  dplyr::select(hgnc_symbol, P.Value, logFC)
 
@@
 
+DEXP$MAYO.CER = DEXP$MAYO.CER %>% 
+  dplyr::select(hgnc_symbol, P.Value, logFC)
+
 DEXP$MAYO.TC = DEXP$MAYO.TC %>% 
-  dplyr::select(hgnc_symbol, P.Value)
+  dplyr::select(hgnc_symbol, P.Value, logFC)
 
 DEXP$MAYO.CER.AR = DEXP$MAYO.CER.AR %>% 
-  dplyr::select(hgnc_symbol, P.Value)
+  dplyr::select(hgnc_symbol, P.Value, logFC)
 
 DEXP$MAYO.TC.AR = DEXP$MAYO.TC.AR %>% 
-  dplyr::select(hgnc_symbol, P.Value)
+  dplyr::select(hgnc_symbol, P.Value, logFC)
+```
 
+```{r plotDE}
+tmp = lapply(DEXP, function(x){
+  x$observed.P.val = p.adjust(x$P.Value, 'fdr')
+  x = arrange(x, observed.P.val)
+  x = mutate(x, expected.P.val = seq(0,1,1/(dim(x)[1]-1)))
+  x$dexp = 0;
+  x$dexp[x$observed.P.val <= 0.05 & x$logFC > 0] = 1;
+  x$dexp[x$observed.P.val <= 0.05 & x$logFC < 0] = -1;
+  dplyr::select(x, observed.P.val, expected.P.val, dexp)
+})
+
+p = list()
+for (i in 1:length(tmp)){
+  p[[i]] = ggplot(tmp[[i]], aes(x = -log10(expected.P.val), y = -log10(observed.P.val))) + geom_point(aes(color = factor(dexp))) 
+  p[[i]] = p[[i]] + scale_colour_manual(values=c('green','grey','red')) + theme(legend.position="none")
+  x = -log10(tmp[[i]]$expected.P.val)
+  x[4]-x[1]/
+  p[[i]] = p[[i]] + geom_abline(slope=1, intercept=0) + ggtitle(names(tmp)[i])
+}
+multiplot(plotlist = p, cols = 4)
+```
+
+```{r uniqGenes}
 GNames = DEXP[[1]]$hgnc_symbol
@@
 
+writeLines('Genes that are differentially expressed at a FDR <= 0.05 in 8/8 datasets')
+paste(rownames(Adj.P.Value)[which(rowSums(Adj.P.Value<=0.05)==8)], collapse = ',')
+
 writeLines('Genes that are differentially expressed at a FDR <= 0.05 in 7/8 datasets')
@@
 
-writeLines('Genes that are differentially expressed at a FDR <= 0.05 in 6/8 datasets')
-paste(rownames(Adj.P.Value)[which(rowSums(Adj.P.Value<=0.05)==6)], collapse = ',')
+writeLines('Genes that are differentially expressed at a FDR <= 0.05 in 6/6 RNASeq datasets')
+paste(rownames(Adj.P.Value)[which(rowSums(Adj.P.Value[,-(4:5)]<=0.05)==6)], collapse = ',')
 ```

### 2016-04-19 code/Rmd/ROSMAP_RNASeq.Rmd
Minor modifications

--- 2016-01-29 code/Rmd/ROSMAP_RNASeq.Rmd
@@
 library(ggplot2)
-library(WGCNA)
 
@@
 
-plotdata <- left_join(plotdata, METADATA)
+plotdata <- left_join(plotdata, METADATA) %>%
+  dplyr::mutate(Batch = factor(Batch),
+                age_death = as.numeric(age_death),
+                cogdx = factor(cogdx))
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=factor(Batch), shape=as.factor(cogdx), size=as.numeric(age_death)))
+p <- p + geom_point(aes(color=Batch, shape=cogdx, size=age_death))
 p <- p + theme_bw() + theme(legend.position="top") 
 # p <- p + geom_text(aes(label= Sampleid), size=4, hjust=1)
+p <- p + xlab(paste0('PC1(',round((PC$sdev^2/sum(PC$sdev^2))[1]*100),'%)'))
+p <- p + ylab(paste0('PC2(',round((PC$sdev^2/sum(PC$sdev^2))[2]*100),'%)'))
 p
@@
                     dendroLabels = FALSE, 
-                    abHeight = 0.80, 
+                    abHeight = 5, 
                     main = "Sample dendrogram",
@@
 tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION[,-(1:2)])))
-WGCNA::plotDendroAndColors(tree, 
+plotDendroAndColors(tree, 
                     colors = labels2colors(COVARIATES.tmp), 
@@
 
-DE = DE %>% plyr::ldply(.id = 'Comparison') %>%
-  tidyr::separate(ensembl_gene_id, into=c("ensemble_id","position")) %>%
-  dplyr::rename(ensembl_gene_id = ensemble_id) %>%
-  left_join(Ensemble2HGNC)
+DE = DE %>% plyr::ldply(.id = 'Comparison')
 

### 2016-05-01 code/Rmd/MSSM_BM_44_RNASeq.Rmd
Added markdown to process BM 44 region of MSBB RNASeq samples

---
title: "Covariates analysis of MSBB RNASeq data of region BM 44"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knitr)
library(githubr)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./MSSM_BM_44_RNASeq.Rmd",
                                 parentId ="syn5570248",
                                 entityName = 'MSBB BM 44 RNASeq')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

### Clear R console screen output
cat("\014")

## Load required libraries
library(CovariateAnalysis)
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(reshape2)
library(limma)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(xlsx)

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

synapseLogin()

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570248';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of MSBB BM 44 RNASeq samples';

thisFileName <- 'MSSM_BM_44_RNASeq.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
Data were obtained from AMP-AD portal of MSSM, and following brain regions were analysed for differential expression between case and control:
  | Region | Name                        | 
  | :----- | --------------------------- | 
  | BM_44  | inferior frontal Gyrus (IFG)|

* BM 44: pars opercularis (of the inferior frontal gyrus), and it refers to a subdivision of the frontal region of cerebral cortex.

Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_ID <- 'syn5898488'
ALL_USED_IDs <- COUNT_ID
COUNT_OBJ <- synGet(COUNT_ID)
COUNT <- fread(COUNT_OBJ@filePath, data.table = F)
rownames(COUNT) = COUNT$Geneid

# Get technical metadata
METADATA_ID <- 'syn5898489'
METADATA_OBJ <- synGet(METADATA_ID)
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA <- read.table(METADATA_OBJ@filePath,sep='\t',header=T, row.names=1)

# Get clinical metadata
METADATA_TECHNICAL_ID <- 'syn5475828'
METADATA_TECHNICAL_OBJ <- synGet(METADATA_TECHNICAL_ID)
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECHNICAL_ID
METADATA_TECHNICAL <- read.xlsx(METADATA_TECHNICAL_OBJ@filePath, sheetIndex = 1)

# Manually fix 4 sample IDs
METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1132), "BM36.Fnl"] = 397
METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1117), "BM22.Fnl"] = 144
METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1320), "BM22.Fnl"] = 159
METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1475), "BM22.Fnl"] = 186

METADATA = left_join(METADATA %>%
  dplyr::mutate(ID = gsub('hB_RNA_','',NewBarcode),
                ID = as.numeric(ID)),
  METADATA_TECHNICAL %>%
  dplyr::select(BB, BM.44.Barcode, BM36.Fnl, BM22.Fnl, BM10.Fnl, PMI, RACE, 
                AOD, CDR, SEX, NP.1, PlaqueMean, bbscore) %>%
  dplyr::rename(BM_44 = BM.44.Barcode,
                BM_10 = BM10.Fnl,
                BM_22 = BM22.Fnl,
                BM_36 = BM36.Fnl) %>%
  tidyr::gather(BrainRegion, ID, -BB, -PMI, -RACE, 
                -AOD, -CDR, -SEX, -NP.1, -PlaqueMean, -bbscore) %>%
  filter(!is.na(ID)) %>%
  dplyr::mutate(ID = as.numeric(ID))) %>%
  dplyr::filter(BrainRegion == 'BM_44')
rownames(METADATA) = METADATA$NewBarcode

# Filter data
ind <- intersect(rownames(METADATA), colnames(COUNT))
ind <- setdiff(ind, 'hB_RNA_10622')
METADATA <- METADATA[ind,]
COUNT <- COUNT[, ind]
```

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering, cache=TRUE}
METADATA$AOD = gsub('\\+','',as.character(METADATA$AOD))
METADATA$RIN2 = METADATA$RIN^2

METADATA$Dx = 'Control'
METADATA$Dx[METADATA$CDR > 1] = 'AD'

FactorCovariates <- c('RACE','SEX','NP.1','bbscore','Dx','Batch')
ContCovariates <- c('PMI','AOD','PlaqueMean','RIN', 'RIN2')

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- rownames(METADATA)

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)

rownames(COVARIATES) <- rownames(METADATA)

# Sub sample the analysis
COVARIATES = rownameToFirstColumn(COVARIATES, 'ID')
COVARIATES = COVARIATES %>% 
  group_by(BrainRegion.CDR, SEX, RACE, Batch) %>% 
  top_n(2) %>%
  data.frame
COUNT = COUNT[,COVARIATES$ID]
rownames(COVARIATES) = COVARIATES$ID
COVARIATES$ID = NULL
```
Covariate correlation
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.3, row.width=0.15)
```
### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

Remove genes that have less than 1 cpm counts in at least 50% of samples.
```{r cpmnormalisation, cache=TRUE}
PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT,MIN_GENE_CPM=1, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
```
### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T)
```
### Outlier analysis
Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T)

# Plot first 2 PCs
plotdata <- data.frame(ID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'ID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=BrainRegion.CDR, shape=SEX, size=PMI))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= ID), size=4, hjust=1)
p

COVARIATES.tmp = data.matrix(COVARIATES)
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);
tmp = gsub('\\..*','',cols)
WGCNA::plotDendroAndColors(tree, 
                           colors = tmp, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
dev.off()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Initial design matrix with covariates under consideration
DM = getDesignMatrix(COVARIATES, Intercept = F)

# Find correlation between PC's of gene expression with covariates
DESIGN = getDesignMat2Fact(DM$design,FactorCovariates)
rownames(DESIGN) = rownames(COVARIATES)
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, DESIGN,'NULL design(voom-normalized)', isKeyPlot=TRUE)

# Find significant covariates
adjustCovars = designMatVarsToCovars(getCovariatesMap(DM), preAdjustedSigCovars$significantCovars)
```
Significant covariates to adjust at FDR 0.1 are `r adjustCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix. Here BB(Donor Information) is chosen as random effect
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = 'Dx'

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars

significantCovars = designMatVarsToCovars(getCovariatesMap(DM), 
                                          residualSigCovars$significantCovars)

covariatesEffects = designMat2CovEffects(getCovariatesMap(DM),
                                         residualSigCovars$Effects.significantCovars)

covariatesEffects = covariatesEffects[setdiff(significantCovars, postAdjustCovars)]

postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate correlation of random effects
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)

  # Fit a linear model
  FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), postAdjustCovars)

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                dplyr::select(DESIGN,-one_of(colnames(DM1$design))),
                                                'all adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  significantCovars = designMatVarsToCovars(getCovariatesMap(DM), residualSigCovars$significantCovars)
  covariatesEffects = designMat2CovEffects(getCovariatesMap(DM),residualSigCovars$Effects.significantCovars)
  covariatesEffects = covariatesEffects[setdiff(significantCovars, postAdjustCovars)]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`

Degenerate variables are:
```{r degenVar}
# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F], Intercept = F)
tmp = linColumnFinder(DM1$design)
print(tmp$relations)
DM1$design = DM1$design[,tmp$indepCols]

# Recalculate weights with new design
VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, 
                                     design=DM1$design, plot=F)

# Linear model with new weights and design
VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
rownames(DESIGN) <- rownames(COVARIATES)
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION,
                                              DESIGN,
                                              'all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "CDR" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = c( "DxControl","DxAD")
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T)

# Plot first 2 PCs
plotdata <- data.frame(ID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'ID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(Batch), shape=as.factor(CDR), size=as.numeric(PMI)))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p

COVARIATES.tmp = data.matrix(COVARIATES)
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);
tmp = gsub('\\..*','',cols)
WGCNA::plotDendroAndColors(tree, 
                           colors = tmp, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```
```{r get.gene.symbols, cache=TRUE}
# Define biomart object
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = rownames(RESIDUAL.GENE_EXPRESSION),
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id"),
                                      Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = as.data.frame(summary(COVARIATES$CDR))
colnames(tmp) = 'No. of samples'
kable(tmp)
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=18}
# Fit contrast
contrast = colnames(VOOM.ADJUSTED.FIT$coefficients) %>%
  grep('Dx',.,value=T) %>%
  combn(2) %>%
  apply(2,paste,collapse='-') %>%
  makeContrasts(contrasts=.,
                levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT = eBayes(FIT)

# Get differnetial expression
DE = list()
p = list()
for (i in 1:dim(contrast)[2]){
  DE[[i]] = topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id") %>%
    left_join(Ensemble2HGNC)
  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
  p[[i]] <- p[[i]] + ggtitle(gsub('TISSUE.Dx','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red')
}
# multiplot(plotlist=p, cols = 3)
names(DE) = gsub('CDR','',colnames(contrast))

DE = DE %>% 
  rbindlist(idcol = 'Comparison', use.names=T, fill=T)

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(Comparison, hgnc_symbol) %>%
  group_by(Comparison) %>%
  dplyr::summarise(nDiffExp = length(unique(hgnc_symbol)), 
            DiffExpGenes = paste(unique(hgnc_symbol), collapse = ','))

kable(tmp)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "MSBB BM 44 RNASeq", parentId = parentId)
CODE <- synStore(CODE)

# Expression
expr = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'EnsembleIDs')
write.table(expr, file='expression.tsv', sep = '\t', row.names=F, quote=F)
obj = File('expression.tsv', 
           name = 'Voom Adjusted Weighted Residual Expression (Dx added)', 
           parentId = CODE$properties$id)
expr_obj = synStore(obj,
               used = ALL_USED_IDs, 
               activityName = activityName, 
               executed = thisFile, 
               activityDescription = activityDescription)

# Covariates
covar = rownameToFirstColumn(COVARIATES, 'SampleId')
write.table(covar, file='covariates.tsv', sep = '\t', row.names=F, quote=F)
obj = File('covariates.tsv', 
           name = 'Covariates', 
           parentId = CODE$properties$id)
covar_obj = synStore(obj,
               used = ALL_USED_IDs, 
               activityName = activityName, 
               executed = thisFile, 
               activityDescription = activityDescription)

# Store differential expression results
write.table(DE, file = 'diffExpression.tsv', sep = '\t', row.names=F, quote=F)
dexp_obj = File('diffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
dexp_obj = synStore(dexp_obj, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
```
|  *Results*            |  *SynapseID*                    |
|  -------              |   ---------                     |
|  Covariates           |  `r covar_obj$properties$id`    |
|  Residual Expression  |  `r expr_obj$properties$id`     |
|  Differential Expression  |  `r dexp_obj$properties$id` |

### 2016-05-01 code/Rmd/MSSM_BM_44_RNASeq.Rmd
Minor bug fix

@@
 rownames(COVARIATES) <- rownames(METADATA)
-
-# Sub sample the analysis
-COVARIATES = rownameToFirstColumn(COVARIATES, 'ID')
-COVARIATES = COVARIATES %>% 
-  group_by(BrainRegion.CDR, SEX, RACE, Batch) %>% 
-  top_n(2) %>%
-  data.frame
-COUNT = COUNT[,COVARIATES$ID]
-rownames(COVARIATES) = COVARIATES$ID
-COVARIATES$ID = NULL
 ```
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=BrainRegion.CDR, shape=SEX, size=PMI))
+p <- p + geom_point(aes(color=Dx, shape=SEX, size=PMI))
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=factor(Batch), shape=as.factor(CDR), size=as.numeric(PMI)))
+p <- p + geom_point(aes(color=factor(Batch), shape=as.factor(Dx), size=as.numeric(PMI)))
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
@@
 writeLines('Number of samples in each category are')
-tmp = as.data.frame(summary(COVARIATES$CDR))
+tmp = as.data.frame(summary(COVARIATES$Dx))
 colnames(tmp) = 'No. of samples'

### 2016-05-01 code/Rmd/MSSM_BM_36_RNASeq.Rmd
Added markdown to process BM 36

--- 2016-05-01 code/Rmd/MSSM_BM_44_RNASeq.Rmd
@@
 ---
-title: "Covariates analysis of MSBB RNASeq data of region BM 44"
+title: "Covariates analysis of MSBB RNASeq data of region BM 36"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_BM_44_RNASeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_BM_36_RNASeq.Rmd",
                                  parentId ="syn5570248",
-                                 entityName = 'MSBB BM 44 RNASeq')
+                                 entityName = 'MSBB BM 36 RNASeq')
 ```
@@
 activityName = 'Covariate analysis';
-activityDescription = 'Covariate analysis of MSBB BM 44 RNASeq samples';
+activityDescription = 'Covariate analysis of MSBB BM 36 RNASeq samples';
 
-thisFileName <- 'MSSM_BM_44_RNASeq.Rmd'
+thisFileName <- 'MSSM_BM_36_RNASeq.Rmd'
 
@@
   | :----- | --------------------------- | 
-  | BM_44  | inferior frontal Gyrus (IFG)|
+  | BM_36  | Parahipocampal Gyrus (PHG)  |
 
-* BM 44: pars opercularis (of the inferior frontal gyrus), and it refers to a subdivision of the frontal region of cerebral cortex.
+  * BM 36: Ectorhinal area 36 is a subdivision of the cytoarchitecturally defined temporal region of cerebral cortex.
 
@@
                        dplyr::mutate(ID = as.numeric(ID))) %>%
-  dplyr::filter(BrainRegion == 'BM_44')
+  dplyr::filter(BrainRegion == 'BM_36')
 rownames(METADATA) = METADATA$NewBarcode
@@
 # Code
-CODE <- Folder(name = "MSBB BM 44 RNASeq", parentId = parentId)
+CODE <- Folder(name = "MSBB BM 36 RNASeq", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-05-01 code/Rmd/MSSM_BM_36_RNASeq.Rmd
Minor fix

@@
 
-FactorCovariates <- c('RACE','SEX','NP.1','bbscore','Dx','Batch')
+FactorCovariates <- c('RACE','SEX','Dx','Batch')
 ContCovariates <- c('PMI','AOD','PlaqueMean','RIN', 'RIN2')
@@
   # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
+  tmp = sapply(dplyr::select(DESIGN,-one_of(colnames(DM1$design))), as.numeric)
   residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                                dplyr::select(DESIGN,-one_of(colnames(DM1$design))),
-                                                'all adjusted design(voom-normalized)',
+                                                tmp, 'all adjusted design(voom-normalized)',
                                                 isKeyPlot=TRUE)

### 2016-05-06 code/Rmd/HBTRC_Array.Rmd

--- 2016-01-07 code/Rmd/MSSM_Array.Rmd
@@
 ---
-title: "Covariate analysis of MSBB microarray data"
+title: "Covariate analysis of HBTRC microarray data"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_Array.Rmd",
-                                 parentId ="syn5570248",
-                                 entityName = 'MSBB Microarray')
+knit2synapse::knitToFolderEntity(file = "./HBTRC_Array.Rmd",
+                                 parentId ="syn5570324",
+                                 entityName = "HBTRC Microarray")
 ```
@@
 # Synapse parameters
-parentId = 'syn5570248';
+parentId = 'syn5570324';
 activityName = 'Covariate analysis';
-activityDescription = 'Covariate analysis of MSBB microarrays';
+activityDescription = 'Covariate analysis of HBTRC microarrays';
 
-thisFileName <- 'MSSM_Array.Rmd'
+thisFileName <- 'HBTRC_Array.Rmd'
 
@@
 ```
-Following brain regions were analysed for differential expression between case and control
+Following brain regions were analysed for differential expression
 
@@
 | ----------------------------------- | ---------- |
-| middle temporal                     | syn3191101 |
-| superior temporal                   | syn3191103 |
-| inferior temporal                   | syn3191099 |
-| precuneus superior parietal lobule  | syn3191119 |
-| caudal anterior cingulate           | syn3191107 |
-| frontal pole                        | syn3191095 |
-| para hippocampal gyrus              | syn3191109 |
+| Cerebellum                          | syn3275756 |
+| Prefrontal cortex                   | syn3275756 |
+| Visual cortex                       | syn3275756 |
+### Data download
+#### Obtain expression matrix and metadata from synapse.
+```{r download.data, cache=TRUE}
+# Download expression data
+FOLDER_ID <- 'syn3275756';
+ALL_USED_IDs <- FOLDER_ID
+FOLDER_OBJ <- synGet(FOLDER_ID)
+
+# Unzip all files
+FNames = unzip(FOLDER_OBJ@filePath)
+
+# Get expression data
+brain_regions = c('cerebellum', 'visualcortex', 'prefrontalcortex')
+EXPR = lapply(brain_regions, function(region, fileNames){
+  ind = intersect(grep(region, fileNames),grep('gene_expression.txt', fileNames))
+  expr = read.table(fileNames[ind], header=T, row.names = 1)
+
+#   ind = intersect(grep(region, fileNames),grep('gene_expression_pvals.txt', fileNames))
+#   pvals = read.table(fileNames[ind], header=T, row.names = 1)
+#   pvals = pvals[rownames(expr), colnames(expr)]
+#   
+#   expr[pvals > 0.05] = NA
+#   ind = (rowSums(is.na(expr))/dim(expr)[2]) <= 0.5
+#   expr = expr[ind,]
+
+  return(expr)
+}, FNames)
+names(EXPR) = brain_regions
+
+# Get metadata
+METADATA1 = fread(FNames[grep('phenotype.txt',FNames)], header=T, data.table=F)
+METADATA2 = read.table(pipe("cut -f1,1,2 ./hbtrc/phenotype/phenotype_description.txt"), header=T)
+METADATA = full_join(METADATA2, METADATA1)
+
+# Arrange metadata and expression
+allDATA = lapply(EXPR, function(expr, metadata){
+  rownames(metadata) = metadata$trait_name
+  metadata = t(metadata[,-(1:2)])
+
+  colnames(expr) = gsub('X','',colnames(expr))
+  cnames.expr = sapply(colnames(expr), function(x){str_split(x,'\\.')[[1]][1]})
+
+  expr = expr[,names(cnames.expr)[cnames.expr %in% rownames(metadata)]]
+  cnames.expr = cnames.expr[cnames.expr %in% rownames(metadata)]
+
+  newMetadata = metadata[cnames.expr,]
+  rownames(newMetadata) = colnames(expr)
+
+  return(list(expr = expr, metadata = newMetadata))
+}, METADATA)
+
+# Seperate expression
+EXPR = lapply(brain_regions, function(region, allData){
+  expr = allData[[region]]$expr
+  colnames(expr) = paste(colnames(expr),region,sep='.')
+  expr = rownameToFirstColumn(expr, 'probeIDs')
+  return(expr)
+}, allDATA) %>%
+  plyr::join_all(type = 'full')
+
+# Seperate metadata
+ALL.METADATA = lapply(brain_regions, function(region, allData){
+  metadata = allData[[region]]$metadata
+  metadata = cbind(data.frame(metadata), data.frame(donorID = rownames(metadata)))
+
+  rownames(metadata) = paste(rownames(metadata), region, sep = '.')
+  return(metadata)
+}, allDATA) %>%
+  rbindlist
+
+# Store results in synapse
+write.table(EXPR, file = 'rawExpression.tsv', sep = '\t', quote=F, row.names = F)
+obj = File('rawExpression.tsv', name = 'Raw Expression', parentId = CODE$properties$id)
+obj = synStore(obj, used = as.character(ALL_USED_IDs), executed = ThisFile, 
+               activityDescription = activityDescription, activityName = activityName)
+
+# Store results in synapse
+write.table(ALL.METADATA, file = 'rawMetaData.tsv', sep = '\t', quote=F, row.names = F)
+obj = File('rawMetaData.tsv', name = 'Raw Metadata', parentId = CODE$properties$id)
+obj = synStore(obj, used = as.character(ALL_USED_IDs), executed = ThisFile, 
+               activityDescription = activityDescription, activityName = activityName)
+```
 

### 2016-05-06 code/Rmd/MSSM_BM_10_RNASeq.Rmd

--- 2016-05-01 code/Rmd/MSSM_BM_36_RNASeq.Rmd
@@
 ---
-title: "Covariates analysis of MSBB RNASeq data of region BM 36"
+title: "Covariates analysis of MSBB RNASeq data of region BM 10"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_BM_36_RNASeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_BM_10_RNASeq.Rmd",
                                  parentId ="syn5570248",
-                                 entityName = 'MSBB BM 36 RNASeq')
+                                 entityName = 'MSBB BM 10 RNASeq')
 ```
@@
 activityName = 'Covariate analysis';
-activityDescription = 'Covariate analysis of MSBB BM 36 RNASeq samples';
+activityDescription = 'Covariate analysis of MSBB BM 10 RNASeq samples';
 
-thisFileName <- 'MSSM_BM_36_RNASeq.Rmd'
+thisFileName <- 'MSSM_BM_10_RNASeq.Rmd'
 
@@
   | :----- | --------------------------- | 
-  | BM_36  | Parahipocampal Gyrus (PHG)  |
+  | BM_10  | Frontal Pole (FP)           |
 
-  * BM 36: Ectorhinal area 36 is a subdivision of the cytoarchitecturally defined temporal region of cerebral cortex.
+* BM 10: frontopolar prefrontal cortex, rostrolateral prefrontal cortex, or anterior prefrontal cortex,
 
@@
                        dplyr::mutate(ID = as.numeric(ID))) %>%
-  dplyr::filter(BrainRegion == 'BM_36')
+  dplyr::filter(BrainRegion == 'BM_10')
 rownames(METADATA) = METADATA$NewBarcode
@@
 # Code
-CODE <- Folder(name = "MSBB BM 36 RNASeq", parentId = parentId)
+CODE <- Folder(name = "MSBB BM 10 RNASeq", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-05-06 code/Rmd/MSSM_BM_22_RNASeq.Rmd

--- 2016-05-06 code/Rmd/MSSM_BM_10_RNASeq.Rmd
@@
 ---
-title: "Covariates analysis of MSBB RNASeq data of region BM 10"
+title: "Covariates analysis of MSBB RNASeq data of region BM 22"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_BM_10_RNASeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_BM_22_RNASeq.Rmd",
                                  parentId ="syn5570248",
-                                 entityName = 'MSBB BM 10 RNASeq')
+                                 entityName = 'MSBB BM 22 RNASeq')
 ```
@@
 activityName = 'Covariate analysis';
-activityDescription = 'Covariate analysis of MSBB BM 10 RNASeq samples';
+activityDescription = 'Covariate analysis of MSBB BM 22 RNASeq samples';
 
-thisFileName <- 'MSSM_BM_10_RNASeq.Rmd'
+thisFileName <- 'MSSM_BM_22_RNASeq.Rmd'
 
@@
   | :----- | --------------------------- | 
-  | BM_10  | Frontal Pole (FP)           |
+  | BM_22  | Superior Temporal Gyrus     |
 
-* BM 10: frontopolar prefrontal cortex, rostrolateral prefrontal cortex, or anterior prefrontal cortex,
+* BM 22: In humans it corresponds approximately to the lateral and caudal two thirds of the superior temporal gyrus,
 
@@
                        dplyr::mutate(ID = as.numeric(ID))) %>%
-  dplyr::filter(BrainRegion == 'BM_10')
+  dplyr::filter(BrainRegion == 'BM_22')
 rownames(METADATA) = METADATA$NewBarcode
@@
 # Code
-CODE <- Folder(name = "MSBB BM 10 RNASeq", parentId = parentId)
+CODE <- Folder(name = "MSBB BM 22 RNASeq", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-05-06 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd

@@
 METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
-METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
+METADATA$Diagnosis = gsub("Pathologic Aging", "Pathologic_Aging", METADATA$Diagnosis)
 METADATA <- METADATA %>%

### 2016-05-06 code/R/collateDiffExp.R
Added R code to collate all the differential expression results

@@ -1,0 +1,66 @@
+## It is assumed your working directory is where this file is
+
+# Clear R console screen output
+cat("\014")  
+
+# Load required libraries
+library(CovariateAnalysis)
+library(data.table)
+library(tidyr)
+library(plyr)
+library(dplyr)
+library(stringr)
+
+library(synapseClient)
+library(knitr)
+library(githubr)
+
+library(biomaRt)
+library(ComplexHeatmap)
+
+synapseLogin()
+
+# Get the latest commit of the code from git
+thisFileName <- 'collateDiffExp.R'
+
+# Github link
+thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
+thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/R/',thisFileName))
+
+# Get protein coding genes from biomaRt
+ensembl=useMart("ensembl")
+ensemblHSapiens = useDataset("hsapiens_gene_ensembl",mart=ensembl)
+humanProteinCodingGenes = getBM(attributes=c("ensembl_gene_id","external_gene_name","description"), 
+                                filters='biotype', 
+                                values=c('protein_coding'), 
+                                mart=ensemblHSapiens)
+
+# Get all differential expresison results from synapse
+diffExp.ids = c('ROSMAP DLPFC' = 'syn5608845',
+                'MSBB Frontal Pole' = 'syn6037382',
+                'MSBB Superior Temporal Gyrus'	= 'syn6037356',
+                'MSBB Para Hippocampal Gyrus' = 'syn6037276',
+                'MSBB Inferior Frontal Gyrus' = 'syn6037095',
+                'Mayo Cerebellum' = 'syn5609850',
+                'Mayo Temporal Cortex' = 'syn5609813')
+diffExp = lapply(diffExp.ids, downloadFile) %>%
+  rbindlist(idcol = 'DataSetName', fill=T, use.names=T) %>%
+  dplyr::select(-genes, -position) %>%
+  filter(ensembl_gene_id %in% humanProteinCodingGenes$ensembl_gene_id)
+
+# Store differential expression results in synapse
+write.table(diffExp, file = 'differentialExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('differentialExpression.tsv', 
+                name = 'Differential expression results of all the RNASeq data (protein coding only)', 
[…]

### 2016-05-07 code/R/collateDiffExp.R
Intermediate commit

@@ -24,7 +24,7 @@ synapseLogin()
 thisFileName <- 'collateDiffExp.R'
 
 # Github link
-thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
+thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='metaAnal')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/R/',thisFileName))
 
 # Get protein coding genes from biomaRt
@@ -59,7 +59,7 @@ DEXP_OBJ = synStore(DEXP_OBJ,
                    executed = thisFile)
 
 logFC = diffExp %>%
-  filter(adj.P.Val <= 0.05, abs(logFC) <= 1) %>%
+  filter(adj.P.Val <= 0.05, abs(logFC) >= 2) %>%
   dplyr::select(DataSetName, Comparison, ensembl_gene_id, logFC) %>%
   unite(DataSet.Comparison, DataSetName, Comparison, sep = ' ') %>%
   spread(DataSet.Comparison, logFC)

### 2016-05-14 code/Rmd/MSSM_RNASeq.Rmd
Minor changes

@@
 library(synapseClient)
+library(githubr)
 
@@
 ## Load required libraries
-library(synapseClient)
-library(rGithubClient) ## Needs the dev branch
+library(CovariateAnalysis)
+library(data.table)
+library(tidyr)
+library(plyr)
+library(dplyr)
+library(stringr)
+
 library(ggplot2)
 library(reshape2)
-library(data.table)
-library(plyr)
-library(dplyr)
-library(tidyr)
-library(stringr)
-library(knitr)
 library(limma)
@@
 library(biomaRt)
+library(xlsx)
+
+# library(doParallel)
+# library(foreach)
+# 
+# cl = makeCluster(12)
+# registerDoParallel(cl)
 
@@
 
-synapseLogin()
-
-# source utility files from ../R/lib folder
-file.sources = list.files('../R/lib',pattern="*.R", full.names=T)
-tmp = sapply(file.sources,source,.GlobalEnv)
+# Source required R files
+source('../R/lib/parallelDuplicateCorrelation.R')
 
@@
   | BM_36  | Parahipocampal Gyrus (PHG)|
+  | BM_44  | inferior frontal Gyrus (IFG)|
 
@@
 * BM 22: In humans it corresponds approximately to the lateral and caudal two thirds of the superior temporal gyrus, 
-* BM 36: Ectorhinal area 36 is a subdivision of the cytoarchitecturally defined temporal region of cerebral cortex.
+* BM 36: Ectorhinal area 36 is a subdivision of the cytoarchitecturally defined temporal region of cerebral cortex,
+* BM 44: pars opercularis (of the inferior frontal gyrus), and it refers to a subdivision of the frontal region of cerebral cortex.
 
@@
 # Download expression data
-COUNT_ID <- 'syn3205353'
+COUNT_ID <- 'syn5898488'
 ALL_USED_IDs <- COUNT_ID
 COUNT_OBJ <- synGet(COUNT_ID)
-COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F, row.names = 1)
+COUNT <- fread(COUNT_OBJ@filePath, data.table = F)
+rownames(COUNT) = COUNT$Geneid
 
-# Get metadata
-METADATA_ID <- 'syn3205337'
+# Get technical metadata
+METADATA_ID <- 'syn5898489'
 METADATA_OBJ <- synGet(METADATA_ID)
@@
 METADATA <- read.table(METADATA_OBJ@filePath,sep='\t',header=T, row.names=1)
-METADATA <- cbind(METADATA, data.frame(Dx = ifelse(METADATA$CDR<1,'Control','AD')))
+
+# Get clinical metadata
+METADATA_TECHNICAL_ID <- 'syn5475828'
+METADATA_TECHNICAL_OBJ <- synGet(METADATA_TECHNICAL_ID)
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECHNICAL_ID
+METADATA_TECHNICAL <- read.xlsx(METADATA_TECHNICAL_OBJ@filePath, sheetIndex = 1)
+
+# Manually fix 4 sample IDs
+METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1132), "BM36.Fnl"] = 397
+METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1117), "BM22.Fnl"] = 144
+METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1320), "BM22.Fnl"] = 159
+METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1475), "BM22.Fnl"] = 186
+
+METADATA = left_join(METADATA %>%
+                       dplyr::mutate(ID = gsub('hB_RNA_','',NewBarcode),
+                                     ID = as.numeric(ID)),
+                     METADATA_TECHNICAL %>%
+                       dplyr::select(BB, BM.44.Barcode, BM36.Fnl, BM22.Fnl, BM10.Fnl, PMI, RACE, 
+                                     AOD, CDR, SEX, NP.1, PlaqueMean, bbscore) %>%
+                       dplyr::rename(BM_44 = BM.44.Barcode,
+                                     BM_10 = BM10.Fnl,
+                                     BM_22 = BM22.Fnl,
+                                     BM_36 = BM36.Fnl) %>%
+                       tidyr::gather(BrainRegion, ID, -BB, -PMI, -RACE, 
+                                     -AOD, -CDR, -SEX, -NP.1, -PlaqueMean, -bbscore) %>%
+                       filter(!is.na(ID)) %>%
+                       dplyr::mutate(ID = as.numeric(ID)))
+rownames(METADATA) = METADATA$NewBarcode
+
+# Filter data
+ind <- intersect(rownames(METADATA), colnames(COUNT))
+ind <- setdiff(ind, 'hB_RNA_10622')
+METADATA <- METADATA[ind,]
+COUNT <- COUNT[, ind]
+```
+### Covariate clustering
+Determine relationship between covariates. 
+```{r covariates.clustering, cache=TRUE}
+# Max out age at 90
+METADATA$AOD = gsub('\\+','',as.character(METADATA$AOD))
+
+# Create RIN^2 column
+METADATA$RIN2 = METADATA$RIN^2
+
+# Create diagnosis column (classified into 3 groups ND, MD, SD)
+METADATA$Dx = 'ND'
+METADATA$Dx[METADATA$CDR == 1 | METADATA$CDR == 2] = 'MD'
+METADATA$Dx[METADATA$CDR >2 ] = 'SD'
+
+# Create brain region x diagnosis column
+METADATA$BrainRegion.Dx = paste(METADATA$BrainRegion, METADATA$Dx, sep = '.')
+
+# Select covariates of interest
+FactorCovariates <- c('RACE', 'SEX', 'BrainRegion.Dx', 'Batch', 'BB')
+ContCovariates <- c('PMI', 'AOD', 'RIN', 'RIN2')
+
+# Find inter relation between factor covariates
+COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
+rownames(COVARIATES) <- METADATA$NewBarcode
+
+# Convert factor covariates to factors
+COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
+COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
+COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
+
+# Arrange expression and covariates
+COUNT = COUNT[, rownames(COVARIATES)]
+```
+Covariate correlation
+```{r covariates.correlation, fig.width=10, fig.height=10}
+COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
+ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.3, row.width=0.15)
 ```
@@
 ```{r cpmnormalisation, cache=TRUE}
-ind <- intersect(rownames(METADATA), colnames(COUNT))
-METADATA <- METADATA[ind,]
-COUNT <- COUNT[, ind]
 PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT,MIN_GENE_CPM=1, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
-```
-### Covariate clustering
-Determine relationship between covariates. 
-```{r covariates.clustering, cache=TRUE}
-FactorCovariates <- c('TISSUE','Donor','RACE','CDR','CERJ','SEX','NP.1','bbscore','Dx')
-ContCovariates <- c('PMI','AOD','PlaqueMean')
-
-# Find inter relation between factor covariates
-COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
-rownames(COVARIATES) <- rownames(METADATA)
-
-# Convert factor covariates to factors
-COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
-COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
-COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
-
-COVARIATES <- dplyr::mutate(COVARIATES, TISSUE.Dx = paste(TISSUE,Dx,sep='.'))
-COVARIATES$TISSUE.Dx <- factor(COVARIATES$TISSUE.Dx)
-rownames(COVARIATES) <- rownames(METADATA)
-```
-Covariate correlation
-```{r covariates.correlation, fig.width=10, fig.height=10}
-COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 1)
-ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.3, row.width=0.15)
 ```
@@
 VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T)
-
-# Initial design matrix with covariates under consideration
-DM = getDesignMatrix(COVARIATES, Intercept = F)
 ```
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=factor(TISSUE.Dx), shape=as.factor(CDR), size=as.numeric(PMI)))
+p <- p + geom_point(aes(color=BrainRegion.Dx, shape=SEX, size=PMI))
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
-# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
+# p <- p + geom_text(aes(label= ID), size=4, hjust=1)
 p
@@
 ```
-
 ### Significant Covariates
@@
 # Find correlation between PC's of gene expression with covariates
-DESIGN = getDesignMat2Fact(DM$design,FactorCovariates)
-rownames(DESIGN) = rownames(METADATA)
-preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, DESIGN,'NULL design(voom-normalized)', isKeyPlot=TRUE)
-
-# Find significant covariates
-adjustCovars = designMatVarsToCovars(getCovariatesMap(DM),preAdjustedSigCovars$significantCovars)
+COVARIATES.tmp = COVARIATES
+COVARIATES.tmp[, FactorCovariates] = sapply(COVARIATES.tmp[, FactorCovariates], as.numeric)
+preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, COVARIATES.tmp,'NULL design(voom-normalized)', isKeyPlot=TRUE, MIN_PVE_PCT_PC = 1)
 ```
-Significant covariates to adjust at FDR 0.1 are `r adjustCovars`
+Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
 ```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
@@
 ```
-### Normalisation (custom/manual design)
-Re-perform normalisation with custom design matrix (manually with "TISSUE.Dx","SEX","RACE","PMI" as fixed effects and Donor as random effects)
-```{r reperform.normalisation, cache=TRUE}
+
+### Normalisation (iterative design)
+Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix. Here BB (Donor Information) is chosen as random effect
+```{r iterativeNorm}
+# Primary variable of interest
+postAdjustCovars = 'BrainRegion.Dx'
+
+# Assign residual covariates
+residualSigCovars = preAdjustedSigCovars
+
+covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, 'BB'))]
+
+postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
+
+loopCount = 0 
+while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
+  writeLines(paste('Using following covariates in the model:',
+                   paste(postAdjustCovars, collapse=', '),
+                   'as fixed effects'))
+
   # Post adjusted design matrix
-DM1 = getDesignMatrix(COVARIATES[,c("TISSUE.Dx","SEX","RACE","PMI"),drop=F],Intercept = F)
+  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
   DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
@@
   VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)
-correlation <- duplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION$E, design=DM1$design, block=COVARIATES$Donor)
+  correlation <- parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block=COVARIATES$BB)
 
@@
   VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F,
-                                     block=COVARIATES$Donor, correlation = correlation$cor)
+                                       block=COVARIATES$BB, correlation = correlation$cor)
 
   # Re-calculate linear fit using new weights and new design
-VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION$E, design=DM1$design, weights = VOOM.ADJUSTED.GENE_EXPRESSION$weights,
-                          block=COVARIATES$Donor, correlation = correlation$cor)
+  VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
 
@@
 
+  # Residual covariates to choose from
+  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, 'BB'))
+
   # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-rownames(DESIGN) <- rownames(METADATA)
   residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                              dplyr::select(DESIGN,-starts_with('Donor')),
-                                              'all adjusted design(voom-normalized)',
+                                                COVARIATES.tmp[, residCovars, drop=F], 'all adjusted design(voom-normalized)',
                                                 isKeyPlot=TRUE)
+
+  # Add postadjusted covariates (if any)
+  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, 'BB'))]
+
+  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
+  loopCount = loopCount + 1
+}
+modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
+                  'as fixed effects')
+
+tmp <- paste('Using following covariates in the final model:', modelStr)
 ```
-
+`r tmp`
+### Sanity check
 ```{r residualSigCovars.manual, fig.width=20, fig.height=12}
@@
 ### Residual calculation
-Calculate weighted residuals and add back "TISSUE.Dx" to the residuals
+Calculate weighted residuals and add back "BrainRegion.Dx" to the residuals
 ```{r varsToAddBack}
 # Add variable of interest back to the residuals
-varsToAddIn = c("TISSUE.DxBM_10.AD", "TISSUE.DxBM_10.Control", "TISSUE.DxBM_22.AD",
-                "TISSUE.DxBM_22.Control", "TISSUE.DxBM_36.AD","TISSUE.DxBM_36.Control")
+varsToAddIn = grep("BrainRegion.Dx", colnames(DM1$design), value = T)
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=factor(TISSUE.Dx), shape=as.factor(CDR), size=as.numeric(PMI)))
-p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
+p <- p + geom_point(aes(color=factor(BrainRegion.Dx),size=as.numeric(PMI)))
+p <- p + theme_bw() + theme(legend.position="top") 
 # p <- p + geom_text(aes(label= ID), size=4, hjust=0)
@@
 ```
-```{r get.gene.symbols, cache=TRUE}
+```{r get.gene.symbols, cache=TRUE, eval=FALSE}
 # Define biomart object
-mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl", host = )
 
@@
 writeLines('Number of samples in each category are')
-tmp = as.data.frame(summary(COVARIATES$TISSUE.Dx))
+tmp = as.data.frame(summary(COVARIATES$BrainRegion.Dx))
 colnames(tmp) = 'No. of samples'
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("TISSUE.DxBM_10.Control-TISSUE.DxBM_10.AD",
-                                     "TISSUE.DxBM_22.Control-TISSUE.DxBM_22.AD",
-                                     "TISSUE.DxBM_36.Control-TISSUE.DxBM_36.AD"), 
+contrast = makeContrasts(contrasts=c("BrainRegion.DxBM_10.ND-BrainRegion.DxBM_10.MD", 
+                                     "BrainRegion.DxBM_10.ND-BrainRegion.DxBM_10.SD", 
+                                     "BrainRegion.DxBM_10.MD-BrainRegion.DxBM_10.SD",
+                                     "BrainRegion.DxBM_22.ND-BrainRegion.DxBM_22.MD", 
+                                     "BrainRegion.DxBM_22.ND-BrainRegion.DxBM_22.SD", 
+                                     "BrainRegion.DxBM_22.MD-BrainRegion.DxBM_22.SD",
+                                     "BrainRegion.DxBM_36.ND-BrainRegion.DxBM_36.MD", 
+                                     "BrainRegion.DxBM_36.ND-BrainRegion.DxBM_36.SD", 
+                                     "BrainRegion.DxBM_36.MD-BrainRegion.DxBM_36.SD",
+                                     "BrainRegion.DxBM_44.ND-BrainRegion.DxBM_44.MD", 
+                                     "BrainRegion.DxBM_44.ND-BrainRegion.DxBM_44.SD", 
+                                     "BrainRegion.DxBM_44.MD-BrainRegion.DxBM_44.SD"), 
                          levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
@@
 p = list()
-for (i in 1:3){
+for (i in 1:dim(contrast)[2]){
   DE[[i]] = topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
-  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id") %>%
-    left_join(Ensemble2HGNC)
+  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id")
   p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
-  p[[i]] <- p[[i]] + ggtitle(gsub('TISSUE.Dx','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
+  p[[i]] <- p[[i]] + ggtitle(gsub('BrainRegion.Dx','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
   p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red')
@@
 multiplot(plotlist=p, cols = 3)
-names(DE) = gsub('TISSUE.Dx','',colnames(contrast))
+names(DE) = gsub('BrainRegion.Dx','',colnames(contrast))
 
@@
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(Comparison, hgnc_symbol) %>%
+  dplyr::select(Comparison, ensembl_gene_id) %>%
   group_by(Comparison) %>%
-  dplyr::summarise(nDiffExp = length(unique(hgnc_symbol)), 
-            DiffExpGenes = paste(unique(hgnc_symbol), collapse = ','))
+  dplyr::summarise(nDiffExp = length(unique(ensembl_gene_id)), 
+            DiffExpGenes = paste(unique(ensembl_gene_id), collapse = ','))
 
@@
 obj = File('expression.tsv', 
-           name = 'Voom Adjusted Weighted Residual Expression (Tissue.Dx added)', 
+           name = 'Voom Adjusted Weighted Residual Expression (BrainRegion.Dx added)', 
            parentId = CODE$properties$id)

### 2016-05-14 code/R/lib/parallelDuplicateCorrelation.R
Added R code to calculate correlation coefficient between models parallelly

@@ -1,0 +1,97 @@
+parallelDuplicateCorrelation <- function(object, design = NULL, ndups = 2, spacing = 1, block = NULL,
+                                         trim = 0.15, weights = NULL) {  
+  y <- getEAWP(object)
+  M <- y$exprs
+  ngenes <- nrow(M)
+  narrays <- ncol(M)
+  if (is.null(design)) 
+    design <- y$design
+  if (is.null(design)) 
+    design <- matrix(1, ncol(y$exprs), 1)
+  else {
+    design <- as.matrix(design)
+    if (mode(design) != "numeric") 
+      stop("design must be a numeric matrix")
+  }
+  if (nrow(design) != narrays) 
+    stop("Number of rows of design matrix does not match number of arrays")
+  ne <- nonEstimable(design)
+  if (!is.null(ne)) 
+    cat("Coefficients not estimable:", paste(ne, collapse = " "), 
+        "\n")
+  nbeta <- ncol(design)
+  if (missing(ndups) && !is.null(y$printer$ndups)) 
+    ndups <- y$printer$ndups
+  if (missing(spacing) && !is.null(y$printer$spacing)) 
+    spacing <- y$printer$spacing
+  if (missing(weights) && !is.null(y$weights)) 
+    weights <- y$weights
+  if (!is.null(weights)) {
+    weights <- as.matrix(weights)
+    if (any(dim(weights) != dim(M))) 
+      weights <- array(weights, dim(M))
+    M[weights < 1e-15] <- NA
+    weights[weights < 1e-15] <- NA
+  }
+  if (is.null(block)) {
+    if (ndups < 2) {
+      warning("No duplicates: correlation between duplicates not estimable")
+      return(list(cor = NA, cor.genes = rep(NA, nrow(M))))
+    }
+    if (is.character(spacing)) {
+      if (spacing == "columns") 
+        spacing <- 1
+      if (spacing == "rows") 
+        spacing <- object$printer$nspot.c
+      if (spacing == "topbottom") 
+        spacing <- nrow(M)/2
+    }
+    Array <- rep(1:narrays, rep(ndups, narrays))
+  }
+  else {
+    ndups <- 1
+    nspacing <- 1
+    Array <- block
+  }
+  if (is.null(block)) {
+    M <- unwrapdups(M, ndups = ndups, spacing = spacing)
+    ngenes <- nrow(M)
[…]

### 2016-05-16 code/R/lib/calcCompleteCorAndPlot.R
Modified factor continuous association statistics code

@@ -8,29 +8,28 @@ calcCompleteCorAndPlot <- function(COMPARE_data, COVAR_data, correlationType, ti
   FactorCovariates <- colnames(COVAR_data)[sapply(COVAR_data,is.factor)]
   ContCovariates <- setdiff(colnames(COVAR_data),FactorCovariates)
   
-  # Convert factor covariates to numeric vector
-  COVAR_data[,FactorCovariates] <- apply(COVAR_data[,FactorCovariates],2,
-                                          function(x){x <- unclass(x)})
-    
   # Calculate correlation between compare_data and factor covariates
   if (length(FactorCovariates) > 0){
     comb <- expand.grid(colnames(COMPARE_data),FactorCovariates)
-    factCont_cor <- apply(comb,1,
-                          getFactorContAssociationStatistics,
-                          cbind(COMPARE_data,COVAR_data[rownames(COMPARE_data),FactorCovariates]),
-                          alpha=MAX_FDR)
-    factCont_cor_vals <- matrix(factCont_cor['Estimate',],
-                                nrow = length(colnames(COMPARE_data)),
-                                ncol = length(FactorCovariates))
-    factCont_cor_p <- matrix(factCont_cor['Pval',],
-                             nrow = length(colnames(COMPARE_data)),
-                             ncol = length(FactorCovariates))
+    factCont_cor <- plyr::ddply(comb,
+                                .(Var1, Var2),
+                                .fun = getFactorContAssociationStatistics,
+                                cbind(COMPARE_data,COVAR_data[rownames(COMPARE_data),FactorCovariates,drop=F]),
+                                alpha=MAX_FDR)
+    factCont_cor_vals <- factCont_cor %>%
+      dplyr::select(Var1, Var2, Estimate) %>%
+      tidyr::spread(Var2, Estimate)
+    rownames(factCont_cor_vals) <- factCont_cor_vals$Var1
+    factCont_cor_vals$Var1 <- NULL
+    factCont_cor_vals = factCont_cor_vals %>% data.matrix()
     
-    rownames(factCont_cor_vals) <- colnames(COMPARE_data)
-    colnames(factCont_cor_vals) <- FactorCovariates
+    factCont_cor_p <- factCont_cor %>%
+      dplyr::select(Var1, Var2, Pval) %>%
+      tidyr::spread(Var2, Pval)
+    rownames(factCont_cor_p) <- factCont_cor_p$Var1
+    factCont_cor_p$Var1 <- NULL
[…]

### 2016-05-16 code/R/lib/getFactorContAssociationStatistics.R

@@ -2,17 +2,16 @@
 # Inspired from http://stats.stackexchange.com/questions/108007/correlations-with-categorical-variables
 getFactorContAssociationStatistics <- function(factorContNames,COVARIATES, na.action='remove', 
                                                alpha = 0.05){
-  
   require(psych)
+  
+  factorContNames = factorContNames %>% unlist %>% as.character()
   
   if (na.action == "remove")
     COVARIATES = na.omit(COVARIATES[,factorContNames])
   
-  stats = ICC(COVARIATES[,factorContNames], alpha = alpha)
+  COVARIATES = sapply(COVARIATES, as.numeric)
+  stats = psych::ICC(COVARIATES, missing = T, alpha = alpha)
   
-  Pval = summary(aov(COVARIATES[,factorContNames[1]]~COVARIATES[,factorContNames[2]]))[[1]][["Pr(>F)"]][1]
-  
-  
-  return(c(Estimate = stats$results['Single_raters_absolute','ICC'],
-           Pval = Pval))
+  return(data.frame(Estimate = stats$results['Single_fixed_raters','ICC'],
+                    Pval = stats$results['Single_fixed_raters','p']))
 }

### 2016-05-16 code/R/lib/runPCAandPlotCorrelations.R

@@ -1,8 +1,9 @@
 # Function to run principal component analysis and plot correlations
 runPCAandPlotCorrelations <- function(genesBySamples, samplesByCovariates, dataName, isKeyPlot=FALSE, 
-                                      SCALE_DATA_FOR_PCA = TRUE, MIN_PVE_PCT_PC = 1.0, CORRELATION_TYPE = "pearson",
+                                      SCALE_DATA_FOR_PCA = TRUE, MIN_PVE_PCT_PC = 1.0, CORRELATION_TYPE = 'pearson',
                                       ALSO_PLOT_ALL_COVARS_VS_PCA = TRUE, MAX_NUM_LEVELS_PER_COVAR = 50) {
       
+  
   title = paste(ifelse(SCALE_DATA_FOR_PCA, "S", "Un-s"), "caled ", dataName, " ", " data in PCA; PVE >= ", MIN_PVE_PCT_PC, "%; ", CORRELATION_TYPE, " correlations ", sep="")
   writeLines(paste("\nRunning PCA and calculating correlations for:\n", title, sep=""))
  

### 2016-05-16 code/Rmd/Mayo_RNASeq.Rmd
Added processing markdown for Mayo RNASeq. This version has both brain regions modeled together with donor id as random effects

---
title: "Covariate analysis of UFL-Mayo-ISB RNASeq (both cerebellum and temporal cortex)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)
library(knitr)
library(githubr)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq.Rmd",
                                 parentId = "syn5570325",
                                 entityName = 'UFL_Mayo_ISB RNASeq')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis)
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(reshape2)
library(stringr)
library(limma)
library(Biobase)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(fpc)

library(doParallel)
library(foreach)

cl = makeCluster(14)
registerDoParallel(cl)

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

# source utility files from ../R/lib folder
file.sources = list.files('../R/lib',pattern="*.R", full.names=T)
tmp = sapply(file.sources,source,.GlobalEnv)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
dev.off()
```

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn5570325';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of UFL_Mayo_ISB RNASeq data';

thisFileName <- 'Mayo_RNASeq.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_TC_ID <- 'syn4650257';
ALL_USED_IDs <- COUNT_TC_ID
COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F)

# Download expression data
COUNT_CE_ID <- 'syn5201012';
ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_CE_ID
COUNT_CE <- read.table(synGet(COUNT_CE_ID)@filePath, header=T, sep='\t', check.names = F)

COUNT = plyr::join_all(list(COUNT_TC, COUNT_CE)) %>%
  as.data.frame

# Get clinical metadata
METADATA_TC_ID <- 'syn3817650'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TC_ID
METADATA_TC <- read.table(synGet(METADATA_TC_ID)@filePath,sep=',',header=T, stringsAsFactors=F) %>%
  tidyr::separate(ID, c('Donor_ID', 'BrainRegion'), sep = '_') %>%
  dplyr::mutate(ID = paste(Donor_ID, BrainRegion, sep = '_'))

# Get clinical metadata
METADATA_CE_ID <- 'syn5223705'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CE_ID
METADATA_CE <- read.table(synGet(METADATA_CE_ID)@filePath,sep=',',header=T, stringsAsFactors=F) %>%
  tidyr::separate(SampleID, c('Donor_ID', 'BrainRegion')) %>%
  dplyr::mutate(ID = paste(Donor_ID, BrainRegion, sep = '_')) %>%
  dplyr::rename(Gender = Sex, FLOWCELL = Flowcell)

METADATA = rbindlist(list(METADATA_TC, METADATA_CE), use.names = T, fill = T) %>%
  as.data.frame
```

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
METADATA$BrainRegion.Diagnosis = paste(METADATA$BrainRegion, METADATA$Diagnosis, sep = '.')
METADATA <- METADATA %>%
  dplyr::filter(!(ID %in% c("742_CER", "11386_CER","11311_CER","1923_CER","11477_CER", "11344_CER", 
                            "1923_TCX", "11423_TCX", "1940_TCX", "7095_TCX", "11408_TCX")))
```
Following samples were considered outliers and removed based on PCA and/or eucledian distance:
```{r preprocess.data1}
# Match covariates to expression data
indToRetain = intersect(METADATA$ID, colnames(COUNT))

COUNT <- COUNT[, c('ensembl_id', intersect(colnames(COUNT), indToRetain))]

rownames(METADATA) = METADATA$ID
METADATA = METADATA[rownames(METADATA) %in% indToRetain,]

COUNT = COUNT[, c('ensembl_id', indToRetain)]
METADATA = METADATA[indToRetain,]

rownames(COUNT) = COUNT$ensembl_id
COUNT$ensembl_id = NULL
```
`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD portal

```{r diagnosisPlots, fig.height=6, fig.width = 12}
# RIN
p = list()
p[[1]] = ggplot(METADATA, aes(x = Diagnosis, y = RIN, fill = Tissue)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')

# AgeAtDeath
p[[2]] = ggplot(METADATA, aes(x = Diagnosis, y = as.numeric(AgeAtDeath), fill = Tissue)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeAtDeath') + theme(legend.position = 'top')

multiplot(plotlist = p, cols = 2)

# Source
vcd::mosaic(~ Source + Diagnosis + Tissue, data = METADATA)
```

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "AgeAtDeath")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$ID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Covariate correlation
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = CovariateAnalysis::getAssociationStatistics(COVARIATES, PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.4, row.width=0.1)
```

### Gene Filtering
Retain genes with atleast 50% of the samples have >= 1 gene counts
```{r gene.filtering, results='asis', echo=FALSE}
# Remove genes
EXPR <- getGeneFilteredGeneExprMatrix(COUNT, 
                                      MIN_GENE_CPM=1,
                                      MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5,
                                      verbose=T)
```

### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=NULL, plot=T)

# Initial design matrix with covariates under consideration
DM = getDesignMatrix(COVARIATES, Intercept = F)
DESIGN = getDesignMat2Fact(DM$design,FactorCovariates)
colnames(DESIGN) = substr(colnames(DESIGN),1,50)
```

Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data1, fig.height=10, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E)

# Plot first 2 PCs
plotdata <- data.frame(ID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2],
                       PC3=PC$rotation[,3],
                       PC4=PC$rotation[,4])

plotdata <- left_join(plotdata, METADATA, by="ID")

pl = list()
p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Tissue), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
p <- p + geom_text(aes(label= ID), size=4, hjust=0)
pl[[1]] <- p

p <- ggplot(plotdata, aes(x=PC3, y=PC4))
p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Diagnosis), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
pl[[2]] <- p

multiplot(plotlist = pl, cols = 1)
```
```{r decompse.normalise.data2, fig.height=9, fig.width=25, results='asis'}
# Tree based clustering
sampleTrees = flashClust::hclust(dist(t(VOOM.GENE_EXPRESSION$E)), method = "average")
tree = WGCNA::cutreeStatic(sampleTrees, cutHeight = 135, minSize=10)

COVARIATES.tmp = COVARIATES
COVARIATES.tmp[,FactorCovariates] = lapply(COVARIATES.tmp[,FactorCovariates], as.numeric)

WGCNA::plotDendroAndColors(sampleTrees, 
                           labels2colors(cbind(tree,COVARIATES.tmp)),
                           groupLabels = c('Cluster',colnames(COVARIATES.tmp)))
gc()
dev.off()

tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[tree == 0], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[tree == 3], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
```
### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, results='asis', echo=FALSE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, COVARIATES,
                                                 'NULL design(voom-normalized)',
                                                 isKeyPlot=TRUE)

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
```{r iterative.adjusted.voom.normalisation, results='asis', echo=FALSE}
postAdjustCovars = c('BrainRegion.Diagnosis')
randomCovariates = 'Donor_ID'

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
significantCovars = residualSigCovars$significantCovars
covariatesEffects = residualSigCovars$Effects.significantCovars
covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates))]

postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))

loopCount = 0 
while(length(significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:', 
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=DM1$design, plot=F)

  # Calculate correlation values of random effects
  correlation = parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block = COVARIATES$Donor_ID)

  # Re-estimate voom weights
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=DM1$design, plot=F, 
                                       block = COVARIATES$Donor_ID, correlation = correlation$cor)

  # Fit a mixed effect linear model
  FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, randomCovariates))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES,
                                                'all adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  significantCovars = residualSigCovars$significantCovars
  covariatesEffects = residualSigCovars$Effects.significantCovars
  covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars, randomCovariates))]

  postAdjustCovars = c(postAdjustCovars,names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
modelStr <-paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=','), 'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`

### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Adding "BrainRegion.Diagnosis" back to the residuals
```{r addVarsBack}
varsToAddBackIn <-  grep("BrainRegion.Diagnosis", colnames(FIT$coefficients), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'ensembl_gene_id') 
```

### Clustering residual gene expression
```{r decompse.normalise.data3, fig.height=10, fig.width=9, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1)], scale.=T)

# Tree based clustering
sampleTrees = flashClust::hclust(dist(t(RESIDUAL.GENE_EXPRESSION[,-(1)])), method = "average")

# Cluster expression
clust = cutreeStatic(sampleTrees, cutHeight = 100, minSize = 10)

# Plot first 2 PCs
plotdata <- data.frame(ID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2],
                       PC3=PC$rotation[,3], 
                       PC4=PC$rotation[,4],
                       clust = clust)

plotdata <- left_join(plotdata, METADATA, by="ID")

pl = list()
p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Tissue), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
p <- p + geom_text(aes(label= ID), size=4, hjust=0)
pl[[1]] <- p

p <- ggplot(plotdata, aes(x=PC3, y=PC4))
p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Diagnosis), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
pl[[2]] <- p

multiplot(plotlist = pl, cols = 1)

names(clust) = sampleTrees$labels
clust = clust[rownames(COVARIATES)]
```
```{r decompse.normalise.data4, fig.height=10, fig.width=25, results='asis'}
# Tree based clustering
plotDendroAndColors(sampleTrees, 
                    colors = labels2colors(cbind(data.frame(clust = clust), COVARIATES)), 
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES))
gc()
dev.off()

tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[tree == 0], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
tmp = paste('Samples in cluster 1 are', paste(sampleTrees$labels[tree == 1], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[tree == 3], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
```

```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = data.frame(summary(COVARIATES$BrainRegion.Diagnosis))
colnames(tmp) = "No. of samples"
kable(tmp)
```

```{r get.gene.symbols}
# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "www.ensembl.org")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = RESIDUAL.GENE_EXPRESSION$ensembl_gene_id,
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(RESIDUAL.GENE_EXPRESSION, Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=16}
# Fit contrast
contrast = makeContrasts(contrasts=c(
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.Pathologic_Aging",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.PSP-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.Pathologic_Aging",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.PSP",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.PSP"), 
                         levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = list()
p = list()
for (i in 1:12){
  DE[[i]] = topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
  DE[[i]] = rownameToFirstColumn(DE[[i]], 'ensembl_gene_id') %>%
    left_join(Ensemble2HGNC)
  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red') 
  p[[i]] <- p[[i]] + ggtitle(gsub('BrainRegion.Diagnosis','',colnames(contrast)[i]))
}
multiplot(plotlist=p, cols = 3)
names(DE) = gsub('BrainRegion.Diagnosis','',colnames(contrast))

DE = plyr::ldply(DE, .id = 'Comparison')
tmp = DE %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(Comparison) %>%
  dplyr::summarize(nDiffExpGenes = length(unique(hgnc_symbol)))
kable(tmp)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB RNASeq', parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'ID')
write.table(COVARIATES, file = 'covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store covariates
write.table(RESIDUAL.GENE_EXPRESSION, file = 'expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('expression.tsv', name = 'Voom Adjusted Residual Expression (BrainRegion.Diagnosis added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store differential expression results
write.table(DE, file = 'diffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('diffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
stopCluster(cl)
```
|  *Results*            |  *SynapseID*                    |
|  -------              |   ---------                     |
|  Covariates           |  `r COV_OBJ$properties$id`      |
|  Residual Expression  |  `r EXP_OBJ$properties$id`      |
|  Differential Expression  |  `r DEXP_OBJ$properties$id`      |

### Source R Code
[This File](`r thisFile`)

### 2016-05-16 code/Rmd/Mayo_RNASeq.Rmd
Added RIN2 to the Mayo pipeline

@@
 METADATA$BrainRegion.Diagnosis = paste(METADATA$BrainRegion, METADATA$Diagnosis, sep = '.')
+METADATA$RIN2 = METADATA$RIN^2
 METADATA <- METADATA %>%
@@
 FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
-ContCovariates <- c("RIN", "AgeAtDeath")
+ContCovariates <- c("RIN", "RIN2", "AgeAtDeath")
 

### 2016-05-16 code/Rmd/Mayo_RNASeq.Rmd
Minor sanity fix

@@
 
-```{r diagnosisPlots, fig.height=6, fig.width = 12}
+```{r diagnosisPlots, fig.height=10, fig.width = 15}
 # RIN
@@
 Clustering of initial normalised data (with NULL design)
-```{r decompse.normalise.data1, fig.height=10, fig.width=10, results='asis'}
+```{r decompse.normalise.data1, fig.height=15, fig.width=10, results='asis'}
 # Find principal components of expression to plot
@@
                            groupLabels = c('Cluster',colnames(COVARIATES.tmp)))
-gc()
 dev.off()
@@
 ```{r preAdjusted.covariates, results='asis', echo=FALSE}
+gc()
 # Find correlation between PC's of gene expression with covariates
@@
 # Tree based clustering
+COVARIATES.tmp = COVARIATES
+COVARIATES.tmp[, FactorCovariates] = sapply(COVARIATES.tmp[,FactorCovariates], as.numeric)
 plotDendroAndColors(sampleTrees, 
-                    colors = labels2colors(cbind(data.frame(clust = clust), COVARIATES)), 
+                    colors = labels2colors(cbind(data.frame(clust = clust), COVARIATES.tmp)), 
                     abHeight = 0.80, 
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=10, fig.width=16}
+```{r diffExp, fig.height=25, fig.width=25}
 # Fit contrast

### 2016-05-16 code/Rmd/Mayo_MayoBrainBankDickson_RNASeq.Rmd
Added markdown to process MayoBrainBank samples alone (removed BannerHealth samples)

--- 2016-05-16 code/Rmd/Mayo_RNASeq.Rmd
@@
 ---
-title: "Covariate analysis of UFL-Mayo-ISB RNASeq (both cerebellum and temporal cortex)"
+title: "Covariate analysis of UFL-Mayo-ISB RNASeq (both cerebellum and temporal cortex) samples from MayoBrainBank_Dickson alone"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./Mayo_MayoBrainBankDickson_RNASeq.Rmd",
                                  parentId = "syn5570325",
-                                 entityName = 'UFL_Mayo_ISB RNASeq')
+                                 entityName = 'UFL_Mayo_ISB Mayo Brain Bank Dickson RNASeq')
 ```
@@
 
-thisFileName <- 'Mayo_RNASeq.Rmd'
+thisFileName <- 'Mayo_MayoBrainBankDickson_RNASeq.Rmd'
 
@@
 METADATA$BrainRegion.Diagnosis = paste(METADATA$BrainRegion, METADATA$Diagnosis, sep = '.')
+METADATA$RIN2 = METADATA$RIN^2
 METADATA <- METADATA %>%
   dplyr::filter(!(ID %in% c("742_CER", "11386_CER","11311_CER","1923_CER","11477_CER", "11344_CER", 
-                            "1923_TCX", "11423_TCX", "1940_TCX", "7095_TCX", "11408_TCX")))
+                            "1923_TCX", "11423_TCX", "1940_TCX", "7095_TCX", "11408_TCX"))) %>%
+  filter(Source == 'MayoBrainBank_Dickson')
 ```
@@
 ```{r covariates.clustering}
-FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
-ContCovariates <- c("RIN", "AgeAtDeath")
+FactorCovariates <- c("Donor_ID", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
+ContCovariates <- c("RIN", "RIN2", "AgeAtDeath")
 
@@
                     groupLabels = colnames(COVARIATES))
-gc()
-dev.off()
 
@@
 writeLines(tmp)
-tmp = paste('Samples in cluster 1 are', paste(sampleTrees$labels[tree == 1], collapse = ','))
+tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[tree == 3], collapse = ','))
 tmp = base::gsub('_','\\\\_',tmp)
 writeLines(tmp)
-tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[tree == 3], collapse = ','))
+tmp = paste('Samples in cluster 4 are', paste(sampleTrees$labels[tree == 4], collapse = ','))
 tmp = base::gsub('_','\\\\_',tmp)
@@
   "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.AD",
-  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.Pathologic_Aging",
   "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.PSP",
-  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.AD",
   "BrainRegion.DiagnosisCER.PSP-BrainRegion.DiagnosisCER.AD",
-  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.PSP",
   "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.AD",
-  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.Pathologic_Aging",
   "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.PSP",
-  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.AD",
-  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD",
-  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.PSP"), 
+  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD"), 
                          levels = colnames(FIT$coefficients))
@@
 p = list()
-for (i in 1:12){
+for (i in 1:6){
   DE[[i]] = topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
@@
 # Code
-CODE <- Folder(name = 'UFL_Mayo_ISB RNASeq', parentId = parentId)
+CODE <- Folder(name = 'UFL_Mayo_ISB Mayo Brain Bank Dickson RNASeq', parentId = parentId)
 CODE <- synStore(CODE)

### 2016-05-16 code/Rmd/Mayo_MayoBrainBankDickson_RNASeq.Rmd
Minor sanity changes

@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=25, fig.width=25}
+```{r diffExp, fig.height=16, fig.width=25}
 # Fit contrast

### 2016-05-16 code/Rmd/Mayo_RNASeq_CellMarkersAdjusted.Rmd
Added markdown to adjust for cell types in Mayo RNASeq data

--- 2016-05-16 code/Rmd/Mayo_RNASeq.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_CellMarkersAdjusted.Rmd",
                                  parentId = "syn5570325",
-                                 entityName = 'UFL_Mayo_ISB RNASeq')
+                                 entityName = 'UFL_Mayo_ISB RNASeq Cell Markers Adjusted')
 ```
@@
   cache = TRUE)
-dev.off()
 ```
@@
 
-thisFileName <- 'Mayo_RNASeq.Rmd'
+thisFileName <- 'Mayo_RNASeq_CellMarkersAdjusted.Rmd'
 
@@
 
+### Cell type markers (i.e., eigen genes of cell types) as additional covariates
+```{r cellMark}
+load(synGet('syn4893059')@filePath)
+GeneSets = GeneSets[grep("Zhang", names(GeneSets))] %>%
+  ldply(.fun = function(x){
+    data.frame(hgnc_symbol = x)
+  }, .id = 'CellType') %>%
+  dplyr::mutate(CellType = gsub("Zhang:","",CellType)) %>%
+  left_join(Ensemble2HGNC) %>%
+  filter(!is.na(ensembl_gene_id))
+
+kable(data.frame(table(GeneSets$CellType)) %>%
+        dplyr::rename(CellType = Var1, NumberOfGenes = Freq))
+
+COVARIATES.cellmarkers = dlply(GeneSets, .(CellType), .fun = function(gs){
+  EXPR = COUNT[rownames(COUNT) %in% gs$ensembl_gene_id, ]
+  PC <- prcomp(EXPR, scale.=T, center = T)
+  rownameToFirstColumn(PC$rotation[,1, drop=F], 'ID') %>%
+    plyr::rename(c('PC1' = unique(gs$CellType)))
+}) %>%
+  plyr::join_all(by = 'ID')
+
+METADATA = join_all(list(METADATA, COVARIATES.cellmarkers))
+
+p = METADATA %>%
+  dplyr::select(Diagnosis, Tissue, one_of(colnames(COVARIATES.cellmarkers))) %>%
+  gather(CellType, EigenGeneExpression, -ID, -Diagnosis, -Tissue) %>%
+  ggplot(aes(x = Diagnosis, y = abs(EigenGeneExpression), fill = Tissue)) + geom_boxplot() + facet_grid(.~CellType)
+p                         
+```
+
 ### Covariate clustering
@@
 FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
-ContCovariates <- c("RIN", "RIN2", "AgeAtDeath")
+ContCovariates <- c("RIN", "RIN2", "AgeAtDeath", "Astrocyte", "Endothelial", "Microglia", "MyelinOligos", "Neuron", "NewOligos", "OPC")
 
@@
 Clustering of initial normalised data (with NULL design)
-```{r decompse.normalise.data1, fig.height=15, fig.width=10, results='asis'}
+```{r decompse.normalise.data1, fig.height=15, fig.width=8, results='asis'}
 # Find principal components of expression to plot
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Tissue), size=as.numeric(AgeAtDeath)))
+p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Tissue), size=as.numeric(OPC)))
 p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
-p <- p + geom_text(aes(label= ID), size=4, hjust=0)
+# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
 pl[[1]] <- p
@@
 p <- ggplot(plotdata, aes(x=PC3, y=PC4))
-p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Diagnosis), size=as.numeric(AgeAtDeath)))
+p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Diagnosis), size=as.numeric(OPC)))
 p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
@@
 WGCNA::plotDendroAndColors(sampleTrees, 
-                           labels2colors(cbind(tree,COVARIATES.tmp)),
-                           groupLabels = c('Cluster',colnames(COVARIATES.tmp)))
+                           labels2colors(cbind(tree,COVARIATES.tmp[, FactorCovariates])),
+                           groupLabels = c('Cluster',colnames(COVARIATES.tmp[, FactorCovariates])))
 dev.off()
@@
 ```{r preAdjusted.covariates, results='asis', echo=FALSE}
-gc()
 # Find correlation between PC's of gene expression with covariates
@@
 p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
-p <- p + geom_text(aes(label= ID), size=4, hjust=0)
+# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
 pl[[1]] <- p
@@
 plotDendroAndColors(sampleTrees, 
-                    colors = labels2colors(cbind(data.frame(clust = clust), COVARIATES.tmp)), 
+                    colors = labels2colors(cbind(data.frame(clust = clust), COVARIATES.tmp[,FactorCovariates])), 
                     abHeight = 0.80, 
@@
                     groupLabels = colnames(COVARIATES))
-gc()
-dev.off()
 
 tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[tree == 0], collapse = ','))
-tmp = base::gsub('_','\\\\_',tmp)
-writeLines(tmp)
-tmp = paste('Samples in cluster 1 are', paste(sampleTrees$labels[tree == 1], collapse = ','))
 tmp = base::gsub('_','\\\\_',tmp)
@@
 # Code
-CODE <- Folder(name = 'UFL_Mayo_ISB RNASeq', parentId = parentId)
+CODE <- Folder(name = 'UFL_Mayo_ISB RNASeq Cell Markers Adjusted', parentId = parentId)
 CODE <- synStore(CODE)

### 2016-05-16 code/Rmd/Mayo_RNASeq_CellMarkersAdjusted.Rmd
Minor changes

@@
 ### Cell type markers (i.e., eigen genes of cell types) as additional covariates
-```{r cellMark}
+```{r cellMark, fig.height=10, fig.width = 15}
 load(synGet('syn4893059')@filePath)
@@
   ggplot(aes(x = Diagnosis, y = abs(EigenGeneExpression), fill = Tissue)) + geom_boxplot() + facet_grid(.~CellType)
+p = p +  theme(axis.text.x = element_text(angle = 90, hjust = 1))   
 p
@@
 # p <- p + geom_text(aes(label= ID), size=4, hjust=0)
-pl[[1]] <- p
+pl[[1]] <- p + xlab(paste0('PC1(', round(PC$sdev[1]^2/sum(PC$sdev^2)*100), '%)')) + ylab(paste0('PC2(', round(PC$sdev[2]^2/sum(PC$sdev^2)*100), '%)'))
 
@@
 # p <- p + geom_text(aes(label= ID), size=4, hjust=0)
-pl[[2]] <- p
+pl[[2]] <- p + xlab(paste0('PC3(', round(PC$sdev[3]^2/sum(PC$sdev^2)*100), '%)')) + ylab(paste0('PC4(', round(PC$sdev[4]^2/sum(PC$sdev^2)*100), '%)'))
 
-multiplot(plotlist = pl, cols = 1)
+multiplot(plotlist = pl, cols = 2)
 ```
@@
 ### Clustering residual gene expression
-```{r decompse.normalise.data3, fig.height=10, fig.width=9, results='asis'}
+```{r decompse.normalise.data3, fig.height=15, fig.width=8, results='asis'}
 # Find principal components of expression to plot
@@
   p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red') 
+  p[[i]] <- p[[i]] + geom_vline(xintercept = -log2(1.5), color = 'red')
+  p[[i]] <- p[[i]] + geom_vline(xintercept = log2(1.5), color = 'red')
   p[[i]] <- p[[i]] + ggtitle(gsub('BrainRegion.Diagnosis','',colnames(contrast)[i]))
@@
 tmp = DE %>% 
-  dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.5)) %>%
   group_by(Comparison) %>%

### 2016-05-18 code/Rmd/MSSM_RNASeq.Rmd
Minor modifications

@@
 ```
+
+```{r dataExplore, fig.width = 10, fig.height = 10}
+# RIN
+p = list()
+p[[1]] = ggplot(METADATA, aes(x = Dx, y = RIN, fill = BrainRegion)) + geom_boxplot()
+p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')
+
+# AgeAtDeath
+p[[2]] = ggplot(METADATA, aes(x = Dx, y = as.numeric(AOD), fill = BrainRegion)) + geom_boxplot()
+p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + theme(legend.position = 'top')
+
+# Mean Plaque
+p[[3]] = ggplot(METADATA, aes(x = Dx, y = as.numeric(PlaqueMean), fill = BrainRegion)) + geom_boxplot()
+p[[3]] = p[[3]] + ggtitle('PlaqueMean') + theme(legend.position = 'top')
+
+# Mean Plaque
+p[[4]] = ggplot(METADATA, aes(x = Dx, y = as.numeric(NP.1), fill = BrainRegion)) + geom_boxplot()
+p[[4]] = p[[4]] + ggtitle('NP.1') + theme(legend.position = 'top')
+
+multiplot(plotlist = p, cols = 2)
+
+# Sex
+vcd::mosaic(~ SEX + Dx + BrainRegion, data = METADATA)
+```
+
 ### CPM Normalisation

### 2016-05-20 code/R/lib/g_legend.R
Added R code to plot ggheatmap

@@ -1,0 +1,8 @@
+require(ggplot2)
+g_legend<-function(a.gplot){
+  ## from http://stackoverflow.com/questions/11883844/inserting-a-table-under-the-legend-in-a-ggplot2-histogram
+  tmp <- ggplot2::ggplot_gtable(ggplot2::ggplot_build(a.gplot))
+  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
+  legend <- tmp$grobs[[leg]]
+  return(legend)
+}

### 2016-05-20 code/R/lib/getAssociationStatistics.R

@@ -1,18 +1,25 @@
 getAssociationStatistics <- function(COVARIATES, PVAL=0.05){
 
+  require(plyr)
   # Get factor and continuous covariates
   FactorCovariates <- names(COVARIATES)[sapply(COVARIATES,is.factor)]
   ContCovariates <- setdiff(names(COVARIATES),FactorCovariates)
   
-  # Convert factor covariates to numeric vector
-  COVARIATES[,FactorCovariates] <- lapply(COVARIATES[,FactorCovariates],
-                                          function(x){x <- as.numeric(unclass(x))})
-
+  # Convert factor covariates to numeric vector of factors
+  COVARIATES[,FactorCovariates] <- lapply(COVARIATES[,FactorCovariates],as.numeric)
+  COVARIATES[,FactorCovariates] <- lapply(COVARIATES[,FactorCovariates],as.factor)
+  
   # Find association between factor covariates
-  if (length(FactorCovariates) != 0 & length(FactorCovariates) != 1){
-    COVARIATES.FACTOR.CORRELATION = apply(expand.grid(FactorCovariates,FactorCovariates),1,getFactorAssociationStatistics,COVARIATES[,FactorCovariates])
-    COVARIATES.FACTOR.CORRELATION.ESTIMATE <- matrix(COVARIATES.FACTOR.CORRELATION['Estimate',],nrow=length(FactorCovariates),ncol=length(FactorCovariates))
-    COVARIATES.FACTOR.CORRELATION.PVAL <- matrix(COVARIATES.FACTOR.CORRELATION['Pval',],nrow=length(FactorCovariates),ncol=length(FactorCovariates))
+  if (length(FactorCovariates) > 1){
+    COVARIATES.FACTOR.CORRELATION = apply(expand.grid(FactorCovariates,FactorCovariates),
+                                          1, getFactorAssociationStatistics,
+                                          COVARIATES[,FactorCovariates])
+    COVARIATES.FACTOR.CORRELATION.ESTIMATE <- matrix(COVARIATES.FACTOR.CORRELATION['Estimate',],
+                                                     nrow=length(FactorCovariates),
+                                                     ncol=length(FactorCovariates))
+    COVARIATES.FACTOR.CORRELATION.PVAL <- matrix(COVARIATES.FACTOR.CORRELATION['Pval',],
[…]

### 2016-05-20 code/R/lib/getFactorAssociationStatistics.R

@@ -8,7 +8,7 @@ getFactorAssociationStatistics <- function(factorNames,COVARIATES, na.action='re
   fac1 = as.factor(COVARIATES[,1])
   fac2 = as.factor(COVARIATES[,2])
   
-  stats = assocstats(xtabs(~fac1+fac2))
+  stats = vcd::assocstats(xtabs(~fac1+fac2))
   
   return(c(Estimate=stats$cramer,Pval=stats$chisq_tests['Pearson','P(> X^2)']))
 }

### 2016-05-20 code/R/lib/getFactorContAssociationStatistics.R

@@ -3,7 +3,6 @@
 getFactorContAssociationStatistics <- function(factorContNames,COVARIATES, na.action='remove', 
                                                alpha = 0.05){
   require(psych)
-  
   factorContNames = factorContNames %>% unlist %>% as.character()
   
   if (na.action == "remove")

### 2016-05-20 code/R/lib/ggheatmap.R

@@ -1,0 +1,62 @@
+##' generate a heatmap + dendrograms, ggplot2 style
+##'
+##' @param x data matrix
+##' @param hm.colours vector of colours (optional)
+##' @return invisibly returns a list of ggplot2 objects. Display them with ggheatmap.show()
+##' @author Chris Wallace
+##' @export
+##' @examples
+##' ## test run
+##' ## simulate data
+##' library(mvtnorm)
+##' sigma=matrix(0,10,10)
+##' sigma[1:4,1:4] <- 0.6
+##' sigma[6:10,6:10] <- 0.8
+##' diag(sigma) <- 1
+##' X <- rmvnorm(n=100,mean=rep(0,10),sigma=sigma)
+##'  
+##' ## make plot
+##' p <- ggheatmap(X)
+##'  
+##' ## display plot
+##' ggheatmap.show(p)
+require(ggdendro)
+require(ggplot2)
+ggheatmap <- function(x,
+                      hm.colours=my.colours) {
+  # Sourced from https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R
+  if(is.null(colnames(x)))
+    colnames(x) <- sprintf("col%s",1:ncol(x))
+  if(is.null(rownames(x)))
+    rownames(x) <- sprintf("row%s",1:nrow(x))
+  ## plot a heatmap
+  ## x is an expression matrix
+  row.hc <- hclust(dist(x), "ward")
+  col.hc <- hclust(dist(t(x)), "ward")
+  row.dendro <- ggdendro::dendro_data(as.dendrogram(row.hc),type="rectangle")
+  col.dendro <- ggdendro::dendro_data(as.dendrogram(col.hc),type="rectangle")
+  
+  ## dendro plots
+  col.plot <- mydplot(col.dendro, col=TRUE, labels=TRUE) +
+    ggplot2::scale_x_continuous(breaks = 1:ncol(x),labels=col.hc$labels[col.hc$order]) +
+    ggplot2::theme(plot.margin = unit(c(0,0,0,0), "lines"))
+  row.plot <- mydplot(row.dendro, row=TRUE, labels=FALSE) +
+    ggplot2::theme(plot.margin = unit(rep(0, 4), "lines"))
+  
+  ## order of the dendros
+  col.ord <- match(col.dendro$labels$label, colnames(x))
+  row.ord <- match(row.dendro$labels$label, rownames(x))
[…]

### 2016-05-20 code/R/lib/ggheatmap.show.R

@@ -1,0 +1,37 @@
+##' Display a ggheatmap
+##'
+##' this function sets up some viewports, and tries to plot the dendrograms to line up with the heatmap
+##' @param L a list with 3 named plots: col, row, centre, generated by ggheatmap
+##' @param col.width,row.width number between 0 and 1, fraction of the device devoted to the column or row-wise dendrogram. If 0, don't print the dendrogram
+##' @return no return value, side effect of displaying plot in current device
+##' @export
+##' @author Chris Wallace
+require(grid)
+ggheatmap.show <- function(L, col.width=0.2, row.width=0.2) {
+  # Function sourced from https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R
+  grid::grid.newpage()
+  top.layout <- grid::grid.layout(nrow = 2, ncol = 2,
+                                  widths = unit(c(1-row.width,row.width), "null"),
+                                  heights = unit(c(col.width,1-row.width), "null"))
+  grid::pushViewport(grid::viewport(layout=top.layout))
+  if(col.width>0)
+    print(L$col, vp=grid::viewport(layout.pos.col=1, layout.pos.row=1))
+  if(row.width>0)
+    print(L$row, vp=grid::viewport(layout.pos.col=2, layout.pos.row=2))
+  ## print centre without legend
+  print(L$centre +
+          ggplot2::theme(axis.line=element_blank(),
+                       axis.text.x=element_blank(),axis.text.y=element_blank(),
+                       axis.ticks=element_blank(),
+                       axis.title.x=element_blank(),axis.title.y=element_blank(),
+                       legend.position="none",
+                       panel.background=element_blank(),
+                       panel.border=element_blank(),panel.grid.major=element_blank(),
+                       panel.grid.minor=element_blank(),plot.background=element_blank()),
+        vp=grid::viewport(layout.pos.col=1, layout.pos.row=2))
+  ## add legend
+  legend <- g_legend(L$centre)
+  grid::pushViewport(grid::viewport(layout.pos.col=2, layout.pos.row=1))
+  grid::grid.draw(legend)
+  grid::upViewport(0)
+}

### 2016-05-20 code/R/lib/mydplot.R

@@ -1,0 +1,22 @@
+require(ggplot2)
+require(ggdendro)
+mydplot <- function(ddata, row=!col, col=!row, labels=col) {
+  ## plot a dendrogram
+  yrange <- range(ddata$segments$y)
+  yd <- yrange[2] - yrange[1]
+  nc <- max(nchar(as.character(ddata$labels$label)))
+  tangle <- if(row) { 0 } else { 90 }
+  tshow <- col
+  p <- ggplot2::ggplot() +
+    ggplot2::geom_segment(data=segment(ddata), aes(x=x, y=y, xend=xend, yend=yend)) +
+    ggplot2::labs(x = NULL, y = NULL) + ggdendro::theme_dendro()
+  if(row) {
+    p <- p +
+      ggplot2::scale_x_continuous(expand=c(0.5/length(ddata$labels$x),0)) +
+      ggplot2::coord_flip()
+  } else {
+    p <- p +
+      ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1))
+  }
+  return(p)
+}

### 2016-05-20 code/Rmd/ROSMAP_RNASeq_Reprocessed.Rmd
Added markdown to process ROSMAP reprocessed RNASeq data

---
title: "Covariate analysis of ROSMAP RNASeq"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq.Rmd",
                                 parentId ="syn5570291",
                                 entityName = 'ROSMAP RNASeq Reprocessed Counts (Based on cogdx)')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get it from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(reshape2)
library(limma)
library(Biobase)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570291';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of ROSMAP reprocessed RNASeq data';

thisFileName <- 'ROSMAP_RNASeq_Reprocessed.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_ID <- 'syn6112813';
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F)

# Get clinical metadata
METADATA.CLINICAL_ID <- 'syn3191087'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.CLINICAL_ID
METADATA.CLINICAL_OBJ <- synGet(METADATA.CLINICAL_ID)
METADATA.CLINICAL <- read.table(METADATA.CLINICAL_OBJ@filePath,sep=',',header=T)

# Get technical covariates
METADATA.TECH_ID <- 'syn4300313'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.TECH_ID
METADATA.TECH_OBJ <- synGet(METADATA.TECH_ID)
METADATA.TECH <- read.table(METADATA.TECH_OBJ@filePath,sep='\t',header=T)

# Fix error in technical covariates data
KEY_ID <- 'syn3382527'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = KEY_ID  
KEY <- read.csv(synGet(KEY_ID)@filePath) %>% 
  filter(mrna_data == 1) %>%
  dplyr::select(projid, mrna_id) %>%
  separate(mrna_id, c('a','b','batch'), sep = '_') %>% 
  unite(Sampleid, a, b) %>%
  dplyr::select(-batch) %>%
  unique

# Match technical and clinical covariates
METADATA <- METADATA.TECH %>% 
  dplyr::select(-projid) %>%
  dplyr::left_join(KEY) %>%
  dplyr::left_join(METADATA.CLINICAL)

# Pick higher quality RIN batch for sample 492_120515
METADATA <- METADATA %>%
  group_by(Sampleid) %>%
  top_n(1, RINcontinuous)
```

### Data preprocessing
```{r preprocess.data,cache=TRUE, echo=TRUE}
# Remove samples with no cogdx, RIN and PMI scores
METADATA <- METADATA %>%
  filter(Sampleid %in% colnames(COUNT)) %>%
  filter(!is.na(cogdx)) %>%
  filter(!is.na(RINcontinuous)) %>%
  filter(!is.na(pmi)) %>%
  as.data.frame %>%
  filter(!(Sampleid %in% c('380_120503')))
```

```{r preprocess.data1,cache=TRUE}
# Match covariates to expression data
indToRetain = intersect(METADATA$Sampleid, colnames(COUNT))
removedIDs = setdiff(colnames(COUNT), METADATA$Sampleid)

COUNT = COUNT[,indToRetain]

rownames(METADATA) = METADATA$Sampleid
METADATA = METADATA[indToRetain,]
```
Dorsolateral prefrontal cortex of `r dim(COUNT)[2]` subjects from the ROS and MAP cohorts are used for the analysis. Following sample are removed `r paste(removedIDs, collapse = ',')`

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering, cache=TRUE}
#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv"
FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx")
ContCovariates <- c("RINcontinuous","age_first_ad_dx","age_death","age_at_visit_max","pmi","educ")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES <- data.frame(lapply(COVARIATES,function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'\\+','')})}))
rownames(COVARIATES) <- METADATA$Sampleid

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)

# Add in RIN^2 values
COVARIATES$RINcontinuous2 = COVARIATES$RINcontinuous^2
ContCovariates = c(ContCovariates, 'RINcontinuous2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.1)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.5, row.width=0.1)
```

```{r dataExplore, fig.width = 10, fig.height = 10}
# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = cogdx, y = RINcontinuous)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')

# AgeAtDeath
p[[2]] = ggplot(COVARIATES, aes(x = cogdx, y = age_death)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + theme(legend.position = 'top')

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = cogdx, y = pmi)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI') + theme(legend.position = 'top')

# Education
p[[4]] = ggplot(COVARIATES, aes(x = cogdx, y = educ)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Education') + theme(legend.position = 'top')

multiplot(plotlist = p, cols = 2)

# Batch
vcd::mosaic(~ Batch + cogdx, data = COVARIATES)

# Sex
vcd::mosaic(~ msex + cogdx, data = COVARIATES)
```

### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

Remove genes that have less than 1 cpm counts in at least 50% of samples.
```{r cpmnormalisation, cache=TRUE}
PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT,MIN_GENE_CPM=1, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
```
### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T)
```
### Outlier analysis
Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(ID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'ID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=cogdx, shape=Batch, size=age_death))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p

# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,FactorCovariates]))
dev.off()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```
Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "cogdx" is chosen as the primary variable of interest
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = 'cogdx'

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate correlation of random effects
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), postAdjustCovars)

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "cogdx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("cogdx", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T)

# Plot first 2 PCs
plotdata <- data.frame(ID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'ID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=cogdx, shape = Batch, size=RINcontinuous))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p

# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                          colors = cols, 
                          dendroLabels = FALSE, 
                          abHeight = 0.80, 
                          main = "Sample dendrogram",
                          groupLabels = colnames(COVARIATES.tmp[,FactorCovariates]))
dev.off()
```

```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = data.frame(cogdx = 1:6, No.OfSamples = summary(COVARIATES$cogdx))
kable(tmp)
```

Interpretation of cogdx scores

1.  NCI, No cognitive impairment (No impaired domains)
2.  MCI, Mild cognitive impairment (One impaired domain) and NO other cause of CI
3.  MCI, Mild cognitive impairment (One impaired domain) AND another cause of CI
4.  AD, Alzheimer's disease and NO other cause of CI (NINCDS PROB AD)
5.  AD, Alzheimer's disease AND another cause of CI (NINCDS POSS AD)
6.  Other dementia. Other primary cause of dementia

```{r get.gene.symbols, cache=TRUE, eval=T}
row.id = data.frame(GeneID = rownames(RESIDUAL.GENE_EXPRESSION),
                    ensembl_gene_id = rownames(RESIDUAL.GENE_EXPRESSION)) %>%
  separate(ensembl_gene_id, c('ensembl_gene_id','position'), sep = '\\.')

# Define biomart object
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = row.id$ensembl_gene_id,
                       mart = mart)

# Add gene names
RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('GeneID') %>%
  left_join(row.id) %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('GeneID') %>%
  right_join(row.id) %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=8, fig.width=16}
# Fit contrast
contrast = makeContrasts(contrasts=c("cogdx1-cogdx2",
                                     "cogdx1-cogdx4",
                                     "cogdx2-cogdx4"), 
                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = list()
p = list()
for (i in 1:3){
  DE[[i]] = topTable(FIT.CONTR, coef=i, number = dim(COUNT)[1])
  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id") 
}
names(DE) = colnames(contrast)

DE = DE %>% plyr::ldply(.id = 'Comparison') %>%
  tidyr::separate(ensembl_gene_id, into=c("ensemble_id","position")) %>%
  dplyr::rename(ensembl_gene_id = ensemble_id) %>%
  left_join(Ensemble2HGNC)

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(Comparison, hgnc_symbol) %>%
  group_by(Comparison) %>%
  dplyr::summarise(nDiffExp = length(unique(hgnc_symbol)), 
            DiffExpGenes = paste(unique(hgnc_symbol), collapse = ','))

kable(tmp)

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') + facet_grid(.~Comparison, scales = 'fixed')
p
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts (Based on cogdx)", parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'Sampleid')
write.table(COVARIATES, file = 'ROSMAPCovariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('ROSMAPCovariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store logCPM
write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'ROSMAPlogCPM.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('ROSMAPlogCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                     executed = thisFile, activityDescription = activityDescription)

# Store residual gene expression
write.table(RESIDUAL.GENE_EXPRESSION, file = 'ROSMAPExpression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('ROSMAPExpression.tsv', name = 'Normalised Adjusted Residual Expression (cogdx added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store differential expression results
write.table(DE, file = 'ROSMAPDiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('ROSMAPDiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
```
|  *Results*                |  *SynapseID*                     |
|  -------                  |   ---------                      |
|  Covariates               |  `r COV_OBJ$properties$id`       |
|  Raw counts (log CPM)     |  `r COUNT_OBJ$properties$id`     |
|  Residual Expression      |  `r EXP_OBJ$properties$id`       |
|  Differential Expression  |  `r DEXP_OBJ$properties$id`      |

### 2016-05-20 code/Rmd/ROSMAP_RNASeq_Reprocessed.Rmd
Minor modifications

@@
 ---
-title: "Covariate analysis of ROSMAP RNASeq"
+title: "Covariate analysis of ROSMAP Reprocessed RNASeq"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed.Rmd",
                                  parentId ="syn5570291",
@@
 ```
-
 ### CPM Normalisation
@@
 ### Outlier analysis
-Clustering of initial normalised data (with NULL design)
+Clustering of voom adjusted data (with NULL design)
 ```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}

### 2016-05-24 code/Rmd/MSSM_RNASeq.Rmd
Added section to store raw log cpm values to synapse

@@
 ## Load required libraries
-library(CovariateAnalysis)
+library(CovariateAnalysis) # Get the utilities package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
 library(data.table)
@@
 COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
-ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.3, row.width=0.15)
+ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.5, row.width=0.1)
 ```
@@
 
-# Mean Plaque
+# NP.1
 p[[4]] = ggplot(METADATA, aes(x = Dx, y = as.numeric(NP.1), fill = BrainRegion)) + geom_boxplot()
@@
 # Find principal components of expression to plot
-PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T)
+PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)
 
@@
 
-COVARIATES.tmp = data.matrix(COVARIATES)
+# Eucledian tree based analysis
+COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
@@
 cols = WGCNA::labels2colors(COVARIATES.tmp);
-tmp = gsub('\\..*','',cols)
+
 WGCNA::plotDendroAndColors(tree, 
-                           colors = tmp, 
+                           colors = cols, 
                            dendroLabels = FALSE, 
@@
                            main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp))
+                           groupLabels = colnames(COVARIATES.tmp[,FactorCovariates]))
 dev.off()
@@
 # Find correlation between PC's of gene expression with covariates
-COVARIATES.tmp = COVARIATES
-COVARIATES.tmp[, FactorCovariates] = sapply(COVARIATES.tmp[, FactorCovariates], as.numeric)
-preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, COVARIATES.tmp,'NULL design(voom-normalized)', isKeyPlot=TRUE, MIN_PVE_PCT_PC = 1)
+preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, COVARIATES,'NULL design(voom-normalized)', isKeyPlot=TRUE, MIN_PVE_PCT_PC = 1)
 ```
@@
   residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                                COVARIATES.tmp[, residCovars, drop=F], 'all adjusted design(voom-normalized)',
+                                                COVARIATES[, residCovars, drop=F], 'all adjusted design(voom-normalized)',
                                                 isKeyPlot=TRUE)
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T)
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale. = T, center = T)
 
@@
 RESIDUAL.GENE_EXPRESSION <- left_join(rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id"),
+                                      Ensemble2HGNC) %>%
+  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+VOOM.ADJUSTED.GENE_EXPRESSION$E <- left_join(rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, "ensembl_gene_id"),
                                              Ensemble2HGNC) %>%
@@
 }
-multiplot(plotlist=p, cols = 3)
+multiplot(plotlist=p, cols = 4)
 names(DE) = gsub('BrainRegion.Dx','',colnames(contrast))
@@
 # Expression
-expr = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'EnsembleIDs')
-write.table(expr, file='expression.tsv', sep = '\t', row.names=F, quote=F)
+write.table(RESIDUAL.GENE_EXPRESSION, file='expression.tsv', sep = '\t', row.names=F, quote=F)
 obj = File('expression.tsv', 
@@
 expr_obj = synStore(obj,
+                    used = ALL_USED_IDs, 
+                    activityName = activityName, 
+                    executed = thisFile, 
+                    activityDescription = activityDescription)
+
+# Raw Expression
+write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file='MSSMlogCPM.tsv', sep = '\t', row.names=F, quote=F)
+obj = File('MSSMlogCPM.tsv', 
+           name = 'Raw logCPM',
+           parentId = CODE$properties$id)
+rawcpm_obj = synStore(obj,
                     used = ALL_USED_IDs, 

### 2016-05-25 code/Rmd/MSSM_RNASeq_PlaqueMean.Rmd
Added markdown to process MSBB data with mean plaque adjustments

--- 2016-05-24 code/Rmd/MSSM_RNASeq.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_PlaqueMean.Rmd",
                                  parentId ="syn5570248",
-                                 entityName = 'MSBB RNASeq')
+                                 entityName = 'MSBB RNASeq (with adjustments for mean plaque)')
 ```
@@
 
-# library(doParallel)
-# library(foreach)
-# 
-# cl = makeCluster(12)
-# registerDoParallel(cl)
+library(doParallel)
+library(foreach)
+
+cl = makeCluster(12)
+registerDoParallel(cl)
 
@@
 
-thisFileName <- 'MSSM_RNASeq.Rmd'
+thisFileName <- 'MSSM_RNASeq_PlaqueMean.Rmd'
 
@@
 FactorCovariates <- c('RACE', 'SEX', 'BrainRegion.Dx', 'Batch', 'BB')
-ContCovariates <- c('PMI', 'AOD', 'RIN', 'RIN2')
+ContCovariates <- c('PMI', 'AOD', 'RIN', 'RIN2', 'PlaqueMean')
 
@@
 # Code
-CODE <- Folder(name = "MSBB RNASeq", parentId = parentId)
+CODE <- Folder(name = "MSBB RNASeq (with adjustments for mean plaque)", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-05-25 code/Rmd/MSSM_RNASeq_cdr.Rmd
Added markdown to process MSSM data based on cdr scores with a mixed effect model

--- 2016-05-24 code/Rmd/MSSM_RNASeq.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_cdr.Rmd",
                                  parentId ="syn5570248",
-                                 entityName = 'MSBB RNASeq')
+                                 entityName = 'MSBB RNASeq (based on CDR)')
 ```
@@
 
-thisFileName <- 'MSSM_RNASeq.Rmd'
+thisFileName <- 'MSSM_RNASeq_cdr.Rmd'
 
@@
 
-# Create diagnosis column (classified into 3 groups ND, MD, SD)
-METADATA$Dx = 'ND'
-METADATA$Dx[METADATA$CDR == 1 | METADATA$CDR == 2] = 'MD'
-METADATA$Dx[METADATA$CDR >2 ] = 'SD'
-
 # Create brain region x diagnosis column
-METADATA$BrainRegion.Dx = paste(METADATA$BrainRegion, METADATA$Dx, sep = '.')
+METADATA$BrainRegion.CDR = paste(METADATA$BrainRegion, METADATA$CDR, sep = '.')
 
 # Select covariates of interest
-FactorCovariates <- c('RACE', 'SEX', 'BrainRegion.Dx', 'Batch', 'BB')
-ContCovariates <- c('PMI', 'AOD', 'RIN', 'RIN2')
+FactorCovariates <- c('RACE', 'SEX', 'BrainRegion.CDR', 'Batch', 'BB')
+ContCovariates <- c('PMI', 'AOD', 'RIN', 'RIN2', 'PlaqueMean')
 
@@
 p = list()
-p[[1]] = ggplot(METADATA, aes(x = Dx, y = RIN, fill = BrainRegion)) + geom_boxplot()
+p[[1]] = ggplot(METADATA, aes(x = CDR, y = RIN, fill = BrainRegion)) + geom_boxplot()
 p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')
@@
 # AgeAtDeath
-p[[2]] = ggplot(METADATA, aes(x = Dx, y = as.numeric(AOD), fill = BrainRegion)) + geom_boxplot()
+p[[2]] = ggplot(METADATA, aes(x = CDR, y = as.numeric(AOD), fill = BrainRegion)) + geom_boxplot()
 p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + theme(legend.position = 'top')
@@
 # Mean Plaque
-p[[3]] = ggplot(METADATA, aes(x = Dx, y = as.numeric(PlaqueMean), fill = BrainRegion)) + geom_boxplot()
+p[[3]] = ggplot(METADATA, aes(x = CDR, y = as.numeric(PlaqueMean), fill = BrainRegion)) + geom_boxplot()
 p[[3]] = p[[3]] + ggtitle('PlaqueMean') + theme(legend.position = 'top')
@@
 # NP.1
-p[[4]] = ggplot(METADATA, aes(x = Dx, y = as.numeric(NP.1), fill = BrainRegion)) + geom_boxplot()
+p[[4]] = ggplot(METADATA, aes(x = CDR, y = as.numeric(NP.1), fill = BrainRegion)) + geom_boxplot()
 p[[4]] = p[[4]] + ggtitle('NP.1') + theme(legend.position = 'top')
@@
 # Sex
-vcd::mosaic(~ SEX + Dx + BrainRegion, data = METADATA)
+vcd::mosaic(~ SEX + CDR + BrainRegion, data = METADATA)
 ```
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=BrainRegion.Dx, shape=SEX, size=PMI))
+p <- p + geom_point(aes(color=BrainRegion.CDR, shape=SEX, size=PMI))
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
@@
 # Primary variable of interest
-postAdjustCovars = 'BrainRegion.Dx'
+postAdjustCovars = 'BrainRegion.CDR'
 
@@
 ### Residual calculation
-Calculate weighted residuals and add back "BrainRegion.Dx" to the residuals
+Calculate weighted residuals and add back "BrainRegion.CDR" to the residuals
 ```{r varsToAddBack}
 # Add variable of interest back to the residuals
-varsToAddIn = grep("BrainRegion.Dx", colnames(DM1$design), value = T)
+varsToAddIn = grep("BrainRegion.CDR", colnames(DM1$design), value = T)
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=factor(BrainRegion.Dx),size=as.numeric(PMI)))
+p <- p + geom_point(aes(color=factor(BrainRegion.CDR),size=as.numeric(PMI)))
 p <- p + theme_bw() + theme(legend.position="top") 
@@
 writeLines('Number of samples in each category are')
-tmp = as.data.frame(summary(COVARIATES$BrainRegion.Dx))
+tmp = as.data.frame(summary(COVARIATES$BrainRegion.CDR))
 colnames(tmp) = 'No. of samples'
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("BrainRegion.DxBM_10.ND-BrainRegion.DxBM_10.MD", 
-                                     "BrainRegion.DxBM_10.ND-BrainRegion.DxBM_10.SD", 
-                                     "BrainRegion.DxBM_10.MD-BrainRegion.DxBM_10.SD",
-                                     "BrainRegion.DxBM_22.ND-BrainRegion.DxBM_22.MD", 
-                                     "BrainRegion.DxBM_22.ND-BrainRegion.DxBM_22.SD", 
-                                     "BrainRegion.DxBM_22.MD-BrainRegion.DxBM_22.SD",
-                                     "BrainRegion.DxBM_36.ND-BrainRegion.DxBM_36.MD", 
-                                     "BrainRegion.DxBM_36.ND-BrainRegion.DxBM_36.SD", 
-                                     "BrainRegion.DxBM_36.MD-BrainRegion.DxBM_36.SD",
-                                     "BrainRegion.DxBM_44.ND-BrainRegion.DxBM_44.MD", 
-                                     "BrainRegion.DxBM_44.ND-BrainRegion.DxBM_44.SD", 
-                                     "BrainRegion.DxBM_44.MD-BrainRegion.DxBM_44.SD"), 
+contrast = makeContrasts(contrasts = sapply(unique(METADATA$BrainRegion), function(x){
+  apply(combn(paste0('BrainRegion.CDR',x,'.',unique(METADATA$CDR)),2), 2, paste, collapse = '-')
+}) %>%
+  as.character(), 
 levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
@@
   p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
-  p[[i]] <- p[[i]] + ggtitle(gsub('BrainRegion.Dx','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
+  p[[i]] <- p[[i]] + ggtitle(gsub('BrainRegion.CDR','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
   p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red')
 }
-multiplot(plotlist=p, cols = 4)
-names(DE) = gsub('BrainRegion.Dx','',colnames(contrast))
+# multiplot(plotlist=p, cols = 4)
+names(DE) = gsub('BrainRegion.CDR','',colnames(contrast))
 
@@
 
-kable(tmp)
+kable(tmp[,1:2])
 ```
@@
 # Code
-CODE <- Folder(name = "MSBB RNASeq", parentId = parentId)
+CODE <- Folder(name = "MSBB RNASeq (based on CDR)", parentId = parentId)
 CODE <- synStore(CODE)
@@
 # Expression
-write.table(RESIDUAL.GENE_EXPRESSION, file='expression.tsv', sep = '\t', row.names=F, quote=F)
-obj = File('expression.tsv', 
-           name = 'Voom Adjusted Weighted Residual Expression (BrainRegion.Dx added)', 
+write.table(RESIDUAL.GENE_EXPRESSION, file='MSSMexpression.tsv', sep = '\t', row.names=F, quote=F)
+obj = File('MSSMexpression.tsv', 
+           name = 'Voom Adjusted Weighted Residual Expression (BrainRegion.CDR added)', 
            parentId = CODE$properties$id)
@@
 covar = rownameToFirstColumn(COVARIATES, 'SampleId')
-write.table(covar, file='covariates.tsv', sep = '\t', row.names=F, quote=F)
-obj = File('covariates.tsv', 
+write.table(covar, file='MSSMcovariates.tsv', sep = '\t', row.names=F, quote=F)
+obj = File('MSSMcovariates.tsv', 
            name = 'Covariates', 
@@
 # Store differential expression results
-write.table(DE, file = 'diffExpression.tsv', sep = '\t', row.names=F, quote=F)
-dexp_obj = File('diffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
+write.table(DE, file = 'MSSMdiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+dexp_obj = File('MSSMdiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
 dexp_obj = synStore(dexp_obj, used = ALL_USED_IDs, activityName = activityName, 

### 2016-05-25 code/R/collateDiffExp.R
Minor changes

@@ -35,32 +35,60 @@ humanProteinCodingGenes = getBM(attributes=c("ensembl_gene_id","external_gene_na
                                 values=c('protein_coding'), 
                                 mart=ensemblHSapiens)
 
+#### Get all covariates
+covariates.id = c('ROSMAP' = 'syn6114444', 
+                  'MSBB' = 'syn5573110',
+                  'Mayo' = 'syn6088521')
+covariates = lapply(covariates.id, downloadFile)
+
+covariates$ROSMAP = covariates$ROSMAP %>%
+  dplyr::select(Sampleid, msex, cogdx) %>%
+  dplyr::rename(ID = Sampleid, Gender = msex, Status = cogdx) %>%
+  dplyr::mutate(Gender = factor(Gender, labels = c('1' = 'male','2'='female')),
+                Status = factor(Status, labels = c('1' = 'NCI', '2' = 'MCI', '3' = 'MCI.CI', 
+                                                   '4' = 'AD', '5' = 'AD.CI', '6' = 'OD')),
+                BrainRegion = 'DLPFC') %>%
+  dplyr::filter(Status %in% c('NCI', 'AD'))
+
+covariates$MSBB = covariates$MSBB %>%
+  dplyr::select(SampleId, SEX, BrainRegion.Dx) %>%
+  tidyr::separate(BrainRegion.Dx, c('BrainRegion', 'Dx'), sep = '\\.') %>%
+  dplyr::rename(ID = SampleId, Gender = SEX, Status = Dx) %>%
+  dplyr::mutate(Gender = factor(Gender, labels = c('M' = 'male','F'='female')),
+                Status = factor(Status)) %>%
+  dplyr::filter(Status %in% c('ND', 'SD'))
+
+covariates = rbindlist(covariate, use.names = T, fill = T, idcol = 'Study')
+
+#### Get raw expression values
+logcpm.id = c('ROSMAP' = 'syn6114447', 'MSBB' = 'syn6117295')
+logcpm = lapply(logcpm.id, downloadFile)
+
 # Get all differential expresison results from synapse
-diffExp.ids = c('ROSMAP DLPFC' = 'syn5608845',
-                'MSBB Frontal Pole' = 'syn6037382',
-                'MSBB Superior Temporal Gyrus'	= 'syn6037356',
-                'MSBB Para Hippocampal Gyrus' = 'syn6037276',
-                'MSBB Inferior Frontal Gyrus' = 'syn6037095',
-                'Mayo Cerebellum' = 'syn5609850',
[…]

### 2016-05-25 code/Rmd/Mayo_RNASeq.Rmd
Minor changes

@@
   dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+VOOM.ADJUSTED.GENE_EXPRESSION$E <- left_join(rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, "ensembl_gene_id"),
+                                             Ensemble2HGNC) %>%
+  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+
 ```
@@
 
-# Store covariates
+# Raw Expression
+write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file='MAYOlogCPM.tsv', sep = '\t', row.names=F, quote=F)
+obj = File('MAYOlogCPM.tsv', 
+           name = 'Raw logCPM',
+           parentId = CODE$properties$id)
+rawcpm_obj = synStore(obj,
+                    used = ALL_USED_IDs, 
+                    activityName = activityName, 
+                    executed = thisFile, 
+                    activityDescription = activityDescription)
+
+# Store residual gene expression
 write.table(RESIDUAL.GENE_EXPRESSION, file = 'expression.tsv', sep = '\t', row.names=F, quote=F)

### 2016-05-26 code/R/collateDiffExp.R
Added heatmap for logcpm data

@@ -24,27 +24,27 @@ synapseLogin()
 thisFileName <- 'collateDiffExp.R'
 
 # Github link
-thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
+thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='metaAnal')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/R/',thisFileName))
 
 # Get protein coding genes from biomaRt
-ensembl=useMart("ensembl")
-ensemblHSapiens = useDataset("hsapiens_gene_ensembl",mart=ensembl)
+# ensembl=useMart(biomart = "ensembl", host="www.ensembl.org", dataset = "hsapiens_gene_ensembl")
+ensembl=useMart(biomart = "ENSEMBL_MART_ENSEMBL", host="dec2015.archive.ensembl.org", dataset = "hsapiens_gene_ensembl") # Use old version if ensembl is down
 humanProteinCodingGenes = getBM(attributes=c("ensembl_gene_id","external_gene_name","description"), 
                                 filters='biotype', 
                                 values=c('protein_coding'), 
-                                mart=ensemblHSapiens)
+                                mart=ensembl)
 
 #### Get all covariates
 covariates.id = c('ROSMAP' = 'syn6114444', 
                   'MSBB' = 'syn5573110',
-                  'Mayo' = 'syn6088521')
+                  'MAYO' = 'syn6088521')
 covariates = lapply(covariates.id, downloadFile)
 
 covariates$ROSMAP = covariates$ROSMAP %>%
   dplyr::select(Sampleid, msex, cogdx) %>%
   dplyr::rename(ID = Sampleid, Gender = msex, Status = cogdx) %>%
-  dplyr::mutate(Gender = factor(Gender, labels = c('1' = 'male','2'='female')),
+  dplyr::mutate(Gender = factor(Gender, labels = c('1' = 'MALE','2'='FEMALE')),
                 Status = factor(Status, labels = c('1' = 'NCI', '2' = 'MCI', '3' = 'MCI.CI', 
                                                    '4' = 'AD', '5' = 'AD.CI', '6' = 'OD')),
                 BrainRegion = 'DLPFC') %>%
[…]

### 2016-05-27 code/Rmd/ROSMAP_RNASeq_Reprocessed.Rmd
Added both single and paired end reads to ROSMAP processing code

@@
 ```{r download.data, cache=TRUE}
-# Download expression data
-COUNT_ID <- 'syn6112813';
-ALL_USED_IDs <- COUNT_ID
-COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F)
+# Download reprocessed counts (paired end)
+COUNT_PE_ID <- 'syn6112813';
+ALL_USED_IDs <- COUNT_PE_ID
+COUNT_PE <- read.table(synGet(COUNT_PE_ID)@filePath, header=T, sep='\t', check.names = F)
+
+# Download reprocessed counts (single end)
+COUNT_SE_ID <- 'syn6117775';
+ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_SE_ID
+COUNT_SE <- read.table(synGet(COUNT_SE_ID)@filePath, header=T, sep='\t', check.names = F)
+
+COUNT = COUNT_PE + COUNT_SE[rownames(COUNT_PE), colnames(COUNT_PE)]
 

### 2016-05-28 code/Rmd/ROSMAP_RNASeq_SVA_Reprocessed.Rmd
Added markdown to process reprocessed RNASeq of ROSMAP data with SVA

--- 2016-05-27 code/Rmd/ROSMAP_RNASeq_Reprocessed.Rmd
@@
 ---
-title: "Covariate analysis of ROSMAP Reprocessed RNASeq"
+title: "Covariate analysis of ROSMAP Reprocessed RNASeq (with SVA)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed.Rmd",
+knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_SVA_Reprocessed.Rmd",
                                  parentId ="syn5570291",
-                                 entityName = 'ROSMAP RNASeq Reprocessed Counts (Based on cogdx)')
+                                 entityName = 'ROSMAP RNASeq Reprocessed Counts (Based on cogdx with SVA)')
 ```
@@
 
+library(sva)
+
+# library(doParallel)
+# library(foreach)
+# 
+# cl = makeCluster(14)
+# registerDoParallel(cl)
+
 options(xtable.type="html")
@@
 
-thisFileName <- 'ROSMAP_RNASeq_Reprocessed.Rmd'
+thisFileName <- 'ROSMAP_RNASeq_SVA_Reprocessed.Rmd'
 
@@
 ```
+
+### Estimate surrogate variables using SVA package
+```{r sva}
+MODEL0 = DM1$design[,-grep('cogdx',colnames(DM1$design))] # Get null model by removing variable of interest
+MODEL1 = DM1$design
+
+## Calculate the number of surrogate variables
+# Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
+SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
+NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
+SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
+SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]
+
+# Permute and calculate the permuted singular values
+PERM.SING.VALS <- foreach (i = 1:100,
+         .combine = rbind,
+         .packages = c('limma','CovariateAnalysis'),
+         .export = c('PROCESSED_COUNTS','MODEL1')) %dopar% {
+  PERM.COUNTS <- t(apply(PROCESSED_COUNTS$filteredExprMatrix, 1, sample, replace = FALSE))
+  # Estimate voom weights
+  VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)
+
+  # Fit linear model using new weights and new design
+  VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)
+
+  # Residuals after normalisation
+  RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)
+
+  PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
+  PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
+}
+num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])
+
+num.sv2 = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456)
+NUM.SV = min(num.sv1, num.sv2)
+
+SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, method = "irw", n.sv = NUM.SV, B = 20)$sv
+colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
+
+# Estimate voom weights
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=cbind(MODEL1, SURR.VAR), plot=F)
+
+# Fit linear model using new weights and new design
+VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
+
+# Residuals after normalisation
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
+```
+
 ### Residual calculation

### 2016-05-29 code/Rmd/Mayo_RNASeq_Reprocessed.Rmd
Added markdown to process reprocessed RNASeq from Mayo brain bank data

--- 2016-05-16 code/Rmd/Mayo_MayoBrainBankDickson_RNASeq.Rmd
@@
 ---
-title: "Covariate analysis of UFL-Mayo-ISB RNASeq (both cerebellum and temporal cortex) samples from MayoBrainBank_Dickson alone"
+title: "Covariate analysis of UFL-Mayo-ISB RNASeq reprocessed counts (both cerebellum and temporal cortex)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./Mayo_MayoBrainBankDickson_RNASeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed.Rmd",
                                  parentId = "syn5570325",
-                                 entityName = 'UFL_Mayo_ISB Mayo Brain Bank Dickson RNASeq')
+                                 entityName = 'UFL_Mayo_ISB RNASeq Reprocessed Counts')
 ```
@@
 
-library(doParallel)
-library(foreach)
-
-cl = makeCluster(14)
-registerDoParallel(cl)
-
-## Requires ggdendro
-devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")
-
-# source utility files from ../R/lib folder
-file.sources = list.files('../R/lib',pattern="*.R", full.names=T)
-tmp = sapply(file.sources,source,.GlobalEnv)
+# library(doParallel)
+# library(foreach)
+# 
+# cl = makeCluster(14)
+# registerDoParallel(cl)
 
@@
 
-thisFileName <- 'Mayo_MayoBrainBankDickson_RNASeq.Rmd'
+thisFileName <- 'Mayo_RNASeq_Reprocessed.Rmd'
 
@@
 ```{r download.data, cache=TRUE}
-# Download expression data
-COUNT_TC_ID <- 'syn4650257';
-ALL_USED_IDs <- COUNT_TC_ID
-COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F)
+# Get reprocessed paired end sample counts for temporal cortex
+COUNT_TC_PE_ID <- 'syn6114408';
+ALL_USED_IDs <- COUNT_TC_PE_ID
+COUNT_TC_PE <- read.table(synGet(COUNT_TC_PE_ID)@filePath, header=T, sep='\t', check.names = F)
 
-# Download expression data
-COUNT_CE_ID <- 'syn5201012';
+# Get reprocessed single end sample counts for temporal cortex
+COUNT_TC_SE_ID <- 'syn6114409';
+ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_TC_SE_ID
+COUNT_TC_SE <- read.table(synGet(COUNT_TC_SE_ID)@filePath, header=T, sep='\t', check.names = F)
+
+COUNT_TC = COUNT_TC_PE + COUNT_TC_SE[rownames(COUNT_TC_PE), colnames(COUNT_TC_PE)]
+colnames(COUNT_TC) = gsub('.snap','',colnames(COUNT_TC))
+colnames(COUNT_TC) = gsub('11423_R0212989', '11423_TCX',colnames(COUNT_TC))
+colnames(COUNT_TC) = gsub("11427_R0211812", "11427_TCX", colnames(COUNT_TC))
+
+# Get reprocessed sample counts for cerebellum
+COUNT_CE_ID <- 'syn6124945';
 ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_CE_ID
@@
 
-COUNT = plyr::join_all(list(COUNT_TC, COUNT_CE)) %>%
-  as.data.frame
+COUNT = cbind(COUNT_TC, COUNT_CE[rownames(COUNT_TC),])
 
@@
 METADATA <- METADATA %>%
-  dplyr::filter(!(ID %in% c("742_CER", "11386_CER","11311_CER","1923_CER","11477_CER", "11344_CER", 
-                            "1923_TCX", "11423_TCX", "1940_TCX", "7095_TCX", "11408_TCX"))) %>%
-  filter(Source == 'MayoBrainBank_Dickson')
+  filter(!is.na(Source)) %>%
+  dplyr::filter(!(ID %in% c("11386_CER","1923_CER", "11311_CER", "11477_CER", "11408_TCX", "11294_TCX", "11396_TCX")))
 ```
@@
 indToRetain = intersect(METADATA$ID, colnames(COUNT))
+indRemoved = setdiff(colnames(COUNT), METADATA$ID)
 
-COUNT <- COUNT[, c('ensembl_id', intersect(colnames(COUNT), indToRetain))]
+COUNT <- COUNT[, indToRetain]
 
 rownames(METADATA) = METADATA$ID
-METADATA = METADATA[rownames(METADATA) %in% indToRetain,]
-
-COUNT = COUNT[, c('ensembl_id', indToRetain)]
 METADATA = METADATA[indToRetain,]
-
-rownames(COUNT) = COUNT$ensembl_id
-COUNT$ensembl_id = NULL
 ```
-`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD portal
+`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project. Following samples were removed `r indRemoved`
 
@@
 ```{r covariates.clustering}
-FactorCovariates <- c("Donor_ID", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
+FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
 ContCovariates <- c("RIN", "RIN2", "AgeAtDeath")
@@
 # Find principal components of expression to plot
-PC <- prcomp(VOOM.GENE_EXPRESSION$E)
+PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale. = T, center = T)
 
@@
                            groupLabels = c('Cluster',colnames(COVARIATES.tmp)))
-dev.off()
 
@@
 writeLines(tmp)
-tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[tree == 3], collapse = ','))
-tmp = base::gsub('_','\\\\_',tmp)
-writeLines(tmp)
 ```
+
+```{r temp, include=FALSE}
+dev.off()
+gc()
+```
+
 ### Significant Covariates
@@
   t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
-RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'ensembl_gene_id') 
 ```
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1)], scale.=T)
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1)], scale.=T, center = T)
 
@@
 
-tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[tree == 0], collapse = ','))
+tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[clust == 0], collapse = ','))
 tmp = base::gsub('_','\\\\_',tmp)
 writeLines(tmp)
-tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[tree == 3], collapse = ','))
+tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[clust == 3], collapse = ','))
 tmp = base::gsub('_','\\\\_',tmp)
 writeLines(tmp)
-tmp = paste('Samples in cluster 4 are', paste(sampleTrees$labels[tree == 4], collapse = ','))
+tmp = paste('Samples in cluster 4 are', paste(sampleTrees$labels[clust == 4], collapse = ','))
 tmp = base::gsub('_','\\\\_',tmp)
 writeLines(tmp)
+```
+
+```{r temp1, include=FALSE}
+gc()
+dev.off()
 ```
@@
 
+# Get background gene ids
+backgroundGenes = data.frame(Gene.ID = rownames(COUNT), ID = rownames(COUNT)) %>%
+  tidyr::separate(ID, c('ensembl_gene_id','position'), sep = '\\.')
+
 # Query biomart
 Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
-                       filters = "ensembl_gene_id", values = RESIDUAL.GENE_EXPRESSION$ensembl_gene_id,
+                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                        mart = mart)
-RESIDUAL.GENE_EXPRESSION <- left_join(RESIDUAL.GENE_EXPRESSION, Ensemble2HGNC) %>%
-  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+
+# Add gene names
+RESIDUAL.GENE_EXPRESSION <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'Gene.ID') %>%
+  left_join(backgroundGenes) %>% left_join(Ensemble2HGNC) %>%
+  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+
+VOOM.ADJUSTED.GENE_EXPRESSION$E <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, 'Gene.ID') %>%
+  left_join(backgroundGenes) %>% left_join(Ensemble2HGNC) %>%
+  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
 ```
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=16, fig.width=25}
+```{r diffExp, fig.height=25, fig.width=25}
 # Fit contrast
@@
   "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.AD",
+  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.Pathologic_Aging",
   "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.PSP",
+  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.AD",
   "BrainRegion.DiagnosisCER.PSP-BrainRegion.DiagnosisCER.AD",
+  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.PSP",
   "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.AD",
+  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.Pathologic_Aging",
   "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.PSP",
-  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD"), 
+  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.AD",
+  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD",
+  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.PSP"), 
                          levels = colnames(FIT$coefficients))
@@
 # Get differnetial expression
-DE = list()
-p = list()
-for (i in 1:6){
-  DE[[i]] = topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
-  DE[[i]] = rownameToFirstColumn(DE[[i]], 'ensembl_gene_id') %>%
-    left_join(Ensemble2HGNC)
-  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
-  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red') 
-  p[[i]] <- p[[i]] + ggtitle(gsub('BrainRegion.Diagnosis','',colnames(contrast)[i]))
-}
-multiplot(plotlist=p, cols = 3)
+DE = lapply(1:12, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
+  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
+}, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
 names(DE) = gsub('BrainRegion.Diagnosis','',colnames(contrast))
 
-DE = plyr::ldply(DE, .id = 'Comparison')
+DE = plyr::ldply(DE, .id = 'Comparison') %>%
+  dplyr::mutate(Comparison1 = Comparison) %>%
+  tidyr::separate(Comparison1, c('from.state','to.state'), sep = '-') %>%
+  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
+  tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
+  tidyr::unite(Name, reference, against, sep = '-')
+
+p <- ggplot(DE, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
+p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
+p <- p + geom_vline(xintercept = log2(1.2), color = 'red')
+p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
+p <- p + facet_grid(BrainRegion~.+Name)
+p
+
 tmp = DE %>% 
@@
   group_by(Comparison) %>%
-  dplyr::summarize(nDiffExpGenes = length(unique(hgnc_symbol)))
+  dplyr::summarize(nDiffExpGenes = length(unique(genes)))
 kable(tmp)
@@
 # Code
-CODE <- Folder(name = 'UFL_Mayo_ISB Mayo Brain Bank Dickson RNASeq', parentId = parentId)
+CODE <- Folder(name = 'UFL_Mayo_ISB RNASeq Reprocessed Counts', parentId = parentId)
 CODE <- synStore(CODE)
@@
 COVARIATES = rownameToFirstColumn(COVARIATES, 'ID')
-write.table(COVARIATES, file = 'covariates.tsv', sep = '\t', row.names=F, quote=F)
-COV_OBJ = File('covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+write.table(COVARIATES, file = 'MAYOcovariates.tsv', sep = '\t', row.names=F, quote=F)
+COV_OBJ = File('MAYOcovariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
 COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 
-# Store covariates
-write.table(RESIDUAL.GENE_EXPRESSION, file = 'expression.tsv', sep = '\t', row.names=F, quote=F)
-EXP_OBJ = File('expression.tsv', name = 'Voom Adjusted Residual Expression (BrainRegion.Diagnosis added)', parentId = CODE$properties$id)
+# Raw Expression
+write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file='MAYOlogCPM.tsv', sep = '\t', row.names=F, quote=F)
+RAWCPM_OBJ = File('MAYOlogCPM.tsv', 
+                  name = 'Raw logCPM',
+                  parentId = CODE$properties$id)
+RAWCPM_OBJ = synStore(RAWCPM_OBJ,
+                      used = ALL_USED_IDs, 
+                      activityName = activityName, 
+                      executed = thisFile, 
+                      activityDescription = activityDescription)
+
+# Store residual gene expression
+write.table(RESIDUAL.GENE_EXPRESSION, file = 'MAYOexpression.tsv', sep = '\t', row.names=F, quote=F)
+EXP_OBJ = File('MAYOexpression.tsv', name = 'Voom Adjusted Residual Expression (BrainRegion.Diagnosis added)', parentId = CODE$properties$id)
 EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 # Store differential expression results
-write.table(DE, file = 'diffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('diffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
+write.table(DE, file = 'MAYOdiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('MAYOdiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
 DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 |  Covariates               |  `r COV_OBJ$properties$id`      |
+|  Raw counts (logCPM)      |  `r RAWCPM_OBJ$properties$id`   |
 |  Residual Expression      |  `r EXP_OBJ$properties$id`      |

### 2016-05-29 code/Rmd/Mayo_RNASeq_Reprocessed.Rmd
Minor bug fix

@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1)], scale.=T, center = T)
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
 
 # Tree based clustering
-sampleTrees = flashClust::hclust(dist(t(RESIDUAL.GENE_EXPRESSION[,-(1)])), method = "average")
+sampleTrees = flashClust::hclust(dist(t(RESIDUAL.GENE_EXPRESSION)), method = "average")
 

### 2016-05-29 code/Rmd/Mayo_RNASeq_SVA_Reprocessed.Rmd
Added markdown to process reprocessed RNASeq data from Mayo brain bank with two-step sva adjustment

--- 2016-05-29 code/Rmd/Mayo_RNASeq_Reprocessed.Rmd
@@
 ---
-title: "Covariate analysis of UFL-Mayo-ISB RNASeq reprocessed counts (both cerebellum and temporal cortex)"
+title: "Covariate analysis of UFL-Mayo-ISB RNASeq reprocessed counts (both cerebellum and temporal cortex) with SVA adjustments"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed.Rmd",
+knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_SVA_Reprocessed.Rmd",
                                  parentId = "syn5570325",
-                                 entityName = 'UFL_Mayo_ISB RNASeq Reprocessed Counts')
+                                 entityName = 'UFL_Mayo_ISB RNASeq Reprocessed Counts (with SVA adjustments)')
 ```
@@
 
-thisFileName <- 'Mayo_RNASeq_Reprocessed.Rmd'
+thisFileName <- 'Mayo_RNASeq_SVA_Reprocessed.Rmd'
 
@@
 VOOM.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=NULL, plot=T)
-
-# Initial design matrix with covariates under consideration
-DM = getDesignMatrix(COVARIATES, Intercept = F)
-DESIGN = getDesignMat2Fact(DM$design,FactorCovariates)
-colnames(DESIGN) = substr(colnames(DESIGN),1,50)
 ```
@@
 `r tmp`
-
 ### Sanity Check
@@
 ```
+### Estimate surrogate variables using SVA package
+```{r sva}
+MODEL0 = getDesignMatrix(COVARIATES[,setdiff(postAdjustCovars, "BrainRegion.Diagnosis"),drop=F],Intercept = F)$design # Get null model by removing variable of interest
+MODEL1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
+
+## Calculate the number of surrogate variables
+# Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
+SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
+NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
+SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
+SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]
+
+# Permute and calculate the permuted singular values
+PERM.SING.VALS <- foreach (i = 1:100,
+         .combine = rbind,
+         .packages = c('limma','CovariateAnalysis'),
+         .export = c('EXPR','MODEL1')) %dopar% {
+
+           # Permute expression counts
+           PERM.COUNTS <- t(apply(EXPR$filteredExprMatrix, 1, sample, replace = FALSE))
+
+           # Estimate voom weights
+           VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)
+
+           # Fit linear model using new weights and new design
+           VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)
+
+           # Residuals after normalisation
+           RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)
+
+           PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
+           PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
+        }
+num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])
+
+num.sv2 = num.sv(EXPR$filteredExprMatrix$COUNT, MODEL1, method = 'be', seed = 123456)
+NUM.SV = min(num.sv1, num.sv2)
+
+SURR.VAR = svaseq(EXPR$filteredExprMatrix$counts, MODEL1, MODEL0, n.sv = NUM.SV, method = "two-step", B = 30)$sv
+colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
+
+# Estimate voom weights
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=cbind(DM1$design, SURR.VAR), plot=F)
+
+# Calculate correlation values of random effects
+correlation = parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block = COVARIATES$Donor_ID)
+
+# Re-estimate voom weights
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix$counts, design=cbind(DM1$design, SURR.VAR), plot=F, 
+                                     block = COVARIATES$Donor_ID, correlation = correlation$cor)
+
+# Fit a mixed effect linear model
+FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
+
+# Residuals after normalisation
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
+```
+
+### Residual calculation
 Adding "BrainRegion.Diagnosis" back to the residuals
@@
 DE = lapply(1:12, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
-  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
+  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
+    rownameToFirstColumn('Gene.ID')
 }, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
@@
   group_by(Comparison) %>%
-  dplyr::summarize(nDiffExpGenes = length(unique(genes)))
+  dplyr::summarize(nDiffExpGenes = length(unique(Gene.ID)))
 kable(tmp)
@@
 # Code
-CODE <- Folder(name = 'UFL_Mayo_ISB RNASeq Reprocessed Counts', parentId = parentId)
+CODE <- Folder(name = 'UFL_Mayo_ISB RNASeq Reprocessed Counts (with SVA adjustments)', parentId = parentId)
 CODE <- synStore(CODE)

### 2016-05-29 code/Rmd/Mayo_RNASeq_SVA_Reprocessed.Rmd
Minor bug fix

@@
 
-num.sv2 = num.sv(EXPR$filteredExprMatrix$COUNT, MODEL1, method = 'be', seed = 123456)
+num.sv2 = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456)
 NUM.SV = min(num.sv1, num.sv2)
 
-SURR.VAR = svaseq(EXPR$filteredExprMatrix$counts, MODEL1, MODEL0, n.sv = NUM.SV, method = "two-step", B = 30)$sv
+SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, n.sv = NUM.SV, method = "two-step", B = 30)$sv
 colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])

### 2016-05-30 code/Rmd/Mayo_RNASeq_SVA_Reprocessed.Rmd
Minor changes

@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=25, fig.width=25}
+```{r diffExp, fig.height=15, fig.width=25}
 # Fit contrast

### 2016-05-30 code/Rmd/ROSMAP_RNASeq_SVA_Reprocessed.Rmd
Changed SVA method to two-step in ROSMAP reprocessed rnaseq analysis

@@
 
-SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, method = "irw", n.sv = NUM.SV, B = 20)$sv
+SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, method = "two-step", n.sv = NUM.SV, B = 30)$sv
 colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
@@
 # Code
-CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts (Based on cogdx)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts (Based on cogdx with SVA)", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-05-31 code/Rmd/MSSM_RNASeq_SVA_cdr.Rmd
Added two-step SVA based processing for MSSM RNASeq (all four brain regions)

--- 2016-05-25 code/Rmd/MSSM_RNASeq_cdr.Rmd
@@
 ---
-title: "Covariates analysis of MSBB RNASeq data"
+title: "Covariates analysis of MSBB RNASeq (with SVA)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_cdr.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_SVA_cdr.Rmd",
                                  parentId ="syn5570248",
-                                 entityName = 'MSBB RNASeq (based on CDR)')
+                                 entityName = 'MSBB RNASeq (based on CDR with SVA)')
 ```
@@
 
-# library(doParallel)
-# library(foreach)
-# 
-# cl = makeCluster(12)
-# registerDoParallel(cl)
+library(doParallel)
+library(foreach)
 
-## Requires ggdendro
-devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")
-
-# Source required R files
-source('../R/lib/parallelDuplicateCorrelation.R')
+cl = makeCluster(12)
+registerDoParallel(cl)
 
@@
 
-thisFileName <- 'MSSM_RNASeq_cdr.Rmd'
+thisFileName <- 'MSSM_RNASeq_SVA_cdr.Rmd'
 
@@
 FactorCovariates <- c('RACE', 'SEX', 'BrainRegion.CDR', 'Batch', 'BB')
-ContCovariates <- c('PMI', 'AOD', 'RIN', 'RIN2', 'PlaqueMean')
+ContCovariates <- c('PMI', 'AOD', 'RIN', 'RIN2')
 
@@
 # Primary variable of interest
-postAdjustCovars = 'BrainRegion.CDR'
+postAdjustCovars = c('BrainRegion.CDR', 'RIN', 'PMI', 'AOD', 'Batch') # Based on a separate analysis
 
@@
 
-covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, 'BB'))]
-
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
+# covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, 'BB'))]
+# 
+# postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
 
@@
 ```
+### Estimate surrogate variables using SVA package
+```{r sva}
+MODEL0 = DM1$design[,-grep('BrainRegion.CDR',colnames(DM1$design))] # Get null model by removing variable of interest
+MODEL1 = DM1$design
+
+## Calculate the number of surrogate variables
+NUM.SV = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456)
+
+SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, method = "two-step", n.sv = NUM.SV, B = 30)$sv
+colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
+
+# Estimate correlation of random effects
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=cbind(MODEL1, SURR.VAR), plot=F)
+correlation <- parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block=COVARIATES$BB)
+
+# Re-calculate voom weights with correlation of random effects
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=cbind(MODEL1, SURR.VAR), plot=F,
+                                     block=COVARIATES$BB, correlation = correlation$cor)
+
+# Re-calculate linear fit using new weights and new design
+VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
+
+# Residuals after normalisation
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
+```
+
 ### Residual calculation
@@
 ```
+```{r temp1, include=FALSE}
+dev.off()
+gc()
+```
+
 ```{r get.gene.symbols, cache=TRUE, eval=FALSE}
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=10, fig.width=18}
+```{r diffExp, fig.height=10, fig.width=25}
 # Fit contrast
@@
 # Get differnetial expression
-DE = list()
-p = list()
-for (i in 1:dim(contrast)[2]){
-  DE[[i]] = topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
-  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id")
-  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
-  p[[i]] <- p[[i]] + ggtitle(gsub('BrainRegion.CDR','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
-  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red')
-}
-# multiplot(plotlist=p, cols = 4)
+DE = lapply(1:dim(contrast)[2], function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
+  topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
+    rownameToFirstColumn("ensembl_gene_id") %>%
+    left_join(Ensemble2HGNC)
+}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC)
 names(DE) = gsub('BrainRegion.CDR','',colnames(contrast))
 
-DE = DE %>% ldply(.id = 'Comparison')
+DE = DE %>% 
+  ldply(.id = 'Comparison') %>%
+  dplyr::mutate(Comparison1 = Comparison) %>%
+  tidyr::separate(Comparison1, c('from.state','to.state'), sep = '-') %>%
+  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
+  tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
+  tidyr::unite(Name, reference, against, sep = '-')
+
+p <- ggplot(DE, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
+p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
+p <- p + geom_vline(xintercept = log2(1.2), color = 'red')
+p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
+p <- p + facet_grid(BrainRegion~.+Name)
+p
 
@@
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(Comparison, ensembl_gene_id) %>%
-  group_by(Comparison) %>%
-  dplyr::summarise(nDiffExp = length(unique(ensembl_gene_id)),
-                   DiffExpGenes = paste(unique(ensembl_gene_id), collapse = ','))
+  dplyr::select(BrainRegion, Name, ensembl_gene_id) %>%
+  group_by(BrainRegion, Name) %>%
+  dplyr::summarise(nDiffExp = length(unique(ensembl_gene_id))) %>%
+  tidyr::spread(Name, nDiffExp)
 
-kable(tmp[,1:2])
+kable(tmp)
 ```
@@
 # Code
-CODE <- Folder(name = "MSBB RNASeq (based on CDR)", parentId = parentId)
+CODE <- Folder(name = "MSBB RNASeq (based on CDR with SVA)", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-05-31 code/Rmd/ROSMAP_RNASeq_SVA_Reprocessed.Rmd
Added two-step SVA based processing for MSSM RNASeq (all four brain regions)

@@
                                  parentId ="syn5570291",
-                                 entityName = 'ROSMAP RNASeq Reprocessed Counts (Based on cogdx with SVA)')
+                                 entityName = 'ROSMAP RNASeq Reprocessed Counts (Based on cogdx with SVA IRW)')
 ```
@@
 
-SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, method = "two-step", n.sv = NUM.SV, B = 30)$sv
+SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, method = "irw", n.sv = NUM.SV, B = 30)$sv
 colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
@@
 # Code
-CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts (Based on cogdx with SVA)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts (Based on cogdx with SVA IRW)", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-05-31 code/Rmd/MSSM_RNASeq_cdr_Reprocessed.Rmd
Added markdown to process reprocessed rnaseq data from MSSM batch 1,2 and 3

--- 2016-05-25 code/Rmd/MSSM_RNASeq_cdr.Rmd
@@
 ---
-title: "Covariates analysis of MSBB RNASeq data"
+title: "Covariates analysis of MSBB reprocessed RNASeq data"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_cdr.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_cdr_Reprocessed.Rmd",
                                  parentId ="syn5570248",
-                                 entityName = 'MSBB RNASeq (based on CDR)')
+                                 entityName = 'MSBB Reprocessed RNASeq (based on CDR)')
 ```
@@
 library(xlsx)
+library(ggdendro)
 
-# library(doParallel)
-# library(foreach)
-# 
-# cl = makeCluster(12)
-# registerDoParallel(cl)
+library(doParallel)
+library(foreach)
 
-## Requires ggdendro
+cl = makeCluster(2)
+registerDoParallel(cl)
+
 devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")
-
-# Source required R files
-source('../R/lib/parallelDuplicateCorrelation.R')
 
@@
 
-thisFileName <- 'MSSM_RNASeq_cdr.Rmd'
+thisFileName <- 'MSSM_RNASeq_cdr_Reprocessed.Rmd'
 
@@
 # Download expression data
-COUNT_ID <- 'syn5898488'
+COUNT_ID <- 'syn6114414'
 ALL_USED_IDs <- COUNT_ID
-COUNT_OBJ <- synGet(COUNT_ID)
-COUNT <- fread(COUNT_OBJ@filePath, data.table = F)
-rownames(COUNT) = COUNT$Geneid
+COUNT <- read.table(synGet(COUNT_ID)@filePath, quote = '', header = T, sep = '\t')
 
@@
 METADATA_ID <- 'syn5898489'
-METADATA_OBJ <- synGet(METADATA_ID)
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
-METADATA <- read.table(METADATA_OBJ@filePath,sep='\t',header=T, row.names=1)
+METADATA_TECHNICAL <- read.table(synGet(METADATA_ID)@filePath,sep='\t',header=T, row.names=1) %>%
+  rownameToFirstColumn('OldBarcode')
 
 # Get clinical metadata
-METADATA_TECHNICAL_ID <- 'syn5475828'
-METADATA_TECHNICAL_OBJ <- synGet(METADATA_TECHNICAL_ID)
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECHNICAL_ID
-METADATA_TECHNICAL <- read.xlsx(METADATA_TECHNICAL_OBJ@filePath, sheetIndex = 1)
+METADATA_CLINICAL_ID <- 'syn5475828'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CLINICAL_ID
+METADATA_CLINICAL <- read.xlsx(synGet(METADATA_CLINICAL_ID)@filePath, sheetIndex = 1)
 
 # Manually fix 4 sample IDs
-METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1132), "BM36.Fnl"] = 397
-METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1117), "BM22.Fnl"] = 144
-METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1320), "BM22.Fnl"] = 159
-METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1475), "BM22.Fnl"] = 186
+METADATA_CLINICAL[which(METADATA_CLINICAL$BB == 1132), "BM36.Fnl"] = 397
+METADATA_CLINICAL[which(METADATA_CLINICAL$BB == 1117), "BM22.Fnl"] = 144
+METADATA_CLINICAL[which(METADATA_CLINICAL$BB == 1320), "BM22.Fnl"] = 159
+METADATA_CLINICAL[which(METADATA_CLINICAL$BB == 1475), "BM22.Fnl"] = 186
 
-METADATA = left_join(METADATA %>%
+METADATA = left_join(METADATA_TECHNICAL %>%
                        dplyr::mutate(ID = gsub('hB_RNA_','',NewBarcode),
                                      ID = as.numeric(ID)),
-                     METADATA_TECHNICAL %>%
+                     METADATA_CLINICAL %>%
                        dplyr::select(BB, BM.44.Barcode, BM36.Fnl, BM22.Fnl, BM10.Fnl, PMI, RACE, 
@@
                        dplyr::mutate(ID = as.numeric(ID)))
-rownames(METADATA) = METADATA$NewBarcode
+rownames(METADATA) = METADATA$OldBarcode
 
@@
 ind <- intersect(rownames(METADATA), colnames(COUNT))
-ind <- setdiff(ind, 'hB_RNA_10622')
+# ind <- setdiff(ind, 'hB_RNA_10622')
 METADATA <- METADATA[ind,]
@@
 COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
-rownames(COVARIATES) <- METADATA$NewBarcode
+rownames(COVARIATES) <- METADATA$OldBarcode
 
@@
 p = list()
-p[[1]] = ggplot(METADATA, aes(x = CDR, y = RIN, fill = BrainRegion)) + geom_boxplot()
+p[[1]] = ggplot(METADATA, aes(x = as.factor(CDR), y = as.numeric(RIN), fill = as.factor(BrainRegion))) + geom_boxplot()
 p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')
@@
 # AgeAtDeath
-p[[2]] = ggplot(METADATA, aes(x = CDR, y = as.numeric(AOD), fill = BrainRegion)) + geom_boxplot()
+p[[2]] = ggplot(METADATA, aes(x = as.factor(CDR), y = as.numeric(AOD), fill = as.factor(BrainRegion))) + geom_boxplot()
 p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + theme(legend.position = 'top')
@@
 # Mean Plaque
-p[[3]] = ggplot(METADATA, aes(x = CDR, y = as.numeric(PlaqueMean), fill = BrainRegion)) + geom_boxplot()
+p[[3]] = ggplot(METADATA, aes(x = as.factor(CDR), y = as.numeric(PlaqueMean), fill = as.factor(BrainRegion))) + geom_boxplot()
 p[[3]] = p[[3]] + ggtitle('PlaqueMean') + theme(legend.position = 'top')
@@
 # NP.1
-p[[4]] = ggplot(METADATA, aes(x = CDR, y = as.numeric(NP.1), fill = BrainRegion)) + geom_boxplot()
+p[[4]] = ggplot(METADATA, aes(x = as.factor(CDR), y = as.numeric(NP.1), fill = as.factor(BrainRegion))) + geom_boxplot()
 p[[4]] = p[[4]] + ggtitle('NP.1') + theme(legend.position = 'top')
@@
                            groupLabels = colnames(COVARIATES.tmp[,FactorCovariates]))
+```
+```{r temp, include=F}
 dev.off()
+gc()
 ```
+
 ### Significant Covariates
@@
 
+backgroundGenes = data.frame(Gene.ID = rownames(RESIDUAL.GENE_EXPRESSION),
+                             ID = rownames(RESIDUAL.GENE_EXPRESSION)) %>%
+  tidyr::separate(ID, c('ensembl_gene_id','position'), sep = '\\.')
+
 # Query biomart
 Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
-                       filters = "ensembl_gene_id", values = rownames(RESIDUAL.GENE_EXPRESSION),
+                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                        mart = mart)
-RESIDUAL.GENE_EXPRESSION <- left_join(rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id"),
-                                      Ensemble2HGNC) %>%
-  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
-VOOM.ADJUSTED.GENE_EXPRESSION$E <- left_join(rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, "ensembl_gene_id"),
-                                             Ensemble2HGNC) %>%
-  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+RESIDUAL.GENE_EXPRESSION <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "Gene.ID") %>%
+  left_join(backgroundGenes) %>%
+  left_join(Ensemble2HGNC) %>% 
+  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+VOOM.ADJUSTED.GENE_EXPRESSION$E <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, "Gene.ID") %>%
+  left_join(backgroundGenes) %>%
+  left_join(Ensemble2HGNC) %>% 
+  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
 ```
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=10, fig.width=18}
+```{r diffExp, fig.height=12, fig.width=25}
 # Fit contrast
@@
 # Get differnetial expression
-DE = list()
-p = list()
-for (i in 1:dim(contrast)[2]){
-  DE[[i]] = topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
-  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id")
-  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
-  p[[i]] <- p[[i]] + ggtitle(gsub('BrainRegion.CDR','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
-  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red')
-}
-# multiplot(plotlist=p, cols = 4)
+DE = lapply(1:dim(contrast)[2], function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC, backgroundGenes){
+  topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
+    rownameToFirstColumn("Gene.ID") %>%
+    left_join(backgroundGenes) %>%
+    left_join(Ensemble2HGNC)
+}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC, backgroundGenes)
 names(DE) = gsub('BrainRegion.CDR','',colnames(contrast))
 
-DE = DE %>% ldply(.id = 'Comparison')
+DE = DE %>% 
+  ldply(.id = 'Comparison') %>%
+  dplyr::mutate(Comparison1 = Comparison) %>%
+  tidyr::separate(Comparison1, c('from.state','to.state'), sep = '-') %>%
+  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
+  tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
+  tidyr::unite(Name, reference, against, sep = '-')
+
+p <- ggplot(DE, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
+p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
+p <- p + geom_vline(xintercept = log2(1.2), color = 'red')
+p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
+p <- p + facet_grid(BrainRegion~.+Name)
+p
 
@@
 # Code
-CODE <- Folder(name = "MSBB RNASeq (based on CDR)", parentId = parentId)
+CODE <- Folder(name = "MSBB Reprocessed RNASeq (based on CDR)", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-05-31 code/Rmd/MSSM_RNASeq_cdr_Reprocessed.Rmd
Minor fix

@@
 library(dplyr)
-library(stringr)
 
@@
 ```
-```{r get.gene.symbols, cache=TRUE, eval=FALSE}
+```{r get.gene.symbols, cache=TRUE, eval=TRUE}
 # Define biomart object
-mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl", host = )
+mart <- useMart('ENSEMBL_MART_ENSEMBL', dataset = "hsapiens_gene_ensembl", host = 'dec2015.archive.ensembl.org')
 

### 2016-05-31 code/Rmd/MSSM_RNASeq_SVA_cdr.Rmd
Minor fix

@@
 ---
-title: "Covariates analysis of MSBB RNASeq (with SVA)"
+title: "Covariates analysis of MSBB RNASeq (with two-step SVA)"
 author: "Thanneer Perumal"
@@
 registerDoParallel(cl)
+
+source('../R/lib/num.sv')
 
@@
 ## Calculate the number of surrogate variables
-NUM.SV = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456)
+NUM.SV = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456, LINPACK = TRUE, B = 10)
 
@@
 
-```{r get.gene.symbols, cache=TRUE, eval=FALSE}
+```{r get.gene.symbols, cache=TRUE, eval=TRUE}
 # Define biomart object

### 2016-06-01 code/Rmd/ROSMAP_RNASeq_Reprocessed.Rmd
Minor modifications in the reprocessed RNASeq processing pipeline

@@
 ```{r covariates.correlation, fig.width=10, fig.height=10}
-COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.1)
+COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
 ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.5, row.width=0.1)
@@
                            groupLabels = colnames(COVARIATES.tmp[,FactorCovariates]))
+```
+```{r temp, include = F}
 dev.off()
+gc()
 ```
@@
                           groupLabels = colnames(COVARIATES.tmp[,FactorCovariates]))
+```
+```{r temp1, include=F}
 dev.off()
+gc()
 ```
@@
 ```{r get.gene.symbols, cache=TRUE, eval=T}
-row.id = data.frame(GeneID = rownames(RESIDUAL.GENE_EXPRESSION),
-                    ensembl_gene_id = rownames(RESIDUAL.GENE_EXPRESSION)) %>%
-  separate(ensembl_gene_id, c('ensembl_gene_id','position'), sep = '\\.')
+backgroundGenes = data.frame(Gene.ID = rownames(COUNT),
+                             ID = rownames(COUNT)) %>%
+  separate(ID, c('ensembl_gene_id','position'), sep = '\\.')
 
@@
 Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
-                       filters = "ensembl_gene_id", values = row.id$ensembl_gene_id,
+                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                        mart = mart)
@@
 RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
-  rownameToFirstColumn('GeneID') %>%
-  left_join(row.id) %>%
+  rownameToFirstColumn('Gene.ID') %>%
+  left_join(backgroundGenes) %>%
   left_join(Ensemble2HGNC) %>%
-  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
 
 VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
-  rownameToFirstColumn('GeneID') %>%
-  right_join(row.id) %>%
+  rownameToFirstColumn('Gene.ID') %>%
+  left_join(backgroundGenes) %>%
   left_join(Ensemble2HGNC) %>%
-  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
 ```
@@
 # Get differnetial expression
-DE = list()
-p = list()
-for (i in 1:3){
-  DE[[i]] = topTable(FIT.CONTR, coef=i, number = dim(COUNT)[1])
-  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id") 
-}
+DE = lapply(1:3, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, backgroundGenes, Ensemble2HGNC){
+  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
+    rownameToFirstColumn('Gene.ID') %>%
+    left_join(backgroundGenes) %>%
+    left_join(Ensemble2HGNC) %>%
+    dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, logFC, AveExpr, t, P.Value, adj.P.Val, B)
+}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, backgroundGenes, Ensemble2HGNC)
 names(DE) = colnames(contrast)
 
-DE = DE %>% plyr::ldply(.id = 'Comparison') %>%
-  tidyr::separate(ensembl_gene_id, into=c("ensemble_id","position")) %>%
-  dplyr::rename(ensembl_gene_id = ensemble_id) %>%
-  left_join(Ensemble2HGNC)
+DE = DE %>% 
+  plyr::ldply(.id = 'Comparison')
 
@@
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(Comparison, hgnc_symbol) %>%
+  dplyr::select(Gene.ID, Comparison) %>%
   group_by(Comparison) %>%
-  dplyr::summarise(nDiffExp = length(unique(hgnc_symbol)), 
-            DiffExpGenes = paste(unique(hgnc_symbol), collapse = ','))
+  dplyr::summarise(FDR_0_05 = length(unique(Gene.ID)))
 
-kable(tmp)
+tmp1 = DE %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
+  dplyr::select(Gene.ID, Comparison) %>%
+  group_by(Comparison) %>%
+  dplyr::summarise(FDR_0_05_FC_1_2 = length(unique(Gene.ID)))
+
+kable(full_join(tmp,tmp1))
 
@@
 p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
-p = p + geom_vline(xintercept = -log2(1.2), color = 'red') + facet_grid(.~Comparison, scales = 'fixed')
+p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
+p = p + facet_grid(.~Comparison, scales = 'fixed')
 p
@@
 
+# Store design matrix
+write.table(DM1$design, file = 'ROSMAPDesign.tsv', sep = '\t', row.names=F, quote=F)
+DM_OBJ = File('ROSMAPDesign.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                   executed = thisFile, activityDescription = activityDescription)
+
 # Store residual gene expression
@@
 |  Raw counts (log CPM)     |  `r COUNT_OBJ$properties$id`     |
+|  Design Matrix            |  `r DM_OBJ$properties$id`        |
 |  Residual Expression      |  `r EXP_OBJ$properties$id`       |

### 2016-06-01 code/Rmd/MSSM_RNASeq_cdr_Reprocessed.Rmd
Major update on the reprocessed RNASeq processing code for MSSM

@@
 
-cl = makeCluster(2)
+cl = makeCluster(14)
 registerDoParallel(cl)
@@
 # Download expression data
-COUNT_ID <- 'syn6114414'
-ALL_USED_IDs <- COUNT_ID
-COUNT <- read.table(synGet(COUNT_ID)@filePath, quote = '', header = T, sep = '\t')
+COUNT_B1_ID <- 'syn6114414'
+ALL_USED_IDs <- COUNT_B1_ID
+COUNT_B1 <- read.table(synGet(COUNT_B1_ID)@filePath, quote = '', header = T, sep = '\t')
+
+COUNT_B2_3_ID <- 'syn6125203'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_B2_3_ID
+COUNT_B2_3 <- read.table(synGet(COUNT_B2_3_ID)@filePath, quote = '', header = T, sep = '\t')
+
+# Download expression data
+COUNT_B4_ID <- 'syn6130073'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_B4_ID
+COUNT_B4 <- read.table(synGet(COUNT_B4_ID)@filePath, quote = '', header = T, sep = '\t')
+
+# Remove samples in Batch 2&3 if they present in batch 4
+COUNT_B2_3 = COUNT_B2_3[,setdiff(colnames(COUNT_B2_3), colnames(COUNT_B4))]
+
+COUNT = cbind(COUNT_B1, COUNT_B2_3[rownames(COUNT_B1),], COUNT_B4[rownames(COUNT_B1),])
+
+# Get sample ids
+SampleID = data.frame(SampleID = colnames(COUNT),
+                 ID = colnames(COUNT)) %>% 
+  tidyr::separate(ID, c('A','B','ID'), sep = '_') 
+SampleID$ID[SampleID$ID == 'HISEQ'] = SampleID$B[SampleID$ID == 'HISEQ']
+
+SampleID = SampleID %>%
+  dplyr::select(SampleID, ID) %>%
+  dplyr::mutate(ID = as.numeric(ID))
 
@@
                        filter(!is.na(ID)) %>%
-                       dplyr::mutate(ID = as.numeric(ID)))
-rownames(METADATA) = METADATA$OldBarcode
+                       dplyr::mutate(ID = as.numeric(ID))) %>%
+  left_join(SampleID) %>%
+  filter(!is.na(SampleID)) %>%
+  unique
+rownames(METADATA) = METADATA$SampleID
 
 # Filter data
-ind <- intersect(rownames(METADATA), colnames(COUNT))
+ind <- intersect(METADATA$SampleID, colnames(COUNT))
+indRemoved <- setdiff(colnames(COUNT), METADATA$SampleID)
 # ind <- setdiff(ind, 'hB_RNA_10622')
@@
 COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
-rownames(COVARIATES) <- METADATA$OldBarcode
+rownames(COVARIATES) <- METADATA$SampleID
 
@@
 ```
+### Quality Control
+```{r check.sex}
+# Query biomart for Y chromosome genes
+ensembl = useMart('ensembl', dataset = 'hsapiens_gene_ensembl', host = 'www.ensembl.org')
+EnsembleInY <- getBM(attributes = c("ensembl_gene_id"),
+                       filters = "chromosome_name", 
+                       values = 'Y',
+                       mart = ensembl)
 
+YChromGenes = data.frame(Gene.ID = rownames(COUNT),
+                             ID = rownames(COUNT)) %>%
+  tidyr::separate(ID, c('ensembl_gene_id','position'), sep = '\\.') %>%
+  inner_join(EnsembleInY)
+
+tmp1 = t(COUNT[YChromGenes$Gene.ID,]) %>%
+  rowSums %>% rownameToFirstColumn('SampleID') %>%
+  dplyr::rename(CPM.Sum = DF) %>%
+  left_join(COVARIATES %>% 
+              rownameToFirstColumn('SampleID'))
+
+ggplot(tmp1, aes(x = SEX, y= log2(CPM.Sum))) + geom_boxplot()
+
+EnsembleInX <- getBM(attributes = c("ensembl_gene_id"),
+                       filters = "chromosome_name", 
+                       values = 'X',
+                       mart = ensembl)
+
+XChromGenes = data.frame(Gene.ID = rownames(COUNT),
+                             ID = rownames(COUNT)) %>%
+  tidyr::separate(ID, c('ensembl_gene_id','position'), sep = '\\.') %>%
+  inner_join(EnsembleInX)
+
+tmp1 = t(COUNT[XChromGenes$Gene.ID,]) %>%
+  rowSums %>% rownameToFirstColumn('SampleID') %>%
+  dplyr::rename(CPM.Sum = DF) %>%
+  left_join(COVARIATES %>% 
+              rownameToFirstColumn('SampleID'))
+
+ggplot(tmp1, aes(x = SEX, y= log2(CPM.Sum))) + geom_boxplot()
+```
 ### CPM Normalisation
@@
 # Find correlation between PC's of gene expression with covariates
-preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, COVARIATES,'NULL design(voom-normalized)', isKeyPlot=TRUE, MIN_PVE_PCT_PC = 1)
+preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
+                                                 COVARIATES,
+                                                 'NULL design(voom-normalized)', 
+                                                 isKeyPlot=TRUE, 
+                                                 MIN_PVE_PCT_PC = 1)
 ```
@@
 
+# Design Matrix
+write.table(DM1$design, file='MSSMDesignMat.tsv', sep = '\t', row.names=F, quote=F)
+obj = File('MSSMDesignMat.tsv', 
+           name = 'Design Matrix',
+           parentId = CODE$properties$id)
+dm_obj = synStore(obj,
+                    used = ALL_USED_IDs, 
+                    activityName = activityName, 
+                    executed = thisFile, 
+                    activityDescription = activityDescription)
+
 # Raw Expression
@@
 |  Covariates               |  `r covar_obj$properties$id`    |
+|  Raw counts (log CPM)     |  `r rawcpm_obj$properties$id`   |
+|  Design Matrix            |  `r dm_obj$properties$id`       |
 |  Residual Expression      |  `r expr_obj$properties$id`     |

### 2016-06-02 code/Rmd/MSSM_RNASeq_Reprocessed.Rmd
Added markdown to process reprocessed RNASeq from MSBB (grouped CDR to ND,MD,SD)

--- 2016-06-01 code/Rmd/MSSM_RNASeq_cdr_Reprocessed.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_cdr_Reprocessed.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed.Rmd",
                                  parentId ="syn5570248",
-                                 entityName = 'MSBB Reprocessed RNASeq (based on CDR)')
+                                 entityName = 'MSBB Reprocessed RNASeq')
 ```
@@
 
-thisFileName <- 'MSSM_RNASeq_cdr_Reprocessed.Rmd'
+thisFileName <- 'MSSM_RNASeq_Reprocessed.Rmd'
 
@@
 ```{r covariates.clustering, cache=TRUE}
+# Add a diagnosis
+METADATA$Diagnosis = 'MD'
+METADATA$Diagnosis[METADATA$CDR < 1] = 'ND'
+METADATA$Diagnosis[METADATA$CDR > 2] = 'SD'
+
 # Max out age at 90
@@
 # Create brain region x diagnosis column
-METADATA$BrainRegion.CDR = paste(METADATA$BrainRegion, METADATA$CDR, sep = '.')
+METADATA$BrainRegion.Diagnosis = paste(METADATA$BrainRegion, METADATA$Diagnosis, sep = '.')
 
 # Select covariates of interest
-FactorCovariates <- c('RACE', 'SEX', 'BrainRegion.CDR', 'Batch', 'BB')
+FactorCovariates <- c('RACE', 'SEX', 'BrainRegion.Diagnosis', 'Batch', 'BB')
 ContCovariates <- c('PMI', 'AOD', 'RIN', 'RIN2')
@@
 # AgeAtDeath
-p[[2]] = ggplot(METADATA, aes(x = as.factor(CDR), y = as.numeric(AOD), fill = as.factor(BrainRegion))) + geom_boxplot()
+p[[2]] = ggplot(METADATA, aes(x = as.factor(Diagnosis), y = as.numeric(AOD), fill = as.factor(BrainRegion))) + geom_boxplot()
 p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + theme(legend.position = 'top')
@@
 # Mean Plaque
-p[[3]] = ggplot(METADATA, aes(x = as.factor(CDR), y = as.numeric(PlaqueMean), fill = as.factor(BrainRegion))) + geom_boxplot()
+p[[3]] = ggplot(METADATA, aes(x = as.factor(Diagnosis), y = as.numeric(PlaqueMean), fill = as.factor(BrainRegion))) + geom_boxplot()
 p[[3]] = p[[3]] + ggtitle('PlaqueMean') + theme(legend.position = 'top')
@@
 # NP.1
-p[[4]] = ggplot(METADATA, aes(x = as.factor(CDR), y = as.numeric(NP.1), fill = as.factor(BrainRegion))) + geom_boxplot()
+p[[4]] = ggplot(METADATA, aes(x = as.factor(Diagnosis), y = as.numeric(NP.1), fill = as.factor(BrainRegion))) + geom_boxplot()
 p[[4]] = p[[4]] + ggtitle('NP.1') + theme(legend.position = 'top')
@@
 # Sex
-vcd::mosaic(~ SEX + CDR + BrainRegion, data = METADATA)
+vcd::mosaic(~ SEX + Diagnosis + BrainRegion, data = METADATA)
 ```
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=BrainRegion.CDR, shape=SEX, size=PMI))
+p <- p + geom_point(aes(color=BrainRegion.Diagnosis, shape=SEX, size=PMI))
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
@@
 # Primary variable of interest
-postAdjustCovars = 'BrainRegion.CDR'
+postAdjustCovars = 'BrainRegion.Diagnosis'
 
@@
 ### Residual calculation
-Calculate weighted residuals and add back "BrainRegion.CDR" to the residuals
+Calculate weighted residuals and add back "BrainRegion.Diagnosis" to the residuals
 ```{r varsToAddBack}
 # Add variable of interest back to the residuals
-varsToAddIn = grep("BrainRegion.CDR", colnames(DM1$design), value = T)
+varsToAddIn = grep("BrainRegion.Diagnosis", colnames(DM1$design), value = T)
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=factor(BrainRegion.CDR),size=as.numeric(PMI)))
+p <- p + geom_point(aes(color=factor(BrainRegion.Diagnosis),size=as.numeric(PMI)))
 p <- p + theme_bw() + theme(legend.position="top") 
@@
 writeLines('Number of samples in each category are')
-tmp = as.data.frame(summary(COVARIATES$BrainRegion.CDR))
-colnames(tmp) = 'No. of samples'
+tmp = summary(COVARIATES$BrainRegion.Diagnosis) %>%
+  rownameToFirstColumn('BrainRegion.Diagnosis') %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion','Diagnosis'), sep = '\\.') %>%
+  tidyr::spread(Diagnosis, DF)
 kable(tmp)
@@
 contrast = makeContrasts(contrasts = sapply(unique(METADATA$BrainRegion), function(x){
-  apply(combn(paste0('BrainRegion.CDR',x,'.',unique(METADATA$CDR)),2), 2, paste, collapse = '-')
+  apply(combn(paste0('BrainRegion.Diagnosis',x,'.',unique(METADATA$Diagnosis)),2), 2, paste, collapse = '-')
 }) %>%
@@
 }, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC, backgroundGenes)
-names(DE) = gsub('BrainRegion.CDR','',colnames(contrast))
+names(DE) = gsub('BrainRegion.Diagnosis','',colnames(contrast))
 
@@
 
+writeLines('Differentially expressed at an FDR 0.05')
 tmp = DE %>%
   dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(BrainRegion, Name, ensembl_gene_id) %>%
+  group_by(BrainRegion, Name) %>%
+  dplyr::summarise(nDiffExp = length(unique(ensembl_gene_id))) %>%
+  tidyr::spread(Name, nDiffExp)
+
+kable(tmp)
+
+writeLines('Differentially expressed at an FDR 0.05 and FC greater than 1.2')
+tmp = DE %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
   dplyr::select(BrainRegion, Name, ensembl_gene_id) %>%
@@
 # Code
-CODE <- Folder(name = "MSBB Reprocessed RNASeq (based on CDR)", parentId = parentId)
+CODE <- Folder(name = "MSBB Reprocessed RNASeq", parentId = parentId)
 CODE <- synStore(CODE)
@@
 obj = File('MSSMexpression.tsv', 
-           name = 'Voom Adjusted Weighted Residual Expression (BrainRegion.CDR added)', 
+           name = 'Voom Adjusted Weighted Residual Expression (BrainRegion.Diagnosis added)', 
            parentId = CODE$properties$id)

### 2016-06-02 code/Rmd/Mayo_RNASeq_Reprocessed.Rmd
Minor modifications

@@
 
-# library(doParallel)
-# library(foreach)
-# 
-# cl = makeCluster(14)
-# registerDoParallel(cl)
+library(doParallel)
+library(foreach)
+
+cl = makeCluster(2)
+registerDoParallel(cl)
 
@@
 writeLines('Number of samples in each category are')
-tmp = data.frame(summary(COVARIATES$BrainRegion.Diagnosis))
-colnames(tmp) = "No. of samples"
+tmp = summary(COVARIATES$BrainRegion.Diagnosis) %>%
+  rownameToFirstColumn('SampleName') %>%
+  tidyr::separate(SampleName, c('BrainRegion','Diagnosis'), sep = '\\.') %>%
+  tidyr::spread(Diagnosis, DF)
 kable(tmp)
@@
 
+# Design Matrix
+write.table(DM1$design, file='MAYOdesign.tsv', sep = '\t', row.names=F, quote=F)
+DM_OBJ = File('MAYOdesign.tsv', 
+                  name = 'Design Matrix',
+                  parentId = CODE$properties$id)
+DM_OBJ = synStore(DM_OBJ,
+                      used = ALL_USED_IDs, 
+                      activityName = activityName, 
+                      executed = thisFile, 
+                      activityDescription = activityDescription)
+
 # Store residual gene expression
@@
 |  Raw counts (logCPM)      |  `r RAWCPM_OBJ$properties$id`   |
+|  Design Matrix            |  `r DM_OBJ$properties$id`       |
 |  Residual Expression      |  `r EXP_OBJ$properties$id`      |

### 2016-06-02 code/Rmd/MSSM_RNASeq_Reprocessed.Rmd
Minor modifications

@@
 # Sex
-vcd::mosaic(~ SEX + Diagnosis + BrainRegion, data = METADATA)
+vcd::mosaic( Diagnosis + SEX ~ BrainRegion, data = METADATA)
 ```
@@
 
-tmp1 = t(COUNT[YChromGenes$Gene.ID,]) %>%
-  rowSums %>% rownameToFirstColumn('SampleID') %>%
-  dplyr::rename(CPM.Sum = DF) %>%
+PROCESSED_COUNTS = COUNT
+PROCESSED_COUNTS = PROCESSED_COUNTS[rowSums(COUNT) != 0,]
+
+VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS, design=NULL, plot=F)
+
+tmp1 = apply(VOOM.GENE_EXPRESSION$E[intersect(rownames(VOOM.GENE_EXPRESSION$E), YChromGenes$Gene.ID),], 2, mean) %>%
+  rownameToFirstColumn('SampleID') %>%
+  dplyr::rename(mean.log2.CPM = DF) %>%
   left_join(COVARIATES %>% 
@@
 
-ggplot(tmp1, aes(x = SEX, y= log2(CPM.Sum))) + geom_boxplot()
+pl = list()
+pl[[1]] = ggplot(tmp1, aes(x = SEX, y= mean.log2.CPM)) + geom_boxplot() + ggtitle('Mean expression of Y chromosome genes')
 
@@
 
-tmp1 = t(COUNT[XChromGenes$Gene.ID,]) %>%
-  rowSums %>% rownameToFirstColumn('SampleID') %>%
-  dplyr::rename(CPM.Sum = DF) %>%
+tmp1 = apply(VOOM.GENE_EXPRESSION$E[intersect(rownames(VOOM.GENE_EXPRESSION$E), XChromGenes$Gene.ID),], 2, mean) %>%
+  rownameToFirstColumn('SampleID') %>%
+  dplyr::rename(mean.log2.CPM = DF) %>%
   left_join(COVARIATES %>% 
@@
 
-ggplot(tmp1, aes(x = SEX, y= log2(CPM.Sum))) + geom_boxplot()
+pl[[2]] = ggplot(tmp1, aes(x = SEX, y= mean.log2.CPM)) + geom_boxplot() + ggtitle('Mean expression of X chromosome genes')
+
+multiplot(plotlist = pl, cols = 2)
 ```

### 2016-06-02 code/Rmd/MSSM_RNASeq_SVA_Reprocessed.Rmd
Added IRW SVA processing pipeline for MSSM reprocessed rnaseq

--- 2016-06-02 code/Rmd/MSSM_RNASeq_Reprocessed.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_SVA_Reprocessed.Rmd",
                                  parentId ="syn5570248",
-                                 entityName = 'MSBB Reprocessed RNASeq')
+                                 entityName = 'MSBB Reprocessed RNASeq (with IRW SVA)')
 ```
@@
 
-thisFileName <- 'MSSM_RNASeq_Reprocessed.Rmd'
+thisFileName <- 'MSSM_RNASeq_SVA_Reprocessed.Rmd'
 
@@
 p = list()
-p[[1]] = ggplot(METADATA, aes(x = as.factor(CDR), y = as.numeric(RIN), fill = as.factor(BrainRegion))) + geom_boxplot()
+p[[1]] = ggplot(METADATA, aes(x = as.factor(Diagnosis), y = as.numeric(RIN), fill = as.factor(BrainRegion))) + geom_boxplot()
 p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')
@@
 ```
+### Estimate surrogate variables using SVA package
+```{r sva}
+MODEL0 = DM1$design[,-grep('BrainRegion.Diagnosis',colnames(DM1$design))] # Get null model by removing variable of interest
+MODEL1 = DM1$design
+
+## Calculate the number of surrogate variables
+NUM.SV = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456, LINPACK = TRUE, B = 10)
+
+SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, method = "irw", n.sv = NUM.SV, B = 30)$sv
+colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
+
+DM1$design = cbind(MODEL1, SURR.VAR)
+
+# Estimate correlation of random effects
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)
+correlation <- parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block=COVARIATES$BB)
+
+# Re-calculate voom weights with correlation of random effects
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F,
+                                     block=COVARIATES$BB, correlation = correlation$cor)
+
+# Re-calculate linear fit using new weights and new design
+VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
+
+# Residuals after normalisation
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
+```
+
 ### Residual calculation
@@
 # Code
-CODE <- Folder(name = "MSBB Reprocessed RNASeq", parentId = parentId)
+CODE <- Folder(name = "MSBB Reprocessed RNASeq (with IRW SVA)", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-06-06 code/Rmd/MetaAnalysis.Rmd
minor fix

@@
                                  parentId ="syn5569102",
-                                 entityName = 'MetaAnalysis of Control vs AD Data')
+                                 entityName = 'MetaAnalysis of Control vs AD from all three brain banks')
 ```
@@
 # Load required libraries
-library(synapseClient)
+library(CovariateAnalysis)
+library(data.table)
+library(tidyr)
+library(plyr)
+library(dplyr)
+library(stringr)
+
 library(RColorBrewer)
 library(ggplot2)
-library(data.table)
-library(plyr)
-library(dplyr)
-library(psych)
-library(tidyr)
-library(Matrix)
+library(ComplexHeatmap)
 
-library(limma)
-library(sva)
 library(RankProd)
-
-library(knitr)
-library(rGithubClient) ## Needs the dev branch
-
-## Requires ggdendro
-devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")
+library(githubr)
 
 synapseLogin()
-
-# source utility files from ../R/lib folder
-source('../R/lib/rownameToFirstColumn.R')
-# file.sources = list.files('../R/lib/rownfile.sources',pattern="row*.R", full.names = T)
-# tmp = sapply(file.sources,source,.GlobalEnv)
 
@@
 ```{r fxns}
-# Set ggplot theme
-my.theme = theme(title = element_text(size = 24), 
-                 axis.text = element_text(size = 20), 
-                 axis.text.x = element_text(angle = 90, hjust=1, vjust = 0.5))
-
-# Utility function to download files from synapse
-download.file <- function(id){ tmp = fread(synGet(as.character(id))@filePath, data.table=F, header=T);  }
-
 # Utility function for combining p-values using Fishers inverse chi-square method
@@
 
+### Get covariates/sample metadata from synapse
+```{r covariates}
+covariates.id = c('ROSMAP' = 'syn6114444',
+                  'MSBB' = 'syn6131421',
+                  'MAYO' = 'syn6128267')
+
+covariates = lapply(covariates.id, downloadFile)
+
+covariates$ROSMAP = covariates$ROSMAP %>%
+  dplyr::select(Sampleid, msex, cogdx) %>%
+  dplyr::rename(ID = Sampleid, Gender = msex, Status = cogdx) %>%
+  dplyr::mutate(Gender = factor(Gender, labels = c('1' = 'MALE','2'='FEMALE')),
+                Status = factor(Status, labels = c('1' = 'NCI', '2' = 'MCI', '3' = 'MCI.CI', 
+                                                   '4' = 'AD', '5' = 'AD.CI', '6' = 'OD')),
+                BrainRegion = 'DLPFC') %>%
+  dplyr::filter(Status %in% c('NCI', 'AD'))
+
+covariates$MSBB = covariates$MSBB %>%
+  dplyr::select(SampleId, SEX, BrainRegion.Diagnosis) %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Dx'), sep = '\\.') %>%
+  dplyr::rename(ID = SampleId, Gender = SEX, Status = Dx) %>%
+  dplyr::mutate(Gender = factor(Gender, labels = c('M' = 'MALE','F'='FEMALE')),
+                Status = factor(Status),
+                BrainRegion = factor(BrainRegion, labels = c('BM_10' = 'FP', 'BM_22' = 'STG', 
+                                                             'BM_36' = 'PHG', 'BM_44' = 'IFG'))) %>%
+  dplyr::filter(Status %in% c('ND', 'SD'))
+
+covariates$MAYO = covariates$MAYO %>%
+  dplyr::select(ID, Gender, BrainRegion.Diagnosis) %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Dx'), sep = '\\.') %>%
+  dplyr::rename(Status = Dx) %>%
+  dplyr::mutate(Gender = factor(Gender, labels = c('M' = 'MALE','F'='FEMALE')),
+                Status = factor(Status)) %>%
+  dplyr::filter(Status %in% c('Control', 'AD'))
+
+covariates = rbindlist(covariates, use.names = T, fill = T, idcol = 'Study')
+
+```
+
+### Get logCPM counts from synapse
+```{r counts}
+logcpm.id = c('ROSMAP' = 'syn6114447' ,
+              'MSBB' = 'syn6131419',
+              'MAYO' = 'syn6128269')
+
+logcpm = lapply(logcpm.id, downloadFile) %>%
+  lapply(function(x){
+    x = dplyr::select(x, -hgnc_symbol)
+    }) %>%
+  join_all(type = 'full') %>%
+  dplyr::select(Gene.ID, ensembl_gene_id, one_of(covariates$ID)) 
+logcpm[is.na(logcpm)] = dplyr::select(logcpm, -Gene.ID, -ensembl_gene_id) %>%
+  min
+```
+
 ### Download differential expression data
 ```{r download.data}
-# Get all differential expression files and their synIds
-DEXP_ID = c(ROSMAP.PFC = 'syn5608845',
-            MSSB = 'syn5609009',
-            MAYO.CER = 'syn5609850',
-            MAYO.TC = 'syn5609813',
-            MAYO.CER.AR = 'syn5607457',
-            MAYO.TC.AR = 'syn5607475')
+diffexp.id = c('ROSMAP' = 'syn6114453',
+               'MSBB' = 'syn6131423',
+               'MAYO' = 'syn6128273')
 
-# Download differential expression data
-DEXP = lapply(DEXP_ID, download.file)
+diffexp = lapply(diffexp.id, downloadFile)
 
-# Extract subset of differential expression data
-DEXP$ROSMAP.PFC = DEXP$ROSMAP.PFC %>% 
-  dplyr::filter(Comparison == "cogdx1-cogdx4") %>%
-  dplyr::select(hgnc_symbol, P.Value, logFC)
+diffexp$ROSMAP = diffexp$ROSMAP %>%
+  filter(Comparison == 'cogdx1-cogdx4') %>%
+  dplyr::select(Gene.ID, logFC, P.Value, adj.P.Val) %>%
+  dplyr::mutate(Tissue = 'DLPFC', Contrast = 'NCI-AD')
 
-DEXP$MSSB.FP = DEXP$MSSB %>% 
-  dplyr::filter(Comparison == "BM_10.Control-BM_10.AD") %>%
-  dplyr::select(hgnc_symbol, P.Value, logFC)
+diffexp$MSBB = diffexp$MSBB %>%
+  dplyr::filter(Name == 'ND-SD') %>%
+  dplyr::select(Gene.ID, logFC, P.Value, adj.P.Val, BrainRegion, Name) %>%
+  dplyr::mutate(BrainRegion = factor(BrainRegion, labels = c('BM_10' = 'FP', 'BM_22' = 'STG', 
+                                                             'BM_36' = 'PHG', 'BM_44' = 'IFG'))) %>%
+  dplyr::rename(Tissue = BrainRegion, Contrast = Name)
 
-DEXP$MSSB.STG = DEXP$MSSB %>% 
-  dplyr::filter(Comparison == "BM_22.Control-BM_22.AD") %>%
-  dplyr::select(hgnc_symbol, P.Value, logFC)
+diffexp$MAYO = diffexp$MAYO %>%
+  dplyr::filter(Name == 'Control-AD') %>%
+  dplyr::select(genes, logFC, P.Value, adj.P.Val, BrainRegion, Name) %>%
+  dplyr::rename(Tissue = BrainRegion, Gene.ID = genes, Contrast = Name)
 
-DEXP$MSSB.PHG = DEXP$MSSB %>% 
-  dplyr::filter(Comparison == "BM_36.Control-BM_36.AD") %>%
-  dplyr::select(hgnc_symbol, P.Value, logFC)
+diffexp = rbindlist(diffexp, use.names = T, fill = T, idcol = 'Study') %>%
+  tidyr::separate(Gene.ID, c('ensembl_gene_id','position'),sep = '\\.')
 
-DEXP$MSSB = NULL
+diffexp$DE = 'NONE'
+diffexp$DE[diffexp$adj.P.Val <= 0.05 & diffexp$logFC >= log2(1.2)] = 'UP'
+diffexp$DE[diffexp$adj.P.Val <= 0.05 & diffexp$logFC <= -log2(1.2)] = 'DOWN'
 
-DEXP$MAYO.CER = DEXP$MAYO.CER %>% 
-  dplyr::select(hgnc_symbol, P.Value, logFC)
+pl = list()
+for (study in c('ROSMAP', 'MSBB', 'MAYO')){
+  p = filter(diffexp, Study == study) %>%
+    ggplot(aes(x = logFC, y = -log10(adj.P.Val), color = DE)) 
+  p = p + scale_colour_manual(values = c("green","grey", "red"))
+  p = p + geom_point() + theme(legend.position = 'None')
+  p = p + facet_grid(Study~.+Tissue)
+  pl[[study]] = p
+}
+multiplot(plotlist = pl, cols = 1)
+```
 
-DEXP$MAYO.TC = DEXP$MAYO.TC %>% 
-  dplyr::select(hgnc_symbol, P.Value, logFC)
+### Get hgnc_symbol mapping from biomaRt
+```{r biomart}
+# Define biomart object
+mart <- useMart('ENSEMBL_MART_ENSEMBL', dataset = "hsapiens_gene_ensembl", host = 'dec2015.archive.ensembl.org')
 
-DEXP$MAYO.CER.AR = DEXP$MAYO.CER.AR %>% 
-  dplyr::select(hgnc_symbol, P.Value, logFC)
+# Query biomart
+Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
+                       filters = "ensembl_gene_id", 
+                       values = unique(diffexp$ensembl_gene_id),
+                       mart = mart) %>%
+  data.table()
 
-DEXP$MAYO.TC.AR = DEXP$MAYO.TC.AR %>% 
-  dplyr::select(hgnc_symbol, P.Value, logFC)
+# Add hgnc_symbols to the logcpm matrix
+logcpm = logcpm %>%
+  left_join(Ensemble2HGNC)
+
+# Add hgnc_symbols to the diffexp matrix
+diffexp = diffexp %>%
+  left_join(Ensemble2HGNC)
+```
+
+### Pathway enrichment
+```{r pathway.enrichment}
+# Differentially expressed in ROSMAP
+## Up regulated
+tmp = filter(diffexp, adj.P.Val <= 0.05, logFC >0, Tissue == 'DLPFC') %>%
+  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
+  filter(hgnc_symbol != '')
+### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6x3h
+
+## Down regulated
+tmp = filter(diffexp, adj.P.Val <= 0.05, logFC<0, Tissue == 'DLPFC') %>%
+  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
+  filter(hgnc_symbol != '')
+### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6x4e
+
+# Differentially expressed in MSSM FP
+## Up regulated
+tmp = filter(diffexp, adj.P.Val <= 0.05, logFC >0, Tissue == 'FP') %>%
+  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
+  filter(hgnc_symbol != '')
+### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6xcm
+
+## Down regulated
+tmp = filter(diffexp, adj.P.Val <= 0.05, logFC < 0, Tissue == 'FP') %>%
+  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
+  filter(hgnc_symbol != '')
+
+# Differentially expressed in MSSM STG
+## Up regulated
+tmp = filter(diffexp, adj.P.Val <= 0.05, logFC > 0, Tissue == 'STG') %>%
+  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
+  filter(hgnc_symbol != '')
+
+## Down regulated
+tmp = filter(diffexp, adj.P.Val <= 0.05, logFC < 0, Tissue == 'STG') %>%
+  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
+  filter(hgnc_symbol != '')
+
+# Differentially expressed in MSSM PHG
+## Up regulated
+tmp = filter(diffexp, adj.P.Val <= 0.05, logFC > 0, Tissue == 'PHG') %>%
+  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
+  filter(hgnc_symbol != '')
+### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6xco
+
+## Down regulated
+tmp = filter(diffexp, adj.P.Val <= 0.05, logFC < 0, Tissue == 'PHG') %>%
+  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
+  filter(hgnc_symbol != '')
+### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6xge
+
+# Differentially expressed in MAYO CER
+## Up regulated
+tmp = filter(diffexp, adj.P.Val <= 0.05, logFC > 0, Tissue == 'CER') %>%
+  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
+  filter(hgnc_symbol != '')
+### Enrichr results are here:
+
+## Down regulated
+tmp = filter(diffexp, adj.P.Val <= 0.05, logFC < 0, Tissue == 'PHG') %>%
+  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
+  filter(hgnc_symbol != '')
+### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6xge
+```
+
+### Meta Analysis
+```{r metaanalysis}
+P.Value = diffexp %>% 
+  # filter(Tissue %in% c('DLPFC', 'PHG', 'TCX')) %>%
+  dplyr::select(ensembl_gene_id, P.Value, Tissue, Contrast) %>%
+  tidyr::unite(Tissue.Contrast, Tissue, Contrast, sep = '_') %>%
+  tidyr::spread(Tissue.Contrast, P.Value)
+P.Value[is.na(P.Value)] = 1
+
+min.pval = dplyr::select(P.Value, -ensembl_gene_id) %>%
+  data.matrix %>% rowMin %>%
+  p.adjust(method = 'BH') %>%
+  data.frame(ensembl_gene_id = P.Value$ensembl_gene_id,
+             min.pval = .)
+
+max.pval = dplyr::select(P.Value, -ensembl_gene_id) %>%
+  data.matrix %>% rowMax %>%
+  p.adjust(method = 'BH') %>%
+  data.frame(ensembl_gene_id = P.Value$ensembl_gene_id,
+             max.pval = .)
+
+fb.pval = dplyr::select(P.Value, -ensembl_gene_id) %>%
+  data.matrix %>% fishersPvalue %>%
+  p.adjust(method = 'BH') %>%
+  data.frame(ensembl_gene_id = P.Value$ensembl_gene_id,
+             fb.pval = .)
+
+metaAnal = join_all(list(min.pval, max.pval, fb.pval), type = 'full')
+tmp = metaAnal %>% gather(method, pval, -ensembl_gene_id) %>% group_by(method) %>% filter(pval <= 0.05, method == 'fb.pval') %>%
+  left_join(Ensemble2HGNC) %>%
+  dplyr::select(hgnc_symbol) %>%
+  unique
 ```
@@
 
-### Fisher-Brown's meta analysis
 ```{r fisher.brown}
@@
 
-# Store pvalues
-tmp = join_all(list(Pmin,Pfb))
-write.table(tmp, file = './Pval.tsv', sep = '\t', row.names=F, quote=F)
-FILE_OBJ = File("./Pval.tsv", name = 'Meta Differential Expression Analysis FDR', parentId = CODE$properties$id)
-FILE_OBJ = synStore(FILE_OBJ, 
-                    used = as.character(DEXP_ID),
+# Store covariates in synapse
+write.table(covariates, file = 'ALLCovariates.tsv', sep = '\t', row.names=F, quote=F)
+OBJ = File('ALLCovariates.tsv', 
+           name = 'Covariates', 
+           parentId = CODE$properties$id)
+OBJ = synStore(OBJ,
+               used = as.character(c(covariates.id)),
                executed = thisFile,
-                    activityName = 'Differential expression analysis of control-AD')
+               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
 
-# Store differentially expressed genes
-write.table(rownameToFirstColumn(Adj.P.Value, 'hgnc_symbol'), file = './DiffGenes.tsv', sep = '\n', row.names=F, quote=F, col.names = F)
-FILE_OBJ = File("./DiffGenes.tsv", name = 'Differentially expressed genes FDR (individual studies)', parentId = CODE$properties$id)
-FILE_OBJ = synStore(FILE_OBJ, 
-                    used = as.character(DEXP_ID),
+# Store logcpm in synapse
+write.table(logcpm, file = 'ALLLogCPM.tsv', sep = '\t', row.names=F, quote=F)
+OBJ = File('ALLLogCPM.tsv', 
+           name = 'Log CPM (AD and Control)', 
+           parentId = CODE$properties$id)
+OBJ = synStore(OBJ,
+               used = as.character(c(covariates.id)),
                executed = thisFile,
-                    activityName = 'Differential expression analysis of control-AD')
+               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
+
+# Store diffexp in synapse
+write.table(diffexp, file = 'ALLDiffExp.tsv', sep = '\t', row.names=F, quote=F)
+OBJ = File('ALLDiffExp.tsv', 
+           name = 'Differential Expression (AD and Control)', 
+           parentId = CODE$properties$id)
+OBJ = synStore(OBJ,
+               used = as.character(c(covariates.id)),
+               executed = thisFile,
+               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
 ```

### 2016-06-10 code/Rmd/ROSMAP_RNASeq_SVA_Reprocessed.Rmd
Minor modifications

@@
 
-# library(doParallel)
-# library(foreach)
-# 
-# cl = makeCluster(14)
-# registerDoParallel(cl)
+library(doParallel)
+library(foreach)
+
+cl = makeCluster(14)
+registerDoParallel(cl)
 
@@
 
+# Store SVs
+write.table(SURR.VAR, file = 'ROSMAPsurrogatevar.tsv', sep = '\t', row.names=F, quote=F)
+SURR_OBJ = File('ROSMAPsurrogatevar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
+SURR_OBJ = synStore(SURR_OBJ, used = ALL_USED_IDs, activityName = activityName,
+                    executed = thisFile, activityDescription = activityDescription)
+
 # Store residual gene expression
@@
 |  Raw counts (log CPM)     |  `r COUNT_OBJ$properties$id`     |
+|  Surrogate variable       |  `r SURR_OBJ$properties$id`      |
 |  Residual Expression      |  `r EXP_OBJ$properties$id`       |

### 2016-06-11 code/Rmd/PredictPMI.Rmd
Added markdown to identify features which can predict PMI in ROSMAP study

---
title: "Prediction model for PMI using ROSMAP Reprocessed RNASeq"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---

```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./PredictPMI.Rmd",
                                 parentId ="syn5570291",
                                 entityName = 'Predict PMI')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get it from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(reshape2)
library(limma)
library(Biobase)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570291';
activityName = 'Predict PMI';
activityDescription = 'Predict PMI using ROSMAP data';

thisFileName <- 'PredictPMI.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r rosmap.download.data, cache=TRUE}
# Download reprocessed counts (paired end)
COUNT_PE_ID <- 'syn6112813';
ALL_USED_IDs <- COUNT_PE_ID
COUNT_PE <- read.table(synGet(COUNT_PE_ID)@filePath, header=T, sep='\t', check.names = F)

# Download reprocessed counts (single end)
COUNT_SE_ID <- 'syn6117775';
ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_SE_ID
COUNT_SE <- read.table(synGet(COUNT_SE_ID)@filePath, header=T, sep='\t', check.names = F)

COUNT = COUNT_PE + COUNT_SE[rownames(COUNT_PE), colnames(COUNT_PE)]
rownames(COUNT) = sapply(rownames(COUNT), function(x){unlist(str_split(x, '\\.'))[1]})

# Download filtered logCPM
LOG.CPM_ID <- 'syn6129370';
ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_ID
LOG.CPM <- read.table(synGet(LOG.CPM_ID)@filePath, header=T, sep='\t', check.names = F)

COUNT = COUNT[intersect(rownames(COUNT), LOG.CPM$ensembl_gene_id), intersect(colnames(COUNT), colnames(LOG.CPM))]

# Get covariates
COVARIATES_ID <- 'syn6129368'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = COVARIATES_ID
COVARIATES <- read.table(synGet(COVARIATES_ID)@filePath,sep='\t',header=T, row.names = 1)

FactorCovariates <- c("cogdx")
ContCovariates <- c("RINcontinuous","age_death","pmi")

COVARIATES = COVARIATES[,c(FactorCovariates,ContCovariates),drop=F]
# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates,drop=F], factor)

# Get surrogate variables
SURR.VAR_ID <- 'syn6140323'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = SURR.VAR_ID
SURR.VAR <- read.table(synGet(SURR.VAR_ID)@filePath,sep='\t',header=T)
rownames(SURR.VAR) = rownames(COVARIATES)
COVARIATES = cbind(COVARIATES, SURR.VAR)

COUNT = COUNT[,rownames(COVARIATES)]
```
Using dorsolateral prefrontal cortex of `r dim(COUNT)[2]` subjects from the ROS and MAP cohorts to predict PMI. Following sample are removed `r paste(removedIDs, collapse = ',')`
### Fit model with PMI
```{r w.pmi}
writeLines(paste('Using following covariates in the model:',paste(colnames(COVARIATES), collapse=', '),'as fixed effects'))
# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES,Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

# Estimate sample weights using voom methodology
VOOM.ADJUSTED.GENE_EXPRESSION = voom(COUNT, design=DM1$design, plot=F)

# Fit linear model using estimated weights and calculate BIC's
BIC.WPMI = sapply(1:dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], function(gene){
  y = VOOM.ADJUSTED.GENE_EXPRESSION$E[gene,]
  w = VOOM.ADJUSTED.GENE_EXPRESSION$weights[gene,]
  x = DM1$design
  obj = lm(y~0+., data = cbind(x,data.frame(y=y)), weights = w)
  BIC(obj)
})
names(BIC.WPMI) = rownames(VOOM.ADJUSTED.GENE_EXPRESSION$E)
```
### Fit model with out PMI
```{r wo.pmi}
COVARIATES.tmp = COVARIATES[,-(4)]
writeLines(paste('Using following covariates in the model:',paste(colnames(COVARIATES.tmp), collapse=', '),'as fixed effects'))

# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES.tmp,Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

# Estimate sample weights using voom methodology
VOOM.ADJUSTED.GENE_EXPRESSION = voom(COUNT, design=DM1$design, plot=F)

# Fit a linear model
VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)

# Calculate residual
RESIDUAL.EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT, VOOM.ADJUSTED.GENE_EXPRESSION$E)

# Fit linear model using estimated weights and calculate BIC's
BIC.WOPMI = sapply(1:dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], function(gene){
  y = VOOM.ADJUSTED.GENE_EXPRESSION$E[gene,]
  w = VOOM.ADJUSTED.GENE_EXPRESSION$weights[gene,]
  x = DM1$design
  obj = lm(y~0+., data = cbind(x,data.frame(y=y)), weights = w)
  BIC(obj)
})
names(BIC.WOPMI) = rownames(VOOM.ADJUSTED.GENE_EXPRESSION$E)
```
### Predict PMI
```{rpredict.pmi}
# Extract features that show improvement in the model with pmi when compared to without pmi
genes = rownames(VOOM.ADJUSTED.GENE_EXPRESSION$E)[which((BIC.WOPMI-BIC.WPMI) >= 6)]

# Extract x and y
x = t(RESIDUAL.EXPRESSION[genes,])
y = COVARIATES$pmi

# Remove samples with 0 pmi
ind = y == 0
x = x[!ind,]
y = log(y[!ind])

# Linear regression model
glm.model = foreach(i = 1:100,
                .export = c('x','y'),
                .packages = c('glmnet', 'CovariateAnalysis', 'dplyr', 'plyr')) %dopar% {
  ind = sample(1:length(y), round(length(y)*0.8))
  ind.comp = setdiff(1:length(y), ind)

  obj = cv.glmnet(x[ind,],y[ind], family = 'gaussian')
  coeff = as.matrix(coef.glmnet(obj, s = 'lambda.min'))
  ind.var1 = rownames(coeff)
  ind.var2 = setdiff(rownames(coeff)[coeff != 0], '(Intercept)')

  # Prune model based on least variable of importance
  while(length(ind.var1) > length(ind.var2)){
    obj1 = cv.glmnet(x[ind,ind.var2],y[ind], family = 'gaussian')
    coeff = as.matrix(coef.glmnet(obj1, s = 'lambda.min'))
    ind.var1 = ind.var2
    ind.var2 = setdiff(rownames(coeff)[coeff != 0], '(Intercept)')
  }

  mse = sum((predict(obj1, x[ind.comp,ind.var2], s = 'lambda.min', type = 'response') - y[ind.comp])^2)/length(ind.comp)
  coeff = as.matrix(coef.glmnet(obj1, s = 'lambda.min')) %>%
    rownameToFirstColumn('Coefficients') %>%
    plyr::rename(c('1' = i))
  list(mse = mse, coeff = coeff)
}
boxplot(sapply(glm.model, function(x){x$mse}), main = 'glmnet', ylab = 'mse')
coeff1 = lapply(glm.model, function(x){x$coeff}) %>%
  join_all(type = 'full')
coeff1[is.na(coeff1)] = 0
coeff1 = coeff1[which(rowSums(coeff1[,-(1)] == 0) < 80),]

# Random forest model
rf.model = foreach(i = 1:100,
                .export = c('x','y'),
                .packages = c('ranger', 'CovariateAnalysis', 'dplyr', 'plyr')) %dopar% {
  ind = sample(1:length(y), round(length(y)*0.8))
  ind.comp = setdiff(1:length(y), ind)

  obj = ranger(formula = y ~., data = cbind(data.frame(y = y[ind]), x[ind,]), 
               importance = 'permutation',
               write.forest = TRUE)
  coeff = obj$variable.importance
  ind.var1 = names(coeff)
  ind.var2 = names(coeff)[coeff >= abs(min(coeff))]

  # Prune model based on least variable of importance
  while(length(ind.var1) > length(ind.var2)){
    obj1 = ranger(formula = y ~., data = cbind(data.frame(y = y[ind]), x[ind,ind.var2]),
                  importance = 'permutation',
                  write.forest = TRUE)
    coeff = obj1$variable.importance
    ind.var1 = ind.var2
    ind.var2 = names(coeff)[coeff >= abs(min(coeff))]
  }

  mse = obj1$prediction.error
  coeff = obj1$variable.importance %>%
    rownameToFirstColumn('Coefficients') %>%
    plyr::rename(c('DF' = i))
  list(mse = mse, coeff = coeff)
}
boxplot(sapply(rf.model, function(x){x$mse}), main = 'random forest', ylab = 'mse')
coeff2 = lapply(rf.model, function(x){x$coeff}) %>%
  join_all(type = 'full')
coeff2[is.na(coeff2)] = 0
coeff2 = coeff2[which(rowSums(coeff2[,-(1)] == 0) < 80),]
```

### Store results in synapse
```{r syn.store}
# Create folder to store results in synapse
CODE = Folder(name = 'Predict PMI', parentId ="syn5570291")
CODE = synStore(CODE)

# Store features in synapse (union of both the models)
tmp = setdiff(unique(c(coeff1$Coefficients, coeff2$coefficients)), '(Intercept)')
write.table(tmp, file = 'features.tsv', sep = '\n', row.names = F, quote=F)
obj = File('features.tsv', name = 'Predictors of PMI', parentId = CODE$properties$id)
obj = synStore(obj, used = as.character(ALL_USED_IDs), executed = thisFile, 
               activityName = activityName, activityDescription = activityDescription)
```

### 2016-06-11 code/Rmd/PredictPMI.Rmd
Minor fix

@@
 library(biomaRt)
+
+library(doParallel)
+library(foreach)
+
+cl = makeCluster(14)
+registerDoParallel(cl)
 
@@
 ### Predict PMI
-```{rpredict.pmi}
+```{r predict.pmi, fig.height=6, fig.width=6}
 # Extract features that show improvement in the model with pmi when compared to without pmi
@@
                activityName = activityName, activityDescription = activityDescription)
+stopCluster(cl)
 ```

### 2016-06-11 code/Rmd/Mayo_RNASeq_Reprocessed_AdjustPMI.Rmd
Added markdown to process Mayo RNASeq based on PMI guided SVA

--- 2016-05-30 code/Rmd/Mayo_RNASeq_SVA_Reprocessed.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_SVA_Reprocessed.Rmd",
+knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed_AdjustPMI.Rmd",
                                  parentId = "syn5570325",
-                                 entityName = 'UFL_Mayo_ISB RNASeq Reprocessed Counts (with SVA adjustments)')
+                                 entityName = 'UFL_Mayo_ISB RNASeq Reprocessed Counts (with SVA based PMI adjustments)')
 ```
@@
 
-thisFileName <- 'Mayo_RNASeq_SVA_Reprocessed.Rmd'
+thisFileName <- 'Mayo_RNASeq_Reprocessed_AdjustPMI.Rmd'
 
@@
 ```{r sva}
+# Get genes that are predictors of PMI and use them as control probes
+CONTROL.PROBES_ID = 'syn6145639'
+ALL_USED_IDs = c(ALL_USED_IDs, CONTROL.PROBES_ID)
+CONTROL.PROBES = fread(synGet(CONTROL.PROBES_ID)@filePath, data.table = F)
+cntrl = rep(0.5, dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
+names(cntrl) = sapply(rownames(VOOM.ADJUSTED.GENE_EXPRESSION$E), function(x){unlist(str_split(x,'\\.'))[1]})
+cntrl[intersect(unlist(CONTROL.PROBES), names(cntrl))] = 1
+
+# Get actual and null model
 MODEL0 = getDesignMatrix(COVARIATES[,setdiff(postAdjustCovars, "BrainRegion.Diagnosis"),drop=F],Intercept = F)$design # Get null model by removing variable of interest
@@
 
-## Calculate the number of surrogate variables
-# Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
-SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
-NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
-SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
-SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]
-
-# Permute and calculate the permuted singular values
-PERM.SING.VALS <- foreach (i = 1:100,
-         .combine = rbind,
-         .packages = c('limma','CovariateAnalysis'),
-         .export = c('EXPR','MODEL1')) %dopar% {
-
-           # Permute expression counts
-           PERM.COUNTS <- t(apply(EXPR$filteredExprMatrix, 1, sample, replace = FALSE))
-
-           # Estimate voom weights
-           VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)
-
-           # Fit linear model using new weights and new design
-           VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)
-
-           # Residuals after normalisation
-           RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)
-
-           PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
-           PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
-        }
-num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])
+# ## Calculate the number of surrogate variables
+# # Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
+# SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
+# NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
+# SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
+# SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]
+# 
+# # Permute and calculate the permuted singular values
+# PERM.SING.VALS <- foreach (i = 1:100,
+#          .combine = rbind,
+#          .packages = c('limma','CovariateAnalysis'),
+#          .export = c('EXPR','MODEL1')) %dopar% {
+#            
+#            # Permute expression counts
+#            PERM.COUNTS <- t(apply(EXPR$filteredExprMatrix, 1, sample, replace = FALSE))
+#            
+#            # Estimate voom weights
+#            VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)
+#            
+#            # Fit linear model using new weights and new design
+#            VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)
+#            
+#            # Residuals after normalisation
+#            RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)
+#            
+#            PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
+#            PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
+#         }
+# num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])
 
 num.sv2 = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456)
-NUM.SV = min(num.sv1, num.sv2)
+NUM.SV = num.sv2; #min(num.sv1, num.sv2)
 
-SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, n.sv = NUM.SV, method = "two-step", B = 30)$sv
+SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, 
+               MODEL1, MODEL0, 
+               n.sv = NUM.SV, 
+               controls = cntrl,
+               method = "supervised", 
+               B = 30)
+
+SURR.VAR = SURR.VAR$sv
 colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
@@
 kable(tmp)
+
+DE = DE %>%
+  left_join(backgroundGenes) %>% left_join(Ensemble2HGNC) %>%
+  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, Comparison, BrainRegion, Name, 
+                logFC, AveExpr, t, P.Value, adj.P.Val, B)
 ```
@@
 # Code
-CODE <- Folder(name = 'UFL_Mayo_ISB RNASeq Reprocessed Counts (with SVA adjustments)', parentId = parentId)
+CODE <- Folder(name = 'UFL_Mayo_ISB RNASeq Reprocessed Counts (with SVA based PMI adjustments)', parentId = parentId)
 CODE <- synStore(CODE)
@@
 COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                   executed = thisFile, activityDescription = activityDescription)
+
+# Store surrogate variables
+rownames(SURR.VAR) = rownames(COVARIATES)
+SURR.VAR = rownameToFirstColumn(SURR.VAR, 'ID')
+write.table(SURR.VAR, file = 'MAYOsurrogatevariable.tsv', sep = '\t', row.names=F, quote=F)
+SURR_OBJ = File('MAYOsurrogatevariable.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
+SURR_OBJ = synStore(SURR_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                     executed = thisFile, activityDescription = activityDescription)
@@
 |  Covariates               |  `r COV_OBJ$properties$id`      |
+|  Surrogate Variables      |  `r SURR_OBJ$properties$id`     |
 |  Raw counts (logCPM)      |  `r RAWCPM_OBJ$properties$id`   |

### 2016-07-13 code/Rmd/ROSMAP_RNASeq_SVA_Reprocessed.Rmd
Minor modifications

@@
 
-# Permute and calculate the permuted singular values
-PERM.SING.VALS <- foreach (i = 1:100,
-         .combine = rbind,
-         .packages = c('limma','CovariateAnalysis'),
-         .export = c('PROCESSED_COUNTS','MODEL1')) %dopar% {
-  PERM.COUNTS <- t(apply(PROCESSED_COUNTS$filteredExprMatrix, 1, sample, replace = FALSE))
-  # Estimate voom weights
-  VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)
-
-  # Fit linear model using new weights and new design
-  VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)
-
-  # Residuals after normalisation
-  RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)
-
-  PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
-  PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
-}
-num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])
+# # Permute and calculate the permuted singular values
+# PERM.SING.VALS <- foreach (i = 1:100,
+#          .combine = rbind,
+#          .packages = c('limma','CovariateAnalysis'),
+#          .export = c('PROCESSED_COUNTS','MODEL1')) %dopar% {
+#   PERM.COUNTS <- t(apply(PROCESSED_COUNTS$filteredExprMatrix, 1, sample, replace = FALSE))
+#   # Estimate voom weights
+#   VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)
+#   
+#   # Fit linear model using new weights and new design
+#   VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)
+#   
+#   # Residuals after normalisation
+#   RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)
+#   
+#   PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
+#   PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
+# }
+# num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])
 
 num.sv2 = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456)
-NUM.SV = min(num.sv1, num.sv2)
+NUM.SV = num.sv2 #min(num.sv1, num.sv2)
 
@@
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
-  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(VOOM.ADJUSTED.FIT$design[,varsToAddIn])
 ```
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T)
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
 

### 2016-07-13 code/Rmd/Mayo_RNASeq_Reprocessed.Rmd
Included PMI predicitons based on the availability of PMI values

@@
 ## Load required libraries
-library(CovariateAnalysis)
+library(CovariateAnalysis) ## install_github('th1vairam/CovariateAnalysis@dev')
 library(data.table)
@@
 
+library(glmnet)
+
 library(doParallel)
@@
 ```
-Following samples were considered outliers and removed based on PCA and/or eucledian distance:
+Following samples were considered outliers and removed based on PCA and/or eucledian distance: `r paste(c("11386_CER","1923_CER", "11311_CER", "11477_CER", "11408_TCX", "11294_TCX", "11396_TCX"), collapse = ',')`
 ```{r preprocess.data1}
@@
 `r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project. Following samples were removed `r indRemoved`
+### Predict missing PMI
+```{r predict.pmi}
+PMI = read.table('Mayo_PMI.tsv', header=T, fill=T) %>%
+  dplyr::select(-Source) %>% 
+  right_join(METADATA %>%
+               dplyr::mutate(SubjectId_Synpase = paste(Donor_ID, BrainRegion, sep = '_')))
+rownames(PMI) = PMI$SubjectId_Synpase
+
+# Get genes that are predictors of PMI and use them as control probes
+CONTROL.PROBES_ID = 'syn6145639'
+ALL_USED_IDs = c(ALL_USED_IDs, CONTROL.PROBES_ID)
+CONTROL.PROBES = fread(synGet(CONTROL.PROBES_ID)@filePath, data.table = F)
+
+# Remove genes
+EXPR <- getGeneFilteredGeneExprMatrix(COUNT, 
+                                      MIN_GENE_CPM=0,
+                                      MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0,
+                                      verbose=T)
+
+# Initial normalisation of gene expression
+VOOM.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=NULL, plot=F)
+
+ind = sapply(rownames(VOOM.GENE_EXPRESSION$E), function(x){str_split(x, '\\.')[[1]][1]}) %in% CONTROL.PROBES$x 
+EXPR = VOOM.GENE_EXPRESSION$E[ind,]
+PMI = PMI[colnames(EXPR),]
+
+expr = dlply(PMI, .(Donor_ID), function(mtd, expr){
+  if (dim(mtd)[1] > 1)
+    expr = rowMeans(expr[, mtd$SubjectId_Synpase]) %>% rownameToFirstColumn('GeneID')
+  else
+    expr = expr[, mtd$SubjectId_Synpase] %>% rownameToFirstColumn('GeneID')
+  expr = expr %>% plyr::rename(c('DF' = unique(mtd$Donor_ID)))
+  return(expr)
+}, EXPR) %>%
+  join_all
+rownames(expr) = expr$GeneID
+
+PMI = PMI[,colnames(PMI) %in% c('Donor_ID','PMI')] %>% unique
+rownames(PMI) = PMI$Donor_ID
+
+PMI = PMI[intersect(rownames(PMI), colnames(expr)), ]
+expr = expr[,intersect(rownames(PMI), colnames(expr))]
+
+# Split data into trainining and prediction set
+x.train = expr[,!is.na(PMI$PMI)]
+x.predict = expr[,is.na(PMI$PMI)]
+
+y.train = PMI[!is.na(PMI$PMI),]
+
+plr.model = cv.glmnet(t(x.train), y.train$PMI, family = 'gaussian')
+coeff = as.matrix(coef.glmnet(plr.model, s = 'lambda.min'))
+mse = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train$PMI)^2)/length(y.train$PMI)
+y.predict = predict(plr.model, t(x.predict), s = 'lambda.min', type = 'response') %>%
+  rownameToFirstColumn('Donor_ID') %>%
+  plyr::rename(c('1' = 'PMI'))
+
+PMI = rbindlist(list(y.train, y.predict), use.names = T, fill = T)
+
+METADATA = left_join(METADATA, PMI)
+```
 
@@
 
-multiplot(plotlist = p, cols = 2)
+# PMI
+p[[3]] = ggplot(METADATA, aes(x = Diagnosis, y = as.numeric(PMI), fill = Tissue)) + geom_boxplot()
+p[[3]] = p[[3]] + ggtitle('PMI') + theme(legend.position = 'top')
+
+multiplot(plotlist = p, cols = 3)
 
@@
 FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
-ContCovariates <- c("RIN", "RIN2", "AgeAtDeath")
+ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI")
 
@@
 COVARIATES.tmp[,FactorCovariates] = lapply(COVARIATES.tmp[,FactorCovariates], as.numeric)
+col.mat = labels2colors(cbind(tree,COVARIATES.tmp))
+col.mat = gsub('\\.1', '', col.mat)
 
 WGCNA::plotDendroAndColors(sampleTrees, 
-                           labels2colors(cbind(tree,COVARIATES.tmp)),
+                           col.mat,
                            groupLabels = c('Cluster',colnames(COVARIATES.tmp)))
@@
 Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
-```{r iterative.adjusted.voom.normalisation, results='asis', echo=FALSE}
+```{r iterative.adjusted.voom.normalisation, results='asis'}
 postAdjustCovars = c('BrainRegion.Diagnosis')
@@
 `r tmp`
+### Normalisation (manual design)
+Since there is a strong association of case control status with Source and flowcell information we are manually including them.
+Also, to be consistent with other brain banks we are including AgeAtDeath as a covariate.
+```{r manual.adjusted.voom.normalisation, results='asis', echo=FALSE}
+postAdjustCovars = c(postAdjustCovars, c('Source','FLOWCELL','AgeAtDeath')) %>% unique
+randomCovariates = 'Donor_ID'
+
+writeLines(paste('Using following covariates in the model:', 
+                 paste(postAdjustCovars, collapse=', '),
+                 'as fixed effects and',
+                 paste(randomCovariates, collapse=', '),
+                 'random effect'))
+
+# Post adjusted design matrix
+DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
+DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
+
+# Estimate voom weights
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=DM1$design, plot=F)
+
+# Calculate correlation values of random effects
+correlation = parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block = COVARIATES$Donor_ID)
+
+# Re-estimate voom weights
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=DM1$design, plot=F, 
+                                     block = COVARIATES$Donor_ID, correlation = correlation$cor)
+
+# Fit a mixed effect linear model
+FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
+
+# Residuals after normalisation
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
+
+# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
+residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+                                              COVARIATES,
+                                              'all adjusted design(voom-normalized)',
+                                              isKeyPlot=TRUE)
+```
 
@@
 COVARIATES.tmp[, FactorCovariates] = sapply(COVARIATES.tmp[,FactorCovariates], as.numeric)
+col.mat = labels2colors(cbind(tree,COVARIATES.tmp))
+col.mat = gsub('\\.1', '', col.mat)
+
 plotDendroAndColors(sampleTrees, 
-                    colors = labels2colors(cbind(data.frame(clust = clust), COVARIATES.tmp)), 
+                    colors = col.mat, 
                     abHeight = 0.80, 
@@
 tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[clust == 3], collapse = ','))
-tmp = base::gsub('_','\\\\_',tmp)
-writeLines(tmp)
-tmp = paste('Samples in cluster 4 are', paste(sampleTrees$labels[clust == 4], collapse = ','))
 tmp = base::gsub('_','\\\\_',tmp)
@@
 # Store covariates
-COVARIATES = rownameToFirstColumn(COVARIATES, 'ID')
+COVARIATES = rownameToFirstColumn(COVARIATES, 'ID') %>%
+  dplyr::select(-PMI)
 write.table(COVARIATES, file = 'MAYOcovariates.tsv', sep = '\t', row.names=F, quote=F)
@@
 # Design Matrix
+DM1$design = DM1$design[,!(colnames(DM1$design) %in% c('PMI'))]
 write.table(DM1$design, file='MAYOdesign.tsv', sep = '\t', row.names=F, quote=F)

### 2016-07-13 code/Rmd/Mayo_RNASeq_Reprocessed.Rmd
Added PMI data to SVA based adjustments with Mayo reprocessed counts

@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=25, fig.width=25}
+```{r diffExp, fig.height=10, fig.width=20}
 # Fit contrast

### 2016-07-14 code/Rmd/MSSM_RNASeq_SVA_Reprocessed.Rmd
Removed 206 samples from Noam's analysis in MSSM SVA markdown

@@
 
+# Get sample ids that are to be retained (data from Noam on 29th June) 
+tmp = read.table('MSSM_SampleIDs.txt', header = T)
+removeID = setdiff(SampleID$ID, tmp$x)
+writeLines(paste('Removed',length(removeID),'samples from MSSM'))
+SampleID = filter(SampleID, ID %in% tmp$x)
+
 # Get technical metadata
@@
                        dplyr::mutate(ID = as.numeric(ID))) %>%
-  left_join(SampleID) %>%
+  inner_join(SampleID) %>%
   filter(!is.na(SampleID)) %>%
@@
 ```
+`r paste('In total',length(indRemoved),'samples were removed')`
+`r paste(indRemoved, collapse = ',')`
 ### Covariate clustering
@@
 
-### Normalisation (iterative design)
+### Normalisation (manual design)
 Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix. Here BB (Donor Information) is chosen as random effect
-```{r iterativeNorm}
+```{r manualNorm}
 # Primary variable of interest
-postAdjustCovars = 'BrainRegion.Diagnosis'
+postAdjustCovars = c("BrainRegion.Diagnosis", "RIN", "Batch", "PMI", "AOD")
 
-# Assign residual covariates
-residualSigCovars = preAdjustedSigCovars
-
-covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, 'BB'))]
-
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
-
-loopCount = 0 
-while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
 writeLines(paste('Using following covariates in the model:',
@@
 residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                                COVARIATES[, residCovars, drop=F], 'all adjusted design(voom-normalized)',
+                                              COVARIATES, 'all adjusted design(voom-normalized)',
                                               isKeyPlot=TRUE)
-
-  # Add postadjusted covariates (if any)
-  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, 'BB'))]
-
-  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
-  loopCount = loopCount + 1
-}
-modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
-                  'as fixed effects')
-
-tmp <- paste('Using following covariates in the final model:', modelStr)
 ```

### 2016-07-14 code/Rmd/MSSM_RNASeq_SVA_Reprocessed.Rmd
Added parallel version of num.sv and minor modifications in MSSM SVA reprocessed counts

@@
 devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")
+source('../R/num.sv.R') # Source parallel version of num.sv from local repo
 
@@
 ```
-`r tmp`
 ### Sanity check
@@
 ## Calculate the number of surrogate variables
-NUM.SV = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456, LINPACK = TRUE, B = 10)
+NUM.SV = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456, B = 10)
 

### 2016-07-14 code/R/num.sv.R

@@ -1,0 +1,83 @@
+num.sv <- function (dat, mod, method = c("be", "leek"), vfilter = NULL, 
+                    B = 20, seed = NULL, LINPACK = F) 
+{
+  if (!is.null(vfilter)) {
+    if (vfilter < 100 | vfilter > dim(dat)[1]) {
+      stop(paste("The number of genes used in the analysis must be between 100 and", 
+                 dim(dat)[1], "\n"))
+    }
+    tmpv = rowVars(dat)
+    ind = which(rank(-tmpv) < vfilter)
+    dat = dat[ind, ]
+  }
+  method <- match.arg(method)
+  if (method == "be") {
+    if (!is.null(seed)) {
+      set.seed(seed)
+    }
+    warn <- NULL
+    n <- ncol(dat)
+    m <- nrow(dat)
+    H <- mod %*% solve(t(mod) %*% mod) %*% t(mod)
+    res <- dat - t(H %*% t(dat))
+    uu <- svd(res, LINPACK = LINPACK)
+    ndf <- min(m, n) - ceiling(sum(diag(H)))
+    dstat <- uu$d[1:ndf]^2/sum(uu$d[1:ndf]^2)
+    dstat0 <- foreach (i = 1:B,
+                       .combine = rbind,
+                       .export = c('res', 'H', 'LINPACK', 'ndf')) %dopar% {
+                         res0 <- t(apply(res, 1, sample, replace = FALSE))
+                         res0 <- res0 - t(H %*% t(res0))
+                         tryCatch({
+                           uu0 <- svd(res0, LINPACK = LINPACK)
+                           dstat0 <- uu0$d[1:ndf]^2/sum(uu0$d[1:ndf]^2)
+                         }, error = function(x) {
+                           dstat0 = matrix(0,1,ndf)
+                         }) 
+                       }
+    psv <- rep(1, n)
+    for (i in 1:ndf) {
+      psv[i] <- mean(dstat0[, i] >= dstat[i])
+    }
+    for (i in 2:ndf) {
+      psv[i] <- max(psv[(i - 1)], psv[i])
+    }
+    nsv <- sum(psv <= 0.1)
+    return(as.numeric(list(n.sv = nsv)))
+  }
+  else {
+    dat <- as.matrix(dat)
+    dims <- dim(dat)
+    a <- seq(0, 2, length = 100)
+    n <- floor(dims[1]/10)
[…]

### 2016-07-20 code/Rmd/MSSM_RNASeq_Reprocessed.Rmd

@@
 `r tmp`
+### Normalisation (manual design)
+Adding AOD as a covariate to the identified covariates (for cross brain bank comparison)
+```{r manualNorm}
+# Primary variable of interest
+postAdjustCovars = c(postAdjustCovars, 'Batch', 'PMI', 'AOD') %>% unique
+
+writeLines(paste('Using following covariates in the model:',
+                 paste(postAdjustCovars, collapse=', '),
+                 'as fixed effects'))
+
+# Post adjusted design matrix
+DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
+DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
+
+# Estimate correlation of random effects
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)
+correlation <- parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block=COVARIATES$BB)
+
+# Re-calculate voom weights with correlation of random effects
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F,
+                                     block=COVARIATES$BB, correlation = correlation$cor)
+
+# Re-calculate linear fit using new weights and new design
+VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
+
+# Residuals after normalisation
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
+
+# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
+residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+                                              COVARIATES, 'all adjusted design(voom-normalized)',
+                                              isKeyPlot=TRUE)
+```
 ### Sanity check
@@
 ```
+
 ```{r get.gene.symbols, cache=TRUE, eval=TRUE}
@@
 ```
+
 ```{r summary, results='asis', echo=FALSE}

### 2016-07-20 code/Rmd/Mayo_RNASeq_Reprocessed_AdjustPMI.Rmd

@@
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  group_by(Comparison) %>%
-  dplyr::summarize(nDiffExpGenes = length(unique(Gene.ID)))
+  dplyr::mutate(D = logFC/abs(logFC)) %>%
+  group_by(Comparison, D) %>%
+  dplyr::summarize(nDiffExpGenes = length(unique(Gene.ID))) %>%
+  spread(D, nDiffExpGenes)
 kable(tmp)

### 2016-07-20 code/Rmd/Mayo_RNASeq_SVA_Reprocessed.Rmd

@@
 
-# library(doParallel)
-# library(foreach)
-# 
-# cl = makeCluster(14)
-# registerDoParallel(cl)
+library(glmnet)
+
+library(doParallel)
+library(foreach)
+
+cl = makeCluster(14)
+registerDoParallel(cl)
 
@@
 p[[2]] = p[[2]] + ggtitle('AgeAtDeath') + theme(legend.position = 'top')
+# PMI
+p[[3]] = ggplot(METADATA, aes(x = Diagnosis, y = as.numeric(PMI), fill = Tissue)) + geom_boxplot()
+p[[3]] = p[[3]] + ggtitle('PMI') + theme(legend.position = 'top')
 
-multiplot(plotlist = p, cols = 2)
+multiplot(plotlist = p, cols = 3)
 
@@
 FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
-ContCovariates <- c("RIN", "RIN2", "AgeAtDeath")
+ContCovariates <- c("RIN", "RIN2", "AgeAtDeath", "PMI")
 
@@
 ```
-### Normalisation (iterative)
-Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
-```{r iterative.adjusted.voom.normalisation, results='asis', echo=FALSE}
-postAdjustCovars = c('BrainRegion.Diagnosis')
+### Normalisation (manual design)
+RIN and PMI.
+Since there is a strong association of case control status with Source and flowcell information we are manually including them.
+Also, to be consistent with other brain banks we are including AgeAtDeath as a covariate.
+```{r manual.adjusted.voom.normalisation, results='asis', echo=FALSE}
+postAdjustCovars = c('BrainRegion.Diagnosis', 'RIN', 'PMI', 'Source', 'FLOWCELL', 'AgeAtDeath')
 randomCovariates = 'Donor_ID'
 
-# Assign residual covariates
-residualSigCovars = preAdjustedSigCovars
-significantCovars = residualSigCovars$significantCovars
-covariatesEffects = residualSigCovars$Effects.significantCovars
-covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates))]
-
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
-
-loopCount = 0 
-while(length(significantCovars)!=0 && loopCount <= 100){
 writeLines(paste('Using following covariates in the model:', 
                  paste(postAdjustCovars, collapse=', '),
-                   'as fixed effects'))
+                 'as fixed effects and',
+                 paste(randomCovariates, collapse=', '),
+                 'random effect'))
 
@@
 
-  # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, randomCovariates))
-
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
@@
                                               isKeyPlot=TRUE)
-
-  # Add postadjusted covariates (if any)
-  significantCovars = residualSigCovars$significantCovars
-  covariatesEffects = residualSigCovars$Effects.significantCovars
-  covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars, randomCovariates))]
-
-  postAdjustCovars = c(postAdjustCovars,names(which.max(covariatesEffects)))
-  loopCount = loopCount + 1
-}
-modelStr <-paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=','), 'as fixed effects')
-
-tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
 ```
-`r tmp`
 ### Sanity Check
@@
 MODEL1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
+MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
 
-## Calculate the number of surrogate variables
-# Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
-SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
-NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
-SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
-SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]
-
-# Permute and calculate the permuted singular values
-PERM.SING.VALS <- foreach (i = 1:100,
-         .combine = rbind,
-         .packages = c('limma','CovariateAnalysis'),
-         .export = c('EXPR','MODEL1')) %dopar% {
-
-           # Permute expression counts
-           PERM.COUNTS <- t(apply(EXPR$filteredExprMatrix, 1, sample, replace = FALSE))
-
-           # Estimate voom weights
-           VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)
-
-           # Fit linear model using new weights and new design
-           VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)
-
-           # Residuals after normalisation
-           RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)
-
-           PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
-           PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
-        }
-num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])
+# ## Calculate the number of surrogate variables
+# # Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
+# SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
+# NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
+# SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
+# SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]
+# 
+# # Permute and calculate the permuted singular values
+# PERM.SING.VALS <- foreach (i = 1:100,
+#          .combine = rbind,
+#          .packages = c('limma','CovariateAnalysis'),
+#          .export = c('EXPR','MODEL1')) %dopar% {
+#            
+#            # Permute expression counts
+#            PERM.COUNTS <- t(apply(EXPR$filteredExprMatrix, 1, sample, replace = FALSE))
+#            
+#            # Estimate voom weights
+#            VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)
+#            
+#            # Fit linear model using new weights and new design
+#            VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)
+#            
+#            # Residuals after normalisation
+#            RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)
+#            
+#            PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
+#            PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
+#         }
+# num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])
 
 num.sv2 = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456)
-NUM.SV = min(num.sv1, num.sv2)
+NUM.SV = num.sv2#min(num.sv1, num.sv2)
 
@@
 COVARIATES.tmp[, FactorCovariates] = sapply(COVARIATES.tmp[,FactorCovariates], as.numeric)
+col.mat = labels2colors(cbind(tree,COVARIATES.tmp))
+col.mat = gsub('\\.1', '', col.mat)
+
 plotDendroAndColors(sampleTrees, 
-                    colors = labels2colors(cbind(data.frame(clust = clust), COVARIATES.tmp)), 
+                    colors = col.mat,
                     abHeight = 0.80, 
@@
                     groupLabels = colnames(COVARIATES))
-
-tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[clust == 0], collapse = ','))
-tmp = base::gsub('_','\\\\_',tmp)
-writeLines(tmp)
-tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[clust == 3], collapse = ','))
-tmp = base::gsub('_','\\\\_',tmp)
-writeLines(tmp)
-tmp = paste('Samples in cluster 4 are', paste(sampleTrees$labels[clust == 4], collapse = ','))
-tmp = base::gsub('_','\\\\_',tmp)
-writeLines(tmp)
 ```
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=15, fig.width=25}
+```{r diffExp, fig.height=10, fig.width=20}
 # Fit contrast
@@
 RAWCPM_OBJ = synStore(RAWCPM_OBJ,
+                      used = ALL_USED_IDs, 
+                      activityName = activityName, 
+                      executed = thisFile, 
+                      activityDescription = activityDescription)
+
+# Design Matrix
+DM = VOOM.ADJUSTED.GENE_EXPRESSION$design
+DM = DM[,!(colnames(DM) %in% c('PMI'))]
+write.table(DM, file='MAYOdesign.tsv', sep = '\t', row.names=F, quote=F)
+DM_OBJ = File('MAYOdesign.tsv', 
+                  name = 'Design Matrix',
+                  parentId = CODE$properties$id)
+DM_OBJ = synStore(DM_OBJ,
                       used = ALL_USED_IDs, 

### 2016-07-20 code/Rmd/ROSMAP_RNASeq_Reprocessed_NonConditionedCogdx.Rmd

--- 2016-06-01 code/Rmd/ROSMAP_RNASeq_Reprocessed.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed.Rmd",
+knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_NonConditionedCogdx.Rmd",
                                  parentId ="syn5570291",
-                                 entityName = 'ROSMAP RNASeq Reprocessed Counts (Based on cogdx)')
+                                 entityName = 'ROSMAP RNASeq Reprocessed Counts 2 (Based on cogdx)')
 ```
@@
 
+library(ggdendro)
+library(RColorBrewer)
+
 options(xtable.type="html")
+
+devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")
 
@@
 
-thisFileName <- 'ROSMAP_RNASeq_Reprocessed.Rmd'
+thisFileName <- 'ROSMAP_RNASeq_Reprocessed_NonConditionedCogdx.Rmd'
 
@@
 # Primary variable of interest
-postAdjustCovars = 'cogdx'
+postAdjustCovars = NULL
 
@@
   DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
+  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols, drop=F]
 
@@
                        mart = mart)
-
-# Add gene names
-RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
-  rownameToFirstColumn('Gene.ID') %>%
-  left_join(backgroundGenes) %>%
-  left_join(Ensemble2HGNC) %>%
-  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
-
-VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
-  rownameToFirstColumn('Gene.ID') %>%
-  left_join(backgroundGenes) %>%
-  left_join(Ensemble2HGNC) %>%
-  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
 ```
@@
 ```{r diffExp, fig.height=8, fig.width=16}
+# Post adjusted design matrix
+DM1 = getDesignMatrix(COVARIATES[,'cogdx',drop=F],Intercept = F)
+DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols, drop=F]
+
+# Fit linear model using residuals and new design matrix
+FIT = lmFit(RESIDUAL.GENE_EXPRESSION, DM1$design)
+
 # Fit contrast
@@
                                      "cogdx2-cogdx4"), 
-                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
-FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
+                         levels = colnames(FIT$coefficients))
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
 FIT.CONTR = eBayes(FIT.CONTR)
@@
 ```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
+# Add gene names
+RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
+  rownameToFirstColumn('Gene.ID') %>%
+  left_join(backgroundGenes) %>%
+  left_join(Ensemble2HGNC) %>%
+  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+
+VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
+  rownameToFirstColumn('Gene.ID') %>%
+  left_join(backgroundGenes) %>%
+  left_join(Ensemble2HGNC) %>%
+  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+
 # Code
-CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts (Based on cogdx)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts 2 (Based on cogdx)", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-07-29 code/Rmd/InvestigateCoexpIssue.Rmd
Added secondary codes

---
title: "Investigate co-expression issue in reprocessed RNASeq"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
# Investigate co-expression issue
# Knitting code (commented)
knit2synapse::knitToFolderEntity(file = "./InvestigateCoexpIssue.Rmd",
                                 parentId = 'syn5569102',
                                 entityName = 'RNASeq Reprocessing Coexpression Quality Control')
```

```{r libload}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get it from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(reshape2)
library(limma)
library(Biobase)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)

library(synapseClient)
library(knitr)
library(githubr)

synapseLogin()
```
### Data download
#### Obtain adjusted FPKM/logcpm residual matrix from synapse
```{r data.download}
original.residuals.id = c('ROSMAP' = 'syn5581268', 'MAYO' = 'syn6088523', 'MSSM' = 'syn6120103')
all.used.ids = original.residuals.id
original.residuals = lapply(original.residuals.id, downloadFile)
original.residuals.cor = lapply(original.residuals, function(x){cor(t(x[,-(1:2)]))})

h = mapply(function(x,y){ hist(x, main = y) }, original.residuals.cor,  names(original.residuals.cor), SIMPLIFY = F)
writeLines('Adjusted residuals of original data deposited from the individual sites')
plot(h[[1]], main = names(original.residuals.cor)[1])
plot(h[[2]], main = names(original.residuals.cor)[2])
plot(h[[3]], main = names(original.residuals.cor)[3])
```
#### Obtain adjusted residual matrix from reprocessed counts in synapse
```{r data.download2}
reproc.residuals.id = c('ROSMAP' = 'syn6114451', 'MAYO' = 'syn6128271', 'MSSM' = 'syn6131415')
all.used.ids = c(all.used.ids, reproc.residuals.id)
reproc.residuals = lapply(reproc.residuals.id, downloadFile)
reproc.residuals.cor = lapply(reproc.residuals, function(x){cor(t(x[,-(1:3)]))})

h = lapply(reproc.residuals.cor, hist)
writeLines('Adjusted residuals of reprocessed data deposited by KKDang')
plot(h[[1]], main = names(reproc.residuals.cor)[1])
plot(h[[2]], main = names(reproc.residuals.cor)[2])
plot(h[[3]], main = names(reproc.residuals.cor)[3])
```
### Extract PC's and find the loading genes in the reprocessed data from each site
#### From the reprocessed data
```{r sample}
reproc.pc.samples = lapply(reproc.residuals, function(x){
  x = x[!duplicated(x$Gene.ID),]
  rownames(x) = x$Gene.ID
  x = x[,-(1:3)]

  PC = prcomp(x, scale. = T, center = T)
  writeLines('Percent variance explained by top 10 PCs are')
  print(((PC$sdev^2/sum(PC$sdev^2)) * 100)[1:10])

  return(PC)
})
pairs(reproc.pc.samples$ROSMAP$rotation[,1:4], main = 'ROSMAP')
pairs(reproc.pc.samples$MSSM$rotation[,1:4], main = 'MSSM')
pairs(reproc.pc.samples$MAYO$rotation[,1:4], main = 'MAYO')
```
##### Find all samples deviating from 2 SD in the first PC
```{r gene}
dev.samples = lapply(reproc.pc.samples, function(x){
  mn = mean(x$rotation[,1]); sdev = sd(x$rotation[,1]);
  zscr = (x$rotation[,1] - mn)/sdev
  print(names(zscr)[abs(zscr) >= 2])
})

reproc.pc.genes = lapply(reproc.residuals, function(x){
  x = x[!duplicated(x$Gene.ID),]
  rownames(x) = x$Gene.ID
  x = x[,-(1:3)]

  PC = prcomp(t(x), scale. = T, center = T)

  writeLines('Percent variance explained by top 10 PCs are')
  print(((PC$sdev^2/sum(PC$sdev^2)) * 100)[1:10])

  return(PC)
})
pairs(reproc.pc.genes$ROSMAP$rotation[,1:4], main = 'ROSMAP')
pairs(reproc.pc.genes$MSSM$rotation[,1:4], main = 'MSSM')
pairs(reproc.pc.genes$MAYO$rotation[,1:4], main = 'MAYO')

##### Find all samples deviating from 2 SD in the first PC
dev.genes = mapply(function(x,y){
  mn = mean(x$rotation[,y]); sdev = sd(x$rotation[,y]);
  zscr = (x$rotation[,y] - mn)/sdev
  genes = (names(zscr)[abs(zscr) >= 2])
}, reproc.pc.genes, c(1,2,1), SIMPLIFY = F)
tmp = Venn(dev.genes)
writeLines('Number of top deviating genes (2 SD)')
plot(tmp, doWeights = F)
writeLines('14 genes that are common to all three data sets are')
genes = intersect(intersect(dev.genes$ROSMAP,dev.genes$MAYO), dev.genes$MSSM)
paste(genes, collapse = ',') %>% print()
```
## Store results in Synapse
```{r syn.store}
parentId = 'syn5569102';
activityName = 'Quality control';

thisFileName <- 'InvestigateCoexpIssue.Rmd'

### Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))

### Store results in synapse
CODE = Folder(name='RNASeq Reprocessing Coexpression Quality Control', parentId = parentId)
CODE = synStore(CODE)

save(list = ls(), file = 'investigate.RData')
obj = File(file = 'investigate.RData', name = 'RNASeq Reprocessing Coexpression Quality Control', parentId = CODE$properties$id)
obj = synStore(obj, activityName = activityName, used = as.character(all.used.ids), executed = thisFile)
```

### 2016-08-17 code/Rmd/MetaAnalysis.Rmd
Minor modifications

@@
 
+library(Vennerable)
+library(colorfulVennPlot)
+
 synapseLogin()
+
+library(doParallel)
+library(foreach)
+
+clust = makeCluster(14)
+registerDoParallel(clust)
 
@@
 ```{r covariates}
-covariates.id = c('ROSMAP' = 'syn6114444',
-                  'MSBB' = 'syn6131421',
-                  'MAYO' = 'syn6128267')
+# covariates.id = c('ROSMAP' = 'syn6114444',
+#                   'MSBB' = 'syn6131421',
+#                   'MAYO' = 'syn6128267')
+# SVA Normalised
+covariates.id = c('ROSMAP' = 'syn6129368',
+                  'MSBB' = 'syn7059296',
+                  'MAYO' = 'syn6128630')
 
@@
 covariates = rbindlist(covariates, use.names = T, fill = T, idcol = 'Study')
-
 ```
@@
 ```{r counts}
-logcpm.id = c('ROSMAP' = 'syn6114447' ,
-              'MSBB' = 'syn6131419',
-              'MAYO' = 'syn6128269')
+# logcpm.id = c('ROSMAP' = 'syn6114447' ,
+#               'MSBB' = 'syn6131419',
+#               'MAYO' = 'syn6128269')
+# Get SVA adjusted data
+logcpm.id = c('ROSMAP' = 'syn6129370',
+              'MSBB' = 'syn7059294',
+              'MAYO' = 'syn6128632')
 
@@
 
+### Get residual expression from synapse
+```{r counts1}
+# Get SVA adjusted data
+resid.expr.id = c('ROSMAP' = 'syn6129372',
+                  'MSBB' = 'syn7059289',
+                  'MAYO' = 'syn6128634')
+
+resid.expr = lapply(resid.expr.id, downloadFile) %>%
+  lapply(function(x){
+    x = dplyr::select(x, -hgnc_symbol)
+    }) %>%
+  join_all(type = 'full') %>%
+  dplyr::select(Gene.ID, ensembl_gene_id, one_of(covariates$ID)) 
+resid.expr[is.na(resid.expr)] = dplyr::select(resid.expr, -Gene.ID, -ensembl_gene_id) %>%
+  min(na.rm=T)
+```
+
 ### Download differential expression data
 ```{r download.data}
-diffexp.id = c('ROSMAP' = 'syn6114453',
-               'MSBB' = 'syn6131423',
-               'MAYO' = 'syn6128273')
+# diffexp.id = c('ROSMAP' = 'syn6114453',
+#                'MSBB' = 'syn6131423',
+#                'MAYO' = 'syn6128273')
+# Get SVA adjusted differential expression
+diffexp.id = c('ROSMAP' = 'syn6129374',
+               'MSBB' = 'syn7059298',
+               'MAYO' = 'syn6128636')
 
@@
   filter(Comparison == 'cogdx1-cogdx4') %>%
-  dplyr::select(Gene.ID, logFC, P.Value, adj.P.Val) %>%
+  dplyr::select(ensembl_gene_id, logFC, P.Value, adj.P.Val) %>%
+  # dplyr::select(Gene.ID, logFC, P.Value, adj.P.Val) %>%
   dplyr::mutate(Tissue = 'DLPFC', Contrast = 'NCI-AD')
@@
   dplyr::filter(Name == 'ND-SD') %>%
-  dplyr::select(Gene.ID, logFC, P.Value, adj.P.Val, BrainRegion, Name) %>%
+  dplyr::select(ensembl_gene_id, logFC, P.Value, adj.P.Val, BrainRegion, Name) %>%
   dplyr::mutate(BrainRegion = factor(BrainRegion, labels = c('BM_10' = 'FP', 'BM_22' = 'STG', 
@@
   dplyr::filter(Name == 'Control-AD') %>%
-  dplyr::select(genes, logFC, P.Value, adj.P.Val, BrainRegion, Name) %>%
-  dplyr::rename(Tissue = BrainRegion, Gene.ID = genes, Contrast = Name)
+  tidyr::separate(Gene.ID, c('ensembl_gene_id', 'position'), sep = '\\.') %>%
+  dplyr::select(ensembl_gene_id, logFC, P.Value, adj.P.Val, BrainRegion, Name) %>%
+  dplyr::rename(Tissue = BrainRegion, Contrast = Name)
 
-diffexp = rbindlist(diffexp, use.names = T, fill = T, idcol = 'Study') %>%
-  tidyr::separate(Gene.ID, c('ensembl_gene_id','position'),sep = '\\.')
+diffexp = rbindlist(diffexp, use.names = T, fill = T, idcol = 'Study')
 
@@
 
+# Add hgnc_symbols to the resid.mat matrix
+resid.expr = resid.expr %>%
+  left_join(Ensemble2HGNC)
+
 # Add hgnc_symbols to the diffexp matrix
@@
   left_join(Ensemble2HGNC)
+
+diffexp$Direction = 'UP'
+diffexp$Direction[diffexp$logFC < 0] = 'DOWN'
+
+writeLines('Differentially expressed at an fdr of 0.05')
+diffexp %>%
+  filter(adj.P.Val <= 0.05) %>%
+  group_by(Tissue, Direction) %>%
+  summarise(n = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, n)
+
+writeLines('Differentially expressed at an fdr of 0.05 and fold change of 1.2')
+diffexp %>%
+  filter(adj.P.Val <= 0.05, (logFC >= log2(1.2) | logFC <= -log2(1.2))) %>%
+  group_by(Tissue, Direction) %>%
+  summarise(n = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, n)
 ```
@@
   # filter(Tissue %in% c('DLPFC', 'PHG', 'TCX')) %>%
+  filter(ensembl_gene_id != '') %>%
   dplyr::select(ensembl_gene_id, P.Value, Tissue, Contrast) %>%
@@
 
-metaAnal = join_all(list(min.pval, max.pval, fb.pval), type = 'full')
-tmp = metaAnal %>% gather(method, pval, -ensembl_gene_id) %>% group_by(method) %>% filter(pval <= 0.05, method == 'fb.pval') %>%
-  left_join(Ensemble2HGNC) %>%
-  dplyr::select(hgnc_symbol) %>%
-  unique
+# Rank prod based analysis
+expr = resid.expr[!duplicated(resid.expr$ensembl_gene_id),-c(1:2)]
+rownames(expr) = resid.expr$ensembl_gene_id[!duplicated(resid.expr$ensembl_gene_id)]
+
+cl = covariates$Status
+names(cl) = covariates$ID
+cl[cl == 'NCI'] = 'Control'
+cl[cl == 'ND'] = 'Control'
+cl[cl == 'SD'] = 'AD'
+cl = cl[cl %in% c('Control','AD')] %>% droplevels 
+cl.tmp = -as.numeric(cl)+2
+names(cl.tmp) = names(cl)
+
+origin = covariates$BrainRegion %>% as.numeric
+names(origin) = covariates$ID
+origin = origin[names(cl)]
+
+expr = expr[, names(cl)]
+
+rankprod.pfp = RPadvance(expr, cl.tmp, origin, num.perm = 100, gene.names = rownames(expr), plot=T, rand = 123456, huge=T)
+rankprod.pval = apply(rankprod.pfp$pfp, 1, min) %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  dplyr::rename(rankprod.pval = DF)
+
+metaAnal = join_all(list(min.pval, max.pval, fb.pval, rankprod.pval), type = 'full')
+metaAnal$weight = -log10(metaAnal[,-c(1,6)]) %>% rowSums %>% rank
+metaAnal = metaAnal %>% arrange(desc(weight))
+
+writeLines('Number of genes differentially expressed between AD and control at an FDR of 0.05 are')
+tmp = metaAnal %>% 
+  dplyr::select(-weight) %>%
+  gather(method, pval, -ensembl_gene_id) %>% 
+  group_by(method) %>% 
+  filter(pval <= 0.05) %>%
+  dplyr::select(ensembl_gene_id) %>%
+  unique %>%
+  dlply(.(method), .fun = function(x){unique(x$ensembl_gene_id)})
+kable(sapply(tmp, length) %>% as.data.frame)
+
+tmp1 = Venn(tmp)
+
+plotVenn3d(tmp1@IndicatorWeight[-(1),'.Weight'], labels = names(tmp))
+```
+### Heatmap of expression
+```{r heatmap}
+# Get the overlapping gene sets from all three methods
+tmp.expr = t(scale(t(expr)))
+tmp.expr = scale(tmp.expr)
+tmp.expr = tmp.expr[intersect(intersect(tmp$fb.pval, tmp$min.pval), tmp$rankprod.pval),]
+
+tmp.covariates = dplyr::select(covariates, BrainRegion, Status)
+tmp.covariates$Status[tmp.covariates$Status == 'NCI'] = 'Control'
+tmp.covariates$Status[tmp.covariates$Status == 'ND'] = 'Control'
+tmp.covariates$Status[tmp.covariates$Status == 'SD'] = 'AD'
+tmp.covariates = tmp.covariates %>% droplevels
+rownames(tmp.covariates) = covariates$ID
+
+h1 = HeatmapAnnotation(tmp.covariates)
+Heatmap(tmp.expr, top_annotation = h1, show_row_names = F, show_column_names = F,
+        show_row_dend = F, show_column_dend = F)
 ```
 
-```{r plotDE}
-tmp = lapply(DEXP, function(x){
-  x$observed.P.val = p.adjust(x$P.Value, 'fdr')
-  x = arrange(x, observed.P.val)
-  x = mutate(x, expected.P.val = seq(0,1,1/(dim(x)[1]-1)))
-  x$dexp = 0;
-  x$dexp[x$observed.P.val <= 0.05 & x$logFC > 0] = 1;
-  x$dexp[x$observed.P.val <= 0.05 & x$logFC < 0] = -1;
-  dplyr::select(x, observed.P.val, expected.P.val, dexp)
+### Pathway enrichment analysis
+```{r pathway.enrich}
+setsToTest = diffexp %>%
+  filter(adj.P.Val <= 0.05) %>%
+  dlply(.(Tissue, Direction), .fun = function(x){
+    unique(x$ensembl_gene_id)
   })
 
-p = list()
-for (i in 1:length(tmp)){
-  p[[i]] = ggplot(tmp[[i]], aes(x = -log10(expected.P.val), y = -log10(observed.P.val))) + geom_point(aes(color = factor(dexp))) 
-  p[[i]] = p[[i]] + scale_colour_manual(values=c('green','grey','red')) + theme(legend.position="none")
-  x = -log10(tmp[[i]]$expected.P.val)
-  x[4]-x[1]/
-  p[[i]] = p[[i]] + geom_abline(slope=1, intercept=0) + ggtitle(names(tmp)[i])
+tmp1 = rankprod.pfp$pfp[intersect(intersect(tmp$fb.pval, tmp$min.pval), tmp$rankprod.pval),] <= 0.05
+setsToTest$META.UP = rownames(tmp1)[which(tmp1[,1])]
+setsToTest$META.DOWN = rownames(tmp1)[which(tmp1[,2])]
+
+# Convert ensembl id to hgnc_symbols
+setsToTest = lapply(setsToTest, function(x, Ensemble2HGNC){
+  Ensemble2HGNC$hgnc_symbol[Ensemble2HGNC$ensembl_gene_id %in% x] %>%
+    unique()
+}, Ensemble2HGNC)
+
+# Get Background genes
+backgroundGenes = Ensemble2HGNC$hgnc_symbol[Ensemble2HGNC$ensembl_gene_id %in% rownames(expr)] %>% unique()
+
+############################################################################################################
+# Function to perform Fishers enrichment analysis
+fisherEnrichment <- function(genesInSignificantSet, # A character vector of differentially expressed or some significant genes to test
+                             genesInGeneSet, # A character vector of genes in gene set like GO annotations, pathways etc...
+                             genesInBackground # Background genes that are 
+){
+  genesInSignificantSet = intersect(genesInSignificantSet, genesInBackground) # back ground filtering
+  genesInNonSignificantSet = base::setdiff(genesInBackground, genesInSignificantSet)
+  genesInGeneSet = intersect(genesInGeneSet, genesInBackground) # back ground filtering
+  genesOutGeneSet = base::setdiff(genesInBackground,genesInGeneSet)
+
+  pval = fisher.test(
+    matrix(c(length(intersect(genesInGeneSet, genesInSignificantSet)),             
+             length(intersect(genesInGeneSet, genesInNonSignificantSet)),
+             length(intersect(genesOutGeneSet, genesInSignificantSet)),
+             length(intersect(genesOutGeneSet, genesInNonSignificantSet))), 
+           nrow=2, ncol=2), alternative="greater")
+  OR = (length(intersect(genesInGeneSet, genesInSignificantSet)) * length(intersect(genesOutGeneSet, genesInNonSignificantSet))) / (length(intersect(genesInGeneSet, genesInNonSignificantSet)) * length(intersect(genesOutGeneSet, genesInSignificantSet))) %>% as.numeric()
+  return(data.frame(pval = pval$p.value,
+                    ngenes = length(genesInGeneSet),
+                    noverlap = length(intersect(genesInGeneSet, genesInSignificantSet)),
+                    Odds.Ratio = OR,
+                    Genes = paste(intersect(genesInGeneSet, genesInSignificantSet), collapse = '|')))}
+############################################################################################################
+
+############################################################################################################
+#### Get gene sets ####
+# Download enrichr gene sets from synapse
+GL_OBJ = synGet('syn4867851')
+ALL_USED_IDs = c(ALL_USED_IDs, GL_OBJ$properties$id)
+load(GL_OBJ@filePath)
+
+gsets = c("BioCarta_2015", "GO_Biological_Process", "Reactome_2015", "TargetScan_microRNA", "WikiPathways_2015")
+GeneSets.Enrichr = GeneSets[gsets]
+
+# Download AD related gene sets from synapse
+GL_OBJ = synGet('syn4893059');
+ALL_USED_IDs = c(ALL_USED_IDs, GL_OBJ$properties$id)
+load(GL_OBJ@filePath)
+
+GeneSets.CM = list(CellTypeMarkers = GeneSets[grep('Zhang', names(GeneSets))],
+                   AD = GeneSets['AD:GeneticLoci'])
+GeneSets = c(GeneSets.Enrichr, GeneSets.CM)
+
+# Filter gene list
+GeneSets = filterGeneSets(GeneSets, backgroundGenes, minSize = 3, maxSize = 5000)
+
+enrichResults = list()
+for (name in names(setsToTest)){
+  genesInModule = setsToTest[[name]]  
+  enrichResults[[name]] = lapply(GeneSets,
+                                    function(x, genesInModule, genesInBackground, fisherEnrichment, rownameToFirstColumn){
+                                      tmp = as.data.frame(t(sapply(x, fisherEnrichment, genesInModule, genesInBackground)))
+                                      tmp = rownameToFirstColumn(tmp,'GeneSetName')
+                                      return(tmp)
+                                    }, unique(genesInModule), unique(backgroundGenes), fisherEnrichment, rownameToFirstColumn) %>%
+    rbindlist(use.names=TRUE, idcol = 'Category') %>%
+    dplyr::mutate(fdr = p.adjust(pval, 'fdr'))
+  writeLines(paste0('Completed ',name))  
 }
-multiplot(plotlist = p, cols = 4)
-```
 
-```{r uniqGenes}
-GNames = DEXP[[1]]$hgnc_symbol
-for (i in 2: length(DEXP)){
-  GNames = intersect(GNames, DEXP[[i]]$hgnc_symbol)
-}
-GNames = GNames[!is.na(GNames)]
-GNames = GNames[GNames != '']
+# Write results to file
+tmp1 = rbindlist(enrichResults, use.names = TRUE, idcol = 'ComparisonName', fill = TRUE) %>%
+  filter(!is.na(Category))
 
-DEXP = lapply(DEXP, function(dexp, GNames){
-  dexp = dexp %>%
-    dplyr::filter(hgnc_symbol %in% GNames) %>%
-    group_by(hgnc_symbol) %>%
-    dplyr::summarise(P.Value = min(P.Value)) %>%
-    data.frame
-  rownames(dexp) = dexp$hgnc_symbol
-  dexp = dexp[GNames,]
-  }, GNames)
+tmp1$pval = unlist(tmp1$pval)
+tmp1$ngenes = unlist(tmp1$ngenes)
+tmp1$noverlap = unlist(tmp1$noverlap)
+tmp1$Odds.Ratio = unlist(tmp1$Odds.Ratio)
+tmp1$Genes = unlist(tmp1$Genes)
 
-DEXP = mapply(function(dexp, dataName){
-  colnames(dexp) = c('hgnc_symbol', dataName)
-  return(dexp)
-}, DEXP, names(DEXP), SIMPLIFY = F)
-
-DEXP = join_all(DEXP, by = "hgnc_symbol")
-```
-Overall `r dim(DEXP)[1]` genes in `r dim(DEXP)[2]-1` studies are analysed
-### Differential expression analysis
-```{r diffExp, results = 'asis'}
-Adj.P.Value = apply(DEXP[,-(1)], 2, p.adjust, method = "fdr")
-rownames(Adj.P.Value) = DEXP$hgnc_symbol
-
-writeLines('Total number of genes that are differentially expressed at a FDR <= 0.05 are')
-tmp = (data.frame(colSums(Adj.P.Value <= 0.05)))
-colnames(tmp) = 'No. of Genes'
-kable(tmp)
-
-writeLines('Genes that are differentially expressed at a FDR <= 0.05 in 8/8 datasets')
-paste(rownames(Adj.P.Value)[which(rowSums(Adj.P.Value<=0.05)==8)], collapse = ',')
-
-writeLines('Genes that are differentially expressed at a FDR <= 0.05 in 7/8 datasets')
-paste(rownames(Adj.P.Value)[which(rowSums(Adj.P.Value<=0.05)==7)], collapse = ',')
-
-writeLines('Genes that are differentially expressed at a FDR <= 0.05 in 6/6 RNASeq datasets')
-paste(rownames(Adj.P.Value)[which(rowSums(Adj.P.Value[,-(4:5)]<=0.05)==6)], collapse = ',')
-```
-
-### Minimum Pvalue meta analysis
-```{r minP}
-Pval = DEXP
-
-Pmin = data.frame(hgnc_symbols = Pval[,1], Pmin = rowMin(data.matrix(Pval[,-(1)])))
-Pmin$Pmin = p.adjust(Pmin$Pmin, method = 'fdr')
-
-tmp = dplyr::filter(Pmin, Pmin <= 0.05) 
-tmp = data.frame(No.of.Genes = dim(tmp)[1], 
-                 GeneList = tmp %>%
-                   dplyr::select(hgnc_symbols) %>%
-                   unlist %>% paste(collapse = ','))
-writeLines("Genes that are atleast differentially expressed at an FDR <= 0.05 in one study are")
-kable(tmp)
-```
-
-```{r fisher.brown}
-Pfb = data.frame(hgnc_symbols = Pval[,1], 
-                 Pfb = p.adjust(fishersPvalue(Pval[,-(1)]), method = 'fdr'))
-
-Pfb$Pfb = p.adjust(Pfb$Pfb, method = 'fdr')
-
-tmp = dplyr::filter(Pfb, Pfb <= 0.05) 
-tmp = data.frame(No.of.Genes = dim(tmp)[1], 
-                 GeneList = tmp %>%
-                   dplyr::select(hgnc_symbols) %>%
-                   unlist %>% paste(collapse = ','))
-
-writeLines("Genes that are differentially expressed in the meta analysis at an FDR <= 0.05 are")
-kable(tmp)
+# Store logcpm in synapse
+write.table(tmp1, file = 'enrich.tsv', sep = '\t', row.names=F, quote=F)
+OBJ = File('enrich.tsv', 
+           name = 'Pathway Enrichment Results (AD and Control)', 
+           parentId = CODE$properties$id)
+OBJ = synStore(OBJ,
+               used = as.character(c(covariates.id)),
+               executed = thisFile,
+               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
 ```
@@
 
+# Store residual expression in synapse
+write.table(expr, file = 'ALLExpr.tsv', sep = '\t', row.names=F, quote=F)
+OBJ = File('ALLExpr.tsv', 
+           name = 'Residual Expression (AD and Control)', 
+           parentId = CODE$properties$id)
+OBJ = synStore(OBJ,
+               used = as.character(c(covariates.id)),
+               executed = thisFile,
+               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
+
 # Store diffexp in synapse
@@
                activityName = 'MetaAnalysis of AMP-AD reprocessed data')
+
+# Store meta analysis in synapse
+write.table(metaAnal, file = 'ALLMetaAnal.tsv', sep = '\t', row.names=F, quote=F)
+OBJ = File('ALLMetaAnal.tsv', 
+           name = 'Meta Analysis (AD and Control)', 
+           parentId = CODE$properties$id)
+OBJ = synStore(OBJ,
+               used = as.character(c(covariates.id)),
+               executed = thisFile,
+               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
 ```

### 2016-09-02 code/Rmd/InvestigateCoexpIssue.Rmd
Temp commit

@@
 
+library(Vennerable)
 synapseLogin()
+
+options(xtable.type="html")
+
+knitr::opts_chunk$set(
+  echo=FALSE,
+  warning=FALSE,
+  message=FALSE,
+  error = FALSE,
+  tidy = FALSE,
+  cache = FALSE)
 ```
@@
 #### Obtain adjusted FPKM/logcpm residual matrix from synapse
-```{r data.download}
+```{r data.download, cache = F}
 original.residuals.id = c('ROSMAP' = 'syn5581268', 'MAYO' = 'syn6088523', 'MSSM' = 'syn6120103')
@@
 #### Obtain adjusted residual matrix from reprocessed counts in synapse
-```{r data.download2}
+```{r data.download2, cache=F}
 reproc.residuals.id = c('ROSMAP' = 'syn6114451', 'MAYO' = 'syn6128271', 'MSSM' = 'syn6131415')
@@
 save(list = ls(), file = 'investigate.RData')
-obj = File(file = 'investigate.RData', name = 'RNASeq Reprocessing Coexpression Quality Control', parentId = CODE$properties$id)
-obj = synStore(obj, activityName = activityName, used = as.character(all.used.ids), executed = thisFile)
+# system('zip investigate investigate.RData')
+# obj = File('./InvestigateCoexpIssue.Rmd', name = 'RNASeq Reprocessing Coexpression Quality Control', parentId = CODE$properties$id)
+# obj = synStore(obj, activityName = activityName, used = as.character(all.used.ids), executed = thisFile)
 ```

### 2016-09-02 code/Rmd/PredictPMI.Rmd
Minor changes

@@
 ---
-
 ```{r knit2synapse, eval=FALSE}

### 2016-09-02 code/Rmd/ROSMAP_RNASeq_Reprocessed_NonConditionedCogdx.Rmd

@@
                                  parentId ="syn5570291",
-                                 entityName = 'ROSMAP RNASeq Reprocessed Counts 2 (Based on cogdx)')
+                                 entityName = 'ROSMAP RNASeq Reprocessed Counts UnConditioned (Based on cogdx)')
 ```
@@
 # Code
-CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts 2 (Based on cogdx)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts UnConditioned (Based on cogdx)", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-09-02 code/Rmd/ROSMAP_RNASeq_Reprocessed_StarHTSeq.Rmd
Added markdown to process Star-HTSeq aligned ROSMAP data

--- 2016-06-01 code/Rmd/ROSMAP_RNASeq_Reprocessed.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed.Rmd",
+knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_StarHTSeq.Rmd",
                                  parentId ="syn5570291",
-                                 entityName = 'ROSMAP RNASeq Reprocessed Counts (Based on cogdx)')
+                                 entityName = 'ROSMAP RNASeq Reprocessed Counts Star-HTSeq (Based on cogdx)')
 ```
@@
 
+## Requires ggdendro
+devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")
+
 options(xtable.type="html")
@@
 
-thisFileName <- 'ROSMAP_RNASeq_Reprocessed.Rmd'
+thisFileName <- 'ROSMAP_RNASeq_Reprocessed_StarHTSeq.Rmd'
 
@@
 # Download reprocessed counts (paired end)
-COUNT_PE_ID <- 'syn6112813';
-ALL_USED_IDs <- COUNT_PE_ID
-COUNT_PE <- read.table(synGet(COUNT_PE_ID)@filePath, header=T, sep='\t', check.names = F)
-
-# Download reprocessed counts (single end)
-COUNT_SE_ID <- 'syn6117775';
-ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_SE_ID
-COUNT_SE <- read.table(synGet(COUNT_SE_ID)@filePath, header=T, sep='\t', check.names = F)
-
-COUNT = COUNT_PE + COUNT_SE[rownames(COUNT_PE), colnames(COUNT_PE)]
+COUNT_ID <- 'syn7211360';
+ALL_USED_IDs <- COUNT_ID
+COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F)
+COUNT = COUNT[-(1:4), !(colnames(COUNT) %in% c('111_120418','14_120410','323_120501','513_120515','586_120522','665_120530','701_120605'))]
 
@@
 ```
+Coexpression of genes 
+```{r coexp1}
+cr = cor(t(VOOM.GENE_EXPRESSION$E))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
+```
+
 ### Outlier analysis
@@
   VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+```
+
+Coexpression of genes 
+```{r coexp2}
+cr = cor(t(RESIDUAL.GENE_EXPRESSION))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
 ```
@@
 # Code
-CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts (Based on cogdx)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts Star-HTSeq (Based on cogdx)", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-09-03 code/Rmd/ROSMAP_RNASeq_Reprocessed_StarHTSeq_Qntl.Rmd
Added markdown to process ROSMAP reprocessed RNASeq data with quantile normalisation

--- 2016-09-02 code/Rmd/ROSMAP_RNASeq_Reprocessed_StarHTSeq.Rmd
@@
 ---
-title: "Covariate analysis of ROSMAP Reprocessed RNASeq"
+title: "Covariate analysis of ROSMAP Reprocessed RNASeq data aligned using STAR and quantified using HTSeq with quantile normalisation"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_StarHTSeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_StarHTSeq_Qntl.Rmd",
                                  parentId ="syn5570291",
-                                 entityName = 'ROSMAP RNASeq Reprocessed Counts Star-HTSeq (Based on cogdx)')
+                                 entityName = 'ROSMAP RNASeq Reprocessed Counts Star-HTSeq Qntl (Based on cogdx)')
 ```
@@
 
-thisFileName <- 'ROSMAP_RNASeq_Reprocessed_StarHTSeq.Rmd'
+thisFileName <- 'ROSMAP_RNASeq_Reprocessed_StarHTSeq_Qntl.Rmd'
 
@@
 # Initial normalisation of gene expression
-VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T)
+VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T, normalize.method = "quantile")
 ```
 Coexpression of genes 
-```{r coexp1}
+```{r coexp1, cache=FALSE}
 cr = cor(t(VOOM.GENE_EXPRESSION$E))
@@
   # Estimate correlation of random effects
-  VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)
+  VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F, normalize.method = "quantile")
 
@@
 Coexpression of genes 
-```{r coexp2}
+```{r coexp2, cache = F}
 cr = cor(t(RESIDUAL.GENE_EXPRESSION))
@@
 # Code
-CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts Star-HTSeq (Based on cogdx)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts Star-HTSeq Qntl (Based on cogdx)", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-09-03 code/Rmd/ROSMAP_RNASeq.Rmd
Minor modifications

--- 2016-02-09 code/Rmd/ROSMAP_RNASeq.Rmd
@@
 ## Load required libraries
-library(synapseClient)
+library(CovariateAnalysis) # get it from devtools::install_github('th1vairam/CovariateAnalysis@dev')
+library(data.table)
+library(tidyr)
+library(plyr)
+library(dplyr)
+library(stringr)
+
 library(ggplot2)
 library(reshape2)
-library(data.table)
-library(dplyr)
-library(stringr)
 library(limma)
-library(knitr)
 library(Biobase)
-library(GEOquery)
-library(RColorBrewer)
 library(gplots)
@@
 library(biomaRt)
-library(ggplot2)
-library(WGCNA)
-
-## Needs the dev branch
-library(rGithubClient)
 
@@
 devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")
-
-synapseLogin()
-
-# source utility files from ../R/lib folder
-file.sources = list.files('../R/lib',pattern="*.R", full.names=T)
-tmp = sapply(file.sources,source,.GlobalEnv)
 
@@
 
+# Fix error in technical covariates data
+KEY_ID <- 'syn3382527'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = KEY_ID  
+KEY <- read.csv(synGet(KEY_ID)@filePath) %>% 
+  filter(mrna_data == 1) %>%
+  dplyr::select(projid, mrna_id) %>%
+  separate(mrna_id, c('a','b','batch'), sep = '_') %>% 
+  unite(Sampleid, a, b) %>%
+  dplyr::select(-batch) %>%
+  unique
+
 # Match technical and clinical covariates
-METADATA <- dplyr::inner_join(METADATA.CLINICAL,METADATA.TECH)
+METADATA <- METADATA.TECH %>% 
+  dplyr::select(-projid) %>%
+  dplyr::left_join(KEY) %>%
+  dplyr::left_join(METADATA.CLINICAL)
+
+# Pick higher quality RIN batch for sample 492_120515
+METADATA <- METADATA %>%
+  group_by(Sampleid) %>%
+  top_n(1, RINcontinuous)
 ```
@@
 METADATA <- METADATA %>%
+  ungroup %>%
   mutate(Sampleid = paste(Sampleid, Batch, sep = '_')) %>%
@@
   filter(!is.na(pmi)) %>%
-  filter(!(Sampleid %in% c('380_120503_1')))
+  filter(!(Sampleid %in% c('380_120503_1'))) %>%
+  as.data.frame()
 ```
@@
 `r dim(EXPR)[1]` number of genes and `r dim(EXPR)[2]` number of samples are considered for the analysis
+
+Coexpression of genes 
+```{r coexp1, cache=FALSE}
+cr = cor(t(EXPR))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
+```
 
@@
 
+Coexpression of genes 
+```{r coexp2, cache=FALSE}
+cr = cor(t(RESIDUAL.GENE_EXPRESSION[, !(colnames(RESIDUAL.GENE_EXPRESSION) %in% c('ensembl_gene_id','position'))]))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
+```
+
 ### Clustering residual gene expression
@@
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
-```
-```{r decompse.normalise.data4, fig.height=8, fig.width=16, results='asis'}
-tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION[,-(1:2)])))
-WGCNA::plotDendroAndColors(tree, 
-                    colors = labels2colors(COVARIATES.tmp), 
-                    dendroLabels = FALSE, 
-                    abHeight = 0.80, 
-                    main = "Sample dendrogram",
-                    groupLabels = colnames(COVARIATES.tmp))
-dev.off()
 ```

### 2016-09-03 code/Rmd/ROSMAP_RNASeq_Qntl.Rmd
Added markdown for processign ROSMAP FPKM matrix using quantile normalisation

--- 2016-09-03 code/Rmd/ROSMAP_RNASeq.Rmd
@@
                                  parentId ="syn5570291",
-                                 entityName = 'ROSMAP RNASeq (Based on cogdx)')
+                                 entityName = 'ROSMAP RNASeq Qntl (Based on cogdx)')
 ```
@@
 
-thisFileName <- 'ROSMAP_RNASeq.Rmd'
+thisFileName <- 'ROSMAP_RNASeq_Qntl.Rmd'
 
@@
 EXPR <- log2(EXPR+1e-3)
+
+# Between sample quantile normalisation
+EXPR = normalizeBetweenArrays(data.matrix(EXPR), method = 'quantile')
 ```
@@
 # Code
-CODE <- Folder(name = "ROSMAP RNASeq (Based on cogdx)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP RNASeq Qntl (Based on cogdx)", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-09-04 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM.Rmd
Added markdown to process ROSMAP with voom TMM normalisation

--- 2016-06-01 code/Rmd/ROSMAP_RNASeq_Reprocessed.Rmd
@@
 ---
-title: "Covariate analysis of ROSMAP Reprocessed RNASeq"
+title: "Covariate analysis of ROSMAP Reprocessed RNASeq data with TMM"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed.Rmd",
+knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_TMM.Rmd",
                                  parentId ="syn5570291",
-                                 entityName = 'ROSMAP RNASeq Reprocessed Counts (Based on cogdx)')
+                                 entityName = 'ROSMAP RNASeq Reprocessed Counts TMM (Based on cogdx)')
 ```
@@
 
-thisFileName <- 'ROSMAP_RNASeq_Reprocessed.Rmd'
+thisFileName <- 'ROSMAP_RNASeq_Reprocessed_TMM.Rmd'
 
@@
 ```{r initial.voom.normalisation}
+# TMM normalisation
+TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix)
+
 # Initial normalisation of gene expression
-VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T)
+VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
+```
+Coexpression of genes 
+```{r coexp1, cache=FALSE}
+cr = cor(t(VOOM.GENE_EXPRESSION$E))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
 ```
@@
   # Estimate correlation of random effects
-  VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)
+  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
 
@@
   VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+```
+Coexpression of genes 
+```{r coexp2, cache=FALSE}
+cr = cor(t(RESIDUAL.GENE_EXPRESSION))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
 ```
@@
 # Code
-CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts (Based on cogdx)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts TMM (Based on cogdx)", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-09-07 code/Rmd/ROSMAP_RNASeq_Qntl_SVA.Rmd
Added markdown to process FPKM from ROSMAP using quantile normalisation and SVA

--- 2016-09-03 code/Rmd/ROSMAP_RNASeq_Qntl.Rmd
@@
 ---
-title: "Covariate analysis of ROSMAP RNASeq"
+title: "Covariate analysis of ROSMAP RNASeq with Quantile and SVA normalisation"
 author: "Thanneer Perumal"
@@
 library(synapseClient)
+library(knitr)
 
@@
 
-knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Qntl_SVA.Rmd",
                                  parentId ="syn5570291",
-                                 entityName = 'ROSMAP RNASeq Qntl (Based on cogdx)')
+                                 entityName = 'ROSMAP RNASeq Qntl SVA (Based on cogdx)')
 ```
@@
 library(biomaRt)
+library(Rcolorbrewer)
+
+library(sva)
 
@@
 activityName = 'Covariate analysis';
-activityDescription = 'Covariate analysis of ROSMAP RNASeq data';
+activityDescription = 'Covariate analysis of ROSMAP RNASeq data with quantile normalisation and SVA adjustments';
 
-thisFileName <- 'ROSMAP_RNASeq_Qntl.Rmd'
+thisFileName <- 'ROSMAP_RNASeq_Qntl_SVA.Rmd'
 
@@
 ```
-### Normalisation (custom design)
-Fitting a linear model with cogdx, Batch, RINcontinuous, and pmi 
-```{r iterative.normalisation, cache=TRUE, results='asis', echo=FALSE}
-postAdjustCovars <- c('cogdx', 'Batch', 'RINcontinuous', 'pmi')
+### Normalisation (iterative design)
+Since many covariates are correlated, re-normalising FPKM with an iterative design matrix. Here "cogdx" is chosen as the primary variable of interest
+```{r iterativeNorm}
+# Primary variable of interest
+postAdjustCovars = 'cogdx'
 
-writeLines(paste('Using following covariates in the final model',paste(postAdjustCovars,collapse=','),'as fixed effects',sep=' '))
+# Assign residual covariates
+residualSigCovars = preAdjustedSigCovars
+covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
+postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
+
+loopCount = 0 
+while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
+  writeLines(paste('Using following covariates in the model:',
+                   paste(postAdjustCovars, collapse=', '),
+                   'as fixed effects'))
 
@@
                                                 isKeyPlot=TRUE)
+
+  # Add postadjusted covariates (if any)
+  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
+
+  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
+  loopCount = loopCount + 1
+}
+modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
+                  'as fixed effects')
+
+tmp <- paste('Using following covariates in the final model:', modelStr)
+```
+`r tmp`
+### Sanity check
+```{r residualSigCovars.manual, fig.width=20, fig.height=12}
+residualSigCovars[["PC_res"]][[2]]$plotData
 ```
 
-```{r residualSigCovars.manual.ALL, fig.width=15, fig.height=8}
-residualSigCovars[["PC_res"]][[2]]$plotData
+### Estimate surrogate variables using SVA package
+```{r sva}
+MODEL0 = DM1$design[,-grep('cogdx',colnames(DM1$design))] # Get null model by removing variable of interest
+MODEL1 = DM1$design
+
+## Calculate the number of surrogate variables
+# Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
+# SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
+# NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
+# SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
+# SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]
+
+# # Permute and calculate the permuted singular values
+# PERM.SING.VALS <- foreach (i = 1:100,
+#          .combine = rbind,
+#          .packages = c('limma','CovariateAnalysis'),
+#          .export = c('PROCESSED_COUNTS','MODEL1')) %dopar% {
+#   PERM.COUNTS <- t(apply(PROCESSED_COUNTS$filteredExprMatrix, 1, sample, replace = FALSE))
+#   # Estimate voom weights
+#   VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)
+#   
+#   # Fit linear model using new weights and new design
+#   VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)
+#   
+#   # Residuals after normalisation
+#   RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)
+#   
+#   PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
+#   PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
+# }
+# num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])
+
+num.sv2 = num.sv(EXPR, MODEL1, method = 'be', seed = 123456)
+NUM.SV = num.sv2 #min(num.sv1, num.sv2)
+
+## Compute surrogate variables 
+SURR.VAR = sva(EXPR, MODEL1, MODEL0, method = "irw", n.sv = NUM.SV, B = 30)$sv
+colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
+
+# Fit linear model using estimated surrogate variables and postadjust covariates
+FIT = lmFit(EXPR, design = cbind(MODEL1, SURR.VAR))
+
+# Residuals after normalisation
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,EXPR)
 ```
-Adding cogdx back to the residuals
-```{r addVarsBack}
-varsToAddBackIn <- c("cogdx1",  "cogdx2", "cogdx3", "cogdx4", "cogdx5", "cogdx6")
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + t(data.matrix(DESIGN[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
-RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'EnsemblGeneID') %>%
-  tidyr::separate(col = EnsemblGeneID, into = c('ensembl_gene_id','position'), sep = '\\.')
+
+### Residual calculation
+Calculate weighted residuals and add back "cogdx" to the residuals
+```{r varsToAddBack}
+# Add variable of interest back to the residuals
+varsToAddIn = grep("cogdx", colnames(DM1$design), value = T)
+RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
+  FIT$coefficients[,varsToAddIn] %*% t(FIT$design[,varsToAddIn])
 ```
@@
 ```{r coexp2, cache=FALSE}
-cr = cor(t(RESIDUAL.GENE_EXPRESSION[, !(colnames(RESIDUAL.GENE_EXPRESSION) %in% c('ensembl_gene_id','position'))]))
+cr = cor(t(RESIDUAL.GENE_EXPRESSION))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1:2)], scale.=T)
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T)
 
@@
 p
-
-COVARIATES.tmp = data.matrix(COVARIATES)
-COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
 ```
@@
 mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
+
+RESIDUAL.GENE_EXPRESSION <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'ensembl_gene_id.position') %>%
+  tidyr::separate(ensembl_gene_id.position, c('ensembl_gene_id', 'position'), sep = '\\.')
 
@@
 # Code
-CODE <- Folder(name = "ROSMAP RNASeq Qntl (Based on cogdx)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP RNASeq Qntl SVA (Based on cogdx)", parentId = parentId)
 CODE <- synStore(CODE)
@@
 
-# Store covariates
+# Store design matrix
+MODEL1 = rownameToFirstColumn(MODEL1, 'Sampleid_batch')
+write.table(MODEL1, file = 'designMatrix.tsv', sep = '\t', row.names=F, quote=F)
+DES_OBJ = File('designMatrix.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+DES_OBJ = synStore(DES_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                   executed = thisFile, activityDescription = activityDescription)
+
+# Store surrogate variables 
+rownames(SURR.VAR) = rownames(MODEL1)
+SURR.VAR = rownameToFirstColumn(SURR.VAR, 'Sampleid_batch')
+write.table(SURR.VAR, file = 'SurrogateVariable.tsv', sep = '\t', row.names=F, quote=F)
+SV_OBJ = File('SurrogateVariable.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
+SV_OBJ = synStore(SV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                  executed = thisFile, activityDescription = activityDescription)
+
+# Store residualised expression
 write.table(RESIDUAL.GENE_EXPRESSION, file = 'expression.tsv', sep = '\t', row.names=F, quote=F)

### 2016-09-07 code/Rmd/ROSMAP_RNASeq_Qntl_SVA.Rmd
Minor fix

@@
 
-```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
+```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache = FALSE}
 ## It is assumed your working directory is where this file
@@
 library(biomaRt)
-library(Rcolorbrewer)
+library(RColorBrewer)
 

### 2016-09-09 code/Rmd/MSSM_BM_44_RNASeq_Realigned_StarHTSeq.Rmd
Added markdown to process BM_44 samples from MSSM aligned from source FASTQ using star htseq

--- 2016-05-06 code/Rmd/MSSM_BM_44_RNASeq.Rmd
@@
 ---
-title: "Covariates analysis of MSBB RNASeq data of region BM 44"
+title: "Covariates analysis of MSBB RNASeq data of region BM 44 (realigned from source FASTQ with STAR-HTSeq)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_BM_44_RNASeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_BM_44_RNASeq_Realigned_StarHTSeq.Rmd",
                                  parentId ="syn5570248",
-                                 entityName = 'MSBB BM 44 RNASeq')
+                                 entityName = 'MSBB BM 44 RNASeq RealignedFromSource StarHTSeq')
 ```
@@
 
-thisFileName <- 'MSSM_BM_44_RNASeq.Rmd'
+thisFileName <- 'MSSM_BM_44_RNASeq_Realigned_StarHTSeq.Rmd'
 
@@
 # Download expression data
-COUNT_ID <- 'syn5898488'
+COUNT_ID <- 'syn7221166'
 ALL_USED_IDs <- COUNT_ID
-COUNT_OBJ <- synGet(COUNT_ID)
-COUNT <- fread(COUNT_OBJ@filePath, data.table = F)
-rownames(COUNT) = COUNT$Geneid
+COUNT <- read.table(synGet(COUNT_ID)@filePath, quote = '', header = T, sep = '\t')
+
+# Get sample ids
+SampleID = data.frame(SampleID = colnames(COUNT),
+                      ID = colnames(COUNT)) %>% 
+  tidyr::separate(ID, c('A','B','ID', 'C'), sep = '_')
+
+SampleID = SampleID %>%
+  dplyr::select(SampleID, ID) %>%
+  dplyr::mutate(ID = as.numeric(ID))
+
+# # Get sample ids that are to be retained (data from Noam on 29th June) 
+# tmp = read.table('MSSM_SampleIDs.txt', header = T)
+# removeID = setdiff(SampleID$ID, tmp$x)
+# writeLines(paste('Removed',length(removeID),'samples from MSSM'))
+# SampleID = filter(SampleID, ID %in% tmp$x)
 
@@
 METADATA_ID <- 'syn5898489'
-METADATA_OBJ <- synGet(METADATA_ID)
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
-METADATA <- read.table(METADATA_OBJ@filePath,sep='\t',header=T, row.names=1)
+METADATA_TECHNICAL <- read.table(synGet(METADATA_ID)@filePath,sep='\t',header=T, row.names=1) %>%
+  rownameToFirstColumn('OldBarcode')
 
 # Get clinical metadata
-METADATA_TECHNICAL_ID <- 'syn5475828'
-METADATA_TECHNICAL_OBJ <- synGet(METADATA_TECHNICAL_ID)
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECHNICAL_ID
-METADATA_TECHNICAL <- read.xlsx(METADATA_TECHNICAL_OBJ@filePath, sheetIndex = 1)
+METADATA_CLINICAL_ID <- 'syn5475828'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CLINICAL_ID
+METADATA_CLINICAL <- read.xlsx(synGet(METADATA_CLINICAL_ID)@filePath, sheetIndex = 1)
 
 # Manually fix 4 sample IDs
-METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1132), "BM36.Fnl"] = 397
-METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1117), "BM22.Fnl"] = 144
-METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1320), "BM22.Fnl"] = 159
-METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1475), "BM22.Fnl"] = 186
+METADATA_CLINICAL[which(METADATA_CLINICAL$BB == 1132), "BM36.Fnl"] = 397
+METADATA_CLINICAL[which(METADATA_CLINICAL$BB == 1117), "BM22.Fnl"] = 144
+METADATA_CLINICAL[which(METADATA_CLINICAL$BB == 1320), "BM22.Fnl"] = 159
+METADATA_CLINICAL[which(METADATA_CLINICAL$BB == 1475), "BM22.Fnl"] = 186
 
-METADATA = left_join(METADATA %>%
+METADATA = left_join(METADATA_TECHNICAL %>%
                        dplyr::mutate(ID = gsub('hB_RNA_','',NewBarcode),
                                      ID = as.numeric(ID)),
-                     METADATA_TECHNICAL %>%
-                       dplyr::select(BB, BM.44.Barcode, BM36.Fnl, BM22.Fnl, BM10.Fnl, PMI, RACE, 
+                     METADATA_CLINICAL %>%
+                       dplyr::select(BB, BM.44.Barcode, PMI, RACE, 
                                      AOD, CDR, SEX, NP.1, PlaqueMean, bbscore) %>%
-                       dplyr::rename(BM_44 = BM.44.Barcode,
-                                     BM_10 = BM10.Fnl,
-                                     BM_22 = BM22.Fnl,
-                                     BM_36 = BM36.Fnl) %>%
+                       dplyr::rename(BM_44 = BM.44.Barcode) %>%
                        tidyr::gather(BrainRegion, ID, -BB, -PMI, -RACE, 
@@
                        dplyr::mutate(ID = as.numeric(ID))) %>%
-  dplyr::filter(BrainRegion == 'BM_44')
-rownames(METADATA) = METADATA$NewBarcode
+  inner_join(SampleID) %>%
+  filter(!is.na(SampleID)) %>%
+  unique
+rownames(METADATA) = METADATA$SampleID
 
 # Filter data
-ind <- intersect(rownames(METADATA), colnames(COUNT))
-ind <- setdiff(ind, 'hB_RNA_10622')
+ind <- intersect(METADATA$SampleID, colnames(COUNT))
+indRemoved <- setdiff(colnames(COUNT), METADATA$SampleID)
+# ind <- setdiff(ind, 'hB_RNA_10622')
 METADATA <- METADATA[ind,]
@@
 
-METADATA$Dx = 'Control'
-METADATA$Dx[METADATA$CDR > 1] = 'AD'
-
-FactorCovariates <- c('RACE','SEX','Dx','Batch')
+FactorCovariates <- c('RACE','SEX','CDR','Batch')
 ContCovariates <- c('PMI','AOD','PlaqueMean','RIN', 'RIN2')
@@
 ```
+Coexpression of genes
+```{r coexp1}
+cr = cor(t(VOOM.GENE_EXPRESSION$E))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
+```
+
 ### Outlier analysis
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Dx, shape=SEX, size=PMI))
+p <- p + geom_point(aes(color=CDR, shape=SEX, size=PMI))
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
@@
 # Primary variable of interest
-postAdjustCovars = 'Dx'
+postAdjustCovars = 'CDR'
 
@@
 # Add variable of interest back to the residuals
-varsToAddIn = c( "DxControl","DxAD")
+varsToAddIn = c("CDR0","CDR0.5","CDR1","CDR2","CDR3","CDR4","CDR5")
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
@@
 ```
-
+Coexpression of genes
+```{r coexp2}
+cr = cor(t(RESIDUAL.GENE_EXPRESSION))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
+```
 ### Clustering residual data
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=factor(Batch), shape=as.factor(Dx), size=as.numeric(PMI)))
+p <- p + geom_point(aes(color=factor(Batch), shape=as.factor(CDR), size=as.numeric(PMI)))
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
@@
 # Define biomart object
-mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(host = "jul2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl", biomart = "ENSEMBL_MART_ENSEMBL")
+
+RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'GeneID') %>%
+  dplyr::mutate(gene.id = GeneID) %>%
+  tidyr::separate(GeneID, c('ensembl_gene_id','position'),sep = '\\.')
 
@@
 Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
-                       filters = "ensembl_gene_id", values = rownames(RESIDUAL.GENE_EXPRESSION),
+                       filters = "ensembl_gene_id", values = RESIDUAL.GENE_EXPRESSION$ensembl_gene_id,
                        mart = mart)
-RESIDUAL.GENE_EXPRESSION <- left_join(rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id"),
-                                      Ensemble2HGNC) %>%
-  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+RESIDUAL.GENE_EXPRESSION <- left_join(RESIDUAL.GENE_EXPRESSION, Ensemble2HGNC) %>%
+  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
 ```
+
 ```{r summary, results='asis', echo=FALSE}
 writeLines('Number of samples in each category are')
-tmp = as.data.frame(summary(COVARIATES$Dx))
+tmp = as.data.frame(summary(COVARIATES$CDR))
 colnames(tmp) = 'No. of samples'
@@
 contrast = colnames(VOOM.ADJUSTED.FIT$coefficients) %>%
-  grep('Dx',.,value=T) %>%
+  grep('CDR',.,value=T) %>%
   combn(2) %>%
@@
   DE[[i]] = topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
-  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id") %>%
+  DE[[i]] = rownameToFirstColumn(DE[[i]], "gene.id") %>%
+    dplyr::mutate(GeneID = gene.id) %>%
+    tidyr::separate(GeneID, c('ensembl_gene_id','position'),sep = '\\.') %>%
     left_join(Ensemble2HGNC)
   p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
-  p[[i]] <- p[[i]] + ggtitle(gsub('TISSUE.Dx','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
+  p[[i]] <- p[[i]] + ggtitle(gsub('TISSUE.CDR','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
   p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red')
@@
 # Code
-CODE <- Folder(name = "MSBB BM 44 RNASeq", parentId = parentId)
+CODE <- Folder(name = "MSBB BM 44 RNASeq  RealignedFromSource StarHTSeq", parentId = parentId)
 CODE <- synStore(CODE)
@@
 # Expression
-expr = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'EnsembleIDs')
-write.table(expr, file='expression.tsv', sep = '\t', row.names=F, quote=F)
+write.table(RESIDUAL.GENE_EXPRESSION, file='expression.tsv', sep = '\t', row.names=F, quote=F)
 obj = File('expression.tsv', 
-           name = 'Voom Adjusted Weighted Residual Expression (Dx added)', 
+           name = 'Voom Adjusted Weighted Residual Expression (CDR added)', 
            parentId = CODE$properties$id)

### 2016-09-10 code/Rmd/MSSM_BM_44_RNASeq_Reprocessed_LillyPipeline.Rmd
Added markdown to process BM 44 in MSSM (reprocessed by eli lilly)

--- 2016-09-09 code/Rmd/MSSM_BM_44_RNASeq_Realigned_StarHTSeq.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_BM_44_RNASeq_Realigned_StarHTSeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_BM_44_RNASeq_Reprocessed_LillyPipeline.Rmd",
                                  parentId ="syn5570248",
-                                 entityName = 'MSBB BM 44 RNASeq RealignedFromSource StarHTSeq')
+                                 entityName = 'MSBB BM 44 RNASeq Reprocessed LillyPipeline')
 ```
@@
 
-thisFileName <- 'MSSM_BM_44_RNASeq_Realigned_StarHTSeq.Rmd'
+thisFileName <- 'MSSM_BM_44_RNASeq_Reprocessed_LillyPipeline.Rmd'
 
@@
 # Download expression data
-COUNT_ID <- 'syn7221166'
-ALL_USED_IDs <- COUNT_ID
-COUNT <- read.table(synGet(COUNT_ID)@filePath, quote = '', header = T, sep = '\t')
+COUNT_B1_ID <- 'syn6114414'
+ALL_USED_IDs <- COUNT_B1_ID
+COUNT_B1 <- read.table(synGet(COUNT_B1_ID)@filePath, quote = '', header = T, sep = '\t')
+
+COUNT_B2_3_ID <- 'syn6125203'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_B2_3_ID
+COUNT_B2_3 <- read.table(synGet(COUNT_B2_3_ID)@filePath, quote = '', header = T, sep = '\t')
+
+# Download expression data
+COUNT_B4_ID <- 'syn6130073'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_B4_ID
+COUNT_B4 <- read.table(synGet(COUNT_B4_ID)@filePath, quote = '', header = T, sep = '\t')
+
+# Remove samples in Batch 2&3 if they present in batch 4
+COUNT_B2_3 = COUNT_B2_3[,setdiff(colnames(COUNT_B2_3), colnames(COUNT_B4))]
+
+COUNT = cbind(COUNT_B1, COUNT_B2_3[rownames(COUNT_B1),], COUNT_B4[rownames(COUNT_B1),])
 
@@
                  ID = colnames(COUNT)) %>% 
-  tidyr::separate(ID, c('A','B','ID', 'C'), sep = '_')
+  tidyr::separate(ID, c('A','B','ID'), sep = '_') 
+SampleID$ID[SampleID$ID == 'HISEQ'] = SampleID$B[SampleID$ID == 'HISEQ']
 
@@
                      METADATA_CLINICAL %>%
-                       dplyr::select(BB, BM.44.Barcode, PMI, RACE, 
+                       dplyr::select(BB, BM.44.Barcode, BM36.Fnl, BM22.Fnl, BM10.Fnl, PMI, RACE, 
                                      AOD, CDR, SEX, NP.1, PlaqueMean, bbscore) %>%
-                       dplyr::rename(BM_44 = BM.44.Barcode) %>%
+                       dplyr::rename(BM_44 = BM.44.Barcode,
+                                     BM_10 = BM10.Fnl,
+                                     BM_22 = BM22.Fnl,
+                                     BM_36 = BM36.Fnl) %>%
                        tidyr::gather(BrainRegion, ID, -BB, -PMI, -RACE, 
@@
   filter(!is.na(SampleID)) %>%
-  unique
+  unique %>%
+  filter(BrainRegion == 'BM_44', !(SampleID %in% c('hB_RNA_8565', 'hB_RNA_4919', 'hB_RNA_9005')))
 rownames(METADATA) = METADATA$SampleID
@@
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
-# p <- p + geom_text(aes(label= ID), size=4, hjust=1)
+p <- p + geom_text(aes(label= ID), size=4, hjust=0)
 p
@@
 # Code
-CODE <- Folder(name = "MSBB BM 44 RNASeq  RealignedFromSource StarHTSeq", parentId = parentId)
+CODE <- Folder(name = 'MSBB BM 44 RNASeq Reprocessed LillyPipeline', parentId = parentId)
 CODE <- synStore(CODE)

### 2016-09-12 code/Rmd/MSSM_BM_36_RNASeq_Realigned_StarHTSeq.Rmd
Added markdown to process Star Htseq aligned BM 36

--- 2016-09-09 code/Rmd/MSSM_BM_44_RNASeq_Realigned_StarHTSeq.Rmd
@@
 ---
-title: "Covariates analysis of MSBB RNASeq data of region BM 44 (realigned from source FASTQ with STAR-HTSeq)"
+title: "Covariates analysis of MSBB RNASeq data of region BM 36 (realigned from source FASTQ with STAR-HTSeq)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_BM_44_RNASeq_Realigned_StarHTSeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_BM_36_RNASeq_Realigned_StarHTSeq.Rmd",
                                  parentId ="syn5570248",
-                                 entityName = 'MSBB BM 44 RNASeq RealignedFromSource StarHTSeq')
+                                 entityName = 'MSBB BM 36 RNASeq RealignedFromSource StarHTSeq')
 ```
@@
 activityName = 'Covariate analysis';
-activityDescription = 'Covariate analysis of MSBB BM 44 RNASeq samples';
+activityDescription = 'Covariate analysis of MSBB BM 36 RNASeq samples';
 
-thisFileName <- 'MSSM_BM_44_RNASeq_Realigned_StarHTSeq.Rmd'
+thisFileName <- 'MSSM_BM_36_RNASeq_Realigned_StarHTSeq.Rmd'
 
@@
 ```
-Data were obtained from AMP-AD portal of MSSM, and following brain regions were analysed for differential expression between case and control:
-  | Region | Name                        | 
-  | :----- | --------------------------- | 
-  | BM_44  | inferior frontal Gyrus (IFG)|
-
-* BM 44: pars opercularis (of the inferior frontal gyrus), and it refers to a subdivision of the frontal region of cerebral cortex.
 
@@
 # Download expression data
-COUNT_ID <- 'syn7221166'
+COUNT_ID <- 'syn7224673'
 ALL_USED_IDs <- COUNT_ID
@@
                      METADATA_CLINICAL %>%
-                       dplyr::select(BB, BM.44.Barcode, PMI, RACE, 
+                       dplyr::select(BB, BM36.Fnl, PMI, RACE, 
                                      AOD, CDR, SEX, NP.1, PlaqueMean, bbscore) %>%
-                       dplyr::rename(BM_44 = BM.44.Barcode) %>%
+                       dplyr::rename(BM_36 =  BM36.Fnl) %>%
                        tidyr::gather(BrainRegion, ID, -BB, -PMI, -RACE, 
@@
   filter(!is.na(SampleID)) %>%
+  filter(!(ID %in% c(335:341, 10567, 10622, 10577))) %>%
   unique
@@
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
-# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
+p <- p + geom_text(aes(label= ID), size=4, hjust=1)
 p
@@
 # Code
-CODE <- Folder(name = "MSBB BM 44 RNASeq  RealignedFromSource StarHTSeq", parentId = parentId)
+CODE <- Folder(name = "MSBB BM 36 RNASeq  RealignedFromSource StarHTSeq", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-09-16 code/Rmd/curateDREAM5DataSets.Rmd
Added markdown to process DREAM5 datasets

---
title: "Format DREAM5 network prediction data for reuse with metanetwork pipeline"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)
library(knitr)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./curateDREAM5DataSets.Rmd",
                                 parentId ="syn5569099",
                                 entityName = 'DREAM5 Networks')
```
#### Initialise libraries
```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(synapseClient)
library(knitr)
library(githubr)

library(CovariateAnalysis)
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(githubr) ## Needs the dev branch

synapseLogin()

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5569099';
activityName = 'Curate DREAM5 networks';
activityDescription = 'Curate DREAM5 data for reuse';

thisFileName <- 'curateDREAM5DataSets.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='synNet')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))

# Create folder to store results
FOLD = Folder(name = 'DREAM5 Networks', parentId = parentId)
FOLD = synStore(FOLD)
```
#### Extract data and network
```{r extractNet, cache=FALSE}
all.folders = synQuery('select * from Folder where parentId == "syn2787212"')
all.expr.obj = dlply(all.folders, .(Folder.id), .fun = function(x){
  all.files = synQuery(paste0('select * from File where parentId == "',x$Folder.id,'"'))

  # Get expression data
  ind.expr = grep('expression', all.files$File.name)
  expr = read.table(synGet(all.files$File.id[ind.expr])@filePath, sep = '\t', header = T, fill=NA)

  # Get gene names
  ind.genes = grep('gene', all.files$File.name)
  genes = read.table(synGet(all.files$File.id[ind.genes])@filePath, sep = '\t', header = F, fill=NA)

  colnames(expr) = genes$V2
  rownames(expr) = paste0('EXPR',1:dim(expr)[1])
  expr = t(expr)

  # Store expression matrix in synapse
  fold.obj = Folder(name = x$Folder.name, parentId = FOLD@properties$id)
  fold.obj = synStore(fold.obj)

  write.table(expr %>% rownameToFirstColumn('GeneNames'), 'Expression.tsv',sep = '\t',row.names = F, quote=F)
  expr.obj = File('Expression.tsv', name = 'Expression Matrix', parentId = fold.obj@properties$id)
  expr.obj = synStore(expr.obj, activityName = activityName, activityDescription = activityDescription,
                      executed = thisFile, used = all.files$File.id)
})
```

### 2016-09-16 code/Rmd/synNetGen.Rmd
Minor modifications

@@
 #### Extract network
-```{r extractNet}
+```{r extractNet, cache=FALSE}
 # Extract settings text file
@@
 
-netSize = c(a = 1e1, b = 1e2, c = 1e3, d = 1e4)
-for (net in names(netSize))){
+settings.txt[grep('ssDualKnockouts', settings.txt)] = "ssDualKnockouts = 0"
+settings.txt[grep('ssMultifactorial', settings.txt)] = "ssMultifactorial = 0"
+settings.txt[grep('tsDREAM4TimeSeries', settings.txt)] = "tsDREAM4TimeSeries = 0"
+settings.txt[grep('simulateSDE', settings.txt)] = "simulateSDE = 1"
+
+netSize = c(a = 1e2, b = 1500)
+for (net in names(netSize)){
   # Number of minimum regulators
-  minReg = max(5, round(netSize[[net]]*0.5))
+  minReg = max(5, round(netSize[[net]]*0.3))
 
@@
 
-  # Extract 10 subnetworks of size n from ecoli data
+  # Extract 1 subnetworks of size n from ecoli data
   system(paste('java','-jar','../../../gnw/gnw-3.1.2b.jar',
@@
                  '../../../gnw/sandbox/ecoli_transcriptional_network_regulonDB_6_7.tsv',
-                 '--random-seed','--greedy-selection','--num-subnets=10',
+                 '--random-seed','--greedy-selection','--num-subnets=1',
                  paste0('--subnet-size=',netSize[[net]]),'--output-net-format=4',
@@
 
-  # Change settings for simulations  
-  tmp = settings.txt
-  tmp[grep('outputDirectory = ', tmp)] = paste('outputDirectory =',outputDir)
-  tmp[grep('ssDualKnockouts',tmp)] = "ssDualKnockouts = 0"
-  tmp[grep('tsDREAM4TimeSeries',tmp)] = "tsDREAM4TimeSeries = 0"
-
-  con = file(paste0(outputDir,'/settings.txt'))
-  writeLines(tmp, con)
-  close(con)
-
+  # # Change settings for simulations  
+  # tmp = settings.txt
+  # tmp[grep('outputDirectory = ', tmp)] = paste('outputDirectory =',outputDir)
+  # tmp[grep('ssDualKnockouts',tmp)] = "ssDualKnockouts = 0"
+  # tmp[grep('tsDREAM4TimeSeries',tmp)] = "tsDREAM4TimeSeries = 0"
+  # 
+  # con = file(paste0(outputDir,'/settings.txt'))
+  # writeLines(tmp, con)
+  # close(con)
+  # 
   # Simulate from extracted networks
@@
     extractData <- function(pat,allfiles, file.name){
-      read.table(allfiles[grep(paste0(file.name,pat),allfiles)], header=T, sep = '\t', quote='')
+      if (length(grep(paste0(file.name,pat),allfiles)) != 0)
+        return(read.table(allfiles[grep(paste0(file.name,pat),allfiles)], header=T, sep = '\t', quote=''))
+      else
+        return(NULL)
     }
@@
 
-    # Set private acl to the gold standard file
-    resAcc = ResourceAccess(principalId = 3319864, 
-                            accessType = c('CREATE', 'READ', 'UPDATE', 'DELETE', 'CHANGE_PERMISSIONS', 'DOWNLOAD', 'UPLOAD', 'PARTICIPATE', 'SUBMIT', 'READ_PRIVATE_SUBMISSION', 'UPDATE_SUBMISSION', 'DELETE_SUBMISSION', 'TEAM_MEMBERSHIP_UPDATE', 'SEND_MESSAGE', 'CHANGE_SETTINGS'))
-    acl = AccessControlList(id = OBJ$properties$id, resourceAccess = ResourceAccessList(resAcc))
-    tmp = synCreateEntityACL(acl)
+    # # Set private acl to the gold standard file
+    # resAcc = ResourceAccess(principalId = 3319864, 
+    #                         accessType = c('CREATE', 'READ', 'UPDATE', 'DELETE', 'CHANGE_PERMISSIONS', 'DOWNLOAD', 'UPLOAD', 'PARTICIPATE', 'SUBMIT', 'READ_PRIVATE_SUBMISSION', 'UPDATE_SUBMISSION', 'DELETE_SUBMISSION', 'TEAM_MEMBERSHIP_UPDATE', 'SEND_MESSAGE', 'CHANGE_SETTINGS'))
+    # acl = AccessControlList(id = OBJ$properties$id, resourceAccess = ResourceAccessList(resAcc))
+    # tmp = synCreateEntityACL(acl)
 

### 2016-09-18 code/Rmd/CMC_DLPFC_StarAligned_TMM.Rmd
Added markdown to process DLPFC RNASeq data reprocessed and aligned with STAR pipeline

---
title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (DLPFC with TMM normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./CMC_DLPFC_StarAligned_TMM.Rmd",
                                 parentId ="syn5569976",
                                 entityName = 'CMC - DLPFC - STAR Reprocessed - with TMM')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get it from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(reshape2)
library(limma)
library(Biobase)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5569976';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM normalisation (DLPFC)';

thisFileName <- 'CMC_DLPFC_StarAligned_TMM.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download reprocessed counts (DLPFC)
COUNT_ID <- 'syn7217916';
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F, nrows = 56632) %>%
  dplyr::select(-chr, -start, -end, -strand, -length)
rownames(COUNT) = COUNT$geneID
COUNT$geneID = NULL
COUNT = data.matrix(COUNT)

# Get merged metadata
METADATA_ID <- 'syn2299154'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA <- read.csv(synGet(METADATA_ID)@filePath, fill = T, header = F)

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Ethnicity", "Age of Death", "PMI (in hours)", 
                  "pH", "Dx", "CDR", "CauseOfDeath", "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                  "DLPFC_RNA_isolation: RNA Isolation Batch", "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?",
                  "DLPFC_RNA_report: Library Batch")

# Filter metadata 
ind = as.character(METADATA[1,] %>% unlist) %in% metadataVarOfInterest
METADATA = METADATA[,ind]
colnames(METADATA) = gsub('[^[:alnum:]]','_',as.character(METADATA[1,] %>% unlist))

METADATA = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, PMI__in_hours_, pH, Dx, CDR, CauseOfDeath,
                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RNA_Isolation_Batch,
                DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Library_Batch) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                IsolationBatch = DLPFC_RNA_isolation__RNA_Isolation_Batch,
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_, 
                LibraryBatch = DLPFC_RNA_report__Library_Batch) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch),
                IsolationBatch = as.factor(IsolationBatch))
```

### Data preprocessing
```{r preprocess.data, echo=TRUE}
# Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
METADATA <- METADATA %>%
  filter(SampleID %in% colnames(COUNT)) %>%
  filter(IsolationExclude == 0, ReportExclude == 0) %>%
  filter(!is.na(PMI)) %>%
  filter(!is.na(Gender)) %>%
  filter(!is.na(Ethnicity)) %>%
  filter(!is.na(RIN)) %>%
  filter(!(SampleID %in% c('MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_60', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_361', 
                           'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_389', 'PENN_RNA_PFC_56', 'PENN_RNA_PFC_59'))) %>%
  as.data.frame()
```

```{r preprocess.data1, results='asis'}
# Match covariates to expression data
indToRetain = intersect(METADATA$SampleID, colnames(COUNT))
removedIDs = setdiff(colnames(COUNT), METADATA$SampleID)

COUNT = COUNT[,indToRetain]

rownames(METADATA) = METADATA$SampleID
METADATA = METADATA[indToRetain,]

METADATA$Dx %>% 
  table() %>% 
  as.data.frame %>% 
  setnames(c('Dx','Total.Number.of.Samples')) %>% 
  kable()
```
Following sample are removed `r paste(removedIDs, collapse = ',')`

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Institution", "Gender", "Ethnicity", "CDR", "CauseOfDeath", "IsolationBatch", "LibraryBatch", "Dx")
ContCovariates <- c("Age_of_Death", "PMI", "pH", "RIN")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES <- data.frame(lapply(COVARIATES,function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'\\+','')})}))
rownames(COVARIATES) <- METADATA$SampleID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.5, row.width=0.1)
```
Explore metatdata
```{r dataExplore, fig.width = 10, fig.height = 10}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = Dx, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = Dx, y = Age_of_Death)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = Dx, y = PMI)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme

# pH
p[[4]] = ggplot(COVARIATES, aes(x = Dx, y = pH)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('pH') + my.theme

multiplot(plotlist = p, cols = 2)
```
### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.
```{r cpmnormalisation}
genesToAnalyze = dlply(METADATA, .(Dx), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID %>% droplevels()],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)
genesToAnalyze = unlist(genesToAnalyze) %>% unique()

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation}
# TMM normalisation
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix)

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
```
Coexpression of genes 
```{r coexp1, cache=FALSE}
cr = cor(t(VOOM.GENE_EXPRESSION$E))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Dx, shape=Institution, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p

# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx','Institution')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Dx','Institution')]))
```

```{r temp, include = F}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```
Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "Dx" is chosen as the primary variable of interest
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = 'RIN';#'Dx'

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
                                                                          c(postAdjustCovars))]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

# Add primary variable to the list (in case not chosen by iterative method)
postAdjustCovars = unique(c(postAdjustCovars, 'Dx'))

writeLines(paste('Using following covariates in the model:',
                 paste(postAdjustCovars, collapse=', '),
                 'as fixed effects'))

# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

# Estimate voom weights
VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)

# Fit linear model using new weights and new design
VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

# Residual covariates to choose from
residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars))

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES[, residCovars, drop=F], 
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Dx, shape = Institution, size=RIN))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p

# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx','Institution')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Dx','Institution')]))
```

```{r temp1, include=F}
dev.off()
gc()
```

```{r get.gene.symbols, cache=TRUE, eval=T}
backgroundGenes = data.frame(ensembl_gene_id = rownames(COUNT))

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "jul2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

# Add gene names
RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=8, fig.width=16}
# Fit contrast
contrast = makeContrasts(contrasts=c("DxControl-DxAFF", "DxControl-DxBP",
                                     "DxControl-DxSCZ", "DxAFF-DxBP",
                                     "DxAFF-DxSCZ", "DxBP-DxSCZ"),
                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:6, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
    rownameToFirstColumn('ensembl_gene_id') %>%
    left_join(Ensemble2HGNC) %>%
    dplyr::select(ensembl_gene_id, hgnc_symbol, logFC, AveExpr, t, P.Value, adj.P.Val, B)
}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC)
names(DE) = colnames(contrast)

DE = DE %>% rbindlist(idcol = 'Comparison')

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))

kable(full_join(tmp,tmp1))

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
p = p + facet_grid(.~Comparison, scales = 'fixed')
p
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "CMC - DLPFC - STAR Reprocessed - with TMM", parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'CMC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('CMC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store logCPM
write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'CMC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_DLPFC_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                     executed = thisFile, activityDescription = activityDescription)

# Store design matrix
write.table(DM1$design, file = 'CMC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('CMC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)

# Store residual gene expression
write.table(RESIDUAL.GENE_EXPRESSION, file = 'CMC_DLPFC_Expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('CMC_DLPFC_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store differential expression results
write.table(DE, file = 'CMC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_DLPFC_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
```
|  *Results*                |  *SynapseID*                     |
|  -------                  |   ---------                      |
|  Covariates               |  `r COV_OBJ$properties$id`       |
|  Raw counts (log CPM)     |  `r COUNT_OBJ$properties$id`     |
|  Design Matrix            |  `r DM_OBJ$properties$id`        |
|  Residual Expression      |  `r EXP_OBJ$properties$id`       |
|  Differential Expression  |  `r DEXP_OBJ$properties$id`      |

### 2016-09-19 code/Rmd/CMC_DLPFC_StarAligned_TMM.Rmd
Minor modifications

@@
                                  parentId ="syn5569976",
-                                 entityName = 'CMC - DLPFC - STAR Reprocessed - with TMM')
+                                 entityName = 'DLPFC - STAR Reprocessed - with TMM')
 ```
@@
 # Add primary variable to the list (in case not chosen by iterative method)
-postAdjustCovars = unique(c(postAdjustCovars, 'Dx'))
+postAdjustCovars = unique(c('Dx', postAdjustCovars))
 
@@
 # Code
-CODE <- Folder(name = "CMC - DLPFC - STAR Reprocessed - with TMM", parentId = parentId)
+CODE <- Folder(name = "DLPFC - STAR Reprocessed - with TMM", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-09-19 code/Rmd/CMC_ACC_StarAligned_TMM.Rmd
Added markdown to process STAR aligned ACC samples

--- 2016-09-19 code/Rmd/CMC_DLPFC_StarAligned_TMM.Rmd
@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (DLPFC with TMM normlaisation)"
+title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (ACC with TMM normlaisation)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_DLPFC_StarAligned_TMM.Rmd",
+knit2synapse::knitToFolderEntity(file = "./CMC_ACC_StarAligned_TMM.Rmd",
                                  parentId ="syn5569976",
-                                 entityName = 'DLPFC - STAR Reprocessed - with TMM')
+                                 entityName = 'ACC - STAR Reprocessed - with TMM')
 ```
@@
 activityName = 'Covariate adjustments';
-activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM normalisation (DLPFC)';
+activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM normalisation (ACC)';
 
-thisFileName <- 'CMC_DLPFC_StarAligned_TMM.Rmd'
+thisFileName <- 'CMC_ACC_StarAligned_TMM.Rmd'
 
@@
 ```{r download.data, cache=TRUE}
-# Download reprocessed counts (DLPFC)
-COUNT_ID <- 'syn7217916';
+# Download reprocessed counts (ACC)
+COUNT_ID <- 'syn7217917';
 ALL_USED_IDs <- COUNT_ID
@@
 metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Ethnicity", "Age of Death", "PMI (in hours)", 
-                  "pH", "Dx", "CDR", "CauseOfDeath", "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
-                  "DLPFC_RNA_isolation: RNA Isolation Batch", "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?",
-                  "DLPFC_RNA_report: Library Batch")
+                          "pH", "Dx", "CDR", "CauseOfDeath", "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
+                          "ACC_RNA_isolation: RNA Isolation Batch", "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?",
+                          "ACC_RNA_report: Library Batch")
 
@@
   dplyr::select(Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, PMI__in_hours_, pH, Dx, CDR, CauseOfDeath,
-                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RNA_Isolation_Batch,
-                DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Library_Batch) %>%
+                ACC_RNA_isolation__Exclude_, ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RNA_Isolation_Batch,
+                ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_, ACC_RNA_report__Library_Batch) %>%
   dplyr::rename(PMI = PMI__in_hours_, 
-                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
-                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
-                IsolationBatch = DLPFC_RNA_isolation__RNA_Isolation_Batch,
-                RIN = DLPFC_RNA_isolation__RIN, 
-                ReportExclude = DLPFC_RNA_report__Exclude_, 
-                LibraryBatch = DLPFC_RNA_report__Library_Batch) %>%
+                IsolationExclude = ACC_RNA_isolation__Exclude_, 
+                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
+                IsolationBatch = ACC_RNA_isolation__RNA_Isolation_Batch,
+                RIN = ACC_RNA_isolation__RIN, 
+                ReportExclude = ACC_RNA_report__Exclude_, 
+                LibraryBatch = ACC_RNA_report__Library_Batch) %>%
   dplyr::filter(SampleID %in% colnames(COUNT)) %>%
@@
   filter(!is.na(RIN)) %>%
-  filter(!(SampleID %in% c('MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_60', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_361', 
-                           'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_389', 'PENN_RNA_PFC_56', 'PENN_RNA_PFC_59'))) %>%
+  filter(!(SampleID %in% c('MSSM_RNA_ACC_250', 'MSSM_RNA_ACC_170'))) %>%
   as.data.frame()
@@
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
-p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
 p
@@
 # Code
-CODE <- Folder(name = "DLPFC - STAR Reprocessed - with TMM", parentId = parentId)
+CODE <- Folder(name = "ACC - STAR Reprocessed - with TMM", parentId = parentId)
 CODE <- synStore(CODE)
@@
 COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
-write.table(COVARIATES, file = 'CMC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
-COV_OBJ = File('CMC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+write.table(COVARIATES, file = 'CMC_ACC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
+COV_OBJ = File('CMC_ACC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
 COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 # Store logCPM
-write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'CMC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('CMC_DLPFC_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
+write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'CMC_ACC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
+COUNT_OBJ = File('CMC_ACC_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
 COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 # Store design matrix
-write.table(DM1$design, file = 'CMC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('CMC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+write.table(DM1$design, file = 'CMC_ACC_Design.tsv', sep = '\t', row.names=F, quote=F)
+DM_OBJ = File('CMC_ACC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
 DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 # Store residual gene expression
-write.table(RESIDUAL.GENE_EXPRESSION, file = 'CMC_DLPFC_Expression.tsv', sep = '\t', row.names=F, quote=F)
-EXP_OBJ = File('CMC_DLPFC_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', parentId = CODE$properties$id)
+write.table(RESIDUAL.GENE_EXPRESSION, file = 'CMC_ACC_Expression.tsv', sep = '\t', row.names=F, quote=F)
+EXP_OBJ = File('CMC_ACC_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', parentId = CODE$properties$id)
 EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 # Store differential expression results
-write.table(DE, file = 'CMC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_DLPFC_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
+write.table(DE, file = 'CMC_ACC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('CMC_ACC_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
 DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 

### 2016-09-19 code/Rmd/CMC_ACC_StarAligned_TMM.Rmd
Minor modifications

@@
   droplevels() %>%
-  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch),
+  dplyr::mutate(LibraryBatch = gsub('[^[:alnum:]]','_',LibraryBatch),
+                LibraryBatch = as.factor(LibraryBatch),
                 IsolationBatch = as.factor(IsolationBatch))

### 2016-09-19 code/Rmd/CMC_ACC_StarAligned_TMM_SVA.Rmd
Added markdown to process CMC ACC data with SVA normalisation

--- 2016-09-19 code/Rmd/CMC_ACC_StarAligned_TMM.Rmd
@@
                                  parentId ="syn5569976",
-                                 entityName = 'ACC - STAR Reprocessed - with TMM')
+                                 entityName = 'ACC - STAR Reprocessed - with TMM and SVA')
 ```
@@
 
-thisFileName <- 'CMC_ACC_StarAligned_TMM.Rmd'
+thisFileName <- 'CMC_ACC_StarAligned_TMM_SVA.Rmd'
 
@@
 
-### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "Dx" is chosen as the primary variable of interest
-```{r iterativeNorm}
+### Normalisation (Custom design)
+Since many covariates are correlated, re-normalising COUNTS with a custom design matrix. Here "Dx" is chosen as the primary variable of interest
+```{r custom.norm}
 # Primary variable of interest
-postAdjustCovars = 'RIN';#'Dx'
-
-# Assign residual covariates
-residualSigCovars = preAdjustedSigCovars
-covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
-
-loopCount = 0 
-while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
-  writeLines(paste('Using following covariates in the model:',
-                   paste(postAdjustCovars, collapse=', '),
-                   'as fixed effects'))
-
-  # Post adjusted design matrix
-  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
-
-  # Estimate voom weights
-  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
-
-  # Fit linear model using new weights and new design
-  VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
-
-  # Residuals after normalisation
-  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
-
-  # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars))
-
-  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                                COVARIATES[, residCovars, drop=F], 
-                                                'adjusted design(voom-normalized)',
-                                                isKeyPlot=TRUE)
-
-  # Add postadjusted covariates (if any)
-  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
-                                                                          c(postAdjustCovars))]
-
-  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
-  loopCount = loopCount + 1
-}
-
-# Add primary variable to the list (in case not chosen by iterative method)
-postAdjustCovars = unique(c('Dx', postAdjustCovars))
+postAdjustCovars = c('Dx', 'RIN', 'RIN2', 'Age_of_Death', 'PMI', 'LibraryBatch')
 
@@
 residualSigCovars[["PC_res"]][[2]]$plotData
+```
+Estimate surrogate variables using SVA package
+```{r sva}
+MODEL0 = DM1$design[,-grep('Dx',colnames(DM1$design))] # Get null model by removing variable of interest
+MODEL1 = DM1$design
+
+## Calculate the number of surrogate variables
+NUM.SV = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456, B = 10)
+
+SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, method = "irw", n.sv = NUM.SV, B = 30)$sv
+colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
+
+DM1$design = cbind(MODEL1, SURR.VAR)
+
+# Estimate correlation of random effects
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)
+
+# Re-calculate linear fit using new weights and new design
+VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
+
+# Residuals after normalisation
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
 ```
@@
 # Code
-CODE <- Folder(name = "ACC - STAR Reprocessed - with TMM", parentId = parentId)
+CODE <- Folder(name = "ACC - STAR Reprocessed - with TMM and SVA", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-09-19 code/Rmd/CMC_DLPFC_StarAligned_TMM_SVA.Rmd
Added markdown to process star aligned DLPFC data with TMM and SVA based normalisation

--- 2016-09-19 code/Rmd/CMC_DLPFC_StarAligned_TMM.Rmd
@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (DLPFC with TMM normlaisation)"
+title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (DLPFC with TMM and SVA normlaisation)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_DLPFC_StarAligned_TMM.Rmd",
+knit2synapse::knitToFolderEntity(file = "./CMC_DLPFC_StarAligned_TMM_SVA.Rmd",
                                  parentId ="syn5569976",
-                                 entityName = 'DLPFC - STAR Reprocessed - with TMM')
+                                 entityName = 'DLPFC - STAR Reprocessed - with TMM and SVA')
 ```
@@
 activityName = 'Covariate adjustments';
-activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM normalisation (DLPFC)';
+activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM and SVA normalisation (DLPFC)';
 
-thisFileName <- 'CMC_DLPFC_StarAligned_TMM.Rmd'
+thisFileName <- 'CMC_DLPFC_StarAligned_TMM_SVA.Rmd'
 
@@
 # Primary variable of interest
-postAdjustCovars = 'RIN';#'Dx'
-
-# Assign residual covariates
-residualSigCovars = preAdjustedSigCovars
-covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
-
-loopCount = 0 
-while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
-  writeLines(paste('Using following covariates in the model:',
-                   paste(postAdjustCovars, collapse=', '),
-                   'as fixed effects'))
-
-  # Post adjusted design matrix
-  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
-
-  # Estimate voom weights
-  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
-
-  # Fit linear model using new weights and new design
-  VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
-
-  # Residuals after normalisation
-  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
-
-  # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars))
-
-  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                                COVARIATES[, residCovars, drop=F], 
-                                                'adjusted design(voom-normalized)',
-                                                isKeyPlot=TRUE)
-
-  # Add postadjusted covariates (if any)
-  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
-                                                                          c(postAdjustCovars))]
-
-  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
-  loopCount = loopCount + 1
-}
-
-# Add primary variable to the list (in case not chosen by iterative method)
-postAdjustCovars = unique(c('Dx', postAdjustCovars))
+postAdjustCovars = c('Dx', 'RIN', 'RIN2', 'Age_of_Death', 'PMI', 'IsolationBatch')
 
@@
 residualSigCovars[["PC_res"]][[2]]$plotData
+```
+### Estimate surrogate variables using SVA package
+```{r sva}
+MODEL0 = DM1$design[,-grep('Dx',colnames(DM1$design))] # Get null model by removing variable of interest
+MODEL1 = DM1$design
+
+## Calculate the number of surrogate variables
+NUM.SV = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456, B = 10)
+
+SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, method = "irw", n.sv = NUM.SV, B = 30)$sv
+colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
+
+DM1$design = cbind(MODEL1, SURR.VAR)
+
+# Estimate correlation of random effects
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)
+
+# Re-calculate linear fit using new weights and new design
+VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
+
+# Residuals after normalisation
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
 ```
@@
 # Code
-CODE <- Folder(name = "DLPFC - STAR Reprocessed - with TMM", parentId = parentId)
+CODE <- Folder(name = "DLPFC - STAR Reprocessed - with TMM and SVA", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-09-20 code/Rmd/MSSM_BM_36_RNASeq_Realigned_StarHTSeq.Rmd
Minor modifications

@@
   dplyr::select(SampleID, ID) %>%
-  dplyr::mutate(ID = as.numeric(ID))
+  dplyr::mutate(ID = as.numeric(ID),
+                Sample.ID = SampleID,
+                SampleID = paste0('BM_36_',ID))
+
+# Sum counts from multiple lanes
+COUNT = dlply(SampleID, .(ID), .fun = function(x, count){
+  rowSums(count[,x$Sample.ID,drop=F]) %>%
+    rownameToFirstColumn('Gene.ID') %>%
+    plyr::rename(c('DF' = paste0('BM_36_',unique(x$ID)))) %>%
+    dplyr::select(-Gene.ID)
+}, COUNT) %>%
+  Reduce(cbind, .)
+COUNT = data.matrix(COUNT)
 
@@
 # Get technical metadata
-METADATA_ID <- 'syn5898489'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
+METADATA_TECHNICAL_ID <- 'syn5898489'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECHNICAL_ID
 METADATA_TECHNICAL <- read.table(synGet(METADATA_ID)@filePath,sep='\t',header=T, row.names=1) %>%
@@
                        dplyr::mutate(ID = as.numeric(ID))) %>%
-  inner_join(SampleID) %>%
+  inner_join(SampleID %>% dplyr::select(-Sample.ID)) %>%
   filter(!is.na(SampleID)) %>%
-  filter(!(ID %in% c(335:341, 10567, 10622, 10577))) %>%
+  filter(!(ID %in% c(335, 338, 339, 341))) %>%
   unique
@@
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
-# p <- p + geom_text(aes(label= ID), size=4, hjust=1)
+p <- p + geom_text(aes(label= ID), size=4, hjust=1)
 p
@@
                    DiffExpGenes = paste(unique(hgnc_symbol), collapse = ','))
-
 kable(tmp)

### 2016-09-20 code/Rmd/MSSM_BM_36_RNASeq_Realigned_StarHTSeq2.Rmd
Added markdown to process BM36 region from MSSM (started from source BAM in minerva converted to FASTQ using picard realigned using star and quantified using HTseq)

--- 2016-09-20 code/Rmd/MSSM_BM_36_RNASeq_Realigned_StarHTSeq.Rmd
@@
 ---
-title: "Covariates analysis of MSBB RNASeq data of region BM 36 (realigned from source FASTQ with STAR-HTSeq)"
+title: "Covariates analysis of MSBB RNASeq data of region BM 36 (realigned from FASTQ extracted source BAM in minerva with STAR-HTSeq)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_BM_36_RNASeq_Realigned_StarHTSeq.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_BM_36_RNASeq_Realigned_StarHTSeq2.Rmd",
                                  parentId ="syn5570248",
-                                 entityName = 'MSBB BM 36 RNASeq RealignedFromSource StarHTSeq')
+                                 entityName = 'MSBB BM 36 RNASeq RealignedFromSource BAM Picard StarHTSeq')
 ```
@@
 
-thisFileName <- 'MSSM_BM_36_RNASeq_Realigned_StarHTSeq.Rmd'
+thisFileName <- 'MSSM_BM_36_RNASeq_Realigned_StarHTSeq2.Rmd'
 
@@
 # Download expression data
-COUNT_ID <- 'syn7224673'
+COUNT_ID <- 'syn7254146'
 ALL_USED_IDs <- COUNT_ID
@@
 # Code
-CODE <- Folder(name = "MSBB BM 36 RNASeq  RealignedFromSource StarHTSeq", parentId = parentId)
+CODE <- Folder(name = "MSBB BM 36 RNASeq RealignedFromSource BAM Picard StarHTSeq", parentId = parentId)
 CODE <- synStore(CODE)
@@
                     executed = thisFile, activityDescription = activityDescription)
+
+# Store logcpm values
+write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'logCPM.tsv', sep = '\t', row.names=F, quote=F)
+exp1_obj = File('logCPM.tsv', name = 'Raw logCPM', parentId = CODE$properties$id)
+exp1_obj = synStore(exp1_obj, used = ALL_USED_IDs, activityName = activityName, 
+                   executed = thisFile, activityDescription = activityDescription)
+
 ```
@@
 |  Covariates               |  `r covar_obj$properties$id`    |
+|  logCPM                   |  `r exp1_obj$properties$id`     |
 |  Residual Expression      |  `r expr_obj$properties$id`     |

### 2016-09-20 code/Rmd/MSSM_BM_36_RNASeq_Realigned_StarHTSeq.Rmd
Minor modifications

@@
                     executed = thisFile, activityDescription = activityDescription)
+
+# Store logcpm values
+write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'logCPM.tsv', sep = '\t', row.names=F, quote=F)
+exp1_obj = File('logCPM.tsv', name = 'Raw logCPM', parentId = CODE$properties$id)
+exp1_obj = synStore(exp1_obj, used = ALL_USED_IDs, activityName = activityName, 
+                   executed = thisFile, activityDescription = activityDescription)
 ```

### 2016-09-20 code/Rmd/MSSM_BM_36_RNASeq.Rmd
Added markdown to process BM36 from MSSM aligned and processed by Eli-Lilly

@@
                     executed = thisFile, activityDescription = activityDescription)
+
+# Store logcpm values
+write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'logCPM.tsv', sep = '\t', row.names=T, quote=F)
+exp1_obj = File('logCPM.tsv', name = 'Raw logCPM', parentId = CODE$properties$id)
+exp1_obj = synStore(exp1_obj, used = ALL_USED_IDs, activityName = activityName, 
+                   executed = thisFile, activityDescription = activityDescription)
 ```

### 2016-09-20 code/Rmd/MSSM_BM_36_RNASeq_Realigned_StarHTSeq2.Rmd
Added markdown to process BM36 from MSSM aligned and processed by Eli-Lilly

@@
 # Store logcpm values
-write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'logCPM.tsv', sep = '\t', row.names=F, quote=F)
+write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'logCPM.tsv', sep = '\t', row.names=T, quote=F)
 exp1_obj = File('logCPM.tsv', name = 'Raw logCPM', parentId = CODE$properties$id)

### 2016-09-20 code/Rmd/MSSM_BM_36_RNASeq_Reprocessed_LillyPipeline.Rmd
Added markdown to process BM36 from MSSM aligned and processed by Eli-Lilly

--- 2016-09-10 code/Rmd/MSSM_BM_44_RNASeq_Reprocessed_LillyPipeline.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_BM_44_RNASeq_Reprocessed_LillyPipeline.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_BM_36_RNASeq_Reprocessed_LillyPipeline.Rmd",
                                  parentId ="syn5570248",
-                                 entityName = 'MSBB BM 44 RNASeq Reprocessed LillyPipeline')
+                                 entityName = 'MSBB BM 36 RNASeq Reprocessed LillyPipeline')
 ```
@@
 activityName = 'Covariate analysis';
-activityDescription = 'Covariate analysis of MSBB BM 44 RNASeq samples';
+activityDescription = 'Covariate analysis of MSBB BM 36 RNASeq samples';
 
-thisFileName <- 'MSSM_BM_44_RNASeq_Reprocessed_LillyPipeline.Rmd'
+thisFileName <- 'MSSM_BM_36_RNASeq_Reprocessed_LillyPipeline.Rmd'
 
@@
   unique %>%
-  filter(BrainRegion == 'BM_44', !(SampleID %in% c('hB_RNA_8565', 'hB_RNA_4919', 'hB_RNA_9005')))
+  filter(BrainRegion == 'BM_36', !(SampleID %in% c('hB_RNA_8565', 'hB_RNA_4919', 'hB_RNA_9005')))
 rownames(METADATA) = METADATA$SampleID
@@
 # Code
-CODE <- Folder(name = 'MSBB BM 44 RNASeq Reprocessed LillyPipeline', parentId = parentId)
+CODE <- Folder(name = 'MSBB BM 36 RNASeq Reprocessed LillyPipeline', parentId = parentId)
 CODE <- synStore(CODE)

### 2016-09-22 code/Rmd/CMC_DLPFC_StarAligned_TMM.Rmd
Modified DLPFC processing to include custom design

@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (DLPFC with TMM normlaisation)"
+title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (DLPFC with TMM normlaisation)"
 author: "Thanneer Perumal"
@@
 library(biomaRt)
+library(RColorBrewer)
+
+library(synapseClient)
+library(knitr)
+library(githubr)
+
+synapseLogin()
+
+## Requires ggdendro
+devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")
 
@@
 
+# Get ancestry vector calculated using gemtools
+ANCESTRY_ID <- 'syn2511399'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = ANCESTRY_ID
+ANCESTRY <- read.table(synGet(ANCESTRY_ID)@filePath, fill = T, header = T, sep = '\t') %>%
+  plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'GenotypingSampleID'))
+
 ## Metadata with specific information for RNASeq processing
-metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Ethnicity", "Age of Death", "PMI (in hours)", 
-                  "pH", "Dx", "CDR", "CauseOfDeath", "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
-                  "DLPFC_RNA_isolation: RNA Isolation Batch", "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?",
-                  "DLPFC_RNA_report: Library Batch")
+metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", 
+                          "Dx", "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
+                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
+                          'DNA_genotyping_report: Genotyping Sample_ID')
 
@@
 METADATA = METADATA %>%
-  dplyr::select(Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, PMI__in_hours_, pH, Dx, CDR, CauseOfDeath,
-                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RNA_Isolation_Batch,
-                DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Library_Batch) %>%
+  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, DLPFC_RNA_isolation__Exclude_,
+                DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_,
+                DLPFC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
   dplyr::rename(PMI = PMI__in_hours_, 
@@
                 SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
-                IsolationBatch = DLPFC_RNA_isolation__RNA_Isolation_Batch,
                 RIN = DLPFC_RNA_isolation__RIN, 
                 ReportExclude = DLPFC_RNA_report__Exclude_, 
-                LibraryBatch = DLPFC_RNA_report__Library_Batch) %>%
+                LibraryBatch = DLPFC_RNA_report__Library_Batch,
+                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
   dplyr::filter(SampleID %in% colnames(COUNT)) %>%
   droplevels() %>%
-  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch),
-                IsolationBatch = as.factor(IsolationBatch))
+  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
+  left_join(ANCESTRY)
+
+# Replace AFF and BP samples as other
+levels(METADATA$Dx) = c("Other", "Other", "Control", "SCZ")
 ```
@@
 ### Data preprocessing
-```{r preprocess.data, echo=TRUE}
+```{r preprocess.data}
 # Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
-METADATA <- METADATA %>%
-  filter(SampleID %in% colnames(COUNT)) %>%
-  filter(IsolationExclude == 0, ReportExclude == 0) %>%
-  filter(!is.na(PMI)) %>%
-  filter(!is.na(Gender)) %>%
-  filter(!is.na(Ethnicity)) %>%
-  filter(!is.na(RIN)) %>%
-  filter(!(SampleID %in% c('MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_60', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_361', 
-                           'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_389', 'PENN_RNA_PFC_56', 'PENN_RNA_PFC_59'))) %>%
+writeLines('Following counts are missing any metadata')
+writeLines(paste(setdiff(colnames(COUNT), levels(METADATA$SampleID)), collapse = ', '))
+METADATA <- METADATA %>% filter(SampleID %in% colnames(COUNT))
+
+writeLines('Following samples are marked exclude')
+writeLines(paste(METADATA$SampleID[METADATA$IsolationExclude == 1 | METADATA$ReportExclude == 1], collapse = ', '))
+METADATA <- METADATA  %>% filter(IsolationExclude == 0, ReportExclude == 0)
+
+writeLines('Following samples are missing PMI information')
+writeLines(paste(METADATA$SampleID[is.na(METADATA$PMI)], collapse = ', '))
+METADATA <- METADATA  %>% filter(!is.na(PMI))
+
+writeLines('Following samples are missing gender information')
+writeLines(paste(METADATA$SampleID[is.na(METADATA$Gender)], collapse = ', '))
+METADATA <- METADATA  %>% filter(!is.na(Gender))
+
+writeLines('Following samples are missing RIN information')
+writeLines(paste(METADATA$SampleID[is.na(METADATA$RIN)], collapse = ', '))
+METADATA <- METADATA  %>% filter(!is.na(RIN))
+
+writeLines('Following samples are missing EV information')
+writeLines(paste(METADATA$SampleID[is.na(METADATA$EV.1)], collapse = ', '))
+METADATA <- METADATA  %>% filter(!is.na(METADATA$EV.1))
+
+writeLines('Following samples are tagged as outliers')
+METADATA <- METADATA  %>% filter(!(SampleID %in% c('MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_60', 
+                                                   'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_361', 
+                                                   'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_389', 
+                                                   'PENN_RNA_PFC_56', 'PENN_RNA_PFC_59'))) %>%
   as.data.frame()
@@
 ```{r covariates.clustering}
-FactorCovariates <- c("Institution", "Gender", "Ethnicity", "CDR", "CauseOfDeath", "IsolationBatch", "LibraryBatch", "Dx")
-ContCovariates <- c("Age_of_Death", "PMI", "pH", "RIN")
+FactorCovariates <- c("Institution", "Gender", "LibraryBatch", "Dx")
+ContCovariates <- c("Age_of_Death", "PMI", "RIN", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")
 
@@
 COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
-COVARIATES <- data.frame(lapply(COVARIATES,function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'\\+','')})}))
+COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
 rownames(COVARIATES) <- METADATA$SampleID
@@
 
-# pH
-p[[4]] = ggplot(COVARIATES, aes(x = Dx, y = pH)) + geom_boxplot()
-p[[4]] = p[[4]] + ggtitle('pH') + my.theme
-
-multiplot(plotlist = p, cols = 2)
+multiplot(plotlist = p, cols = 3)
 ```
@@
 ```{r initial.voom.normalisation}
+# TMM normalisation
+TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix)
+
+# Initial normalisation of gene expression
+VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
+
+# Set gene counts in specific samples that are deviating 3 sd from other samples to NA
+log.mat = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
+  mn = mean(x, na.rm = T)
+  std.dev = sd(x, na.rm = T)
+  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
+}) %>% t
+PROCESSED_COUNTS$filteredExprMatrix[log.mat] = NA
+
 # TMM normalisation
@@
 # Primary variable of interest
-postAdjustCovars = 'RIN';#'Dx'
+postAdjustCovars = 'RIN';
 
@@
 }
+```
 
+### Normlaisation (Custom design)
+To be consistent with DLPFC Nature Neuroscience article we performed a custom adjustment with Dx, RIN, RIN2, Age of Death, PMI, Library Batch, Institution, Gender, EV1, EV2, EV3, EV4, EV5
+```{r custom.design}
 # Add primary variable to the list (in case not chosen by iterative method)
-postAdjustCovars = unique(c('Dx', postAdjustCovars))
+postAdjustCovars = c('Dx', c(FactorCovariates, ContCovariates)) %>% unique()
 
@@
 RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
-
-# Residual covariates to choose from
-residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars))
-
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                              COVARIATES[, residCovars, drop=F], 
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
 
@@
 # Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "jul2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "mar2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
 
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("DxControl-DxAFF", "DxControl-DxBP",
-                                     "DxControl-DxSCZ", "DxAFF-DxBP",
-                                     "DxAFF-DxSCZ", "DxBP-DxSCZ"),
+contrast = makeContrasts(contrasts=c("DxControl-DxOther",
+                                     "DxControl-DxSCZ",
+                                     "DxOther-DxSCZ"),
                          levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
@@
 # Get differnetial expression
-DE = lapply(1:6, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
+DE = lapply(1:3, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
   topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%

### 2016-09-22 code/Rmd/CMC_DLPFC_StarAligned_TMM_SVA.Rmd
Modified DLPFC with SVA pipeline to incorporate custom desing

@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (DLPFC with TMM and SVA normlaisation)"
+title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (DLPFC with TMM and SVA normlaisation)"
 author: "Thanneer Perumal"
@@
 ## Metadata with specific information for RNASeq processing
-metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Ethnicity", "Age of Death", "PMI (in hours)", 
-                  "pH", "Dx", "CDR", "CauseOfDeath", "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
-                  "DLPFC_RNA_isolation: RNA Isolation Batch", "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?",
-                  "DLPFC_RNA_report: Library Batch")
+metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", 
+                          "Dx", "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
+                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
+                          'DNA_genotyping_report: Genotyping Sample_ID')
 
@@
                 SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
-                IsolationBatch = DLPFC_RNA_isolation__RNA_Isolation_Batch,
                 RIN = DLPFC_RNA_isolation__RIN, 
                 ReportExclude = DLPFC_RNA_report__Exclude_, 
-                LibraryBatch = DLPFC_RNA_report__Library_Batch) %>%
+                LibraryBatch = DLPFC_RNA_report__Library_Batch,
+                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
   dplyr::filter(SampleID %in% colnames(COUNT)) %>%
   droplevels() %>%
-  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch),
-                IsolationBatch = as.factor(IsolationBatch))
+  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch))
+
+# Replace AFF and BP samples as other
+levels(METADATA$Dx) = c("Other", "Other", "Control", "SCZ")
 ```
@@
 ### Data preprocessing
-```{r preprocess.data, echo=TRUE}
+```{r preprocess.data}
 # Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
-METADATA <- METADATA %>%
-  filter(SampleID %in% colnames(COUNT)) %>%
-  filter(IsolationExclude == 0, ReportExclude == 0) %>%
-  filter(!is.na(PMI)) %>%
-  filter(!is.na(Gender)) %>%
-  filter(!is.na(Ethnicity)) %>%
-  filter(!is.na(RIN)) %>%
-  filter(!(SampleID %in% c('MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_60', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_361', 
-                           'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_389', 'PENN_RNA_PFC_56', 'PENN_RNA_PFC_59'))) %>%
+writeLines('Following counts are missing any metadata')
+writeLines(paste(setdiff(colnames(COUNT), levels(METADATA$SampleID)), collapse = ', '))
+METADATA <- METADATA %>% filter(SampleID %in% colnames(COUNT))
+
+writeLines('Following samples are marked exclude')
+writeLines(paste(METADATA$SampleID[METADATA$IsolationExclude == 1 | METADATA$ReportExclude == 1], collapse = ', '))
+METADATA <- METADATA  %>% filter(IsolationExclude == 0, ReportExclude == 0)
+
+writeLines('Following samples are missing PMI information')
+writeLines(paste(METADATA$SampleID[is.na(METADATA$PMI)], collapse = ', '))
+METADATA <- METADATA  %>% filter(!is.na(PMI))
+
+writeLines('Following samples are missing gender information')
+writeLines(paste(METADATA$SampleID[is.na(METADATA$Gender)], collapse = ', '))
+METADATA <- METADATA  %>% filter(!is.na(Gender))
+
+writeLines('Following samples are missing RIN information')
+writeLines(paste(METADATA$SampleID[is.na(METADATA$RIN)], collapse = ', '))
+METADATA <- METADATA  %>% filter(!is.na(RIN))
+
+writeLines('Following samples are tagged as outliers')
+METADATA <- METADATA  %>% filter(!(SampleID %in% c('MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_60', 
+                                                   'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_361', 
+                                                   'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_389', 
+                                                   'PENN_RNA_PFC_56', 'PENN_RNA_PFC_59'))) %>%
   as.data.frame()
@@
 ```{r covariates.clustering}
-FactorCovariates <- c("Institution", "Gender", "Ethnicity", "CDR", "CauseOfDeath", "IsolationBatch", "LibraryBatch", "Dx")
-ContCovariates <- c("Age_of_Death", "PMI", "pH", "RIN")
+FactorCovariates <- c("Institution", "Gender", "LibraryBatch", "Dx")
+ContCovariates <- c("Age_of_Death", "PMI", "RIN")
 
@@
 
-### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "Dx" is chosen as the primary variable of interest
-```{r iterativeNorm}
-# Primary variable of interest
-postAdjustCovars = c('Dx', 'RIN', 'RIN2', 'Age_of_Death', 'PMI', 'IsolationBatch')
+### Normlaisation (Custom design)
+To be consistent with DLPFC Nature Neuroscience article we performed a custom adjustment with Dx, RIN, RIN2, Age of Death, PMI, Library Batch, Institution, Gender
+```{r custom.design}
+# Add primary variable to the list (in case not chosen by iterative method)
+postAdjustCovars = c('Dx', c(FactorCovariates, ContCovariates)) %>% unique()
 
@@
 
-# Residual covariates to choose from
-residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars))
-
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                              COVARIATES[, residCovars, drop=F], 
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
-
 modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
@@
 `r tmp`
-### Sanity check
-```{r residualSigCovars.manual, fig.width=20, fig.height=12}
-residualSigCovars[["PC_res"]][[2]]$plotData
-```
 ### Estimate surrogate variables using SVA package
@@
 RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
+```
+### Sanity check
+```{r residualSigCovars.manual, fig.width=20, fig.height=12}
+residualSigCovars[["PC_res"]][[2]]$plotData
 ```

### 2016-09-22 code/Rmd/CMC_ACC_StarAligned_TMM.Rmd
Major modifications in processing ACC

--- 2016-09-19 code/Rmd/CMC_ACC_StarAligned_TMM.Rmd
@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (ACC with TMM normlaisation)"
+title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC with TMM normlaisation)"
 author: "Thanneer Perumal"
@@
 
+# Get ancestry vector calculated using gemtools
+ANCESTRY_ID <- 'syn2511399'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = ANCESTRY_ID
+ANCESTRY <- read.table(synGet(ANCESTRY_ID)@filePath, fill = T, header = T, sep = '\t') %>%
+  plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'GenotypingSampleID'))
+
 ## Metadata with specific information for RNASeq processing
-metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Ethnicity", "Age of Death", "PMI (in hours)", 
-                          "pH", "Dx", "CDR", "CauseOfDeath", "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
-                          "ACC_RNA_isolation: RNA Isolation Batch", "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?",
-                          "ACC_RNA_report: Library Batch")
+metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", 
+                          "Dx", "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
+                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?", "ACC_RNA_report: Library Batch",
+                          'DNA_genotyping_report: Genotyping Sample_ID')
 
@@
 METADATA = METADATA %>%
-  dplyr::select(Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, PMI__in_hours_, pH, Dx, CDR, CauseOfDeath,
-                ACC_RNA_isolation__Exclude_, ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RNA_Isolation_Batch,
-                ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_, ACC_RNA_report__Library_Batch) %>%
+  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, ACC_RNA_isolation__Exclude_,
+                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_,
+                ACC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
   dplyr::rename(PMI = PMI__in_hours_, 
@@
                 SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
-                IsolationBatch = ACC_RNA_isolation__RNA_Isolation_Batch,
                 RIN = ACC_RNA_isolation__RIN, 
                 ReportExclude = ACC_RNA_report__Exclude_, 
-                LibraryBatch = ACC_RNA_report__Library_Batch) %>%
+                LibraryBatch = ACC_RNA_report__Library_Batch,
+                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
   dplyr::filter(SampleID %in% colnames(COUNT)) %>%
   droplevels() %>%
-  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch),
-                IsolationBatch = as.factor(IsolationBatch))
+  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
+  left_join(ANCESTRY)
+
+# Replace AFF and BP samples as other
+levels(METADATA$Dx) = c("Other", "Other", "Control", "SCZ")
 ```
@@
 ### Data preprocessing
-```{r preprocess.data, echo=TRUE}
+```{r preprocess.data}
 # Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
-METADATA <- METADATA %>%
-  filter(SampleID %in% colnames(COUNT)) %>%
-  filter(IsolationExclude == 0, ReportExclude == 0) %>%
-  filter(!is.na(PMI)) %>%
-  filter(!is.na(Gender)) %>%
-  filter(!is.na(Ethnicity)) %>%
-  filter(!is.na(RIN)) %>%
-  filter(!(SampleID %in% c('MSSM_RNA_ACC_250', 'MSSM_RNA_ACC_170'))) %>%
-  as.data.frame()
+writeLines('Following counts are missing any metadata')
+writeLines(paste(setdiff(colnames(COUNT), levels(METADATA$SampleID)), collapse = ', '))
+METADATA <- METADATA %>% filter(SampleID %in% colnames(COUNT))
+
+writeLines('Following samples are marked exclude')
+writeLines(paste(METADATA$SampleID[METADATA$IsolationExclude == 1 | METADATA$ReportExclude == 1], collapse = ', '))
+METADATA <- METADATA  %>% filter(IsolationExclude == 0, ReportExclude == 0)
+
+writeLines('Following samples are missing PMI information')
+writeLines(paste(METADATA$SampleID[is.na(METADATA$PMI)], collapse = ', '))
+METADATA <- METADATA  %>% filter(!is.na(PMI))
+
+writeLines('Following samples are missing gender information')
+writeLines(paste(METADATA$SampleID[is.na(METADATA$Gender)], collapse = ', '))
+METADATA <- METADATA  %>% filter(!is.na(Gender))
+
+writeLines('Following samples are missing RIN information')
+writeLines(paste(METADATA$SampleID[is.na(METADATA$RIN)], collapse = ', '))
+METADATA <- METADATA  %>% filter(!is.na(RIN))
+
+writeLines('Following samples are missing EV information')
+writeLines(paste(METADATA$SampleID[is.na(METADATA$EV.1)], collapse = ', '))
+METADATA <- METADATA  %>% filter(!is.na(METADATA$EV.1)) 
 ```
@@
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
 p
@@
 }
+```
 
+### Normlaisation (Custom design)
+Adding Dx, Institution, Gender, EV1, EV3, EV4 and EV5 to the list of postadjust covariates RIN, RIN2, Age of Death, PMI and EV2
+```{r custom.design}
 # Add primary variable to the list (in case not chosen by iterative method)
-postAdjustCovars = unique(c('Dx', postAdjustCovars))
+postAdjustCovars = c('Dx', c(FactorCovariates, ContCovariates)) %>% unique()
 
@@
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)
 p

### 2016-09-22 code/Rmd/CMC_ACC_StarAligned_TMM_SVA.Rmd
Modified ACC pipeline with SVA to account for change in covariates

@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (ACC with TMM normlaisation)"
+title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC with TMM and SVA normlaisation)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_ACC_StarAligned_TMM.Rmd",
+knit2synapse::knitToFolderEntity(file = "./CMC_ACC_StarAligned_TMM_SVA.Rmd",
                                  parentId ="syn5569976",
@@
 activityName = 'Covariate adjustments';
-activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM normalisation (ACC)';
+activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM and SVA normalisation (ACC)';
 
@@
 ## Metadata with specific information for RNASeq processing
-metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Ethnicity", "Age of Death", "PMI (in hours)", 
-                          "pH", "Dx", "CDR", "CauseOfDeath", "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
-                          "ACC_RNA_isolation: RNA Isolation Batch", "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?",
-                          "ACC_RNA_report: Library Batch")
+metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", 
+                          "Dx", "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
+                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?", "ACC_RNA_report: Library Batch",
+                          'DNA_genotyping_report: Genotyping Sample_ID')
 
@@
                 SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
-                IsolationBatch = ACC_RNA_isolation__RNA_Isolation_Batch,
                 RIN = ACC_RNA_isolation__RIN, 
                 ReportExclude = ACC_RNA_report__Exclude_, 
-                LibraryBatch = ACC_RNA_report__Library_Batch) %>%
+                LibraryBatch = ACC_RNA_report__Library_Batch,
+                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
   dplyr::filter(SampleID %in% colnames(COUNT)) %>%
   droplevels() %>%
-  dplyr::mutate(LibraryBatch = gsub('[^[:alnum:]]','_',LibraryBatch),
-                LibraryBatch = as.factor(LibraryBatch),
-                IsolationBatch = as.factor(IsolationBatch))
+  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch))
+
+# Replace AFF and BP samples as other
+levels(METADATA$Dx) = c("Other", "Other", "Control", "SCZ")
 ```
@@
 ### Data preprocessing
-```{r preprocess.data, echo=TRUE}
+```{r preprocess.data}
 # Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
-METADATA <- METADATA %>%
-  filter(SampleID %in% colnames(COUNT)) %>%
-  filter(IsolationExclude == 0, ReportExclude == 0) %>%
-  filter(!is.na(PMI)) %>%
-  filter(!is.na(Gender)) %>%
-  filter(!is.na(Ethnicity)) %>%
-  filter(!is.na(RIN)) %>%
-  filter(!(SampleID %in% c('MSSM_RNA_ACC_250', 'MSSM_RNA_ACC_170'))) %>%
-  as.data.frame()
+writeLines('Following counts are missing any metadata')
+writeLines(paste(setdiff(colnames(COUNT), levels(METADATA$SampleID)), collapse = ', '))
+METADATA <- METADATA %>% filter(SampleID %in% colnames(COUNT))
+
+writeLines('Following samples are marked exclude')
+writeLines(paste(METADATA$SampleID[METADATA$IsolationExclude == 1 | METADATA$ReportExclude == 1], collapse = ', '))
+METADATA <- METADATA  %>% filter(IsolationExclude == 0, ReportExclude == 0)
+
+writeLines('Following samples are missing PMI information')
+writeLines(paste(METADATA$SampleID[is.na(METADATA$PMI)], collapse = ', '))
+METADATA <- METADATA  %>% filter(!is.na(PMI))
+
+writeLines('Following samples are missing gender information')
+writeLines(paste(METADATA$SampleID[is.na(METADATA$Gender)], collapse = ', '))
+METADATA <- METADATA  %>% filter(!is.na(Gender))
+
+writeLines('Following samples are missing RIN information')
+writeLines(paste(METADATA$SampleID[is.na(METADATA$RIN)], collapse = ', '))
+METADATA <- METADATA  %>% filter(!is.na(RIN)) 
 ```
@@
 
-### Normalisation (Custom design)
-Since many covariates are correlated, re-normalising COUNTS with a custom design matrix. Here "Dx" is chosen as the primary variable of interest
-```{r custom.norm}
-# Primary variable of interest
-postAdjustCovars = c('Dx', 'RIN', 'RIN2', 'Age_of_Death', 'PMI', 'LibraryBatch')
+### Normlaisation (Custom design)
+To be consistent with ACC Nature Neuroscience article we performed a custom adjustment with Dx, RIN, RIN2, Age of Death, PMI, Library Batch, Institution, Gender
+```{r custom.design}
+# Add primary variable to the list (in case not chosen by iterative method)
+postAdjustCovars = c('Dx', c(FactorCovariates, ContCovariates)) %>% unique()
 
@@
 `r tmp`
-### Sanity check
-```{r residualSigCovars.manual, fig.width=20, fig.height=12}
-residualSigCovars[["PC_res"]][[2]]$plotData
-```
-Estimate surrogate variables using SVA package
+### Estimate surrogate variables using SVA package
 ```{r sva}

### 2016-09-26 code/Rmd/GTEx_SkeletalMuscle.Rmd
Added markdown to process GTEx skeletal muscle samples

---
title: "Covariate adjustments on GTEx data (Skeletal Muscle)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)
library(knitr)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./GTEx_SkeletalMuscle.Rmd",
                                 parentId = "syn7250653",
                                 entityName = "GTEx (Skeletal Muscle)")
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get it from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(RColorBrewer)
library(ggplot2)
library(reshape2)
library(limma)
library(Biobase)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)

devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn7250653';
activityName = 'Covariate analysis of GTEx';
activityDescription = 'Covariate analysis of GTEx RNASeq data from skeletal muscle';

thisFileName <- 'GTEx_SkeletalMuscle.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='GTEX')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse (Phil's downloaded version from syn6172827)
```{r download.data, cache=TRUE}
# Download reprocessed counts (paired end)
COUNT_ID <- 'syn7250659'
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(gzfile(synGet(COUNT_ID)@filePath), header=T, fill = T, skip = 2, sep = '\t', check.names = F)
Gene.Description <- COUNT[,c('Name', 'Description')]
rownames(COUNT) = COUNT$Name
COUNT = COUNT[,-(1:2)]

# Get sample phenotypes
METADATA_CLINICAL_ID <- 'syn7250661'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CLINICAL_ID
METADATA_CLINICAL <- read.table(synGet(METADATA_CLINICAL_ID)@filePath, sep='\t', header=T)

# Get sample metadata
METADATA_TECHNICAL_ID <- 'syn7250662'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECHNICAL_ID
METADATA_TECHNICAL <- read.table(synGet(METADATA_TECHNICAL_ID)@filePath, sep = '\t', header=T)

# Match technical and clinical covariates
METADATA <- METADATA_TECHNICAL %>%
  filter(SAMPID %in% colnames(COUNT)) %>%
  tidyr::separate(SAMPID, c('a','b','c','d','e'), sep = '-') %>%
  dplyr::mutate(SUBJID = paste(a,b, sep = '-')) %>%
  tidyr::unite(SAMPID, a,b,c,d,e, sep = '-') %>%
  dplyr::left_join(METADATA_CLINICAL)
```

### Data preprocessing
```{r preprocess.data,cache=TRUE, echo=TRUE}
# Remove samples that are not from brain tissue and has missing RIN scores
METADATA <- METADATA %>%
  filter(SMTSD == 'Muscle - Skeletal') %>%
  filter(!is.na(SMRIN))
```

```{r preprocess.data1,cache=TRUE}
# Match covariates to expression data
indToRetain = intersect(METADATA$SAMPID, colnames(COUNT))
removedIDs = setdiff(colnames(COUNT), METADATA$SAMPID)

COUNT = COUNT[,indToRetain]

rownames(METADATA) = METADATA$SAMPID
METADATA = METADATA[indToRetain,]
```

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering, cache=TRUE}
FactorCovariates <- c("SMNABTCH", "AGE", "GENDER")
ContCovariates <- c("SMTSISCH", "SMRIN", "SMMPPD", "SMATSSCR")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES <- data.frame(lapply(COVARIATES,function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'\\+','')})}))
rownames(COVARIATES) <- METADATA$SAMPID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates,drop=F], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)

# Add in RIN^2 values
COVARIATES$SMRIN2 = COVARIATES$SMRIN^2
ContCovariates = c(ContCovariates, 'SMRIN2')
```
Explanation of covariates

1. SAMPID:	Sample ID, GTEx Public Sample ID
2. SMNABTCH:	Nucleic Acid Isolation Batch ID
3. SMATSSCR:	Autolysis Score
4. SMRIN:	RIN Number
5. SMTSISCH:	Total Ischemic time for a sample in 4 hour intervals
6. SMMPPD:	Total number of reads aligned/mapped
7. AGE:
8. GENDER:

Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.5, row.width=0.1)
```

### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

Remove genes that have less than 1 cpm counts in at least 50% of samples.
```{r cpmnormalisation, cache=TRUE}
PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT,MIN_GENE_CPM=1, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
```
### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T)
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(ID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'ID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color = as.factor(SMCENTER), size = SMRIN))
p <- p + theme_bw() + theme(legend.position="top") 
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p

# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,FactorCovariates]))
```
```{r temp, include = F}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```
Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix.
```{r iterativeNorm}
# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
covariatesEffects = residualSigCovars$Effects.significantCovars
postAdjustCovars = c('AGE','GENDER', names(which.max(covariatesEffects)))

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate correlation of random effects
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), postAdjustCovars)

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(ID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'ID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=SMNABTCH, shape = GENDER, size=SMRIN))
p <- p + theme_bw() + theme(legend.position="none")
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p

# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,FactorCovariates]))
```
```{r temp1, include=F}
dev.off()
gc()
```

```{r get.gene.symbols, cache=TRUE, eval=T}
# Add gene names
RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('Name') %>%
  left_join(Gene.Description) %>%
  dplyr::select(Name, Description, one_of(rownames(COVARIATES)))

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('Name') %>%
  left_join(Gene.Description) %>%
  dplyr::select(Name, Description, one_of(rownames(COVARIATES)))
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "GTEx (Skeletal Muscle)", parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SAMPLEID')
write.table(COVARIATES, file = 'GTExSMCovariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('GTExSMCovariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store logCPM
write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'GTExSMlogCPM.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('GTExSMlogCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                     executed = thisFile, activityDescription = activityDescription)

# Store design matrix
write.table(DM1$design, file = 'GTExSMDesign.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('GTExSMDesign.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)

# Store residual gene expression
write.table(RESIDUAL.GENE_EXPRESSION, file = 'GTExSMExpression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('GTExSMExpression.tsv', name = 'Normalised Adjusted Residual Expression', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
```
|  *Results*                |  *SynapseID*                     |
|  -------                  |   ---------                      |
|  Covariates               |  `r COV_OBJ$properties$id`       |
|  Raw counts (log CPM)     |  `r COUNT_OBJ$properties$id`     |
|  Design Matrix            |  `r DM_OBJ$properties$id`        |
|  Residual Expression      |  `r EXP_OBJ$properties$id`       |

### 2016-09-26 code/Rmd/GTEx_SkeletalMuscle.Rmd
Minor modifications

@@
 library(biomaRt)
+
+library(synapseClient)
+library(knitr)
+library(githubr)
+
+synapseLogin()
 
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color = as.factor(SMCENTER), size = SMRIN))
+p <- p + geom_point(aes(color = SMNABTCH, shape = GENDER, size = SMRIN))
 p <- p + theme_bw() + theme(legend.position="top") 

### 2016-09-26 code/Rmd/ROSMAP_RNASeq_Qntl_SVA.Rmd
Minor changes (included gene based filtering)

@@
   filter(!is.na(pmi)) %>%
-  filter(!(Sampleid %in% c('380_120503_1'))) %>%
+  filter(!(Sampleid %in% c('380_120503_1', '792_130530_7', '15_120410_3', '256_120426_5'))) %>%
   as.data.frame()
@@
 
+# Set gene counts in specific samples that are deviating 3 sd from other samples to NA
+log.mat = apply(EXPR, 1, function(x){
+  mn = mean(x, na.rm = T)
+  std.dev = sd(x, na.rm = T)
+  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
+}) %>% t
+EXPR[log.mat] = NA
+
 # Between sample quantile normalisation
@@
 ```{r coexp1, cache=FALSE}
-cr = cor(t(EXPR))
+EXPR.tmp = EXPR
+EXPR.tmp[is.na(EXPR)] = 0
+cr = cor(t(EXPR.tmp))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(EXPR, scale.=T)
+PC <- prcomp(EXPR.tmp, scale.=T, center = T)
 
@@
 ```{r decompse.normalise.data2, fig.height=8, fig.width=16, results='asis'}
-tree = hclust(as.dist(t(EXPR)))
+tree = hclust(as.dist(t(EXPR.tmp)))
 plotDendroAndColors(tree, 
@@
 dev.off()
+gc()
 ```
@@
 # Find correlation between PC's of gene expression with covariates
-DM = getDesignMatrix(COVARIATES, Intercept = F)
-DESIGN = getDesignMat2Fact(DM$design,FactorCovariates)
-rownames(DESIGN) = rownames(COVARIATES)
-preAdjustedSigCovars = runPCAandPlotCorrelations(EXPR, DESIGN,'non-normalized', isKeyPlot=TRUE)
+preAdjustedSigCovars = runPCAandPlotCorrelations(EXPR.tmp, COVARIATES,'non-normalized', isKeyPlot=TRUE)
 
 # Find significant covariates
-adjustCovars = designMatVarsToCovars(getCovariatesMap(DM),preAdjustedSigCovars$significantCovars)
+adjustCovars = preAdjustedSigCovars$significantCovars
 ```
@@
   # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-  rownames(DESIGN) <- rownames(COVARIATES)
-  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION,
-                                                DESIGN,
+  RESIDUAL.GENE_EXPRESSION.tmp = RESIDUAL.GENE_EXPRESSION
+  RESIDUAL.GENE_EXPRESSION.tmp[is.na(RESIDUAL.GENE_EXPRESSION)] = 0
+  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION.tmp,
+                                                COVARIATES,
                                                 'all adjusted design',
@@
 
-## Calculate the number of surrogate variables
-# Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
-# SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
-# NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
-# SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
-# SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]
-
-# # Permute and calculate the permuted singular values
-# PERM.SING.VALS <- foreach (i = 1:100,
-#          .combine = rbind,
-#          .packages = c('limma','CovariateAnalysis'),
-#          .export = c('PROCESSED_COUNTS','MODEL1')) %dopar% {
-#   PERM.COUNTS <- t(apply(PROCESSED_COUNTS$filteredExprMatrix, 1, sample, replace = FALSE))
-#   # Estimate voom weights
-#   VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)
-#   
-#   # Fit linear model using new weights and new design
-#   VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)
-#   
-#   # Residuals after normalisation
-#   RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)
-#   
-#   PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
-#   PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
-# }
-# num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])
-
-num.sv2 = num.sv(EXPR, MODEL1, method = 'be', seed = 123456)
-NUM.SV = num.sv2 #min(num.sv1, num.sv2)
+num.sv = num.sv(EXPR.tmp, MODEL1, method = 'be', seed = 123456)
 
 ## Compute surrogate variables 
-SURR.VAR = sva(EXPR, MODEL1, MODEL0, method = "irw", n.sv = NUM.SV, B = 30)$sv
+SURR.VAR = sva(EXPR.tmp, MODEL1, MODEL0, method = "irw", n.sv = num.sv, B = 30)$sv
 colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
@@
 ```{r coexp2, cache=FALSE}
-cr = cor(t(RESIDUAL.GENE_EXPRESSION))
+RESIDUAL.GENE_EXPRESSION.tmp = RESIDUAL.GENE_EXPRESSION
+RESIDUAL.GENE_EXPRESSION.tmp[is.na(RESIDUAL.GENE_EXPRESSION)] = 0
+cr = cor(t(RESIDUAL.GENE_EXPRESSION.tmp))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T)
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION.tmp, scale.=T, center = T)
 

### 2016-09-27 code/Rmd/ROSMAP_RNASeq_Qntl.Rmd
Minor modifications

@@
   filter(!is.na(pmi)) %>%
-  filter(!(Sampleid %in% c('380_120503_1'))) %>%
+  filter(!(Sampleid %in% c('380_120503_1', '792_130530_7', '15_120410_3', '256_120426_5'))) %>%
   as.data.frame()
 ```
-Sample id 380\_120503\_1 is considered as outlier and removed
+
 ```{r preprocess.data1,cache=TRUE}
@@
   # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-rownames(DESIGN) <- rownames(COVARIATES)
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION,
-                                              DESIGN,
+  RESIDUAL.GENE_EXPRESSION.tmp = RESIDUAL.GENE_EXPRESSION
+  RESIDUAL.GENE_EXPRESSION.tmp[is.na(RESIDUAL.GENE_EXPRESSION)] = 0
+  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION.tmp,
+                                                COVARIATES,
                                                 'all adjusted design',
                                                 isKeyPlot=TRUE)
+
+  # Add postadjusted covariates (if any)
+  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
+
+  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
+  loopCount = loopCount + 1
+}
+modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
+                  'as fixed effects')
+
+tmp <- paste('Using following covariates in the final model:', modelStr)
+```
+`r tmp`
+### Sanity check
+```{r residualSigCovars.manual, fig.width=20, fig.height=12}
+residualSigCovars[["PC_res"]][[2]]$plotData
 ```
 
-```{r residualSigCovars.manual.ALL, fig.width=15, fig.height=8}
-residualSigCovars[["PC_res"]][[2]]$plotData
-```
-Adding cogdx back to the residuals
-```{r addVarsBack}
-varsToAddBackIn <- c("cogdx1",  "cogdx2", "cogdx3", "cogdx4", "cogdx5", "cogdx6")
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + t(data.matrix(DESIGN[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
-RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'EnsemblGeneID') %>%
-  tidyr::separate(col = EnsemblGeneID, into = c('ensembl_gene_id','position'), sep = '\\.')
+### Residual calculation
+Calculate weighted residuals and add back "cogdx" to the residuals
+```{r varsToAddBack}
+# Add variable of interest back to the residuals
+varsToAddIn = grep("cogdx", colnames(DM1$design), value = T)
+RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
+  FIT$coefficients[,varsToAddIn] %*% t(FIT$design[,varsToAddIn])
 ```
@@
 ```{r coexp2, cache=FALSE}
-cr = cor(t(RESIDUAL.GENE_EXPRESSION[, !(colnames(RESIDUAL.GENE_EXPRESSION) %in% c('ensembl_gene_id','position'))]))
+RESIDUAL.GENE_EXPRESSION.tmp = RESIDUAL.GENE_EXPRESSION
+RESIDUAL.GENE_EXPRESSION.tmp[is.na(RESIDUAL.GENE_EXPRESSION)] = 0
+cr = cor(t(RESIDUAL.GENE_EXPRESSION.tmp))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1:2)], scale.=T)
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION.tmp, scale.=T, center = T)
 
@@
 # Define biomart object
-mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = 'mar2016.archive.ensembl.org')
+
+RESIDUAL.GENE_EXPRESSION <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'ensembl_gene_id.position') %>%
+  tidyr::separate(ensembl_gene_id.position, c('ensembl_gene_id', 'position'), sep = '\\.')
 
@@
 ```
+
 ### Store files in synapse
@@
 
+# Store design matrix
+MODEL1 = rownameToFirstColumn(MODEL1, 'Sampleid_batch')
+write.table(MODEL1, file = 'designMatrix.tsv', sep = '\t', row.names=F, quote=F)
+DES_OBJ = File('designMatrix.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+DES_OBJ = synStore(DES_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                   executed = thisFile, activityDescription = activityDescription)
+
 # Store differential expression results

### 2016-09-27 code/Rmd/ROSMAP_RNASeq_Qntl_SVA.Rmd
Minor modifications

@@
 # Define biomart object
-mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = 'mar2016.archive.ensembl.org')
 

### 2016-10-25 code/Rmd/ROSMAP_RNASeq_Qntl.Rmd
Minor modifications

@@
 for (i in 1:3){
-  DE[[i]] = topTable(FIT.CONTR, coef=i, number = dim(EXPR)[1])
+  DE[[i]] = topTable(FIT.CONTR, coef=i, number = dim(EXPR)[1], confint = T)
   DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id") 

### 2016-10-25 code/R/lib/calcCompleteCorAndPlot.R

@@ -1,112 +1,0 @@
-# Function to calculate correlation and plot
-calcCompleteCorAndPlot <- function(COMPARE_data, COVAR_data, correlationType, title, 
-                                   WEIGHTS = NULL, PLOT_ALL_COVARS=FALSE, EXCLUDE_VARS_FROM_FDR=NULL, MAX_FDR = 0.1) {
-  
-  # require(plyr)
-  
-  # Get factor and continuous covariates
-  FactorCovariates <- colnames(COVAR_data)[sapply(COVAR_data,is.factor)]
-  ContCovariates <- setdiff(colnames(COVAR_data),FactorCovariates)
-  
-  # Calculate correlation between compare_data and factor covariates
-  if (length(FactorCovariates) > 0){
-    comb <- expand.grid(colnames(COMPARE_data),FactorCovariates)
-    factCont_cor <- plyr::ddply(comb,
-                                .(Var1, Var2),
-                                .fun = getFactorContAssociationStatistics,
-                                cbind(COMPARE_data,COVAR_data[rownames(COMPARE_data),FactorCovariates,drop=F]),
-                                alpha=MAX_FDR)
-    factCont_cor_vals <- factCont_cor %>%
-      dplyr::select(Var1, Var2, Estimate) %>%
-      tidyr::spread(Var2, Estimate)
-    rownames(factCont_cor_vals) <- factCont_cor_vals$Var1
-    factCont_cor_vals$Var1 <- NULL
-    factCont_cor_vals = factCont_cor_vals %>% data.matrix()
-    
-    factCont_cor_p <- factCont_cor %>%
-      dplyr::select(Var1, Var2, Pval) %>%
-      tidyr::spread(Var2, Pval)
-    rownames(factCont_cor_p) <- factCont_cor_p$Var1
-    factCont_cor_p$Var1 <- NULL
-    factCont_cor_p = factCont_cor_p %>% data.matrix()
-    
-  } else {
-    factCont_cor_vals <- NULL
-    factCont_cor_p <- NULL
-  }
-  
-  # Calculate correlation between compare_data and factor covariates
-  if (length(ContCovariates) > 0){
-    cont_cor <- corr.test(COMPARE_data,
-                          COVAR_data[,ContCovariates, drop=F],
-                          use='pairwise.complete.obs',
[…]

### 2016-10-25 code/R/lib/calcResiduals.R

@@ -1,65 +1,0 @@
-# Function to optain residual matrix (modified from covairates pipeline of Menachem Former)
-calcResiduals <- function(geneBySampleValues, samplesByCovariates, factorCovariates = NULL, varsToAddBackIn=NULL, sampleWeights=NULL) {  
-  
-  # Convert factor covariates data frame to numeric matrix
-  if (!is.null(factorCovariates)){
-    colNames = intersect(colnames(samplesByCovariates),factorCovariates)
-    samplesByCovariates[,colNames] = apply(samplesByCovariates[,colNames,drop=F],2, function(cols){unclass(factor(cols))})
-  }    
-  samplesByCovariates = as.matrix(samplesByCovariates)
-  
-  ##########################################################################################################################################################
-  #### If sampleWeights are given as matrix use calcResiduals in a for loop (this is done because lm function take wt as only vector and not as matrix) ####
-  ##########################################################################################################################################################
-  if (is.matrix(sampleWeights)) {
-    residualizedMat = matrix(NA, nrow=nrow(geneBySampleValues), ncol=ncol(geneBySampleValues), dimnames=dimnames(geneBySampleValues))
-    for (gInd in 1:nrow(geneBySampleValues)) {
-      gRow = calcResiduals(geneBySampleValues[gInd, , drop=FALSE], samplesByCovariates, factorCovariates = NULL, varsToAddBackIn = varsToAddBackIn, sampleWeights = sampleWeights[gInd, ])
-      residualizedMat[gInd, ] = gRow
-    }
-    return(residualizedMat)
-  }
-  #################################################################################
-  
-  # If lest square model is needed (uncomment the following line)
-  # result.lm = lsfit(x=samplesByCovariates, y=t(geneBySampleValues), wt=sampleWeights, intercept=FALSE)
-  
-  # Formula of "y ~ 0 + x" means no intercept:
-  result.lm = lm(t(geneBySampleValues) ~ 0 + samplesByCovariates, weights=sampleWeights)
-  coef = result.lm$coefficients
-  
-  # Assign covariate names to coefficients matrix
-  covarNames = colnames(samplesByCovariates)
-  isMatrixForm = is.matrix(coef)
-  if (isMatrixForm) {
-    rownames(coef) = covarNames
-  }
-  else {
[…]

### 2016-10-25 code/R/lib/corMatFDRthreshFunc.R

@@ -1,9 +1,0 @@
-# Function for FDR thresholding of correlation matrix
-corMatFDRthreshFunc <- function(cor_mat, indicesMask=NULL, MAX_FDR = 0.1) {    
-  if (is.null(indicesMask)) {indicesMask = 1:nrow(cor_mat)}
-  
-  fdr = rep(1.0, nrow(cor_mat))
-  fdr[indicesMask] = p.adjust(cor_mat$pvalue[indicesMask], method="fdr")
-  
-  return (fdr <= MAX_FDR)
-}

### 2016-10-25 code/R/lib/designMat2CovEffects.R

@@ -1,6 +1,0 @@
-designMat2CovEffects <- function(COVARIATES_MAP, sigCovars.Effects, type='mean'){
-  if (type == 'mean')
-    sapply(COVARIATES_MAP,function(x,y = sigCovars.Effects){return(mean(y[names(y) %in% x]))})
-  else if (type == 'sum') 
-    sapply(COVARIATES_MAP,function(x,y = sigCovars.Effects){return(sum(y[names(y) %in% x]))})
-}

### 2016-10-25 code/R/lib/designMatVarsToCovars.R

@@ -1,3 +1,0 @@
-designMatVarsToCovars <- function(COVARIATES_MAP, designMatVars) {
-  names(which(sapply(COVARIATES_MAP, function(v) any(v %in% designMatVars))))
-}

### 2016-10-25 code/R/lib/filterGeneSets.R

@@ -1,28 +1,0 @@
-filterGeneSets <- function(GeneLists, # List of lists
-                           genesInBackground, # background set of genes
-                           minSize = 10,
-                           maxSize = 1000){
- GeneLists = lapply(GeneLists, 
-                    function(x, genesInBackground){
-                      x = lapply(x, 
-                                 function(x, genesInBackground){
-                                   return(intersect(x, genesInBackground))
-                                 },
-                                 genesInBackground)
-                      return(x)
-                    }, 
-                    genesInBackground)
- 
- GeneLists = lapply(GeneLists, 
-                    function(x, minSize, maxSize){
-                      len = sapply(x, length)
-                      x = x[len>minSize && len<maxSize]
-                      return(x)
-                    },
-                    minSize,
-                    maxSize)
- len = sapply(GeneLists, length)
- GeneLists = GeneLists[len != 0]
- 
- return(GeneLists)
-}

### 2016-10-25 code/R/lib/g_legend.R

@@ -1,8 +1,0 @@
-require(ggplot2)
-g_legend<-function(a.gplot){
-  ## from http://stackoverflow.com/questions/11883844/inserting-a-table-under-the-legend-in-a-ggplot2-histogram
-  tmp <- ggplot2::ggplot_gtable(ggplot2::ggplot_build(a.gplot))
-  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
-  legend <- tmp$grobs[[leg]]
-  return(legend)
-}

### 2016-10-25 code/R/lib/getAssociationStatistics.R

@@ -1,111 +1,0 @@
-getAssociationStatistics <- function(COVARIATES, PVAL=0.05){
-
-  require(plyr)
-  # Get factor and continuous covariates
-  FactorCovariates <- names(COVARIATES)[sapply(COVARIATES,is.factor)]
-  ContCovariates <- setdiff(names(COVARIATES),FactorCovariates)
-  
-  # Convert factor covariates to numeric vector of factors
-  COVARIATES[,FactorCovariates] <- lapply(COVARIATES[,FactorCovariates],as.numeric)
-  COVARIATES[,FactorCovariates] <- lapply(COVARIATES[,FactorCovariates],as.factor)
-  
-  # Find association between factor covariates
-  if (length(FactorCovariates) > 1){
-    COVARIATES.FACTOR.CORRELATION = apply(expand.grid(FactorCovariates,FactorCovariates),
-                                          1, getFactorAssociationStatistics,
-                                          COVARIATES[,FactorCovariates])
-    COVARIATES.FACTOR.CORRELATION.ESTIMATE <- matrix(COVARIATES.FACTOR.CORRELATION['Estimate',],
-                                                     nrow=length(FactorCovariates),
-                                                     ncol=length(FactorCovariates))
-    COVARIATES.FACTOR.CORRELATION.PVAL <- matrix(COVARIATES.FACTOR.CORRELATION['Pval',],
-                                                 nrow=length(FactorCovariates),
-                                                 ncol=length(FactorCovariates))
-  
-    colnames(COVARIATES.FACTOR.CORRELATION.ESTIMATE) <- FactorCovariates
-    rownames(COVARIATES.FACTOR.CORRELATION.ESTIMATE) <- FactorCovariates
-    
-    colnames(COVARIATES.FACTOR.CORRELATION.PVAL) <- FactorCovariates
-    rownames(COVARIATES.FACTOR.CORRELATION.PVAL) <- FactorCovariates
-  } else if (length(FactorCovariates) == 1){
-    COVARIATES.FACTOR.CORRELATION.ESTIMATE <- as.data.frame(1)
-    colnames(COVARIATES.FACTOR.CORRELATION.ESTIMATE) <- FactorCovariates
[…]

### 2016-10-25 code/R/lib/getCovariatesMap.R

@@ -1,16 +1,0 @@
-getCovariatesMap <- function(DESIGN_AND_COVARS) {
-  FACTORS_AND_LEVELS = DESIGN_AND_COVARS$factorsLevels
-  COVARIATES_MAP = sapply(1:length(FACTORS_AND_LEVELS), function(i) {paste(deparse(as.name(names(FACTORS_AND_LEVELS)[i]), backtick=TRUE), FACTORS_AND_LEVELS[[i]], sep="")}, simplify=FALSE)
-  names(COVARIATES_MAP) = names(FACTORS_AND_LEVELS)
-  
-  numericCovars = DESIGN_AND_COVARS$numericCovars
-  names(numericCovars) = numericCovars
-  numericCovars = getSymbolicNamesList(numericCovars)
-  COVARIATES_MAP = c(COVARIATES_MAP, numericCovars)
-  
-  # Re-order as originally given:
-  COVARIATES_MAP = COVARIATES_MAP[DESIGN_AND_COVARS$covariates]
-  
-  return(COVARIATES_MAP)
-}
-

### 2016-10-25 code/R/lib/getDesignMat2Fact.R

@@ -1,7 +1,0 @@
-getDesignMat2Fact <- function(designMat, FactorCovariates){
-  ind <- grep(paste(FactorCovariates,collapse='|'),colnames(designMat))
-  tmp <- as.data.frame(designMat)
-  
-  tmp[,ind] <- lapply(tmp[,ind],function(cols,FactorCovariates){cols <- factor(as.character(cols))})
-  return(tmp)                      
-}

### 2016-10-25 code/R/lib/getDesignMatrix.R

@@ -1,55 +1,0 @@
-# Function to optain desing matrix (modified from covairates pipeline of Menachem Former)
-getDesignMatrix <- function(covariatesDataFrame, Intercept = T, RELEVELS=list()) {
-  
-  ROWNAMES = rownames(covariatesDataFrame)
-  COLNAMES = colnames(covariatesDataFrame)
-  
-  FACTOR_COVARIATE_NAMES <- names(covariatesDataFrame)[sapply(covariatesDataFrame,is.factor)]
-  FACTOR_COVARIATE_NAMES = setdiff(FACTOR_COVARIATE_NAMES, FACTOR_COVARIATE_NAMES[!(FACTOR_COVARIATE_NAMES %in% colnames(covariatesDataFrame))])
-  NUMERIC_COVARIATE_NAMES = setdiff(COLNAMES, FACTOR_COVARIATE_NAMES)
-  
-  # Ensure the factors are in fact of type factor, and the quantitative variables are numeric:
-  covariatesDataFrame = as.data.frame( lapply(colnames(covariatesDataFrame), function(column) {if (column %in% FACTOR_COVARIATE_NAMES) {fac = as.factor(covariatesDataFrame[, column]); if (column %in% names(RELEVELS)) {fac = relevel(fac, ref=RELEVELS[[column]])}; return(fac)} else {return(as.numeric(covariatesDataFrame[, column]))}}) )
-  rownames(covariatesDataFrame) = ROWNAMES
-  colnames(covariatesDataFrame) = COLNAMES
-  
-  contra = NULL
-  MAX_NUM_CATS = Inf
-  catData = covariatesDataFrame[, FACTOR_COVARIATE_NAMES, drop=FALSE]
-  if (ncol(catData) > 0) {
-    numCats = sapply(colnames(catData), function(col) nlevels(factor(catData[, col])))
-    EXCLUDE_CATEGORICAL_COLS = names(numCats)[numCats <= 1 | numCats > MAX_NUM_CATS]
-    if (!is.null(EXCLUDE_CATEGORICAL_COLS) && length(EXCLUDE_CATEGORICAL_COLS) > 0) {
-      warning(paste("Excluding categorical variables with less than 2", ifelse(is.infinite(MAX_NUM_CATS), "", paste(" or more than ", MAX_NUM_CATS, sep="")), " categories: ", paste(paste("'", EXCLUDE_CATEGORICAL_COLS, "'", sep=""), collapse=", "), sep=""))
[…]

### 2016-10-25 code/R/lib/getFactorAssociationStatistics.R

@@ -1,14 +1,0 @@
-getFactorAssociationStatistics <- function(factorNames,COVARIATES, na.action='remove'){
-  
-  require(vcd)
-  
-  if (na.action == "remove")
-    COVARIATES = na.omit(COVARIATES[,factorNames])
-  
-  fac1 = as.factor(COVARIATES[,1])
-  fac2 = as.factor(COVARIATES[,2])
-  
-  stats = vcd::assocstats(xtabs(~fac1+fac2))
-  
-  return(c(Estimate=stats$cramer,Pval=stats$chisq_tests['Pearson','P(> X^2)']))
-}

### 2016-10-25 code/R/lib/getFactorContAssociationStatistics.R

@@ -1,16 +1,0 @@
-# Find Inter Class Correlation between factor and continuous covariates
-# Inspired from http://stats.stackexchange.com/questions/108007/correlations-with-categorical-variables
-getFactorContAssociationStatistics <- function(factorContNames,COVARIATES, na.action='remove', 
-                                               alpha = 0.05){
-  require(psych)
-  factorContNames = factorContNames %>% unlist %>% as.character()
-  
-  if (na.action == "remove")
-    COVARIATES = na.omit(COVARIATES[,factorContNames])
-  
-  COVARIATES = sapply(COVARIATES, as.numeric)
-  stats = psych::ICC(COVARIATES, missing = T, alpha = alpha)
-  
-  return(data.frame(Estimate = stats$results['Single_fixed_raters','ICC'],
-                    Pval = stats$results['Single_fixed_raters','p']))
-}

### 2016-10-25 code/R/lib/getGeneFilteredGeneExprMatrix.R

@@ -1,40 +1,0 @@
-# Function to convert counts to cpm and filter cpm counts matrix
-getGeneFilteredGeneExprMatrix <- function(genesBySamplesCounts,ONLY_USE_GENES=NULL,
-                                          MIN_GENE_CPM=1,
-                                          MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5,
-                                          EDGER_NORMALIZATION.calcNormFactors.method = "none", # Choices are: "TMM", "RLE", "upperquartile", "none" [see edgeR::calcNormFactors()]
-                                          EDGER_NORMALIZATION.keep.lib.sizes = TRUE,
-                                          verbose=FALSE) {
-  if (!is.null(ONLY_USE_GENES)) {
-    useGenes = colnames(genesBySamplesCounts)
-    useGenes = useGenes[useGenes %in% ONLY_USE_GENES]
-    genesBySamplesCounts = genesBySamplesCounts[, useGenes]
-    if (verbose) {
-      writeLines(paste("\nLimiting expression data to ", length(useGenes), " genes specified by the ONLY_USE_GENES parameter.", sep=""))
-    }
-  }
-  
-  # Make edgeR object
-  MATRIX.ALL_GENES = DGEList(counts=genesBySamplesCounts, genes=rownames(genesBySamplesCounts))
-  
-  # Keep genes with at least MIN_GENE_CPM count-per-million reads (cpm) in at least (MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM)% of the samples:
-  MATRIX.ALL_GENES.CPM = cpm(MATRIX.ALL_GENES)
-  MATRIX.ALL_GENES.CPM[is.nan(MATRIX.ALL_GENES.CPM)] = 0
-  fracSamplesWithMinCPM = rowMeans(MATRIX.ALL_GENES.CPM >= MIN_GENE_CPM)
-  isNonLowExpr = fracSamplesWithMinCPM >= MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM
-  
-  MATRIX.NON_LOW_GENES = MATRIX.ALL_GENES[isNonLowExpr, ,keep.lib.sizes=EDGER_NORMALIZATION.keep.lib.sizes]
[…]

### 2016-10-25 code/R/lib/getSymbolicNamesList.R

@@ -1,3 +1,0 @@
-getSymbolicNamesList <- function(strings) {
-  lapply(strings, function(v) deparse(as.name(v), backtick=TRUE))
-}

### 2016-10-25 code/R/lib/getUpDownGenes.R

@@ -1,23 +1,0 @@
-getUpDownGenes <- function(PVAL,logFC, Genes, contName,PVAL_CUTOFF = 0.05, FC_CUTOFF = 1e-10){
-  ind.pos <- which(PVAL<=PVAL_CUTOFF & logFC >= FC_CUTOFF)
-  if (length(ind.pos) > 0){
-    tmp.pos <- data.frame(GeneSymbol = Genes[ind.pos],
-                          logFC = as.numeric(logFC[ind.pos]),
-                          adj.P.value = as.numeric(PVAL[ind.pos]),
-                          Comparison = paste0(contName,'__up'))
-  } else {
-    tmp.pos <- data.frame()
-  }
-  
-  ind.neg <- which(PVAL<=PVAL_CUTOFF & logFC <= -FC_CUTOFF)
-  if (length(ind.neg) > 0){
-    tmp.neg <- data.frame(GeneSymbol = Genes[ind.neg],
-                          logFC = as.numeric(logFC[ind.neg]),
-                          adj.P.value = as.numeric(PVAL[ind.neg]),
-                          Comparison = paste0(contName,'__down'))  
-  } else {
-    tmp.neg <- data.frame()
-  }
-  
-  return(rbind(tmp.pos,tmp.neg))
-}

### 2016-10-25 code/R/lib/ggMosaicPlot.R

@@ -1,20 +1,0 @@
-# Function to mosaic plot two factor variables
-ggMosaicPlot <- function(var1, var2, var1Name = 'var1', var2Name = 'var2'){
-  library(ggplot2)
-  levVar1 <- length(levels(var1))
-  levVar2 <- length(levels(var2))
-  
-  jointTable <- prop.table(table(var1, var2))
-  plotData <- as.data.frame(jointTable)
-  colnames(plotData)[1:2] <- c(var1Name,var2Name)
-  
-  plotData$marginVar1 <- prop.table(table(var1))
-  plotData$var2Height <- plotData$Freq / plotData$marginVar1
-  plotData$var1Center <- c(0, cumsum(plotData$marginVar1)[1:levVar1 -1]) +  plotData$marginVar1 / 2
-  
-  p <- ggplot(plotData, aes(var1Center, var2Height))
-  p <- p + geom_bar(stat = "identity", aes(width = marginVar1, fill = var2), col = "Black")
-  p <- p + geom_text(aes(label = as.character(var1), x = var1Center, y = 1.05, angle = 30))
-  p <- p + xlab(var1Name) + ylab(var2Name)
-  p
-}

### 2016-10-25 code/R/lib/ggheatmap.R

@@ -1,62 +1,0 @@
-##' generate a heatmap + dendrograms, ggplot2 style
-##'
-##' @param x data matrix
-##' @param hm.colours vector of colours (optional)
-##' @return invisibly returns a list of ggplot2 objects. Display them with ggheatmap.show()
-##' @author Chris Wallace
-##' @export
-##' @examples
-##' ## test run
-##' ## simulate data
-##' library(mvtnorm)
-##' sigma=matrix(0,10,10)
-##' sigma[1:4,1:4] <- 0.6
-##' sigma[6:10,6:10] <- 0.8
-##' diag(sigma) <- 1
-##' X <- rmvnorm(n=100,mean=rep(0,10),sigma=sigma)
-##'  
-##' ## make plot
-##' p <- ggheatmap(X)
-##'  
-##' ## display plot
-##' ggheatmap.show(p)
-require(ggdendro)
-require(ggplot2)
-ggheatmap <- function(x,
-                      hm.colours=my.colours) {
-  # Sourced from https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R
-  if(is.null(colnames(x)))
-    colnames(x) <- sprintf("col%s",1:ncol(x))
-  if(is.null(rownames(x)))
-    rownames(x) <- sprintf("row%s",1:nrow(x))
-  ## plot a heatmap
-  ## x is an expression matrix
-  row.hc <- hclust(dist(x), "ward")
-  col.hc <- hclust(dist(t(x)), "ward")
-  row.dendro <- ggdendro::dendro_data(as.dendrogram(row.hc),type="rectangle")
-  col.dendro <- ggdendro::dendro_data(as.dendrogram(col.hc),type="rectangle")
-  
-  ## dendro plots
-  col.plot <- mydplot(col.dendro, col=TRUE, labels=TRUE) +
-    ggplot2::scale_x_continuous(breaks = 1:ncol(x),labels=col.hc$labels[col.hc$order]) +
-    ggplot2::theme(plot.margin = unit(c(0,0,0,0), "lines"))
-  row.plot <- mydplot(row.dendro, row=TRUE, labels=FALSE) +
-    ggplot2::theme(plot.margin = unit(rep(0, 4), "lines"))
-  
-  ## order of the dendros
-  col.ord <- match(col.dendro$labels$label, colnames(x))
-  row.ord <- match(row.dendro$labels$label, rownames(x))
[…]

### 2016-10-25 code/R/lib/ggheatmap.show.R

@@ -1,37 +1,0 @@
-##' Display a ggheatmap
-##'
-##' this function sets up some viewports, and tries to plot the dendrograms to line up with the heatmap
-##' @param L a list with 3 named plots: col, row, centre, generated by ggheatmap
-##' @param col.width,row.width number between 0 and 1, fraction of the device devoted to the column or row-wise dendrogram. If 0, don't print the dendrogram
-##' @return no return value, side effect of displaying plot in current device
-##' @export
-##' @author Chris Wallace
-require(grid)
-ggheatmap.show <- function(L, col.width=0.2, row.width=0.2) {
-  # Function sourced from https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R
-  grid::grid.newpage()
-  top.layout <- grid::grid.layout(nrow = 2, ncol = 2,
-                                  widths = unit(c(1-row.width,row.width), "null"),
-                                  heights = unit(c(col.width,1-row.width), "null"))
-  grid::pushViewport(grid::viewport(layout=top.layout))
-  if(col.width>0)
-    print(L$col, vp=grid::viewport(layout.pos.col=1, layout.pos.row=1))
-  if(row.width>0)
-    print(L$row, vp=grid::viewport(layout.pos.col=2, layout.pos.row=2))
-  ## print centre without legend
-  print(L$centre +
-          ggplot2::theme(axis.line=element_blank(),
-                       axis.text.x=element_blank(),axis.text.y=element_blank(),
-                       axis.ticks=element_blank(),
-                       axis.title.x=element_blank(),axis.title.y=element_blank(),
-                       legend.position="none",
-                       panel.background=element_blank(),
-                       panel.border=element_blank(),panel.grid.major=element_blank(),
-                       panel.grid.minor=element_blank(),plot.background=element_blank()),
-        vp=grid::viewport(layout.pos.col=1, layout.pos.row=2))
-  ## add legend
-  legend <- g_legend(L$centre)
-  grid::pushViewport(grid::viewport(layout.pos.col=2, layout.pos.row=1))
-  grid::grid.draw(legend)
-  grid::upViewport(0)
-}

### 2016-10-25 code/R/lib/linColumnFinder.R

@@ -1,38 +1,0 @@
-# Modified from http://stackoverflow.com/questions/13088770/how-to-write-linearly-dependent-column-in-a-matrix-in-terms-of-linearly-independ
-# Function to find linearly dependednt columns of a matrix
-linColumnFinder <- function(mat){
-
-  mat[is.na(mat)] = 0
-  
-  # If the matrix is full rank then we're done
-  if(qr(mat)$rank == ncol(mat)){
-    return(list(indepCols = seq(1,ncol(mat),1), relations = "Matrix is of full rank"))
-  }
-
-  m <- ncol(mat)
-  # cols keeps track of which columns are linearly independent
-  cols <- 1
-  All.message <- c()
-  for(i in seq(2, m)){
-    ids <- c(cols, i)
-    mymat <- mat[, ids]
-    if(qr(mymat)$rank != length(ids)){
-      # Regression the column of interest on the previous columns to figure out the relationship
-      o <- lm(mat[,i] ~ as.matrix(mat[,cols]) + 0)
-      # Construct the output message
-      start <- paste0(colnames(mat)[i], " = ")
-      # Which coefs are nonzero
-      nz <- !(abs(coef(o)) <= .Machine$double.eps^0.5)
-      tmp <- colnames(mat)[cols[nz]]
-      vals <- paste(coef(o)[nz], tmp, sep = "*", collapse = " + ")
-      message <- paste0(start, vals)      
-      All.message <- c(All.message,message)
-    } else {
-      # If the matrix subset was of full rank
-      # then the newest column in linearly independent
-      # so add it to the cols list
-      cols <- ids
-    }
-  }
-  return(list(indepCols = cols, relations = All.message))
-}

### 2016-10-25 code/R/lib/makeSynapseLink.R

@@ -1,26 +1,0 @@
-makeSynapseLink <- function(fileId, parentId, linkName = NULL, annotations = list()){
-  library(synapseClient)
-  
-  # Get synapse properties of the source object
-  obj = synapseClient::synGet(fileId, downloadFile=F)
-  if (is.null(linkName))
-    linkName = obj@properties$name
-  
-  # Create a linksTo list
-  linksTo = list(
-    targetVersionNumber = obj@properties$versionNumber,
-    targetId = obj@properties$id
-  )
-  
-  # Create a linkClass list
-  linkClass = list(
-    concreteType  = 'org.sagebionetworks.repo.model.Link',
-    linksTo	= linksTo,
-    entityType = 'org.sagebionetworks.repo.model.Link',
-    parentId	= parentId,
-    name = linkName,
-    annotations	= annotations,
-    linksToClassName = 'org.sagebionetworks.repo.model.FileEntity')
-  
-  synapseClient::synRestPOST('/entity',linkClass)
-}

### 2016-10-25 code/R/lib/mydplot.R

@@ -1,22 +1,0 @@
-require(ggplot2)
-require(ggdendro)
-mydplot <- function(ddata, row=!col, col=!row, labels=col) {
-  ## plot a dendrogram
-  yrange <- range(ddata$segments$y)
-  yd <- yrange[2] - yrange[1]
-  nc <- max(nchar(as.character(ddata$labels$label)))
-  tangle <- if(row) { 0 } else { 90 }
-  tshow <- col
-  p <- ggplot2::ggplot() +
-    ggplot2::geom_segment(data=segment(ddata), aes(x=x, y=y, xend=xend, yend=yend)) +
-    ggplot2::labs(x = NULL, y = NULL) + ggdendro::theme_dendro()
-  if(row) {
-    p <- p +
-      ggplot2::scale_x_continuous(expand=c(0.5/length(ddata$labels$x),0)) +
-      ggplot2::coord_flip()
-  } else {
-    p <- p +
-      ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1))
-  }
-  return(p)
-}

### 2016-10-25 code/R/lib/parallelDuplicateCorrelation.R

@@ -1,97 +1,0 @@
-parallelDuplicateCorrelation <- function(object, design = NULL, ndups = 2, spacing = 1, block = NULL,
-                                         trim = 0.15, weights = NULL) {  
-  y <- getEAWP(object)
-  M <- y$exprs
-  ngenes <- nrow(M)
-  narrays <- ncol(M)
-  if (is.null(design)) 
-    design <- y$design
-  if (is.null(design)) 
-    design <- matrix(1, ncol(y$exprs), 1)
-  else {
-    design <- as.matrix(design)
-    if (mode(design) != "numeric") 
-      stop("design must be a numeric matrix")
-  }
-  if (nrow(design) != narrays) 
-    stop("Number of rows of design matrix does not match number of arrays")
-  ne <- nonEstimable(design)
-  if (!is.null(ne)) 
-    cat("Coefficients not estimable:", paste(ne, collapse = " "), 
-        "\n")
-  nbeta <- ncol(design)
-  if (missing(ndups) && !is.null(y$printer$ndups)) 
-    ndups <- y$printer$ndups
-  if (missing(spacing) && !is.null(y$printer$spacing)) 
-    spacing <- y$printer$spacing
-  if (missing(weights) && !is.null(y$weights)) 
-    weights <- y$weights
-  if (!is.null(weights)) {
-    weights <- as.matrix(weights)
-    if (any(dim(weights) != dim(M))) 
-      weights <- array(weights, dim(M))
-    M[weights < 1e-15] <- NA
-    weights[weights < 1e-15] <- NA
-  }
-  if (is.null(block)) {
-    if (ndups < 2) {
-      warning("No duplicates: correlation between duplicates not estimable")
-      return(list(cor = NA, cor.genes = rep(NA, nrow(M))))
-    }
-    if (is.character(spacing)) {
-      if (spacing == "columns") 
-        spacing <- 1
-      if (spacing == "rows") 
-        spacing <- object$printer$nspot.c
-      if (spacing == "topbottom") 
-        spacing <- nrow(M)/2
-    }
-    Array <- rep(1:narrays, rep(ndups, narrays))
-  }
-  else {
-    ndups <- 1
-    nspacing <- 1
-    Array <- block
-  }
-  if (is.null(block)) {
-    M <- unwrapdups(M, ndups = ndups, spacing = spacing)
-    ngenes <- nrow(M)
[…]

### 2016-10-25 code/R/lib/plotCorWithCompare.R

@@ -1,80 +1,0 @@
-# Function to plot correlation matrix
-plotCorWithCompare <- function(plotCor, title=NULL, MARK_CORRELATIONS_NAME=NULL, markColumnsAsMissing=NULL) {
-  # Mark the X-axis labels based on user-defined pattern:
-  markMissingEntries = rep(FALSE, nrow(plotCor))
-  
-  use.mark.x = FALSE
-  if (!is.null(markColumnsAsMissing)) {
-    COVAR_NAMES = as.character(unique(levels(plotCor$COVAR)))
-    mark.x = rep(FALSE, length(COVAR_NAMES))
-    names(mark.x) = COVAR_NAMES
-    mark.x[markColumnsAsMissing] = TRUE
-    
-    plot.x.labels = levels(plotCor$COVAR)[levels(plotCor$COVAR) %in% plotCor$COVAR]
-    use.mark.x = as.character(mark.x[plot.x.labels])
-    use.mark.x[is.na(use.mark.x)] = FALSE
-    
-    markMissingEntries[plotCor$COVAR %in% markColumnsAsMissing] = TRUE
-  }
-  plotCor$markMissingEntries = markMissingEntries
-  
-  use.face.x = ifelse(use.mark.x, "bold.italic", "plain")
-  use.color.x = ifelse(use.mark.x, "darkgray", "black")
-  
-  plotSingle = FALSE
-  plot_aes = aes(COVAR, COMPARE, fill=r, alpha=as.factor(markMissingEntries))
-  if (length(unique(plotCor$r)) <= 1) {
-    plot_aes = aes(COVAR, COMPARE, fill=factor(r))
-    plotSingle = TRUE
-  }
-  
-  # Reverse the Y-axis:
-  plotCor$COMPARE = factor(plotCor$COMPARE, levels=rev(levels(plotCor$COMPARE)))
-  
-  alphaVals = c('TRUE'=0.85, 'FALSE'=1)
-  gRes = ggplot(plotCor, plot_aes) + geom_tile() + scale_alpha_manual(values=alphaVals, guide="none")
-  
-  gRes = gRes + xlab("") + ylab("")
-  gRes = gRes + theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5, size=14, face=use.face.x, color=use.color.x), axis.text.y=element_text(size=14, color="black", hjust=0))
-  gRes = gRes + theme(panel.grid.major.x=element_line(color="black", linetype="dashed")) + theme(panel.grid.minor.x=element_blank())
[…]

### 2016-10-25 code/R/lib/qc.R

@@ -1,25 +1,0 @@
-corr_two_nominal_vars = function(v, dat){
-    v=as.character(v)
-    x = as.factor(as.character(dat[,v[1]]))
-    y = as.factor(as.character(dat[,v[2]]))
-    dd = data.frame(x=x,y=y)
-    dd = dd[rowSums(is.na(dd))==0,]
-    sm = chisq.test(dd$y, dd$x) # get a p-value
-    r= assocstats(xtabs(~x+y,data=dd))$cramer # get a R^2 value for categorical vars
-    pval = sm$p.value
-    list(r,pval)
-}
-
-most_variable = function(mat, ntop=100){
-    require(statmod)
-    # now let's recalculate the most variable genes with the winsorized matrix (wed)
-    means = rowMeans(mat); vars = apply(mat,1,var); cv2 <- vars/means^2
-    useForFit <- means >= unname( quantile( means[ which( cv2 > .3 ) ], .95 ) )
-    fit <- glmgam.fit( cbind( a0 = 1, a1tilde = 1/means[useForFit] ),cv2[useForFit] )
-    afit <- fit$coef["a1tilde"]/means+fit$coef["a0"]
-    xg <- exp(seq( min(mat), max(mat), length.out=1000 ))
-    vfit <- fit$coef["a1tilde"]/xg+fit$coef["a0"]
-
-    varFitRatio <- vars/(afit*means^2)
-    order(varFitRatio,decreasing=T)[1:ntop]
-}

### 2016-10-25 code/R/lib/rownameToFirstColumn.R

@@ -1,7 +1,0 @@
-rownameToFirstColumn <- function(DF,colname){
-  DF <- as.data.frame(DF)
-  DF[,colname] <- row.names(DF)
-  DF <- DF[,c(dim(DF)[2],1:(dim(DF)[2]-1))]
-  return(DF)
-}
-  

### 2016-10-25 code/R/lib/runPCA.R

@@ -1,15 +1,0 @@
-# Function to run principal component analysis
-runPCA <- function(genesBySamples, SCALE_DATA_FOR_PCA = TRUE, MIN_PVE_PCT_PC = 1.0) {
-  
-  # estimate variance in data by PC:
-  pca.res <- prcomp(t(genesBySamples), center=TRUE, scale.=SCALE_DATA_FOR_PCA, retx=TRUE)
-  
-  # examine how much variance is explained by PCs, and consider those with PVE >= (MIN_PVE_PCT_PC %):
-  pc.var <- pca.res$sdev^2
-  pve <- 100 * (pc.var / sum(pc.var))  
-  npca <- max(1,length(which(pve >= MIN_PVE_PCT_PC)))
-  
-  samplePCvals <- pca.res$x[, 1:npca, drop=FALSE]
-  
-  list(samplePCvals=samplePCvals, pve=pve)
-}

### 2016-10-25 code/R/lib/runPCAandPlotCorrelations.R

@@ -1,48 +1,0 @@
-# Function to run principal component analysis and plot correlations
-runPCAandPlotCorrelations <- function(genesBySamples, samplesByCovariates, dataName, isKeyPlot=FALSE, 
-                                      SCALE_DATA_FOR_PCA = TRUE, MIN_PVE_PCT_PC = 1.0, CORRELATION_TYPE = 'pearson',
-                                      ALSO_PLOT_ALL_COVARS_VS_PCA = TRUE, MAX_NUM_LEVELS_PER_COVAR = 50) {
-      
-  
-  title = paste(ifelse(SCALE_DATA_FOR_PCA, "S", "Un-s"), "caled ", dataName, " ", " data in PCA; PVE >= ", MIN_PVE_PCT_PC, "%; ", CORRELATION_TYPE, " correlations ", sep="")
-  writeLines(paste("\nRunning PCA and calculating correlations for:\n", title, sep=""))
- 
-  pcaRes <- runPCA(genesBySamples=genesBySamples, 
-                   SCALE_DATA_FOR_PCA=SCALE_DATA_FOR_PCA, 
-                   MIN_PVE_PCT_PC=MIN_PVE_PCT_PC)
-  
-  samplePCvals <- pcaRes$samplePCvals
-  pve <- pcaRes$pve
-  
-  npca <- ncol(samplePCvals)
-  
-  colnames(samplePCvals) = paste(colnames(samplePCvals), " (", sprintf("%.2f", pve[1:npca]), "%)", sep="")
-  
-  # Find covariates without any missing data
-  samplesByFullCovariates = samplesByCovariates[, which(apply(samplesByCovariates, 2, 
-                                                              function(dat) all(!is.na(dat))))]
-  EXCLUDE_VARS_FROM_FDR = setdiff(colnames(samplesByCovariates), colnames(samplesByFullCovariates))
-  
-  add_PC_res = list()
-  significantCovars = c()
-  
-  LOOP_PLOT_ALL_COVARS = FALSE
-  if (ALSO_PLOT_ALL_COVARS_VS_PCA) { LOOP_PLOT_ALL_COVARS = unique(c(LOOP_PLOT_ALL_COVARS, TRUE)) }
-  
-  for (PLOT_ALL_COVARS in LOOP_PLOT_ALL_COVARS) {
-    corrRes = calcCompleteCorAndPlot(samplePCvals, 
-                                     samplesByCovariates, 
-                                     CORRELATION_TYPE, 
-                                     title, 
[…]

### 2016-10-25 code/Rmd/MSSM_BM_36_RNASeq_Reprocessed_SynBAM_StarHTSeq.Rmd
Added markdown to process BM36 from synapse BAM with picard star htseq pipeline

--- 2016-09-20 code/Rmd/MSSM_BM_36_RNASeq_Realigned_StarHTSeq2.Rmd
@@
 ---
-title: "Covariates analysis of MSBB RNASeq data of region BM 36 (realigned from FASTQ extracted source BAM in minerva with STAR-HTSeq)"
+title: "Covariates analysis of MSBB RNASeq data of region BM 36 (realigned from synapse BAM with Picard-STAR-HTSeq)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_BM_36_RNASeq_Realigned_StarHTSeq2.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_BM_36_RNASeq_Reprocessed_SynBAM_StarHTSeq.Rmd",
                                  parentId ="syn5570248",
-                                 entityName = 'MSBB BM 36 RNASeq RealignedFromSource BAM Picard StarHTSeq')
+                                 entityName = "MSBB BM 36 RNASeq RealignedFromSynBAM PicardStarHTSeq")
 ```
@@
 
-thisFileName <- 'MSSM_BM_36_RNASeq_Realigned_StarHTSeq2.Rmd'
+thisFileName <- 'MSSM_BM_36_RNASeq_Reprocessed_SynBAM_StarHTSeq.Rmd'
 
@@
 # Download expression data
-COUNT_ID <- 'syn7254146'
+COUNT_ID <- 'syn7425130'
 ALL_USED_IDs <- COUNT_ID
@@
 # Get sample ids
-SampleID = data.frame(SampleID = colnames(COUNT),
-                      ID = colnames(COUNT)) %>% 
-  tidyr::separate(ID, c('A','B','ID', 'C'), sep = '_')
+ind.reseq = grep('resequenced', colnames(COUNT))
+ind.all = setdiff(1:dim(COUNT)[2], ind.reseq)
 
-SampleID = SampleID %>%
+SampleID = rbind(data.frame(SampleID = colnames(COUNT)[ind.all],
+                      ID = colnames(COUNT)[ind.all]) %>% 
+  tidyr::separate(ID, c('A','B','ID', 'C'), sep = '_') %>%
+  dplyr::mutate(ID = gsub('.accepted','',ID)) %>%
   dplyr::select(SampleID, ID) %>%
@@
                 Sample.ID = SampleID,
-                SampleID = paste0('BM_36_',ID))
+                SampleID = paste0('BM_36_',ID)),
+  data.frame(SampleID = colnames(COUNT)[ind.reseq],
+             ID = colnames(COUNT)[ind.reseq]) %>% 
+  tidyr::separate(ID, c('A','B','ID', 'C'), sep = '_') %>%
+  dplyr::mutate(ID = gsub('.accepted','',ID)) %>%
+  dplyr::select(SampleID, ID) %>%
+  dplyr::mutate(ID = as.numeric(ID),
+                Sample.ID = SampleID,
+                SampleID = paste0('BM_36_Reseq_',ID)))
 
 # Sum counts from multiple lanes
-COUNT = dlply(SampleID, .(ID), .fun = function(x, count){
-  rowSums(count[,x$Sample.ID,drop=F]) %>%
-    rownameToFirstColumn('Gene.ID') %>%
-    plyr::rename(c('DF' = paste0('BM_36_',unique(x$ID)))) %>%
-    dplyr::select(-Gene.ID)
-}, COUNT) %>%
-  Reduce(cbind, .)
+# COUNT = dlply(SampleID, .(ID), .fun = function(x, count){
+#   rowSums(count[,x$Sample.ID,drop=F]) %>%
+#     rownameToFirstColumn('Gene.ID') %>%
+#     plyr::rename(c('DF' = paste0('BM_36_',unique(x$ID)))) %>%
+#     dplyr::select(-Gene.ID)
+# }, COUNT) %>%
+#   Reduce(cbind, .)
 COUNT = data.matrix(COUNT)
+
+rownames(SampleID) = SampleID$Sample.ID
+colnames(COUNT) = SampleID[colnames(COUNT), 'SampleID']
 
@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECHNICAL_ID
-METADATA_TECHNICAL <- read.table(synGet(METADATA_ID)@filePath,sep='\t',header=T, row.names=1) %>%
+METADATA_TECHNICAL <- read.table(synGet(METADATA_TECHNICAL_ID)@filePath,sep='\t',header=T, row.names=1) %>%
   rownameToFirstColumn('OldBarcode')
@@
   filter(!is.na(SampleID)) %>%
-  filter(!(ID %in% c(335, 338, 339, 341))) %>%
   unique
@@
 # Code
-CODE <- Folder(name = "MSBB BM 36 RNASeq RealignedFromSource BAM Picard StarHTSeq", parentId = parentId)
+CODE <- Folder(name = "MSBB BM 36 RNASeq RealignedFromSynBAM PicardStarHTSeq", parentId = parentId)
 CODE <- synStore(CODE)
@@
                    executed = thisFile, activityDescription = activityDescription)
-
 ```
@@
 |  Covariates               |  `r covar_obj$properties$id`    |
-|  logCPM                   |  `r exp1_obj$properties$id`     |
 |  Residual Expression      |  `r expr_obj$properties$id`     |

### 2016-10-26 code/Rmd/BM36_Compare_All_Data.Rmd
Added markdown to compare all BM36 datasets

---
title: "Compare different versions of BM 36 from MSSM"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knitr)
library(githubr)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./BM36_Compare_All_Data.Rmd",
                                 parentId ="syn5570248",
                                 entityName = 'MSBB BM 36 Comparison between all versions')
```

### Download all versions of LogCPM from synapse
```{r data.download}
library(CovariateAnalysis)
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(synapseClient)
library(knitr)
library(githubr)

synapseLogin()

lcpm.fromPortal = read.table(synGet('syn7255229')@filePath, row.names = 1, header = T)
lcpm.fromFASTQ = read.table(synGet('syn7255101')@filePath, row.names = 1, header = T)
lcpm.fromBAM = read.table(synGet('syn7255067')@filePath, row.names = 1, header = T)
lcpm.fromSynapseBAM = read.table(synGet('syn7437407')@filePath, row.names = 1, header = T)
```

```{r curate}
Gene.ID = data.frame(Gene.ID = c(rownames(lcpm.fromPortal), 
                                 rownames(lcpm.fromFASTQ), 
                                 rownames(lcpm.fromBAM),
                                 rownames(lcpm.fromSynapseBAM))) %>%
  dplyr::mutate(GeneID = Gene.ID) %>%
  tidyr::separate(Gene.ID, c('ensembl_gene_id','position'), sep = '\\.')

tmp = filter(Gene.ID, GeneID %in% rownames(lcpm.fromBAM)) %>% unique
lcpm.fromBAM = WGCNA::collapseRows(lcpm.fromBAM, tmp$ensembl_gene_id, tmp$GeneID)$datETcollapsed

tmp = filter(Gene.ID, GeneID %in% rownames(lcpm.fromFASTQ)) %>% unique
lcpm.fromFASTQ = WGCNA::collapseRows(lcpm.fromFASTQ, tmp$ensembl_gene_id, tmp$GeneID)$datETcollapsed

tmp = filter(Gene.ID, GeneID %in% rownames(lcpm.fromSynapseBAM)) %>% unique
lcpm.fromSynapseBAM = WGCNA::collapseRows(lcpm.fromSynapseBAM, tmp$ensembl_gene_id, tmp$GeneID)$datETcollapsed

genes = intersect(rownames(lcpm.fromSynapseBAM), rownames(lcpm.fromFASTQ)) %>% 
  intersect(rownames(lcpm.fromBAM)) %>%
  intersect(rownames(lcpm.fromPortal))

colnames(lcpm.fromPortal) = gsub('hB_RNA_','BM_36_',colnames(lcpm.fromPortal))
colnames(lcpm.fromSynapseBAM) = gsub('hB_RNA_','BM_36_',colnames(lcpm.fromSynapseBAM))
samples = intersect(colnames(lcpm.fromPortal), colnames(lcpm.fromFASTQ)) %>%
  intersect(colnames(lcpm.fromBAM)) %>%
  intersect(colnames(lcpm.fromSynapseBAM))

lcpm.fromSynapseBAM = lcpm.fromSynapseBAM[genes, samples]
lcpm.fromBAM = lcpm.fromBAM[genes, samples]
lcpm.fromFASTQ = lcpm.fromFASTQ[genes, samples]
lcpm.fromPortal = lcpm.fromPortal[genes, samples]
```

```{r correlate}
# par(mfrow = c(3,1))
cr.por.bam = stats::cor(lcpm.fromPortal, lcpm.fromBAM)
# hist(diag(cr.dep.bam), main = 'Correlation between sinaiDeposit and fromBAM')
# 
cr.por.fastq = stats::cor(lcpm.fromPortal, lcpm.fromFASTQ)
# hist(diag(cr.dep.fastq), main = 'Correlation between sinaiDeposit and fromFASTQ')
# 
cr.por.synbam = stats::cor(lcpm.fromPortal, lcpm.fromSynapseBAM)
# hist(diag(cr.dep.synbam), main = 'Correlation between sinaiDeposit and fromSynapseBAM')
# 
cr.fastq.synbam = stats::cor(lcpm.fromFASTQ, lcpm.fromSynapseBAM)
# hist(diag(cr.bam.fastq), main = 'Correlation between fromBAM and fromFASTQ')

cr.fastq.bam = stats::cor(lcpm.fromFASTQ, lcpm.fromBAM)

tmp = rbind(data.frame(Comp = 'Portal_vs_BAM', values = diag(cr.por.bam)),
            data.frame(Comp = 'Portal_vs_FASTQ', values = diag(cr.por.fastq)),
            data.frame(Comp = 'Portal_vs_synBAM', values = diag(cr.por.synbam)),
            data.frame(Comp = 'FASTQ_vs_synBAM', values = diag(cr.fastq.synbam)),
            data.frame(Comp = 'FASTQ_vs_BAM', values = diag(cr.fastq.bam)))
p = ggplot(tmp, aes(x = Comp, y = values)) + geom_boxplot()
p = p + ylab('Correlation') + xlab('')
p
```

### 2016-11-04 code/Rmd/BM36_Compare_All_Data.Rmd
Minor changes

@@
 synapseLogin()
+
+knitr::opts_chunk$set(
+  echo=FALSE,
+  warning=FALSE,
+  message=FALSE,
+  error = FALSE,
+  tidy = FALSE,
+  cache = TRUE)
 
@@
 p
+
+plot(diag(cr.fastq.synbam), diag(cr.fastq.bam), xlab = 'FASTQ_vs_synBAM', ylab = 'FASTQ_vs_BAM')
 ```

### 2016-11-04 code/Rmd/CMC_DLPFC_StarAligned_TMM.Rmd
Minor sanity modifications for the DLPFC markdown

@@
 knit2synapse::knitToFolderEntity(file = "./CMC_DLPFC_StarAligned_TMM.Rmd",
-                                 parentId ="syn5569976",
+                                 parentId ="syn7501880",
                                  entityName = 'DLPFC - STAR Reprocessed - with TMM')
@@
 # Synapse parameters
-parentId = 'syn5569976';
+parentId = 'syn7501880';
 activityName = 'Covariate adjustments';
@@
 ```
+
 Correlation/association between covariates at an FDR <= 0.1
-```{r covariates.correlation, fig.width=10, fig.height=10}
+
+```{r covariates.correlation, fig.width=8, fig.height=8}
 COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
@@
 ```
+
 Explore metatdata
-```{r dataExplore, fig.width = 10, fig.height = 10}
+
+```{r dataExplore, fig.width = 6, fig.height = 14}
 my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
@@
 ### CPM Normalisation
-Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.
+Preprocess count matrix and metadata. Determine design matrix for normalisation and differential expression analysis.
 
@@
 ```
+Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples
+
 ### Normalisation (with NULL)
 Initial normalisation usign voom (with NULL design)
-```{r initial.voom.normalisation}
+```{r initial.voom.normalisation, fig.height=5, fig.width=5}
 # TMM normalisation
-TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix)
+TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')
 
@@
 # TMM normalisation
-TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix)
+TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')
 
@@
 Coexpression of genes 
-```{r coexp1, cache=FALSE}
+```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
 cr = cor(t(VOOM.GENE_EXPRESSION$E))
@@
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
-p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
 p
-
+```
+Tree based classification of samples
+```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
 # Eucledian tree based analysis
@@
 ```
+
 Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
@@
 ### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "Dx" is chosen as the primary variable of interest
-```{r iterativeNorm}
+Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix.
+Here "Dx" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis)
+
+Adding Instituion and Gender apriori to variable selection as they are correlated with Dx
+Also adding first 5 eigen vectors computed from ancestry
+```{r iterativeNorm, results='asis'}
 # Primary variable of interest
-postAdjustCovars = 'RIN';
+postAdjustCovars = c('Dx', 'Institution', 'Gender');
 
@@
 }
-```
-
-### Normlaisation (Custom design)
-To be consistent with DLPFC Nature Neuroscience article we performed a custom adjustment with Dx, RIN, RIN2, Age of Death, PMI, Library Batch, Institution, Gender, EV1, EV2, EV3, EV4, EV5
-```{r custom.design}
-# Add primary variable to the list (in case not chosen by iterative method)
-postAdjustCovars = c('Dx', c(FactorCovariates, ContCovariates)) %>% unique()
-
-writeLines(paste('Using following covariates in the model:',
-                 paste(postAdjustCovars, collapse=', '),
-                 'as fixed effects'))
-
-# Post adjusted design matrix
-DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
-
-# Estimate voom weights
-VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
-
-# Fit linear model using new weights and new design
-VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
-
-# Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
 
@@
 p
+```
 
+```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
 # Eucledian tree based analysis

### 2016-11-05 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM.Rmd
Added markdown to process DLPFC and ACC together with a random effects model

---
title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC and DLPFC with TMM normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_And_DLPFC_StarAligned_TMM.Rmd",
                                 parentId ="syn7501880",
                                 entityName = 'ACC and DLPFC - STAR Reprocessed - with TMM')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get it from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(reshape2)
library(limma)
library(Biobase)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)

library(synapseClient)
library(knitr)
library(githubr)

library(foreach)
library(doParallel)

cl = makeCluster(14)
registerDoParallel(cl)

synapseLogin()

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn7501880';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM normalisation (ACC and DLPFC)';

thisFileName <- 'CMC_ACC_And_DLPFC_StarAligned_TMM.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download reprocessed counts (ACC)
COUNT_ACC_ID = 'syn7217917';
ALL_USED_IDs = COUNT_ACC_ID
COUNT_ACC = read.table(synGet(COUNT_ACC_ID)@filePath, header=T, sep='\t', check.names = F, nrows = 56632) %>%
  dplyr::select(-chr, -start, -end, -strand, -length)

# Download reprocessed counts (DLPFC)
COUNT_DLPFC_ID = 'syn7217916';
ALL_USED_IDs[length(ALL_USED_IDs)+1] = COUNT_DLPFC_ID
COUNT_DLPFC = read.table(synGet(COUNT_DLPFC_ID)@filePath, header=T, sep='\t', check.names = F, nrows = 56632) %>%
  dplyr::select(-chr, -start, -end, -strand, -length)

COUNT = join_all(list(COUNT_ACC, COUNT_DLPFC), type = 'full')
rownames(COUNT) = COUNT$geneID
COUNT$geneID = NULL
COUNT = data.matrix(COUNT)

# Get merged metadata
METADATA_ID = 'syn2299154'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA = read.csv(synGet(METADATA_ID)@filePath, fill = T, header = F)

# Get ancestry vector calculated using gemtools
ANCESTRY_ID = 'syn2511399'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = ANCESTRY_ID
ANCESTRY = read.table(synGet(ANCESTRY_ID)@filePath, fill = T, header = T, sep = '\t') %>%
  plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'GenotypingSampleID'))

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", 
                          "Dx", "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?", "ACC_RNA_report: Library Batch",
                          "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
                          'DNA_genotyping_report: Genotyping Sample_ID')

# Filter metadata 
ind = as.character(METADATA[1,] %>% unlist) %in% metadataVarOfInterest
METADATA = METADATA[,ind]
colnames(METADATA) = gsub('[^[:alnum:]]','_',as.character(METADATA[1,] %>% unlist))

METADATA_ACC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, ACC_RNA_isolation__Exclude_,
                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_,
                ACC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = ACC_RNA_isolation__Exclude_, 
                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
                RIN = ACC_RNA_isolation__RIN, 
                ReportExclude = ACC_RNA_report__Exclude_, 
                LibraryBatch = ACC_RNA_report__Library_Batch,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'ACC')

METADATA_DLPFC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, DLPFC_RNA_isolation__Exclude_,
                DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_,
                DLPFC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_, 
                LibraryBatch = DLPFC_RNA_report__Library_Batch,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'DLPFC')

METADATA = rbindlist(list(METADATA_ACC, METADATA_DLPFC), use.names = T, fill = T) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
  left_join(ANCESTRY)

# Replace AFF and BP samples as other
levels(METADATA$Dx) = c("Other", "Other", "Control", "SCZ")
```

### Data preprocessing
```{r preprocess.data}
# Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
writeLines('Following counts are missing any metadata')
writeLines(paste(setdiff(colnames(COUNT), levels(METADATA$SampleID)), collapse = ', '))
METADATA <- METADATA %>% filter(SampleID %in% colnames(COUNT))

writeLines('Following samples are marked exclude')
writeLines(paste(METADATA$SampleID[METADATA$IsolationExclude == 1 | METADATA$ReportExclude == 1], collapse = ', '))
METADATA <- METADATA  %>% filter(IsolationExclude == 0, ReportExclude == 0)

writeLines('Following samples are missing PMI information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$PMI)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(PMI))

writeLines('Following samples are missing gender information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$Gender)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(Gender))

writeLines('Following samples are missing RIN information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$RIN)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(RIN))

writeLines('Following samples are missing EV information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$EV.1)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(METADATA$EV.1)) 
```

```{r preprocess.data1, results='asis'}
# Match covariates to expression data
indToRetain = intersect(METADATA$SampleID, colnames(COUNT))
removedIDs = setdiff(colnames(COUNT), METADATA$SampleID)

COUNT = COUNT[,indToRetain]

rownames(METADATA) = METADATA$SampleID
METADATA = METADATA[indToRetain,]

METADATA %>% 
  group_by(Dx, Tissue) %>% 
  summarise(count = n()) %>% 
  spread(Tissue, count) %>%
  kable()

METADATA = METADATA %>%
  dplyr::mutate(Dx.Tissue = paste(Dx, Tissue, sep = '.'))
```
Following sample are removed `r paste(removedIDs, collapse = ',')`

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Individual_ID", "Institution", "Gender", "LibraryBatch", "Dx.Tissue")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
rownames(COVARIATES) <- METADATA$SampleID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```

Correlation/association between covariates at an FDR <= 0.1

```{r covariates.correlation, fig.width=8, fig.height=8}
ind.cov = setdiff(colnames(COVARIATES), 'Individual_ID')
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES[,ind.cov], PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.5, row.width=0.1)
```

Explore metatdata

```{r dataExplore, fig.width = 6, fig.height = 15}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = Age_of_Death)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme

multiplot(plotlist = p, cols = 1)
```
### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue
```{r cpmnormalisation}
genesToAnalyze = dlply(METADATA, .(Dx.Tissue), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID %>% droplevels()],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)
genesToAnalyze = unlist(genesToAnalyze) %>% unique()

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# TMM normalisation
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to NA
log.mat = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix[log.mat] = NA

# TMM normalisation
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(VOOM.GENE_EXPRESSION$E))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Dx.Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="top")
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Institution', 'Dx.Tissue')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Institution', 'Dx.Tissue')]))
```

```{r temp, include = F}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
                                                 COVARIATES[,ind.cov],
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "Dx.Tissue" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue)

Adding Instituion and Gender a priori to variable selection as they are correlated with Dx.Tissue
```{r iterativeNorm, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender');

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
  correlation <- parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block = COVARIATES$Individual_ID)

  # Re-calculate voom weights with correlation of random effects
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F,
                                       block = COVARIATES$Individual_ID, 
                                       correlation = correlation$cor)

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, 'Individual_ID'))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
                                                                          c(postAdjustCovars))]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2)) 
p <- p + geom_point(aes(color=Institution, shape = Dx.Tissue, size=RIN)) 
p <- p + theme_bw() + theme(legend.position="top")
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)
p
```

```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue','Institution')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue','Institution')]))
```

```{r temp1, include=F}
dev.off()
gc()
```

```{r get.gene.symbols, cache=TRUE, eval=T}
backgroundGenes = data.frame(ensembl_gene_id = rownames(COUNT))

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "mar2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

# Add gene names
RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=16, fig.width=16}
# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueControl_ACC-Dx.TissueOther_ACC",
                                     "Dx.TissueControl_ACC-Dx.TissueSCZ_ACC",
                                     "Dx.TissueOther_ACC-Dx.TissueSCZ_ACC",
                                     "Dx.TissueControl_DLPFC-Dx.TissueOther_DLPFC",
                                     "Dx.TissueControl_DLPFC-Dx.TissueSCZ_DLPFC",
                                     "Dx.TissueOther_DLPFC-Dx.TissueSCZ_DLPFC"),
                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:6, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
    rownameToFirstColumn('ensembl_gene_id') %>%
    left_join(Ensemble2HGNC) %>%
    dplyr::select(ensembl_gene_id, hgnc_symbol, logFC, AveExpr, t, P.Value, adj.P.Val, B)
}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC) 
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison)) %>%
  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
  tidyr::separate(ref.state, into = c('Dx1','Tissue'), sep = '_') %>%
  tidyr::separate(to.state, into = c('Dx2','Tissue'), sep = '_') %>%
  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_')

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
  group_by(Comparison, Tissue) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
  group_by(Comparison, Tisue) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))

kable(full_join(tmp,tmp1))

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
p = p + facet_grid(Tissue+.~Comparison, scales = 'fixed')
p
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "ACC and DLPFC - STAR Reprocessed - with TMM", parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'CMC_ACC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('CMC_ACC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store logCPM
write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'CMC_ACC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_ACC_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                     executed = thisFile, activityDescription = activityDescription)

# Store design matrix
write.table(DM1$design, file = 'CMC_ACC_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('CMC_ACC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)

# Store residual gene expression
write.table(RESIDUAL.GENE_EXPRESSION, file = 'CMC_ACC_Expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('CMC_ACC_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store differential expression results
write.table(DE, file = 'CMC_ACC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_ACC_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
stopCluster(cl)
```
|  *Results*                |  *SynapseID*                     |
|  -------                  |   ---------                      |
|  Covariates               |  `r COV_OBJ$properties$id`       |
|  Raw counts (log CPM)     |  `r COUNT_OBJ$properties$id`     |
|  Design Matrix            |  `r DM_OBJ$properties$id`        |
|  Residual Expression      |  `r EXP_OBJ$properties$id`       |
  |  Differential Expression  |  `r DEXP_OBJ$properties$id`      |

### 2016-11-05 code/Rmd/CMC_ACC_StarAligned_TMM.Rmd
Added markdown to process DLPFC and ACC together with a random effects model

@@
 knit2synapse::knitToFolderEntity(file = "./CMC_ACC_StarAligned_TMM.Rmd",
-                                 parentId ="syn5569976",
+                                 parentId ="syn7501880",
                                  entityName = 'ACC - STAR Reprocessed - with TMM')
@@
 ```
+
 Explore metatdata
-```{r dataExplore, fig.width = 10, fig.height = 10}
+
+```{r dataExplore, fig.width = 14, fig.height = 6}
 my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
@@
 ```
+
 Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
-```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
+```{r preAdjustedSigCovars.NULL, fig.width=10, fig.height=6}
 preAdjustedSigCovars[["PC_res"]][[2]]$plotData
@@
 ### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "Dx" is chosen as the primary variable of interest
-```{r iterativeNorm}
+Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix.
+Here "Dx" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis)
+
+Adding Instituion and Gender a priori to variable selection as they are correlated with Dx
+```{r iterativeNorm, results='asis'}
 # Primary variable of interest
-postAdjustCovars = 'RIN';
+postAdjustCovars = c('Dx', 'Institution', 'Gender');
 
@@
 }
-```
-
-### Normlaisation (Custom design)
-Adding Dx, Institution, Gender, EV1, EV3, EV4 and EV5 to the list of postadjust covariates RIN, RIN2, Age of Death, PMI and EV2
-```{r custom.design}
-# Add primary variable to the list (in case not chosen by iterative method)
-postAdjustCovars = c('Dx', c(FactorCovariates, ContCovariates)) %>% unique()
-
-writeLines(paste('Using following covariates in the model:',
-                 paste(postAdjustCovars, collapse=', '),
-                 'as fixed effects'))
-
-# Post adjusted design matrix
-DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
-
-# Estimate voom weights
-VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
-
-# Fit linear model using new weights and new design
-VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
-
-# Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
 
@@
 `r tmp`
+
 ### Sanity check
-```{r residualSigCovars.manual, fig.width=20, fig.height=12}
+```{r residualSigCovars.manual, fig.width=10, fig.height=6}
 residualSigCovars[["PC_res"]][[2]]$plotData
@@
 Coexpression of genes 
-```{r coexp2, cache=FALSE}
+```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
 cr = cor(t(RESIDUAL.GENE_EXPRESSION))
@@
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
-p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)
+# p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)
 p
+```
 
+```{r decompse.normalise.data2.1, fig.height=6, fig.width=10, results='asis'}
 # Eucledian tree based analysis
@@
 ```
+
 ### Differential expression analysis

### 2016-11-06 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM_CD1.Rmd
Added markdown to process DLPFC and ACC together with a random effects model with a custom desing with all important variables

--- 2016-11-05 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM.Rmd
@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC and DLPFC with TMM normlaisation)"
+title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC and DLPFC with TMM normlaisation and Custom Design 1)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_ACC_And_DLPFC_StarAligned_TMM.Rmd",
+knit2synapse::knitToFolderEntity(file = "./CMC_ACC_And_DLPFC_StarAligned_TMM_CD1.Rmd",
                                  parentId ="syn7501880",
-                                 entityName = 'ACC and DLPFC - STAR Reprocessed - with TMM')
+                                 entityName = 'ACC and DLPFC - STAR Reprocessed - with TMM - Custom Design 1')
 ```
@@
 
-thisFileName <- 'CMC_ACC_And_DLPFC_StarAligned_TMM.Rmd'
+thisFileName <- 'CMC_ACC_And_DLPFC_StarAligned_TMM_CD1.Rmd'
 
@@
 
-### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix.
-Here "Dx.Tissue" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue)
-
-Adding Instituion and Gender a priori to variable selection as they are correlated with Dx.Tissue
+### Normalisation (custom design)
+Normalising COUNTS with a custom design matrix.
+Here Instituiton, Gender, LibraryBatch, Age\\_of\\_Death, PMI, RIN, RIN2, EV.1, EV.2, EV.3, EV.4, EV.5 are adjusted conditioned on Dx.Tissue with Individual\\_ID as random effects.
 ```{r iterativeNorm, results='asis'}
 # Primary variable of interest
-postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender');
+postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'LibraryBatch', 'Age\\_of\\_Death',
+                     'PMI', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5');
 
-# Assign residual covariates
-residualSigCovars = preAdjustedSigCovars
-covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
-
-loopCount = 0 
-while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
 writeLines(paste('Using following covariates in the model:',
                  paste(postAdjustCovars, collapse=', '),
-                   'as fixed effects'))
+                 'as fixed effects and Individual_ID as random effect'))
 
@@
                                               isKeyPlot=TRUE)
-
-  # Add postadjusted covariates (if any)
-  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
-                                                                          c(postAdjustCovars))]
-
-  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
-  loopCount = loopCount + 1
-}
-
-modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
-                  'as fixed effects')
-
-tmp <- paste('Using following covariates in the final model:', modelStr)
 ```
-`r tmp`
 
@@
   dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
-  group_by(Comparison, Tisue) %>%
+  group_by(Comparison, Tissue) %>%
   dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
@@
 # Code
-CODE <- Folder(name = "ACC and DLPFC - STAR Reprocessed - with TMM", parentId = parentId)
+CODE <- Folder(name = "ACC and DLPFC - STAR Reprocessed - with TMM - Custom Design 1", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-11-07 code/Rmd/CMC_DLPFC_StarAligned_TMM_CD1.Rmd
Added markdown to adjust DLPFC data with TMM normalisation and custom design matrix

--- 2016-11-04 code/Rmd/CMC_DLPFC_StarAligned_TMM.Rmd
@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (DLPFC with TMM normlaisation)"
+title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (DLPFC with TMM normlaisation and Custom Design 1)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_DLPFC_StarAligned_TMM.Rmd",
+knit2synapse::knitToFolderEntity(file = "./CMC_DLPFC_StarAligned_TMM_CD1.Rmd",
                                  parentId ="syn7501880",
-                                 entityName = 'DLPFC - STAR Reprocessed - with TMM')
+                                 entityName = 'DLPFC - STAR Reprocessed - with TMM - Custom Design 1')
 ```
@@
 
-thisFileName <- 'CMC_DLPFC_StarAligned_TMM.Rmd'
+thisFileName <- 'CMC_DLPFC_StarAligned_TMM_CD1.Rmd'
 
@@
 ### Covariate clustering
-Determine relationship between covariates. 
+Determine relationship between covariates
 ```{r covariates.clustering}
@@
 
-```{r dataExplore, fig.width = 6, fig.height = 14}
+```{r dataExplore, fig.width = 14, fig.height = 6}
 my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
@@
 Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
-```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
+```{r preAdjustedSigCovars.NULL, fig.width=12, fig.height=8}
 preAdjustedSigCovars[["PC_res"]][[2]]$plotData
@@
 
-### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix.
-Here "Dx" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis)
-
-Adding Instituion and Gender apriori to variable selection as they are correlated with Dx
-Also adding first 5 eigen vectors computed from ancestry
-```{r iterativeNorm, results='asis'}
+### Normalisation (custom design)
+Normalising COUNTS with a custom design matrix. Here Instituiton, Gender, LibraryBatch, Age\\_of\\_Death, PMI, RIN, RIN2, EV.1, EV.2, EV.3, EV.4, EV.5 are adjusted conditioned on Dx.Tissue
+```{r customNorm, results='asis'}
 # Primary variable of interest
-postAdjustCovars = c('Dx', 'Institution', 'Gender');
+postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'LibraryBatch', 'Age_of_Death',
+                     'PMI', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5');
 
@@
 
-loopCount = 0 
-while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
 writeLines(paste('Using following covariates in the model:',
@@
 
-  # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars))
-
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
 residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                                COVARIATES[, residCovars, drop=F], 
+                                              COVARIATES[, ind.cov, drop=F], 
                                               'adjusted design(voom-normalized)',
                                               isKeyPlot=TRUE)
+```
 
-  # Add postadjusted covariates (if any)
-  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
-                                                                          c(postAdjustCovars))]
-
-  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
-  loopCount = loopCount + 1
-}
-
-modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
-                  'as fixed effects')
-
-tmp <- paste('Using following covariates in the final model:', modelStr)
-```
-`r tmp`
 ### Sanity check
-```{r residualSigCovars.manual, fig.width=20, fig.height=12}
+```{r residualSigCovars.manual, fig.width=12, fig.height=8}
 residualSigCovars[["PC_res"]][[2]]$plotData
 ```
+
 ### Residual calculation
@@
 # Code
-CODE <- Folder(name = "DLPFC - STAR Reprocessed - with TMM", parentId = parentId)
+CODE <- Folder(name = "DLPFC - STAR Reprocessed - with TMM - Custom Design 1", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-11-07 code/Rmd/CMC_DLPFC_StarAligned_TMM_CD1.Rmd
Minor bug fix

@@
 ### Normalisation (custom design)
-Normalising COUNTS with a custom design matrix. Here Instituiton, Gender, LibraryBatch, Age\\_of\\_Death, PMI, RIN, RIN2, EV.1, EV.2, EV.3, EV.4, EV.5 are adjusted conditioned on Dx.Tissue
+Normalising COUNTS with a custom design matrix. Here Instituiton, Gender, LibraryBatch, Age\\_of\\_Death, PMI, RIN, RIN2, EV.1, EV.2, EV.3, EV.4, EV.5 are adjusted conditioned on Dx
 ```{r customNorm, results='asis'}
 # Primary variable of interest
-postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'LibraryBatch', 'Age_of_Death',
+postAdjustCovars = c('Dx', 'Institution', 'Gender', 'LibraryBatch', 'Age_of_Death',
                      'PMI', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5');
-
-# Assign residual covariates
-residualSigCovars = preAdjustedSigCovars
-covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
 

### 2016-11-07 code/Rmd/CMC_DLPFC_StarAligned_TMM_CD1.Rmd
Minor bug fix

@@
 residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                              COVARIATES[, ind.cov, drop=F], 
+                                              COVARIATES, 
                                               'adjusted design(voom-normalized)',

### 2016-11-07 code/Rmd/CMC_ACC_StarAligned_TMM_CD1.Rmd
Added markdown to process ACC data with custom design

--- 2016-11-05 code/Rmd/CMC_ACC_StarAligned_TMM.Rmd
@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC with TMM normlaisation)"
+title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC with TMM normlaisation and Custom Design 1)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_ACC_StarAligned_TMM.Rmd",
+knit2synapse::knitToFolderEntity(file = "./CMC_ACC_StarAligned_TMM_CD1.Rmd",
                                  parentId ="syn7501880",
-                                 entityName = 'ACC - STAR Reprocessed - with TMM')
+                                 entityName = 'ACC - STAR Reprocessed - with TMM - Custom Design 1')
 ```
@@
 
-thisFileName <- 'CMC_ACC_StarAligned_TMM.Rmd'
+thisFileName <- 'CMC_ACC_StarAligned_TMM_CD1.Rmd'
 
@@
 
-### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix.
-Here "Dx" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis)
+### Normalisation (custom design)
+Normalising COUNTS with a custom design matrix. Here Instituiton, Gender, LibraryBatch, Age\\_of\\_Death, PMI, RIN, RIN2, EV.1, EV.2, EV.3, EV.4, EV.5 are adjusted conditioned on Dx
+```{r customNorm, results='asis'}
+# Primary variable of interest
+postAdjustCovars = c('Dx', 'Institution', 'Gender', 'LibraryBatch', 'Age_of_Death',
+                     'PMI', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5');
 
-Adding Instituion and Gender a priori to variable selection as they are correlated with Dx
-```{r iterativeNorm, results='asis'}
-# Primary variable of interest
-postAdjustCovars = c('Dx', 'Institution', 'Gender');
-
-# Assign residual covariates
-residualSigCovars = preAdjustedSigCovars
-covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
-
-loopCount = 0 
-while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
 writeLines(paste('Using following covariates in the model:',
@@
 
-  # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars))
-
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
 residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                                COVARIATES[, residCovars, drop=F], 
+                                              COVARIATES, 
                                               'adjusted design(voom-normalized)',
                                               isKeyPlot=TRUE)
-
-  # Add postadjusted covariates (if any)
-  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
-                                                                          c(postAdjustCovars))]
-
-  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
-  loopCount = loopCount + 1
-}
-
-modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
-                  'as fixed effects')
-
-tmp <- paste('Using following covariates in the final model:', modelStr)
 ```
-`r tmp`
 
@@
 ```
+
 ### Residual calculation
@@
 ```
+
 Coexpression of genes 
@@
 # Code
-CODE <- Folder(name = "ACC - STAR Reprocessed - with TMM", parentId = parentId)
+CODE <- Folder(name = "ACC - STAR Reprocessed - with TMM - Custom Design 1", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-11-09 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM.Rmd
Added enrichment analysis

@@
 p
+```
+
+### Gene set enrichment analysis
+Using Fisher's exact test for gene set enrichment analysis
+```{r enrich.de, fig.height=16, fig.width=16}
+diffexp.genes = DE %>%
+  dlply(.(Comparison, Tissue), .fun = function(x){
+    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
+    tmp = tmp[!is.na(tmp)]
+  })
+diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
+backgroundGenes = unique(DE$hgnc_symbol)
+
+# Download all related genesets from synapse
+geneset.files = synQuery('select * from file where parentId == "syn3240583"')
+ALL_USED_IDs = c(ALL_USED_IDs, geneset.files$file.id)
+genesets.to.test = sapply(geneset.files$file.id, function(id){
+  tmp = downloadFile(id) %>%
+    dlply(.(Name), .fun = function(x){
+      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
+    })
+  }) %>%
+  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
+  do.call(c,.)
+
+# Perform enrichment analysis
+enrich.Results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
+   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
+     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
+    }, x, backgroundGenes, .parallel = TRUE) %>%
+    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
+}, genesets.to.test, backgroundGenes) %>%
+  rbindlist(idcol = 'Comparison', use.names = T, fill = T)
 ```
@@
                     executed = thisFile, activityDescription = activityDescription)
+
+# Store results of enrichment analysis
+write.table(enrich.Results, file = 'CMC_DLPFC_ACC_Enrichment.tsv', sep = '\t', row.names=F, quote=F)
+ENR_OBJ = File('CMC_DLPFC_ACC_Enrichment.tsv', name = 'Enrichment Analysis', parentId = CODE$properties$id)
+ENR_OBJ = synStore(ENR_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                    executed = thisFile, activityDescription = activityDescription)
 stopCluster(cl)

### 2016-11-10 code/Rmd/Mayo_CBE_RNASeq_Realigned_SynBAM_StarHTSeq.Rmd
Added markdown to process reprocessed mayo cerebellum data

---
title: "Covariates analysis of UFL-Mayo-ISB RNASeq data of cerebellum tissue (realigned from synapse BAM with Picard-STAR-HTSeq)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)
library(knitr)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./Mayo_CBE_RNASeq_Realigned_SynBAM_StarHTSeq.Rmd",
                                 parentId = "syn5570325",
                                 entityName = 'UFL_Mayo_ISB Cerebellum RNASeq RealignedFromSynBAM PicardStarHTSeq')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
## It is assumed your working directory is where this file

### Clear R console screen output
cat("\014")

## Load required libraries
library(CovariateAnalysis)
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(githubr)

library(ggplot2)
library(reshape2)
library(limma)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(xlsx)

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

synapseLogin()

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn5570325';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of UFL_Mayo_ISB RNASeq data';

thisFileName <- 'Mayo_CBE_RNASeq_Realigned_SynBAM_StarHTSeq.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Get reprocessed paired end sample counts for cerebellum 
COUNT_ID <- 'syn7513807'
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(synGet(COUNT_ID)@filePath, header = T, row.names = 1, quote = '', check.names = F)

# Get clinical metadata
METADATA_ID <- 'syn5223705'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA <- read.table(synGet(METADATA_ID)@filePath,sep=',',header=T, stringsAsFactors=F) %>%
  tidyr::separate(SampleID, c('Donor_ID', 'BrainRegion')) %>%
  dplyr::mutate(ID = paste(Donor_ID, BrainRegion, sep = '_')) %>%
  dplyr::rename(Gender = Sex, FLOWCELL = Flowcell)
```

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
METADATA$BrainRegion.Diagnosis = paste(METADATA$BrainRegion, METADATA$Diagnosis, sep = '.')
METADATA$RIN2 = METADATA$RIN^2
METADATA <- METADATA %>%
  filter(!is.na(Source)) %>%
  filter(!(ID %in% c('1923_CER', '11311_CER', '742_CER')))
```

```{r preprocess.data1, results='asis'}
# Match covariates to expression data
indToRetain = intersect(METADATA$ID, colnames(COUNT))
removedIDs = setdiff(colnames(COUNT), METADATA$ID)

COUNT = COUNT[,indToRetain]

rownames(METADATA) = METADATA$ID
METADATA = METADATA[indToRetain,]

METADATA %>% 
  group_by(BrainRegion, Diagnosis) %>% 
  summarise(count = n()) %>% 
  spread(BrainRegion, count) %>%
  kable()
```
Following sample are removed `r paste(removedIDs, collapse = ',')`

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "AgeAtDeath")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES[,FactorCovariates] <- lapply(COVARIATES[,FactorCovariates],function(x){
  x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})
}) %>% data.frame()
rownames(COVARIATES) <- METADATA$ID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```

Correlation/association between covariates at an FDR <= 0.1

```{r covariates.correlation, fig.width=8, fig.height=8}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.5, row.width=0.1)
```

Explore metatdata

```{r dataExplore, fig.width = 6, fig.height = 15}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = AgeAtDeath)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('Age At Death') + my.theme

multiplot(plotlist = p, cols = 2)
```
### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion.Diagnosis
```{r cpmnormalisation}
genesToAnalyze = dlply(METADATA, .(BrainRegion.Diagnosis), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$ID %>% unique()],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)
genesToAnalyze = unlist(genesToAnalyze) %>% unique()

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to NA
log.mat = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix[log.mat] = NA

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(VOOM.GENE_EXPRESSION$E))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(ID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'ID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Source, shape=BrainRegion.Diagnosis, size=AgeAtDeath))
p <- p + theme_bw() + theme(legend.position="top")
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Source', 'Gender', 'FLOWCELL', 'BrainRegion.Diagnosis')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Source', 'Gender', 'FLOWCELL', 'BrainRegion.Diagnosis')]))
```

```{r temp, include = F}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix
```{r iterative.norm}
# Primary variable of interest
postAdjustCovars = 'BrainRegion.Diagnosis'

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
significantCovars = residualSigCovars$significantCovars
covariatesEffects = residualSigCovars$Effects.significantCovars
covariatesEffects = covariatesEffects[setdiff(significantCovars, postAdjustCovars)]

postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate correlation of random effects
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)

  # Fit a linear model
  FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), postAdjustCovars)

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES, 
                                                'all adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  significantCovars = residualSigCovars$significantCovars
  covariatesEffects = residualSigCovars$Effects.significantCovars
  covariatesEffects = covariatesEffects[setdiff(significantCovars, postAdjustCovars)]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```

### Residual calculation
Calculate weighted residuals and add back "Dx.Tissue" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("BrainRegion.Diagnosis", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

# Plot first 4 PCs
plotdata <- data.frame(ID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'ID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2)) 
p <- p + geom_point(aes(color=Source, shape = BrainRegion.Diagnosis, size=RIN)) 
p <- p + theme_bw() + theme(legend.position="top")
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)
p
```

```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Source', 'Gender', 'FLOWCELL', 'BrainRegion.Diagnosis')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Source', 'Gender', 'FLOWCELL', 'BrainRegion.Diagnosis')]))
```

```{r temp1, include=F}
dev.off()
gc()
```

```{r get.gene.symbols, cache=TRUE}
# Define biomart object
mart <- useMart(host = "jul2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl", biomart = "ENSEMBL_MART_ENSEMBL")

RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'GeneID') %>%
  dplyr::mutate(gene.id = GeneID) %>%
  tidyr::separate(GeneID, c('ensembl_gene_id','position'),sep = '\\.')

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = RESIDUAL.GENE_EXPRESSION$ensembl_gene_id,
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(RESIDUAL.GENE_EXPRESSION, Ensemble2HGNC) %>%
  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=16, fig.width=16}
# Fit contrast
contrast = makeContrasts(contrasts=c("BrainRegion.DiagnosisCER_Control-BrainRegion.DiagnosisCER_AD",
                                     "BrainRegion.DiagnosisCER_Control-BrainRegion.DiagnosisCER_Pathologic_Aging",
                                     "BrainRegion.DiagnosisCER_Control-BrainRegion.DiagnosisCER_PSP",
                                     "BrainRegion.DiagnosisCER_Pathologic_Aging-BrainRegion.DiagnosisCER_AD",
                                     "BrainRegion.DiagnosisCER_PSP-BrainRegion.DiagnosisCER_AD",
                                     "BrainRegion.DiagnosisCER_PSP-BrainRegion.DiagnosisCER_Pathologic_Aging"),
                         levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:6, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
    rownameToFirstColumn('GeneID') %>%
    dplyr::mutate(gene.id = GeneID) %>%
    tidyr::separate(GeneID, c('ensembl_gene_id','position'),sep = '\\.') %>%
    left_join(Ensemble2HGNC) %>%
    dplyr::select(ensembl_gene_id, hgnc_symbol, logFC, AveExpr, t, P.Value, adj.P.Val, B)
}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC) 
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis', '', Comparison)) %>%
  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
  tidyr::separate(ref.state, into = c('Tissue','Dx1'), sep = '_') %>%
  tidyr::separate(to.state, into = c('Tissue','Dx2'), sep = '_') %>%
  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_') %>%
  dplyr::mutate(Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')))

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue, Direction) %>%
  group_by(Comparison, Tissue, Direction) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05)

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue, Direction) %>%
  group_by(Comparison, Tissue, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2)

kable(full_join(tmp %>%
                  dplyr::rename(DOWN_FDR_0_05 = DOWN, UP_FDR_0_05 = UP),
                tmp1 %>%
                  dplyr::rename(DOWN_FDR_0_05_FC_1.2 = DOWN, UP_FDR_0_05_FC_1.2 = UP)))

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
p = p + facet_grid(Tissue+.~Comparison, scales = 'fixed')
p
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB Cerebellum RNASeq RealignedFromSynBAM PicardStarHTSeq', parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'ID')
write.table(COVARIATES, file = 'MAYOcovariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('MAYOcovariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Raw Expression
write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file='MAYOlogCPM.tsv', sep = '\t', row.names=F, quote=F)
RAWCPM_OBJ = File('MAYOlogCPM.tsv', 
                  name = 'Raw logCPM',
                  parentId = CODE$properties$id)
RAWCPM_OBJ = synStore(RAWCPM_OBJ,
                      used = ALL_USED_IDs, 
                      activityName = activityName, 
                      executed = thisFile, 
                      activityDescription = activityDescription)

# Design Matrix
DM1$design = DM1$design[,!(colnames(DM1$design) %in% c('PMI'))]
write.table(DM1$design, file='MAYOdesign.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('MAYOdesign.tsv', 
                  name = 'Design Matrix',
                  parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ,
                      used = ALL_USED_IDs, 
                      activityName = activityName, 
                      executed = thisFile, 
                      activityDescription = activityDescription)

# Store residual gene expression
write.table(RESIDUAL.GENE_EXPRESSION, file = 'MAYOexpression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('MAYOexpression.tsv', name = 'Voom Adjusted Residual Expression (BrainRegion.Diagnosis added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store differential expression results
write.table(DE, file = 'MAYOdiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('MAYOdiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
```
|  *Results*                |  *SynapseID*                    |
|  -------                  |   ---------                     |
|  Covariates               |  `r COV_OBJ$properties$id`      |
|  Raw counts (logCPM)      |  `r RAWCPM_OBJ$properties$id`   |
|  Design Matrix            |  `r DM_OBJ$properties$id`       |
|  Residual Expression      |  `r EXP_OBJ$properties$id`      |
|  Differential Expression  |  `r DEXP_OBJ$properties$id`     |

### Source R Code
[This File](`r thisFile`)

### 2016-11-10 code/Rmd/Mayo_CBE_RNASeq_Realigned_SynBAM_StarHTSeq.Rmd
Minor change

@@
 
-```{r dataExplore, fig.width = 6, fig.height = 15}
+```{r dataExplore, fig.width = 10, fig.height = 6}
 my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=16, fig.width=16}
+```{r diffExp, fig.height=8, fig.width=20}
 # Fit contrast

### 2016-11-11 code/Rmd/Mayo_CBE_RNASeq_Realigned_SynBAM_StarHTSeq.Rmd
Minor changes

@@
 
+### Comparison of reprocessed counts with source counts
+```{r comp, fig.height=5, fig.width=5}
+COUNT_OR = downloadFile('syn5201012')
+rownames(COUNT_OR) = COUNT_OR$ensembl_id
+COUNT_OR$ensembl_id = NULL
+
+COUNT = COUNT %>%
+  rownameToFirstColumn('GeneID') %>%
+  dplyr::mutate(gene.id = GeneID) %>%
+  tidyr::separate(GeneID, c('ensembl_gene_id','position'),sep = '\\.') %>%
+  filter(ensembl_gene_id %in% rownames(COUNT_OR))
+rownames(COUNT) = COUNT$ensembl_gene_id
+COUNT = COUNT[intersect(rownames(COUNT), rownames(COUNT_OR)),intersect(colnames(COUNT), colnames(COUNT_OR))]
+COUNT_OR = COUNT_OR[intersect(rownames(COUNT), rownames(COUNT_OR)),intersect(colnames(COUNT), colnames(COUNT_OR))]
+
+cr = cor(COUNT, COUNT_OR, use = 'pairwise.complete.obs')
+boxplot(diag(cr), main = 'Individual sample level correlation', ylab = 'Correlation', xlab = 'n=271')
+```
+
 ### Store files in synapse

### 2016-11-15 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM_SVA.Rmd
Added markdown to process ACC and DLPFC data with SVA

--- 2016-11-06 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM_CD1.Rmd
@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC and DLPFC with TMM normlaisation and Custom Design 1)"
+title: "Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC and DLPFC with TMM normlaisation and SVA)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_ACC_And_DLPFC_StarAligned_TMM_CD1.Rmd",
+knit2synapse::knitToFolderEntity(file = "./CMC_ACC_And_DLPFC_StarAligned_TMM_SVA.Rmd",
                                  parentId ="syn7501880",
-                                 entityName = 'ACC and DLPFC - STAR Reprocessed - with TMM - Custom Design 1')
+                                 entityName = 'ACC and DLPFC - STAR Reprocessed - with TMM - SVA')
 ```
@@
 library(RColorBrewer)
+library(sva)
 
@@
 activityName = 'Covariate adjustments';
-activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM normalisation (ACC and DLPFC)';
+activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM normalisation and SVA (ACC and DLPFC)';
 
-thisFileName <- 'CMC_ACC_And_DLPFC_StarAligned_TMM_CD1.Rmd'
+thisFileName <- 'CMC_ACC_And_DLPFC_StarAligned_TMM_SVA.Rmd'
 
@@
 ### Normalisation (custom design)
-Normalising COUNTS with a custom design matrix.
-Here Instituiton, Gender, LibraryBatch, Age\\_of\\_Death, PMI, RIN, RIN2, EV.1, EV.2, EV.3, EV.4, EV.5 are adjusted conditioned on Dx.Tissue with Individual\\_ID as random effects.
-```{r iterativeNorm, results='asis'}
+Normalising COUNTS with a design matrix obtained from an iterative normalisation analysis (see syn7503251 for more details).
+
+Here Institution, Gender, RIN, Age\\_of\\_Death, PMI, EV.4 were used as fixed effects conditioned on Dx.Tissue with Individual\\_ID as random effects.
+```{r customNorm, results='asis'}
 # Primary variable of interest
-postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'LibraryBatch', 'Age\\_of\\_Death',
-                     'PMI', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5');
+postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'RIN', 'Age_of_Death', 'PMI', 'EV.4');
 
@@
 
-# Residual covariates to choose from
-residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, 'Individual_ID'))
-
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
 residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                              COVARIATES[, residCovars, drop=F], 
+                                              COVARIATES[, ind.cov, drop=F], 
                                               'adjusted design(voom-normalized)',
@@
 ```
+
+### Estimate surrogate variables using SVA package
+```{r svaAdjust}
+MODEL0 = DM1$design[,-grep('Dx.Tissue',colnames(DM1$design))] # Get null model by removing variable of interest
+MODEL1 = DM1$design
+
+## Calculate the number of surrogate variables
+NUM.SV = sva::num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1,
+                     method = 'be', seed = 123456, B = 20)
+
+SURR.VAR = sva::sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0,
+                    block = COVARIATES$Individual_ID, 
+                    n.sv = NUM.SV, B = 30)$sv
+colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
+
+DM1$design = cbind(MODEL1, SURR.VAR)
+
+# Estimate voom weights
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
+correlation <- parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
+
+# Re-calculate voom weights with correlation of random effects
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F,
+                                     block = COVARIATES$Individual_ID, 
+                                     correlation = correlation$cor)
+
+# Fit linear model using new weights and new design
+VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
+
+# Residuals after normalisation
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
+
+# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
+residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+                                              COVARIATES[, ind.cov, drop=F], 
+                                              'adjusted design(voom-normalized)',
+                                              isKeyPlot=TRUE)
+```
 ### Residual calculation
-Calculate weighted residuals and add back "Dx" to the residuals
+Calculate weighted residuals and add back "Dx.Tissue" to the residuals
 ```{r varsToAddBack}
@@
   tidyr::separate(to.state, into = c('Dx2','Tissue'), sep = '_') %>%
-  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_')
+  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_') %>%
+  dplyr::mutate(Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')))
 
@@
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
-  group_by(Comparison, Tissue) %>%
-  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
+  dplyr::select(ensembl_gene_id, Comparison, Tissue, Direction) %>%
+  group_by(Comparison, Tissue, Direction) %>%
+  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05)
 
@@
   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
-  group_by(Comparison, Tissue) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
+  dplyr::select(ensembl_gene_id, Comparison, Tissue, Direction) %>%
+  group_by(Comparison, Tissue, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05_FC_1.2)
 
-kable(full_join(tmp,tmp1))
+kable(full_join(tmp %>%
+                  dplyr::rename(DOWN_FDR_0_05 = DOWN, UP_FDR_0_05 = UP),
+                tmp1 %>%
+                  dplyr::rename(DOWN_FDR_0_05_FC_1.2 = DOWN, UP_FDR_0_05_FC_1.2 = UP)))
 
@@
 # Code
-CODE <- Folder(name = "ACC and DLPFC - STAR Reprocessed - with TMM - Custom Design 1", parentId = parentId)
+CODE <- Folder(name = "ACC and DLPFC - STAR Reprocessed - with TMM - SVA", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-11-15 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM_SVA.Rmd
Minor bug fix

@@
 SURR.VAR = sva::sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0,
-                    block = COVARIATES$Individual_ID, 
                     n.sv = NUM.SV, B = 30)$sv

### 2016-11-22 code/Rmd/ROSMAP_RNASeq_Realigned_SynBAM_StarHTSeq.Rmd
Added markdown to process ROSMAP realigned from synBAM using STAR-HTSeq pipeline

--- 2016-06-01 code/Rmd/ROSMAP_RNASeq_Reprocessed.Rmd
@@
 ---
-title: "Covariate analysis of ROSMAP Reprocessed RNASeq"
+title: "Covariate analysis of ROSMAP RNASeq data processed using Picard-STAR-HTSeq pipeline from synapse BAM deposited by RUSH"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed.Rmd",
+knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Realigned_SynBAM_StarHTSeq.Rmd",
                                  parentId ="syn5570291",
-                                 entityName = 'ROSMAP RNASeq Reprocessed Counts (Based on cogdx)')
+                                 entityName = 'ROSMAP RNASeq RealignedFromSynBAM PicardStarHTSeq')
 ```
@@
 
-thisFileName <- 'ROSMAP_RNASeq_Reprocessed.Rmd'
+thisFileName <- 'ROSMAP_RNASeq_Realigned_SynBAM_StarHTSeq.Rmd'
 
@@
 # Download reprocessed counts (paired end)
-COUNT_PE_ID <- 'syn6112813';
-ALL_USED_IDs <- COUNT_PE_ID
-COUNT_PE <- read.table(synGet(COUNT_PE_ID)@filePath, header=T, sep='\t', check.names = F)
-
-# Download reprocessed counts (single end)
-COUNT_SE_ID <- 'syn6117775';
-ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_SE_ID
-COUNT_SE <- read.table(synGet(COUNT_SE_ID)@filePath, header=T, sep='\t', check.names = F)
-
-COUNT = COUNT_PE + COUNT_SE[rownames(COUNT_PE), colnames(COUNT_PE)]
+COUNT_ID <- 'syn7750270';
+ALL_USED_IDs <- COUNT_ID
+COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F)
 
@@
 
-Remove genes that have less than 1 cpm counts in at least 50% of samples.
-```{r cpmnormalisation, cache=TRUE}
-PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT,MIN_GENE_CPM=1, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
+Remove genes that have less than 1 cpm counts in at least 50% of samples per "cogdx"
+```{r cpmnormalisation}
+genesToAnalyze = dlply(METADATA, .(cogdx), .fun = function(mtd, count){
+  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$Sampleid %>% unique()],
+                                                   MIN_GENE_CPM=1, 
+                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
+  processed.counts$filteredExprMatrix$genes
+}, COUNT)
+genesToAnalyze = unlist(genesToAnalyze) %>% unique()
+
+PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
 ```
+Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples
+
 ### Normalisation (with NULL)
 Initial normalisation usign voom (with NULL design)
-```{r initial.voom.normalisation}
+```{r initial.voom.normalisation, fig.height=5, fig.width=5}
 # Initial normalisation of gene expression
 VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T)
+
+# Set gene counts in specific samples that are deviating 3 sd from other samples to NA
+log.mat = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
+  mn = mean(x, na.rm = T)
+  std.dev = sd(x, na.rm = T)
+  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
+}) %>% t
+PROCESSED_COUNTS$filteredExprMatrix[log.mat] = NA
+
+# Initial normalisation of gene expression
+VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T)
+```
+Coexpression of genes 
+```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
+cr = cor(t(VOOM.GENE_EXPRESSION$E))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
+
+boxplot(VOOM.GENE_EXPRESSION$E, main = 'Distribution of expression')
 ```
@@
 p
-
+```
+Tree based classification of samples
+```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
 # Eucledian tree based analysis
@@
 ### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "cogdx" is chosen as the primary variable of interest
-```{r iterativeNorm}
+Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix
+```{r iterative.norm}
 # Primary variable of interest
@@
 residualSigCovars = preAdjustedSigCovars
-covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
+significantCovars = residualSigCovars$significantCovars
+covariatesEffects = residualSigCovars$Effects.significantCovars
+covariatesEffects = covariatesEffects[setdiff(significantCovars, postAdjustCovars)]
+
 postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
@@
 
-  # Fit linear model using new weights and new design
-  VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
+  # Fit a linear model
+  FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
 
   # Residuals after normalisation
-  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
+  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
 
@@
   residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                                COVARIATES[, residCovars, drop=F], 
-                                                'adjusted design(voom-normalized)',
+                                                COVARIATES, 
+                                                'all adjusted design(voom-normalized)',
                                                 isKeyPlot=TRUE)
@@
   # Add postadjusted covariates (if any)
-  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
+  significantCovars = residualSigCovars$significantCovars
+  covariatesEffects = residualSigCovars$Effects.significantCovars
+  covariatesEffects = covariatesEffects[setdiff(significantCovars, postAdjustCovars)]
 
@@
 ### Sanity check
-```{r residualSigCovars.manual, fig.width=20, fig.height=12}
+```{r residualSigCovars.manual, fig.width=12, fig.height=8}
 residualSigCovars[["PC_res"]][[2]]$plotData
 ```
+
 ### Residual calculation
@@
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
-  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+  FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+```
+Coexpression of genes 
+```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
+cr = cor(t(RESIDUAL.GENE_EXPRESSION))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
 ```
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=8, fig.width=16}
+```{r diffExp, fig.height=6, fig.width=12}
 # Fit contrast
@@
 # Code
-CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts (Based on cogdx)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP RNASeq RealignedFromSynBAM PicardStarHTSeq", parentId = parentId)
 CODE <- synStore(CODE)

### 2016-11-22 code/Rmd/ROSMAP_RNASeq_Realigned_SynBAM_StarHTSeq.Rmd
Minor modifications

@@
 
-boxplot(VOOM.GENE_EXPRESSION$E, main = 'Distribution of expression')
+for (i in 0:8){
+  expr2 = VOOM.GENE_EXPRESSION$E[,rownames(COVARIATES)[(COVARIATES$Batch %in% c(i))]]
+  cr = cor(t(expr2))
+  hist(cr, main = paste('Batch',i))
+}
 ```
@@
 
-p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=cogdx, shape=Batch, size=age_death))
+colours <- RColorBrewer::brewer.pal(9, 'Set1')
+p <- ggplot(plotdata, aes(x=PC1, y=PC2, shape=cogdx, color = Batch))
+p <- p + geom_point(aes(size=RINcontinuous))
+p <- p + scale_color_manual(values=colours)
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
@@
 # Primary variable of interest
-postAdjustCovars = 'cogdx'
+postAdjustCovars = c('cogdx', 'Batch')
 
@@
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
+
+for (i in 0:8){
+  expr2 = RESIDUAL.GENE_EXPRESSION[,rownames(COVARIATES)[(COVARIATES$Batch %in% c(i))]]
+  cr = cor(t(expr2))
+  hist(cr, main = paste('Batch',i))
+}
 ```
@@
 
-p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=cogdx, shape = Batch, size=RINcontinuous))
+colours <- RColorBrewer::brewer.pal(9, 'Set1')
+p <- ggplot(plotdata, aes(x=PC1, y=PC2, shape=cogdx, color = Batch))
+p <- p + geom_point(aes(size=RINcontinuous))
+p <- p + scale_color_manual(values=colours)
 p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
@@
 # Define biomart object
-mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(host = "jul2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl", biomart = "ENSEMBL_MART_ENSEMBL")
 
@@
 ```
+
+```{r residual.analysis}
+expr = RESIDUAL.GENE_EXPRESSION[!duplicated(RESIDUAL.GENE_EXPRESSION$Gene.ID),-c(1:3)]
+rownames(expr) = RESIDUAL.GENE_EXPRESSION$Gene.ID[!(duplicated(RESIDUAL.GENE_EXPRESSION$Gene.ID))]
+
+PC = prcomp(expr, scale. = T, center = T)
+
+# Plot first 5 PCs
+plotdata <- rownameToFirstColumn(PC$rotation, 'ID') %>% 
+    left_join(rownameToFirstColumn(COVARIATES, 'ID'))
+
+barplot((PC$sdev^2/sum(PC$sdev^2))[1:10], main = 'Fraction variance explained (top 10 PC)')
+p <- GGally::ggpairs(plotdata, mapping = aes(shape=cogdx, color = Batch, size=RINcontinuous),
+                     columns = 2:5, diag = list('continuous' = 'densityDiag'),
+                     title = 'First 4 PCs of residuals')
+p
+
+lb = median(PC$rotation[,1]) - 2*sd(PC$rotation[,1])
+ub = median(PC$rotation[,1]) + 2*sd(PC$rotation[,1])
+
+ind = c(which(PC$rotation[,1] <= lb), which(PC$rotation[,1] >= ub))
+writeLines('Samples deviating 2 SD from median of PC1 are')
+writeLines(paste(rownames(PC$rotation)[ind], collapse = ', '))
+
+writeLines('Distribution of batches of deviating samples')
+kable(rownameToFirstColumn(summary(COVARIATES$Batch[ind]), 'Batch Number'))
+```
+
 ### Store files in synapse

### 2016-11-24 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM.Rmd
Added additional metadata (RNAseq technical) to the DLPFC ACC covariates pipeline

@@
 # Get merged metadata
-METADATA_ID = 'syn2299154'
+METADATA_ID = 'syn7574833';#'syn2299154'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
@@
 ## Metadata with specific information for RNASeq processing
-metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", 
-                          "Dx", "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
+metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", "Dx",
+
+                          "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                           "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?", "ACC_RNA_report: Library Batch",
+                          "ACC_RNA_report: Ribozero Batch", "ACC_RNA_report: Flowcell Batch", 
+                          "ACC_RNA_report: Mapped Reads", "ACC_RNA_report: Intragenic Rate",
+                          "ACC_RNA_report: Intronic Rate", "ACC_RNA_report: Intergenic Rate",
+                          "ACC_RNA_report: Genes Detected", "ACC_RNA_report: Transcripts Detected",
+                          "ACC_RNA_report: Expression Profiling Efficiency", "ACC_RNA_report: rRNA Rate",
+                          "ACC_RNA_report: Total Reads", "ACC_RNA_report: Percent Aligned",
+
                           "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                           "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
-                          'DNA_genotyping_report: Genotyping Sample_ID')
+                          "DLPFC_RNA_report: Ribozero Batch", "DLPFC_RNA_report: Flowcell Batch",
+                          "DLPFC_RNA_report: Mapped Reads", "DLPFC_RNA_report: Intragenic Rate",
+                          "DLPFC_RNA_report: Intronic Rate", "DLPFC_RNA_report: Intergenic Rate", 
+                          "DLPFC_RNA_report: Genes Detected", "DLPFC_RNA_report: Transcripts Detected", 
+                          "DLPFC_RNA_report: Expression Profiling Efficiency", "DLPFC_RNA_report: rRNA Rate",
+                          "DLPFC_RNA_report: Total Reads", "DLPFC_RNA_report: Percent Aligned",
+
+                          "DNA_genotyping_report: Genotyping Sample_ID")
 
@@
 METADATA_ACC = METADATA %>%
-  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, ACC_RNA_isolation__Exclude_,
-                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_,
-                ACC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
+  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
+                ACC_RNA_isolation__Exclude_, ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN,
+                ACC_RNA_report__Exclude_, ACC_RNA_report__Ribozero_Batch, ACC_RNA_report__Library_Batch, 
+                ACC_RNA_report__Flowcell_Batch, ACC_RNA_report__Mapped_Reads, ACC_RNA_report__Intragenic_Rate,
+                ACC_RNA_report__Intronic_Rate, ACC_RNA_report__Intergenic_Rate, ACC_RNA_report__Genes_Detected,
+                ACC_RNA_report__Transcripts_Detected, ACC_RNA_report__Expression_Profiling_Efficiency, 
+                ACC_RNA_report__rRNA_Rate, ACC_RNA_report__Total_Reads, ACC_RNA_report__Percent_Aligned,
+                DNA_genotyping_report__Genotyping_Sample_ID) %>%
   dplyr::rename(PMI = PMI__in_hours_, 
@@
                 ReportExclude = ACC_RNA_report__Exclude_, 
+                RibozeroBatch = ACC_RNA_report__Ribozero_Batch,
                 LibraryBatch = ACC_RNA_report__Library_Batch,
+                FlowcellBatch = ACC_RNA_report__Flowcell_Batch, 
+                MappedReads = ACC_RNA_report__Mapped_Reads, 
+                IntragenicRate = ACC_RNA_report__Intragenic_Rate,
+                IntronicRate = ACC_RNA_report__Intronic_Rate, 
+                IntergenicRate = ACC_RNA_report__Intergenic_Rate, 
+                GenesDetected = ACC_RNA_report__Genes_Detected,
+                TranscriptsDetected = ACC_RNA_report__Transcripts_Detected, 
+                ExpProfEfficiency = ACC_RNA_report__Expression_Profiling_Efficiency, 
+                rRNARate = ACC_RNA_report__rRNA_Rate, 
+                TotalReads = ACC_RNA_report__Total_Reads, 
+                PercentAligned = ACC_RNA_report__Percent_Aligned,
                 GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
@@
 METADATA_DLPFC = METADATA %>%
-  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, DLPFC_RNA_isolation__Exclude_,
-                DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_,
-                DLPFC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
+  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
+                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN,
+                DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Ribozero_Batch, DLPFC_RNA_report__Library_Batch, 
+                DLPFC_RNA_report__Flowcell_Batch, DLPFC_RNA_report__Mapped_Reads, DLPFC_RNA_report__Intragenic_Rate,
+                DLPFC_RNA_report__Intronic_Rate, DLPFC_RNA_report__Intergenic_Rate, DLPFC_RNA_report__Genes_Detected,
+                DLPFC_RNA_report__Transcripts_Detected, DLPFC_RNA_report__Expression_Profiling_Efficiency, 
+                DLPFC_RNA_report__rRNA_Rate, DLPFC_RNA_report__Total_Reads, DLPFC_RNA_report__Percent_Aligned,
+                DNA_genotyping_report__Genotyping_Sample_ID) %>%
   dplyr::rename(PMI = PMI__in_hours_, 
@@
                 ReportExclude = DLPFC_RNA_report__Exclude_, 
+                RibozeroBatch = DLPFC_RNA_report__Ribozero_Batch,
                 LibraryBatch = DLPFC_RNA_report__Library_Batch,
+                FlowcellBatch = DLPFC_RNA_report__Flowcell_Batch, 
+                MappedReads = DLPFC_RNA_report__Mapped_Reads, 
+                IntragenicRate = DLPFC_RNA_report__Intragenic_Rate,
+                IntronicRate = DLPFC_RNA_report__Intronic_Rate, 
+                IntergenicRate = DLPFC_RNA_report__Intergenic_Rate, 
+                GenesDetected = DLPFC_RNA_report__Genes_Detected,
+                TranscriptsDetected = DLPFC_RNA_report__Transcripts_Detected, 
+                ExpProfEfficiency = DLPFC_RNA_report__Expression_Profiling_Efficiency, 
+                rRNARate = DLPFC_RNA_report__rRNA_Rate, 
+                TotalReads = DLPFC_RNA_report__Total_Reads, 
+                PercentAligned = DLPFC_RNA_report__Percent_Aligned,
                 GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
@@
   dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
-  left_join(ANCESTRY)
+  left_join(ANCESTRY) %>%
+  unique
 
@@
 ```{r covariates.clustering}
-FactorCovariates <- c("Individual_ID", "Institution", "Gender", "LibraryBatch", "Dx.Tissue")
-ContCovariates <- c("Age_of_Death", "PMI", "RIN", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")
+FactorCovariates <- c('Individual_ID', "Institution", "Gender", "RibozeroBatch", "LibraryBatch", "FlowcellBatch", "Dx.Tissue")
+ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
+                    "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
+                    "PercentAligned", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")
 
@@
 COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
-COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
+COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){
+  x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
 rownames(COVARIATES) <- METADATA$SampleID
@@
 ContCovariates = c(ContCovariates, 'RIN2')
+
+## Get GC content from biomart
+backgroundGenes = data.frame(ensembl_gene_id = rownames(COUNT))
+
+# Define biomart object
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "mar2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
+
+# Query biomart
+Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content"),
+                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
+                       mart = mart)
+GC.CONT = Ensemble2HGNC %>%
+  dplyr::select(ensembl_gene_id, percentage_gc_content) %>%
+  unique
+rownames(GC.CONT) = GC.CONT$ensembl_gene_id
 ```
@@
 
-```{r dataExplore, fig.width = 6, fig.height = 15}
+```{r dataExplore, fig.width = 5, fig.height = 5}
 my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
@@
 p[[1]] = p[[1]] + ggtitle('RIN') + my.theme
+p[[1]]
 
@@
 p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme
+p[[2]]
 
@@
 p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme
+p[[3]]
 
-multiplot(plotlist = p, cols = 1)
+# Intronic Rate
+p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate)) + geom_boxplot()
+p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme
+p[[4]]
+
+# IntergenicRate
+p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate)) + geom_boxplot()
+p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme
+p[[5]]
+
+# Transcripts Detected
+p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TranscriptsDetected)) + geom_boxplot()
+p[[6]] = p[[6]] + ggtitle('Transcripts Detected') + my.theme
+p[[6]]
+
+# Mapped Reads
+p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot()
+p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme
+p[[7]]
+
+# PercentAligned
+p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PercentAligned)) + geom_boxplot()
+p[[8]] = p[[8]] + ggtitle('Percent Aligned') + my.theme
+p[[8]]
+
+# rRNARate
+p[[9]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate)) + geom_boxplot()
+p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme
+p[[9]]
+
+# Institution
+vcd::mosaic(~ Institution + Dx.Tissue, data = COVARIATES)
+
+# Gender
+vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
 ```
@@
 ```{r get.gene.symbols, cache=TRUE, eval=T}
-backgroundGenes = data.frame(ensembl_gene_id = rownames(COUNT))
-
-# Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "mar2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
-
-# Query biomart
-Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
-                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
-                       mart = mart)
-
 # Add gene names

### 2016-11-24 code/Rmd/ROSMAP_Comparison_All_Data.Rmd
Added markdown to compare all versions of ROSMAP

---
title: "Compare different versions of ROSMAP data"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)
library(githubr)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./ROSMAP_Comparison_All_Data.Rmd",
                                 parentId ="syn5570291",
                                 entityName = 'ROSMAP RNASeq Comparisons')
```

### Download all versions of LogCPM from synapse
```{r data.download}
library(CovariateAnalysis)
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)
library(ggplot2)
library(Vennerable)
library(edgeR)
library(geneLenDataBase)

library(synapseClient)
library(knitr)
library(githubr)

synapseLogin()

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r get.data}
fpkm = downloadFile('syn3505720')
rownames(fpkm) = fpkm$tracking_id
fpkm = fpkm[,-c(1:2)]

cnames = data.frame(cnames = colnames(fpkm)) %>%
  dplyr::mutate(cn = cnames) %>%
  separate(cn, c('c1','c2','c3'), sep = '\\_') %>%
  unite(cn, c1, c2)
fpkm = fpkm[,which(!duplicated(cnames$cn))]
colnames(fpkm) = cnames$cn[which(!duplicated(cnames$cn))]

count = list(Lilly = read.table(synGet('syn6112813')@filePath, sep = '\t', quote = '', check.names = F),
             Old = read.table(synGet('syn7211360')@filePath, sep = '\t', quote = '', check.names = F),
             New = read.table(synGet('syn7750270')@filePath, sep = '\t', quote = '', check.names = F))

lcpm = list(Lilly = downloadFile('syn6114447'),
            Old = downloadFile('syn7211581'),
            New = downloadFile('syn7765662'))
```

```{r compute.rpkm}
# Define biomart object
mart <- useMart(host = "jul2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl", biomart = "ENSEMBL_MART_ENSEMBL")

# Get genes to query in biomart
Gene.ID = lapply(count, rownames) %>%
  Reduce(c, .) %>%
  unique %>% 
  data.frame(Gene.ID = .) %>%
  dplyr::mutate(GeneID = Gene.ID) %>%
  tidyr::separate(Gene.ID, c('ensembl_gene_id','position'), sep = '\\.') %>% 
  unique

# Query biomart for transcript length
Ensemble2GeneLength <- getBM(attributes = c("ensembl_gene_id", "transcript_length"),
                             filters = "ensembl_gene_id", values = Gene.ID$ensembl_gene_id,
                             mart = mart) %>%
  group_by(ensembl_gene_id) %>%
  summarise(gene.len = median(transcript_length, na.rm = T)) %>%
  left_join(Gene.ID)

# Compute fpkm from counts
all.fpkm = lapply(count, function(x, ens.len){
  len = ens.len$gene.len
  names(len) = ens.len$GeneID
  len = len[rownames(x)]

  x = edgeR::rpkm(x, gene.length = len)
  rs = rowSums(x, na.rm = T) 
  x = x[!is.na(rs) & (rs != 0),]
  return(x)
}, Ensemble2GeneLength)

# Convert lcpm and fpkm matrices to ensemble ids
all.fpkm = lapply(all.fpkm, function(x){
  GeneID = data.frame(Gene.ID = rownames(x)) %>%
    dplyr::mutate(GeneID = Gene.ID) %>%
    tidyr::separate(Gene.ID, c('ensembl_gene_id','position'), sep = '\\.')

  y = WGCNA::collapseRows(x, GeneID$ensembl_gene_id, rownames(x))$datETcollapsed
  return(y)
})

lcpm = lapply(lcpm, function(x){
  x = x[!duplicated(x$Gene.ID),]
  rownames(x) = x$Gene.ID
  y = WGCNA::collapseRows(x[,-c(1:3)], x$ensembl_gene_id, x$Gene.ID)$datETcollapsed
  return(y)
})

GeneID = data.frame(Gene.ID = rownames(fpkm)) %>%
  dplyr::mutate(GeneID = Gene.ID) %>%
  tidyr::separate(Gene.ID, c('ensembl_gene_id','position'), sep = '\\.')
all.fpkm$Portal = WGCNA::collapseRows(fpkm, GeneID$ensembl_gene_id, rownames(fpkm))$datETcollapsed

###################################################
# Find unique IDs and samples across all data
GeneID = lapply(all.fpkm, rownames) %>%
  Reduce(intersect, .) %>%
  intersect(lapply(lcpm, rownames) %>%
              Reduce(intersect,.)) %>%
  unique

SampleID = lapply(all.fpkm, colnames) %>%
  Reduce(intersect, .) %>%
  intersect(lapply(lcpm, colnames) %>%
              Reduce(intersect, .)) %>%
  unique

all.fpkm = lapply(all.fpkm, function(x, GeneID, SampleID){
  x = x[GeneID, SampleID]
}, GeneID, SampleID)

lcpm = lapply(lcpm, function(x, GeneID, SampleID){
  x = x[GeneID, SampleID]
}, GeneID, SampleID)
```

```{r correlate.fpkm, fig.height=5, fig.width=8}
cr1 = stats::cor(all.fpkm[['Lilly']],all.fpkm[['Old']], use = 'pairwise.complete.obs') 
all.cr = data.frame(data = 'Lilly_vs_Old', values = diag(cr1))

cr2 = stats::cor(all.fpkm[['Lilly']],all.fpkm[['New']], use = 'pairwise.complete.obs') 
all.cr = rbind(all.cr, data.frame(data = 'Lilly_vs_New', values = diag(cr2)))

cr3 = stats::cor(all.fpkm[['Old']],all.fpkm[['New']], use = 'pairwise.complete.obs') 
all.cr = rbind(all.cr, data.frame(data = 'Old_vs_New', values = diag(cr3)))

cr4 = stats::cor(all.fpkm[['Lilly']],all.fpkm[['Portal']], use = 'pairwise.complete.obs') 
all.cr = rbind(all.cr, data.frame(data = 'Lilly_vs_Portal', values = diag(cr4)))

cr5 = stats::cor(all.fpkm[['Old']],all.fpkm[['Portal']], use = 'pairwise.complete.obs') 
all.cr = rbind(all.cr, data.frame(data = 'Old_vs_Portal', values = diag(cr5)))

cr6 = stats::cor(all.fpkm[['New']],all.fpkm[['Portal']], use = 'pairwise.complete.obs') 
all.cr = rbind(all.cr, data.frame(data = 'New_vs_Portal', values = diag(cr6)))

writeLines('')
p = ggplot(all.cr, aes(x = data, y = values)) + geom_boxplot()
p = p + ylab('Correlation') + xlab('') + ggtitle('Correlation between FPKM values')
p
```

```{r correlate.lcpm, fig.height=5, fig.width=5}
cr1 = stats::cor(lcpm[['Lilly']],lcpm[['Old']], use = 'pairwise.complete.obs') 
all.cr = data.frame(data = 'Lilly_vs_Old', values = diag(cr1))

cr2 = stats::cor(lcpm[['Lilly']],lcpm[['New']], use = 'pairwise.complete.obs') 
all.cr = rbind(all.cr, data.frame(data = 'Lilly_vs_New', values = diag(cr2)))

cr3 = stats::cor(lcpm[['Old']],lcpm[['New']], use = 'pairwise.complete.obs') 
all.cr = rbind(all.cr, data.frame(data = 'Old_vs_New', values = diag(cr3)))

writeLines('')
p = ggplot(all.cr, aes(x = data, y = values)) + geom_boxplot()
p = p + ylab('Correlation') + xlab('') + ggtitle('Correlation between lcpm values')
p
```

```{r sample.dev, results='asis'}
PC = lapply(lcpm, function(x){ prcomp(x, center = T, scale. = T)})

writeLines('Percent variance explained by first 10 PCs')
tmp = sapply(PC, function(x){
  pv = (x$sdev^2/sum(x$sdev^2))*100
  pv[1:10]
}) 
rownames(tmp) = paste0('PC',1:10)
kable(tmp)

writeLines('Deviating samples (2SD from PC1 median) in all three cases')
dev.samples = lapply(PC, function(x){ 
  lb = median(x$rotation[,1]) - 2*sd(x$rotation[,1]) 
  ub = median(x$rotation[,1]) + 2*sd(x$rotation[,1]) 
  rownames(x$rotation)[c(which(x$rotation[,1]<lb), which(x$rotation[,1] > ub))]
})
plot(Venn(dev.samples), doWeights = F)
tmp = mapply(function(x, y){writeLines(y); writeLines(paste(x, collapse = ','))}, dev.samples, names(dev.samples))
```

### 2016-11-24 code/Rmd/ROSMAP_RNASeq_Realigned_SynBAM_StarHTSeq.Rmd
Added markdown to compare all versions of ROSMAP

@@
 library(biomaRt)
+library(GGally)
 
@@
 
-```{r residual.analysis}
+```{r residual.analysis, fig.height=5, fig.width=5}
 expr = RESIDUAL.GENE_EXPRESSION[!duplicated(RESIDUAL.GENE_EXPRESSION$Gene.ID),-c(1:3)]
@@
 barplot((PC$sdev^2/sum(PC$sdev^2))[1:10], main = 'Fraction variance explained (top 10 PC)')
+```
+```{r residual.analysis1, fig.height=15, fig.width=15}
 p <- GGally::ggpairs(plotdata, mapping = aes(shape=cogdx, color = Batch, size=RINcontinuous),

### 2016-11-30 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM_CD1.Rmd
Added more technical variables to the custom design matrix

@@
 
-cl = makeCluster(14)
+cl = makeCluster(38)
 registerDoParallel(cl)
@@
 
-```{r dataExplore, fig.width = 6, fig.height = 15}
-my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
+```{r dataExplore, fig.width = 5, fig.height = 5}
+my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1),
+                  plot.title=element_text(hjust=0.5))
 
@@
 ### Normalisation (custom design)
-Normalising COUNTS with a custom design matrix.
-Here Instituiton, Gender, LibraryBatch, Age\\_of\\_Death, PMI, RIN, RIN2, EV.1, EV.2, EV.3, EV.4, EV.5 are adjusted conditioned on Dx.Tissue with Individual\\_ID as random effects.
 ```{r iterativeNorm, results='asis'}
 # Primary variable of interest
-postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'LibraryBatch', 'Age\\_of\\_Death',
-                     'PMI', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5');
+postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'IntronicRate', 'IntergenicRate', 'MappedReads', 'PercentAligned',
+                     'RIN', 'PMI', 'Age_of_Death', 'rRNARate', 'RibozeroBatch', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5',
+                     'GenesDetected');
 
@@
 ### Residual calculation
-Calculate weighted residuals and add back "Dx" to the residuals
+Calculate weighted residuals and add back "Dx.Tissue" to the residuals
 ```{r varsToAddBack}
@@
 # Query biomart
-Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
+Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content", "transcript_length"),
                        filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                        mart = mart)
+
+GC.CONT = Ensemble2HGNC %>%
+  dplyr::select(ensembl_gene_id, percentage_gc_content) %>%
+  unique
+rownames(GC.CONT) = GC.CONT$ensembl_gene_id
+
+Gx.LEN = Ensemble2HGNC %>%
+  dplyr::select(ensembl_gene_id, transcript_length) %>%
+  group_by(ensembl_gene_id) %>%
+  summarise(transcript_length = max(transcript_length, na.rm=T)) %>%
+  unique() %>% as.data.frame()
+rownames(Gx.LEN) = Gx.LEN$ensembl_gene_id
 
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  left_join(Ensemble2HGNC) %>%
-  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+  left_join(Ensemble2HGNC %>%
+              dplyr::select(ensembl_gene_id, hgnc_symbol) %>%
+              unique()) %>%
+  unique
 
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  left_join(Ensemble2HGNC) %>%
-  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+    left_join(Ensemble2HGNC %>%
+              dplyr::select(ensembl_gene_id, hgnc_symbol) %>%
+              unique()) %>%
+  unique
 ```
@@
 tmp = DE %>%
+  dplyr::mutate(dir = logFC/abs(logFC)) %>%
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
-  group_by(Comparison, Tissue) %>%
+  dplyr::select(ensembl_gene_id, Comparison, Tissue, dir) %>%
+  group_by(Comparison, Tissue, dir) %>%
   dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
@@
 tmp1 = DE %>%
+  dplyr::mutate(dir = logFC/abs(logFC)) %>%
   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
-  group_by(Comparison, Tissue) %>%
+  dplyr::select(ensembl_gene_id, Comparison, Tissue, dir) %>%
+  group_by(Comparison, Tissue, dir) %>%
   dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
@@
 p
+```
+
+```{r gc.gx, fig.width = 7, fig.height = 14}
+tmp = left_join(DE, GC.CONT) %>%
+  left_join(Gx.LEN) %>%
+  unique
+
+p = list()
+p[[1]] = ggplot(tmp, aes(x = log10(transcript_length), y = logFC)) + geom_point() + geom_smooth(method = 'lm')
+p[[1]] = p[[1]] + xlab('Log10 of Maximum Tx Length')
+
+p[[2]] = ggplot(tmp, aes(x = log10(percentage_gc_content), y = logFC)) + geom_point() + geom_smooth(method = 'lm')
+p[[2]] = p[[2]] + xlab('Percent GC Content')
+CovariateAnalysis::multiplot(plotlist = p, cols = 2)
 ```
@@
 # Store design matrix
-write.table(DM1$design, file = 'CMC_ACC_Design.tsv', sep = '\t', row.names=F, quote=F)
+write.table(DM1$design, file = 'CMC_ACC_Design.tsv', sep = '\t', row.names=T, quote=F)
 DM_OBJ = File('CMC_ACC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)

### 2016-12-06 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM.Rmd
Added new desing for SVA based adjustments

@@
 # Query biomart
-Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content"),
+Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content", "transcript_length"),
                        filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
@@
 rownames(GC.CONT) = GC.CONT$ensembl_gene_id
+
+Gx.LEN = Ensemble2HGNC %>%
+  dplyr::select(ensembl_gene_id, transcript_length) %>%
+  group_by(ensembl_gene_id) %>%
+  summarise(transcript_length = max(transcript_length, na.rm=T)) %>%
+  as.data.frame()
+rownames(Gx.LEN) = Gx.LEN$ensembl_gene_id
 ```
@@
 ```{r dataExplore, fig.width = 5, fig.height = 5}
-my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
+my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1),
+                  plot.title=element_text(hjust=0.5))
 
@@
 p <- p + geom_point(aes(color=Institution, shape = Dx.Tissue, size=RIN)) 
-p <- p + theme_bw() + theme(legend.position="top")
+p <- p + theme_bw() + theme(legend.position = "top")
 # p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)

### 2016-12-06 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM_SVA.Rmd
Added new desing for SVA based adjustments

--- 2016-11-15 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM_SVA.Rmd
@@
 #### Obtain count matrix and metadata from synapse.
+### Data download
+#### Obtain count matrix and metadata from synapse.
 ```{r download.data, cache=TRUE}
@@
 ## Metadata with specific information for RNASeq processing
-metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", 
-                          "Dx", "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
+metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", "Dx",
+
+                          "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                           "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?", "ACC_RNA_report: Library Batch",
+                          "ACC_RNA_report: Ribozero Batch", "ACC_RNA_report: Flowcell Batch", 
+                          "ACC_RNA_report: Mapped Reads", "ACC_RNA_report: Intragenic Rate",
+                          "ACC_RNA_report: Intronic Rate", "ACC_RNA_report: Intergenic Rate",
+                          "ACC_RNA_report: Genes Detected", "ACC_RNA_report: Transcripts Detected",
+                          "ACC_RNA_report: Expression Profiling Efficiency", "ACC_RNA_report: rRNA Rate",
+                          "ACC_RNA_report: Total Reads", "ACC_RNA_report: Percent Aligned",
+
                           "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                           "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
-                          'DNA_genotyping_report: Genotyping Sample_ID')
+                          "DLPFC_RNA_report: Ribozero Batch", "DLPFC_RNA_report: Flowcell Batch",
+                          "DLPFC_RNA_report: Mapped Reads", "DLPFC_RNA_report: Intragenic Rate",
+                          "DLPFC_RNA_report: Intronic Rate", "DLPFC_RNA_report: Intergenic Rate", 
+                          "DLPFC_RNA_report: Genes Detected", "DLPFC_RNA_report: Transcripts Detected", 
+                          "DLPFC_RNA_report: Expression Profiling Efficiency", "DLPFC_RNA_report: rRNA Rate",
+                          "DLPFC_RNA_report: Total Reads", "DLPFC_RNA_report: Percent Aligned",
+
+                          "DNA_genotyping_report: Genotyping Sample_ID",
+                          "DNA_genotyping_report: Exclude?")
 
@@
 METADATA_ACC = METADATA %>%
-  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, ACC_RNA_isolation__Exclude_,
-                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_,
-                ACC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
+  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
+                ACC_RNA_isolation__Exclude_, ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN,
+                ACC_RNA_report__Exclude_, ACC_RNA_report__Ribozero_Batch, ACC_RNA_report__Library_Batch, 
+                ACC_RNA_report__Flowcell_Batch, ACC_RNA_report__Mapped_Reads, ACC_RNA_report__Intragenic_Rate,
+                ACC_RNA_report__Intronic_Rate, ACC_RNA_report__Intergenic_Rate, ACC_RNA_report__Genes_Detected,
+                ACC_RNA_report__Transcripts_Detected, ACC_RNA_report__Expression_Profiling_Efficiency, 
+                ACC_RNA_report__rRNA_Rate, ACC_RNA_report__Total_Reads, ACC_RNA_report__Percent_Aligned,
+                DNA_genotyping_report__Genotyping_Sample_ID, DNA_genotyping_report__Exclude_) %>%
   dplyr::rename(PMI = PMI__in_hours_, 
@@
                 ReportExclude = ACC_RNA_report__Exclude_, 
+                RibozeroBatch = ACC_RNA_report__Ribozero_Batch,
                 LibraryBatch = ACC_RNA_report__Library_Batch,
-                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
+                FlowcellBatch = ACC_RNA_report__Flowcell_Batch, 
+                MappedReads = ACC_RNA_report__Mapped_Reads, 
+                IntragenicRate = ACC_RNA_report__Intragenic_Rate,
+                IntronicRate = ACC_RNA_report__Intronic_Rate, 
+                IntergenicRate = ACC_RNA_report__Intergenic_Rate, 
+                GenesDetected = ACC_RNA_report__Genes_Detected,
+                TranscriptsDetected = ACC_RNA_report__Transcripts_Detected, 
+                ExpProfEfficiency = ACC_RNA_report__Expression_Profiling_Efficiency, 
+                rRNARate = ACC_RNA_report__rRNA_Rate, 
+                TotalReads = ACC_RNA_report__Total_Reads, 
+                PercentAligned = ACC_RNA_report__Percent_Aligned,
+                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID,
+                GenotypingExclude = DNA_genotyping_report__Exclude_) %>%
   dplyr::filter(SampleID %in% colnames(COUNT)) %>%
@@
 METADATA_DLPFC = METADATA %>%
-  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, DLPFC_RNA_isolation__Exclude_,
-                DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_,
-                DLPFC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
+  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
+                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN,
+                DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Ribozero_Batch, DLPFC_RNA_report__Library_Batch, 
+                DLPFC_RNA_report__Flowcell_Batch, DLPFC_RNA_report__Mapped_Reads, DLPFC_RNA_report__Intragenic_Rate,
+                DLPFC_RNA_report__Intronic_Rate, DLPFC_RNA_report__Intergenic_Rate, DLPFC_RNA_report__Genes_Detected,
+                DLPFC_RNA_report__Transcripts_Detected, DLPFC_RNA_report__Expression_Profiling_Efficiency, 
+                DLPFC_RNA_report__rRNA_Rate, DLPFC_RNA_report__Total_Reads, DLPFC_RNA_report__Percent_Aligned,
+                DNA_genotyping_report__Genotyping_Sample_ID, DNA_genotyping_report__Exclude_) %>%
   dplyr::rename(PMI = PMI__in_hours_, 
@@
                 ReportExclude = DLPFC_RNA_report__Exclude_, 
+                RibozeroBatch = DLPFC_RNA_report__Ribozero_Batch,
                 LibraryBatch = DLPFC_RNA_report__Library_Batch,
-                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
+                FlowcellBatch = DLPFC_RNA_report__Flowcell_Batch, 
+                MappedReads = DLPFC_RNA_report__Mapped_Reads, 
+                IntragenicRate = DLPFC_RNA_report__Intragenic_Rate,
+                IntronicRate = DLPFC_RNA_report__Intronic_Rate, 
+                IntergenicRate = DLPFC_RNA_report__Intergenic_Rate, 
+                GenesDetected = DLPFC_RNA_report__Genes_Detected,
+                TranscriptsDetected = DLPFC_RNA_report__Transcripts_Detected, 
+                ExpProfEfficiency = DLPFC_RNA_report__Expression_Profiling_Efficiency, 
+                rRNARate = DLPFC_RNA_report__rRNA_Rate, 
+                TotalReads = DLPFC_RNA_report__Total_Reads, 
+                PercentAligned = DLPFC_RNA_report__Percent_Aligned,
+                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID,
+                GenotypingExclude = DNA_genotyping_report__Exclude_) %>%
   dplyr::filter(SampleID %in% colnames(COUNT)) %>%
@@
 METADATA <- METADATA  %>% filter(IsolationExclude == 0, ReportExclude == 0)
+
+writeLines('Following samples are marked excluded in genotyping')
+writeLines(paste(METADATA$SampleID[METADATA$GenotypingExclude == 1], collapse = ', '))
+METADATA <- METADATA  %>% filter(GenotypingExclude == 0)
 
@@
 ContCovariates = c(ContCovariates, 'RIN2')
+
+## Get GC content from biomart
+backgroundGenes = data.frame(ensembl_gene_id = rownames(COUNT))
+
+# Define biomart object
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "mar2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
+
+# Query biomart
+Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content", "transcript_length"),
+                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
+                       mart = mart)
+GC.CONT = Ensemble2HGNC %>%
+  dplyr::select(ensembl_gene_id, percentage_gc_content) %>%
+  unique
+rownames(GC.CONT) = GC.CONT$ensembl_gene_id
+
+Gx.LEN = Ensemble2HGNC %>%
+  dplyr::select(ensembl_gene_id, transcript_length) %>%
+  group_by(ensembl_gene_id) %>%
+  summarise(transcript_length = max(transcript_length, na.rm=T)) %>%
+  as.data.frame()
+rownames(Gx.LEN) = Gx.LEN$ensembl_gene_id
 ```
@@
 ```
-
 ### Normalisation (custom design)
@@
 
-Here Institution, Gender, RIN, Age\\_of\\_Death, PMI, EV.4 were used as fixed effects conditioned on Dx.Tissue with Individual\\_ID as random effects.
+Here Dx.Tissue, Institution, Gender, Intronic Rate, Intergenic Rate, Transcripts Detected, Mapped Reads, Percent Aligned, RIN, PMI, Age\\_of\\_Death, rRNA Rate, Ribozero Batch and EV.1, EV.2, EV.3, EV.4 and EV.5 were chosen as fixed effects and Individual\\_ID as random effects.
 ```{r customNorm, results='asis'}
 # Primary variable of interest
-postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'RIN', 'Age_of_Death', 'PMI', 'EV.4');
+postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'IntronicRate', 'IntergenicRate', 'TranscriptsDetected', 
+                     'MappedReads', 'PercentAligned', 'RIN', 'PMI', 'Age_of_Death', 'rRNARate', 'RibozeroBatch',
+                     'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5');
 

### 2017-01-10 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM_SVA.Rmd
Minor modifications

@@
 MODEL0 = DM1$design[,-grep('Dx.Tissue',colnames(DM1$design))] # Get null model by removing variable of interest
+MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
 MODEL1 = DM1$design
+MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
 
 ## Calculate the number of surrogate variables
-NUM.SV = sva::num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1,
-                     method = 'be', seed = 123456, B = 20)
+source('../R/num.sv.R')
+NUM.SV = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1,
+                method = 'be', seed = 123456, B = 20, tol = 1e-22)
 
 SURR.VAR = sva::sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0,
-                    block = COVARIATES$Individual_ID, 
                     n.sv = NUM.SV, B = 30)$sv

### 2017-01-10 code/R/num.sv.R
temp commit

@@ -1,0 +1,77 @@
+num.sv <- function (dat, mod, method = c("be", "leek"), vfilter = NULL, B = 20, seed = NULL, tol = 1e-22) 
+{
+  if (!is.null(vfilter)) {
+    if (vfilter < 100 | vfilter > dim(dat)[1]) {
+      stop(paste("The number of genes used in the analysis must be between 100 and", 
+                 dim(dat)[1], "\n"))
+    }
+    tmpv = rowVars(dat)
+    ind = which(rank(-tmpv) < vfilter)
+    dat = dat[ind, ]
+  }
+  method <- match.arg(method)
+  if (method == "be") {
+    if (!is.null(seed)) {
+      set.seed(seed)
+    }
+    warn <- NULL
+    n <- ncol(dat)
+    m <- nrow(dat)
+    H <- mod %*% solve(t(mod) %*% mod, tol = tol) %*% t(mod)
+    res <- dat - t(H %*% t(dat))
+    uu <- svd(res)
+    ndf <- min(m, n) - ceiling(sum(diag(H)))
+    dstat <- uu$d[1:ndf]^2/sum(uu$d[1:ndf]^2)
+    dstat0 <- matrix(0, nrow = B, ncol = ndf)
+    for (i in 1:B) {
+      res0 <- t(apply(res, 1, sample, replace = FALSE))
+      res0 <- res0 - t(H %*% t(res0))
+      uu0 <- svd(res0)
+      dstat0[i, ] <- uu0$d[1:ndf]^2/sum(uu0$d[1:ndf]^2)
+    }
+    psv <- rep(1, n)
+    for (i in 1:ndf) {
+      psv[i] <- mean(dstat0[, i] >= dstat[i])
+    }
+    for (i in 2:ndf) {
+      psv[i] <- max(psv[(i - 1)], psv[i])
+    }
+    nsv <- sum(psv <= 0.1)
+    return(as.numeric(list(n.sv = nsv)))
+  }
+  else {
+    dat <- as.matrix(dat)
+    dims <- dim(dat)
+    a <- seq(0, 2, length = 100)
+    n <- floor(dims[1]/10)
+    rhat <- matrix(0, nrow = 100, ncol = 10)
+    P <- (diag(dims[2]) - mod %*% solve(t(mod) %*% mod, tol = tol) %*% 
+            t(mod))
+    for (j in 1:10) {
+      dats <- dat[1:(j * n), ]
[…]

### 2017-01-12 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM.Rmd
Minor modifications on ROSMAP reprocessed RNASeq pipeline with TMM normalisation

--- 2016-06-01 code/Rmd/ROSMAP_RNASeq_Reprocessed.Rmd
@@
 ---
-title: "Covariate analysis of ROSMAP Reprocessed RNASeq"
+title: "Covariate analysis of ROSMAP reprocessed counts (with TMM normalisation)"
 author: "Thanneer Perumal"
@@
 ```{r knit2synapse, eval=FALSE}
-library(knitr)
+library(synapseClient)
 library(knit2synapse)
-library(synapseClient)
 
@@
 
-knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed.Rmd",
-                                 parentId ="syn5570291",
-                                 entityName = 'ROSMAP RNASeq Reprocessed Counts (Based on cogdx)')
+knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_TMM.Rmd",
+                                 parentId = "syn5570291",
+                                 entityName = 'ROSMAP Reprocessed RNASeq TMM (Based on cogdx)')
 ```
@@
 ```{r download.data, cache=TRUE}
-# Download reprocessed counts (paired end)
-COUNT_PE_ID <- 'syn6112813';
-ALL_USED_IDs <- COUNT_PE_ID
-COUNT_PE <- read.table(synGet(COUNT_PE_ID)@filePath, header=T, sep='\t', check.names = F)
-
-# Download reprocessed counts (single end)
-COUNT_SE_ID <- 'syn6117775';
-ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_SE_ID
-COUNT_SE <- read.table(synGet(COUNT_SE_ID)@filePath, header=T, sep='\t', check.names = F)
-
-COUNT = COUNT_PE + COUNT_SE[rownames(COUNT_PE), colnames(COUNT_PE)]
+# Download expression data
+COUNT_ID <- 'syn7750270';
+ALL_USED_IDs <- COUNT_ID
+COUNT_OBJ <- synGet(COUNT_ID)
+COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F)
 
@@
 METADATA <- METADATA %>%
+  ungroup %>%
   filter(Sampleid %in% colnames(COUNT)) %>%
@@
   filter(!is.na(pmi)) %>%
-  as.data.frame %>%
-  filter(!(Sampleid %in% c('380_120503')))
+  filter(!is.na(Sampleid)) %>%
+  filter(!(Sampleid %in% c('380_120503'))) %>%
+  as.data.frame()
 ```
@@
 
-Remove genes that have less than 1 cpm counts in at least 50% of samples.
-```{r cpmnormalisation, cache=TRUE}
-PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT,MIN_GENE_CPM=1, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
+Remove genes that have less than 1 cpm counts in at least 50% of samples per cogdx
+```{r cpmnormalisation}
+genesToAnalyze = COVARIATES %>%
+  rownameToFirstColumn('Sampleid') %>%
+  dlply(.(cogdx), .fun = function(mtd, count){
+  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$Sampleid],
+                                                   MIN_GENE_CPM=1, 
+                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
+  processed.counts$filteredExprMatrix$genes
+}, COUNT)
+genesToAnalyze = unlist(genesToAnalyze) %>% unique()
+
+PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
 ```
+Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples
+
 ### Normalisation (with NULL)
 Initial normalisation usign voom (with NULL design)
-```{r initial.voom.normalisation}
+```{r initial.voom.normalisation, fig.height=5, fig.width=5}
+# TMM normalisation
+TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')
+
 # Initial normalisation of gene expression
-VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T)
+VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
+
+# Set gene counts in specific samples that are deviating 3 sd from other samples to NA
+log.mat = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
+  mn = mean(x, na.rm = T)
+  std.dev = sd(x, na.rm = T)
+  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
+}) %>% t
+PROCESSED_COUNTS$filteredExprMatrix[log.mat] = NA
+
+# TMM normalisation
+TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')
+
+# Initial normalisation of gene expression
+VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
+```
+Coexpression of genes 
+```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
+cr = cor(t(VOOM.GENE_EXPRESSION$E))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
 ```
@@
 # Plot first 2 PCs
-plotdata <- data.frame(ID=rownames(PC$rotation), 
+plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                        PC1=PC$rotation[,1], 
@@
 
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'ID'))
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=cogdx, shape=Batch, size=age_death))
-p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
-# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
+p <- p + geom_point(aes(color=Batch, shape=cogdx, size=RINcontinuous))
+p <- p + theme_bw() + theme(legend.position="top")
+# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
 p
-
+```
+Tree based classification of samples
+```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
+COVARIATES.tmp = data.matrix(COVARIATES[,c('Batch', 'msex', 'cogdx')])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
@@
                            main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp[,FactorCovariates]))
+                           groupLabels = colnames(COVARIATES.tmp[,c('Batch', 'msex', 'cogdx')]))
 ```
+
 ```{r temp, include = F}
@@
 ```
+### Normalisation (iterative design)
+Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix.
+Here "cogdx" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on cognitive scores)
 
-### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "cogdx" is chosen as the primary variable of interest
-```{r iterativeNorm}
+Adding Batch and msex a priori to variable selection
+```{r customNorm, results='asis'}
 # Primary variable of interest
-postAdjustCovars = 'cogdx'
+postAdjustCovars = c('cogdx', 'Batch', 'msex');
 
@@
 covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
+postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
 
@@
 
-  # Estimate correlation of random effects
-  VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)
+  # Estimate voom weights
+  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
 
@@
   # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates), postAdjustCovars)
+  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars))
 
@@
   # Add postadjusted covariates (if any)
-  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
+  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
+                                                                          c(postAdjustCovars))]
 
@@
 }
+
 modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
@@
 ```
-`r tmp`
+
 ### Sanity check
-```{r residualSigCovars.manual, fig.width=20, fig.height=12}
+```{r residualSigCovars.manual, fig.width=12, fig.height=8}
 residualSigCovars[["PC_res"]][[2]]$plotData
 ```
+
 ### Residual calculation
@@
 ```
+Coexpression of genes 
+```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
+cr = cor(t(RESIDUAL.GENE_EXPRESSION))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
+```
 
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T)
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
 
-# Plot first 2 PCs
-plotdata <- data.frame(ID=rownames(PC$rotation), 
+# Plot first 4 PCs
+plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                        PC1=PC$rotation[,1], 
@@
 
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'ID'))
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2)) 
-p <- p + geom_point(aes(color=cogdx, shape = Batch, size=RINcontinuous))
-p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
-# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
+p <- p + geom_point(aes(color=Batch, shape=cogdx, size=RINcontinuous))
+p <- p + theme_bw() + theme(legend.position="top")
+# p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)
 p
+```
 
+```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
+COVARIATES.tmp = data.matrix(COVARIATES[,c('Batch', 'msex', 'cogdx')])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
@@
                            main = "Sample dendrogram",
-                          groupLabels = colnames(COVARIATES.tmp[,FactorCovariates]))
+                           groupLabels = colnames(COVARIATES.tmp))
 ```
+
 ```{r temp1, include=F}
@@
 
-```{r summary, results='asis', echo=FALSE}
-writeLines('Number of samples in each category are')
-tmp = data.frame(cogdx = 1:6, No.OfSamples = summary(COVARIATES$cogdx))
-kable(tmp)
+```{r get.gene.symbols, cache=TRUE, eval=T}
+backgroundGenes = data.frame(gene.id = rownames(COUNT)) %>%
+  dplyr::mutate(ensembl_gene_id = gene.id) %>%
+  tidyr::separate(ensembl_gene_id, c('ensembl_gene_id', 'position'), sep = '\\.')
+
+# Define biomart object
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "mar2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
+
+# Query biomart
+Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
+                       filters = "ensembl_gene_id", 
+                       values = backgroundGenes$ensembl_gene_id,
+                       mart = mart)
+
+# Add gene names
+RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
+  rownameToFirstColumn('gene.id') %>%
+  left_join(backgroundGenes) %>%
+  left_join(Ensemble2HGNC) %>%
+  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+
+VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
+  rownameToFirstColumn('gene.id') %>%
+  left_join(backgroundGenes) %>%
+  left_join(Ensemble2HGNC) %>%
+  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
 ```
@@
 
-```{r get.gene.symbols, cache=TRUE, eval=T}
-backgroundGenes = data.frame(Gene.ID = rownames(COUNT),
-                             ID = rownames(COUNT)) %>%
-  separate(ID, c('ensembl_gene_id','position'), sep = '\\.')
-
-# Define biomart object
-mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
-
-# Query biomart
-Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
-                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
-                       mart = mart)
-
-# Add gene names
-RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
-  rownameToFirstColumn('Gene.ID') %>%
-  left_join(backgroundGenes) %>%
-  left_join(Ensemble2HGNC) %>%
-  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
-
-VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
-  rownameToFirstColumn('Gene.ID') %>%
-  left_join(backgroundGenes) %>%
-  left_join(Ensemble2HGNC) %>%
-  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
-```
 ### Differential expression analysis
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=8, fig.width=16}
+```{r diffExp, fig.height=16, fig.width=16}
 # Fit contrast
@@
 # Get differnetial expression
-DE = lapply(1:3, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, backgroundGenes, Ensemble2HGNC){
+DE = lapply(1:3, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
   topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
-    rownameToFirstColumn('Gene.ID') %>%
+    rownameToFirstColumn('gene.id') %>%
     left_join(backgroundGenes) %>%
     left_join(Ensemble2HGNC) %>%
-    dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, logFC, AveExpr, t, P.Value, adj.P.Val, B)
-}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, backgroundGenes, Ensemble2HGNC)
+    dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, logFC, AveExpr, t, P.Value, adj.P.Val, B)
+}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC) 
 names(DE) = colnames(contrast)
@@
 DE = DE %>% 
-  plyr::ldply(.id = 'Comparison')
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')))
 
@@
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(Gene.ID, Comparison) %>%
-  group_by(Comparison) %>%
-  dplyr::summarise(FDR_0_05 = length(unique(Gene.ID)))
+  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
+  group_by(Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05)
 
@@
   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(Gene.ID, Comparison) %>%
-  group_by(Comparison) %>%
-  dplyr::summarise(FDR_0_05_FC_1_2 = length(unique(Gene.ID)))
+  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
+  group_by(Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05_FC_1.2)
 
-kable(full_join(tmp,tmp1))
+kable(full_join(tmp %>%
+                  dplyr::rename(DOWN_FDR_0_05 = DOWN, UP_FDR_0_05 = UP),
+                tmp1 %>%
+                  dplyr::rename(DOWN_FDR_0_05_FC_1.2 = DOWN, UP_FDR_0_05_FC_1.2 = UP)))
 
@@
 # Code
-CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts (Based on cogdx)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP Reprocessed RNASeq TMM (Based on cogdx)", parentId = parentId)
 CODE <- synStore(CODE)
@@
 COVARIATES = rownameToFirstColumn(COVARIATES, 'Sampleid')
-write.table(COVARIATES, file = 'ROSMAPCovariates.tsv', sep = '\t', row.names=F, quote=F)
-COV_OBJ = File('ROSMAPCovariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+write.table(COVARIATES, file = 'ROSMAP_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
+COV_OBJ = File('ROSMAP_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
 COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 
-# Store logCPM
-write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'ROSMAPlogCPM.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('ROSMAPlogCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
+# Store filtered counts
+PROCESSED_COUNTS$filteredExprMatrix %>%
+  rownameToFirstColumn('gene.id') %>%
+  write.table(file = 'ROSMAP_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
+COUNT_OBJ = File('ROSMAP_DLPFC_Counts.tsv', name = 'Raw counts (filtered)', parentId = CODE$properties$id)
 COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 
+# Store logCPM
+write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'ROSMAP_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
+LCOUNT_OBJ = File('ROSMAP_DLPFC_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
+LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                      executed = thisFile, activityDescription = activityDescription)
+
 # Store design matrix
-write.table(DM1$design, file = 'ROSMAPDesign.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('ROSMAPDesign.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+write.table(DM1$design, file = 'ROSMAP_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
+DM_OBJ = File('ROSMAP_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
 DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 # Store residual gene expression
-write.table(RESIDUAL.GENE_EXPRESSION, file = 'ROSMAPExpression.tsv', sep = '\t', row.names=F, quote=F)
-EXP_OBJ = File('ROSMAPExpression.tsv', name = 'Normalised Adjusted Residual Expression (cogdx added)', parentId = CODE$properties$id)
+write.table(RESIDUAL.GENE_EXPRESSION, file = 'ROSMAP_DLPFC_Expression.tsv', sep = '\t', row.names=F, quote=F)
+EXP_OBJ = File('ROSMAP_DLPFC_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', parentId = CODE$properties$id)
 EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 # Store differential expression results
-write.table(DE, file = 'ROSMAPDiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('ROSMAPDiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
+write.table(DE, file = 'ROSMAP_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('ROSMAP_DLPFC_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
 DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 

### 2017-01-13 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM.Rmd
Minor modifications

@@
 PROCESSED_COUNTS$filteredExprMatrix %>%
-  rownameToFirstColumn('gene.id') %>%
+  as.matrix() %>% rownameToFirstColumn('gene.id') %>%
   write.table(file = 'ROSMAP_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)

### 2017-01-13 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM_SVA.Rmd
Added markdown to process reprocessed RNASeq data from ROSMAP cohort with TMM and be based SVA normalisation

--- 2017-01-13 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM.Rmd
@@
 ---
-title: "Covariate analysis of ROSMAP reprocessed counts (with TMM normalisation)"
+title: "Covariate analysis of ROSMAP reprocessed counts (with TMM and SVA normalisation)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_TMM.Rmd",
+knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_TMM_SVA.Rmd",
                                  parentId = "syn5570291",
-                                 entityName = 'ROSMAP Reprocessed RNASeq TMM (Based on cogdx)')
+                                 entityName = 'ROSMAP Reprocessed RNASeq TMM SVA (Based on cogdx)')
 ```
@@
 
-thisFileName <- 'ROSMAP_RNASeq_Reprocessed_TMM.Rmd'
+thisFileName <- 'ROSMAP_RNASeq_Reprocessed_TMM_SVA.Rmd'
 
@@
 ```
+### SVA 
+Estimate surrogate variables using SVA package
+```{r sva}
+MODEL0 = DM1$design[,-grep('cogdx',colnames(DM1$design))] # Get null model by removing variable of interest
+MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
+MODEL1 = DM1$design
+MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
+
+## Calculate the number of surrogate variables
+NUM.SV = sva::num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1,
+                     method = 'be', seed = 123456, B = 30)
+
+SURR.VAR = sva::sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0,
+                    n.sv = NUM.SV, B = 30)$sv
+
+colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
+rownames(SURR.VAR) = rownames(DM1$design)
+
+DM1$design = cbind(MODEL1, SURR.VAR)
+
+# Estimate voom weights
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
+
+# Fit linear model using new weights and new design
+VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
+
+# Residuals after normalisation
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
+```
 
@@
 # Code
-CODE <- Folder(name = "ROSMAP Reprocessed RNASeq TMM (Based on cogdx)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP Reprocessed RNASeq TMM SVA (Based on cogdx)", parentId = parentId)
 CODE <- synStore(CODE)

### 2017-01-16 code/Rmd/Mayo_RNASeq_Reprocessed_TMM.Rmd
Added markdown to process reprocessed RNASeq from Mayo UFL brainbank with TMM normalisation

---
title: "Covariate analysis of MAYO reprocessed counts (with TMM and SVA normalisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed_TMM.Rmd",
                                 parentId = "syn5570325",
                                 entityName = 'UFL_Mayo_ISB Reprocessed RNASeq TMM')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get it from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(reshape2)
library(limma)
library(Biobase)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)
library(glmnet)

library(synapseClient)
library(knitr)
library(githubr)

library(parallel)
library(foreach)

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

synapseLogin()

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn5570325';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of UFL_Mayo_ISB RNASeq data';

thisFileName <- 'Mayo_RNASeq_Reprocessed_TMM.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Get reprocessed sample counts for temporal cortex
COUNT_TC_ID <- 'syn7987765';
ALL_USED_IDs <- COUNT_TC_ID
COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F)
colnames(COUNT_TC) = paste(colnames(COUNT_TC), 'TCX', sep = '_')

# Get reprocessed sample counts for cerebellum
COUNT_CE_ID <- 'syn7513807';
ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_CE_ID
COUNT_CE <- read.table(synGet(COUNT_CE_ID)@filePath, header=T, sep='\t', check.names = F)

COUNT = cbind(COUNT_TC, COUNT_CE[rownames(COUNT_TC),])

# Get clinical metadata
METADATA_TC_ID <- 'syn3817650'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TC_ID
METADATA_TC <- read.table(synGet(METADATA_TC_ID)@filePath,sep='\t',header=T, stringsAsFactors=F) %>%
  tidyr::separate(ID, c('Donor_ID', 'BrainRegion'), sep = '_') %>%
  dplyr::mutate(ID = paste(Donor_ID, BrainRegion, sep = '_'))

# Get clinical metadata
METADATA_CE_ID <- 'syn5223705'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CE_ID
METADATA_CE <- read.table(synGet(METADATA_CE_ID)@filePath,sep='\t',header=T, stringsAsFactors=F) %>%
  tidyr::separate(SampleID, c('Donor_ID', 'BrainRegion')) %>%
  dplyr::mutate(ID = paste(Donor_ID, BrainRegion, sep = '_')) %>%
  dplyr::rename(Gender = Sex, FLOWCELL = Flowcell)

METADATA = rbindlist(list(METADATA_TC, METADATA_CE), use.names = T, fill = T) %>%
  as.data.frame
```

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
METADATA$BrainRegion.Diagnosis = paste(METADATA$BrainRegion, METADATA$Diagnosis, sep = '.')
METADATA$RIN2 = METADATA$RIN^2
METADATA <- METADATA %>%
  filter(!is.na(Source)) %>%
  dplyr::filter(!(ID %in% c("11311_CER", "1923_CER", "1923_TCX", 
                            "11396_TCX", "11294_TCX", "11408_TCX")))
```
Following samples were considered outliers and removed based on PCA and/or eucledian distance: 11311\\_CER, 1923\\_CER, 1923\\_TCX, 11396\\_TCX, 11294\\_TCX, 11408\\_TCX
```{r preprocess.data1}
# Match covariates to expression data
indToRetain = intersect(METADATA$ID, colnames(COUNT))
indRemoved = setdiff(colnames(COUNT), METADATA$ID)

COUNT <- COUNT[, indToRetain]

rownames(METADATA) = METADATA$ID
METADATA = METADATA[indToRetain,]
```
`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project. Following samples were removed `r indRemoved`
### Predict missing PMI
```{r predict.pmi}
PMI = METADATA[,c('PMI', 'ID'), drop = F]
rownames(PMI) = METADATA$ID

# Get genes that are predictors of PMI and use them as control probes
CONTROL.PROBES_ID = 'syn6145639'
ALL_USED_IDs = c(ALL_USED_IDs, CONTROL.PROBES_ID)
CONTROL.PROBES = fread(synGet(CONTROL.PROBES_ID)@filePath, data.table = F)

# Remove genes
EXPR <- getGeneFilteredGeneExprMatrix(COUNT, 
                                      MIN_GENE_CPM=0,
                                      MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0,
                                      verbose=T)

# TMM normalisation
TMM.GENE_EXPRESSION = calcNormFactors(EXPR$filteredExprMatrix, method = 'TMM')

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)

ind = sapply(rownames(VOOM.GENE_EXPRESSION$E), function(x){str_split(x, '\\.')[[1]][1]}) %in% CONTROL.PROBES$x 
EXPR = VOOM.GENE_EXPRESSION$E[ind,]
PMI = PMI[colnames(EXPR),,drop=F]

expr = dlply(PMI, .(ID), function(mtd, expr){
  if (dim(mtd)[1] > 1)
    expr = rowMeans(expr[, mtd$ID]) %>% rownameToFirstColumn('GeneID')
  else
    expr = expr[, mtd$ID] %>% rownameToFirstColumn('GeneID')
  expr = expr %>% plyr::rename(c('DF' = unique(mtd$ID)))
  return(expr)
}, EXPR) %>%
  join_all
rownames(expr) = expr$GeneID

PMI = PMI[,colnames(PMI) %in% c('ID','PMI')] %>% unique
rownames(PMI) = PMI$ID

PMI = PMI[intersect(rownames(PMI), colnames(expr)), ]
expr = expr[,intersect(rownames(PMI), colnames(expr))]

# Split data into trainining and prediction set
x.train = expr[,!is.na(PMI$PMI)]
x.predict = expr[,is.na(PMI$PMI)]

y.train = PMI[!is.na(PMI$PMI),]

plr.model = cv.glmnet(t(x.train), y.train$PMI, family = 'gaussian')
coeff = as.matrix(coef.glmnet(plr.model, s = 'lambda.min'))
mse = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train$PMI)^2)/length(y.train$PMI)
y.predict = predict(plr.model, t(x.predict), s = 'lambda.min', type = 'response') %>%
  rownameToFirstColumn('ID') %>%
  plyr::rename(c('1' = 'PMI'))

PMI = rbindlist(list(y.train, y.predict), use.names = T, fill = T)

METADATA = METADATA %>% 
  dplyr::select(-PMI) %>%
  left_join(PMI)
```

```{r diagnosisPlots, fig.height=10, fig.width = 15}
# RIN
p = list()
p[[1]] = ggplot(METADATA, aes(x = BrainRegion.Diagnosis, y = RIN, fill = Tissue)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# AgeAtDeath
p[[2]] = ggplot(METADATA, aes(x = BrainRegion.Diagnosis, y = as.numeric(AgeAtDeath), fill = Tissue)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeAtDeath') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# PMI
p[[3]] = ggplot(METADATA, aes(x = BrainRegion.Diagnosis, y = as.numeric(PMI), fill = Tissue)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

multiplot(plotlist = p, cols = 3)

# Source
vcd::mosaic(~ Source + Diagnosis + Tissue, data = METADATA)
```

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$ID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Covariate correlation
```{r covariates.correlation, fig.width=8, fig.height=8}
COVARIATES.CORRELATION = CovariateAnalysis::getAssociationStatistics(COVARIATES, PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.4, row.width=0.1)
```
### Gene Filtering
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion.Diagnosis
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('Sampleid') %>%
  dlply(.(BrainRegion.Diagnosis), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$Sampleid],
                                                     MIN_GENE_CPM=1, 
                                                     MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
    processed.counts$filteredExprMatrix$genes
    }, COUNT)
genesToAnalyze = unlist(genesToAnalyze) %>% unique()

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

Sample breakup
```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = summary(COVARIATES$BrainRegion.Diagnosis) %>%
  rownameToFirstColumn('SampleName') %>%
  tidyr::separate(SampleName, c('BrainRegion','Diagnosis'), sep = '\\.') %>%
  tidyr::spread(Diagnosis, DF)
kable(tmp)
```

### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation}
# TMM normalisation
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to NA
log.mat = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix[log.mat] = NA

# TMM normalisation
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
```

Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data1, fig.height=15, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale. = T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(ID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2],
                       PC3=PC$rotation[,3],
                       PC4=PC$rotation[,4])

plotdata <- left_join(plotdata, METADATA, by="ID")

pl = list()

plotdata.cer = plotdata[grep('CER', plotdata$ID),]
p <- ggplot(plotdata.cer, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Tissue), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
pl[[1]] <- p

plotdata.tcx = plotdata[grep('TCX', plotdata$ID),]
p <- ggplot(plotdata.tcx, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Tissue), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
pl[[2]] <- p

multiplot(plotlist = pl, cols = 1)
```
```{r decompse.normalise.data2, fig.height=9, fig.width=25, results='asis'}
# Tree based clustering
sampleTrees = flashClust::hclust(dist(t(VOOM.GENE_EXPRESSION$E)), method = "average")
tree = WGCNA::cutreeStatic(sampleTrees, cutHeight = 135, minSize=10)

COVARIATES.tmp = COVARIATES
COVARIATES.tmp[,FactorCovariates] = lapply(COVARIATES.tmp[,FactorCovariates], as.numeric)
col.mat = labels2colors(cbind(tree,COVARIATES.tmp))
col.mat = gsub('\\.1', '', col.mat)

WGCNA::plotDendroAndColors(sampleTrees, 
                           col.mat,
                           groupLabels = c('Cluster',colnames(COVARIATES.tmp)))

tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[tree == 0], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
```

```{r temp, include=FALSE}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, results='asis', echo=FALSE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, COVARIATES,
                                                 'NULL design(voom-normalized)',
                                                 isKeyPlot=TRUE)

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
```{r iterative.adjusted.voom.normalisation, results='asis'}
postAdjustCovars = c('BrainRegion.Diagnosis', 'Source', 'FLOWCELL', 'Gender')
randomCovariates = 'Donor_ID'

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
significantCovars = residualSigCovars$significantCovars
covariatesEffects = residualSigCovars$Effects.significantCovars
covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates))]

postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))

loopCount = 0 
while(length(significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:', 
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)

  # Calculate correlation values of random effects
  correlation = parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block = COVARIATES$Donor_ID)

  # Re-estimate voom weights
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F, 
                                       block = COVARIATES$Donor_ID, correlation = correlation$cor)

  # Fit a mixed effect linear model
  FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, randomCovariates))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES,
                                                'all adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  significantCovars = residualSigCovars$significantCovars
  covariatesEffects = residualSigCovars$Effects.significantCovars
  covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars, randomCovariates))]

  postAdjustCovars = c(postAdjustCovars,names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
modelStr <-paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=','), 'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Adding "BrainRegion.Diagnosis" back to the residuals
```{r addVarsBack}
varsToAddBackIn <-  grep("BrainRegion.Diagnosis", colnames(FIT$coefficients), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
```

### Clustering residual gene expression
```{r decompse.normalise.data3, fig.height=10, fig.width=9, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

# Tree based clustering
sampleTrees = flashClust::hclust(dist(t(RESIDUAL.GENE_EXPRESSION)), method = "average")

# Cluster expression
clust = cutreeStatic(sampleTrees, cutHeight = 100, minSize = 10)

# Plot first 2 PCs
plotdata <- data.frame(ID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2],
                       PC3=PC$rotation[,3], 
                       PC4=PC$rotation[,4],
                       clust = clust)

plotdata <- left_join(plotdata, METADATA, by="ID")

pl = list()
p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Tissue), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
pl[[1]] <- p

p <- ggplot(plotdata, aes(x=PC3, y=PC4))
p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Diagnosis), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
pl[[2]] <- p

multiplot(plotlist = pl, cols = 1)

names(clust) = sampleTrees$labels
clust = clust[rownames(COVARIATES)]
```
```{r decompse.normalise.data4, fig.height=10, fig.width=25, results='asis'}
# Tree based clustering
COVARIATES.tmp = COVARIATES
COVARIATES.tmp[, FactorCovariates] = sapply(COVARIATES.tmp[,FactorCovariates], as.numeric)
col.mat = labels2colors(cbind(tree,COVARIATES.tmp))
col.mat = gsub('\\.1', '', col.mat)

plotDendroAndColors(sampleTrees, 
                    colors = col.mat, 
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES))
```

```{r temp1, include=FALSE}
gc()
dev.off()
```

```{r get.gene.symbols}
# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "www.ensembl.org")

# Get background gene ids
backgroundGenes = data.frame(gene.id = rownames(COUNT), ID = rownames(COUNT)) %>%
  tidyr::separate(ID, c('ensembl_gene_id','position'), sep = '\\.')

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

# Add gene names
RESIDUAL.GENE_EXPRESSION <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'gene.id') %>%
  left_join(backgroundGenes) %>% 
  left_join(Ensemble2HGNC) %>%
  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))

VOOM.ADJUSTED.GENE_EXPRESSION$E <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, 'gene.id') %>%
  left_join(backgroundGenes) %>% 
  left_join(Ensemble2HGNC) %>%
  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=20}
# Fit contrast
contrast = makeContrasts(contrasts=c(
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.Pathologic_Aging",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisCER.PSP-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.Pathologic_Aging",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.PSP",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.PSP",
  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD"),
  levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:12, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
}, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
names(DE) = gsub('BrainRegion.Diagnosis','',colnames(contrast))

DE = plyr::ldply(DE, .id = 'Comparison') %>%
  dplyr::mutate(Comparison1 = Comparison) %>%
  tidyr::separate(Comparison1, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
  tidyr::unite(Name, reference, against, sep = '-')

p <- ggplot(DE, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
p <- p + geom_vline(xintercept = log2(1.2), color = 'red')
p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
p <- p + facet_grid(BrainRegion~.+Name)
p

tmp = DE %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(Comparison) %>%
  dplyr::summarize(nDiffExpGenes = length(unique(genes)))
kable(tmp)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB Reprocessed RNASeq TMM', parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'Sampleid')
write.table(COVARIATES, file = 'MAYO_CBE_TCX_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('MAYO_CBE_TCX_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix %>%
  as.matrix() %>% rownameToFirstColumn('gene.id') %>%
  write.table(file = 'MAYO_CBE_TCX_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('MAYO_CBE_TCX_Counts.tsv', name = 'Raw counts (filtered)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                     executed = thisFile, activityDescription = activityDescription)

# Store logCPM
write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'MAYO_CBE_TCX_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('MAYO_CBE_TCX_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)

# Store design matrix
write.table(DM1$design, file = 'MAYO_CBE_TCX_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('MAYO_CBE_TCX_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)

# Store residual gene expression
write.table(RESIDUAL.GENE_EXPRESSION, file = 'MAYO_CBE_TCX_Expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('MAYO_CBE_TCX_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', 
               parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store differential expression results
write.table(DE, file = 'MAYO_CBE_TCX_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('MAYO_CBE_TCX_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)

stopCluster(cl)
```
|  *Results*                |  *SynapseID*                    |
|  -------                  |   ---------                     |
|  Covariates               |  `r COV_OBJ$properties$id`      |
|  Raw counts (logCPM)      |  `r RAWCPM_OBJ$properties$id`   |
|  Design Matrix            |  `r DM_OBJ$properties$id`       |
|  Residual Expression      |  `r EXP_OBJ$properties$id`      |
|  Differential Expression  |  `r DEXP_OBJ$properties$id`     |

### Source R Code
[This File](`r thisFile`)

### 2017-01-16 code/Rmd/Mayo_RNASeq_Reprocessed_TMM.Rmd
Minor modifications

@@
 ```
-Following samples were considered outliers and removed based on PCA and/or eucledian distance: 11311\\_CER, 1923\\_CER, 1923\\_TCX, 11396\\_TCX, 11294\\_TCX, 11408\\_TCX
+Following samples were considered outliers and removed based on PCA and/or eucledian distance: 11311\_CER, 1923\_CER, 1923\_TCX, 11396\_TCX, 11294\_TCX, 11408\_TCX
 ```{r preprocess.data1}
@@
 ```
-`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project. Following samples were removed `r indRemoved`
+
+`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project.
+
+Following samples were removed `r indRemoved`
+
 ### Predict missing PMI
@@
 # Initial normalisation of gene expression
-VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
+VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=F)
 
@@
 
-```{r diagnosisPlots, fig.height=10, fig.width = 15}
+```{r diagnosisPlots, fig.height=6, fig.width = 10}
 # RIN
@@
 
-Sample breakup
 ```{r summary, results='asis', echo=FALSE}
@@
 Initial normalisation usign voom (with NULL design)
-```{r initial.voom.normalisation}
+```{r initial.voom.normalisation, fig.height=5, fig.width=5}
 # TMM normalisation
@@
 Clustering of initial normalised data (with NULL design)
-```{r decompse.normalise.data1, fig.height=15, fig.width=10, results='asis'}
+```{r decompse.normalise.data1, fig.height=14, fig.width=7, results='asis'}
 # Find principal components of expression to plot
@@
 |  Covariates               |  `r COV_OBJ$properties$id`      |
-|  Raw counts (logCPM)      |  `r RAWCPM_OBJ$properties$id`   |
+|  Raw counts (logCPM)      |  `r LCOUNT_OBJ$properties$id`   |
 |  Design Matrix            |  `r DM_OBJ$properties$id`       |

### 2017-01-16 code/Rmd/Mayo_RNASeq_Reprocessed_TMM_SVA.Rmd
Added markdown to process reprocessed RNASeq from Mayo Brain bank with TMM and SVA normalisation

--- 2017-01-16 code/Rmd/Mayo_RNASeq_Reprocessed_TMM.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed_TMM.Rmd",
+knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed_TMM_SVA.Rmd",
                                  parentId = "syn5570325",
-                                 entityName = 'UFL_Mayo_ISB Reprocessed RNASeq TMM')
+                                 entityName = 'UFL_Mayo_ISB Reprocessed RNASeq TMM SVA')
 ```
@@
 
-thisFileName <- 'Mayo_RNASeq_Reprocessed_TMM.Rmd'
+thisFileName <- 'Mayo_RNASeq_Reprocessed_TMM_SVA.Rmd'
 
@@
 `r tmp`
+### SVA 
+Estimate surrogate variables using SVA package
+```{r sva}
+MODEL0 = DM1$design[,-grep('BrainRegion.Diagnosis',colnames(DM1$design))] # Get null model by removing variable of interest
+MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
+MODEL1 = DM1$design
+MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
+
+## Calculate the number of surrogate variables
+NUM.SV = sva::num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1,
+                     method = 'be', seed = 123456, B = 30)
+
+SURR.VAR = sva::sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0,
+                    n.sv = NUM.SV, B = 30)$sv
+
+colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
+rownames(SURR.VAR) = rownames(DM1$design)
+
+DM1$design = cbind(MODEL1, SURR.VAR)
+
+# Estimate voom weights
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
+
+# Calculate correlation values of random effects
+correlation = parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block = COVARIATES$Donor_ID)
+
+# Re-estimate voom weights
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F, 
+                                     block = COVARIATES$Donor_ID, correlation = correlation$cor)
+
+# Fit a mixed effect linear model
+FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
+
+# Residuals after normalisation
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
+
+# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
+residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+                                              COVARIATES,
+                                              'all adjusted design(voom-normalized)',
+                                              isKeyPlot=TRUE)
+```
 ### Sanity Check
@@
 # Code
-CODE <- Folder(name = 'UFL_Mayo_ISB Reprocessed RNASeq TMM', parentId = parentId)
+CODE <- Folder(name = 'UFL_Mayo_ISB Reprocessed RNASeq TMM SVA', parentId = parentId)
 CODE <- synStore(CODE)

### 2017-01-21 code/Rmd/MSSM_RNASeq_Reprocessed_TMM.Rmd
Added markdown to process reprocessed RNASeq from Mt.Sinai Brain bank with TMM normalisation

---
title: "Covariate analysis of MSSM reprocessed counts (with TMM normalisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM.Rmd",
                                 parentId = "syn5570248",
                                 entityName = 'MSBB Reprocessed RNASeq TMM')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get it from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(reshape2)
library(limma)
library(Biobase)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)

library(synapseClient)
library(knitr)
library(githubr)

library(parallel)
library(foreach)

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

synapseLogin()

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn5570248';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of MSBB RNASeq data';

thisFileName <- 'MSBB_RNASeq_Reprocessed_TMM.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
Data were obtained from AMP-AD portal of MSSM, and following brain regions were analysed for differential expression between case and control:
  | Region | Name                      | 
  | :----- | ------------------------- | 
  | BM_10  | Frontal Pole (FP)         |
  | BM_22  | Superior Temporal Gyrus   |
  | BM_36  | Parahipocampal Gyrus (PHG)|
  | BM_44  | inferior frontal Gyrus (IFG)|

* BM 10: frontopolar prefrontal cortex, rostrolateral prefrontal cortex, or anterior prefrontal cortex, 
* BM 22: In humans it corresponds approximately to the lateral and caudal two thirds of the superior temporal gyrus, 
* BM 36: Ectorhinal area 36 is a subdivision of the cytoarchitecturally defined temporal region of cerebral cortex,
* BM 44: pars opercularis (of the inferior frontal gyrus), and it refers to a subdivision of the frontal region of cerebral cortex.

#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Get reprocessed sample counts for temporal cortex
COUNT_ID <- 'syn7994853';
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F)

# Get sample ids
SampleID = data.frame(SampleID = colnames(COUNT),
                 ID = colnames(COUNT)) %>% 
  tidyr::separate(ID, c('A','B','ID'), sep = '_')

SampleID = SampleID %>%
  dplyr::select(SampleID, ID) %>%
  dplyr::mutate(ID = as.numeric(ID))

# Get technical metadata
METADATA_TECH_ID <- 'syn6100548'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECH_ID
METADATA_TECH <- read.table(synGet(METADATA_TECH_ID)@filePath,sep=',',header=T, row.names=1) %>%
  group_by(sampleIdentifier) %>%
  top_n(1, -rRNA.rate) %>%
  dplyr::select(sampleIdentifier, BrodmannArea, barcode, individualIdentifier, batch, RIN, TotalReads, Mapped, rRNA.rate) %>%
  unique

# Get clinical metadata
METADATA_CLINICAL_ID <- 'syn6101474'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CLINICAL_ID
METADATA_CLINICAL <- read.csv(synGet(METADATA_CLINICAL_ID)@filePath, header = T)

METADATA <- full_join(METADATA_TECH, METADATA_CLINICAL)
```

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA$AOD = gsub('\\+','',METADATA$AOD)
METADATA$BrodmannArea.CDR = paste(METADATA$BrodmannArea, METADATA$CDR, sep = '.')
METADATA$RIN2 = METADATA$RIN^2
levels(METADATA$batch)[levels(METADATA$batch) == ''] = 'NoBatch'

METADATA <- METADATA %>%
  filter(!is.na(BrodmannArea), !is.na(PMI), !is.na(RIN))
```

```{r preprocess.data1}
# Match covariates to expression data
indToRetain = intersect(METADATA$sampleIdentifier, colnames(COUNT))
indRemoved = setdiff(colnames(COUNT), METADATA$sampleIdentifier)

COUNT <- COUNT[, indToRetain]

METADATA = as.data.frame(METADATA)
rownames(METADATA) = METADATA$sampleIdentifier
METADATA = METADATA[indToRetain,]
```
`r dim(COUNT)[2]` samples from `r length(unique(METADATA$individualIdentifier))` subjects were obtained from the MSBB cohorts in AMP-AD reprocessed rnaseq project

Following `r length(indRemoved)` samples are removed:`r paste(indRemoved, collapse = ', ')`
```{r diagnosisPlots, fig.height=15, fig.width = 10}
# RIN
p = list()
p[[1]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = RIN, fill = BrodmannArea)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# AOD
p[[2]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = as.numeric(AOD), fill = BrodmannArea)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('Age At Death') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# PMI
p[[3]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = as.numeric(PMI), fill = BrodmannArea)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# Total Reads
p[[4]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = as.numeric(TotalReads), fill = BrodmannArea)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Total Reads') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# Mapped Reads
p[[5]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = as.numeric(Mapped), fill = BrodmannArea)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Mapped') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# Mapped Reads
p[[6]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = as.numeric(rRNA.rate), fill = BrodmannArea)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('rRNA Rate') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

multiplot(plotlist = p, cols = 2)
```

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrodmannArea.CDR")
ContCovariates <- c("RIN", "RIN2", "AOD","PMI", "TotalReads", "Mapped", "rRNA.rate")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$sampleIdentifier

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Covariate correlation
```{r covariates.correlation, fig.width=8, fig.height=8}
COVARIATES.CORRELATION = CovariateAnalysis::getAssociationStatistics(COVARIATES, PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.4, row.width=0.1)
```
### Gene Filtering
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

Remove genes that have less than 1 cpm counts in at least 50% of samples per BrodmannArea.CDR
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('sampleIdentifier') %>%
  dlply(.(BrodmannArea.CDR), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$sampleIdentifier],
                                                     MIN_GENE_CPM=1, 
                                                     MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
    processed.counts$filteredExprMatrix$genes
    }, COUNT)
genesToAnalyze = unlist(genesToAnalyze) %>% unique()

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = summary(COVARIATES$BrodmannArea.CDR) %>%
  rownameToFirstColumn('BrodmannArea.CDR') %>%
  tidyr::separate(BrodmannArea.CDR, c('BrodmannArea', 'CDR'), sep = 5) %>%
  tidyr::spread(CDR, DF)
kable(tmp)
```

### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# TMM normalisation
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to NA
log.mat = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix[log.mat] = NA

# TMM normalisation
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
```

Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data1, fig.height=14, fig.width=7, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale. = T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(sampleIdentifier=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2],
                       PC3=PC$rotation[,3],
                       PC4=PC$rotation[,4])

plotdata <- left_join(plotdata, METADATA)

pl = list()

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=factor(CDR), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
pl[[1]] <- p

p <- ggplot(plotdata, aes(x=PC2, y=PC3))
p <- p + geom_point(aes(shape=factor(CDR), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
pl[[2]] <- p

multiplot(plotlist = pl, cols = 1)
```
```{r decompse.normalise.data2, fig.height=9, fig.width=25, results='asis'}
# Tree based clustering
sampleTrees = flashClust::hclust(dist(t(VOOM.GENE_EXPRESSION$E)), method = "average")
tree = WGCNA::cutreeStatic(sampleTrees, cutHeight = 100, minSize=10)

COVARIATES.tmp = COVARIATES[, c('batch','RACE','SEX','BrodmannArea.CDR')]
COVARIATES.tmp = sapply(COVARIATES.tmp, as.numeric)
col.mat = labels2colors(cbind(tree,COVARIATES.tmp))
col.mat = gsub('\\.1', '', col.mat)

WGCNA::plotDendroAndColors(sampleTrees, 
                           col.mat,
                           groupLabels = c('Cluster',colnames(COVARIATES.tmp)))
```

```{r temp, include=FALSE}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, results='asis', echo=FALSE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, COVARIATES,
                                                 'NULL design(voom-normalized)',
                                                 isKeyPlot=TRUE)

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
```{r iterative.adjusted.voom.normalisation, results='asis'}
postAdjustCovars = c('BrodmannArea.CDR', 'batch', 'SEX')
randomCovariates = 'individualIdentifier'

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
significantCovars = residualSigCovars$significantCovars
covariatesEffects = residualSigCovars$Effects.significantCovars
covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates))]

postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))

loopCount = 0 
while(length(significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:', 
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects and individualIdentifier as random effects'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)

  # Calculate correlation values of random effects
  correlation = parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block = COVARIATES$individualIdentifier)

  # Re-estimate voom weights
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F, 
                                       block = COVARIATES$individualIdentifier, 
                                       correlation = correlation$cor)

  # Fit a mixed effect linear model
  FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, randomCovariates))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES,
                                                'all adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  significantCovars = residualSigCovars$significantCovars
  covariatesEffects = residualSigCovars$Effects.significantCovars
  covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars, randomCovariates))]

  postAdjustCovars = c(postAdjustCovars,names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
modelStr <-paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=','), 'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Adding "BrodmannArea.CDR" back to the residuals
```{r addVarsBack}
varsToAddBackIn <-  grep("BrainRegion.Diagnosis", colnames(FIT$coefficients), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
```

### Clustering residual gene expression
```{r decompse.normalise.data3, fig.height=10, fig.width=9, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

# Tree based clustering
sampleTrees = flashClust::hclust(dist(t(RESIDUAL.GENE_EXPRESSION)), method = "average")

# Cluster expression
clust = cutreeStatic(sampleTrees, cutHeight = 100, minSize = 10)

# Plot first 2 PCs
plotdata <- data.frame(sampleIdentifier=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2],
                       PC3=PC$rotation[,3], 
                       PC4=PC$rotation[,4],
                       clust = clust)

plotdata <- left_join(plotdata, METADATA)

pl = list()
p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=factor(CDR), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
pl[[1]] <- p

p <- ggplot(plotdata, aes(x=PC2, y=PC3))
p <- p + geom_point(aes(shape=factor(CDR), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
pl[[2]] <- p

multiplot(plotlist = pl, cols = 1)

names(clust) = sampleTrees$labels
clust = clust[rownames(COVARIATES)]
```

```{r decompse.normalise.data4, fig.height=10, fig.width=25, results='asis'}
# Tree based clustering
COVARIATES.tmp = COVARIATES
COVARIATES.tmp = sapply(COVARIATES.tmp[,setdiff(FactorCovariates, 'individualIdentifier')], as.numeric)
col.mat = labels2colors(cbind(tree,COVARIATES.tmp))
col.mat = gsub('\\.1', '', col.mat)

plotDendroAndColors(sampleTrees, 
                    colors = col.mat, 
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES))
```

```{r temp1, include=FALSE}
gc()
dev.off()
```

```{r get.gene.symbols}
backgroundGenes = data.frame(gene.id = rownames(COUNT)) %>%
  dplyr::mutate(ensembl_gene_id = gene.id) %>%
  tidyr::separate(ensembl_gene_id, c('ensembl_gene_id', 'position'), sep = '\\.')

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "www.ensembl.org")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", 
                       values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

# Add gene names
RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('gene.id') %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('gene.id') %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=20}
# Fit contrast
cntr = c()
for (BA in unique(METADATA$BrodmannArea)){
  tmp = combn(paste0('BrodmannArea.CDR', BA, '.', c(0,0.5,1,2,3,4,5)),2)
  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
}

contrast = makeContrasts(contrasts=cntr,
                         levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:84, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
}, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
names(DE) = gsub('BrodmannArea.CDR','',colnames(contrast))

DE1 = plyr::ldply(DE, .id = 'Comparison') %>%
  dplyr::mutate(Comparison1 = Comparison) %>%
  tidyr::separate(Comparison1, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrodmannArea','From.CDR'), sep = 5) %>%
  tidyr::separate(to.state, c('BrodmannArea','To.CDR'), sep = 5) %>%
  tidyr::unite(Name, From.CDR, To.CDR, sep = '-')

p <- ggplot(DE1, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
p <- p + geom_vline(xintercept = log2(1.2), color = 'red')
p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
p <- p + facet_grid(BrodmannArea~.+Name)
p

tmp = DE1 %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(BrodmannArea, Name) %>%
  summarise(count = length(unique(genes))) %>%
  tidyr::spread(BrodmannArea, count)
kable(tmp)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'MSBB Reprocessed RNASeq TMM', parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'sampleIdentifier')
write.table(COVARIATES, file = 'MSBBCovariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('MSBBCovariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix %>%
  as.matrix() %>% rownameToFirstColumn('gene.id') %>%
  write.table(file = 'MSBBCounts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('MSBBCounts.tsv', name = 'Raw counts (filtered)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                     executed = thisFile, activityDescription = activityDescription)

# Store logCPM
write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'MSBBlogCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('MSBBlogCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)

# Store design matrix
write.table(DM1$design, file = 'MSBBDesign.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('MSBBDesign.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)

# Store residual gene expression
write.table(RESIDUAL.GENE_EXPRESSION, file = 'MSBBExpression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('MSBBExpression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', 
               parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store differential expression results
write.table(DE, file = 'MSBBDiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('MSBBDiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)

stopCluster(cl)
```
|  *Results*                |  *SynapseID*                    |
|  -------                  |   ---------                     |
|  Covariates               |  `r COV_OBJ$properties$id`      |
|  Raw counts (logCPM)      |  `r LCOUNT_OBJ$properties$id`   |
|  Design Matrix            |  `r DM_OBJ$properties$id`       |
|  Residual Expression      |  `r EXP_OBJ$properties$id`      |
|  Differential Expression  |  `r DEXP_OBJ$properties$id`     |

### Source R Code
[This File](`r thisFile`)

### 2017-01-23 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_SVA.Rmd
Added markdown to process reprocessed RNASeq from MSBB with TMM and SVA normalization

--- 2017-01-21 code/Rmd/MSSM_RNASeq_Reprocessed_TMM.Rmd
@@
 ---
-title: "Covariate analysis of MSSM reprocessed counts (with TMM normalisation)"
+title: "Covariate analysis of MSSM reprocessed counts (with TMM and SVA normalisation)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM_SVA.Rmd",
                                  parentId = "syn5570248",
-                                 entityName = 'MSBB Reprocessed RNASeq TMM')
+                                 entityName = 'MSBB Reprocessed RNASeq TMM SVA')
 ```
@@
 
-thisFileName <- 'MSBB_RNASeq_Reprocessed_TMM.Rmd'
+thisFileName <- 'MSSM_RNASeq_Reprocessed_TMM_SVA.Rmd'
 
@@
 ```
-### Normalisation (iterative)
-Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
+### Normalisation (custom design)
+Since many covariates are correlated, re-normalising COVARIATES with a custom design matrix (see syn8049649 for details)
 ```{r iterative.adjusted.voom.normalisation, results='asis'}
-postAdjustCovars = c('BrodmannArea.CDR', 'batch', 'SEX')
+postAdjustCovars = c('BrodmannArea.CDR', 'batch', 'SEX', 'RIN', 'rRNA.rate', 'PMI', 'AOD')
 randomCovariates = 'individualIdentifier'
 
-# Assign residual covariates
-residualSigCovars = preAdjustedSigCovars
-significantCovars = residualSigCovars$significantCovars
-covariatesEffects = residualSigCovars$Effects.significantCovars
-covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates))]
-
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
-
-loopCount = 0 
-while(length(significantCovars)!=0 && loopCount <= 100){
 writeLines(paste('Using following covariates in the model:', 
@@
 RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
+```
+`r tmp`
+### SVA 
+Estimate surrogate variables using SVA package
+```{r sva}
+MODEL0 = DM1$design[,-grep('BrodmannArea.CDR',colnames(DM1$design))] # Get null model by removing variable of interest
+MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
+MODEL1 = DM1$design
+MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
 
-  # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, randomCovariates))
+## Calculate the number of surrogate variables
+NUM.SV = sva::num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1,
+                     method = 'be', seed = 123456, B = 30)
+
+SURR.VAR = sva::sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0,
+                    n.sv = NUM.SV, B = 30)$sv
+
+colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
+rownames(SURR.VAR) = rownames(DM1$design)
+
+DM1$design = cbind(MODEL1, SURR.VAR)
+
+# Estimate voom weights
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
+
+# Calculate correlation values of random effects
+correlation = parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block = COVARIATES$Donor_ID)
+
+# Re-estimate voom weights
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F, 
+                                     block = COVARIATES$Donor_ID, correlation = correlation$cor)
+
+# Fit a mixed effect linear model
+FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
+
+# Residuals after normalisation
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
 
@@
 ```{r addVarsBack}
-varsToAddBackIn <-  grep("BrainRegion.Diagnosis", colnames(FIT$coefficients), value = T)
+varsToAddBackIn <-  grep("BrodmannArea.CDR", colnames(FIT$coefficients), value = T)
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
@@
 # Code
-CODE <- Folder(name = 'MSBB Reprocessed RNASeq TMM', parentId = parentId)
+CODE <- Folder(name = 'MSBB Reprocessed RNASeq TMM SVA', parentId = parentId)
 CODE <- synStore(CODE)

### 2017-01-23 code/Rmd/MSSM_RNASeq_Reprocessed_TMM.Rmd
Minor bug fix
Added folder annotations to the markdown code

@@
 
-thisFileName <- 'MSBB_RNASeq_Reprocessed_TMM.Rmd'
+thisFileName <- 'MSSM_RNASeq_Reprocessed_TMM.Rmd'
 
@@
   | Region | Name                           | 
-  | :----- | ------------------------- | 
+  | :----- | ------------------------------ | 
   | BM_10  | Frontal Pole (FP)              |
-  | BM_22  | Superior Temporal Gyrus   |
-  | BM_36  | Parahipocampal Gyrus (PHG)|
-  | BM_44  | inferior frontal Gyrus (IFG)|
+  | BM_22  | Superior Temporal Gyrus (STG)  |
+  | BM_36  | Parahipocampal Gyrus (PHG)     |
+  | BM_44  | inferior frontal Gyrus (IFG)   |
 
@@
 CODE <- Folder(name = 'MSBB Reprocessed RNASeq TMM', parentId = parentId)
+annotations(CODE) = list(
+  dataType = 'mRNA',
+  dataSubType = 'geneExp',
+  summaryLevel = 'gene',
+  assay	 = 'Reprocessed RNAseq',
+
+  tissueTypeAbrv	= 'FP, STG, PHG, IFG', 
+  study = 'MSSM', 
+  center = 'MSBB',
+
+  organism = 'HomoSapiens',
+  consortium	= 'AMP-AD',
+
+  normalizationStatus	= TRUE,
+  modelSystem	= FALSE,
+
+  normalizationType	= 'TMM'
+)
 CODE <- synStore(CODE)

### 2017-01-23 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_SVA.Rmd
Minor bug fix
Added folder annotations to the markdown code

@@
 library(foreach)
+library(doMC)
+library(doParallel)
 
@@
 ```
-`r tmp`
 ### SVA 
@@
 # Calculate correlation values of random effects
-correlation = parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block = COVARIATES$Donor_ID)
+correlation = parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block = COVARIATES$individualIdentifier)
 
@@
 CODE <- Folder(name = 'MSBB Reprocessed RNASeq TMM SVA', parentId = parentId)
+annotations(CODE) = list(
+  dataType = 'mRNA',
+  dataSubType = 'geneExp',
+  summaryLevel = 'gene',
+  assay	 = 'Reprocessed RNAseq',
+
+  tissueTypeAbrv	= 'FP, STG, PHG, IFG', 
+  study = 'MSSM', 
+  center = 'MSBB',
+
+  organism = 'HomoSapiens',
+  consortium	= 'AMP-AD',
+
+  normalizationStatus	= TRUE,
+  modelSystem	= FALSE,
+
+  normalizationType	= 'TMM and SVA'
+)
 CODE <- synStore(CODE)

### 2017-01-23 code/Rmd/Mayo_RNASeq_Reprocessed_TMM.Rmd
Added folder annotations to the markdown code

@@
 CODE <- Folder(name = 'UFL_Mayo_ISB Reprocessed RNASeq TMM', parentId = parentId)
+annotations(CODE) = list(
+  dataType = 'mRNA',
+  dataSubType = 'geneExp',
+  summaryLevel = 'gene',
+  assay	 = 'Reprocessed RNAseq',
+
+  tissueTypeAbrv	= 'CBE, TCX', 
+  study = 'MayoRNASeq', 
+  center = 'UFL-MAYO-ISB',
+
+  organism = 'HomoSapiens',
+  consortium	= 'AMP-AD',
+
+  normalizationStatus	= TRUE,
+  modelSystem	= FALSE,
+
+  normalizationType	= 'TMM'
+)
 CODE <- synStore(CODE)

### 2017-01-23 code/Rmd/Mayo_RNASeq_Reprocessed_TMM_SVA.Rmd
Added folder annotations to the markdown code

@@
 CODE <- Folder(name = 'UFL_Mayo_ISB Reprocessed RNASeq TMM SVA', parentId = parentId)
+annotations(CODE) = list(
+  dataType = 'mRNA',
+  dataSubType = 'geneExp',
+  summaryLevel = 'gene',
+  assay	 = 'Reprocessed RNAseq',
+
+  tissueTypeAbrv	= 'CBE, TCX', 
+  study = 'MayoRNASeq', 
+  center = 'UFL-MAYO-ISB',
+
+  organism = 'HomoSapiens',
+  consortium	= 'AMP-AD',
+
+  normalizationStatus	= TRUE,
+  modelSystem	= FALSE,
+
+  normalizationType	= 'TMM and SVA'
+)
 CODE <- synStore(CODE)

### 2017-01-23 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM.Rmd
Added folder annotations to the markdown code

@@
 CODE <- Folder(name = "ROSMAP Reprocessed RNASeq TMM (Based on cogdx)", parentId = parentId)
+annotations(CODE) = list(
+  dataType = 'mRNA',
+  dataSubType = 'geneExp',
+  summaryLevel = 'gene',
+  assay	 = 'Reprocessed RNAseq',
+
+  tissueTypeAbrv	= 'DLPFC', 
+  study = 'ROSMAP', 
+  center = 'BROAD-RUSH',
+
+  organism = 'HomoSapiens',
+  consortium	= 'AMP-AD',
+
+  normalizationStatus	= TRUE,
+  modelSystem	= FALSE,
+
+  normalizationType	= 'TMM'
+)
 CODE <- synStore(CODE)

### 2017-01-23 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM_SVA.Rmd
Added folder annotations to the markdown code

@@
 CODE <- Folder(name = "ROSMAP Reprocessed RNASeq TMM SVA (Based on cogdx)", parentId = parentId)
+annotations(CODE) = list(
+  dataType = 'mRNA',
+  dataSubType = 'geneExp',
+  summaryLevel = 'gene',
+  assay	 = 'Reprocessed RNAseq',
+
+  tissueTypeAbrv	= 'DLPFC', 
+  study = 'ROSMAP', 
+  center = 'BROAD-RUSH',
+
+  organism = 'HomoSapiens',
+  consortium	= 'AMP-AD',
+
+  normalizationStatus	= TRUE,
+  modelSystem	= FALSE,
+
+  normalizationType	= 'TMM and SVA'
+)
 CODE <- synStore(CODE)

### 2017-01-25 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd
Added markdown to process reprocessed RNASeq of MSSM samples with binned CDR using TMM normalization

--- 2017-01-23 code/Rmd/MSSM_RNASeq_Reprocessed_TMM.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd",
                                  parentId = "syn5570248",
-                                 entityName = 'MSBB Reprocessed RNASeq TMM')
+                                 entityName = 'MSBB Reprocessed RNASeq TMM Binned CDR')
 ```
@@
 library(foreach)
+library(doMC)
+library(foreach)
+library(doParallel)
 
@@
 
-thisFileName <- 'MSSM_RNASeq_Reprocessed_TMM.Rmd'
+thisFileName <- 'MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd'
 
@@
 METADATA$AOD = gsub('\\+','',METADATA$AOD)
-METADATA$BrodmannArea.CDR = paste(METADATA$BrodmannArea, METADATA$CDR, sep = '.')
+METADATA$Dx[METADATA$CDR <= 1] = 'ND'
+METADATA$Dx[METADATA$CDR == 2 | METADATA$CDR == 3] = 'MD'
+METADATA$Dx[METADATA$CDR >= 4] = 'SD'
+METADATA$BrodmannArea.Dx = paste(METADATA$BrodmannArea, METADATA$Dx, sep = '.')
 METADATA$RIN2 = METADATA$RIN^2
@@
 p = list()
-p[[1]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = RIN, fill = BrodmannArea)) + geom_boxplot()
+p[[1]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = RIN, fill = BrodmannArea)) + geom_boxplot()
 p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
@@
 # AOD
-p[[2]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = as.numeric(AOD), fill = BrodmannArea)) + geom_boxplot()
+p[[2]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(AOD), fill = BrodmannArea)) + geom_boxplot()
 p[[2]] = p[[2]] + ggtitle('Age At Death') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
@@
 # PMI
-p[[3]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = as.numeric(PMI), fill = BrodmannArea)) + geom_boxplot()
+p[[3]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(PMI), fill = BrodmannArea)) + geom_boxplot()
 p[[3]] = p[[3]] + ggtitle('PMI') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
@@
 # Total Reads
-p[[4]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = as.numeric(TotalReads), fill = BrodmannArea)) + geom_boxplot()
+p[[4]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(TotalReads), fill = BrodmannArea)) + geom_boxplot()
 p[[4]] = p[[4]] + ggtitle('Total Reads') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
@@
 # Mapped Reads
-p[[5]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = as.numeric(Mapped), fill = BrodmannArea)) + geom_boxplot()
+p[[5]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(Mapped), fill = BrodmannArea)) + geom_boxplot()
 p[[5]] = p[[5]] + ggtitle('Mapped') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
@@
 # Mapped Reads
-p[[6]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = as.numeric(rRNA.rate), fill = BrodmannArea)) + geom_boxplot()
+p[[6]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(rRNA.rate), fill = BrodmannArea)) + geom_boxplot()
 p[[6]] = p[[6]] + ggtitle('rRNA Rate') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
@@
 ```{r covariates.clustering}
-FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrodmannArea.CDR")
+FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrodmannArea.Dx")
 ContCovariates <- c("RIN", "RIN2", "AOD","PMI", "TotalReads", "Mapped", "rRNA.rate")
@@
 
-Remove genes that have less than 1 cpm counts in at least 50% of samples per BrodmannArea.CDR
+Remove genes that have less than 1 cpm counts in at least 50% of samples per BrodmannArea.Dx
 ```{r cpmnormalisation}
@@
   rownameToFirstColumn('sampleIdentifier') %>%
-  dlply(.(BrodmannArea.CDR), .fun = function(mtd, count){
+  dlply(.(BrodmannArea.Dx), .fun = function(mtd, count){
     processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$sampleIdentifier],
@@
 writeLines('Number of samples in each category are')
-tmp = summary(COVARIATES$BrodmannArea.CDR) %>%
-  rownameToFirstColumn('BrodmannArea.CDR') %>%
-  tidyr::separate(BrodmannArea.CDR, c('BrodmannArea', 'CDR'), sep = 5) %>%
-  tidyr::spread(CDR, DF)
+tmp = summary(COVARIATES$BrodmannArea.Dx) %>%
+  rownameToFirstColumn('BrodmannArea.Dx') %>%
+  tidyr::separate(BrodmannArea.Dx, c('BrodmannArea', 'Dx'), sep = 5) %>%
+  tidyr::spread(Dx, DF)
 kable(tmp)
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=factor(CDR), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
-p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:7)
+p <- p + geom_point(aes(shape=factor(Dx), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
+p <- p + theme_bw() + theme(legend.position="left") + scale_shape_manual(values = 1:7)
 # p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
@@
 p <- ggplot(plotdata, aes(x=PC2, y=PC3))
-p <- p + geom_point(aes(shape=factor(CDR), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
-p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:7)
+p <- p + geom_point(aes(shape=factor(Dx), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
+p <- p + theme_bw() + theme(legend.position="left") + scale_shape_manual(values = 1:7)
 # p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
@@
 
-COVARIATES.tmp = COVARIATES[, c('batch','RACE','SEX','BrodmannArea.CDR')]
+COVARIATES.tmp = COVARIATES[, c('batch','RACE','SEX','BrodmannArea.Dx')]
 COVARIATES.tmp = sapply(COVARIATES.tmp, as.numeric)
@@
 ```{r iterative.adjusted.voom.normalisation, results='asis'}
-postAdjustCovars = c('BrodmannArea.CDR', 'batch', 'SEX')
+postAdjustCovars = c('BrodmannArea.Dx', 'batch', 'SEX')
 randomCovariates = 'individualIdentifier'
@@
 ```
-Adding "BrodmannArea.CDR" back to the residuals
+Adding "BrodmannArea.Dx" back to the residuals
 ```{r addVarsBack}
-varsToAddBackIn <-  grep("BrainRegion.Diagnosis", colnames(FIT$coefficients), value = T)
+varsToAddBackIn <-  grep("BrodmannArea.Dx", colnames(FIT$coefficients), value = T)
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=factor(CDR), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
-p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:7)
+p <- p + geom_point(aes(shape=factor(Dx), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
+p <- p + theme_bw() + theme(legend.position="left") +scale_shape_manual(values = 1:7)
 # p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
@@
 p <- ggplot(plotdata, aes(x=PC2, y=PC3))
-p <- p + geom_point(aes(shape=factor(CDR), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
-p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:7)
+p <- p + geom_point(aes(shape=factor(Dx), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
+p <- p + theme_bw() + theme(legend.position="left") +scale_shape_manual(values = 1:7)
 # p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
@@
 # Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "www.ensembl.org")
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "dec2016.archive.ensembl.org")
 
@@
 for (BA in unique(METADATA$BrodmannArea)){
-  tmp = combn(paste0('BrodmannArea.CDR', BA, '.', c(0,0.5,1,2,3,4,5)),2)
+  tmp = combn(paste0('BrodmannArea.Dx', BA, '.', c('ND','MD','SD')),2)
   cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
@@
 # Get differnetial expression
-DE = lapply(1:84, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
-  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
+DE = lapply(1:12, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
+  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T)
 }, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
-names(DE) = gsub('BrodmannArea.CDR','',colnames(contrast))
+names(DE) = gsub('BrodmannArea.Dx','',colnames(contrast))
 
@@
   tidyr::separate(Comparison1, c('from.state','to.state'), sep = '-') %>%
-  tidyr::separate(from.state, c('BrodmannArea','From.CDR'), sep = 5) %>%
-  tidyr::separate(to.state, c('BrodmannArea','To.CDR'), sep = 5) %>%
-  tidyr::unite(Name, From.CDR, To.CDR, sep = '-')
+  tidyr::separate(from.state, c('BrodmannArea','From.Dx'), sep = 5) %>%
+  tidyr::separate(to.state, c('BrodmannArea','To.Dx'), sep = 5) %>%
+  tidyr::unite(Dx, From.Dx, To.Dx, sep = '-')
 
@@
 p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
-p <- p + facet_grid(BrodmannArea~.+Name)
+p <- p + facet_grid(BrodmannArea~.+Dx)
 p
@@
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  group_by(BrodmannArea, Name) %>%
+  group_by(BrodmannArea, Dx) %>%
   summarise(count = length(unique(genes))) %>%
@@
 # Code
-CODE <- Folder(name = 'MSBB Reprocessed RNASeq TMM', parentId = parentId)
+CODE <- Folder(name = 'MSBB Reprocessed RNASeq TMM Binned CDR', parentId = parentId)
 annotations(CODE) = list(

### 2017-01-25 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_SVA_BinnedCDR.Rmd
Added markdown to process reprocessed RNASeq of MSSM samples with binned CDR using TMM and SVA normalization

--- 2017-01-25 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM_SVA_BinnedCDR.Rmd",
                                  parentId = "syn5570248",
-                                 entityName = 'MSBB Reprocessed RNASeq TMM Binned CDR')
+                                 entityName = 'MSBB Reprocessed RNASeq TMM SVA Binned CDR')
 ```
@@
 
-thisFileName <- 'MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd'
+thisFileName <- 'MSSM_RNASeq_Reprocessed_TMM_SVA_BinnedCDR.Rmd'
 
@@
 ```
-`r tmp`
+### SVA 
+Estimate surrogate variables using SVA package
+```{r sva}
+MODEL0 = DM1$design[,-grep('BrodmannArea.Dx',colnames(DM1$design))] # Get null model by removing variable of interest
+MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
+MODEL1 = DM1$design
+MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
+
+## Calculate the number of surrogate variables
+NUM.SV = sva::num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1,
+                     method = 'be', seed = 123456, B = 30)
+
+SURR.VAR = sva::sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0,
+                    n.sv = NUM.SV, B = 30)$sv
+
+colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
+rownames(SURR.VAR) = rownames(DM1$design)
+
+DM1$design = cbind(MODEL1, SURR.VAR)
+
+# Estimate voom weights
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
+
+# Calculate correlation values of random effects
+correlation = parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block = COVARIATES$individualIdentifier)
+
+# Re-estimate voom weights
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F, 
+                                     block = COVARIATES$Donor_ID, correlation = correlation$cor)
+
+# Fit a mixed effect linear model
+FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
+
+# Residuals after normalisation
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
+
+# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
+residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+                                              COVARIATES,
+                                              'all adjusted design(voom-normalized)',
+                                              isKeyPlot=TRUE)
+```
+
 ### Sanity Check
@@
 # Code
-CODE <- Folder(name = 'MSBB Reprocessed RNASeq TMM Binned CDR', parentId = parentId)
+CODE <- Folder(name = 'MSBB Reprocessed RNASeq TMM SVA Binned CDR', parentId = parentId)
 annotations(CODE) = list(
@@
 
-  normalizationType	= 'TMM'
+  normalizationType	= 'TMM and SVA'
 )

### 2017-01-26 code/R/getGSE67835.R
Added R code to extract and process expression data from GEO

@@ -1,0 +1,89 @@
+## R code to extract and curate GSE67835 expression data from GEO
+library(CovariateAnalysis)
+library(data.table)
+library(tidyr)
+library(plyr)
+library(dplyr)
+library(stringr)
+
+library(synapseClient)
+library(knitr)
+library(githubr)
+
+library(GEOquery)
+
+synapseLogin()
+
+# Get commit from github
+thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='celltype')
+thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/R/','getGSE67835.R'))
+
+# Get raw data from geo
+geo.gse = getGEO('GSE67835')
+pheno.data = rbind(pData(geo.gse[[2]]),
+                   pData(geo.gse[[1]])) %>%
+  dplyr::select(title, geo_accession, characteristics_ch1, characteristics_ch1.1, 
+                characteristics_ch1.2, characteristics_ch1.3, characteristics_ch1.4) %>%
+  tidyr::separate(characteristics_ch1, c('tmp1', 'Tissue'), sep = '\\: ') %>%
+  tidyr::separate(characteristics_ch1.1, c('tmp2', 'CellType'), sep = '\\: ') %>%
+  tidyr::separate(characteristics_ch1.2, c('tmp3', 'Age'), sep = '\\: ') %>%
+  tidyr::separate(characteristics_ch1.3, c('tmp4', 'ChipID'), sep = '\\: ') %>%
+  tidyr::separate(characteristics_ch1.4, c('tmp5', 'SampleName'), sep = '\\: ') %>%
+  dplyr::select(-starts_with('tmp')) %>%
+  filter(CellType %in% c('oligodendrocytes', 'astrocytes', 'OPC', 'microglia', 'neurons', 'endothelial'))
+
+# Get supplementary files from GEO
+geo.files = getGEOSuppFiles('GSE67835')
+ind = grep('_RAW.tar', rownames(geo.files))
+untar(rownames(geo.files)[ind])
+all.files = dir(pattern = '*.csv.gz')
+counts = lapply(all.files, function(x){
+  dat = read.table(x, sep = '\t', header = F)
+  x = stringr::str_split(x, '\\_')[[1]][1]
+  colnames(dat) = c('Gene.Names',x)
+  return(dat)
+}) %>%
+  plyr::join_all(type = 'full') %>%
[…]

### 2017-01-30 code/Rmd/MSSM_RNASeq_Reprocessed_TMM.Rmd
Minor bug and sanity fixes

@@
 Clustering of initial normalised data (with NULL design)
-```{r decompse.normalise.data1, fig.height=14, fig.width=7, results='asis'}
+```{r decompse.normalise.data1, fig.height=10, fig.width=10, results='asis'}
 # Find principal components of expression to plot
@@
 p <- p + geom_point(aes(shape=factor(CDR), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
-p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:7)
+p <- p + theme_bw() + theme(legend.position="right") +scale_shape_manual(values = 1:7)
 # p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
@@
 ### Clustering residual gene expression
-```{r decompse.normalise.data3, fig.height=10, fig.width=9, results='asis'}
+```{r decompse.normalise.data3, fig.height=10, fig.width=10, results='asis'}
 # Find principal components of expression to plot
@@
 DE = lapply(1:84, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
-  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
+  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T)
 }, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
@@
 # Store differential expression results
-write.table(DE, file = 'MSBBDiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+write.table(DE1, file = 'MSBBDiffExpression.tsv', sep = '\t', row.names=F, quote=F)
 DEXP_OBJ = File('MSBBDiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)

### 2017-01-30 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd
Minor bug and sanity fixes

@@
   tidyr::separate(to.state, c('BrodmannArea','To.Dx'), sep = 5) %>%
-  tidyr::unite(Dx, From.Dx, To.Dx, sep = '-')
+  tidyr::unite(Dx, From.Dx, To.Dx, sep = '-') %>%
+  dplyr::select(-Comparison) %>%
+  dplyr::rename(Region = BrodmannArea, Comparison = Dx, gene.id = genes) %>%
+  dplyr::mutate(Region = factor(Region, labels = c('FP', 'STG', 'PHG', 'IFG')),
+                Study = 'MSBB') %>%
+  left_join(backgroundGenes) %>%
+  left_join(Ensemble2HGNC) %>%
+  dplyr::select(-position)
 
@@
 p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
-p <- p + facet_grid(BrodmannArea~.+Dx)
+p <- p + facet_grid(Region~.+Comparison)
 p
@@
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  group_by(BrodmannArea, Dx) %>%
+  group_by(Region, Comparison) %>%
   summarise(count = length(unique(genes))) %>%
-  tidyr::spread(BrodmannArea, count)
+  tidyr::spread(Region, Comparison)
 kable(tmp)

### 2017-01-30 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_SVA_BinnedCDR.Rmd
Minor bug and sanity fixes

@@
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  group_by(BrodmannArea, Dx) %>%
-  summarise(count = length(unique(genes))) %>%
-  tidyr::spread(BrodmannArea, count)
+  group_by(Region, Comparison) %>%
+  summarise(count = length(unique(gene.id))) %>%
+  tidyr::spread(Region, count)
 kable(tmp)

### 2017-01-30 code/Rmd/Mayo_RNASeq_Reprocessed_TMM.Rmd
Minor bug and sanity fixes

@@
 ---
-title: "Covariate analysis of MAYO reprocessed counts (with TMM and SVA normalisation)"
+title: "Covariate analysis of MAYO reprocessed counts (with TMM normalisation)"
 author: "Thanneer Perumal"

### 2017-01-30 code/Rmd/PredictPMI.Rmd
Minor bug and sanity fixes

--- 2016-06-11 code/Rmd/PredictPMI.Rmd
@@
 ---
-title: "Prediction model for PMI using ROSMAP Reprocessed RNASeq"
+title: "Prediction model for PMI using ROSMAP and MSSM Reprocessed RNASeq"
 author: "Thanneer Perumal"
@@
 knit2synapse::knitToFolderEntity(file = "./PredictPMI.Rmd",
-                                 parentId ="syn5570291",
+                                 parentId ="syn5569102",
                                  entityName = 'Predict PMI')
@@
 library(biomaRt)
+library(glmnet)
+
+library(doParallel)
+library(foreach)
+library(doMC)
+library(doParallel)
+
+cl = makeCluster(detectCores() - 2)
+registerDoParallel(cl)
 
@@
 # Synapse parameters
-parentId = 'syn5570291';
+parentId = 'syn5569102';
 activityName = 'Predict PMI';
-activityDescription = 'Predict PMI using ROSMAP data';
+activityDescription = 'Predict PMI using ROSMAP and MSSM raw counts data';
 
@@
 ```{r rosmap.download.data, cache=TRUE}
-# Download reprocessed counts (paired end)
-COUNT_PE_ID <- 'syn6112813';
-ALL_USED_IDs <- COUNT_PE_ID
-COUNT_PE <- read.table(synGet(COUNT_PE_ID)@filePath, header=T, sep='\t', check.names = F)
+# Download reprocessed counts (ROSMAP)
+COUNT_ROSMAP_ID <- 'syn8020058';
+ALL_USED_IDs <- COUNT_ROSMAP_ID
+COUNT_ROSMAP <- downloadFile(COUNT_ROSMAP_ID)
 
-# Download reprocessed counts (single end)
-COUNT_SE_ID <- 'syn6117775';
-ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_SE_ID
-COUNT_SE <- read.table(synGet(COUNT_SE_ID)@filePath, header=T, sep='\t', check.names = F)
+# Get design matrix (ROSMAP)
+DESIGN_ROSMAP_ID <- 'syn8020068'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = DESIGN_ROSMAP_ID
+DESIGN_ROSMAP <- downloadFile(DESIGN_ROSMAP_ID)
 
-COUNT = COUNT_PE + COUNT_SE[rownames(COUNT_PE), colnames(COUNT_PE)]
-rownames(COUNT) = sapply(rownames(COUNT), function(x){unlist(str_split(x, '\\.'))[1]})
+# Download reprocessed counts (MSBB)
+COUNT_MSBB_ID <- 'syn8074324';
+ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_MSBB_ID
+COUNT_MSBB <- downloadFile(COUNT_MSBB_ID)
 
-# Download filtered logCPM
-LOG.CPM_ID <- 'syn6129370';
-ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_ID
-LOG.CPM <- read.table(synGet(LOG.CPM_ID)@filePath, header=T, sep='\t', check.names = F)
+# Get design matrix (ROSMAP)
+DESIGN_MSBB_ID <- 'syn8074328'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = DESIGN_MSBB_ID
+DESIGN_MSBB <- downloadFile(DESIGN_MSBB_ID)
 
-COUNT = COUNT[intersect(rownames(COUNT), LOG.CPM$ensembl_gene_id), intersect(colnames(COUNT), colnames(LOG.CPM))]
+# Intersect genes that are common to both ROSMAP and MSBB
+ind = intersect(COUNT_ROSMAP$gene.id, COUNT_MSBB$gene.id)
 
-# Get covariates
-COVARIATES_ID <- 'syn6129368'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = COVARIATES_ID
-COVARIATES <- read.table(synGet(COVARIATES_ID)@filePath,sep='\t',header=T, row.names = 1)
+COUNT_ROSMAP = filter(COUNT_ROSMAP, gene.id %in% ind)
+rownames(COUNT_ROSMAP) = COUNT_ROSMAP$gene.id
+COUNT_ROSMAP$gene.id = NULL
 
-FactorCovariates <- c("cogdx")
-ContCovariates <- c("RINcontinuous","age_death","pmi")
+COUNT_MSBB = filter(COUNT_MSBB, gene.id %in% ind)
+rownames(COUNT_MSBB) = COUNT_MSBB$gene.id
+COUNT_MSBB$gene.id = NULL
+COUNT_MSBB = COUNT_MSBB[rownames(COUNT_ROSMAP),]
 
-COVARIATES = COVARIATES[,c(FactorCovariates,ContCovariates),drop=F]
-# Convert factor covariates to factors
-COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates,drop=F], factor)
-
-# Get surrogate variables
-SURR.VAR_ID <- 'syn6140323'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = SURR.VAR_ID
-SURR.VAR <- read.table(synGet(SURR.VAR_ID)@filePath,sep='\t',header=T)
-rownames(SURR.VAR) = rownames(COVARIATES)
-COVARIATES = cbind(COVARIATES, SURR.VAR)
-
-COUNT = COUNT[,rownames(COVARIATES)]
+rownames(DESIGN_ROSMAP) = colnames(COUNT_ROSMAP)
+rownames(DESIGN_MSBB) = colnames(COUNT_MSBB)
 ```
-Using dorsolateral prefrontal cortex of `r dim(COUNT)[2]` subjects from the ROS and MAP cohorts to predict PMI. Following sample are removed `r paste(removedIDs, collapse = ',')`
+Using `r dim(COUNT_ROSMAP)[2]` subjects from the ROSMAP cohorts and `r dim(COUNT_MSBB)[2]` subjects from MSBB cohorts to predict PMI
 ### Fit model with PMI
 ```{r w.pmi}
-writeLines(paste('Using following covariates in the model:',paste(colnames(COVARIATES), collapse=', '),'as fixed effects'))
-# Post adjusted design matrix
-DM1 = getDesignMatrix(COVARIATES,Intercept = F)
-DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
+# TMM normalisation
+TMM.GENE_EXPRESSION_ROSMAP = calcNormFactors(COUNT_ROSMAP, method = 'TMM')
+TMM.GENE_EXPRESSION_MSBB = calcNormFactors(COUNT_MSBB, method = 'TMM')
 
-# Estimate sample weights using voom methodology
-VOOM.ADJUSTED.GENE_EXPRESSION = voom(COUNT, design=DM1$design, plot=F)
+# Voom normalisation of gene expression
+VOOM.GENE_EXPRESSION_ROSMAP = voom(COUNT_ROSMAP, design = DESIGN_ROSMAP,
+                                   lib.size = TMM.GENE_EXPRESSION_ROSMAP,
+                                   normalize.method = 'quantile')
+VOOM.GENE_EXPRESSION_MSBB = voom(COUNT_MSBB, design = DESIGN_MSBB,
+                                 lib.size = TMM.GENE_EXPRESSION_MSBB,
+                                 normalize.method = 'quantile')
+
+# Fit a linear model
+FIT_ROSMAP = lmFit(VOOM.GENE_EXPRESSION_ROSMAP)
+FIT_MSBB = lmFit(VOOM.GENE_EXPRESSION_MSBB)
+
+# Find residuals 
+RESIDUAL_ROSMAP = residuals.MArrayLM(FIT_ROSMAP, VOOM.GENE_EXPRESSION_ROSMAP$E)
+RESIDUAL_MSBB = residuals.MArrayLM(FIT_MSBB, VOOM.GENE_EXPRESSION_MSBB$E)
+
+# Add variable of interest back to the residuals
+RESIDUAL_ROSMAP = RESIDUAL_ROSMAP + FIT_ROSMAP$coefficients[,'pmi'] %*% t(DESIGN_ROSMAP[,'pmi'])
+RESIDUAL_MSBB = RESIDUAL_MSBB + FIT_MSBB$coefficients[,'PMI'] %*% t(DESIGN_MSBB[,'PMI'])
 
 # Fit linear model using estimated weights and calculate BIC's
-BIC.WPMI = sapply(1:dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], function(gene){
-  y = VOOM.ADJUSTED.GENE_EXPRESSION$E[gene,]
-  w = VOOM.ADJUSTED.GENE_EXPRESSION$weights[gene,]
-  x = DM1$design
+BIC.WPMI_ROSMAP = sapply(1:dim(VOOM.GENE_EXPRESSION_ROSMAP$E)[1], function(gene){
+  y = VOOM.GENE_EXPRESSION_ROSMAP$E[gene,]
+  w = VOOM.GENE_EXPRESSION_ROSMAP$weights[gene,]
+  x = DESIGN_ROSMAP
   obj = lm(y~0+., data = cbind(x,data.frame(y=y)), weights = w)
@@
 })
-names(BIC.WPMI) = rownames(VOOM.ADJUSTED.GENE_EXPRESSION$E)
+names(BIC.WPMI_ROSMAP) = rownames(VOOM.GENE_EXPRESSION_ROSMAP$E)
+
+BIC.WPMI_MSBB = sapply(1:dim(VOOM.GENE_EXPRESSION_MSBB$E)[1], function(gene){
+  y = VOOM.GENE_EXPRESSION_MSBB$E[gene,]
+  w = VOOM.GENE_EXPRESSION_MSBB$weights[gene,]
+  x = DESIGN_MSBB
+  obj = lm(y~0+., data = cbind(x,data.frame(y=y)), weights = w)
+  BIC(obj)
+})
+names(BIC.WPMI_MSBB) = rownames(VOOM.GENE_EXPRESSION_MSBB$E)
 ```
@@
 ```{r wo.pmi}
-COVARIATES.tmp = COVARIATES[,-(4)]
-writeLines(paste('Using following covariates in the model:',paste(colnames(COVARIATES.tmp), collapse=', '),'as fixed effects'))
-
-# Post adjusted design matrix
-DM1 = getDesignMatrix(COVARIATES.tmp,Intercept = F)
-DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
-
-# Estimate sample weights using voom methodology
-VOOM.ADJUSTED.GENE_EXPRESSION = voom(COUNT, design=DM1$design, plot=F)
-
-# Fit a linear model
-VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
-
-# Calculate residual
-RESIDUAL.EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT, VOOM.ADJUSTED.GENE_EXPRESSION$E)
+# Voom normalisation of gene expression
+DESIGN_ROSMAP1 = DESIGN_ROSMAP[,setdiff(colnames(DESIGN_ROSMAP), 'pmi')]
+VOOM.GENE_EXPRESSION_ROSMAP = voom(COUNT_ROSMAP, design = DESIGN_ROSMAP1, 
+                                   lib.size = TMM.GENE_EXPRESSION_ROSMAP,
+                                   normalize.method = 'quantile')
+DESIGN_MSBB1 = DESIGN_MSBB[,setdiff(colnames(DESIGN_MSBB), 'PMI')]
+VOOM.GENE_EXPRESSION_MSBB = voom(COUNT_MSBB, design = DESIGN_MSBB1, 
+                                   lib.size = TMM.GENE_EXPRESSION_MSBB,
+                                   normalize.method = 'quantile')
 
 # Fit linear model using estimated weights and calculate BIC's
-BIC.WOPMI = sapply(1:dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], function(gene){
-  y = VOOM.ADJUSTED.GENE_EXPRESSION$E[gene,]
-  w = VOOM.ADJUSTED.GENE_EXPRESSION$weights[gene,]
-  x = DM1$design
+BIC.WOPMI_ROSMAP = sapply(1:dim(VOOM.GENE_EXPRESSION_ROSMAP$E)[1], function(gene){
+  y = VOOM.GENE_EXPRESSION_ROSMAP$E[gene,]
+  w = VOOM.GENE_EXPRESSION_ROSMAP$weights[gene,]
+  x = DESIGN_ROSMAP1
   obj = lm(y~0+., data = cbind(x,data.frame(y=y)), weights = w)
@@
 })
-names(BIC.WOPMI) = rownames(VOOM.ADJUSTED.GENE_EXPRESSION$E)
+names(BIC.WOPMI_ROSMAP) = rownames(VOOM.GENE_EXPRESSION_ROSMAP$E)
+
+BIC.WOPMI_MSBB = sapply(1:dim(VOOM.GENE_EXPRESSION_MSBB$E)[1], function(gene){
+  y = VOOM.GENE_EXPRESSION_MSBB$E[gene,]
+  w = VOOM.GENE_EXPRESSION_MSBB$weights[gene,]
+  x = DESIGN_MSBB1
+  obj = lm(y~0+., data = cbind(x,data.frame(y=y)), weights = w)
+  BIC(obj)
+})
+names(BIC.WOPMI_MSBB) = rownames(VOOM.GENE_EXPRESSION_MSBB$E)
 ```
 ### Predict PMI
-```{rpredict.pmi}
+```{r predict.pmi, fig.height=6, fig.width=6}
 # Extract features that show improvement in the model with pmi when compared to without pmi
-genes = rownames(VOOM.ADJUSTED.GENE_EXPRESSION$E)[which((BIC.WOPMI-BIC.WPMI) >= 6)]
+genes_ROSMAP = names(BIC.WOPMI_ROSMAP)[which((BIC.WOPMI_ROSMAP-BIC.WPMI_ROSMAP) >= 6)]
+genes_MSBB = names(BIC.WOPMI_MSBB)[which((BIC.WOPMI_MSBB-BIC.WPMI_MSBB) >= 6)]
+writeLines(paste(length(genes_ROSMAP),'genes from ROSMAP and',
+                 length(genes_MSBB),'genes from MSBB are used as features to predict PMI of which', 
+                 length(intersect(genes_MSBB, genes_ROSMAP)), 'are common between ROSMAP and MSBB'))
 
+#### ROSMAP ####
 # Extract x and y
-x = t(RESIDUAL.EXPRESSION[genes,])
-y = COVARIATES$pmi
+x = t(RESIDUAL_ROSMAP[genes_ROSMAP,])
+y = DESIGN_ROSMAP$pmi
 
@@
 coeff2 = coeff2[which(rowSums(coeff2[,-(1)] == 0) < 80),]
+
+REG_ROSMAP = coeff1
+RF_ROSMAP = coeff2
+#### ####
+
+#### MSBB ####
+# Extract x and y
+x = t(RESIDUAL_MSBB[genes_MSBB,])
+y = DESIGN_MSBB$PMI
+
+# Remove samples with 0 pmi
+ind = y == 0
+x = x[!ind,]
+y = log(y[!ind])
+
+# Linear regression model
+glm.model = foreach(i = 1:100,
+                    .export = c('x','y'),
+                    .packages = c('glmnet', 'CovariateAnalysis', 'dplyr', 'plyr')) %dopar% {
+  tryCatch({
+    ind = sample(1:length(y), round(length(y)*0.8))
+    ind.comp = setdiff(1:length(y), ind)
+
+    obj = cv.glmnet(x[ind,],y[ind], family = 'gaussian')
+    coeff = as.matrix(coef.glmnet(obj, s = 'lambda.min'))
+    ind.var1 = rownames(coeff)
+    ind.var2 = setdiff(rownames(coeff)[coeff != 0], '(Intercept)')
+
+    # Prune model based on least variable of importance
+    while(length(ind.var1) > length(ind.var2) & length(ind.var2) > 2){
+      obj1 = cv.glmnet(x[ind,ind.var2],y[ind], family = 'gaussian')
+      coeff = as.matrix(coef.glmnet(obj1, s = 'lambda.min'))
+      ind.var1 = ind.var2
+      ind.var2 = setdiff(rownames(coeff)[coeff != 0], '(Intercept)')
+    }
+
+    mse = sum((predict(obj1, x[ind.comp,ind.var2, drop = F], s = 'lambda.min', type = 'response') - y[ind.comp])^2)/length(ind.comp)
+    coeff = as.matrix(coef.glmnet(obj1, s = 'lambda.min')) %>%
+      rownameToFirstColumn('Coefficients') %>%
+      plyr::rename(c('1' = i))
+    list(mse = mse, coeff = coeff)
+  }, error = function(e){
+     list(mse = NA, coeff = NA) 
+  })
+}
+boxplot(sapply(glm.model, function(x){x$mse}), main = 'glmnet', ylab = 'mse')
+coeff1 = lapply(glm.model, function(x){x$coeff}) %>%
+  join_all(type = 'full')
+coeff1[is.na(coeff1)] = 0
+coeff1 = coeff1[which(rowSums(coeff1[,-(1)] == 0) < 80),]
+
+# Random forest model
+rf.model = foreach(i = 1:100,
+                .export = c('x','y'),
+                .packages = c('ranger', 'CovariateAnalysis', 'dplyr', 'plyr')) %dopar% {
+  ind = sample(1:length(y), round(length(y)*0.8))
+  ind.comp = setdiff(1:length(y), ind)
+
+  obj = ranger(formula = y ~., data = cbind(data.frame(y = y[ind]), x[ind,]), 
+               importance = 'permutation',
+               write.forest = TRUE)
+  coeff = obj$variable.importance
+  ind.var1 = names(coeff)
+  ind.var2 = names(coeff)[coeff >= abs(min(coeff))]
+
+  # Prune model based on least variable of importance
+  while(length(ind.var1) > length(ind.var2)){
+    obj1 = ranger(formula = y ~., data = cbind(data.frame(y = y[ind]), x[ind,ind.var2]),
+                  importance = 'permutation',
+                  write.forest = TRUE)
+    coeff = obj1$variable.importance
+    ind.var1 = ind.var2
+    ind.var2 = names(coeff)[coeff >= abs(min(coeff))]
+  }
+
+  mse = obj1$prediction.error
+  coeff = obj1$variable.importance %>%
+    rownameToFirstColumn('Coefficients') %>%
+    plyr::rename(c('DF' = i))
+  list(mse = mse, coeff = coeff)
+}
+boxplot(sapply(rf.model, function(x){x$mse}), main = 'random forest', ylab = 'mse')
+coeff2 = lapply(rf.model, function(x){x$coeff}) %>%
+  join_all(type = 'full')
+coeff2[is.na(coeff2)] = 0
+coeff2 = coeff2[which(rowSums(coeff2[,-(1)] == 0) < 80),]
+
+REG_MSBB = coeff1
+RF_MSBB = coeff2
+#### ####
 ```
@@
 # Create folder to store results in synapse
-CODE = Folder(name = 'Predict PMI', parentId ="syn5570291")
+CODE = Folder(name = 'Predict PMI', parentId = parentId)
 CODE = synStore(CODE)
 
-# Store features in synapse (union of both the models)
-tmp = setdiff(unique(c(coeff1$Coefficients, coeff2$coefficients)), '(Intercept)')
+# Store features in synapse (union of all models)
+tmp = setdiff(unique(c(REG_ROSMAP$Coefficients, REG_MSBB$Coefficients, RF_ROSMAP$Coefficients, RF_MSBB$Coefficients)), '(Intercept)')
 write.table(tmp, file = 'features.tsv', sep = '\n', row.names = F, quote=F)
@@
                activityName = activityName, activityDescription = activityDescription)
+
+# Store coefficients in synapse
+save(REG_ROSMAP, REG_MSBB, RF_ROSMAP, RF_MSBB, file = 'coeffPMI.RData')
+obj = File('coeffPMI.RData', name = 'Coefficients of all models', parentId = CODE$properties$id)
+obj = synStore(obj, used = as.character(ALL_USED_IDs), executed = thisFile, 
+               activityName = activityName, activityDescription = activityDescription)
+stopCluster(cl)
 ```

### 2017-01-30 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM.Rmd
Minor bug and sanity fixes

@@
 p <- p + geom_point(aes(color=Batch, shape=cogdx, size=RINcontinuous))
-p <- p + theme_bw() + theme(legend.position="top")
+p <- p + theme_bw() + theme(legend.position="right")
 # p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
@@
 p <- p + geom_point(aes(color=Batch, shape=cogdx, size=RINcontinuous))
-p <- p + theme_bw() + theme(legend.position="top")
+p <- p + theme_bw() + theme(legend.position="right")
 # p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)
@@
 # Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "mar2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "dec2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
 
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=16, fig.width=16}
+```{r diffExp, fig.height=6, fig.width=18}
 # Fit contrast
@@
 DE = lapply(1:3, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
-  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
+  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T) %>%
     rownameToFirstColumn('gene.id') %>%
@@
     left_join(Ensemble2HGNC) %>%
-    dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, logFC, AveExpr, t, P.Value, adj.P.Val, B)
+    dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, logFC, CI.L, CI.R, AveExpr, t, P.Value, adj.P.Val, B)
 }, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC) 
@@
   rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Direction = logFC/abs(logFC),
+  dplyr::mutate(Study = 'ROSMAP', Region = 'DLPFC',
+                Direction = logFC/abs(logFC),
                 Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')))

### 2017-01-30 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM_SVA.Rmd
Minor bug and sanity fixes

@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=16, fig.width=16}
+```{r diffExp, fig.height=8, fig.width=16}
 # Fit contrast
@@
   rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Direction = logFC/abs(logFC),
+  dplyr::mutate(Study = 'ROSMAP',
+                Region = 'DLPFC',
+                Direction = logFC/abs(logFC),
                 Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')))

### 2017-01-30 code/Rmd/Mayo_RNASeq_Reprocessed_TMM.Rmd
Major sanity modifications in MAYO reprocessed RNASeq markdowns

@@
 
-ind = sapply(rownames(VOOM.GENE_EXPRESSION$E), function(x){str_split(x, '\\.')[[1]][1]}) %in% CONTROL.PROBES$x 
+ind = rownames(VOOM.GENE_EXPRESSION$E) %in% CONTROL.PROBES$x 
 EXPR = VOOM.GENE_EXPRESSION$E[ind,]
@@
 Clustering of initial normalised data (with NULL design)
-```{r decompse.normalise.data1, fig.height=14, fig.width=7, results='asis'}
+```{r decompse.normalise.data1, fig.height=14, fig.width=9, results='asis'}
 # Find principal components of expression to plot
@@
 p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Tissue), size=as.numeric(AgeAtDeath)))
-p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
+p <- p + theme_bw() + theme(legend.position="right") +scale_shape_manual(values = 1:2)
 # p <- p + geom_text(aes(label= ID), size=4, hjust=0)
@@
 p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Diagnosis), size=as.numeric(AgeAtDeath)))
-p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
+p <- p + theme_bw() + theme(legend.position="right") +scale_shape_manual(values = 1:2)
 # p <- p + geom_text(aes(label= ID), size=4, hjust=0)
@@
 DE = lapply(1:12, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
-  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
+  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T)
 }, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
@@
 
-DE = plyr::ldply(DE, .id = 'Comparison') %>%
+DE1 = plyr::ldply(DE, .id = 'Comparison') %>%
   dplyr::mutate(Comparison1 = Comparison) %>%
@@
   tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
-  tidyr::unite(Name, reference, against, sep = '-')
+  tidyr::unite(Name, reference, against, sep = '-') %>%
+  dplyr::mutate(Comparison = Name, Study = 'MAYO') %>%
+  dplyr::rename(gene.id = genes, Region = BrainRegion) %>%
+  dplyr::select(-Name) %>%
+  left_join(backgroundGenes) %>%
+  left_join(Ensemble2HGNC)
 
-p <- ggplot(DE, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
+p <- ggplot(DE1, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
 p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
@@
 p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
-p <- p + facet_grid(BrainRegion~.+Name)
+p <- p + facet_grid(Region~.+Comparison)
 p
 
-tmp = DE %>% 
+tmp = DE1 %>% 
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  group_by(Comparison) %>%
-  dplyr::summarize(nDiffExpGenes = length(unique(genes)))
+  group_by(Region, Comparison) %>%
+  dplyr::summarize(nDiffExpGenes = length(unique(gene.id))) %>%
+  spread(Region, nDiffExpGenes)
 kable(tmp)
@@
 # Store design matrix
-write.table(DM1$design, file = 'MAYO_CBE_TCX_Design.tsv', sep = '\t', row.names=F, quote=F)
+write.table(DM1$design, file = 'MAYO_CBE_TCX_Design.tsv', sep = '\t', row.names=T, quote=F)
 DM_OBJ = File('MAYO_CBE_TCX_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
@@
 # Store differential expression results
-write.table(DE, file = 'MAYO_CBE_TCX_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+write.table(DE1, file = 'MAYO_CBE_TCX_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
 DEXP_OBJ = File('MAYO_CBE_TCX_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)

### 2017-02-02 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd
Added markdown to deconvolute cell typr fraction in reprocessed RNASeq from ROSMAP

--- 2017-01-13 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM_SVA.Rmd
@@
 ---
-title: "Covariate analysis of ROSMAP reprocessed counts (with TMM and SVA normalisation)"
+title: "Markdown to estimate the abundance of brain cell types with the reprocessed counts from ROSMAP cohort (with TMM-VOOM normalisation)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_TMM_SVA.Rmd",
+knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd",
                                  parentId = "syn5570291",
-                                 entityName = 'ROSMAP Reprocessed RNASeq TMM SVA (Based on cogdx)')
+                                 entityName = 'ROSMAP Cell Type Abundance with Reprocessed RNASeq')
 ```
@@
 library(RColorBrewer)
+library(xlsx)
 
@@
 
+library(parallel)
+library(doParallel)
+library(doMC)
+library(foreach)
+
+cl = makeCluster(detectCores() - 2)
+registerDoParallel(cl)
+
 synapseLogin()
+
+# Source external files
+source('../R/CIBERSORT.R')
 
@@
 parentId = 'syn5570291';
-activityName = 'Covariate analysis';
-activityDescription = 'Covariate analysis of ROSMAP reprocessed RNASeq data';
+activityName = 'Cell type abundance in ROSMAP';
+activityDescription = 'Cell type abundance in ROSMAP with TMM-VOOM normalised reprocessed RNASeq counts';
 
-thisFileName <- 'ROSMAP_RNASeq_Reprocessed_TMM_SVA.Rmd'
+thisFileName <- 'ROSMAP_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd'
 
 # Github link
-thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
+thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='celltype')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
@@
 
-### Covariate clustering
-Determine relationship between covariates. 
+### Get Covariates
 ```{r covariates.clustering, cache=TRUE}
@@
 ```
-Correlation/association between covariates at an FDR <= 0.1
-```{r covariates.correlation, fig.width=10, fig.height=10}
-COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
-ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.5, row.width=0.1)
-```
 
-```{r dataExplore, fig.width = 10, fig.height = 10}
-# RIN
-p = list()
-p[[1]] = ggplot(COVARIATES, aes(x = cogdx, y = RINcontinuous)) + geom_boxplot()
-p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')
-
-# AgeAtDeath
-p[[2]] = ggplot(COVARIATES, aes(x = cogdx, y = age_death)) + geom_boxplot()
-p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + theme(legend.position = 'top')
-
-# PMI
-p[[3]] = ggplot(COVARIATES, aes(x = cogdx, y = pmi)) + geom_boxplot()
-p[[3]] = p[[3]] + ggtitle('PMI') + theme(legend.position = 'top')
-
-# Education
-p[[4]] = ggplot(COVARIATES, aes(x = cogdx, y = educ)) + geom_boxplot()
-p[[4]] = p[[4]] + ggtitle('Education') + theme(legend.position = 'top')
-
-multiplot(plotlist = p, cols = 2)
-
-# Batch
-vcd::mosaic(~ Batch + cogdx, data = COVARIATES)
-
-# Sex
-vcd::mosaic(~ msex + cogdx, data = COVARIATES)
-```
 ### CPM Normalisation
@@
 ```
-Coexpression of genes 
-```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(VOOM.GENE_EXPRESSION$E))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
-### Outlier analysis
-Clustering of voom adjusted data (with NULL design)
-```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
-# Find principal components of expression to plot
-PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)
 
-# Plot first 2 PCs
-plotdata <- data.frame(SampleID=rownames(PC$rotation), 
-                       PC1=PC$rotation[,1], 
-                       PC2=PC$rotation[,2])
+### Get cell type specific signatures and expression data from synapse
+```{r get.signatures}
+ALL_USED_IDs = c(ALL_USED_IDs, 'syn8068511', 'syn8068475')
+# Signatures from Zhang,Y et al., 2014
+sig.zhang = downloadFile('syn8068511') %>%
+  dlply(.(Set), .fun = function(x){return(unique(x$Gene))})
 
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
+# Signatures from Darmanis, S et al., 2015
+sig.dar.obj = synGet('syn8068475')  
+sig.dar.unbiased = read.xlsx(sig.dar.obj@filePath, sheetName = 'Unbiased') %>%
+  lapply(function(x){as.character(unique(x))})
+sig.dar.unbiased = sig.dar.unbiased[1:10]
 
-p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Batch, shape=cogdx, size=RINcontinuous))
-p <- p + theme_bw() + theme(legend.position="top")
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
-p
-```
-Tree based classification of samples
-```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
-# Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Batch', 'msex', 'cogdx')])
-COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
+# Extract all signatures
+sig = list()
+sig$OPC = unique(c(sig.zhang$OPC, sig.dar.unbiased$Cluster.1))
+sig$Oligodendrocyte = unique(c(sig.zhang$MyelinOligos, sig.zhang$NewOligos, sig.dar.unbiased$Cluster.3))
+sig$Neuron = unique(c(sig.zhang$Neuron, sig.dar.unbiased$Cluster.6, sig.dar.unbiased$Cluster.7,
+                      sig.dar.unbiased$Cluster.10, sig.dar.unbiased$Cluster.2))
+sig$Endothelial = unique(c(sig.zhang$Endothelial, sig.dar.unbiased$Cluster.8))
+sig$Astrocyte = unique(c(sig.zhang$Astrocyte, sig.dar.unbiased$Cluster.4))
+sig$Microglia = unique(c(sig.zhang$Astrocyte, sig.dar.unbiased$Cluster.5))
 
-tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
-cols = WGCNA::labels2colors(COVARIATES.tmp);
+sig = ldply(sig, .fun = function(x){as.data.frame(x)})
+colnames(sig) = c('CellType','Gene.Name')
 
-WGCNA::plotDendroAndColors(tree, 
-                           colors = cols, 
-                           dendroLabels = FALSE, 
-                           abHeight = 0.80, 
-                           main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp[,c('Batch', 'msex', 'cogdx')]))
-```
+# Get reference expression data from Darmanis., S et al., 2015
+ALL_USED_IDs = c(ALL_USED_IDs, 'syn8077142', 'syn8077191')
+ref.expr = downloadFile('syn8077142')
+rownames(ref.expr) = ref.expr$hgnc_symbol
+ref.expr$hgnc_symbol = NULL
 
-```{r temp, include = F}
-dev.off()
-gc()
-```
+ref.cellTypes = downloadFile('syn8077191') %>%
+  dplyr::select(SampleID, CellType)
 
-### Significant Covariates
-Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
-```{r preAdjusted.covariates, cache=TRUE}
-# Find correlation between PC's of gene expression with covariates
-preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
-                                                 COVARIATES,
-                                                 'NULL design(voom-normalized)', 
-                                                 isKeyPlot=TRUE, 
-                                                 MIN_PVE_PCT_PC = 1)
-```
-
-Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
-```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
-preAdjustedSigCovars[["PC_res"]][[2]]$plotData
-```
-### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix.
-Here "cogdx" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on cognitive scores)
-
-Adding Batch and msex a priori to variable selection
-```{r customNorm, results='asis'}
-# Primary variable of interest
-postAdjustCovars = c('cogdx', 'Batch', 'msex');
-
-# Assign residual covariates
-residualSigCovars = preAdjustedSigCovars
-covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
-
-loopCount = 0 
-while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
-  writeLines(paste('Using following covariates in the model:',
-                   paste(postAdjustCovars, collapse=', '),
-                   'as fixed effects'))
-
-  # Post adjusted design matrix
-  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
-
-  # Estimate voom weights
-  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
-
-  # Fit linear model using new weights and new design
-  VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
-
-  # Residuals after normalisation
-  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
-
-  # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars))
-
-  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                                COVARIATES[, residCovars, drop=F], 
-                                                'adjusted design(voom-normalized)',
-                                                isKeyPlot=TRUE)
-
-  # Add postadjusted covariates (if any)
-  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
-                                                                          c(postAdjustCovars))]
-
-  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
-  loopCount = loopCount + 1
+ref.expr1 = list()
+for(celltype in unique(ref.cellTypes$CellType)){
+  ids = ref.cellTypes$SampleID[ref.cellTypes$CellType == celltype]
+  ref.expr1[[celltype]] = log2(rowSums(2^ref.expr[,ids], na.rm = T)) %>%
+    rownameToFirstColumn('hgnc_symbol') %>%
+    plyr::rename(c('DF' = celltype))
 }
-
-modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
-                  'as fixed effects')
-
-tmp <- paste('Using following covariates in the final model:', modelStr)
-```
-
-### Sanity check
-```{r residualSigCovars.manual, fig.width=12, fig.height=8}
-residualSigCovars[["PC_res"]][[2]]$plotData
-```
-### SVA 
-Estimate surrogate variables using SVA package
-```{r sva}
-MODEL0 = DM1$design[,-grep('cogdx',colnames(DM1$design))] # Get null model by removing variable of interest
-MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
-MODEL1 = DM1$design
-MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
-
-## Calculate the number of surrogate variables
-NUM.SV = sva::num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1,
-                     method = 'be', seed = 123456, B = 30)
-
-SURR.VAR = sva::sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0,
-                    n.sv = NUM.SV, B = 30)$sv
-
-colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
-rownames(SURR.VAR) = rownames(DM1$design)
-
-DM1$design = cbind(MODEL1, SURR.VAR)
-
-# Estimate voom weights
-VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
-
-# Fit linear model using new weights and new design
-VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
-
-# Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
-```
-
-### Residual calculation
-Calculate weighted residuals and add back "cogdx" to the residuals
-```{r varsToAddBack}
-# Add variable of interest back to the residuals
-varsToAddIn = grep("cogdx", colnames(DM1$design), value = T)
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
-  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
-```
-Coexpression of genes 
-```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.GENE_EXPRESSION))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
-
-### Clustering residual data
-```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
-# Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
-
-# Plot first 4 PCs
-plotdata <- data.frame(SampleID=rownames(PC$rotation), 
-                       PC1=PC$rotation[,1], 
-                       PC2=PC$rotation[,2])
-
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
-
-p <- ggplot(plotdata, aes(x=PC1, y=PC2)) 
-p <- p + geom_point(aes(color=Batch, shape=cogdx, size=RINcontinuous))
-p <- p + theme_bw() + theme(legend.position="top")
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)
-p
-```
-
-```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
-# Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Batch', 'msex', 'cogdx')])
-COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
-
-tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
-cols = WGCNA::labels2colors(COVARIATES.tmp);
-
-WGCNA::plotDendroAndColors(tree, 
-                           colors = cols, 
-                           dendroLabels = FALSE, 
-                           abHeight = 0.80, 
-                           main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp))
-```
-
-```{r temp1, include=F}
-dev.off()
-gc()
+ref.expr1 = join_all(ref.expr1)
 ```
@@
 ```{r get.gene.symbols, cache=TRUE, eval=T}
-backgroundGenes = data.frame(gene.id = rownames(COUNT)) %>%
+backgroundGenes = data.frame(gene.id = rownames(VOOM.GENE_EXPRESSION$E)) %>%
   dplyr::mutate(ensembl_gene_id = gene.id) %>%
@@
 # Add gene names
-RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
+EXPR <- VOOM.GENE_EXPRESSION$E %>%
   rownameToFirstColumn('gene.id') %>%
@@
   left_join(Ensemble2HGNC) %>%
-  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+  filter(!duplicated(gene.id))
+rownames(EXPR) = EXPR$gene.id
 
-VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
-  rownameToFirstColumn('gene.id') %>%
-  left_join(backgroundGenes) %>%
-  left_join(Ensemble2HGNC) %>%
-  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+EXPR1 = WGCNA::collapseRows(EXPR %>% dplyr::select(-ensembl_gene_id, -position, -hgnc_symbol, -gene.id), 
+                            EXPR$hgnc_symbol, 
+                            EXPR$gene.id)$datETcollapsed
+
+ind = intersect(rownames(EXPR1), ref.expr1$hgnc_symbol)
+EXPR1 = EXPR1 %>% 
+  rownameToFirstColumn('hgnc_symbol') %>%
+  filter(hgnc_symbol %in% ind)
+ref.expr1 = ref.expr1 %>%
+  filter(hgnc_symbol %in% ind)
+
+write.table(ref.expr1, file = 'ref.expr.txt', sep = '\t', quote=F, row.names = F)
+write.table(EXPR1[,1:20], file = 'mixture.expr.txt', sep = '\t', quote=F, row.names = F)
+
+cellprop = CIBERSORT('ref.expr.txt', 'mixture.expr.txt', perm = 100) # Obtain CIBERSORT.R function from http://cibersort.stanford.edu/
+stopCluster(cl)
 ```
 
-Interpretation of cogdx scores
+```{r plot.cellprop}
+cellprop1 = as.data.frame(cellprop)
+cellprop1$cogdx = COVARIATES[rownames(cellprop), 'cogdx']
+cellprop1 = cellprop1 %>% 
+  rownameToFirstColumn('SampleID') %>%
+  dplyr::select(-`P-value`,-Correlation, -RMSE) %>%
+  tidyr::gather(Celltype, Fraction, -SampleID, -cogdx) %>%
+  dplyr::arrange(desc(cogdx))
 
-1. NCI, No cognitive impairment (No impaired domains)
-2. MCI, Mild cognitive impairment (One impaired domain) and NO other cause of CI
-3. MCI, Mild cognitive impairment (One impaired domain) AND another cause of CI
-4. AD, Alzheimer's disease and NO other cause of CI (NINCDS PROB AD)
-5. AD, Alzheimer's disease AND another cause of CI (NINCDS POSS AD)
-6. Other dementia. Other primary cause of dementia
+p = ggplot(cellprop1, aes(x = SampleID, y = Fraction, fill = Celltype, color = cogdx)) + geom_bar(stat = 'identity')
+p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
+# p = p + facet_grid(.~cogdx)
+p
 
-### Differential expression analysis
-Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=16, fig.width=16}
-# Fit contrast
-contrast = makeContrasts(contrasts=c("cogdx1-cogdx2",
-                                     "cogdx1-cogdx4",
-                                     "cogdx2-cogdx4"),
-                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
-FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
-FIT.CONTR = eBayes(FIT.CONTR)
-
-# Get differnetial expression
-DE = lapply(1:3, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
-  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
-    rownameToFirstColumn('gene.id') %>%
-    left_join(backgroundGenes) %>%
-    left_join(Ensemble2HGNC) %>%
-    dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, logFC, AveExpr, t, P.Value, adj.P.Val, B)
-}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC) 
-names(DE) = colnames(contrast)
-
-DE = DE %>% 
-  rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Direction = logFC/abs(logFC),
-                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')))
-
-tmp = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
-  group_by(Comparison, Direction) %>%
-  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id))) %>%
-  spread(Direction, FDR_0_05)
-
-tmp1 = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
-  group_by(Comparison, Direction) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
-  spread(Direction, FDR_0_05_FC_1.2)
-
-kable(full_join(tmp %>%
-                  dplyr::rename(DOWN_FDR_0_05 = DOWN, UP_FDR_0_05 = UP),
-                tmp1 %>%
-                  dplyr::rename(DOWN_FDR_0_05_FC_1.2 = DOWN, UP_FDR_0_05_FC_1.2 = UP)))
-
-p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
-p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
-p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
-p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
-p = p + facet_grid(.~Comparison, scales = 'fixed')
+p = ggplot(cellprop1, aes(x = Celltype, y = Fraction, color = cogdx)) + geom_boxplot()
+p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
 p
 ```
+### Synapse store
+```{r syn.store, cache=TRUE}
+CODE = Folder(name = 'ROSMAP Cell Type Abundance with Reprocessed RNASeq', parentId = parentId)
+CODE = synStore(CODE)
 
-### Store files in synapse
-```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
-# Code
-CODE <- Folder(name = "ROSMAP Reprocessed RNASeq TMM SVA (Based on cogdx)", parentId = parentId)
-CODE <- synStore(CODE)
-
-# Store covariates
-COVARIATES = rownameToFirstColumn(COVARIATES, 'Sampleid')
-write.table(COVARIATES, file = 'ROSMAP_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
-COV_OBJ = File('ROSMAP_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
-COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                   executed = thisFile, activityDescription = activityDescription)
-
-# Store filtered counts
-PROCESSED_COUNTS$filteredExprMatrix %>%
-  as.matrix() %>% rownameToFirstColumn('gene.id') %>%
-  write.table(file = 'ROSMAP_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('ROSMAP_DLPFC_Counts.tsv', name = 'Raw counts (filtered)', parentId = CODE$properties$id)
-COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                     executed = thisFile, activityDescription = activityDescription)
-
-# Store logCPM
-write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'ROSMAP_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('ROSMAP_DLPFC_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
-LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                      executed = thisFile, activityDescription = activityDescription)
-
-# Store design matrix
-write.table(DM1$design, file = 'ROSMAP_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('ROSMAP_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
-DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                  executed = thisFile, activityDescription = activityDescription)
-
-# Store residual gene expression
-write.table(RESIDUAL.GENE_EXPRESSION, file = 'ROSMAP_DLPFC_Expression.tsv', sep = '\t', row.names=F, quote=F)
-EXP_OBJ = File('ROSMAP_DLPFC_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', parentId = CODE$properties$id)
-EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                   executed = thisFile, activityDescription = activityDescription)
-
-# Store differential expression results
-write.table(DE, file = 'ROSMAP_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('ROSMAP_DLPFC_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
-DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                    executed = thisFile, activityDescription = activityDescription)
+cellprop %>%
+  rownameToFirstColumn('SampleID') %>%
+  write.table(file = 'ROSMAP_CellFraction_Reproc.tsv', row.names = F, sep = '\t', quote = F)
+obj = File('ROSMAP_CellFraction_Reproc.tsv', name = 'ROSMAP Cell Fraction', parentId = CODE$properties$id)
+obj = synStore(obj, activityName = activityName, activityDescription = activityDescription,
+               used = ALL_USED_IDs, executed = thisFile)
 ```
-|  *Results*                |  *SynapseID*                     |
-|  -------                  |   ---------                      |
-|  Covariates               |  `r COV_OBJ$properties$id`       |
-|  Raw counts (log CPM)     |  `r COUNT_OBJ$properties$id`     |
-|  Design Matrix            |  `r DM_OBJ$properties$id`        |
-|  Residual Expression      |  `r EXP_OBJ$properties$id`       |
-|  Differential Expression  |  `r DEXP_OBJ$properties$id`      |

### 2017-02-03 code/Rmd/Mayo_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd
Added markdown to deconvolute celltype fraction from Mayo data

--- 2017-02-02 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd
@@
 ---
-title: "Markdown to estimate the abundance of brain cell types with the reprocessed counts from ROSMAP cohort (with TMM-VOOM normalisation)"
+title: "Markdown to estimate the abundance of brain cell types with the reprocessed counts from Mayo cohorts (with TMM-VOOM normalisation)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd",
-                                 parentId = "syn5570291",
-                                 entityName = 'ROSMAP Cell Type Abundance with Reprocessed RNASeq')
+knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd",
+                                 parentId = "syn5570325",
+                                 entityName = 'MAYO Cell Type Abundance with Reprocessed RNASeq')
 ```
@@
 # Synapse parameters
-parentId = 'syn5570291';
-activityName = 'Cell type abundance in ROSMAP';
-activityDescription = 'Cell type abundance in ROSMAP with TMM-VOOM normalised reprocessed RNASeq counts';
+parentId = 'syn5570325';
+activityName = 'Cell type abundance in Mayo';
+activityDescription = 'Cell type abundance in Mayo with TMM-VOOM normalised reprocessed RNASeq counts';
 
-thisFileName <- 'ROSMAP_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd'
+thisFileName <- 'Mayo_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd'
 
@@
 ```{r download.data, cache=TRUE}
-# Download expression data
-COUNT_ID <- 'syn7750270';
-ALL_USED_IDs <- COUNT_ID
-COUNT_OBJ <- synGet(COUNT_ID)
-COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F)
+# Get reprocessed sample counts for temporal cortex
+COUNT_TC_ID <- 'syn7987765';
+ALL_USED_IDs <- COUNT_TC_ID
+COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F)
+colnames(COUNT_TC) = paste(colnames(COUNT_TC), 'TCX', sep = '_')
+
+# Get reprocessed sample counts for cerebellum
+COUNT_CE_ID <- 'syn7513807';
+ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_CE_ID
+COUNT_CE <- read.table(synGet(COUNT_CE_ID)@filePath, header=T, sep='\t', check.names = F)
+
+COUNT = cbind(COUNT_TC, COUNT_CE[rownames(COUNT_TC),])
 
 # Get clinical metadata
-METADATA.CLINICAL_ID <- 'syn3191087'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.CLINICAL_ID
-METADATA.CLINICAL_OBJ <- synGet(METADATA.CLINICAL_ID)
-METADATA.CLINICAL <- read.table(METADATA.CLINICAL_OBJ@filePath,sep=',',header=T)
+METADATA_TC_ID <- 'syn3817650'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TC_ID
+METADATA_TC <- read.table(synGet(METADATA_TC_ID)@filePath,sep='\t',header=T, stringsAsFactors=F) %>%
+  tidyr::separate(ID, c('Donor_ID', 'BrainRegion'), sep = '_') %>%
+  dplyr::mutate(ID = paste(Donor_ID, BrainRegion, sep = '_'))
 
-# Get technical covariates
-METADATA.TECH_ID <- 'syn4300313'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.TECH_ID
-METADATA.TECH_OBJ <- synGet(METADATA.TECH_ID)
-METADATA.TECH <- read.table(METADATA.TECH_OBJ@filePath,sep='\t',header=T)
+# Get clinical metadata
+METADATA_CE_ID <- 'syn5223705'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CE_ID
+METADATA_CE <- read.table(synGet(METADATA_CE_ID)@filePath,sep='\t',header=T, stringsAsFactors=F) %>%
+  tidyr::separate(SampleID, c('Donor_ID', 'BrainRegion')) %>%
+  dplyr::mutate(ID = paste(Donor_ID, BrainRegion, sep = '_')) %>%
+  dplyr::rename(Gender = Sex, FLOWCELL = Flowcell)
 
-# Fix error in technical covariates data
-KEY_ID <- 'syn3382527'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = KEY_ID  
-KEY <- read.csv(synGet(KEY_ID)@filePath) %>% 
-  filter(mrna_data == 1) %>%
-  dplyr::select(projid, mrna_id) %>%
-  separate(mrna_id, c('a','b','batch'), sep = '_') %>% 
-  unite(Sampleid, a, b) %>%
-  dplyr::select(-batch) %>%
-  unique
-
-# Match technical and clinical covariates
-METADATA <- METADATA.TECH %>% 
-  dplyr::select(-projid) %>%
-  dplyr::left_join(KEY) %>%
-  dplyr::left_join(METADATA.CLINICAL)
-
-# Pick higher quality RIN batch for sample 492_120515
-METADATA <- METADATA %>%
-  group_by(Sampleid) %>%
-  top_n(1, RINcontinuous)
+METADATA = rbindlist(list(METADATA_TC, METADATA_CE), use.names = T, fill = T) %>%
+  as.data.frame
 ```
@@
 ### Data preprocessing
-```{r preprocess.data,cache=TRUE, echo=TRUE}
-# Remove samples with no cogdx, RIN and PMI scores
+```{r preprocess.data,cache=FALSE, echo=TRUE}
+# Fix metadata
+METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
+METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
+METADATA$BrainRegion.Diagnosis = paste(METADATA$BrainRegion, METADATA$Diagnosis, sep = '.')
+METADATA$RIN2 = METADATA$RIN^2
 METADATA <- METADATA %>%
-  ungroup %>%
-  filter(Sampleid %in% colnames(COUNT)) %>%
-  filter(!is.na(cogdx)) %>%
-  filter(!is.na(RINcontinuous)) %>%
-  filter(!is.na(pmi)) %>%
-  filter(!is.na(Sampleid)) %>%
-  filter(!(Sampleid %in% c('380_120503'))) %>%
-  as.data.frame()
+  filter(!is.na(Source)) %>%
+  dplyr::filter(!(ID %in% c("11311_CER", "1923_CER", "1923_TCX", 
+                            "11396_TCX", "11294_TCX", "11408_TCX")))
 ```
+Following samples were considered outliers and removed based on PCA and/or eucledian distance: 11311\_CER, 1923\_CER, 1923\_TCX, 11396\_TCX, 11294\_TCX, 11408\_TCX
+```{r preprocess.data1}
+# Match covariates to expression data
+indToRetain = intersect(METADATA$ID, colnames(COUNT))
+indRemoved = setdiff(colnames(COUNT), METADATA$ID)
 
-```{r preprocess.data1,cache=TRUE}
-# Match covariates to expression data
-indToRetain = intersect(METADATA$Sampleid, colnames(COUNT))
-removedIDs = setdiff(colnames(COUNT), METADATA$Sampleid)
+COUNT <- COUNT[, indToRetain]
 
-COUNT = COUNT[,indToRetain]
-
-rownames(METADATA) = METADATA$Sampleid
+rownames(METADATA) = METADATA$ID
 METADATA = METADATA[indToRetain,]
 ```
-Dorsolateral prefrontal cortex of `r dim(COUNT)[2]` subjects from the ROS and MAP cohorts are used for the analysis. Following sample are removed `r paste(removedIDs, collapse = ',')`
+`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project.
 
-### Get Covariates
-```{r covariates.clustering, cache=TRUE}
-#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv"
-FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx")
-ContCovariates <- c("RINcontinuous","age_first_ad_dx","age_death","age_at_visit_max","pmi","educ")
+Following samples were removed `r indRemoved`
+
+### Covariate clustering
+Determine relationship between covariates. 
+```{r covariates.clustering}
+FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
+ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI")
 
@@
 COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
-COVARIATES <- data.frame(lapply(COVARIATES,function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'\\+','')})}))
-rownames(COVARIATES) <- METADATA$Sampleid
+rownames(COVARIATES) <- METADATA$ID
 
@@
 COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
-
-# Add in RIN^2 values
-COVARIATES$RINcontinuous2 = COVARIATES$RINcontinuous^2
-ContCovariates = c(ContCovariates, 'RINcontinuous2')
 ```
-
-### CPM Normalisation
+### Gene Filtering
 Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.
 
-Remove genes that have less than 1 cpm counts in at least 50% of samples per cogdx
+Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion.Diagnosis
 ```{r cpmnormalisation}
@@
   rownameToFirstColumn('Sampleid') %>%
-  dlply(.(cogdx), .fun = function(mtd, count){
+  dlply(.(BrainRegion.Diagnosis), .fun = function(mtd, count){
     processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$Sampleid],
@@
 Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples
-
 ### Normalisation (with NULL)
@@
 ```
-
 ### Get cell type specific signatures and expression data from synapse
@@
 
-```{r get.gene.symbols, cache=TRUE, eval=T}
+```{r get.gene.symbols, cache=TRUE, include=FALSE}
 backgroundGenes = data.frame(gene.id = rownames(VOOM.GENE_EXPRESSION$E)) %>%
@@
 write.table(ref.expr1, file = 'ref.expr.txt', sep = '\t', quote=F, row.names = F)
-write.table(EXPR1[,1:20], file = 'mixture.expr.txt', sep = '\t', quote=F, row.names = F)
+write.table(EXPR1, file = 'mixture.expr.txt', sep = '\t', quote=F, row.names = F)
 
@@
 # p = p + facet_grid(.~cogdx)
-p
+# p
 
@@
 ```{r syn.store, cache=TRUE}
-CODE = Folder(name = 'ROSMAP Cell Type Abundance with Reprocessed RNASeq', parentId = parentId)
+CODE = Folder(name = 'MAYO Cell Type Abundance with Reprocessed RNASeq', parentId = parentId)
 CODE = synStore(CODE)
@@
   rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'ROSMAP_CellFraction_Reproc.tsv', row.names = F, sep = '\t', quote = F)
-obj = File('ROSMAP_CellFraction_Reproc.tsv', name = 'ROSMAP Cell Fraction', parentId = CODE$properties$id)
+  write.table(file = 'MAYO_CellFraction_Reproc.tsv', row.names = F, sep = '\t', quote = F)
+obj = File('MAYO_CellFraction_Reproc.tsv', name = 'MAYO Cell Fraction', parentId = CODE$properties$id)
 obj = synStore(obj, activityName = activityName, activityDescription = activityDescription,

### 2017-02-03 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd
Added markdown to deconvolute celltype fraction from MSSM data

--- 2017-02-03 code/Rmd/Mayo_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd
@@
 ---
-title: "Markdown to estimate the abundance of brain cell types with the reprocessed counts from Mayo cohorts (with TMM-VOOM normalisation)"
+title: "Markdown to estimate the abundance of brain cell types with the reprocessed counts from MSSM cohorts (with TMM-VOOM normalisation)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd",
                                  parentId = "syn5570325",
-                                 entityName = 'MAYO Cell Type Abundance with Reprocessed RNASeq')
+                                 entityName = 'MSSM Cell Type Abundance with Reprocessed RNASeq')
 ```
 
-```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
+```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
 ## It is assumed your working directory is where this file
@@
 parentId = 'syn5570325';
-activityName = 'Cell type abundance in Mayo';
-activityDescription = 'Cell type abundance in Mayo with TMM-VOOM normalised reprocessed RNASeq counts';
+activityName = 'Cell type abundance in MSSM';
+activityDescription = 'Cell type abundance in MSSM with TMM-VOOM normalised reprocessed RNASeq counts';
 
-thisFileName <- 'Mayo_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd'
+thisFileName <- 'MSSM_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd'
 
@@
 ### Data download
+Data were obtained from AMP-AD portal of MSSM, and following brain regions were analysed for differential expression between case and control:
+  | Region | Name                           | 
+  | :----- | ------------------------------ | 
+  | BM_10  | Frontal Pole (FP)              |
+  | BM_22  | Superior Temporal Gyrus (STG)  |
+  | BM_36  | Parahipocampal Gyrus (PHG)     |
+  | BM_44  | inferior frontal Gyrus (IFG)   |
+
+* BM 10: frontopolar prefrontal cortex, rostrolateral prefrontal cortex, or anterior prefrontal cortex, 
+* BM 22: In humans it corresponds approximately to the lateral and caudal two thirds of the superior temporal gyrus, 
+* BM 36: Ectorhinal area 36 is a subdivision of the cytoarchitecturally defined temporal region of cerebral cortex,
+* BM 44: pars opercularis (of the inferior frontal gyrus), and it refers to a subdivision of the frontal region of cerebral cortex.
+
 #### Obtain count matrix and metadata from synapse.
@@
 # Get reprocessed sample counts for temporal cortex
-COUNT_TC_ID <- 'syn7987765';
-ALL_USED_IDs <- COUNT_TC_ID
-COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F)
-colnames(COUNT_TC) = paste(colnames(COUNT_TC), 'TCX', sep = '_')
+COUNT_ID <- 'syn7994853';
+ALL_USED_IDs <- COUNT_ID
+COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F)
 
-# Get reprocessed sample counts for cerebellum
-COUNT_CE_ID <- 'syn7513807';
-ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_CE_ID
-COUNT_CE <- read.table(synGet(COUNT_CE_ID)@filePath, header=T, sep='\t', check.names = F)
+# Get sample ids
+SampleID = data.frame(SampleID = colnames(COUNT),
+                 ID = colnames(COUNT)) %>% 
+  tidyr::separate(ID, c('A','B','ID'), sep = '_')
 
-COUNT = cbind(COUNT_TC, COUNT_CE[rownames(COUNT_TC),])
+SampleID = SampleID %>%
+  dplyr::select(SampleID, ID) %>%
+  dplyr::mutate(ID = as.numeric(ID))
+
+# Get technical metadata
+METADATA_TECH_ID <- 'syn6100548'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECH_ID
+METADATA_TECH <- read.table(synGet(METADATA_TECH_ID)@filePath,sep=',',header=T, row.names=1) %>%
+  group_by(sampleIdentifier) %>%
+  top_n(1, -rRNA.rate) %>%
+  dplyr::select(sampleIdentifier, BrodmannArea, barcode, individualIdentifier, batch, RIN, TotalReads, Mapped, rRNA.rate) %>%
+  unique
 
 # Get clinical metadata
-METADATA_TC_ID <- 'syn3817650'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TC_ID
-METADATA_TC <- read.table(synGet(METADATA_TC_ID)@filePath,sep='\t',header=T, stringsAsFactors=F) %>%
-  tidyr::separate(ID, c('Donor_ID', 'BrainRegion'), sep = '_') %>%
-  dplyr::mutate(ID = paste(Donor_ID, BrainRegion, sep = '_'))
+METADATA_CLINICAL_ID <- 'syn6101474'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CLINICAL_ID
+METADATA_CLINICAL <- read.csv(synGet(METADATA_CLINICAL_ID)@filePath, header = T)
 
-# Get clinical metadata
-METADATA_CE_ID <- 'syn5223705'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CE_ID
-METADATA_CE <- read.table(synGet(METADATA_CE_ID)@filePath,sep='\t',header=T, stringsAsFactors=F) %>%
-  tidyr::separate(SampleID, c('Donor_ID', 'BrainRegion')) %>%
-  dplyr::mutate(ID = paste(Donor_ID, BrainRegion, sep = '_')) %>%
-  dplyr::rename(Gender = Sex, FLOWCELL = Flowcell)
-
-METADATA = rbindlist(list(METADATA_TC, METADATA_CE), use.names = T, fill = T) %>%
-  as.data.frame
+METADATA <- full_join(METADATA_TECH, METADATA_CLINICAL)
 ```
@@
 # Fix metadata
-METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
-METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
-METADATA$BrainRegion.Diagnosis = paste(METADATA$BrainRegion, METADATA$Diagnosis, sep = '.')
+METADATA$AOD = gsub('\\+','',METADATA$AOD)
+METADATA$Dx[METADATA$CDR <= 1] = 'ND'
+METADATA$Dx[METADATA$CDR == 2 | METADATA$CDR == 3] = 'MD'
+METADATA$Dx[METADATA$CDR >= 4] = 'SD'
+METADATA$BrodmannArea.Dx = paste(METADATA$BrodmannArea, METADATA$Dx, sep = '.')
 METADATA$RIN2 = METADATA$RIN^2
+levels(METADATA$batch)[levels(METADATA$batch) == ''] = 'NoBatch'
+
 METADATA <- METADATA %>%
-  filter(!is.na(Source)) %>%
-  dplyr::filter(!(ID %in% c("11311_CER", "1923_CER", "1923_TCX", 
-                            "11396_TCX", "11294_TCX", "11408_TCX")))
+  filter(!is.na(BrodmannArea), !is.na(PMI), !is.na(RIN))
 ```
-Following samples were considered outliers and removed based on PCA and/or eucledian distance: 11311\_CER, 1923\_CER, 1923\_TCX, 11396\_TCX, 11294\_TCX, 11408\_TCX
+
 ```{r preprocess.data1}
 # Match covariates to expression data
-indToRetain = intersect(METADATA$ID, colnames(COUNT))
-indRemoved = setdiff(colnames(COUNT), METADATA$ID)
+indToRetain = intersect(METADATA$sampleIdentifier, colnames(COUNT))
+indRemoved = setdiff(colnames(COUNT), METADATA$sampleIdentifier)
 
@@
 
-rownames(METADATA) = METADATA$ID
+METADATA = as.data.frame(METADATA)
+rownames(METADATA) = METADATA$sampleIdentifier
 METADATA = METADATA[indToRetain,]
 ```
-`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project.
+`r dim(COUNT)[2]` samples from `r length(unique(METADATA$individualIdentifier))` subjects were obtained from the MSBB cohorts in AMP-AD reprocessed rnaseq project
 
-Following samples were removed `r indRemoved`
-
+Following `r length(indRemoved)` samples are removed:`r paste(indRemoved, collapse = ', ')`
 ### Covariate clustering
@@
 ```{r covariates.clustering}
-FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
-ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI")
+FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrodmannArea.Dx")
+ContCovariates <- c("RIN", "RIN2", "AOD","PMI", "TotalReads", "Mapped", "rRNA.rate")
 
@@
 COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
-rownames(COVARIATES) <- METADATA$ID
+rownames(COVARIATES) <- METADATA$sampleIdentifier
 
@@
 
-Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion.Diagnosis
+Remove genes that have less than 1 cpm counts in at least 50% of samples per BrodmannArea.Dx
 ```{r cpmnormalisation}
 genesToAnalyze = COVARIATES %>%
-  rownameToFirstColumn('Sampleid') %>%
-  dlply(.(BrainRegion.Diagnosis), .fun = function(mtd, count){
-    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$Sampleid],
+  rownameToFirstColumn('sampleIdentifier') %>%
+  dlply(.(BrodmannArea.Dx), .fun = function(mtd, count){
+    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$sampleIdentifier],
                                                      MIN_GENE_CPM=1, 
@@
 
-```{r get.gene.symbols, cache=TRUE, include=FALSE}
+```{r get.gene.symbols, include=FALSE}
 backgroundGenes = data.frame(gene.id = rownames(VOOM.GENE_EXPRESSION$E)) %>%
@@
 
-```{r plot.cellprop}
+```{r plot.cellprop, fig.width=16, fig.height=8}
 cellprop1 = as.data.frame(cellprop)
-cellprop1$cogdx = COVARIATES[rownames(cellprop), 'cogdx']
+cellprop1$BrodmannArea.Dx = COVARIATES[rownames(cellprop), 'BrodmannArea.Dx']
 cellprop1 = cellprop1 %>% 
@@
   dplyr::select(-`P-value`,-Correlation, -RMSE) %>%
-  tidyr::gather(Celltype, Fraction, -SampleID, -cogdx) %>%
-  dplyr::arrange(desc(cogdx))
+  tidyr::gather(Celltype, Fraction, -SampleID, -BrodmannArea.Dx) %>%
+  dplyr::arrange(desc(BrodmannArea.Dx))
 
-p = ggplot(cellprop1, aes(x = SampleID, y = Fraction, fill = Celltype, color = cogdx)) + geom_bar(stat = 'identity')
+p = ggplot(cellprop1, aes(x = SampleID, y = Fraction, fill = Celltype, color = BrodmannArea.Dx)) + geom_bar(stat = 'identity')
 p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
-# p = p + facet_grid(.~cogdx)
+# p = p + facet_grid(.~BrodmannArea.Dx)
 # p
 
-p = ggplot(cellprop1, aes(x = Celltype, y = Fraction, color = cogdx)) + geom_boxplot()
+p = ggplot(cellprop1, aes(x = Celltype, y = Fraction, color = BrodmannArea.Dx)) + geom_boxplot()
 p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
@@
 ```{r syn.store, cache=TRUE}
-CODE = Folder(name = 'MAYO Cell Type Abundance with Reprocessed RNASeq', parentId = parentId)
+CODE = Folder(name = 'MSSM Cell Type Abundance with Reprocessed RNASeq', parentId = parentId)
 CODE = synStore(CODE)
@@
   rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'MAYO_CellFraction_Reproc.tsv', row.names = F, sep = '\t', quote = F)
-obj = File('MAYO_CellFraction_Reproc.tsv', name = 'MAYO Cell Fraction', parentId = CODE$properties$id)
+  write.table(file = 'MSSM_CellFraction_Reproc.tsv', row.names = F, sep = '\t', quote = F)
+obj = File('MSSM_CellFraction_Reproc.tsv', name = 'MSSM Cell Fraction', parentId = CODE$properties$id)
 obj = synStore(obj, activityName = activityName, activityDescription = activityDescription,

### 2017-02-03 code/Rmd/Mayo_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd

@@
 # Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "dec2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "mar2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
 
@@
 cellprop1 = as.data.frame(cellprop)
-cellprop1$cogdx = COVARIATES[rownames(cellprop), 'cogdx']
+cellprop1$BrainRegion.Diagnosis = COVARIATES[rownames(cellprop), 'BrainRegion.Diagnosis']
 cellprop1 = cellprop1 %>% 
@@
   dplyr::select(-`P-value`,-Correlation, -RMSE) %>%
-  tidyr::gather(Celltype, Fraction, -SampleID, -cogdx) %>%
-  dplyr::arrange(desc(cogdx))
+  tidyr::gather(Celltype, Fraction, -SampleID, -BrainRegion.Diagnosis) %>%
+  dplyr::arrange(desc(BrainRegion.Diagnosis))
 
-p = ggplot(cellprop1, aes(x = SampleID, y = Fraction, fill = Celltype, color = cogdx)) + geom_bar(stat = 'identity')
+p = ggplot(cellprop1, aes(x = SampleID, y = Fraction, fill = Celltype, color = BrainRegion.Diagnosis)) + geom_bar(stat = 'identity')
 p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
-# p = p + facet_grid(.~cogdx)
+# p = p + facet_grid(.~BrainRegion.Diagnosis)
 # p
 
-p = ggplot(cellprop1, aes(x = Celltype, y = Fraction, color = cogdx)) + geom_boxplot()
+p = ggplot(cellprop1, aes(x = Celltype, y = Fraction, color = BrainRegion.Diagnosis)) + geom_boxplot()
 p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))

### 2017-02-03 code/R/computeCellTypeSpecificity.R
Added R code to compute cell type specificity metric tau for each gene

@@ -1,0 +1,71 @@
+## Compute cell type specificity metrics
+## NOTE: Metrics were obtained from A benchmark of gene expression tissue-specificity metrics
+##       DOI: 10.1093/bib/bbw008
+
+# Load libraries
+library(CovariateAnalysis)
+library(data.table)
+library(tidyr)
+library(plyr)
+library(dplyr)
+library(stringr)
+
+library(synapseClient)
+library(knitr)
+library(githubr)
+
+synapseLogin()
+
+# Get reference expression and cell type labels from Darmanis., S et al., 2015
+ALL_USED_IDs = c('syn8077142', 'syn8077191')
+ref.expr = downloadFile('syn8077142')
+rownames(ref.expr) = ref.expr$hgnc_symbol
+ref.expr$hgnc_symbol = NULL
+
+ref.cellTypes = downloadFile('syn8077191') %>%
+  dplyr::select(SampleID, CellType)
+
+# Summarise reference expression interms of cell types
+ref.expr.sum = list()
+for(celltype in unique(ref.cellTypes$CellType)){
+  ids = ref.cellTypes$SampleID[ref.cellTypes$CellType == celltype]
+  ref.expr.sum[[celltype]] = log2(rowSums(2^ref.expr[,ids], na.rm = T)) %>%
+    rownameToFirstColumn('hgnc_symbol') %>%
+    plyr::rename(c('DF' = celltype))
+}
+ref.expr.sum = join_all(ref.expr.sum)
+
+# Compute tau
+tmp = data.matrix(ref.expr.sum[,-(1)])
+rownames(tmp) = ref.expr.sum$hgnc_symbol
+tau = rowSums(1 - (tmp/rowMax(tmp)), na.rm = T)
+tau = (tau/(dim(tmp)[2]-1)) %>%
+  rownameToFirstColumn('hgnc_symbol') %>%
+  dplyr::rename(tau = DF)
+
+# Compute TSI
+TSI = rowMax(tmp)/rowSums(tmp) %>%
+  as.data.frame()
+colnames(TSI) = 'TSI'
+TSI$hgnc_symbol = rownames(TSI)
+
+# Compute Hg
+pi = tmp/rowSums(tmp)
+Hg = -rowSums(pi*log2(pi)) %>%
+  rownameToFirstColumn('hgnc_symbol') %>%
+  dplyr::rename(Hg = DF)
+
+# Get github commit
+thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", 
[…]

### 2017-02-06 code/R/computeCellTypeSpecificity.R
Minor changes

@@ -65,7 +65,7 @@ thisFile <- getPermlink(repository = thisRepo,
 # Store results to synapse
 plyr::join_all(list(tau, TSI, Hg), type = 'full', match = 'all') %>%
   write.table(file = 'CellSpecificityMetrics.tsv', sep = '\t', row.names = F, quote=F)
-obj = File('CellSpecificityMetrics.tsv', parentId = 'syn8068475')
+obj = File('CellSpecificityMetrics.tsv', parentId = 'syn8077138')
 obj = synStore(obj, activityName = 'Compute cell type specificity',
                used = ALL_USED_IDs,
                executed = thisFile)

### 2017-02-13 code/Rmd/MetaAnalysis.Rmd
Major modifications

@@
 ---
-title: "Meta analysis of differential expression results from AMP-AD (Control vs AD)"
+title: "Meta analysis of differential expression results from AMP-AD (Control vs AD, reprocessed RNASeq)"
 author: "Thanneer Perumal"
@@
                                  parentId ="syn5569102",
-                                 entityName = 'MetaAnalysis of Control vs AD from all three brain banks')
+                                 entityName = 'MetaAnalysis of Control vs AD from all three brain banks (Reprocessed RNASeq)')
 ```
@@
 # Load required libraries
-library(CovariateAnalysis)
+library(CovariateAnalysis) # install from th1vairam repo using devtools::install_github('th1vairam/CovariateAnalysis@devtools')
 library(data.table)
@@
 library(ComplexHeatmap)
+library(biomaRt)
+library(Matrix)
+library(org.Hs.eg.db)
 
+library(limma)
+library(edgeR)
 library(RankProd)
@@
 library(foreach)
+library(doParallel)
 
-clust = makeCluster(14)
+clust = makeCluster(detectCores() - 2)
 registerDoParallel(clust)
@@
 ```{r covariates}
-# covariates.id = c('ROSMAP' = 'syn6114444',
-#                   'MSBB' = 'syn6131421',
-#                   'MAYO' = 'syn6128267')
-# SVA Normalised
-covariates.id = c('ROSMAP' = 'syn6129368',
-                  'MSBB' = 'syn7059296',
-                  'MAYO' = 'syn6128630')
+covariates.ids = c(ROSMAP = 'syn8018344',
+                   MSBB = 'syn8073656',
+                   MAYO = 'syn8028561')
+covariates = lapply(covariates.ids, function(id){
+  x = downloadFile(id)
+  colnames(x)[1] = 'SampleID'
+  return(x)
+})
 
-covariates = lapply(covariates.id, downloadFile)
+covariates$ROSMAP = dplyr::filter(covariates$ROSMAP, cogdx %in% c(1,4)) %>%
+  dplyr::mutate(Dx = factor(cogdx, levels = c(1,4), labels = c('1' = 'Control', '4' = 'AD')),
+                Study = 'ROSMAP',
+                BrainRegion = 'DLPFC') %>%
+  dplyr::select(SampleID, Study, BrainRegion, Dx)
 
-covariates$ROSMAP = covariates$ROSMAP %>%
-  dplyr::select(Sampleid, msex, cogdx) %>%
-  dplyr::rename(ID = Sampleid, Gender = msex, Status = cogdx) %>%
-  dplyr::mutate(Gender = factor(Gender, labels = c('1' = 'MALE','2'='FEMALE')),
-                Status = factor(Status, labels = c('1' = 'NCI', '2' = 'MCI', '3' = 'MCI.CI', 
-                                                   '4' = 'AD', '5' = 'AD.CI', '6' = 'OD')),
-                BrainRegion = 'DLPFC') %>%
-  dplyr::filter(Status %in% c('NCI', 'AD'))
+covariates$MSBB = dplyr::filter(covariates$MSBB, BrodmannArea.Dx %in% c('BM10.SD', 'BM22.SD', 'BM36.SD', 'BM44.SD', 'BM36.ND', 'BM10.ND', 'BM44.ND', 'BM22.ND')) %>%
+  dplyr::select(SampleID, BrodmannArea.Dx) %>%
+  tidyr::separate(BrodmannArea.Dx, c('BrainRegion', 'Dx')) %>%
+  dplyr::mutate(Study = 'MSSM',
+                Dx = factor(Dx, levels = c('ND','SD'), labels = c('ND' = 'Control', 'SD' = 'AD')),
+                BrainRegion = factor(BrainRegion, labels = c('FP', 'STG', 'PHG', 'IFG'))) %>%
+  dplyr::select(SampleID, Study, BrainRegion, Dx)
 
-covariates$MSBB = covariates$MSBB %>%
-  dplyr::select(SampleId, SEX, BrainRegion.Diagnosis) %>%
-  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Dx'), sep = '\\.') %>%
-  dplyr::rename(ID = SampleId, Gender = SEX, Status = Dx) %>%
-  dplyr::mutate(Gender = factor(Gender, labels = c('M' = 'MALE','F'='FEMALE')),
-                Status = factor(Status),
-                BrainRegion = factor(BrainRegion, labels = c('BM_10' = 'FP', 'BM_22' = 'STG', 
-                                                             'BM_36' = 'PHG', 'BM_44' = 'IFG'))) %>%
-  dplyr::filter(Status %in% c('ND', 'SD'))
+covariates$MAYO = dplyr::filter(covariates$MAYO, BrainRegion.Diagnosis %in% c('TCX.AD', 'TCX.Control', 'CER.AD', 'CER.Control')) %>%
+  dplyr::select(SampleID, BrainRegion.Diagnosis) %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Dx')) %>%
+  dplyr::mutate(Study = 'MAYO',
+                Dx = factor(Dx, levels = c('Control','AD')),
+                BrainRegion = factor(BrainRegion, levels = c('CER', 'TCX'))) %>%
+  dplyr::select(SampleID, Study, BrainRegion, Dx)
 
-covariates$MAYO = covariates$MAYO %>%
-  dplyr::select(ID, Gender, BrainRegion.Diagnosis) %>%
-  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Dx'), sep = '\\.') %>%
-  dplyr::rename(Status = Dx) %>%
-  dplyr::mutate(Gender = factor(Gender, labels = c('M' = 'MALE','F'='FEMALE')),
-                Status = factor(Status)) %>%
-  dplyr::filter(Status %in% c('Control', 'AD'))
-
-covariates = rbindlist(covariates, use.names = T, fill = T, idcol = 'Study')
+covariates = covariates %>%
+  rbindlist(use.names = T, fill = T)
 ```
 
-### Get logCPM counts from synapse
+### Get filtered gene.id and sample.id from synapse
 ```{r counts}
-# logcpm.id = c('ROSMAP' = 'syn6114447' ,
-#               'MSBB' = 'syn6131419',
-#               'MAYO' = 'syn6128269')
-# Get SVA adjusted data
-logcpm.id = c('ROSMAP' = 'syn6129370',
-              'MSBB' = 'syn7059294',
-              'MAYO' = 'syn6128632')
+logcpm.ids = c('ROSMAP' = 'syn8018352',
+               'MSBB' = 'syn8073666',
+               'MAYO' = 'syn8028565')
 
-logcpm = lapply(logcpm.id, downloadFile) %>%
+logcpm = lapply(logcpm.ids, downloadFile) %>%
   lapply(function(x){
-    x = dplyr::select(x, -hgnc_symbol)
+    x = dplyr::select(x, -hgnc_symbol, -ensembl_gene_id)
   }) %>%
   join_all(type = 'full') %>%
-  dplyr::select(Gene.ID, ensembl_gene_id, one_of(covariates$ID)) 
-logcpm[is.na(logcpm)] = dplyr::select(logcpm, -Gene.ID, -ensembl_gene_id) %>%
-  min
+  dplyr::select(gene.id, one_of(covariates$SampleID)) 
+logcpm[is.na(logcpm)] = min(logcpm[,-(1)], na.rm = T)
+```
+
+### Get raw counts from synapse
+```{r raw.counts, eval = F}
+raw.counts.ids = c('MAYO_CBE' = 'syn7513807',
+                   'MSBB' = 'syn7994853',
+                   'MAYO_TCX' = 'syn7987765',
+                   'ROSMAP' = 'syn7750270')
+
+lcpm = lapply(raw.counts.ids, function(id, lcpm){
+  expr = read.table(synGet(id)@filePath, sep = '\t', check.names = F) 
+  lib.size = calcNormFactors(expr, method = 'TMM')
+  expr = voom(expr, lib.size = lib.size, design = NULL, plot = F, normalize.method = 'quantile')$E
+  expr = expr[row.names(expr) %in% lcpm$gene.id, ] %>%
+    rownameToFirstColumn('gene.id')
+}, logcpm) %>%
+  join_all(type = 'full')
 ```
@@
 ### Get residual expression from synapse
-```{r counts1}
+```{r resid.expr}
 # Get SVA adjusted data
@@
 
-### Download differential expression data
-```{r download.data}
-# diffexp.id = c('ROSMAP' = 'syn6114453',
-#                'MSBB' = 'syn6131423',
-#                'MAYO' = 'syn6128273')
-# Get SVA adjusted differential expression
-diffexp.id = c('ROSMAP' = 'syn6129374',
-               'MSBB' = 'syn7059298',
-               'MAYO' = 'syn6128636')
+### Get hgnc_symbol mapping from biomaRt
+```{r biomart}
+# Define biomart object
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "dec2016.archive.ensembl.org")
 
-diffexp = lapply(diffexp.id, downloadFile)
+# Get background gene ids
+backgroundGenes = logcpm[,'gene.id', drop = F] %>% 
+  dplyr::mutate(ID = gene.id) %>%
+  tidyr::separate(ID, c('ensembl_gene_id','position'), sep = '\\.')
 
-diffexp$ROSMAP = diffexp$ROSMAP %>%
-  filter(Comparison == 'cogdx1-cogdx4') %>%
-  dplyr::select(ensembl_gene_id, logFC, P.Value, adj.P.Val) %>%
-  # dplyr::select(Gene.ID, logFC, P.Value, adj.P.Val) %>%
-  dplyr::mutate(Tissue = 'DLPFC', Contrast = 'NCI-AD')
+# Query biomart
+Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
+                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
+                       mart = mart)
 
-diffexp$MSBB = diffexp$MSBB %>%
-  dplyr::filter(Name == 'ND-SD') %>%
-  dplyr::select(ensembl_gene_id, logFC, P.Value, adj.P.Val, BrainRegion, Name) %>%
-  dplyr::mutate(BrainRegion = factor(BrainRegion, labels = c('BM_10' = 'FP', 'BM_22' = 'STG', 
-                                                             'BM_36' = 'PHG', 'BM_44' = 'IFG'))) %>%
-  dplyr::rename(Tissue = BrainRegion, Contrast = Name)
+Entrez2Ensembl <- as.data.frame(org.Hs.egENSEMBL2EG)
+Entrez2HGNC <- as.data.frame(org.Hs.egSYMBOL2EG)
+HGNC2Ensembl <- left_join(Entrez2HGNC, Entrez2Ensembl)
+HGNC2Ensembl = HGNC2Ensembl[grep('^LOC',HGNC2Ensembl$symbol),c('ensembl_id', 'symbol')] %>%
+  filter(!is.na(ensembl_id), !is.na(symbol)) %>%
+  dplyr::rename(ensembl_gene_id = ensembl_id, hgnc_symbol = symbol)
 
-diffexp$MAYO = diffexp$MAYO %>%
-  dplyr::filter(Name == 'Control-AD') %>%
-  tidyr::separate(Gene.ID, c('ensembl_gene_id', 'position'), sep = '\\.') %>%
-  dplyr::select(ensembl_gene_id, logFC, P.Value, adj.P.Val, BrainRegion, Name) %>%
-  dplyr::rename(Tissue = BrainRegion, Contrast = Name)
+Ensemble2HGNC = left_join(backgroundGenes,
+                          rbindlist(list(Ensemble2HGNC, HGNC2Ensembl), use.names = T, fill = T) %>%
+                            unique) %>%
+  unique
+```
 
-diffexp = rbindlist(diffexp, use.names = T, fill = T, idcol = 'Study')
+### Get differential expression results from synapse
+```{r diffexp, fig.height=16, fig.width=10}
+diff.exp.ids = c(ROSMAP = 'syn8018358',
+                 MSBB = 'syn8073678',
+                 MAYO = 'syn8028572')
+diff.exp = lapply(diff.exp.ids, downloadFile)
 
-diffexp$DE = 'NONE'
-diffexp$DE[diffexp$adj.P.Val <= 0.05 & diffexp$logFC >= log2(1.2)] = 'UP'
-diffexp$DE[diffexp$adj.P.Val <= 0.05 & diffexp$logFC <= -log2(1.2)] = 'DOWN'
+# Filter case-control differences alone
+diff.exp$ROSMAP = filter(diff.exp$ROSMAP, Comparison == 'cogdx1-cogdx4')
+diff.exp$MSBB = filter(diff.exp$MSBB, Comparison == 'ND-SD')
+diff.exp$MAYO = filter(diff.exp$MAYO, Comparison == 'Control-AD')
+
+writeLines('Total number of genes differentially expressed between Control-AD at an FDR of 0.05 and foldchange of 1.2 are')
+lapply(diff.exp, function(x){
+  x = filter(x, adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) 
+  x$Direction = 'NONE'
+  x$Direction[x$logFC > 0] = 'UP'
+  x$Direction[x$logFC < 0] = 'DOWN'
+  x = x %>% group_by(Study, Region, Direction) %>%
+    summarise(count = length(unique(gene.id)))
+}) %>%
+  rbindlist() %>%
+  spread(Direction, count) %>%
+  kable
+
+tmp = diff.exp %>% 
+  rbindlist(use.names = T, fill = T) 
+tmp$Direction = 'NONE'
+tmp$Direction[tmp$logFC > log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
+tmp$Direction[tmp$logFC < -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'
 
 pl = list()
-for (study in c('ROSMAP', 'MSBB', 'MAYO')){
-  p = filter(diffexp, Study == study) %>%
-    ggplot(aes(x = logFC, y = -log10(adj.P.Val), color = DE)) 
-  p = p + scale_colour_manual(values = c("green","grey", "red"))
-  p = p + geom_point() + theme(legend.position = 'None')
-  p = p + facet_grid(Study~.+Tissue)
+for (study in c('ROSMAP','MSBB','MAYO')){
+  p = ggplot(tmp %>% filter(Study == study) %>% droplevels, aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point()
+  p = p + scale_color_manual(values = c('green','grey','red'))
+  p = p + facet_grid(Study~.+Region) + theme_bw() + theme(legend.position = 'top')
   pl[[study]] = p
@@
 multiplot(plotlist = pl, cols = 1)
-```
 
-### Get hgnc_symbol mapping from biomaRt
-```{r biomart}
-# Define biomart object
-mart <- useMart('ENSEMBL_MART_ENSEMBL', dataset = "hsapiens_gene_ensembl", host = 'dec2015.archive.ensembl.org')
+# Find genes that are differentially expressed in all brain regions
+diff.exp.gl = dlply(tmp, .(Region), .fun = function(x){
+  dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05) %>%
+    dplyr::select(gene.id) %>%
+    unique %>% unlist
+})
 
-# Query biomart
-Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
-                       filters = "ensembl_gene_id", 
-                       values = unique(diffexp$ensembl_gene_id),
-                       mart = mart) %>%
-  data.table()
+tmp1 = filter(tmp, gene.id %in% Reduce(intersect, diff.exp.gl)) %>%
+  dplyr::select(-ensembl_gene_id, -hgnc_symbol) %>%
+  left_join(Ensemble2HGNC %>% dplyr::select(-position))
+tmp1$hgnc_symbol[tmp1$ensembl_gene_id == 'ENSG00000275830'] = 'LOC101927646'
 
-# Add hgnc_symbols to the logcpm matrix
-logcpm = logcpm %>%
-  left_join(Ensemble2HGNC)
+tmp.eff.sz = group_by(tmp1, gene.id) %>% summarise(es = sum(logFC, na.rm = T)) %>% top_n(6, abs(es))
 
-# Add hgnc_symbols to the resid.mat matrix
-resid.expr = resid.expr %>%
-  left_join(Ensemble2HGNC)
+writeLines('Top six differentially expressed genes are')
+pl = list()
+p = ggplot(tmp1 %>% filter(gene.id %in% tmp.eff.sz$gene.id[tmp.eff.sz$es>0]), aes(x = Region, y = logFC, fill = Study)) + geom_bar(stat = 'identity')
+p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
+p = p + theme(legend.position = 'top')
+pl[[1]] = p
 
-# Add hgnc_symbols to the diffexp matrix
-diffexp = diffexp %>%
-  left_join(Ensemble2HGNC)
+p = ggplot(tmp1 %>% filter(gene.id %in% tmp.eff.sz$gene.id[tmp.eff.sz$es<0]), aes(x = Region, y = logFC, fill = Study)) + geom_bar(stat = 'identity')
+p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
+p = p + theme(legend.position = 'top')
+pl[[2]] = p
 
-diffexp$Direction = 'UP'
-diffexp$Direction[diffexp$logFC < 0] = 'DOWN'
+multiplot(plotlist = pl, cols = 2)
 
-writeLines('Differentially expressed at an fdr of 0.05')
-diffexp %>%
-  filter(adj.P.Val <= 0.05) %>%
-  group_by(Tissue, Direction) %>%
-  summarise(n = length(unique(ensembl_gene_id))) %>%
-  spread(Direction, n)
+# Genes differentially expressed in all brain regions
+tmp1 = dlply(tmp, .(Region), .fun = function(x){
+  dplyr::filter(x, abs(logFC) >= log2(1) & adj.P.Val <= 0.05) %>%
+    dplyr::select(gene.id) %>%
+    unique %>% unlist
+}) %>%
+  Reduce(intersect, .)
+writeLines('Genes that are differentially expressed between Control-AD at FDR 0.05 in all brain regions')
+kable(data.frame(ngenes = length(tmp1),
+                 hgnc_symbol = paste(unique(Ensemble2HGNC$hgnc_symbol[Ensemble2HGNC$gene.id %in% tmp1]), collapse = ', ')))
 
-writeLines('Differentially expressed at an fdr of 0.05 and fold change of 1.2')
-diffexp %>%
-  filter(adj.P.Val <= 0.05, (logFC >= log2(1.2) | logFC <= -log2(1.2))) %>%
-  group_by(Tissue, Direction) %>%
-  summarise(n = length(unique(ensembl_gene_id))) %>%
-  spread(Direction, n)
-```
-
-### Pathway enrichment
-```{r pathway.enrichment}
-# Differentially expressed in ROSMAP
-## Up regulated
-tmp = filter(diffexp, adj.P.Val <= 0.05, logFC >0, Tissue == 'DLPFC') %>%
-  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
-  filter(hgnc_symbol != '')
-### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6x3h
-
-## Down regulated
-tmp = filter(diffexp, adj.P.Val <= 0.05, logFC<0, Tissue == 'DLPFC') %>%
-  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
-  filter(hgnc_symbol != '')
-### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6x4e
-
-# Differentially expressed in MSSM FP
-## Up regulated
-tmp = filter(diffexp, adj.P.Val <= 0.05, logFC >0, Tissue == 'FP') %>%
-  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
-  filter(hgnc_symbol != '')
-### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6xcm
-
-## Down regulated
-tmp = filter(diffexp, adj.P.Val <= 0.05, logFC < 0, Tissue == 'FP') %>%
-  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
-  filter(hgnc_symbol != '')
-
-# Differentially expressed in MSSM STG
-## Up regulated
-tmp = filter(diffexp, adj.P.Val <= 0.05, logFC > 0, Tissue == 'STG') %>%
-  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
-  filter(hgnc_symbol != '')
-
-## Down regulated
-tmp = filter(diffexp, adj.P.Val <= 0.05, logFC < 0, Tissue == 'STG') %>%
-  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
-  filter(hgnc_symbol != '')
-
-# Differentially expressed in MSSM PHG
-## Up regulated
-tmp = filter(diffexp, adj.P.Val <= 0.05, logFC > 0, Tissue == 'PHG') %>%
-  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
-  filter(hgnc_symbol != '')
-### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6xco
-
-## Down regulated
-tmp = filter(diffexp, adj.P.Val <= 0.05, logFC < 0, Tissue == 'PHG') %>%
-  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
-  filter(hgnc_symbol != '')
-### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6xge
-
-# Differentially expressed in MAYO CER
-## Up regulated
-tmp = filter(diffexp, adj.P.Val <= 0.05, logFC > 0, Tissue == 'CER') %>%
-  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
-  filter(hgnc_symbol != '')
-### Enrichr results are here:
-
-## Down regulated
-tmp = filter(diffexp, adj.P.Val <= 0.05, logFC < 0, Tissue == 'PHG') %>%
-  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
-  filter(hgnc_symbol != '')
-### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6xge
+tmp1 = dlply(tmp, .(Region), .fun = function(x){
+  dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05) %>%
+    dplyr::select(gene.id) %>%
+    unique %>% unlist
+}) %>%
+  Reduce(intersect, .)
+writeLines('Genes that are differentially expressed between Control-AD at FDR 0.05 and foldchange of 1.2 in all brain regions')
+kable(data.frame(ngenes = length(tmp1),
+                 hgnc_symbol = paste(unique(Ensemble2HGNC$hgnc_symbol[Ensemble2HGNC$gene.id %in% tmp1]), collapse = ', ')))
 ```
@@
 ### Meta Analysis
-```{r metaanalysis}
-P.Value = diffexp %>% 
-  # filter(Tissue %in% c('DLPFC', 'PHG', 'TCX')) %>%
-  filter(ensembl_gene_id != '') %>%
-  dplyr::select(ensembl_gene_id, P.Value, Tissue, Contrast) %>%
-  tidyr::unite(Tissue.Contrast, Tissue, Contrast, sep = '_') %>%
-  tidyr::spread(Tissue.Contrast, P.Value)
+```{r metaanalysis.min.pval}
+# Get raw p.values
+P.Value = diff.exp %>% 
+  rbindlist(use.names = T, fill = T) %>%
+  dplyr::select(gene.id, P.Value, Region) %>%
+  unique() %>%
+  tidyr::spread(Region, P.Value)
 P.Value[is.na(P.Value)] = 1
+rownames(P.Value) = P.Value$gene.id
+P.Value$gene.id = NULL
+P.Value = data.matrix(P.Value)
 
-min.pval = dplyr::select(P.Value, -ensembl_gene_id) %>%
-  data.matrix %>% rowMin %>%
-  p.adjust(method = 'BH') %>%
-  data.frame(ensembl_gene_id = P.Value$ensembl_gene_id,
-             min.pval = .)
+# Perform minimum p-value based meta-analysis
+min.pval = P.Value %>% 
+  apply(1, min, na.rm = T) %>%
+  p.adjust(method = 'bonferroni') %>%
+  rownameToFirstColumn('gene.id') %>%
+  dplyr::rename(min.pval = DF)
 
-max.pval = dplyr::select(P.Value, -ensembl_gene_id) %>%
-  data.matrix %>% rowMax %>%
-  p.adjust(method = 'BH') %>%
-  data.frame(ensembl_gene_id = P.Value$ensembl_gene_id,
-             max.pval = .)
+# Perform fisher p-value based meta-analysis
+fb.pval = P.Value %>%
+  fishersPvalue %>%
+  p.adjust(method = 'bonferroni') %>%
+  rownameToFirstColumn('gene.id') %>%
+  dplyr::rename(fb.pval = DF)
+```
 
-fb.pval = dplyr::select(P.Value, -ensembl_gene_id) %>%
-  data.matrix %>% fishersPvalue %>%
-  p.adjust(method = 'BH') %>%
-  data.frame(ensembl_gene_id = P.Value$ensembl_gene_id,
-             fb.pval = .)
-
+```{r }
 # Rank prod based analysis

### 2017-02-13 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd
Temp commit

@@
 knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd",
-                                 parentId = "syn5570325",
+                                 parentId = "syn5570248",
                                  entityName = 'MSSM Cell Type Abundance with Reprocessed RNASeq')
@@
 # Synapse parameters
-parentId = 'syn5570325';
+parentId = 'syn5570248';
 activityName = 'Cell type abundance in MSSM';

### 2017-02-13 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM_CellTypeFracControl.Rmd
Added markdown to process reprocessed RNASeq from ROSMAP samples controlled for cell type fraction with TMM normalisation

--- 2017-01-30 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM.Rmd
@@
 ---
-title: "Covariate analysis of ROSMAP reprocessed counts (with TMM normalisation)"
+title: "Covariate analysis of ROSMAP reprocessed counts with controlled cell type variation (with TMM normalisation)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_TMM.Rmd",
+knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_TMM_CellTypeFracControl.Rmd",
                                  parentId = "syn5570291",
-                                 entityName = 'ROSMAP Reprocessed RNASeq TMM (Based on cogdx)')
+                                 entityName = 'ROSMAP Reprocessed RNASeq TMM WithCellTypeFractionControlled (Based on cogdx)')
 ```
@@
 parentId = 'syn5570291';
-activityName = 'Covariate analysis';
+activityName = 'Covariate analysis with celltype controlled';
 activityDescription = 'Covariate analysis of ROSMAP reprocessed RNASeq data';
 
-thisFileName <- 'ROSMAP_RNASeq_Reprocessed_TMM.Rmd'
+thisFileName <- 'ROSMAP_RNASeq_Reprocessed_TMM_CellTypeFracControl.Rmd'
 
@@
   top_n(1, RINcontinuous)
+
+# Get deconvoluted cell type fractions of ROSMAP samples from synapse 
+CELL_TYPE_FRAC = downloadFile('syn8124520') %>%
+  dplyr::select(-OPC, -`P-value`, -Correlation, -RMSE)
+rownames(CELL_TYPE_FRAC) = CELL_TYPE_FRAC$SampleID
+CELL_TYPE_FRAC$SampleID = NULL
+CELL_TYPE_FRAC = data.matrix(CELL_TYPE_FRAC)
 ```
@@
 METADATA = METADATA[indToRetain,]
+
+CELL_TYPE_FRAC = CELL_TYPE_FRAC[indToRetain,]
 ```
@@
 ContCovariates = c(ContCovariates, 'RINcontinuous2')
+
+# Add cell type fractions to the covariates
+COVARIATES = cbind(COVARIATES, CELL_TYPE_FRAC[rownames(COVARIATES),])
 ```
@@
 multiplot(plotlist = p, cols = 2)
+
+# Cell type fractions
+p = CELL_TYPE_FRAC %>%
+  rownameToFirstColumn('SampleID') %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID') %>%
+              dplyr::select(cogdx, SampleID)) %>%
+  tidyr::gather(CellType, Fraction, -SampleID, -cogdx) %>%
+  ggplot(aes(x = cogdx, y = Fraction, color = CellType))
+p = p + geom_boxplot()
+p
 
@@
 Clustering of voom adjusted data (with NULL design)
-```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
+```{r decompse.normalise.data, fig.height=5, fig.width=8, results='asis'}
 # Find principal components of expression to plot
@@
 ```
-### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix.
-Here "cogdx" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on cognitive scores)
+### Normalisation (custom design)
+Since many covariates are correlated, re-normalising COUNTS with an custom design matrix.
+Here "cogdx" times cell type fractions are chosen as the primary variables of interest (i.e., covariate selection is conditioned on cognitive scores times cell type proportion)
 
-Adding Batch and msex a priori to variable selection
+Final model:
+    Gx = Batch + msex + RINcontinuous + pmi + age_death + cogdx.oligodendrocytes + cogdx.astrocytes + cogdx.microglia + cogdx.neurons + cogdx.endothelial
 ```{r customNorm, results='asis'}
-# Primary variable of interest
-postAdjustCovars = c('cogdx', 'Batch', 'msex');
+# Covariates of interest
+postAdjustCovars = c('cogdx', 'Batch', 'msex', 'RINcontinuous', 'pmi', 'age_death');
 
-# Assign residual covariates
-residualSigCovars = preAdjustedSigCovars
-covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
+# Post adjusted design matrix
+DESIGN.MATRIX = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
 
-loopCount = 0 
-while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
+ind.design = grep('cogdx', colnames(DESIGN.MATRIX), value = T)
+ind.celltypes = colnames(CELL_TYPE_FRAC)
+for(i in ind.celltypes){
+  for(j in ind.design){
+    tmp = DESIGN.MATRIX[, j, drop = F] * COVARIATES[,i, drop = F]
+    colnames(tmp) = paste0(j,'.',i)
+    DESIGN.MATRIX = cbind(tmp, DESIGN.MATRIX)
+  }
+}
+DESIGN.MATRIX = DESIGN.MATRIX[, setdiff(colnames(DESIGN.MATRIX), ind.design)]
+DESIGN.MATRIX = DESIGN.MATRIX[,linColumnFinder(DESIGN.MATRIX)$indepCols]
+
 writeLines(paste('Using following covariates in the model:',
-                   paste(postAdjustCovars, collapse=', '),
+                 paste(colnames(DESIGN.MATRIX), collapse=', '),
                  'as fixed effects'))
 
-  # Post adjusted design matrix
-  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
-
 # Estimate voom weights
-  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DESIGN.MATRIX, plot=F)
 
@@
 
-  # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars))
-
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
 residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                                COVARIATES[, residCovars, drop=F], 
+                                              COVARIATES, 
                                               'adjusted design(voom-normalized)',
                                               isKeyPlot=TRUE)
-
-  # Add postadjusted covariates (if any)
-  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
-                                                                          c(postAdjustCovars))]
-
-  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
-  loopCount = loopCount + 1
-}
-
-modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
-                  'as fixed effects')
-
-tmp <- paste('Using following covariates in the final model:', modelStr)
 ```
@@
 ### Residual calculation
-Calculate weighted residuals and add back "cogdx" to the residuals
+Calculate weighted residuals and add back "cogdx.celltypes" to the residuals
 ```{r varsToAddBack}
 # Add variable of interest back to the residuals
-varsToAddIn = grep("cogdx", colnames(DM1$design), value = T)
+varsToAddIn = grep("cogdx", colnames(DESIGN.MATRIX), value = T)
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
-  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(VOOM.ADJUSTED.FIT$design[,varsToAddIn])
 ```
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("cogdx1-cogdx2",
-                                     "cogdx1-cogdx4",
-                                     "cogdx2-cogdx4"),
+contrast = makeContrasts(contrasts=c("cogdx1.oligodendrocytes-cogdx2.oligodendrocytes",
+                                     "cogdx1.oligodendrocytes-cogdx4.oligodendrocytes",
+                                     "cogdx2.oligodendrocytes-cogdx4.oligodendrocytes",
+                                     "cogdx1.astrocytes-cogdx2.astrocytes",
+                                     "cogdx1.astrocytes-cogdx4.astrocytes",
+                                     "cogdx2.astrocytes-cogdx4.astrocytes",
+                                     "cogdx1.microglia-cogdx2.microglia",
+                                     "cogdx1.microglia-cogdx4.microglia",
+                                     "cogdx2.microglia-cogdx4.microglia",
+                                     "cogdx1.neurons-cogdx2.neurons",
+                                     "cogdx1.neurons-cogdx4.neurons",
+                                     "cogdx2.neurons-cogdx4.neurons",
+                                     "cogdx1.endothelial-cogdx2.endothelial",
+                                     "cogdx1.endothelial-cogdx4.endothelial",
+                                     "cogdx2.endothelial-cogdx4.endothelial"),
                          levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
@@
 # Get differnetial expression
-DE = lapply(1:3, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
-  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T) %>%
+DE = lapply(1:dim(contrast)[2], function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
+  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T,
+           adjust.method = 'BH') %>%
     rownameToFirstColumn('gene.id') %>%
@@
 
-DE = DE %>% 
+DE1 = DE %>% 
   rbindlist(idcol = 'Comparison') %>%
+  tidyr::separate(Comparison, c('c1','c2'), sep = '-') %>%
+  tidyr::separate(c1, c('cond1', 'celltype'), sep = '\\.') %>%
+  tidyr::separate(c2, c('cond2', 'celltype2'), sep = '\\.') %>%
+  tidyr::unite(Comparison, cond1, cond2, sep = '_vs_') %>%
+  dplyr::select(-celltype2) %>%
   dplyr::mutate(Study = 'ROSMAP', Region = 'DLPFC',
@@
                 Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')))
+DE1$Direction = as.character(DE1$Direction)
+DE1$Direction[DE1$logFC < log2(1.2) & DE1$adj.P.Val > 0.05] = 'NONE'
+DE1$Direction[DE1$logFC > -log2(1.2) & DE1$adj.P.Val > 0.05] = 'NONE'
+DE1$Direction = factor(DE1$Direction, levels = c('DOWN','NONE','UP'))
 
-tmp = DE %>%
+writeLines('Number of differentially expressed genes at an FDR of 0.05')
+DE1 %>%
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
-  group_by(Comparison, Direction) %>%
-  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id))) %>%
-  spread(Direction, FDR_0_05)
+  dplyr::select(gene.id, Comparison, celltype, Direction) %>%
+  group_by(Comparison, celltype, Direction) %>%
+  dplyr::summarise(FDR_0_05 = length(unique(gene.id))) %>%
+  tidyr::unite('celltype.direction', celltype, Direction, sep = '.') %>%
+  spread(celltype.direction, FDR_0_05) %>%
+  kable
 
-tmp1 = DE %>%
+writeLines('Number of differentially expressed genes at an FDR of 0.05 and fold change of 1.2')
+DE1 %>%
   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
-  group_by(Comparison, Direction) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
-  spread(Direction, FDR_0_05_FC_1.2)
+  dplyr::select(gene.id, Comparison, celltype, Direction) %>%
+  group_by(Comparison, celltype, Direction) %>%
+  dplyr::summarise(FDR_0_05 = length(unique(gene.id))) %>%
+  tidyr::unite('celltype.direction', celltype, Direction, sep = '.') %>%
+  spread(celltype.direction, FDR_0_05) %>%
+  kable
 
-kable(full_join(tmp %>%
-                  dplyr::rename(DOWN_FDR_0_05 = DOWN, UP_FDR_0_05 = UP),
-                tmp1 %>%
-                  dplyr::rename(DOWN_FDR_0_05_FC_1.2 = DOWN, UP_FDR_0_05_FC_1.2 = UP)))
-
-p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
+p = ggplot(DE1 %>% filter(Comparison != 'cogdx1_vs_cogdx2'), aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point()
+p = p + scale_color_manual(values = c('green','grey','red'))
 p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
@@
 p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
-p = p + facet_grid(.~Comparison, scales = 'fixed')
+p = p + facet_grid(Comparison~.+celltype, scales = 'free_x')
 p
@@
 # Code
-CODE <- Folder(name = "ROSMAP Reprocessed RNASeq TMM (Based on cogdx)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP Reprocessed RNASeq TMM WithCellTypeFractionControlled (Based on cogdx)", parentId = parentId)
 annotations(CODE) = list(
@@
 
-  normalizationType	= 'TMM'
+  normalizationType	= 'TMM and CellType'
 )
@@
 # Store design matrix
-write.table(DM1$design, file = 'ROSMAP_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
+write.table(DESIGN.MATRIX, file = 'ROSMAP_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
 DM_OBJ = File('ROSMAP_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
@@
 # Store differential expression results
-write.table(DE, file = 'ROSMAP_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+write.table(DE1, file = 'ROSMAP_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
 DEXP_OBJ = File('ROSMAP_DLPFC_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)

### 2017-02-13 code/Rmd/Mayo_RNASeq_Reprocessed_TMM_CellTypeFracControl.Rmd
Added markdown to process reprocessed RNASeq from Mayo samples controlled for cell type fraction with TMM normalisation

--- 2017-01-30 code/Rmd/Mayo_RNASeq_Reprocessed_TMM.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed_TMM.Rmd",
+knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed_TMM_CellTypeFracControl.Rmd",
                                  parentId = "syn5570325",
-                                 entityName = 'UFL_Mayo_ISB Reprocessed RNASeq TMM')
+                                 entityName = 'UFL_Mayo_ISB Reprocessed RNASeq TMM WithCellTypeFractionControlled')
 ```
@@
 parentId = 'syn5570325';
-activityName = 'Covariate analysis';
+activityName = 'Covariate analysis with celltype controlled';
 activityDescription = 'Covariate analysis of UFL_Mayo_ISB RNASeq data';
 
-thisFileName <- 'Mayo_RNASeq_Reprocessed_TMM.Rmd'
+thisFileName <- 'Mayo_RNASeq_Reprocessed_TMM_CellTypeFracControl.Rmd'
 
@@
   as.data.frame
+
+# Get deconvoluted cell type fractions of MAYO samples from synapse 
+CELL_TYPE_FRAC = downloadFile('syn8135250') %>%
+  dplyr::select(-OPC, -`P-value`, -Correlation, -RMSE)
+rownames(CELL_TYPE_FRAC) = CELL_TYPE_FRAC$SampleID
+CELL_TYPE_FRAC$SampleID = NULL
+CELL_TYPE_FRAC = data.matrix(CELL_TYPE_FRAC)
 ```
@@
 METADATA = METADATA[indToRetain,]
+
+CELL_TYPE_FRAC = CELL_TYPE_FRAC[indToRetain,]
 ```
-
 `r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project.
@@
 COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
+
+# Add cell type fractions to the covariates
+COVARIATES = cbind(COVARIATES, CELL_TYPE_FRAC[rownames(COVARIATES),])
 ```
@@
 ```
+cell type fractions
+```{r cell.frac, fig.height=6, fig.width=12}
+# Cell type fractions
+p = CELL_TYPE_FRAC %>%
+  rownameToFirstColumn('SampleID') %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID') %>%
+              dplyr::select(BrainRegion.Diagnosis, SampleID)) %>%
+  tidyr::gather(CellType, Fraction, -SampleID, -BrainRegion.Diagnosis) %>%
+  ggplot(aes(x = BrainRegion.Diagnosis, y = Fraction, color = CellType))
+p = p + geom_boxplot()
+p
+```
+
 ### Gene Filtering
@@
 ```
-### Normalisation (iterative)
-Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
+### Normalisation (custom design)
+Since many covariates are correlated, re-normalising COVARIATES with a custom design matrix. Here BrainRegion.Diagnosis.Celltype is chose as the primary variable of interest.
+
+Final model:
+    Gx = Source + FLOWCELL + Gender + RIN + PMI + BrainRegion.Diagnosis.oligodendrocytes + BrainRegion.Diagnosis.astrocytes + BrainRegion.Diagnosis.microglia + BrainRegion.Diagnosis.neurons + BrainRegion.Diagnosis.endothelial + (1/Donor_ID)
+
 ```{r iterative.adjusted.voom.normalisation, results='asis'}
-postAdjustCovars = c('BrainRegion.Diagnosis', 'Source', 'FLOWCELL', 'Gender')
+postAdjustCovars = c('BrainRegion.Diagnosis', 'Source', 'FLOWCELL', 'Gender', 'RIN', 'PMI')
 randomCovariates = 'Donor_ID'
 
-# Assign residual covariates
-residualSigCovars = preAdjustedSigCovars
-significantCovars = residualSigCovars$significantCovars
-covariatesEffects = residualSigCovars$Effects.significantCovars
-covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates))]
+# Post adjusted design matrix
+DESIGN.MATRIX = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
 
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
+ind.design = grep('BrainRegion.Diagnosis', colnames(DESIGN.MATRIX), value = T)
+ind.celltypes = colnames(CELL_TYPE_FRAC)
+for(i in ind.celltypes){
+  for(j in ind.design){
+    tmp = DESIGN.MATRIX[, j, drop = F] * COVARIATES[,i, drop = F]
+    colnames(tmp) = paste0(j,'.',i)
+    DESIGN.MATRIX = cbind(tmp, DESIGN.MATRIX)
+  }
+}
+DESIGN.MATRIX = DESIGN.MATRIX[, setdiff(colnames(DESIGN.MATRIX), ind.design)]
+DESIGN.MATRIX = DESIGN.MATRIX[,linColumnFinder(DESIGN.MATRIX)$indepCols]
 
-loopCount = 0 
-while(length(significantCovars)!=0 && loopCount <= 100){
 writeLines(paste('Using following covariates in the model:', 
-                   paste(postAdjustCovars, collapse=', '),
-                   'as fixed effects'))
-
-  # Post adjusted design matrix
-  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
+                 paste(colnames(DESIGN.MATRIX), collapse=', '),
+                 'as fixed effects and Donor_ID as random effect'))
 
 # Estimate voom weights
-  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DESIGN.MATRIX, plot=F)
 
@@
 # Re-estimate voom weights
-  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F, 
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DESIGN.MATRIX, plot=F, 
                                      block = COVARIATES$Donor_ID, correlation = correlation$cor)
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c(
-  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.AD",
-  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.Pathologic_Aging",
-  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.PSP",
-  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.AD",
-  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.PSP",
-  "BrainRegion.DiagnosisCER.PSP-BrainRegion.DiagnosisCER.AD",
-  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.AD",
-  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.Pathologic_Aging",
-  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.PSP",
-  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.AD",
-  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.PSP",
-  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD"),
-  levels = colnames(FIT$coefficients))
+contrast = c()
+for (region in unique(METADATA$BrainRegion)){
+  for (celltype in ind.celltypes){
+    contrast = cbind(contrast, 
+                     apply(combn(unique(METADATA$Diagnosis),2),2,paste, celltype, sep = '.') %>%
+                       apply(2, function(x,y){paste(y,x,sep = '.')}, region))
+  }
+}
+contrast = apply(contrast, 2, function(x,y){paste0(y,x)}, 'BrainRegion.Diagnosis') %>%
+  apply(2, paste, collapse = '-')
+
+contrast = makeContrasts(contrasts = contrast, levels = colnames(FIT$coefficients))
 FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
@@
 # Get differnetial expression
-DE = lapply(1:12, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
-  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T)
+DE = lapply(1:dim(contrast)[2], function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
+  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T) %>%
+    rownameToFirstColumn('gene.id') %>%
+    left_join(backgroundGenes) %>%
+    left_join(Ensemble2HGNC) %>%
+    dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, logFC, CI.L, CI.R, AveExpr, t, P.Value, adj.P.Val, B)
 }, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
@@
 DE1 = plyr::ldply(DE, .id = 'Comparison') %>%
-  dplyr::mutate(Comparison1 = Comparison) %>%
-  tidyr::separate(Comparison1, c('from.state','to.state'), sep = '-') %>%
-  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
-  tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
-  tidyr::unite(Name, reference, against, sep = '-') %>%
-  dplyr::mutate(Comparison = Name, Study = 'MAYO') %>%
-  dplyr::rename(gene.id = genes, Region = BrainRegion) %>%
-  dplyr::select(-Name) %>%
-  left_join(backgroundGenes) %>%
-  left_join(Ensemble2HGNC)
+  tidyr::separate(Comparison, c('c1','c2'), sep = '-') %>%
+  tidyr::separate(c1, c('Region', 'cond1', 'celltype'), sep = '\\.') %>%
+  tidyr::separate(c2, c('Region2', 'cond2', 'celltype2'), sep = '\\.') %>%
+  tidyr::unite(Comparison, cond1, cond2, sep = '_vs_') %>%
+  dplyr::select(-celltype2, - Region2) %>%
+  dplyr::mutate(Study = 'MAYO',
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')))
+DE1$Direction = as.character(DE1$Direction)
+DE1$Direction[DE1$logFC < log2(1.2) & DE1$adj.P.Val > 0.05] = 'NONE'
+DE1$Direction[DE1$logFC > -log2(1.2) & DE1$adj.P.Val > 0.05] = 'NONE'
+DE1$Direction = factor(DE1$Direction, levels = c('DOWN','NONE','UP'))
 
-p <- ggplot(DE1, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
-p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
-p <- p + geom_vline(xintercept = log2(1.2), color = 'red')
-p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
-p <- p + facet_grid(Region~.+Comparison)
-p
+writeLines('Number of differentially expressed genes at an FDR of 0.05')
+DE1 %>%
+  dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(gene.id, Comparison, Region, celltype, Direction) %>%
+  group_by(Comparison, Region, celltype, Direction) %>%
+  dplyr::summarise(FDR_0_05 = length(unique(gene.id))) %>%
+  tidyr::unite('region.celltype.direction', Region, celltype, Direction, sep = '.') %>%
+  spread(region.celltype.direction, FDR_0_05) %>%
+  kable
 
-tmp = DE1 %>% 
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  group_by(Region, Comparison) %>%
-  dplyr::summarize(nDiffExpGenes = length(unique(gene.id))) %>%
-  spread(Region, nDiffExpGenes)
-kable(tmp)
+writeLines('Number of differentially expressed genes at an FDR of 0.05 and fold change of 1.2')
+DE1 %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
+  dplyr::select(gene.id, Comparison, Region, celltype, Direction) %>%
+  group_by(Comparison, Region, celltype, Direction) %>%
+  dplyr::summarise(FDR_0_05 = length(unique(gene.id))) %>%
+  tidyr::unite('region.celltype.direction', Region, celltype, Direction, sep = '.') %>%
+  spread(region.celltype.direction, FDR_0_05) %>%
+  kable
+
+pl = list()
+for (comp in unique(DE1$Comparison)){
+  p = ggplot(DE1 %>% filter(Comparison == comp), aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point()
+  p = p + scale_color_manual(values = c('green','grey','red'))
+  p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
+  p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
+  p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
+  p = p + facet_grid(Comparison+Region~.+celltype, scales = 'free_x')
+  pl[[comp]] = p
+}
+
+pl$AD_vs_Control
+
+pl$PSP_vs_AD
+
+pl$AD_vs_Pathologic_Aging
 ```
@@
 # Code
-CODE <- Folder(name = 'UFL_Mayo_ISB Reprocessed RNASeq TMM', parentId = parentId)
+CODE <- Folder(name = 'UFL_Mayo_ISB Reprocessed RNASeq TMM WithCellTypeFractionControlled', parentId = parentId)
 annotations(CODE) = list(

### 2017-02-14 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_BinnedCDR_CellTypeFracControl.Rmd
Added markdown to process reprocessed RNASeq from MSSM samples controlled for cell type fraction with TMM normalisation

--- 2017-01-25 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM_BinnedCDR_CellTypeFracControl.Rmd",
                                  parentId = "syn5570248",
-                                 entityName = 'MSBB Reprocessed RNASeq TMM Binned CDR')
+                                 entityName = 'MSBB Reprocessed RNASeq TMM Binned CDR WithCellTypeFractionControlled')
 ```
@@
 parentId = 'syn5570248';
-activityName = 'Covariate analysis';
+activityName = 'Covariate analysis with celltype controlled';
 activityDescription = 'Covariate analysis of MSBB RNASeq data';
 
-thisFileName <- 'MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd'
+thisFileName <- 'MSSM_RNASeq_Reprocessed_TMM_BinnedCDR_CellTypeFracControl.Rmd'
 
@@
 METADATA <- full_join(METADATA_TECH, METADATA_CLINICAL)
+
+# Get deconvoluted cell type fractions of MSBB samples from synapse 
+CELL_TYPE_FRAC = downloadFile('syn8146376') %>%
+  dplyr::select(-OPC, -`P-value`, -Correlation, -RMSE)
+rownames(CELL_TYPE_FRAC) = CELL_TYPE_FRAC$SampleID
+CELL_TYPE_FRAC$SampleID = NULL
+CELL_TYPE_FRAC = data.matrix(CELL_TYPE_FRAC)
 ```
@@
 ```{r covariates.clustering}
-FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrodmannArea.Dx")
-ContCovariates <- c("RIN", "RIN2", "AOD","PMI", "TotalReads", "Mapped", "rRNA.rate")
+FactorCovariates <- c("individualIdentifier","batch", "RACE", "SEX", "BrodmannArea.Dx")
+ContCovariates <- c("RIN", "RIN2", "AOD","PMI")
 
@@
 ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.4, row.width=0.1)
+```
+Cell type fractions
+```{r cell.frac, fig.height=6, fig.width=12}
+# Cell type fractions
+p = CELL_TYPE_FRAC %>%
+  rownameToFirstColumn('SampleID') %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID') %>%
+              dplyr::select(BrodmannArea.Dx, SampleID)) %>%
+  tidyr::gather(CellType, Fraction, -SampleID, -BrodmannArea.Dx) %>%
+  ggplot(aes(x = BrodmannArea.Dx, y = Fraction, color = CellType))
+p = p + geom_boxplot()
+p
 ```
@@
 ```
-### Normalisation (iterative)
-Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
-```{r iterative.adjusted.voom.normalisation, results='asis'}
-postAdjustCovars = c('BrodmannArea.Dx', 'batch', 'SEX')
+### Normalisation (custom design)
+Since many covariates are correlated, re-normalising COVARIATES with a custom design matrix. Here BrodmannArea.Dx.Celltype is chose as the primary variable of interest.
+
+Final model:
+    Gx = batch + SEX + RIN + PMI + AOD + BrodmannArea.Dx.oligodendrocytes + BrodmannArea.Dx.astrocytes + BrodmannArea.Dx.microglia + BrodmannArea.Dx.neurons + BrodmannArea.Dx.endothelial + (1/individualIdentifier)
+
+```{r custom.design, results='asis'}
+postAdjustCovars = c('BrodmannArea.Dx', 'batch', 'SEX', 'RIN', 'PMI', 'AOD')
 randomCovariates = 'individualIdentifier'
 
-# Assign residual covariates
-residualSigCovars = preAdjustedSigCovars
-significantCovars = residualSigCovars$significantCovars
-covariatesEffects = residualSigCovars$Effects.significantCovars
-covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates))]
+# Post adjusted design matrix
+DESIGN.MATRIX = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
 
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
+ind.design = grep('BrodmannArea.Dx', colnames(DESIGN.MATRIX), value = T)
+ind.celltypes = colnames(CELL_TYPE_FRAC)
+for(i in ind.celltypes){
+  for(j in ind.design){
+    tmp = DESIGN.MATRIX[, j, drop = F] * COVARIATES[,i, drop = F]
+    colnames(tmp) = paste0(j,'.',i)
+    DESIGN.MATRIX = cbind(tmp, DESIGN.MATRIX)
+  }
+}
+DESIGN.MATRIX = DESIGN.MATRIX[, setdiff(colnames(DESIGN.MATRIX), ind.design)]
+DESIGN.MATRIX = DESIGN.MATRIX[,linColumnFinder(DESIGN.MATRIX)$indepCols]
 
-loopCount = 0 
-while(length(significantCovars)!=0 && loopCount <= 100){
 writeLines(paste('Using following covariates in the model:', 
-                   paste(postAdjustCovars, collapse=', '),
-                   'as fixed effects and individualIdentifier as random effects'))
-
-  # Post adjusted design matrix
-  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
+                 paste(colnames(DESIGN.MATRIX), collapse=', '),
+                 'as fixed effects and individualIdentifier as random effect'))
 
 # Estimate voom weights
-  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DESIGN.MATRIX, plot=F)
 
@@
 # Re-estimate voom weights
-  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F, 
+VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DESIGN.MATRIX, plot=F, 
                                      block = COVARIATES$individualIdentifier, 
@@
                                               isKeyPlot=TRUE)
+```
 
-  # Add postadjusted covariates (if any)
-  significantCovars = residualSigCovars$significantCovars
-  covariatesEffects = residualSigCovars$Effects.significantCovars
-  covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars, randomCovariates))]
-
-  postAdjustCovars = c(postAdjustCovars,names(which.max(covariatesEffects)))
-  loopCount = loopCount + 1
-}
-modelStr <-paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=','), 'as fixed effects')
-
-tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
-```
-`r tmp`
 ### Sanity Check
@@
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
-  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
+  t(data.matrix(FIT$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
 ```
@@
 # Fit contrast
-cntr = c()
-for (BA in unique(METADATA$BrodmannArea)){
-  tmp = combn(paste0('BrodmannArea.Dx', BA, '.', c('ND','MD','SD')),2)
-  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
+contrast = c()
+for (region in unique(METADATA$BrodmannArea)){
+  for (celltype in ind.celltypes){
+    contrast = cbind(contrast, 
+                     apply(combn(c('ND', 'MD', 'SD'),2),2,paste, celltype, sep = '.') %>%
+                       apply(2, function(x,y){paste(y,x,sep = '.')}, region))
   }
+}
+contrast = apply(contrast, 2, function(x,y){paste0(y,x)}, 'BrodmannArea.Dx') %>%
+  apply(2, paste, collapse = '-')
+contrast = makeContrasts(contrasts=contrast, levels = colnames(FIT$coefficients))
 
-contrast = makeContrasts(contrasts=cntr,
-                         levels = colnames(FIT$coefficients))
 FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
@@
 DE1 = plyr::ldply(DE, .id = 'Comparison') %>%
-  dplyr::mutate(Comparison1 = Comparison) %>%
-  tidyr::separate(Comparison1, c('from.state','to.state'), sep = '-') %>%
-  tidyr::separate(from.state, c('BrodmannArea','From.Dx'), sep = 5) %>%
-  tidyr::separate(to.state, c('BrodmannArea','To.Dx'), sep = 5) %>%
-  tidyr::unite(Dx, From.Dx, To.Dx, sep = '-')
+  tidyr::separate(Comparison, c('c1','c2'), sep = '-') %>%
+  tidyr::separate(c1, c('Region', 'cond1', 'celltype'), sep = '\\.') %>%
+  tidyr::separate(c2, c('Region2', 'cond2', 'celltype2'), sep = '\\.') %>%
+  tidyr::unite(Comparison, cond1, cond2, sep = '_vs_') %>%
+  dplyr::select(-celltype2, - Region2) %>%
+  dplyr::mutate(Study = 'MSBB',
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')))
+DE1$Direction = as.character(DE1$Direction)
+DE1$Direction[DE1$logFC < log2(1.2) & DE1$adj.P.Val > 0.05] = 'NONE'
+DE1$Direction[DE1$logFC > -log2(1.2) & DE1$adj.P.Val > 0.05] = 'NONE'
+DE1$Direction = factor(DE1$Direction, levels = c('DOWN','NONE','UP'))
 
-p <- ggplot(DE1, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
-p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
-p <- p + geom_vline(xintercept = log2(1.2), color = 'red')
-p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
-p <- p + facet_grid(BrodmannArea~.+Dx)
-p
+writeLines('Number of differentially expressed genes at an FDR of 0.05')
+DE1 %>%
+  dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(gene.id, Comparison, Region, celltype, Direction) %>%
+  group_by(Comparison, Region, celltype, Direction) %>%
+  dplyr::summarise(FDR_0_05 = length(unique(gene.id))) %>%
+  tidyr::unite('region.celltype.direction', Region, celltype, Direction, sep = '.') %>%
+  spread(region.celltype.direction, FDR_0_05) %>%
+  kable
 
-tmp = DE1 %>% 
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  group_by(BrodmannArea, Dx) %>%
-  summarise(count = length(unique(genes))) %>%
-  tidyr::spread(BrodmannArea, count)
-kable(tmp)
+writeLines('Number of differentially expressed genes at an FDR of 0.05 and fold change of 1.2')
+DE1 %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
+  dplyr::select(gene.id, Comparison, Region, celltype, Direction) %>%
+  group_by(Comparison, Region, celltype, Direction) %>%
+  dplyr::summarise(FDR_0_05 = length(unique(gene.id))) %>%
+  tidyr::unite('region.celltype.direction', Region, celltype, Direction, sep = '.') %>%
+  spread(region.celltype.direction, FDR_0_05) %>%
+  kable
+
+pl = list()
+for (comp in unique(DE1$Comparison)){
+  p = ggplot(DE1 %>% filter(Comparison == comp), aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point()
+  p = p + scale_color_manual(values = c('green','grey','red'))
+  p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
+  p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
+  p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
+  p = p + facet_grid(Comparison+Region~.+celltype, scales = 'free_x')
+  p = p + theme(legend.position = 'top')
+  pl[[comp]] = p
+}
+
+pl$ND_vs_SD
+
+pl$ND_vs_MD
+
+pl$MD_vs_SD
 ```
@@
 # Code
-CODE <- Folder(name = 'MSBB Reprocessed RNASeq TMM Binned CDR', parentId = parentId)
+CODE <- Folder(name = 'MSBB Reprocessed RNASeq TMM Binned CDR WithCellTypeFractionControlled', parentId = parentId)
 annotations(CODE) = list(

### 2017-02-21 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd
Minor modifications

@@
 COVARIATES.CORRELATION = CovariateAnalysis::getAssociationStatistics(COVARIATES, PVAL = 0.05)
-ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.4, row.width=0.1)
+COVARIATES.CORRELATION$plot
 ```

### 2017-02-21 code/Rmd/Mayo_RNASeq_Reprocessed_TMM.Rmd
Minor modifications

@@
 
-plotdata.cer = plotdata[grep('CER', plotdata$ID),]
-p <- ggplot(plotdata.cer, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Tissue), size=as.numeric(AgeAtDeath)))
-p <- p + theme_bw() + theme(legend.position="right") +scale_shape_manual(values = 1:2)
+plotdata[grep('CER', plotdata$ID),'Region'] = 'CER'
+plotdata[grep('TCX', plotdata$ID),'Region'] = 'TCX'
+plotdata$Region = factor(plotdata$Region)
+plotdata$Source = factor(plotdata$Source)
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(shape = Source, color = Region))
+p <- p + theme_bw() + theme(legend.position = "left") +scale_shape_manual(values = 1:2)
+p <- p + facet_grid(Region~.)
 # p <- p + geom_text(aes(label= ID), size=4, hjust=0)
-pl[[1]] <- p
-
-plotdata.tcx = plotdata[grep('TCX', plotdata$ID),]
-p <- ggplot(plotdata.tcx, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Tissue), size=as.numeric(AgeAtDeath)))
-p <- p + theme_bw() + theme(legend.position="right") +scale_shape_manual(values = 1:2)
-# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
-pl[[2]] <- p
-
-multiplot(plotlist = pl, cols = 1)
+p
 ```

### 2017-02-21 code/Rmd/combineResults.Rmd
Added markdown to combine resulat and perform enrichment analysis of pathways

---
title: "Combine differential expression and celltype fraction results from all brain banks"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)
synapseLogin()

knit2synapse::knitToFolderEntity(file = "./combineResults.Rmd", 
                                 parentId ="syn5569102",
                                 entityName = 'All Differential Expression Results')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file is

# Clear R console screen output
cat("\014")

# Load required libraries
library(CovariateAnalysis)
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(RColorBrewer)
library(ggplot2)
library(ComplexHeatmap)

library(RankProd)
library(githubr)

library(Vennerable)
library(colorfulVennPlot)

synapseLogin()

options(xtable.type="html")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

### Get covariates/sample metadata from synapse
```{r covariates}
covariates.ids = c(ROSMAP = 'syn8018344',
                   MSBB = 'syn8073656',
                   MAYO = 'syn8028561')
covariates = lapply(covariates.ids, function(id){
  x = downloadFile(id)
  colnames(x)[1] = 'SampleID'
  return(x)
})

covariates$ROSMAP = dplyr::filter(covariates$ROSMAP, cogdx %in% c(1,4)) %>%
  dplyr::mutate(Dx = factor(cogdx, levels = c(1,4), labels = c('1' = 'NCI', '4' = 'AD')),
                Study = 'ROSMAP',
                BrainRegion = 'DLPFC') %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)

covariates$MSBB = dplyr::filter(covariates$MSBB, BrodmannArea.Dx %in% c('BM10.SD', 'BM22.SD', 'BM36.SD', 'BM44.SD', 'BM36.ND', 'BM10.ND', 'BM44.ND', 'BM22.ND')) %>%
  dplyr::select(SampleID, BrodmannArea.Dx) %>%
  tidyr::separate(BrodmannArea.Dx, c('BrainRegion', 'Dx')) %>%
  dplyr::mutate(Study = 'MSSM',
                Dx = factor(Dx, levels = c('ND','SD')),
                BrainRegion = factor(BrainRegion, labels = c('FP', 'STG', 'PHG', 'IFG'))) %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)

covariates$MAYO = dplyr::filter(covariates$MAYO, BrainRegion.Diagnosis %in% c('TCX.AD', 'TCX.Control', 'CER.AD', 'CER.Control')) %>%
  dplyr::select(SampleID, BrainRegion.Diagnosis) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Dx')) %>%
  dplyr::mutate(Study = 'MAYO',
                Dx = factor(Dx, levels = c('Control','AD')),
                BrainRegion = factor(BrainRegion, levels = c('CER', 'TCX'))) %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)

diff.exp.ids = c(ROSMAP = 'syn8018358',
                 MSBB = 'syn8073678',
                 MAYO = 'syn8028572')
diff.exp = lapply(diff.exp.ids, downloadFile)

diff.exp$ROSMAP = filter(diff.exp$ROSMAP, Comparison == 'cogdx1-cogdx4')
diff.exp$MSBB = filter(diff.exp$MSBB, Comparison == 'ND-SD')
diff.exp$MAYO = filter(diff.exp$MAYO, Comparison == 'Control-AD')

celltype.fraction.ids = c(ROSMAP = 'syn8124520',
                      MSBB = 'syn8146376',
                      MAYO = 'syn8135250')
celltype.fraction = lapply(celltype.fraction.ids, downloadFile) 
```

```{r anlz.celltype}
cell.frac = mapply(function(cov, frc){
  x = filter(frc, SampleID %in% cov$SampleID) %>%
    dplyr::select(-`P-value`,-RMSE, -Correlation) %>%
    tidyr::gather(celltype, fraction, -SampleID) %>%
    left_join(cov)
}, covariates, celltype.fraction, SIMPLIFY = F) %>%
  rbindlist(use.names = T, fill = T) %>%
  dplyr::mutate(Dx = factor(Dx, levels = c('Control', 'NCI', 'ND', 'SD', 'AD')))

head(cell.frac)

p = ggplot(cell.frac, aes(x = BrainRegion, y = fraction, color = Dx)) + geom_boxplot()
p = p + facet_grid(celltype~., scales = 'free_y')
p = p + xlab('Brain Region') + ylab('Fraction')
p = p + theme(axis.text.x = element_text(hjust = 1, angle = 90))
p
```

```{r anlz.diff.exp}
tmp = lapply(diff.exp, function(x){
  x = filter(x, adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) 
  x$Direction = 'NONE'
  x$Direction[x$logFC > 0] = 'UP'
  x$Direction[x$logFC < 0] = 'DOWN'
  x = x %>% group_by(Study, Region, Direction) %>%
    summarise(count = length(unique(gene.id)))
}) %>%
  rbindlist() %>%
  spread(Direction, count)

tmp = diff.exp %>% 
  rbindlist(use.names = T, fill = T) 
tmp$Direction = 'NONE'
tmp$Direction[x$logFC > 0] = 'UP'
tmp$Direction[x$logFC < 0] = 'DOWN'

p = ggplot(tmp, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
p
```

### 2017-02-21 code/Rmd/enrichmentAnalysis.Rmd
Added markdown to combine resulat and perform enrichment analysis of pathways

---
title: "Enrichment analysis of differential expression results from AMP-AD (Control vs AD, reprocessed RNASeq)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)
synapseLogin()

knit2synapse::knitToFolderEntity(file = "./enrichmentAnalysis.Rmd", 
                                 parentId ="syn5569102",
                                 entityName = 'Enrichment Analysis of Control vs AD from all three brain banks (Reprocessed RNASeq)')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file is

# Clear R console screen output
cat("\014")

# Load required libraries
library(CovariateAnalysis) # install from th1vairam repo using devtools::install_github('th1vairam/CovariateAnalysis@devtools')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(RColorBrewer)
library(ggplot2)
library(ComplexHeatmap)
library(biomaRt)
library(Matrix)
library(org.Hs.eg.db)

library(limma)
library(edgeR)
library(RankProd)
library(githubr)

library(Vennerable)
library(colorfulVennPlot)

synapseLogin()

library(doParallel)
library(foreach)
library(doParallel)

clust = makeCluster(detectCores() - 2)
registerDoParallel(clust)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r fxns}
# Function to perform Fishers enrichment analysis
fisherEnrichment <- function(genesInSignificantSet, # A character vector of differentially expressed or some significant genes to test
                             genesInGeneSet, # A character vector of genes in gene set like GO annotations, pathways etc...
                             genesInBackground # Background genes that are 
){
  genesInSignificantSet = intersect(genesInSignificantSet, genesInBackground) # back ground filtering
  genesInNonSignificantSet = base::setdiff(genesInBackground, genesInSignificantSet)
  genesInGeneSet = intersect(genesInGeneSet, genesInBackground) # back ground filtering
  genesOutGeneSet = base::setdiff(genesInBackground,genesInGeneSet)

  pval = fisher.test(
    matrix(c(length(intersect(genesInGeneSet, genesInSignificantSet)),             
             length(intersect(genesInGeneSet, genesInNonSignificantSet)),
             length(intersect(genesOutGeneSet, genesInSignificantSet)),
             length(intersect(genesOutGeneSet, genesInNonSignificantSet))), 
           nrow=2, ncol=2), alternative="greater")
  OR = (length(intersect(genesInGeneSet, genesInSignificantSet)) * length(intersect(genesOutGeneSet, genesInNonSignificantSet))) / (length(intersect(genesInGeneSet, genesInNonSignificantSet)) * length(intersect(genesOutGeneSet, genesInSignificantSet))) %>% as.numeric()
  return(data.frame(pval = pval$p.value,
                    ngenes = length(genesInGeneSet),
                    noverlap = length(intersect(genesInGeneSet, genesInSignificantSet)),
                    Odds.Ratio = OR,
                    Genes = paste(intersect(genesInGeneSet, genesInSignificantSet), collapse = '|')))}
```

### Get covariates/sample metadata from synapse
```{r covariates}
covariates.ids = c(ROSMAP = 'syn8018344',
                   MSBB = 'syn8073656',
                   MAYO = 'syn8028561')
covariates = lapply(covariates.ids, function(id){
  x = downloadFile(id)
  colnames(x)[1] = 'SampleID'
  return(x)
})

covariates$ROSMAP = dplyr::filter(covariates$ROSMAP, cogdx %in% c(1,4)) %>%
  dplyr::mutate(Dx = factor(cogdx, levels = c(1,4), labels = c('1' = 'NCI', '4' = 'AD')),
                Study = 'ROSMAP',
                BrainRegion = 'DLPFC') %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)

covariates$MSBB = dplyr::filter(covariates$MSBB, BrodmannArea.Dx %in% c('BM10.SD', 'BM22.SD', 'BM36.SD', 
                                                                        'BM44.SD', 'BM36.ND', 'BM10.ND', 
                                                                        'BM44.ND', 'BM22.ND')) %>%
  dplyr::select(SampleID, BrodmannArea.Dx) %>%
  tidyr::separate(BrodmannArea.Dx, c('BrainRegion', 'Dx')) %>%
  dplyr::mutate(Study = 'MSSM',
                Dx = factor(Dx, levels = c('ND','SD')),
                BrainRegion = factor(BrainRegion, labels = c('FP', 'STG', 'PHG', 'IFG'))) %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)

covariates$MAYO = dplyr::filter(covariates$MAYO, BrainRegion.Diagnosis %in% c('TCX.AD', 'TCX.Control', 'CER.AD', 'CER.Control')) %>%
  dplyr::select(SampleID, BrainRegion.Diagnosis) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Dx')) %>%
  dplyr::mutate(Study = 'MAYO',
                Dx = factor(Dx, levels = c('Control','AD')),
                BrainRegion = factor(BrainRegion, levels = c('CER', 'TCX'))) %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)
```

### Get differential expression data from synapse
```{r diffexp}
diff.exp.ids = c(ROSMAP = 'syn8018358',
                 MSBB = 'syn8073678',
                 MAYO = 'syn8028572')
diff.exp = lapply(diff.exp.ids, downloadFile)

diff.exp$ROSMAP = filter(diff.exp$ROSMAP, Comparison == 'cogdx1-cogdx4')
diff.exp$MSBB = filter(diff.exp$MSBB, Comparison == 'ND-SD')
diff.exp$MAYO = filter(diff.exp$MAYO, Comparison == 'Control-AD')

tmp = diff.exp %>% 
  rbindlist(use.names = T, fill = T) 
tmp$Direction = 'NONE'
tmp$Direction[tmp$logFC >= log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
tmp$Direction[tmp$logFC <= -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'

writeLines('Number of differentially expressed genes at an FDR of 0.05 and fold change of 1.2 are')
tmp %>%
  group_by(Study, Region, Direction, Comparison) %>%
  summarise(count = n()) %>%
  spread(Direction, count) %>%
  kable

pl = list()
for(region in c('ROSMAP', 'MSBB', 'MAYO')){
  p = ggplot(tmp %>% filter(Study == region), aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point()
  p = p + scale_color_manual(values = c('green','grey','red'))
  p = p + facet_grid(Study~.+Region) + theme(legend.position = 'top')
  pl[[region]] = p
}
multiplot(plotlist = pl, cols = 1)
```

### Get hgnc_symbol mapping from biomaRt
```{r biomart}
# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "dec2016.archive.ensembl.org")

# Get background gene ids
backgroundGenes = sapply(diff.exp, function(x) x$ensembl_gene_id) %>%
  Reduce(c,.)

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = backgroundGenes,
                       mart = mart)

Entrez2Ensembl <- as.data.frame(org.Hs.egENSEMBL2EG)
Entrez2HGNC <- as.data.frame(org.Hs.egSYMBOL2EG)
HGNC2Ensembl <- left_join(Entrez2HGNC, Entrez2Ensembl)
HGNC2Ensembl = HGNC2Ensembl[grep('^LOC',HGNC2Ensembl$symbol),c('ensembl_id', 'symbol')] %>%
  filter(!is.na(ensembl_id), !is.na(symbol)) %>%
  dplyr::rename(ensembl_gene_id = ensembl_id, hgnc_symbol = symbol)

Ensemble2HGNC = rbindlist( list(Ensemble2HGNC, HGNC2Ensembl), use.names = T, fill = T) %>%
  unique
```

### Pathway enrichment analysis
```{r enrich.all, fig.height=16, fig.width=10}
# Find genes that are differentially expressed in each brain region
diff.exp.gl = lapply(diff.exp, function(x){
  dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05 & sign(CI.L) == sign(CI.R)) %>%
    dplyr::select(Region, hgnc_symbol) %>%
    unique
}) %>%
  rbindlist(use.names = T, fill = T) %>%
  dlply(.(Region), .fun = function(x){unique(x$hgnc_symbol)})

# Get Background genes
backgroundGenes = Ensemble2HGNC$hgnc_symbol[Ensemble2HGNC$ensembl_gene_id %in% backgroundGenes] %>% unique()

#### Get gene sets ####
# Download enrichr gene sets from synapse
GL_OBJ = synGet('syn4867851')
ALL_USED_IDs = c(ALL_USED_IDs, GL_OBJ$properties$id)
load(GL_OBJ@filePath)

gsets = c("BioCarta_2015", "GO_Biological_Process", "Reactome_2015", "TargetScan_microRNA", "WikiPathways_2015")
GeneSets.Enrichr = GeneSets[gsets]

# Download AD related gene sets from synapse
GL_OBJ = synGet('syn4893059');
ALL_USED_IDs = c(ALL_USED_IDs, GL_OBJ$properties$id)
load(GL_OBJ@filePath)

GeneSets.CM = list(CellTypeMarkers = GeneSets[grep('Zhang', names(GeneSets))],
                   AD = GeneSets['AD:GeneticLoci'])
GeneSets = c(GeneSets.Enrichr, GeneSets.CM)

# Filter gene list
GeneSets = filterGeneSets(GeneSets, backgroundGenes, minSize = 3, maxSize = 5000)

# Enrichment Analysis
enrichResults = plyr::ldply(diff.exp.gl, .fun = function(gs, geneSets, backgroundGenes, fisherEnrichment){
  tmp = plyr::ldply(geneSets, .fun = function(x, gs, backgroundGenes, fisherEnrichment){
    plyr::ldply(x, .fun = fisherEnrichment, gs, backgroundGenes, .id = 'Name', .parallel = T)
  }, gs, backgroundGenes, fisherEnrichment, .id = 'Category', .parallel = T, .paropts = list(.packages = c('dplyr'))) %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'bonferroni'))
}, GeneSets, backgroundGenes, fisherEnrichment, .parallel = T, .paropts = list(.packages = c('dplyr', 'plyr')))

# Filter results
tmp1 = filter(enrichResults, fdr <= 0.05)
```

### Store results in synapse
```{r synapse.store, include=FALSE, cache=FALSE}
thisFileName <- 'enrichmentAnalysis.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='metaAnal')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))

# Store consolidated data in synapse
CODE = Folder(name = 'Pathway Enrichment Analysis of Control vs AD Data (all brain banks)', parentId = "syn5569102")
CODE = synStore(CODE)

# Store enrichment analysis in synapse
write.table(tmp1, file = 'enrichmentResults.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('enrichmentResults.tsv', 
           name = 'Pathway Enrichment Results (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(ALL_USED_IDs),
               executed = thisFile,
               activityName = 'Pathway enrichment analysis')
stopCluster(cl)
```

### 2017-03-01 code/Rmd/MetaAnalysis.Rmd
Minor changes

@@
 covariates$ROSMAP = dplyr::filter(covariates$ROSMAP, cogdx %in% c(1,4)) %>%
-  dplyr::mutate(Dx = factor(cogdx, levels = c(1,4), labels = c('1' = 'Control', '4' = 'AD')),
+  dplyr::mutate(Dx = factor(cogdx, levels = c(1,4), labels = c('1' = 'NCI', '4' = 'AD')),
                 Study = 'ROSMAP',
@@
 
-covariates$MSBB = dplyr::filter(covariates$MSBB, BrodmannArea.Dx %in% c('BM10.SD', 'BM22.SD', 'BM36.SD', 'BM44.SD', 'BM36.ND', 'BM10.ND', 'BM44.ND', 'BM22.ND')) %>%
+covariates$MSBB = dplyr::filter(covariates$MSBB, BrodmannArea.Dx %in% c('BM10.SD', 'BM22.SD', 'BM36.SD', 
+                                                                        'BM44.SD', 'BM36.ND', 'BM10.ND', 
+                                                                        'BM44.ND', 'BM22.ND')) %>%
   dplyr::select(SampleID, BrodmannArea.Dx) %>%
@@
   dplyr::mutate(Study = 'MSSM',
-                Dx = factor(Dx, levels = c('ND','SD'), labels = c('ND' = 'Control', 'SD' = 'AD')),
+                Dx = factor(Dx, levels = c('ND','SD')),
                 BrainRegion = factor(BrainRegion, labels = c('FP', 'STG', 'PHG', 'IFG'))) %>%
@@
   dplyr::select(SampleID, Study, BrainRegion, Dx)
-
-covariates = covariates %>%
-  rbindlist(use.names = T, fill = T)
 ```
 
-### Get filtered gene.id and sample.id from synapse
-```{r counts}
-logcpm.ids = c('ROSMAP' = 'syn8018352',
-               'MSBB' = 'syn8073666',
-               'MAYO' = 'syn8028565')
+### Get differential expression data from synapse
+```{r diffexp}
+diff.exp.ids = c(ROSMAP = 'syn8018358',
+                 MSBB = 'syn8073678',
+                 MAYO = 'syn8028572')
+diff.exp = lapply(diff.exp.ids, downloadFile)
 
-logcpm = lapply(logcpm.ids, downloadFile) %>%
-  lapply(function(x){
-    x = dplyr::select(x, -hgnc_symbol, -ensembl_gene_id)
-  }) %>%
-  join_all(type = 'full') %>%
-  dplyr::select(gene.id, one_of(covariates$SampleID)) 
-logcpm[is.na(logcpm)] = min(logcpm[,-(1)], na.rm = T)
+diff.exp$ROSMAP = filter(diff.exp$ROSMAP, Comparison == 'cogdx1-cogdx4')
+diff.exp$MSBB = filter(diff.exp$MSBB, Comparison == 'ND-SD')
+diff.exp$MAYO = filter(diff.exp$MAYO, Comparison == 'Control-AD')
+
+tmp = diff.exp %>% 
+  rbindlist(use.names = T, fill = T) 
+tmp$Direction = 'NONE'
+tmp$Direction[tmp$logFC >= log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
+tmp$Direction[tmp$logFC <= -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'
+
+writeLines('Number of differentially expressed genes at an FDR of 0.05 and fold change of 1.2 are')
+tmp %>%
+  group_by(Study, Region, Direction, Comparison) %>%
+  summarise(count = n()) %>%
+  spread(Direction, count) %>%
+  kable
+
+pl = list()
+for(region in c('ROSMAP', 'MSBB', 'MAYO')){
+  p = ggplot(tmp %>% filter(Study == region), aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point()
+  p = p + scale_color_manual(values = c('green','grey','red'))
+  p = p + facet_grid(Study~.+Region) + theme(legend.position = 'top')
+  pl[[region]] = p
+}
+multiplot(plotlist = pl, cols = 1)
 ```
 
-### Get raw counts from synapse
-```{r raw.counts, eval = F}
-raw.counts.ids = c('MAYO_CBE' = 'syn7513807',
-                   'MSBB' = 'syn7994853',
-                   'MAYO_TCX' = 'syn7987765',
-                   'ROSMAP' = 'syn7750270')
+### Get cell type fraction from synapse
+```{r celltype.frac}
+celltype.fraction.ids = c(ROSMAP = 'syn8124520',
+                          MSBB = 'syn8146376',
+                          MAYO = 'syn8135250')
+celltype.fraction = lapply(celltype.fraction.ids, downloadFile)
 
-lcpm = lapply(raw.counts.ids, function(id, lcpm){
-  expr = read.table(synGet(id)@filePath, sep = '\t', check.names = F) 
-  lib.size = calcNormFactors(expr, method = 'TMM')
-  expr = voom(expr, lib.size = lib.size, design = NULL, plot = F, normalize.method = 'quantile')$E
-  expr = expr[row.names(expr) %in% lcpm$gene.id, ] %>%
-    rownameToFirstColumn('gene.id')
-}, logcpm) %>%
-  join_all(type = 'full')
-```
+cell.frac = mapply(function(cov, frc){
+  x = filter(frc, SampleID %in% cov$SampleID) %>%
+    dplyr::select(-`P-value`,-RMSE, -Correlation) %>%
+    tidyr::gather(celltype, fraction, -SampleID) %>%
+    left_join(cov)
+}, covariates, celltype.fraction, SIMPLIFY = F) %>%
+  rbindlist(use.names = T, fill = T) %>%
+  dplyr::mutate(Dx = factor(Dx, levels = c('Control', 'NCI', 'ND', 'SD', 'AD')))
 
-### Get residual expression from synapse
-```{r resid.expr}
-# Get SVA adjusted data
-resid.expr.id = c('ROSMAP' = 'syn6129372',
-                  'MSBB' = 'syn7059289',
-                  'MAYO' = 'syn6128634')
+pl = list()
 
-resid.expr = lapply(resid.expr.id, downloadFile) %>%
-  lapply(function(x){
-    x = dplyr::select(x, -hgnc_symbol)
-    }) %>%
-  join_all(type = 'full') %>%
-  dplyr::select(Gene.ID, ensembl_gene_id, one_of(covariates$ID)) 
-resid.expr[is.na(resid.expr)] = dplyr::select(resid.expr, -Gene.ID, -ensembl_gene_id) %>%
-  min(na.rm=T)
+p = ggplot(cell.frac %>% filter(celltype %in% c('neurons', 'microglia', 'astrocytes')), 
+           aes(x = BrainRegion, y = fraction, color = Dx)) + geom_boxplot()
+p = p + facet_grid(celltype~., scales = 'free_y')
+p = p + xlab('Brain Region') + ylab('Fraction')
+p = p + theme(axis.text.x = element_text(hjust = 1, angle = 90), legend.position = 'top')
+pl[[1]] = p
+
+p = ggplot(cell.frac %>% filter(celltype %in% c('endothelial', 'oligodendrocytes', 'OPC')), 
+           aes(x = BrainRegion, y = fraction, color = Dx)) + geom_boxplot()
+p = p + facet_grid(celltype~., scales = 'free_y')
+p = p + xlab('Brain Region') + ylab('Fraction')
+p = p + theme(axis.text.x = element_text(hjust = 1, angle = 90), legend.position = 'top')
+pl[[2]] = p
+
+multiplot(plotlist = pl, cols = 2)
 ```
@@
 # Get background gene ids
-backgroundGenes = logcpm[,'gene.id', drop = F] %>% 
-  dplyr::mutate(ID = gene.id) %>%
-  tidyr::separate(ID, c('ensembl_gene_id','position'), sep = '\\.')
+backgroundGenes = sapply(diff.exp, function(x) x$ensembl_gene_id) %>%
+  Reduce(c,.)
 
@@
 Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
-                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
+                       filters = "ensembl_gene_id", values = backgroundGenes,
                        mart = mart)
@@
 
-Ensemble2HGNC = left_join(backgroundGenes,
-                          rbindlist(list(Ensemble2HGNC, HGNC2Ensembl), use.names = T, fill = T) %>%
-                            unique) %>%
+Ensemble2HGNC = rbindlist( list(Ensemble2HGNC, HGNC2Ensembl), use.names = T, fill = T) %>%
   unique
@@
 
-### Get differential expression results from synapse
-```{r diffexp, fig.height=16, fig.width=10}
-diff.exp.ids = c(ROSMAP = 'syn8018358',
-                 MSBB = 'syn8073678',
-                 MAYO = 'syn8028572')
-diff.exp = lapply(diff.exp.ids, downloadFile)
+### Consistently differentially expressed genes (in all 7 brain regions)
+```{r diffexp.all, fig.height=16, fig.width=10}
+# Find genes that are differentially expressed in each brain region
+diff.exp.gl = lapply(diff.exp, function(x){
+  dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05 & sign(CI.L) == sign(CI.R)) %>%
+    dplyr::select(Region, gene.id) %>%
+    unique
+}) %>%
+  rbindlist(use.names = T, fill = T) %>%
+  dlply(.(Region), .fun = function(x){x$gene.id})
 
-# Filter case-control differences alone
-diff.exp$ROSMAP = filter(diff.exp$ROSMAP, Comparison == 'cogdx1-cogdx4')
-diff.exp$MSBB = filter(diff.exp$MSBB, Comparison == 'ND-SD')
-diff.exp$MAYO = filter(diff.exp$MAYO, Comparison == 'Control-AD')
+tmp1 = rbindlist(diff.exp, use.names = T, fill = T) %>%
+  dplyr::filter(gene.id %in% Reduce(intersect, diff.exp.gl)) %>%
+  dplyr::select(-hgnc_symbol) %>%
+  left_join(Ensemble2HGNC)
+tmp1$hgnc_symbol[tmp1$ensembl_gene_id == 'ENSG00000275830'] = 'LOC101927646'
 
-writeLines('Total number of genes differentially expressed between Control-AD at an FDR of 0.05 and foldchange of 1.2 are')
-lapply(diff.exp, function(x){
-  x = filter(x, adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) 
-  x$Direction = 'NONE'
-  x$Direction[x$logFC > 0] = 'UP'
-  x$Direction[x$logFC < 0] = 'DOWN'
-  x = x %>% group_by(Study, Region, Direction) %>%
-    summarise(count = length(unique(gene.id)))
-}) %>%
-  rbindlist() %>%
-  spread(Direction, count) %>%
-  kable
+tmp = group_by(tmp1, gene.id) %>%
+  summarise(count = sum(logFC/abs(logFC))) %>%
+  filter(abs(count) == 7) %>%
+  dplyr::mutate(GeneID = gene.id) %>%
+  tidyr::separate(gene.id, c('ensembl_gene_id','position'), sep = '\\.') %>%
+  left_join(Ensemble2HGNC)
+writeLines(paste('Differentially expressed genes in all 7 brain regions at an FDR of 0.05 and fold change of 1.2 are',
+                 length(unique(tmp$GeneID))))
+kable(data.frame(ngenes = length(unique(tmp$GeneID)),
+                 hgnc_symbol = tmp$hgnc_symbol %>% unique %>% paste(collapse = ',')))
 
-tmp = diff.exp %>% 
-  rbindlist(use.names = T, fill = T) 
-tmp$Direction = 'NONE'
-tmp$Direction[tmp$logFC > log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
-tmp$Direction[tmp$logFC < -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'
+writeLines('Top six differentially expressed genes are')
+tmp.eff.sz = group_by(tmp1, gene.id) %>% 
+  summarise(es = sum(logFC, na.rm = T)) %>% 
+  top_n(6, abs(es))
 
 pl = list()
-for (study in c('ROSMAP','MSBB','MAYO')){
-  p = ggplot(tmp %>% filter(Study == study) %>% droplevels, aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point()
-  p = p + scale_color_manual(values = c('green','grey','red'))
-  p = p + facet_grid(Study~.+Region) + theme_bw() + theme(legend.position = 'top')
-  pl[[study]] = p
-}
-multiplot(plotlist = pl, cols = 1)
-
-# Find genes that are differentially expressed in all brain regions
-diff.exp.gl = dlply(tmp, .(Region), .fun = function(x){
-  dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05) %>%
-    dplyr::select(gene.id) %>%
-    unique %>% unlist
-})
-
-tmp1 = filter(tmp, gene.id %in% Reduce(intersect, diff.exp.gl)) %>%
-  dplyr::select(-ensembl_gene_id, -hgnc_symbol) %>%
-  left_join(Ensemble2HGNC %>% dplyr::select(-position))
-tmp1$hgnc_symbol[tmp1$ensembl_gene_id == 'ENSG00000275830'] = 'LOC101927646'
-
-tmp.eff.sz = group_by(tmp1, gene.id) %>% summarise(es = sum(logFC, na.rm = T)) %>% top_n(6, abs(es))
-
-writeLines('Top six differentially expressed genes are')
-pl = list()
-p = ggplot(tmp1 %>% filter(gene.id %in% tmp.eff.sz$gene.id[tmp.eff.sz$es>0]), aes(x = Region, y = logFC, fill = Study)) + geom_bar(stat = 'identity')
+p = ggplot(tmp1 %>% 
+             filter(gene.id %in% tmp.eff.sz$gene.id[tmp.eff.sz$es>0]), 
+           aes(x = Region, y = logFC, fill = Study)) + geom_bar(stat = 'identity')
 p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
@@
 multiplot(plotlist = pl, cols = 2)
-
-# Genes differentially expressed in all brain regions
-tmp1 = dlply(tmp, .(Region), .fun = function(x){
-  dplyr::filter(x, abs(logFC) >= log2(1) & adj.P.Val <= 0.05) %>%
-    dplyr::select(gene.id) %>%
-    unique %>% unlist
-}) %>%
-  Reduce(intersect, .)
-writeLines('Genes that are differentially expressed between Control-AD at FDR 0.05 in all brain regions')
-kable(data.frame(ngenes = length(tmp1),
-                 hgnc_symbol = paste(unique(Ensemble2HGNC$hgnc_symbol[Ensemble2HGNC$gene.id %in% tmp1]), collapse = ', ')))
-
-tmp1 = dlply(tmp, .(Region), .fun = function(x){
-  dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05) %>%
-    dplyr::select(gene.id) %>%
-    unique %>% unlist
-}) %>%
-  Reduce(intersect, .)
-writeLines('Genes that are differentially expressed between Control-AD at FDR 0.05 and foldchange of 1.2 in all brain regions')
-kable(data.frame(ngenes = length(tmp1),
-                 hgnc_symbol = paste(unique(Ensemble2HGNC$hgnc_symbol[Ensemble2HGNC$gene.id %in% tmp1]), collapse = ', ')))
 ```

### 2017-03-01 code/Rmd/enrichmentAnalysis.Rmd
Minor changes

@@
 
-```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
+```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
 ## It is assumed your working directory is where this file is
@@
 library(Vennerable)
-library(colorfulVennPlot)
 
@@
 tmp1 = filter(enrichResults, fdr <= 0.05)
+
+# Plot results
+tmp = enrichResults %>%
+  dplyr::filter(Name %in% tmp1$Name) %>%
+  dplyr::select(Region, Name, fdr) %>% 
+  tidyr::spread(Region, fdr)
+rownames(tmp) = tmp$Name
+tmp = data.matrix(tmp[,-(1)])
+tmp[is.na(tmp)] = 1
+tmp = tmp[setdiff(1:dim(tmp)[1], grep('Mus', rownames(tmp))),]
+
+Heatmap(tmp, name = 'fdr', show_row_names = F)
+
+# In any 3
+h = Heatmap(tmp[rowSums(tmp <= 0.05, na.rm = T) >= 3,], name = 'fdr',
+        col = colorRamp2(c(0,1), c('red', 'white')),
+        show_row_dend = F, row_names_side = 'right',
+        row_names_gp = gpar(fontsize = 8))
+draw(h, heatmap_legend_side = 'left')
+
+# Overlap with AMP-AD targets
+targets = downloadFile('syn7555804') %>%
+  dplyr::select(GENE_SYMBOL) %>%
+  unlist() %>% unique()
+
+plyr::ldply(diff.exp.gl, .fun = fisherEnrichment, targets, backgroundGenes, .id = 'Name', .parallel = T) %>%
+  kable
+
 ```
@@
                activityName = 'Pathway enrichment analysis')
-stopCluster(cl)
+# stopCluster(cl)
 ```

### 2017-03-07 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM.Rmd
Minor modifications in ROSMAP reprocessed RNASeq to include picard metrics

@@
 
-## Requires ggdendro
-devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")
-
 options(xtable.type="html")
@@
 COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F)
+COUNT[,grep('150_120419', colnames(COUNT))[2]] = NULL
 
@@
 METADATA.TECH <- read.table(METADATA.TECH_OBJ@filePath,sep='\t',header=T)
+
+# Get picard metrics from synapse
+METADATA.PICARD_ID <- 'syn8380366'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
+METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE)
+METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
+METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)
+
+METADATA.PICARD$Sampleid = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
+  gsub('Aligned.out/picard.CombinedMetrics.csv','',.)
+colnames(METADATA.PICARD) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD))
+colnames(METADATA.PICARD) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD))
 
@@
 METADATA <- METADATA.TECH %>%
+  left_join(METADATA.PICARD) %>%
   dplyr::select(-projid) %>%
@@
 ### Covariate clustering
-Determine relationship between covariates. 
+Determine relationship between covariates
 ```{r covariates.clustering, cache=TRUE}
-#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv"
+#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv", "PF_INDEL_RATE", "PF_MISMATCH_RATE"
 FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx")
-ContCovariates <- c("RINcontinuous","age_first_ad_dx","age_death","age_at_visit_max","pmi","educ")
+ContCovariates <- c("RINcontinuous","age_first_ad_dx","age_death","age_at_visit_max","pmi","educ", 
+                    "PCT_PF_READS_ALIGNED", "PCT_CORRECT_STRAND_READS", "PCT_INTERGENIC_BASES", 
+                    "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")
 
@@
 COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
-ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.5, row.width=0.1)
+draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
 ```
 
-```{r dataExplore, fig.width = 10, fig.height = 10}
+```{r dataExplore, fig.width = 10, fig.height = 14}
 # RIN
@@
 p[[4]] = p[[4]] + ggtitle('Education') + theme(legend.position = 'top')
+
+# Intronic bases
+p[[5]] = ggplot(COVARIATES, aes(x = cogdx, y = PCT_INTRONIC_BASES)) + geom_boxplot()
+p[[5]] = p[[5]] + ggtitle('Fraction Intronic Bases') + theme(legend.position = 'top')
+
+# Ribosomal bases
+p[[6]] = ggplot(COVARIATES, aes(x = cogdx, y = PCT_RIBOSOMAL_BASES)) + geom_boxplot()
+p[[6]] = p[[6]] + ggtitle('Fraction Ribosomal Bases') + theme(legend.position = 'top')
 
@@
 
-PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
+PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], 
+                                                 MIN_GENE_CPM=0, 
+                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
 ```
@@
 ### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix.
-Here "cogdx" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on cognitive scores)
-
-Adding Batch and msex a priori to variable selection
+Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Adding Batch and msex a priori to variable selection
 ```{r customNorm, results='asis'}
 # Primary variable of interest
-postAdjustCovars = c('cogdx', 'Batch', 'msex');
+postAdjustCovars = c('Batch', 'msex');
 
@@
 residualSigCovars = preAdjustedSigCovars
-covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
+covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars,
+                                                                        c(postAdjustCovars, 'cogdx'))]
 postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
@@
   # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars))
+  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, 'cogdx'))
 
@@
   covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
-                                                                          c(postAdjustCovars))]
+                                                                          c(postAdjustCovars, 'cogdx'))]
 
@@
 ```{r residualSigCovars.manual, fig.width=12, fig.height=8}
+# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
+residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+                                              COVARIATES,
+                                              'adjusted design(voom-normalized)',
+                                              isKeyPlot=TRUE)
+
 residualSigCovars[["PC_res"]][[2]]$plotData
-```
-
-### Residual calculation
-Calculate weighted residuals and add back "cogdx" to the residuals
-```{r varsToAddBack}
-# Add variable of interest back to the residuals
-varsToAddIn = grep("cogdx", colnames(DM1$design), value = T)
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
-  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
 ```
@@
                        mart = mart)
-
-# Add gene names
-RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
-  rownameToFirstColumn('gene.id') %>%
-  left_join(backgroundGenes) %>%
-  left_join(Ensemble2HGNC) %>%
-  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
-
-VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
-  rownameToFirstColumn('gene.id') %>%
-  left_join(backgroundGenes) %>%
-  left_join(Ensemble2HGNC) %>%
-  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
 ```
@@
 ```{r diffExp, fig.height=6, fig.width=18}
+# Get design matrix
+DESIGN = getDesignMatrix(COVARIATES[, 'cogdx', drop = F], Intercept = F)
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
+# Fit linear model with residual expression
+FIT = lmFit(RESIDUAL.GENE_EXPRESSION, design = DESIGN$design)
+
 # Fit contrast
@@
                                      "cogdx2-cogdx4"),
-                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
-FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
+                         levels = colnames(FIT$coefficients))
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
 FIT.CONTR = eBayes(FIT.CONTR)
@@
 # Get differnetial expression
-DE = lapply(1:3, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
-  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T) %>%
+DE = lapply(1:3, function(i, FIT, RESID.EXPR, backgroundGenes, Ensemble2HGNC){
+  topTable(FIT.CONTR, coef=i, number = dim(RESID.EXPR)[1], confint = T) %>%
     rownameToFirstColumn('gene.id') %>%
@@
     dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, logFC, CI.L, CI.R, AveExpr, t, P.Value, adj.P.Val, B)
-}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC) 
+}, FIT, RESIDUAL.GENE_EXPRESSION, backgroundGenes, Ensemble2HGNC) 
 names(DE) = colnames(contrast)
@@
                 Direction = logFC/abs(logFC),
-                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')))
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction))
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
 
@@
 
-p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
-p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
-p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
-p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
+p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red'))
 p = p + facet_grid(.~Comparison, scales = 'fixed')
@@
 
+### Associate differential expression results with gc content and gene length
+```{r associate.de}
+geneLength.GCcontent = EDASeq::getGeneLengthAndGCContent(unique(DE$ensembl_gene_id), 
+                                                        org = 'hsa', 
+                                                        mode = "biomart") %>%
+  rownameToFirstColumn('ensembl_gene_id')
+
+tmp = left_join(DE, geneLenth.GCcontent)
+
+pl = list()
+pl[[1]] = ggplot(tmp, aes(y = log10(length), x = abs(logFC))) + 
+  geom_point() + 
+  stat_smooth(method=glm, se = TRUE, na.rm = T)
+pl[[2]] = ggplot(tmp, aes(y = gc, x = abs(logFC))) + 
+  geom_point() + 
+  stat_smooth(method=glm, se = TRUE, na.rm = T)
+pl[[3]] = ggplot(tmp, aes(y = log10(length), x = adj.P.Val)) + 
+  geom_point() + 
+  stat_smooth(method=glm, se = TRUE, na.rm = T)
+pl[[4]] = ggplot(tmp, aes(y = gc, x = adj.P.Val)) + 
+  geom_point() + 
+  stat_smooth(method=glm, se = TRUE, na.rm = T)
+
+multiplot(plotlist = pl, cols = 2)
+```
+
 ### Store files in synapse
 ```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
+# Add gene names
+RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
+  rownameToFirstColumn('gene.id') %>%
+  left_join(backgroundGenes) %>%
+  left_join(Ensemble2HGNC) %>%
+  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+
+VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
+  rownameToFirstColumn('gene.id') %>%
+  left_join(backgroundGenes) %>%
+  left_join(Ensemble2HGNC) %>%
+  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+
 # Code
@@
 COV_OBJ = File('ROSMAP_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+annotations(COV_OBJ) = annotations(CODE)
+annotations(COV_OBJ)$dataSubType = 'covariates'
 COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 COUNT_OBJ = File('ROSMAP_DLPFC_Counts.tsv', name = 'Raw counts (filtered)', parentId = CODE$properties$id)
+annotations(COUNT_OBJ) = annotations(CODE)
+annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
 COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 LCOUNT_OBJ = File('ROSMAP_DLPFC_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
+annotations(LCOUNT_OBJ) = annotations(CODE)
+annotations(LCOUNT_OBJ)$dataSubType = 'logcpm'
 LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 DM_OBJ = File('ROSMAP_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+annotations(DM_OBJ) = annotations(CODE)
+annotations(DM_OBJ)$dataSubType = 'designMatrix'
 DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 write.table(RESIDUAL.GENE_EXPRESSION, file = 'ROSMAP_DLPFC_Expression.tsv', sep = '\t', row.names=F, quote=F)
-EXP_OBJ = File('ROSMAP_DLPFC_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', parentId = CODE$properties$id)
+EXP_OBJ = File('ROSMAP_DLPFC_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', 
+               parentId = CODE$properties$id)
+annotations(EXP_OBJ) = annotations(CODE)
+annotations(EXP_OBJ)$dataSubType = 'residualGeneExp'
 EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 DEXP_OBJ = File('ROSMAP_DLPFC_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
+annotations(DEXP_OBJ) = annotations(CODE)
+annotations(DEXP_OBJ)$dataSubType = 'diffExp'
 DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 

### 2017-03-07 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd
Added picard metrics to MSSM reprocessed rnaseq analysis

@@
 library(foreach)
-library(doMC)
-library(foreach)
 library(doParallel)
@@
 synapseLogin()
-
-## Requires ggdendro
-devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")
 
@@
   top_n(1, -rRNA.rate) %>%
-  dplyr::select(sampleIdentifier, BrodmannArea, barcode, individualIdentifier, batch, RIN, TotalReads, Mapped, rRNA.rate) %>%
+  dplyr::select(sampleIdentifier, BrodmannArea, barcode, individualIdentifier, batch, RIN) %>%
   unique
@@
 
-METADATA <- full_join(METADATA_TECH, METADATA_CLINICAL)
+# Get picard metrics from synapse
+METADATA.PICARD_ID <- 'syn8380379'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
+METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
+METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
+METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
+METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)
+
+METADATA.PICARD$sampleIdentifier = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
+  gsub('.accepted_hits.sort.coord/picard.CombinedMetrics.csv','',.)
+colnames(METADATA.PICARD) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD))
+colnames(METADATA.PICARD) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD))
+
+METADATA <- full_join(METADATA_TECH, METADATA_CLINICAL) %>%
+  inner_join(METADATA.PICARD)
 ```
@@
 
-# Total Reads
-p[[4]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(TotalReads), fill = BrodmannArea)) + geom_boxplot()
-p[[4]] = p[[4]] + ggtitle('Total Reads') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
+# RIBOSOMAL FRACTION
+p[[4]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(PCT_RIBOSOMAL_BASES), fill = BrodmannArea)) + geom_boxplot()
+p[[4]] = p[[4]] + ggtitle('Fraction of Ribosomal Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
+
+# INTRONIC BASES
+p[[5]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(PCT_INTRONIC_BASES), fill = BrodmannArea)) + geom_boxplot()
+p[[5]] = p[[5]] + ggtitle('Frction Intronic Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
 
 # Mapped Reads
-p[[5]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(Mapped), fill = BrodmannArea)) + geom_boxplot()
-p[[5]] = p[[5]] + ggtitle('Mapped') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
-
-# Mapped Reads
-p[[6]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(rRNA.rate), fill = BrodmannArea)) + geom_boxplot()
-p[[6]] = p[[6]] + ggtitle('rRNA Rate') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
+p[[6]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(METADATA$PCT_INTERGENIC_BASES), fill = BrodmannArea)) + geom_boxplot()
+p[[6]] = p[[6]] + ggtitle('Fraction Intergenic Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
 
@@
 FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrodmannArea.Dx")
-ContCovariates <- c("RIN", "RIN2", "AOD","PMI", "TotalReads", "Mapped", "rRNA.rate")
+ContCovariates <- c("RIN", "RIN2", "AOD","PMI", "PCT_PF_READS", "PCT_PF_READS_ALIGNED", 
+                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", 
+                    "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")
 
@@
 ```{r covariates.correlation, fig.width=8, fig.height=8}
-COVARIATES.CORRELATION = CovariateAnalysis::getAssociationStatistics(COVARIATES, PVAL = 0.05)
-COVARIATES.CORRELATION$plot
+COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES[,-(11)])
+draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
 ```
@@
 ### Normalisation (iterative)
-Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
+Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix
 ```{r iterative.adjusted.voom.normalisation, results='asis'}
-postAdjustCovars = c('BrodmannArea.Dx', 'batch', 'SEX')
+postAdjustCovars = c('batch', 'SEX')
 randomCovariates = 'individualIdentifier'
@@
 covariatesEffects = residualSigCovars$Effects.significantCovars
-covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates))]
+covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates,'BrodmannArea.Dx'))]
 
@@
   # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, randomCovariates))
+  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, randomCovariates, 'BrodmannArea.Dx'))
 
@@
   covariatesEffects = residualSigCovars$Effects.significantCovars
-  covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars, randomCovariates))]
+  covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars, randomCovariates, 'BrodmannArea.Dx'))]
 
@@
 ```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
+# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
+residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+                                              COVARIATES,
+                                              'adjusted design(voom-normalized)',
+                                              isKeyPlot=TRUE)
+
 residualSigCovars[["PC_res"]][[2]]$plotData
-```
-Adding "BrodmannArea.Dx" back to the residuals
-```{r addVarsBack}
-varsToAddBackIn <-  grep("BrodmannArea.Dx", colnames(FIT$coefficients), value = T)
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
-  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
 ```
@@
 
-```{r get.gene.symbols}
+```{r get.gene.symbols, cache=TRUE, eval=T}
 backgroundGenes = data.frame(gene.id = rownames(COUNT)) %>%
@@
 # Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "dec2016.archive.ensembl.org")
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "dec2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
 
@@
 ```{r diffExp, fig.height=10, fig.width=20}
+# Post adjusted design matrix
+DESIGN = getDesignMatrix(COVARIATES[,'BrodmannArea.Dx',drop=F], Intercept = F)
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
+# Fit linear model with residual expression
+FIT = lmFit(RESIDUAL.GENE_EXPRESSION, design = DESIGN$design)
+
 # Fit contrast
@@
 # Get differnetial expression
-DE = lapply(1:12, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
-  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T)
-}, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
+DE = lapply(1:12, function(i, FIT.CONTR, RESIDUAL.GENE_EXPRESSION){
+  topTable(FIT.CONTR, coef=i, number = dim(RESIDUAL.GENE_EXPRESSION)[1], confint = T) %>%
+    rownameToFirstColumn('genes')
+}, FIT.CONTR, RESIDUAL.GENE_EXPRESSION)
 names(DE) = gsub('BrodmannArea.Dx','',colnames(contrast))
@@
   left_join(Ensemble2HGNC) %>%
-  dplyr::select(-position)
+  dplyr::select(-position) %>%
+  dplyr::mutate(Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction))
+DE1$Direction[DE1$adj.P.Val > 0.05 | abs(DE1$logFC) < log2(1.2)] = 'NONE'
 
-p <- ggplot(DE1, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
-p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
-p <- p + geom_vline(xintercept = log2(1.2), color = 'red')
-p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
-p <- p + facet_grid(Region~.+Comparison)
+p = ggplot(DE1, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red'))
+p = p + facet_grid(Region~.+Comparison, scales = 'fixed')
 p
@@
 
+### Associate differential expression results with gc content and gene length
+```{r associate.de}
+geneLength.GCcontent = EDASeq::getGeneLengthAndGCContent(unique(DE1$ensembl_gene_id), 
+                                                         org = 'hsa', 
+                                                         mode = "biomart") %>%
+  rownameToFirstColumn('ensembl_gene_id')
+
+tmp = left_join(DE1, geneLength.GCcontent)
+
+pl = list()
+pl[[1]] = ggplot(tmp, aes(y = log10(length), x = abs(logFC))) + 
+  geom_point() + 
+  stat_smooth(method=glm, se = TRUE, na.rm = T)
+pl[[2]] = ggplot(tmp, aes(y = gc, x = abs(logFC))) + 
+  geom_point() + 
+  stat_smooth(method=glm, se = TRUE, na.rm = T)
+pl[[3]] = ggplot(tmp, aes(y = log10(length), x = adj.P.Val)) + 
+  geom_point() + 
+  stat_smooth(method=glm, se = TRUE, na.rm = T)
+pl[[4]] = ggplot(tmp, aes(y = gc, x = adj.P.Val)) + 
+  geom_point() + 
+  stat_smooth(method=glm, se = TRUE, na.rm = T)
+
+multiplot(plotlist = pl, cols = 2)
+```
+
 ### Store files in synapse
 ```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
+# Add gene names
+RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
+  rownameToFirstColumn('gene.id') %>%
+  left_join(backgroundGenes) %>%
+  left_join(Ensemble2HGNC) %>%
+  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+
+VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
+  rownameToFirstColumn('gene.id') %>%
+  left_join(backgroundGenes) %>%
+  left_join(Ensemble2HGNC) %>%
+  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+
 # Code
@@
 COVARIATES = rownameToFirstColumn(COVARIATES, 'sampleIdentifier')
-write.table(COVARIATES, file = 'MSBBCovariates.tsv', sep = '\t', row.names=F, quote=F)
-COV_OBJ = File('MSBBCovariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+write.table(COVARIATES, file = 'MSBB_Covariates.tsv', sep = '\t', row.names=F, quote=F)
+COV_OBJ = File('MSBB_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+annotations(COV_OBJ) = annotations(CODE)
+annotations(COV_OBJ)$dataSubType = 'covariates'
 COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
   as.matrix() %>% rownameToFirstColumn('gene.id') %>%
-  write.table(file = 'MSBBCounts.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('MSBBCounts.tsv', name = 'Raw counts (filtered)', parentId = CODE$properties$id)
+  write.table(file = 'MSBB_Counts.tsv', sep = '\t', row.names=F, quote=F)
+COUNT_OBJ = File('MSBB_Counts.tsv', name = 'Raw counts (filtered)', parentId = CODE$properties$id)
+annotations(COUNT_OBJ) = annotations(CODE)
+annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
 COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 # Store logCPM
-write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'MSBBlogCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('MSBBlogCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
+write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'MSBB_logCPM.tsv', sep = '\t', row.names=F, quote=F)
+LCOUNT_OBJ = File('MSBB_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
+annotations(LCOUNT_OBJ) = annotations(CODE)
+annotations(LCOUNT_OBJ)$dataSubType = 'logcpm'
 LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 # Store design matrix
-write.table(DM1$design, file = 'MSBBDesign.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('MSBBDesign.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+write.table(DM1$design, file = 'MSBB_Design.tsv', sep = '\t', row.names=F, quote=F)
+DM_OBJ = File('MSBB_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+annotations(DM_OBJ) = annotations(CODE)
+annotations(DM_OBJ)$dataSubType = 'designMatrix'
 DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 # Store residual gene expression
-write.table(RESIDUAL.GENE_EXPRESSION, file = 'MSBBExpression.tsv', sep = '\t', row.names=F, quote=F)
-EXP_OBJ = File('MSBBExpression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', 
+write.table(RESIDUAL.GENE_EXPRESSION, file = 'MSBB_Expression.tsv', sep = '\t', row.names=F, quote=F)
+EXP_OBJ = File('MSBB_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', 
                parentId = CODE$properties$id)
+annotations(EXP_OBJ) = annotations(CODE)
+annotations(EXP_OBJ)$dataSubType = 'residualGeneExp'
 EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 # Store differential expression results
-write.table(DE1, file = 'MSBBDiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('MSBBDiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
+write.table(DE1, file = 'MSBB_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('MSBB_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
+annotations(DEXP_OBJ) = annotations(CODE)
+annotations(DEXP_OBJ)$dataSubType = 'diffExp'
 DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                     executed = thisFile, activityDescription = activityDescription)
-
 stopCluster(cl)

### 2017-03-07 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM.Rmd
Added picard metrics to MSSM reprocessed rnaseq analysis

@@
 ContCovariates <- c("RINcontinuous","age_first_ad_dx","age_death","age_at_visit_max","pmi","educ", 
-                    "PCT_PF_READS_ALIGNED", "PCT_CORRECT_STRAND_READS", "PCT_INTERGENIC_BASES", 
-                    "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")
+                    "PCT_PF_READS", "PCT_PF_READS_ALIGNED", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", 
+                    "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")
 

### 2017-03-07 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM.Rmd
Minor modifications

@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
-METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE)
+METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
+METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
 METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]

### 2017-03-08 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd
Minor bug fix

@@
 
-# RIBOSOMAL FRACTION
+# Ribosomal bases
 p[[4]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(PCT_RIBOSOMAL_BASES), fill = BrodmannArea)) + geom_boxplot()
@@
 
-# INTRONIC BASES
+# Intronic bases
 p[[5]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(PCT_INTRONIC_BASES), fill = BrodmannArea)) + geom_boxplot()
@@
 
-# Mapped Reads
-p[[6]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(METADATA$PCT_INTERGENIC_BASES), fill = BrodmannArea)) + geom_boxplot()
+# Intergenic bases
+p[[6]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(PCT_INTERGENIC_BASES), fill = BrodmannArea)) + geom_boxplot()
 p[[6]] = p[[6]] + ggtitle('Fraction Intergenic Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

### 2017-03-08 code/Rmd/Mayo_RNASeq_Reprocessed_TMM.Rmd
Minor bug fix

@@
 
+# Get picard metrics from synapse
+METADATA.PICARD_CE_ID <- 'syn8380392'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_CE_ID
+METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = FALSE, exdir = getwd())
+METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = TRUE, exdir = getwd())
+METADATA.PICARD_CE_FPATHS = METADATA.PICARD_CE_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_CE_FPATHS)]
+METADATA.PICARD_CE <- plyr::ldply(METADATA.PICARD_CE_FPATHS, read.csv)
+
+METADATA.PICARD_CE$ID = gsub('picard/', '', METADATA.PICARD_CE_FPATHS) %>%
+  gsub('/picard.CombinedMetrics.csv','',.) %>%
+  sapply(function(x){str_split(x, '\\.')[[1]][1]})
+colnames(METADATA.PICARD_CE) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_CE))
+colnames(METADATA.PICARD_CE) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD_CE))
+
+METADATA.PICARD_TC_ID <- 'syn8383233'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_TC_ID
+METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = FALSE, exdir = getwd())
+METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = TRUE, exdir = getwd())
+METADATA.PICARD_TC_FPATHS = METADATA.PICARD_TC_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_TC_FPATHS)]
+METADATA.PICARD_TC <- plyr::ldply(METADATA.PICARD_TC_FPATHS, read.csv)
+
+METADATA.PICARD_TC$ID = gsub('picard/', '', METADATA.PICARD_TC_FPATHS) %>%
+  gsub('/picard.CombinedMetrics.csv','',.) %>%
+  sapply(function(x){str_split(x, '\\.')[[1]][1]}) %>%
+  gsub('Aligned','',.)
+colnames(METADATA.PICARD_TC) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_TC))
+colnames(METADATA.PICARD_TC) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD_TC))
+
+# Merge all metadata
 METADATA = rbindlist(list(METADATA_TC, METADATA_CE), use.names = T, fill = T) %>%
+  inner_join(rbindlist(list(METADATA.PICARD_TC, METADATA.PICARD_CE), use.names = T, fill = T)) %>%
   as.data.frame
@@
 
-```{r diagnosisPlots, fig.height=6, fig.width = 10}
+```{r diagnosisPlots, fig.height=14, fig.width = 8}
 # RIN
@@
 
-multiplot(plotlist = p, cols = 3)
+# Ribosomal bases
+p[[4]] = ggplot(METADATA, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_RIBOSOMAL_BASES), fill = BrainRegion)) + geom_boxplot()
+p[[4]] = p[[4]] + ggtitle('Fraction of Ribosomal Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
+
+# Intronic bases
+p[[5]] = ggplot(METADATA, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTRONIC_BASES), fill = BrainRegion)) + geom_boxplot()
+p[[5]] = p[[5]] + ggtitle('Frction Intronic Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
+
+# Intergenic bases
+p[[6]] = ggplot(METADATA, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTERGENIC_BASES), fill = BrainRegion)) + geom_boxplot()
+p[[6]] = p[[6]] + ggtitle('Fraction Intergenic Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
+
+multiplot(plotlist = p, cols = 2)
 
@@
 FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
-ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI")
+ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI", 
+                    "PCT_PF_READS", "PCT_PF_READS_ALIGNED", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", 
+                    "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")
 
@@
 COVARIATES.CORRELATION = CovariateAnalysis::getAssociationStatistics(COVARIATES, PVAL = 0.05)
-ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.4, row.width=0.1)
+draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
 ```
@@
 p <- p + theme_bw() + theme(legend.position = "left") +scale_shape_manual(values = 1:2)
-p <- p + facet_grid(Region~.)
+p <- p + facet_grid(Region~., scale = 'free_y')
 # p <- p + geom_text(aes(label= ID), size=4, hjust=0)
@@
 ### Normalisation (iterative)
-Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
+Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix
 ```{r iterative.adjusted.voom.normalisation, results='asis'}
-postAdjustCovars = c('BrainRegion.Diagnosis', 'Source', 'FLOWCELL', 'Gender')
+postAdjustCovars = c('Source', 'FLOWCELL', 'Gender')
 randomCovariates = 'Donor_ID'
@@
 covariatesEffects = residualSigCovars$Effects.significantCovars
-covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates))]
+covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates, 'BrainRegion.Diagnosis'))]
 
@@
   # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, randomCovariates))
+  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, randomCovariates, 'BrainRegion.Diagnosis'))
 
@@
   residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                                COVARIATES,
+                                                COVARIATES[, residCovars, drop=F],
                                                 'all adjusted design(voom-normalized)',
@@
   covariatesEffects = residualSigCovars$Effects.significantCovars
-  covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars, randomCovariates))]
+  covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars, randomCovariates, 'BrainRegion.Diagnosis'))]
 
@@
 ```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
+# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
+residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+                                              COVARIATES,
+                                              'adjusted design(voom-normalized)',
+                                              isKeyPlot=TRUE)
+
 residualSigCovars[["PC_res"]][[2]]$plotData
 ```
-Adding "BrainRegion.Diagnosis" back to the residuals
-```{r addVarsBack}
-varsToAddBackIn <-  grep("BrainRegion.Diagnosis", colnames(FIT$coefficients), value = T)
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
-  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
+Coexpression of genes 
+```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
+cr = cor(t(RESIDUAL.GENE_EXPRESSION))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
 ```
@@
 
-```{r get.gene.symbols}
+```{r get.gene.symbols, cache=TRUE, eval=T}
+backgroundGenes = data.frame(gene.id = rownames(COUNT)) %>%
+  dplyr::mutate(ensembl_gene_id = gene.id) %>%
+  tidyr::separate(ensembl_gene_id, c('ensembl_gene_id', 'position'), sep = '\\.')
+
 # Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "dec2016.archive.ensembl.org")
-
-# Get background gene ids
-backgroundGenes = data.frame(gene.id = rownames(COUNT), ID = rownames(COUNT)) %>%
-  tidyr::separate(ID, c('ensembl_gene_id','position'), sep = '\\.')
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "dec2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
 
@@
 Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
-                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
+                       filters = "ensembl_gene_id", 
+                       values = backgroundGenes$ensembl_gene_id,
                        mart = mart)
-
-# Add gene names
-RESIDUAL.GENE_EXPRESSION <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'gene.id') %>%
-  left_join(backgroundGenes) %>% 
-  left_join(Ensemble2HGNC) %>%
-  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
-
-VOOM.ADJUSTED.GENE_EXPRESSION$E <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, 'gene.id') %>%
-  left_join(backgroundGenes) %>% 
-  left_join(Ensemble2HGNC) %>%
-  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
 ```
@@
 ```{r diffExp, fig.height=10, fig.width=20}
+# Get design matrix
+DESIGN = getDesignMatrix(COVARIATES[, 'BrainRegion.Diagnosis', drop = F], Intercept = F)
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
+# Fit linear model with residual expression
+FIT = lmFit(RESIDUAL.GENE_EXPRESSION, design = DESIGN$design)
+
 # Fit contrast
@@
 # Get differnetial expression
-DE = lapply(1:12, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
-  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T)
-}, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
+DE = lapply(1:12, function(i, FIT.CONTR, RESIDUAL.GENE_EXPRESSION){
+  topTable(FIT.CONTR, coef=i, number = dim(RESIDUAL.GENE_EXPRESSION)[1], confint = T) %>%
+    rownameToFirstColumn('genes')
+}, FIT.CONTR, RESIDUAL.GENE_EXPRESSION)
 names(DE) = gsub('BrainRegion.Diagnosis','',colnames(contrast))
@@
   left_join(backgroundGenes) %>%
-  left_join(Ensemble2HGNC)
+  left_join(Ensemble2HGNC) %>%
+  dplyr::mutate(Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction))
+DE1$Direction[DE1$adj.P.Val > 0.05 | abs(DE1$logFC) < log2(1.2)] = 'NONE'
 
-p <- ggplot(DE1, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
-p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
-p <- p + geom_vline(xintercept = log2(1.2), color = 'red')
-p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
-p <- p + facet_grid(Region~.+Comparison)
+p = ggplot(DE1, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red'))
+p = p + facet_grid(Region~.+Comparison, scales = 'fixed')
 p
@@
 ```
+### Associate differential expression results with gc content and gene length
+```{r associate.de}
+geneLength.GCcontent = EDASeq::getGeneLengthAndGCContent(unique(DE1$ensembl_gene_id), 
+                                                         org = 'hsa', 
+                                                         mode = "biomart") %>%
+  rownameToFirstColumn('ensembl_gene_id')
+
+tmp = left_join(DE1, geneLength.GCcontent)
+
+pl = list()
+pl[[1]] = ggplot(tmp, aes(y = log10(length), x = abs(logFC))) + 
+  geom_point() + 
+  stat_smooth(method=glm, se = TRUE, na.rm = T)
+pl[[2]] = ggplot(tmp, aes(y = gc, x = abs(logFC))) + 
+  geom_point() + 
+  stat_smooth(method=glm, se = TRUE, na.rm = T)
+pl[[3]] = ggplot(tmp, aes(y = log10(length), x = adj.P.Val)) + 
+  geom_point() + 
+  stat_smooth(method=glm, se = TRUE, na.rm = T)
+pl[[4]] = ggplot(tmp, aes(y = gc, x = adj.P.Val)) + 
+  geom_point() + 
+  stat_smooth(method=glm, se = TRUE, na.rm = T)
+
+multiplot(plotlist = pl, cols = 2)
+```
 
@@
 ```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
-# Code
+# Add gene names
+RESIDUAL.GENE_EXPRESSION <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'gene.id') %>%
+  left_join(backgroundGenes) %>% 
+  left_join(Ensemble2HGNC) %>%
+  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+
+VOOM.ADJUSTED.GENE_EXPRESSION$E <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, 'gene.id') %>%
+  left_join(backgroundGenes) %>% 
+  left_join(Ensemble2HGNC) %>%
+  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
+
 CODE <- Folder(name = 'UFL_Mayo_ISB Reprocessed RNASeq TMM', parentId = parentId)
@@
 # Store covariates
-COVARIATES = rownameToFirstColumn(COVARIATES, 'Sampleid')
-write.table(COVARIATES, file = 'MAYO_CBE_TCX_Covariates.tsv', sep = '\t', row.names=F, quote=F)
-COV_OBJ = File('MAYO_CBE_TCX_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+COVARIATES = rownameToFirstColumn(COVARIATES, 'ID')
+write.table(COVARIATES, file = 'MAYO_Covariates.tsv', sep = '\t', row.names=F, quote=F)
+COV_OBJ = File('MAYO_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+annotations(COV_OBJ) = annotations(CODE)
+annotations(COV_OBJ)$dataSubType = 'covariates'
 COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
   as.matrix() %>% rownameToFirstColumn('gene.id') %>%
-  write.table(file = 'MAYO_CBE_TCX_Counts.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('MAYO_CBE_TCX_Counts.tsv', name = 'Raw counts (filtered)', parentId = CODE$properties$id)
+  write.table(file = 'MAYO_Counts.tsv', sep = '\t', row.names=F, quote=F)
+COUNT_OBJ = File('MAYO_Counts.tsv', name = 'Raw counts (filtered)', parentId = CODE$properties$id)
+annotations(COUNT_OBJ) = annotations(CODE)
+annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
 COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 # Store logCPM
-write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'MAYO_CBE_TCX_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('MAYO_CBE_TCX_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
+write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'MAYO_logCPM.tsv', sep = '\t', row.names=F, quote=F)
+LCOUNT_OBJ = File('MAYO_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
+annotations(LCOUNT_OBJ) = annotations(CODE)
+annotations(LCOUNT_OBJ)$dataSubType = 'logcpm'
 LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 # Store design matrix
-write.table(DM1$design, file = 'MAYO_CBE_TCX_Design.tsv', sep = '\t', row.names=T, quote=F)
-DM_OBJ = File('MAYO_CBE_TCX_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+write.table(DM1$design, file = 'MAYO_Design.tsv', sep = '\t', row.names=F, quote=F)
+DM_OBJ = File('MAYO_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+annotations(DM_OBJ) = annotations(CODE)
+annotations(DM_OBJ)$dataSubType = 'designMatrix'
 DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 # Store residual gene expression
-write.table(RESIDUAL.GENE_EXPRESSION, file = 'MAYO_CBE_TCX_Expression.tsv', sep = '\t', row.names=F, quote=F)
-EXP_OBJ = File('MAYO_CBE_TCX_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', 
+write.table(RESIDUAL.GENE_EXPRESSION, file = 'MAYO_Expression.tsv', sep = '\t', row.names=F, quote=F)
+EXP_OBJ = File('MAYO_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', 
                parentId = CODE$properties$id)
+annotations(EXP_OBJ) = annotations(CODE)
+annotations(EXP_OBJ)$dataSubType = 'residualGeneExp'
 EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 # Store differential expression results
-write.table(DE1, file = 'MAYO_CBE_TCX_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('MAYO_CBE_TCX_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
+write.table(DE1, file = 'MAYO_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('MAYO_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
+annotations(DEXP_OBJ) = annotations(CODE)
+annotations(DEXP_OBJ)$dataSubType = 'diffExp'
 DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 

### 2017-03-08 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd
Minor bug fix

--- 2017-03-07 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd
@@
   residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                                COVARIATES,
+                                                COVARIATES[,residCovars,drop = F],
                                                 'all adjusted design(voom-normalized)',
@@
   group_by(Region, Comparison) %>%
-  summarise(count = length(unique(genes))) %>%
-  tidyr::spread(Region, Comparison)
+  summarise(count = length(unique(gene.id))) %>%
+  tidyr::spread(Region, count)
 kable(tmp)

### 2017-03-08 code/Rmd/Mayo_RNASeq_Reprocessed_TMM.Rmd
Minor bug fix

--- 2017-02-21 code/Rmd/Mayo_RNASeq_Reprocessed_TMM.Rmd
@@
 
+library(ComplexHeatmap)
+library(circlize)
+
 library(synapseClient)
@@
 METADATA <- METADATA %>%
-  filter(!is.na(Source)) %>%
-  dplyr::filter(!(ID %in% c("11311_CER", "1923_CER", "1923_TCX", 
-                            "11396_TCX", "11294_TCX", "11408_TCX")))
+  filter(!is.na(Source))
+# %>%
+#   dplyr::filter(!(ID %in% c("11311_CER", "1923_CER", "1923_TCX", 
+#                             "11396_TCX", "11294_TCX", "11408_TCX")))
 ```
@@
 plotdata$Source = factor(plotdata$Source)
+plotdata$ID[!(plotdata$ID %in% c("11311_CER", "1923_CER", "1923_TCX", "11396_TCX", "11294_TCX", "11408_TCX"))] = ''
 
@@
 p <- p + geom_point(aes(shape = Source, color = Region))
-p <- p + theme_bw() + theme(legend.position = "left") +scale_shape_manual(values = 1:2)
+p <- p + theme_bw() + theme(legend.position = "right") +scale_shape_manual(values = 1:2)
 p <- p + facet_grid(Region~.)
-# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
+p <- p + geom_text(aes(label= ID), size=4, hjust=1, vjust=-1)
 p

### 2017-03-08 code/Rmd/Mayo_RNASeq_Reprocessed_TMM.Rmd

--- 2017-03-08 code/Rmd/Mayo_RNASeq_Reprocessed_TMM.Rmd
@@
 library(glmnet)
+
+library(ComplexHeatmap)
+library(circlize)
 

### 2017-03-08 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd
Minor modifications

@@
 ### Associate differential expression results with gc content and gene length
-```{r associate.de}
+```{r associate.de, eval = FALSE}
 geneLength.GCcontent = EDASeq::getGeneLengthAndGCContent(unique(DE1$ensembl_gene_id), 

### 2017-03-10 code/R/computeCellTypeSpecificity.R
Minor modifications in summarising the cell type specific data

@@ -9,6 +9,7 @@ library(tidyr)
 library(plyr)
 library(dplyr)
 library(stringr)
+library(matrixStats)
 
 library(synapseClient)
 library(knitr)
@@ -29,7 +30,7 @@ ref.cellTypes = downloadFile('syn8077191') %>%
 ref.expr.sum = list()
 for(celltype in unique(ref.cellTypes$CellType)){
   ids = ref.cellTypes$SampleID[ref.cellTypes$CellType == celltype]
-  ref.expr.sum[[celltype]] = log2(rowSums(2^ref.expr[,ids], na.rm = T)) %>%
+  ref.expr.sum[[celltype]] = log2(apply((2^ref.expr[,ids]+0.5), 1, median, na.rm = T)) %>%
     rownameToFirstColumn('hgnc_symbol') %>%
     plyr::rename(c('DF' = celltype))
 }
@@ -38,13 +39,13 @@ ref.expr.sum = join_all(ref.expr.sum)
 # Compute tau
 tmp = data.matrix(ref.expr.sum[,-(1)])
 rownames(tmp) = ref.expr.sum$hgnc_symbol
-tau = rowSums(1 - (tmp/rowMax(tmp)), na.rm = T)
+tau = rowSums(1 - (tmp/rowMaxs(tmp)), na.rm = T)
 tau = (tau/(dim(tmp)[2]-1)) %>%
   rownameToFirstColumn('hgnc_symbol') %>%
   dplyr::rename(tau = DF)
 
 # Compute TSI
-TSI = rowMax(tmp)/rowSums(tmp) %>%
+TSI = rowMaxs(tmp)/rowSums(tmp) %>%
   as.data.frame()
 colnames(TSI) = 'TSI'
 TSI$hgnc_symbol = rownames(TSI)

### 2017-03-13 code/R/computeCellTypeSpecificity.R
Minor bug fix

@@ -30,7 +30,7 @@ ref.cellTypes = downloadFile('syn8077191') %>%
 ref.expr.sum = list()
 for(celltype in unique(ref.cellTypes$CellType)){
   ids = ref.cellTypes$SampleID[ref.cellTypes$CellType == celltype]
-  ref.expr.sum[[celltype]] = log2(apply((2^ref.expr[,ids]+0.5), 1, median, na.rm = T)) %>%
+  ref.expr.sum[[celltype]] = log2(apply(2^ref.expr[,ids]+1, 1, median, na.rm = T)) %>%
     rownameToFirstColumn('hgnc_symbol') %>%
     plyr::rename(c('DF' = celltype))
 }

### 2017-03-13 code/R/computeCellTypeSpecificity.R
Minor modifications

@@ -10,6 +10,7 @@ library(plyr)
 library(dplyr)
 library(stringr)
 library(matrixStats)
+library(edgeR)
 
 library(synapseClient)
 library(knitr)
@@ -18,10 +19,12 @@ library(githubr)
 synapseLogin()
 
 # Get reference expression and cell type labels from Darmanis., S et al., 2015
-ALL_USED_IDs = c('syn8077142', 'syn8077191')
-ref.expr = downloadFile('syn8077142')
-rownames(ref.expr) = ref.expr$hgnc_symbol
+ALL_USED_IDs = c('syn8077189', 'syn8077191')
+ref.expr = downloadFile('syn8077189')
+rownames(ref.expr) = ref.expr$Gene.Names
 ref.expr$hgnc_symbol = NULL
+ref.expr$Gene.Names = NULL
+ref.expr = cpm(ref.expr)
 
 ref.cellTypes = downloadFile('syn8077191') %>%
   dplyr::select(SampleID, CellType)
@@ -30,7 +33,7 @@ ref.cellTypes = downloadFile('syn8077191') %>%
 ref.expr.sum = list()
 for(celltype in unique(ref.cellTypes$CellType)){
   ids = ref.cellTypes$SampleID[ref.cellTypes$CellType == celltype]
-  ref.expr.sum[[celltype]] = log2(apply(2^ref.expr[,ids]+1, 1, median, na.rm = T)) %>%
+  ref.expr.sum[[celltype]] = log2(apply(ref.expr[,ids], 1, median, na.rm = T) + 1) %>%
     rownameToFirstColumn('hgnc_symbol') %>%
     plyr::rename(c('DF' = celltype))
 }

### 2017-03-13 code/R/computeCellTypeSpecificity.R
Changed mean to median in cell type cpecificity computation

@@ -33,7 +33,7 @@ ref.cellTypes = downloadFile('syn8077191') %>%
 ref.expr.sum = list()
 for(celltype in unique(ref.cellTypes$CellType)){
   ids = ref.cellTypes$SampleID[ref.cellTypes$CellType == celltype]
-  ref.expr.sum[[celltype]] = log2(apply(ref.expr[,ids], 1, median, na.rm = T) + 1) %>%
+  ref.expr.sum[[celltype]] = log2(apply(ref.expr[,ids], 1, mean, na.rm = T) + 1) %>%
     rownameToFirstColumn('hgnc_symbol') %>%
     plyr::rename(c('DF' = celltype))
 }

### 2017-03-14 code/Rmd/MetaAnalysis.Rmd
temp change

@@
 tmp1 = rbindlist(diff.exp, use.names = T, fill = T) %>%
-  dplyr::filter(gene.id %in% Reduce(intersect, diff.exp.gl)) %>%
+  dplyr::filter(gene.id %in% Reduce(c, diff.exp.gl)) %>%
   dplyr::select(-hgnc_symbol) %>%
@@
 
-tmp = group_by(tmp1, gene.id) %>%
-  summarise(count = sum(logFC/abs(logFC))) %>%
-  filter(abs(count) == 7) %>%
-  dplyr::mutate(GeneID = gene.id) %>%
-  tidyr::separate(gene.id, c('ensembl_gene_id','position'), sep = '\\.') %>%
-  left_join(Ensemble2HGNC)
-writeLines(paste('Differentially expressed genes in all 7 brain regions at an FDR of 0.05 and fold change of 1.2 are',
-                 length(unique(tmp$GeneID))))
-kable(data.frame(ngenes = length(unique(tmp$GeneID)),
+tmp1$Direction = 'NONE'
+tmp1$Direction[tmp1$adj.P.Val <= 0.05 & tmp1$logFC >= log2(1.2)] = 'UP'
+tmp1$Direction[tmp1$adj.P.Val <= 0.05 & tmp1$logFC <= -log2(1.2)] = 'DOWN'
+
+tmp = tmp1 %>%
+  dplyr::filter(hgnc_symbol != '') %>%
+  group_by(hgnc_symbol) %>%
+  summarise(count = sum(Direction != 'NONE')) %>%
+  filter(count >= 5) 
+writeLines(paste('Differentially expressed genes in any 5 brain regions at an FDR of 0.05 and fold change of 1.2 are',
+                 length(unique(tmp$hgnc_symbol))))
+kable(data.frame(ngenes = length(unique(tmp$hgnc_symbol)),
                  hgnc_symbol = tmp$hgnc_symbol %>% unique %>% paste(collapse = ',')))
@@
 writeLines('Top six differentially expressed genes are')
-tmp.eff.sz = group_by(tmp1, gene.id) %>% 
+tmp.eff.sz = filter(tmp1, hgnc_symbol %in% tmp$hgnc_symbol) %>%
+  group_by(hgnc_symbol) %>% 
   summarise(es = sum(logFC, na.rm = T)) %>% 
@@
 p = ggplot(tmp1 %>% 
-             filter(gene.id %in% tmp.eff.sz$gene.id[tmp.eff.sz$es>0]), 
-           aes(x = Region, y = logFC, fill = Study)) + geom_bar(stat = 'identity')
+             filter(hgnc_symbol %in% tmp.eff.sz$hgnc_symbol[tmp.eff.sz$es>2.6]), 
+           aes(x = Region, y = logFC, color = Study, fill = Direction)) + geom_bar(stat = 'identity')
 p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
-p = p + theme(legend.position = 'top')
+p = p + theme(legend.position = 'top') + scale_fill_manual(values = c('grey', 'red'))
 pl[[1]] = p
 
-p = ggplot(tmp1 %>% filter(gene.id %in% tmp.eff.sz$gene.id[tmp.eff.sz$es<0]), aes(x = Region, y = logFC, fill = Study)) + geom_bar(stat = 'identity')
+p = ggplot(tmp1 %>% 
+             filter(hgnc_symbol %in% tmp.eff.sz$hgnc_symbol[tmp.eff.sz$es<2.6]), 
+           aes(x = Region, y = logFC, color = Study, fill = Direction)) + geom_bar(stat = 'identity')
 p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
-p = p + theme(legend.position = 'top')
+p = p + theme(legend.position = 'top')  + scale_fill_manual(values = c('green','grey','red'))
 pl[[2]] = p
@@
 ```{r pathway.enrich}
-setsToTest = diffexp %>%
-  filter(adj.P.Val <= 0.05) %>%
-  dlply(.(Tissue, Direction), .fun = function(x){
-    unique(x$ensembl_gene_id)
+setsToTest = rbindlist(diff.exp, use.names = T, fill = T) %>%
+  filter(adj.P.Val <= 0.05, Direction != 'NONE') %>%
+  dlply(.(Region, Direction), .fun = function(x){
+    unique(x$hgnc_symbol)
   })
@@
 # Get Background genes
-backgroundGenes = Ensemble2HGNC$hgnc_symbol[Ensemble2HGNC$ensembl_gene_id %in% rownames(expr)] %>% unique()
+backgroundGenes = unique(rbindlist(diff.exp, use.names = T, fill = T) %>% 
+                           dplyr::select(hgnc_symbol) %>%
+                           unlist)
 

### 2017-03-15 code/Rmd/ROSMAP_REPROCESSED.Rmd
Added markdown to process ROSMAP reprocessed rnaseq with cqn normalisation and covariates adjustments

---
title: "Covariate analysis of ROSMAP reprocessed counts (with CQN normalisation and covariates adjusted)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./ROSMAP_REPROCESSED.Rmd",
                                 parentId = "syn8403743",
                                 entityName = 'ROSMAP Reprocessed RNASeq (Normalised)')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(reshape2)
library(limma)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)
library(cqn)

library(synapseClient)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

synapseLogin()

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn8403743';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of ROSMAP reprocessed RNASeq data';

thisFileName <- 'ROSMAP_REPROCESSED.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_ID <- 'syn7750270';
ALL_USED_IDs <- COUNT_ID
COUNT_OBJ <- synGet(COUNT_ID)
COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F)
COUNT[,grep('150_120419', colnames(COUNT))[2]] = NULL

# Convert rownames of counts from tracking id to ensemble gene id
tmp = data.frame(Gene.ID = rownames(COUNT)) %>%
  dplyr::mutate(ID = Gene.ID) %>%
  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')
rownames(tmp) = tmp$Gene.ID
rownames(COUNT) = tmp[rownames(COUNT), 'ensembl_gene_id']

# Get clinical metadata
METADATA.CLINICAL_ID <- 'syn3191087'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.CLINICAL_ID
METADATA.CLINICAL_OBJ <- synGet(METADATA.CLINICAL_ID)
METADATA.CLINICAL <- read.table(METADATA.CLINICAL_OBJ@filePath,sep=',',header=T)

# Get technical covariates
METADATA.TECH_ID <- 'syn4300313'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.TECH_ID
METADATA.TECH_OBJ <- synGet(METADATA.TECH_ID)
METADATA.TECH <- read.table(METADATA.TECH_OBJ@filePath,sep='\t',header=T)

# Get picard metrics from synapse
METADATA.PICARD_ID <- 'syn8380366'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)

METADATA.PICARD$Sampleid = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
  gsub('Aligned.out/picard.CombinedMetrics.csv','',.)
colnames(METADATA.PICARD) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD))
colnames(METADATA.PICARD) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD))

# Fix error in technical covariates data
KEY_ID <- 'syn3382527'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = KEY_ID  
KEY <- read.csv(synGet(KEY_ID)@filePath) %>% 
  filter(mrna_data == 1) %>%
  dplyr::select(projid, mrna_id) %>%
  separate(mrna_id, c('a','b','batch'), sep = '_') %>% 
  unite(Sampleid, a, b) %>%
  dplyr::select(-batch) %>%
  unique

# Match technical and clinical covariates
METADATA <- METADATA.TECH %>%
  left_join(METADATA.PICARD) %>%
  dplyr::select(-projid) %>%
  dplyr::left_join(KEY) %>%
  dplyr::left_join(METADATA.CLINICAL)

# Pick higher quality RIN batch for sample 492_120515
METADATA <- METADATA %>%
  group_by(Sampleid) %>%
  top_n(1, RINcontinuous)

# Get gene specific parameters from synapse
GENE.PARAM = downloadFile('syn8449369')
ALL_USED_IDs = c(ALL_USED_IDs, 'syn8449369')

GENE.LEN = dplyr::select(GENE.PARAM, ensembl_gene_id, gene.length) %>% unique()
rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id

GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% unique()
rownames(GENE.GC) = GENE.GC$ensembl_gene_id
```

### Data preprocessing
```{r preprocess.data,cache=TRUE, echo=TRUE}
# Remove samples with no cogdx, RIN and PMI scores
METADATA <- METADATA %>%
  ungroup %>%
  filter(Sampleid %in% colnames(COUNT)) %>%
  filter(!is.na(cogdx)) %>%
  filter(!is.na(RINcontinuous)) %>%
  filter(!is.na(pmi)) %>%
  filter(!is.na(Sampleid)) %>%
  filter(!(Sampleid %in% c('380_120503'))) %>%
  as.data.frame()
```

```{r preprocess.data1,cache=TRUE}
# Match covariates to expression data
indToRetain = intersect(METADATA$Sampleid, colnames(COUNT))
removedIDs = setdiff(colnames(COUNT), METADATA$Sampleid)

COUNT = COUNT[,indToRetain]

rownames(METADATA) = METADATA$Sampleid
METADATA = METADATA[indToRetain,]
```
Dorsolateral prefrontal cortex of `r dim(COUNT)[2]` subjects from the ROS and MAP cohorts are used for the analysis. Following sample are removed `r paste(removedIDs, collapse = ',')`

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering, cache=TRUE}
#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv", "PF_INDEL_RATE", "PF_MISMATCH_RATE"
primaryVariable <- 'cogdx'
FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx")
ContCovariates <- c("RINcontinuous","age_first_ad_dx","age_death","age_at_visit_max","pmi","educ", 
                    "PCT_PF_READS", "PCT_PF_READS_ALIGNED", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", 
                    "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES <- data.frame(lapply(COVARIATES,function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'\\+','')})}))
rownames(COVARIATES) <- METADATA$Sampleid

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)

# Add in RIN^2 values
COVARIATES$RINcontinuous2 = COVARIATES$RINcontinuous^2
ContCovariates = c(ContCovariates, 'RINcontinuous2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
```

```{r dataExplore, fig.width = 10, fig.height = 14}
# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = cogdx, y = RINcontinuous)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')

# AgeAtDeath
p[[2]] = ggplot(COVARIATES, aes(x = cogdx, y = age_death)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + theme(legend.position = 'top')

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = cogdx, y = pmi)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI') + theme(legend.position = 'top')

# Education
p[[4]] = ggplot(COVARIATES, aes(x = cogdx, y = educ)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Education') + theme(legend.position = 'top')

# Intronic bases
p[[5]] = ggplot(COVARIATES, aes(x = cogdx, y = PCT_INTRONIC_BASES)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Fraction Intronic Bases') + theme(legend.position = 'top')

# Ribosomal bases
p[[6]] = ggplot(COVARIATES, aes(x = cogdx, y = PCT_RIBOSOMAL_BASES)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Fraction Ribosomal Bases') + theme(legend.position = 'top')

multiplot(plotlist = p, cols = 2)

# Batch
vcd::mosaic(~ Batch + cogdx, data = COVARIATES)

# Sex
vcd::mosaic(~ msex + cogdx, data = COVARIATES)
```
### CPM Normalisation
Preprocess counts matrix using CQN (Conditional Quantile Normlaisation accounting for both GC content and gene length)

Remove genes that have less than 1 cpm counts in at least 50% of samples per cogdx. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('Sampleid') %>%
  dlply(.(cogdx), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$Sampleid],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  setdiff(c(GENE.GC$ensembl_gene_id[is.na(GENE.GC$percentage_gc_content)],
            GENE.LEN$ensembl_gene_id[is.na(GENE.LEN$gene.length)]))

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], 
                                                 MIN_GENE_CPM=0, 
                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

### Normalisation (with NULL)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Batch, shape=cogdx, size=RINcontinuous))
p <- p + theme_bw() + theme(legend.position="right")
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Batch', 'msex', primaryVariable)])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Batch', 'msex', primaryVariable)]))
```

```{r temp, include = F}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative design)
Since many covariates are correlated, re-normalising and re-adjusting COUNTS with an iterative design matrix. Adding Batch and msex a priori to variable selection
```{r customNorm, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Batch', 'msex');

# Assign residual covariates
residualCovars = setdiff(preAdjustedSigCovars$significantCovars, c(postAdjustCovars, primaryVariable))
covariatesEffects = preAdjustedSigCovars$Effects.significantCovars[residualCovars]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 20){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights for dispersion control
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
                            design = DM1$design,
                            weights = VOOM.GENE_EXPRESSION$weights)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, primaryVariable))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, primaryVariable))
  covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

residualSigCovars[["PC_res"]][[2]]$plotData
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2)) 
p <- p + geom_point(aes(color=Batch, shape=cogdx, size=RINcontinuous))
p <- p + theme_bw() + theme(legend.position="right")
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)
p
```

```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Batch', 'msex', primaryVariable)])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

```{r temp1, include=F}
dev.off()
gc()
```

### Differential expression analysis (with cogdx as primary variable)
Differential expression is performed on the primary variable by controlling for covariates identified above

Interpretation of cogdx scores
1. NCI, No cognitive impairment (No impaired domains)
2. MCI, Mild cognitive impairment (One impaired domain) and NO other cause of CI
3. MCI, Mild cognitive impairment (One impaired domain) AND another cause of CI
4. AD, Alzheimer's disease and NO other cause of CI (NINCDS PROB AD)
5. AD, Alzheimer's disease AND another cause of CI (NINCDS POSS AD)
6. Other dementia. Other primary cause of dementia

Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=6, fig.width=18}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable, postAdjustCovars), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)

# Fit linear model using new weights and new design
FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
            design = DESIGN$design,
            weights = VOOM.WEIGHTS$weights)

# Fit contrast
contrast = makeContrasts(contrasts=c("cogdx1-cogdx2",
                                     "cogdx1-cogdx4",
                                     "cogdx2-cogdx4"),
                         levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:3, function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('ensembl_gene_id')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique()) %>%
  dplyr::mutate(Study = 'ROSMAP', Region = 'DLPFC',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction))
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
  group_by(Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05)

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
  group_by(Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2)

kable(full_join(tmp %>%
                  dplyr::rename(DOWN_FDR_0_05 = DOWN, UP_FDR_0_05 = UP),
                tmp1 %>%
                  dplyr::rename(DOWN_FDR_0_05_FC_1.2 = DOWN, UP_FDR_0_05_FC_1.2 = UP)))

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red'))
p = p + facet_grid(.~Comparison, scales = 'fixed')
p
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de}
pl = list()
pl[[1]] = ggplot(DE, aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point()
pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE, aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE, aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

multiplot(plotlist = pl, cols = 3)
```

### SVA Adjustments for eQTL analysis
Conditioned on primary variable (cogdx) and identified covariates estimate surrogate variables using SVA package
```{r sva.adjust}
# Get (null) design matrix
MODEL0 = DESIGN$design[,-grep(primaryVariable,colnames(DESIGN$design))] # Get null model by removing variable of interest
MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
MODEL1 = DESIGN$design
MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]

# Calculate the number of surrogate variables
NUM.SV = sva::num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                     MODEL1, method = 'be', seed = 123456, B = 30)

# Estimate surrogate variables
SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                    MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv

colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
rownames(SURR.VAR) = rownames(MODEL1)

# Re-estimate voom weights
VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design = cbind(MODEL1, SURR.VAR), plot=F)

# Fit linear model using new weights and new design
VOOM.SVA.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                     design = cbind(MODEL1, SURR.VAR),
                     weights = VOOM.SVA.WEIGHTS$weights)

# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
```

### Covariates Adjustments for network analysis
Identified covariates are regressed out from the expression matrix for network analysis
```{r network.adjust}
# Get design matrix
DESIGN.NET = getDesignMatrix(COVARIATES[, postAdjustCovars, drop = F], Intercept = F)
DESIGN.NET = DESIGN.NET$design[,linColumnFinder(DESIGN.NET$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design=DESIGN.NET, plot=F)

# Fit linear model using new weights and new design
VOOM.NET.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
                     design = DESIGN.NET,
                     weights = VOOM.NET.WEIGHTS$weights)

# Residuals after normalisation
RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Set annotations
all.annotations = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay	 = 'Reprocessed RNAseq',

  tissueTypeAbrv	= 'DLPFC', 
  study = 'ROSMAP', 
  center = 'BROAD-RUSH',

  organism = 'HomoSapiens',
  consortium	= 'AMP-AD',

  normalizationStatus	= TRUE,
  modelSystem	= FALSE,

  normalizationType	= 'TMM'
)

# Code
CODE <- Folder(name = "ROSMAP Reprocessed RNASeq (Normalised)", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'ROSMAP_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('ROSMAP_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
annotations(COV_OBJ) = annotations(CODE)
annotations(COV_OBJ)$dataSubType = 'covariates'
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'ROSMAP_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('ROSMAP_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store logCPM
CQN.GENE_EXPRESSION$y %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'ROSMAP_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('ROSMAP_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store cqn offsets
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'ROSMAP_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('ROSMAP_DLPFC_offset.tsv', name = 'Gene length and GC content offset)', parentId = CODE$properties$id)
annotations(OFFSET_OBJ) = annotations(CODE)
annotations(OFFSET_OBJ)$dataSubType = 'offset'
OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store design matrix
MODEL1 %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'ROSMAP_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('ROSMAP_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
annotations(DM_OBJ) = annotations(CODE)
annotations(DM_OBJ)$dataSubType = 'designMatrix'
DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store surrogate variables
SURR.VAR %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'ROSMAP_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
SURR_OBJ = File('ROSMAP_DLPFC_SurVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
annotations(SURR_OBJ) = annotations(CODE)
annotations(SURR_OBJ)$dataSubType = 'surrogateVariables'
SURR_OBJ = synStore(SURR_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store residual gene expression for eQTL analysis
RESIDUAL.SVA.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'ROSMAP_DLPFC_eQTLResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
eEXP_OBJ = File('ROSMAP_DLPFC_eQTLResidualExpression.tsv', 
                name = 'Normalised, Covariates and SVA Adjusted Residual Expression (for eQTL)', 
               parentId = CODE$properties$id)
annotations(eEXP_OBJ) = annotations(CODE)
annotations(eEXP_OBJ)$dataSubType = 'residualGeneExpForeQTL'
eEXP_OBJ = synStore(eEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store residual gene expression for eQTL analysis
RESIDUAL.NET.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'ROSMAP_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('ROSMAP_DLPFC_netResidualExpression.tsv', 
                name = 'Normalised, Covariates Removed Residual Expression (for network analysis)', 
               parentId = CODE$properties$id)
annotations(nEXP_OBJ) = annotations(CODE)
annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store differential expression results
write.table(DE, file = 'ROSMAP_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('ROSMAP_DLPFC_DiffExpression.tsv', name = 'Differential Expression Results (cogdx)', 
                parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene lengeth and gc content)   |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

### 2017-03-15 code/Rmd/ROSMAP_REPROCESSED.Rmd
Minor bug fix in ROSMAP reprocessing code

@@
 ```{r covariates.clustering, cache=TRUE}
-#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv", "PF_INDEL_RATE", "PF_MISMATCH_RATE"
-primaryVariable <- 'cogdx'
+#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv", "PF_INDEL_RATE", "PF_MISMATCH_RATE", "age_first_ad_dx","age_at_visit_max",
+primaryVariable <- "cogdx"
 FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx")
-ContCovariates <- c("RINcontinuous","age_first_ad_dx","age_death","age_at_visit_max","pmi","educ", 
-                    "PCT_PF_READS", "PCT_PF_READS_ALIGNED", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", 
-                    "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")
+ContCovariates <- c("RINcontinuous", "age_death", "pmi", "educ", "PCT_PF_READS", "PCT_PF_READS_ALIGNED",
+                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")
 
@@
 Correlation/association between covariates at an FDR <= 0.1
-```{r covariates.correlation, fig.width=10, fig.height=10}
+```{r covariates.correlation, fig.width=6, fig.height=6}
 COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
@@
 
-```{r dataExplore, fig.width = 10, fig.height = 14}
+```{r data.explore, fig.width = 10, fig.height = 14}
 # RIN
@@
 Clustering of voom adjusted data (with NULL design)
-```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
+```{r decompse.normalise.data1.1, fig.height=6, fig.width=8, results='asis'}
 # Find principal components of expression to plot
@@
 Tree based classification of samples
-```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
+```{r decompse.normalise.data1.2, fig.height=6, fig.width=10, results='asis'}
 # Eucledian tree based analysis
@@
 Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
-```{r preAdjusted.covariates, cache=TRUE}
+```{r preadj.covariates, cache=TRUE}
 # Find correlation between PC's of gene expression with covariates
@@
 Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
-```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
+```{r preadj.covariates.plot, fig.width=12, fig.height=8}
 preAdjustedSigCovars[["PC_res"]][[2]]$plotData
@@
 Since many covariates are correlated, re-normalising and re-adjusting COUNTS with an iterative design matrix. Adding Batch and msex a priori to variable selection
-```{r customNorm, results='asis'}
+```{r iterative.norm, results='asis'}
 # Primary variable of interest
@@
 residualCovars = setdiff(preAdjustedSigCovars$significantCovars, c(postAdjustCovars, primaryVariable))
+residualSigCovars = preAdjustedSigCovars
 covariatesEffects = preAdjustedSigCovars$Effects.significantCovars[residualCovars]
@@
 ### Sanity check
-```{r residualSigCovars.manual, fig.width=12, fig.height=8}
+```{r residual.adj, fig.width=12, fig.height=8}
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
@@
 ### Clustering residual data
-```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
+```{r decompse.normalise.data2.1, fig.height=6, fig.width=8, results='asis'}
 # Find principal components of expression to plot
@@
 
-```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
+```{r decompse.normalise.data2.2, fig.height=8, fig.width=12, results='asis'}
 # Eucledian tree based analysis
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=6, fig.width=18}
+```{r diff.exp, fig.height=6, fig.width=18}
 # Get design matrix
@@
 ### Associate differential expression results with gc content, gene length and average expression
-```{r associate.de}
+```{r associate.de, fig.height=10, fig.width=15}
 pl = list()
 pl[[1]] = ggplot(DE, aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
+pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
 pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
@@
 pl[[2]] = ggplot(DE, aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
+pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
 pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
@@
 pl[[3]] = ggplot(DE, aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
+pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
 pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
@@
   write.table(file = 'ROSMAP_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('ROSMAP_DLPFC_offset.tsv', name = 'Gene length and GC content offset)', parentId = CODE$properties$id)
+OFFSET_OBJ = File('ROSMAP_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
 annotations(OFFSET_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'ROSMAP_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
+  write.table(file = 'ROSMAP_DLPFC_SurVar.tsv', sep = '\t', row.names=F, quote=F)
 SURR_OBJ = File('ROSMAP_DLPFC_SurVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
@@
 eEXP_OBJ = File('ROSMAP_DLPFC_eQTLResidualExpression.tsv', 
-                name = 'Normalised, Covariates and SVA Adjusted Residual Expression (for eQTL)', 
+                name = 'Normalised, covariates and surrogate variable adjusted residual expression (for eQTL)',
                 parentId = CODE$properties$id)
@@
 nEXP_OBJ = File('ROSMAP_DLPFC_netResidualExpression.tsv', 
-                name = 'Normalised, Covariates Removed Residual Expression (for network analysis)', 
+                name = 'Normalised, covariates removed residual expression (for network analysis)', 
                 parentId = CODE$properties$id)
@@
 |  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
-|  Offset (for gene lengeth and gc content)   |  `r OFFSET_OBJ$properties$id`    |
+|  Offset (for gene length and gc content)   |  `r OFFSET_OBJ$properties$id`    |
 |  Design Matrix                              |  `r DM_OBJ$properties$id`        |

### 2017-03-17 code/Rmd/MAYO_REPROCESSED.Rmd
Added markdown to process MAYO reprocessed rnaseq with cqn and covariates adjustments with and without sva

---
title: "Covariate analysis of MAYO reprocessed counts (with CQN normalisation and covariates adjusted)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED.Rmd",
                                 parentId = "syn7554308",
                                 entityName = 'MAYO Reprocessed RNASeq (Normalised)')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(reshape2)
library(limma)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)
library(cqn)

library(synapseClient)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

library(parallel)
library(foreach)

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

synapseLogin()

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn8403743';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of MAYO reprocessed RNASeq data';

thisFileName <- 'MAYO_REPROCESSED.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Get reprocessed sample counts for temporal cortex
COUNT_TC_ID <- 'syn7987765';
ALL_USED_IDs <- COUNT_TC_ID
COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F)
colnames(COUNT_TC) = paste(colnames(COUNT_TC), 'TCX', sep = '_')

# Get reprocessed sample counts for cerebellum
COUNT_CE_ID <- 'syn7513807';
ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_CE_ID
COUNT_CE <- read.table(synGet(COUNT_CE_ID)@filePath, header=T, sep='\t', check.names = F)

COUNT = cbind(COUNT_TC, COUNT_CE[rownames(COUNT_TC),])

# Convert rownames of counts from tracking id to ensemble gene id
tmp = data.frame(Gene.ID = rownames(COUNT)) %>%
  dplyr::mutate(ID = Gene.ID) %>%
  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')
rownames(tmp) = tmp$Gene.ID
rownames(COUNT) = tmp[rownames(COUNT), 'ensembl_gene_id']

# Get clinical metadata
METADATA_TC_ID <- 'syn3817650'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TC_ID
METADATA_TC <- read.table(synGet(METADATA_TC_ID)@filePath,sep='\t',header=T, stringsAsFactors=F) %>%
  tidyr::separate(ID, c('Donor_ID', 'BrainRegion'), sep = '_') %>%
  dplyr::mutate(ID = paste(Donor_ID, BrainRegion, sep = '_'))

# Get clinical metadata
METADATA_CE_ID <- 'syn5223705'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CE_ID
METADATA_CE <- read.table(synGet(METADATA_CE_ID)@filePath,sep='\t',header=T, stringsAsFactors=F) %>%
  tidyr::separate(SampleID, c('Donor_ID', 'BrainRegion')) %>%
  dplyr::mutate(ID = paste(Donor_ID, BrainRegion, sep = '_')) %>%
  dplyr::rename(Gender = Sex, FLOWCELL = Flowcell)

# Get picard metrics from synapse
METADATA.PICARD_CE_ID <- 'syn8380392'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_CE_ID
METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = FALSE, exdir = getwd())
METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = TRUE, exdir = getwd())
METADATA.PICARD_CE_FPATHS = METADATA.PICARD_CE_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_CE_FPATHS)]
METADATA.PICARD_CE <- plyr::ldply(METADATA.PICARD_CE_FPATHS, read.csv)

METADATA.PICARD_CE$ID = gsub('picard/', '', METADATA.PICARD_CE_FPATHS) %>%
  gsub('/picard.CombinedMetrics.csv','',.) %>%
  sapply(function(x){str_split(x, '\\.')[[1]][1]})
colnames(METADATA.PICARD_CE) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_CE))
colnames(METADATA.PICARD_CE) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD_CE))

METADATA.PICARD_TC_ID <- 'syn8383233'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_TC_ID
METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = FALSE, exdir = getwd())
METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = TRUE, exdir = getwd())
METADATA.PICARD_TC_FPATHS = METADATA.PICARD_TC_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_TC_FPATHS)]
METADATA.PICARD_TC <- plyr::ldply(METADATA.PICARD_TC_FPATHS, read.csv)

METADATA.PICARD_TC$ID = gsub('picard/', '', METADATA.PICARD_TC_FPATHS) %>%
  gsub('/picard.CombinedMetrics.csv','',.) %>%
  sapply(function(x){str_split(x, '\\.')[[1]][1]}) %>%
  gsub('Aligned','',.)
colnames(METADATA.PICARD_TC) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_TC))
colnames(METADATA.PICARD_TC) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD_TC))

# Merge all metadata
METADATA = rbindlist(list(METADATA_TC, METADATA_CE), use.names = T, fill = T) %>%
  inner_join(rbindlist(list(METADATA.PICARD_TC, METADATA.PICARD_CE), use.names = T, fill = T)) %>%
  as.data.frame

# Get gene specific parameters from synapse
GENE.PARAM = downloadFile('syn8449369')
ALL_USED_IDs = c(ALL_USED_IDs, 'syn8449369')

GENE.LEN = dplyr::select(GENE.PARAM, ensembl_gene_id, gene.length) %>% unique()
rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id

GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% unique()
rownames(GENE.GC) = GENE.GC$ensembl_gene_id
```

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
METADATA$BrainRegion.Diagnosis = paste(METADATA$BrainRegion, METADATA$Diagnosis, sep = '.')
METADATA$RIN2 = METADATA$RIN^2
METADATA <- METADATA %>%
  filter(!is.na(Source)) %>%
  dplyr::filter(!(ID %in% c("11311_CER", "1923_CER", "1923_TCX",
                            "11396_TCX", "11294_TCX", "11408_TCX")))
```
Following samples were considered outliers and removed based on PCA and/or eucledian distance: 11311\_CER, 1923\_CER, 1923\_TCX, 11396\_TCX, 11294\_TCX, 11408\_TCX
```{r preprocess.data1}
# Match covariates to expression data
indToRetain = intersect(METADATA$ID, colnames(COUNT))
indRemoved = setdiff(colnames(COUNT), METADATA$ID)

COUNT <- COUNT[, indToRetain]

rownames(METADATA) = METADATA$ID
METADATA = METADATA[indToRetain,]
```
`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project.

Following samples were removed `r indRemoved`

### Predict missing PMI
```{r predict.pmi}
PMI = METADATA[,c('PMI', 'ID'), drop = F]
PMI$PMI = PMI$PMI * 60
rownames(PMI) = METADATA$ID

# Get genes that are predictors of PMI and use them as control probes
CONTROL.PROBES_ID = 'syn6145639'
ALL_USED_IDs = c(ALL_USED_IDs, CONTROL.PROBES_ID)
CONTROL.PROBES = fread(synGet(CONTROL.PROBES_ID)@filePath, data.table = F) %>%
  dplyr::mutate(ID = x) %>%
  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')

# Conditional quantile normalisation
tmp.count = COUNT
genesToAnalyze = unique(intersect(rownames(GENE.GC)[!is.na(GENE.GC$percentage_gc_content)], 
                                  rownames(GENE.LEN)[!is.na(GENE.LEN$gene.length)]))
tmp.count = tmp.count[rownames(tmp.count) %in% genesToAnalyze,]
CQN.GENE_EXPRESSION = cqn(tmp.count,
                          x = GENE.GC[rownames(tmp.count), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(tmp.count), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

# Initial normalisation of gene expression
ind = rownames(CQN.GENE_EXPRESSION$y) %in% CONTROL.PROBES$ensembl_gene_id 
EXPR = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
EXPR = EXPR[ind,]
PMI = PMI[colnames(EXPR),,drop=F]

expr = dlply(PMI, .(ID), function(mtd, expr){
  if (dim(mtd)[1] > 1)
    expr = rowMeans(expr[, mtd$ID]) %>% rownameToFirstColumn('GeneID')
  else
    expr = expr[, mtd$ID] %>% rownameToFirstColumn('GeneID')
  expr = expr %>% plyr::rename(c('DF' = unique(mtd$ID)))
  return(expr)
}, EXPR) %>%
  join_all
rownames(expr) = expr$GeneID

PMI = PMI[,colnames(PMI) %in% c('ID','PMI')] %>% unique
rownames(PMI) = PMI$ID

PMI = PMI[intersect(rownames(PMI), colnames(expr)), ]
expr = expr[,intersect(rownames(PMI), colnames(expr))]

# Split data into trainining and prediction set
x.train = expr[,!is.na(PMI$PMI)]
x.predict = expr[,is.na(PMI$PMI)]

y.train = PMI[!is.na(PMI$PMI),]

plr.model = cv.glmnet(t(x.train), y.train$PMI, family = 'gaussian')
coeff = as.matrix(coef.glmnet(plr.model, s = 'lambda.min'))
mse = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train$PMI)^2)/length(y.train$PMI)
y.predict = predict(plr.model, t(x.predict), s = 'lambda.min', type = 'response') %>%
  rownameToFirstColumn('ID') %>%
  plyr::rename(c('1' = 'PMI'))

PMI = rbindlist(list(y.train, y.predict), use.names = T, fill = T)
PMI$PMI[PMI$PMI < 0] = 0

METADATA = METADATA %>% 
  dplyr::select(-PMI) %>%
  left_join(PMI)
```

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering, cache=TRUE}
primaryVariable <- "BrainRegion.Diagnosis"
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI", "PCT_PF_READS", "PCT_PF_READS_ALIGNED",
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$ID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=6, fig.height=6}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
```

```{r data.explore, fig.width = 12, fig.height = 8}
my.theme = theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# AgeAtDeath
p[[2]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(AgeAtDeath))) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeAtDeath') + my.theme

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PMI))) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI') + my.theme

# Ribosomal bases
p[[4]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_RIBOSOMAL_BASES))) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Fraction of Ribosomal Bases') + my.theme

# Intronic bases
p[[5]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTRONIC_BASES))) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Fraction Intronic Bases') + my.theme

# Intergenic bases
p[[6]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTERGENIC_BASES))) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Fraction Intergenic Bases') + my.theme

multiplot(plotlist = p, cols = 3)
```
### CPM Normalisation
Preprocess counts matrix using CQN (Conditional Quantile Normlaisation accounting for both GC content and gene length)

Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion x Diagnosis. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('SampleID') %>%
  dlply(.(BrainRegion.Diagnosis), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC$ensembl_gene_id[!is.na(GENE.GC$percentage_gc_content)]) %>%
  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$gene.length)]) %>%
  setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous"))

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], 
                                                 MIN_GENE_CPM=0, 
                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

### Normalisation (with NULL)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=6, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('Region','Diagnosis'), sep = '\\.') %>% 
  dplyr::mutate(Region = factor(Region),
                Source = factor(Source))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Region~., scales = 'free_y')
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data1.2, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = (COVARIATES[,c(FactorCovariates), drop = F])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

```{r temp, include = F}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
```{r preadj.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preadj.covariates.plot, fig.width=12, fig.height=8}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative design)
Since many covariates are correlated, re-normalising and re-adjusting COUNTS with an iterative design matrix.

NOTE:
1. Using a mixed effect model where random effect is chosen as Donor_ID
2. Adding Source, FLOWCELL and Gender a priori to variable selection
3. Primary variable of interest BrainRegion.Diagnosis is excluded from the pool of available covariates for selection
```{r iterative.norm, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Source', 'FLOWCELL', 'Gender');

# Assign residual covariates
residualCovars = setdiff(preAdjustedSigCovars$significantCovars, 
                         c(postAdjustCovars, primaryVariable, 'Donor_ID'))
residualSigCovars = preAdjustedSigCovars
covariatesEffects = preAdjustedSigCovars$Effects.significantCovars[residualCovars]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 20){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects and Donor_ID as random effect'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights for dispersion control
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)

  # Calculate correlation values of random effects
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
                                             block = COVARIATES$Donor_ID)

  # Re-estimate voom weights
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                              design=DM1$design, plot=F, 
                              block = COVARIATES$Donor_ID, 
                              correlation = correlation$cor)

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
                            design = DM1$design,
                            weights = VOOM.GENE_EXPRESSION$weights,
                            block = COVARIATES$Donor_ID, 
                            correlation = correlation$cor)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates),
                         c(postAdjustCovars, primaryVariable, 'Donor_ID'))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, primaryVariable))
  covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```

### Sanity check
```{r residual.adj, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

residualSigCovars[["PC_res"]][[2]]$plotData
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2.1, fig.height=10, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('Region','Diagnosis'), sep = '\\.') %>% 
  dplyr::mutate(Region = factor(Region),
                Source = factor(Source))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Region~., scales = 'free_y')
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```

```{r decompse.normalise.data2.2, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

```{r temp1, include=F}
dev.off()
gc()
```

### Differential expression analysis (with BrainRegion.Diagnosis as primary variable)
Differential expression is performed on the primary variable by controlling for covariates identified above

Interpretation of BrainRegion.Diagnosis
1. TCX.PSP: Progressive Supranuclear Palsy samples from Temporal cortex region
2. TCX.AD: Alzheimers Disease samples from Temporal cortex region
3. TCX.Control: Control samples from Temporal cortex region
4. TCX.Pathologic_Aging: Pathological Aging samples from Temporal cortex region
5. CER.PSP: Progressive Supranuclear Palsy samples from Cerebellum region
6. CER.AD: Alzheimers Disease samples from Cerebellum region
7. CER.Control: Control samples from Cerebellum region
8. CER.Pathologic_Aging: Pathological Aging samples from Cerebellum region

Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp, fig.height=6, fig.width=18}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable, postAdjustCovars), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)

# Calculate correlation values of random effects
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Donor_ID)

# Re-estimate voom weights
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$Donor_ID, 
                    correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
FIT = lmFit(VOOM.WEIGHTS)

# Fit contrast
contrast = makeContrasts(contrasts=c(
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.Pathologic_Aging",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisCER.PSP-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.Pathologic_Aging",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.PSP",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.PSP",
  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD"),
  levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:12, function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('ensembl_gene_id')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
                Study = 'MAYO',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('Region','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('Region','against'), sep = '\\.') %>%
  tidyr::unite(Comparison, reference, against, sep = '-') %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction, Region) %>%
  group_by(Region, Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05)

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction, Region) %>%
  group_by(Region, Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2)

kable(full_join(tmp %>%
                  dplyr::rename(DOWN_FDR_0_05 = DOWN, UP_FDR_0_05 = UP),
                tmp1 %>%
                  dplyr::rename(DOWN_FDR_0_05_FC_1.2 = DOWN, UP_FDR_0_05_FC_1.2 = UP)))

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red'))
p = p + facet_grid(Region+.~Comparison, scales = 'fixed')
p
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=10, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE %>%
                   dplyr::filter(Comparison == 'Control-AD'), 
                 aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE %>%
                   dplyr::filter(Comparison == 'Control-AD'), 
                 aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE %>%
                   dplyr::filter(Comparison == 'Control-AD'), 
                 aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

multiplot(plotlist = pl, cols = 3)
```

### SVA Adjustments for eQTL analysis
Conditioned on primary variable (BrainRegion.Diagnosis) and identified covariates estimate surrogate variables using SVA package
```{r sva.adjust}
# Get (null) design matrix
MODEL0 = DESIGN$design[,-grep(primaryVariable,colnames(DESIGN$design))] # Get null model by removing variable of interest
MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
MODEL1 = DESIGN$design
MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]

# Source modified functions from sva package
source('../R/num.sv.R')
source('../R/irwsva.build.R')
source('../R/f.pvalue.R')

# Calculate the number of surrogate variables
NUM.SV = num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                     MODEL1, method = 'be', seed = 123456, B = 30, tol = 1e-20)

# Estimate surrogate variables
SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
                        MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv

colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
rownames(SURR.VAR) = rownames(MODEL1)

# Estimate voom weights for dispersion control
VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design=cbind(MODEL1, SURR.VAR), plot=F)

# Calculate correlation values of random effects
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$Donor_ID)

# Re-estimate voom weights
VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design = cbind(MODEL1, SURR.VAR), plot=F,
                        block = COVARIATES$Donor_ID, 
                        correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)

# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
```

### Covariates Adjustments for network analysis
Identified covariates are regressed out from the expression matrix for network analysis
```{r network.adjust}
# Get design matrix
DESIGN.NET = getDesignMatrix(COVARIATES[, postAdjustCovars, drop = F], Intercept = F)
DESIGN.NET = DESIGN.NET$design[,linColumnFinder(DESIGN.NET$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design=DESIGN.NET, plot=F)

# Calculate correlation values of random effects
VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$Donor_ID)

# Re-estimate voom weights
VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design = DESIGN.NET, plot=F,
                        block = COVARIATES$Donor_ID, 
                        correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)

# Residuals after normalisation
RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Set annotations
all.annotations = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay	 = 'Reprocessed RNAseq',

  tissueTypeAbrv	= 'CBE, TCX', 
  study = 'MayoRNASeq', 
  center = 'UFL-MAYO-ISB',

  organism = 'HomoSapiens',
  consortium	= 'AMP-AD',

  normalizationStatus	= TRUE,
  modelSystem	= FALSE,

  normalizationType	= 'CQN'
)

# Code
CODE <- Folder(name = "MAYO Reprocessed RNASeq (Normalised)", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'MAYO_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('MAYO_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
annotations(COV_OBJ) = annotations(CODE)
annotations(COV_OBJ)$dataSubType = 'covariates'
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MAYO_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('MAYO_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store logCPM
CQN.GENE_EXPRESSION$y %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MAYO_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('MAYO_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store cqn offsets
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MAYO_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('MAYO_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
annotations(OFFSET_OBJ) = annotations(CODE)
annotations(OFFSET_OBJ)$dataSubType = 'offset'
OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store design matrix
MODEL1 %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'MAYO_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('MAYO_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
annotations(DM_OBJ) = annotations(CODE)
annotations(DM_OBJ)$dataSubType = 'designMatrix'
DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store surrogate variables
SURR.VAR %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'MAYO_DLPFC_SurVar.tsv', sep = '\t', row.names=F, quote=F)
SURR_OBJ = File('MAYO_DLPFC_SurVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
annotations(SURR_OBJ) = annotations(CODE)
annotations(SURR_OBJ)$dataSubType = 'surrogateVariables'
SURR_OBJ = synStore(SURR_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store residual gene expression for eQTL analysis
RESIDUAL.SVA.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MAYO_DLPFC_eQTLResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
eEXP_OBJ = File('MAYO_DLPFC_eQTLResidualExpression.tsv', 
                name = 'Normalised, covariates and surrogate variable adjusted residual expression (for eQTL)',
                parentId = CODE$properties$id)
annotations(eEXP_OBJ) = annotations(CODE)
annotations(eEXP_OBJ)$dataSubType = 'residualGeneExpForeQTL'
eEXP_OBJ = synStore(eEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store residual gene expression for eQTL analysis
RESIDUAL.NET.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MAYO_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('MAYO_DLPFC_netResidualExpression.tsv', 
                name = 'Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
annotations(nEXP_OBJ) = annotations(CODE)
annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store differential expression results
write.table(DE, file = 'MAYO_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('MAYO_DLPFC_DiffExpression.tsv', 
                name = 'Differential Expression Results (BrainRegion.Diagnosis)', 
                parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
stopCluster(cl)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

### 2017-03-17 code/Rmd/MAYO_REPROCESSED.Rmd
Modified irwsva.build, num.sv and f.pvalue codes from sva package to include tolerance

@@
 knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED.Rmd",
-                                 parentId = "syn7554308",
+                                 parentId = "syn8403743",
                                  entityName = 'MAYO Reprocessed RNASeq (Normalised)')
@@
 # Synapse parameters
-parentId = 'syn8403743';
+parentId = 'syn8403743'
 activityName = 'Covariate analysis';
@@
 ### Associate differential expression results with gc content, gene length and average expression
-```{r associate.de, fig.height=10, fig.width=15}
+```{r associate.de, fig.height=5, fig.width=15}
 pl = list()

### 2017-03-17 code/R/irwsva.build .R

@@ -1,0 +1,40 @@
+irwsva.build <- function (dat, mod, mod0 = NULL, n.sv, B = 5, tol = 1e-18) 
+{
+  n <- ncol(dat)
+  m <- nrow(dat)
+  if (is.null(mod0)) {
+    mod0 <- mod[, 1]
+  }
+  Id <- diag(n)
+  resid <- dat %*% (Id - mod %*% solve(t(mod) %*% mod, tol = tol) %*% 
+                      t(mod))
+  uu <- eigen(t(resid) %*% resid)
+  vv <- uu$vectors
+  ndf <- n - dim(mod)[2]
+  pprob <- rep(1, m)
+  one <- rep(1, n)
+  Id <- diag(n)
+  df1 <- dim(mod)[2] + n.sv
+  df0 <- dim(mod0)[2] + n.sv
+  rm(resid)
+  cat(paste("Iteration (out of", B, "):"))
+  for (i in 1:B) {
+    mod.b <- cbind(mod, uu$vectors[, 1:n.sv])
+    mod0.b <- cbind(mod0, uu$vectors[, 1:n.sv])
+    ptmp <- f.pvalue(dat, mod.b, mod0.b)
+    pprob.b <- (1 - edge.lfdr(ptmp))
+    mod.gam <- cbind(mod0, uu$vectors[, 1:n.sv])
+    mod0.gam <- cbind(mod0)
+    ptmp <- f.pvalue(dat, mod.gam, mod0.gam)
+    pprob.gam <- (1 - edge.lfdr(ptmp))
+    pprob <- pprob.gam * (1 - pprob.b)
+    dats <- dat * pprob
+    dats <- dats - rowMeans(dats)
+    uu <- eigen(t(dats) %*% dats)
+    cat(paste(i, " "))
+  }
+  sv = svd(dats)$v[, 1:n.sv]
+  retval <- list(sv = sv, pprob.gam = pprob.gam, pprob.b = pprob.b, 
+                 n.sv = n.sv)
+  return(retval)
+}

### 2017-03-17 code/R/num.sv.R

@@ -1,5 +1,5 @@
 num.sv <- function (dat, mod, method = c("be", "leek"), vfilter = NULL, 
-                    B = 20, seed = NULL, LINPACK = F) 
+                    B = 20, seed = NULL, LINPACK = F, tol = 1e-18) 
 {
   if (!is.null(vfilter)) {
     if (vfilter < 100 | vfilter > dim(dat)[1]) {
@@ -18,7 +18,7 @@ num.sv <- function (dat, mod, method = c("be", "leek"), vfilter = NULL,
     warn <- NULL
     n <- ncol(dat)
     m <- nrow(dat)
-    H <- mod %*% solve(t(mod) %*% mod) %*% t(mod)
+    H <- mod %*% solve(t(mod) %*% mod, tol = tol) %*% t(mod)
     res <- dat - t(H %*% t(dat))
     uu <- svd(res, LINPACK = LINPACK)
     ndf <- min(m, n) - ceiling(sum(diag(H)))
@@ -51,7 +51,7 @@ num.sv <- function (dat, mod, method = c("be", "leek"), vfilter = NULL,
     a <- seq(0, 2, length = 100)
     n <- floor(dims[1]/10)
     rhat <- matrix(0, nrow = 100, ncol = 10)
-    P <- (diag(dims[2]) - mod %*% solve(t(mod) %*% mod) %*% 
+    P <- (diag(dims[2]) - mod %*% solve(t(mod) %*% mod, tol = tol) %*% 
             t(mod))
     for (j in 1:10) {
       dats <- dat[1:(j * n), ]

### 2017-03-17 code/Rmd/MAYO_REPROCESSED.Rmd
Changed destination location for MAYO reprocessed RNASeq analysis

@@
 knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED.Rmd",
-                                 parentId = "syn8403743",
+                                 parentId = "syn7554308",
                                  entityName = 'MAYO Reprocessed RNASeq (Normalised)')
@@
 # Synapse parameters
-parentId = 'syn8403743'
+parentId = 'syn7554308'
 activityName = 'Covariate analysis';
@@
 COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
-write.table(COVARIATES, file = 'MAYO_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
-COV_OBJ = File('MAYO_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+write.table(COVARIATES, file = 'MAYO_CBE_TCX_Covariates.tsv', sep = '\t', row.names=F, quote=F)
+COV_OBJ = File('MAYO_CBE_TCX_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
 annotations(COV_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MAYO_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('MAYO_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
+  write.table(file = 'MAYO_CBE_TCX_Counts.tsv', sep = '\t', row.names=F, quote=F)
+COUNT_OBJ = File('MAYO_CBE_TCX_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
 annotations(COUNT_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MAYO_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('MAYO_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
+  write.table(file = 'MAYO_CBE_TCX_logCPM.tsv', sep = '\t', row.names=F, quote=F)
+LCOUNT_OBJ = File('MAYO_CBE_TCX_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
 annotations(LCOUNT_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MAYO_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('MAYO_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
+  write.table(file = 'MAYO_CBE_TCX_offset.tsv', sep = '\t', row.names=F, quote=F)
+OFFSET_OBJ = File('MAYO_CBE_TCX_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
 annotations(OFFSET_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'MAYO_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('MAYO_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+  write.table(file = 'MAYO_CBE_TCX_Design.tsv', sep = '\t', row.names=F, quote=F)
+DM_OBJ = File('MAYO_CBE_TCX_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
 annotations(DM_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'MAYO_DLPFC_SurVar.tsv', sep = '\t', row.names=F, quote=F)
-SURR_OBJ = File('MAYO_DLPFC_SurVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
+  write.table(file = 'MAYO_CBE_TCX_SurVar.tsv', sep = '\t', row.names=F, quote=F)
+SURR_OBJ = File('MAYO_CBE_TCX_SurVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
 annotations(SURR_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MAYO_DLPFC_eQTLResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-eEXP_OBJ = File('MAYO_DLPFC_eQTLResidualExpression.tsv', 
+  write.table(file = 'MAYO_CBE_TCX_eQTLResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+eEXP_OBJ = File('MAYO_CBE_TCX_eQTLResidualExpression.tsv', 
                 name = 'Normalised, covariates and surrogate variable adjusted residual expression (for eQTL)',
@@
 
-# Store residual gene expression for eQTL analysis
+# Store residual gene expression for network analysis
 RESIDUAL.NET.GENE_EXPRESSION %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MAYO_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('MAYO_DLPFC_netResidualExpression.tsv', 
+  write.table(file = 'MAYO_CBE_TCX_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+nEXP_OBJ = File('MAYO_CBE_TCX_netResidualExpression.tsv', 
                 name = 'Normalised, covariates removed residual expression (for network analysis)', 
@@
 # Store differential expression results
-write.table(DE, file = 'MAYO_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('MAYO_DLPFC_DiffExpression.tsv', 
+write.table(DE, file = 'MAYO_CBE_TCX_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('MAYO_CBE_TCX_DiffExpression.tsv', 
                 name = 'Differential Expression Results (BrainRegion.Diagnosis)', 

### 2017-03-17 code/Rmd/ROSMAP_REPROCESSED.Rmd
Changed destination location for MAYO reprocessed RNASeq analysis

@@
   unique() %>%
-  setdiff(c(GENE.GC$ensembl_gene_id[is.na(GENE.GC$percentage_gc_content)],
-            GENE.LEN$ensembl_gene_id[is.na(GENE.LEN$gene.length)]))
+  intersect(GENE.GC$ensembl_gene_id[!is.na(GENE.GC$percentage_gc_content)]) %>%
+  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$gene.length)]) %>%
+  setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous"))
 
@@
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Batch', 'msex', primaryVariable)])
+COVARIATES.tmp = (COVARIATES[,c(FactorCovariates), drop = F])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
 
-tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
+tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
                            main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp[,c('Batch', 'msex', primaryVariable)]))
+                           groupLabels = colnames(COVARIATES.tmp))
 ```
@@
 ### Significant Covariates
-Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
+Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
 ```{r preadj.covariates, cache=TRUE}
@@
 ### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising and re-adjusting COUNTS with an iterative design matrix. Adding Batch and msex a priori to variable selection
+Since many covariates are correlated, re-normalising and re-adjusting COUNTS with an iterative design matrix.
+1. Adding FLOWCELL and Gender a priori to variable selection
+2. Primary variable of interest BrainRegion.Diagnosis is excluded from the pool of available covariates for selection
 ```{r iterative.norm, results='asis'}
@@
 
-  normalizationType	= 'TMM'
+  normalizationType	= 'CQN'
 )
@@
 
-# Store residual gene expression for eQTL analysis
+# Store residual gene expression for network analysis
 RESIDUAL.NET.GENE_EXPRESSION %>%

### 2017-03-17 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Added markdown to process STAR feature counts aligned to Ensembl v75 with CQN normalisation accounting for gc content and gene length

---
title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
                                 parentId ="syn7501880",
                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(reshape2)
library(limma)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)
library(cqn)

library(synapseClient)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

synapseLogin()

library(doParallel)
library(foreach)

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn7501880';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (ACC and DLPFC)';

thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download reprocessed counts (ACC)
COUNT_ACC_ID = 'syn8414424';
ALL_USED_IDs = COUNT_ACC_ID
COUNT_ACC = downloadFile(COUNT_ACC_ID)

# Download reprocessed counts (DLPFC)
COUNT_DLPFC_ID = 'syn8413221';
ALL_USED_IDs[length(ALL_USED_IDs)+1] = COUNT_DLPFC_ID
COUNT_DLPFC = downloadFile(COUNT_DLPFC_ID)

# Join ACC and DLPFC together
COUNT = join_all(list(COUNT_ACC, COUNT_DLPFC), type = 'full')

GENE.LEN = dplyr::select(COUNT_ACC, Geneid, Length) %>%
  dplyr::rename(ensembl_gene_id = Geneid) %>%
  unique()
rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id

rownames(COUNT) = COUNT$Geneid
COUNT[,c('Geneid', 'Chr', 'Start', 'End', 'Strand', 'Length')] = NULL
COUNT = data.matrix(COUNT)

# Get merged metadata
METADATA_ID = 'syn2299154' #'syn7574833';#
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA = read.csv(synGet(METADATA_ID)@filePath, fill = T, header = F)

# Get ancestry vector calculated using gemtools
ANCESTRY_ID = 'syn2511399'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = ANCESTRY_ID
ANCESTRY = read.table(synGet(ANCESTRY_ID)@filePath, fill = T, header = T, sep = '\t') %>%
  plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'GenotypingSampleID'))

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", "Dx",

                          "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?", "ACC_RNA_report: Library Batch",
                          "ACC_RNA_report: Ribozero Batch", "ACC_RNA_report: Flowcell Batch", 
                          "ACC_RNA_report: Mapped Reads", "ACC_RNA_report: Intragenic Rate",
                          "ACC_RNA_report: Intronic Rate", "ACC_RNA_report: Intergenic Rate",
                          "ACC_RNA_report: Genes Detected", "ACC_RNA_report: Transcripts Detected",
                          "ACC_RNA_report: Expression Profiling Efficiency", "ACC_RNA_report: rRNA Rate",
                          "ACC_RNA_report: Total Reads", "ACC_RNA_report: Percent Aligned",

                          "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
                          "DLPFC_RNA_report: Ribozero Batch", "DLPFC_RNA_report: Flowcell Batch",
                          "DLPFC_RNA_report: Mapped Reads", "DLPFC_RNA_report: Intragenic Rate",
                          "DLPFC_RNA_report: Intronic Rate", "DLPFC_RNA_report: Intergenic Rate", 
                          "DLPFC_RNA_report: Genes Detected", "DLPFC_RNA_report: Transcripts Detected", 
                          "DLPFC_RNA_report: Expression Profiling Efficiency", "DLPFC_RNA_report: rRNA Rate",
                          "DLPFC_RNA_report: Total Reads", "DLPFC_RNA_report: Percent Aligned",

                          "DNA_genotyping_report: Genotyping Sample_ID")

# Filter metadata 
ind = as.character(METADATA[1,] %>% unlist) %in% metadataVarOfInterest
METADATA = METADATA[,ind]
colnames(METADATA) = gsub('[^[:alnum:]]','_',as.character(METADATA[1,] %>% unlist))

METADATA_ACC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                ACC_RNA_isolation__Exclude_, ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN,
                ACC_RNA_report__Exclude_, ACC_RNA_report__Ribozero_Batch, ACC_RNA_report__Library_Batch, 
                ACC_RNA_report__Flowcell_Batch, ACC_RNA_report__Mapped_Reads, ACC_RNA_report__Intragenic_Rate,
                ACC_RNA_report__Intronic_Rate, ACC_RNA_report__Intergenic_Rate, ACC_RNA_report__Genes_Detected,
                ACC_RNA_report__Transcripts_Detected, ACC_RNA_report__Expression_Profiling_Efficiency, 
                ACC_RNA_report__rRNA_Rate, ACC_RNA_report__Total_Reads, ACC_RNA_report__Percent_Aligned,
                DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = ACC_RNA_isolation__Exclude_, 
                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
                RIN = ACC_RNA_isolation__RIN, 
                ReportExclude = ACC_RNA_report__Exclude_, 
                RibozeroBatch = ACC_RNA_report__Ribozero_Batch,
                LibraryBatch = ACC_RNA_report__Library_Batch,
                FlowcellBatch = ACC_RNA_report__Flowcell_Batch, 
                MappedReads = ACC_RNA_report__Mapped_Reads, 
                IntragenicRate = ACC_RNA_report__Intragenic_Rate,
                IntronicRate = ACC_RNA_report__Intronic_Rate, 
                IntergenicRate = ACC_RNA_report__Intergenic_Rate, 
                GenesDetected = ACC_RNA_report__Genes_Detected,
                TranscriptsDetected = ACC_RNA_report__Transcripts_Detected, 
                ExpProfEfficiency = ACC_RNA_report__Expression_Profiling_Efficiency, 
                rRNARate = ACC_RNA_report__rRNA_Rate, 
                TotalReads = ACC_RNA_report__Total_Reads, 
                PercentAligned = ACC_RNA_report__Percent_Aligned,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'ACC')

METADATA_DLPFC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN,
                DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Ribozero_Batch, DLPFC_RNA_report__Library_Batch, 
                DLPFC_RNA_report__Flowcell_Batch, DLPFC_RNA_report__Mapped_Reads, DLPFC_RNA_report__Intragenic_Rate,
                DLPFC_RNA_report__Intronic_Rate, DLPFC_RNA_report__Intergenic_Rate, DLPFC_RNA_report__Genes_Detected,
                DLPFC_RNA_report__Transcripts_Detected, DLPFC_RNA_report__Expression_Profiling_Efficiency, 
                DLPFC_RNA_report__rRNA_Rate, DLPFC_RNA_report__Total_Reads, DLPFC_RNA_report__Percent_Aligned,
                DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_, 
                RibozeroBatch = DLPFC_RNA_report__Ribozero_Batch,
                LibraryBatch = DLPFC_RNA_report__Library_Batch,
                FlowcellBatch = DLPFC_RNA_report__Flowcell_Batch, 
                MappedReads = DLPFC_RNA_report__Mapped_Reads, 
                IntragenicRate = DLPFC_RNA_report__Intragenic_Rate,
                IntronicRate = DLPFC_RNA_report__Intronic_Rate, 
                IntergenicRate = DLPFC_RNA_report__Intergenic_Rate, 
                GenesDetected = DLPFC_RNA_report__Genes_Detected,
                TranscriptsDetected = DLPFC_RNA_report__Transcripts_Detected, 
                ExpProfEfficiency = DLPFC_RNA_report__Expression_Profiling_Efficiency, 
                rRNARate = DLPFC_RNA_report__rRNA_Rate, 
                TotalReads = DLPFC_RNA_report__Total_Reads, 
                PercentAligned = DLPFC_RNA_report__Percent_Aligned,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'DLPFC')

METADATA = rbindlist(list(METADATA_ACC, METADATA_DLPFC), use.names = T, fill = T) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
  left_join(ANCESTRY) %>%
  unique

# Replace AFF and BP samples as other
levels(METADATA$Dx) = c("Other", "Other", "Control", "SCZ")
```

### Data preprocessing
```{r preprocess.data}
# Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
writeLines('Following counts are missing any metadata')
writeLines(paste(setdiff(colnames(COUNT), levels(METADATA$SampleID)), collapse = ', '))
METADATA <- METADATA %>% filter(SampleID %in% colnames(COUNT))

writeLines('Following samples are marked exclude')
writeLines(paste(METADATA$SampleID[METADATA$IsolationExclude == 1 | METADATA$ReportExclude == 1], collapse = ', '))
METADATA <- METADATA  %>% filter(IsolationExclude == 0, ReportExclude == 0)

writeLines('Following samples are missing PMI information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$PMI)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(PMI))

writeLines('Following samples are missing gender information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$Gender)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(Gender))

writeLines('Following samples are missing RIN information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$RIN)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(RIN))

writeLines('Following samples are missing EV information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$EV.1)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(METADATA$EV.1))

writeLines('Following samples are tagged outliers and removed')
writeLines(paste(c('MSSM_RNA_ACC_BP_24'), collapse = ', '))
METADATA <- METADATA  %>% filter(!(SampleID %in% c('MSSM_RNA_ACC_BP_24')))
```

```{r preprocess.data1, results='asis'}
# Match covariates to expression data
indToRetain = intersect(METADATA$SampleID, colnames(COUNT))
removedIDs = setdiff(colnames(COUNT), METADATA$SampleID)

COUNT = COUNT[,indToRetain]

rownames(METADATA) = METADATA$SampleID
METADATA = METADATA[indToRetain,]

METADATA %>% 
  group_by(Dx, Tissue) %>% 
  summarise(count = n()) %>% 
  spread(Tissue, count) %>%
  kable()

METADATA = METADATA %>%
  dplyr::mutate(Dx.Tissue = paste(Dx, Tissue, sep = '.'))
```
Following sample are removed `r paste(removedIDs, collapse = ',')`

```{r gene.param}
## Get GC content from biomart
backgroundGenes = data.frame(ensembl_gene_id = rownames(COUNT))

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "dec2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::select(ensembl_gene_id, percentage_gc_content) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$ensembl_gene_id
```

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Gender", "RibozeroBatch", "LibraryBatch", "FlowcellBatch", "Dx.Tissue")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
                    "PercentAligned", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){
  x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
rownames(COVARIATES) <- METADATA$SampleID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=8, fig.height=8}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
```

## Explore metatdata
```{r data.explore, fig.width = 12, fig.height = 12}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = Age_of_Death)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme

# Intronic Rate
p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme

# IntergenicRate
p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme

# Transcripts Detected
p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TranscriptsDetected)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Transcripts Detected') + my.theme

# Mapped Reads
p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot()
p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme

# PercentAligned
p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PercentAligned)) + geom_boxplot()
p[[8]] = p[[8]] + ggtitle('Percent Aligned') + my.theme

# rRNARate
p[[9]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate)) + geom_boxplot()
p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme

multiplot(plotlist = p, cols = 3)

# Institution
# vcd::mosaic(~ Institution + Dx.Tissue, data = COVARIATES)

# Gender
# vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
```
### Filter genes
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = dlply(METADATA, .(Dx.Tissue), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID %>% droplevels()],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC.CONT$ensembl_gene_id[!is.na(GENE.GC.CONT$percentage_gc_content)]) %>%
  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$Length)])

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

### Normalisation (with NULL)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

```{r temp, include = F}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. 
1. "Dx.Tissue" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue)
2. Adding Instituion and Gender a priori to variable selection as they are correlated with Dx.Tissue
```{r iterative.normalisation, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender');

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, 'Individual_ID'))
covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects and Individual_ID is chosen as random effect'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
  correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)

  # Re-calculate voom weights with correlation of random effects
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                              design=DM1$design, plot=F,
                              block = COVARIATES$Individual_ID, 
                              correlation = correlation$cor)

  # Fit linear model using new weights and new design
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
  ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, 'Individual_ID'))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
                                                                          c(postAdjustCovars, 'Individual_ID'))]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```

```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue','Institution')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue','Institution')]))
```

```{r temp1, include=F}
dev.off()
gc()
```

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=16, fig.width=16}
# Post adjusted design matrix
DESIGN = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
correlation <- parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Individual_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$Individual_ID, 
                    correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
FIT = lmFit(VOOM.WEIGHTS)

# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueControl_ACC-Dx.TissueOther_ACC",
                                     "Dx.TissueControl_ACC-Dx.TissueSCZ_ACC",
                                     "Dx.TissueOther_ACC-Dx.TissueSCZ_ACC",
                                     "Dx.TissueControl_DLPFC-Dx.TissueOther_DLPFC",
                                     "Dx.TissueControl_DLPFC-Dx.TissueSCZ_DLPFC",
                                     "Dx.TissueOther_DLPFC-Dx.TissueSCZ_DLPFC"),
                         levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:6, function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR) 
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
  tidyr::separate(ref.state, into = c('Dx1','Tissue'), sep = '_') %>%
  tidyr::separate(to.state, into = c('Dx2','Tissue'), sep = '_') %>%
  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_') %>%
  left_join(Ensemble2HGNC) %>%
  left_join(GENE.LEN)
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
  group_by(Comparison, Tissue) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
  group_by(Comparison, Tissue) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))

kable(full_join(tmp,tmp1))

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red'))
p = p + facet_grid(Tissue+.~Comparison, scales = 'fixed')
p
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=10, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE, aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE, aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE, aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

multiplot(plotlist = pl, cols = 3)
```

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16}
diffexp.genes = DE %>%
  dlply(.(Comparison, Tissue), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
backgroundGenes = unique(DE$hgnc_symbol)

# Download all related genesets from synapse
geneset.files = synQuery('select * from file where parentId == "syn3240583"')
ALL_USED_IDs = c(ALL_USED_IDs, geneset.files$file.id)
genesets.to.test = sapply(geneset.files$file.name, function(dataSource, geneset.files){
  tmp = downloadFile(geneset.files$file.id[geneset.files$file.name == dataSource]) %>%
    dlply(.(Name), .fun = function(x){
      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
    })
  }, geneset.files) %>%
  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
  do.call(c,.)

# Perform enrichment analysis
enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
    }, x, backgroundGenes, .parallel = TRUE) %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
}, genesets.to.test, backgroundGenes) %>%
  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
  tidyr::separate(Comparison, c('Name', 'Region'), sep = '\\.') %>%
  dplyr::rename(Comparison = Name, Category = .id)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Set annotations
all.annotations = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay	 = 'RNAseq',

  tissueTypeAbrv	= 'DLPFC, ACC', 
  study = 'CMC',

  organism = 'HomoSapiens',
  consortium	= 'CMC',

  normalizationStatus	= TRUE,

  normalizationType	= 'CQN'
)

# Code
CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'CMC_ACC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('CMC_ACC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
annotations(COV_OBJ) = annotations(CODE)
annotations(COV_OBJ)$dataSubType = 'covariates'
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store logCPM
CQN.GENE_EXPRESSION$y %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store cqn offsets
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('CMC_ACC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
annotations(OFFSET_OBJ) = annotations(CODE)
annotations(OFFSET_OBJ)$dataSubType = 'offset'
OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store design matrix
DESIGN$design %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'CMC_ACC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('CMC_ACC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
annotations(DM_OBJ) = annotations(CODE)
annotations(DM_OBJ)$dataSubType = 'designMatrix'
DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store residual gene expression for network analysis
RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('CMC_ACC_DLPFC_netResidualExpression.tsv', 
                name = 'Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
annotations(nEXP_OBJ) = annotations(CODE)
annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store differential expression results
write.table(DE, file = 'CMC_ACC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_ACC_DLPFC_DiffExpression.tsv', 
                name = 'Differential Expression Results (Dx.Tissue)', 
                parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
stopCluster(cl)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

### 2017-03-18 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor bug fix

@@
 ```{r residualSigCovars.manual, fig.width=12, fig.height=8}
+# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
+residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+                                              COVARIATES, 
+                                              'adjusted design(voom-normalized)',
+                                              isKeyPlot=TRUE)
+
 residualSigCovars[["PC_res"]][[2]]$plotData
@@
 p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
-p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
+p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
 # p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=16, fig.width=16}
+```{r diffExp, fig.height=10, fig.width=15}
 # Post adjusted design matrix
@@
 DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
+# Store enrichment results
+write.table(enrichment.results, file = 'CMC_ACC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
+ENRICH_OBJ = File('CMC_ACC_DLPFC_EnrichResults.tsv', 
+                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
+                  parentId = CODE$properties$id)
+annotations(ENRICH_OBJ) = annotations(CODE)
+annotations(ENRICH_OBJ)$dataSubType = 'enrichResults'
+ENRICH_OBJ = synStore(ENRICH_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 stopCluster(cl)
@@
 |  Design Matrix                              |  `r DM_OBJ$properties$id`        |
-|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
-|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
 |  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
 |  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
+|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |

### 2017-03-19 code/Rmd/MAYO_REPROCESSED.Rmd
Minor bug fix

@@
 
-library(parallel)
+library(doParallel)
 library(foreach)
@@
 cl = makeCluster(detectCores()-2)
-registerDoParallel(cl)
+doParallel::registerDoParallel(cl)
 

### 2017-03-19 code/Rmd/MSSM_REPROCESSED.Rmd
Added markdown to process reprocessed RNASeq from MSSM cohorts with CQN and SVA normalisation

---
title: "Covariate analysis of MSSM reprocessed counts (with CQN normalisation and covariates adjusted)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./MSSM_REPROCESSED.Rmd",
                                 parentId = "syn7554308",
                                 entityName = 'MSSM Reprocessed RNASeq (Normalised)')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(reshape2)
library(limma)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)
library(cqn)

library(synapseClient)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

library(doParallel)
library(foreach)

cl = makeCluster(detectCores()-2)
doParallel::registerDoParallel(cl)

synapseLogin()

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn7554308'
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of MSSM reprocessed RNASeq data';

thisFileName <- 'MSSM_REPROCESSED.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Get reprocessed sample counts for temporal cortex
COUNT_ID <- 'syn7994853';
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F)

# Convert rownames of counts from tracking id to ensemble gene id
tmp = data.frame(Gene.ID = rownames(COUNT)) %>%
  dplyr::mutate(ID = Gene.ID) %>%
  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')
rownames(tmp) = tmp$Gene.ID
rownames(COUNT) = tmp[rownames(COUNT), 'ensembl_gene_id']

# Get sample ids
SampleID = data.frame(SampleID = colnames(COUNT),
                 ID = colnames(COUNT)) %>% 
  tidyr::separate(ID, c('A','B','ID'), sep = '_') %>%
  dplyr::select(SampleID, ID) %>%
  dplyr::mutate(ID = as.numeric(ID))

# Get technical metadata
METADATA_TECH_ID <- 'syn6100548'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECH_ID
METADATA_TECH <- read.table(synGet(METADATA_TECH_ID)@filePath,sep=',',header=T, row.names=1) %>%
  group_by(sampleIdentifier) %>%
  top_n(1, -rRNA.rate) %>%
  dplyr::select(sampleIdentifier, BrodmannArea, barcode, individualIdentifier, batch, RIN) %>%
  unique

# Get clinical metadata
METADATA_CLINICAL_ID <- 'syn6101474'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CLINICAL_ID
METADATA_CLINICAL <- read.csv(synGet(METADATA_CLINICAL_ID)@filePath, header = T)

# Get picard metrics from synapse
METADATA.PICARD_ID <- 'syn8380379'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)

METADATA.PICARD$sampleIdentifier = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
  gsub('.accepted_hits.sort.coord/picard.CombinedMetrics.csv','',.)
colnames(METADATA.PICARD) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD))
colnames(METADATA.PICARD) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD))

METADATA <- full_join(METADATA_TECH, METADATA_CLINICAL) %>%
  inner_join(METADATA.PICARD)

# Get gene specific parameters from synapse
GENE.PARAM = downloadFile('syn8449369')
ALL_USED_IDs = c(ALL_USED_IDs, 'syn8449369')

GENE.LEN = dplyr::select(GENE.PARAM, ensembl_gene_id, gene.length) %>% unique()
rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id

GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% unique()
rownames(GENE.GC) = GENE.GC$ensembl_gene_id
```

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA = METADATA %>%
  dplyr::mutate(AOD = gsub('\\+','',AOD),
                BrainRegion = factor(BrodmannArea,
                                     levels = c('BM10', 'BM22', 'BM36', 'BM44'),
                                     labels = c('BM10' = 'FP', 'BM22' = 'STG', 
                                                'BM36' = 'PHG', 'BM44' = 'IFG')),
                Diagnosis = 'NONE', 
                RIN2 = RIN^2)
METADATA$Diagnosis[METADATA$CDR <= 1] = 'ND'
METADATA$Diagnosis[METADATA$CDR == 2 | METADATA$CDR == 3] = 'MD'
METADATA$Diagnosis[METADATA$CDR >= 4] = 'SD'
levels(METADATA$batch)[levels(METADATA$batch) == ''] = 'NoBatch'

METADATA = METADATA %>%
  tidyr::unite(BrainRegion.Diagnosis, BrainRegion, Diagnosis, sep = '.')

METADATA <- METADATA %>%
  filter(!is.na(BrainRegion.Diagnosis), !is.na(PMI), !is.na(RIN))
```
Following samples were considered outliers and removed based on PCA and/or eucledian distance: 11311\_CER, 1923\_CER, 1923\_TCX, 11396\_TCX, 11294\_TCX, 11408\_TCX
```{r preprocess.data1}
# Match covariates to expression data
indToRetain = intersect(METADATA$sampleIdentifier, colnames(COUNT))
indRemoved = setdiff(colnames(COUNT), METADATA$sampleIdentifier)

COUNT <- COUNT[, indToRetain]

METADATA = as.data.frame(METADATA)
rownames(METADATA) = METADATA$sampleIdentifier
METADATA = METADATA[indToRetain,]
```
`r dim(COUNT)[2]` samples from `r length(unique(METADATA$individualIdentifier))` subjects were obtained from the MSSM cohorts in AMP-AD reprocessed rnaseq project.

Following `r length(indRemoved)` samples were removed: `r indRemoved`

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering, cache=TRUE}
primaryVariable <- "BrainRegion.Diagnosis"
FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AOD","PMI", "PCT_PF_READS", "PCT_PF_READS_ALIGNED", 
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", 
                    "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$sampleIdentifier

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=6, fig.height=6}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
```

```{r data.explore, fig.width = 12, fig.height = 8}
my.theme = theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# Age Of Death
p[[2]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(AOD))) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AOD') + my.theme

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PMI))) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI') + my.theme

# Ribosomal bases
p[[4]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_RIBOSOMAL_BASES))) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Fraction of Ribosomal Bases') + my.theme

# Intronic bases
p[[5]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTRONIC_BASES))) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Fraction Intronic Bases') + my.theme

# Intergenic bases
p[[6]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTERGENIC_BASES))) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Fraction Intergenic Bases') + my.theme

multiplot(plotlist = p, cols = 3)
```
### CPM Normalisation
Preprocess counts matrix using CQN (Conditional Quantile Normlaisation accounting for both GC content and gene length)

Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion x Diagnosis. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('SampleID') %>%
  dlply(.(BrainRegion.Diagnosis), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                     MIN_GENE_CPM=1, 
                                                     MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC$ensembl_gene_id[!is.na(GENE.GC$percentage_gc_content)]) %>%
  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$gene.length)]) %>%
  setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous"))

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], 
                                                 MIN_GENE_CPM=0, 
                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

### Normalisation (with NULL)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=5, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('Region','Diagnosis'), sep = '\\.') %>% 
  dplyr::mutate(Region = factor(Region), batch = factor(batch))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Diagnosis, size=RIN, shape = SEX))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(.~Region, scales = 'free_y')
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data1.2, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = (COVARIATES[,c(FactorCovariates), drop = F])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

```{r temp, include = F}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
```{r preadj.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preadj.covariates.plot, fig.width=12, fig.height=8}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative design)
Since many covariates are correlated, re-normalising and re-adjusting COUNTS with an iterative design matrix.

NOTE:
1. Using a mixed effect model where random effect is chosen as individualIdentifier
2. Adding batch and SEX a priori to variable selection
3. Primary variable of interest BrainRegion.Diagnosis is excluded from the pool of available covariates for selection
```{r iterative.norm, results='asis'}
# Primary variable of interest
postAdjustCovars = c('batch', 'SEX');

# Assign residual covariates
residualCovars = setdiff(preAdjustedSigCovars$significantCovars, 
                         c(postAdjustCovars, primaryVariable, 'individualIdentifier'))
residualSigCovars = preAdjustedSigCovars
covariatesEffects = preAdjustedSigCovars$Effects.significantCovars[residualCovars]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 20){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects and individualIdentifier as random effect'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights for dispersion control
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)

  # Calculate correlation values of random effects
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
                                             block = COVARIATES$individualIdentifier)

  # Re-estimate voom weights
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                              design=DM1$design, plot=F, 
                              block = COVARIATES$individualIdentifier, 
                              correlation = correlation$cor)

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
                            design = DM1$design,
                            weights = VOOM.GENE_EXPRESSION$weights,
                            block = COVARIATES$individualIdentifier, 
                            correlation = correlation$cor)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates),
                         c(postAdjustCovars, primaryVariable, 'individualIdentifier'))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  residCovars = setdiff(residualSigCovars$significantCovars, 
                        c(postAdjustCovars, primaryVariable, 'individualIdentifier'))
  covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```

### Sanity check
```{r residual.adj, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

residualSigCovars[["PC_res"]][[2]]$plotData
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2.1, fig.height=10, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('Region','Diagnosis'), sep = '\\.') %>% 
  dplyr::mutate(Region = factor(Region))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Diagnosis, size=RIN, shape = SEX))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(.~Region, scales = 'free_y')
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```

```{r decompse.normalise.data2.2, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

```{r temp1, include=F}
dev.off()
gc()
```

### Differential expression analysis (with BrainRegion.Diagnosis as primary variable)
Differential expression is performed on the primary variable by controlling for covariates identified above

Interpretation of BrainRegion.Diagnosis
1. FP.MD: Mild dementia samples from frontal pole with CDR equal to 2 or 3
2. FP.ND: No dementia samples from frontal pole with CDR less than 2
3. FP.SD: Severe dementia samples from frontal pole with CDR grater than 3
4. IFG.MD: Mild dementia samples from inferior frontal gyrus with CDR equal to 2 or 3
5. IFG.ND: No dementia samples from inferior frontal gyrus with CDR less than 2
6. IFG.SD: Severe dementia samples from inferior frontal gyrus with CDR grater than 3
7. PHG.MD: Mild dementia samples from para hippocampal gyrus with CDR equal to 2 or 3
8. PHG.ND: No dementia samples from para hippocampal gyrus with CDR less than 2
9. PHG.SD: Severe dementia samples from para hippocampal gyrus with CDR grater than 3
10. STG.MD: Mild dementia samples from superior temporal gyrus with CDR equal to 2 or 3
11. STG.ND: No dementia samples from superior temporal gyrus with CDR less than 2
12. STG.SD: Severe dementia samples from superior temporal gyrus with CDR grater than 3

Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp, fig.height=15, fig.width=13}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable, postAdjustCovars), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)

# Calculate correlation values of random effects
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)

# Re-estimate voom weights
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$individualIdentifier, 
                    correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
FIT = lmFit(VOOM.WEIGHTS)

# Fit contrast
cntr = c()
for (region in c('FP', 'STG', 'PHG', 'IFG')){
  tmp = combn(paste0('BrainRegion.Diagnosis', region, '.', c('ND','MD','SD')),2)
  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
}

contrast = makeContrasts(contrasts=cntr,
                         levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:12, function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('ensembl_gene_id')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
                Study = 'MSSM',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('Region','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('Region','against'), sep = '\\.') %>%
  tidyr::unite(Comparison, reference, against, sep = '-') %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction, Region) %>%
  group_by(Region, Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05)

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction, Region) %>%
  group_by(Region, Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2)

kable(full_join(tmp %>%
                  dplyr::rename(DOWN_FDR_0_05 = DOWN, UP_FDR_0_05 = UP),
                tmp1 %>%
                  dplyr::rename(DOWN_FDR_0_05_FC_1.2 = DOWN, UP_FDR_0_05_FC_1.2 = UP)))

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red'))
p = p + facet_grid(Region+.~Comparison, scales = 'fixed')
p
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=15, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE %>% dplyr::filter(Comparison == 'ND-SD'), 
                 aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(Region~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE %>% dplyr::filter(Comparison == 'ND-SD'), 
                 aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(Region~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE %>% dplyr::filter(Comparison == 'ND-SD'), 
                 aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(Region~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

multiplot(plotlist = pl, cols = 3)
```

### SVA Adjustments for eQTL analysis
Conditioned on primary variable (BrainRegion.Diagnosis) and identified covariates estimate surrogate variables using SVA package
```{r sva.adjust}
# Get (null) design matrix
MODEL0 = DESIGN$design[,-grep(primaryVariable,colnames(DESIGN$design))] # Get null model by removing variable of interest
MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
MODEL1 = DESIGN$design
MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]

# Calculate the number of surrogate variables
NUM.SV = sva::num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                     MODEL1, method = 'be', seed = 123456, B = 30)

# Estimate surrogate variables
SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
                    MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv

colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
rownames(SURR.VAR) = rownames(MODEL1)

# Estimate voom weights for dispersion control
VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design=cbind(MODEL1, SURR.VAR), plot=F)

# Calculate correlation values of random effects
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$individualIdentifier)

# Re-estimate voom weights
VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design = cbind(MODEL1, SURR.VAR), plot=F,
                        block = COVARIATES$individualIdentifier, 
                        correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)

# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
```

### Covariates Adjustments for network analysis
Identified covariates are regressed out from the expression matrix for network analysis
```{r network.adjust}
# Get design matrix
DESIGN.NET = getDesignMatrix(COVARIATES[, postAdjustCovars, drop = F], Intercept = F)
DESIGN.NET = DESIGN.NET$design[,linColumnFinder(DESIGN.NET$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design=DESIGN.NET, plot=F)

# Calculate correlation values of random effects
VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$individualIdentifier)

# Re-estimate voom weights
VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design = DESIGN.NET, plot=F,
                        block = COVARIATES$individualIdentifier, 
                        correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)

# Residuals after normalisation
RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Set annotations
all.annotations = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay	 = 'Reprocessed RNAseq',

  tissueTypeAbrv	= 'FP, STG, PHG, IFG', 
  study = 'MSSMRNASeq', 
  center = 'MSSM',

  organism = 'HomoSapiens',
  consortium	= 'AMP-AD',

  normalizationStatus	= TRUE,
  modelSystem	= FALSE,

  normalizationType	= 'CQN'
)

# Code
CODE <- Folder(name = "MSSM Reprocessed RNASeq (Normalised)", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'MSSM_FP_STG_PHG_IFG_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('MSSM_FP_STG_PHG_IFG_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
annotations(COV_OBJ) = annotations(CODE)
annotations(COV_OBJ)$dataSubType = 'covariates'
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store logCPM
CQN.GENE_EXPRESSION$y %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store cqn offsets
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('MSSM_FP_STG_PHG_IFG_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
annotations(OFFSET_OBJ) = annotations(CODE)
annotations(OFFSET_OBJ)$dataSubType = 'offset'
OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store design matrix
MODEL1 %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('MSSM_FP_STG_PHG_IFG_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
annotations(DM_OBJ) = annotations(CODE)
annotations(DM_OBJ)$dataSubType = 'designMatrix'
DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store surrogate variables
SURR.VAR %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_SurVar.tsv', sep = '\t', row.names=F, quote=F)
SURR_OBJ = File('MSSM_FP_STG_PHG_IFG_SurVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
annotations(SURR_OBJ) = annotations(CODE)
annotations(SURR_OBJ)$dataSubType = 'surrogateVariables'
SURR_OBJ = synStore(SURR_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store residual gene expression for eQTL analysis
RESIDUAL.SVA.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_eQTLResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
eEXP_OBJ = File('MSSM_FP_STG_PHG_IFG_eQTLResidualExpression.tsv', 
                name = 'Normalised, covariates and surrogate variable adjusted residual expression (for eQTL)',
                parentId = CODE$properties$id)
annotations(eEXP_OBJ) = annotations(CODE)
annotations(eEXP_OBJ)$dataSubType = 'residualGeneExpForeQTL'
eEXP_OBJ = synStore(eEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store residual gene expression for network analysis
RESIDUAL.NET.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('MSSM_FP_STG_PHG_IFG_netResidualExpression.tsv', 
                name = 'Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
annotations(nEXP_OBJ) = annotations(CODE)
annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store differential expression results
write.table(DE, file = 'MSSM_FP_STG_PHG_IFG_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('MSSM_FP_STG_PHG_IFG_DiffExpression.tsv', 
                name = 'Differential Expression Results (BrainRegion.Diagnosis)', 
                parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
stopCluster(cl)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

### 2017-03-19 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Added markdown to process CMC ACC and DLPFC reprocessed RNAseq with CQN and SVA normalisation

--- 2017-03-18 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)"
+title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN and SVA normlaisation)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
+knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd",
                                  parentId ="syn7501880",
-                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
+                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - SVA')
 ```
@@
 activityName = 'Covariate adjustments';
-activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (ACC and DLPFC)';
+activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN and SVA normalisation (ACC and DLPFC)';
 
-thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd'
+thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd'
 
@@
 
-### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. 
+### Normalisation (custom design)
+Since many covariates are correlated, re-normalising COUNTS with covariates identified here in syn8468116. 
 1. "Dx.Tissue" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue)
-2. Adding Instituion and Gender a priori to variable selection as they are correlated with Dx.Tissue
+2.
+Institution, Gender, IntronicRate, RibozeroBatch, PMI, RIN, TranscriptsDetected, Age\\_of\\_Death, TotalReads, IntergenicRate, PercentAligned, LibraryBatch were chosen as fixed effects and Individual_ID is chosen as random effect
 ```{r iterative.normalisation, results='asis'}
 # Primary variable of interest
-postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender');
-
-# Assign residual covariates
-residualSigCovars = preAdjustedSigCovars
-residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, 'Individual_ID'))
-covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
-
-loopCount = 0 
-while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
-  writeLines(paste('Using following covariates in the model:',
-                   paste(postAdjustCovars, collapse=', '),
-                   'as fixed effects and Individual_ID is chosen as random effect'))
+postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'IntronicRate', 'RibozeroBatch', 'PMI', 'RIN', 'TranscriptsDetected',
+                     'Age_of_Death', 'TotalReads', 'IntergenicRate', 'PercentAligned', 'LibraryBatch');
 
@@
 
-  # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, 'Individual_ID'))
-
-  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                                COVARIATES[, residCovars, drop=F], 
-                                                'adjusted design(voom-normalized)',
-                                                isKeyPlot=TRUE)
-
-  # Add postadjusted covariates (if any)
-  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
-                                                                          c(postAdjustCovars, 'Individual_ID'))]
-
-  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
-  loopCount = loopCount + 1
-}
-
-modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
-                  'as fixed effects')
-
-tmp <- paste('Using following covariates in the final model:', modelStr)
-```
-`r tmp`
-
-### Sanity check
-```{r residualSigCovars.manual, fig.width=12, fig.height=8}
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
@@
 ```
+### SVA
+```{r sva}
+# Get (null) design matrix
+MODEL0 = DM1$design[,-grep('Dx.Tissue',colnames(DM1$design))] # Get null model by removing variable of interest
+MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
+MODEL1 = DM1$design
+MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
+
+# Source modified functions from sva package
+source('../R/num.sv.R')
+source('../R/irwsva.build.R')
+source('../R/f.pvalue.R')
+
+# Calculate the number of surrogate variables
+NUM.SV = num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
+                MODEL1, method = 'be', seed = 123456, B = 30, tol = 1e-22)
+
+# Estimate surrogate variables
+SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
+                        MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-22)$sv
+
+colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
+rownames(SURR.VAR) = rownames(MODEL1)
+
+# Estimate voom weights for dispersion control
+VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+                        design=cbind(MODEL1, SURR.VAR), plot=F)
+
+# Calculate correlation values of random effects
+VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
+correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$Individual_ID)
+
+# Re-estimate voom weights
+VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+                        design = cbind(MODEL1, SURR.VAR), plot=F,
+                        block = COVARIATES$individualIdentifier, 
+                        correlation = correlation$cor)
+
+# Fit linear model using new weights and new design
+VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
+VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)
+
+# Residuals after normalisation
+RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
+                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
+```
+
 ### Residual calculation
@@
 varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
-  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
+  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
 ```
@@
 ```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.GENE_EXPRESSION))
+cr = cor(t(RESIDUAL.SVA.GENE_EXPRESSION))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
+PC <- prcomp(RESIDUAL.SVA.GENE_EXPRESSION, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
+tree = hclust(as.dist(t(RESIDUAL.SVA.GENE_EXPRESSION)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 ```{r diffExp, fig.height=10, fig.width=15}
-# Post adjusted design matrix
-DESIGN = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
-
-# Estimate voom weights
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-correlation <- parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Individual_ID)
-
-# Re-calculate voom weights with correlation of random effects
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                    design=DESIGN$design, plot=F,
-                    block = COVARIATES$Individual_ID, 
-                    correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-FIT = lmFit(VOOM.WEIGHTS)
-
 # Fit contrast
@@
                                      "Dx.TissueOther_DLPFC-Dx.TissueSCZ_DLPFC"),
-                         levels = colnames(FIT$coefficients))
-FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+                         levels = colnames(VOOM.SVA.FIT$coefficients))
+FIT.CONTR = contrasts.fit(VOOM.SVA.FIT, contrasts=contrast)
 FIT.CONTR = eBayes(FIT.CONTR)
@@
 # Code
-CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
+CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - SVA", parentId = parentId)
 annotations(CODE) = all.annotations
@@
 
-# Store residual gene expression for network analysis
-RESIDUAL.GENE_EXPRESSION %>%
+# Store surrogate variables
+SURR.VAR %>%
+  rownameToFirstColumn('SampleID') %>%
+  write.table(file = 'CMC_ACC_DLPFC_SurVar.tsv', sep = '\t', row.names=F, quote=F)
+SURR_OBJ = File('CMC_ACC_DLPFC_SurVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
+annotations(SURR_OBJ) = annotations(CODE)
+annotations(SURR_OBJ)$dataSubType = 'surrogateVariables'
+SURR_OBJ = synStore(SURR_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
+# Store residual gene expression for eQTL analysis
+RESIDUAL.SVA.GENE_EXPRESSION %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_ACC_DLPFC_netResidualExpression.tsv', 
-                name = 'Normalised, covariates removed residual expression (for network analysis)', 
+  write.table(file = 'CMC_ACC_DLPFC_eQTLResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+eEXP_OBJ = File('CMC_ACC_DLPFC_eQTLResidualExpression.tsv', 
+                name = 'Normalised, covariates and surrogate variable adjusted residual expression (for eQTL)',
                 parentId = CODE$properties$id)
-annotations(nEXP_OBJ) = annotations(CODE)
-annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
-nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+annotations(eEXP_OBJ) = annotations(CODE)
+annotations(eEXP_OBJ)$dataSubType = 'residualGeneExpForeQTL'
+eEXP_OBJ = synStore(eEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 
@@
 |  Design Matrix                              |  `r DM_OBJ$properties$id`        |
-|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
+|  Residual Expression (for eQTL analysis)    |  `r eEXP_OBJ$properties$id`      |
 |  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

### 2017-03-19 code/R/num.sv.R

@@ -1,0 +1,83 @@
+num.sv <- function (dat, mod, method = c("be", "leek"), vfilter = NULL, 
+                    B = 20, seed = NULL, LINPACK = F, tol = 1e-18) 
+{
+  if (!is.null(vfilter)) {
+    if (vfilter < 100 | vfilter > dim(dat)[1]) {
+      stop(paste("The number of genes used in the analysis must be between 100 and", 
+                 dim(dat)[1], "\n"))
+    }
+    tmpv = rowVars(dat)
+    ind = which(rank(-tmpv) < vfilter)
+    dat = dat[ind, ]
+  }
+  method <- match.arg(method)
+  if (method == "be") {
+    if (!is.null(seed)) {
+      set.seed(seed)
+    }
+    warn <- NULL
+    n <- ncol(dat)
+    m <- nrow(dat)
+    H <- mod %*% solve(t(mod) %*% mod, tol = tol) %*% t(mod)
+    res <- dat - t(H %*% t(dat))
+    uu <- svd(res, LINPACK = LINPACK)
+    ndf <- min(m, n) - ceiling(sum(diag(H)))
+    dstat <- uu$d[1:ndf]^2/sum(uu$d[1:ndf]^2)
+    dstat0 <- foreach (i = 1:B,
+                       .combine = rbind,
+                       .export = c('res', 'H', 'LINPACK', 'ndf')) %dopar% {
+                         res0 <- t(apply(res, 1, sample, replace = FALSE))
+                         res0 <- res0 - t(H %*% t(res0))
+                         tryCatch({
+                           uu0 <- svd(res0, LINPACK = LINPACK)
+                           dstat0 <- uu0$d[1:ndf]^2/sum(uu0$d[1:ndf]^2)
+                         }, error = function(x) {
+                           dstat0 = matrix(0,1,ndf)
+                         }) 
+                       }
+    psv <- rep(1, n)
+    for (i in 1:ndf) {
+      psv[i] <- mean(dstat0[, i] >= dstat[i])
+    }
+    for (i in 2:ndf) {
+      psv[i] <- max(psv[(i - 1)], psv[i])
+    }
+    nsv <- sum(psv <= 0.1)
+    return(as.numeric(list(n.sv = nsv)))
+  }
+  else {
+    dat <- as.matrix(dat)
+    dims <- dim(dat)
+    a <- seq(0, 2, length = 100)
+    n <- floor(dims[1]/10)
[…]

### 2017-03-20 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Minor bug fix

@@
 # Store design matrix
-DESIGN$design %>%
+MODEL1 %>%
   rownameToFirstColumn('SampleID') %>%

### 2017-03-20 code/Rmd/ROSMAP_REPROCESSED_TMM.Rmd
Added markdown to process ROSMAP reprocessed RNASeq with TMM and SVA normalaisation

--- 2017-03-17 code/Rmd/ROSMAP_REPROCESSED.Rmd
@@
 ---
-title: "Covariate analysis of ROSMAP reprocessed counts (with CQN normalisation and covariates adjusted)"
+title: "Covariate analysis of ROSMAP reprocessed counts (with TMM normalisation and covariates adjusted)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./ROSMAP_REPROCESSED.Rmd",
+knit2synapse::knitToFolderEntity(file = "./ROSMAP_REPROCESSED_TMM.Rmd",
                                  parentId = "syn8403743",
-                                 entityName = 'ROSMAP Reprocessed RNASeq (Normalised)')
+                                 entityName = 'ROSMAP Reprocessed RNASeq (Normalised - With out gene length and gc content adjustment)')
 ```
@@
 library(RColorBrewer)
-library(cqn)
 
@@
 
-thisFileName <- 'ROSMAP_REPROCESSED.Rmd'
+thisFileName <- 'ROSMAP_REPROCESSED_TMM.Rmd'
 
@@
 ### CPM Normalisation
-Preprocess counts matrix using CQN (Conditional Quantile Normlaisation accounting for both GC content and gene length)
+Preprocess counts matrix using TMM normalisation
 
@@
 ```{r initial.voom.normalisation, fig.height=5, fig.width=5}
-# Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
-                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
+# Normalise using TMM
+TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')
+VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION)
 
 # Set gene counts in specific samples that are deviating 3 sd from other samples to 0
-log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
+log.mat = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
   mn = mean(x, na.rm = T)
@@
 
-# Re-compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
-                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
+# Re-normalise using TMM
+TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')
+VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION)
 ```
@@
 ```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
+cr = cor(t(VOOM.GENE_EXPRESSION$E))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)
+PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset)))
+tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 # Find correlation between PC's of gene expression with covariates
-preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
+preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
                                                  COVARIATES,
@@
   # Estimate voom weights for dispersion control
-  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)
+  VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
 
   # Fit linear model using new weights and new design
-  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
-                            design = DM1$design,
-                            weights = VOOM.GENE_EXPRESSION$weights)
+  VOOM.ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
 
@@
   RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
-                                                CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
+                                                VOOM.GENE_EXPRESSION$E)
 
@@
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(TMM.GENE_EXPRESSION, design=DESIGN$design, plot=F)
 
 # Fit linear model using new weights and new design
-FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
-            design = DESIGN$design,
-            weights = VOOM.WEIGHTS$weights)
+FIT = lmFit(VOOM.WEIGHTS)
 
@@
 # Calculate the number of surrogate variables
-NUM.SV = sva::num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
-                     MODEL1, method = 'be', seed = 123456, B = 30)
+NUM.SV = sva::num.sv(VOOM.WEIGHTS$E, MODEL1, method = 'be', seed = 123456, B = 30)
 
 # Estimate surrogate variables
-SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
-                    MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv
+SURR.VAR = sva::sva(VOOM.WEIGHTS$E, MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv
 
@@
 # Re-estimate voom weights
-VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+VOOM.SVA.WEIGHTS = voom(TMM.GENE_EXPRESSION, 
                         design = cbind(MODEL1, SURR.VAR), plot=F)
@@
 # Fit linear model using new weights and new design
-VOOM.SVA.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
-                     design = cbind(MODEL1, SURR.VAR),
-                     weights = VOOM.SVA.WEIGHTS$weights)
+VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)
 
@@
 RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
-                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
+                                                  VOOM.SVA.WEIGHTS$E)
 ```
@@
 # Estimate voom weights for dispersion control
-VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                        design=DESIGN.NET, plot=F)
+VOOM.NET.WEIGHTS = voom(TMM.GENE_EXPRESSION, design=DESIGN.NET, plot=F)
 
 # Fit linear model using new weights and new design
-VOOM.NET.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
-                     design = DESIGN.NET,
-                     weights = VOOM.NET.WEIGHTS$weights)
+VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)
 
@@
 RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
-                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
+                                                  VOOM.NET.WEIGHTS$E)
 ```
@@
 
-  normalizationType	= 'CQN'
+  normalizationType	= 'TMM'
 )
@@
 # Code
-CODE <- Folder(name = "ROSMAP Reprocessed RNASeq (Normalised)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP Reprocessed RNASeq (Normalised - With out gene length and gc content adjustment)", parentId = parentId)
 annotations(CODE) = all.annotations
@@
 # Store logCPM
-CQN.GENE_EXPRESSION$y %>%
+VOOM.NET.WEIGHTS$E %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
@@
 LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store cqn offsets
-CQN.GENE_EXPRESSION$offset %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'ROSMAP_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('ROSMAP_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
-annotations(OFFSET_OBJ) = annotations(CODE)
-annotations(OFFSET_OBJ)$dataSubType = 'offset'
-OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 

### 2017-03-20 code/Rmd/MAYO_REPROCESSED_TMM.Rmd
Added markdown to process MAYO reprocessed RNASeq with TMM and SVA normalaisation

--- 2017-03-19 code/Rmd/MAYO_REPROCESSED.Rmd
@@
 ---
-title: "Covariate analysis of MAYO reprocessed counts (with CQN normalisation and covariates adjusted)"
+title: "Covariate analysis of MAYO reprocessed counts (with TMM normalisation and covariates adjusted)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED_TMM.Rmd",
                                  parentId = "syn7554308",
-                                 entityName = 'MAYO Reprocessed RNASeq (Normalised)')
+                                 entityName = 'MAYO Reprocessed RNASeq (Normalised - With out gene length and gc content adjustment)')
 ```
@@
 
-thisFileName <- 'MAYO_REPROCESSED.Rmd'
+thisFileName <- 'MAYO_REPROCESSED_TMM.Rmd'
 
@@
   # Estimate voom weights for dispersion control
-  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)
+  VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
 
   # Calculate correlation values of random effects
-  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
   correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
@@
   # Re-estimate voom weights
-  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+  VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, 
                               design=DM1$design, plot=F, 
@@
   # Fit linear model using new weights and new design
-  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
-                            design = DM1$design,
-                            weights = VOOM.GENE_EXPRESSION$weights,
-                            block = COVARIATES$Donor_ID, 
-                            correlation = correlation$cor)
+  VOOM.ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
 
   # Residuals after normalisation
-  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
-                                                CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
+  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.GENE_EXPRESSION$E)
 
@@
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(TMM.GENE_EXPRESSION, design=DESIGN$design, plot=F)
 
 # Calculate correlation values of random effects
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
 correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Donor_ID)
@@
 # Re-estimate voom weights
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
+VOOM.WEIGHTS = voom(TMM.GENE_EXPRESSION,
                     design=DESIGN$design, plot=F,
@@
 # Fit linear model using new weights and new design
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
 FIT = lmFit(VOOM.WEIGHTS)
@@
 # Calculate the number of surrogate variables
-NUM.SV = num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
-                     MODEL1, method = 'be', seed = 123456, B = 30, tol = 1e-20)
+NUM.SV = num.sv(VOOM.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456, B = 30, tol = 1e-20)
 
 # Estimate surrogate variables
-SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
-                        MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
+SURR.VAR = irwsva.build(VOOM.GENE_EXPRESSION$E, MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
 
@@
 # Estimate voom weights for dispersion control
-VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                        design=cbind(MODEL1, SURR.VAR), plot=F)
+VOOM.SVA.WEIGHTS = voom(TMM.GENE_EXPRESSION, design=cbind(MODEL1, SURR.VAR), plot=F)
 
 # Calculate correlation values of random effects
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
 correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$Donor_ID)
@@
 # Re-estimate voom weights
-VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+VOOM.SVA.WEIGHTS = voom(TMM.GENE_EXPRESSION, 
                         design = cbind(MODEL1, SURR.VAR), plot=F,
@@
 # Fit linear model using new weights and new design
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
 VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)
@@
 # Estimate voom weights for dispersion control
-VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                        design=DESIGN.NET, plot=F)
+VOOM.NET.WEIGHTS = voom(TMM.GENE_EXPRESSION, design=DESIGN.NET, plot=F)
 
 # Calculate correlation values of random effects
-VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
 correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$Donor_ID)
@@
 # Re-estimate voom weights
-VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+VOOM.NET.WEIGHTS = voom(TMM.GENE_EXPRESSION, 
                         design = DESIGN.NET, plot=F,
@@
 # Fit linear model using new weights and new design
-VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
 VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)
@@
 # Code
-CODE <- Folder(name = "MAYO Reprocessed RNASeq (Normalised)", parentId = parentId)
+CODE <- Folder(name = "MAYO Reprocessed RNASeq (Normalised - With out gene length and gc content adjustment)", parentId = parentId)
 annotations(CODE) = all.annotations
@@
 # Store logCPM
-CQN.GENE_EXPRESSION$y %>%
+VOOM.GENE_EXPRESSION$E %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
@@
 LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store cqn offsets
-CQN.GENE_EXPRESSION$offset %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MAYO_CBE_TCX_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('MAYO_CBE_TCX_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
-annotations(OFFSET_OBJ) = annotations(CODE)
-annotations(OFFSET_OBJ)$dataSubType = 'offset'
-OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 
@@
 |  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
-|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
 |  Design Matrix                              |  `r DM_OBJ$properties$id`        |

### 2017-03-20 code/Rmd/MSSM_REPROCESSED_TMM.Rmd
Added markdown to process MSSM reprocessed RNASeq with TMM and SVA normalaisation

--- 2017-03-19 code/Rmd/MSSM_REPROCESSED.Rmd
@@
 ---
-title: "Covariate analysis of MSSM reprocessed counts (with CQN normalisation and covariates adjusted)"
+title: "Covariate analysis of MSSM reprocessed counts (with TMM normalisation and covariates adjusted)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_REPROCESSED.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_REPROCESSED_TMM.Rmd",
                                  parentId = "syn7554308",
-                                 entityName = 'MSSM Reprocessed RNASeq (Normalised)')
+                                 entityName = 'MSSM Reprocessed RNASeq (Normalised - With out gene length and gc content adjustment)')
 ```
@@
 
-thisFileName <- 'MSSM_REPROCESSED.Rmd'
+thisFileName <- 'MSSM_REPROCESSED_TMM.Rmd'
 
@@
 ### Normalisation (with NULL)
-Initial normalisation usign cqn and dispersion control with voom (with NULL design)
+Initial normalisation usign TMM and dispersion control with voom (with NULL design)
 ```{r initial.voom.normalisation, fig.height=5, fig.width=5}
-# Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
-                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
+# Normalise using TMM
+TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')
+VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION)
 
 # Set gene counts in specific samples that are deviating 3 sd from other samples to 0
-log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
+log.mat = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
   mn = mean(x, na.rm = T)
@@
   # Fit linear model using new weights and new design
-  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
-                            design = DM1$design,
-                            weights = VOOM.GENE_EXPRESSION$weights,
-                            block = COVARIATES$individualIdentifier, 
-                            correlation = correlation$cor)
+  VOOM.ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
 
@@
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(TMM.GENE_EXPRESSION, design=DESIGN$design, plot=F)
 
 # Calculate correlation values of random effects
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
 correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)
@@
 # Calculate the number of surrogate variables
-NUM.SV = sva::num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
-                     MODEL1, method = 'be', seed = 123456, B = 30)
+NUM.SV = sva::num.sv(VOOM.WEIGHTS$E, MODEL1, method = 'be', seed = 123456, B = 30)
 
 # Estimate surrogate variables
-SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
-                    MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv
+SURR.VAR = sva::sva(VOOM.WEIGHTS$E, MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv
 
@@
 # Estimate voom weights for dispersion control
-VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                        design=cbind(MODEL1, SURR.VAR), plot=F)
+VOOM.SVA.WEIGHTS = voom(TMM.GENE_EXPRESSION, design=cbind(MODEL1, SURR.VAR), plot=F)
 
 # Calculate correlation values of random effects
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
 correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$individualIdentifier)
@@
 # Estimate voom weights for dispersion control
-VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+VOOM.NET.WEIGHTS = voom(TMM.GENE_EXPRESSION, 
                         design=DESIGN.NET, plot=F)
@@
 # Calculate correlation values of random effects
-VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
 correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$individualIdentifier)
@@
 # Code
-CODE <- Folder(name = "MSSM Reprocessed RNASeq (Normalised)", parentId = parentId)
+CODE <- Folder(name = "MSSM Reprocessed RNASeq (Normalised - With out gene length and gc content adjustment)", parentId = parentId)
 annotations(CODE) = all.annotations
@@
 LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store cqn offsets
-CQN.GENE_EXPRESSION$offset %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MSSM_FP_STG_PHG_IFG_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('MSSM_FP_STG_PHG_IFG_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
-annotations(OFFSET_OBJ) = annotations(CODE)
-annotations(OFFSET_OBJ)$dataSubType = 'offset'
-OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 

### 2017-03-20 code/R/computeCellTypeSpecificity.R
Minor modifications to the cell tyoe specificty computation code

@@ -73,3 +73,10 @@ obj = File('CellSpecificityMetrics.tsv', parentId = 'syn8077138')
 obj = synStore(obj, activityName = 'Compute cell type specificity',
                used = ALL_USED_IDs,
                executed = thisFile)
+
+# Store results to synapse
+write.table(ref.expr.sum, file = 'CombinedExpression.tsv', sep = '\t', row.names = F, quote=F)
+obj = File('CombinedExpression.tsv', name = 'Combined expression (log mean cpm)', parentId = 'syn8077138')
+obj = synStore(obj, activityName = 'Compute cell type specificity',
+               used = ALL_USED_IDs,
+               executed = thisFile)

### 2017-03-21 code/Rmd/MAYO_REPROCESSED_TMM.Rmd
fixed reprocessed metadata issue with MSSM cohorts

@@
 tmp.count = tmp.count[rownames(tmp.count) %in% genesToAnalyze,]
-CQN.GENE_EXPRESSION = cqn(tmp.count,
-                          x = GENE.GC[rownames(tmp.count), 'percentage_gc_content'],
-                          lengths = GENE.LEN[rownames(tmp.count), 'gene.length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
+VOOM.GENE_EXPRESSION = voom(tmp.count)
 
 # Initial normalisation of gene expression
-ind = rownames(CQN.GENE_EXPRESSION$y) %in% CONTROL.PROBES$ensembl_gene_id 
-EXPR = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+ind = rownames(VOOM.GENE_EXPRESSION$E) %in% CONTROL.PROBES$ensembl_gene_id 
+EXPR = VOOM.GENE_EXPRESSION$E
 EXPR = EXPR[ind,]
@@
 ### CPM Normalisation
-Preprocess counts matrix using CQN (Conditional Quantile Normlaisation accounting for both GC content and gene length)
+Preprocess counts matrix using TMM
 
@@
 ### Normalisation (with NULL)
-Initial normalisation usign cqn and dispersion control with voom (with NULL design)
+Initial normalisation usign TMM and dispersion control with voom (with NULL design)
 ```{r initial.voom.normalisation, fig.height=5, fig.width=5}

### 2017-03-21 code/Rmd/MSSM_REPROCESSED.Rmd
fixed reprocessed metadata issue with MSSM cohorts

@@
   dplyr::select(sampleIdentifier, BrodmannArea, barcode, individualIdentifier, batch, RIN) %>%
-  unique
+  unique %>%
+  ddply(.(barcode), .fun = function(x){
+    x = dplyr::mutate(x, batch = paste(batch, collapse = ''), RIN = sum(RIN, na.rm = T))  
+  })
 
@@
 ```
-Following samples were considered outliers and removed based on PCA and/or eucledian distance: 11311\_CER, 1923\_CER, 1923\_TCX, 11396\_TCX, 11294\_TCX, 11408\_TCX
+
 ```{r preprocess.data1}

### 2017-03-22 code/Rmd/MetaAnalysis.Rmd
temp commit

@@
 ```{r covariates}
-covariates.ids = c(ROSMAP = 'syn8018344',
-                   MSBB = 'syn8073656',
-                   MAYO = 'syn8028561')
-covariates = lapply(covariates.ids, function(id){
-  x = downloadFile(id)
-  colnames(x)[1] = 'SampleID'
-  return(x)
-})
+covariates.ids = c(ROSMAP = 'syn8456631',
+                   MSBB = 'syn8484996',
+                   MAYO = 'syn8466814')
+covariates = lapply(covariates.ids, downloadFile)
 
@@
 
-covariates$MSBB = dplyr::filter(covariates$MSBB, BrodmannArea.Dx %in% c('BM10.SD', 'BM22.SD', 'BM36.SD', 
-                                                                        'BM44.SD', 'BM36.ND', 'BM10.ND', 
-                                                                        'BM44.ND', 'BM22.ND')) %>%
-  dplyr::select(SampleID, BrodmannArea.Dx) %>%
-  tidyr::separate(BrodmannArea.Dx, c('BrainRegion', 'Dx')) %>%
+covariates$MSBB = covariates$MSBB %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
+  dplyr::filter(Diagnosis %in% c('SD', 'ND')) %>%
   dplyr::mutate(Study = 'MSSM',
-                Dx = factor(Dx, levels = c('ND','SD')),
+                Dx = factor(Diagnosis, levels = c('ND','SD')),
                 BrainRegion = factor(BrainRegion, labels = c('FP', 'STG', 'PHG', 'IFG'))) %>%
@@
 ```{r diffexp}
-diff.exp.ids = c(ROSMAP = 'syn8018358',
-                 MSBB = 'syn8073678',
-                 MAYO = 'syn8028572')
+diff.exp.ids = c(ROSMAP = 'syn8456721',
+                 MSBB = 'syn8502894',
+                 MAYO = 'syn8468023')
 diff.exp = lapply(diff.exp.ids, downloadFile)
@@
 pl = list()
-for(region in c('ROSMAP', 'MSBB', 'MAYO')){
+for(region in c('ROSMAP', 'MSSM', 'MAYO')){
   p = ggplot(tmp %>% filter(Study == region), aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point()
@@
 ```{r biomart}
-# Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "dec2016.archive.ensembl.org")
-
 # Get background gene ids
-backgroundGenes = sapply(diff.exp, function(x) x$ensembl_gene_id) %>%
-  Reduce(c,.)
-
-# Query biomart
-Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
-                       filters = "ensembl_gene_id", values = backgroundGenes,
-                       mart = mart)
-
-Entrez2Ensembl <- as.data.frame(org.Hs.egENSEMBL2EG)
-Entrez2HGNC <- as.data.frame(org.Hs.egSYMBOL2EG)
-HGNC2Ensembl <- left_join(Entrez2HGNC, Entrez2Ensembl)
-HGNC2Ensembl = HGNC2Ensembl[grep('^LOC',HGNC2Ensembl$symbol),c('ensembl_id', 'symbol')] %>%
-  filter(!is.na(ensembl_id), !is.na(symbol)) %>%
-  dplyr::rename(ensembl_gene_id = ensembl_id, hgnc_symbol = symbol)
-
-Ensemble2HGNC = rbindlist( list(Ensemble2HGNC, HGNC2Ensembl), use.names = T, fill = T) %>%
-  unique
+backgroundGenes = sapply(diff.exp, function(x) x$hgnc_symbol) %>%
+  Reduce(c,.) %>% unique() %>% setdiff(c(''))
 ```
@@
   dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05 & sign(CI.L) == sign(CI.R)) %>%
-    dplyr::select(Region, gene.id) %>%
+    dplyr::select(Region, ensembl_gene_id) %>%
     unique
@@
   rbindlist(use.names = T, fill = T) %>%
-  dlply(.(Region), .fun = function(x){x$gene.id})
+  dlply(.(Region), .fun = function(x){setdiff(unique(x$ensembl_gene_id),c(''))})
 
 tmp1 = rbindlist(diff.exp, use.names = T, fill = T) %>%
-  dplyr::filter(gene.id %in% Reduce(c, diff.exp.gl)) %>%
-  dplyr::select(-hgnc_symbol) %>%
-  left_join(Ensemble2HGNC)
-tmp1$hgnc_symbol[tmp1$ensembl_gene_id == 'ENSG00000275830'] = 'LOC101927646'
+  dplyr::filter(ensembl_gene_id %in% Reduce(c, diff.exp.gl))
 
@@
 tmp = tmp1 %>%
-  dplyr::filter(hgnc_symbol != '') %>%
-  group_by(hgnc_symbol) %>%
+  group_by(ensembl_gene_id) %>%
   summarise(count = sum(Direction != 'NONE')) %>%
-  filter(count >= 5) 
-writeLines(paste('Differentially expressed genes in any 5 brain regions at an FDR of 0.05 and fold change of 1.2 are',
+  filter(count == 7) %>%
+  left_join(dplyr::select(tmp1, ensembl_gene_id, hgnc_symbol) %>% unique)
+
+writeLines(paste('Differentially expressed genes in all 7 brain regions at an FDR of 0.05 and fold change of 1.2 are',
                  length(unique(tmp$hgnc_symbol))))
@@
 p = ggplot(tmp1 %>% 
-             filter(hgnc_symbol %in% tmp.eff.sz$hgnc_symbol[tmp.eff.sz$es>2.6]), 
+             filter(hgnc_symbol %in% c('MEIS3', 'PCSK1', 'ADAMTS2')), 
            aes(x = Region, y = logFC, color = Study, fill = Direction)) + geom_bar(stat = 'identity')
 p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
-p = p + theme(legend.position = 'top') + scale_fill_manual(values = c('grey', 'red'))
+p = p + theme(legend.position = 'top') + scale_fill_manual(values = c('green', 'red'))
 pl[[1]] = p
@@
 p = ggplot(tmp1 %>% 
-             filter(hgnc_symbol %in% tmp.eff.sz$hgnc_symbol[tmp.eff.sz$es<2.6]), 
+             filter(hgnc_symbol %in% c('FOXJ1', 'PRELP', 'S100A4')), 
            aes(x = Region, y = logFC, color = Study, fill = Direction)) + geom_bar(stat = 'identity')
 p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
-p = p + theme(legend.position = 'top')  + scale_fill_manual(values = c('green','grey','red'))
+p = p + theme(legend.position = 'top')  + scale_fill_manual(values = c('green','grey', 'red'))
 pl[[2]] = p
@@
   dlply(.(Region, Direction), .fun = function(x){
-    unique(x$hgnc_symbol)
+    setdiff(unique(x$hgnc_symbol),c(''))
   })
@@
 tmp1 = rbindlist(enrichResults, use.names = TRUE, idcol = 'ComparisonName', fill = TRUE) %>%
-  filter(!is.na(Category))
+  filter(!is.na(Category)) %>%
+  filter(fdr <= 0.01)
+
+tmp2 = dplyr::select(tmp1, ComparisonName, GeneSetName, Odds.Ratio) %>%
+  tidyr::spread(ComparisonName,Odds.Ratio)
+rownames(tmp2) = tmp2[,1]
 

### 2017-03-22 code/Rmd/ROSMAP_REPROCESSED.Rmd
Modified ROSMAP reprocessing to include new picard metrics and sva processing

@@
 synapseLogin()
+
+library(doParallel)
+library(foreach)
+
+cl = makeCluster(detectCores()-2)
+registerDoParallel(cl)
 
@@
 # Get picard metrics from synapse
-METADATA.PICARD_ID <- 'syn8380366'
+METADATA.PICARD_ID <- 'syn8512191'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
@@
   filter(Sampleid %in% colnames(COUNT)) %>%
-  filter(!is.na(cogdx)) %>%
+  filter(!is.na(cogdx), !is.na(braaksc), !is.na(ceradsc)) %>%
   filter(!is.na(RINcontinuous)) %>%
   filter(!is.na(pmi)) %>%
-  filter(!is.na(Sampleid)) %>%
   filter(!(Sampleid %in% c('380_120503'))) %>%
   as.data.frame()
+
+METADATA$Diagnosis = 'NONE'
+METADATA$Diagnosis[METADATA$cogdx == 1 & METADATA$braaksc <= 3 & METADATA$ceradsc > 2] = 'Control'
+METADATA$Diagnosis[METADATA$cogdx == 4 & METADATA$braaksc >= 4 & METADATA$ceradsc <= 2] = 'AD'
 ```
@@
 FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx")
-ContCovariates <- c("RINcontinuous", "age_death", "pmi", "educ", "PCT_PF_READS", "PCT_PF_READS_ALIGNED",
-                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")
+ContCovariates <- c("RINcontinuous", "age_death", "pmi", "educ", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
+                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
 
@@
 
-# Calculate the number of surrogate variables
-NUM.SV = sva::num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
-                     MODEL1, method = 'be', seed = 123456, B = 30)
+# Calculate the number of surrogate variables (be method)
+# NUM.SV = sva::num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, MODEL1, method = 'be', seed = 123456, B = 30)
+
+# Get residuals from differential expression model
+RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, VOOM.WEIGHTS$E)
+
+# Compute actual variance of all principal components
+tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
+actual.var = tmp$d^2/sum(tmp$d^2)
+
+# Compute permuted variance of all principal components
+permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
+  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
+  tmp = svd(tmp.residual)
+  permuted.var = tmp$d^2/sum(tmp$d^2)
+}
+permuted.var = apply(permuted.var, 2, mean)
+NUM.SV = sum(actual.var >= permuted.var[1])
+
+# Plot variance components
+var.comp = data.frame(component = 1:length(actual.var), 
+                      residual = actual.var, 
+                      permuted.residual = permuted.var) %>%
+  tidyr::gather(data, value, -component)
+p = ggplot(var.comp %>% filter(component <= 10), aes(x = component, y = value, color = data)) + geom_point()
+p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
+p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
+p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
+p
 
@@
                     MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv
+SURR.VAR = data.frame(SURR.VAR)
 
@@
 DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+stopCluster(cl)
 ```

### 2017-03-23 code/Rmd/ROSMAP_REPROCESSED.Rmd
Minor sanity fix in ROSMAP reprocessing

@@
 
+writeLines('Differentially expressed genes at an FDR 0.05 and logFC 1.2')
 tmp = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
-  group_by(Comparison, Direction) %>%
-  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id))) %>%
-  spread(Direction, FDR_0_05)
-
-tmp1 = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
   dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
@@
   spread(Direction, FDR_0_05_FC_1.2) 
-
-kable(full_join(tmp %>%
-                  dplyr::rename(DOWN_FDR_0_05 = DOWN, UP_FDR_0_05 = UP),
-                tmp1 %>%
-                  dplyr::rename(DOWN_FDR_0_05_FC_1.2 = DOWN, UP_FDR_0_05_FC_1.2 = UP)))
+kable(tmp)
 

### 2017-03-23 code/Rmd/MAYO_REPROCESSED.Rmd
Modified Mayo reprocessing to include modified algorithms for SVA and samples with picard reruns

@@
 FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
-ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI", "PCT_PF_READS", "PCT_PF_READS_ALIGNED",
-                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")
+ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
+                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
 
@@
                 Source = factor(Source))
+plotdata$SampleID[!(plotdata$SampleID %in% c("11311_CER", "1923_CER", "1923_TCX", "11396_TCX", "11294_TCX", "11408_TCX"))] = ''
 
@@
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Region~., scales = 'free_y')
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
 p
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diff.exp, fig.height=6, fig.width=18}
+```{r diff.exp, fig.height=6, fig.width=18, cache=FALSE}
 # Get design matrix
@@
 tmp = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction, Region) %>%
-  group_by(Region, Comparison, Direction) %>%
-  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id))) %>%
-  spread(Direction, FDR_0_05)
-
-tmp1 = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
   dplyr::select(ensembl_gene_id, Comparison, Direction, Region) %>%
@@
 # Source modified functions from sva package
-source('../R/num.sv.R')
+# source('../R/num.sv.R')
 source('../R/irwsva.build.R')
@@
 # Calculate the number of surrogate variables
-NUM.SV = num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
-                     MODEL1, method = 'be', seed = 123456, B = 30, tol = 1e-20)
+# NUM.SV = num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
+#                      MODEL1, method = 'be', seed = 123456, B = 30, tol = 1e-20)
+
+# Get residuals from differential expression model
+RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, VOOM.WEIGHTS$E)
+
+# Compute actual variance of all principal components
+tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
+actual.var = tmp$d^2/sum(tmp$d^2)
+
+# Compute permuted variance of all principal components
+permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
+  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
+  tmp = svd(tmp.residual)
+  permuted.var = tmp$d^2/sum(tmp$d^2)
+}
+permuted.var = apply(permuted.var, 2, mean)
+NUM.SV = sum(actual.var >= permuted.var[1])
+
+# Plot variance components
+var.comp = data.frame(component = 1:length(actual.var), 
+                      residual = actual.var, 
+                      permuted.residual = permuted.var) %>%
+  tidyr::gather(data, value, -component)
+p = ggplot(var.comp %>% filter(component <= 10), aes(x = component, y = value, color = data)) + geom_point()
+p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
+p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
+p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
+p
 

### 2017-03-23 code/Rmd/MSSM_REPROCESSED.Rmd
Modified MSSM reprocessing to include modified algorithms for SVA and samples with picard reruns

@@
 #### Obtain count matrix and metadata from synapse.
-```{r download.data, cache=TRUE}
+```{r download.data}
 # Get reprocessed sample counts for temporal cortex
@@
 ```{r preprocess.data,cache=FALSE, echo=TRUE}
+# Choose samples with rRNA rate less than 5% and more PF_READS_ALIGNED
+METADATA = METADATA %>%
+  ddply(.(barcode), .fun = function(x){
+    x %>%
+      dplyr::filter(PCT_RIBOSOMAL_BASES < 0.05) %>%
+      top_n(1, PF_READS_ALIGNED)
+  })
+
 # Fix metadata
@@
 Determine relationship between covariates
-```{r covariates.clustering, cache=TRUE}
+```{r covariates.clustering}
 primaryVariable <- "BrainRegion.Diagnosis"
 FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrainRegion.Diagnosis")
-ContCovariates <- c("RIN", "RIN2", "AOD","PMI", "PCT_PF_READS", "PCT_PF_READS_ALIGNED", 
-                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", 
-                    "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")
+ContCovariates <- c("RIN", "RIN2", "AOD","PMI",  "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
+                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
 
@@
 Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
-```{r preadj.covariates, cache=TRUE}
+```{r preadj.covariates}
 # Find correlation between PC's of gene expression with covariates
@@
 
+# Source modified functions from sva package
+# source('../R/num.sv.R')
+source('../R/irwsva.build.R')
+source('../R/f.pvalue.R')
+
 # Calculate the number of surrogate variables
-NUM.SV = sva::num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
-                     MODEL1, method = 'be', seed = 123456, B = 30)
+# NUM.SV = num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
+#                      MODEL1, method = 'be', seed = 123456, B = 30, tol = 1e-20)
+
+# Get residuals from differential expression model
+RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, VOOM.WEIGHTS$E)
+
+# Compute actual variance of all principal components
+tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
+actual.var = tmp$d^2/sum(tmp$d^2)
+
+# Compute permuted variance of all principal components
+permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
+  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
+  tmp = svd(tmp.residual)
+  permuted.var = tmp$d^2/sum(tmp$d^2)
+}
+permuted.var = apply(permuted.var, 2, mean)
+NUM.SV = sum(actual.var >= permuted.var[1])
+
+# Plot variance components
+var.comp = data.frame(component = 1:length(actual.var), 
+                      residual = actual.var, 
+                      permuted.residual = permuted.var) %>%
+  tidyr::gather(data, value, -component)
+p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.2)), aes(x = component, y = value, color = data)) + geom_point()
+p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
+p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
+p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
+p
 
@@
   study = 'MSSMRNASeq', 
-  center = 'MSSM',
+  center = 'MSBB',
 

### 2017-03-23 code/Rmd/MAYO_REPROCESSED.Rmd
Minor modifications

@@
 library(cqn)
+library(glmnet)
 
@@
 ### Data preprocessing
-```{r preprocess.data,cache=FALSE, echo=TRUE}
+```{r preprocess.data, echo=TRUE}
 # Fix metadata
@@
 Determine relationship between covariates
-```{r covariates.clustering, cache=TRUE}
+```{r covariates.clustering}
 primaryVariable <- "BrainRegion.Diagnosis"
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diff.exp, fig.height=6, fig.width=18, cache=FALSE}
+```{r diff.exp, fig.height=6, fig.width=18}
 # Get design matrix
@@
   tidyr::gather(data, value, -component)
-p = ggplot(var.comp %>% filter(component <= 10), aes(x = component, y = value, color = data)) + geom_point()
+p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.2)), aes(x = component, y = value, color = data)) + geom_point()
 p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')

### 2017-03-23 code/Rmd/MAYO_REPROCESSED.Rmd
Excluded samples tagged as outliers from MAYO group

@@
 rownames(GENE.GC) = GENE.GC$ensembl_gene_id
+
+# Samples excluded by investigators
+SAMPLES.EXCLUDE = c(downloadFile('syn6126119')$`Sample Name`,
+                    downloadFile('syn6126114')$`Sample Name`)
+ALL_USED_IDs = c(ALL_USED_IDs, 'syn6126119', 'syn6126114')
 ```
+Following `r length(SAMPLES.EXCLUDE)` samples were removed from analysis as original investigators tagged them as outliers for various reasons as described here syn6126119, syn6126114
+Samples removed are `r paste(SAMPLES.EXCLUDE, collapse = ', ')`
 
@@
   dplyr::filter(!(ID %in% c("11311_CER", "1923_CER", "1923_TCX",
-                            "11396_TCX", "11294_TCX", "11408_TCX")))
+                            "11396_TCX", "11294_TCX", "11408_TCX",
+                            SAMPLES.EXCLUDE)))
 ```

### 2017-04-06 code/Rmd/harmonisedCaseControl.Rmd
Added markdown to harmonise case control definitions

--- 2017-03-23 code/Rmd/MAYO_REPROCESSED.Rmd
@@
 ---
-title: "Covariate analysis of MAYO reprocessed counts (with CQN normalisation and covariates adjusted)"
+title: "Analysing number of case and control based on new harmonised definition"
 author: "Thanneer Perumal"
@@
 ---
-```{r knit2synapse, eval=FALSE}
-library(synapseClient)
-library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')
-
-synapseLogin()
-
-knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED.Rmd",
-                                 parentId = "syn7554308",
-                                 entityName = 'MAYO Reprocessed RNASeq (Normalised)')
-```
-
 ```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
@@
 
-cl = makeCluster(detectCores()-2)
-doParallel::registerDoParallel(cl)
-
 synapseLogin()
@@
 
-```{r synapse.parameters, include=FALSE}
-# Synapse parameters
-parentId = 'syn7554308'
-activityName = 'Covariate analysis';
-activityDescription = 'Covariate analysis of MAYO reprocessed RNASeq data';
-
-thisFileName <- 'MAYO_REPROCESSED.Rmd'
-
-# Github link
-thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
-thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
-```
 ### Data download
@@
 METADATA = METADATA[indToRetain,]
+
+all.counts = list()
+all.mtd = list()
+
+all.counts$MAYO = COUNT
+all.mtd$MAYO = METADATA
 ```
-`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project.
 
-Following samples were removed `r indRemoved`
+### Data download
+#### Obtain count matrix and metadata from synapse.
+```{r download.data}
+# Get reprocessed sample counts for temporal cortex
+COUNT_ID <- 'syn7994853';
+ALL_USED_IDs <- COUNT_ID
+COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F)
 
-### Predict missing PMI
-```{r predict.pmi}
-PMI = METADATA[,c('PMI', 'ID'), drop = F]
-PMI$PMI = PMI$PMI * 60
-rownames(PMI) = METADATA$ID
+# Convert rownames of counts from tracking id to ensemble gene id
+tmp = data.frame(Gene.ID = rownames(COUNT)) %>%
+  dplyr::mutate(ID = Gene.ID) %>%
+  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')
+rownames(tmp) = tmp$Gene.ID
+rownames(COUNT) = tmp[rownames(COUNT), 'ensembl_gene_id']
 
-# Get genes that are predictors of PMI and use them as control probes
-CONTROL.PROBES_ID = 'syn6145639'
-ALL_USED_IDs = c(ALL_USED_IDs, CONTROL.PROBES_ID)
-CONTROL.PROBES = fread(synGet(CONTROL.PROBES_ID)@filePath, data.table = F) %>%
-  dplyr::mutate(ID = x) %>%
-  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')
+# Get sample ids
+SampleID = data.frame(SampleID = colnames(COUNT),
+                 ID = colnames(COUNT)) %>% 
+  tidyr::separate(ID, c('A','B','ID'), sep = '_') %>%
+  dplyr::select(SampleID, ID) %>%
+  dplyr::mutate(ID = as.numeric(ID))
 
-# Conditional quantile normalisation
-tmp.count = COUNT
-genesToAnalyze = unique(intersect(rownames(GENE.GC)[!is.na(GENE.GC$percentage_gc_content)], 
-                                  rownames(GENE.LEN)[!is.na(GENE.LEN$gene.length)]))
-tmp.count = tmp.count[rownames(tmp.count) %in% genesToAnalyze,]
-CQN.GENE_EXPRESSION = cqn(tmp.count,
-                          x = GENE.GC[rownames(tmp.count), 'percentage_gc_content'],
-                          lengths = GENE.LEN[rownames(tmp.count), 'gene.length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
+# Get technical metadata
+METADATA_TECH_ID <- 'syn6100548'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECH_ID
+METADATA_TECH <- read.table(synGet(METADATA_TECH_ID)@filePath,sep=',',header=T, row.names=1) %>%
+  group_by(sampleIdentifier) %>%
+  top_n(1, -rRNA.rate) %>%
+  dplyr::select(sampleIdentifier, BrodmannArea, barcode, individualIdentifier, batch, RIN) %>%
+  unique %>%
+  ddply(.(barcode), .fun = function(x){
+    x = dplyr::mutate(x, batch = paste(batch, collapse = ''), RIN = sum(RIN, na.rm = T))  
+  })
 
-# Initial normalisation of gene expression
-ind = rownames(CQN.GENE_EXPRESSION$y) %in% CONTROL.PROBES$ensembl_gene_id 
-EXPR = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-EXPR = EXPR[ind,]
-PMI = PMI[colnames(EXPR),,drop=F]
+# Get clinical metadata
+METADATA_CLINICAL_ID <- 'syn6101474'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CLINICAL_ID
+METADATA_CLINICAL <- read.csv(synGet(METADATA_CLINICAL_ID)@filePath, header = T)
 
-expr = dlply(PMI, .(ID), function(mtd, expr){
-  if (dim(mtd)[1] > 1)
-    expr = rowMeans(expr[, mtd$ID]) %>% rownameToFirstColumn('GeneID')
-  else
-    expr = expr[, mtd$ID] %>% rownameToFirstColumn('GeneID')
-  expr = expr %>% plyr::rename(c('DF' = unique(mtd$ID)))
-  return(expr)
-}, EXPR) %>%
-  join_all
-rownames(expr) = expr$GeneID
+# Get picard metrics from synapse
+METADATA.PICARD_ID <- 'syn8380379'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
+METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
+METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
+METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
+METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)
 
-PMI = PMI[,colnames(PMI) %in% c('ID','PMI')] %>% unique
-rownames(PMI) = PMI$ID
+METADATA.PICARD$sampleIdentifier = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
+  gsub('.accepted_hits.sort.coord/picard.CombinedMetrics.csv','',.)
+colnames(METADATA.PICARD) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD))
+colnames(METADATA.PICARD) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD))
 
-PMI = PMI[intersect(rownames(PMI), colnames(expr)), ]
-expr = expr[,intersect(rownames(PMI), colnames(expr))]
+METADATA <- full_join(METADATA_TECH, METADATA_CLINICAL) %>%
+  inner_join(METADATA.PICARD)
 
-# Split data into trainining and prediction set
-x.train = expr[,!is.na(PMI$PMI)]
-x.predict = expr[,is.na(PMI$PMI)]
+# Get gene specific parameters from synapse
+GENE.PARAM = downloadFile('syn8449369')
+ALL_USED_IDs = c(ALL_USED_IDs, 'syn8449369')
 
-y.train = PMI[!is.na(PMI$PMI),]
+GENE.LEN = dplyr::select(GENE.PARAM, ensembl_gene_id, gene.length) %>% unique()
+rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id
 
-plr.model = cv.glmnet(t(x.train), y.train$PMI, family = 'gaussian')
-coeff = as.matrix(coef.glmnet(plr.model, s = 'lambda.min'))
-mse = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train$PMI)^2)/length(y.train$PMI)
-y.predict = predict(plr.model, t(x.predict), s = 'lambda.min', type = 'response') %>%
-  rownameToFirstColumn('ID') %>%
-  plyr::rename(c('1' = 'PMI'))
+GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% unique()
+rownames(GENE.GC) = GENE.GC$ensembl_gene_id
+```
 
-PMI = rbindlist(list(y.train, y.predict), use.names = T, fill = T)
-PMI$PMI[PMI$PMI < 0] = 0
+### Data preprocessing
+```{r preprocess.data,cache=FALSE, echo=TRUE}
+# Choose samples with rRNA rate less than 5% and more PF_READS_ALIGNED
+METADATA = METADATA %>%
+  ddply(.(barcode), .fun = function(x){
+    x %>%
+      dplyr::filter(PCT_RIBOSOMAL_BASES < 0.05) %>%
+      top_n(1, PF_READS_ALIGNED)
+  })
+
+# Fix metadata
+METADATA = METADATA %>%
+  dplyr::mutate(AOD = gsub('\\+','',AOD),
+                BrainRegion = factor(BrodmannArea,
+                                     levels = c('BM10', 'BM22', 'BM36', 'BM44'),
+                                     labels = c('BM10' = 'FP', 'BM22' = 'STG', 
+                                                'BM36' = 'PHG', 'BM44' = 'IFG')),
+                Diagnosis = 'NONE', 
+                RIN2 = RIN^2)
+METADATA$Diagnosis[METADATA$CDR <= 1] = 'ND'
+METADATA$Diagnosis[METADATA$CDR == 2 | METADATA$CDR == 3] = 'MD'
+METADATA$Diagnosis[METADATA$CDR >= 4] = 'SD'
+levels(METADATA$batch)[levels(METADATA$batch) == ''] = 'NoBatch'
 
 METADATA = METADATA %>%
-  dplyr::select(-PMI) %>%
-  left_join(PMI)
+  tidyr::unite(BrainRegion.Diagnosis, BrainRegion, Diagnosis, sep = '.')
+
+METADATA <- METADATA %>%
+  filter(!is.na(BrainRegion.Diagnosis), !is.na(PMI), !is.na(RIN))
 ```
 
-### Covariate clustering
-Determine relationship between covariates
-```{r covariates.clustering}
-primaryVariable <- "BrainRegion.Diagnosis"
-FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
-ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
-                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
+```{r preprocess.data1}
+# Match covariates to expression data
+indToRetain = intersect(METADATA$sampleIdentifier, colnames(COUNT))
+indRemoved = setdiff(colnames(COUNT), METADATA$sampleIdentifier)
 
-# Find inter relation between factor covariates
-COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
-rownames(COVARIATES) <- METADATA$ID
+COUNT <- COUNT[, indToRetain]
 
-# Convert factor covariates to factors
-COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
-COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
-COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
+METADATA = as.data.frame(METADATA)
+rownames(METADATA) = METADATA$sampleIdentifier
+METADATA = METADATA[indToRetain,]
+
+all.counts$MSSM = COUNT
+all.mtd$MSSM = METADATA
 ```
-Correlation/association between covariates at an FDR <= 0.1
-```{r covariates.correlation, fig.width=6, fig.height=6}
-COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
-draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
+### Data download
+#### Obtain count matrix and metadata from synapse.
+```{r download.data, cache=TRUE}
+# Download expression data
+COUNT_ID <- 'syn7750270';
+ALL_USED_IDs <- COUNT_ID
+COUNT_OBJ <- synGet(COUNT_ID)
+COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F)
+COUNT[,grep('150_120419', colnames(COUNT))[2]] = NULL
+
+# Convert rownames of counts from tracking id to ensemble gene id
+tmp = data.frame(Gene.ID = rownames(COUNT)) %>%
+  dplyr::mutate(ID = Gene.ID) %>%
+  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')
+rownames(tmp) = tmp$Gene.ID
+rownames(COUNT) = tmp[rownames(COUNT), 'ensembl_gene_id']
+
+# Get clinical metadata
+METADATA.CLINICAL_ID <- 'syn3191087'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.CLINICAL_ID
+METADATA.CLINICAL_OBJ <- synGet(METADATA.CLINICAL_ID)
+METADATA.CLINICAL <- read.table(METADATA.CLINICAL_OBJ@filePath,sep=',',header=T)
+
+# Get technical covariates
+METADATA.TECH_ID <- 'syn4300313'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.TECH_ID
+METADATA.TECH_OBJ <- synGet(METADATA.TECH_ID)
+METADATA.TECH <- read.table(METADATA.TECH_OBJ@filePath,sep='\t',header=T)
+
+# Get picard metrics from synapse
+METADATA.PICARD_ID <- 'syn8512191'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
+METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
+METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
+METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
+METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)
+
+METADATA.PICARD$Sampleid = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
+  gsub('Aligned.out/picard.CombinedMetrics.csv','',.)
+colnames(METADATA.PICARD) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD))
+colnames(METADATA.PICARD) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD))
+
+# Fix error in technical covariates data
+KEY_ID <- 'syn3382527'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = KEY_ID  
+KEY <- read.csv(synGet(KEY_ID)@filePath) %>% 
+  filter(mrna_data == 1) %>%
+  dplyr::select(projid, mrna_id) %>%
+  separate(mrna_id, c('a','b','batch'), sep = '_') %>% 
+  unite(Sampleid, a, b) %>%
+  dplyr::select(-batch) %>%
+  unique
+
+# Match technical and clinical covariates
+METADATA <- METADATA.TECH %>%
+  left_join(METADATA.PICARD) %>%
+  dplyr::select(-projid) %>%
+  dplyr::left_join(KEY) %>%
+  dplyr::left_join(METADATA.CLINICAL)
+
+# Pick higher quality RIN batch for sample 492_120515
+METADATA <- METADATA %>%
+  group_by(Sampleid) %>%
+  top_n(1, RINcontinuous)
+
+# Get gene specific parameters from synapse
+GENE.PARAM = downloadFile('syn8449369')
+ALL_USED_IDs = c(ALL_USED_IDs, 'syn8449369')
+
+GENE.LEN = dplyr::select(GENE.PARAM, ensembl_gene_id, gene.length) %>% unique()
+rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id
+
+GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% unique()
+rownames(GENE.GC) = GENE.GC$ensembl_gene_id
 ```
 
-```{r data.explore, fig.width = 12, fig.height = 8}
-my.theme = theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
+### Data preprocessing
+```{r preprocess.data,cache=TRUE, echo=TRUE}
+# Remove samples with no cogdx, RIN and PMI scores
+METADATA <- METADATA %>%
+  ungroup %>%
+  filter(Sampleid %in% colnames(COUNT)) %>%
+  filter(!is.na(cogdx), !is.na(braaksc), !is.na(ceradsc)) %>%
+  filter(!is.na(RINcontinuous)) %>%
+  filter(!is.na(pmi)) %>%
+  filter(!(Sampleid %in% c('380_120503'))) %>%
+  as.data.frame()
 
-# RIN
-p = list()
-p[[1]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = RIN)) + geom_boxplot()
-p[[1]] = p[[1]] + ggtitle('RIN') + my.theme
-
-# AgeAtDeath
-p[[2]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(AgeAtDeath))) + geom_boxplot()
-p[[2]] = p[[2]] + ggtitle('AgeAtDeath') + my.theme
-
-# PMI
-p[[3]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PMI))) + geom_boxplot()
-p[[3]] = p[[3]] + ggtitle('PMI') + my.theme
-
-# Ribosomal bases
-p[[4]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_RIBOSOMAL_BASES))) + geom_boxplot()
-p[[4]] = p[[4]] + ggtitle('Fraction of Ribosomal Bases') + my.theme
-
-# Intronic bases
-p[[5]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTRONIC_BASES))) + geom_boxplot()
-p[[5]] = p[[5]] + ggtitle('Fraction Intronic Bases') + my.theme
-
-# Intergenic bases
-p[[6]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTERGENIC_BASES))) + geom_boxplot()
-p[[6]] = p[[6]] + ggtitle('Fraction Intergenic Bases') + my.theme
-
-multiplot(plotlist = p, cols = 3)
-```
-### CPM Normalisation
-Preprocess counts matrix using CQN (Conditional Quantile Normlaisation accounting for both GC content and gene length)
-
-Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion x Diagnosis. Also remove genes with missing gene length and percentage GC content
-```{r cpmnormalisation}
-genesToAnalyze = COVARIATES %>%
-  rownameToFirstColumn('SampleID') %>%
-  dlply(.(BrainRegion.Diagnosis), .fun = function(mtd, count){
-    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
-                                                   MIN_GENE_CPM=1, 
-                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
-  processed.counts$filteredExprMatrix$genes
-}, COUNT)
-
-genesToAnalyze = unlist(genesToAnalyze) %>% 
-  unique() %>%
-  intersect(GENE.GC$ensembl_gene_id[!is.na(GENE.GC$percentage_gc_content)]) %>%
-  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$gene.length)]) %>%
-  setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous"))
-
-PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], 
-                                                 MIN_GENE_CPM=0, 
-                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
-```
-Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples
-
-### Normalisation (with NULL)
-Initial normalisation usign cqn and dispersion control with voom (with NULL design)
-```{r initial.voom.normalisation, fig.height=5, fig.width=5}
-# Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
-                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
-
-# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
-log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
-  mn = mean(x, na.rm = T)
-  std.dev = sd(x, na.rm = T)
-  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
-}) %>% t
-PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0
-
-# Re-compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
-                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
-```
-Coexpression of genes 
-```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
-### Outlier analysis
-Clustering of voom adjusted data (with NULL design)
-```{r decompse.normalise.data1.1, fig.height=6, fig.width=8, results='asis'}
-# Find principal components of expression to plot
-PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)
-
-# Plot first 2 PCs
-plotdata <- data.frame(SampleID=rownames(PC$rotation), 
-                       PC1=PC$rotation[,1], 
-                       PC2=PC$rotation[,2])
-
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  tidyr::separate(BrainRegion.Diagnosis, c('Region','Diagnosis'), sep = '\\.') %>% 
-  dplyr::mutate(Region = factor(Region),
-                Source = factor(Source))
-plotdata$SampleID[!(plotdata$SampleID %in% c("11311_CER", "1923_CER", "1923_TCX", "11396_TCX", "11294_TCX", "11408_TCX"))] = ''
-
-p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
-p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Region~., scales = 'free_y')
-p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
-p
-```
-Tree based classification of samples
-```{r decompse.normalise.data1.2, fig.height=6, fig.width=10, results='asis'}
-# Eucledian tree based analysis
-COVARIATES.tmp = (COVARIATES[,c(FactorCovariates), drop = F])
-COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
-
-tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset)))
-cols = WGCNA::labels2colors(COVARIATES.tmp);
-
-WGCNA::plotDendroAndColors(tree, 
-                           colors = cols, 
-                           dendroLabels = FALSE, 
-                           abHeight = 0.80, 
-                           main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp))
+METADATA$Diagnosis = 'NONE'
+METADATA$Diagnosis[METADATA$cogdx == 1 & METADATA$braaksc <= 3 & METADATA$ceradsc > 2] = 'Control'
+METADATA$Diagnosis[METADATA$cogdx == 4 & METADATA$braaksc >= 4 & METADATA$ceradsc <= 2] = 'AD'
 ```
 
-```{r temp, include = F}
-dev.off()
-gc()
+```{r preprocess.data1,cache=TRUE}
+# Match covariates to expression data
+indToRetain = intersect(METADATA$Sampleid, colnames(COUNT))
+removedIDs = setdiff(colnames(COUNT), METADATA$Sampleid)
+
+COUNT = COUNT[,indToRetain]
+
+rownames(METADATA) = METADATA$Sampleid
+METADATA = METADATA[indToRetain,]
+
+all.counts$ROSMAP = COUNT
+all.mtd$ROSMAP = METADATA
 ```
 
-### Significant Covariates
-Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
-```{r preadj.covariates}
-# Find correlation between PC's of gene expression with covariates
-preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
-                                                 COVARIATES,
-                                                 'NULL design(voom-normalized)', 
-                                                 isKeyPlot=TRUE, 
-                                                 MIN_PVE_PCT_PC = 1)
+```{r new.dx}
+all.mtd$ROSMAP$nDx = 'OTHER'
+all.mtd$ROSMAP$nDx[all.mtd$ROSMAP$cogdx==1 & all.mtd$ROSMAP$braaksc <= 3 & all.mtd$ROSMAP$ceradsc <= 1] = 'CONTROL'
+all.mtd$ROSMAP$nDx[all.mtd$ROSMAP$cogdx==4 & all.mtd$ROSMAP$braaksc >= 4 & all.mtd$ROSMAP$ceradsc >= 2] = 'AD'
+
+table(all.mtd$ROSMAP$nDx)
+
+all.mtd$MSSM$nDx = 'OTHER'
+all.mtd$MSSM$nDx[all.mtd$MSSM$CDR <= 0.5 & all.mtd$MSSM$bbscore <= 3 & all.mtd$MSSM$NP.1 <= 1] = 'CONTROL'
+all.mtd$MSSM$nDx[all.mtd$MSSM$CDR >= 1 & all.mtd$MSSM$bbscore >= 4 & all.mtd$MSSM$NP.1 >= 2] = 'AD'
+
+table(all.mtd$MSSM$nDx, all.mtd$MSSM$BrodmannArea)
+
+table(all.mtd$MAYO$Diagnosis, all.mtd$MAYO$Tissue)
 ```
-
-Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
-```{r preadj.covariates.plot, fig.width=12, fig.height=8}
-preAdjustedSigCovars[["PC_res"]][[2]]$plotData
-```
-### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising and re-adjusting COUNTS with an iterative design matrix.
-
-NOTE:
-1. Using a mixed effect model where random effect is chosen as Donor_ID
-2. Adding Source, FLOWCELL and Gender a priori to variable selection
-3. Primary variable of interest BrainRegion.Diagnosis is excluded from the pool of available covariates for selection
-```{r iterative.norm, results='asis'}
-# Primary variable of interest
-postAdjustCovars = c('Source', 'FLOWCELL', 'Gender');
-
-# Assign residual covariates
-residualCovars = setdiff(preAdjustedSigCovars$significantCovars, 
-                         c(postAdjustCovars, primaryVariable, 'Donor_ID'))
-residualSigCovars = preAdjustedSigCovars
-covariatesEffects = preAdjustedSigCovars$Effects.significantCovars[residualCovars]
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
-
-loopCount = 0 
-while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 20){
-  writeLines(paste('Using following covariates in the model:',
-                   paste(postAdjustCovars, collapse=', '),
-                   'as fixed effects and Donor_ID as random effect'))
-
-  # Post adjusted design matrix
-  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
-
-  # Estimate voom weights for dispersion control
-  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)
-
-  # Calculate correlation values of random effects
-  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
-  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
-                                             block = COVARIATES$Donor_ID)
-
-  # Re-estimate voom weights
-  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                              design=DM1$design, plot=F, 
-                              block = COVARIATES$Donor_ID, 
-                              correlation = correlation$cor)
-
-  # Fit linear model using new weights and new design
-  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
-                            design = DM1$design,
-                            weights = VOOM.GENE_EXPRESSION$weights,
-                            block = COVARIATES$Donor_ID, 
-                            correlation = correlation$cor)
-
-  # Residuals after normalisation
-  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
-                                                CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
-
-  # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates),
-                         c(postAdjustCovars, primaryVariable, 'Donor_ID'))
-
-  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                                COVARIATES[, residCovars, drop=F], 
-                                                'adjusted design(voom-normalized)',
-                                                isKeyPlot=TRUE)
-
-  # Add postadjusted covariates (if any)
-  residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, primaryVariable))
-  covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]
-
-  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
-  loopCount = loopCount + 1
-}
-
-modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
-                  'as fixed effects')
-
-tmp <- paste('Using following covariates in the final model:', modelStr)
-```
-
-### Sanity check
-```{r residual.adj, fig.width=12, fig.height=8}
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                              COVARIATES,
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
-
-residualSigCovars[["PC_res"]][[2]]$plotData
-```
-Coexpression of genes 
-```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.GENE_EXPRESSION))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
-
-### Clustering residual data
-```{r decompse.normalise.data2.1, fig.height=10, fig.width=8, results='asis'}
-# Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
-
-# Plot first 4 PCs
-plotdata <- data.frame(SampleID=rownames(PC$rotation), 
-                       PC1=PC$rotation[,1], 
-                       PC2=PC$rotation[,2])
-
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  tidyr::separate(BrainRegion.Diagnosis, c('Region','Diagnosis'), sep = '\\.') %>% 
-  dplyr::mutate(Region = factor(Region),
-                Source = factor(Source))
-
-p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
-p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Region~., scales = 'free_y')
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
-p
-```
-
-```{r decompse.normalise.data2.2, fig.height=8, fig.width=12, results='asis'}
-# Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
-COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
-
-tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
-cols = WGCNA::labels2colors(COVARIATES.tmp);
-
-WGCNA::plotDendroAndColors(tree, 
-                           colors = cols, 
-                           dendroLabels = FALSE, 
-                           abHeight = 0.80, 
-                           main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp))
-```
-
-```{r temp1, include=F}
-dev.off()
-gc()
-```
-
-### Differential expression analysis (with BrainRegion.Diagnosis as primary variable)
-Differential expression is performed on the primary variable by controlling for covariates identified above
-
-Interpretation of BrainRegion.Diagnosis
-1. TCX.PSP: Progressive Supranuclear Palsy samples from Temporal cortex region
-2. TCX.AD: Alzheimers Disease samples from Temporal cortex region
-3. TCX.Control: Control samples from Temporal cortex region
-4. TCX.Pathologic_Aging: Pathological Aging samples from Temporal cortex region
-5. CER.PSP: Progressive Supranuclear Palsy samples from Cerebellum region
-6. CER.AD: Alzheimers Disease samples from Cerebellum region
-7. CER.Control: Control samples from Cerebellum region
-8. CER.Pathologic_Aging: Pathological Aging samples from Cerebellum region
-
-Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diff.exp, fig.height=6, fig.width=18}
-# Get design matrix
-DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable, postAdjustCovars), drop = F], Intercept = F)
-DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
-
-# Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
-
-# Calculate correlation values of random effects
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
-correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Donor_ID)
-
-# Re-estimate voom weights
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
-                    design=DESIGN$design, plot=F,
-                    block = COVARIATES$Donor_ID, 
-                    correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
-FIT = lmFit(VOOM.WEIGHTS)
-
-# Fit contrast
-contrast = makeContrasts(contrasts=c(
-  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.AD",
-  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.Pathologic_Aging",
-  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.PSP",
-  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.AD",
-  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.PSP",
-  "BrainRegion.DiagnosisCER.PSP-BrainRegion.DiagnosisCER.AD",
-  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.AD",
-  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.Pathologic_Aging",
-  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.PSP",
-  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.AD",
-  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.PSP",
-  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD"),
-  levels = colnames(FIT$coefficients))
-FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
-FIT.CONTR = eBayes(FIT.CONTR)
-
-# Get differnetial expression
-DE = lapply(1:12, function(i, FIT){
-  topTable(FIT, coef=i, number = 50000, confint = T) %>%
-    rownameToFirstColumn('ensembl_gene_id')
-}, FIT.CONTR)
-names(DE) = colnames(contrast)
-
-DE = DE %>% 
-  rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
-                Study = 'MAYO',
-                Direction = logFC/abs(logFC),
-                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
-                Direction = as.character(Direction)) %>%
-  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
-  tidyr::separate(from.state, c('Region','reference'), sep = '\\.') %>%
-  tidyr::separate(to.state, c('Region','against'), sep = '\\.') %>%
-  tidyr::unite(Comparison, reference, against, sep = '-') %>%
-  left_join(GENE.PARAM %>%
-              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
-              unique())
-DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
-
-tmp = DE %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction, Region) %>%
-  group_by(Region, Comparison, Direction) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
-  spread(Direction, FDR_0_05_FC_1.2) 
-kable(tmp)
-
-p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
-p = p + scale_color_manual(values = c('green','grey','red'))
-p = p + facet_grid(Region+.~Comparison, scales = 'fixed')
-p
-```
-
-### Associate differential expression results with gc content, gene length and average expression
-```{r associate.de, fig.height=5, fig.width=15}
-pl = list()
-pl[[1]] = ggplot(DE %>%
-                   dplyr::filter(Comparison == 'Control-AD'), 
-                 aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
-pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
-pl[[1]] = pl[[1]] + theme(legend.position = 'top')
-
-pl[[2]] = ggplot(DE %>%
-                   dplyr::filter(Comparison == 'Control-AD'), 
-                 aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
-pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
-pl[[2]] = pl[[2]] + theme(legend.position = 'top')
-
-pl[[3]] = ggplot(DE %>%
-                   dplyr::filter(Comparison == 'Control-AD'), 
-                 aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
-pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
-pl[[3]] = pl[[3]] + theme(legend.position = 'top')
-
-multiplot(plotlist = pl, cols = 3)
-```
-
-### SVA Adjustments for eQTL analysis
-Conditioned on primary variable (BrainRegion.Diagnosis) and identified covariates estimate surrogate variables using SVA package
-```{r sva.adjust}
-# Get (null) design matrix
-MODEL0 = DESIGN$design[,-grep(primaryVariable,colnames(DESIGN$design))] # Get null model by removing variable of interest
-MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
-MODEL1 = DESIGN$design
-MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
-
-# Source modified functions from sva package
-# source('../R/num.sv.R')
-source('../R/irwsva.build.R')
-source('../R/f.pvalue.R')
-
-# Calculate the number of surrogate variables
-# NUM.SV = num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
-#                      MODEL1, method = 'be', seed = 123456, B = 30, tol = 1e-20)
-
-# Get residuals from differential expression model
-RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, VOOM.WEIGHTS$E)
-
-# Compute actual variance of all principal components
-tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
-actual.var = tmp$d^2/sum(tmp$d^2)
-
-# Compute permuted variance of all principal components
-permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
-  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
-  tmp = svd(tmp.residual)
-  permuted.var = tmp$d^2/sum(tmp$d^2)
-}
-permuted.var = apply(permuted.var, 2, mean)
-NUM.SV = sum(actual.var >= permuted.var[1])
-
-# Plot variance components
-var.comp = data.frame(component = 1:length(actual.var), 
-                      residual = actual.var, 
-                      permuted.residual = permuted.var) %>%
-  tidyr::gather(data, value, -component)
-p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.2)), aes(x = component, y = value, color = data)) + geom_point()
-p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
-p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
-p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
-p
-
-# Estimate surrogate variables
-SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
-                        MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
-
-colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
-rownames(SURR.VAR) = rownames(MODEL1)
-
-# Estimate voom weights for dispersion control
-VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                        design=cbind(MODEL1, SURR.VAR), plot=F)
-
-# Calculate correlation values of random effects
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
-correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$Donor_ID)
-
-# Re-estimate voom weights
-VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                        design = cbind(MODEL1, SURR.VAR), plot=F,
-                        block = COVARIATES$Donor_ID, 
-                        correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
-VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)
-
-# Residuals after normalisation
-RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
-                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
-```
-
-### Covariates Adjustments for network analysis
-Identified covariates are regressed out from the expression matrix for network analysis
-```{r network.adjust}
-# Get design matrix
-DESIGN.NET = getDesignMatrix(COVARIATES[, postAdjustCovars, drop = F], Intercept = F)
-DESIGN.NET = DESIGN.NET$design[,linColumnFinder(DESIGN.NET$design)$indepCols]
-
-# Estimate voom weights for dispersion control
-VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                        design=DESIGN.NET, plot=F)
-
-# Calculate correlation values of random effects
-VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
-correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$Donor_ID)
-
-# Re-estimate voom weights
-VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                        design = DESIGN.NET, plot=F,
-                        block = COVARIATES$Donor_ID, 
-                        correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
-VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)
-
-# Residuals after normalisation
-RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
-                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
-```
-
-### Store files in synapse
-```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
-# Set annotations
-all.annotations = list(
-  dataType = 'mRNA',
-  dataSubType = 'geneExp',
-  summaryLevel = 'gene',
-  assay	 = 'Reprocessed RNAseq',
-
-  tissueTypeAbrv	= 'CBE, TCX', 
-  study = 'MayoRNASeq', 
-  center = 'UFL-MAYO-ISB',
-
-  organism = 'HomoSapiens',
-  consortium	= 'AMP-AD',
-
-  normalizationStatus	= TRUE,
-  modelSystem	= FALSE,
-
-  normalizationType	= 'CQN'
-)
-
-# Code
-CODE <- Folder(name = "MAYO Reprocessed RNASeq (Normalised)", parentId = parentId)
-annotations(CODE) = all.annotations
-CODE <- synStore(CODE)
-
-# Store covariates
-COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
-write.table(COVARIATES, file = 'MAYO_CBE_TCX_Covariates.tsv', sep = '\t', row.names=F, quote=F)
-COV_OBJ = File('MAYO_CBE_TCX_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
-annotations(COV_OBJ) = annotations(CODE)
-annotations(COV_OBJ)$dataSubType = 'covariates'
-COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                      executed = thisFile, activityDescription = activityDescription)
-
-# Store filtered counts
-PROCESSED_COUNTS$filteredExprMatrix$counts %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MAYO_CBE_TCX_Counts.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('MAYO_CBE_TCX_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
-annotations(COUNT_OBJ) = annotations(CODE)
-annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
-COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store logCPM
-CQN.GENE_EXPRESSION$y %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MAYO_CBE_TCX_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('MAYO_CBE_TCX_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
-annotations(LCOUNT_OBJ) = annotations(CODE)
-annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
-LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store cqn offsets
-CQN.GENE_EXPRESSION$offset %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MAYO_CBE_TCX_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('MAYO_CBE_TCX_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
-annotations(OFFSET_OBJ) = annotations(CODE)
-annotations(OFFSET_OBJ)$dataSubType = 'offset'
-OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store design matrix
-MODEL1 %>%
-  rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'MAYO_CBE_TCX_Design.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('MAYO_CBE_TCX_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
-annotations(DM_OBJ) = annotations(CODE)
-annotations(DM_OBJ)$dataSubType = 'designMatrix'
-DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store surrogate variables
-SURR.VAR %>%
-  rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'MAYO_CBE_TCX_SurVar.tsv', sep = '\t', row.names=F, quote=F)
-SURR_OBJ = File('MAYO_CBE_TCX_SurVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
-annotations(SURR_OBJ) = annotations(CODE)
-annotations(SURR_OBJ)$dataSubType = 'surrogateVariables'
-SURR_OBJ = synStore(SURR_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store residual gene expression for eQTL analysis
-RESIDUAL.SVA.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MAYO_CBE_TCX_eQTLResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-eEXP_OBJ = File('MAYO_CBE_TCX_eQTLResidualExpression.tsv', 
-                name = 'Normalised, covariates and surrogate variable adjusted residual expression (for eQTL)',
-                parentId = CODE$properties$id)
-annotations(eEXP_OBJ) = annotations(CODE)
-annotations(eEXP_OBJ)$dataSubType = 'residualGeneExpForeQTL'
-eEXP_OBJ = synStore(eEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store residual gene expression for network analysis
-RESIDUAL.NET.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MAYO_CBE_TCX_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('MAYO_CBE_TCX_netResidualExpression.tsv', 
-                name = 'Normalised, covariates removed residual expression (for network analysis)', 
-                parentId = CODE$properties$id)
-annotations(nEXP_OBJ) = annotations(CODE)
-annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
-nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store differential expression results
-write.table(DE, file = 'MAYO_CBE_TCX_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('MAYO_CBE_TCX_DiffExpression.tsv', 
-                name = 'Differential Expression Results (BrainRegion.Diagnosis)', 
-                parentId = CODE$properties$id)
-annotations(DEXP_OBJ) = annotations(CODE)
-annotations(DEXP_OBJ)$dataSubType = 'diffExp'
-DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-stopCluster(cl)
-```
-|  *Results*                                  |  *SynapseID*                     |
-|  -----------------------------------------  |   ---------                      |
-|  Covariates                                 |  `r COV_OBJ$properties$id`       |
-|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
-|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
-|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
-|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
-|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
-|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
-|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
-|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

### 2017-04-06 code/Rmd/ROSMAP_REPROCESSED.Rmd

@@
 # Get residuals from differential expression model
-RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, VOOM.WEIGHTS$E)
+RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
 
@@
   tidyr::gather(data, value, -component)
-p = ggplot(var.comp %>% filter(component <= 10), aes(x = component, y = value, color = data)) + geom_point()
+p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.3)), aes(x = component, y = value, color = data)) + geom_point()
 p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')

### 2017-04-06 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Modified data and metadata source for v75 covariate analysis

@@
 # Download reprocessed counts (ACC)
-COUNT_ACC_ID = 'syn8414424';
+COUNT_ACC_ID = 'syn8414424'
 ALL_USED_IDs = COUNT_ACC_ID
@@
 
-# Get merged metadata
-METADATA_ID = 'syn2299154' #'syn7574833';#
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
-METADATA = read.csv(synGet(METADATA_ID)@filePath, fill = T, header = F)
-
 # Get ancestry vector calculated using gemtools
@@
 
+# Get RNA qc metrics from synapse
+METADATA_QC_ACC_ID = 'syn2929053'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_ACC_ID
+METADATA_QC_ACC = read.table(synGet(METADATA_QC_ACC_ID)@filePath, fill = T, header = T, sep = ',') %>%
+  lapply(as.character)
+
+# Get RNA qc metrics from synapse
+METADATA_QC_DLPFC_ID = 'syn2279446'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
+METADATA_QC_DLPFC = read.table(synGet(METADATA_QC_DLPFC_ID)@filePath, fill = T, header = T, sep = ',') %>%
+  lapply(as.character)
+
+METADATA_QC = rbindlist(list(METADATA_QC_ACC, METADATA_QC_DLPFC), use.names = T, fill = T) %>%
+  dplyr::filter(is.na(Exclude.)) %>%
+  dplyr::select(Sample.RNA.ID, Ribozero.Batch, Library.Batch, Flowcell.Batch, Mapped.Reads, 
+                Intragenic.Rate, Exonic.Rate, Intronic.Rate, Intergenic.Rate, Genes.Detected, 
+                Transcripts.Detected, Expression.Profiling.Efficiency, rRNA.Rate, Total.Reads, Percent.Aligned) %>%
+  dplyr::rename(SampleID = Sample.RNA.ID, RibozeroBatch = Ribozero.Batch, LibraryBatch = Library.Batch, 
+                FlowcellBatch = Flowcell.Batch, MappedReads = Mapped.Reads, IntragenicRate = Intragenic.Rate, 
+                ExonicRate = Exonic.Rate, IntronicRate = Intronic.Rate, IntergenicRate = Intergenic.Rate, 
+                GenesDetected = Genes.Detected, TranscriptsDetected = Transcripts.Detected, 
+                ExpProfEfficiency = Expression.Profiling.Efficiency, rRNARate = rRNA.Rate, 
+                TotalReads = Total.Reads, PercentAligned = Percent.Aligned)
+
+# Get merged metadata
+METADATA_ID = 'syn4484273'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
+METADATA = read.csv(synGet(METADATA_ID)@filePath, fill = T, header = T, check.names = F)
+
 ## Metadata with specific information for RNASeq processing
@@
                           "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
-                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?", "ACC_RNA_report: Library Batch",
-                          "ACC_RNA_report: Ribozero Batch", "ACC_RNA_report: Flowcell Batch", 
-                          "ACC_RNA_report: Mapped Reads", "ACC_RNA_report: Intragenic Rate",
-                          "ACC_RNA_report: Intronic Rate", "ACC_RNA_report: Intergenic Rate",
-                          "ACC_RNA_report: Genes Detected", "ACC_RNA_report: Transcripts Detected",
-                          "ACC_RNA_report: Expression Profiling Efficiency", "ACC_RNA_report: rRNA Rate",
-                          "ACC_RNA_report: Total Reads", "ACC_RNA_report: Percent Aligned",
+                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?",
 
                           "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
-                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
-                          "DLPFC_RNA_report: Ribozero Batch", "DLPFC_RNA_report: Flowcell Batch",
-                          "DLPFC_RNA_report: Mapped Reads", "DLPFC_RNA_report: Intragenic Rate",
-                          "DLPFC_RNA_report: Intronic Rate", "DLPFC_RNA_report: Intergenic Rate", 
-                          "DLPFC_RNA_report: Genes Detected", "DLPFC_RNA_report: Transcripts Detected", 
-                          "DLPFC_RNA_report: Expression Profiling Efficiency", "DLPFC_RNA_report: rRNA Rate",
-                          "DLPFC_RNA_report: Total Reads", "DLPFC_RNA_report: Percent Aligned",
+                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?",
 
@@
 # Filter metadata 
-ind = as.character(METADATA[1,] %>% unlist) %in% metadataVarOfInterest
+ind = colnames(METADATA) %in% metadataVarOfInterest
 METADATA = METADATA[,ind]
-colnames(METADATA) = gsub('[^[:alnum:]]','_',as.character(METADATA[1,] %>% unlist))
+colnames(METADATA) = gsub('[^[:alnum:]]','_',colnames(METADATA))
 
@@
   dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
-                ACC_RNA_isolation__Exclude_, ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN,
-                ACC_RNA_report__Exclude_, ACC_RNA_report__Ribozero_Batch, ACC_RNA_report__Library_Batch, 
-                ACC_RNA_report__Flowcell_Batch, ACC_RNA_report__Mapped_Reads, ACC_RNA_report__Intragenic_Rate,
-                ACC_RNA_report__Intronic_Rate, ACC_RNA_report__Intergenic_Rate, ACC_RNA_report__Genes_Detected,
-                ACC_RNA_report__Transcripts_Detected, ACC_RNA_report__Expression_Profiling_Efficiency, 
-                ACC_RNA_report__rRNA_Rate, ACC_RNA_report__Total_Reads, ACC_RNA_report__Percent_Aligned,
-                DNA_genotyping_report__Genotyping_Sample_ID) %>%
+                DNA_genotyping_report__Genotyping_Sample_ID, ACC_RNA_isolation__Exclude_, 
+                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_) %>%
   dplyr::rename(PMI = PMI__in_hours_, 
@@
                 ReportExclude = ACC_RNA_report__Exclude_,
-                RibozeroBatch = ACC_RNA_report__Ribozero_Batch,
-                LibraryBatch = ACC_RNA_report__Library_Batch,
-                FlowcellBatch = ACC_RNA_report__Flowcell_Batch, 
-                MappedReads = ACC_RNA_report__Mapped_Reads, 
-                IntragenicRate = ACC_RNA_report__Intragenic_Rate,
-                IntronicRate = ACC_RNA_report__Intronic_Rate, 
-                IntergenicRate = ACC_RNA_report__Intergenic_Rate, 
-                GenesDetected = ACC_RNA_report__Genes_Detected,
-                TranscriptsDetected = ACC_RNA_report__Transcripts_Detected, 
-                ExpProfEfficiency = ACC_RNA_report__Expression_Profiling_Efficiency, 
-                rRNARate = ACC_RNA_report__rRNA_Rate, 
-                TotalReads = ACC_RNA_report__Total_Reads, 
-                PercentAligned = ACC_RNA_report__Percent_Aligned,
                 GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
@@
   dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
-                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN,
-                DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Ribozero_Batch, DLPFC_RNA_report__Library_Batch, 
-                DLPFC_RNA_report__Flowcell_Batch, DLPFC_RNA_report__Mapped_Reads, DLPFC_RNA_report__Intragenic_Rate,
-                DLPFC_RNA_report__Intronic_Rate, DLPFC_RNA_report__Intergenic_Rate, DLPFC_RNA_report__Genes_Detected,
-                DLPFC_RNA_report__Transcripts_Detected, DLPFC_RNA_report__Expression_Profiling_Efficiency, 
-                DLPFC_RNA_report__rRNA_Rate, DLPFC_RNA_report__Total_Reads, DLPFC_RNA_report__Percent_Aligned,
-                DNA_genotyping_report__Genotyping_Sample_ID) %>%
+                DNA_genotyping_report__Genotyping_Sample_ID, DLPFC_RNA_isolation__Exclude_, 
+                DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_) %>%
   dplyr::rename(PMI = PMI__in_hours_, 
@@
                 ReportExclude = DLPFC_RNA_report__Exclude_,
-                RibozeroBatch = DLPFC_RNA_report__Ribozero_Batch,
-                LibraryBatch = DLPFC_RNA_report__Library_Batch,
-                FlowcellBatch = DLPFC_RNA_report__Flowcell_Batch, 
-                MappedReads = DLPFC_RNA_report__Mapped_Reads, 
-                IntragenicRate = DLPFC_RNA_report__Intragenic_Rate,
-                IntronicRate = DLPFC_RNA_report__Intronic_Rate, 
-                IntergenicRate = DLPFC_RNA_report__Intergenic_Rate, 
-                GenesDetected = DLPFC_RNA_report__Genes_Detected,
-                TranscriptsDetected = DLPFC_RNA_report__Transcripts_Detected, 
-                ExpProfEfficiency = DLPFC_RNA_report__Expression_Profiling_Efficiency, 
-                rRNARate = DLPFC_RNA_report__rRNA_Rate, 
-                TotalReads = DLPFC_RNA_report__Total_Reads, 
-                PercentAligned = DLPFC_RNA_report__Percent_Aligned,
                 GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
@@
   droplevels() %>%
+  list(.,METADATA_QC) %>%
+  join_all %>%
   dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
@@
 METADATA <- METADATA  %>% filter(!is.na(METADATA$EV.1)) 
-
-writeLines('Following samples are tagged outliers and removed')
-writeLines(paste(c('MSSM_RNA_ACC_BP_24'), collapse = ', '))
-METADATA <- METADATA  %>% filter(!(SampleID %in% c('MSSM_RNA_ACC_BP_24')))
 ```
@@
 COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
-COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
-COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
+COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], function(x){
+  x = as.numeric(as.character(gsub('[\\,\\%]','',x)))
+})
 
@@
 COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
-draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
+tmp = COVARIATES.CORRELATION$ESTIMATE
+tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
+h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
+draw(h, heatmap_legend_side = 'left')
 ```
@@
 ```
-Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples
+Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals
+
+### Outlier analysis
+Detect outlier samples based on their expression (logCPM) pattern
+```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
+# Find principal components of expression to plot
+PC <- prcomp(voom(PROCESSED_COUNTS$fi)$E, scale.=T, center = T)
+
+# Plot first 2 PCs
+plotdata <- data.frame(SampleID=rownames(PC$rotation), 
+                       PC1=PC$rotation[,1], 
+                       PC2=PC$rotation[,2])
+
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
+p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
+# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+p
+```
 

### 2017-04-06 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor modifications

@@
 # Query biomart
-Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content"),
+Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content", "gene_biotype"),
                        filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
@@
 Correlation/association between covariates at an FDR <= 0.1
-```{r covariates.correlation, fig.width=8, fig.height=8}
+```{r covariates.correlation, fig.width=10, fig.height=8}
 COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
@@
 ### Filter genes
-Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.
-
 Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
@@
 PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
+pct.pc = PROCESSED_COUNTS$filteredExprMatrix$genes %>%
+  dplyr::rename(ensembl_gene_id = genes) %>%
+  left_join(Ensemble2HGNC) %>%
+  group_by(gene_biotype) %>%
+  summarise(fraction  = n()) %>%
+  filter(fraction > 100) %>%
+  dplyr::mutate(fraction = fraction/length(PROCESSED_COUNTS$filteredExprMatrix$genes[,1]))
 ```
-Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals
+`r dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]` genes are used for the analysis.
+
+Following is the distribution of biotypes
+`r kable(pct.pc)`
 
@@
 Detect outlier samples based on their expression (logCPM) pattern
-```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
+```{r detect.outliers, fig.height=8, fig.width=8, results='asis'}
 # Find principal components of expression to plot
-PC <- prcomp(voom(PROCESSED_COUNTS$fi)$E, scale.=T, center = T)
+PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)
 
@@
 plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
+  dplyr::mutate(label = SampleID)
+plotdata$label[!(plotdata$SampleID %in% c('MSSM_RNA_ACC_BP_24', 
+                                          'MSSM_RNA_ACC_375'))] = ''
 
@@
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+p <- p + geom_text(aes(label= label), size=4, hjust=0)
 p
+
+# Plot abberent distribution of logcpm counts
+tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID') %>%
+              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))
+
+p = ggplot(tmp1 %>%
+             dplyr::filter(SampleID %in% c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375')),
+           aes(x = logCPM, color = SampleID)) + geom_density() 
+p = p + theme(legend.position = 'top') + facet_grid(Dx+.~Tissue, scale = 'free')
+p
+
+indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375')
+indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
+PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
 ```
+Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals
+
+Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`
 
@@
 # p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+p
+
+# Plot abberent distribution of logcpm counts
+tmp1 = (CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset) %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID') %>%
+              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))
+
+p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
+p = p + theme(legend.position = 'NONE') + facet_grid(Dx+.~Tissue, scale = 'free')
 p

### 2017-04-08 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor modifications

@@
 Detect outlier samples based on their expression (logCPM) pattern
-```{r detect.outliers, fig.height=8, fig.width=8, results='asis'}
+```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
+indToRemove = c('MSSM_RNA_ACC_BP_24',
+
+                'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250', 'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 
+                'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220', 'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133',
+
+                'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 
+                'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56')
+
+# Outliers from Menahce's DLPFC analysis syn3493755
+# outliers = c('MSSM_RNA_BP_PFC_1', 'MSSM_RNA_PFC_21', 'MSSM_RNA_BP_PFC_3', 'MSSM_RNA_PFC_103', 'MSSM_RNA_PFC_105', 'MSSM_RNA_PFC_115', 'MSSM_RNA_PFC_169',  'MSSM_RNA_PFC_17',
+#   'MSSM_RNA_PFC_18', 'MSSM_RNA_PFC_19', 'MSSM_RNA_PFC_213', 'MSSM_RNA_PFC_214', 'MSSM_RNA_PFC_215', 'MSSM_RNA_PFC_216', 'MSSM_RNA_PFC_217', 'MSSM_RNA_PFC_218', 
+#   'MSSM_RNA_PFC_219', 'MSSM_RNA_PFC_222', 'MSSM_RNA_PFC_24', 'MSSM_RNA_PFC_252', 'MSSM_RNA_PFC_259', 'MSSM_RNA_PFC_263', 'MSSM_RNA_PFC_267', 'MSSM_RNA_PFC_275',
+#   'MSSM_RNA_PFC_307', 'MSSM_RNA_PFC_322', 'MSSM_RNA_PFC_325', 'MSSM_RNA_PFC_27', 'MSSM_RNA_PFC_332', 'MSSM_RNA_PFC_333', 'MSSM_RNA_PFC_334', 'MSSM_RNA_PFC_335',
+#   'MSSM_RNA_PFC_340', 'MSSM_RNA_PFC_35', 'MSSM_RNA_PFC_362', 'MSSM_RNA_PFC_372', 'MSSM_RNA_PFC_373', 'MSSM_RNA_PFC_385', 'MSSM_RNA_PFC_393', 'MSSM_RNA_PFC_395',
+#   'MSSM_RNA_PFC_49', 'MSSM_RNA_PFC_54', 'MSSM_RNA_PFC_70', 'MSSM_RNA_PFC_76', 'MSSM_RNA_PFC_77', 'MSSM_RNA_PFC_78', 'MSSM_RNA_PFC_88', 'MSSM_RNA_PFC_93',
+#   'FAKE_DLPFC_CMC_MSSM_426', 'PENN_RNA_PFC_29', 'FAKE_DLPFC_CMC_PENN_055', 'PENN_RNA_PFC_75', 'PENN_RNA_PFC_110', 'FAKE_DLPFC_CMC_PENN_106', 'FAKE_DLPFC_CMC_PENN_107',
+#   'FAKE_DLPFC_CMC_PENN_108', 'MSSM_RNA_PFC_14', 'MSSM_RNA_BP_PFC_21', 'MSSM_RNA_PFC_293', 'MSSM_RNA_PFC_349', 'MSSM_RNA_PFC_353', 'MSSM_RNA_PFC_363', 'MSSM_RNA_PFC_369',
+#   'PENN_RNA_PFC_82', 'PENN_RNA_PFC_89', 'PITT_RNA_PFC_1159', 'PITT_RNA_PFC_1296', 'PITT_RNA_PFC_537',  'PITT_RNA_PFC_567', 'PITT_RNA_PFC_871', 'PITT_RNA_PFC_878',
+#   'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_210',  'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_278', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_361',
+#   'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_56', 'PITT_RNA_BP_PFC_10003', 'PITT_RNA_BP_PFC_1047', 'PITT_RNA_BP_PFC_1086', 'PITT_RNA_BP_PFC_1092', 'PITT_RNA_BP_PFC_1247',
+#   'PITT_RNA_BP_PFC_1282', 'PITT_RNA_BP_PFC_1324', 'PITT_RNA_BP_PFC_1374', 'PITT_RNA_BP_PFC_1391', 'PITT_RNA_BP_PFC_686')
+
 # Find principal components of expression to plot
@@
   dplyr::mutate(label = SampleID)
-plotdata$label[!(plotdata$SampleID %in% c('MSSM_RNA_ACC_BP_24', 
-                                          'MSSM_RNA_ACC_375'))] = ''
+plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''
 
@@
 p = ggplot(tmp1 %>%
-             dplyr::filter(SampleID %in% c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375')),
+             dplyr::filter(SampleID %in% indToRemove),
            aes(x = logCPM, color = SampleID)) + geom_density() 
@@
 
-indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375')
 indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
 PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
+COVARIATES = COVARIATES[indToRetain,]
+
+tmp = tidyr::separate(COVARIATES, Dx.Tissue, c('Dx', 'Tissue'), sep = '_') %>%
+  group_by(Dx, Tissue) %>%
+  summarise(count = n()) %>%
+  spread(Tissue, count)
 ```
@@
 Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`
+
+Distribution of samples are:
+`r kable(tmp)`
 
@@
 Clustering of voom adjusted data (with NULL design)
-```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
+```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
 # Find principal components of expression to plot
@@
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
 p

### 2017-04-10 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Minor modifications

--- 2017-03-19 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
@@
 ```
+
 ### Data download
@@
 Correlation/association between covariates at an FDR <= 0.1
-```{r covariates.correlation, fig.width=8, fig.height=8}
+```{r covariates.correlation, fig.width=10, fig.height=8}
 COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
-draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
+tmp = COVARIATES.CORRELATION$ESTIMATE
+tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
+h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
+draw(h, heatmap_legend_side = 'left')
 ```
@@
 PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
+pct.pc = PROCESSED_COUNTS$filteredExprMatrix$genes %>%
+  dplyr::rename(ensembl_gene_id = genes) %>%
+  left_join(Ensemble2HGNC) %>%
+  group_by(gene_biotype) %>%
+  summarise(fraction  = n()) %>%
+  filter(fraction > 100) %>%
+  dplyr::mutate(fraction = fraction/length(PROCESSED_COUNTS$filteredExprMatrix$genes[,1]))
 ```
-Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples
+`r dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]` genes are used for the analysis.
+
+Following is the distribution of biotypes
+`r kable(pct.pc)`
+
+### Outlier analysis
+Detect outlier samples based on their expression (logCPM) pattern
+```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
+indToRemove = c('MSSM_RNA_ACC_BP_24',
+
+                'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250', 'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 
+                'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220', 'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133',
+
+                'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 
+                'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56')
+
+# Outliers from Menahce's DLPFC analysis syn3493755
+# outliers = c('MSSM_RNA_BP_PFC_1', 'MSSM_RNA_PFC_21', 'MSSM_RNA_BP_PFC_3', 'MSSM_RNA_PFC_103', 'MSSM_RNA_PFC_105', 'MSSM_RNA_PFC_115', 'MSSM_RNA_PFC_169',  'MSSM_RNA_PFC_17',
+#   'MSSM_RNA_PFC_18', 'MSSM_RNA_PFC_19', 'MSSM_RNA_PFC_213', 'MSSM_RNA_PFC_214', 'MSSM_RNA_PFC_215', 'MSSM_RNA_PFC_216', 'MSSM_RNA_PFC_217', 'MSSM_RNA_PFC_218', 
+#   'MSSM_RNA_PFC_219', 'MSSM_RNA_PFC_222', 'MSSM_RNA_PFC_24', 'MSSM_RNA_PFC_252', 'MSSM_RNA_PFC_259', 'MSSM_RNA_PFC_263', 'MSSM_RNA_PFC_267', 'MSSM_RNA_PFC_275',
+#   'MSSM_RNA_PFC_307', 'MSSM_RNA_PFC_322', 'MSSM_RNA_PFC_325', 'MSSM_RNA_PFC_27', 'MSSM_RNA_PFC_332', 'MSSM_RNA_PFC_333', 'MSSM_RNA_PFC_334', 'MSSM_RNA_PFC_335',
+#   'MSSM_RNA_PFC_340', 'MSSM_RNA_PFC_35', 'MSSM_RNA_PFC_362', 'MSSM_RNA_PFC_372', 'MSSM_RNA_PFC_373', 'MSSM_RNA_PFC_385', 'MSSM_RNA_PFC_393', 'MSSM_RNA_PFC_395',
+#   'MSSM_RNA_PFC_49', 'MSSM_RNA_PFC_54', 'MSSM_RNA_PFC_70', 'MSSM_RNA_PFC_76', 'MSSM_RNA_PFC_77', 'MSSM_RNA_PFC_78', 'MSSM_RNA_PFC_88', 'MSSM_RNA_PFC_93',
+#   'FAKE_DLPFC_CMC_MSSM_426', 'PENN_RNA_PFC_29', 'FAKE_DLPFC_CMC_PENN_055', 'PENN_RNA_PFC_75', 'PENN_RNA_PFC_110', 'FAKE_DLPFC_CMC_PENN_106', 'FAKE_DLPFC_CMC_PENN_107',
+#   'FAKE_DLPFC_CMC_PENN_108', 'MSSM_RNA_PFC_14', 'MSSM_RNA_BP_PFC_21', 'MSSM_RNA_PFC_293', 'MSSM_RNA_PFC_349', 'MSSM_RNA_PFC_353', 'MSSM_RNA_PFC_363', 'MSSM_RNA_PFC_369',
+#   'PENN_RNA_PFC_82', 'PENN_RNA_PFC_89', 'PITT_RNA_PFC_1159', 'PITT_RNA_PFC_1296', 'PITT_RNA_PFC_537',  'PITT_RNA_PFC_567', 'PITT_RNA_PFC_871', 'PITT_RNA_PFC_878',
+#   'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_210',  'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_278', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_361',
+#   'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_56', 'PITT_RNA_BP_PFC_10003', 'PITT_RNA_BP_PFC_1047', 'PITT_RNA_BP_PFC_1086', 'PITT_RNA_BP_PFC_1092', 'PITT_RNA_BP_PFC_1247',
+#   'PITT_RNA_BP_PFC_1282', 'PITT_RNA_BP_PFC_1324', 'PITT_RNA_BP_PFC_1374', 'PITT_RNA_BP_PFC_1391', 'PITT_RNA_BP_PFC_686')
+
+# Find principal components of expression to plot
+PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)
+
+# Plot first 2 PCs
+plotdata <- data.frame(SampleID=rownames(PC$rotation), 
+                       PC1=PC$rotation[,1], 
+                       PC2=PC$rotation[,2])
+
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
+  dplyr::mutate(label = SampleID)
+plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
+p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
+p <- p + geom_text(aes(label= label), size=4, hjust=0)
+p
+
+# Plot abberent distribution of logcpm counts
+tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID') %>%
+              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))
+
+p = ggplot(tmp1 %>%
+             dplyr::filter(SampleID %in% indToRemove),
+           aes(x = logCPM, color = SampleID)) + geom_density() 
+p = p + theme(legend.position = 'NONE') + facet_grid(Dx+.~Tissue, scale = 'free')
+p
+
+indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
+PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
+COVARIATES = COVARIATES[indToRetain,]
+
+tmp = tidyr::separate(COVARIATES, Dx.Tissue, c('Dx', 'Tissue'), sep = '_') %>%
+  group_by(Dx, Tissue) %>%
+  summarise(count = n()) %>%
+  spread(Tissue, count)
+```
+Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals
+
+Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`
+
+Distribution of samples are:
+`r kable(tmp)`
 
@@
 ### Normalisation (custom design)
-Since many covariates are correlated, re-normalising COUNTS with covariates identified here in syn8468116. 
+Since many covariates are correlated, re-normalising COUNTS with a custom design matrix. 
 1. "Dx.Tissue" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue)
-2.
-Institution, Gender, IntronicRate, RibozeroBatch, PMI, RIN, TranscriptsDetected, Age\\_of\\_Death, TotalReads, IntergenicRate, PercentAligned, LibraryBatch were chosen as fixed effects and Individual_ID is chosen as random effect
-```{r iterative.normalisation, results='asis'}
+2.  Model:
+    Gx = Dx.Tissue + Institution + Gender + IntronicRate + RibozeroBatch + PMI + RIN + TranscriptsDetected +
+    Age\_of\_Death + IntergenicRate + MappedReads + PercentAligned + (1/Individual\_ID)    
+```{r custom.normalisation, results='asis'}
 # Primary variable of interest
-postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'IntronicRate', 'RibozeroBatch', 'PMI', 'RIN', 'TranscriptsDetected',
-                     'Age_of_Death', 'TotalReads', 'IntergenicRate', 'PercentAligned', 'LibraryBatch');
+postAdjustCovars = c("Dx.Tissue", "Institution", "Gender", "IntronicRate", "RibozeroBatch", "PMI", "RIN",
+                     "TranscriptsDetected", "Age_of_Death", "IntergenicRate", "MappedReads", "PercentAligned");
+
+writeLines(paste('Using following covariates in the model:',
+                 paste(postAdjustCovars, collapse=', '),
+                 'as fixed effects and Individual_ID is chosen as random effect'))
 
@@
 ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
+```
 
-# Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset)
-
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                              COVARIATES, 
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
-
-residualSigCovars[["PC_res"]][[2]]$plotData
-```
 ### SVA
@@
 # Get (null) design matrix
-MODEL0 = DM1$design[,-grep('Dx.Tissue',colnames(DM1$design))] # Get null model by removing variable of interest
+MODEL0 = DESIGN$design[,-grep('Dx.Tissue',colnames(DESIGN$design))] # Get null model by removing variable of interest
 MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
-MODEL1 = DM1$design
+MODEL1 = DESIGN$design
 MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
@@
 
-# Calculate the number of surrogate variables
-NUM.SV = num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
-                MODEL1, method = 'be', seed = 123456, B = 30, tol = 1e-22)
+# Get residuals from differential expression model
+RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, VOOM.GENE_EXPRESSION$E)
+
+# Compute actual variance of all principal components
+tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
+actual.var = tmp$d^2/sum(tmp$d^2)
+
+# Compute permuted variance of all principal components
+permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
+  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
+  tmp = svd(tmp.residual)
+  permuted.var = tmp$d^2/sum(tmp$d^2)
+}
+permuted.var = apply(permuted.var, 2, mean)
+NUM.SV = sum(actual.var >= permuted.var[1])
+
+# Plot variance components
+var.comp = data.frame(component = 1:length(actual.var), 
+                      residual = actual.var, 
+                      permuted.residual = permuted.var) %>%
+  tidyr::gather(data, value, -component)
+p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.2)), aes(x = component, y = value, color = data)) + geom_point()
+p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
+p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
+p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
+p
 
@@
 SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
-                        MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-22)$sv
+                        MODEL1, MODEL0, n.sv = NUM.SV, B = 3, tol = 1e-22)$sv
 
@@
                         design = cbind(MODEL1, SURR.VAR), plot=F,
-                        block = COVARIATES$individualIdentifier, 
+                        block = COVARIATES$Donor_ID, 
                         correlation = correlation$cor)
@@
 # Residuals after normalisation
-RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                               CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
@@
 # Add variable of interest back to the residuals
-varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
-RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
+varsToAddIn = grep("Dx.Tissue", colnames(VOOM.SVA.FIT$coefficients), value = T)
+RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
   VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
@@
 ```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.SVA.GENE_EXPRESSION))
+cr = cor(t(RESIDUAL.GENE_EXPRESSION))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.SVA.GENE_EXPRESSION, scale.=T, center = T)
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(RESIDUAL.SVA.GENE_EXPRESSION)))
+tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 ```{r diffExp, fig.height=10, fig.width=15}
+# Post adjusted design matrix
+DESIGN = getDesignMatrix(cbind(COVARIATES[,postAdjustCovars,drop=F], SURR.VAR),
+                               Intercept = F)
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
+# Estimate voom weights
+VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+correlation <- parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Individual_ID)
+
+# Re-calculate voom weights with correlation of random effects
+VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+                    design=DESIGN$design, plot=F,
+                    block = COVARIATES$Individual_ID, 
+                    correlation = correlation$cor)
+
+# Fit linear model using new weights and new design
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+FIT = lmFit(VOOM.WEIGHTS)
+
 # Fit contrast
@@
                                      "Dx.TissueOther_DLPFC-Dx.TissueSCZ_DLPFC"),
-                         levels = colnames(VOOM.SVA.FIT$coefficients))
-FIT.CONTR = contrasts.fit(VOOM.SVA.FIT, contrasts=contrast)
+                         levels = colnames(FIT$coefficients))
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
 FIT.CONTR = eBayes(FIT.CONTR)
@@
 
-  normalizationType	= 'CQN'
+  normalizationType	= 'CQN, SVA'
 )
@@
 
-# Store surrogate variables
-SURR.VAR %>%
-  rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'CMC_ACC_DLPFC_SurVar.tsv', sep = '\t', row.names=F, quote=F)
-SURR_OBJ = File('CMC_ACC_DLPFC_SurVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
-annotations(SURR_OBJ) = annotations(CODE)
-annotations(SURR_OBJ)$dataSubType = 'surrogateVariables'
-SURR_OBJ = synStore(SURR_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store residual gene expression for eQTL analysis
-RESIDUAL.SVA.GENE_EXPRESSION %>%
+# Store residual gene expression for eqtl analysis
+RESIDUAL.GENE_EXPRESSION %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_eQTLResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-eEXP_OBJ = File('CMC_ACC_DLPFC_eQTLResidualExpression.tsv', 
-                name = 'Normalised, covariates and surrogate variable adjusted residual expression (for eQTL)',
+  write.table(file = 'CMC_ACC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+nEXP_OBJ = File('CMC_ACC_DLPFC_netResidualExpression.tsv', 
+                name = 'Normalised, covariates removed residual expression (for eQTL analysis)', 
                 parentId = CODE$properties$id)
-annotations(eEXP_OBJ) = annotations(CODE)
-annotations(eEXP_OBJ)$dataSubType = 'residualGeneExpForeQTL'
-eEXP_OBJ = synStore(eEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+annotations(nEXP_OBJ) = annotations(CODE)
+annotations(nEXP_OBJ)$dataSubType = 'residualeQTLAnalysis'
+nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 
@@
 |  Design Matrix                              |  `r DM_OBJ$properties$id`        |
-|  Residual Expression (for eQTL analysis)    |  `r eEXP_OBJ$properties$id`      |
+|  Residual Expression (for eQTL analysis)    |  `r nEXP_OBJ$properties$id`      |
 |  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

### 2017-04-10 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Minor bug fix

@@
 # Get (null) design matrix
-MODEL0 = DESIGN$design[,-grep('Dx.Tissue',colnames(DESIGN$design))] # Get null model by removing variable of interest
+MODEL0 = DM1$design[,-grep('Dx.Tissue',colnames(DM1$design))] # Get null model by removing variable of interest
 MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
-MODEL1 = DESIGN$design
+MODEL1 = DM1$design
 MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
@@
 SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
-                        MODEL1, MODEL0, n.sv = NUM.SV, B = 3, tol = 1e-22)$sv
+                        MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-22)$sv
 
@@
 
+# Store surrogate variables
+SURR.VAR %>%
+  rownameToFirstColumn('SampleID') %>%
+  write.table(file = 'CMC_ACC_DLPFC_SurrVar.tsv', sep = '\t', row.names=F, quote=F)
+SURR.VAR_OBJ = File('CMC_ACC_DLPFC_SurrVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
+annotations(SURR.VAR_OBJ) = annotations(CODE)
+annotations(SURR.VAR_OBJ)$dataSubType = 'SurrogateVariables'
+SURR.VAR_OBJ = synStore(SURR.VAR_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
 # Store residual gene expression for eqtl analysis
@@
 |  Design Matrix                              |  `r DM_OBJ$properties$id`        |
+|  Surrogate Variables                        |  `r SURR.VAR_OBJ$properties$id`  |
 |  Residual Expression (for eQTL analysis)    |  `r nEXP_OBJ$properties$id`      |

### 2017-04-14 code/Rmd/ROSMAP_REPROCESSED.Rmd
Modifications in ROSMAP to include second round of reprocessed samples

@@
 # Download expression data
-COUNT_ID <- 'syn7750270';
+COUNT_ID <- 'syn8691134';#'syn7750270';
 ALL_USED_IDs <- COUNT_ID
 COUNT_OBJ <- synGet(COUNT_ID)
-COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F)
+COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F, row.names = 1)
 COUNT[,grep('150_120419', colnames(COUNT))[2]] = NULL
@@
 # Get picard metrics from synapse
-METADATA.PICARD_ID <- 'syn8512191'
+METADATA.PICARD_ID <- 'syn8698240';#'syn8512191'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
-METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
-METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
-METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
-METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)
+# METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
+# METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
+# METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
+# METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)
+# METADATA.PICARD$Sampleid = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
+#   gsub('Aligned.out/picard.CombinedMetrics.csv','',.)
 
-METADATA.PICARD$Sampleid = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
-  gsub('Aligned.out/picard.CombinedMetrics.csv','',.)
+METADATA.PICARD <- CovariateAnalysis::downloadFile(METADATA.PICARD_ID) %>%
+  dplyr::rename(Sampleid = sample)
 colnames(METADATA.PICARD) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD))
@@
 
-GENE.LEN = dplyr::select(GENE.PARAM, ensembl_gene_id, gene.length) %>% unique()
-rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id
+GENE.LEN = dplyr::select(GENE.PARAM, ensembl_gene_id, gene.length) %>% 
+  unique() %>%
+  left_join(data.frame(Gene.ID = rownames(COUNT)) %>%
+              dplyr::mutate(tmp = Gene.ID) %>%
+              tidyr::separate(tmp, c('ensembl_gene_id', 'position'), sep = '\\.'))
+rownames(GENE.LEN) = GENE.LEN$Gene.ID
 
-GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% unique()
-rownames(GENE.GC) = GENE.GC$ensembl_gene_id
+GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% 
+  unique() %>%
+  left_join(data.frame(Gene.ID = rownames(COUNT)) %>%
+              dplyr::mutate(tmp = Gene.ID) %>%
+              tidyr::separate(tmp, c('ensembl_gene_id', 'position'), sep = '\\.'))
+rownames(GENE.GC) = GENE.GC$Gene.ID
 ```
@@
   filter(!is.na(pmi)) %>%
-  filter(!(Sampleid %in% c('380_120503'))) %>%
+  filter(!is.na(PCT_INTRONIC_BASES)) %>%
   as.data.frame()
 
-METADATA$Diagnosis = 'NONE'
-METADATA$Diagnosis[METADATA$cogdx == 1 & METADATA$braaksc <= 3 & METADATA$ceradsc > 2] = 'Control'
+METADATA$Diagnosis = 'OTHER'
+METADATA$Diagnosis[METADATA$cogdx == 1 & METADATA$braaksc <= 3 & METADATA$ceradsc >= 3] = 'CONTROL'
 METADATA$Diagnosis[METADATA$cogdx == 4 & METADATA$braaksc >= 4 & METADATA$ceradsc <= 2] = 'AD'
@@
 #"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv", "PF_INDEL_RATE", "PF_MISMATCH_RATE", "age_first_ad_dx","age_at_visit_max",
-primaryVariable <- "cogdx"
-FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx")
+primaryVariable <- c("cogdx", "Diagnosis")
+FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx", "Diagnosis")
 ContCovariates <- c("RINcontinuous", "age_death", "pmi", "educ", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
@@
 ```
-
+### Explore metadata
 ```{r data.explore, fig.width = 10, fig.height = 14}
@@
 p = list()
-p[[1]] = ggplot(COVARIATES, aes(x = cogdx, y = RINcontinuous)) + geom_boxplot()
+p[[1]] = ggplot(COVARIATES, aes(x = Diagnosis, y = RINcontinuous)) + geom_boxplot()
 p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')
@@
 # AgeAtDeath
-p[[2]] = ggplot(COVARIATES, aes(x = cogdx, y = age_death)) + geom_boxplot()
+p[[2]] = ggplot(COVARIATES, aes(x = Diagnosis, y = age_death)) + geom_boxplot()
 p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + theme(legend.position = 'top')
@@
 # PMI
-p[[3]] = ggplot(COVARIATES, aes(x = cogdx, y = pmi)) + geom_boxplot()
+p[[3]] = ggplot(COVARIATES, aes(x = Diagnosis, y = pmi)) + geom_boxplot()
 p[[3]] = p[[3]] + ggtitle('PMI') + theme(legend.position = 'top')
@@
 # Education
-p[[4]] = ggplot(COVARIATES, aes(x = cogdx, y = educ)) + geom_boxplot()
+p[[4]] = ggplot(COVARIATES, aes(x = Diagnosis, y = educ)) + geom_boxplot()
 p[[4]] = p[[4]] + ggtitle('Education') + theme(legend.position = 'top')
@@
 # Intronic bases
-p[[5]] = ggplot(COVARIATES, aes(x = cogdx, y = PCT_INTRONIC_BASES)) + geom_boxplot()
+p[[5]] = ggplot(COVARIATES, aes(x = Diagnosis, y = PCT_INTRONIC_BASES)) + geom_boxplot()
 p[[5]] = p[[5]] + ggtitle('Fraction Intronic Bases') + theme(legend.position = 'top')
@@
 # Ribosomal bases
-p[[6]] = ggplot(COVARIATES, aes(x = cogdx, y = PCT_RIBOSOMAL_BASES)) + geom_boxplot()
+p[[6]] = ggplot(COVARIATES, aes(x = Diagnosis, y = PCT_RIBOSOMAL_BASES)) + geom_boxplot()
 p[[6]] = p[[6]] + ggtitle('Fraction Ribosomal Bases') + theme(legend.position = 'top')
@@
 # Batch
-vcd::mosaic(~ Batch + cogdx, data = COVARIATES)
+# vcd::mosaic(~ Batch + Diagnosis, data = COVARIATES)
 
 # Sex
-vcd::mosaic(~ msex + cogdx, data = COVARIATES)
+# vcd::mosaic(~ msex + Diagnosis, data = COVARIATES)
 ```
-### CPM Normalisation
-Preprocess counts matrix using CQN (Conditional Quantile Normlaisation accounting for both GC content and gene length)
-
+### Filter genes
 Remove genes that have less than 1 cpm counts in at least 50% of samples per cogdx. Also remove genes with missing gene length and percentage GC content
@@
   rownameToFirstColumn('Sampleid') %>%
-  dlply(.(cogdx), .fun = function(mtd, count){
+  dlply(.(Diagnosis), .fun = function(mtd, count){
   processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$Sampleid],
@@
   unique() %>%
-  intersect(GENE.GC$ensembl_gene_id[!is.na(GENE.GC$percentage_gc_content)]) %>%
-  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$gene.length)]) %>%
+  intersect(GENE.GC$Gene.ID[!is.na(GENE.GC$percentage_gc_content)]) %>%
+  intersect(GENE.LEN$Gene.ID[!is.na(GENE.LEN$gene.length)]) %>%
   setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous"))
@@
 
-### Normalisation (with NULL)
+### Outlier Analysis
+```{r outlier.analysis, cache = FALSE}
+indToRemove = c('380_120503', '500_120515')
+
+# Find principal components of expression to plot
+PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)
+
+# Plot first 2 PCs
+plotdata <- data.frame(SampleID=rownames(PC$rotation), 
+                       PC1=PC$rotation[,1], 
+                       PC2=PC$rotation[,2])
+
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
+  dplyr::mutate(label = SampleID)
+plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(color=Batch, shape=Diagnosis, size=RINcontinuous))
+p <- p + theme_bw() + theme(legend.position="right")
+p <- p + geom_text(aes(label= label), size=4, hjust=0)
+p
+
+# Plot abberent distribution of logcpm counts
+tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
+  rownameToFirstColumn('Gene.ID') %>%
+  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID'))
+
+p = ggplot(tmp1 %>%
+             dplyr::filter(SampleID %in% indToRemove),
+           aes(x = logCPM, color = SampleID)) + geom_density() 
+p = p + theme(legend.position = 'top')
+p
+
+indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
+PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
+COVARIATES = COVARIATES[indToRetain,]
+
+tmp = COVARIATES %>%
+  group_by(Diagnosis, cogdx) %>%
+  summarise(count = n()) %>%
+  spread(Diagnosis, count)
+```
+Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples
+
+Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`
+
+Distribution of samples are: `r kable(tmp)`
+
+### Conditional Quantile Normalisation (CQN)
 Initial normalisation usign cqn and dispersion control with voom (with NULL design)
@@
 ```
+
+```{r explore.data, cache=FALSE, fig.height=5, fig.width=10}
+# Plot abberent distribution of logcpm counts
+tmp1 = (CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset) %>%
+  rownameToFirstColumn('Gene.ID') %>%
+  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID'))
+
+p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
+p = p + theme(legend.position = 'NONE') + facet_grid(.~Diagnosis, scale = 'free')
+p
+```
+
 Coexpression of genes 
@@
 ```
-### Outlier analysis
-Clustering of voom adjusted data (with NULL design)
+### Decomposition of expression
+Clustering of library normalised unadjusted data (with NULL design)
 ```{r decompse.normalise.data1.1, fig.height=6, fig.width=8, results='asis'}
@@
 p <- p + theme_bw() + theme(legend.position="right")
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
 p
@@
 1. Adding FLOWCELL and Gender a priori to variable selection
-2. Primary variable of interest BrainRegion.Diagnosis is excluded from the pool of available covariates for selection
+2. Primary variable of interest Diagnosis is excluded from the pool of available covariates for selection
 ```{r iterative.norm, results='asis'}
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2)) 
-p <- p + geom_point(aes(color=Batch, shape=cogdx, size=RINcontinuous))
+p <- p + geom_point(aes(color=Batch, shape=Diagnosis, size=RINcontinuous))
 p <- p + theme_bw() + theme(legend.position="right")
@@
 # Get design matrix
-DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable, postAdjustCovars), drop = F], Intercept = F)
+DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[1], postAdjustCovars), drop = F], Intercept = F)
 DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
@@
 p
+
+all.diff.exp = list(cogdx = DE)
+```
+
+### Differential expression analysis (with Diagnosis as primary variable)
+Differential expression is performed on the primary variable by controlling for covariates identified above
+
+METADATA$Diagnosis = 'OTHER'
+METADATA$Diagnosis[METADATA$cogdx == 1 & METADATA$braaksc <= 3 & METADATA$ceradsc >= 3] = 'CONTROL'
+METADATA$Diagnosis[METADATA$cogdx == 4 & METADATA$braaksc >= 4 & METADATA$ceradsc <= 2] = 'AD'
+
+Interpretation of Diagnosis
+1. AD: cogdx = 4, braaksc >= 4 and ceradsc <= 2
+2. CONTROL: cogdx = 1, braaksc <= 3 and ceradsc >= 3
+3. OTHER: All the other samples
+
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diff.exp1, fig.height=6, fig.width=18, cache=FALSE}
+# Get design matrix
+DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[2], postAdjustCovars), drop = F], Intercept = F)
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
+
+# Fit linear model using new weights and new design
+FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
+            design = DESIGN$design,
+            weights = VOOM.WEIGHTS$weights)
+
+# Fit contrast
+contrast = makeContrasts(contrasts=c("DiagnosisAD-DiagnosisCONTROL",
+                                     "DiagnosisOTHER-DiagnosisAD",
+                                     "DiagnosisOTHER-DiagnosisCONTROL"),
+                         levels = colnames(FIT$coefficients))
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = lapply(1:3, function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = T) %>%
+    rownameToFirstColumn('ensembl_gene_id')
+}, FIT.CONTR)
+names(DE) = gsub('Diagnosis','',colnames(contrast))
+
+DE = DE %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  left_join(GENE.PARAM %>%
+              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
+              unique()) %>%
+  dplyr::mutate(Study = 'ROSMAP', Region = 'DLPFC',
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction))
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
+
+writeLines('Differentially expressed genes at an FDR 0.05 and logFC 1.2')
+tmp = DE %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
+  group_by(Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05_FC_1.2) 
+kable(tmp)
+
+p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red'))
+p = p + facet_grid(.~Comparison, scales = 'fixed')
+p
+
+all.diff.exp = c(all.diff.exp, list(Diagnosis = DE))
 ```
@@
 ### SVA Adjustments for eQTL analysis
-Conditioned on primary variable (cogdx) and identified covariates estimate surrogate variables using SVA package
+Conditioned on primary variable (Diagnosis) and identified covariates estimate surrogate variables using SVA package
 ```{r sva.adjust}
 # Get (null) design matrix
-MODEL0 = DESIGN$design[,-grep(primaryVariable,colnames(DESIGN$design))] # Get null model by removing variable of interest
+MODEL0 = DESIGN$design[,-grep(primaryVariable[2],colnames(DESIGN$design))] # Get null model by removing variable of interest
 MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
@@
 SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
-                    MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv
+                    MODEL1, MODEL0, n.sv = NUM.SV, B = 3)$sv
 SURR.VAR = data.frame(SURR.VAR)
@@
 
-### Covariates Adjustments for network analysis
+### Covariates Adjustments for Network Analysis
 Identified covariates are regressed out from the expression matrix for network analysis

### 2017-04-14 code/Rmd/ROSMAP_REPROCESSED.Rmd
Minor sanity changes

@@
 SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
-                    MODEL1, MODEL0, n.sv = NUM.SV, B = 3)$sv
+                    MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv
 SURR.VAR = data.frame(SURR.VAR)
@@
   summaryLevel = 'gene',
-  assay	 = 'Reprocessed RNAseq',
+  assay	 = 'RNAseq',
+  assaySubType = 'Reprocessed',
 
@@
 # Store differential expression results
-write.table(DE, file = 'ROSMAP_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+rbindlist(all.diff.exp, use.names = T, fill = T) %>%
+  write.table(file = 'ROSMAP_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
 DEXP_OBJ = File('ROSMAP_DLPFC_DiffExpression.tsv', name = 'Differential Expression Results (cogdx)', 

### 2017-04-14 code/Rmd/ROSMAP_REPROCESSED.Rmd
Minor modifications

--- 2017-04-14 code/Rmd/ROSMAP_REPROCESSED.Rmd
@@
 knit2synapse::knitToFolderEntity(file = "./ROSMAP_REPROCESSED.Rmd",
-                                 parentId = "syn8403743",
+                                 parentId = "syn8672910",
                                  entityName = 'ROSMAP Reprocessed RNASeq (Normalised)')
@@
 # Synapse parameters
-parentId = 'syn8403743';
+parentId = 'syn8672910';
 activityName = 'Covariate analysis';
@@
 Correlation/association between covariates at an FDR <= 0.1
-```{r covariates.correlation, fig.width=6, fig.height=6}
+```{r covariates.correlation, fig.width=9, fig.height=6}
 COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
@@
 ### Filter genes
-Remove genes that have less than 1 cpm counts in at least 50% of samples per cogdx. Also remove genes with missing gene length and percentage GC content
+Remove genes that have less than 1 cpm counts in at least 50% of samples per Diagnosis. Also remove genes with missing gene length and percentage GC content
 ```{r cpmnormalisation}
@@
 ### Outlier Analysis
-```{r outlier.analysis, cache = FALSE}
+```{r outlier.analysis, cache = FALSE, fig.width = 10}
 indToRemove = c('380_120503', '500_120515')
@@
 
-METADATA$Diagnosis = 'OTHER'
-METADATA$Diagnosis[METADATA$cogdx == 1 & METADATA$braaksc <= 3 & METADATA$ceradsc >= 3] = 'CONTROL'
-METADATA$Diagnosis[METADATA$cogdx == 4 & METADATA$braaksc >= 4 & METADATA$ceradsc <= 2] = 'AD'
-
 Interpretation of Diagnosis

### 2017-04-14 code/Rmd/ROSMAP_REPROCESSED.Rmd
Minor bug fix

@@
 GENE.LEN = dplyr::select(GENE.PARAM, ensembl_gene_id, gene.length) %>% 
-  unique() %>%
-  left_join(data.frame(Gene.ID = rownames(COUNT)) %>%
-              dplyr::mutate(tmp = Gene.ID) %>%
-              tidyr::separate(tmp, c('ensembl_gene_id', 'position'), sep = '\\.'))
-rownames(GENE.LEN) = GENE.LEN$Gene.ID
+  unique() 
+rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id
 
 GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% 
-  unique() %>%
-  left_join(data.frame(Gene.ID = rownames(COUNT)) %>%
-              dplyr::mutate(tmp = Gene.ID) %>%
-              tidyr::separate(tmp, c('ensembl_gene_id', 'position'), sep = '\\.'))
-rownames(GENE.GC) = GENE.GC$Gene.ID
+  unique() 
+rownames(GENE.GC) = GENE.GC$ensembl_gene_id 
 ```
@@
   unique() %>%
-  intersect(GENE.GC$Gene.ID[!is.na(GENE.GC$percentage_gc_content)]) %>%
-  intersect(GENE.LEN$Gene.ID[!is.na(GENE.LEN$gene.length)]) %>%
+  intersect(GENE.GC$ensembl_gene_id[!is.na(GENE.GC$percentage_gc_content)]) %>%
+  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$gene.length)]) %>%
   setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous"))
@@
 CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
-                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
+                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix$counts), 'percentage_gc_content'],
+                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix$counts), 'gene.length'],
                           lengthMethod = "smooth", 

### 2017-04-15 code/Rmd/MAYO_REPROCESSED.Rmd
Modified Mayo reprocessing to include new samples for April freeze

@@
 knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED.Rmd",
-                                 parentId = "syn7554308",
+                                 parentId = "syn8672415",
                                  entityName = 'MAYO Reprocessed RNASeq (Normalised)')
@@
 # Synapse parameters
-parentId = 'syn7554308'
+parentId = 'syn8672415'
 activityName = 'Covariate analysis';
@@
 # Get reprocessed sample counts for temporal cortex
-COUNT_TC_ID <- 'syn7987765';
+COUNT_TC_ID <- 'syn8690799';#'syn7987765';
 ALL_USED_IDs <- COUNT_TC_ID
-COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F)
-colnames(COUNT_TC) = paste(colnames(COUNT_TC), 'TCX', sep = '_')
+COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F, row.names = 1)
 
 # Get reprocessed sample counts for cerebellum
-COUNT_CE_ID <- 'syn7513807';
+COUNT_CE_ID <- 'syn8690904';#'syn7513807';
 ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_CE_ID
-COUNT_CE <- read.table(synGet(COUNT_CE_ID)@filePath, header=T, sep='\t', check.names = F)
+COUNT_CE <- read.table(synGet(COUNT_CE_ID)@filePath, header=T, sep='\t', check.names = F, row.names = 1)
 
@@
 # Get picard metrics from synapse
-METADATA.PICARD_CE_ID <- 'syn8380392'
+METADATA.PICARD_CE_ID <- 'syn8698214';#'syn8380392'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_CE_ID
-METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = FALSE, exdir = getwd())
-METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = TRUE, exdir = getwd())
-METADATA.PICARD_CE_FPATHS = METADATA.PICARD_CE_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_CE_FPATHS)]
-METADATA.PICARD_CE <- plyr::ldply(METADATA.PICARD_CE_FPATHS, read.csv)
-
-METADATA.PICARD_CE$ID = gsub('picard/', '', METADATA.PICARD_CE_FPATHS) %>%
-  gsub('/picard.CombinedMetrics.csv','',.) %>%
-  sapply(function(x){str_split(x, '\\.')[[1]][1]})
+# METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = FALSE, exdir = getwd())
+# METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = TRUE, exdir = getwd())
+# METADATA.PICARD_CE_FPATHS = METADATA.PICARD_CE_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_CE_FPATHS)]
+# METADATA.PICARD_CE <- plyr::ldply(METADATA.PICARD_CE_FPATHS, read.csv)
+# 
+# METADATA.PICARD_CE$ID = gsub('picard/', '', METADATA.PICARD_CE_FPATHS) %>%
+#   gsub('/picard.CombinedMetrics.csv','',.) %>%
+#   sapply(function(x){str_split(x, '\\.')[[1]][1]})
+METADATA.PICARD_CE <- CovariateAnalysis::downloadFile(METADATA.PICARD_CE_ID) %>%
+  dplyr::rename(ID = sample)
 colnames(METADATA.PICARD_CE) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_CE))
@@
 
-METADATA.PICARD_TC_ID <- 'syn8383233'
+METADATA.PICARD_TC_ID <- 'syn8698211';#'syn8383233'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_TC_ID
-METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = FALSE, exdir = getwd())
-METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = TRUE, exdir = getwd())
-METADATA.PICARD_TC_FPATHS = METADATA.PICARD_TC_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_TC_FPATHS)]
-METADATA.PICARD_TC <- plyr::ldply(METADATA.PICARD_TC_FPATHS, read.csv)
+# METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = FALSE, exdir = getwd())
+# METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = TRUE, exdir = getwd())
+# METADATA.PICARD_TC_FPATHS = METADATA.PICARD_TC_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_TC_FPATHS)]
+# METADATA.PICARD_TC <- plyr::ldply(METADATA.PICARD_TC_FPATHS, read.csv)
+# METADATA.PICARD_TC$ID = gsub('picard/', '', METADATA.PICARD_TC_FPATHS) %>%
+#   gsub('/picard.CombinedMetrics.csv','',.) %>%
+#   sapply(function(x){str_split(x, '\\.')[[1]][1]}) %>%
+#   gsub('Aligned','',.)
 
-METADATA.PICARD_TC$ID = gsub('picard/', '', METADATA.PICARD_TC_FPATHS) %>%
-  gsub('/picard.CombinedMetrics.csv','',.) %>%
-  sapply(function(x){str_split(x, '\\.')[[1]][1]}) %>%
-  gsub('Aligned','',.)
+METADATA.PICARD_TC <- CovariateAnalysis::downloadFile(METADATA.PICARD_TC_ID) %>%
+  dplyr::rename(ID = sample)
 colnames(METADATA.PICARD_TC) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_TC))
@@
 # Fix metadata
-METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
-METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
-METADATA$BrainRegion.Diagnosis = paste(METADATA$BrainRegion, METADATA$Diagnosis, sep = '.')
-METADATA$RIN2 = METADATA$RIN^2
-METADATA <- METADATA %>%
+METADATA = METADATA %>%
+  dplyr::mutate(AgeAtDeath = gsub("_or_above", "", AgeAtDeath),
+                Diagnosis = factor(Diagnosis, 
+                                   levels = c("AD", "Control", "Pathologic Aging", "PSP"), 
+                                   labels = c("AD" = "AD", "Control" = "CONTROL", "Pathologic Aging" = "OTHER", "PSP" = "OTHER")),
+                BrainRegion.Diagnosis = paste(BrainRegion, Diagnosis, sep = '.'),
+                RIN2 = RIN^2) %>%
   filter(!is.na(Source)) %>%
-  dplyr::filter(!(ID %in% c("11311_CER", "1923_CER", "1923_TCX",
-                            "11396_TCX", "11294_TCX", "11408_TCX",
-                            SAMPLES.EXCLUDE)))
+  dplyr::filter(!(ID %in% SAMPLES.EXCLUDE))
 ```
-Following samples were considered outliers and removed based on PCA and/or eucledian distance: 11311\_CER, 1923\_CER, 1923\_TCX, 11396\_TCX, 11294\_TCX, 11408\_TCX
+
 ```{r preprocess.data1}
@@
 ```{r predict.pmi}
-PMI = METADATA[,c('PMI', 'ID'), drop = F]
-PMI$PMI = PMI$PMI * 60
-rownames(PMI) = METADATA$ID
+# Get PMI
+PMI = METADATA[,c('PMI', 'ID', 'Donor_ID'), drop = F]
+PMI$PMI = PMI$PMI
+
+# Get expression (convert counts to cpm)
+expr = voom(COUNT, design = NULL)$E
+ind = which(rowSums(expr>=1)/dim(expr)[2] >= 0.5)
+expr = expr[ind,]
+
+# Combine expression from different regions of brain of an individual into one
+expr = plyr::dlply(PMI, .(Donor_ID), .fun = function(x, expr){
+  data.frame(value = rowMeans(expr[,x$ID,drop = F], na.rm = T)) %>%
+    plyr::rename(c('value' = unique(x$Donor_ID)))
+}, expr, .parallel = F) %>%
+  do.call(cbind,.)
+
+# Split data into trainining and prediction set
+PMI = dplyr::select(PMI, Donor_ID, PMI) %>% unique()
+rownames(PMI) = PMI$Donor_ID
+PMI = PMI[colnames(expr),]
+
+# Fit a logistic regression model
+x.train = expr[,!is.na(PMI$PMI)]
+x.predict = expr[,is.na(PMI$PMI)]
+
+y.train = PMI[!is.na(PMI$PMI),]
+y.train = y.train[colnames(x.train), 'PMI']
+
+plr.model = cv.glmnet(t(x.train), y.train, family = 'gaussian')
+coeff = as.matrix(coef.glmnet(plr.model, s = 'lambda.min'))
+ssres = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train)^2)
+sstot = sum((y.train-mean(y.train, na.rm = T))^2, na.rm = T)
+R2.allgenes = 1-(ssres/sstot)
+
+y.predict.all = predict(plr.model, t(x.predict), s = 'lambda.min', type = 'response') %>%
+  rownameToFirstColumn('Donor_ID') %>%
+  plyr::rename(c('1' = 'PMI'))
 
@@
 
-# Conditional quantile normalisation
-tmp.count = COUNT
-genesToAnalyze = unique(intersect(rownames(GENE.GC)[!is.na(GENE.GC$percentage_gc_content)], 
-                                  rownames(GENE.LEN)[!is.na(GENE.LEN$gene.length)]))
-tmp.count = tmp.count[rownames(tmp.count) %in% genesToAnalyze,]
-CQN.GENE_EXPRESSION = cqn(tmp.count,
-                          x = GENE.GC[rownames(tmp.count), 'percentage_gc_content'],
-                          lengths = GENE.LEN[rownames(tmp.count), 'gene.length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
+# Initial normalisation of gene expression
+ind = rownames(expr) %in% CONTROL.PROBES$ensembl_gene_id 
+expr = expr[ind,]
 
-# Initial normalisation of gene expression
-ind = rownames(CQN.GENE_EXPRESSION$y) %in% CONTROL.PROBES$ensembl_gene_id 
-EXPR = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-EXPR = EXPR[ind,]
-PMI = PMI[colnames(EXPR),,drop=F]
-
-expr = dlply(PMI, .(ID), function(mtd, expr){
-  if (dim(mtd)[1] > 1)
-    expr = rowMeans(expr[, mtd$ID]) %>% rownameToFirstColumn('GeneID')
-  else
-    expr = expr[, mtd$ID] %>% rownameToFirstColumn('GeneID')
-  expr = expr %>% plyr::rename(c('DF' = unique(mtd$ID)))
-  return(expr)
-}, EXPR) %>%
-  join_all
-rownames(expr) = expr$GeneID
-
-PMI = PMI[,colnames(PMI) %in% c('ID','PMI')] %>% unique
-rownames(PMI) = PMI$ID
-
-PMI = PMI[intersect(rownames(PMI), colnames(expr)), ]
-expr = expr[,intersect(rownames(PMI), colnames(expr))]
-
-# Split data into trainining and prediction set
+# Fit a logistic regression model
 x.train = expr[,!is.na(PMI$PMI)]
@@
 y.train = PMI[!is.na(PMI$PMI),]
+y.train = y.train[colnames(x.train), 'PMI']
 
-plr.model = cv.glmnet(t(x.train), y.train$PMI, family = 'gaussian')
+plr.model = cv.glmnet(t(x.train), y.train, family = 'gaussian')
 coeff = as.matrix(coef.glmnet(plr.model, s = 'lambda.min'))
-mse = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train$PMI)^2)/length(y.train$PMI)
+ssres = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train)^2)
+sstot = sum((y.train-mean(y.train, na.rm = T))^2, na.rm = T)
+R2.smallgenes = 1-(ssres/sstot)
+
 y.predict = predict(plr.model, t(x.predict), s = 'lambda.min', type = 'response') %>%
-  rownameToFirstColumn('ID') %>%
+  rownameToFirstColumn('Donor_ID') %>%
   plyr::rename(c('1' = 'PMI'))
 
-PMI = rbindlist(list(y.train, y.predict), use.names = T, fill = T)
+PMI = PMI %>% 
+  dplyr::select(Donor_ID, PMI) %>%
+  dplyr::filter(!is.na(PMI)) %>%
+  list(y.predict) %>%
+  rbindlist(use.names = T, fill = T)
 PMI$PMI[PMI$PMI < 0] = 0
@@
 ```
+Fitted R2 values are, (i) with all gx features `r round(R2.allgenes, 3)`, (ii) with small set of inferred gx features `r round(R2.smallgenes,3)`
 
@@
 Correlation/association between covariates at an FDR <= 0.1
-```{r covariates.correlation, fig.width=6, fig.height=6}
+```{r covariates.correlation, fig.width=10, fig.height=6}
 COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
@@
 ```
-### CPM Normalisation
-Preprocess counts matrix using CQN (Conditional Quantile Normlaisation accounting for both GC content and gene length)
-
+### Filter genes
 Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion x Diagnosis. Also remove genes with missing gene length and percentage GC content
@@
 
-### Normalisation (with NULL)
+### Outlier Analysis
+```{r outlier.analysis, cache = FALSE, fig.width = 10}
+indToRemove = c("11311_CER", "1923_CER", "1923_TCX", "11396_TCX", "11294_TCX", "11408_TCX")
+
+# Find principal components of expression to plot
+PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)
+
+# Plot first 2 PCs
+plotdata <- data.frame(SampleID=rownames(PC$rotation), 
+                       PC1=PC$rotation[,1], 
+                       PC2=PC$rotation[,2])
+
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
+  dplyr::mutate(label = SampleID) %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
+plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
+p <- p + theme_bw() + theme(legend.position="top") 
+p <- p + facet_grid(BrainRegion~.+Diagnosis, scales = 'free')
+p <- p + geom_text(aes(label= label), size=4, hjust=0)
+p
+
+# Plot abberent distribution of logcpm counts
+tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
+  rownameToFirstColumn('Gene.ID') %>%
+  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID')) %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
+
+p = ggplot(tmp1 %>%
+             dplyr::filter(SampleID %in% indToRemove),
+           aes(x = logCPM, color = SampleID)) + geom_density() 
+p = p + theme(legend.position = 'top') + facet_grid(BrainRegion~.+Diagnosis)
+p
+
+indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
+PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
+COVARIATES = COVARIATES[indToRetain,]
+
+tmp = COVARIATES %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
+  group_by(BrainRegion, Diagnosis) %>%
+  summarise(count = n()) %>%
+  spread(Diagnosis, count)
+```
+Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples
+
+Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`
+
+Distribution of samples are: `r kable(tmp)`
+
+### Conditional Quantile Normalisation (CQN)
 Initial normalisation usign cqn and dispersion control with voom (with NULL design)
@@
 ```
+
+```{r explore.data, cache=FALSE, fig.height=5, fig.width=10}
+# Plot abberent distribution of logcpm counts
+tmp1 = (CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset) %>%
+  rownameToFirstColumn('Gene.ID') %>%
+  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID')) %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
+
+p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
+p = p + theme(legend.position = 'NONE') + facet_grid(BrainRegion+.~Diagnosis, scale = 'free')
+p
+```
 Coexpression of genes 
@@
 ```
-### Outlier analysis
-Clustering of voom adjusted data (with NULL design)
-```{r decompse.normalise.data1.1, fig.height=6, fig.width=8, results='asis'}
+### Decomposition of expression
+Clustering of library normalised unadjusted data (with NULL design)
+```{r decompse.normalise.data1.1, fig.height=6, fig.width=10, results='asis'}
 # Find principal components of expression to plot
@@
 plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  tidyr::separate(BrainRegion.Diagnosis, c('Region','Diagnosis'), sep = '\\.') %>% 
-  dplyr::mutate(Region = factor(Region),
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion','Diagnosis'), sep = '\\.') %>% 
+  dplyr::mutate(BrainRegion = factor(BrainRegion),
                 Source = factor(Source))
-plotdata$SampleID[!(plotdata$SampleID %in% c("11311_CER", "1923_CER", "1923_TCX", "11396_TCX", "11294_TCX", "11408_TCX"))] = ''
 
@@
 p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
-p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Region~., scales = 'free_y')
-p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+p <- p + theme_bw() + theme(legend.position="top") + facet_grid(BrainRegion+.~Diagnosis, scales = 'free_y')
 p
@@
 Interpretation of BrainRegion.Diagnosis
-1. TCX.PSP: Progressive Supranuclear Palsy samples from Temporal cortex region
-2. TCX.AD: Alzheimers Disease samples from Temporal cortex region
-3. TCX.Control: Control samples from Temporal cortex region
-4. TCX.Pathologic_Aging: Pathological Aging samples from Temporal cortex region
-5. CER.PSP: Progressive Supranuclear Palsy samples from Cerebellum region
-6. CER.AD: Alzheimers Disease samples from Cerebellum region
-7. CER.Control: Control samples from Cerebellum region
-8. CER.Pathologic_Aging: Pathological Aging samples from Cerebellum region
+1. CER.AD: Alzheimers Disease samples from Cerebellum region
+2. CER.CONTROL: Control samples from Cerebellum region
+3. CER.OTHER: Progressive Supranuclear Palsy and Pathological Aging samples from Cerebellum region
+4. TCX.AD: Alzheimers Disease samples from Temporal cortex region
+5. TCX.CONTROL: Control samples from Temporal cortex region
+6. TCX.OTHER: Progressive Supranuclear Palsy and Pathological Aging samples from Temporal cortex region
 
@@
 contrast = makeContrasts(contrasts=c(
-  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.AD",
-  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.Pathologic_Aging",
-  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.PSP",
-  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.AD",
-  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.PSP",
-  "BrainRegion.DiagnosisCER.PSP-BrainRegion.DiagnosisCER.AD",
-  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.AD",
-  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.Pathologic_Aging",
-  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.PSP",
-  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.AD",
-  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.PSP",
-  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD"),
+  "BrainRegion.DiagnosisCER.AD-BrainRegion.DiagnosisCER.CONTROL",
+  "BrainRegion.DiagnosisCER.OTHER-BrainRegion.DiagnosisCER.CONTROL",
+  "BrainRegion.DiagnosisCER.OTHER-BrainRegion.DiagnosisCER.AD",
+  "BrainRegion.DiagnosisTCX.AD-BrainRegion.DiagnosisTCX.CONTROL",
+  "BrainRegion.DiagnosisTCX.OTHER-BrainRegion.DiagnosisTCX.CONTROL",
+  "BrainRegion.DiagnosisTCX.OTHER-BrainRegion.DiagnosisTCX.AD"),
   levels = colnames(FIT$coefficients))
@@
 # Get differnetial expression
-DE = lapply(1:12, function(i, FIT){
+DE = lapply(1:6, function(i, FIT){
   topTable(FIT, coef=i, number = 50000, confint = T) %>%
@@
   tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
-  tidyr::separate(from.state, c('Region','reference'), sep = '\\.') %>%
-  tidyr::separate(to.state, c('Region','against'), sep = '\\.') %>%
+  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
+  tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
   tidyr::unite(Comparison, reference, against, sep = '-') %>%
@@
 tmp = DE %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction, Region) %>%
-  group_by(Region, Comparison, Direction) %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion) %>%
+  group_by(BrainRegion, Comparison, Direction) %>%
   dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
@@
 p = p + scale_color_manual(values = c('green','grey','red'))
-p = p + facet_grid(Region+.~Comparison, scales = 'fixed')
+p = p + facet_grid(BrainRegion+.~Comparison, scales = 'fixed')
 p
@@
 pl[[1]] = ggplot(DE %>%
-                   dplyr::filter(Comparison == 'Control-AD'), 
+                   dplyr::filter(Comparison == 'AD-CONTROL'), 
                  aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
 pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[1]] = pl[[1]] + facet_grid(BrainRegion+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
 pl[[1]] = pl[[1]] + theme(legend.position = 'top')
@@
 pl[[2]] = ggplot(DE %>%
-                   dplyr::filter(Comparison == 'Control-AD'), 
+                   dplyr::filter(Comparison == 'AD-CONTROL'), 
                  aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
 pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[2]] = pl[[2]] + facet_grid(BrainRegion+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
 pl[[2]] = pl[[2]] + theme(legend.position = 'top')
@@
 pl[[3]] = ggplot(DE %>%
-                   dplyr::filter(Comparison == 'Control-AD'), 
+                   dplyr::filter(Comparison == 'AD-CONTROL'), 
                  aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
 pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[3]] = pl[[3]] + facet_grid(BrainRegion+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
 pl[[3]] = pl[[3]] + theme(legend.position = 'top')
@@
 # Source modified functions from sva package
-# source('../R/num.sv.R')
 source('../R/irwsva.build.R')
 source('../R/f.pvalue.R')
-
-# Calculate the number of surrogate variables
-# NUM.SV = num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
-#                      MODEL1, method = 'be', seed = 123456, B = 30, tol = 1e-20)
 
@@
 SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
-                        MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
+                        MODEL1, MODEL0, n.sv = NUM.SV, B = 3, tol = 1e-20)$sv
 
@@
 DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-stopCluster(cl)
+# stopCluster(cl)
 ```

### 2017-04-15 code/Rmd/harmonisedCaseControl.Rmd
Minor sanity fix

@@
 all.mtd$ROSMAP$nDx = 'OTHER'
-all.mtd$ROSMAP$nDx[all.mtd$ROSMAP$cogdx==1 & all.mtd$ROSMAP$braaksc <= 3 & all.mtd$ROSMAP$ceradsc <= 1] = 'CONTROL'
-all.mtd$ROSMAP$nDx[all.mtd$ROSMAP$cogdx==4 & all.mtd$ROSMAP$braaksc >= 4 & all.mtd$ROSMAP$ceradsc >= 2] = 'AD'
+all.mtd$ROSMAP$nDx[all.mtd$ROSMAP$cogdx==1 & all.mtd$ROSMAP$braaksc <= 3 & all.mtd$ROSMAP$ceradsc >= 3] = 'CONTROL'
+all.mtd$ROSMAP$nDx[all.mtd$ROSMAP$cogdx==4 & all.mtd$ROSMAP$braaksc >= 4 & all.mtd$ROSMAP$ceradsc <= 2] = 'AD'
 

### 2017-04-15 code/Rmd/MSSM_REPROCESSED.Rmd
Modified MSSM markdown for April 2017 freeze

@@
 knit2synapse::knitToFolderEntity(file = "./MSSM_REPROCESSED.Rmd",
-                                 parentId = "syn7554308",
+                                 parentId = "syn8672415",
                                  entityName = 'MSSM Reprocessed RNASeq (Normalised)')
@@
 # Get reprocessed sample counts for temporal cortex
-COUNT_ID <- 'syn7994853';
+COUNT_ID <- 'syn8691099';#'syn7994853';
 ALL_USED_IDs <- COUNT_ID
-COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F)
+COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F, row.names = 1)
 
@@
 # Get picard metrics from synapse
-METADATA.PICARD_ID <- 'syn8380379'
+METADATA.PICARD_ID <- 'syn8698270';#'syn8380379'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
-METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
-METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
-METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
-METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)
+# METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
+# METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
+# METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
+# METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)
+# METADATA.PICARD$sampleIdentifier = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
+#   gsub('.accepted_hits.sort.coord/picard.CombinedMetrics.csv','',.)
 
-METADATA.PICARD$sampleIdentifier = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
-  gsub('.accepted_hits.sort.coord/picard.CombinedMetrics.csv','',.)
+METADATA.PICARD <- CovariateAnalysis::downloadFile(METADATA.PICARD_ID) %>%
+  dplyr::rename(sampleIdentifier = sample)
 colnames(METADATA.PICARD) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD))
@@
                 RIN2 = RIN^2)
-METADATA$Diagnosis[METADATA$CDR <= 1] = 'ND'
-METADATA$Diagnosis[METADATA$CDR == 2 | METADATA$CDR == 3] = 'MD'
-METADATA$Diagnosis[METADATA$CDR >= 4] = 'SD'
+
+METADATA$Diagnosis = 'OTHER'
+METADATA$Diagnosis[METADATA$CDR <= 0.5 & METADATA$bbscore <= 3 & METADATA$NP.1 <= 1] = 'CONTROL'
+METADATA$Diagnosis[METADATA$CDR >= 1 & METADATA$bbscore >= 4 & METADATA$NP.1 >= 2] = 'AD'
+
 levels(METADATA$batch)[levels(METADATA$batch) == ''] = 'NoBatch'
@@
 ```{r covariates.clustering}
-primaryVariable <- "BrainRegion.Diagnosis"
-FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrainRegion.Diagnosis")
+primaryVariable <- c("CDR", "BrainRegion.Diagnosis")
+FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrainRegion.Diagnosis", "CDR")
 ContCovariates <- c("RIN", "RIN2", "AOD","PMI",  "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
@@
 ```
+
+### Outlier Analysis
+```{r outlier.analysis, cache = FALSE, fig.width = 10}
+indToRemove = c('hB_RNA_10622')
+
+# Find principal components of expression to plot
+PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)
+
+# Plot first 2 PCs
+plotdata <- data.frame(SampleID=rownames(PC$rotation), 
+                       PC1=PC$rotation[,1], 
+                       PC2=PC$rotation[,2])
+
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
+  dplyr::mutate(label = SampleID) %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
+plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(color=Diagnosis, size=RIN, shape = SEX))
+p <- p + theme_bw() + theme(legend.position="right")
+p <- p + geom_text(aes(label= label), size=4, hjust=0)
+p
+
+# Plot abberent distribution of logcpm counts
+tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
+  rownameToFirstColumn('Gene.ID') %>%
+  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID')) %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
+
+p = ggplot(tmp1 %>%
+             dplyr::filter(SampleID %in% indToRemove),
+           aes(x = logCPM, color = SampleID)) + geom_density() 
+p = p + theme(legend.position = 'top') + facet_grid(Diagnosis+.~BrainRegion)
+p
+
+indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
+PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
+COVARIATES = COVARIATES[indToRetain,]
+
+tmp = COVARIATES %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
+  group_by(BrainRegion, Diagnosis) %>%
+  summarise(count = n()) %>%
+  spread(Diagnosis, count)
+```
 Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples
 
-### Normalisation (with NULL)
+Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`
+
+Distribution of samples are: `r kable(tmp)`
+
+### Conditional Quantile Normalisation (CQN)
 Initial normalisation usign cqn and dispersion control with voom (with NULL design)
@@
 ```
-### Outlier analysis
-Clustering of voom adjusted data (with NULL design)
+### Decomposition of expression
+Clustering of library normalised unadjusted data (with NULL design)
 ```{r decompse.normalise.data1.1, fig.height=5, fig.width=10, results='asis'}
@@
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(.~Region, scales = 'free_y')
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
 p
@@
 ```
-
+`r tmp`
 ### Sanity check
@@
 
-# Plot first 4 PCs
+# Plot first 2 PCs
 plotdata <- data.frame(SampleID=rownames(PC$rotation), 
@@
 
+### Differential expression analysis (with BrainRegion.CDR as primary variable)
+Differential expression is performed on the primary variable by controlling for covariates identified above
+
+Interpretation of BrainRegion.CDR
+1. FP.CDR: change in CDR in frontal pole where 0 no cognitive decline and 5 severe cognitive decline
+2. IFG.CDR: change in CDR in inferior frontal gyrus where 0 no cognitive decline and 5 severe cognitive decline
+3. PHG.CDR: change in CDR in parahippocampal gyrus where 0 no cognitive decline and 5 severe cognitive decline
+4. STG.CDR: change in CDR in superior temporal gyrus where 0 no cognitive decline and 5 severe cognitive decline
+
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diff.exp, fig.height=6, fig.width=15}
+# Get design matrix
+COVARIATES.tmp = COVARIATES %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
+  dplyr::mutate(BrainRegion = factor(BrainRegion))
+DESIGN = getDesignMatrix(COVARIATES.tmp[, c('BrainRegion', postAdjustCovars), drop = F], Intercept = F)
+ind = grep('BrainRegion', colnames(DESIGN$design))
+DESIGN$design[,ind] = DESIGN$design[,ind] * as.numeric(COVARIATES.tmp[,'CDR'])
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
+
+# Calculate correlation values of random effects
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
+correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)
+
+# Re-estimate voom weights
+VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
+                    design=DESIGN$design, plot=F,
+                    block = COVARIATES$individualIdentifier, 
+                    correlation = correlation$cor)
+
+# Fit linear model using new weights and new design
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
+FIT = lmFit(VOOM.WEIGHTS)
+
+# Fit contrast
+contrast = makeContrasts(contrasts = c("BrainRegionFP", "BrainRegionIFG", "BrainRegionPHG", "BrainRegionSTG"),
+                         levels = colnames(FIT$coefficients))
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = lapply(1:4, function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = T) %>%
+    rownameToFirstColumn('ensembl_gene_id')
+}, FIT.CONTR)
+names(DE) = colnames(contrast)
+
+DE = DE %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Comparison = gsub('BrainRegion','',Comparison),
+                Study = 'MSSM',
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  dplyr::rename(BrainRegion = Comparison) %>%
+  dplyr::mutate(Comparison = 'CDR') %>%
+  left_join(GENE.PARAM %>%
+              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
+              unique())
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion) %>%
+  group_by(BrainRegion, Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05_FC_1.2) 
+kable(tmp)
+
+p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
+p = p + facet_grid(Comparison+.~BrainRegion, scales = 'free')
+p
+```
+
 ### Differential expression analysis (with BrainRegion.Diagnosis as primary variable)
@@
 
+METADATA$Diagnosis[METADATA$CDR <= 0.5 & METADATA$bbscore <= 3 & METADATA$NP.1 <= 1] = 'CONTROL'
+METADATA$Diagnosis[METADATA$CDR >= 1 & METADATA$bbscore >= 4 & METADATA$NP.1 >= 2] = 'AD'
+
 Interpretation of BrainRegion.Diagnosis
-1. FP.MD: Mild dementia samples from frontal pole with CDR equal to 2 or 3
-2. FP.ND: No dementia samples from frontal pole with CDR less than 2
-3. FP.SD: Severe dementia samples from frontal pole with CDR grater than 3
-4. IFG.MD: Mild dementia samples from inferior frontal gyrus with CDR equal to 2 or 3
-5. IFG.ND: No dementia samples from inferior frontal gyrus with CDR less than 2
-6. IFG.SD: Severe dementia samples from inferior frontal gyrus with CDR grater than 3
-7. PHG.MD: Mild dementia samples from para hippocampal gyrus with CDR equal to 2 or 3
-8. PHG.ND: No dementia samples from para hippocampal gyrus with CDR less than 2
-9. PHG.SD: Severe dementia samples from para hippocampal gyrus with CDR grater than 3
-10. STG.MD: Mild dementia samples from superior temporal gyrus with CDR equal to 2 or 3
-11. STG.ND: No dementia samples from superior temporal gyrus with CDR less than 2
-12. STG.SD: Severe dementia samples from superior temporal gyrus with CDR grater than 3
+1. FP.AD: Samples from frontal pole region with CDR >= 1 and bbscore >= 4 and NP.1 >= 2
+2. FP.CONTROL: Samples from frontal pole region with CDR <= 0.5 and bbscore <= 3 and NP.1 <= 1
+3. FP.OTHER: All the other frontal pole samples
+4. PHG.AD: Samples from parahippocampal gyrus region with CDR >= 1 and bbscore >= 4 and NP.1 >= 2
+5. PHG.CONTROL: Samples from parahippocampal gyrus region with CDR <= 0.5 and bbscore <= 3 and NP.1 <= 1
+6. PHG.OTHER:All the other parahippocampal gyrus samples
+7. STG.AD: Samples from frontal superior temporal gyrus with CDR >= 1 and bbscore >= 4 and NP.1 >= 2
+8. STG.CONTROL: Samples from superior temporal gyrus region with CDR <= 0.5 and bbscore <= 3 and NP.1 <= 1
+9. STG.OTHER: All the other superior temporal gyrus samples
+10. IFG.AD: Samples from inferior frontal gyrus region with CDR >= 1 and bbscore >= 4 and NP.1 >= 2
+11. IFG.CONTROL: Samples from inferior frontal gyrus region with CDR <= 0.5 and bbscore <= 3 and NP.1 <= 1
+12. IFG.OTHER: All the other inferior frontal gyrus samples
 
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diff.exp, fig.height=15, fig.width=13}
+```{r diff.exp1, fig.height=15, fig.width=13}
 # Get design matrix
-DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable, postAdjustCovars), drop = F], Intercept = F)
+DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[2], postAdjustCovars), drop = F], Intercept = F)
 DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
@@
 for (region in c('FP', 'STG', 'PHG', 'IFG')){
-  tmp = combn(paste0('BrainRegion.Diagnosis', region, '.', c('ND','MD','SD')),2)
+  tmp = combn(paste0('BrainRegion.Diagnosis', region, '.', c('OTHER','AD','CONTROL')),2)
   cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
@@
 p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
-p = p + scale_color_manual(values = c('green','grey','red'))
-p = p + facet_grid(Region+.~Comparison, scales = 'fixed')
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position='top')
+p = p + facet_grid(BrainRegion+.~Comparison, scales = 'fixed')
 p
@@
 pl = list()
-pl[[1]] = ggplot(DE %>% dplyr::filter(Comparison == 'ND-SD'), 
+pl[[1]] = ggplot(DE %>% dplyr::filter(Comparison == 'AD-CONTROL'), 
                  aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
 pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(Region~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
+pl[[1]] = pl[[1]] + facet_grid(BrainRegion~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
 pl[[1]] = pl[[1]] + theme(legend.position = 'top')
 
-pl[[2]] = ggplot(DE %>% dplyr::filter(Comparison == 'ND-SD'), 
+pl[[2]] = ggplot(DE %>% dplyr::filter(Comparison == 'AD-CONTROL'), 
                  aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
 pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(Region~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
+pl[[2]] = pl[[2]] + facet_grid(BrainRegion~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
 pl[[2]] = pl[[2]] + theme(legend.position = 'top')
 
-pl[[3]] = ggplot(DE %>% dplyr::filter(Comparison == 'ND-SD'), 
+pl[[3]] = ggplot(DE %>% dplyr::filter(Comparison == 'AD-CONTROL'), 
                  aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
 pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(Region~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
+pl[[3]] = pl[[3]] + facet_grid(BrainRegion~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
 pl[[3]] = pl[[3]] + theme(legend.position = 'top')
@@
 # Get (null) design matrix
-MODEL0 = DESIGN$design[,-grep(primaryVariable,colnames(DESIGN$design))] # Get null model by removing variable of interest
+MODEL0 = DESIGN$design[,-grep(primaryVariable[2],colnames(DESIGN$design))] # Get null model by removing variable of interest
 MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]

### 2017-04-15 code/Rmd/MAYO_REPROCESSED.Rmd

@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diff.exp, fig.height=6, fig.width=18}
+```{r diff.exp, fig.height=10, fig.width=18}
 # Get design matrix
@@
 ### Associate differential expression results with gc content, gene length and average expression
-```{r associate.de, fig.height=5, fig.width=15}
+```{r associate.de, fig.height=10, fig.width=15}
 pl = list()
@@
 SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
-                        MODEL1, MODEL0, n.sv = NUM.SV, B = 3, tol = 1e-20)$sv
+                        MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
 

### 2017-04-15 code/Rmd/MSSM_REPROCESSED.Rmd
Minor changes

@@
 ### Clustering residual data
-```{r decompse.normalise.data2.1, fig.height=10, fig.width=8, results='asis'}
+```{r decompse.normalise.data2.1, fig.height=6, fig.width=10, results='asis'}
 # Find principal components of expression to plot
@@
 Differential expression is performed on the primary variable by controlling for covariates identified above
-
-METADATA$Diagnosis[METADATA$CDR <= 0.5 & METADATA$bbscore <= 3 & METADATA$NP.1 <= 1] = 'CONTROL'
-METADATA$Diagnosis[METADATA$CDR >= 1 & METADATA$bbscore >= 4 & METADATA$NP.1 >= 2] = 'AD'
 

### 2017-04-21 code/Rmd/ROSMAP_REPROCESSED.Rmd
Major modifications in ROSMAP to account for winzorisation

@@
                           verbose = FALSE)
+CQN.GENE_EXPRESSION$E.NA = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
+CQN.GENE_EXPRESSION$E.zero = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
 
 # Set gene counts in specific samples that are deviating 3 sd from other samples to 0
-log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
+log.mat = apply(CQN.GENE_EXPRESSION$E.zero, 1, function(x){
   mn = mean(x, na.rm = T)
@@
 }) %>% t
-PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0
+PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = NA
 
-# Re-compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix$counts), 'percentage_gc_content'],
-                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix$counts), 'gene.length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
+CQN.GENE_EXPRESSION$E.NA[log.mat] = NA
+CQN.GENE_EXPRESSION$E.zero[log.mat] = 0
 ```
@@
 # Plot abberent distribution of logcpm counts
-tmp1 = (CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset) %>%
+tmp1 = (CQN.GENE_EXPRESSION$E.NA) %>%
   rownameToFirstColumn('Gene.ID') %>%
@@
 ```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
+cr = cor(t(CQN.GENE_EXPRESSION$E.zero))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)
+PC <- prcomp(CQN.GENE_EXPRESSION$E.zero, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset)))
+tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E.zero)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 # Find correlation between PC's of gene expression with covariates
-preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
+preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E.zero, 
                                                  COVARIATES,
@@
   # Estimate voom weights for dispersion control
-  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)
+  counts = PROCESSED_COUNTS$filteredExprMatrix$counts
+  counts[is.na(counts)] = 0
+  VOOM.GENE_EXPRESSION = voom(counts, 
+                              design=DM1$design, 
+                              plot=F,
+                              na.rm = T)
 
   # Fit linear model using new weights and new design
-  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
+  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$E.NA,
                             design = DM1$design,
@@
   # Residuals after normalisation
-  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
-                                                CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
+  RESIDUAL.GENE_EXPRESSION = list()
+  RESIDUAL.GENE_EXPRESSION$expr.NA = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
+                                                        CQN.GENE_EXPRESSION$E.NA)
+  RESIDUAL.GENE_EXPRESSION$expr.zero = RESIDUAL.GENE_EXPRESSION$expr.NA
+  RESIDUAL.GENE_EXPRESSION$expr.zero[is.na(RESIDUAL.GENE_EXPRESSION$expr.NA)] = 0
 
@@
   # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION$expr.zero, 
                                                 COVARIATES[, residCovars, drop=F], 
@@
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION$expr.zero, 
                                               COVARIATES,
@@
 ```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.GENE_EXPRESSION))
+cr = cor(t(RESIDUAL.GENE_EXPRESSION$expr.zero))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION$expr.zero, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
+tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION$expr.zero)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
+counts = PROCESSED_COUNTS$filteredExprMatrix$counts
+counts[is.na(counts)] = 0
+VOOM.WEIGHTS = voom(counts, design=DESIGN$design, plot=F)
 
 # Fit linear model using new weights and new design
-FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
+FIT = lmFit(CQN.GENE_EXPRESSION$E.NA,
             design = DESIGN$design,
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("cogdx1-cogdx2",
-                                     "cogdx1-cogdx4",
-                                     "cogdx2-cogdx4"),
+contrast = makeContrasts(contrasts=c("cogdx2-cogdx1",
+                                     "cogdx4-cogdx1",
+                                     "cogdx4-cogdx2"),
                          levels = colnames(FIT$coefficients))
@@
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(counts, design=DESIGN$design, plot=F)
 
 # Fit linear model using new weights and new design
-FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
+FIT = lmFit(CQN.GENE_EXPRESSION$E.NA,
             design = DESIGN$design,
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("DiagnosisAD-DiagnosisCONTROL",
-                                     "DiagnosisOTHER-DiagnosisAD",
-                                     "DiagnosisOTHER-DiagnosisCONTROL"),
+contrast = makeContrasts(contrasts=c("DiagnosisCONTROL-DiagnosisAD",
+                                     "DiagnosisAD-DiagnosisOTHER",
+                                     "DiagnosisCONTROL-DiagnosisOTHER"),
                          levels = colnames(FIT$coefficients))
@@
 # Calculate the number of surrogate variables (be method)
-# NUM.SV = sva::num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, MODEL1, method = 'be', seed = 123456, B = 30)
+# NUM.SV = sva::num.sv(CQN.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456, B = 30)
 
 # Get residuals from differential expression model
-RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
+RESIDUAL.SVA.GENE_EXPRESSION = list()
+RESIDUAL.SVA.GENE_EXPRESSION$expr.NA = residuals.MArrayLM(FIT, CQN.GENE_EXPRESSION$E.NA)
+RESIDUAL.SVA.GENE_EXPRESSION$expr.zero = RESIDUAL.SVA.GENE_EXPRESSION$expr.NA
+RESIDUAL.SVA.GENE_EXPRESSION$expr.zero[is.na(RESIDUAL.SVA.GENE_EXPRESSION$expr.NA)] = 0
 
 # Compute actual variance of all principal components
-tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
+tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION$expr.zero)
 actual.var = tmp$d^2/sum(tmp$d^2)
@@
 permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
-  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
+  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION$expr.zero, 1, sample, replace = FALSE))
   tmp = svd(tmp.residual)
@@
 # Estimate surrogate variables
-SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
+SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$E.zero, 
                     MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv
@@
 # Re-estimate voom weights
-VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+VOOM.SVA.WEIGHTS = voom(counts, 
                         design = cbind(MODEL1, SURR.VAR), plot=F)
@@
 # Fit linear model using new weights and new design
-VOOM.SVA.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
+VOOM.SVA.FIT = lmFit(CQN.GENE_EXPRESSION$E.NA, 
                      design = cbind(MODEL1, SURR.VAR),
@@
 RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
-                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
+                                                  CQN.GENE_EXPRESSION$E.NA)
+
+# Add variable of interest back to the residuals
+varsToAddIn = grep("Diagnosis", colnames(MODEL1), value = T)
+RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
+  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
 ```
@@
 # Estimate voom weights for dispersion control
-VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+VOOM.NET.WEIGHTS = voom(counts, 
                         design=DESIGN.NET, plot=F)
@@
 # Fit linear model using new weights and new design
-VOOM.NET.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
+VOOM.NET.FIT = lmFit(CQN.GENE_EXPRESSION$E.NA,
                      design = DESIGN.NET,
@@
 RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
-                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
+                                                  CQN.GENE_EXPRESSION$E.NA)
 ```

### 2017-04-23 code/Rmd/MAYO_REPROCESSED.Rmd
Minor modifications

@@
                           verbose = FALSE)
+CQN.GENE_EXPRESSION$E.NA = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
 
@@
 }) %>% t
+
+PROCESSED_COUNTS$filteredExprMatrix$counts.NA = PROCESSED_COUNTS$filteredExprMatrix$counts 
+PROCESSED_COUNTS$filteredExprMatrix$counts.NA[log.mat] = NA
 PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0
 
-# Re-compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
-                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
+CQN.GENE_EXPRESSION$E.NA[log.mat] = NA
+CQN.GENE_EXPRESSION$E.zero = CQN.GENE_EXPRESSION$E.NA
+CQN.GENE_EXPRESSION$E.zero[log.mat] = 0
 ```
@@
   # Calculate correlation values of random effects
-  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
+  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E.NA
   correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
@@
   # Fit linear model using new weights and new design
-  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
+  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$E.NA,
                             design = DM1$design,
@@
   # Residuals after normalisation
-  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
-                                                CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
+  RESIDUAL.GENE_EXPRESSION = list()
+  RESIDUAL.GENE_EXPRESSION$expr.NA = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
+                                                CQN.GENE_EXPRESSION$E.NA)
+  RESIDUAL.GENE_EXPRESSION$expr.zero = RESIDUAL.GENE_EXPRESSION$expr.NA
+  RESIDUAL.GENE_EXPRESSION$expr.zero[is.na(RESIDUAL.GENE_EXPRESSION$expr.zero)] = 0
 
@@
 # Calculate correlation values of random effects
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
 correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Donor_ID)
@@
 # Fit linear model using new weights and new design
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
 FIT = lmFit(VOOM.WEIGHTS)
@@
 contrast = makeContrasts(contrasts=c(
-  "BrainRegion.DiagnosisCER.AD-BrainRegion.DiagnosisCER.CONTROL",
-  "BrainRegion.DiagnosisCER.OTHER-BrainRegion.DiagnosisCER.CONTROL",
-  "BrainRegion.DiagnosisCER.OTHER-BrainRegion.DiagnosisCER.AD",
-  "BrainRegion.DiagnosisTCX.AD-BrainRegion.DiagnosisTCX.CONTROL",
-  "BrainRegion.DiagnosisTCX.OTHER-BrainRegion.DiagnosisTCX.CONTROL",
-  "BrainRegion.DiagnosisTCX.OTHER-BrainRegion.DiagnosisTCX.AD"),
+  "BrainRegion.DiagnosisCER.CONTROL-BrainRegion.DiagnosisCER.AD",
+  "BrainRegion.DiagnosisCER.CONTROL-BrainRegion.DiagnosisCER.OTHER",
+  "BrainRegion.DiagnosisCER.AD-BrainRegion.DiagnosisCER.OTHER",
+  "BrainRegion.DiagnosisTCX.CONTROL-BrainRegion.DiagnosisTCX.AD",
+  "BrainRegion.DiagnosisTCX.CONTROL-BrainRegion.DiagnosisTCX.OTHER",
+  "BrainRegion.DiagnosisTCX.AD-BrainRegion.DiagnosisTCX.OTHER"),
   levels = colnames(FIT$coefficients))
@@
 pl[[1]] = ggplot(DE %>%
-                   dplyr::filter(Comparison == 'AD-CONTROL'), 
+                   dplyr::filter(Comparison == 'CONTROL-AD'), 
                  aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
@@
 pl[[2]] = ggplot(DE %>%
-                   dplyr::filter(Comparison == 'AD-CONTROL'), 
+                   dplyr::filter(Comparison == 'CONTROL-AD'), 
                  aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
@@
 pl[[3]] = ggplot(DE %>%
-                   dplyr::filter(Comparison == 'AD-CONTROL'), 
+                   dplyr::filter(Comparison == 'CONTROL-AD'), 
                  aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
@@
 # Get residuals from differential expression model
-RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, VOOM.WEIGHTS$E)
+RESIDUAL.SVA.GENE_EXPRESSION = list()
+RESIDUAL.SVA.GENE_EXPRESSION$expr.NA = residuals.MArrayLM(FIT, CQN.GENE_EXPRESSION$E.NA)
+RESIDUAL.SVA.GENE_EXPRESSION$expr.zero = RESIDUAL.SVA.GENE_EXPRESSION$expr.NA
+RESIDUAL.SVA.GENE_EXPRESSION$expr.zero[is.na(RESIDUAL.SVA.GENE_EXPRESSION$expr.NA)] = 0
 
 # Compute actual variance of all principal components
-tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
+tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION$expr.zero)
 actual.var = tmp$d^2/sum(tmp$d^2)
@@
 # Estimate surrogate variables
-SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
-                        MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
+SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E.zero,
+                        MODEL1, MODEL0, n.sv = NUM.SV, 
+                        B = 30, tol = 1e-20)$sv
 
@@
 # Calculate correlation values of random effects
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
+VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
 correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$Donor_ID)
@@
 # Fit linear model using new weights and new design
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
+VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
 VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)
@@
 RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
-                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
+                                                  CQN.GENE_EXPRESSION$E.NA)
+
+# Add variable of interest back to the residuals
+varsToAddIn = grep("BrainRegion.Diagnosis", colnames(MODEL1), value = T)
+RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
+  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
 ```
@@
 # Calculate correlation values of random effects
-VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
+VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
 correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$Donor_ID)
@@
 # Fit linear model using new weights and new design
-VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
+VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
 VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)

### 2017-04-24 code/Rmd/MAYO_REPROCESSED.Rmd
Fixed outlier expression/ winsorised MSSM data

@@
 # Store filtered counts
-PROCESSED_COUNTS$filteredExprMatrix$counts %>%
+PROCESSED_COUNTS$filteredExprMatrix$counts.NA %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
@@
 # Store logCPM
-CQN.GENE_EXPRESSION$y %>%
+tmp = CQN.GENE_EXPRESSION$y
+tmp[log.mat] = NA
+tmp %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
@@
 # Store cqn offsets
-CQN.GENE_EXPRESSION$offset %>%
+tmp = CQN.GENE_EXPRESSION$offset
+tmp[log.mat] = NA
+tmp %>%
   rownameToFirstColumn('ensembl_gene_id') %>%

### 2017-04-24 code/Rmd/MSSM_REPROCESSED.Rmd
Fixed outlier expression/ winsorised MSSM data

@@
 }) %>% t
+PROCESSED_COUNTS$filteredExprMatrix$counts.NA = PROCESSED_COUNTS$filteredExprMatrix$counts 
+PROCESSED_COUNTS$filteredExprMatrix$counts.NA[log.mat] = NA
 PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0
 
-# Re-compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
-                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
+CQN.GENE_EXPRESSION$E.NA[log.mat] = NA
+CQN.GENE_EXPRESSION$E.zero = CQN.GENE_EXPRESSION$E.NA
+CQN.GENE_EXPRESSION$E.zero[log.mat] = 0
 ```
@@
   # Residuals after normalisation
-  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
-                                                CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
+  RESIDUAL.GENE_EXPRESSION = list()
+  RESIDUAL.GENE_EXPRESSION$expr.NA = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
+                                                        CQN.GENE_EXPRESSION$E.NA)
+  RESIDUAL.GENE_EXPRESSION$expr.zero = RESIDUAL.GENE_EXPRESSION$expr.NA
+  RESIDUAL.GENE_EXPRESSION$expr.zero[is.na(RESIDUAL.GENE_EXPRESSION$expr.zero)] = 0
 
@@
   # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION$E.zero, 
                                                 COVARIATES[, residCovars, drop=F], 
@@
 DESIGN$design[,ind] = DESIGN$design[,ind] * as.numeric(COVARIATES.tmp[,'CDR'])
+colnames(DESIGN$design)[ind] = paste0(colnames(DESIGN$design),'.CDR')
 DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
@@
 # Calculate correlation values of random effects
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
 correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)
@@
 # Fit contrast
-contrast = makeContrasts(contrasts = c("BrainRegionFP", "BrainRegionIFG", "BrainRegionPHG", "BrainRegionSTG"),
+contrast = makeContrasts(contrasts = c("BrainRegionFP.CDR", 
+                                       "BrainRegionIFG.CDR", 
+                                       "BrainRegionPHG.CDR", 
+                                       "BrainRegionSTG.CDR"),
                          levels = colnames(FIT$coefficients))
@@
 for (region in c('FP', 'STG', 'PHG', 'IFG')){
-  tmp = combn(paste0('BrainRegion.Diagnosis', region, '.', c('OTHER','AD','CONTROL')),2)
+  tmp = combn(paste0('BrainRegion.Diagnosis', region, '.', c('CONTROL', 'AD', 'OTHER')),2)
   cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
@@
 pl = list()
-pl[[1]] = ggplot(DE %>% dplyr::filter(Comparison == 'AD-CONTROL'), 
+pl[[1]] = ggplot(DE %>% dplyr::filter(Comparison == 'CONTROL-AD'), 
                  aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
@@
 
-pl[[2]] = ggplot(DE %>% dplyr::filter(Comparison == 'AD-CONTROL'), 
+pl[[2]] = ggplot(DE %>% dplyr::filter(Comparison == 'CONTROL-AD'), 
                  aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
@@
 
-pl[[3]] = ggplot(DE %>% dplyr::filter(Comparison == 'AD-CONTROL'), 
+pl[[3]] = ggplot(DE %>% dplyr::filter(Comparison == 'CONTROL-AD'), 
                  aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
@@
 # Estimate surrogate variables
-SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
+SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$E.zero,
                     MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv
@@
 # Calculate correlation values of random effects
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
+VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
 correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$individualIdentifier)
@@
 # Calculate correlation values of random effects
-VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
+VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
 correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$individualIdentifier)

### 2017-04-25 code/Rmd/MSSM_REPROCESSED.Rmd
Minor bug fix

@@
   # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION$E.zero, 
+  tmp.expr = RESIDUAL.GENE_EXPRESSION$expr.zero
+  tmp.expr = tmp.expr[apply(tmp.expr, 1, var) != 0,]
+  residualSigCovars = runPCAandPlotCorrelations(tmp.expr, 
                                                 COVARIATES[, residCovars, drop=F], 
@@
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION$expr.zero, 
+tmp.expr = RESIDUAL.GENE_EXPRESSION$expr.zero
+tmp.expr = tmp.expr[apply(tmp.expr, 1, var) != 0,]
+residualSigCovars = runPCAandPlotCorrelations(tmp.expr, 
                                               COVARIATES,

### 2017-04-25 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Modified covariates code to winsorize expression data and removed ribozero and flowcell batch variables

@@
 ```{r covariates.clustering}
-FactorCovariates <- c('Individual_ID', "Institution", "Gender", "RibozeroBatch", "LibraryBatch", "FlowcellBatch", "Dx.Tissue")
+#"RibozeroBatch", "FlowcellBatch",
+FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx.Tissue")
 ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
@@
 ```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
-indToRemove = c('MSSM_RNA_ACC_BP_24',
-
-                'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250', 'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 
-                'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220', 'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133',
-
-                'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 
-                'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56')
+indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250',
+                'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220',
+                'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
+                'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
+                'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_ACC_310",
+                "MSSM_RNA_ACC_51",  "MSSM_RNA_ACC_254", "MSSM_RNA_ACC_28", "MSSM_RNA_ACC_293", "MSSM_RNA_ACC_233",
+                "MSSM_RNA_ACC_167", "MSSM_RNA_ACC_353", "MSSM_RNA_ACC_355", "MSSM_RNA_ACC_211", "PENN_RNA_ACC_122",
+                "PENN_RNA_ACC_049", "PENN_RNA_ACC_052", "PENN_RNA_ACC_068", "PENN_RNA_ACC_070", "PITT_RNA_ACC_10020",
+                "PITT_RNA_ACC_1240", "PITT_RNA_ACC_1284", "PITT_RNA_ACC_700", "PITT_RNA_ACC_857", "PITT_RNA_ACC_BP_1589",
+                "MSSM_RNA_PFC_230", "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
+                "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")
 
@@
 }) %>% t
+PROCESSED_COUNTS$filteredExprMatrix$counts.NA = PROCESSED_COUNTS$filteredExprMatrix$counts
+PROCESSED_COUNTS$filteredExprMatrix$counts.NA[log.mat] = NA
 PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0
 
-# Re-compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
-                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
+CQN.GENE_EXPRESSION$E.NA[log.mat] = NA
+CQN.GENE_EXPRESSION$E.zero = CQN.GENE_EXPRESSION$E.NA
+CQN.GENE_EXPRESSION$E.zero[log.mat] = 0
 ```
@@
 # Plot abberent distribution of logcpm counts
-tmp1 = (CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset) %>%
+tmp1 = (CQN.GENE_EXPRESSION$E.NA) %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
@@
 
-tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)))
+tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E.zero)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
   VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)
-  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E.NA
   correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
@@
   # Fit linear model using new weights and new design
-  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E.NA
   ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
@@
   # Residuals after normalisation
-  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset)
+  RESIDUAL.GENE_EXPRESSION = list()
+  RESIDUAL.GENE_EXPRESSION$E.NA = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E.NA)
+  RESIDUAL.GENE_EXPRESSION$E.zero = RESIDUAL.GENE_EXPRESSION$E.NA
+  RESIDUAL.GENE_EXPRESSION$E.zero[is.na(RESIDUAL.GENE_EXPRESSION$E.NA)] = 0
 
@@
   # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+  tmp.expr = RESIDUAL.GENE_EXPRESSION$E.zero
+  tmp.expr = tmp.expr[apply(tmp.expr,1,var) !=0,]
+  residualSigCovars = runPCAandPlotCorrelations(tmp.expr, 
                                                 COVARIATES[, residCovars, drop=F], 
@@
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION$E.zero, 
                                               COVARIATES, 
@@
 varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
+RESIDUAL.GENE_EXPRESSION$E.NA = RESIDUAL.GENE_EXPRESSION$E.NA + 
   ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
@@
 ```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.GENE_EXPRESSION))
+cr = cor(t(RESIDUAL.GENE_EXPRESSION$E.zero))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION$E.zero, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
+tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION$E.zero)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
 correlation <- parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Individual_ID)
@@
 # Fit linear model using new weights and new design
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
 FIT = lmFit(VOOM.WEIGHTS)
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueControl_ACC-Dx.TissueOther_ACC",
-                                     "Dx.TissueControl_ACC-Dx.TissueSCZ_ACC",
-                                     "Dx.TissueOther_ACC-Dx.TissueSCZ_ACC",
-                                     "Dx.TissueControl_DLPFC-Dx.TissueOther_DLPFC",
-                                     "Dx.TissueControl_DLPFC-Dx.TissueSCZ_DLPFC",
-                                     "Dx.TissueOther_DLPFC-Dx.TissueSCZ_DLPFC"),
+contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
+                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
+                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
+                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC"),
                          levels = colnames(FIT$coefficients))
@@
 # Store logCPM
-CQN.GENE_EXPRESSION$y %>%
+CQN.GENE_EXPRESSION$E.NA %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
@@
 # Store residual gene expression for network analysis
-RESIDUAL.GENE_EXPRESSION %>%
+RESIDUAL.GENE_EXPRESSION$E.NA %>%
   rownameToFirstColumn('ensembl_gene_id') %>%

### 2017-04-26 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor bug fix

@@
 # Join ACC and DLPFC together
-COUNT = join_all(list(COUNT_ACC, COUNT_DLPFC), type = 'full')
+COUNT = join_all(list(COUNT_ACC, COUNT_DLPFC), type = 'full') %>%
+  as.data.frame()
 
@@
 rownames(COUNT) = COUNT$Geneid
-COUNT[,c('Geneid', 'Chr', 'Start', 'End', 'Strand', 'Length')] = NULL
-COUNT = data.matrix(COUNT)
+ind = setdiff(colnames(COUNT), c('Geneid', 'Chr', 'Start', 'End', 'Strand', 'Length'))
+COUNT = data.matrix(COUNT[,ind])
 
@@
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION$E.zero, 
+tmp.expr = RESIDUAL.GENE_EXPRESSION$E.zero
+tmp.expr = tmp.expr[apply(tmp.expr,1,var) !=0,]
+residualSigCovars = runPCAandPlotCorrelations(tmp.expr, 
                                               COVARIATES, 
@@
 ENRICH_OBJ = synStore(ENRICH_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-stopCluster(cl)
+# stopCluster(cl)
 ```

### 2017-05-07 code/Rmd/combineAllRNASeqData.Rmd
Markdown to combine all the RNASeq data from 3 brain banks

---
title: "Combine filtered data from all three studies in AMP-AD for metanalysis"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)
synapseLogin()

knit2synapse::knitToFolderEntity(file = "./combineAllRNASeqData.Rmd", 
                                 parentId ="syn9622393",
                                 entityName = "All RNASeq data combined")
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file is

# Clear R console screen output
cat("\014")

# Load required libraries
library(CovariateAnalysis) # install from th1vairam repo using devtools::install_github('th1vairam/CovariateAnalysis@devtools')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

synapseLogin()

options(xtable.type="html")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn9622393';
activityName = 'Combine Data';
activityDescription = 'Combine all RNASeq and covariates data from all three brain banks';

thisFileName <- 'combineAllRNASeqData.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='metaAnal')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```

### Get filtered covariates/sample metadata from synapse
```{r covariates, fig.height=6, fig.width=10}
covariates.ids = c(ROSMAP = 'syn8456631',
                   MSBB = 'syn8484996',
                   MAYO = 'syn8466814')
all.used.ids = as.character(covariates.ids)
covariates = lapply(covariates.ids, downloadFile)

# Modify ROSMAP covariates
covariates$ROSMAP = covariates$ROSMAP %>%
  dplyr::mutate(Gender = factor(msex, levels = c(0,1), labels = c('0' = 'FEMALE', '1' = 'MALE')),
                RIN = RINcontinuous,
                AAD = age_death,
                PMI = pmi,
                BrainRegion = 'DLPFC',
                Study = 'ROSMAP') %>%
  dplyr::select(SampleID, Study, BrainRegion, Diagnosis, Batch, Gender, RIN, AAD, PMI, PCT_PF_READS_ALIGNED,
                PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES)

# Modify MSBB covariates
covariates$MSBB = covariates$MSBB %>%
  dplyr::mutate(Gender = factor(SEX, levels = c('F', 'M'), labels = c('F' = 'FEMALE', 'M' = 'MALE')),
                Batch = batch,
                PMI = PMI/60,
                AAD = AOD,
                Donor_ID = individualIdentifier,
                Study = 'MSBB') %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
  dplyr::select(SampleID, Study, BrainRegion, Diagnosis, Batch, Gender, RIN, AAD, PMI, PCT_PF_READS_ALIGNED,
                PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES)

# Modify MAYO covariates
covariates$MAYO = covariates$MAYO %>%
  dplyr::mutate(Gender = factor(Gender, levels = c('F', 'M'), labels = c('F' = 'FEMALE', 'M' = 'MALE')),
                Batch = FLOWCELL,
                AAD = AgeAtDeath,
                Study = factor(Source, 
                               levels = c('BannerSunHealth_TomBeach', 'MayoBrainBank_Dickson'),
                               labels = c('BannerSunHealth_TomBeach' = 'BANNER', 'MAYO'))) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
  dplyr::select(SampleID, Study, BrainRegion, Diagnosis, Batch, Gender, RIN, AAD, PMI, PCT_PF_READS_ALIGNED,
                PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES)

### All plots
covariates1 = rbindlist(covariates, use.names = T, fill = T) %>%
  dplyr::mutate(BrainRegion = factor(BrainRegion, levels = c('DLPFC', 'IFG', 'STG', 'FP', 'PHG', 'CER', 'TCX')))

plotlist = list()
p = ggplot(covariates1, aes(x = Diagnosis, y = PMI, color = BrainRegion)) + geom_boxplot()
p

p = ggplot(covariates1, aes(x = Diagnosis, y = AAD, color = BrainRegion)) + geom_boxplot()
p

p = ggplot(covariates1, aes(x = Diagnosis, y = RIN, color = BrainRegion)) + geom_boxplot()
p

p = ggplot(covariates1, aes(x = Diagnosis, y = PCT_INTERGENIC_BASES, color = BrainRegion)) + geom_boxplot()
p
```

### Get raw reprocessed counts from synapse
```{r counts}
# Download sample counts
counts.id = c(ROSMAP = 'syn8691134',
              MSBB = 'syn8691099',
              MAYO1 = 'syn8690799', 
              MAYO2 = 'syn8690904')
all.used.ids = c(all.used.ids, as.character(counts.id))
counts = lapply(counts.id, function(id){
  tmp = read.table(synGet(id)@filePath, header=T, sep='\t', check.names = F, row.names = 1)
})

covariates$MAYO1 = covariates$MAYO
covariates$MAYO2 = covariates$MAYO
counts = mapply(function(cnt, cov){
  cnt[,intersect(colnames(cnt), cov$SampleID)] %>%
    CovariateAnalysis::rownameToFirstColumn('Gene.ID')
}, counts, covariates[names(counts)], SIMPLIFY = F) %>%
  plyr::join_all(type = 'full')
```

### Store results in synapse
```{r synapse.store, include=FALSE, cache=FALSE}
# Store consolidated data in synapse
CODE = Folder(name = 'All RNASeq data combined', parentId = parentId)
CODE = synStore(CODE)

# Store covariates in synapse
write.table(covariates1, file = 'allRNASeqCovariates.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('allRNASeqCovariates.tsv', 
           name = 'All RNASeq covariates', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = all.used.ids,
               executed = thisFile,
               activityName = activityName,
               activityDescription = activityDescription)

# Store raw counts in synapse
write.table(logcpm, file = 'allReprocessedCounts.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('allReprocessedCounts.tsv', 
           name = 'All RNASeq reprocessed raw counts', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = all.used.ids,
               executed = thisFile,
               activityName = activityName,
               activityDescription = activityDescription)
```
### Source code in github
[Source Code](`r thisFile`)

### 2017-05-07 code/Rmd/combineAllRNASeqData.Rmd
Minor modifications

@@
 # Modify ROSMAP covariates
+rosmap.key = downloadFile('syn3382527') %>%
+  dplyr::select(projid, mrna_id) %>%
+  tidyr::separate(mrna_id, c('a','b','c'), sep = '_') %>%
+  tidyr::unite(SampleID, a,b, sep = '_') %>%
+  dplyr::rename(Donor_ID = projid)
+all.used.ids = c(all.used.ids, 'syn3382527')
+
 covariates$ROSMAP = covariates$ROSMAP %>%
+  left_join(rosmap.key) %>%
   dplyr::mutate(Gender = factor(msex, levels = c(0,1), labels = c('0' = 'FEMALE', '1' = 'MALE')),
@@
                 Study = 'ROSMAP') %>%
-  dplyr::select(SampleID, Study, BrainRegion, Diagnosis, Batch, Gender, RIN, AAD, PMI, PCT_PF_READS_ALIGNED,
+  dplyr::select(SampleID, Donor_ID, Study, BrainRegion, Diagnosis, Batch, Gender, RIN, AAD, PMI, PCT_PF_READS_ALIGNED,
                 PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES)
@@
   tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
-  dplyr::select(SampleID, Study, BrainRegion, Diagnosis, Batch, Gender, RIN, AAD, PMI, PCT_PF_READS_ALIGNED,
+  dplyr::select(SampleID, Donor_ID, Study, BrainRegion, Diagnosis, Batch, Gender, RIN, AAD, PMI, PCT_PF_READS_ALIGNED,
                 PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES)
@@
 p = ggplot(covariates1, aes(x = Diagnosis, y = PCT_INTERGENIC_BASES, color = BrainRegion)) + geom_boxplot()
+p
+
+p = ggplot(covariates1, aes(x = Diagnosis, y = PCT_INTRONIC_BASES, color = BrainRegion)) + geom_boxplot()
+p
+
+p = ggplot(covariates1, aes(x = Diagnosis, y = PCT_RIBOSOMAL_BASES, color = BrainRegion)) + geom_boxplot()
+p
+
+p = ggplot(covariates1, aes(x = Diagnosis, y = PCT_CODING_BASES, color = BrainRegion)) + geom_boxplot()
+p
+
+p = ggplot(covariates1, aes(x = Diagnosis, y = PCT_PF_READS_ALIGNED, color = BrainRegion)) + geom_boxplot()
 p
@@
 # Store raw counts in synapse
-write.table(logcpm, file = 'allReprocessedCounts.tsv', sep = '\t', row.names=F, quote=F)
+write.table(counts, file = 'allReprocessedCounts.tsv', sep = '\t', row.names=F, quote=F)
 OBJ = File('allReprocessedCounts.tsv', 

### 2017-05-08 code/Rmd/combineAllRNASeqData.Rmd
Added markdown to perform differential expression analysis in every brain region

@@
 # Modify ROSMAP covariates
-rosmap.key = downloadFile('syn3382527') %>%
-  dplyr::select(projid, mrna_id) %>%
-  tidyr::separate(mrna_id, c('a','b','c'), sep = '_') %>%
-  tidyr::unite(SampleID, a,b, sep = '_') %>%
-  dplyr::rename(Donor_ID = projid)
-all.used.ids = c(all.used.ids, 'syn3382527')
+rosmap.donorinfo <- downloadFile('syn4300313') %>%
+  dplyr::rename(SampleID = Sampleid, Donor_ID = projid) %>%
+  dplyr::select(SampleID, Donor_ID) %>%
+  unique %>%
+  dplyr::filter(!(SampleID == '492_120515' & is.na(Donor_ID)))
+all.used.ids = c(all.used.ids, 'syn4300313')
 
 covariates$ROSMAP = covariates$ROSMAP %>%
-  left_join(rosmap.key) %>%
+  left_join(rosmap.donorinfo) %>%
   dplyr::mutate(Gender = factor(msex, levels = c(0,1), labels = c('0' = 'FEMALE', '1' = 'MALE')),
@@
                 PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES) 
+covariates$ROSMAP$Donor_ID[is.na(covariates$ROSMAP$Donor_ID)] = paste0('NotAva',1:sum(is.na(covariates$ROSMAP$Donor_ID)))
 

### 2017-05-08 code/Rmd/regionLevelDiffExpAnlz.Rmd
Added markdown to perform differential expression analysis in every brain region

---
title: "Brain region specific differential expression analysis (all 7 regions/one region at a time)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)
synapseLogin()

knit2synapse::knitToFolderEntity(file = "./regionLevelDiffExpAnlz.Rmd", 
                                 parentId ="syn9622393",
                                 entityName = "Differential Expression Analysis (per brain region)")
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file is

# Clear R console screen output
cat("\014")

# Load required libraries
library(CovariateAnalysis) # install from th1vairam repo using devtools::install_github('th1vairam/CovariateAnalysis@devtools')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(limma)
library(edgeR)
library(cqn)

synapseLogin()

options(xtable.type="html")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn9622393';
activityName = 'Differential expression analysis';
activityDescription = 'Differential expression analysis in all 7 brain regions performed in a region specific manner';

thisFileName <- 'regionLevelDiffExpAnlz.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='metaAnal')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```

### Get merged covariates and raw counts from synapse
```{r download.data, fig.height=6, fig.width=10}
# Get covariates
covariates = downloadFile('syn9765937') %>%
  dplyr::filter(Diagnosis %in% c('AD', 'CONTROL')) %>%
  dplyr::mutate(Diagnosis.Gender = paste(Diagnosis, Gender, sep = '.'))
all.used.ids = 'syn9765625'

kable(table(covariates$BrainRegion, covariates$Diagnosis))

# Get raw counts
counts = downloadFile('syn9765938') %>%
  tidyr::separate(Gene.ID, c('ensembl_gene_id', 'position'), sep = '\\.')
rownames(counts) = counts$ensembl_gene_id
counts = counts[, covariates$SampleID]
all.used.ids = c(all.used.ids, 'syn9765629')

# Get gene specific parameters from synapse
gene.param = downloadFile('syn8449369')
all.used.ids = c(all.used.ids, 'syn8449369')

gene.len = dplyr::select(gene.param, ensembl_gene_id, gene.length) %>% 
  unique()
rownames(gene.len) = gene.len$ensembl_gene_id

gene.gc = dplyr::select(gene.param, ensembl_gene_id, percentage_gc_content) %>% 
  unique()
rownames(gene.gc) = gene.gc$ensembl_gene_id
```

### Filter genes
```{r filter.genes}
# Filter genes (remove genes that are expressed less than 1 cpm in 50% of samples per diagnosis per brain region)
lcpm = cpm(counts)

filtered.gene.ids = plyr::dlply(covariates, .(BrainRegion, Diagnosis), .fun = function(cov, logcpm){
  rownames(logcpm)[which(rowSums(logcpm[,cov$SampleID] >= 1)/dim(logcpm[,cov$SampleID])[2] >= 0.5)]
}, lcpm) %>%
  Reduce(c, .) %>%
  unique %>%
  setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous")) %>%
  intersect(gene.len$ensembl_gene_id[!is.na(gene.len$gene.length)]) %>%
  intersect(gene.gc$ensembl_gene_id[!is.na(gene.gc$percentage_gc_content)])

counts = counts[filtered.gene.ids, ]
```
Choosing `r length(filtered.gene.ids)` genes expressed at 1 cpm in 50% of samples in at least one brain region and diagnosis category for further analysis

```{r fxn}
performDifferentialExpression <- function(cov, cnt, contVar, varOfInterest, n){

  diffExp = foreach(i = 1:n, 
                    .combine = rbind,
                    .packages = c('limma','edgeR','cqn','dplyr','plyr','CovariateAnalysis')) %dopar% {

                    # Sub-sample data for balanced comparison
                    minSampSz = min(summary(cov[,contVar]), na.rm = T)
                    innerCov = rownameToFirstColumn(cov, 'SampleID')  %>%
                      plyr::ddply(.variables = contVar, .fun = function(x, minSampSz){
                        ind = sample(1:dim(x)[1], minSampSz)
                        return(x[ind,])
                      }, minSampSz) %>%
                      droplevels()
                    rownames(innerCov) = innerCov$SampleID

                    # Get design matrix
                    designMat = getDesignMatrix(innerCov[,varOfInterest], Intercept = F)$design
                    designMat = designMat[,linColumnFinder(designMat)$indepCols]

                    # Compute offset for gene length and gc content
                    cqn.gene.expression = cqn(cnt[,innerCov$SampleID],
                                              x = gene.gc[rownames(cnt), 'percentage_gc_content'],
                                              lengths = gene.len[rownames(cnt), 'gene.length'],
                                              lengthMethod = "smooth",
                                              verbose = FALSE)
                    cqn.gene.expression$E = cqn.gene.expression$y+cqn.gene.expression$offset

                    # Identify outlier expression and assign NA values
                    log.mat = apply(cqn.gene.expression$E, 1, function(x){
                      mn = mean(x, na.rm = T)
                      std.dev = sd(x, na.rm = T)
                      return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
                    }) %>% t
                    cqn.gene.expression$E[log.mat] = NA

                    # Estimate voom weights for dispersion control
                    voom.weights = voom(cnt[,innerCov$SampleID], design = designMat, plot=F, na.rm = T)

                    # Fit linear model using new weights and new design
                    baseModel = lmFit(cqn.gene.expression$E,
                                      design = designMat,
                                      weights = voom.weights$weights)

                    # Perform differential expression
                    contrasts = levels(cov[,contVar]) %>%
                      paste(contVar,.,sep = '') %>%
                      combn(2) %>%
                      apply(2,paste, collapse = '-') %>%
                      makeContrasts(contrasts = ., levels = colnames(baseModel$coefficients))
                    baseModel.cntr =  contrasts.fit(baseModel, contrasts)
                    baseModel.cntr = eBayes(baseModel.cntr)
                    innerDiffExp = lapply(1:dim(contrasts)[2], function(i, baseModel.cntr){
                      topTable(baseModel.cntr, coef = i, number = 50000, confint = T) %>%
                        rownameToFirstColumn('ensembl_gene_id')
                    }, baseModel.cntr)
                    names(innerDiffExp) = colnames(contrasts)
                    innerDiffExp = innerDiffExp %>%
                      rbindlist(use.names = T, fill = T, idcol = 'Comparison') %>%
                      dplyr::mutate(iter = i,
                                    Comparison = gsub(contVar,'',Comparison))

                    return(innerDiffExp)
                    }

  diffExp = diffExp %>%
    dplyr::group_by(Comparison, ensembl_gene_id) %>%
    tidyr::gather(feature, value, -Comparison, -ensembl_gene_id) %>%
    dplyr::group_by(Comparison, ensembl_gene_id, feature) %>%
    dplyr::summarise(value = median(value, na.rm = T)) %>%
    tidyr::spread(feature, value)

  return(diffExp)
}
```

### Perform differential expression (per brain region)
```{r diff.exp}
all.results = plyr::dlply(covariates, .(BrainRegion), .fun = function(cov, cnt){
  n = 5
  # Filter counts and covariates
  cnt = cnt[,cov$SampleID]

  # Formulate covariates matrix
  FactorCov = c('Diagnosis', 'Batch', 'Gender')
  ContCov = c('RIN', 'AAD', 'PMI',  'PCT_PF_READS_ALIGNED', 'PCT_CODING_BASES', 
              'PCT_INTERGENIC_BASES', 'PCT_INTRONIC_BASES', 'PCT_RIBOSOMAL_BASES')
  cov[,FactorCov] = lapply(cov[,FactorCov], as.factor)
  cov[,ContCov] = lapply(cov[,ContCov], as.numeric)
  rownames(cov) = cov$SampleID
  cov = cov[,c(FactorCov, ContCov)]
  assocStat = getAssociationStatistics(cov)

  # Perform differential expression between case-control
  varOfInterest = c('Diagnosis', 'Batch', 'Gender', 'RIN', 'AAD', 'PMI', 
                    'PCT_PF_READS_ALIGNED', 'PCT_INTERGENIC_BASES', 'PCT_INTRONIC_BASES', 
                    'PCT_RIBOSOMAL_BASES')
  contVar = 'Diagnosis'

  baseDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, n)

  # Perform differential expression between case-control in a gender specific manner
  cov$Diagnosis.Gender = factor(paste(cov$Diagnosis, cov$Gender, sep = '.'))
  varOfInterest = c('Diagnosis.Gender', 'Batch', 'RIN', 'AAD', 'PMI', 
                    'PCT_PF_READS_ALIGNED', 'PCT_INTERGENIC_BASES', 'PCT_INTRONIC_BASES', 
                    'PCT_RIBOSOMAL_BASES')
  contVar = 'Diagnosis.Gender'

  genderDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, n)

  return(rbindlist(list(baseDiffExp, genderDiffExp), use.names = T, fill = T))
}, counts)

# Display results
writeLines('Number of differentially expressed genes between AD-CONTROL at an FDR of 0.05 and fold change of 1.2 are')
sapply(all.results, function(x){
  sum(x$adj.P.Val <= 0.05 & abs(x$logFC) >= log2(1.2) & x$Comparison == 'AD-CONTROL', na.rm = T)
}) %>% rownameToFirstColumn('Region') %>% dplyr::rename(AD_vs_CONTROL = DF) %>%
  kable(row.names=F)

writeLines('Number of differentially expressed genes between AD.FEMALE-CONTROL.FEMALE at an FDR of 0.05 and fold change of 1.2 are')
sapply(all.results, function(x){
  sum(x$adj.P.Val <= 0.05 & abs(x$logFC) >= log2(1.2) & x$Comparison == 'AD.FEMALE-CONTROL.FEMALE', na.rm = T)
}) %>% rownameToFirstColumn('Region') %>% dplyr::rename(AD.FEMALE_vs_CONTROL.FEMALE = DF) %>%
  kable(row.names=F)

writeLines('Number of differentially expressed genes between AD.MALE-CONTROL.MALE at an FDR of 0.05 and fold change of 1.2 are')
sapply(all.results, function(x){
  sum(x$adj.P.Val <= 0.05 & abs(x$logFC) >= log2(1.2) & x$Comparison == 'AD.MALE-CONTROL.MALE', na.rm = T)
}) %>% rownameToFirstColumn('Region') %>% dplyr::rename(AD.MALE_vs_CONTROL.MALE = DF) %>%
  kable(row.names=F)

writeLines('Number of differentially expressed genes between AD.FEMALE-AD.MALE at an FDR of 0.05 and fold change of 1.2 are')
sapply(all.results, function(x){
  sum(x$adj.P.Val <= 0.05 & abs(x$logFC) >= log2(1.2) & x$Comparison == 'AD.FEMALE-AD.MALE', na.rm = T)
}) %>% rownameToFirstColumn('Region') %>% dplyr::rename(AD.FEMALE_vs_AD.MALE = DF) %>%
  kable(row.names=F)

writeLines('Number of differentially expressed genes between CONTROL.FEMALE-CONTROL.MALE at an FDR of 0.05 and fold change of 1.2 are')
sapply(all.results, function(x){
  sum(x$adj.P.Val <= 0.05 & abs(x$logFC) >= log2(1.2) & x$Comparison == 'CONTROL.FEMALE-CONTROL.MALE', na.rm = T)
}) %>% rownameToFirstColumn('Region') %>% dplyr::rename(CONTROL.FEMALE_vs_CONTROL.MALE = DF) %>%
  kable(row.names=F)
```

### Store results in synapse
```{r synapse.store, include=FALSE, cache=FALSE}
# Store consolidated data in synapse
CODE = Folder(name = 'Differential Expression Analysis (per brain region)', parentId = parentId)
CODE = synStore(CODE)

# Store results in synapse
save(list = 'all.results', file = 'AllDifferentialExpressionResults.RData')
OBJ = File('AllDifferentialExpressionResults.RData', 
           name = 'All Differential Expression Results', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = all.used.ids,
               executed = thisFile,
               activityName = activityName,
               activityDescription = activityDescription)
```
### Source code in github
[Source Code](`r thisFile`)

### 2017-05-08 code/Rmd/regionLevelDiffExpAnlz.Rmd

@@
 synapseLogin()
+
+library(foreach)
+library(doParallel)
+
+cl = makeCluster(detectCores()-2)
+registerDoParallel(cl)
 
@@
 ```{r fxn}
-performDifferentialExpression <- function(cov, cnt, contVar, varOfInterest, n){
+performDifferentialExpression <- function(cov, cnt, contVar, varOfInterest, n, gene.len, gene.gc){
 
@@
                     .combine = rbind,
-                    .packages = c('limma','edgeR','cqn','dplyr','plyr','CovariateAnalysis')) %dopar% {
+                    .packages = c('limma','edgeR','cqn','dplyr','plyr','CovariateAnalysis','data.table'),
+                    .export = c('gene.len', 'gene.gc')) %dopar% {
 
@@
 ```{r diff.exp}
-all.results = plyr::dlply(covariates, .(BrainRegion), .fun = function(cov, cnt){
-  n = 5
+all.results = plyr::dlply(covariates, .(BrainRegion), .fun = function(cov, cnt, gene.len, gene.gc){
+  n = 100
   # Filter counts and covariates
@@
 
-  baseDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, n)
+  baseDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, n, gene.len, gene.gc)
 
@@
 
-  genderDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, n)
+  genderDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, n, gene.len, gene.gc)
 
   return(rbindlist(list(baseDiffExp, genderDiffExp), use.names = T, fill = T))
-}, counts)
+}, counts, gene.len, gene.gc)
 

### 2017-05-10 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Added markdown to select covariates based on weighted random effect based modeling with BIC criteria

--- 2017-04-26 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
+knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd",
                                  parentId ="syn7501880",
-                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
+                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)')
 ```
@@
 library(ggplot2)
-library(reshape2)
 library(limma)
-library(gplots)
-library(WGCNA)
 library(psych)
 library(edgeR)
+library(cqn)
+library(lme4)
 library(biomaRt)
-library(RColorBrewer)
-library(cqn)
 
@@
 
-thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd'
+thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd'
 
@@
 ```{r covariates.clustering}
-#"RibozeroBatch", "FlowcellBatch",
-FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx.Tissue")
+FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
 ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
@@
 ### Filter genes
-Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
+Remove genes that have less than 5 cpm counts in at least 100% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
 ```{r cpmnormalisation}
@@
   processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID %>% droplevels()],
-                                                   MIN_GENE_CPM=1, 
-                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
+                                                   MIN_GENE_CPM=5, 
+                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=1)
   processed.counts$filteredExprMatrix$genes
@@
 
-### Normalisation (with NULL)
-Initial normalisation usign cqn and dispersion control with voom (with NULL design)
-```{r initial.voom.normalisation, fig.height=5, fig.width=5}
+### Winsorise counts (Gene outliers in samples)
+```{r winsorise.data}
+LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)
+
+# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
+LOG.CPM = apply(LOG.CPM, 1, function(x){
+  mn = mean(x, na.rm = T)
+  std.dev = sd(x, na.rm = T)
+
+  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
+  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
+
+  x[x < (mn-3*std.dev)] = x[ind.low]
+  x[x > (mn+3*std.dev)] = x[ind.high]
+  return(x)
+}) %>% t
+
+# Back calculate counts for further analysis
+LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
+NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
+```
+
+### Library Normalisation
+Initial library normalisation usign cqn
+```{r cqn}
 # Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                           x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
@@
                           verbose = FALSE)
-CQN.GENE_EXPRESSION$E.NA = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
 
 # Set gene counts in specific samples that are deviating 3 sd from other samples to 0
-log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
+log.mat = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
   mn = mean(x, na.rm = T)
@@
 }) %>% t
-PROCESSED_COUNTS$filteredExprMatrix$counts.NA = PROCESSED_COUNTS$filteredExprMatrix$counts
-PROCESSED_COUNTS$filteredExprMatrix$counts.NA[log.mat] = NA
-PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0
+NEW.COUNTS[log.mat] = NA
+CQN.GENE_EXPRESSION$E[log.mat] = NA
+```
 
-CQN.GENE_EXPRESSION$E.NA[log.mat] = NA
-CQN.GENE_EXPRESSION$E.zero = CQN.GENE_EXPRESSION$E.NA
-CQN.GENE_EXPRESSION$E.zero[log.mat] = 0
-```
+### Co-expression distribution
 Coexpression of genes 
 ```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(CQN.GENE_EXPRESSION$E.zero))
+tmp = CQN.GENE_EXPRESSION$E
+tmp[is.na(tmp)] = 0
+cr = cor(t(tmp))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
 ```
-### Outlier analysis
-Clustering of voom adjusted data (with NULL design)
+
+### Sample clustering
 ```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
 # Find principal components of expression to plot
-PC <- prcomp(CQN.GENE_EXPRESSION$E.zero, scale.=T, center = T)
+tmp = CQN.GENE_EXPRESSION$E
+tmp[is.na(tmp)] = 0
+PC <- prcomp(tmp, scale.=T, center = T)
 
@@
 p
+```
+Tree based clustering of samples
+```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
+# Eucledian tree based analysis
+COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
+COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
 
+tmp = CQN.GENE_EXPRESSION$E
+tmp[is.na(tmp)] = 0
+tree = hclust(as.dist(t(tmp)))
+cols = WGCNA::labels2colors(COVARIATES.tmp);
+
+WGCNA::plotDendroAndColors(tree, 
+                           colors = cols, 
+                           dendroLabels = FALSE, 
+                           abHeight = 0.80, 
+                           main = "Sample dendrogram",
+                           groupLabels = colnames(COVARIATES.tmp))
+```
+```{r temp, include = F}
+dev.off()
+gc()
+```
+
+### Distribution of samples (log cpm)
+```{r lcpm.dist}
 # Plot abberent distribution of logcpm counts
-tmp1 = (CQN.GENE_EXPRESSION$E.NA) %>%
+tmp1 = (CQN.GENE_EXPRESSION$E) %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
@@
 ```
-Tree based classification of samples
-```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
-# Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
-COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
-
-tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E.zero)))
-cols = WGCNA::labels2colors(COVARIATES.tmp);
-
-WGCNA::plotDendroAndColors(tree, 
-                           colors = cols, 
-                           dendroLabels = FALSE, 
-                           abHeight = 0.80, 
-                           main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp))
-```
-
-```{r temp, include = F}
-dev.off()
-gc()
-```
 
@@
 # Find correlation between PC's of gene expression with covariates
-preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E.zero, 
+tmp = CQN.GENE_EXPRESSION$E
+tmp[is.na(tmp)] = 0
+preAdjustedSigCovars = runPCAandPlotCorrelations(tmp, 
                                                  COVARIATES,
@@
 
-### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. 
-1. "Dx.Tissue" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue)
-2. Adding Instituion and Gender a priori to variable selection as they are correlated with Dx.Tissue
+### Model Identification
+```{r fxn}
+# Function to fit a mixed effect model and return BIC
+fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){
+
+  # Get design matrix
+  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
+  designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]
+
+  # Estimate voom weights
+  cnt[is.na(cnt)] = 0
+  voomWeights = limma::voom(cnt, design = designMat, plot=F)$weights
+
+  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
+                              .combine = rbind,
+                              .packages = c('lme4', 'dplyr'),
+                              .export = c('voomWeights')) %dopar% {
+
+                                # Perform weighted mixed linear models for every gene
+                                expr = data.frame(Gx = lcpm[i,])
+                                data = cbind(expr,
+                                             designMat[rownames(expr),],
+                                             Individual_ID = groupingVar)
+                                formula = paste(setdiff(colnames(data), c('Gx', 'Individual_ID')), collapse = '+') %>%
+                                  paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')
+
+                                weights = voomWeights[i,]
+                                BIC = NA
+                                tryCatch({
+                                  mdl = lmer(formula = formula, 
+                                             data = data, 
+                                             weights = weights)
+                                  BIC = BIC(mdl)
+                                }, error = function(e){
+                                  BIC = NA
+                                })
+                              return(data.frame(ensembl_gene_id = rownames(lcpm)[i],
+                                                BIC = BIC))
+                            }
+  return(modelBIC)
+  }
+
+# Function to select covariates in a step-wise regression fashion
+selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){
+
+  varsSelected = c()
+  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
+           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
+  while(length(covarsToSelect) != 0){
+    allModelBIC = foreach(i = 1:length(covarsToSelect),
+                          .combine = c,
+                          .packages = pkgs,
+                          .export = c('fitMixedEffectModel')) %dopar% {
+
+                            # Fit a mixed effect linear model and get all BIC
+                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
+                                                           covars[colnames(cnt),'Individual_ID'], 
+                                                           lcpm, 
+                                                           cnt, 
+                                                           covars) %>%
+                            plyr::rename(c('BIC' = covarsToSelect[i]))
+
+                            return(list(modelBIC))
+                          }
+    allModelBIC = plyr::join_all(allModelBIC)
+
+    # Find number of genes that are changed based on BIC
+    nChange = allModelBIC %>%
+      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
+      dplyr::left_join(baseModelBIC) %>%
+      dplyr::mutate(diff = baseModel - Value) %>%
+      dplyr::group_by(ModelName) %>%
+      dplyr::summarise(better = sum(diff >= 5), 
+                       worse = sum(diff <= -5)) %>%
+      dplyr::mutate(nchange = better - worse)
+
+    kable(nChange)
+
+    # Select variable
+    tmp = nChange %>%
+      dplyr::filter(nchange > round(dim(cnt)[1]*0.05)) %>%
+      dplyr::top_n(1, nchange)
+
+    if (dim(tmp)[1] == 0){
+      covarsToSelect = c()
+    } else {
+      baseModelBIC = allModelBIC %>%
+        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
+      colnames(baseModelBIC)[2] = 'baseModel'
+      varsSelected = c(varsSelected, tmp$ModelName)
+      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
+      print(paste('Variable chosen to be added:',tmp$ModelName))
+    }
+  }
+
+  return(varsSelected)
+}
+```
+#### Phase I (clinical, ancestry and sample specific technical variables)
+```{r phase1}
+# Fit a base model with the following covariates
+adjust.covars = c('Dx.Tissue', 'Institution', 'Gender')
+baseModelBIC = fitMixedEffectModel(adjust.covars, 
+                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
+                                   CQN.GENE_EXPRESSION$E, 
+                                   NEW.COUNTS, 
+                                   COVARIATES) %>%
+  dplyr::rename(baseModel = BIC)
+
+# Iteratively identify covariates in phase I
+phase1.covars = c('PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
+vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
+                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
+                                     NEW.COUNTS, baseModelBIC)
+
+adjust.covars = c(adjust.covars, vars.selected)
+```
+Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`
+
+#### Phase II (batch effects)
+```{r phase2}
+# Fit a base model with the following covariates
+baseModelBIC = fitMixedEffectModel(adjust.covars, 
+                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
+                                   CQN.GENE_EXPRESSION$E, 
+                                   NEW.COUNTS, 
+                                   COVARIATES) %>%
+  dplyr::rename(baseModel = BIC)
+
+# Iteratively identify covariates in phase I
+phase2.covars = c('LibraryBatch', 'RibozeroBatch', 'FlowcellBatch')
+vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
+                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
+                                     NEW.COUNTS, baseModelBIC)
+
+adjust.covars = c(adjust.covars, vars.selected)
+```
+Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`
+
+#### Phase III (RNASeq alignment specific covariates)
+```{r phase3}
+# Fit a base model with the following covariates
+baseModelBIC = fitMixedEffectModel(adjust.covars, 
+                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
+                                   CQN.GENE_EXPRESSION$E, 
+                                   NEW.COUNTS, 
+                                   COVARIATES) %>%
+  dplyr::rename(baseModel = BIC)
+
+# Iteratively identify covariates in phase I
+phase3.covars = c("MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate", "GenesDetected",
+                  "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", "PercentAligned")
+vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
+                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
+                                     NEW.COUNTS, baseModelBIC)
+
+adjust.covars = c(adjust.covars, vars.selected)
+```
+Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`
+
+### Normalisation
+1. Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
+2. Individual_ID is chosen as random effect
 ```{r iterative.normalisation, results='asis'}
-# Primary variable of interest
-postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender');
-
-# Assign residual covariates
-residualSigCovars = preAdjustedSigCovars
-residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, 'Individual_ID'))
-covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
-
-loopCount = 0 
-while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
 writeLines(paste('Using following covariates in the model:',
-                   paste(postAdjustCovars, collapse=', '),
+                 paste(adjust.covars, collapse=', '),
                  'as fixed effects and Individual_ID is chosen as random effect'))
@@
 # Post adjusted design matrix
-  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
+DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
 DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
@@
 # Estimate voom weights
-  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)
-  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E.NA
+tmp = NEW.COUNTS
+tmp[is.na(tmp)] = 0
+VOOM.GENE_EXPRESSION = voom(tmp, design=DM1$design, plot=F)
+
+# Estimate consensus correlation coefficient
 correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
@@
 # Re-calculate voom weights with correlation of random effects
-  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+VOOM.GENE_EXPRESSION = voom(tmp, 
                             design=DM1$design, plot=F,
@@
 # Fit linear model using new weights and new design
-  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E.NA
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
 ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
@@
 # Residuals after normalisation
-  RESIDUAL.GENE_EXPRESSION = list()
-  RESIDUAL.GENE_EXPRESSION$E.NA = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E.NA)
-  RESIDUAL.GENE_EXPRESSION$E.zero = RESIDUAL.GENE_EXPRESSION$E.NA
-  RESIDUAL.GENE_EXPRESSION$E.zero[is.na(RESIDUAL.GENE_EXPRESSION$E.NA)] = 0
-
-  # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, 'Individual_ID'))
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
 
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-  tmp.expr = RESIDUAL.GENE_EXPRESSION$E.zero
-  tmp.expr = tmp.expr[apply(tmp.expr,1,var) !=0,]
-  residualSigCovars = runPCAandPlotCorrelations(tmp.expr, 
-                                                COVARIATES[, residCovars, drop=F], 
+tmp = RESIDUAL.GENE_EXPRESSION
+tmp[is.na(tmp)] = 0
+residualSigCovars = runPCAandPlotCorrelations(tmp, COVARIATES,
                                               'adjusted design(voom-normalized)',
                                               isKeyPlot=TRUE)
-
-  # Add postadjusted covariates (if any)
-  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
-                                                                          c(postAdjustCovars, 'Individual_ID'))]
-
-  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
-  loopCount = loopCount + 1
-}
-
-modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
-                  'as fixed effects')
-
-tmp <- paste('Using following covariates in the final model:', modelStr)
 ```
-`r tmp`
 
@@
 ```{r residualSigCovars.manual, fig.width=12, fig.height=8}
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-tmp.expr = RESIDUAL.GENE_EXPRESSION$E.zero
-tmp.expr = tmp.expr[apply(tmp.expr,1,var) !=0,]
-residualSigCovars = runPCAandPlotCorrelations(tmp.expr, 
-                                              COVARIATES, 
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
-
 residualSigCovars[["PC_res"]][[2]]$plotData
 ```
+
 ### Residual calculation
@@
 varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
-RESIDUAL.GENE_EXPRESSION$E.NA = RESIDUAL.GENE_EXPRESSION$E.NA + 
+RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
   ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
 ```
-Coexpression of genes 
+
+### Coexpression of residual expression 
 ```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.GENE_EXPRESSION$E.zero))
+tmp = RESIDUAL.GENE_EXPRESSION
+tmp[is.na(tmp)] = 0
+cr = cor(t(tmp))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION$E.zero, scale.=T, center = T)
+PC <- prcomp(tmp, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION$E.zero)))
+tree = hclust(as.dist(t(tmp)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 ```{r diffExp, fig.height=10, fig.width=15}
-# Post adjusted design matrix
-DESIGN = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
-
-# Estimate voom weights
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
-correlation <- parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Individual_ID)
-
-# Re-calculate voom weights with correlation of random effects
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                    design=DESIGN$design, plot=F,
-                    block = COVARIATES$Individual_ID, 
-                    correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
-FIT = lmFit(VOOM.WEIGHTS)
-
 # Fit contrast
@@
                                      "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC"),
-                         levels = colnames(FIT$coefficients))
-FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+                         levels = colnames(ADJUSTED.FIT$coefficients))
+FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
 FIT.CONTR = eBayes(FIT.CONTR)
@@
 
-  normalizationType	= 'CQN'
+  normalizationType	= 'CQNandMS'
 )
@@
 # Code
-CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
+CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)", parentId = parentId)
 annotations(CODE) = all.annotations
@@
 # Store filtered counts
-PROCESSED_COUNTS$filteredExprMatrix$counts.NA %>%
+PROCESSED_COUNTS$filteredExprMatrix$counts %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
@@
 # Store logCPM
-CQN.GENE_EXPRESSION$E.NA %>%
+CQN.GENE_EXPRESSION$E %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
@@
 # Store residual gene expression for network analysis
-RESIDUAL.GENE_EXPRESSION$E.NA %>%
+RESIDUAL.GENE_EXPRESSION %>%
   rownameToFirstColumn('ensembl_gene_id') %>%

### 2017-05-10 code/Rmd/MetaAnalysis.Rmd
Temp commit

@@
 
+### Get curated gene list
+```{r curate.genelist}
+# Download enrichr gene sets from synapse
+GL_OBJ = synGet('syn4867851')
+ALL_USED_IDs = c(ALL_USED_IDs, GL_OBJ$properties$id)
+load(GL_OBJ@filePath)
+
+gsets = c("GO_Biological_Process", "Reactome_2015", "WikiPathways_2015")
+GeneSets.Enrichr = GeneSets[gsets]
+
+# Download AD related gene sets from synapse
+GL_OBJ = synGet('syn4893059');
+ALL_USED_IDs = c(ALL_USED_IDs, GL_OBJ$properties$id)
+load(GL_OBJ@filePath)
+
+GeneSets.CM = list(CellTypeMarkers = GeneSets[grep('Zhang', names(GeneSets))],
+                   AD = GeneSets['AD:GeneticLoci'])
+GeneSets = c(GeneSets.Enrichr, GeneSets.CM) # 
+```
+
 ### Get covariates/sample metadata from synapse
@@
                    MAYO = 'syn8466814')
+all.used.ids = covariates.ids
 covariates = lapply(covariates.ids, downloadFile)
 
-covariates$ROSMAP = dplyr::filter(covariates$ROSMAP, cogdx %in% c(1,4)) %>%
-  dplyr::mutate(Dx = factor(cogdx, levels = c(1,4), labels = c('1' = 'NCI', '4' = 'AD')),
-                Study = 'ROSMAP',
-                BrainRegion = 'DLPFC') %>%
-  dplyr::select(SampleID, Study, BrainRegion, Dx)
+covariates$ROSMAP$BrainRegion = 'DLPFC'
+covariates$MSBB = covariates$MSBB %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
+covariates$MAYO = covariates$MAYO %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
 
-covariates$MSBB = covariates$MSBB %>%
-  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
-  dplyr::filter(Diagnosis %in% c('SD', 'ND')) %>%
-  dplyr::mutate(Study = 'MSSM',
-                Dx = factor(Diagnosis, levels = c('ND','SD')),
-                BrainRegion = factor(BrainRegion, labels = c('FP', 'STG', 'PHG', 'IFG'))) %>%
-  dplyr::select(SampleID, Study, BrainRegion, Dx)
+covariates = covariates %>% 
+  rbindlist(use.names = T, fill = T, idcol = 'Study') %>%
+  dplyr::select(SampleID, Diagnosis, BrainRegion)
 
-covariates$MAYO = dplyr::filter(covariates$MAYO, BrainRegion.Diagnosis %in% c('TCX.AD', 'TCX.Control', 'CER.AD', 'CER.Control')) %>%
-  dplyr::select(SampleID, BrainRegion.Diagnosis) %>%
-  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Dx')) %>%
-  dplyr::mutate(Study = 'MAYO',
-                Dx = factor(Dx, levels = c('Control','AD')),
-                BrainRegion = factor(BrainRegion, levels = c('CER', 'TCX'))) %>%
-  dplyr::select(SampleID, Study, BrainRegion, Dx)
+kable(table(covariates$Diagnosis, covariates$BrainRegion))
 ```
@@
 ### Get differential expression data from synapse
-```{r diffexp}
+```{r diffexp, fig.width=10, fig.height=15}
 diff.exp.ids = c(ROSMAP = 'syn8456721',
@@
                  MAYO = 'syn8468023')
+all.used.ids = c(all.used.ids, diff.exp.ids)
 diff.exp = lapply(diff.exp.ids, downloadFile)
 
-diff.exp$ROSMAP = filter(diff.exp$ROSMAP, Comparison == 'cogdx1-cogdx4')
-diff.exp$MSBB = filter(diff.exp$MSBB, Comparison == 'ND-SD')
-diff.exp$MAYO = filter(diff.exp$MAYO, Comparison == 'Control-AD')
+diff.exp$ROSMAP = diff.exp$ROSMAP %>%
+  setnames(c('Region'), c('BrainRegion'))
 
-tmp = diff.exp %>% 
-  rbindlist(use.names = T, fill = T) 
+tmp = rbindlist(diff.exp, use.names = T, fill = T) %>%
+  dplyr::filter(Comparison == 'CONTROL-AD')
+
 tmp$Direction = 'NONE'
-tmp$Direction[tmp$logFC >= log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
-tmp$Direction[tmp$logFC <= -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'
+tmp$Direction[tmp$logFC > log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
+tmp$Direction[tmp$logFC < -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'
 
-writeLines('Number of differentially expressed genes at an FDR of 0.05 and fold change of 1.2 are')
+writeLines('Number of differentially expressed genes at an FDR of 0.05')
 tmp %>%
-  group_by(Study, Region, Direction, Comparison) %>%
+  group_by(Study, BrainRegion, Direction, Comparison) %>%
   summarise(count = n()) %>%
@@
   p = p + scale_color_manual(values = c('green','grey','red'))
-  p = p + facet_grid(Study~.+Region) + theme(legend.position = 'top')
+  p = p + facet_grid(Study~.+BrainRegion) + theme(legend.position = 'top')
   pl[[region]] = p
@@
 
+### Pathway enrichment analysis (individual brain regions)
+```{r pathway.enrich}
+# setsToTest = rbindlist(diff.exp, use.names = T, fill = T) %>%
+#   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
+#   dlply(.(BrainRegion), .fun = function(x){
+#     setdiff(unique(x$hgnc_symbol),c(''))
+#   })
+
+setsToTest = rbindlist(diff.exp, use.names = T, fill = T) %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) > log2(1.2)) %>%
+  dplyr::mutate(Direction = sign(logFC),
+                Direction = factor(Direction, levels = c(-1,1),
+                                   labels = c('-1' = 'DOWN', '1' = 'UP'))) %>%
+  dlply(.(BrainRegion, Direction), .fun = function(x){
+    setdiff(unique(x$hgnc_symbol),c(''))
+  })
+
+# Filter gene list
+GeneSets = mapply(function(x,y){
+  names(x) = paste(y,names(x),sep = ':')
+  return(x)
+},GeneSets, names(GeneSets), SIMPLIFY = F) %>%
+  Reduce(c,.)
+GeneSets = GeneSets[setdiff(seq(1,length(GeneSets)), c(grep('Mus musculus', names(GeneSets))))]
+
+GeneSets = lapply(GeneSets, function(x, backgroundGenes){
+  x = intersect(x,backgroundGenes)
+}, backgroundGenes)
+GeneSets = GeneSets[sapply(GeneSets, length) >= 15 & sapply(GeneSets, length) <= 1000]
+
+enrichResults = plyr::ldply(setsToTest, .fun = function(genesInGeneSet, geneSets, backgroundGenes){
+  plyr::ldply(geneSets, .fun = fisherEnrichment, genesInGeneSet, backgroundGenes, .parallel = T) %>%
+    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
+}, GeneSets, backgroundGenes, 
+.parallel = F, .progress = 'text', .paropts = list(.packages = c('plyr','dplyr','CovariateAnalysis')))
+
+tmp = dplyr::select(enrichResults, BrainRegion, .id, fdr) %>%
+  dplyr::filter(fdr <= 0.05) %>%
+  tidyr::spread(BrainRegion, fdr)
+
+# Store logcpm in synapse
+write.table(tmp, file = 'enrich.tsv', sep = '\t', row.names=F, quote=F)
+OBJ = File('enrich.tsv', 
+           name = 'Pathway Enrichment Results (AD and Control)', 
+           parentId = CODE$properties$id)
+OBJ = synStore(OBJ,
+               used = as.character(c(covariates.id)),
+               executed = thisFile,
+               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
+```
+
 ### Get cell type fraction from synapse
-```{r celltype.frac}
+```{r celltype.frac, eval = FALSE}
 celltype.fraction.ids = c(ROSMAP = 'syn8124520',
@@
 
-### Pathway enrichment analysis
-```{r pathway.enrich}
-setsToTest = rbindlist(diff.exp, use.names = T, fill = T) %>%
-  filter(adj.P.Val <= 0.05, Direction != 'NONE') %>%
-  dlply(.(Region, Direction), .fun = function(x){
-    setdiff(unique(x$hgnc_symbol),c(''))
-  })
-
-tmp1 = rankprod.pfp$pfp[intersect(intersect(tmp$fb.pval, tmp$min.pval), tmp$rankprod.pval),] <= 0.05
-setsToTest$META.UP = rownames(tmp1)[which(tmp1[,1])]
-setsToTest$META.DOWN = rownames(tmp1)[which(tmp1[,2])]
-
-# Convert ensembl id to hgnc_symbols
-setsToTest = lapply(setsToTest, function(x, Ensemble2HGNC){
-  Ensemble2HGNC$hgnc_symbol[Ensemble2HGNC$ensembl_gene_id %in% x] %>%
-    unique()
-}, Ensemble2HGNC)
-
-# Get Background genes
-backgroundGenes = unique(rbindlist(diff.exp, use.names = T, fill = T) %>% 
-                           dplyr::select(hgnc_symbol) %>%
-                           unlist)
-
-############################################################################################################
-# Function to perform Fishers enrichment analysis
-fisherEnrichment <- function(genesInSignificantSet, # A character vector of differentially expressed or some significant genes to test
-                             genesInGeneSet, # A character vector of genes in gene set like GO annotations, pathways etc...
-                             genesInBackground # Background genes that are 
-){
-  genesInSignificantSet = intersect(genesInSignificantSet, genesInBackground) # back ground filtering
-  genesInNonSignificantSet = base::setdiff(genesInBackground, genesInSignificantSet)
-  genesInGeneSet = intersect(genesInGeneSet, genesInBackground) # back ground filtering
-  genesOutGeneSet = base::setdiff(genesInBackground,genesInGeneSet)
-
-  pval = fisher.test(
-    matrix(c(length(intersect(genesInGeneSet, genesInSignificantSet)),             
-             length(intersect(genesInGeneSet, genesInNonSignificantSet)),
-             length(intersect(genesOutGeneSet, genesInSignificantSet)),
-             length(intersect(genesOutGeneSet, genesInNonSignificantSet))), 
-           nrow=2, ncol=2), alternative="greater")
-  OR = (length(intersect(genesInGeneSet, genesInSignificantSet)) * length(intersect(genesOutGeneSet, genesInNonSignificantSet))) / (length(intersect(genesInGeneSet, genesInNonSignificantSet)) * length(intersect(genesOutGeneSet, genesInSignificantSet))) %>% as.numeric()
-  return(data.frame(pval = pval$p.value,
-                    ngenes = length(genesInGeneSet),
-                    noverlap = length(intersect(genesInGeneSet, genesInSignificantSet)),
-                    Odds.Ratio = OR,
-                    Genes = paste(intersect(genesInGeneSet, genesInSignificantSet), collapse = '|')))}
-############################################################################################################
-
-############################################################################################################
-#### Get gene sets ####
-# Download enrichr gene sets from synapse
-GL_OBJ = synGet('syn4867851')
-ALL_USED_IDs = c(ALL_USED_IDs, GL_OBJ$properties$id)
-load(GL_OBJ@filePath)
-
-gsets = c("BioCarta_2015", "GO_Biological_Process", "Reactome_2015", "TargetScan_microRNA", "WikiPathways_2015")
-GeneSets.Enrichr = GeneSets[gsets]
-
-# Download AD related gene sets from synapse
-GL_OBJ = synGet('syn4893059');
-ALL_USED_IDs = c(ALL_USED_IDs, GL_OBJ$properties$id)
-load(GL_OBJ@filePath)
-
-GeneSets.CM = list(CellTypeMarkers = GeneSets[grep('Zhang', names(GeneSets))],
-                   AD = GeneSets['AD:GeneticLoci'])
-GeneSets = c(GeneSets.Enrichr, GeneSets.CM)
-
-# Filter gene list
-GeneSets = filterGeneSets(GeneSets, backgroundGenes, minSize = 3, maxSize = 5000)
-
-enrichResults = list()
-for (name in names(setsToTest)){
-  genesInModule = setsToTest[[name]]  
-  enrichResults[[name]] = lapply(GeneSets,
-                                    function(x, genesInModule, genesInBackground, fisherEnrichment, rownameToFirstColumn){
-                                      tmp = as.data.frame(t(sapply(x, fisherEnrichment, genesInModule, genesInBackground)))
-                                      tmp = rownameToFirstColumn(tmp,'GeneSetName')
-                                      return(tmp)
-                                    }, unique(genesInModule), unique(backgroundGenes), fisherEnrichment, rownameToFirstColumn) %>%
-    rbindlist(use.names=TRUE, idcol = 'Category') %>%
-    dplyr::mutate(fdr = p.adjust(pval, 'fdr'))
-  writeLines(paste0('Completed ',name))  
-}
-
-# Write results to file
-tmp1 = rbindlist(enrichResults, use.names = TRUE, idcol = 'ComparisonName', fill = TRUE) %>%
-  filter(!is.na(Category)) %>%
-  filter(fdr <= 0.01)
-
-tmp2 = dplyr::select(tmp1, ComparisonName, GeneSetName, Odds.Ratio) %>%
-  tidyr::spread(ComparisonName,Odds.Ratio)
-rownames(tmp2) = tmp2[,1]
-
-tmp1$pval = unlist(tmp1$pval)
-tmp1$ngenes = unlist(tmp1$ngenes)
-tmp1$noverlap = unlist(tmp1$noverlap)
-tmp1$Odds.Ratio = unlist(tmp1$Odds.Ratio)
-tmp1$Genes = unlist(tmp1$Genes)
-
-# Store logcpm in synapse
-write.table(tmp1, file = 'enrich.tsv', sep = '\t', row.names=F, quote=F)
-OBJ = File('enrich.tsv', 
-           name = 'Pathway Enrichment Results (AD and Control)', 
-           parentId = CODE$properties$id)
-OBJ = synStore(OBJ,
-               used = as.character(c(covariates.id)),
-               executed = thisFile,
-               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
-```
-
 ### Store results in synapse

### 2017-05-10 code/Rmd/enrichmentAnalysis.Rmd
Temp commit

--- 2017-02-21 code/Rmd/enrichmentAnalysis.Rmd
@@
                                  parentId ="syn5569102",
-                                 entityName = 'Enrichment Analysis of Control vs AD from all three brain banks (Reprocessed RNASeq)')
+                                 entityName = 'MetaAnalysis of Control vs AD from all three brain banks (Reprocessed RNASeq)')
 ```
@@
 ```{r fxns}
-# Function to perform Fishers enrichment analysis
-fisherEnrichment <- function(genesInSignificantSet, # A character vector of differentially expressed or some significant genes to test
-                             genesInGeneSet, # A character vector of genes in gene set like GO annotations, pathways etc...
-                             genesInBackground # Background genes that are 
-){
-  genesInSignificantSet = intersect(genesInSignificantSet, genesInBackground) # back ground filtering
-  genesInNonSignificantSet = base::setdiff(genesInBackground, genesInSignificantSet)
-  genesInGeneSet = intersect(genesInGeneSet, genesInBackground) # back ground filtering
-  genesOutGeneSet = base::setdiff(genesInBackground,genesInGeneSet)
+# Utility function for combining p-values using Fishers inverse chi-square method
+fishersPvalue <- function(pvalue){
+  pvalue = -2*log(pvalue, base = exp(1))
+  S <- 0
+  N <- dim(pvalue)[2]
 
-  pval = fisher.test(
-    matrix(c(length(intersect(genesInGeneSet, genesInSignificantSet)),             
-             length(intersect(genesInGeneSet, genesInNonSignificantSet)),
-             length(intersect(genesOutGeneSet, genesInSignificantSet)),
-             length(intersect(genesOutGeneSet, genesInNonSignificantSet))), 
-           nrow=2, ncol=2), alternative="greater")
-  OR = (length(intersect(genesInGeneSet, genesInSignificantSet)) * length(intersect(genesOutGeneSet, genesInNonSignificantSet))) / (length(intersect(genesInGeneSet, genesInNonSignificantSet)) * length(intersect(genesOutGeneSet, genesInSignificantSet))) %>% as.numeric()
-  return(data.frame(pval = pval$p.value,
-                    ngenes = length(genesInGeneSet),
-                    noverlap = length(intersect(genesInGeneSet, genesInSignificantSet)),
-                    Odds.Ratio = OR,
-                    Genes = paste(intersect(genesInGeneSet, genesInSignificantSet), collapse = '|')))}
+  correlation <- WGCNA::bicor(pvalue,use = 'pairwise.complete.obs');
+
+  for (i in 1:(N-1))
+    for (j in (i+1):N){
+      if (correlation[i,j] > 0){
+        S <- S + correlation[i,j]*(3.25+0.75*correlation[i,j])
+      } else if (correlation[i,j] <= 0){
+        S <- S + correlation[i,j]*(3.27+0.71*correlation[i,j])
+      }
+    }
+
+  combinedPvalue.chi <- rowSums(pvalue)*(4*N)/(4*N+2*S)
+  combinedPvalue <- sapply(combinedPvalue.chi, function(x,N){
+    pchisq(x, df = 2*N, lower.tail = FALSE)
+  },N)
+  names(combinedPvalue) = rownames(pvalue)
+
+  return(combinedPvalue)
+}
 ```
 
-### Get covariates/sample metadata from synapse
-```{r covariates}
-covariates.ids = c(ROSMAP = 'syn8018344',
-                   MSBB = 'syn8073656',
-                   MAYO = 'syn8028561')
-covariates = lapply(covariates.ids, function(id){
-  x = downloadFile(id)
-  colnames(x)[1] = 'SampleID'
-  return(x)
-})
-
-covariates$ROSMAP = dplyr::filter(covariates$ROSMAP, cogdx %in% c(1,4)) %>%
-  dplyr::mutate(Dx = factor(cogdx, levels = c(1,4), labels = c('1' = 'NCI', '4' = 'AD')),
-                Study = 'ROSMAP',
-                BrainRegion = 'DLPFC') %>%
-  dplyr::select(SampleID, Study, BrainRegion, Dx)
-
-covariates$MSBB = dplyr::filter(covariates$MSBB, BrodmannArea.Dx %in% c('BM10.SD', 'BM22.SD', 'BM36.SD', 
-                                                                        'BM44.SD', 'BM36.ND', 'BM10.ND', 
-                                                                        'BM44.ND', 'BM22.ND')) %>%
-  dplyr::select(SampleID, BrodmannArea.Dx) %>%
-  tidyr::separate(BrodmannArea.Dx, c('BrainRegion', 'Dx')) %>%
-  dplyr::mutate(Study = 'MSSM',
-                Dx = factor(Dx, levels = c('ND','SD')),
-                BrainRegion = factor(BrainRegion, labels = c('FP', 'STG', 'PHG', 'IFG'))) %>%
-  dplyr::select(SampleID, Study, BrainRegion, Dx)
-
-covariates$MAYO = dplyr::filter(covariates$MAYO, BrainRegion.Diagnosis %in% c('TCX.AD', 'TCX.Control', 'CER.AD', 'CER.Control')) %>%
-  dplyr::select(SampleID, BrainRegion.Diagnosis) %>%
-  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Dx')) %>%
-  dplyr::mutate(Study = 'MAYO',
-                Dx = factor(Dx, levels = c('Control','AD')),
-                BrainRegion = factor(BrainRegion, levels = c('CER', 'TCX'))) %>%
-  dplyr::select(SampleID, Study, BrainRegion, Dx)
-```
-
-### Get differential expression data from synapse
-```{r diffexp}
-diff.exp.ids = c(ROSMAP = 'syn8018358',
-                 MSBB = 'syn8073678',
-                 MAYO = 'syn8028572')
-diff.exp = lapply(diff.exp.ids, downloadFile)
-
-diff.exp$ROSMAP = filter(diff.exp$ROSMAP, Comparison == 'cogdx1-cogdx4')
-diff.exp$MSBB = filter(diff.exp$MSBB, Comparison == 'ND-SD')
-diff.exp$MAYO = filter(diff.exp$MAYO, Comparison == 'Control-AD')
-
-tmp = diff.exp %>% 
-  rbindlist(use.names = T, fill = T) 
-tmp$Direction = 'NONE'
-tmp$Direction[tmp$logFC >= log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
-tmp$Direction[tmp$logFC <= -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'
-
-writeLines('Number of differentially expressed genes at an FDR of 0.05 and fold change of 1.2 are')
-tmp %>%
-  group_by(Study, Region, Direction, Comparison) %>%
-  summarise(count = n()) %>%
-  spread(Direction, count) %>%
-  kable
-
-pl = list()
-for(region in c('ROSMAP', 'MSBB', 'MAYO')){
-  p = ggplot(tmp %>% filter(Study == region), aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point()
-  p = p + scale_color_manual(values = c('green','grey','red'))
-  p = p + facet_grid(Study~.+Region) + theme(legend.position = 'top')
-  pl[[region]] = p
-}
-multiplot(plotlist = pl, cols = 1)
-```
-
-### Get hgnc_symbol mapping from biomaRt
-```{r biomart}
-# Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "dec2016.archive.ensembl.org")
-
-# Get background gene ids
-backgroundGenes = sapply(diff.exp, function(x) x$ensembl_gene_id) %>%
-  Reduce(c,.)
-
-# Query biomart
-Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
-                       filters = "ensembl_gene_id", values = backgroundGenes,
-                       mart = mart)
-
-Entrez2Ensembl <- as.data.frame(org.Hs.egENSEMBL2EG)
-Entrez2HGNC <- as.data.frame(org.Hs.egSYMBOL2EG)
-HGNC2Ensembl <- left_join(Entrez2HGNC, Entrez2Ensembl)
-HGNC2Ensembl = HGNC2Ensembl[grep('^LOC',HGNC2Ensembl$symbol),c('ensembl_id', 'symbol')] %>%
-  filter(!is.na(ensembl_id), !is.na(symbol)) %>%
-  dplyr::rename(ensembl_gene_id = ensembl_id, hgnc_symbol = symbol)
-
-Ensemble2HGNC = rbindlist( list(Ensemble2HGNC, HGNC2Ensembl), use.names = T, fill = T) %>%
-  unique
-```
-
-### Pathway enrichment analysis
-```{r enrich.all, fig.height=16, fig.width=10}
-# Find genes that are differentially expressed in each brain region
-diff.exp.gl = lapply(diff.exp, function(x){
-  dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05 & sign(CI.L) == sign(CI.R)) %>%
-    dplyr::select(Region, hgnc_symbol) %>%
-    unique
-}) %>%
-  rbindlist(use.names = T, fill = T) %>%
-  dlply(.(Region), .fun = function(x){unique(x$hgnc_symbol)})
-
-# Get Background genes
-backgroundGenes = Ensemble2HGNC$hgnc_symbol[Ensemble2HGNC$ensembl_gene_id %in% backgroundGenes] %>% unique()
-
-#### Get gene sets ####
+### Get curated gene list
+```{r curate.genelist}
 # Download enrichr gene sets from synapse
 GL_OBJ = synGet('syn4867851')
-ALL_USED_IDs = c(ALL_USED_IDs, GL_OBJ$properties$id)
+all.used.ids = GL_OBJ$properties$id
 load(GL_OBJ@filePath)
 
-gsets = c("BioCarta_2015", "GO_Biological_Process", "Reactome_2015", "TargetScan_microRNA", "WikiPathways_2015")
+gsets = c("GO_Biological_Process", "Reactome_2015", "WikiPathways_2015")
 GeneSets.Enrichr = GeneSets[gsets]
@@
                    AD = GeneSets['AD:GeneticLoci'])
-GeneSets = c(GeneSets.Enrichr, GeneSets.CM)
+GeneSets = c(GeneSets.Enrichr, GeneSets.CM) # 
+```
+
+### Get covariates/sample metadata from synapse
+```{r covariates}
+covariates.ids = c(ROSMAP = 'syn8456631',
+                   MSBB = 'syn8484996',
+                   MAYO = 'syn8466814')
+all.used.ids = covariates.ids
+covariates = lapply(covariates.ids, downloadFile)
+
+covariates$ROSMAP$BrainRegion = 'DLPFC'
+covariates$MSBB = covariates$MSBB %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
+covariates$MAYO = covariates$MAYO %>%
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
+
+covariates = covariates %>% 
+  rbindlist(use.names = T, fill = T, idcol = 'Study') %>%
+  dplyr::select(SampleID, Diagnosis, BrainRegion)
+
+kable(table(covariates$Diagnosis, covariates$BrainRegion))
+```
+
+### Get differential expression data from synapse
+```{r diffexp, fig.width=10, fig.height=15}
+diff.exp.ids = c(ROSMAP = 'syn8456721',
+                 MSBB = 'syn8502894',
+                 MAYO = 'syn8468023')
+all.used.ids = c(all.used.ids, diff.exp.ids)
+diff.exp = lapply(diff.exp.ids, downloadFile)
+
+diff.exp$ROSMAP = diff.exp$ROSMAP %>%
+  setnames(c('Region'), c('BrainRegion'))
+
+tmp = rbindlist(diff.exp, use.names = T, fill = T) %>%
+  dplyr::filter(Comparison == 'CONTROL-AD')
+
+tmp$Direction = 'NONE'
+tmp$Direction[tmp$logFC > log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
+tmp$Direction[tmp$logFC < -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'
+
+writeLines('Number of differentially expressed genes at an FDR of 0.05')
+tmp %>%
+  group_by(Study, BrainRegion, Direction, Comparison) %>%
+  summarise(count = n()) %>%
+  spread(Direction, count) %>%
+  kable
+
+pl = list()
+for(region in c('ROSMAP', 'MSSM', 'MAYO')){
+  p = ggplot(tmp %>% filter(Study == region), aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point()
+  p = p + scale_color_manual(values = c('green','grey','red'))
+  p = p + facet_grid(Study~.+BrainRegion) + theme(legend.position = 'top')
+  pl[[region]] = p
+}
+multiplot(plotlist = pl, cols = 1)
+```
+
+### Pathway enrichment analysis (individual brain regions)
+```{r pathway.enrich}
+# setsToTest = rbindlist(diff.exp, use.names = T, fill = T) %>%
+#   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
+#   dlply(.(BrainRegion), .fun = function(x){
+#     setdiff(unique(x$hgnc_symbol),c(''))
+#   })
+
+setsToTest = rbindlist(diff.exp, use.names = T, fill = T) %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) > log2(1.2)) %>%
+  dplyr::mutate(Direction = sign(logFC),
+                Direction = factor(Direction, levels = c(-1,1),
+                                   labels = c('-1' = 'DOWN', '1' = 'UP'))) %>%
+  dlply(.(BrainRegion, Direction), .fun = function(x){
+    setdiff(unique(x$hgnc_symbol),c(''))
+  })
 
 # Filter gene list
-GeneSets = filterGeneSets(GeneSets, backgroundGenes, minSize = 3, maxSize = 5000)
+GeneSets = mapply(function(x,y){
+  names(x) = paste(y,names(x),sep = ':')
+  return(x)
+},GeneSets, names(GeneSets), SIMPLIFY = F) %>%
+  Reduce(c,.)
+GeneSets = GeneSets[setdiff(seq(1,length(GeneSets)), c(grep('Mus musculus', names(GeneSets))))]
 
-# Enrichment Analysis
-enrichResults = plyr::ldply(diff.exp.gl, .fun = function(gs, geneSets, backgroundGenes, fisherEnrichment){
-  tmp = plyr::ldply(geneSets, .fun = function(x, gs, backgroundGenes, fisherEnrichment){
-    plyr::ldply(x, .fun = fisherEnrichment, gs, backgroundGenes, .id = 'Name', .parallel = T)
-  }, gs, backgroundGenes, fisherEnrichment, .id = 'Category', .parallel = T, .paropts = list(.packages = c('dplyr'))) %>%
-    dplyr::mutate(fdr = p.adjust(pval, method = 'bonferroni'))
-}, GeneSets, backgroundGenes, fisherEnrichment, .parallel = T, .paropts = list(.packages = c('dplyr', 'plyr')))
+GeneSets = lapply(GeneSets, function(x, backgroundGenes){
+  x = intersect(x,backgroundGenes)
+}, backgroundGenes)
+GeneSets = GeneSets[sapply(GeneSets, length) >= 15 & sapply(GeneSets, length) <= 1000]
 
-# Filter results
-tmp1 = filter(enrichResults, fdr <= 0.05)
+enrichResults = plyr::ldply(setsToTest, .fun = function(genesInGeneSet, geneSets, backgroundGenes){
+  plyr::ldply(geneSets, .fun = fisherEnrichment, genesInGeneSet, backgroundGenes, .parallel = T) %>%
+    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
+}, GeneSets, backgroundGenes, 
+.parallel = F, .progress = 'text', .paropts = list(.packages = c('plyr','dplyr','CovariateAnalysis')))
+
+tmp = dplyr::select(enrichResults, BrainRegion, .id, fdr) %>%
+  dplyr::filter(fdr <= 0.05) %>%
+  tidyr::spread(BrainRegion, fdr)
+
+# Store logcpm in synapse
+write.table(tmp, file = 'enrich.tsv', sep = '\t', row.names=F, quote=F)
+OBJ = File('enrich.tsv', 
+           name = 'Pathway Enrichment Results (AD and Control)', 
+           parentId = CODE$properties$id)
+OBJ = synStore(OBJ,
+               used = as.character(c(covariates.id)),
+               executed = thisFile,
+               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
+```
+
+### Get cell type fraction from synapse
+```{r celltype.frac, eval = FALSE}
+celltype.fraction.ids = c(ROSMAP = 'syn8124520',
+                          MSBB = 'syn8146376',
+                          MAYO = 'syn8135250')
+celltype.fraction = lapply(celltype.fraction.ids, downloadFile)
+
+cell.frac = mapply(function(cov, frc){
+  x = filter(frc, SampleID %in% cov$SampleID) %>%
+    dplyr::select(-`P-value`,-RMSE, -Correlation) %>%
+    tidyr::gather(celltype, fraction, -SampleID) %>%
+    left_join(cov)
+}, covariates, celltype.fraction, SIMPLIFY = F) %>%
+  rbindlist(use.names = T, fill = T) %>%
+  dplyr::mutate(Dx = factor(Dx, levels = c('Control', 'NCI', 'ND', 'SD', 'AD')))
+
+pl = list()
+
+p = ggplot(cell.frac %>% filter(celltype %in% c('neurons', 'microglia', 'astrocytes')), 
+           aes(x = BrainRegion, y = fraction, color = Dx)) + geom_boxplot()
+p = p + facet_grid(celltype~., scales = 'free_y')
+p = p + xlab('Brain Region') + ylab('Fraction')
+p = p + theme(axis.text.x = element_text(hjust = 1, angle = 90), legend.position = 'top')
+pl[[1]] = p
+
+p = ggplot(cell.frac %>% filter(celltype %in% c('endothelial', 'oligodendrocytes', 'OPC')), 
+           aes(x = BrainRegion, y = fraction, color = Dx)) + geom_boxplot()
+p = p + facet_grid(celltype~., scales = 'free_y')
+p = p + xlab('Brain Region') + ylab('Fraction')
+p = p + theme(axis.text.x = element_text(hjust = 1, angle = 90), legend.position = 'top')
+pl[[2]] = p
+
+multiplot(plotlist = pl, cols = 2)
+```
+
+### Get hgnc_symbol mapping from biomaRt
+```{r biomart}
+# Get background gene ids
+backgroundGenes = sapply(diff.exp, function(x) x$hgnc_symbol) %>%
+  Reduce(c,.) %>% unique() %>% setdiff(c(''))
+```
+
+### Consistently differentially expressed genes (in all 7 brain regions)
+```{r diffexp.all, fig.height=16, fig.width=10}
+# Find genes that are differentially expressed in each brain region
+diff.exp.gl = lapply(diff.exp, function(x){
+  dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05 & sign(CI.L) == sign(CI.R)) %>%
+    dplyr::select(Region, ensembl_gene_id) %>%
+    unique
+}) %>%
+  rbindlist(use.names = T, fill = T) %>%
+  dlply(.(Region), .fun = function(x){setdiff(unique(x$ensembl_gene_id),c(''))})
+
+tmp1 = rbindlist(diff.exp, use.names = T, fill = T) %>%
+  dplyr::filter(ensembl_gene_id %in% Reduce(c, diff.exp.gl))
+
+tmp1$Direction = 'NONE'
+tmp1$Direction[tmp1$adj.P.Val <= 0.05 & tmp1$logFC >= log2(1.2)] = 'UP'
+tmp1$Direction[tmp1$adj.P.Val <= 0.05 & tmp1$logFC <= -log2(1.2)] = 'DOWN'
+
+tmp = tmp1 %>%
+  group_by(ensembl_gene_id) %>%
+  summarise(count = sum(Direction != 'NONE')) %>%
+  filter(count == 7) %>%
+  left_join(dplyr::select(tmp1, ensembl_gene_id, hgnc_symbol) %>% unique)
+
+writeLines(paste('Differentially expressed genes in all 7 brain regions at an FDR of 0.05 and fold change of 1.2 are',
+                 length(unique(tmp$hgnc_symbol))))
+kable(data.frame(ngenes = length(unique(tmp$hgnc_symbol)),
+                 hgnc_symbol = tmp$hgnc_symbol %>% unique %>% paste(collapse = ',')))
+
+writeLines('Top six differentially expressed genes are')
+tmp.eff.sz = filter(tmp1, hgnc_symbol %in% tmp$hgnc_symbol) %>%
+  group_by(hgnc_symbol) %>% 
+  summarise(es = sum(logFC, na.rm = T)) %>% 
+  top_n(6, abs(es))
+
+pl = list()
+p = ggplot(tmp1 %>% 
+             filter(hgnc_symbol %in% c('MEIS3', 'PCSK1', 'ADAMTS2')), 
+           aes(x = Region, y = logFC, color = Study, fill = Direction)) + geom_bar(stat = 'identity')
+p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
+p = p + theme(legend.position = 'top') + scale_fill_manual(values = c('green', 'red'))
+pl[[1]] = p
+
+p = ggplot(tmp1 %>% 
+             filter(hgnc_symbol %in% c('FOXJ1', 'PRELP', 'S100A4')), 
+           aes(x = Region, y = logFC, color = Study, fill = Direction)) + geom_bar(stat = 'identity')
+p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
+p = p + theme(legend.position = 'top')  + scale_fill_manual(values = c('green','grey', 'red'))
+pl[[2]] = p
+
+multiplot(plotlist = pl, cols = 2)
+```
+
+### Meta Analysis
+```{r metaanalysis.min.pval}
+# Get raw p.values
+P.Value = diff.exp %>% 
+  rbindlist(use.names = T, fill = T) %>%
+  dplyr::select(gene.id, P.Value, Region) %>%
+  unique() %>%
+  tidyr::spread(Region, P.Value)
+P.Value[is.na(P.Value)] = 1
+rownames(P.Value) = P.Value$gene.id
+P.Value$gene.id = NULL
+P.Value = data.matrix(P.Value)
+
+# Perform minimum p-value based meta-analysis
+min.pval = P.Value %>% 
+  apply(1, min, na.rm = T) %>%
+  p.adjust(method = 'bonferroni') %>%
+  rownameToFirstColumn('gene.id') %>%
+  dplyr::rename(min.pval = DF)
+
+# Perform fisher p-value based meta-analysis
+fb.pval = P.Value %>%
+  fishersPvalue %>%
+  p.adjust(method = 'bonferroni') %>%
+  rownameToFirstColumn('gene.id') %>%
+  dplyr::rename(fb.pval = DF)
+```
+
+```{r }
+# Rank prod based analysis
+expr = resid.expr[!duplicated(resid.expr$ensembl_gene_id),-c(1:2)]
+rownames(expr) = resid.expr$ensembl_gene_id[!duplicated(resid.expr$ensembl_gene_id)]
+
+cl = covariates$Status
+names(cl) = covariates$ID
+cl[cl == 'NCI'] = 'Control'
+cl[cl == 'ND'] = 'Control'
+cl[cl == 'SD'] = 'AD'
+cl = cl[cl %in% c('Control','AD')] %>% droplevels 
+cl.tmp = -as.numeric(cl)+2
+names(cl.tmp) = names(cl)
+
+origin = covariates$BrainRegion %>% as.numeric
+names(origin) = covariates$ID
+origin = origin[names(cl)]
+
+expr = expr[, names(cl)]
+
+rankprod.pfp = RPadvance(expr, cl.tmp, origin, num.perm = 100, gene.names = rownames(expr), plot=T, rand = 123456, huge=T)
+rankprod.pval = apply(rankprod.pfp$pfp, 1, min) %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  dplyr::rename(rankprod.pval = DF)
+
+metaAnal = join_all(list(min.pval, max.pval, fb.pval, rankprod.pval), type = 'full')
+metaAnal$weight = -log10(metaAnal[,-c(1,6)]) %>% rowSums %>% rank
+metaAnal = metaAnal %>% arrange(desc(weight))
+
+writeLines('Number of genes differentially expressed between AD and control at an FDR of 0.05 are')
+tmp = metaAnal %>% 
+  dplyr::select(-weight) %>%
+  gather(method, pval, -ensembl_gene_id) %>% 
+  group_by(method) %>% 
+  filter(pval <= 0.05) %>%
+  dplyr::select(ensembl_gene_id) %>%
+  unique %>%
+  dlply(.(method), .fun = function(x){unique(x$ensembl_gene_id)})
+kable(sapply(tmp, length) %>% as.data.frame)
+
+tmp1 = Venn(tmp)
+
+plotVenn3d(tmp1@IndicatorWeight[-(1),'.Weight'], labels = names(tmp))
+```
+### Heatmap of expression
+```{r heatmap}
+# Get the overlapping gene sets from all three methods
+tmp.expr = t(scale(t(expr)))
+tmp.expr = scale(tmp.expr)
+tmp.expr = tmp.expr[intersect(intersect(tmp$fb.pval, tmp$min.pval), tmp$rankprod.pval),]
+
+tmp.covariates = dplyr::select(covariates, BrainRegion, Status)
+tmp.covariates$Status[tmp.covariates$Status == 'NCI'] = 'Control'
+tmp.covariates$Status[tmp.covariates$Status == 'ND'] = 'Control'
+tmp.covariates$Status[tmp.covariates$Status == 'SD'] = 'AD'
+tmp.covariates = tmp.covariates %>% droplevels
+rownames(tmp.covariates) = covariates$ID
+
+h1 = HeatmapAnnotation(tmp.covariates)
+Heatmap(tmp.expr, top_annotation = h1, show_row_names = F, show_column_names = F,
+        show_row_dend = F, show_column_dend = F)
 ```
@@
 ```{r synapse.store, include=FALSE, cache=FALSE}
-thisFileName <- 'enrichmentAnalysis.Rmd'
+thisFileName <- 'MetaAnalysis.Rmd'
 
@@
 # Store consolidated data in synapse
-CODE = Folder(name = 'Pathway Enrichment Analysis of Control vs AD Data (all brain banks)', parentId = "syn5569102")
+CODE = Folder(name = 'MetaAnalysis of Control vs AD Data', parentId = "syn5569102")
 CODE = synStore(CODE)
 
-# Store enrichment analysis in synapse
-write.table(tmp1, file = 'enrichmentResults.tsv', sep = '\t', row.names=F, quote=F)
-OBJ = File('enrichmentResults.tsv', 
-           name = 'Pathway Enrichment Results (AD and Control)', 
+# Store covariates in synapse
+write.table(covariates, file = 'ALLCovariates.tsv', sep = '\t', row.names=F, quote=F)
+OBJ = File('ALLCovariates.tsv', 
+           name = 'Covariates', 
            parentId = CODE$properties$id)
 OBJ = synStore(OBJ,
-               used = as.character(ALL_USED_IDs),
+               used = as.character(c(covariates.id)),
                executed = thisFile,
-               activityName = 'Pathway enrichment analysis')
-stopCluster(cl)
+               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
+
+# Store logcpm in synapse
+write.table(logcpm, file = 'ALLLogCPM.tsv', sep = '\t', row.names=F, quote=F)
+OBJ = File('ALLLogCPM.tsv', 
+           name = 'Log CPM (AD and Control)', 
+           parentId = CODE$properties$id)
+OBJ = synStore(OBJ,
+               used = as.character(c(covariates.id)),
+               executed = thisFile,
+               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
+
+# Store residual expression in synapse
+write.table(expr, file = 'ALLExpr.tsv', sep = '\t', row.names=F, quote=F)
+OBJ = File('ALLExpr.tsv', 
+           name = 'Residual Expression (AD and Control)', 
+           parentId = CODE$properties$id)
+OBJ = synStore(OBJ,
+               used = as.character(c(covariates.id)),
+               executed = thisFile,
+               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
+
+# Store diffexp in synapse
+write.table(diffexp, file = 'ALLDiffExp.tsv', sep = '\t', row.names=F, quote=F)
+OBJ = File('ALLDiffExp.tsv', 
+           name = 'Differential Expression (AD and Control)', 
+           parentId = CODE$properties$id)
+OBJ = synStore(OBJ,
+               used = as.character(c(covariates.id)),
+               executed = thisFile,
+               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
+
+# Store meta analysis in synapse
+write.table(metaAnal, file = 'ALLMetaAnal.tsv', sep = '\t', row.names=F, quote=F)
+OBJ = File('ALLMetaAnal.tsv', 
+           name = 'Meta Analysis (AD and Control)', 
+           parentId = CODE$properties$id)
+OBJ = synStore(OBJ,
+               used = as.character(c(covariates.id)),
+               executed = thisFile,
+               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
 ```

### 2017-05-10 code/Rmd/regionLevelDiffExpAnlz.Rmd
Temp commit

@@
 library(cqn)
+library(knitr)
+library(psych)
 
@@
   cov = cov[,c(FactorCov, ContCov)]
-  assocStat = getAssociationStatistics(cov)
+  # assocStat = getAssociationStatistics(cov)
 

### 2017-05-11 code/R/extractGeneSets.R
Added R function to extract and convert all enrichr genesets to R list format

@@ -1,0 +1,58 @@
+#### Function to extract gene sets from ENRICHR  source files ####
+## It is assumed your working directory is where this file
+
+### Clear R console screen output
+cat("\014") 
+
+## Load required libraries
+library(CovariateAnalysis)
+library(data.table)
+library(tidyr)
+library(plyr)
+library(dplyr)
+library(stringr)
+
+library(synapseClient)
+library(knitr)
+library(githubr)
+
+library(parallel)
+library(doParallel)
+library(foreach)
+
+cl = makeCluster(detectCores() - 2)
+registerDoParallel(cl)
+
+synapseLogin()
+
+#### Synapse specific parameters ####
+parentId = 'syn5570248';
+activityName = 'Extract Genesets';
+activityDescription = 'Extract genesets from enrichr';
+
+# Github link
+thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='geneSetCuration')
+thisFile <- getPermlink(repository = thisRepo, repositoryPath='code/R/extractGeneSets.R')
+
+#### Query Synapse for source files ####
+all.files = synQuery('select id,name from file where parentId == "syn4598359"')
+
+GeneSets = plyr::dlply(all.files, .(file.id), .fun = function(y){
+  print(paste('Started',y$file.name))
+  tmp = read.table(synGet(y$file.id)@filePath, fill = NA, row.names = NULL, header = FALSE)
+  tmp1 = plyr::dlply(tmp, .(V1), .fun = function(x){
+    as.character(unlist(x[-(1)])) %>% unique() %>% setdiff('')
+  }, 
+  .parallel = T,
+  .paropts = list(.packages = c('dplyr')))
+  print(paste('Completed',y$file.name))
+  return(tmp1)
+ }, .progress = 'text')
+names(GeneSets) = gsub('.txt','',all.files$file.name)
+save(list = 'GeneSets', file = 'allEnrichrGeneSets.RData')
+stopCluster(cl)
+
+#### Store in synapse ####
+obj = File('allEnrichrGeneSets.RData', name = 'Gene Sets in RList Format ', parentId = syn4867780)
+obj = synStore(obj, used = all.files$file.id, activityName = activityName, 
+               executed = thisFile, activityDescription = activityDescription)

### 2017-05-12 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Minor modifications

@@
                                  parentId ="syn7501880",
-                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)')
+                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection 1)')
 ```
@@
 ### Filter genes
-Remove genes that have less than 5 cpm counts in at least 100% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
+Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
 ```{r cpmnormalisation}
@@
   processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID %>% droplevels()],
-                                                   MIN_GENE_CPM=5, 
-                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=1)
+                                                   MIN_GENE_CPM=1, 
+                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
   processed.counts$filteredExprMatrix$genes
@@
 ### Distribution of samples (log cpm)
-```{r lcpm.dist}
+```{r lcpm.dist, cache=FALSE}
 # Plot abberent distribution of logcpm counts
@@
 
-    kable(nChange)
+    print(nChange)
 
@@
     tmp = nChange %>%
-      dplyr::filter(nchange > round(dim(cnt)[1]*0.05)) %>%
+      dplyr::filter(nchange > round(dim(cnt)[1]*0.01)) %>%
       dplyr::top_n(1, nchange)
@@
 # Fit a base model with the following covariates
-adjust.covars = c('Dx.Tissue', 'Institution', 'Gender')
+adjust.covars = c('Dx.Tissue')
 baseModelBIC = fitMixedEffectModel(adjust.covars, 
@@
 # Iteratively identify covariates in phase I
-phase1.covars = c('PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
+phase1.covars = c('Institution', 'Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
 vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
@@
 # Code
-CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)", parentId = parentId)
+CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection 1)", parentId = parentId)
 annotations(CODE) = all.annotations
@@
 # Store design matrix
-DESIGN$design %>%
+DM1$design %>%
   rownameToFirstColumn('SampleID') %>%

### 2017-05-12 code/R/extractGeneSets.R
Minor modifications

@@ -39,11 +39,15 @@ all.files = synQuery('select id,name from file where parentId == "syn4598359"')
 
 GeneSets = plyr::dlply(all.files, .(file.id), .fun = function(y){
   print(paste('Started',y$file.name))
-  tmp = read.table(synGet(y$file.id)@filePath, fill = NA, row.names = NULL, header = FALSE)
+  tmp = read.table(synGet(y$file.id)@filePath, fill = NA, row.names = NULL, header = FALSE, sep = '\t')
   tmp1 = plyr::dlply(tmp, .(V1), .fun = function(x){
-    as.character(unlist(x[-(1)])) %>% unique() %>% setdiff('')
+    sapply(x[-(1)], function(x){
+      str_split(x, ',')[[1]][1]
+    }) %>% 
+      unique() %>% 
+      setdiff('')
   }, 
-  .parallel = T,
+  .parallel = F,
   .paropts = list(.packages = c('dplyr')))
   print(paste('Completed',y$file.name))
   return(tmp1)
@@ -53,6 +57,6 @@ save(list = 'GeneSets', file = 'allEnrichrGeneSets.RData')
 stopCluster(cl)
 
 #### Store in synapse ####
-obj = File('allEnrichrGeneSets.RData', name = 'Gene Sets in RList Format ', parentId = syn4867780)
+obj = File('allEnrichrGeneSets.RData', name = 'Gene Sets in RList Format ', parentId = 'syn4867780')
 obj = synStore(obj, used = all.files$file.id, activityName = activityName, 
                executed = thisFile, activityDescription = activityDescription)

### 2017-05-14 code/Rmd/regionLevelDiffExpAnlz.Rmd
Added AAD differential expression function to region level differential expression analysis

@@
 ```{r fxn}
-performDifferentialExpression <- function(cov, cnt, contVar, varOfInterest, n, gene.len, gene.gc){
-
-  diffExp = foreach(i = 1:n, 
-                    .combine = rbind,
-                    .packages = c('limma','edgeR','cqn','dplyr','plyr','CovariateAnalysis','data.table'),
-                    .export = c('gene.len', 'gene.gc')) %dopar% {
-
-                    # Sub-sample data for balanced comparison
-                    minSampSz = min(summary(cov[,contVar]), na.rm = T)
-                    innerCov = rownameToFirstColumn(cov, 'SampleID')  %>%
-                      plyr::ddply(.variables = contVar, .fun = function(x, minSampSz){
-                        ind = sample(1:dim(x)[1], minSampSz)
-                        return(x[ind,])
-                      }, minSampSz) %>%
-                      droplevels()
-                    rownames(innerCov) = innerCov$SampleID
+performDifferentialExpression <- function(cov, cnt, contVar, varOfInterest, gene.len, gene.gc){
 
   # Get design matrix
-                    designMat = getDesignMatrix(innerCov[,varOfInterest], Intercept = F)$design
+  designMat = getDesignMatrix(cov[,varOfInterest], Intercept = F)$design
   designMat = designMat[,linColumnFinder(designMat)$indepCols]
@@
   # Compute offset for gene length and gc content
-                    cqn.gene.expression = cqn(cnt[,innerCov$SampleID],
+  cqn.gene.expression = cqn(cnt,
                             x = gene.gc[rownames(cnt), 'percentage_gc_content'],
@@
   # Estimate voom weights for dispersion control
-                    voom.weights = voom(cnt[,innerCov$SampleID], design = designMat, plot=F, na.rm = T)
+  voom.weights = voom(cnt, design = designMat, plot=F, na.rm = T)
 
@@
   baseModel.cntr = eBayes(baseModel.cntr)
-                    innerDiffExp = lapply(1:dim(contrasts)[2], function(i, baseModel.cntr){
+  diffExp = lapply(1:dim(contrasts)[2], function(i, baseModel.cntr){
     topTable(baseModel.cntr, coef = i, number = 50000, confint = T) %>%
@@
     }, baseModel.cntr)
-                    names(innerDiffExp) = colnames(contrasts)
-                    innerDiffExp = innerDiffExp %>%
+  names(diffExp) = colnames(contrasts)
+  diffExp = diffExp %>%
     rbindlist(use.names = T, fill = T, idcol = 'Comparison') %>%
-                      dplyr::mutate(iter = i,
-                                    Comparison = gsub(contVar,'',Comparison))
+    dplyr::mutate(Comparison = gsub(contVar,'',Comparison))
 
-                    return(innerDiffExp)
+  return(diffExp)
 }
 
+performDifferentialExpressionAAD <- function(cov, cnt, contVar, varOfInterest, gene.len, gene.gc){
+
+  # Get design matrix
+  designMat = getDesignMatrix(cov[,varOfInterest], Intercept = F)$design
+  designMat[,grep('Diagnosis',colnames(designMat))] = designMat[,grep('Diagnosis',colnames(designMat))] * designMat[,'AAD']
+  colnames(designMat)[grep('Diagnosis',colnames(designMat))] = paste(grep('Diagnosis',colnames(designMat), value = T), 'AAD', sep = '.')
+  designMat = designMat[,setdiff(colnames(designMat), 'AAD')]
+  designMat = designMat[,linColumnFinder(designMat)$indepCols]
+
+  # Compute offset for gene length and gc content
+  cqn.gene.expression = cqn(cnt,
+                            x = gene.gc[rownames(cnt), 'percentage_gc_content'],
+                            lengths = gene.len[rownames(cnt), 'gene.length'],
+                            lengthMethod = "smooth",
+                            verbose = FALSE)
+  cqn.gene.expression$E = cqn.gene.expression$y+cqn.gene.expression$offset
+
+  # Identify outlier expression and assign NA values
+  log.mat = apply(cqn.gene.expression$E, 1, function(x){
+    mn = mean(x, na.rm = T)
+    std.dev = sd(x, na.rm = T)
+    return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
+  }) %>% t
+  cqn.gene.expression$E[log.mat] = NA
+
+  # Estimate voom weights for dispersion control
+  voom.weights = voom(cnt, design = designMat, plot=F, na.rm = T)
+
+  # Fit linear model using new weights and new design
+  baseModel = lmFit(cqn.gene.expression$E,
+                    design = designMat,
+                    weights = voom.weights$weights)
+
+  # Perform differential expression
+  contrasts = levels(cov[,contVar]) %>%
+    paste(contVar,.,sep = '') %>%
+    paste('AAD',sep = '.') %>%
+    combn(2) %>%
+    apply(2,paste, collapse = '-') %>%
+    c('DiagnosisAD.AAD', 'DiagnosisCONTROL.AAD') %>%
+    makeContrasts(contrasts = ., levels = colnames(baseModel$coefficients))
+  baseModel.cntr =  contrasts.fit(baseModel, contrasts)
+  baseModel.cntr = eBayes(baseModel.cntr)
+  diffExp = lapply(1:dim(contrasts)[2], function(i, baseModel.cntr){
+    topTable(baseModel.cntr, coef = i, number = 50000, confint = T) %>%
+      rownameToFirstColumn('ensembl_gene_id')
+    }, baseModel.cntr)
+  names(diffExp) = colnames(contrasts)
   diffExp = diffExp %>%
-    dplyr::group_by(Comparison, ensembl_gene_id) %>%
-    tidyr::gather(feature, value, -Comparison, -ensembl_gene_id) %>%
-    dplyr::group_by(Comparison, ensembl_gene_id, feature) %>%
-    dplyr::summarise(value = median(value, na.rm = T)) %>%
-    tidyr::spread(feature, value)
+    rbindlist(use.names = T, fill = T, idcol = 'Comparison') %>%
+    dplyr::mutate(Comparison = gsub(contVar,'',Comparison))
 
@@
 all.results = plyr::dlply(covariates, .(BrainRegion), .fun = function(cov, cnt, gene.len, gene.gc){
-  n = 100
+
   # Filter counts and covariates
@@
 
-  baseDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, n, gene.len, gene.gc)
+  baseDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, gene.len, gene.gc)
 
@@
 
-  genderDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, n, gene.len, gene.gc)
+  genderDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, gene.len, gene.gc)
 
-  return(rbindlist(list(baseDiffExp, genderDiffExp), use.names = T, fill = T))
+  # Perform differential expression between case-control in a AAD specific manner
+  varOfInterest = c('Diagnosis', 'Gender', 'Batch', 'RIN', 'AAD', 'PMI', 
+                    'PCT_PF_READS_ALIGNED', 'PCT_INTERGENIC_BASES', 'PCT_INTRONIC_BASES', 
+                    'PCT_RIBOSOMAL_BASES')
+  contVar = 'Diagnosis'
+
+  aadDiffExp <- performDifferentialExpressionAAD(cov, cnt, contVar, varOfInterest, gene.len, gene.gc)
+
+  return(rbindlist(list(baseDiffExp, genderDiffExp, aadDiffExp), use.names = T, fill = T))
 }, counts, gene.len, gene.gc)
@@
   kable(row.names=F)
+
+writeLines('Number of differentially expressed genes between AD.AAD-CONTROL.AAD at an FDR of 0.05 and fold change of 1.2 are')
+sapply(all.results, function(x){
+  sum(x$adj.P.Val <= 0.05 & abs(x$logFC) >= log2(1.2) & x$Comparison == 'AD.AAD-CONTROL.AAD', na.rm = T)
+}) %>% rownameToFirstColumn('Region') %>% dplyr::rename(AD.AAD_vs_CONTROL.AAD = DF) %>%
+  kable(row.names=F)
+
+writeLines('Number of differentially expressed genes between AD.AAD at an FDR of 0.05 and fold change of 1.2 are')
+sapply(all.results, function(x){
+  sum(x$adj.P.Val <= 0.05 & abs(x$logFC) >= log2(1.2) & x$Comparison == 'AD.AAD', na.rm = T)
+}) %>% rownameToFirstColumn('Region') %>% dplyr::rename(AD.AAD = DF) %>%
+  kable(row.names=F)
+
+writeLines('Number of differentially expressed genes between CONTROL.AAD at an FDR of 0.05 and fold change of 1.2 are')
+sapply(all.results, function(x){
+  sum(x$adj.P.Val <= 0.05 & abs(x$logFC) >= log2(1.2) & x$Comparison == 'CONTROL.AAD', na.rm = T)
+}) %>% rownameToFirstColumn('Region') %>% dplyr::rename(CONTROL.AAD = DF) %>%
+  kable(row.names=F)
 ```

### 2017-05-14 code/Rmd/regionLevelDiffExpAnlzSampled.Rmd
Added markdown to perform differential expression in a region specific manner (balanced sampling)

--- 2017-05-10 code/Rmd/regionLevelDiffExpAnlz.Rmd
@@
 ---
-title: "Brain region specific differential expression analysis (all 7 regions/one region at a time)"
+title: "Brain region specific differential expression analysis (all 7 regions/one region at a time/balanced)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./regionLevelDiffExpAnlz.Rmd", 
+knit2synapse::knitToFolderEntity(file = "./regionLevelDiffExpAnlzSampled.Rmd", 
                                  parentId ="syn9622393",
-                                 entityName = "Differential Expression Analysis (per brain region)")
+                                 entityName = "Differential Expression Analysis (per brain region - balanced)")
 ```
@@
 
-thisFileName <- 'regionLevelDiffExpAnlz.Rmd'
+thisFileName <- 'regionLevelDiffExpAnlzSampled.Rmd'
 
@@
 kable(table(covariates$BrainRegion, covariates$Diagnosis))
+kable(table(covariates$BrainRegion, paste(covariates$Diagnosis, covariates$Gender, sep = '.')))
 
@@
 # Store consolidated data in synapse
-CODE = Folder(name = 'Differential Expression Analysis (per brain region)', parentId = parentId)
+CODE = Folder(name = 'Differential Expression Analysis (per brain region - balanced)', parentId = parentId)
 CODE = synStore(CODE)

### 2017-05-14 code/Rmd/studyLevelDiffExpAnlz.Rmd
Added markdown to perform differential expression in a study specific manner

---
title: "Study specific differential expression analysis (all 3 studies/one study at a time)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)
synapseLogin()

knit2synapse::knitToFolderEntity(file = "./studyLevelDiffExpAnlz.Rmd", 
                                 parentId ="syn9622393",
                                 entityName = "Differential Expression Analysis (per study)")
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file is

# Clear R console screen output
cat("\014")

# Load required libraries
library(CovariateAnalysis) # install from th1vairam repo using devtools::install_github('th1vairam/CovariateAnalysis@devtools')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(limma)
library(edgeR)
library(cqn)
library(knitr)
library(psych)

synapseLogin()

library(foreach)
library(doParallel)

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn9622393';
activityName = 'Differential expression analysis';
activityDescription = 'Differential expression analysis in all 7 brain regions performed in a study specific manner';

thisFileName <- 'studyLevelDiffExpAnlz.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='metaAnal')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```

### Get merged covariates and raw counts from synapse
```{r download.data, fig.height=6, fig.width=10}
# Get covariates
covariates = downloadFile('syn9765937') %>%
  dplyr::filter(Diagnosis %in% c('AD', 'CONTROL')) %>%
  dplyr::mutate(Diagnosis.Gender = paste(Diagnosis, Gender, sep = '.'))
all.used.ids = 'syn9765625'

kable(table(covariates$BrainRegion, covariates$Diagnosis))
kable(table(covariates$BrainRegion, paste(covariates$Diagnosis, covariates$Gender, sep = '.')))

# Get raw counts
counts = downloadFile('syn9765938') %>%
  tidyr::separate(Gene.ID, c('ensembl_gene_id', 'position'), sep = '\\.')
rownames(counts) = counts$ensembl_gene_id
counts = counts[, covariates$SampleID]
all.used.ids = c(all.used.ids, 'syn9765629')

# Get gene specific parameters from synapse
gene.param = downloadFile('syn8449369')
all.used.ids = c(all.used.ids, 'syn8449369')

gene.len = dplyr::select(gene.param, ensembl_gene_id, gene.length) %>% 
  unique()
rownames(gene.len) = gene.len$ensembl_gene_id

gene.gc = dplyr::select(gene.param, ensembl_gene_id, percentage_gc_content) %>% 
  unique()
rownames(gene.gc) = gene.gc$ensembl_gene_id
```

### Filter genes
```{r filter.genes}
# Filter genes (remove genes that are expressed less than 1 cpm in 50% of samples per diagnosis per brain region)
lcpm = cpm(counts)

filtered.gene.ids = plyr::dlply(covariates, .(BrainRegion, Diagnosis), .fun = function(cov, logcpm){
  rownames(logcpm)[which(rowSums(logcpm[,cov$SampleID] >= 1)/dim(logcpm[,cov$SampleID])[2] >= 0.5)]
}, lcpm) %>%
  Reduce(c, .) %>%
  unique %>%
  setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous")) %>%
  intersect(gene.len$ensembl_gene_id[!is.na(gene.len$gene.length)]) %>%
  intersect(gene.gc$ensembl_gene_id[!is.na(gene.gc$percentage_gc_content)])

counts = counts[filtered.gene.ids, ]
```
Choosing `r length(filtered.gene.ids)` genes expressed at 1 cpm in 50% of samples in at least one brain region and diagnosis category for further analysis

```{r fxn}
performDifferentialExpression <- function(cov, cnt, contVar, varOfInterest, grouping, gene.len, gene.gc){

  # Get design matrix
  designMat = getDesignMatrix(cov[,varOfInterest], Intercept = F)$design
  designMat = designMat[,linColumnFinder(designMat)$indepCols]

  # Compute offset for gene length and gc content
  cqn.gene.expression = cqn(cnt,
                            x = gene.gc[rownames(cnt), 'percentage_gc_content'],
                            lengths = gene.len[rownames(cnt), 'gene.length'],
                            lengthMethod = "smooth",
                            verbose = FALSE)
  cqn.gene.expression$E = cqn.gene.expression$y+cqn.gene.expression$offset

  # Identify outlier expression and assign NA values
  log.mat = apply(cqn.gene.expression$E, 1, function(x){
    mn = mean(x, na.rm = T)
    std.dev = sd(x, na.rm = T)
    return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
  }) %>% t
  cqn.gene.expression$E[log.mat] = NA

  # Estimate voom weights for dispersion control
  voom.weights = voom(cnt, design = designMat, plot=F, na.rm = T)

  # Estimate correlated effects
  correlation = parallelDuplicateCorrelation(voom.weights, design = designMat, block = grouping)

  # Fit linear model using new weights and new design
  baseModel = lmFit(cqn.gene.expression$E,
                    design = designMat,
                    weights = voom.weights$weights,
                    correlation = correlation$cor)

  # Perform differential expression
  contrasts = levels(cov[,contVar]) %>%
    paste(contVar,.,sep = '') %>%
    combn(2) %>%
    apply(2,paste, collapse = '-') %>%
    makeContrasts(contrasts = ., levels = colnames(baseModel$coefficients))
  baseModel.cntr =  contrasts.fit(baseModel, contrasts)
  baseModel.cntr = eBayes(baseModel.cntr)
  diffExp = lapply(1:dim(contrasts)[2], function(i, baseModel.cntr){
    topTable(baseModel.cntr, coef = i, number = 50000, confint = T) %>%
      rownameToFirstColumn('ensembl_gene_id')
    }, baseModel.cntr)
  names(diffExp) = colnames(contrasts)
  diffExp = diffExp %>%
    rbindlist(use.names = T, fill = T, idcol = 'Comparison') %>%
    dplyr::mutate(Comparison = gsub(contVar,'',Comparison))

  return(diffExp)
}
```

### Perform differential expression (per brain region)
```{r diff.exp}
all.results = plyr::dlply(covariates, .(Study), .fun = function(cov, cnt, gene.len, gene.gc){

  # Filter counts and covariates
  cnt = cnt[,cov$SampleID]
  grouping = cov$Donor_ID

  # Formulate covariates matrix
  cov$BrainRegion.Diagnosis = paste(cov$BrainRegion, cov$Diagnosis, sep = '.')
  FactorCov = c('BrainRegion.Diagnosis', 'Batch', 'Gender')
  ContCov = c('RIN', 'AAD', 'PMI',  'PCT_PF_READS_ALIGNED', 'PCT_CODING_BASES', 
              'PCT_INTERGENIC_BASES', 'PCT_INTRONIC_BASES', 'PCT_RIBOSOMAL_BASES')
  cov[,FactorCov] = lapply(cov[,FactorCov], as.factor)
  cov[,ContCov] = lapply(cov[,ContCov], as.numeric)
  rownames(cov) = cov$SampleID
  cov = cov[,c(FactorCov, ContCov)]
  # assocStat = getAssociationStatistics(cov)

  # Perform differential expression between case-control
  varOfInterest = c('BrainRegion.Diagnosis', 'Batch', 'Gender', 'RIN', 'AAD', 'PMI', 
                    'PCT_PF_READS_ALIGNED', 'PCT_INTERGENIC_BASES', 'PCT_INTRONIC_BASES', 
                    'PCT_RIBOSOMAL_BASES')
  contVar = 'BrainRegion.Diagnosis'

  baseDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, grouping, gene.len, gene.gc)

  # Perform differential expression between case-control in a gender specific manner
  cov$BrainRegion.Diagnosis.Gender = factor(paste(cov$BrainRegion, cov$Diagnosis, sep = '.') %>%
                                              paste(cov$Gender, sep = '.'))
  varOfInterest = c('BrainRegion.Diagnosis.Gender', 'Batch', 'RIN', 'AAD', 'PMI', 
                    'PCT_PF_READS_ALIGNED', 'PCT_INTERGENIC_BASES', 'PCT_INTRONIC_BASES', 
                    'PCT_RIBOSOMAL_BASES')
  contVar = 'BrainRegion.Diagnosis.Gender'

  genderDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, grouping, gene.len, gene.gc)

  return(rbindlist(list(baseDiffExp, genderDiffExp), use.names = T, fill = T))
}, counts, gene.len, gene.gc)

# Display results
writeLines('Number of differentially expressed genes between AD-CONTROL at an FDR of 0.05 and fold change of 1.2 are')
plyr::ldply(all.results, function(x){
  x %>%
    tidyr::separate(Comparison, c('from','to'), sep = '-') %>%
    tidyr::separate(from, c('from.BrainRegion', 'from.Diagnosis','from.a','from.Gender'), sep = '\\.') %>%
    tidyr::separate(to, c('to.BrainRegion', 'to.Diagnosis','to.a','to.Gender'), sep = '\\.') %>%
    dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), from.Diagnosis == 'AD', to.Diagnosis == 'CONTROL',
                  is.na(from.Gender), is.na(to.Gender), from.BrainRegion == to.BrainRegion) %>%
    tidyr::unite(Comparison, from.BrainRegion, from.Diagnosis, to.Diagnosis, sep = '.') %>%
    dplyr::group_by(Comparison) %>%
    dplyr::summarise(ncount = length(unique(ensembl_gene_id)))
}) %>% 
  kable(row.names=F)

writeLines('Number of differentially expressed genes between AD.FEMALE-CONTROL.FEMALE at an FDR of 0.05 and fold change of 1.2 are')
plyr::ldply(all.results, function(x){
  x %>%
    tidyr::separate(Comparison, c('from','to'), sep = '-') %>%
    tidyr::separate(from, c('from.BrainRegion', 'from.Diagnosis','from.a','from.Gender'), sep = '\\.') %>%
    tidyr::separate(to, c('to.BrainRegion', 'to.Diagnosis','to.a','to.Gender'), sep = '\\.') %>%
    dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), 
                  from.Diagnosis == 'AD', to.Diagnosis == 'CONTROL',
                  from.Gender == 'FEMALE', to.Gender == 'FEMALE', 
                  from.BrainRegion == to.BrainRegion) %>%
    tidyr::unite(Comparison, from.BrainRegion, from.Gender, from.Diagnosis, to.Diagnosis, sep = '.') %>%
    dplyr::group_by(Comparison) %>%
    dplyr::summarise(ncount = length(unique(ensembl_gene_id)))
}) %>% 
  kable(row.names=F)

writeLines('Number of differentially expressed genes between AD.MALE-CONTROL.MALE at an FDR of 0.05 and fold change of 1.2 are')
plyr::ldply(all.results, function(x){
  x %>%
    tidyr::separate(Comparison, c('from','to'), sep = '-') %>%
    tidyr::separate(from, c('from.BrainRegion', 'from.Diagnosis','from.a','from.Gender'), sep = '\\.') %>%
    tidyr::separate(to, c('to.BrainRegion', 'to.Diagnosis','to.a','to.Gender'), sep = '\\.') %>%
    dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), 
                  from.Diagnosis == 'AD', to.Diagnosis == 'CONTROL',
                  from.Gender == 'MALE', to.Gender == 'MALE', 
                  from.BrainRegion == to.BrainRegion) %>%
    tidyr::unite(Comparison, from.BrainRegion, from.Gender, from.Diagnosis, to.Diagnosis, sep = '.') %>%
    dplyr::group_by(Comparison) %>%
    dplyr::summarise(ncount = length(unique(ensembl_gene_id)))
}) %>% 
  kable(row.names=F)

writeLines('Number of differentially expressed genes between AD.FEMALE-AD.MALE at an FDR of 0.05 and fold change of 1.2 are')
plyr::ldply(all.results, function(x){
  x %>%
    tidyr::separate(Comparison, c('from','to'), sep = '-') %>%
    tidyr::separate(from, c('from.BrainRegion', 'from.Diagnosis','from.a','from.Gender'), sep = '\\.') %>%
    tidyr::separate(to, c('to.BrainRegion', 'to.Diagnosis','to.a','to.Gender'), sep = '\\.') %>%
    dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), 
                  from.Diagnosis == 'AD', to.Diagnosis == 'AD',
                  from.Gender == 'FEMALE', to.Gender == 'MALE', 
                  from.BrainRegion == to.BrainRegion) %>%
    tidyr::unite(Comparison, from.BrainRegion, from.Diagnosis, from.Gender, to.Gender, sep = '.') %>%
    dplyr::group_by(Comparison) %>%
    dplyr::summarise(ncount = length(unique(ensembl_gene_id)))
}) %>% 
  kable(row.names=F)

writeLines('Number of differentially expressed genes between CONTROL.FEMALE-CONTROL.MALE at an FDR of 0.05 and fold change of 1.2 are')
plyr::ldply(all.results, function(x){
  x %>%
    tidyr::separate(Comparison, c('from','to'), sep = '-') %>%
    tidyr::separate(from, c('from.BrainRegion', 'from.Diagnosis','from.a','from.Gender'), sep = '\\.') %>%
    tidyr::separate(to, c('to.BrainRegion', 'to.Diagnosis','to.a','to.Gender'), sep = '\\.') %>%
    dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), 
                  from.Diagnosis == 'CONTROL', to.Diagnosis == 'CONTROL',
                  from.Gender == 'FEMALE', to.Gender == 'MALE', 
                  from.BrainRegion == to.BrainRegion) %>%
    tidyr::unite(Comparison, from.BrainRegion, from.Diagnosis, from.Gender, to.Gender, sep = '.') %>%
    dplyr::group_by(Comparison) %>%
    dplyr::summarise(ncount = length(unique(ensembl_gene_id)))
}) %>% 
  kable(row.names=F)
```

### Store results in synapse
```{r synapse.store, include=FALSE, cache=FALSE}
# Store consolidated data in synapse
CODE = Folder(name = 'Differential Expression Analysis (per study)', parentId = parentId)
CODE = synStore(CODE)

# Store results in synapse
save(list = 'all.results', file = 'AllDifferentialExpressionResults.RData')
OBJ = File('AllDifferentialExpressionResults.RData', 
           name = 'All Differential Expression Results', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = all.used.ids,
               executed = thisFile,
               activityName = activityName,
               activityDescription = activityDescription)
```
### Source code in github
[Source Code](`r thisFile`)

### 2017-05-16 code/Rmd/MetaAnalysis.Rmd
Temp Commit

@@
 library(knit2synapse)
+
 synapseLogin()
@@
 knit2synapse::knitToFolderEntity(file = "./MetaAnalysis.Rmd", 
-                                 parentId ="syn5569102",
-                                 entityName = 'MetaAnalysis of Control vs AD from all three brain banks (Reprocessed RNASeq)')
+                                 parentId ="syn9622393",
+                                 entityName = 'MetaAnalysis of Control vs AD (Reprocessed RNASeq)')
 ```
@@
 GL_OBJ = synGet('syn4867851')
-ALL_USED_IDs = c(ALL_USED_IDs, GL_OBJ$properties$id)
+all.used.ids = 'syn4867851'
 load(GL_OBJ@filePath)
 
-gsets = c("GO_Biological_Process", "Reactome_2015", "WikiPathways_2015")
+gsets = c("GO_Biological_Process_2015", "Reactome_2016", "WikiPathways_2016")
 GeneSets.Enrichr = GeneSets[gsets]
 
-# Download AD related gene sets from synapse
+# Download AD genetic loci from synapse
+AD = list(GeneticLoci = synGet('syn4891675')@filePath %>%
+            read.table(header = F) %>% 
+            unlist() %>%
+            as.character())
+
+list()
 GL_OBJ = synGet('syn4893059');

### 2017-05-16 code/Rmd/regionLevelDiffExpAnlz.Rmd
Minor modifications

@@
   cov = cov[,c(FactorCov, ContCov)]
+  cov$RIN2 = cov$RIN^2
   # assocStat = getAssociationStatistics(cov)
@@
   # Perform differential expression between case-control
+  if(any(unique(cov$BrainRegion) %in% c('DLPFC'))){
     varOfInterest = c('Diagnosis', 'Batch', 'Gender', 'RIN', 'AAD', 'PMI', 
-                    'PCT_PF_READS_ALIGNED', 'PCT_INTERGENIC_BASES', 'PCT_INTRONIC_BASES', 
-                    'PCT_RIBOSOMAL_BASES')
+                    'PCT_CODING_BASES', 'PCT_INTERGENIC_BASES')
+  } else {
+    if (any(unique(cov$BrainRegion) %in% c('FP', 'STG', 'PHG', 'IFG'))){
+      varOfInterest = c('Diagnosis', 'Batch', 'Gender', 'RIN', 'AAD', 'PMI', 
+                        'PCT_CODING_BASES', 'PCT_PF_READS_ALIGNED', 'PCT_INTERGENIC_BASES', 
+                        'PCT_INTRONIC_BASES', 'RIN2')
+    } else (any(unique(cov$BrainRegion) %in% c('CER', 'TCX'))) {
+      varOfInterest = c('Diagnosis', 'Batch', 'Gender', 'RIN', 'AAD', 'PMI',
+                        'PCT_INTERGENIC_BASES', 'PCT_INTRONIC_BASES', 
+                        'PCT_RIBOSOMAL_BASES', 'PCT_CODING_BASES')
+    }
+  } 
   contVar = 'Diagnosis'
@@
 
+  # Perform differential expression between case-control in a AAD specific manner
+  aadDiffExp <- performDifferentialExpressionAAD(cov, cnt, contVar, varOfInterest, gene.len, gene.gc)
+
   # Perform differential expression between case-control in a gender specific manner
   cov$Diagnosis.Gender = factor(paste(cov$Diagnosis, cov$Gender, sep = '.'))
-  varOfInterest = c('Diagnosis.Gender', 'Batch', 'RIN', 'AAD', 'PMI', 
-                    'PCT_PF_READS_ALIGNED', 'PCT_INTERGENIC_BASES', 'PCT_INTRONIC_BASES', 
-                    'PCT_RIBOSOMAL_BASES')
+  if(any(unique(cov$BrainRegion) %in% c('DLPFC'))){
+    varOfInterest = c('Diagnosis.Gender', 'Batch', 'Gender', 'RIN', 'AAD', 'PMI', 
+                    'PCT_CODING_BASES', 'PCT_INTERGENIC_BASES')
+  } else {
+    if (any(unique(cov$BrainRegion) %in% c('FP', 'STG', 'PHG', 'IFG'))){
+      varOfInterest = c('Diagnosis.Gender', 'Batch', 'Gender', 'RIN', 'AAD', 'PMI', 
+                        'PCT_CODING_BASES', 'PCT_PF_READS_ALIGNED', 'PCT_INTERGENIC_BASES', 
+                        'PCT_INTRONIC_BASES', 'RIN2')
+    } else (any(unique(cov$BrainRegion) %in% c('CER', 'TCX'))) {
+      varOfInterest = c('Diagnosis.Gender', 'Batch', 'Gender', 'RIN', 'AAD', 'PMI',
+                        'PCT_INTERGENIC_BASES', 'PCT_INTRONIC_BASES', 
+                        'PCT_RIBOSOMAL_BASES', 'PCT_CODING_BASES')
+    }
+  }
   contVar = 'Diagnosis.Gender'
@@
   genderDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, gene.len, gene.gc)
-
-  # Perform differential expression between case-control in a AAD specific manner
-  varOfInterest = c('Diagnosis', 'Gender', 'Batch', 'RIN', 'AAD', 'PMI', 
-                    'PCT_PF_READS_ALIGNED', 'PCT_INTERGENIC_BASES', 'PCT_INTRONIC_BASES', 
-                    'PCT_RIBOSOMAL_BASES')
-  contVar = 'Diagnosis'
-
-  aadDiffExp <- performDifferentialExpressionAAD(cov, cnt, contVar, varOfInterest, gene.len, gene.gc)
 

### 2017-05-16 code/Rmd/MAYO_REPROCESSED.Rmd
Minor changes

@@
 COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
-draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
+draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left', padding  = unit(c(18,2,2,18), 'mm'))
 ```
@@
                                                  MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
+
+# Check gene biotype
+## Define biomart object
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "dec2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
+
+## Query biomart
+Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "gene_biotype"),
+                       filters = "ensembl_gene_id", 
+                       values = PROCESSED_COUNTS$filteredExprMatrix$genes,
+                       mart = mart)
+
+summary(factor(Ensemble2HGNC$gene_biotype)) %>%
+  rownameToFirstColumn('Biotype') %>%
+  dplyr::rename(fraction = DF) %>%
+  dplyr::mutate(fraction = fraction/dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]) %>%
+  dplyr::filter(fraction >= 0.01) %>%
+  kable
 ```

### 2017-05-16 code/Rmd/ROSMAP_REPROCESSED.Rmd
Minor changes

@@
 Since many covariates are correlated, re-normalising and re-adjusting COUNTS with an iterative design matrix.
-1. Adding FLOWCELL and Gender a priori to variable selection
+1. Adding Batch and msex a priori to variable selection
 2. Primary variable of interest Diagnosis is excluded from the pool of available covariates for selection

### 2017-05-17 code/Rmd/MetaAnalysis.Rmd
metananlysis temp commit

@@
 library(colorfulVennPlot)
+library(xlsx)
 
@@
             as.character())
+all.used.ids = c(all.used.ids, 'syn4891675')
 
-list()
-GL_OBJ = synGet('syn4893059');
-ALL_USED_IDs = c(ALL_USED_IDs, GL_OBJ$properties$id)
-load(GL_OBJ@filePath)
+# Download cell type markers from synapse (list from GSE73721 by Mariet)
+load(synGet('syn9818080')@filePath)
+all.used.ids = c(all.used.ids, 'syn9818080')
 
-GeneSets.CM = list(CellTypeMarkers = GeneSets[grep('Zhang', names(GeneSets))],
-                   AD = GeneSets['AD:GeneticLoci'])
-GeneSets = c(GeneSets.Enrichr, GeneSets.CM) # 
+GeneSets = c(list(CellTypeMarkers = CellTypeList,
+                AD = AD),
+                GeneSets.Enrichr)
 ```

### 2017-05-17 code/Rmd/regionLevelDiffExpAnlz.Rmd

@@
   cnt = cnt[,cov$SampleID]
+  BrainRegion = unique(cov$BrainRegion)
+  print(BrainRegion)
 
@@
   # Perform differential expression between case-control
-  if(any(unique(cov$BrainRegion) %in% c('DLPFC'))){
+  if(any(BrainRegion %in% c('DLPFC'))){
     varOfInterest = c('Diagnosis', 'Batch', 'Gender', 'RIN', 'AAD', 'PMI', 
@@
   } else {
-    if (any(unique(cov$BrainRegion) %in% c('FP', 'STG', 'PHG', 'IFG'))){
+    if (any(BrainRegion %in% c('FP', 'STG', 'PHG', 'IFG'))){
       varOfInterest = c('Diagnosis', 'Batch', 'Gender', 'RIN', 'AAD', 'PMI', 
@@
                         'PCT_INTRONIC_BASES', 'RIN2')
-    } else (any(unique(cov$BrainRegion) %in% c('CER', 'TCX'))) {
+    } else {
+      if (any(BrainRegion %in% c('CER', 'TCX'))) {
         varOfInterest = c('Diagnosis', 'Batch', 'Gender', 'RIN', 'AAD', 'PMI',
@@
                           'PCT_RIBOSOMAL_BASES', 'PCT_CODING_BASES')
+      }
     }
@@
   cov$Diagnosis.Gender = factor(paste(cov$Diagnosis, cov$Gender, sep = '.'))
-  if(any(unique(cov$BrainRegion) %in% c('DLPFC'))){
-    varOfInterest = c('Diagnosis.Gender', 'Batch', 'Gender', 'RIN', 'AAD', 'PMI', 
+  if(any(BrainRegion %in% c('DLPFC'))){
+    varOfInterest = c('Diagnosis.Gender', 'Batch', 'RIN', 'AAD', 'PMI', 
                       'PCT_CODING_BASES', 'PCT_INTERGENIC_BASES')
   } else {
-    if (any(unique(cov$BrainRegion) %in% c('FP', 'STG', 'PHG', 'IFG'))){
-      varOfInterest = c('Diagnosis.Gender', 'Batch', 'Gender', 'RIN', 'AAD', 'PMI', 
+    if (any(BrainRegion %in% c('FP', 'STG', 'PHG', 'IFG'))){
+      varOfInterest = c('Diagnosis.Gender', 'Batch', 'RIN', 'AAD', 'PMI', 
                         'PCT_CODING_BASES', 'PCT_PF_READS_ALIGNED', 'PCT_INTERGENIC_BASES', 
                         'PCT_INTRONIC_BASES', 'RIN2')
-    } else (any(unique(cov$BrainRegion) %in% c('CER', 'TCX'))) {
-      varOfInterest = c('Diagnosis.Gender', 'Batch', 'Gender', 'RIN', 'AAD', 'PMI',
+    } else {
+      if (any(BrainRegion %in% c('CER', 'TCX'))) {
+        varOfInterest = c('Diagnosis.Gender', 'Batch', 'RIN', 'AAD', 'PMI',
                           'PCT_INTERGENIC_BASES', 'PCT_INTRONIC_BASES', 
                           'PCT_RIBOSOMAL_BASES', 'PCT_CODING_BASES')
+      }
     }

### 2017-05-17 code/Rmd/MetaAnalysis.Rmd
Major changes

@@
 library(colorfulVennPlot)
-library(xlsx)
+library(rmeta)
 
@@
 
-gsets = c("GO_Biological_Process_2015", "Reactome_2016", "WikiPathways_2016")
+gsets = c("GO_Biological_Process_2015", "Reactome_2016","KEGG_2016")
 GeneSets.Enrichr = GeneSets[gsets]
@@
 
-# Download cell type markers from synapse (list from GSE73721 by Mariet)
+# Download celltype specific genesets from synapse
 load(synGet('syn9818080')@filePath)
@@
 
-GeneSets = c(list(CellTypeMarkers = CellTypeList,
-                AD = AD),
-                GeneSets.Enrichr)
+GeneSets = c(GeneSets.Enrichr, 
+             list(CellTypeMarkers = CellTypeList,
+                  AD = AD))
+
+GeneSets = do.call(c, GeneSets)
 ```
@@
 ```{r covariates}
-covariates.ids = c(ROSMAP = 'syn8456631',
-                   MSBB = 'syn8484996',
-                   MAYO = 'syn8466814')
-all.used.ids = covariates.ids
-covariates = lapply(covariates.ids, downloadFile)
-
-covariates$ROSMAP$BrainRegion = 'DLPFC'
-covariates$MSBB = covariates$MSBB %>%
-  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
-covariates$MAYO = covariates$MAYO %>%
-  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
-
-covariates = covariates %>% 
-  rbindlist(use.names = T, fill = T, idcol = 'Study') %>%
-  dplyr::select(SampleID, Diagnosis, BrainRegion)
+covariates = downloadFile('syn9765937')
+all.used.ids = c(all.used.ids, 'syn9765937')
 
@@
 tmp$Direction = 'NONE'
-tmp$Direction[tmp$logFC > log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
-tmp$Direction[tmp$logFC < -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'
+tmp$Direction[tmp$logFC >= log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
+tmp$Direction[tmp$logFC <= -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'
 
-writeLines('Number of differentially expressed genes at an FDR of 0.05')
+writeLines('Number of differentially expressed genes at an FDR of 0.05 and fold change of 1.2')
 tmp %>%
@@
 multiplot(plotlist = pl, cols = 1)
+
+diffExpSet = plyr::dlply(tmp, .(BrainRegion), .fun = function(x){
+  dplyr::filter(x, abs(logFC) >= log2(1.2), adj.P.Val <= 0.05) %>%
+    dplyr::select(hgnc_symbol) %>%
+    unlist %>%
+    as.character() %>%
+    unique() %>%
+    setdiff('')
+})
+```
+
+### Meta Analysis
+```{r metaanalysis.min.pval}
+library(metamisc)
+metaAnal <- function(diff.exp, region){
+  # Get raw p.values
+  P.Value = diff.exp %>% 
+    rbindlist(use.names = T, fill = T) %>%
+    dplyr::filter(Comparison == 'CONTROL-AD', BrainRegion %in% region) %>%
+    dplyr::select(ensembl_gene_id, P.Value, BrainRegion) %>%
+    unique() %>%
+    tidyr::spread(BrainRegion, P.Value)
+  P.Value[is.na(P.Value)] = 1
+  rownames(P.Value) = P.Value$ensembl_gene_id
+  P.Value$ensembl_gene_id = NULL
+  P.Value = data.matrix(P.Value)
+
+  # Perform fisher p-value based meta-analysis
+  fb.pval = P.Value %>%
+    fishersPvalue %>%
+    p.adjust(method = 'fdr') %>%
+    rownameToFirstColumn('gene.id') %>%
+    dplyr::rename(fb.pval = DF)
+  sum(fb.pval$fb.pval <= 0.05)
+
+  # Dersimonian based meta analysis
+  effect.size = diff.exp %>% 
+    rbindlist(use.names = T, fill = T) %>%
+    dplyr::filter(Comparison == 'CONTROL-AD', BrainRegion %in% region) %>%
+    dplyr::select(ensembl_gene_id, logFC, BrainRegion) %>%
+    unique() %>%
+    tidyr::spread(BrainRegion, logFC)
+  rownames(effect.size) = effect.size$ensembl_gene_id
+  effect.size$ensembl_gene_id = NULL
+  effect.size[is.na(effect.size)] = 0
+
+  effect.var = diff.exp %>% 
+    rbindlist(use.names = T, fill = T) %>%
+    dplyr::filter(Comparison == 'CONTROL-AD', BrainRegion %in% region) %>%
+    dplyr::mutate(sd = (CI.R - logFC)/3,
+                  var = sd^2) %>%
+    dplyr::select(ensembl_gene_id, var, BrainRegion) %>%
+    unique() %>%
+    tidyr::spread(BrainRegion, var)
+  rownames(effect.var) = effect.var$ensembl_gene_id
+  effect.var$ensembl_gene_id = NULL
+  effect.var[is.na(effect.var)] = 0
+
+  ds.metaAnal = ldply(1:dim(effect.size)[1], function(i, effect.size, effect.var){
+    es = t(effect.size[i,]); es = es[es!=0]
+    ev = t(effect.var[i,]); ev = ev[ev!=0]
+  fit = uvmeta(es, ev, model = "random", method = "MOM")
+  data.frame(ensembl_gene_id = rownames(effect.size)[i],
+             logFC = fit$results[1,1],
+             CI.L = fit$results[1,1] - 3 * sqrt(fit$results[1,2]),
+             CI.R = fit$results[1,1] + 3 * sqrt(fit$results[1,2]))
+  }, effect.size, effect.var)
+
+  tmp = dplyr::rename(fb.pval, ensembl_gene_id = gene.id) %>%
+    left_join(ds.metaAnal) %>%
+    dplyr::filter(fb.pval <= 0.05, sign(CI.L) == sign(CI.R), abs(logFC) >= log2(1.2)) %>%
+    dplyr::arrange(desc(logFC)) %>%
+    left_join(diff.exp %>%
+                rbindlist(use.names = T, fill = T) %>%
+                dplyr::select(ensembl_gene_id, hgnc_symbol) %>%
+                unique())
+}
+
+meta.anal.all= metaAnal(diff.exp, c('DLPFC', 'CER', 'TCX', 'FP', 'STG', 'PHG', 'IFG'))
+meta.anal.imp= metaAnal(diff.exp, c('DLPFC', 'TCX', 'PHG'))
+diffExpSet = c(diffExpSet, list(METANLZ.ALL = setdiff(meta.anal.all$hgnc_symbol, ''),
+                                METANLZ.T3 = setdiff(meta.anal.imp$hgnc_symbol, '')))
 ```
@@
 ```{r pathway.enrich}
-# setsToTest = rbindlist(diff.exp, use.names = T, fill = T) %>%
-#   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-#   dlply(.(BrainRegion), .fun = function(x){
-#     setdiff(unique(x$hgnc_symbol),c(''))
-#   })
+backgroundGenes = diff.exp %>%
+  rbindlist(use.names = T, fill = T) %>%
+  dplyr::select(hgnc_symbol) %>%
+  unlist() %>%
+  setdiff('')
 
-setsToTest = rbindlist(diff.exp, use.names = T, fill = T) %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) > log2(1.2)) %>%
-  dplyr::mutate(Direction = sign(logFC),
-                Direction = factor(Direction, levels = c(-1,1),
-                                   labels = c('-1' = 'DOWN', '1' = 'UP'))) %>%
-  dlply(.(BrainRegion, Direction), .fun = function(x){
-    setdiff(unique(x$hgnc_symbol),c(''))
-  })
-
-# Filter gene list
-GeneSets = mapply(function(x,y){
-  names(x) = paste(y,names(x),sep = ':')
-  return(x)
-},GeneSets, names(GeneSets), SIMPLIFY = F) %>%
-  Reduce(c,.)
-GeneSets = GeneSets[setdiff(seq(1,length(GeneSets)), c(grep('Mus musculus', names(GeneSets))))]
-
+# Filter geneset
 GeneSets = lapply(GeneSets, function(x, backgroundGenes){
@@
 
-enrichResults = plyr::ldply(setsToTest, .fun = function(genesInGeneSet, geneSets, backgroundGenes){
-  plyr::ldply(geneSets, .fun = fisherEnrichment, genesInGeneSet, backgroundGenes, .parallel = T) %>%
+enrichResults = plyr::ldply(diffExpSet, .fun = function(genesInGeneSet, geneSets, backgroundGenes){
+  plyr::ldply(geneSets, .fun = fisherEnrichment, genesInGeneSet, backgroundGenes, .parallel = T, .id = 'Category') %>%
     dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
 }, GeneSets, backgroundGenes, 
-.parallel = F, .progress = 'text', .paropts = list(.packages = c('plyr','dplyr','CovariateAnalysis')))
+.parallel = F, .progress = 'text', .paropts = list(.packages = c('plyr','dplyr','CovariateAnalysis')),
+.id = 'BrainRegion')
 
-tmp = dplyr::select(enrichResults, BrainRegion, .id, fdr) %>%
-  dplyr::filter(fdr <= 0.05) %>%
-  tidyr::spread(BrainRegion, fdr)
+tmp = dplyr::select(enrichResults, BrainRegion, Category, fdr, Odds.Ratio)
+tmp = tmp[c(grep('CellTypeMarkers', tmp$Category),
+            grep('GeneticLoci', tmp$Category)), ]
+tmp$Odds.Ratio[tmp$fdr > 0.05] = 0
+tmp = tmp %>%
+  dplyr::select(BrainRegion, Category, Odds.Ratio) %>%
+  tidyr::spread(BrainRegion, Odds.Ratio)
+rownames(tmp) = gsub('CellTypeMarkers.','',tmp$Category) %>%
+  gsub('_Zhang_Neuron_2015','',.)
+tmp$Category = NULL
 
-# Store logcpm in synapse
-write.table(tmp, file = 'enrich.tsv', sep = '\t', row.names=F, quote=F)
-OBJ = File('enrich.tsv', 
-           name = 'Pathway Enrichment Results (AD and Control)', 
-           parentId = CODE$properties$id)
-OBJ = synStore(OBJ,
-               used = as.character(c(covariates.id)),
-               executed = thisFile,
-               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
+h = ComplexHeatmap::Heatmap(tmp, name = 'Odds Ratio', col = wesanderson::wes_palette('Zissou', 4))
+draw(h, heatmap_legend_side = 'left', padding = unit(c(4,2,2,4), 'mm'))
+
+tmp = dplyr::select(enrichResults, BrainRegion, Category, fdr, Odds.Ratio)
+tmp = tmp[c(grep('Reactome', tmp$Category),
+            grep('GO', tmp$Category)), ] %>%
+  unique()
+tmp$Odds.Ratio[tmp$fdr > 0.05] = 0
+tmp = tmp %>%
+  dplyr::select(BrainRegion, Category, Odds.Ratio) %>%
+  tidyr::spread(BrainRegion, Odds.Ratio)
+tmp = tmp[rowSums(tmp[,-(1)]) != 0, c(1, which(colSums(tmp[,-(1)]) != 0) + 1)]
+rownames(tmp) = gsub('GO_Biological_Process_2015.','',tmp$Category) %>%
+  gsub('Reactome_2016.','',.)
+tmp$Category = NULL
+
+h = ComplexHeatmap::Heatmap(tmp, name = 'Odds Ratio', col = wesanderson::wes_palette('Zissou', 4))
+draw(h, heatmap_legend_side = 'left', padding = unit(c(4,2,2,4), 'mm'))
 ```
 
-### Get cell type fraction from synapse
-```{r celltype.frac, eval = FALSE}
-celltype.fraction.ids = c(ROSMAP = 'syn8124520',
-                          MSBB = 'syn8146376',
-                          MAYO = 'syn8135250')
-celltype.fraction = lapply(celltype.fraction.ids, downloadFile)
+### Gender specific differential expression (individual brain regions)
+```{r gender.specific}
+load(synGet('syn9766351')@filePath)
 
-cell.frac = mapply(function(cov, frc){
-  x = filter(frc, SampleID %in% cov$SampleID) %>%
-    dplyr::select(-`P-value`,-RMSE, -Correlation) %>%
-    tidyr::gather(celltype, fraction, -SampleID) %>%
-    left_join(cov)
-}, covariates, celltype.fraction, SIMPLIFY = F) %>%
-  rbindlist(use.names = T, fill = T) %>%
-  dplyr::mutate(Dx = factor(Dx, levels = c('Control', 'NCI', 'ND', 'SD', 'AD')))
+tmp = ldply(all.results, .fun = function(x){
+  x$Direction = 'NONE'
+  x$Direction[x$adj.P.Val <= 0.05 & x$logFC <= -log2(1.2)] = 'DOWN'
+  x$Direction[x$adj.P.Val <= 0.05 & x$logFC >= log2(1.2)] = 'UP'
 
-pl = list()
+  dplyr::filter(x, adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), Comparison == 'AD.FEMALE-CONTROL.FEMALE') %>%
+    dplyr::group_by(Direction) %>%
+    dplyr::summarise(count = n())
+}) %>%
+  dplyr::mutate(Direction = paste(Direction,'FEMALE',sep = '.')) %>%
+  tidyr::spread(Direction, count) %>%
+  left_join(
+    ldply(all.results, .fun = function(x){
+      x$Direction = 'NONE'
+      x$Direction[x$adj.P.Val <= 0.05 & x$logFC <= -log2(1.2)] = 'DOWN'
+      x$Direction[x$adj.P.Val <= 0.05 & x$logFC >= log2(1.2)] = 'UP'
 
-p = ggplot(cell.frac %>% filter(celltype %in% c('neurons', 'microglia', 'astrocytes')), 
-           aes(x = BrainRegion, y = fraction, color = Dx)) + geom_boxplot()
-p = p + facet_grid(celltype~., scales = 'free_y')
-p = p + xlab('Brain Region') + ylab('Fraction')
-p = p + theme(axis.text.x = element_text(hjust = 1, angle = 90), legend.position = 'top')
-pl[[1]] = p
+      dplyr::filter(x, adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), Comparison == 'AD.MALE-CONTROL.MALE') %>%
+        dplyr::group_by(Direction) %>%
+        dplyr::summarise(count = n())
+    }) %>%
+      dplyr::mutate(Direction = paste(Direction,'MALE',sep = '.')) %>%
+      tidyr::spread(Direction, count))
 
-p = ggplot(cell.frac %>% filter(celltype %in% c('endothelial', 'oligodendrocytes', 'OPC')), 
-           aes(x = BrainRegion, y = fraction, color = Dx)) + geom_boxplot()
-p = p + facet_grid(celltype~., scales = 'free_y')
-p = p + xlab('Brain Region') + ylab('Fraction')
-p = p + theme(axis.text.x = element_text(hjust = 1, angle = 90), legend.position = 'top')
-pl[[2]] = p
-
-multiplot(plotlist = pl, cols = 2)
-```
-
-### Get hgnc_symbol mapping from biomaRt
-```{r biomart}
-# Get background gene ids
-backgroundGenes = sapply(diff.exp, function(x) x$hgnc_symbol) %>%
-  Reduce(c,.) %>% unique() %>% setdiff(c(''))
-```
-
-### Consistently differentially expressed genes (in all 7 brain regions)
-```{r diffexp.all, fig.height=16, fig.width=10}
-# Find genes that are differentially expressed in each brain region
-diff.exp.gl = lapply(diff.exp, function(x){
-  dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05 & sign(CI.L) == sign(CI.R)) %>%
-    dplyr::select(Region, ensembl_gene_id) %>%
-    unique
-}) %>%
-  rbindlist(use.names = T, fill = T) %>%
-  dlply(.(Region), .fun = function(x){setdiff(unique(x$ensembl_gene_id),c(''))})
-
-tmp1 = rbindlist(diff.exp, use.names = T, fill = T) %>%
-  dplyr::filter(ensembl_gene_id %in% Reduce(c, diff.exp.gl))
-
-tmp1$Direction = 'NONE'
-tmp1$Direction[tmp1$adj.P.Val <= 0.05 & tmp1$logFC >= log2(1.2)] = 'UP'
-tmp1$Direction[tmp1$adj.P.Val <= 0.05 & tmp1$logFC <= -log2(1.2)] = 'DOWN'
-
-tmp = tmp1 %>%
-  group_by(ensembl_gene_id) %>%
-  summarise(count = sum(Direction != 'NONE')) %>%
-  filter(count == 7) %>%
-  left_join(dplyr::select(tmp1, ensembl_gene_id, hgnc_symbol) %>% unique)
-
-writeLines(paste('Differentially expressed genes in all 7 brain regions at an FDR of 0.05 and fold change of 1.2 are',
-                 length(unique(tmp$hgnc_symbol))))
-kable(data.frame(ngenes = length(unique(tmp$hgnc_symbol)),
-                 hgnc_symbol = tmp$hgnc_symbol %>% unique %>% paste(collapse = ',')))
-
-writeLines('Top six differentially expressed genes are')
-tmp.eff.sz = filter(tmp1, hgnc_symbol %in% tmp$hgnc_symbol) %>%
-  group_by(hgnc_symbol) %>% 
-  summarise(es = sum(logFC, na.rm = T)) %>% 
-  top_n(6, abs(es))
-
-pl = list()
-p = ggplot(tmp1 %>% 
-             filter(hgnc_symbol %in% c('MEIS3', 'PCSK1', 'ADAMTS2')), 
-           aes(x = Region, y = logFC, color = Study, fill = Direction)) + geom_bar(stat = 'identity')
-p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
-p = p + theme(legend.position = 'top') + scale_fill_manual(values = c('green', 'red'))
-pl[[1]] = p
-
-p = ggplot(tmp1 %>% 
-             filter(hgnc_symbol %in% c('FOXJ1', 'PRELP', 'S100A4')), 
-           aes(x = Region, y = logFC, color = Study, fill = Direction)) + geom_bar(stat = 'identity')
-p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
-p = p + theme(legend.position = 'top')  + scale_fill_manual(values = c('green','grey', 'red'))
-pl[[2]] = p
-
-multiplot(plotlist = pl, cols = 2)
-```
-
-### Meta Analysis
-```{r metaanalysis.min.pval}
+metaAnal.gender <- function(allResults, comp, region){
   # Get raw p.values
-P.Value = diff.exp %>% 
-  rbindlist(use.names = T, fill = T) %>%
-  dplyr::select(gene.id, P.Value, Region) %>%
+  P.Value = allResults %>%
+    rbindlist(use.names = T, fill = T, idcol = 'BrainRegion') %>%
+    dplyr::filter(Comparison == comp, BrainRegion %in% region) %>%
+    dplyr::select(ensembl_gene_id, P.Value, BrainRegion) %>%
     unique() %>%
-  tidyr::spread(Region, P.Value)
+    tidyr::spread(BrainRegion, P.Value)
   P.Value[is.na(P.Value)] = 1
-rownames(P.Value) = P.Value$gene.id
-P.Value$gene.id = NULL
+  rownames(P.Value) = P.Value$ensembl_gene_id
+  P.Value$ensembl_gene_id = NULL
   P.Value = data.matrix(P.Value)
-
-# Perform minimum p-value based meta-analysis
-min.pval = P.Value %>% 
-  apply(1, min, na.rm = T) %>%
-  p.adjust(method = 'bonferroni') %>%
-  rownameToFirstColumn('gene.id') %>%
-  dplyr::rename(min.pval = DF)
 
@@
     fishersPvalue %>%
-  p.adjust(method = 'bonferroni') %>%
+    p.adjust(method = 'fdr') %>%
     rownameToFirstColumn('gene.id') %>%
     dplyr::rename(fb.pval = DF)
-```
+  sum(fb.pval$fb.pval <= 0.05)
 
-```{r }
-# Rank prod based analysis
-expr = resid.expr[!duplicated(resid.expr$ensembl_gene_id),-c(1:2)]
-rownames(expr) = resid.expr$ensembl_gene_id[!duplicated(resid.expr$ensembl_gene_id)]
+  # Dersimonian based meta analysis
+  effect.size = allResults %>% 
+    rbindlist(use.names = T, fill = T, idcol = 'BrainRegion') %>%
+    dplyr::filter(Comparison == comp, BrainRegion %in% region) %>%
+    dplyr::select(ensembl_gene_id, logFC, BrainRegion) %>%
+    unique() %>%
+    tidyr::spread(BrainRegion, logFC)
+  rownames(effect.size) = effect.size$ensembl_gene_id
+  effect.size$ensembl_gene_id = NULL
+  effect.size[is.na(effect.size)] = 0
 
-cl = covariates$Status
-names(cl) = covariates$ID
-cl[cl == 'NCI'] = 'Control'
-cl[cl == 'ND'] = 'Control'
-cl[cl == 'SD'] = 'AD'
-cl = cl[cl %in% c('Control','AD')] %>% droplevels 
-cl.tmp = -as.numeric(cl)+2
-names(cl.tmp) = names(cl)
+  effect.var = allResults %>% 
+    rbindlist(use.names = T, fill = T, idcol = 'BrainRegion') %>%
+    dplyr::filter(Comparison == comp, BrainRegion %in% region) %>%
+    dplyr::mutate(sd = (CI.R - logFC)/3,
+                  var = sd^2) %>%
+    dplyr::select(ensembl_gene_id, var, BrainRegion) %>%
+    unique() %>%
+    tidyr::spread(BrainRegion, var)
+  rownames(effect.var) = effect.var$ensembl_gene_id
+  effect.var$ensembl_gene_id = NULL
+  effect.var[is.na(effect.var)] = 0
 
-origin = covariates$BrainRegion %>% as.numeric
-names(origin) = covariates$ID
-origin = origin[names(cl)]
+  ds.metaAnal = ldply(1:dim(effect.size)[1], function(i, effect.size, effect.var){
+    es = t(effect.size[i,]); 
+    ev = t(effect.var[i,]); 
+    es1 = es[es != 0 & ev != 0]; ev1 = ev[ev != 0 & es != 0]
+    fit = uvmeta(es1, ev1, model = "random", method = "MOM")
+    data.frame(ensembl_gene_id = rownames(effect.size)[i],
+               logFC = fit$results[1,1],
+               CI.L = fit$results[1,1] - 3 * sqrt(fit$results[1,2]),
+               CI.R = fit$results[1,1] + 3 * sqrt(fit$results[1,2]))
+  }, effect.size, effect.var)
 
-expr = expr[, names(cl)]
+  tmp = dplyr::rename(fb.pval, ensembl_gene_id = gene.id) %>%
+    left_join(ds.metaAnal) %>%
+    dplyr::filter(fb.pval <= 0.05, sign(CI.L) == sign(CI.R)) %>%
+    dplyr::arrange(desc(logFC)) %>%
+    left_join(diff.exp %>%
+                rbindlist(use.names = T, fill = T) %>%
+                dplyr::select(ensembl_gene_id, hgnc_symbol) %>%
+                unique())
+}
 
-rankprod.pfp = RPadvance(expr, cl.tmp, origin, num.perm = 100, gene.names = rownames(expr), plot=T, rand = 123456, huge=T)
-rankprod.pval = apply(rankprod.pfp$pfp, 1, min) %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  dplyr::rename(rankprod.pval = DF)
+meta.female = metaAnal.gender(all.results, comp = 'AD.FEMALE-CONTROL.FEMALE', 
+                              region = c('DLPFC', 'CER', 'TCX', 'FP', 'STG', 'PHG', 'IFG'))
+meta.male = metaAnal.gender(all.results, comp = 'AD.MALE-CONTROL.MALE',
+                            region = c('DLPFC', 'CER', 'TCX', 'FP', 'STG', 'PHG', 'IFG'))
+meta.female.imp = metaAnal.gender(all.results, comp = 'AD.FEMALE-CONTROL.FEMALE', 
+                                  region = c('DLPFC', 'TCX', 'PHG'))
+meta.male.imp = metaAnal.gender(all.results, comp = 'AD.MALE-CONTROL.MALE',
+                                region = c('DLPFC', 'TCX', 'PHG'))
 
-metaAnal = join_all(list(min.pval, max.pval, fb.pval, rankprod.pval), type = 'full')
-metaAnal$weight = -log10(metaAnal[,-c(1,6)]) %>% rowSums %>% rank
-metaAnal = metaAnal %>% arrange(desc(weight))
+diffExpSet.gender = llply(all.results, .fun = function(x){
+  dplyr::filter(x, adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), Comparison == 'AD.FEMALE-CONTROL.FEMALE') %>%
+    dplyr::left_join(diff.exp %>%
+                       rbindlist(use.names = T, fill = T) %>%
+                       dplyr::select(ensembl_gene_id, hgnc_symbol) %>%
+                       unique) %>%
+    dplyr::select(hgnc_symbol) %>%
+    unlist() %>% setdiff('')
+})
 
-writeLines('Number of genes differentially expressed between AD and control at an FDR of 0.05 are')
-tmp = metaAnal %>% 
-  dplyr::select(-weight) %>%
-  gather(method, pval, -ensembl_gene_id) %>% 
-  group_by(method) %>% 
-  filter(pval <= 0.05) %>%
-  dplyr::select(ensembl_gene_id) %>%
-  unique %>%
-  dlply(.(method), .fun = function(x){unique(x$ensembl_gene_id)})
-kable(sapply(tmp, length) %>% as.data.frame)
+diffExpSet.gender = c(diffExpSet.gender, llply(all.results, .fun = function(x){
+  dplyr::filter(x, adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), Comparison == 'AD.MALE-CONTROL.MALE') %>%
+    dplyr::left_join(diff.exp %>%
+                       rbindlist(use.names = T, fill = T) %>%
+                       dplyr::select(ensembl_gene_id, hgnc_symbol) %>%
+                       unique) %>%
+    dplyr::select(hgnc_symbol) %>%
+    unlist() %>% setdiff('')
+}))
+names(diffExpSet.gender)[1:7] = paste('FEMALE',names(diffExpSet.gender)[1:7],sep = '.')
+names(diffExpSet.gender)[8:14] = paste('MALE',names(diffExpSet.gender)[8:14],sep = '.')
+diffExpSet.gender$FEMALE.METANALZ.ALL = setdiff(meta.female$hgnc_symbol,'')
+diffExpSet.gender$MALE.METANALZ.ALL = setdiff(meta.male$hgnc_symbol,'')
+diffExpSet.gender$FEMALE.METANALZ.T3 = setdiff(meta.female.imp$hgnc_symbol,'')
+diffExpSet.gender$MALE.METANALZ.T3 = setdiff(meta.male.imp$hgnc_symbol,'')
 
-tmp1 = Venn(tmp)
+enrichResults.gender = plyr::ldply(diffExpSet.gender, .fun = function(genesInGeneSet, geneSets, backgroundGenes){
+  plyr::ldply(geneSets, .fun = fisherEnrichment, genesInGeneSet, backgroundGenes, .parallel = T, .id = 'Category') %>%
+    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
+}, GeneSets, backgroundGenes, 
+.parallel = F, .progress = 'text', .paropts = list(.packages = c('plyr','dplyr','CovariateAnalysis')),
+.id = 'BrainRegion')
 
-plotVenn3d(tmp1@IndicatorWeight[-(1),'.Weight'], labels = names(tmp))
-```
-### Heatmap of expression
-```{r heatmap}
-# Get the overlapping gene sets from all three methods
-tmp.expr = t(scale(t(expr)))
-tmp.expr = scale(tmp.expr)
-tmp.expr = tmp.expr[intersect(intersect(tmp$fb.pval, tmp$min.pval), tmp$rankprod.pval),]
+tmp = dplyr::select(enrichResults.gender, BrainRegion, Category, fdr, Odds.Ratio)
+tmp = tmp[c(grep('CellTypeMarkers', tmp$Category),
+            grep('GeneticLoci', tmp$Category)), ]
+tmp$Odds.Ratio[tmp$fdr > 0.05] = 0
+tmp = tmp %>%
+  dplyr::select(BrainRegion, Category, Odds.Ratio) %>%
+  tidyr::spread(BrainRegion, Odds.Ratio)
+rownames(tmp) = gsub('CellTypeMarkers.','',tmp$Category) %>%
+  gsub('_Zhang_Neuron_2015','',.)
+tmp$Category = NULL
 
-tmp.covariates = dplyr::select(covariates, BrainRegion, Status)
-tmp.covariates$Status[tmp.covariates$Status == 'NCI'] = 'Control'
-tmp.covariates$Status[tmp.covariates$Status == 'ND'] = 'Control'
-tmp.covariates$Status[tmp.covariates$Status == 'SD'] = 'AD'
-tmp.covariates = tmp.covariates %>% droplevels
-rownames(tmp.covariates) = covariates$ID
+h = ComplexHeatmap::Heatmap(t(tmp), name = 'Odds Ratio', col = wesanderson::wes_palette('Zissou', 4))
+draw(h, heatmap_legend_side = 'left', padding = unit(c(4,2,2,10), 'mm'))
 
-h1 = HeatmapAnnotation(tmp.covariates)
-Heatmap(tmp.expr, top_annotation = h1, show_row_names = F, show_column_names = F,
-        show_row_dend = F, show_column_dend = F)
 ```
@@
 
-# Store covariates in synapse
-write.table(covariates, file = 'ALLCovariates.tsv', sep = '\t', row.names=F, quote=F)
-OBJ = File('ALLCovariates.tsv', 
-           name = 'Covariates', 
-           parentId = CODE$properties$id)
-OBJ = synStore(OBJ,
-               used = as.character(c(covariates.id)),
-               executed = thisFile,
-               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
-
 # Store logcpm in synapse
-write.table(logcpm, file = 'ALLLogCPM.tsv', sep = '\t', row.names=F, quote=F)
-OBJ = File('ALLLogCPM.tsv', 
-           name = 'Log CPM (AD and Control)', 
-           parentId = CODE$properties$id)
-OBJ = synStore(OBJ,
-               used = as.character(c(covariates.id)),
-               executed = thisFile,
-               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
-
-# Store residual expression in synapse
-write.table(expr, file = 'ALLExpr.tsv', sep = '\t', row.names=F, quote=F)
-OBJ = File('ALLExpr.tsv', 
-           name = 'Residual Expression (AD and Control)', 
-           parentId = CODE$properties$id)
-OBJ = synStore(OBJ,
-               used = as.character(c(covariates.id)),
-               executed = thisFile,
-               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
-
-# Store diffexp in synapse
-write.table(diffexp, file = 'ALLDiffExp.tsv', sep = '\t', row.names=F, quote=F)
-OBJ = File('ALLDiffExp.tsv', 
-           name = 'Differential Expression (AD and Control)', 
-           parentId = CODE$properties$id)
-OBJ = synStore(OBJ,
-               used = as.character(c(covariates.id)),
-               executed = thisFile,
-               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
-
-# Store meta analysis in synapse
-write.table(metaAnal, file = 'ALLMetaAnal.tsv', sep = '\t', row.names=F, quote=F)
-OBJ = File('ALLMetaAnal.tsv', 
-           name = 'Meta Analysis (AD and Control)', 
+write.table(tmp, file = 'enrich.tsv', sep = '\t', row.names=F, quote=F)
+OBJ = File('enrich.tsv', 
+           name = 'Pathway Enrichment Results (AD and Control)', 
            parentId = CODE$properties$id)

### 2017-05-17 code/Rmd/MetaAnalysis.Rmd

@@
 library(rmeta)
+library(xlsx)
 
@@
 
-# Download celltype specific genesets from synapse
+# Download cell type markers from synapse (list from GSE73721 by Mariet)
 load(synGet('syn9818080')@filePath)
@@
 
-GeneSets = c(GeneSets.Enrichr, 
-             list(CellTypeMarkers = CellTypeList,
-                  AD = AD))
-
-GeneSets = do.call(c, GeneSets)
+GeneSets = c(list(CellTypeMarkers = CellTypeList,
+                AD = AD),
+                GeneSets.Enrichr)
 ```

### 2017-05-19 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Major modifications to account for new model selection

--- 2017-04-08 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
@@
 ```{r covariates.clustering}
-FactorCovariates <- c('Individual_ID', "Institution", "Gender", "RibozeroBatch", "LibraryBatch", "FlowcellBatch", "Dx.Tissue")
+FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
 ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
@@
 ```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
-indToRemove = c('MSSM_RNA_ACC_BP_24',
+# Based on PC
+indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250',
+                'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220',
+                'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
+                'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
+                'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_ACC_310",
+                "MSSM_RNA_ACC_51",  "MSSM_RNA_ACC_254", "MSSM_RNA_ACC_28", "MSSM_RNA_ACC_293", "MSSM_RNA_ACC_233",
+                "MSSM_RNA_ACC_167", "MSSM_RNA_ACC_353", "MSSM_RNA_ACC_355", "MSSM_RNA_ACC_211", "PENN_RNA_ACC_122",
+                "PENN_RNA_ACC_049", "PENN_RNA_ACC_052", "PENN_RNA_ACC_068", "PENN_RNA_ACC_070", "PITT_RNA_ACC_10020",
+                "PITT_RNA_ACC_1240", "PITT_RNA_ACC_1284", "PITT_RNA_ACC_700", "PITT_RNA_ACC_857", "PITT_RNA_ACC_BP_1589",
+                "MSSM_RNA_PFC_230", "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
+                "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")
 
-                'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250', 'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 
-                'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220', 'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133',
+# Based on number of gene outliers
+indToRemove = c(indToRemove, "MSSM_RNA_ACC_BP_15", "MSSM_RNA_ACC_175", "MSSM_RNA_ACC_224", "MSSM_RNA_ACC_240", 
+                "MSSM_RNA_ACC_285", "MSSM_RNA_ACC_344", "MSSM_RNA_ACC_376", "MSSM_RNA_ACC_374", "PENN_RNA_ACC_003", 
+                "PITT_RNA_ACC_539", "MSSM_RNA_PFC_201", "MSSM_RNA_PFC_234", "MSSM_RNA_PFC_355", "MSSM_RNA_PFC_389",
+                "MSSM_RNA_PFC_80", "MSSM_RNA_PFC_83", "PENN_RNA_PFC_22", "PENN_RNA_PFC_59", "PENN_RNA_PFC_62", 
+                "PENN_RNA_PFC_80", "PENN_RNA_PFC_105", "PITT_RNA_PFC_1472", "PITT_RNA_PFC_566", "PITT_RNA_PFC_727", 
+                "PITT_RNA_PFC_802")
 
-                'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 
-                'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56')
-
-# Outliers from Menahce's DLPFC analysis syn3493755
-# outliers = c('MSSM_RNA_BP_PFC_1', 'MSSM_RNA_PFC_21', 'MSSM_RNA_BP_PFC_3', 'MSSM_RNA_PFC_103', 'MSSM_RNA_PFC_105', 'MSSM_RNA_PFC_115', 'MSSM_RNA_PFC_169',  'MSSM_RNA_PFC_17',
-#   'MSSM_RNA_PFC_18', 'MSSM_RNA_PFC_19', 'MSSM_RNA_PFC_213', 'MSSM_RNA_PFC_214', 'MSSM_RNA_PFC_215', 'MSSM_RNA_PFC_216', 'MSSM_RNA_PFC_217', 'MSSM_RNA_PFC_218', 
-#   'MSSM_RNA_PFC_219', 'MSSM_RNA_PFC_222', 'MSSM_RNA_PFC_24', 'MSSM_RNA_PFC_252', 'MSSM_RNA_PFC_259', 'MSSM_RNA_PFC_263', 'MSSM_RNA_PFC_267', 'MSSM_RNA_PFC_275',
-#   'MSSM_RNA_PFC_307', 'MSSM_RNA_PFC_322', 'MSSM_RNA_PFC_325', 'MSSM_RNA_PFC_27', 'MSSM_RNA_PFC_332', 'MSSM_RNA_PFC_333', 'MSSM_RNA_PFC_334', 'MSSM_RNA_PFC_335',
-#   'MSSM_RNA_PFC_340', 'MSSM_RNA_PFC_35', 'MSSM_RNA_PFC_362', 'MSSM_RNA_PFC_372', 'MSSM_RNA_PFC_373', 'MSSM_RNA_PFC_385', 'MSSM_RNA_PFC_393', 'MSSM_RNA_PFC_395',
-#   'MSSM_RNA_PFC_49', 'MSSM_RNA_PFC_54', 'MSSM_RNA_PFC_70', 'MSSM_RNA_PFC_76', 'MSSM_RNA_PFC_77', 'MSSM_RNA_PFC_78', 'MSSM_RNA_PFC_88', 'MSSM_RNA_PFC_93',
-#   'FAKE_DLPFC_CMC_MSSM_426', 'PENN_RNA_PFC_29', 'FAKE_DLPFC_CMC_PENN_055', 'PENN_RNA_PFC_75', 'PENN_RNA_PFC_110', 'FAKE_DLPFC_CMC_PENN_106', 'FAKE_DLPFC_CMC_PENN_107',
-#   'FAKE_DLPFC_CMC_PENN_108', 'MSSM_RNA_PFC_14', 'MSSM_RNA_BP_PFC_21', 'MSSM_RNA_PFC_293', 'MSSM_RNA_PFC_349', 'MSSM_RNA_PFC_353', 'MSSM_RNA_PFC_363', 'MSSM_RNA_PFC_369',
-#   'PENN_RNA_PFC_82', 'PENN_RNA_PFC_89', 'PITT_RNA_PFC_1159', 'PITT_RNA_PFC_1296', 'PITT_RNA_PFC_537',  'PITT_RNA_PFC_567', 'PITT_RNA_PFC_871', 'PITT_RNA_PFC_878',
-#   'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_210',  'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_278', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_361',
-#   'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_56', 'PITT_RNA_BP_PFC_10003', 'PITT_RNA_BP_PFC_1047', 'PITT_RNA_BP_PFC_1086', 'PITT_RNA_BP_PFC_1092', 'PITT_RNA_BP_PFC_1247',
-#   'PITT_RNA_BP_PFC_1282', 'PITT_RNA_BP_PFC_1324', 'PITT_RNA_BP_PFC_1374', 'PITT_RNA_BP_PFC_1391', 'PITT_RNA_BP_PFC_686')
+# # Outliers from Menahcem's DLPFC analysis syn3493755
+# outliers = c('MSSM_RNA_BP_PFC_1', 'MSSM_RNA_PFC_21', 'MSSM_RNA_BP_PFC_3', 'MSSM_RNA_PFC_103', 
+#              'MSSM_RNA_PFC_105', 'MSSM_RNA_PFC_115', 'MSSM_RNA_PFC_169',  'MSSM_RNA_PFC_17',
+#              'MSSM_RNA_PFC_18', 'MSSM_RNA_PFC_19', 'MSSM_RNA_PFC_213', 'MSSM_RNA_PFC_214', 
+#              'MSSM_RNA_PFC_215', 'MSSM_RNA_PFC_216', 'MSSM_RNA_PFC_217', 'MSSM_RNA_PFC_218',
+#              'MSSM_RNA_PFC_219', 'MSSM_RNA_PFC_222', 'MSSM_RNA_PFC_24', 'MSSM_RNA_PFC_252', 
+#              'MSSM_RNA_PFC_259', 'MSSM_RNA_PFC_263', 'MSSM_RNA_PFC_267', 'MSSM_RNA_PFC_275',
+#              'MSSM_RNA_PFC_307', 'MSSM_RNA_PFC_322', 'MSSM_RNA_PFC_325', 'MSSM_RNA_PFC_27', 
+#              'MSSM_RNA_PFC_332', 'MSSM_RNA_PFC_333', 'MSSM_RNA_PFC_334', 'MSSM_RNA_PFC_335',
+#              'MSSM_RNA_PFC_340', 'MSSM_RNA_PFC_35', 'MSSM_RNA_PFC_362', 'MSSM_RNA_PFC_372', 
+#              'MSSM_RNA_PFC_373', 'MSSM_RNA_PFC_385', 'MSSM_RNA_PFC_393', 'MSSM_RNA_PFC_395',
+#              'MSSM_RNA_PFC_49', 'MSSM_RNA_PFC_54', 'MSSM_RNA_PFC_70', 'MSSM_RNA_PFC_76', 
+#              'MSSM_RNA_PFC_77', 'MSSM_RNA_PFC_78', 'MSSM_RNA_PFC_88', 'MSSM_RNA_PFC_93',
+#              'FAKE_DLPFC_CMC_MSSM_426', 'PENN_RNA_PFC_29', 'FAKE_DLPFC_CMC_PENN_055', 
+#              'PENN_RNA_PFC_75', 'PENN_RNA_PFC_110', 'FAKE_DLPFC_CMC_PENN_106', 'FAKE_DLPFC_CMC_PENN_107',
+#              'FAKE_DLPFC_CMC_PENN_108', 'MSSM_RNA_PFC_14', 'MSSM_RNA_BP_PFC_21', 'MSSM_RNA_PFC_293', 
+#              'MSSM_RNA_PFC_349', 'MSSM_RNA_PFC_353', 'MSSM_RNA_PFC_363', 'MSSM_RNA_PFC_369',
+#              'PENN_RNA_PFC_82', 'PENN_RNA_PFC_89', 'PITT_RNA_PFC_1159', 'PITT_RNA_PFC_1296', 'PITT_RNA_PFC_537', 
+#              'PITT_RNA_PFC_567', 'PITT_RNA_PFC_871', 'PITT_RNA_PFC_878', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139',
+#              'MSSM_RNA_PFC_210',  'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_278', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_348',
+#              'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_56', 'PITT_RNA_BP_PFC_10003', 'PITT_RNA_BP_PFC_1047',
+#              'PITT_RNA_BP_PFC_1086', 'PITT_RNA_BP_PFC_1092', 'PITT_RNA_BP_PFC_1247', 'PITT_RNA_BP_PFC_1282', 
+#              'PITT_RNA_BP_PFC_1324', 'PITT_RNA_BP_PFC_1374', 'PITT_RNA_BP_PFC_1391', 'PITT_RNA_BP_PFC_686')
 
@@
 
-Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`
+Based on the expression pattern following `r length(indToRemove)` samples were tagged as outliers based on their mrna expression: `r paste(indToRemove, collapse = ', ')`
 
@@
 
-### Normalisation (with NULL)
-Initial normalisation usign cqn and dispersion control with voom (with NULL design)
-```{r initial.voom.normalisation, fig.height=5, fig.width=5}
+### Identify gene outliers
+Set outlier expression of genes in each samples to NA
+```{r winsorise.data}
+LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)
+
+# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
+log.mat = apply(LOG.CPM, 1, function(x){
+  mn = mean(x, na.rm = T)
+  std.dev = sd(x, na.rm = T)
+  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
+}) %>% t
+PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = NA
+
+# Distribution of fraction NA genes in samples
+tmp.frac = apply(log.mat, 2, sum)/dim(LOG.CPM)[1]
+writeLines('Distribution of fraction NAs in samples')
+quantile(tmp.frac, seq(0, 1, 0.1)) %>%
+  round(3) 
+top.nsamples = names(tmp.frac)[tmp.frac >= 0.02]
+```
+Only `r length(top.nsamples)` samples have more than 2% of NA genes. They are: `r paste(top.nsamples, collapse = ', ')`
+
+### Library Normalisation
+Initial library normalisation usign cqn
+```{r cqn}
+# Function to convert NA values to zero
+changeNAtoZero <- function(mtrx){
+  mtrx[is.na(mtrx)] = 0
+  return(mtrx)
+}
+
 # Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
                           x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
@@
                           verbose = FALSE)
+CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+CQN.GENE_EXPRESSION$E[is.na(PROCESSED_COUNTS$filteredExprMatrix$counts)] = NA
+```
 
-# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
-log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
-  mn = mean(x, na.rm = T)
-  std.dev = sd(x, na.rm = T)
-  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
-}) %>% t
-PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0
-
-# Re-compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
-                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
-```
+### Co-expression distribution
 Coexpression of genes 
 ```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
+cr = cor(t(CQN.GENE_EXPRESSION$E %>% changeNAtoZero))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
 ```
-### Outlier analysis
-Clustering of voom adjusted data (with NULL design)
+
+### Sample clustering
 ```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
 # Find principal components of expression to plot
-PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)
+PC <- prcomp(CQN.GENE_EXPRESSION$E %>% changeNAtoZero, scale.=T, center = T)
 
@@
   tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
+plotdata$label = plotdata$SampleID
+plotdata$label[!(plotdata$label %in% top.nasamples)] = ''
 
@@
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+p <- p + geom_text(aes(label= label), size=4, hjust=0)
 p
+```
+Tree based clustering of samples
+```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
+# Eucledian tree based analysis
+COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
+COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
 
+tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E %>% changeNAtoZero)))
+cols = WGCNA::labels2colors(COVARIATES.tmp);
+
+WGCNA::plotDendroAndColors(tree, 
+                           colors = cols, 
+                           dendroLabels = FALSE, 
+                           abHeight = 0.80, 
+                           main = "Sample dendrogram",
+                           groupLabels = colnames(COVARIATES.tmp))
+```
+```{r temp, include = F}
+dev.off()
+gc()
+```
+
+### Distribution of samples (log cpm)
+```{r lcpm.dist, cache=FALSE}
 # Plot abberent distribution of logcpm counts
-tmp1 = (CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset) %>%
+tmp1 = (CQN.GENE_EXPRESSION$E) %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
@@
 ```
-Tree based classification of samples
-```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
-# Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
-COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
-
-tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)))
-cols = WGCNA::labels2colors(COVARIATES.tmp);
-
-WGCNA::plotDendroAndColors(tree, 
-                           colors = cols, 
-                           dendroLabels = FALSE, 
-                           abHeight = 0.80, 
-                           main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp))
-```
-
-```{r temp, include = F}
-dev.off()
-gc()
-```
 
@@
 ```{r preAdjusted.covariates, cache=TRUE}
-# Find correlation between PC's of gene expression with covariates
-preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
+preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E %>% changeNAtoZero, 
                                                  COVARIATES,
@@
 
-### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. 
-1. "Dx.Tissue" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue)
-2. Adding Instituion and Gender a priori to variable selection as they are correlated with Dx.Tissue
-```{r iterative.normalisation, results='asis'}
+### Normalisation
+Based on different model selection methodology (as seen here syn8468116, syn9772932, and syn9774491), following variables are used as covariates: Institution, Gender, RIN, and IntronicRate
+
+NOTE: Dx.Tissue is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue), and Individual\_ID is chosen as the random effect
+
+```{r custom.normalisation, results='asis'}
 # Primary variable of interest
-postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender');
+postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'RIN', 'IntronicRate')
 
-# Assign residual covariates
-residualSigCovars = preAdjustedSigCovars
-residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, 'Individual_ID'))
-covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]
-postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
-
-loopCount = 0 
-while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
 writeLines(paste('Using following covariates in the model:',
@@
 # Post adjusted design matrix
-  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
+DESIGN = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
+DESIGN = DESIGN[,linColumnFinder(DESIGN)$indepCols]
 
 # Estimate voom weights
-  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)
-  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
+                            design = DESIGN)
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
 correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
@@
 # Re-calculate voom weights with correlation of random effects
-  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                              design=DM1$design, plot=F,
+VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero,   
+                            design = DESIGN, plot=F,
                             block = COVARIATES$Individual_ID, 
                             correlation = correlation$cor)
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
 
 # Fit linear model using new weights and new design
-  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
 ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
@@
 # Residuals after normalisation
-  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset)
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
+```
 
-  # Residual covariates to choose from
-  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, 'Individual_ID'))
-
-  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                                COVARIATES[, residCovars, drop=F], 
-                                                'adjusted design(voom-normalized)',
-                                                isKeyPlot=TRUE)
-
-  # Add postadjusted covariates (if any)
-  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
-                                                                          c(postAdjustCovars, 'Individual_ID'))]
-
-  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
-  loopCount = loopCount + 1
-}
-
-modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
-                  'as fixed effects')
-
-tmp <- paste('Using following covariates in the final model:', modelStr)
-```
-`r tmp`
-
-### Sanity check
-```{r residualSigCovars.manual, fig.width=12, fig.height=8}
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                              COVARIATES, 
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
-
-residualSigCovars[["PC_res"]][[2]]$plotData
-```
 ### Residual calculation
-Calculate weighted residuals and add back "Dx" to the residuals
+Calculate weighted residuals and add back "Dx.Tissue" to the residuals
 ```{r varsToAddBack}
 # Add variable of interest back to the residuals
-varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
+varsToAddIn = grep("Dx.Tissue", colnames(DESIGN), value = T)
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
-  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DESIGN[,varsToAddIn])
 ```
@@
 ```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.GENE_EXPRESSION))
+cr = cor(t(RESIDUAL.GENE_EXPRESSION %>% changeNAtoZero))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION %>% changeNAtoZero, scale.=T, center = T)
 
@@
 plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
+  dplyr::mutate(label = SampleID)
+plotdata$label[!(plotdata$SampleID %in% top.nsamples)] = ''
 
@@
 p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
-p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
+p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
+p <- p + geom_text(aes(label= label), size=4, hjust=0)
 p
@@
 
-tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
+tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION %>% changeNAtoZero)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 ```{r diffExp, fig.height=10, fig.width=15}
-# Post adjusted design matrix
-DESIGN = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
-
-# Estimate voom weights
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-correlation <- parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Individual_ID)
-
-# Re-calculate voom weights with correlation of random effects
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                    design=DESIGN$design, plot=F,
-                    block = COVARIATES$Individual_ID, 
-                    correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-FIT = lmFit(VOOM.WEIGHTS)
-
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueControl_ACC-Dx.TissueOther_ACC",
-                                     "Dx.TissueControl_ACC-Dx.TissueSCZ_ACC",
-                                     "Dx.TissueOther_ACC-Dx.TissueSCZ_ACC",
-                                     "Dx.TissueControl_DLPFC-Dx.TissueOther_DLPFC",
-                                     "Dx.TissueControl_DLPFC-Dx.TissueSCZ_DLPFC",
-                                     "Dx.TissueOther_DLPFC-Dx.TissueSCZ_DLPFC"),
-                         levels = colnames(FIT$coefficients))
-FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
+                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
+                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
+                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC",
+                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueControl_ACC",
+                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_ACC"),
+                         levels = colnames(ADJUSTED.FIT$coefficients))
+FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
 FIT.CONTR = eBayes(FIT.CONTR)
@@
 # Get differnetial expression
-DE = lapply(1:6, function(i, FIT){
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
   topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
@@
 }, FIT.CONTR) 
-names(DE) = colnames(contrast)
+names(DE) = gsub('Dx.Tissue', '', colnames(contrast))
+names(DE)[7:8] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC')
 
@@
   dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
-                Direction = logFC/abs(logFC),
+                Direction = sign(logFC),
                 Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                 Direction = as.character(Direction)) %>%
-  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
-  tidyr::separate(ref.state, into = c('Dx1','Tissue'), sep = '_') %>%
-  tidyr::separate(to.state, into = c('Dx2','Tissue'), sep = '_') %>%
-  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_') %>%
+  tidyr::separate(Comparison, into = c('to.state','ref.state'), sep = '-') %>%
+  tidyr::separate(to.state, into = c('to.Dx','to.Tissue'), sep = '_') %>%
+  tidyr::separate(ref.state, into = c('ref.Dx','ref.Tissue'), sep = '_') %>%
   left_join(Ensemble2HGNC) %>%
@@
 tmp = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
-  group_by(Comparison, Tissue) %>%
+  dplyr::filter(adj.P.Val <= 0.05, ref.Dx != 'Other', to.Dx != 'Other') %>%
+  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
+  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
   dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
@@
 tmp1 = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
-  group_by(Comparison, Tissue) %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), ref.Dx != 'Other', to.Dx != 'Other') %>%
+  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
+  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
   dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
@@
 
-p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
-p = p + scale_color_manual(values = c('green','grey','red'))
-p = p + facet_grid(Tissue+.~Comparison, scales = 'fixed')
+p = DE %>%
+  dplyr::filter(ref.Dx != 'Other', to.Dx != 'Other', ref.Dx != 'ALL', to.Dx != 'ALL') %>%
+  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
+p = p + facet_grid(ref.Dx+.~to.Dx+to.Tissue, scales = 'fixed')
+p = p + ggtitle('Differential Expression between Schizophrenia and Control')
+p
+
+p = DE %>%
+  dplyr::filter(ref.Dx == 'ALL', to.Dx == 'ALL') %>%
+  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
+p = p + facet_grid(ref.Tissue+.~to.Tissue, scales = 'fixed')
+p = p + ggtitle('Differential Expression between DLPFC and ACC')
 p
@@
 pl = list()
-pl[[1]] = ggplot(DE, aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
+pl[[1]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
+                 aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
 pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[1]] = pl[[1]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
 pl[[1]] = pl[[1]] + theme(legend.position = 'top')
 
-pl[[2]] = ggplot(DE, aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
+pl[[2]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
+                 aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
 pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[2]] = pl[[2]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
 pl[[2]] = pl[[2]] + theme(legend.position = 'top')
 
-pl[[3]] = ggplot(DE, aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
+pl[[3]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
+                 aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
 pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[3]] = pl[[3]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
 pl[[3]] = pl[[3]] + theme(legend.position = 'top')
@@
 diffexp.genes = DE %>%
-  dlply(.(Comparison, Tissue), .fun = function(x){
+  dlply(.(ref.Dx, ref.Tissue, to.Dx, to.Tissue), .fun = function(x){
     tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
@@
   rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
-  tidyr::separate(Comparison, c('Name', 'Region'), sep = '\\.') %>%
-  dplyr::rename(Comparison = Name, Category = .id)
+  dplyr::rename(Category = .id)
 ```
@@
 # Store logCPM
-CQN.GENE_EXPRESSION$y %>%
+CQN.GENE_EXPRESSION$E %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
@@
 # Store design matrix
-DESIGN$design %>%
+DESIGN %>%
   rownameToFirstColumn('SampleID') %>%

### 2017-05-19 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM_SVA.Rmd
Major modifications to account for new model selection with SVA

--- 2016-11-15 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM_SVA.Rmd
@@
 
-cl = makeCluster(14)
+cl = makeCluster(detectCores()-2)
 registerDoParallel(cl)
@@
 # Download reprocessed counts (ACC)
-COUNT_ACC_ID = 'syn7217917';
+COUNT_ACC_ID = 'syn8414424'
 ALL_USED_IDs = COUNT_ACC_ID
-COUNT_ACC = read.table(synGet(COUNT_ACC_ID)@filePath, header=T, sep='\t', check.names = F, nrows = 56632) %>%
-  dplyr::select(-chr, -start, -end, -strand, -length)
+COUNT_ACC = downloadFile(COUNT_ACC_ID)
 
 # Download reprocessed counts (DLPFC)
-COUNT_DLPFC_ID = 'syn7217916';
+COUNT_DLPFC_ID = 'syn8413221';
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = COUNT_DLPFC_ID
-COUNT_DLPFC = read.table(synGet(COUNT_DLPFC_ID)@filePath, header=T, sep='\t', check.names = F, nrows = 56632) %>%
-  dplyr::select(-chr, -start, -end, -strand, -length)
+COUNT_DLPFC = downloadFile(COUNT_DLPFC_ID)
 
-COUNT = join_all(list(COUNT_ACC, COUNT_DLPFC), type = 'full')
-rownames(COUNT) = COUNT$geneID
-COUNT$geneID = NULL
-COUNT = data.matrix(COUNT)
+# Join ACC and DLPFC together
+COUNT = join_all(list(COUNT_ACC, COUNT_DLPFC), type = 'full') %>%
+  as.data.frame()
 
-# Get merged metadata
-METADATA_ID = 'syn2299154'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
-METADATA = read.csv(synGet(METADATA_ID)@filePath, fill = T, header = F)
+GENE.LEN = dplyr::select(COUNT_ACC, Geneid, Length) %>%
+  dplyr::rename(ensembl_gene_id = Geneid) %>%
+  unique()
+rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id
+
+rownames(COUNT) = COUNT$Geneid
+ind = setdiff(colnames(COUNT), c('Geneid', 'Chr', 'Start', 'End', 'Strand', 'Length'))
+COUNT = data.matrix(COUNT[,ind])
 
@@
 
+# Get RNA qc metrics from synapse
+METADATA_QC_ACC_ID = 'syn2929053'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_ACC_ID
+METADATA_QC_ACC = read.table(synGet(METADATA_QC_ACC_ID)@filePath, fill = T, header = T, sep = ',') %>%
+  lapply(as.character)
+
+# Get RNA qc metrics from synapse
+METADATA_QC_DLPFC_ID = 'syn2279446'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
+METADATA_QC_DLPFC = read.table(synGet(METADATA_QC_DLPFC_ID)@filePath, fill = T, header = T, sep = ',') %>%
+  lapply(as.character)
+
+METADATA_QC = rbindlist(list(METADATA_QC_ACC, METADATA_QC_DLPFC), use.names = T, fill = T) %>%
+  dplyr::filter(is.na(Exclude.)) %>%
+  dplyr::select(Sample.RNA.ID, Ribozero.Batch, Library.Batch, Flowcell.Batch, Mapped.Reads, 
+                Intragenic.Rate, Exonic.Rate, Intronic.Rate, Intergenic.Rate, Genes.Detected, 
+                Transcripts.Detected, Expression.Profiling.Efficiency, rRNA.Rate, Total.Reads, Percent.Aligned) %>%
+  dplyr::rename(SampleID = Sample.RNA.ID, RibozeroBatch = Ribozero.Batch, LibraryBatch = Library.Batch, 
+                FlowcellBatch = Flowcell.Batch, MappedReads = Mapped.Reads, IntragenicRate = Intragenic.Rate, 
+                ExonicRate = Exonic.Rate, IntronicRate = Intronic.Rate, IntergenicRate = Intergenic.Rate, 
+                GenesDetected = Genes.Detected, TranscriptsDetected = Transcripts.Detected, 
+                ExpProfEfficiency = Expression.Profiling.Efficiency, rRNARate = rRNA.Rate, 
+                TotalReads = Total.Reads, PercentAligned = Percent.Aligned)
+
+# Get merged metadata
+METADATA_ID = 'syn4484273'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
+METADATA = read.csv(synGet(METADATA_ID)@filePath, fill = T, header = T, check.names = F)
+
 ## Metadata with specific information for RNASeq processing
-metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", 
-                          "Dx", "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
-                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?", "ACC_RNA_report: Library Batch",
+metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", "Dx",
+
+                          "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
+                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?",
+
                           "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
-                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
-                          'DNA_genotyping_report: Genotyping Sample_ID')
+                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?",
+
+                          "DNA_genotyping_report: Genotyping Sample_ID")
 
 # Filter metadata 
-ind = as.character(METADATA[1,] %>% unlist) %in% metadataVarOfInterest
+ind = colnames(METADATA) %in% metadataVarOfInterest
 METADATA = METADATA[,ind]
-colnames(METADATA) = gsub('[^[:alnum:]]','_',as.character(METADATA[1,] %>% unlist))
+colnames(METADATA) = gsub('[^[:alnum:]]','_',colnames(METADATA))
 
 METADATA_ACC = METADATA %>%
-  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, ACC_RNA_isolation__Exclude_,
-                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_,
-                ACC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
+  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
+                DNA_genotyping_report__Genotyping_Sample_ID, ACC_RNA_isolation__Exclude_, 
+                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_) %>%
   dplyr::rename(PMI = PMI__in_hours_, 
@@
                 ReportExclude = ACC_RNA_report__Exclude_,
-                LibraryBatch = ACC_RNA_report__Library_Batch,
                 GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
@@
 METADATA_DLPFC = METADATA %>%
-  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, DLPFC_RNA_isolation__Exclude_,
-                DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_,
-                DLPFC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
+  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
+                DNA_genotyping_report__Genotyping_Sample_ID, DLPFC_RNA_isolation__Exclude_, 
+                DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_) %>%
   dplyr::rename(PMI = PMI__in_hours_, 
@@
                 ReportExclude = DLPFC_RNA_report__Exclude_,
-                LibraryBatch = DLPFC_RNA_report__Library_Batch,
                 GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
@@
   droplevels() %>%
+  list(.,METADATA_QC) %>%
+  join_all %>%
   dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
-  left_join(ANCESTRY)
+  left_join(ANCESTRY) %>%
+  unique
 
@@
 
-### Covariate clustering
+```{r gene.param}
+## Get GC content from biomart
+backgroundGenes = data.frame(ensembl_gene_id = rownames(COUNT))
+
+# Define biomart object
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "dec2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
+
+# Query biomart
+Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content", "gene_biotype"),
+                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
+                       mart = mart)
+
+GENE.GC.CONT = Ensemble2HGNC %>%
+  dplyr::select(ensembl_gene_id, percentage_gc_content) %>%
+  unique
+rownames(GENE.GC.CONT) = GENE.GC.CONT$ensembl_gene_id
+```
+
+### Covariates clustering
 Determine relationship between covariates. 
 ```{r covariates.clustering}
-FactorCovariates <- c("Individual_ID", "Institution", "Gender", "LibraryBatch", "Dx.Tissue")
-ContCovariates <- c("Age_of_Death", "PMI", "RIN", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")
+FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
+ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
+                    "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
+                    "PercentAligned", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")
 
@@
 ```
-
 Correlation/association between covariates at an FDR <= 0.1
-
-```{r covariates.correlation, fig.width=8, fig.height=8}
-ind.cov = setdiff(colnames(COVARIATES), 'Individual_ID')
-COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES[,ind.cov], PVAL = 0.05)
-ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.5, row.width=0.1)
+```{r covariates.correlation, fig.width=10, fig.height=8}
+COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
+tmp = COVARIATES.CORRELATION$ESTIMATE
+tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
+h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
+draw(h, heatmap_legend_side = 'left')
 ```
 
-Explore metatdata
-
-```{r dataExplore, fig.width = 6, fig.height = 15}
-my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
+### Explore metatdata
+```{r data.explore, fig.width = 12, fig.height = 12}
+my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))
 
@@
 
-multiplot(plotlist = p, cols = 1)
+# Intronic Rate
+p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate)) + geom_boxplot()
+p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme
+
+# IntergenicRate
+p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate)) + geom_boxplot()
+p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme
+
+# Transcripts Detected
+p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TranscriptsDetected)) + geom_boxplot()
+p[[6]] = p[[6]] + ggtitle('Transcripts Detected') + my.theme
+
+# Mapped Reads
+p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot()
+p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme
+
+# PercentAligned
+p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PercentAligned)) + geom_boxplot()
+p[[8]] = p[[8]] + ggtitle('Percent Aligned') + my.theme
+
+# rRNARate
+p[[9]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate)) + geom_boxplot()
+p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme
+
+multiplot(plotlist = p, cols = 3)
+
+# Institution
+# vcd::mosaic(~ Institution + Dx.Tissue, data = COVARIATES)
+
+# Gender
+# vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
 ```
-### CPM Normalisation
-Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.
-
-Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue
+### Filter genes
+Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
 ```{r cpmnormalisation}
@@
 }, COUNT)
-genesToAnalyze = unlist(genesToAnalyze) %>% unique()
+
+genesToAnalyze = unlist(genesToAnalyze) %>% 
+  unique() %>%
+  intersect(GENE.GC.CONT$ensembl_gene_id[!is.na(GENE.GC.CONT$percentage_gc_content)]) %>%
+  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$Length)])
 
 PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
+pct.pc = PROCESSED_COUNTS$filteredExprMatrix$genes %>%
+  dplyr::rename(ensembl_gene_id = genes) %>%
+  left_join(Ensemble2HGNC) %>%
+  group_by(gene_biotype) %>%
+  summarise(fraction  = n()) %>%
+  filter(fraction > 100) %>%
+  dplyr::mutate(fraction = fraction/length(PROCESSED_COUNTS$filteredExprMatrix$genes[,1]))
 ```
-Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples
+`r dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]` genes are used for the analysis.
 
-### Normalisation (with NULL)
-Initial normalisation usign voom (with NULL design)
-```{r initial.voom.normalisation, fig.height=5, fig.width=5}
-# TMM normalisation
-TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')
+Following is the distribution of biotypes
+`r kable(pct.pc)`
 
-# Initial normalisation of gene expression
-VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
+### Outlier analysis
+Detect outlier samples based on their expression (logCPM) pattern
+```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
+# Based on PC
+indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250',
+                'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220',
+                'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
+                'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
+                'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_ACC_310",
+                "MSSM_RNA_ACC_51",  "MSSM_RNA_ACC_254", "MSSM_RNA_ACC_28", "MSSM_RNA_ACC_293", "MSSM_RNA_ACC_233",
+                "MSSM_RNA_ACC_167", "MSSM_RNA_ACC_353", "MSSM_RNA_ACC_355", "MSSM_RNA_ACC_211", "PENN_RNA_ACC_122",
+                "PENN_RNA_ACC_049", "PENN_RNA_ACC_052", "PENN_RNA_ACC_068", "PENN_RNA_ACC_070", "PITT_RNA_ACC_10020",
+                "PITT_RNA_ACC_1240", "PITT_RNA_ACC_1284", "PITT_RNA_ACC_700", "PITT_RNA_ACC_857", "PITT_RNA_ACC_BP_1589",
+                "MSSM_RNA_PFC_230", "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
+                "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")
 
-# Set gene counts in specific samples that are deviating 3 sd from other samples to NA
-log.mat = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
-  mn = mean(x, na.rm = T)
-  std.dev = sd(x, na.rm = T)
-  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
-}) %>% t
-PROCESSED_COUNTS$filteredExprMatrix[log.mat] = NA
+# Based on number of gene outliers
+indToRemove = c(indToRemove, "MSSM_RNA_ACC_BP_15", "MSSM_RNA_ACC_175", "MSSM_RNA_ACC_224", "MSSM_RNA_ACC_240", 
+                "MSSM_RNA_ACC_285", "MSSM_RNA_ACC_344", "MSSM_RNA_ACC_376", "MSSM_RNA_ACC_374", "PENN_RNA_ACC_003", 
+                "PITT_RNA_ACC_539", "MSSM_RNA_PFC_201", "MSSM_RNA_PFC_234", "MSSM_RNA_PFC_355", "MSSM_RNA_PFC_389",
+                "MSSM_RNA_PFC_80", "MSSM_RNA_PFC_83", "PENN_RNA_PFC_22", "PENN_RNA_PFC_59", "PENN_RNA_PFC_62", 
+                "PENN_RNA_PFC_80", "PENN_RNA_PFC_105", "PITT_RNA_PFC_1472", "PITT_RNA_PFC_566", "PITT_RNA_PFC_727", 
+                "PITT_RNA_PFC_802")
 
-# TMM normalisation
-TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')
+# # Outliers from Menahcem's DLPFC analysis syn3493755
+# outliers = c('MSSM_RNA_BP_PFC_1', 'MSSM_RNA_PFC_21', 'MSSM_RNA_BP_PFC_3', 'MSSM_RNA_PFC_103', 
+#              'MSSM_RNA_PFC_105', 'MSSM_RNA_PFC_115', 'MSSM_RNA_PFC_169',  'MSSM_RNA_PFC_17',
+#              'MSSM_RNA_PFC_18', 'MSSM_RNA_PFC_19', 'MSSM_RNA_PFC_213', 'MSSM_RNA_PFC_214', 
+#              'MSSM_RNA_PFC_215', 'MSSM_RNA_PFC_216', 'MSSM_RNA_PFC_217', 'MSSM_RNA_PFC_218',
+#              'MSSM_RNA_PFC_219', 'MSSM_RNA_PFC_222', 'MSSM_RNA_PFC_24', 'MSSM_RNA_PFC_252', 
+#              'MSSM_RNA_PFC_259', 'MSSM_RNA_PFC_263', 'MSSM_RNA_PFC_267', 'MSSM_RNA_PFC_275',
+#              'MSSM_RNA_PFC_307', 'MSSM_RNA_PFC_322', 'MSSM_RNA_PFC_325', 'MSSM_RNA_PFC_27', 
+#              'MSSM_RNA_PFC_332', 'MSSM_RNA_PFC_333', 'MSSM_RNA_PFC_334', 'MSSM_RNA_PFC_335',
+#              'MSSM_RNA_PFC_340', 'MSSM_RNA_PFC_35', 'MSSM_RNA_PFC_362', 'MSSM_RNA_PFC_372', 
+#              'MSSM_RNA_PFC_373', 'MSSM_RNA_PFC_385', 'MSSM_RNA_PFC_393', 'MSSM_RNA_PFC_395',
+#              'MSSM_RNA_PFC_49', 'MSSM_RNA_PFC_54', 'MSSM_RNA_PFC_70', 'MSSM_RNA_PFC_76', 
+#              'MSSM_RNA_PFC_77', 'MSSM_RNA_PFC_78', 'MSSM_RNA_PFC_88', 'MSSM_RNA_PFC_93',
+#              'FAKE_DLPFC_CMC_MSSM_426', 'PENN_RNA_PFC_29', 'FAKE_DLPFC_CMC_PENN_055', 
+#              'PENN_RNA_PFC_75', 'PENN_RNA_PFC_110', 'FAKE_DLPFC_CMC_PENN_106', 'FAKE_DLPFC_CMC_PENN_107',
+#              'FAKE_DLPFC_CMC_PENN_108', 'MSSM_RNA_PFC_14', 'MSSM_RNA_BP_PFC_21', 'MSSM_RNA_PFC_293', 
+#              'MSSM_RNA_PFC_349', 'MSSM_RNA_PFC_353', 'MSSM_RNA_PFC_363', 'MSSM_RNA_PFC_369',
+#              'PENN_RNA_PFC_82', 'PENN_RNA_PFC_89', 'PITT_RNA_PFC_1159', 'PITT_RNA_PFC_1296', 'PITT_RNA_PFC_537', 
+#              'PITT_RNA_PFC_567', 'PITT_RNA_PFC_871', 'PITT_RNA_PFC_878', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139',
+#              'MSSM_RNA_PFC_210',  'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_278', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_348',
+#              'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_56', 'PITT_RNA_BP_PFC_10003', 'PITT_RNA_BP_PFC_1047',
+#              'PITT_RNA_BP_PFC_1086', 'PITT_RNA_BP_PFC_1092', 'PITT_RNA_BP_PFC_1247', 'PITT_RNA_BP_PFC_1282', 
+#              'PITT_RNA_BP_PFC_1324', 'PITT_RNA_BP_PFC_1374', 'PITT_RNA_BP_PFC_1391', 'PITT_RNA_BP_PFC_686')
 
-# Initial normalisation of gene expression
-VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
-```
-Coexpression of genes 
-```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(VOOM.GENE_EXPRESSION$E))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
-### Outlier analysis
-Clustering of voom adjusted data (with NULL design)
-```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
 # Find principal components of expression to plot
-PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)
+PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)
 
@@
 
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
+  dplyr::mutate(label = SampleID)
+plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Dx.Tissue, size=Age_of_Death))
-p <- p + theme_bw() + theme(legend.position="top")
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
+p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
+p <- p + geom_text(aes(label= label), size=4, hjust=0)
+p
+
+# Plot abberent distribution of logcpm counts
+tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID') %>%
+              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))
+
+p = ggplot(tmp1 %>%
+             dplyr::filter(SampleID %in% indToRemove),
+           aes(x = logCPM, color = SampleID)) + geom_density() 
+p = p + theme(legend.position = 'top') + facet_grid(Dx+.~Tissue, scale = 'free')
+p
+
+indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
+PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
+COVARIATES = COVARIATES[indToRetain,]
+
+tmp = tidyr::separate(COVARIATES, Dx.Tissue, c('Dx', 'Tissue'), sep = '_') %>%
+  group_by(Dx, Tissue) %>%
+  summarise(count = n()) %>%
+  spread(Tissue, count)
+```
+Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals
+
+Based on the expression pattern following `r length(indToRemove)` samples were tagged as outliers based on their mrna expression: `r paste(indToRemove, collapse = ', ')`
+
+Distribution of samples are:
+`r kable(tmp)`
+
+### Identify gene outliers
+Set outlier expression of genes in each samples to NA
+```{r winsorise.data}
+LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)
+
+# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
+log.mat = apply(LOG.CPM, 1, function(x){
+  mn = mean(x, na.rm = T)
+  std.dev = sd(x, na.rm = T)
+  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
+}) %>% t
+PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = NA
+
+# Distribution of fraction NA genes in samples
+tmp.frac = apply(log.mat, 2, sum)/dim(LOG.CPM)[1]
+writeLines('Distribution of fraction NAs in samples')
+quantile(tmp.frac, seq(0, 1, 0.1)) %>%
+  round(3) 
+top.nsamples = names(tmp.frac)[tmp.frac >= 0.02]
+```
+Only `r length(top.nsamples)` samples have more than 2% of NA genes. They are: `r paste(top.nsamples, collapse = ', ')`
+
+### Library Normalisation
+Initial library normalisation usign cqn
+```{r cqn}
+# Function to convert NA values to zero
+changeNAtoZero <- function(mtrx){
+  mtrx[is.na(mtrx)] = 0
+  return(mtrx)
+}
+
+# Compute offset for gene length and gc content
+CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
+                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
+                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
+                          lengthMethod = "smooth", 
+                          verbose = FALSE)
+CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+CQN.GENE_EXPRESSION$E[is.na(PROCESSED_COUNTS$filteredExprMatrix$counts)] = NA
+```
+
+### Co-expression distribution
+Coexpression of genes 
+```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
+cr = cor(t(CQN.GENE_EXPRESSION$E %>% changeNAtoZero))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
+```
+
+### Sample clustering
+```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
+# Find principal components of expression to plot
+PC <- prcomp(CQN.GENE_EXPRESSION$E %>% changeNAtoZero, scale.=T, center = T)
+
+# Plot first 2 PCs
+plotdata <- data.frame(SampleID=rownames(PC$rotation), 
+                       PC1=PC$rotation[,1], 
+                       PC2=PC$rotation[,2])
+
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
+plotdata$label = plotdata$SampleID
+plotdata$label[!(plotdata$label %in% top.nasamples)] = ''
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
+p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
+p <- p + geom_text(aes(label= label), size=4, hjust=0)
 p
 ```
-Tree based classification of samples
+Tree based clustering of samples
 ```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Institution', 'Dx.Tissue')])
+COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
 
-tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
+tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E %>% changeNAtoZero)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
                            main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp[,c('Institution', 'Dx.Tissue')]))
+                           groupLabels = colnames(COVARIATES.tmp))
 ```
-
 ```{r temp, include = F}
@@
 
+### Distribution of samples (log cpm)
+```{r lcpm.dist, cache=FALSE}
+# Plot abberent distribution of logcpm counts
+tmp1 = (CQN.GENE_EXPRESSION$E) %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID') %>%
+              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))
+
+p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
+p = p + theme(legend.position = 'NONE') + facet_grid(Dx+.~Tissue, scale = 'free')
+p
+```
+
 ### Significant Covariates
@@
 ```{r preAdjusted.covariates, cache=TRUE}
-# Find correlation between PC's of gene expression with covariates
-preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
-                                                 COVARIATES[,ind.cov],
+preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E %>% changeNAtoZero, 
+                                                 COVARIATES,
                                                  'NULL design(voom-normalized)', 
@@
 
-### Normalisation (custom design)
-Normalising COUNTS with a design matrix obtained from an iterative normalisation analysis (see syn7503251 for more details).
+### Normalisation
+Based on different model selection methodology (as seen here syn8468116, syn9772932, and syn9774491), following variables are used as covariates: Institution, Gender, RIN, and IntronicRate
 
-Here Institution, Gender, RIN, Age\\_of\\_Death, PMI, EV.4 were used as fixed effects conditioned on Dx.Tissue with Individual\\_ID as random effects.
-```{r customNorm, results='asis'}
+NOTE: Dx.Tissue is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue), and Individual\_ID is chosen as the random effect
+
+```{r custom.normalisation, results='asis'}
 # Primary variable of interest
-postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'RIN', 'Age_of_Death', 'PMI', 'EV.4');
+postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'RIN', 'IntronicRate')
 
@@
                  paste(postAdjustCovars, collapse=', '),
-                 'as fixed effects and Individual_ID as random effect'))
+                 'as fixed effects and Individual_ID is chosen as random effect'))
 
 # Post adjusted design matrix
-DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
+DESIGN = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
+DESIGN = DESIGN[,linColumnFinder(DESIGN)$indepCols]
 
 # Estimate voom weights
-VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
-correlation <- parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
+VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
+                            design = DESIGN)
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
+correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
 
 # Re-calculate voom weights with correlation of random effects
-VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F,
+VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero,   
+                            design = DESIGN, plot=F,
                             block = COVARIATES$Individual_ID, 
                             correlation = correlation$cor)
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
 
 # Fit linear model using new weights and new design
-VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
+ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
 
 # Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
-
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                              COVARIATES[, ind.cov, drop=F], 
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
 ```
 
-### Sanity check
-```{r residualSigCovars.manual, fig.width=12, fig.height=8}
-residualSigCovars[["PC_res"]][[2]]$plotData
+### SVA Adjustments for eQTL analysis
+Conditioned on primary variable (Dx.Tissue) and covariates estimate surrogate variables using SVA package
+```{r sva}
+# Get (null) design matrix
+MODEL0 = DESIGN[,-grep('Dx.Tissue',colnames(DESIGN))] # Get null model by removing variable of interest
+MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
+MODEL1 = DESIGN
+MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
+
+# Source modified functions from sva package
+source('../R/irwsva.build.R')
+source('../R/f.pvalue.R')
+
+# Compute actual variance of all principal components
+tmp = svd(RESIDUAL.GENE_EXPRESSION %>% changeNAtoZero)
+actual.var = tmp$d^2/sum(tmp$d^2)
+
+# Compute permuted variance of all principal components
+permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
+  tmp.residual = t(apply(RESIDUAL.GENE_EXPRESSION %>% changeNAtoZero, 1, sample, replace = FALSE))
+  tmp = svd(tmp.residual)
+  permuted.var = tmp$d^2/sum(tmp$d^2)
+}
+permuted.var = apply(permuted.var, 2, mean)
+NUM.SV = sum(actual.var >= permuted.var[1])
+
+# Plot variance components
+var.comp = data.frame(component = 1:length(actual.var), 
+                      residual = actual.var, 
+                      permuted.residual = permuted.var) %>%
+  tidyr::gather(data, value, -component)
+p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.2)), aes(x = component, y = value, color = data)) + geom_point()
+p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
+p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
+p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
+p
+
+# Estimate surrogate variables
+SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E %>% changeNAtoZero, MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
+colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
+rownames(SURR.VAR) = rownames(MODEL1)
+
+# Estimate voom weights for dispersion control
+VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
+                        design=cbind(MODEL1, SURR.VAR), plot=F)
+VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+
+# Calculate correlation values of random effects
+correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$Individual_ID)
+
+# Re-estimate voom weights
+VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
+                        design = cbind(MODEL1, SURR.VAR), plot=F,
+                        block = COVARIATES$Individual_ID, 
+                        correlation = correlation$cor)
+VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+
+# Fit linear model using new weights and new design
+VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)
 ```
 
-### Estimate surrogate variables using SVA package
-```{r svaAdjust}
-MODEL0 = DM1$design[,-grep('Dx.Tissue',colnames(DM1$design))] # Get null model by removing variable of interest
-MODEL1 = DM1$design
-
-## Calculate the number of surrogate variables
-NUM.SV = sva::num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1,
-                     method = 'be', seed = 123456, B = 20)
-
-SURR.VAR = sva::sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0,
-                    n.sv = NUM.SV, B = 30)$sv
-colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
-
-DM1$design = cbind(MODEL1, SURR.VAR)
-
-# Estimate voom weights
-VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)
-correlation <- parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
-
-# Re-calculate voom weights with correlation of random effects
-VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F,
-                                     block = COVARIATES$Individual_ID, 
-                                     correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)
-
-# Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
-
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
-                                              COVARIATES[, ind.cov, drop=F], 
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
-```
 ### Residual calculation
@@
 ```{r varsToAddBack}
+# Residuals after normalisation
+RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
+                                                  CQN.GENE_EXPRESSION$E)
+
 # Add variable of interest back to the residuals
-varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
-  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+varsToAddIn = grep("Dx.Tissue", colnames(DESIGN), value = T)
+RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
+  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
 ```
@@
 ```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.GENE_EXPRESSION))
+cr = cor(t(RESIDUAL.SVA.GENE_EXPRESSION %>% changeNAtoZero))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
+PC <- prcomp(RESIDUAL.SVA.GENE_EXPRESSION %>% changeNAtoZero, scale.=T, center = T)
 
@@
 
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
+  dplyr::mutate(label = SampleID)
+plotdata$label[!(plotdata$SampleID %in% top.nsamples)] = ''
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape = Dx.Tissue, size=RIN)) 
-p <- p + theme_bw() + theme(legend.position="top")
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)
+p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
+p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
+p <- p + geom_text(aes(label= label), size=4, hjust=0)
 p
@@
 
-tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
+tree = hclust(as.dist(t(RESIDUAL.SVA.GENE_EXPRESSION %>% changeNAtoZero)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 ```
-
-```{r get.gene.symbols, cache=TRUE, eval=T}
-backgroundGenes = data.frame(ensembl_gene_id = rownames(COUNT))
-
-# Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "mar2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
-
-# Query biomart
-Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
-                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
-                       mart = mart)
-
-# Add gene names
-RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  left_join(Ensemble2HGNC) %>%
-  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
-
-VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  left_join(Ensemble2HGNC) %>%
-  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
-```
-
 ### Differential expression analysis
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=16, fig.width=16}
+```{r diffExp, fig.height=10, fig.width=15}
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueControl_ACC-Dx.TissueOther_ACC",
-                                     "Dx.TissueControl_ACC-Dx.TissueSCZ_ACC",
-                                     "Dx.TissueOther_ACC-Dx.TissueSCZ_ACC",
-                                     "Dx.TissueControl_DLPFC-Dx.TissueOther_DLPFC",
-                                     "Dx.TissueControl_DLPFC-Dx.TissueSCZ_DLPFC",
-                                     "Dx.TissueOther_DLPFC-Dx.TissueSCZ_DLPFC"),
-                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
-FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
+contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
+                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
+                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
+                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC",
+                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueControl_ACC",
+                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_ACC"),
+                         levels = colnames(VOOM.SVA.FIT$coefficients))
+FIT.CONTR = contrasts.fit(VOOM.SVA.FIT, contrasts=contrast)
 FIT.CONTR = eBayes(FIT.CONTR)
@@
 # Get differnetial expression
-DE = lapply(1:6, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
-  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
-    rownameToFirstColumn('ensembl_gene_id') %>%
-    left_join(Ensemble2HGNC) %>%
-    dplyr::select(ensembl_gene_id, hgnc_symbol, logFC, AveExpr, t, P.Value, adj.P.Val, B)
-}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC) 
-names(DE) = colnames(contrast)
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
+    rownameToFirstColumn('ensembl_gene_id') 
+}, FIT.CONTR) 
+names(DE) = gsub('Dx.Tissue', '', colnames(contrast))
+names(DE)[7:8] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC')
 
@@
   rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison)) %>%
-  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
-  tidyr::separate(ref.state, into = c('Dx1','Tissue'), sep = '_') %>%
-  tidyr::separate(to.state, into = c('Dx2','Tissue'), sep = '_') %>%
-  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_') %>%
-  dplyr::mutate(Direction = logFC/abs(logFC),
-                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')))
+  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
+                Direction = sign(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  tidyr::separate(Comparison, into = c('to.state','ref.state'), sep = '-') %>%
+  tidyr::separate(to.state, into = c('to.Dx','to.Tissue'), sep = '_') %>%
+  tidyr::separate(ref.state, into = c('ref.Dx','ref.Tissue'), sep = '_') %>%
+  left_join(Ensemble2HGNC) %>%
+  left_join(GENE.LEN)
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
 
 tmp = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue, Direction) %>%
-  group_by(Comparison, Tissue, Direction) %>%
-  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id))) %>%
-  spread(Direction, FDR_0_05)
+  dplyr::filter(adj.P.Val <= 0.05, ref.Dx != 'Other', to.Dx != 'Other') %>%
+  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
+  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
+  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
 
 tmp1 = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue, Direction) %>%
-  group_by(Comparison, Tissue, Direction) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
-  spread(Direction, FDR_0_05_FC_1.2)
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), ref.Dx != 'Other', to.Dx != 'Other') %>%
+  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
+  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
 
-kable(full_join(tmp %>%
-                  dplyr::rename(DOWN_FDR_0_05 = DOWN, UP_FDR_0_05 = UP),
-                tmp1 %>%
-                  dplyr::rename(DOWN_FDR_0_05_FC_1.2 = DOWN, UP_FDR_0_05_FC_1.2 = UP)))
+kable(full_join(tmp,tmp1))
 
-p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
-p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
-p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
-p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
-p = p + facet_grid(Tissue+.~Comparison, scales = 'fixed')
+p = DE %>%
+  dplyr::filter(ref.Dx != 'Other', to.Dx != 'Other', ref.Dx != 'ALL', to.Dx != 'ALL') %>%
+  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
+p = p + facet_grid(ref.Dx+.~to.Dx+to.Tissue, scales = 'fixed')
+p = p + ggtitle('Differential Expression between Schizophrenia and Control')
 p
+
+p = DE %>%
+  dplyr::filter(ref.Dx == 'ALL', to.Dx == 'ALL') %>%
+  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
+p = p + facet_grid(ref.Tissue+.~to.Tissue, scales = 'fixed')
+p = p + ggtitle('Differential Expression between DLPFC and ACC')
+p
+```
+
+### Associate differential expression results with gc content, gene length and average expression
+```{r associate.de, fig.height=10, fig.width=15}
+pl = list()
+pl[[1]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
+                 aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
+pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
+pl[[1]] = pl[[1]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[1]] = pl[[1]] + theme(legend.position = 'top')
+
+pl[[2]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
+                 aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
+pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
+pl[[2]] = pl[[2]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[2]] = pl[[2]] + theme(legend.position = 'top')
+
+pl[[3]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
+                 aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
+pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
+pl[[3]] = pl[[3]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[3]] = pl[[3]] + theme(legend.position = 'top')
+
+multiplot(plotlist = pl, cols = 3)
+```
+
+### Gene set enrichment analysis
+Using Fisher's exact test for gene set enrichment analysis
+```{r enrich.de, fig.height=16, fig.width=16}
+diffexp.genes = DE %>%
+  dlply(.(ref.Dx, ref.Tissue, to.Dx, to.Tissue), .fun = function(x){
+    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
+    tmp = tmp[!is.na(tmp)]
+  })
+diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
+backgroundGenes = unique(DE$hgnc_symbol)
+
+# Download all related genesets from synapse
+geneset.files = synQuery('select * from file where parentId == "syn3240583"')
+ALL_USED_IDs = c(ALL_USED_IDs, geneset.files$file.id)
+genesets.to.test = sapply(geneset.files$file.name, function(dataSource, geneset.files){
+  tmp = downloadFile(geneset.files$file.id[geneset.files$file.name == dataSource]) %>%
+    dlply(.(Name), .fun = function(x){
+      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
+    })
+  }, geneset.files) %>%
+  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
+  do.call(c,.)
+
+# Perform enrichment analysis
+enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
+   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
+     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
+    }, x, backgroundGenes, .parallel = TRUE) %>%
+    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
+}, genesets.to.test, backgroundGenes) %>%
+  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
+  dplyr::rename(Category = .id)
 ```
@@
 ```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
+# Set annotations
+all.annotations = list(
+  dataType = 'mRNA',
+  dataSubType = 'geneExp',
+  summaryLevel = 'gene',
+  assay	 = 'RNAseq',
+
+  tissueTypeAbrv	= 'DLPFC, ACC', 
+  study = 'CMC',
+
+  organism = 'HomoSapiens',
+  consortium	= 'CMC',
+
+  normalizationStatus	= TRUE,
+
+  normalizationType	= 'CQN+SVA'
+)
+
 # Code
 CODE <- Folder(name = "ACC and DLPFC - STAR Reprocessed - with TMM - SVA", parentId = parentId)
+annotations(CODE) = all.annotations
 CODE <- synStore(CODE)
@@
 COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
-write.table(COVARIATES, file = 'CMC_ACC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
-COV_OBJ = File('CMC_ACC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+write.table(COVARIATES, file = 'CMC_ACC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
+COV_OBJ = File('CMC_ACC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+annotations(COV_OBJ) = annotations(CODE)
+annotations(COV_OBJ)$dataSubType = 'covariates'
 COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
 
+# Store filtered counts
+PROCESSED_COUNTS$filteredExprMatrix$counts %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
+COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
+annotations(COUNT_OBJ) = annotations(CODE)
+annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
+COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
 # Store logCPM
-write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'CMC_ACC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('CMC_ACC_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
-COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                     executed = thisFile, activityDescription = activityDescription)
+CQN.GENE_EXPRESSION$y[is.na(CQN.GENE_EXPRESSION$E)] = NA
+CQN.GENE_EXPRESSION$y %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
+LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
+annotations(LCOUNT_OBJ) = annotations(CODE)
+annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
+LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
+# Store cqn offsets
+CQN.GENE_EXPRESSION$offset[is.na(CQN.GENE_EXPRESSION$E)] = NA
+CQN.GENE_EXPRESSION$offset %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  write.table(file = 'CMC_ACC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
+OFFSET_OBJ = File('CMC_ACC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
+annotations(OFFSET_OBJ) = annotations(CODE)
+annotations(OFFSET_OBJ)$dataSubType = 'offset'
+OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 
 # Store design matrix
-write.table(DM1$design, file = 'CMC_ACC_Design.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('CMC_ACC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
-DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                  executed = thisFile, activityDescription = activityDescription)
+DESIGN %>%
+  rownameToFirstColumn('SampleID') %>%
+  write.table(file = 'CMC_ACC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
+DM_OBJ = File('CMC_ACC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+annotations(DM_OBJ) = annotations(CODE)
+annotations(DM_OBJ)$dataSubType = 'designMatrix'
+DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 
-# Store residual gene expression
-write.table(RESIDUAL.GENE_EXPRESSION, file = 'CMC_ACC_Expression.tsv', sep = '\t', row.names=F, quote=F)
-EXP_OBJ = File('CMC_ACC_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', parentId = CODE$properties$id)
-EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                   executed = thisFile, activityDescription = activityDescription)
+# Store surrogate variable matrix
+SURR.VAR %>%
+  rownameToFirstColumn('SampleID') %>%
+  write.table(file = 'CMC_ACC_DLPFC_SurrVar.tsv', sep = '\t', row.names=F, quote=F)
+DM1_OBJ = File('CMC_ACC_DLPFC_Design.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
+annotations(DM1_OBJ) = annotations(CODE)
+annotations(DM1_OBJ)$dataSubType = 'surrVar'
+DM1_OBJ = synStore(DM1_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
+# Store residual gene expression for eqtl analysis
+RESIDUAL.SVA.GENE_EXPRESSION %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  write.table(file = 'CMC_ACC_DLPFC_eqtlResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+eEXP_OBJ = File('CMC_ACC_DLPFC_eqtlResidualExpression.tsv', 
+                name = 'Normalised, covariates removed residual expression (for eqtl analysis)', 
+                parentId = CODE$properties$id)
+annotations(eEXP_OBJ) = annotations(CODE)
+annotations(eEXP_OBJ)$dataSubType = 'residualGeneExpForeQTLAnlz'
+eEXP_OBJ = synStore(eEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 
 # Store differential expression results
-write.table(DE, file = 'CMC_ACC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_ACC_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
-DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                    executed = thisFile, activityDescription = activityDescription)
-stopCluster(cl)
+write.table(DE, file = 'CMC_ACC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('CMC_ACC_DLPFC_DiffExpression.tsv', 
+                name = 'Differential Expression Results (Dx.Tissue)', 
+                parentId = CODE$properties$id)
+annotations(DEXP_OBJ) = annotations(CODE)
+annotations(DEXP_OBJ)$dataSubType = 'diffExp'
+DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
+# Store enrichment results
+write.table(enrichment.results, file = 'CMC_ACC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
+ENRICH_OBJ = File('CMC_ACC_DLPFC_EnrichResults.tsv', 
+                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
+                  parentId = CODE$properties$id)
+annotations(ENRICH_OBJ) = annotations(CODE)
+annotations(ENRICH_OBJ)$dataSubType = 'enrichResults'
+ENRICH_OBJ = synStore(ENRICH_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+# stopCluster(cl)
 ```
 |  *Results*                                  |  *SynapseID*                     |
-|  -------                  |   ---------                      |
+|  -----------------------------------------  |   ---------                      |
 |  Covariates                                 |  `r COV_OBJ$properties$id`       |
-|  Raw counts (log CPM)     |  `r COUNT_OBJ$properties$id`     |
+|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
+|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
+|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
 |  Design Matrix                              |  `r DM_OBJ$properties$id`        |
-|  Residual Expression      |  `r EXP_OBJ$properties$id`       |
+|  Surrogate Variables                        |  `r DM1_OBJ$properties$id`       |
+|  Residual Expression (for eQTL analysis)    |  `r eEXP_OBJ$properties$id`      |
 |  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
+|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |

### 2017-05-19 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Major modifications to account for new model selection with SVA

@@
 
-## Explore metatdata
+### Explore metatdata
 ```{r data.explore, fig.width = 12, fig.height = 12}

### 2017-05-19 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Major modifications to account for new model selection with SVA

--- 2017-04-10 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
@@
            aes(x = logCPM, color = SampleID)) + geom_density() 
-p = p + theme(legend.position = 'NONE') + facet_grid(Dx+.~Tissue, scale = 'free')
+p = p + theme(legend.position = 'top') + facet_grid(Dx+.~Tissue, scale = 'free')
 p
@@
 
-### Normalisation (custom design)
-Since many covariates are correlated, re-normalising COUNTS with a custom design matrix. 
-1. "Dx.Tissue" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue)
-2.  Model:
-    Gx = Dx.Tissue + Institution + Gender + IntronicRate + RibozeroBatch + PMI + RIN + TranscriptsDetected +
-    Age\_of\_Death + IntergenicRate + MappedReads + PercentAligned + (1/Individual\_ID)    
+### Normalisation
+Based on different model selection methodology (as seen here syn8468116, syn9772932, and syn9774491), following variables are used as covariates: Institution, Gender, RIN, and IntronicRate
+
+NOTE: Dx.Tissue is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue), and Individual\_ID is chosen as the random effect
+
 ```{r custom.normalisation, results='asis'}
 # Primary variable of interest
-postAdjustCovars = c("Dx.Tissue", "Institution", "Gender", "IntronicRate", "RibozeroBatch", "PMI", "RIN",
-                     "TranscriptsDetected", "Age_of_Death", "IntergenicRate", "MappedReads", "PercentAligned");
+postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'RIN', 'IntronicRate')
 
@@
 # Post adjusted design matrix
-DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
-DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
+DESIGN = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
+DESIGN = DESIGN[,linColumnFinder(DESIGN)$indepCols]
 
 # Estimate voom weights
-VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
+                            design = DESIGN)
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
 correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
@@
 # Re-calculate voom weights with correlation of random effects
-VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                            design=DM1$design, plot=F,
+VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero,   
+                            design = DESIGN, plot=F,
                             block = COVARIATES$Individual_ID, 
                             correlation = correlation$cor)
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
 
 # Fit linear model using new weights and new design
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
 ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
+
+# Residuals after normalisation
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
 ```
 
-### SVA
+### SVA Adjustments for eQTL analysis
+Conditioned on primary variable (Dx.Tissue) and covariates estimate surrogate variables using SVA package
 ```{r sva}
 # Get (null) design matrix
-MODEL0 = DESIGN$design[,-grep('Dx.Tissue',colnames(DESIGN$design))] # Get null model by removing variable of interest
+MODEL0 = DESIGN[,-grep('Dx.Tissue',colnames(DESIGN))] # Get null model by removing variable of interest
 MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
-MODEL1 = DESIGN$design
+MODEL1 = DESIGN
 MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
@@
 # Source modified functions from sva package
-# source('../R/num.sv.R')
 source('../R/irwsva.build.R')
@@
 
-# Get residuals from differential expression model
-RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, VOOM.GENE_EXPRESSION$E)
-
 # Compute actual variance of all principal components
-tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
+tmp = svd(RESIDUAL.GENE_EXPRESSION %>% changeNAtoZero)
 actual.var = tmp$d^2/sum(tmp$d^2)
@@
 # Compute permuted variance of all principal components
-permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
-  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
+permuted.var = foreach(i = 1:20, .combine = rbind, .packages = c('dplyr')) %dopar% {
+  tmp.residual = t(apply(RESIDUAL.GENE_EXPRESSION %>% changeNAtoZero, 1, sample, replace = FALSE))
   tmp = svd(tmp.residual)
@@
 # Estimate surrogate variables
-SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
-                        MODEL1, MODEL0, n.sv = NUM.SV, B = 3, tol = 1e-22)$sv
-
+SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E %>% changeNAtoZero, MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
 colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
@@
 # Estimate voom weights for dispersion control
-VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
                         design=cbind(MODEL1, SURR.VAR), plot=F)
+VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
 
 # Calculate correlation values of random effects
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
 correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$Individual_ID)
@@
 # Re-estimate voom weights
-VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
                         design = cbind(MODEL1, SURR.VAR), plot=F,
-                        block = COVARIATES$Donor_ID, 
+                        block = COVARIATES$Individual_ID, 
                         correlation = correlation$cor)
+VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
 
 # Fit linear model using new weights and new design
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
 VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)
-
-# Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
-                                              CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
 ```
@@
 ### Residual calculation
-Calculate weighted residuals and add back "Dx" to the residuals
+Calculate weighted residuals and add back "Dx.Tissue" to the residuals
 ```{r varsToAddBack}
+# Residuals after normalisation
+RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
+                                                  CQN.GENE_EXPRESSION$E)
+
 # Add variable of interest back to the residuals
-varsToAddIn = grep("Dx.Tissue", colnames(VOOM.SVA.FIT$coefficients), value = T)
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
+varsToAddIn = grep("Dx.Tissue", colnames(DESIGN), value = T)
+RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
   VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
@@
 
-tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
+tree = hclust(as.dist(t(RESIDUAL.SVA.GENE_EXPRESSION %>% changeNAtoZero)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 ```
-
 ### Differential expression analysis
@@
 ```{r diffExp, fig.height=10, fig.width=15}
-# Post adjusted design matrix
-DESIGN = getDesignMatrix(cbind(COVARIATES[,postAdjustCovars,drop=F], SURR.VAR),
-                               Intercept = F)
-DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
-
-# Estimate voom weights
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-correlation <- parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Individual_ID)
-
-# Re-calculate voom weights with correlation of random effects
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                    design=DESIGN$design, plot=F,
-                    block = COVARIATES$Individual_ID, 
-                    correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-FIT = lmFit(VOOM.WEIGHTS)
-
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueControl_ACC-Dx.TissueOther_ACC",
-                                     "Dx.TissueControl_ACC-Dx.TissueSCZ_ACC",
-                                     "Dx.TissueOther_ACC-Dx.TissueSCZ_ACC",
-                                     "Dx.TissueControl_DLPFC-Dx.TissueOther_DLPFC",
-                                     "Dx.TissueControl_DLPFC-Dx.TissueSCZ_DLPFC",
-                                     "Dx.TissueOther_DLPFC-Dx.TissueSCZ_DLPFC"),
-                         levels = colnames(FIT$coefficients))
-FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
+                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
+                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
+                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC",
+                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueControl_ACC",
+                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_ACC"),
+                         levels = colnames(VOOM.SVA.FIT$coefficients))
+FIT.CONTR = contrasts.fit(VOOM.SVA.FIT, contrasts=contrast)
 FIT.CONTR = eBayes(FIT.CONTR)
@@
 # Store logCPM
+CQN.GENE_EXPRESSION$y[is.na(CQN.GENE_EXPRESSION$E)] = NA
 CQN.GENE_EXPRESSION$y %>%
@@
 # Store cqn offsets
+CQN.GENE_EXPRESSION$offset[is.na(CQN.GENE_EXPRESSION$E)] = NA
 CQN.GENE_EXPRESSION$offset %>%
@@
 
+# Store surrogate variable matrix
+SURR.VAR %>%
+  rownameToFirstColumn('SampleID') %>%
+  write.table(file = 'CMC_ACC_DLPFC_SurrVar.tsv', sep = '\t', row.names=F, quote=F)
+DM1_OBJ = File('CMC_ACC_DLPFC_Design.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
+annotations(DM1_OBJ) = annotations(CODE)
+annotations(DM1_OBJ)$dataSubType = 'surrVar'
+DM1_OBJ = synStore(DM1_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
 # Store residual gene expression for eqtl analysis
-RESIDUAL.GENE_EXPRESSION %>%
+RESIDUAL.SVA.GENE_EXPRESSION %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_ACC_DLPFC_netResidualExpression.tsv', 
-                name = 'Normalised, covariates removed residual expression (for eQTL analysis)', 
+  write.table(file = 'CMC_ACC_DLPFC_eqtlResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+eEXP_OBJ = File('CMC_ACC_DLPFC_eqtlResidualExpression.tsv', 
+                name = 'Normalised, covariates removed residual expression (for eqtl analysis)', 
                 parentId = CODE$properties$id)
-annotations(nEXP_OBJ) = annotations(CODE)
-annotations(nEXP_OBJ)$dataSubType = 'residualeQTLAnalysis'
-nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+annotations(eEXP_OBJ) = annotations(CODE)
+annotations(eEXP_OBJ)$dataSubType = 'residualGeneExpForeQTLAnlz'
+eEXP_OBJ = synStore(eEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 
@@
 |  Design Matrix                              |  `r DM_OBJ$properties$id`        |
-|  Residual Expression (for eQTL analysis)    |  `r nEXP_OBJ$properties$id`      |
+|  Surrogate Variables                        |  `r DM1_OBJ$properties$id`       |
+|  Residual Expression (for eQTL analysis)    |  `r eEXP_OBJ$properties$id`      |
 |  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

### 2017-05-22 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor sanity modifications with CMC Phase II analysis

@@
 ### Data download
-#### Obtain count matrix and metadata from synapse.
+Obtain count matrix and metadata from synapse.
 ```{r download.data, cache=TRUE}
@@
 Set outlier expression of genes in each samples to NA
-```{r winsorise.data}
+```{r detect.outlier.genes}
 LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)
@@
 # Store logCPM
-CQN.GENE_EXPRESSION$E %>%
+CQN.GENE_EXPRESSION$y[is.na(CQN.GENE_EXPRESSION$E)] = NA
+CQN.GENE_EXPRESSION$y %>%
   rownameToFirstColumn('ensembl_gene_id') %>%

### 2017-05-22 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Minor sanity modifications with CMC Phase II analysis

@@
 ```
-
 ### Data download
-#### Obtain count matrix and metadata from synapse.
+Obtain count matrix and metadata from synapse.
 ```{r download.data, cache=TRUE}
@@
 ```
-
 ### Data preprocessing
@@
 ```
-
 ```{r preprocess.data1, results='asis'}
@@
 Following sample are removed `r paste(removedIDs, collapse = ',')`
-
 ```{r gene.param}
@@
 ```
-
 ### Covariates clustering
@@
 ```
-
 ### Explore metatdata
@@
 `r kable(pct.pc)`
-
 ### Outlier analysis
@@
 `r kable(tmp)`
-
 ### Identify gene outliers
 Set outlier expression of genes in each samples to NA
-```{r winsorise.data}
+```{r detect.gene.outliers}
 LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)
@@
 Only `r length(top.nsamples)` samples have more than 2% of NA genes. They are: `r paste(top.nsamples, collapse = ', ')`
-
 ### Library Normalisation
@@
 ```
-
 ### Co-expression distribution
@@
 ```
-
 ### Sample clustering
@@
 ```
-
 ### Distribution of samples (log cpm)
@@
 ```
-
 Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
@@
 ```
-
 ### Normalisation
@@
 NOTE: Dx.Tissue is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue), and Individual\_ID is chosen as the random effect
-
 ```{r custom.normalisation, results='asis'}
@@
 ```
-
 ### SVA Adjustments for eQTL analysis
@@
 ```
-
 ### Residual calculation
@@
 ```
-
 ### Clustering residual data
@@
 ```
-
 ```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
@@
 ```
-
 ```{r temp1, include=F}
@@
 ```
-
 ### Associate differential expression results with gc content, gene length and average expression
@@
 ```
-
 ### Gene set enrichment analysis
@@
 ```
-
 ### Store files in synapse

### 2017-05-31 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor modifications

@@
                                      "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC",
+                                     "Dx.TissueSCZ_DLPFC-Dx.TissueSCZ_ACC",
+                                     "Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",
                                      "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueControl_ACC",
@@
 names(DE) = gsub('Dx.Tissue', '', colnames(contrast))
-names(DE)[7:8] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC')
+names(DE)[9:10] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC')
 
@@
 |  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |
+
+### Source Code
+[R markdown](`r thisFile`)

### 2017-05-31 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd

--- 2017-05-12 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)"
+title: "Covariate analysis of HBCC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN normlaisation)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd",
-                                 parentId ="syn7501880",
-                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection 1)')
+knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd",
+                                 parentId ="syn9846623",
+                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)')
 ```
@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = 'syn7501880';
+parentId = 'syn9846623';
 activityName = 'Covariate adjustments';
-activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (ACC and DLPFC)';
-
-thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd'
+activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (DLPFC)';
+
+thisFileName <- 'HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd'
 
@@
 ### Data download
-#### Obtain count matrix and metadata from synapse.
+Obtain count matrix and metadata from synapse.
 ```{r download.data, cache=TRUE}
-# Download reprocessed counts (ACC)
-COUNT_ACC_ID = 'syn8414424'
-ALL_USED_IDs = COUNT_ACC_ID
-COUNT_ACC = downloadFile(COUNT_ACC_ID)
-
 # Download reprocessed counts (DLPFC)
-COUNT_DLPFC_ID = 'syn8413221';
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = COUNT_DLPFC_ID
-COUNT_DLPFC = downloadFile(COUNT_DLPFC_ID)
-
-# Join ACC and DLPFC together
-COUNT = join_all(list(COUNT_ACC, COUNT_DLPFC), type = 'full') %>%
+COUNT_DLPFC_ID = 'syn8465067';
+ALL_USED_IDs = COUNT_DLPFC_ID
+COUNT = downloadFile(COUNT_DLPFC_ID) %>%
   as.data.frame()
 
-GENE.LEN = dplyr::select(COUNT_ACC, Geneid, Length) %>%
-  dplyr::rename(ensembl_gene_id = Geneid) %>%
+GENE.LEN = dplyr::select(COUNT, GeneID, Length) %>%
+  dplyr::rename(ensembl_gene_id = GeneID) %>%
   unique()
@@
 
-rownames(COUNT) = COUNT$Geneid
-ind = setdiff(colnames(COUNT), c('Geneid', 'Chr', 'Start', 'End', 'Strand', 'Length'))
+rownames(COUNT) = COUNT$GeneID
+ind = setdiff(colnames(COUNT), c('GeneID', 'Chr', 'Start', 'End', 'Strand', 'Length'))
 COUNT = data.matrix(COUNT[,ind])
@@
 # Get ancestry vector calculated using gemtools
-ANCESTRY_ID = 'syn2511399'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = ANCESTRY_ID
-ANCESTRY = read.table(synGet(ANCESTRY_ID)@filePath, fill = T, header = T, sep = '\t') %>%
-  plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'GenotypingSampleID'))
+# ANCESTRY_ID = 'syn2511399'
+# ALL_USED_IDs[length(ALL_USED_IDs)+1] = ANCESTRY_ID
+# ANCESTRY = read.table(synGet(ANCESTRY_ID)@filePath, fill = T, header = T, sep = '\t') %>%
+#   plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'GenotypingSampleID'))
 
 # Get RNA qc metrics from synapse
-METADATA_QC_ACC_ID = 'syn2929053'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_ACC_ID
-METADATA_QC_ACC = read.table(synGet(METADATA_QC_ACC_ID)@filePath, fill = T, header = T, sep = ',') %>%
-  lapply(as.character)
-
-# Get RNA qc metrics from synapse
-METADATA_QC_DLPFC_ID = 'syn2279446'
+METADATA_QC_DLPFC_ID = 'syn8539839'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
-METADATA_QC_DLPFC = read.table(synGet(METADATA_QC_DLPFC_ID)@filePath, fill = T, header = T, sep = ',') %>%
-  lapply(as.character)
-
-METADATA_QC = rbindlist(list(METADATA_QC_ACC, METADATA_QC_DLPFC), use.names = T, fill = T) %>%
-  dplyr::filter(is.na(Exclude.)) %>%
-  dplyr::select(Sample.RNA.ID, Ribozero.Batch, Library.Batch, Flowcell.Batch, Mapped.Reads, 
-                Intragenic.Rate, Exonic.Rate, Intronic.Rate, Intergenic.Rate, Genes.Detected, 
-                Transcripts.Detected, Expression.Profiling.Efficiency, rRNA.Rate, Total.Reads, Percent.Aligned) %>%
-  dplyr::rename(SampleID = Sample.RNA.ID, RibozeroBatch = Ribozero.Batch, LibraryBatch = Library.Batch, 
-                FlowcellBatch = Flowcell.Batch, MappedReads = Mapped.Reads, IntragenicRate = Intragenic.Rate, 
-                ExonicRate = Exonic.Rate, IntronicRate = Intronic.Rate, IntergenicRate = Intergenic.Rate, 
-                GenesDetected = Genes.Detected, TranscriptsDetected = Transcripts.Detected, 
-                ExpProfEfficiency = Expression.Profiling.Efficiency, rRNARate = rRNA.Rate, 
-                TotalReads = Total.Reads, PercentAligned = Percent.Aligned)
+METADATA_QC = read.table(synGet(METADATA_QC_DLPFC_ID)@filePath, fill = T, header = T, sep = ',', row.names = 1) %>%
+  rownameToFirstColumn('Sample.RNA.ID') %>%
+  dplyr::select(Sample.RNA.ID, Total_Purity_Filtered_Reads_Sequenced, X5prime_Norm, Mean_Per_Base_Cov.,  
+                Mean_CV, Fragment_Length_StdDev,  Mapped_Pairs, End_1_Mismatch_Rate, End_2_Mismatch_Rate,
+                End_1_Mapping_Rate, End_2_Mapping_Rate, Mapped, Intragenic_Rate, Exonic_Rate, Intergenic_Rate,
+                Intronic_Rate, rRNA_rate, Genes_Detected, Transcripts_Detected, Expression_Profiling_Efficiency,
+                UNIQUE_READS._Uniquely_mapped_reads_percent) %>%
+  dplyr::rename(SampleID = Sample.RNA.ID, 
+                IntragenicRate = Intragenic_Rate, 
+                ExonicRate = Exonic_Rate, 
+                IntronicRate = Intronic_Rate, 
+                IntergenicRate = Intergenic_Rate, 
+                GenesDetected = Genes_Detected, 
+                TranscriptsDetected = Transcripts_Detected, 
+                ExpProfEfficiency = Expression_Profiling_Efficiency, 
+                rRNARate = rRNA_rate, 
+                TotalReads = Total_Purity_Filtered_Reads_Sequenced, 
+                PercentMapped = UNIQUE_READS._Uniquely_mapped_reads_percent)
 
@@
 
-                          "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
-                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?",
+                          "DLPFC_RNA_report: Library Batch", "DLPFC_RNA_report: Ribozero Batch", "DLPFC_RNA_report: Flowcell Batch",
 
                           "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
-                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?",
-
-                          "DNA_genotyping_report: Genotyping Sample_ID")
+                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?")
 
@@
 
-METADATA_ACC = METADATA %>%
+METADATA = METADATA %>%
   dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
-                DNA_genotyping_report__Genotyping_Sample_ID, ACC_RNA_isolation__Exclude_, 
-                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_) %>%
-  dplyr::rename(PMI = PMI__in_hours_, 
-                IsolationExclude = ACC_RNA_isolation__Exclude_, 
-                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
-                RIN = ACC_RNA_isolation__RIN, 
-                ReportExclude = ACC_RNA_report__Exclude_,
-                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
-  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
-  dplyr::mutate(Tissue = 'ACC')
-
-METADATA_DLPFC = METADATA %>%
-  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
-                DNA_genotyping_report__Genotyping_Sample_ID, DLPFC_RNA_isolation__Exclude_, 
-                DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_) %>%
+                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, 
+                DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Ribozero_Batch, 
+                DLPFC_RNA_report__Library_Batch, DLPFC_RNA_report__Flowcell_Batch) %>%
   dplyr::rename(PMI = PMI__in_hours_, 
@@
                 ReportExclude = DLPFC_RNA_report__Exclude_,
-                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
+                RibozeroBatch = DLPFC_RNA_report__Ribozero_Batch, 
+                LibraryBatch = DLPFC_RNA_report__Library_Batch, 
+                FlowcellBatch = DLPFC_RNA_report__Flowcell_Batch) %>%
   dplyr::filter(SampleID %in% colnames(COUNT)) %>%
-  dplyr::mutate(Tissue = 'DLPFC')
-
-METADATA = rbindlist(list(METADATA_ACC, METADATA_DLPFC), use.names = T, fill = T) %>%
-  droplevels() %>%
+  dplyr::mutate(Tissue = 'DLPFC') %>%
   list(.,METADATA_QC) %>%
-  join_all %>%
-  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
-  left_join(ANCESTRY) %>%
-  unique
-
-# Replace AFF and BP samples as other
-levels(METADATA$Dx) = c("Other", "Other", "Control", "SCZ")
+  join_all %>% droplevels() %>%
+  unique %>% as.data.frame()
+METADATA$SampleID = as.character(METADATA$SampleID)
 ```
@@
 writeLines('Following counts are missing any metadata')
-writeLines(paste(setdiff(colnames(COUNT), levels(METADATA$SampleID)), collapse = ', '))
-METADATA <- METADATA %>% filter(SampleID %in% colnames(COUNT))
+writeLines(paste(setdiff(colnames(COUNT), METADATA$SampleID), collapse = ', '))
+METADATA <- METADATA %>% dplyr::filter(SampleID %in% colnames(COUNT))
 
@@
 writeLines(paste(METADATA$SampleID[METADATA$IsolationExclude == 1 | METADATA$ReportExclude == 1], collapse = ', '))
-METADATA <- METADATA  %>% filter(IsolationExclude == 0, ReportExclude == 0)
+METADATA <- METADATA  %>% dplyr::filter(IsolationExclude == 0, ReportExclude == 0)
 
@@
 writeLines(paste(METADATA$SampleID[is.na(METADATA$PMI)], collapse = ', '))
-METADATA <- METADATA  %>% filter(!is.na(PMI))
+METADATA <- METADATA  %>% dplyr::filter(!is.na(PMI))
 
@@
 writeLines(paste(METADATA$SampleID[is.na(METADATA$Gender)], collapse = ', '))
-METADATA <- METADATA  %>% filter(!is.na(Gender))
+METADATA <- METADATA  %>% dplyr::filter(!is.na(Gender))
 
@@
 writeLines(paste(METADATA$SampleID[is.na(METADATA$RIN)], collapse = ', '))
-METADATA <- METADATA  %>% filter(!is.na(RIN))
-
-writeLines('Following samples are missing EV information')
-writeLines(paste(METADATA$SampleID[is.na(METADATA$EV.1)], collapse = ', '))
-METADATA <- METADATA  %>% filter(!is.na(METADATA$EV.1)) 
+METADATA <- METADATA  %>% dplyr::filter(!is.na(RIN))
+
+writeLines(paste('Following',length(METADATA$SampleID[is.na(METADATA$TotalReads)]), 'samples are missing RNASeq QC metrics:'))
+writeLines(paste(METADATA$SampleID[is.na(METADATA$TotalReads)], collapse = ', '))
+METADATA <- METADATA  %>% dplyr::filter(!is.na(TotalReads))
+
+writeLines('Some control samples with Age of death less than 17 are removed')
+writeLines(paste('Following',length(METADATA$SampleID[METADATA$Age_of_Death <= 17]), 'samples are removed:'))
+writeLines(paste(METADATA$SampleID[METADATA$Age_of_Death <= 17], collapse = ', '))
+METADATA <- METADATA  %>% dplyr::filter(Age_of_Death >= 17) 
 ```
@@
 METADATA %>% 
-  group_by(Dx, Tissue) %>% 
-  summarise(count = n()) %>% 
-  spread(Tissue, count) %>%
+  dplyr::group_by(Dx) %>% 
+  dplyr::summarise(count = n()) %>%
   kable()
@@
 ```
-Following sample are removed `r paste(removedIDs, collapse = ',')`
+Following `r length(removedIDs)` sample are removed `r paste(removedIDs, collapse = ',')`
 
@@
 ```{r covariates.clustering}
-FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
-ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
-                    "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
-                    "PercentAligned", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")
+FactorCovariates <- c("Gender", "Dx.Tissue")
+ContCovariates <- c("Age_of_Death", "PMI", "RIN", "TotalReads", "X5prime_Norm", "Mean_Per_Base_Cov.",
+                    "Mean_CV", "Fragment_Length_StdDev", "Mapped_Pairs", "End_1_Mismatch_Rate",
+                    "End_2_Mismatch_Rate", "Mapped", "IntragenicRate", "ExonicRate", "IntergenicRate",
+                    "IntronicRate", "rRNARate", "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency",   
+                    "PercentMapped")
 
@@
 h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
-draw(h, heatmap_legend_side = 'left')
+draw(h, heatmap_legend_side = 'left', padding = unit(c(10,2,2,10), "mm"))
 ```
@@
 # Mapped Reads
-p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot()
+p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = Mapped)) + geom_boxplot()
 p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme
 
-# PercentAligned
-p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PercentAligned)) + geom_boxplot()
-p[[8]] = p[[8]] + ggtitle('Percent Aligned') + my.theme
+# ExonicRate
+p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = ExonicRate)) + geom_boxplot()
+p[[8]] = p[[8]] + ggtitle('Exonic Rate') + my.theme
 
@@
 ```{r cpmnormalisation}
+COUNT = COUNT[setdiff(rownames(COUNT), 'Geneid'),]
 genesToAnalyze = dlply(METADATA, .(Dx.Tissue), .fun = function(mtd, count){
-  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID %>% droplevels()],
+  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                    MIN_GENE_CPM=1, 
@@
 ```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
-indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250',
-                'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220',
-                'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
-                'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
-                'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_ACC_310",
-                "MSSM_RNA_ACC_51",  "MSSM_RNA_ACC_254", "MSSM_RNA_ACC_28", "MSSM_RNA_ACC_293", "MSSM_RNA_ACC_233",
-                "MSSM_RNA_ACC_167", "MSSM_RNA_ACC_353", "MSSM_RNA_ACC_355", "MSSM_RNA_ACC_211", "PENN_RNA_ACC_122",
-                "PENN_RNA_ACC_049", "PENN_RNA_ACC_052", "PENN_RNA_ACC_068", "PENN_RNA_ACC_070", "PITT_RNA_ACC_10020",
-                "PITT_RNA_ACC_1240", "PITT_RNA_ACC_1284", "PITT_RNA_ACC_700", "PITT_RNA_ACC_857", "PITT_RNA_ACC_BP_1589",
-                "MSSM_RNA_PFC_230", "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
-                "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")
-
-# Outliers from Menahce's DLPFC analysis syn3493755
-# outliers = c('MSSM_RNA_BP_PFC_1', 'MSSM_RNA_PFC_21', 'MSSM_RNA_BP_PFC_3', 'MSSM_RNA_PFC_103', 'MSSM_RNA_PFC_105', 'MSSM_RNA_PFC_115', 'MSSM_RNA_PFC_169',  'MSSM_RNA_PFC_17',
-#   'MSSM_RNA_PFC_18', 'MSSM_RNA_PFC_19', 'MSSM_RNA_PFC_213', 'MSSM_RNA_PFC_214', 'MSSM_RNA_PFC_215', 'MSSM_RNA_PFC_216', 'MSSM_RNA_PFC_217', 'MSSM_RNA_PFC_218', 
-#   'MSSM_RNA_PFC_219', 'MSSM_RNA_PFC_222', 'MSSM_RNA_PFC_24', 'MSSM_RNA_PFC_252', 'MSSM_RNA_PFC_259', 'MSSM_RNA_PFC_263', 'MSSM_RNA_PFC_267', 'MSSM_RNA_PFC_275',
-#   'MSSM_RNA_PFC_307', 'MSSM_RNA_PFC_322', 'MSSM_RNA_PFC_325', 'MSSM_RNA_PFC_27', 'MSSM_RNA_PFC_332', 'MSSM_RNA_PFC_333', 'MSSM_RNA_PFC_334', 'MSSM_RNA_PFC_335',
-#   'MSSM_RNA_PFC_340', 'MSSM_RNA_PFC_35', 'MSSM_RNA_PFC_362', 'MSSM_RNA_PFC_372', 'MSSM_RNA_PFC_373', 'MSSM_RNA_PFC_385', 'MSSM_RNA_PFC_393', 'MSSM_RNA_PFC_395',
-#   'MSSM_RNA_PFC_49', 'MSSM_RNA_PFC_54', 'MSSM_RNA_PFC_70', 'MSSM_RNA_PFC_76', 'MSSM_RNA_PFC_77', 'MSSM_RNA_PFC_78', 'MSSM_RNA_PFC_88', 'MSSM_RNA_PFC_93',
-#   'FAKE_DLPFC_CMC_MSSM_426', 'PENN_RNA_PFC_29', 'FAKE_DLPFC_CMC_PENN_055', 'PENN_RNA_PFC_75', 'PENN_RNA_PFC_110', 'FAKE_DLPFC_CMC_PENN_106', 'FAKE_DLPFC_CMC_PENN_107',
-#   'FAKE_DLPFC_CMC_PENN_108', 'MSSM_RNA_PFC_14', 'MSSM_RNA_BP_PFC_21', 'MSSM_RNA_PFC_293', 'MSSM_RNA_PFC_349', 'MSSM_RNA_PFC_353', 'MSSM_RNA_PFC_363', 'MSSM_RNA_PFC_369',
-#   'PENN_RNA_PFC_82', 'PENN_RNA_PFC_89', 'PITT_RNA_PFC_1159', 'PITT_RNA_PFC_1296', 'PITT_RNA_PFC_537',  'PITT_RNA_PFC_567', 'PITT_RNA_PFC_871', 'PITT_RNA_PFC_878',
-#   'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_210',  'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_278', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_361',
-#   'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_56', 'PITT_RNA_BP_PFC_10003', 'PITT_RNA_BP_PFC_1047', 'PITT_RNA_BP_PFC_1086', 'PITT_RNA_BP_PFC_1092', 'PITT_RNA_BP_PFC_1247',
-#   'PITT_RNA_BP_PFC_1282', 'PITT_RNA_BP_PFC_1324', 'PITT_RNA_BP_PFC_1374', 'PITT_RNA_BP_PFC_1391', 'PITT_RNA_BP_PFC_686')
+indToRemove = c("CMC_HBCC_RNA_PFC_3402")
 
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
+p <- p + geom_point(aes(color=Gender, shape=Dx, size=Age_of_Death))
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
@@
 ```
-Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals
+Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples
 
@@
                           x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
-                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
+                          lengths = as.numeric(GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length']),
                           lengthMethod = "smooth", 
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
+p <- p + geom_point(aes(color=Gender, shape=Tissue, size=Age_of_Death))
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
@@
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
+COVARIATES.tmp = data.matrix(COVARIATES[,c("Gender", "Dx.Tissue")])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
@@
 # Function to fit a mixed effect model and return BIC
-fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){
+fitFixedEffectModel <- function(varsToFit, lcpm, cnt, covars){
 
@@
 
-  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
-                              .combine = rbind,
-                              .packages = c('lme4', 'dplyr'),
-                              .export = c('voomWeights')) %dopar% {
-
-                                # Perform weighted mixed linear models for every gene
-                                expr = data.frame(Gx = lcpm[i,])
-                                data = cbind(expr,
-                                             designMat[rownames(expr),],
-                                             Individual_ID = groupingVar)
-                                formula = paste(setdiff(colnames(data), c('Gx', 'Individual_ID')), collapse = '+') %>%
-                                  paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')
-
-                                weights = voomWeights[i,]
-                                BIC = NA
-                                tryCatch({
-                                  mdl = lmer(formula = formula, 
+  modelBIC = plyr::ldply(1:dim(lcpm)[1],
+                         .fun = function(i, lcpm, designMat, voomWeights){
+                           # Fit a linear models for every gene
+                           data = cbind(data.frame(Gx = lcpm[i,]),
+                                        designMat)
+                           wts = voomWeights[i,]
+
+                           formula = paste(setdiff(colnames(data), c('Gx')), collapse = '+') %>%
+                             paste('Gx ~',., collapse = ' ')
+
+                           mdl = lm(formula = formula, 
                                     data = data, 
-                                             weights = weights)
-                                  BIC = BIC(mdl)
-                                }, error = function(e){
-                                  BIC = NA
-                                })
-                              return(data.frame(ensembl_gene_id = rownames(lcpm)[i],
-                                                BIC = BIC))
-                            }
+                                    subset = 1:dim(data)[1],
+                                    weights = wts)
+
+                           data.frame(ensembl_gene_id = rownames(lcpm)[i], BIC = BIC(mdl))
+                         }, 
+                         lcpm, designMat, voomWeights, 
+                         .parallel = F,
+                         .paropts = list(.packages = c('dplyr')))
   return(modelBIC)
@@
   while(length(covarsToSelect) != 0){
-    allModelBIC = foreach(i = 1:length(covarsToSelect),
-                          .combine = c,
-                          .packages = pkgs,
-                          .export = c('fitMixedEffectModel')) %dopar% {
+    allModelBIC = plyr::llply(1:length(covarsToSelect),
+                              .fun = function(i, covarsToSelect, adjustCovars, lcpm, cnt, covars){
 
                                 # Fit a mixed effect linear model and get all BIC
-                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
-                                                           covars[colnames(cnt),'Individual_ID'], 
-                                                           lcpm, 
-                                                           cnt, 
-                                                           covars) %>%
+                                modelBIC = fitFixedEffectModel(c(adjustCovars, covarsToSelect[i]),
+                                                               lcpm, cnt, covars) %>%
                                   plyr::rename(c('BIC' = covarsToSelect[i]))
 
-                            return(list(modelBIC))
-                          }
+                                return(modelBIC)
+                              },
+                              covarsToSelect, adjustCovars, lcpm, cnt, covars, 
+                              .parallel = F,
+                              .paropts = list(.export = c('fitFixedEffectModel')))
+
     allModelBIC = plyr::join_all(allModelBIC)
@@
 adjust.covars = c('Dx.Tissue')
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
+baseModelBIC = fitFixedEffectModel(adjust.covars, 
                                    CQN.GENE_EXPRESSION$E, 
@@
 # Iteratively identify covariates in phase I
-phase1.covars = c('Institution', 'Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
+phase1.covars = c('Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2')
 vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
@@
 #### Phase II (batch effects)
-```{r phase2}
+```{r phase2, eval = F}
 # Fit a base model with the following covariates
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
+baseModelBIC = fitFixedEffectModel(adjust.covars, 
                                    CQN.GENE_EXPRESSION$E, 
@@
 # Fit a base model with the following covariates
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
+baseModelBIC = fitFixedEffectModel(adjust.covars, 
                                    CQN.GENE_EXPRESSION$E, 
@@
 # Iteratively identify covariates in phase I
-phase3.covars = c("MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate", "GenesDetected",
-                  "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", "PercentAligned")
+phase3.covars = c("TotalReads", "X5prime_Norm", "Mean_Per_Base_Cov.", "Mean_CV", "Fragment_Length_StdDev",
+                  "Mapped_Pairs", "End_1_Mismatch_Rate", "End_2_Mismatch_Rate", "Mapped", "IntragenicRate",
+                  "ExonicRate", "IntergenicRate", "IntronicRate", "rRNARate", "GenesDetected",
+                  "TranscriptsDetected", "ExpProfEfficiency", "PercentMapped")
 vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
@@
 ### Normalisation
-1. Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
-2. Individual_ID is chosen as random effect
+Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
 ```{r iterative.normalisation, results='asis'}
@@
 VOOM.GENE_EXPRESSION = voom(tmp, design=DM1$design, plot=F)
-
-# Estimate consensus correlation coefficient
-correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
-
-# Re-calculate voom weights with correlation of random effects
-VOOM.GENE_EXPRESSION = voom(tmp, 
-                            design=DM1$design, plot=F,
-                            block = COVARIATES$Individual_ID, 
-                            correlation = correlation$cor)
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
 
 # Fit linear model using new weights and new design
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
 ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
+p <- p + geom_point(aes(color=Gender, shape=Tissue, size=Age_of_Death))
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
@@
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue','Institution')])
+COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue','Gender')])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
@@
                            main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue','Institution')]))
+                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue','Gender')]))
 ```
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
-                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
-                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
-                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
+contrast = makeContrasts(contrasts=c("Dx.TissueBP_DLPFC-Dx.TissueControl_DLPFC",
                                      "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
-                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC"),
+                                     "Dx.TissueSCZ_DLPFC-Dx.TissueBP_DLPFC"),
                          levels = colnames(ADJUSTED.FIT$coefficients))
@@
 # Get differnetial expression
-DE = lapply(1:6, function(i, FIT){
+DE = lapply(1:3, function(i, FIT){
   topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
@@
 pl = list()
-pl[[1]] = ggplot(DE, aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
-pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
+pl[[1]] = ggplot(DE, aes(x = log10(as.numeric(Length)), y = logFC, color = Direction)) + geom_point() 
+pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(as.numeric(Length)), y = logFC))
 pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
@@
 
-  tissueTypeAbrv	= 'DLPFC, ACC', 
-  study = 'CMC',
+  tissueTypeAbrv	= 'DLPFC', 
+  study = 'HBCC',
 
@@
 # Code
-CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection 1)", parentId = parentId)
+CODE <- Folder(name = "HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)", parentId = parentId)
 annotations(CODE) = all.annotations

### 2017-06-19 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor bug fix

@@
 Set outlier expression of genes in each samples to NA
-```{r detect.outlier.genes}
+```{r detect.outlier.genes, cache = FALSE}
 LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)
@@
 plotdata$label = plotdata$SampleID
-plotdata$label[!(plotdata$label %in% top.nasamples)] = ''
+plotdata$label[!(plotdata$label %in% top.nsamples)] = ''
 

### 2017-06-19 code/Rmd/ROSMAP_REPROCESSED.Rmd
Included ages 90+ for ROSMAP data analysis

@@
 
+METADATA.CLINICAL_ID1 <- 'syn7116000'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.CLINICAL_ID1
+METADATA.CLINICAL_OBJ1 <- synGet(METADATA.CLINICAL_ID1)
+METADATA.CLINICAL1 <- read.table(METADATA.CLINICAL_OBJ1@filePath,sep=',',header=T)
+
 # Get technical covariates
@@
   dplyr::left_join(KEY) %>%
-  dplyr::left_join(METADATA.CLINICAL)
+  dplyr::left_join(METADATA.CLINICAL) %>%
+  dplyr::select(-age_first_ad_dx, -age_death, -age_at_visit_max) %>%
+  dplyr::left_join(METADATA.CLINICAL1)
 
@@
 ```{r preprocess.data,cache=TRUE, echo=TRUE}
-# Remove samples with no cogdx, RIN and PMI scores
+# Remove samples with no cogdx, RIN, PMI scores and age_death
 METADATA <- METADATA %>%
   ungroup %>%
-  filter(Sampleid %in% colnames(COUNT)) %>%
-  filter(!is.na(cogdx), !is.na(braaksc), !is.na(ceradsc)) %>%
-  filter(!is.na(RINcontinuous)) %>%
-  filter(!is.na(pmi)) %>%
-  filter(!is.na(PCT_INTRONIC_BASES)) %>%
+  dplyr::filter(Sampleid %in% colnames(COUNT)) %>%
+  dplyr::filter(!is.na(cogdx), !is.na(braaksc), !is.na(ceradsc)) %>%
+  dplyr::filter(!is.na(RINcontinuous)) %>%
+  dplyr::filter(!is.na(pmi)) %>%
+  dplyr::filter(!is.na(PCT_INTRONIC_BASES)) %>%
+  dplyr::filter(!is.na(age_death)) %>%
   as.data.frame()
@@
 all.diff.exp = list(cogdx = DE)
+all.fit = list(cogdx = FIT)
 ```
@@
 
-Genes that are differentially expressed at an FDR <= 0.05 are
+Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
 ```{r diff.exp1, fig.height=6, fig.width=18, cache=FALSE}
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("DiagnosisCONTROL-DiagnosisAD",
-                                     "DiagnosisAD-DiagnosisOTHER",
-                                     "DiagnosisCONTROL-DiagnosisOTHER"),
+contrast = makeContrasts(contrasts=c("DiagnosisAD-DiagnosisCONTROL"),
                          levels = colnames(FIT$coefficients))
@@
 # Get differnetial expression
-DE = lapply(1:3, function(i, FIT){
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
   topTable(FIT, coef=i, number = 50000, confint = T) %>%
@@
 }, FIT.CONTR)
-names(DE) = gsub('Diagnosis','',colnames(contrast))
+names(DE) = c('AD-CONTROL')
 
@@
 all.diff.exp = c(all.diff.exp, list(Diagnosis = DE))
+all.fit = c(all.fit, list(Diagnosis = FIT))
 ```
@@
 
+### Differential expression analysis (with Diagnosis and Gender)
+Differential expression is performed on the Diagnosis x Gender variable by controlling for covariates identified above
+
+Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
+```{r diff.exp2, fig.height=6, fig.width=18, cache=FALSE}
+## Modeling Diagnosis, msex and age_death conjointly
+COVARIATES$Diagnosis.msex = paste(COVARIATES$Diagnosis, factor(COVARIATES$msex, labels = c('0' = 'FEMALE', '1' = 'MALE')), sep = '.') %>%
+  factor
+
+# Get design matrix
+DESIGN = getDesignMatrix(COVARIATES[, c('Diagnosis.msex', setdiff(postAdjustCovars, 'msex')), drop = F], Intercept = F)
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+colnames(DESIGN$design) = gsub('Diagnosis.msex','',colnames(DESIGN$design))
+
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(counts, design=DESIGN$design, plot=F)
+
+# Fit linear model using new weights and new design
+FIT = lmFit(CQN.GENE_EXPRESSION$E.NA,
+            design = DESIGN$design,
+            weights = VOOM.WEIGHTS$weights)
+
+# Fit contrast
+contrast = makeContrasts(contrasts=c("(AD.FEMALE+AD.MALE)/2-(CONTROL.FEMALE+CONTROL.MALE)/2",
+                                     "(AD.FEMALE+CONTROL.FEMALE)/2-(AD.MALE+CONTROL.MALE)/2",
+                                     "(AD.FEMALE-CONTROL.FEMALE)/2-(AD.MALE-CONTROL.MALE)/2",
+                                     "AD.FEMALE-CONTROL.FEMALE",
+                                     "AD.MALE-CONTROL.MALE"),
+                         levels = colnames(FIT$coefficients))
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = T) %>%
+    rownameToFirstColumn('ensembl_gene_id')
+}, FIT.CONTR)
+names(DE) = c('AD-CONTROL','FEMALE-MALE','AD-CONTROL.IN.FEMALE-MALE','AD-CONTROL.IN.FEMALE', 'AD-CONTROL.IN.MALE')
+
+DE = DE %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  left_join(GENE.PARAM %>%
+              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
+              unique()) %>%
+  dplyr::mutate(Study = 'ROSMAP', Region = 'DLPFC',
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction))
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
+
+writeLines('Differentially expressed genes at an FDR 0.05 and logFC 1.2')
+tmp = DE %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
+  group_by(Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05_FC_1.2) 
+kable(tmp)
+
+p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red'))
+p = p + facet_grid(.~Comparison, scales = 'fixed')
+p
+
+all.diff.exp = c(all.diff.exp, list(Diagnosis.Gender = DE))
+all.fit = c(all.fit, list(Diagnosis.Gender = FIT))
+```
+
+### Differential expression analysis (with Diagnosis and AOD)
+Differential expression is performed on the Diagnosis x AOD variable by controlling for covariates identified above
+
+Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
+```{r diff.exp3, fig.height=6, fig.width=18, cache=FALSE}
+# Get design matrix
+DESIGN = getDesignMatrix(COVARIATES[, c('Diagnosis', setdiff(postAdjustCovars, 'age_death')), drop = F], Intercept = F)
+ind = grep('Diagnosis', colnames(DESIGN$design))
+DESIGN$design[,ind] = DESIGN$design[,ind] * COVARIATES[rownames(DESIGN$design),'age_death']
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+colnames(DESIGN$design) = gsub('Diagnosis','AOD.',colnames(DESIGN$design))
+
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(counts, design=DESIGN$design, plot=F)
+
+# Fit linear model using new weights and new design
+FIT = lmFit(CQN.GENE_EXPRESSION$E.NA,
+            design = DESIGN$design,
+            weights = VOOM.WEIGHTS$weights)
+
+# Fit contrast
+contrast = makeContrasts(contrasts=c("AOD.AD-AOD.CONTROL",
+                                     "(AOD.AD+AOD.CONTROL)/2"),
+                         levels = colnames(FIT$coefficients))
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = T) %>%
+    rownameToFirstColumn('ensembl_gene_id')
+}, FIT.CONTR)
+names(DE) = c('AOD.AD-AOD.CONTROL','AOD')
+
+DE = DE %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  left_join(GENE.PARAM %>%
+              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
+              unique()) %>%
+  dplyr::mutate(Study = 'ROSMAP', Region = 'DLPFC',
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction))
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
+
+writeLines('Differentially expressed genes at an FDR 0.05 and logFC 1.2')
+tmp = DE %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
+  group_by(Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05_FC_1.2) 
+kable(tmp)
+
+p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red'))
+p = p + facet_grid(.~Comparison, scales = 'fixed')
+p
+
+all.diff.exp = c(all.diff.exp, list(Diagnosis.AOD = DE))
+all.fit = c(all.fit, list(Diagnosis.AOD = FIT))
+```
+
 ### SVA Adjustments for eQTL analysis
@@
 ```{r sva.adjust}
+# Get design matrix
+DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[2], postAdjustCovars), drop = F], Intercept = F)
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
 # Get (null) design matrix
@@
 
+# Create seperate acccess list for 90+ data
+changeACL <- function(id){
+  old.acl = synGetEntityACL(METADATA.CLINICAL_OBJ1@properties$id)
+  myaccess = ResourceAccess('3319864', list('UPDATE', 'CHANGE_SETTINGS', 'DELETE', 'READ', 'DOWNLOAD', 'MODERATE', 'CREATE', 'CHANGE_PERMISSIONS'))
+  new.acl = AccessControlList(id = id, resourceAccess = append(old.acl@resourceAccess, myaccess))
+  synCreateEntityACL(new.acl)
+}
+
 # Store covariates
@@
                    executed = thisFile, activityDescription = activityDescription)
+changeACL(COV_OBJ@properties$id)
 
@@
 DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+changeACL(DM_OBJ@properties$id)
 
@@
 # Store differential expression results
-rbindlist(all.diff.exp, use.names = T, fill = T) %>%
+rbindlist(all.diff.exp, use.names = T, fill = T, idcol = 'Model') %>%
   write.table(file = 'ROSMAP_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
@@
 DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
+# Store all differential expression models
+save(all.fit, file = 'ROSMAP_Models.RData')
+FIT_OBJ = File('ROSMAP_Models.RData', 
+               name = 'All Limma Models',
+               parentId = CODE$properties$id)
+annotations(FIT_OBJ) = annotations(CODE)
+annotations(FIT_OBJ)$dataSubType = 'modelFits'
+FIT_OBJ = synStore(FIT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+changeACL(FIT_OBJ@properties$id)
+
 stopCluster(cl)
@@
 |  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
+|  Linear Model Fits                          |  `r FIT_OBJ$properties$id`       |
+
+### R Source Code
+[Github](`r thisFile`)

### 2017-06-21 code/Rmd/ROSMAP_REPROCESSED.Rmd
Minor modifications

@@
 library(githubr) # get the package from devtools::install_github('brian-bot/githubr')
+library(rSynapseUtilities)
 
@@
 
+# Create new resource acls
+new.racl = ResourceAccess(principalId = '3319864', 
+                          accessType = c('UPDATE', 'CHANGE_SETTINGS', 'DELETE', 'READ', 'DOWNLOAD', 'MODERATE', 'CREATE', 'CHANGE_PERMISSIONS')) %>%
+  ResourceAccessList() %>%
+  synapseClient::append(ResourceAccess(principalId = '274008', 
+                                       accessType = c('UPDATE', 'CHANGE_SETTINGS', 'DELETE', 'READ', 'DOWNLOAD', 'MODERATE', 'CREATE', 'CHANGE_PERMISSIONS'))) %>%
+  synapseClient::append(ResourceAccess(principalId = '273995', 
+                                       accessType = c('UPDATE', 'CHANGE_SETTINGS', 'DELETE', 'READ', 'DOWNLOAD', 'MODERATE', 'CREATE', 'CHANGE_PERMISSIONS'))) %>%
+  synapseClient::append(ResourceAccess(principalId = '3344250', 
+                                       accessType = c('READ', 'DOWNLOAD')))
+
 # Code
@@
 CODE <- synStore(CODE)
-
-# Create seperate acccess list for 90+ data
-changeACL <- function(id){
-  old.acl = synGetEntityACL(METADATA.CLINICAL_OBJ1@properties$id)
-  myaccess = ResourceAccess('3319864', list('UPDATE', 'CHANGE_SETTINGS', 'DELETE', 'READ', 'DOWNLOAD', 'MODERATE', 'CREATE', 'CHANGE_PERMISSIONS'))
-  new.acl = AccessControlList(id = id, resourceAccess = append(old.acl@resourceAccess, myaccess))
-  synCreateEntityACL(new.acl)
-}
+synChangeEntityACL(CODE@properties$id, new.racl)
 
@@
                    executed = thisFile, activityDescription = activityDescription)
-changeACL(COV_OBJ@properties$id)
+synChangeEntityACL(COV_OBJ@properties$id, new.racl)
 
@@
 DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-changeACL(DM_OBJ@properties$id)
+synChangeEntityACL(DM_OBJ@properties$id, new.racl)
 
@@
 FIT_OBJ = synStore(FIT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-changeACL(FIT_OBJ@properties$id)
+synChangeEntityACL(FIT_OBJ@properties$id, new.racl)
+
+# Delete folder acl
+synDeleteEntityACL(CODE@properties$id)
 
@@
 |  -----------------------------------------  |   ---------                      |
-|  Covariates                                 |  `r COV_OBJ$properties$id`       |
+|  Covariates (90+ ages)                      |  `r COV_OBJ$properties$id`       |
 |  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |

### 2017-06-21 code/Rmd/ROSMAP_REPROCESSED.Rmd
Minor modifications

@@
 tmp = COVARIATES %>%
-  group_by(Diagnosis, cogdx) %>%
-  summarise(count = n()) %>%
-  spread(Diagnosis, count)
+  dplyr::group_by(Diagnosis, cogdx) %>%
+  dplyr::summarise(count = n()) %>%
+  tidyr::spread(Diagnosis, count)
 ```
@@
 CODE <- synStore(CODE)
+old.acl = synGetEntityACL(CODE@properties$id)
 synChangeEntityACL(CODE@properties$id, new.racl)
@@
 # Delete folder acl
-synDeleteEntityACL(CODE@properties$id)
+synChangeEntityACL(CODE@properties$id, old.acl@resourceAccess)
 

### 2017-06-24 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Minor modifications

@@
 METADATA_QC = read.table(synGet(METADATA_QC_DLPFC_ID)@filePath, fill = T, header = T, sep = ',', row.names = 1) %>%
-  rownameToFirstColumn('Sample.RNA.ID') %>%
-  dplyr::select(Sample.RNA.ID, Total_Purity_Filtered_Reads_Sequenced, X5prime_Norm, Mean_Per_Base_Cov.,  
+  dplyr::select(Sample, Total_Purity_Filtered_Reads_Sequenced, X5prime_Norm, Mean_Per_Base_Cov.,  
                 Mean_CV, Fragment_Length_StdDev,  Mapped_Pairs, End_1_Mismatch_Rate, End_2_Mismatch_Rate,
@@
                 UNIQUE_READS._Uniquely_mapped_reads_percent) %>%
-  dplyr::rename(SampleID = Sample.RNA.ID, 
+  dplyr::rename(SampleID = Sample, 
                 IntragenicRate = Intragenic_Rate, 
@@
 
-writeLines('Some control samples with Age of death less than 17 are removed')
-writeLines(paste('Following',length(METADATA$SampleID[METADATA$Age_of_Death <= 17]), 'samples are removed:'))
-writeLines(paste(METADATA$SampleID[METADATA$Age_of_Death <= 17], collapse = ', '))
-METADATA <- METADATA  %>% dplyr::filter(Age_of_Death >= 17) 
+writeLines('Some control samples with Age of death less than 21 are removed')
+writeLines(paste('Following',length(METADATA$SampleID[METADATA$Age_of_Death <= 21]), 'samples are removed:'))
+writeLines(paste(METADATA$SampleID[METADATA$Age_of_Death <= 21], collapse = ', '))
+METADATA <- METADATA  %>% dplyr::filter(Age_of_Death >= 21) 
 ```
@@
   dplyr::rename(ensembl_gene_id = genes) %>%
-  left_join(Ensemble2HGNC) %>%
-  group_by(gene_biotype) %>%
-  summarise(fraction  = n()) %>%
-  filter(fraction > 100) %>%
+  dplyr::left_join(Ensemble2HGNC) %>%
+  dplyr::group_by(gene_biotype) %>%
+  dplyr::summarise(fraction  = n()) %>%
+  dplyr::filter(fraction > 100) %>%
   dplyr::mutate(fraction = fraction/length(PROCESSED_COUNTS$filteredExprMatrix$genes[,1]))
@@
 tmp = tidyr::separate(COVARIATES, Dx.Tissue, c('Dx', 'Tissue'), sep = '_') %>%
-  group_by(Dx, Tissue) %>%
-  summarise(count = n()) %>%
-  spread(Tissue, count)
+  dplyr::group_by(Dx, Tissue) %>%
+  dplyr::summarise(count = n()) %>%
+  tidyr::spread(Tissue, count)
 ```
@@
                            data = cbind(data.frame(Gx = lcpm[i,]),
-                                        designMat)
-                           wts = voomWeights[i,]
+                                        designMat, 
+                                        weight = voomWeights[i,])
 
-                           formula = paste(setdiff(colnames(data), c('Gx')), collapse = '+') %>%
+                           formula = paste(setdiff(colnames(data), c('Gx', 'weight')), collapse = '+') %>%
                              paste('Gx ~',., collapse = ' ')
 
-                           mdl = lm(formula = formula, 
+                           mdl = stats::lm(formula = formula, 
                                            data = data, 
-                                    subset = 1:dim(data)[1],
-                                    weights = wts)
+                                           weights = weight)
 
@@
 DE = DE %>% 
-  rbindlist(idcol = 'Comparison') %>%
+  data.table::rbindlist(idcol = 'Comparison') %>%
   dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
@@
   tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
-  tidyr::separate(ref.state, into = c('Dx1','Tissue'), sep = '_') %>%
-  tidyr::separate(to.state, into = c('Dx2','Tissue'), sep = '_') %>%
-  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_') %>%
+  tidyr::separate(ref.state, into = c('Dx1','Tissue1'), sep = '_') %>%
+  tidyr::separate(to.state, into = c('Dx2','Tissue2'), sep = '_') %>%
   left_join(Ensemble2HGNC) %>%
-  left_join(GENE.LEN)
+  left_join(GENE.LEN) %>%
+  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_')
+
 DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
@@
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
-  group_by(Comparison, Tissue) %>%
+  dplyr::select(ensembl_gene_id, Comparison, Tissue1) %>%
+  group_by(Comparison, Tissue1) %>%
   dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
@@
   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
-  group_by(Comparison, Tissue) %>%
+  dplyr::select(ensembl_gene_id, Comparison, Tissue1) %>%
+  group_by(Comparison, Tissue1) %>%
   dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
@@
 p = p + scale_color_manual(values = c('green','grey','red'))
-p = p + facet_grid(Tissue+.~Comparison, scales = 'fixed')
+p = p + facet_grid(Tissue1+.~Comparison, scales = 'fixed')
 p

### 2017-06-30 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor modifications

@@
 knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
-                                 parentId ="syn7501880",
+                                 parentId = "syn9872607",
                                  entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = 'syn7501880';
+parentId = 'syn9872607';
 activityName = 'Covariate adjustments';
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
-                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
-                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
-                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
+contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
                                      "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
-                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC",
                                      "Dx.TissueSCZ_DLPFC-Dx.TissueSCZ_ACC",
                                      "Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",
-                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueControl_ACC",
-                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_ACC"),
+                                     "Dx.TissueSCZ_DLPFC+Dx.TissueSCZ_ACC-Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",         
+                                     "Dx.TissueSCZ_DLPFC+Dx.TissueControl_DLPFC-Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC-Dx.TissueSCZ_ACC+Dx.TissueControl_ACC"),
                          levels = colnames(ADJUSTED.FIT$coefficients))
@@
 names(DE) = gsub('Dx.Tissue', '', colnames(contrast))
-names(DE)[9:10] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC')
+names(DE)[5:7] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC', 'ALL_ALL-ALL_ALL')
 
@@
   dplyr::filter(adj.P.Val <= 0.05, ref.Dx != 'Other', to.Dx != 'Other') %>%
-  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
+  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue, logFC) %>%
   group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
-  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
+  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))
 
@@
   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), ref.Dx != 'Other', to.Dx != 'Other') %>%
-  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
+  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue, logFC) %>%
   group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
+  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))
 
@@
 
+```{r diffExp1, fig.height=10, fig.width=15}
+postAdjustCovars = c('Tissue', 'Dx', 'Institution', 'Gender', 'RIN', 'IntronicRate')
+
+# Post adjusted design matrix
+COV1 = COVARIATES
+COV1 = tidyr::separate(COV1, Dx.Tissue, c('Dx', 'Tissue'), sep = '_') %>%
+  dplyr::mutate(Dx = factor(Dx, levels = c('SCZ', 'Control', 'Other')), Tissue = factor(Tissue))
+
+DESIGN = getDesignMatrix(COV1[,postAdjustCovars,drop=F],Intercept = F)$design
+DESIGN = DESIGN[,linColumnFinder(DESIGN)$indepCols]
+
+# Re-calculate voom weights with correlation of random effects
+VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero,   
+                            design = DESIGN, plot=F,
+                            block = COV1$Individual_ID, 
+                            correlation = correlation$cor)
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
+
+# Fit linear model using new weights and new design
+ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
+
+# Fit contrast
+contrast = makeContrasts(contrasts=c("TissueACC-TissueDLPFC",
+                                     "DxSCZ-DxControl"),
+                         levels = colnames(ADJUSTED.FIT$coefficients))
+FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE1 = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
+    rownameToFirstColumn('ensembl_gene_id') 
+}, FIT.CONTR) 
+names(DE1) = c('ACC-DLPFC', 'SCZ-CONTROL')
+
+DE1 = DE1 %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Direction = sign(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  left_join(Ensemble2HGNC) 
+DE1$Direction[DE1$adj.P.Val > 0.05 | abs(DE1$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE1 %>%
+  dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
+  group_by(Comparison) %>%
+  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))
+
+tmp1 = DE1 %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
+  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
+  group_by(Comparison) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))
+
+kable(full_join(tmp,tmp1))
+
+p = DE1 %>%
+  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
+p = p + facet_grid(Comparison~., scales = 'free_y')
+p = p + ggtitle('Differential Expression between Schizophrenia and Control')
+p
+```
+
 ### Gene set enrichment analysis
@@
 ```{r enrich.de, fig.height=16, fig.width=16}
-diffexp.genes = DE %>%
+diffexp.genes.up = DE %>%
   dlply(.(ref.Dx, ref.Tissue, to.Dx, to.Tissue), .fun = function(x){
-    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
+    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC > 0] %>% unique
     tmp = tmp[!is.na(tmp)]
   })
+names(diffexp.genes.up) = paste(names(diffexp.genes.up), 'UP', sep = '.')
+
+diffexp.genes.down = DE %>%
+  dlply(.(ref.Dx, ref.Tissue, to.Dx, to.Tissue), .fun = function(x){
+    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC < 0] %>% unique
+    tmp = tmp[!is.na(tmp)]
+  })
+names(diffexp.genes.down) = paste(names(diffexp.genes.down), 'DOWN', sep = '.')
+
+diffexp.genes.up$ACC.DLPFC.UP = dplyr::filter(DE1, adj.P.Val <= 0.05, logFC > 0, Comparison == 'ACC-DLPFC') %>%
+  dplyr::select(hgnc_symbol) %>%
+  unique() %>% unlist() %>% as.character()
+diffexp.genes.down$ACC.DLPFC.DOWN = dplyr::filter(DE1, adj.P.Val <= 0.05, logFC < 0, Comparison == 'ACC-DLPFC') %>%
+  dplyr::select(hgnc_symbol) %>%
+  unique() %>% unlist() %>% as.character()
+diffexp.genes.up$ACC.DLPFC.UNCHANGED = dplyr::filter(DE1, adj.P.Val > 0.05, Comparison == 'ACC-DLPFC') %>%
+  dplyr::select(hgnc_symbol) %>%
+  unique() %>% unlist() %>% as.character()
+
+diffexp.genes = c(diffexp.genes.up, diffexp.genes.down)
+diffexp.genes = diffexp.genes[c("Control.ACC.SCZ.ACC.UP", "Control.ACC.SCZ.ACC.DOWN",
+                                "Control.DLPFC.SCZ.DLPFC.UP", "Control.DLPFC.SCZ.DLPFC.DOWN",
+                                "ACC.DLPFC.UP", "ACC.DLPFC.DOWN", "ACC.DLPFC.UNCHANGED")]
+diffexp.genes$Control.SCZ.ACC.DLPFC.UP = setdiff(diffexp.genes$Control.ACC.SCZ.ACC.UP, diffexp.genes$Control.DLPFC.SCZ.DLPFC.UP)
+diffexp.genes$Control.SCZ.ACC.DLPFC.DOWN = setdiff(diffexp.genes$Control.ACC.SCZ.ACC.DOWN, diffexp.genes$Control.DLPFC.SCZ.DLPFC.DOWN)
+diffexp.genes$Control.SCZ.DLPFC.ACC.UP = setdiff(diffexp.genes$Control.DLPFC.SCZ.DLPFC.UP, diffexp.genes$Control.ACC.SCZ.ACC.UP)
+diffexp.genes$Control.SCZ.DLPFC.ACC.DOWN = setdiff(diffexp.genes$Control.DLPFC.SCZ.DLPFC.DOWN, diffexp.genes$Control.ACC.SCZ.ACC.DOWN)
+diffexp.genes$Control.SCZ.DLPFC.ACC.BOTH.UP = intersect(diffexp.genes$Control.DLPFC.SCZ.DLPFC.UP, diffexp.genes$Control.ACC.SCZ.ACC.UP)
+diffexp.genes$Control.SCZ.DLPFC.ACC.BOTH.DOWN = intersect(diffexp.genes$Control.DLPFC.SCZ.DLPFC.DOWN, diffexp.genes$Control.ACC.SCZ.ACC.DOWN)
 diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]

### 2017-06-30 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Minor modifications

@@
 tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
-h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
+fdr = COVARIATES.CORRELATION$PVAL
+fdr = format(fdr, digits=1)
+h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate',
+            cell_fun = function(j,i,x,y,wifht,height,fill){
+                grid.text(fdr[i,j], x, y)
+              })
 draw(h, heatmap_legend_side = 'left', padding = unit(c(10,2,2,10), "mm"))
@@
 
-tmp = DE %>%
+tmp = de.hbcc %>%
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue1) %>%
+  dplyr::select(ensembl_gene_id, Comparison, Tissue1, logFC) %>%
   group_by(Comparison, Tissue1) %>%
-  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
+  dplyr::summarise(FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])),
+                   FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])))
 
-tmp1 = DE %>%
+tmp1 = de.hbcc %>%
   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue1) %>%
+  dplyr::select(ensembl_gene_id, Comparison, Tissue1, logFC) %>%
   group_by(Comparison, Tissue1) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
+  dplyr::summarise(FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > log2(1.2)])),
+                   FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < -log2(1.2)])))
 
@@
 ```{r enrich.de, fig.height=16, fig.width=16}
-diffexp.genes = DE %>%
-  dlply(.(Comparison, Tissue), .fun = function(x){
+diffexp.genes = de.hbcc %>%
+  dplyr::mutate(Direction = sign(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP'))) %>%
+  dlply(.(Comparison, Tissue1, Direction), .fun = function(x){
     tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
@@
 # Perform enrichment analysis
-enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
+enrichment.results = plyr::llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
    ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){

### 2017-07-01 code/Rmd/MSSM_REPROCESSED.Rmd
Added MSSM 90+ ages

@@
 
+# Get 90+ age
+METADATA_AGE_ID = 'syn10156693'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_AGE_ID
+METADATA_AGE <- read.table(synGet(METADATA_AGE_ID)@filePath, sep='\t', header=T) %>%
+  dplyr::rename(individualIdentifier = BBid)
+
 # Get clinical metadata
@@
 
-METADATA <- full_join(METADATA_TECH, METADATA_CLINICAL) %>%
-  inner_join(METADATA.PICARD)
+METADATA <- dplyr::full_join(METADATA_TECH, METADATA_CLINICAL) %>%
+  dplyr::inner_join(METADATA.PICARD) %>%
+  dplyr::select(-AOD) %>%
+  dplyr::left_join(METADATA_AGE)
 
@@
   filter(!is.na(BrainRegion.Diagnosis), !is.na(PMI), !is.na(RIN))
+
+# Fix RIN values for some samples (Based on email conversation with Minghui May 22 2017)
+FIX.RIN = c(hB_RNA_12312_resequenced = 3.9,
+            hB_RNA_12342 = 3.9,
+            hB_RNA_12352 = 4.2,
+            hB_RNA_9191 = 5,
+            hB_RNA_12262_resequenced = 5.2,
+            hB_RNA_9216 = 5.7,
+            hB_RNA_12272_resequenced = 5.8,
+            hB_RNA_12382 = 5.8, 
+            hB_RNA_13359 = 6.4)
+rownames(METADATA) = METADATA$sampleIdentifier
+METADATA[names(FIX.RIN),'RIN'] = FIX.RIN
+
+METADATA = METADATA %>%
+  dplyr::filter(RIN >= 5)
 ```
@@
   dplyr::mutate(BrainRegion = factor(BrainRegion))
+
 DESIGN = getDesignMatrix(COVARIATES.tmp[, c('BrainRegion', postAdjustCovars), drop = F], Intercept = F)
@@
 p
+
+all.diff.exp = list(CDR = DE)
+all.fit = list(CDR = FIT)
 ```
@@
   tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
-  tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
-  tidyr::unite(Comparison, reference, against, sep = '-') %>%
+  tidyr::separate(to.state, c('BrainRegion1','against'), sep = '\\.') %>%
+  dplyr::mutate(Comparison = paste(reference, against, sep = '-')) %>%
+  dplyr::select(-BrainRegion1) %>%
   left_join(GENE.PARAM %>%
@@
 p
+
+all.diff.exp = c(all.diff.exp, list(Diagnosis = DE))
+all.fit = c(all.fit, list(Diagnosis = FIT))
 ```
@@
 
+### Differential expression analysis (with BrainRegion.Diagnosis and Gender)
+Differential expression is performed on the BrainRegion.Diagnosis x Gender variable by controlling for covariates identified above
+
+Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
+```{r diff.exp2, fig.height=15, fig.width=13}
+## Modeling Diagnosis, msex and age_death conjointly
+COVARIATES$BrainRegion.Diagnosis.SEX = paste(COVARIATES$BrainRegion.Diagnosis, 
+                                             factor(COVARIATES$SEX, labels = c('F' = 'FEMALE', 'M' = 'MALE')), 
+                                             sep = '.') %>% factor
+
+# Get design matrix
+DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis.SEX', setdiff(postAdjustCovars, 'SEX')), drop = F], Intercept = F)
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
+
+# Calculate correlation values of random effects
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
+correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)
+
+# Re-estimate voom weights
+VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
+                    design=DESIGN$design, plot=F,
+                    block = COVARIATES$individualIdentifier, 
+                    correlation = correlation$cor)
+
+# Fit linear model using new weights and new design
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
+FIT = lmFit(VOOM.WEIGHTS)
+
+# Fit contrast
+cntr = c('BrainRegion.Diagnosis.SEXIFG.AD.MALE-BrainRegion.Diagnosis.SEXIFG.CONTROL.MALE',
+         'BrainRegion.Diagnosis.SEXIFG.AD.FEMALE-BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE',
+         '0.5*BrainRegion.Diagnosis.SEXIFG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXIFG.AD.FEMALE
+         -0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE',
+         '0.5*BrainRegion.Diagnosis.SEXIFG.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE
+         -0.5*BrainRegion.Diagnosis.SEXIFG.AD.MALE-0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE',
+         '0.5*BrainRegion.Diagnosis.SEXIFG.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE
+         -0.5*BrainRegion.Diagnosis.SEXIFG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.MALE',
+         'BrainRegion.Diagnosis.SEXFP.AD.MALE-BrainRegion.Diagnosis.SEXFP.CONTROL.MALE',
+         'BrainRegion.Diagnosis.SEXFP.AD.FEMALE-BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE',
+         '0.5*BrainRegion.Diagnosis.SEXFP.AD.MALE+0.5*BrainRegion.Diagnosis.SEXFP.AD.FEMALE
+         -0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE',
+         '0.5*BrainRegion.Diagnosis.SEXFP.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE
+         -0.5*BrainRegion.Diagnosis.SEXFP.AD.MALE-0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE',
+         '0.5*BrainRegion.Diagnosis.SEXFP.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE
+         -0.5*BrainRegion.Diagnosis.SEXFP.AD.MALE+0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.MALE',
+         'BrainRegion.Diagnosis.SEXPHG.AD.MALE-BrainRegion.Diagnosis.SEXPHG.CONTROL.MALE',
+         'BrainRegion.Diagnosis.SEXPHG.AD.FEMALE-BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE',
+         '0.5*BrainRegion.Diagnosis.SEXPHG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXPHG.AD.FEMALE
+         -0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE',
+         '0.5*BrainRegion.Diagnosis.SEXPHG.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE
+         -0.5*BrainRegion.Diagnosis.SEXPHG.AD.MALE-0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE',
+         '0.5*BrainRegion.Diagnosis.SEXPHG.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE
+         -0.5*BrainRegion.Diagnosis.SEXPHG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.MALE',
+         'BrainRegion.Diagnosis.SEXSTG.AD.MALE-BrainRegion.Diagnosis.SEXSTG.CONTROL.MALE',
+         'BrainRegion.Diagnosis.SEXSTG.AD.FEMALE-BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE',
+         '0.5*BrainRegion.Diagnosis.SEXSTG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXSTG.AD.FEMALE
+         -0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE',
+         '0.5*BrainRegion.Diagnosis.SEXSTG.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE
+         -0.5*BrainRegion.Diagnosis.SEXSTG.AD.MALE-0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE',
+         '0.5*BrainRegion.Diagnosis.SEXSTG.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE
+         -0.5*BrainRegion.Diagnosis.SEXSTG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.MALE')
+contrast = makeContrasts(contrasts=cntr,
+                         levels = colnames(FIT$coefficients))
+colnames(contrast) = c('IFG.AD.MALE-IFG.CONTROL.MALE',
+                       'IFG.AD.FEMALE-IFG.CONTROL.FEMALE',
+                       'IFG.AD.ALL-IFG.CONTROL.ALL',
+                       'IFG.ALL.FEMALE-IFG.ALL.MALE',
+                       'IFG.ALL.ALL-IFG.ALL.ALL',
+                       'FP.AD.MALE-FP.CONTROL.MALE',
+                       'FP.AD.FEMALE-FP.CONTROL.FEMALE',
+                       'FP.AD.ALL-FP.CONTROL.ALL',
+                       'FP.ALL.FEMALE-FP.ALL.MALE',
+                       'FP.ALL.ALL-FP.ALL.ALL',
+                       'PHG.AD.MALE-PHG.CONTROL.MALE',
+                       'PHG.AD.FEMALE-PHG.CONTROL.FEMALE',
+                       'PHG.AD.ALL-PHG.CONTROL.ALL',
+                       'PHG.ALL.FEMALE-PHG.ALL.MALE',
+                       'PHG.ALL.ALL-PHG.ALL.ALL',
+                       'STG.AD.MALE-STG.CONTROL.MALE',
+                       'STG.AD.FEMALE-STG.CONTROL.FEMALE',
+                       'STG.AD.ALL-STG.CONTROL.ALL',
+                       'STG.ALL.FEMALE-STG.ALL.MALE',
+                       'STG.ALL.ALL-STG.ALL.ALL')
+
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = T) %>%
+    rownameToFirstColumn('ensembl_gene_id')
+}, FIT.CONTR)
+names(DE) = colnames(contrast)
+
+DE = DE %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
+                Study = 'MSSM',
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  left_join(GENE.PARAM %>%
+              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
+              unique())
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
+  group_by(Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05_FC_1.2) 
+kable(tmp)
+
+all.diff.exp = c(all.diff.exp, list(Diagnosis.SEX = DE))
+all.fit = c(all.fit, list(Diagnosis.SEX = FIT))
+```
+
+### Differential expression analysis (with BrainRegion.Diagnosis and AOD)
+Differential expression is performed on the BrainRegion.Diagnosis x AOD variable by controlling for covariates identified above
+
+Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
+```{r diff.exp3, fig.height=15, fig.width=13}
+# Get design matrix
+DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[2], setdiff(postAdjustCovars, 'AOD')), drop = F], Intercept = F)
+ind = grep('Diagnosis', colnames(DESIGN$design))
+DESIGN$design[,ind] = DESIGN$design[,ind] * COVARIATES[rownames(DESIGN$design),'AOD']
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+colnames(DESIGN$design)[ind] = gsub('BrainRegion.Diagnosis', '', paste0(colnames(DESIGN$design)[ind], '.AOD'))
+
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
+
+# Calculate correlation values of random effects
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
+correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)
+
+# Re-estimate voom weights
+VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
+                    design=DESIGN$design, plot=F,
+                    block = COVARIATES$individualIdentifier, 
+                    correlation = correlation$cor)
+
+# Fit linear model using new weights and new design
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
+FIT = lmFit(VOOM.WEIGHTS)
+
+# Fit contrast
+cntr = c("FP.AD.AOD-FP.CONTROL.AOD",
+         "0.5*FP.AD.AOD+0.5*FP.CONTROL.AOD",
+         "IFG.AD.AOD-IFG.CONTROL.AOD",
+         "0.5*IFG.AD.AOD+0.5*IFG.CONTROL.AOD",
+         "PHG.AD.AOD-PHG.CONTROL.AOD",
+         "0.5*PHG.AD.AOD+0.5*PHG.CONTROL.AOD",
+         "STG.AD.AOD-STG.CONTROL.AOD",
+         "0.5*STG.AD.AOD+0.5*STG.CONTROL.AOD")
+contrast = makeContrasts(contrasts=cntr,
+                         levels = colnames(FIT$coefficients))
+colnames(contrast) = c("FP.AD.AOD-FP.CONTROL.AOD",
+                       "FP.ALL.AOD",
+                       "IFG.AD.AOD-IFG.CONTROL.AOD",
+                       "IFG.ALL.AOD",
+                       "PHG.AD.AOD-PHG.CONTROL.AOD",
+                       "PHG.ALL.AOD",
+                       "STG.AD.AOD-STG.CONTROL.AOD",
+                       "STG.ALL.AOD")
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = T) %>%
+    rownameToFirstColumn('ensembl_gene_id')
+}, FIT.CONTR)
+names(DE) = colnames(contrast)
+
+DE = DE %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
+                Study = 'MSSM',
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  left_join(GENE.PARAM %>%
+              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
+              unique())
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
+  group_by(Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05_FC_1.2) 
+kable(tmp)
+
+all.diff.exp = c(all.diff.exp, list(Diagnosis.AOD = DE))
+all.fit = c(all.fit, list(Diagnosis.AOD = FIT))
+```
+
 ### SVA Adjustments for eQTL analysis
@@
 
+# Function to change entity ACL (also creates new acls if it doesn't exist)
+synChangeEntityACL <- function(id, new.racl){
+  tryCatch({ 
+    # If old acl exist update
+    old.acl = synGetEntityACL(id)
+    old.acl@resourceAccess = new.racl
+    synUpdateEntityACL(old.acl)
+  }, error = function(e){ 
+    # If not create a new acl
+    new.acl = AccessControlList(id = id, resourceAccess = new.racl)
+    synCreateEntityACL(new.acl)
+  })
+}
+
+# Create new resource acls
+new.racl = ResourceAccess(principalId = '3319864', 
+                          accessType = c('UPDATE', 'CHANGE_SETTINGS', 'DELETE', 'READ', 
+                                         'DOWNLOAD', 'MODERATE', 'CREATE', 'CHANGE_PERMISSIONS')) %>%
+  ResourceAccessList() %>%
+  synapseClient::append(ResourceAccess(principalId = '274008', 
+                                       accessType = c('UPDATE', 'CHANGE_SETTINGS', 'DELETE', 'READ', 
+                                                      'DOWNLOAD', 'MODERATE', 'CREATE', 'CHANGE_PERMISSIONS'))) %>%
+  synapseClient::append(ResourceAccess(principalId = '273995', 
+                                       accessType = c('UPDATE', 'CHANGE_SETTINGS', 'DELETE', 'READ', 
+                                                      'DOWNLOAD', 'MODERATE', 'CREATE', 'CHANGE_PERMISSIONS'))) %>%
+  synapseClient::append(ResourceAccess(principalId = '3359054', 
+                                       accessType = c('READ', 'DOWNLOAD')))
+
 # Code
@@
 CODE <- synStore(CODE)
+synChangeEntityACL(CODE@properties$id, new.racl)
 
@@
                       executed = thisFile, activityDescription = activityDescription)
+synChangeEntityACL(COV_OBJ@properties$id, new.racl)
 
@@
 DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+synChangeEntityACL(DM_OBJ@properties$id, new.racl)
 
@@
 # Store differential expression results
-write.table(DE, file = 'MSSM_FP_STG_PHG_IFG_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('MSSM_FP_STG_PHG_IFG_DiffExpression.tsv', 
-                name = 'Differential Expression Results (BrainRegion.Diagnosis)', 
+rbindlist(all.diff.exp, use.names = T, fill = T, idcol = 'Model') %>%
+  write.table(file = 'MSSM_FP_STG_PHG_IFG_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('MSSM_FP_STG_PHG_IFG_DiffExpression.tsv', name = 'Differential Expression Results (cogdx)', 
                 parentId = CODE$properties$id)
@@
 DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-stopCluster(cl)
+
+# Store all differential expression models
+save(all.fit, file = 'MSSM_FP_STG_PHG_IFG_Models.RData')
+FIT_OBJ = File('MSSM_FP_STG_PHG_IFG_Models.RData', 
+               name = 'All Limma Models',
+               parentId = CODE$properties$id)
+annotations(FIT_OBJ) = annotations(CODE)
+annotations(FIT_OBJ)$dataSubType = 'modelFits'
+FIT_OBJ = synStore(FIT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+synChangeEntityACL(FIT_OBJ@properties$id, new.racl)
+
+# Delete folder acl
+synDeleteEntityACL(CODE@properties$id)
 ```

### 2017-07-04 code/Rmd/ALL_REPROCESSED.Rmd
Added markdown to process all data together

---
title: "Differential expression analysis of reprocessed AMPAD RNASeq data (All 7 brain regions with CQN normlaisation and BIC based model selection)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./ALL_REPROCESSED.Rmd",
                                 parentId = "syn8672415",
                                 entityName = 'ALL Reprocessed RNASeq (Normalised)')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(limma)
library(psych)
library(edgeR)
library(cqn)
library(lme4)
library(biomaRt)

library(synapseClient)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

synapseLogin()

library(doParallel)
library(foreach)

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn8672415';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of reprocessed rnaseq from all 7 brain regions';

thisFileName <- 'ALL_REPROCESSED.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```

### Data download
```{r download.data, cache=TRUE}
# Get covariates from synapse
covariates.ids = c(ROSMAP = 'syn8456631',
                   MSBB = 'syn8484996',
                   MAYO = 'syn8466814')
all.used.ids = as.character(covariates.ids)
covariates = lapply(covariates.ids, function(id){
  read.table(synGet(id)@filePath, header = T, sep = '\t')
})

covariates$ROSMAP = covariates$ROSMAP %>%
  dplyr::mutate(SampleID = as.character(SampleID),
                BrainRegion = 'DLPFC',
                BrainRegion.Diagnosis = factor(paste(BrainRegion, Diagnosis, sep = '.')),
                msex = factor(msex, labels = c('1'='MALE','0'='FEMALE')),
                Batch = factor(Batch)) %>%
  dplyr::rename(Gender = msex, RIN = RINcontinuous, AOD = age_death, PMI = pmi) %>%
  dplyr::select(SampleID, BrainRegion.Diagnosis, Batch, Gender, AOD, PMI, RIN, PCT_PF_READS_ALIGNED, 
                PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES)
covariates$ROSMAP$individualIdentifier = factor(paste0('ROSMAPpseudoid',1:dim(covariates$ROSMAP)[1]))

covariates$MSBB = covariates$MSBB %>%
  dplyr::mutate(SampleID = as.character(SampleID),
                individualIdentifier = factor(paste0('BB',individualIdentifier)),
                SEX = factor(SEX, labels = c('M'='MALE','F'='FEMALE')),
                batch = factor(batch),
                PMI = PMI/60) %>%
  dplyr::rename(Batch = batch, Gender = SEX) %>%
  dplyr::select(SampleID, BrainRegion.Diagnosis, individualIdentifier, Batch, Gender, AOD, PMI, RIN, 
                PCT_PF_READS_ALIGNED, PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES)

covariates$MAYO = covariates$MAYO %>%
  dplyr::mutate(SampleID = as.character(SampleID),
                Donor_ID = factor(paste0('MAYO', Donor_ID)),
                Gender = factor(Gender, labels = c('M'='MALE','F'='FEMALE')),
                Batch = factor(Source)) %>%
  dplyr::rename(individualIdentifier = Donor_ID, AOD = AgeAtDeath) %>%
  dplyr::select(SampleID, BrainRegion.Diagnosis, individualIdentifier, Batch, Gender, AOD, PMI, RIN, 
                PCT_PF_READS_ALIGNED, PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES)

# Download sample counts
counts.id = c(ROSMAP = 'syn8691134',
              MSBB = 'syn8691099',
              MAYO1 = 'syn8690799', 
              MAYO2 = 'syn8690904')
all.used.ids = c(all.used.ids, as.character(counts.id))
counts = lapply(counts.id, function(id){
  tmp = read.table(synGet(id)@filePath, header=T, sep='\t', check.names = F)
}) %>%
  plyr::join_all(type = 'full')
```

```{r preprocess.data1, results='asis'}
# Filter OTHER samples 
covariates = rbindlist(covariates, use.names = T, fill = T, idcol = 'Study') %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
  dplyr::filter(Diagnosis != 'OTHER') %>%
  dplyr::mutate(BrainRegion = factor(BrainRegion, levels = c('DLPFC', 'FP', 'IFG', 'PHG', 'STG', 'CER', 'TCX')),
                BrainRegion.Diagnosis = factor(paste(BrainRegion, Diagnosis, sep = '.')))

# Match covariates to expression data
indToRetain = intersect(covariates$SampleID, colnames(counts))

rownames(counts) = counts$feature
counts = counts[,indToRetain]

rownames(covariates) = covariates$SampleID
covariates = covariates[indToRetain,]

covariates %>% 
  dplyr::group_by(BrainRegion, Diagnosis) %>% 
  dplyr::summarise(count = length(unique(SampleID))) %>% 
  tidyr::spread(BrainRegion, count) %>%
  kable()

COUNT = counts
```

```{r gene.param}
# Get gene specific parameters from synapse
GENE.PARAM = downloadFile('syn8449369')
all.used.ids = c(all.used.ids, 'syn8449369')

GENE.LEN = dplyr::select(GENE.PARAM, Gene.ID, gene.length) %>% unique()
rownames(GENE.LEN) = GENE.LEN$Gene.ID

GENE.GC = dplyr::select(GENE.PARAM, Gene.ID, percentage_gc_content) %>% unique()
rownames(GENE.GC) = GENE.GC$Gene.ID

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "dec2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "gene_biotype"),
                       filters = "ensembl_gene_id", values = GENE.PARAM$ensembl_gene_id,
                       mart = mart)
GENE.PARAM = left_join(GENE.PARAM, Ensemble2HGNC)
```

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("individualIdentifier", "Study", "BrainRegion.Diagnosis", "BrainRegion", "Diagnosis",
                      "Gender", "Batch")
ContCovariates <- c("AOD", "PMI", "RIN", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES",
                    "PCT_RIBOSOMAL_BASES")

# Find inter relation between factor covariates
COVARIATES = covariates[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){
  x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
rownames(COVARIATES) <- covariates$SampleID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], function(x){
  x = as.numeric(as.character(gsub('[\\,\\%]','',x)))
})
COVARIATES$BrainRegion = factor(COVARIATES$BrainRegion, levels = c('DLPFC', 'FP', 'IFG', 'PHG', 'STG', 'CER', 'TCX'))

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=8}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
tmp = COVARIATES.CORRELATION$ESTIMATE
tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
h = Heatmap(tmp, name = 'EffectSize',
            cluster_rows = T, cluster_columns = T, 
            col = colorRamp2(c(-1,0,1),c('blue','white','red')),
            cell_fun = function(j,i,x,y,wifht,height,fill){
              grid.text(COVARIATES.CORRELATION$PVAL[i,j] %>% format(digits = 1), x, y)
            })
draw(h, heatmap_legend_side = 'left', padding = unit(c(20,2,2,15), 'mm'))
```

## Explore metatdata
```{r data.explore, fig.width = 16, fig.height = 8}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 45, hjust = 1), plot.title=element_text(hjust=0.5))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = BrainRegion, y = RIN, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = BrainRegion, y = AOD, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('Age Of Death') + my.theme

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PMI, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme

# Intronic Rate
p[[4]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_INTRONIC_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Intronic Factions') + my.theme

# IntergenicRate
p[[5]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_INTERGENIC_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Intergenic Fractions') + my.theme

# Ribosomal Fractions
p[[6]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_RIBOSOMAL_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Ribosomal Fractions') + my.theme

# Reads Aligned
p[[7]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_PF_READS_ALIGNED, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[7]] = p[[7]] + ggtitle('Fraction Reads Aligned') + my.theme

# Coding Bases
p[[8]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_CODING_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[8]] = p[[8]] + ggtitle('Fraction Coding Bases') + my.theme

multiplot(plotlist = p, cols = 4)
```
```{r data.explore1, fig.width = 6, fig.height = 6}
# BrainRegion
vcd::mosaic(~ BrainRegion + Diagnosis, data = COVARIATES)

# Gender
vcd::mosaic(~ BrainRegion + Gender, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion.Diagnosis. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = plyr::dlply(covariates, .(BrainRegion.Diagnosis), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC$Gene.ID[!is.na(GENE.GC$percentage_gc_content)]) %>%
  intersect(GENE.LEN$Gene.ID[!is.na(GENE.LEN$gene.length)])

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
pct.pc = PROCESSED_COUNTS$filteredExprMatrix$genes %>%
  dplyr::rename(Gene.ID = genes) %>%
  dplyr::left_join(GENE.PARAM) %>%
  dplyr::group_by(gene_biotype) %>%
  dplyr::summarise(fraction  = length(unique(Gene.ID))) %>%
  dplyr::filter(fraction > 100) %>%
  dplyr::mutate(fraction = fraction/length(PROCESSED_COUNTS$filteredExprMatrix$genes[,1]))
```
`r dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]` genes are used for the analysis.

Following is the distribution of biotypes
`r kable(pct.pc)`

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
indToRemove = c('367_120502')

# Find principal components of expression to plot
PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  dplyr::mutate(label = SampleID)
plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=BrainRegion, shape=Study, size=RIN))
p <- p + theme_bw() + theme(legend.position="top")
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p

# Plot abberent distribution of logcpm counts
tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
  rownameToFirstColumn('Gene.ID') %>%
  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID') %>%
              tidyr::separate(BrainRegion.Diagnosis, c('Dx', 'Tissue'), sep = '_'))

p = ggplot(tmp1 %>%
             dplyr::filter(SampleID %in% indToRemove),
           aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'top') + facet_grid(BrainRegion+.~Diagnosis, scale = 'free')
p

indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
COVARIATES = COVARIATES[indToRetain,]

tmp = COVARIATES %>%
  rownameToFirstColumn('SampleID') %>%
  dplyr::group_by(BrainRegion, Diagnosis) %>%
  dplyr::summarise(count = length(unique(SampleID))) %>%
  tidyr::spread(BrainRegion, count)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$individualIdentifier))` unique individuals

Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`

Distribution of samples are:
`r kable(tmp)`

### Winsorise counts (Gene outliers in samples)
```{r winsorise.data}
LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
LOG.CPM = apply(LOG.CPM, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))

  x[x < (mn-3*std.dev)] = x[ind.low]
  x[x > (mn+3*std.dev)] = x[ind.high]
  return(x)
}) %>% t

# Back calculate counts for further analysis
LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
```

### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                          x = GENE.GC[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
LOG.CPM = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))

  x[x < (mn-3*std.dev)] = x[ind.low]
  x[x > (mn+3*std.dev)] = x[ind.high]
  return(x)
}) %>% t
CQN.GENE_EXPRESSION$E = LOG.CPM
```

### Sample clustering
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=BrainRegion, shape=Study, size=RIN))
p <- p + theme_bw() + theme(legend.position="right")
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("BrainRegion", "Gender", "Diagnosis", "Study")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```
```{r temp, include = F}
dev.off()
gc()
```

### Distribution of samples (log cpm)
```{r lcpm.dist, cache=FALSE, fig.height=10, fig.width=20}
# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$E) %>%
  rownameToFirstColumn('Gene.ID') %>%
  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID') %>%
              tidyr::separate(BrainRegion.Diagnosis, c('Diagnosis', 'BrainRegion'), sep = '_'))

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'NONE') + facet_grid(BrainRegion+.~Diagnosis, scale = 'free')
p
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Model Identification
```{r fxn}
# Function to fit a mixed effect model and return BIC
fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){

  # Get design matrix
  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
  designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]

  # Estimate voom weights
  voomWeights = limma::voom(cnt[rownames(lcpm),colnames(lcpm)], design = designMat, plot=F)$weights

  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
                              .combine = rbind,
                              .packages = c('lme4', 'dplyr'),
                              .export = c('voomWeights')) %dopar% {

                                # Perform weighted mixed linear models for every gene
                                expr = data.frame(Gx = lcpm[i,])
                                data = cbind(expr,
                                             designMat[rownames(expr),],
                                             covars[rownames(designMat),groupingVar,drop=F])
                                formula = paste(setdiff(colnames(data), c('Gx', groupingVar)), collapse = '+') %>%
                                  paste('Gx ~',.) %>%
                                  c(.,paste0('(1|',groupingVar,')')) %>%
                                  paste(collapse = '+')

                                weights = voomWeights[i,]
                                BIC = NA
                                tryCatch({
                                  mdl = lmer(formula = formula, 
                                             data = data, 
                                             weights = weights)
                                  BIC = BIC(mdl)
                                }, error = function(e){
                                  BIC = NA
                                })
                              return(data.frame(Gene.ID = rownames(lcpm)[i],
                                                BIC = BIC))
                            }
  return(modelBIC)
  }

# Function to select covariates in a step-wise regression fashion
selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){

  varsSelected = c()
  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
  while(length(covarsToSelect) != 0){
    allModelBIC = foreach(i = 1:length(covarsToSelect),
                          .combine = c,
                          .packages = pkgs,
                          .export = c('fitMixedEffectModel')) %dopar% {

                            # Fit a mixed effect linear model and get all BIC
                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
                                                           c('individualIdentifier'), 
                                                           lcpm, 
                                                           cnt, 
                                                           covars) %>%
                            plyr::rename(c('BIC' = covarsToSelect[i]))

                            return(list(modelBIC))
                          }
    allModelBIC = plyr::join_all(allModelBIC)

    # Find number of genes that are changed based on BIC
    nChange = allModelBIC %>%
      tidyr::gather(ModelName, Value, -Gene.ID) %>%
      dplyr::left_join(baseModelBIC) %>%
      dplyr::mutate(diff = baseModel - Value) %>%
      dplyr::group_by(ModelName) %>%
      dplyr::summarise(better = sum(diff >= 5), 
                       worse = sum(diff <= -5)) %>%
      dplyr::mutate(nchange = (better/(better+worse)) * 100)

    print(nChange)

    # Select variable
    tmp = nChange %>%
      dplyr::filter(nchange > 1) %>%
      dplyr::top_n(1, nchange)

    if (dim(tmp)[1] == 0){
      covarsToSelect = c()
    } else {
      baseModelBIC = allModelBIC %>%
        dplyr::select(Gene.ID, one_of(tmp$ModelName))
      colnames(baseModelBIC)[2] = 'baseModel'
      varsSelected = c(varsSelected, tmp$ModelName)
      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
      print(paste('Variable chosen to be added:',tmp$ModelName))
    }
  }

  return(varsSelected)
}
```
#### Phase I (clinical, ancestry and sample specific technical variables)
```{r phase1}
# Fit a base model with the following covariates
adjust.covars = c('BrainRegion.Diagnosis')
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   c('individualIdentifier'), 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase1.covars = c('Study', 'Gender', 'PMI', 'AOD', 'RIN', 'RIN2')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`

#### Phase II (batch effects)
```{r phase2}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   c('individualIdentifier'), 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase2.covars = c('Batch')
vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`

#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   c('individualIdentifier'), 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase3.covars = c("PCT_PF_READS_ALIGNED", "PCT_CODING_BASES", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`

### Normalisation
1. BrainRegion.Diagnosis is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
2. Individual_ID is chosen as random effect
```{r iterative.normalisation, results='asis'}
writeLines(paste('Using following covariates in the model:',
                 paste(adjust.covars, collapse=', '),
                 'as fixed effects and Individual_ID is chosen as random effect'))

# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

# Estimate voom weights
VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F)

# Estimate consensus correlation coefficient
correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, 
                            design=DM1$design, plot=F,
                            block = COVARIATES$Individual_ID, 
                            correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```

### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("BrainRegion.Diagnosis", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

### Coexpression of residual expression 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=BrainRegion, shape=Study, size=RIN))
p <- p + theme_bw() + theme(legend.position="right")
p
```

```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('BrainRegion', 'Diagnosis', 'Gender', 'Study')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

```{r temp1, include=F}
dev.off()
gc()
```

### Differential expression analysis (with BrainRegion.Diagnosis as primary variable)
Differential expression is performed on the primary variable by controlling for covariates identified above
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp1, fig.height=7, fig.width=24}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, adjust.covars, drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)

# Calculate correlation values of random effects
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)

# Re-estimate voom weights
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$individualIdentifier, 
                    correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
FIT = lmFit(VOOM.WEIGHTS)

# Fit contrast
cntr = c()
for (region in c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
  tmp = combn(paste0('BrainRegion.Diagnosis', region, '_', c('AD', 'CONTROL')),2)
  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
}

contrast = makeContrasts(contrasts=cntr,
                         levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT.CONTR){
  topTable(FIT.CONTR, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('Gene.ID')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
                Study = 'ALL',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(GENE.PARAM %>%
              dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
  group_by(Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2) 
kable(tmp)

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position='top')
p = p + facet_grid(.~Comparison, scales = 'fixed')
p

all.diff.exp = c(all.diff.exp, list(Diagnosis = DE))
all.fit = c(all.fit, list(Diagnosis = FIT))
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=20, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE, aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE, aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE, aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

multiplot(plotlist = pl, cols = 3)
```

### Differential expression analysis (with BrainRegion.Diagnosis and Gender)
Differential expression is performed on the BrainRegion.Diagnosis x Gender variable by controlling for covariates identified above
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp2, fig.height=15, fig.width=13, eval = F}
## Modeling Diagnosis, msex and age_death conjointly
COVARIATES$BrainRegion.Diagnosis.Gender = factor(paste(COVARIATES$BrainRegion.Diagnosis,
                                                       COVARIATES$Gender,
                                                       sep = '_'))

# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis.Gender', setdiff(adjust.covars, 'Gender')), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)

# Calculate correlation values of random effects
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)

# Re-estimate voom weights
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$individualIdentifier, 
                    correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
FIT = lmFit(VOOM.WEIGHTS)

# Fit contrast
# Fit contrast
cntr = c()
for (region in c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
  tmp = combn(paste0('BrainRegion.Diagnosis.Gender', region, '_', c('AD', 'CONTROL')),2)
  tmp1 = combn(paste0())
  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
}

cntr = c('BrainRegion.Diagnosis.SEXIFG.AD.MALE-BrainRegion.Diagnosis.SEXIFG.CONTROL.MALE',
         'BrainRegion.Diagnosis.SEXIFG.AD.FEMALE-BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXIFG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXIFG.AD.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXIFG.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXIFG.AD.MALE-0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXIFG.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXIFG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.MALE',
         'BrainRegion.Diagnosis.SEXFP.AD.MALE-BrainRegion.Diagnosis.SEXFP.CONTROL.MALE',
         'BrainRegion.Diagnosis.SEXFP.AD.FEMALE-BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXFP.AD.MALE+0.5*BrainRegion.Diagnosis.SEXFP.AD.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXFP.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXFP.AD.MALE-0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXFP.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXFP.AD.MALE+0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.MALE',
         'BrainRegion.Diagnosis.SEXPHG.AD.MALE-BrainRegion.Diagnosis.SEXPHG.CONTROL.MALE',
         'BrainRegion.Diagnosis.SEXPHG.AD.FEMALE-BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXPHG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXPHG.AD.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXPHG.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXPHG.AD.MALE-0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXPHG.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXPHG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.MALE',
         'BrainRegion.Diagnosis.SEXSTG.AD.MALE-BrainRegion.Diagnosis.SEXSTG.CONTROL.MALE',
         'BrainRegion.Diagnosis.SEXSTG.AD.FEMALE-BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXSTG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXSTG.AD.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXSTG.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXSTG.AD.MALE-0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXSTG.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXSTG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.MALE')
contrast = makeContrasts(contrasts=cntr,
                         levels = colnames(FIT$coefficients))
colnames(contrast) = c('IFG.AD.MALE-IFG.CONTROL.MALE',
                       'IFG.AD.FEMALE-IFG.CONTROL.FEMALE',
                       'IFG.AD.ALL-IFG.CONTROL.ALL',
                       'IFG.ALL.FEMALE-IFG.ALL.MALE',
                       'IFG.ALL.ALL-IFG.ALL.ALL',
                       'FP.AD.MALE-FP.CONTROL.MALE',
                       'FP.AD.FEMALE-FP.CONTROL.FEMALE',
                       'FP.AD.ALL-FP.CONTROL.ALL',
                       'FP.ALL.FEMALE-FP.ALL.MALE',
                       'FP.ALL.ALL-FP.ALL.ALL',
                       'PHG.AD.MALE-PHG.CONTROL.MALE',
                       'PHG.AD.FEMALE-PHG.CONTROL.FEMALE',
                       'PHG.AD.ALL-PHG.CONTROL.ALL',
                       'PHG.ALL.FEMALE-PHG.ALL.MALE',
                       'PHG.ALL.ALL-PHG.ALL.ALL',
                       'STG.AD.MALE-STG.CONTROL.MALE',
                       'STG.AD.FEMALE-STG.CONTROL.FEMALE',
                       'STG.AD.ALL-STG.CONTROL.ALL',
                       'STG.ALL.FEMALE-STG.ALL.MALE',
                       'STG.ALL.ALL-STG.ALL.ALL')

FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('ensembl_gene_id')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
                Study = 'MSSM',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
  group_by(Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2) 
kable(tmp)

all.diff.exp = c(all.diff.exp, list(Diagnosis.SEX = DE))
all.fit = c(all.fit, list(Diagnosis.SEX = FIT))
```

### Differential expression analysis (with BrainRegion.Diagnosis and AOD)
Differential expression is performed on the BrainRegion.Diagnosis x AOD variable by controlling for covariates identified above

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp3, fig.height=15, fig.width=13, eval = F}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[2], setdiff(postAdjustCovars, 'AOD')), drop = F], Intercept = F)
ind = grep('Diagnosis', colnames(DESIGN$design))
DESIGN$design[,ind] = DESIGN$design[,ind] * COVARIATES[rownames(DESIGN$design),'AOD']
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
colnames(DESIGN$design)[ind] = gsub('BrainRegion.Diagnosis', '', paste0(colnames(DESIGN$design)[ind], '.AOD'))

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)

# Calculate correlation values of random effects
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)

# Re-estimate voom weights
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$individualIdentifier, 
                    correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
FIT = lmFit(VOOM.WEIGHTS)

# Fit contrast
cntr = c("FP.AD.AOD-FP.CONTROL.AOD",
         "0.5*FP.AD.AOD+0.5*FP.CONTROL.AOD",
         "IFG.AD.AOD-IFG.CONTROL.AOD",
         "0.5*IFG.AD.AOD+0.5*IFG.CONTROL.AOD",
         "PHG.AD.AOD-PHG.CONTROL.AOD",
         "0.5*PHG.AD.AOD+0.5*PHG.CONTROL.AOD",
         "STG.AD.AOD-STG.CONTROL.AOD",
         "0.5*STG.AD.AOD+0.5*STG.CONTROL.AOD")
contrast = makeContrasts(contrasts=cntr,
                         levels = colnames(FIT$coefficients))
colnames(contrast) = c("FP.AD.AOD-FP.CONTROL.AOD",
                       "FP.ALL.AOD",
                       "IFG.AD.AOD-IFG.CONTROL.AOD",
                       "IFG.ALL.AOD",
                       "PHG.AD.AOD-PHG.CONTROL.AOD",
                       "PHG.ALL.AOD",
                       "STG.AD.AOD-STG.CONTROL.AOD",
                       "STG.ALL.AOD")
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('ensembl_gene_id')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
                Study = 'MSSM',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
  group_by(Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2) 
kable(tmp)

all.diff.exp = c(all.diff.exp, list(Diagnosis.AOD = DE))
all.fit = c(all.fit, list(Diagnosis.AOD = FIT))
```

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16, eval = F}
diffexp.genes = DE %>%
  dlply(.(Comparison, Tissue), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
backgroundGenes = unique(DE$hgnc_symbol)

# Download all related genesets from synapse
geneset.files = synQuery('select * from file where parentId == "syn3240583"')
ALL_USED_IDs = c(ALL_USED_IDs, geneset.files$file.id)
genesets.to.test = sapply(geneset.files$file.name, function(dataSource, geneset.files){
  tmp = downloadFile(geneset.files$file.id[geneset.files$file.name == dataSource]) %>%
    dlply(.(Name), .fun = function(x){
      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
    })
  }, geneset.files) %>%
  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
  do.call(c,.)

# Perform enrichment analysis
enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
    }, x, backgroundGenes, .parallel = TRUE) %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
}, genesets.to.test, backgroundGenes) %>%
  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
  tidyr::separate(Comparison, c('Name', 'Region'), sep = '\\.') %>%
  dplyr::rename(Comparison = Name, Category = .id)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Set annotations
all.annotations = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay	 = 'RNAseq',
  assaySubType = 'Reprocessed',

  tissueTypeAbrv	= 'DLPFC, FP, STG, PHG, IFG, CER, TCX', 
  study = 'ALL', 
  center = 'ALL',

  organism = 'HomoSapiens',
  consortium	= 'AMP-AD',

  normalizationStatus	= TRUE,
  modelSystem	= FALSE,

  normalizationType	= 'CQN'
)

# Function to change entity ACL (also creates new acls if it doesn't exist)
synChangeEntityACL <- function(id, new.racl){
  tryCatch({ 
    # If old acl exist update
    old.acl = synGetEntityACL(id)
    old.acl@resourceAccess = new.racl
    synUpdateEntityACL(old.acl)
  }, error = function(e){ 
    # If not create a new acl
    new.acl = AccessControlList(id = id, resourceAccess = new.racl)
    synCreateEntityACL(new.acl)
  })
}

# Create new resource acls
new.racl = ResourceAccess(principalId = '3319864', 
                          accessType = c('UPDATE', 'CHANGE_SETTINGS', 'DELETE', 'READ', 
                                         'DOWNLOAD', 'MODERATE', 'CREATE', 'CHANGE_PERMISSIONS')) %>%
  ResourceAccessList()

# Code
CODE <- Folder(name = "ALL Reprocessed RNASeq (Normalised)", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)
synChangeEntityACL(CODE@properties$id, new.racl)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'AMP_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('AMP_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
annotations(COV_OBJ) = annotations(CODE)
annotations(COV_OBJ)$dataSubType = 'covariates'
COV_OBJ = synStore(COV_OBJ, used = all.used.ids, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('Gene.ID') %>%
  write.table(file = 'AMP_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('AMP_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store logCPM
CQN.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('Gene.ID') %>%
  write.table(file = 'AMP_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('AMP_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store cqn offsets
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('Gene.ID') %>%
  write.table(file = 'AMP_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('AMP_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
annotations(OFFSET_OBJ) = annotations(CODE)
annotations(OFFSET_OBJ)$dataSubType = 'offset'
OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store design matrix
DESIGN$design %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'AMP_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('AMP_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
annotations(DM_OBJ) = annotations(CODE)
annotations(DM_OBJ)$dataSubType = 'designMatrix'
DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store residual gene expression for network analysis
RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'AMP_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('AMP_netResidualExpression.tsv', 
                name = 'Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
annotations(nEXP_OBJ) = annotations(CODE)
annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store differential expression results
write.table(DE, file = 'AMP_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('AMP_DiffExpression.tsv', 
                name = 'Differential Expression Results (BrainRegion.Diagnosis)', 
                parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

### 2017-07-04 code/Rmd/MSSM_REPROCESSED.Rmd

@@
 Distribution of samples are: `r kable(tmp)`
+### Winsorise counts (Gene outliers in samples)
+```{r winsorise.data}
+LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)
+
+# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
+LOG.CPM = apply(LOG.CPM, 1, function(x){
+  mn = mean(x, na.rm = T)
+  std.dev = sd(x, na.rm = T)
+
+  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
+  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
+
+  x[x < (mn-3*std.dev)] = x[ind.low]
+  x[x > (mn+3*std.dev)] = x[ind.high]
+  return(x)
+}) %>% t
+
+# Back calculate counts for further analysis
+LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
+NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
+```
 
@@
 # Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
-                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
+CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
+                          x = GENE.GC[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
+                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'gene.length'],
                           lengthMethod = "smooth", 
                           verbose = FALSE)
-CQN.GENE_EXPRESSION$E.NA = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-
-# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
-log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
+CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+
+# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
+LOG.CPM = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
   mn = mean(x, na.rm = T)
   std.dev = sd(x, na.rm = T)
-  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
+
+  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
+  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
+
+  x[x < (mn-3*std.dev)] = x[ind.low]
+  x[x > (mn+3*std.dev)] = x[ind.high]
+  return(x)
 }) %>% t
-PROCESSED_COUNTS$filteredExprMatrix$counts.NA = PROCESSED_COUNTS$filteredExprMatrix$counts 
-PROCESSED_COUNTS$filteredExprMatrix$counts.NA[log.mat] = NA
-PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0
-
-CQN.GENE_EXPRESSION$E.NA[log.mat] = NA
-CQN.GENE_EXPRESSION$E.zero = CQN.GENE_EXPRESSION$E.NA
-CQN.GENE_EXPRESSION$E.zero[log.mat] = 0
+CQN.GENE_EXPRESSION$E = LOG.CPM
+
+PROCESSED_COUNTS$filteredExprMatrix$counts = NEW.COUNTS
 ```
@@
 # Plot abberent distribution of logcpm counts
-tmp1 = (CQN.GENE_EXPRESSION$E.NA) %>%
+tmp1 = (CQN.GENE_EXPRESSION$E) %>%
   rownameToFirstColumn('Gene.ID') %>%
@@
 ```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(CQN.GENE_EXPRESSION$E.zero))
+cr = cor(t(CQN.GENE_EXPRESSION$E))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(CQN.GENE_EXPRESSION$E.zero, scale.=T, center = T)
+PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E.zero)))
+tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 # Find correlation between PC's of gene expression with covariates
-preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E.zero, 
+preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E, 
                                                  COVARIATES,
@@
   # Calculate correlation values of random effects
-  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E.NA
   correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
@@
   # Fit linear model using new weights and new design
-  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$E.NA,
+  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$E,
                             design = DM1$design,
@@
   # Residuals after normalisation
-  RESIDUAL.GENE_EXPRESSION = list()
-  RESIDUAL.GENE_EXPRESSION$expr.NA = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
-                                                        CQN.GENE_EXPRESSION$E.NA)
-  RESIDUAL.GENE_EXPRESSION$expr.zero = RESIDUAL.GENE_EXPRESSION$expr.NA
-  RESIDUAL.GENE_EXPRESSION$expr.zero[is.na(RESIDUAL.GENE_EXPRESSION$expr.zero)] = 0
+  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
+                                                CQN.GENE_EXPRESSION$E)
 
@@
   # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-  tmp.expr = RESIDUAL.GENE_EXPRESSION$expr.zero
-  tmp.expr = tmp.expr[apply(tmp.expr, 1, var) != 0,]
-  residualSigCovars = runPCAandPlotCorrelations(tmp.expr, 
+  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                 COVARIATES[, residCovars, drop=F], 
@@
   loopCount = loopCount + 1
+  print(loopCount)
 }
-
+RESIDUAL.GENE_EXPRESSION
 modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
@@
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-tmp.expr = RESIDUAL.GENE_EXPRESSION$expr.zero
-tmp.expr = tmp.expr[apply(tmp.expr, 1, var) != 0,]
-residualSigCovars = runPCAandPlotCorrelations(tmp.expr, 
+residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                               COVARIATES,
@@
 ```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.GENE_EXPRESSION$expr.zero))
+cr = cor(t(RESIDUAL.GENE_EXPRESSION))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION$expr.zero, scale.=T, center = T)
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION$expr.zero)))
+tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 # Calculate correlation values of random effects
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
 correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)
@@
 # Fit linear model using new weights and new design
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
 FIT = lmFit(VOOM.WEIGHTS)
@@
 # Get residuals from differential expression model
-RESIDUAL.SVA.GENE_EXPRESSION = list()
-RESIDUAL.SVA.GENE_EXPRESSION$expr.NA = residuals.MArrayLM(FIT, CQN.GENE_EXPRESSION$E.NA)
-RESIDUAL.SVA.GENE_EXPRESSION$expr.zero = RESIDUAL.SVA.GENE_EXPRESSION$expr.NA
-RESIDUAL.SVA.GENE_EXPRESSION$expr.zero[is.na(RESIDUAL.SVA.GENE_EXPRESSION$expr.NA)] = 0
+RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, CQN.GENE_EXPRESSION$E)
 
 # Compute actual variance of all principal components
-tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION$expr.zero)
+tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
 actual.var = tmp$d^2/sum(tmp$d^2)
@@
 permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
-  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION$expr.zero, 1, sample, replace = FALSE))
+  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
   tmp = svd(tmp.residual)
@@
 # Estimate surrogate variables
-SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$E.zero,
+SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$E,
                     MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv
@@
 # Calculate correlation values of random effects
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
+VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
 correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$individualIdentifier)
@@
 # Fit linear model using new weights and new design
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
+VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
 VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)
@@
 RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
-                                                  CQN.GENE_EXPRESSION$E.NA)
+                                                  CQN.GENE_EXPRESSION$E)
 
@@
 # Calculate correlation values of random effects
-VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
+VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E
 correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$individualIdentifier)
@@
 # Fit linear model using new weights and new design
-VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
+VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E
 VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)
@@
 RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
-                                                  CQN.GENE_EXPRESSION$E.NA)
+                                                  CQN.GENE_EXPRESSION$E)
 ```
@@
 # Store logCPM
-tmp = CQN.GENE_EXPRESSION$y
-tmp[log.mat] = NA
-tmp %>%
+CQN.GENE_EXPRESSION$E %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
@@
 # Store cqn offsets
-tmp = CQN.GENE_EXPRESSION$offset
-tmp[log.mat] = NA
-tmp %>%
+CQN.GENE_EXPRESSION$offset %>%
   rownameToFirstColumn('ensembl_gene_id') %>%

### 2017-07-05 code/Rmd/ALL_REPROCESSED.Rmd
Minor modifications

@@
 
+```{r decompse.RESIDUALS, fig.height=8, fig.width=8, results='asis'}
+# Find principal components of expression to plot
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
+
+# Plot first 4 PCs
+plotdata <- data.frame(SampleID=rownames(PC$rotation), 
+                       PC1=PC$rotation[,1], 
+                       PC2=PC$rotation[,2])
+
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(color=BrainRegion, shape=Study, size=RIN))
+p <- p + theme_bw() + theme(legend.position="right")
+p
+```
+
 ### Residual calculation
-Calculate weighted residuals and add back "Dx" to the residuals
+Calculate weighted residuals and add back "BrainRegion.Diagnosis" to the residuals
 ```{r varsToAddBack}
@@
 
-all.diff.exp = c(all.diff.exp, list(Diagnosis = DE))
-all.fit = c(all.fit, list(Diagnosis = FIT))
+all.diff.exp = list(Diagnosis = DE)
+all.fit = list(Diagnosis = FIT)
 ```
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diff.exp2, fig.height=15, fig.width=13, eval = F}
+```{r diff.exp2, fig.height=7, fig.width=24}
 ## Modeling Diagnosis, msex and age_death conjointly
@@
 # Get design matrix
-DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis.Gender', setdiff(adjust.covars, 'Gender')), drop = F], Intercept = F)
+DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis.Gender', setdiff(adjust.covars, c('BrainRegion.Diagnosis', 'Gender'))), drop = F], Intercept = F)
 DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
@@
 # Fit contrast
-# Fit contrast
 cntr = c()
 for (region in  c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
-  tmp = combn(paste0('BrainRegion.Diagnosis.Gender', region, '_', c('AD', 'CONTROL')),2)
-  tmp1 = combn(paste0())
-  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
+  cnt = c(paste0('BrainRegion.Diagnosis.Gender', region, '_AD_MALE-BrainRegion.Diagnosis.Gender', region, '_CONTROL_MALE'),
+          paste0('BrainRegion.Diagnosis.Gender', region, '_AD_FEMALE-BrainRegion.Diagnosis.Gender', region, '_CONTROL_FEMALE'),
+          paste0('0.5*BrainRegion.Diagnosis.Gender', region, '_AD_MALE+0.5*BrainRegion.Diagnosis.Gender', region, 
+                 '_AD_FEMALE-0.5*BrainRegion.Diagnosis.Gender', region, '_CONTROL_MALE-0.5*BrainRegion.Diagnosis.Gender',region, '_CONTROL_FEMALE'),
+          paste0('0.5*BrainRegion.Diagnosis.Gender', region, '_AD_FEMALE+0.5*BrainRegion.Diagnosis.Gender', region, 
+                 '_CONTROL_FEMALE-0.5*BrainRegion.Diagnosis.Gender', region, '_AD_MALE-0.5*BrainRegion.Diagnosis.Gender',region, '_CONTROL_MALE'),
+          paste0('0.5*BrainRegion.Diagnosis.Gender', region, '_AD_FEMALE-0.5*BrainRegion.Diagnosis.Gender', region, 
+                 '_CONTROL_FEMALE-0.5*BrainRegion.Diagnosis.Gender', region, '_AD_MALE+0.5*BrainRegion.Diagnosis.Gender',region, '_CONTROL_MALE'))
+  cntr = c(cntr, cnt)
 }
-
-cntr = c('BrainRegion.Diagnosis.SEXIFG.AD.MALE-BrainRegion.Diagnosis.SEXIFG.CONTROL.MALE',
-         'BrainRegion.Diagnosis.SEXIFG.AD.FEMALE-BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE',
-         '0.5*BrainRegion.Diagnosis.SEXIFG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXIFG.AD.FEMALE
-         -0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE',
-         '0.5*BrainRegion.Diagnosis.SEXIFG.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE
-         -0.5*BrainRegion.Diagnosis.SEXIFG.AD.MALE-0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE',
-         '0.5*BrainRegion.Diagnosis.SEXIFG.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE
-         -0.5*BrainRegion.Diagnosis.SEXIFG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.MALE',
-         'BrainRegion.Diagnosis.SEXFP.AD.MALE-BrainRegion.Diagnosis.SEXFP.CONTROL.MALE',
-         'BrainRegion.Diagnosis.SEXFP.AD.FEMALE-BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE',
-         '0.5*BrainRegion.Diagnosis.SEXFP.AD.MALE+0.5*BrainRegion.Diagnosis.SEXFP.AD.FEMALE
-         -0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE',
-         '0.5*BrainRegion.Diagnosis.SEXFP.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE
-         -0.5*BrainRegion.Diagnosis.SEXFP.AD.MALE-0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE',
-         '0.5*BrainRegion.Diagnosis.SEXFP.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE
-         -0.5*BrainRegion.Diagnosis.SEXFP.AD.MALE+0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.MALE',
-         'BrainRegion.Diagnosis.SEXPHG.AD.MALE-BrainRegion.Diagnosis.SEXPHG.CONTROL.MALE',
-         'BrainRegion.Diagnosis.SEXPHG.AD.FEMALE-BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE',
-         '0.5*BrainRegion.Diagnosis.SEXPHG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXPHG.AD.FEMALE
-         -0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE',
-         '0.5*BrainRegion.Diagnosis.SEXPHG.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE
-         -0.5*BrainRegion.Diagnosis.SEXPHG.AD.MALE-0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE',
-         '0.5*BrainRegion.Diagnosis.SEXPHG.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE
-         -0.5*BrainRegion.Diagnosis.SEXPHG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.MALE',
-         'BrainRegion.Diagnosis.SEXSTG.AD.MALE-BrainRegion.Diagnosis.SEXSTG.CONTROL.MALE',
-         'BrainRegion.Diagnosis.SEXSTG.AD.FEMALE-BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE',
-         '0.5*BrainRegion.Diagnosis.SEXSTG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXSTG.AD.FEMALE
-         -0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE',
-         '0.5*BrainRegion.Diagnosis.SEXSTG.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE
-         -0.5*BrainRegion.Diagnosis.SEXSTG.AD.MALE-0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE',
-         '0.5*BrainRegion.Diagnosis.SEXSTG.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE
-         -0.5*BrainRegion.Diagnosis.SEXSTG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.MALE')
 contrast = makeContrasts(contrasts=cntr,
                          levels = colnames(FIT$coefficients))
-colnames(contrast) = c('IFG.AD.MALE-IFG.CONTROL.MALE',
-                       'IFG.AD.FEMALE-IFG.CONTROL.FEMALE',
-                       'IFG.AD.ALL-IFG.CONTROL.ALL',
-                       'IFG.ALL.FEMALE-IFG.ALL.MALE',
-                       'IFG.ALL.ALL-IFG.ALL.ALL',
-                       'FP.AD.MALE-FP.CONTROL.MALE',
-                       'FP.AD.FEMALE-FP.CONTROL.FEMALE',
-                       'FP.AD.ALL-FP.CONTROL.ALL',
-                       'FP.ALL.FEMALE-FP.ALL.MALE',
-                       'FP.ALL.ALL-FP.ALL.ALL',
-                       'PHG.AD.MALE-PHG.CONTROL.MALE',
-                       'PHG.AD.FEMALE-PHG.CONTROL.FEMALE',
-                       'PHG.AD.ALL-PHG.CONTROL.ALL',
-                       'PHG.ALL.FEMALE-PHG.ALL.MALE',
-                       'PHG.ALL.ALL-PHG.ALL.ALL',
-                       'STG.AD.MALE-STG.CONTROL.MALE',
-                       'STG.AD.FEMALE-STG.CONTROL.FEMALE',
-                       'STG.AD.ALL-STG.CONTROL.ALL',
-                       'STG.ALL.FEMALE-STG.ALL.MALE',
-                       'STG.ALL.ALL-STG.ALL.ALL')
+cnames = c()
+for (region in  c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
+  cnt = c(paste0(region, '_AD_MALE-', region, '_CONTROL_MALE'),
+          paste0(region, '_AD_FEMALE-', region, '_CONTROL_FEMALE'),
+          paste0(region, '_AD_ALL-', region, '_CONTROL_ALL'),
+          paste0(region, '_ALL_FEMALE-', region, '_ALL_MALE'),
+          paste0(region, '_ALL_ALL-', region,'_ALL_ALL'))
+  cnames = c(cnames, cnt)
+}
+colnames(contrast) = cnames
 
@@
   topTable(FIT, coef=i, number = 50000, confint = T) %>%
-    rownameToFirstColumn('ensembl_gene_id')
+    rownameToFirstColumn('Gene.ID')
 }, FIT.CONTR)
@@
   rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
-                Study = 'MSSM',
-                Direction = logFC/abs(logFC),
+  dplyr::mutate(Direction = logFC/abs(logFC),
                 Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
@@
   left_join(GENE.PARAM %>%
-              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
+              dplyr::select(Gene.ID, hgnc_symbol, percentage_gc_content, gene.length) %>%
               unique())
@@
 tmp = DE %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
+  dplyr::select(Gene.ID, Comparison, Direction) %>%
   group_by(Comparison, Direction) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(Gene.ID))) %>%
   spread(Direction, FDR_0_05_FC_1.2) 
@@
 Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
-```{r diff.exp3, fig.height=15, fig.width=13, eval = F}
+```{r diff.exp3, fig.height=7, fig.width=24}
 # Get design matrix
-DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[2], setdiff(postAdjustCovars, 'AOD')), drop = F], Intercept = F)
-ind = grep('Diagnosis', colnames(DESIGN$design))
-DESIGN$design[,ind] = DESIGN$design[,ind] * COVARIATES[rownames(DESIGN$design),'AOD']
+DESIGN = getDesignMatrix(COVARIATES[, c(setdiff(adjust.covars, 'AOD')), drop = F], Intercept = F)
+ind = grep('BrainRegion.Diagnosis', colnames(DESIGN$design))
+DESIGN$design[,ind] = DESIGN$design[,ind]*COVARIATES$AOD
+colnames(DESIGN$design)[ind] = paste(colnames(DESIGN$design)[ind], 'AOD', sep = '_')
 DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
-colnames(DESIGN$design)[ind] = gsub('BrainRegion.Diagnosis', '', paste0(colnames(DESIGN$design)[ind], '.AOD'))
 
@@
 # Fit contrast
-cntr = c("FP.AD.AOD-FP.CONTROL.AOD",
-         "0.5*FP.AD.AOD+0.5*FP.CONTROL.AOD",
-         "IFG.AD.AOD-IFG.CONTROL.AOD",
-         "0.5*IFG.AD.AOD+0.5*IFG.CONTROL.AOD",
-         "PHG.AD.AOD-PHG.CONTROL.AOD",
-         "0.5*PHG.AD.AOD+0.5*PHG.CONTROL.AOD",
-         "STG.AD.AOD-STG.CONTROL.AOD",
-         "0.5*STG.AD.AOD+0.5*STG.CONTROL.AOD")
+cntr = c()
+for (region in  c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
+  cnt = c(paste0('BrainRegion.Diagnosis', region, '_AD_AOD-BrainRegion.Diagnosis', region, '_CONTROL_AOD'),
+          paste0('0.5*BrainRegion.Diagnosis', region, '_AD_AOD+0.5*BrainRegion.Diagnosis', region, '_CONTROL_AOD'))
+  cntr = c(cntr, cnt)
+}
 contrast = makeContrasts(contrasts=cntr,
                          levels = colnames(FIT$coefficients))
-colnames(contrast) = c("FP.AD.AOD-FP.CONTROL.AOD",
-                       "FP.ALL.AOD",
-                       "IFG.AD.AOD-IFG.CONTROL.AOD",
-                       "IFG.ALL.AOD",
-                       "PHG.AD.AOD-PHG.CONTROL.AOD",
-                       "PHG.ALL.AOD",
-                       "STG.AD.AOD-STG.CONTROL.AOD",
-                       "STG.ALL.AOD")
+cnames = c()
+for (region in  c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
+  cnt = c(paste0(region, '_AD_AOD-', region, '_CONTROL_AOD'),
+          paste0(region, '_ALL_AOD'))
+  cnames = c(cnames, cnt)
+}
+colnames(contrast) = cnames
+
 FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
@@
 all.fit = c(all.fit, list(Diagnosis.AOD = FIT))
+```
+
+### Differential expression analysis (with BrainRegion.Diagnosis and Celltype Fractions)
+Differential expression is performed on the BrainRegion.Diagnosis x CelltypeFractions variable by controlling for covariates identified above
+
+Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
+```{r diff.exp3, fig.height=7, fig.width=24, eval = F}
 ```

### 2017-07-06 code/Rmd/ALL_REPROCESSED.Rmd
Minor modifications

@@
 
-all.diff.exp = c(all.diff.exp, list(Diagnosis.SEX = DE))
-all.fit = c(all.fit, list(Diagnosis.SEX = FIT))
+all.diff.exp = c(all.diff.exp, list(Diagnosis.Gender = DE))
+all.fit = c(all.fit, list(Diagnosis.Gender = FIT))
 ```
@@
 Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
-```{r diff.exp3, fig.height=7, fig.width=24, eval = F}
+```{r diff.exp4.1, fig.height=10, fig.width=15}
+# Cell type markers
+CELL.FRAC = downloadFile('syn10163089')
+all.used.ids = c(all.used.ids, 'syn10163089')
+
+# Merge covariates to cell type fraction
+NEW.COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID') %>%
+  left_join(CELL.FRAC)
+
+my.theme = theme(legend.position = 'top', axis.text.x = element_text(angle = 45, hjust = 1))
+pl = list()
+pl[[1]] = ggplot(NEW.COVARIATES, aes(x = BrainRegion, y = astrocytes, color = Diagnosis)) + geom_boxplot() + ggtitle('Astrocytes') + my.theme
+pl[[2]] = ggplot(NEW.COVARIATES, aes(x = BrainRegion, y = endothelial, color = Diagnosis)) + geom_boxplot() + ggtitle('Endothelial') + my.theme
+pl[[3]] = ggplot(NEW.COVARIATES, aes(x = BrainRegion, y = microglia, color = Diagnosis)) + geom_boxplot() + ggtitle('Microglia') + my.theme
+pl[[4]] = ggplot(NEW.COVARIATES, aes(x = BrainRegion, y = neurons, color = Diagnosis)) + geom_boxplot() + ggtitle('Neurons') + my.theme
+pl[[5]] = ggplot(NEW.COVARIATES, aes(x = BrainRegion, y = myeloid, color = Diagnosis)) + geom_boxplot() + ggtitle('Myeloid') + my.theme
+pl[[6]] = ggplot(NEW.COVARIATES, aes(x = BrainRegion, y = oligodendrocytes, color = Diagnosis)) + geom_boxplot() + ggtitle('Oligodendrocytes') + my.theme
+multiplot(plotlist = pl, cols = 3)
+
+# Get design matrix
+DESIGN = getDesignMatrix(NEW.COVARIATES[, c(adjust.covars, "astrocytes", "endothelial", "microglia", "myeloid", "neurons", "oligodendrocytes", "OPC"), drop = F], Intercept = F)
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
+
+# Calculate correlation values of random effects
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)
+
+# Re-estimate voom weights
+VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
+                    design=DESIGN$design, plot=F,
+                    block = COVARIATES$individualIdentifier, 
+                    correlation = correlation$cor)
+
+# Fit linear model using new weights and new design
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+FIT = lmFit(VOOM.WEIGHTS)
+
+# Fit contrast
+cntr = c()
+for (region in  c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
+  cnt = c(paste0('BrainRegion.Diagnosis', region, '_AD-BrainRegion.Diagnosis', region, '_CONTROL'))
+  cntr = c(cntr, cnt)
+}
+contrast = makeContrasts(contrasts=cntr,
+                         levels = colnames(FIT$coefficients))
+cnames = c()
+for (region in  c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
+  cnt = c(paste0(region, '_AD_CELLFRAC-', region, '_CONTROL_CELLFRAC'))
+  cnames = c(cnames, cnt)
+}
+colnames(contrast) = cnames
+
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = T) %>%
+    rownameToFirstColumn('Gene.ID')
+}, FIT.CONTR)
+names(DE) = colnames(contrast)
+
+DE = DE %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  left_join(GENE.PARAM %>%
+              dplyr::select(Gene.ID, hgnc_symbol, percentage_gc_content, gene.length) %>%
+              unique())
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE %>%
+  dplyr::select(Gene.ID, Comparison, Direction) %>%
+  group_by(Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(Gene.ID))) %>%
+  spread(Direction, FDR_0_05_FC_1.2) 
+kable(tmp)
+
+all.diff.exp = c(all.diff.exp, list(Diagnosis.CellFraction = DE))
+all.fit = c(all.fit, list(Diagnosis.CellFraction = FIT))
 ```
@@
 Using Fisher's exact test for gene set enrichment analysis
-```{r enrich.de, fig.height=16, fig.width=16, eval = F}
-diffexp.genes = DE %>%
-  dlply(.(Comparison, Tissue), .fun = function(x){
-    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
-    tmp = tmp[!is.na(tmp)]
+```{r enrich.de, fig.height=16, fig.width=16}
+# Perform gene set enrichment analysis using wilcox test from limma
+diffexp.genes = all.diff.exp %>%
+  data.table::rbindlist(idcol='Model', use.names = T, fill = T) %>%
+  dplyr::mutate(Direction = sign(logFC),
+                Direction = factor(Direction, labels = c('-1' = 'DOWN', '1' = 'UP')))
+
+geneSet.test = diffexp.genes %>%
+  plyr::dlply(.(Model, Comparison, Direction), .fun = function(x){
+    unique(x$hgnc_symbol[x$adj.P.Val <= 0.05])
   })
-diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
+
+statistics.test = diffexp.genes %>%
+  plyr::dlply(.(Model, Comparison), .fun = function(x){
+    tmp = dplyr::select(x, hgnc_symbol, adj.P.Val) %>% dplyr::group_by(hgnc_symbol) %>% dplyr::top_n(1,-adj.P.Val) %>% unique() %>% as.data.frame()
+    rownames(tmp) = tmp$hgnc_symbol
+    return(tmp)
+  })
+names(statistics.test) = paste(names(statistics.test),'UP',sep = '.')
+statistics.test1 = statistics.test
+names(statistics.test1) = gsub('.UP','.DOWN', names(statistics.test1))
+statistics.test = c(statistics.test, statistics.test1)
+
 backgroundGenes = unique(DE$hgnc_symbol)
@@
 # Download all related genesets from synapse
-geneset.files = synQuery('select * from file where parentId == "syn3240583"')
-ALL_USED_IDs = c(ALL_USED_IDs, geneset.files$file.id)
-genesets.to.test = sapply(geneset.files$file.name, function(dataSource, geneset.files){
-  tmp = downloadFile(geneset.files$file.id[geneset.files$file.name == dataSource]) %>%
-    dlply(.(Name), .fun = function(x){
-      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
-    })
-  }, geneset.files) %>%
+load(synGet('syn4867851')@filePath)
+all.used.ids = c(all.used.ids, 'syn4867851')
+geneSet.ref = GeneSets[c('KEGG_2016', 'GO_Biological_Process_2015', 'Reactome_2016', 'WikiPathways_2016')] %>%
   CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
@@
 # Perform enrichment analysis
-enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
-   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
-     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
-    }, x, backgroundGenes, .parallel = TRUE) %>%
-    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
-}, genesets.to.test, backgroundGenes) %>%
-  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
-  tidyr::separate(Comparison, c('Name', 'Region'), sep = '\\.') %>%
-  dplyr::rename(Comparison = Name, Category = .id)
+enrichment.results = mapply(function(gs.test, st.test, gs.ref){
+  plyr::ldply(gs.ref, .fun = function(y, gs.test, st.test){
+    tmp = CovariateAnalysis::fisherEnrichment(gs.test, y, st.test$hgnc_symbol)
+    tmp$wilcox.pval = limma::geneSetTest(which(st.test$hgnc_symbol %in% y), st.test[,'adj.P.Val'])
+    return(tmp)
+  }, gs.test, st.test, .parallel = T, .id = 'GeneSetName') %>%
+    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'),
+                  wilcox.fdr = p.adjust(wilcox.pval, method = 'fdr'))
+}, geneSet.test, statistics.test[names(geneSet.test)], 
+MoreArgs = list(geneSet.ref),
+SIMPLIFY = F) %>%
+  data.table::rbindlist(idcol = 'Comparison', use.names = T, fill = T)
 ```
@@
 DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store residual gene expression for network analysis
-RESIDUAL.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'AMP_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('AMP_netResidualExpression.tsv', 
-                name = 'Normalised, covariates removed residual expression (for network analysis)', 
-                parentId = CODE$properties$id)
-annotations(nEXP_OBJ) = annotations(CODE)
-annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
-nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+synChangeEntityACL(DM_OBJ@properties$id, new.racl)
 
 # Store differential expression results
-write.table(DE, file = 'AMP_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+write.table(diffexp.genes, file = 'AMP_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
 DEXP_OBJ = File('AMP_DiffExpression.tsv', 
-                name = 'Differential Expression Results (BrainRegion.Diagnosis)', 
+                name = 'Differential Expression Results', 
                 parentId = CODE$properties$id)
@@
 DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
+# Store enrichment results
+write.table(enrichment.results, file = 'AMP_EnrichmentResults.tsv', sep = '\t', row.names=F, quote=F)
+ENRICH_OBJ = File('AMP_EnrichmentResults.tsv', 
+                  name = 'GeneSet Enrichment Results', 
+                  parentId = CODE$properties$id)
+annotations(ENRICH_OBJ) = annotations(CODE)
+annotations(ENRICH_OBJ)$dataSubType = 'enrichment'
+ENRICH_OBJ = synStore(ENRICH_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
+# Store genesets
+save(list = 'geneSet.test', file = 'AMP_DiffExpGeneSet.RData')
+DEG_OBJ = File('AMP_DiffExpGeneSet.RData',
+               name = 'Differential Expression Gene Set', 
+               parentId = CODE$properties$id)
+annotations(DEG_OBJ) = annotations(CODE)
+annotations(DEG_OBJ)$dataSubType = 'diffExpGS'
+DEG_OBJ = synStore(DEG_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
+synDeleteEntityACL(CODE@properties$id)
 ```
 |  *Results*                                  |  *SynapseID*                     |
-|  -----------------------------------------  |   ---------                      |
+|  -----------------------------------------  |   ----------------------------   |
 |  Covariates                                 |  `r COV_OBJ$properties$id`       |
@@
 |  Design Matrix                              |  `r DM_OBJ$properties$id`        |
-|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
 |  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
+|  Enrichment Results                         |  `r ENRICH_OBJ$properties$id`    |
+|  Differentially Expressed Gene Sets         |  `r DEG_OBJ$properties$id`       |

### 2017-07-07 code/Rmd/ALL_REPROCESSED.Rmd
Minor bug fix

@@
                 BrainRegion.Diagnosis = factor(paste(BrainRegion, Diagnosis, sep = '.')),
-                msex = factor(msex, labels = c('1'='MALE','0'='FEMALE')),
+                msex = factor(msex, levels = c('1', '0'), labels = c('1'='MALE','0'='FEMALE')),
                 Batch = factor(Batch)) %>%
@@
                 individualIdentifier = factor(paste0('BB',individualIdentifier)),
-                SEX = factor(SEX, labels = c('M'='MALE','F'='FEMALE')),
+                SEX = factor(SEX, levels = c('M', 'F'), labels = c('M'='MALE','F'='FEMALE')),
                 batch = factor(batch),
@@
                 Donor_ID = factor(paste0('MAYO', Donor_ID)),
-                Gender = factor(Gender, labels = c('M'='MALE','F'='FEMALE')),
+                Gender = factor(Gender, levels = c('M', 'F'), labels = c('M'='MALE','F'='FEMALE')),
                 Batch = factor(Source)) %>%
@@
 p[[4]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_INTRONIC_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
-p[[4]] = p[[4]] + ggtitle('Intronic Factions') + my.theme
+p[[4]] = p[[4]] + ggtitle('Intronic Fractions') + my.theme
 
@@
 
+sdev = round((PC$sdev^2/sum(PC$sdev^2)) * 100, 0)
+
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
 p <- p + geom_point(aes(color=BrainRegion, shape=Study, size=RIN))
-p <- p + theme_bw() + theme(legend.position="top")
+p <- p + theme_bw() + theme(legend.position="top") + ylab(paste0('PC1(',sdev[2],'%)')) + xlab(paste0('PC1(',sdev[1],'%)'))
 p <- p + geom_text(aes(label= label), size=4, hjust=0)

### 2017-07-22 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Minor changes

@@
 # Get merged metadata
-METADATA_ID = 'syn4484273'
+METADATA_ID = 'syn2299154'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
@@
   dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
+                DLPFC_RNA_report__Library_Batch, DLPFC_RNA_report__Ribozero_Batch, DLPFC_RNA_report__Flowcell_Batch,
                 DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, 
@@
                 DLPFC_RNA_report__Library_Batch, DLPFC_RNA_report__Flowcell_Batch) %>%
-  dplyr::rename(PMI = PMI__in_hours_, 
+  dplyr::rename(LibraryBatch = DLPFC_RNA_report__Library_Batch, 
+                RibozeroBatch = DLPFC_RNA_report__Ribozero_Batch, 
+                FlowcellBatch = DLPFC_RNA_report__Flowcell_Batch,
+                PMI = PMI__in_hours_, 
                 IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
@@
 
-writeLines('Some control samples with Age of death less than 21 are removed')
-writeLines(paste('Following',length(METADATA$SampleID[METADATA$Age_of_Death <= 21]), 'samples are removed:'))
-writeLines(paste(METADATA$SampleID[METADATA$Age_of_Death <= 21], collapse = ', '))
-METADATA <- METADATA  %>% dplyr::filter(Age_of_Death >= 21)
+writeLines('Some control samples with Age of death less than 18 are removed')
+writeLines(paste('Following',length(METADATA$SampleID[METADATA$Age_of_Death <= 18]), 'samples are removed:'))
+writeLines(paste(METADATA$SampleID[METADATA$Age_of_Death <= 18], collapse = ', '))
+METADATA <- METADATA  %>% dplyr::filter(Age_of_Death >= 18)
 ```
@@
 plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''
+plotdata$label[plotdata$Age_of_Death <= 21] = plotdata$SampleID[plotdata$Age_of_Death <= 21]
 

### 2017-07-23 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Minor changes

@@
 knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd",
-                                 parentId ="syn7501880",
-                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection 1)')
+                                 parentId ="syn9846623",
+                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - Sparse Model')
 ```
@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = 'syn7501880';
+parentId = 'syn9846623';
 activityName = 'Covariate adjustments';
@@
 # Code
-CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection 1)", parentId = parentId)
+CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - Sparse Model", parentId = parentId)
 annotations(CODE) = all.annotations

### 2017-07-23 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Minor changes

@@
 
-tmp = de.hbcc %>%
+tmp = DE %>%
   dplyr::filter(adj.P.Val <= 0.05) %>%
@@
 
-tmp1 = de.hbcc %>%
+tmp1 = DE %>%
   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
@@
 ```{r enrich.de, fig.height=16, fig.width=16}
-diffexp.genes = de.hbcc %>%
+diffexp.genes = DE %>%
   dplyr::mutate(Direction = sign(logFC),

### 2017-07-27 code/Rmd/ALL_REPROCESSED_v2.Rmd
Added markdown to process all rnaseq data from three brain banck with study regressed prior to model selection

--- 2017-07-07 code/Rmd/ALL_REPROCESSED.Rmd
@@
 ---
-title: "Differential expression analysis of reprocessed AMPAD RNASeq data (All 7 brain regions with CQN normlaisation and BIC based model selection)"
+title: "Differential expression analysis of reprocessed AMPAD RNASeq data (All 7 brain regions with CQN normlaisation and BIC based model selection with study regressed apriori)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./ALL_REPROCESSED.Rmd",
+knit2synapse::knitToFolderEntity(file = "./ALL_REPROCESSED_v2.Rmd",
                                  parentId = "syn8672415",
-                                 entityName = 'ALL Reprocessed RNASeq (Normalised)')
+                                 entityName = 'ALL Reprocessed RNASeq (Normalised) - study effects removed')
 ```
@@
 
-thisFileName <- 'ALL_REPROCESSED.Rmd'
+thisFileName <- 'ALL_REPROCESSED_v2.Rmd'
 
@@
 p[[4]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_INTRONIC_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
-p[[4]] = p[[4]] + ggtitle('Intronic Fractions') + my.theme
+p[[4]] = p[[4]] + ggtitle('Intronic Fractions') + my.theme + ylab('Fraction')
 
@@
 p[[5]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_INTERGENIC_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
-p[[5]] = p[[5]] + ggtitle('Intergenic Fractions') + my.theme
+p[[5]] = p[[5]] + ggtitle('Intergenic Fractions') + my.theme + ylab('Fraction')
 
@@
 p[[6]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_RIBOSOMAL_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
-p[[6]] = p[[6]] + ggtitle('Ribosomal Fractions') + my.theme
+p[[6]] = p[[6]] + ggtitle('Ribosomal Fractions') + my.theme + ylab('Fraction')
 
@@
 p[[7]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_PF_READS_ALIGNED, color = Study, fill = Diagnosis)) + geom_boxplot()
-p[[7]] = p[[7]] + ggtitle('Fraction Reads Aligned') + my.theme
+p[[7]] = p[[7]] + ggtitle('Fraction Reads Aligned') + my.theme + ylab('Fraction')
 
@@
 p[[8]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_CODING_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
-p[[8]] = p[[8]] + ggtitle('Fraction Coding Bases') + my.theme
+p[[8]] = p[[8]] + ggtitle('Fraction Coding Bases') + my.theme + ylab('Fraction')
 
@@
 p <- p + geom_point(aes(color=BrainRegion, shape=Study, size=RIN))
-p <- p + theme_bw() + theme(legend.position="right")
+p <- p + theme_bw() + theme(legend.position="top") + xlab(paste0('PC1(',round(100*PC$sdev[1]^2/sum(PC$sdev^2)),'%)')) 
+p <- p + ylab(paste0('PC2(',round(100*PC$sdev[2]^2/sum(PC$sdev^2)),'%)')) 
 p
@@
 # Function to fit a mixed effect model and return BIC
-fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){
+fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, voomWeights, covars){
 
@@
   designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]
-
-  # Estimate voom weights
-  voomWeights = limma::voom(cnt[rownames(lcpm),colnames(lcpm)], design = designMat, plot=F)$weights
 
@@
 # Function to select covariates in a step-wise regression fashion
-selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){
+selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, voomWeight, baseModelBIC){
 
@@
                           .packages = pkgs,
-                          .export = c('fitMixedEffectModel')) %dopar% {
+                          .export = c('fitMixedEffectModel','voomWeight')) %dopar% {
 
@@
                                                            lcpm, 
-                                                           cnt, 
+                                                           voomWeight, 
                                                            covars) %>%
@@
     tmp = nChange %>%
-      dplyr::filter(nchange > 1) %>%
+      dplyr::filter(nchange > 5) %>%
       dplyr::top_n(1, nchange)
@@
 ```
+#### Regress out Study 
+```{r regress.study}
+# Get design matrix
+DESIGN.MAT = getDesignMatrix(COVARIATES[colnames(NEW.COUNTS), 'Study', drop=F], Intercept = F)
+DESIGN.MAT = DESIGN.MAT$design[,linColumnFinder(DESIGN.MAT$design)$indepCols]
+
+# Estimate voom weights
+VOOM.WEIGHTS = limma::voom(NEW.COUNTS, design = DESIGN.MAT, plot=F)
+
+# Fit a linear model
+FIT = lmFit(VOOM.WEIGHTS)
+
+# Calculate residuals
+RESIDUAL.EXPR = residuals.MArrayLM(FIT, VOOM.WEIGHTS$E)
+
+# Find principal components of expression to plot
+PC <- prcomp(RESIDUAL.EXPR, scale.=T, center = T)
+
+# Plot first 4 PCs
+plotdata <- data.frame(SampleID=rownames(PC$rotation), 
+                       PC1=PC$rotation[,1], 
+                       PC2=PC$rotation[,2])
+
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(color=BrainRegion, shape=Study, size=RIN))
+p <- p + theme_bw() + theme(legend.position="right")
+p
+```
 #### Phase I (clinical, ancestry and sample specific technical variables)
@@
                                    c('individualIdentifier'), 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
+                                   RESIDUAL.EXPR, 
+                                   VOOM.WEIGHTS$weights, 
                                    COVARIATES) %>%
@@
 # Iteratively identify covariates in phase I
-phase1.covars = c('Study', 'Gender', 'PMI', 'AOD', 'RIN', 'RIN2')
+phase1.covars = c('Gender', 'PMI', 'AOD', 'RIN', 'RIN2')
 vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                      COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
+                                     VOOM.WEIGHTS$weights, baseModelBIC)
 
@@
                                      COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
+                                     VOOM.WEIGHTS$weights, baseModelBIC)
 

### 2017-07-27 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor changes

--- 2017-05-31 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
@@
 knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
-                                 parentId ="syn7501880",
+                                 parentId = "syn7501880",
                                  entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
@@
                 "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")
-
-# Based on number of gene outliers
-indToRemove = c(indToRemove, "MSSM_RNA_ACC_BP_15", "MSSM_RNA_ACC_175", "MSSM_RNA_ACC_224", "MSSM_RNA_ACC_240", 
-                "MSSM_RNA_ACC_285", "MSSM_RNA_ACC_344", "MSSM_RNA_ACC_376", "MSSM_RNA_ACC_374", "PENN_RNA_ACC_003", 
-                "PITT_RNA_ACC_539", "MSSM_RNA_PFC_201", "MSSM_RNA_PFC_234", "MSSM_RNA_PFC_355", "MSSM_RNA_PFC_389",
-                "MSSM_RNA_PFC_80", "MSSM_RNA_PFC_83", "PENN_RNA_PFC_22", "PENN_RNA_PFC_59", "PENN_RNA_PFC_62", 
-                "PENN_RNA_PFC_80", "PENN_RNA_PFC_105", "PITT_RNA_PFC_1472", "PITT_RNA_PFC_566", "PITT_RNA_PFC_727", 
-                "PITT_RNA_PFC_802")
 
@@
 
-### Identify gene outliers
-Set outlier expression of genes in each samples to NA
-```{r detect.outlier.genes}
+### Winsorise counts (Gene outliers in samples)
+```{r winsorise.data}
 LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)
 
-# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
-log.mat = apply(LOG.CPM, 1, function(x){
+# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
+LOG.CPM = apply(LOG.CPM, 1, function(x){
   mn = mean(x, na.rm = T)
   std.dev = sd(x, na.rm = T)
-  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
+
+  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
+  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
+
+  x[x < (mn-3*std.dev)] = x[ind.low]
+  x[x > (mn+3*std.dev)] = x[ind.high]
+  return(x)
 }) %>% t
-PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = NA
-
-# Distribution of fraction NA genes in samples
-tmp.frac = apply(log.mat, 2, sum)/dim(LOG.CPM)[1]
-writeLines('Distribution of fraction NAs in samples')
-quantile(tmp.frac, seq(0, 1, 0.1)) %>%
-  round(3) 
-top.nsamples = names(tmp.frac)[tmp.frac >= 0.02]
-```
-Only `r length(top.nsamples)` samples have more than 2% of NA genes. They are: `r paste(top.nsamples, collapse = ', ')`
+
+# Back calculate counts for further analysis
+LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
+NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
+```
 
@@
 ```{r cqn}
-# Function to convert NA values to zero
-changeNAtoZero <- function(mtrx){
-  mtrx[is.na(mtrx)] = 0
-  return(mtrx)
-}
-
 # Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
+CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                           x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
@@
 CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-CQN.GENE_EXPRESSION$E[is.na(PROCESSED_COUNTS$filteredExprMatrix$counts)] = NA
+
+# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
+CQN.GENE_EXPRESSION$E = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
+  mn = mean(x, na.rm = T)
+  std.dev = sd(x, na.rm = T)
+
+  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
+  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
+
+  x[x < (mn-3*std.dev)] = x[ind.low]
+  x[x > (mn+3*std.dev)] = x[ind.high]
+  return(x)
+}) %>% t
+
+# Back calculate counts for further analysis
+LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
+NEW.COUNTS = (2^CQN.GENE_EXPRESSION$E) * t(replicate(dim(CQN.GENE_EXPRESSION$E)[1], LIB.SIZE))/1e6
 ```
@@
 ```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(CQN.GENE_EXPRESSION$E %>% changeNAtoZero))
+cr = cor(t(CQN.GENE_EXPRESSION$E))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(CQN.GENE_EXPRESSION$E %>% changeNAtoZero, scale.=T, center = T)
+PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)
 
@@
   tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
-plotdata$label = plotdata$SampleID
-plotdata$label[!(plotdata$label %in% top.nasamples)] = ''
 
@@
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
-p <- p + geom_text(aes(label= label), size=4, hjust=0)
+# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
 p
@@
 
-tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E %>% changeNAtoZero)))
+tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 ```{r preAdjusted.covariates, cache=TRUE}
-preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E %>% changeNAtoZero, 
+# Find correlation between PC's of gene expression with covariates
+preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E, 
                                                  COVARIATES,
@@
 # Estimate voom weights
-VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
-                            design = DESIGN)
+VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design = DESIGN)
 VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
@@
 # Re-calculate voom weights with correlation of random effects
-VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero,   
+VOOM.GENE_EXPRESSION = voom(NEW.COUNTS,   
                             design = DESIGN, plot=F,
@@
 ```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.GENE_EXPRESSION %>% changeNAtoZero))
+cr = cor(t(RESIDUAL.GENE_EXPRESSION))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION %>% changeNAtoZero, scale.=T, center = T)
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION %>% changeNAtoZero)))
+tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 
-### Differential expression analysis
+### Differential expression analysis (SCZ - Control)
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=10, fig.width=15}
+```{r diffExp, fig.height=15, fig.width=20}
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
-                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
+contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
                                      "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
-                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
+                                     "Dx.TissueOther_ACC-Dx.TissueControl_ACC",
+
                                      "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                      "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC",
+                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
+
                                      "Dx.TissueSCZ_DLPFC-Dx.TissueSCZ_ACC",
                                      "Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",
-                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueControl_ACC",
-                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_ACC"),
+                                     "Dx.TissueOther_DLPFC-Dx.TissueOther_ACC",
+
+                                     "Dx.TissueSCZ_DLPFC+Dx.TissueSCZ_ACC-Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",         
+                                     "Dx.TissueSCZ_DLPFC+Dx.TissueControl_DLPFC-Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC-Dx.TissueSCZ_ACC+Dx.TissueControl_ACC"),
                          levels = colnames(ADJUSTED.FIT$coefficients))
@@
 names(DE) = gsub('Dx.Tissue', '', colnames(contrast))
-names(DE)[9:10] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC')
+names(DE)[10:12] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC', 'ALL_ALL-ALL_ALL')
 
@@
 p = DE %>%
-  dplyr::filter(ref.Dx != 'Other', to.Dx != 'Other', ref.Dx != 'ALL', to.Dx != 'ALL') %>%
+  dplyr::filter(ref.Dx != 'ALL', to.Dx != 'ALL') %>%
   ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
 p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
-p = p + facet_grid(ref.Dx+.~to.Dx+to.Tissue, scales = 'fixed')
+p = p + facet_grid(ref.Dx+ref.Tissue~.+to.Dx+to.Tissue, scales = 'fixed')
 p = p + ggtitle('Differential Expression between Schizophrenia and Control')
@@
 
+### Differential expression analysis (Gender specific SCZ - Control)
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diffExp1, fig.height=15, fig.width=20}
+# Covariates to adjust
+COVARIATES$Dx.Tissue.Gender = factor(paste(COVARIATES$Dx.Tissue, COVARIATES$Gender, sep = '_'))
+postAdjustCovars = c('Dx.Tissue.Gender', 'Institution', 'RIN', 'IntronicRate')
+
+writeLines(paste('Using following covariates in the model:',
+                 paste(postAdjustCovars, collapse=', '),
+                 'as fixed effects and Individual_ID is chosen as random effect'))
+
+# Post adjusted design matrix
+DESIGN1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
+DESIGN1 = DESIGN1[,linColumnFinder(DESIGN1)$indepCols]
+
+# Estimate voom weights
+VOOM.GENE_EXPRESSION1 = voom(NEW.COUNTS, design = DESIGN1)
+VOOM.GENE_EXPRESSION1$E = CQN.GENE_EXPRESSION$E
+correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION1, block = COVARIATES$Individual_ID)
+
+# Re-calculate voom weights with correlation of random effects
+VOOM.GENE_EXPRESSION1 = voom(NEW.COUNTS,   
+                             design = DESIGN1, plot=F,
+                             block = COVARIATES$Individual_ID, 
+                             correlation = correlation$cor)
+VOOM.GENE_EXPRESSION1$E = CQN.GENE_EXPRESSION$E
+
+# Fit linear model using new weights and new design
+ADJUSTED.FIT1 = lmFit(VOOM.GENE_EXPRESSION1)
+
+# Fit contrast
+contrast = makeContrasts(contrasts=c("Dx.Tissue.GenderSCZ_ACC_Male-Dx.Tissue.GenderControl_ACC_Male",
+                                     "Dx.Tissue.GenderSCZ_ACC_Female-Dx.Tissue.GenderControl_ACC_Female",
+
+                                     "Dx.Tissue.GenderSCZ_DLPFC_Male-Dx.Tissue.GenderControl_DLPFC_Male",
+                                     "Dx.Tissue.GenderSCZ_DLPFC_Female-Dx.Tissue.GenderControl_DLPFC_Female"),
+                         levels = colnames(ADJUSTED.FIT1$coefficients))
+FIT.CONTR1 = contrasts.fit(ADJUSTED.FIT1, contrasts=contrast)
+FIT.CONTR1 = eBayes(FIT.CONTR1)
+
+# Get differnetial expression
+DE1 = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
+    rownameToFirstColumn('ensembl_gene_id') 
+}, FIT.CONTR1) 
+names(DE1) = gsub('Dx.Tissue.Gender', '', colnames(contrast))
+
+DE1 = DE1 %>% 
+  data.table::rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Comparison = gsub('Dx.Tissue.Gender','',Comparison),
+                Direction = sign(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  tidyr::separate(Comparison, into = c('to.state','ref.state'), sep = '-') %>%
+  tidyr::separate(to.state, into = c('to.Dx', 'to.Tissue', 'to.Gender'), sep = '_') %>%
+  tidyr::separate(ref.state, into = c('ref.Dx', 'ref.Tissue', 'ref.Gender'), sep = '_') %>%
+  left_join(Ensemble2HGNC) %>%
+  left_join(GENE.LEN)
+DE1$Direction[DE1$adj.P.Val > 0.05 | abs(DE1$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE1 %>%
+  dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender, logFC) %>%
+  group_by(to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender) %>%
+  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))
+
+tmp1 = DE1 %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
+  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender, logFC) %>%
+  group_by(to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))
+
+kable(full_join(tmp,tmp1))
+
+p = DE1 %>%
+  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
+p = p + facet_grid(ref.Tissue~.+ref.Gender, scales = 'fixed')
+p = p + ggtitle('Differential Expression between Schizophrenia and Control')
+p
+
+table(COVARIATES$Dx.Tissue, COVARIATES$Gender)
+```
+
+### Differential expression analysis (ACC - DLPFC, SCZ - Control, Female - Male)
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diffExp2, fig.height=16, fig.width=8}
+COVARIATES = tidyr::separate(COVARIATES, Dx.Tissue, c('Dx', 'Tissue')) %>%
+  dplyr::mutate(Dx = factor(Dx, levels = c('SCZ', 'Control', 'Other')),
+                Tissue = factor(Tissue))
+postAdjustCovars = c('Tissue', 'Dx', 'Institution', 'Gender', 'RIN', 'IntronicRate')
+
+# Post adjusted design matrix
+DESIGN2 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
+DESIGN2 = DESIGN2[,linColumnFinder(DESIGN2)$indepCols]
+
+# Estimate voom weights
+VOOM.GENE_EXPRESSION2 = voom(NEW.COUNTS, design = DESIGN2)
+VOOM.GENE_EXPRESSION2$E = CQN.GENE_EXPRESSION$E
+correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION2, block = COVARIATES$Individual_ID)
+
+# Re-calculate voom weights with correlation of random effects
+VOOM.GENE_EXPRESSION2 = voom(NEW.COUNTS,   
+                             design = DESIGN2, plot=F,
+                             block = COVARIATES$Individual_ID, 
+                             correlation = correlation$cor)
+VOOM.GENE_EXPRESSION2$E = CQN.GENE_EXPRESSION$E
+
+# Fit linear model using new weights and new design
+ADJUSTED.FIT2 = lmFit(VOOM.GENE_EXPRESSION2)
+
+# Fit contrast
+contrast = makeContrasts(contrasts=c("TissueACC-TissueDLPFC",
+                                     "DxSCZ-DxControl",
+                                     "GenderFemale"),
+                         levels = colnames(ADJUSTED.FIT2$coefficients))
+FIT.CONTR2 = contrasts.fit(ADJUSTED.FIT2, contrasts=contrast)
+FIT.CONTR2 = eBayes(FIT.CONTR2)
+
+# Get differnetial expression
+DE2 = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
+    rownameToFirstColumn('ensembl_gene_id') 
+}, FIT.CONTR2) 
+names(DE2) = c('ACC-DLPFC', 'SCZ-CONTROL', 'FEMALE-MALE')
+
+DE2 = DE2 %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Direction = sign(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  left_join(Ensemble2HGNC) 
+DE2$Direction[DE2$adj.P.Val > 0.05 | abs(DE2$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE2 %>%
+  dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
+  group_by(Comparison) %>%
+  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))
+
+tmp1 = DE2 %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
+  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
+  group_by(Comparison) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))
+
+kable(full_join(tmp,tmp1))
+
+p = DE2 %>%
+  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
+p = p + facet_grid(Comparison~., scales = 'free_y')
+p = p + ggtitle('Differential Expression')
+p
+```
+
 ### Gene set enrichment analysis
@@
 ```{r enrich.de, fig.height=16, fig.width=16}
-diffexp.genes = DE %>%
-  dlply(.(ref.Dx, ref.Tissue, to.Dx, to.Tissue), .fun = function(x){
-    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
+all.diff.exp = list(DE %>%
+                      tidyr::unite(Comparison1, to.Dx, to.Tissue) %>%
+                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue) %>%
+                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
+                    DE1%>%
+                      tidyr::unite(Comparison1, to.Dx, to.Tissue, to.Gender) %>%
+                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue, ref.Gender) %>%
+                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
+                    DE2) %>%
+  data.table::rbindlist(fill = TRUE, use.names = TRUE)
+diffexp.genes.up = all.diff.exp %>%
+  dlply(.(Comparison), .fun = function(x){
+    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC > 0] %>% unique
     tmp = tmp[!is.na(tmp)]
   })
+names(diffexp.genes.up) = paste(names(diffexp.genes.up), 'UP', sep = '.')
+
+diffexp.genes.down = all.diff.exp %>%
+  dlply(.(Comparison), .fun = function(x){
+    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC < 0] %>% unique
+    tmp = tmp[!is.na(tmp)]
+  })
+names(diffexp.genes.down) = paste(names(diffexp.genes.down), 'DOWN', sep = '.')
+
+diffexp.genes = c(diffexp.genes.up, diffexp.genes.down)
 diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
@@
 # Store differential expression results
-write.table(DE, file = 'CMC_ACC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+write.table(all.diff.exp, file = 'CMC_ACC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
 DEXP_OBJ = File('CMC_ACC_DLPFC_DiffExpression.tsv', 

### 2017-07-28 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor modifications

@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = 'syn9872607';
+parentId = "syn7501880";
 activityName = 'Covariate adjustments';
@@
   dplyr::mutate(label = SampleID)
-plotdata$label[!(plotdata$SampleID %in% top.nsamples)] = ''
+# plotdata$label[!(plotdata$SampleID %in% top.nsamples)] = ''
 
@@
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
-p <- p + geom_text(aes(label= label), size=4, hjust=0)
+# p <- p + geom_text(aes(label= label), size=4, hjust=0)
 p

### 2017-07-28 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Minor bug fix

@@
 # Get differnetial expression
-DE = lapply(1:6, function(i, FIT){
-  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
+  limma::topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
     rownameToFirstColumn('ensembl_gene_id') 
@@
   tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
-  tidyr::separate(ref.state, into = c('Dx1','Tissue'), sep = '_') %>%
-  tidyr::separate(to.state, into = c('Dx2','Tissue'), sep = '_') %>%
+  tidyr::separate(ref.state, into = c('Dx1','Tissue1'), sep = '_') %>%
+  tidyr::separate(to.state, into = c('Dx2','Tissue2'), sep = '_') %>%
   tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_') %>%
@@
 diffexp.genes = DE %>%
-  dlply(.(Comparison, Tissue), .fun = function(x){
+  dlply(.(Comparison, Tissue1), .fun = function(x){
     tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique

### 2017-07-28 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Minor bug fix

@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueBP_DLPFC-Dx.TissueControl_DLPFC",
+contrast = makeContrasts(contrasts=c("Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
                                      "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
-                                     "Dx.TissueSCZ_DLPFC-Dx.TissueBP_DLPFC"),
+                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC",
+                                     "Dx.TissueOther_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC"),
                          levels = colnames(ADJUSTED.FIT$coefficients))
@@
 # Get differnetial expression
-DE = lapply(1:3, function(i, FIT){
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
   topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%

### 2017-07-28 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Minor changes

--- 2017-05-19 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN and SVA normlaisation)"
+title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)"
 author: "Thanneer Perumal"
@@
 knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd",
-                                 parentId ="syn7501880",
+                                 parentId = "syn9872607",
                                  entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - SVA')
@@
 library(cqn)
+library(sva)
 
@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = 'syn7501880';
+parentId = "syn9872607";
 activityName = 'Covariate adjustments';
-activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN and SVA normalisation (ACC and DLPFC)';
+activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (ACC and DLPFC)';
 
@@
 
-### Identify gene outliers
-Set outlier expression of genes in each samples to NA
+### Winsorise counts (Gene outliers in samples)
 ```{r winsorise.data}
@@
 
-# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
-log.mat = apply(LOG.CPM, 1, function(x){
+# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
+LOG.CPM = apply(LOG.CPM, 1, function(x){
   mn = mean(x, na.rm = T)
   std.dev = sd(x, na.rm = T)
-  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
+
+  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
+  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
+
+  x[x < (mn-3*std.dev)] = x[ind.low]
+  x[x > (mn+3*std.dev)] = x[ind.high]
+  return(x)
 }) %>% t
-PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = NA
-
-# Distribution of fraction NA genes in samples
-tmp.frac = apply(log.mat, 2, sum)/dim(LOG.CPM)[1]
-writeLines('Distribution of fraction NAs in samples')
-quantile(tmp.frac, seq(0, 1, 0.1)) %>%
-  round(3) 
-top.nsamples = names(tmp.frac)[tmp.frac >= 0.02]
-```
-Only `r length(top.nsamples)` samples have more than 2% of NA genes. They are: `r paste(top.nsamples, collapse = ', ')`
+
+# Back calculate counts for further analysis
+LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
+NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
+```
 
@@
 ```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.SVA.GENE_EXPRESSION %>% changeNAtoZero))
+cr = cor(t(RESIDUAL.SVA.GENE_EXPRESSION))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.SVA.GENE_EXPRESSION %>% changeNAtoZero, scale.=T, center = T)
+PC <- prcomp(RESIDUAL.SVA.GENE_EXPRESSION, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(RESIDUAL.SVA.GENE_EXPRESSION %>% changeNAtoZero)))
+tree = hclust(as.dist(t(RESIDUAL.SVA.GENE_EXPRESSION)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 ```
-### Differential expression analysis
+
+### Differential expression analysis (SCZ - Control)
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=10, fig.width=15}
+```{r diffExp, fig.height=15, fig.width=20}
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
-                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
+contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
                                      "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
-                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
+                                     "Dx.TissueOther_ACC-Dx.TissueControl_ACC",
+
                                      "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                      "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC",
-                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueControl_ACC",
-                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_ACC"),
-                         levels = colnames(VOOM.SVA.FIT$coefficients))
-FIT.CONTR = contrasts.fit(VOOM.SVA.FIT, contrasts=contrast)
+                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
+
+                                     "Dx.TissueSCZ_DLPFC-Dx.TissueSCZ_ACC",
+                                     "Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",
+                                     "Dx.TissueOther_DLPFC-Dx.TissueOther_ACC",
+
+                                     "Dx.TissueSCZ_DLPFC+Dx.TissueSCZ_ACC-Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",         
+                                     "Dx.TissueSCZ_DLPFC+Dx.TissueControl_DLPFC-Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC-Dx.TissueSCZ_ACC+Dx.TissueControl_ACC"),
+                         levels = colnames(ADJUSTED.FIT$coefficients))
+FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
 FIT.CONTR = eBayes(FIT.CONTR)
@@
 names(DE) = gsub('Dx.Tissue', '', colnames(contrast))
-names(DE)[7:8] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC')
+names(DE)[10:12] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC', 'ALL_ALL-ALL_ALL')
 
@@
 ```{r enrich.de, fig.height=16, fig.width=16}
-diffexp.genes = DE %>%
-  dlply(.(ref.Dx, ref.Tissue, to.Dx, to.Tissue), .fun = function(x){
-    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
+all.diff.exp = list(DE %>%
+                      tidyr::unite(Comparison1, to.Dx, to.Tissue) %>%
+                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue) %>%
+                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
+                    DE1%>%
+                      tidyr::unite(Comparison1, to.Dx, to.Tissue, to.Gender) %>%
+                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue, ref.Gender) %>%
+                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
+                    DE2) %>%
+  data.table::rbindlist(fill = TRUE, use.names = TRUE)
+diffexp.genes.up = all.diff.exp %>%
+  dlply(.(Comparison), .fun = function(x){
+    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC > 0] %>% unique
     tmp = tmp[!is.na(tmp)]
     })
+names(diffexp.genes.up) = paste(names(diffexp.genes.up), 'UP', sep = '.')
+
+diffexp.genes.down = all.diff.exp %>%
+  dlply(.(Comparison), .fun = function(x){
+    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC < 0] %>% unique
+    tmp = tmp[!is.na(tmp)]
+    })
+names(diffexp.genes.down) = paste(names(diffexp.genes.down), 'DOWN', sep = '.')
+
+diffexp.genes = c(diffexp.genes.up, diffexp.genes.down)
 diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
@@
 |  Design Matrix                              |  `r DM_OBJ$properties$id`        |
-|  Surrogate Variables                        |  `r DM1_OBJ$properties$id`       |
-|  Residual Expression (for eQTL analysis)    |  `r eEXP_OBJ$properties$id`      |
+|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
 |  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
 |  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |
+
+### Source Code
+[R markdown](`r thisFile`)

### 2017-07-28 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Minor changes

@@
 knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd",
-                                 parentId = "syn9872607",
+                                 parentId = "syn7501880",
                                  entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - SVA')
@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = "syn9872607";
+parentId = "syn7501880";
 activityName = 'Covariate adjustments';
@@
 # Compute actual variance of all principal components
-tmp = svd(RESIDUAL.GENE_EXPRESSION %>% changeNAtoZero)
+tmp = svd(RESIDUAL.GENE_EXPRESSION)
 actual.var = tmp$d^2/sum(tmp$d^2)
@@
 permuted.var = foreach(i = 1:20, .combine = rbind, .packages = c('dplyr')) %dopar% {
-  tmp.residual = t(apply(RESIDUAL.GENE_EXPRESSION %>% changeNAtoZero, 1, sample, replace = FALSE))
+  tmp.residual = t(apply(RESIDUAL.GENE_EXPRESSION, 1, sample, replace = FALSE))
   tmp = svd(tmp.residual)
@@
 # Estimate surrogate variables
-SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E %>% changeNAtoZero, MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
+SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E, MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
 colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
@@
 # Estimate voom weights for dispersion control
-VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
+VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
                         design=cbind(MODEL1, SURR.VAR), plot=F)
@@
 # Re-estimate voom weights
-VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
+VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
                         design = cbind(MODEL1, SURR.VAR), plot=F,

### 2017-07-29 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Minor modifications

@@
                                  parentId ="syn9846623",
-                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - Sparse Model')
+                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - Dense Model')
 ```
@@
                        worse = sum(diff <= -5)) %>%
-      dplyr::mutate(nchange = better - worse)
+      dplyr::mutate(nchange = better)
 
@@
 # Code
-CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - Sparse Model", parentId = parentId)
+CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - Dense Model", parentId = parentId)
 annotations(CODE) = all.annotations

### 2017-07-29 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd

@@
 knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
-                                 parentId = "syn7501880",
+                                 parentId = "syn9872607",
                                  entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = "syn7501880";
+parentId = "syn9872607";
 activityName = 'Covariate adjustments';

### 2017-07-29 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_MSSM.Rmd

--- 2017-07-28 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
+knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_MSSM.Rmd",
                                  parentId = "syn7501880",
-                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
+                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - MSSM')
 ```
@@
 
-thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd'
+thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_MSSM.Rmd'
 
@@
   left_join(ANCESTRY) %>%
+  dplyr::filter(Institution == 'MSSM') %>%
   unique
@@
 ```{r covariates.clustering}
-FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
+FactorCovariates <- c("Individual_ID", "Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
 ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
@@
 ### Normalisation
-Based on different model selection methodology (as seen here syn8468116, syn9772932, and syn9774491), following variables are used as covariates: Institution, Gender, RIN, and IntronicRate
+Based on different model selection methodology (as seen here syn8468116, syn9772932, and syn9774491), following variables are used as covariates: Gender, RIN, and IntronicRate
 
@@
 # Primary variable of interest
-postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'RIN', 'IntronicRate')
+postAdjustCovars = c('Dx.Tissue', 'Gender', 'RIN', 'IntronicRate')
 
@@
 
-  normalizationType	= 'CQN'
+  normalizationType	= 'CQN',
+  institution = 'MSSM'
 )
@@
 # Code
-CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
+CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - MSSM", parentId = parentId)
 annotations(CODE) = all.annotations

### 2017-07-29 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_PENN.Rmd
Added markdown to process penn samples alone

--- 2017-07-29 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_MSSM.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_MSSM.Rmd",
+knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_PENN.Rmd",
                                  parentId = "syn7501880",
-                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - MSSM')
+                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - PENN')
 ```
@@
   left_join(ANCESTRY) %>%
-  dplyr::filter(Institution == 'MSSM') %>%
+  dplyr::filter(Institution == 'Penn') %>%
   unique
@@
 ```{r covariates.clustering}
-FactorCovariates <- c("Individual_ID", "Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
+FactorCovariates <- c("Individual_ID", "Institution", "Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
 ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
@@
 preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E, 
-                                                 COVARIATES,
+                                                 COVARIATES[,setdiff(colnames(COVARIATES), c('Institution'))],
                                                  'NULL design(voom-normalized)', 
@@
   normalizationType	= 'CQN',
-  institution = 'MSSM'
+  institution = 'PENN'
 )
@@
 # Code
-CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - MSSM", parentId = parentId)
+CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - PENN", parentId = parentId)
 annotations(CODE) = all.annotations

### 2017-07-31 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_PENN.Rmd
Minor bug fix

@@
 ```{r covariates.correlation, fig.width=10, fig.height=8}
-COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
+COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES[,setdiff(colnames(COVARIATES), 'Institution')], PVAL = 0.05)
 tmp = COVARIATES.CORRELATION$ESTIMATE
@@
 contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
-                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
-                                     "Dx.TissueOther_ACC-Dx.TissueControl_ACC",
 
                                      "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
-                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC",
-                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
 
@@
                                      "Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",
-                                     "Dx.TissueOther_DLPFC-Dx.TissueOther_ACC",
 
@@
 names(DE) = gsub('Dx.Tissue', '', colnames(contrast))
-names(DE)[10:12] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC', 'ALL_ALL-ALL_ALL')
+names(DE)[5:7] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC', 'ALL_ALL-ALL_ALL')
 
@@
                 Tissue = factor(Tissue))
-postAdjustCovars = c('Tissue', 'Dx', 'Institution', 'Gender', 'RIN', 'IntronicRate')
+postAdjustCovars = c('Tissue', 'Dx', 'Gender', 'RIN', 'IntronicRate')
 
 # Post adjusted design matrix
-DESIGN2 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
+DESIGN2 = getDesignMatrix(droplevels(COVARIATES[,postAdjustCovars,drop=F]),Intercept = T)$design
 DESIGN2 = DESIGN2[,linColumnFinder(DESIGN2)$indepCols]
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("TissueACC-TissueDLPFC",
-                                     "DxSCZ-DxControl",
-                                     "GenderFemale"),
+contrast = makeContrasts(contrasts=c("TissueDLPFC",
+                                     "DxControl",
+                                     "GenderMale"),
                          levels = colnames(ADJUSTED.FIT2$coefficients))

### 2017-07-31 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_PITT.Rmd
Added markdown to process samples from pitt

--- 2017-07-31 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_PENN.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_PENN.Rmd",
+knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_PITT.Rmd",
                                  parentId = "syn7501880",
-                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - PENN')
+                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - PITT')
 ```
@@
 
-thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_MSSM.Rmd'
+thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_PITT.Rmd'
 
@@
   left_join(ANCESTRY) %>%
-  dplyr::filter(Institution == 'Penn') %>%
+  dplyr::filter(Institution == 'Pitt', Dx %in% c('Control','SCZ')) %>%
   unique
@@
 # Replace AFF and BP samples as other
-levels(METADATA$Dx) = c("Other", "Other", "Control", "SCZ")
+levels(METADATA$Dx) = c("Control", "SCZ")
 ```
@@
   normalizationType	= 'CQN',
-  institution = 'PENN'
+  institution = 'PITT'
 )
@@
 # Code
-CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - PENN", parentId = parentId)
+CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - PITT", parentId = parentId)
 annotations(CODE) = all.annotations

### 2017-08-06 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Included batch effects in HBCC model selection

--- 2017-07-23 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
@@
 #   plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'GenotypingSampleID'))
+
+# Get RNA qc metrics from synapse
+METADATA_ALG_REP_DLPFC_ID = 'syn2279446'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ALG_REP_DLPFC_ID
+METADATA_ALG_REP_DLPFC = read.table(synGet(METADATA_ALG_REP_DLPFC_ID)@filePath, 
+                                    fill = T, header = T, sep = ',') %>%
+  dplyr::select(Sample.RNA.ID, Ribozero.Batch, Library.Batch, Flowcell.Batch) %>%
+  dplyr::rename(SampleID = Sample.RNA.ID,
+                LibraryBatch = Library.Batch, 
+                RibozeroBatch = Ribozero.Batch, 
+                FlowcellBatch = Flowcell.Batch) %>%
+  unique()
 
@@
 
-                          "DLPFC_RNA_report: Library Batch", "DLPFC_RNA_report: Ribozero Batch", "DLPFC_RNA_report: Flowcell Batch",
-
                           "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
@@
   dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
-                DLPFC_RNA_report__Library_Batch, DLPFC_RNA_report__Ribozero_Batch, DLPFC_RNA_report__Flowcell_Batch,
                 DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, 
-                DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Ribozero_Batch, 
-                DLPFC_RNA_report__Library_Batch, DLPFC_RNA_report__Flowcell_Batch) %>%
-  dplyr::rename(LibraryBatch = DLPFC_RNA_report__Library_Batch, 
-                RibozeroBatch = DLPFC_RNA_report__Ribozero_Batch, 
-                FlowcellBatch = DLPFC_RNA_report__Flowcell_Batch,
-                PMI = PMI__in_hours_, 
+                DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_) %>%
+  dplyr::rename(PMI = PMI__in_hours_, 
                 IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
@@
                 RIN = DLPFC_RNA_isolation__RIN, 
-                ReportExclude = DLPFC_RNA_report__Exclude_,
-                RibozeroBatch = DLPFC_RNA_report__Ribozero_Batch, 
-                LibraryBatch = DLPFC_RNA_report__Library_Batch, 
-                FlowcellBatch = DLPFC_RNA_report__Flowcell_Batch) %>%
+                ReportExclude = DLPFC_RNA_report__Exclude_) %>%
   dplyr::filter(SampleID %in% colnames(COUNT)) %>%
   dplyr::mutate(Tissue = 'DLPFC') %>%
-  list(.,METADATA_QC) %>%
+  list(.,METADATA_QC, METADATA_ALG_REP_DLPFC) %>%
   join_all %>% droplevels() %>%
@@
 METADATA$SampleID = as.character(METADATA$SampleID)
+
+# Fix missing values
+levels(METADATA$LibraryBatch)[1] = 'NotAvailable'
+levels(METADATA$FlowcellBatch)[1] = 'NotAvailable'
+levels(METADATA$RibozeroBatch)[1] = 'NotAvailable'
 ```
@@
 ```{r covariates.clustering}
-FactorCovariates <- c("Gender", "Dx.Tissue")
+FactorCovariates <- c("Gender", "Dx.Tissue", "LibraryBatch", "RibozeroBatch", "FlowcellBatch")
 ContCovariates <- c("Age_of_Death", "PMI", "RIN", "TotalReads", "X5prime_Norm", "Mean_Per_Base_Cov.",
@@
                           x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
-                          lengths = as.numeric(GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length']),
+                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                           lengthMethod = "smooth", 
@@
 
-# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
-log.mat = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
+# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
+CQN.GENE_EXPRESSION$E = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
   mn = mean(x, na.rm = T)
   std.dev = sd(x, na.rm = T)
-  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
+
+  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
+  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
+
+  x[x < (mn-3*std.dev)] = x[ind.low]
+  x[x > (mn+3*std.dev)] = x[ind.high]
+  return(x)
 }) %>% t
-NEW.COUNTS[log.mat] = NA
-CQN.GENE_EXPRESSION$E[log.mat] = NA
+
+# Back calculate counts for further analysis
+LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
+NEW.COUNTS = (2^CQN.GENE_EXPRESSION$E) * t(replicate(dim(CQN.GENE_EXPRESSION$E)[1], LIB.SIZE))/1e6
 ```
@@
 Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`
-
 #### Phase II (batch effects)
-```{r phase2, eval = F}
+```{r phase2}
 # Fit a base model with the following covariates
@@
 Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`
-
 #### Phase III (RNASeq alignment specific covariates)
@@
 
-### Differential expression analysis
+### Differential expression analysis (SCZ - Control)
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=10, fig.width=15}
+```{r diffExp, fig.height=15, fig.width=20}
+# Covariates to adjust
+postAdjustCovars = adjust.covars
+
+writeLines(paste('Using following covariates in the model:',
+                 paste(postAdjustCovars, collapse=', ')))
+
+# Post adjusted design matrix
+DESIGN = getDesignMatrix(droplevels(COVARIATES[,postAdjustCovars,drop=F]),Intercept = F)$design
+DESIGN = DESIGN[,linColumnFinder(DESIGN)$indepCols]
+
+# Estimate voom weights
+VOOM.GENE_EXPRESSION = limma::voom(NEW.COUNTS, design = DESIGN)
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
+
+# Fit linear model using new weights and new design
+ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
+
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueBP_DLPFC-Dx.TissueControl_DLPFC",
-                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
+contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
+                                     "Dx.TissueBP_DLPFC-Dx.TissueControl_DLPFC",
                                      "Dx.TissueSCZ_DLPFC-Dx.TissueBP_DLPFC"),
@@
 }, FIT.CONTR) 
-names(DE) = colnames(contrast)
+names(DE) = gsub('Dx.Tissue', '', colnames(contrast))
 
 DE = DE %>% 
-  data.table::rbindlist(idcol = 'Comparison') %>%
+  rbindlist(idcol = 'Comparison') %>%
   dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
-                Direction = logFC/abs(logFC),
+                Direction = sign(logFC),
                 Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                 Direction = as.character(Direction)) %>%
-  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
-  tidyr::separate(ref.state, into = c('Dx1','Tissue1'), sep = '_') %>%
-  tidyr::separate(to.state, into = c('Dx2','Tissue2'), sep = '_') %>%
+  tidyr::separate(Comparison, into = c('to.state','ref.state'), sep = '-') %>%
+  tidyr::separate(to.state, into = c('to.Dx','to.Tissue'), sep = '_') %>%
+  tidyr::separate(ref.state, into = c('ref.Dx','ref.Tissue'), sep = '_') %>%
   left_join(Ensemble2HGNC) %>%
-  left_join(GENE.LEN) %>%
-  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_')
-
+  left_join(GENE.LEN %>% dplyr::mutate(Length = as.numeric(Length)))
 DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
@@
 tmp = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue1, logFC) %>%
-  group_by(Comparison, Tissue1) %>%
-  dplyr::summarise(FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])),
-                   FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])))
+  dplyr::filter(adj.P.Val <= 0.05, ref.Dx != 'Other', to.Dx != 'Other') %>%
+  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue, logFC) %>%
+  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
+  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))
 
 tmp1 = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue1, logFC) %>%
-  group_by(Comparison, Tissue1) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > log2(1.2)])),
-                   FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < -log2(1.2)])))
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), ref.Dx != 'Other', to.Dx != 'Other') %>%
+  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue, logFC) %>%
+  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))
 
@@
 
-p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
-p = p + scale_color_manual(values = c('green','grey','red'))
-p = p + facet_grid(Tissue1+.~Comparison, scales = 'fixed')
+p = DE %>%
+  dplyr::filter(ref.Dx != 'ALL', to.Dx != 'ALL') %>%
+  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
+p = p + facet_grid(ref.Dx+ref.Tissue~.+to.Dx+to.Tissue, scales = 'fixed')
+p = p + ggtitle('Differential Expression between Schizophrenia and Control')
 p
@@
 pl = list()
-pl[[1]] = ggplot(DE, aes(x = log10(as.numeric(Length)), y = logFC, color = Direction)) + geom_point() 
-pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(as.numeric(Length)), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[1]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
+                 aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
+pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
+pl[[1]] = pl[[1]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
 pl[[1]] = pl[[1]] + theme(legend.position = 'top')
 
-pl[[2]] = ggplot(DE, aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
+pl[[2]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
+                 aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
 pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[2]] = pl[[2]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
 pl[[2]] = pl[[2]] + theme(legend.position = 'top')
 
-pl[[3]] = ggplot(DE, aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
+pl[[3]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
+                 aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
 pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[3]] = pl[[3]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
 pl[[3]] = pl[[3]] + theme(legend.position = 'top')
@@
 multiplot(plotlist = pl, cols = 3)
+```
+
+### Differential expression analysis (Gender specific SCZ - Control)
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diffExp1, fig.height=15, fig.width=20}
+# Covariates to adjust
+COVARIATES$Dx.Tissue.Gender = factor(paste(COVARIATES$Dx.Tissue, COVARIATES$Gender, sep = '_'))
+postAdjustCovars = c('Dx.Tissue.Gender', 'RIN', 'IntronicRate')
+
+writeLines(paste('Using following covariates in the model:',
+                 paste(postAdjustCovars, collapse=', ')))
+
+# Post adjusted design matrix
+DESIGN1 = getDesignMatrix(droplevels(COVARIATES[,postAdjustCovars,drop=F]),Intercept = F)$design
+DESIGN1 = DESIGN1[,linColumnFinder(DESIGN1)$indepCols]
+
+# Estimate voom weights
+VOOM.GENE_EXPRESSION1 = limma::voom(NEW.COUNTS, design = DESIGN1)
+VOOM.GENE_EXPRESSION1$E = CQN.GENE_EXPRESSION$E
+
+# Fit linear model using new weights and new design
+ADJUSTED.FIT1 = lmFit(VOOM.GENE_EXPRESSION1)
+
+# Fit contrast
+contrast = makeContrasts(contrasts=c("Dx.Tissue.GenderSCZ_DLPFC_Male-Dx.Tissue.GenderControl_DLPFC_Male",
+                                     "Dx.Tissue.GenderSCZ_DLPFC_Female-Dx.Tissue.GenderControl_DLPFC_Female"),
+                         levels = colnames(ADJUSTED.FIT1$coefficients))
+FIT.CONTR1 = contrasts.fit(ADJUSTED.FIT1, contrasts=contrast)
+FIT.CONTR1 = eBayes(FIT.CONTR1)
+
+# Get differnetial expression
+DE1 = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
+    rownameToFirstColumn('ensembl_gene_id') 
+}, FIT.CONTR1) 
+names(DE1) = gsub('Dx.Tissue.Gender', '', colnames(contrast))
+
+DE1 = DE1 %>% 
+  data.table::rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Comparison = gsub('Dx.Tissue.Gender','',Comparison),
+                Direction = sign(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  tidyr::separate(Comparison, into = c('to.state','ref.state'), sep = '-') %>%
+  tidyr::separate(to.state, into = c('to.Dx', 'to.Tissue', 'to.Gender'), sep = '_') %>%
+  tidyr::separate(ref.state, into = c('ref.Dx', 'ref.Tissue', 'ref.Gender'), sep = '_') %>%
+  left_join(Ensemble2HGNC) %>%
+  left_join(GENE.LEN %>% dplyr::mutate(Length = as.numeric(Length)))
+DE1$Direction[DE1$adj.P.Val > 0.05 | abs(DE1$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE1 %>%
+  dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender, logFC) %>%
+  group_by(to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender) %>%
+  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))
+
+tmp1 = DE1 %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
+  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender, logFC) %>%
+  group_by(to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))
+
+kable(full_join(tmp,tmp1))
+
+p = DE1 %>%
+  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
+p = p + facet_grid(ref.Tissue~.+ref.Gender, scales = 'fixed')
+p = p + ggtitle('Differential Expression between Schizophrenia and Control')
+p
+
+table(COVARIATES$Dx.Tissue, COVARIATES$Gender)
+```
+
+### Differential expression analysis (ACC - DLPFC, SCZ - Control, Female - Male)
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diffExp2, fig.height=16, fig.width=8}
+postAdjustCovars = c('Dx.Tissue', 'Gender', 'RIN', 'IntronicRate')
+
+# Post adjusted design matrix
+DESIGN2 = getDesignMatrix(droplevels(COVARIATES[,postAdjustCovars,drop=F]),Intercept = T)$design
+DESIGN2 = DESIGN2[,linColumnFinder(DESIGN2)$indepCols]
+
+# Estimate voom weights
+VOOM.GENE_EXPRESSION2 = voom(NEW.COUNTS, design = DESIGN2)
+VOOM.GENE_EXPRESSION2$E = CQN.GENE_EXPRESSION$E
+
+# Fit linear model using new weights and new design
+ADJUSTED.FIT2 = lmFit(VOOM.GENE_EXPRESSION2)
+
+# Fit contrast
+contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
+                                     "GenderMale"),
+                         levels = colnames(ADJUSTED.FIT2$coefficients))
+FIT.CONTR2 = contrasts.fit(ADJUSTED.FIT2, contrasts=contrast)
+FIT.CONTR2 = eBayes(FIT.CONTR2)
+
+# Get differnetial expression
+DE2 = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
+    rownameToFirstColumn('ensembl_gene_id') 
+}, FIT.CONTR2) 
+names(DE2) = c('SCZ-CONTROL', 'FEMALE-MALE')
+
+DE2 = DE2 %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Direction = sign(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  left_join(Ensemble2HGNC) 
+DE2$Direction[DE2$adj.P.Val > 0.05 | abs(DE2$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE2 %>%
+  dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
+  group_by(Comparison) %>%
+  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))
+
+tmp1 = DE2 %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
+  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
+  group_by(Comparison) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))
+
+kable(full_join(tmp,tmp1))
+
+p = DE2 %>%
+  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
+p = p + facet_grid(Comparison~., scales = 'free_y')
+p = p + ggtitle('Differential Expression')
+p
 ```
@@
 ```{r enrich.de, fig.height=16, fig.width=16}
-diffexp.genes = DE %>%
-  dplyr::mutate(Direction = sign(logFC),
-                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP'))) %>%
-  dlply(.(Comparison, Tissue1, Direction), .fun = function(x){
-    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
+all.diff.exp = list(DE %>%
+                      tidyr::unite(Comparison1, to.Dx, to.Tissue) %>%
+                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue) %>%
+                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
+                    DE1%>%
+                      tidyr::unite(Comparison1, to.Dx, to.Tissue, to.Gender) %>%
+                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue, ref.Gender) %>%
+                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
+                    DE2) %>%
+  data.table::rbindlist(fill = TRUE, use.names = TRUE)
+diffexp.genes.up = all.diff.exp %>%
+  dlply(.(Comparison), .fun = function(x){
+    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC > 0] %>% unique
     tmp = tmp[!is.na(tmp)]
   })
+names(diffexp.genes.up) = paste(names(diffexp.genes.up), 'UP', sep = '.')
+
+diffexp.genes.down = all.diff.exp %>%
+  dlply(.(Comparison), .fun = function(x){
+    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC < 0] %>% unique
+    tmp = tmp[!is.na(tmp)]
+  })
+names(diffexp.genes.down) = paste(names(diffexp.genes.down), 'DOWN', sep = '.')
+
+diffexp.genes = c(diffexp.genes.up, diffexp.genes.down)
 diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
@@
 # Perform enrichment analysis
-enrichment.results = plyr::llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
+enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
    ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
@@
 # Store filtered counts
-PROCESSED_COUNTS$filteredExprMatrix$counts %>%
+NEW.COUNTS %>%
   rownameToFirstColumn('ensembl_gene_id') %>%

### 2017-08-06 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Minor bug fix

@@
 Correlation/association between covariates at an FDR <= 0.1
-```{r covariates.correlation, fig.width=10, fig.height=8}
+```{r covariates.correlation, fig.width=15, fig.height=15}
 COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=15, fig.width=20}
+```{r diffExp, fig.height=8, fig.width=8}
 # Covariates to adjust
-postAdjustCovars = adjust.covars
+postAdjustCovars = c('Dx.Tissue', 'Gender', 'RIN', 'IntronicRate')
 
@@
 ### Associate differential expression results with gc content, gene length and average expression
-```{r associate.de, fig.height=10, fig.width=15}
+```{r associate.de, fig.height= 6, fig.width = 12}
 pl = list()
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp1, fig.height=15, fig.width=20}
+```{r diffExp1, fig.height=6, fig.width=12}
 # Covariates to adjust
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp2, fig.height=16, fig.width=8}
+```{r diffExp2, fig.height=10, fig.width=6}
 postAdjustCovars = c('Dx.Tissue', 'Gender', 'RIN', 'IntronicRate')

### 2017-08-07 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_PITT.Rmd
Included batch effects in HBCC model selection
Minor fix

@@
 # Replace AFF and BP samples as other
-levels(METADATA$Dx) = c("Control", "SCZ")
+levels(METADATA$Dx) = c("Other", "Other", "Control", "SCZ")
 ```
@@
   tissueTypeAbrv	= 'DLPFC, ACC', 
-  study = 'CMC',
+  study = 'PITT',
 
@@
 
-  normalizationType	= 'CQN',
-  institution = 'PITT'
+  normalizationType	= 'CQN'
 )

### 2017-08-07 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_PENN.Rmd
Minor fix

@@
   tissueTypeAbrv	= 'DLPFC, ACC', 
-  study = 'CMC',
+  study = 'PENN',
 
@@
 
-  normalizationType	= 'CQN',
-  institution = 'PENN'
+  normalizationType	= 'CQN'
 )

### 2017-08-07 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_MSSM.Rmd
Minor fix

@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
+p <- p + geom_point(aes(shape=Tissue, size=Age_of_Death))
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
@@
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
+COVARIATES.tmp = data.matrix(COVARIATES[,c("Gender", "LibraryBatch", "Dx.Tissue")])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
@@
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue','Institution')])
+COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue'),drop=F])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
@@
                            main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue','Institution')]))
+                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue'),drop=F]))
 ```
@@
 COVARIATES$Dx.Tissue.Gender = factor(paste(COVARIATES$Dx.Tissue, COVARIATES$Gender, sep = '_'))
-postAdjustCovars = c('Dx.Tissue.Gender', 'Institution', 'RIN', 'IntronicRate')
+postAdjustCovars = c('Dx.Tissue.Gender', 'RIN', 'IntronicRate')
 
@@
   tissueTypeAbrv	= 'DLPFC, ACC', 
-  study = 'CMC',
+  study = 'MSSM',
 
@@
 
-  normalizationType	= 'CQN',
-  institution = 'MSSM'
+  normalizationType	= 'CQN'
 )

### 2017-08-07 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Added markdown to process HBCC DLPFC RNASeq data

--- 2017-08-06 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd",
-                                 parentId ="syn9846623",
-                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)')
+knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
+                                 parentId ="syn7501880",
+                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
 ```
@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = 'syn9846623';
+parentId = 'syn7501880';
 activityName = 'Covariate adjustments';
@@
 
-thisFileName <- 'HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd'
+thisFileName <- 'HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd'
 
@@
 
-### Model Identification
-```{r fxn}
-# Function to fit a mixed effect model and return BIC
-fitFixedEffectModel <- function(varsToFit, lcpm, cnt, covars){
-
-  # Get design matrix
-  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
-  designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]
-
-  # Estimate voom weights
-  cnt[is.na(cnt)] = 0
-  voomWeights = limma::voom(cnt, design = designMat, plot=F)$weights
-
-  modelBIC = plyr::ldply(1:dim(lcpm)[1],
-                         .fun = function(i, lcpm, designMat, voomWeights){
-                           # Fit a linear models for every gene
-                           data = cbind(data.frame(Gx = lcpm[i,]),
-                                        designMat, 
-                                        weight = voomWeights[i,])
-
-                           formula = paste(setdiff(colnames(data), c('Gx', 'weight')), collapse = '+') %>%
-                             paste('Gx ~',., collapse = ' ')
-
-                           mdl = stats::lm(formula = formula, 
-                                           data = data, 
-                                           weights = weight)
-
-                           data.frame(ensembl_gene_id = rownames(lcpm)[i], BIC = BIC(mdl))
-                         }, 
-                         lcpm, designMat, voomWeights, 
-                         .parallel = F,
-                         .paropts = list(.packages = c('dplyr')))
-  return(modelBIC)
-  }
-
-# Function to select covariates in a step-wise regression fashion
-selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){
-
-  varsSelected = c()
-  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
-           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
-  while(length(covarsToSelect) != 0){
-    allModelBIC = plyr::llply(1:length(covarsToSelect),
-                              .fun = function(i, covarsToSelect, adjustCovars, lcpm, cnt, covars){
-
-                                # Fit a mixed effect linear model and get all BIC
-                                modelBIC = fitFixedEffectModel(c(adjustCovars, covarsToSelect[i]),
-                                                               lcpm, cnt, covars) %>%
-                                  plyr::rename(c('BIC' = covarsToSelect[i]))
-
-                                return(modelBIC)
-                              },
-                              covarsToSelect, adjustCovars, lcpm, cnt, covars, 
-                              .parallel = F,
-                              .paropts = list(.export = c('fitFixedEffectModel')))
-
-    allModelBIC = plyr::join_all(allModelBIC)
-
-    # Find number of genes that are changed based on BIC
-    nChange = allModelBIC %>%
-      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
-      dplyr::left_join(baseModelBIC) %>%
-      dplyr::mutate(diff = baseModel - Value) %>%
-      dplyr::group_by(ModelName) %>%
-      dplyr::summarise(better = sum(diff >= 5), 
-                       worse = sum(diff <= -5)) %>%
-      dplyr::mutate(nchange = better - worse)
-
-    print(nChange)
-
-    # Select variable
-    tmp = nChange %>%
-      dplyr::filter(nchange > round(dim(cnt)[1]*0.01)) %>%
-      dplyr::top_n(1, nchange)
-
-    if (dim(tmp)[1] == 0){
-      covarsToSelect = c()
-    } else {
-      baseModelBIC = allModelBIC %>%
-        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
-      colnames(baseModelBIC)[2] = 'baseModel'
-      varsSelected = c(varsSelected, tmp$ModelName)
-      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
-      print(paste('Variable chosen to be added:',tmp$ModelName))
-    }
-  }
-
-  return(varsSelected)
-}
-```
-#### Phase I (clinical, ancestry and sample specific technical variables)
-```{r phase1}
-# Fit a base model with the following covariates
-adjust.covars = c('Dx.Tissue')
-baseModelBIC = fitFixedEffectModel(adjust.covars, 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
-                                   COVARIATES) %>%
-  dplyr::rename(baseModel = BIC)
-
-# Iteratively identify covariates in phase I
-phase1.covars = c('Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2')
-vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
-
-adjust.covars = c(adjust.covars, vars.selected)
-```
-Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`
-#### Phase II (batch effects)
-```{r phase2}
-# Fit a base model with the following covariates
-baseModelBIC = fitFixedEffectModel(adjust.covars, 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
-                                   COVARIATES) %>%
-  dplyr::rename(baseModel = BIC)
-
-# Iteratively identify covariates in phase I
-phase2.covars = c('LibraryBatch', 'RibozeroBatch', 'FlowcellBatch')
-vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
-
-adjust.covars = c(adjust.covars, vars.selected)
-```
-Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`
-#### Phase III (RNASeq alignment specific covariates)
-```{r phase3}
-# Fit a base model with the following covariates
-baseModelBIC = fitFixedEffectModel(adjust.covars, 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
-                                   COVARIATES) %>%
-  dplyr::rename(baseModel = BIC)
-
-# Iteratively identify covariates in phase I
-phase3.covars = c("TotalReads", "X5prime_Norm", "Mean_Per_Base_Cov.", "Mean_CV", "Fragment_Length_StdDev",
-                  "Mapped_Pairs", "End_1_Mismatch_Rate", "End_2_Mismatch_Rate", "Mapped", "IntragenicRate",
-                  "ExonicRate", "IntergenicRate", "IntronicRate", "rRNARate", "GenesDetected",
-                  "TranscriptsDetected", "ExpProfEfficiency", "PercentMapped")
-vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
-
-adjust.covars = c(adjust.covars, vars.selected)
-```
-Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`
-
-### Normalisation
-Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
-```{r iterative.normalisation, results='asis'}
+### Custom Normalisation
+Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue).
+
+Based on the model selection analysis here (syn9907541), following variables were controlled for differential expression: 
+    Gender, RIN, IntronicRate
+```{r custom.normalisation, results='asis'}
+adjust.covars = c('Dx.Tissue', 'Gender', 'RIN', 'IntronicRate')
 writeLines(paste('Using following covariates in the model:',
-                 paste(adjust.covars, collapse=', '),
-                 'as fixed effects and Individual_ID is chosen as random effect'))
+                 paste(adjust.covars, collapse=', ')))
 
@@
 ```{r diffExp, fig.height=8, fig.width=8}
-# Covariates to adjust
-postAdjustCovars = c('Dx.Tissue', 'Gender', 'RIN', 'IntronicRate')
-
-writeLines(paste('Using following covariates in the model:',
-                 paste(postAdjustCovars, collapse=', ')))
-
-# Post adjusted design matrix
-DESIGN = getDesignMatrix(droplevels(COVARIATES[,postAdjustCovars,drop=F]),Intercept = F)$design
-DESIGN = DESIGN[,linColumnFinder(DESIGN)$indepCols]
-
-# Estimate voom weights
-VOOM.GENE_EXPRESSION = limma::voom(NEW.COUNTS, design = DESIGN)
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
-
-# Fit linear model using new weights and new design
-ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
-
 # Fit contrast
@@
 
-  normalizationType	= 'CQNandMS'
+  normalizationType	= 'CQN'
 )
@@
 # Code
-CODE <- Folder(name = "HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)", parentId = parentId)
+CODE <- Folder(name = "HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
 annotations(CODE) = all.annotations
@@
 # Store filtered counts
-NEW.COUNTS %>%
+PROCESSED_COUNTS$filteredExprMatrix$counts %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
@@
 COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
+# Store filtered counts
+NEW.COUNTS %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  write.table(file = 'CMC_ACC_DLPFC_nCounts.tsv', sep = '\t', row.names=F, quote=F)
+NCOUNT_OBJ = File('CMC_ACC_DLPFC_nCounts.tsv', name = 'Counts (estimated)', parentId = CODE$properties$id)
+annotations(NCOUNT_OBJ) = annotations(CODE)
+annotations(NCOUNT_OBJ)$dataSubType = 'estimatedCounts'
+NCOUNT_OBJ = synStore(NCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 
@@
 |  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
+|  Counts (estimated)                         |  `r NCOUNT_OBJ$properties$id`    |
 |  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |

### 2017-08-07 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Added markdown to process HBCC DLPFC RNASeq data

@@
 Correlation/association between covariates at an FDR <= 0.1
-```{r covariates.correlation, fig.width=15, fig.height=15}
+```{r covariates.correlation, fig.width=20, fig.height=10}
 COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
@@
 Detect outlier samples based on their expression (logCPM) pattern
-```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
+```{r detect.outliers, fig.height=5, fig.width=6, results='asis', cache = FALSE}
 indToRemove = c("CMC_HBCC_RNA_PFC_3402")
@@
 ### Sample clustering
-```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
+```{r decompse.normalise.data, fig.height=5, fig.width=5, results='asis', cache=FALSE}
 # Find principal components of expression to plot
@@
 Tree based clustering of samples
-```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
+```{r decompse.normalise.data.1, fig.height=5, fig.width=10, results='asis'}
 # Eucledian tree based analysis
@@
 ### Clustering residual data
-```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
+```{r decompse.normalise.data2, fig.height=5, fig.width=5, results='asis'}
 # Find principal components of expression to plot
@@
 
-```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
+```{r decompse.normalise.data2.1, fig.height=6, fig.width=10, results='asis'}
 # Eucledian tree based analysis

### 2017-08-07 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Added markdown to process HBCC data with surrogate variable analysis

--- 2017-08-07 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
@@
 ---
-title: "Covariate analysis of HBCC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN normlaisation)"
+title: "Covariate analysis of HBCC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN and SVA normlaisation)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
+knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd",
                                  parentId ="syn7501880",
-                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
+                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - SVA')
 ```
@@
 activityName = 'Covariate adjustments';
-activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (DLPFC)';
+activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN and SVA normalisation (DLPFC)';
 
-thisFileName <- 'HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd'
+thisFileName <- 'HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd'
 
@@
 Detect outlier samples based on their expression (logCPM) pattern
-```{r detect.outliers, fig.height=5, fig.width=6, results='asis', cache = FALSE}
+```{r detect.outliers, fig.height=5, fig.width=5, results='asis', cache = FALSE}
 indToRemove = c("CMC_HBCC_RNA_PFC_3402")
@@
 ### Sample clustering
-```{r decompse.normalise.data, fig.height=5, fig.width=5, results='asis', cache=FALSE}
+```{r decompse.normalise.data, fig.height=6, fig.width=8, results='asis', cache=FALSE}
 # Find principal components of expression to plot
@@
 Tree based clustering of samples
-```{r decompse.normalise.data.1, fig.height=5, fig.width=10, results='asis'}
+```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
 # Eucledian tree based analysis
@@
 RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
-
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-tmp = RESIDUAL.GENE_EXPRESSION
-tmp[is.na(tmp)] = 0
-residualSigCovars = runPCAandPlotCorrelations(tmp, COVARIATES,
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
 ```
 
-### Sanity check
-```{r residualSigCovars.manual, fig.width=12, fig.height=8}
-residualSigCovars[["PC_res"]][[2]]$plotData
+### SVA Adjustments for eQTL analysis
+Conditioned on primary variable (Dx.Tissue) and covariates estimate surrogate variables using SVA package
+```{r sva}
+# Get (null) design matrix
+DESIGN = DM1$design
+MODEL0 = DESIGN[,-grep('Dx.Tissue',colnames(DESIGN))] # Get null model by removing variable of interest
+MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
+MODEL1 = DESIGN
+MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
+
+# Source modified functions from sva package
+source('../R/irwsva.build.R')
+source('../R/f.pvalue.R')
+
+# Compute actual variance of all principal components
+tmp = svd(RESIDUAL.GENE_EXPRESSION)
+actual.var = tmp$d^2/sum(tmp$d^2)
+
+# Compute permuted variance of all principal components
+permuted.var = foreach(i = 1:20, .combine = rbind, .packages = c('dplyr')) %dopar% {
+  tmp.residual = t(apply(RESIDUAL.GENE_EXPRESSION, 1, sample, replace = FALSE))
+  tmp = svd(tmp.residual)
+  permuted.var = tmp$d^2/sum(tmp$d^2)
+}
+permuted.var = apply(permuted.var, 2, mean)
+NUM.SV = sum(actual.var >= permuted.var[1])
+
+# Plot variance components
+var.comp = data.frame(component = 1:length(actual.var), 
+                      residual = actual.var, 
+                      permuted.residual = permuted.var) %>%
+  tidyr::gather(data, value, -component)
+p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.2)), aes(x = component, y = value, color = data)) + geom_point()
+p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
+p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
+p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
+p
+
+# Estimate surrogate variables
+SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E, MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
+colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
+rownames(SURR.VAR) = rownames(MODEL1)
+
+# Estimate voom weights for dispersion control
+VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
+                        design=cbind(MODEL1, SURR.VAR), plot=F)
+VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+
+# Fit linear model using new weights and new design
+VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)
 ```
@@
 ### Residual calculation
-Calculate weighted residuals and add back "Dx" to the residuals
+Calculate weighted residuals and add back "Dx.Tissue" to the residuals
 ```{r varsToAddBack}
+# Residuals after normalisation
+RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
+                                                  CQN.GENE_EXPRESSION$E)
+
 # Add variable of interest back to the residuals
-varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
-  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+varsToAddIn = grep("Dx.Tissue", colnames(DESIGN), value = T)
+RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
+  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
 ```
-
-### Coexpression of residual expression 
+Coexpression of genes 
 ```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-tmp = RESIDUAL.GENE_EXPRESSION
-tmp[is.na(tmp)] = 0
-cr = cor(t(tmp))
+cr = cor(t(RESIDUAL.SVA.GENE_EXPRESSION))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
 ```
-
-### Clustering residual data
-```{r decompse.normalise.data2, fig.height=5, fig.width=5, results='asis'}
+Clustering residual data
+```{r decompse.normalise.data2, fig.height=6, fig.width=6, results='asis'}
 # Find principal components of expression to plot
-PC <- prcomp(tmp, scale.=T, center = T)
+PC <- prcomp(RESIDUAL.SVA.GENE_EXPRESSION, scale.=T, center = T)
 
@@
 plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
+  dplyr::mutate(label = SampleID)
+# plotdata$label[!(plotdata$SampleID %in% top.nsamples)] = ''
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Gender, shape=Tissue, size=Age_of_Death))
-p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
+p <- p + geom_point(aes(color = Dx, shape=Tissue, size=Age_of_Death))
+p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
+# p <- p + geom_text(aes(label= label), size=4, hjust=0)
 p
@@
 
-```{r decompse.normalise.data2.1, fig.height=6, fig.width=10, results='asis'}
+```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue','Gender')])
+COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue'),drop=FALSE])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
 
-tree = hclust(as.dist(t(tmp)))
+tree = hclust(as.dist(t(RESIDUAL.SVA.GENE_EXPRESSION)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
                            main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue','Gender')]))
+                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue'),drop=FALSE]))
 ```
@@
 gc()
-```
-
-### Differential expression analysis (SCZ - Control)
-Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=8, fig.width=8}
-# Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
-                                     "Dx.TissueBP_DLPFC-Dx.TissueControl_DLPFC",
-                                     "Dx.TissueSCZ_DLPFC-Dx.TissueBP_DLPFC"),
-                         levels = colnames(ADJUSTED.FIT$coefficients))
-FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
-FIT.CONTR = eBayes(FIT.CONTR)
-
-# Get differnetial expression
-DE = lapply(1:dim(contrast)[2], function(i, FIT){
-  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
-    rownameToFirstColumn('ensembl_gene_id') 
-}, FIT.CONTR) 
-names(DE) = gsub('Dx.Tissue', '', colnames(contrast))
-
-DE = DE %>% 
-  rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
-                Direction = sign(logFC),
-                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
-                Direction = as.character(Direction)) %>%
-  tidyr::separate(Comparison, into = c('to.state','ref.state'), sep = '-') %>%
-  tidyr::separate(to.state, into = c('to.Dx','to.Tissue'), sep = '_') %>%
-  tidyr::separate(ref.state, into = c('ref.Dx','ref.Tissue'), sep = '_') %>%
-  left_join(Ensemble2HGNC) %>%
-  left_join(GENE.LEN %>% dplyr::mutate(Length = as.numeric(Length)))
-DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
-
-tmp = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05, ref.Dx != 'Other', to.Dx != 'Other') %>%
-  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue, logFC) %>%
-  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
-  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
-                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))
-
-tmp1 = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), ref.Dx != 'Other', to.Dx != 'Other') %>%
-  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue, logFC) %>%
-  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
-                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))
-
-kable(full_join(tmp,tmp1))
-
-p = DE %>%
-  dplyr::filter(ref.Dx != 'ALL', to.Dx != 'ALL') %>%
-  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
-p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
-p = p + facet_grid(ref.Dx+ref.Tissue~.+to.Dx+to.Tissue, scales = 'fixed')
-p = p + ggtitle('Differential Expression between Schizophrenia and Control')
-p
-```
-
-### Associate differential expression results with gc content, gene length and average expression
-```{r associate.de, fig.height= 6, fig.width = 12}
-pl = list()
-pl[[1]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
-                 aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
-pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
-pl[[1]] = pl[[1]] + theme(legend.position = 'top')
-
-pl[[2]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
-                 aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
-pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
-pl[[2]] = pl[[2]] + theme(legend.position = 'top')
-
-pl[[3]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
-                 aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
-pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
-pl[[3]] = pl[[3]] + theme(legend.position = 'top')
-
-multiplot(plotlist = pl, cols = 3)
-```
-
-### Differential expression analysis (Gender specific SCZ - Control)
-Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp1, fig.height=6, fig.width=12}
-# Covariates to adjust
-COVARIATES$Dx.Tissue.Gender = factor(paste(COVARIATES$Dx.Tissue, COVARIATES$Gender, sep = '_'))
-postAdjustCovars = c('Dx.Tissue.Gender', 'RIN', 'IntronicRate')
-
-writeLines(paste('Using following covariates in the model:',
-                 paste(postAdjustCovars, collapse=', ')))
-
-# Post adjusted design matrix
-DESIGN1 = getDesignMatrix(droplevels(COVARIATES[,postAdjustCovars,drop=F]),Intercept = F)$design
-DESIGN1 = DESIGN1[,linColumnFinder(DESIGN1)$indepCols]
-
-# Estimate voom weights
-VOOM.GENE_EXPRESSION1 = limma::voom(NEW.COUNTS, design = DESIGN1)
-VOOM.GENE_EXPRESSION1$E = CQN.GENE_EXPRESSION$E
-
-# Fit linear model using new weights and new design
-ADJUSTED.FIT1 = lmFit(VOOM.GENE_EXPRESSION1)
-
-# Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.Tissue.GenderSCZ_DLPFC_Male-Dx.Tissue.GenderControl_DLPFC_Male",
-                                     "Dx.Tissue.GenderSCZ_DLPFC_Female-Dx.Tissue.GenderControl_DLPFC_Female"),
-                         levels = colnames(ADJUSTED.FIT1$coefficients))
-FIT.CONTR1 = contrasts.fit(ADJUSTED.FIT1, contrasts=contrast)
-FIT.CONTR1 = eBayes(FIT.CONTR1)
-
-# Get differnetial expression
-DE1 = lapply(1:dim(contrast)[2], function(i, FIT){
-  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
-    rownameToFirstColumn('ensembl_gene_id') 
-}, FIT.CONTR1) 
-names(DE1) = gsub('Dx.Tissue.Gender', '', colnames(contrast))
-
-DE1 = DE1 %>% 
-  data.table::rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Comparison = gsub('Dx.Tissue.Gender','',Comparison),
-                Direction = sign(logFC),
-                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
-                Direction = as.character(Direction)) %>%
-  tidyr::separate(Comparison, into = c('to.state','ref.state'), sep = '-') %>%
-  tidyr::separate(to.state, into = c('to.Dx', 'to.Tissue', 'to.Gender'), sep = '_') %>%
-  tidyr::separate(ref.state, into = c('ref.Dx', 'ref.Tissue', 'ref.Gender'), sep = '_') %>%
-  left_join(Ensemble2HGNC) %>%
-  left_join(GENE.LEN %>% dplyr::mutate(Length = as.numeric(Length)))
-DE1$Direction[DE1$adj.P.Val > 0.05 | abs(DE1$logFC) < log2(1.2)] = 'NONE'
-
-tmp = DE1 %>%
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender, logFC) %>%
-  group_by(to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender) %>%
-  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
-                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))
-
-tmp1 = DE1 %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender, logFC) %>%
-  group_by(to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
-                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))
-
-kable(full_join(tmp,tmp1))
-
-p = DE1 %>%
-  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
-p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
-p = p + facet_grid(ref.Tissue~.+ref.Gender, scales = 'fixed')
-p = p + ggtitle('Differential Expression between Schizophrenia and Control')
-p
-
-table(COVARIATES$Dx.Tissue, COVARIATES$Gender)
-```
-
-### Differential expression analysis (ACC - DLPFC, SCZ - Control, Female - Male)
-Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp2, fig.height=10, fig.width=6}
-postAdjustCovars = c('Dx.Tissue', 'Gender', 'RIN', 'IntronicRate')
-
-# Post adjusted design matrix
-DESIGN2 = getDesignMatrix(droplevels(COVARIATES[,postAdjustCovars,drop=F]),Intercept = T)$design
-DESIGN2 = DESIGN2[,linColumnFinder(DESIGN2)$indepCols]
-
-# Estimate voom weights
-VOOM.GENE_EXPRESSION2 = voom(NEW.COUNTS, design = DESIGN2)
-VOOM.GENE_EXPRESSION2$E = CQN.GENE_EXPRESSION$E
-
-# Fit linear model using new weights and new design
-ADJUSTED.FIT2 = lmFit(VOOM.GENE_EXPRESSION2)
-
-# Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
-                                     "GenderMale"),
-                         levels = colnames(ADJUSTED.FIT2$coefficients))
-FIT.CONTR2 = contrasts.fit(ADJUSTED.FIT2, contrasts=contrast)
-FIT.CONTR2 = eBayes(FIT.CONTR2)
-
-# Get differnetial expression
-DE2 = lapply(1:dim(contrast)[2], function(i, FIT){
-  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
-    rownameToFirstColumn('ensembl_gene_id') 
-}, FIT.CONTR2) 
-names(DE2) = c('SCZ-CONTROL', 'FEMALE-MALE')
-
-DE2 = DE2 %>% 
-  rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Direction = sign(logFC),
-                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
-                Direction = as.character(Direction)) %>%
-  left_join(Ensemble2HGNC) 
-DE2$Direction[DE2$adj.P.Val > 0.05 | abs(DE2$logFC) < log2(1.2)] = 'NONE'
-
-tmp = DE2 %>%
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
-  group_by(Comparison) %>%
-  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
-                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))
-
-tmp1 = DE2 %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
-  group_by(Comparison) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
-                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))
-
-kable(full_join(tmp,tmp1))
-
-p = DE2 %>%
-  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
-p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
-p = p + facet_grid(Comparison~., scales = 'free_y')
-p = p + ggtitle('Differential Expression')
-p
-```
-
-### Gene set enrichment analysis
-Using Fisher's exact test for gene set enrichment analysis
-```{r enrich.de, fig.height=16, fig.width=16}
-all.diff.exp = list(DE %>%
-                      tidyr::unite(Comparison1, to.Dx, to.Tissue) %>%
-                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue) %>%
-                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
-                    DE1%>%
-                      tidyr::unite(Comparison1, to.Dx, to.Tissue, to.Gender) %>%
-                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue, ref.Gender) %>%
-                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
-                    DE2) %>%
-  data.table::rbindlist(fill = TRUE, use.names = TRUE)
-diffexp.genes.up = all.diff.exp %>%
-  dlply(.(Comparison), .fun = function(x){
-    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC > 0] %>% unique
-    tmp = tmp[!is.na(tmp)]
-  })
-names(diffexp.genes.up) = paste(names(diffexp.genes.up), 'UP', sep = '.')
-
-diffexp.genes.down = all.diff.exp %>%
-  dlply(.(Comparison), .fun = function(x){
-    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC < 0] %>% unique
-    tmp = tmp[!is.na(tmp)]
-  })
-names(diffexp.genes.down) = paste(names(diffexp.genes.down), 'DOWN', sep = '.')
-
-diffexp.genes = c(diffexp.genes.up, diffexp.genes.down)
-diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
-backgroundGenes = unique(DE$hgnc_symbol)
-
-# Download all related genesets from synapse
-geneset.files = synQuery('select * from file where parentId == "syn3240583"')
-ALL_USED_IDs = c(ALL_USED_IDs, geneset.files$file.id)
-genesets.to.test = sapply(geneset.files$file.name, function(dataSource, geneset.files){
-  tmp = downloadFile(geneset.files$file.id[geneset.files$file.name == dataSource]) %>%
-    dlply(.(Name), .fun = function(x){
-      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
-    })
-  }, geneset.files) %>%
-  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
-  do.call(c,.)
-
-# Perform enrichment analysis
-enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
-   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
-     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
-    }, x, backgroundGenes, .parallel = TRUE) %>%
-    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
-}, genesets.to.test, backgroundGenes) %>%
-  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
-  dplyr::rename(Category = .id)
 ```
@@
 
-  normalizationType	= 'CQN'
+  normalizationType	= 'SVA'
 )
@@
 # Code
-CODE <- Folder(name = "HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
+CODE <- Folder(name = "HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - SVA", parentId = parentId)
 annotations(CODE) = all.annotations
@@
 
+# Store design matrix
+SURR.VAR %>%
+  rownameToFirstColumn('SampleID') %>%
+  write.table(file = 'CMC_ACC_DLPFC_SurrVar.tsv', sep = '\t', row.names=F, quote=F)
+SURR.VAR_OBJ = File('CMC_ACC_DLPFC_SurrVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
+annotations(SURR.VAR_OBJ) = annotations(CODE)
+annotations(SURR.VAR_OBJ)$dataSubType = 'surrogateVariable'
+SURR.VAR_OBJ = synStore(SURR.VAR_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
 # Store residual gene expression for network analysis
-RESIDUAL.GENE_EXPRESSION %>%
+RESIDUAL.SVA.GENE_EXPRESSION %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_ACC_DLPFC_netResidualExpression.tsv', 
-                name = 'Normalised, covariates removed residual expression (for network analysis)', 
+  write.table(file = 'CMC_ACC_DLPFC_eqtlResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+eEXP_OBJ = File('CMC_ACC_DLPFC_eqtlResidualExpression.tsv', 
+                name = 'Normalised, covariates removed residual expression (for eQTL analysis)', 
                 parentId = CODE$properties$id)
-annotations(nEXP_OBJ) = annotations(CODE)
-annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
-nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+annotations(eEXP_OBJ) = annotations(CODE)
+annotations(eEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
+eEXP_OBJ = synStore(eEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 
-# Store differential expression results
-write.table(DE, file = 'CMC_ACC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_ACC_DLPFC_DiffExpression.tsv', 
-                name = 'Differential Expression Results (Dx.Tissue)', 
-                parentId = CODE$properties$id)
-annotations(DEXP_OBJ) = annotations(CODE)
-annotations(DEXP_OBJ)$dataSubType = 'diffExp'
-DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store enrichment results
-write.table(enrichment.results, file = 'CMC_ACC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
-ENRICH_OBJ = File('CMC_ACC_DLPFC_EnrichResults.tsv', 
-                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
-                  parentId = CODE$properties$id)
-annotations(ENRICH_OBJ) = annotations(CODE)
-annotations(ENRICH_OBJ)$dataSubType = 'enrichResults'
-ENRICH_OBJ = synStore(ENRICH_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 # stopCluster(cl)
@@
 |  Design Matrix                              |  `r DM_OBJ$properties$id`        |
-|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
-|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
-|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |
+|  Surrogate Variables                        |  `r SURR.VAR_OBJ$properties$id`  |
+|  Residual Expression (for eQTL analysis)    |  `r eEXP_OBJ$properties$id`      |
 

### 2017-08-08 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor sanity fix

@@
 
-### Differential expression analysis (ACC - DLPFC, SCZ - Control, Female - Male)
+### Differential expression analysis (DLPFC, SCZ - Control, Female - Male)
 Genes that are differentially expressed at an FDR <= 0.05 are
@@
 COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
-write.table(COVARIATES, file = 'CMC_ACC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
-COV_OBJ = File('CMC_ACC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+write.table(COVARIATES, file = 'HBCC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
+COV_OBJ = File('HBCC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
 annotations(COV_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
+  write.table(file = 'HBCC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
+COUNT_OBJ = File('HBCC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
 annotations(COUNT_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_nCounts.tsv', sep = '\t', row.names=F, quote=F)
-NCOUNT_OBJ = File('CMC_ACC_DLPFC_nCounts.tsv', name = 'Counts (estimated)', parentId = CODE$properties$id)
+  write.table(file = 'HBCC_DLPFC_nCounts.tsv', sep = '\t', row.names=F, quote=F)
+NCOUNT_OBJ = File('HBCC_DLPFC_nCounts.tsv', name = 'Counts (estimated)', parentId = CODE$properties$id)
 annotations(NCOUNT_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
+  write.table(file = 'HBCC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
+LCOUNT_OBJ = File('HBCC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
 annotations(LCOUNT_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('CMC_ACC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
+  write.table(file = 'HBCC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
+OFFSET_OBJ = File('HBCC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
 annotations(OFFSET_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'CMC_ACC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('CMC_ACC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+  write.table(file = 'HBCC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
+DM_OBJ = File('HBCC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
 annotations(DM_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_ACC_DLPFC_netResidualExpression.tsv', 
+  write.table(file = 'HBCC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+nEXP_OBJ = File('HBCC_DLPFC_netResidualExpression.tsv', 
                 name = 'Normalised, covariates removed residual expression (for network analysis)', 
@@
 # Store differential expression results
-write.table(DE, file = 'CMC_ACC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_ACC_DLPFC_DiffExpression.tsv', 
+write.table(DE, file = 'HBCC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('HBCC_DLPFC_DiffExpression.tsv', 
                 name = 'Differential Expression Results (Dx.Tissue)', 
@@
 # Store enrichment results
-write.table(enrichment.results, file = 'CMC_ACC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
-ENRICH_OBJ = File('CMC_ACC_DLPFC_EnrichResults.tsv', 
+write.table(enrichment.results, file = 'HBCC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
+ENRICH_OBJ = File('HBCC_DLPFC_EnrichResults.tsv', 
                   name = 'Enrichment Analysis Results (Dx.Tissue)', 

### 2017-08-08 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Minor bug fix

@@
   rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'CMC_ACC_DLPFC_SurrVar.tsv', sep = '\t', row.names=F, quote=F)
-SURR.VAR_OBJ = File('CMC_ACC_DLPFC_SurrVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
+  write.table(file = 'HBCC_DLPFC_SurrVar.tsv', sep = '\t', row.names=F, quote=F)
+SURR.VAR_OBJ = File('HBCC_DLPFC_SurrVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
 annotations(SURR.VAR_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_eqtlResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-eEXP_OBJ = File('CMC_ACC_DLPFC_eqtlResidualExpression.tsv', 
+  write.table(file = 'HBCC_DLPFC_eqtlResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+eEXP_OBJ = File('HBCC_DLPFC_eqtlResidualExpression.tsv', 
                 name = 'Normalised, covariates removed residual expression (for eQTL analysis)', 

### 2017-08-09 code/Rmd/ALL_REPROCESSED.Rmd
Minor changes

@@
 CQN.GENE_EXPRESSION$E = LOG.CPM
+NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
 ```
@@
 
+# Store filtered counts
+NEW.COUNTS %>%
+  rownameToFirstColumn('Gene.ID') %>%
+  write.table(file = 'AMP_nCounts.tsv', sep = '\t', row.names=F, quote=F)
+NCOUNT_OBJ = File('AMP_nCounts.tsv', name = 'Counts (estimated)', parentId = CODE$properties$id)
+annotations(NCOUNT_OBJ) = annotations(CODE)
+annotations(NCOUNT_OBJ)$dataSubType = 'estimatedCounts'
+NCOUNT_OBJ = synStore(NCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
 # Store logCPM

### 2017-08-09 code/Rmd/ALL_REPROCESSED_v2.Rmd
Minor changes

@@
 ### Model Identification
-```{r fxn}
+```{r fxn, eval = F}
 # Function to fit a mixed effect model and return BIC
@@
 # Get design matrix
-DESIGN.MAT = getDesignMatrix(COVARIATES[colnames(NEW.COUNTS), 'Study', drop=F], Intercept = F)
+DESIGN.MAT = getDesignMatrix(COVARIATES[colnames(NEW.COUNTS), 'Study', drop=F], Intercept = T)
 DESIGN.MAT = DESIGN.MAT$design[,linColumnFinder(DESIGN.MAT$design)$indepCols]
@@
 p
+
+adjust.covars = c('BrainRegion.Diagnosis', 'RIN', 'RIN2', 'Gender', 'PMI', 'AOD', 'Batch', 'PCT_RIBOSOMAL_BASES', 
+                  'PCT_INTRONIC_BASES', 'PCT_CODING_BASES', 'PCT_PF_READS_ALIGNED', 'PCT_INTERGENIC_BASES')
 ```
 #### Phase I (clinical, ancestry and sample specific technical variables)
-```{r phase1}
+```{r phase1, eval = F}
 # Fit a base model with the following covariates
@@
 #### Phase II (batch effects)
-```{r phase2}
+```{r phase2, eval = F}
 # Fit a base model with the following covariates
@@
 #### Phase III (RNASeq alignment specific covariates)
-```{r phase3}
+```{r phase3, eval = F}
 # Fit a base model with the following covariates
@@
 # Estimate consensus correlation coefficient
-correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
+# correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
+correlation = list()
+correlation$cor = 0.45
 
@@
 # Fit linear model using new weights and new design
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
+VOOM.GENE_EXPRESSION$E = RESIDUAL.EXPR
 ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
@@
 # Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, VOOM.GENE_EXPRESSION$E)
 
@@
 VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)
+# correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)
 
@@
 # Code
-CODE <- Folder(name = "ALL Reprocessed RNASeq (Normalised)", parentId = parentId)
+CODE <- Folder(name = "ALL Reprocessed RNASeq (Normalised) - study effects removed", parentId = parentId)
 annotations(CODE) = all.annotations

### 2017-08-10 code/Rmd/ROSMAP_REPROCESSED.Rmd
Added apoe4 specific differential expression to ROSMAP

@@
 #"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv", "PF_INDEL_RATE", "PF_MISMATCH_RATE", "age_first_ad_dx","age_at_visit_max",
-primaryVariable <- c("cogdx", "Diagnosis")
-FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx", "Diagnosis")
+primaryVariable <- c("cogdx", "Diagnosis", "apoe_genotype")
+FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx", "Diagnosis", "apoe_genotype")
 ContCovariates <- c("RINcontinuous", "age_death", "pmi", "educ", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
@@
 ContCovariates = c(ContCovariates, 'RINcontinuous2')
+
+levels(COVARIATES$apoe_genotype) = c('0','0','1','0','1','2')
 ```
@@
 
-### Conditional Quantile Normalisation (CQN)
-Initial normalisation usign cqn and dispersion control with voom (with NULL design)
-```{r initial.voom.normalisation, fig.height=5, fig.width=5}
+### Winsorise counts (Gene outliers in samples)
+```{r winsorise.data}
+LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)
+
+# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
+LOG.CPM = apply(LOG.CPM, 1, function(x){
+  mn = mean(x, na.rm = T)
+  std.dev = sd(x, na.rm = T)
+
+  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
+  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
+
+  x[x < (mn-3*std.dev)] = x[ind.low]
+  x[x > (mn+3*std.dev)] = x[ind.high]
+  return(x)
+}) %>% t
+
+# Back calculate counts for further analysis
+LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
+NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
+```
+
+### Library Normalisation
+Initial library normalisation usign cqn
+```{r cqn}
 # Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix$counts), 'percentage_gc_content'],
-                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix$counts), 'gene.length'],
+CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
+                          x = GENE.GC[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
+                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'gene.length'],
                           lengthMethod = "smooth", 
                           verbose = FALSE)
-CQN.GENE_EXPRESSION$E.NA = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
-CQN.GENE_EXPRESSION$E.zero = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
-
-# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
-log.mat = apply(CQN.GENE_EXPRESSION$E.zero, 1, function(x){
+CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+
+# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
+LOG.CPM = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
   mn = mean(x, na.rm = T)
   std.dev = sd(x, na.rm = T)
-  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
+
+  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
+  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
+
+  x[x < (mn-3*std.dev)] = x[ind.low]
+  x[x > (mn+3*std.dev)] = x[ind.high]
+  return(x)
 }) %>% t
-PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = NA
-
-CQN.GENE_EXPRESSION$E.NA[log.mat] = NA
-CQN.GENE_EXPRESSION$E.zero[log.mat] = 0
-```
-
-```{r explore.data, cache=FALSE, fig.height=5, fig.width=10}
-# Plot abberent distribution of logcpm counts
-tmp1 = (CQN.GENE_EXPRESSION$E.NA) %>%
-  rownameToFirstColumn('Gene.ID') %>%
-  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
-  left_join(COVARIATES %>%
-              rownameToFirstColumn('SampleID'))
-
-p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
-p = p + theme(legend.position = 'NONE') + facet_grid(.~Diagnosis, scale = 'free')
-p
-```
-
-Coexpression of genes 
-```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(CQN.GENE_EXPRESSION$E.zero))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
-### Decomposition of expression
-Clustering of library normalised unadjusted data (with NULL design)
-```{r decompse.normalise.data1.1, fig.height=6, fig.width=8, results='asis'}
+CQN.GENE_EXPRESSION$E = LOG.CPM
+NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
+```
+
+### Sample clustering
+```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
 # Find principal components of expression to plot
-PC <- prcomp(CQN.GENE_EXPRESSION$E.zero, scale.=T, center = T)
+PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)
 
@@
 ```
-Tree based classification of samples
-```{r decompse.normalise.data1.2, fig.height=6, fig.width=10, results='asis'}
+Tree based clustering of samples
+```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
 # Eucledian tree based analysis
-COVARIATES.tmp = (COVARIATES[,c(FactorCovariates), drop = F])
+COVARIATES.tmp = data.matrix(COVARIATES[,c("Batch", "msex", "Diagnosis")])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
 
-tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E.zero)))
+tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 ```
-
 ```{r temp, include = F}
@@
 gc()
+```
+
+### Distribution of samples (log cpm)
+```{r lcpm.dist, cache=FALSE, fig.height=10, fig.width=20}
+# Plot abberent distribution of logcpm counts
+tmp1 = (CQN.GENE_EXPRESSION$E) %>%
+  rownameToFirstColumn('Gene.ID') %>%
+  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID'))
+
+p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
+p = p + theme(legend.position = 'NONE') + facet_grid(.~Diagnosis, scale = 'free')
+p
+```
+
+Coexpression of genes 
+```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
+cr = cor(t(CQN.GENE_EXPRESSION$E))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
 ```
@@
   # Estimate voom weights for dispersion control
-  counts = PROCESSED_COUNTS$filteredExprMatrix$counts
-  counts[is.na(counts)] = 0
-  VOOM.GENE_EXPRESSION = voom(counts, 
+  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, 
                               design=DM1$design, 
@@
   # Residuals after normalisation
-  RESIDUAL.GENE_EXPRESSION = list()
-  RESIDUAL.GENE_EXPRESSION$expr.NA = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
-                                                        CQN.GENE_EXPRESSION$E.NA)
-  RESIDUAL.GENE_EXPRESSION$expr.zero = RESIDUAL.GENE_EXPRESSION$expr.NA
-  RESIDUAL.GENE_EXPRESSION$expr.zero[is.na(RESIDUAL.GENE_EXPRESSION$expr.NA)] = 0
+  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
+                                                CQN.GENE_EXPRESSION$E)
 
@@
   # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION$expr.zero, 
+  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                 COVARIATES[, residCovars, drop=F], 
@@
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION$expr.zero, 
+residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                               COVARIATES,
@@
 # Estimate voom weights for dispersion control
-counts = PROCESSED_COUNTS$filteredExprMatrix$counts
-counts[is.na(counts)] = 0
-VOOM.WEIGHTS = voom(counts, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
 
 # Fit linear model using new weights and new design
-FIT = lmFit(CQN.GENE_EXPRESSION$E.NA,
+FIT = lmFit(CQN.GENE_EXPRESSION$E,
             design = DESIGN$design,
@@
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(counts, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
 
 # Fit linear model using new weights and new design
-FIT = lmFit(CQN.GENE_EXPRESSION$E.NA,
+FIT = lmFit(CQN.GENE_EXPRESSION$E,
             design = DESIGN$design,
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("DiagnosisAD-DiagnosisCONTROL"),
+contrast = makeContrasts(contrasts=c("DiagnosisAD-DiagnosisCONTROL",
+                                     "msex0"),
                          levels = colnames(FIT$coefficients))
@@
 }, FIT.CONTR)
-names(DE) = c('AD-CONTROL')
+names(DE) = c('AD-CONTROL','MALE-FEMALE')
 
@@
 
-### Differential expression analysis (with Diagnosis and Gender)
-Differential expression is performed on the Diagnosis x Gender variable by controlling for covariates identified above
+### Differential expression analysis (with APOE status as primary variable)
+Differential expression is performed on the primary variable by controlling for covariates identified above
+
+Interpretation of apoe_genotype
+1. 0: No apoe4 allele
+2. 1: 1 apoe4 allele
+3. 2: 2 apoe4 allele
 
@@
 ```{r diff.exp2, fig.height=6, fig.width=18, cache=FALSE}
-## Modeling Diagnosis, msex and age_death conjointly
-COVARIATES$Diagnosis.msex = paste(COVARIATES$Diagnosis, factor(COVARIATES$msex, labels = c('0' = 'FEMALE', '1' = 'MALE')), sep = '.') %>%
-  factor
-
 # Get design matrix
-DESIGN = getDesignMatrix(COVARIATES[, c('Diagnosis.msex', setdiff(postAdjustCovars, 'msex')), drop = F], Intercept = F)
+DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[3], postAdjustCovars), drop = F], Intercept = F)
 DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
-colnames(DESIGN$design) = gsub('Diagnosis.msex','',colnames(DESIGN$design))
+DESIGN$design = na.omit(DESIGN$design)
 
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(counts, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(NEW.COUNTS[,rownames(DESIGN$design)], design=DESIGN$design, plot=F)
 
 # Fit linear model using new weights and new design
-FIT = lmFit(CQN.GENE_EXPRESSION$E.NA,
+FIT = lmFit(CQN.GENE_EXPRESSION$E[,rownames(DESIGN$design)],
             design = DESIGN$design,
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("(AD.FEMALE+AD.MALE)/2-(CONTROL.FEMALE+CONTROL.MALE)/2",
-                                     "(AD.FEMALE+CONTROL.FEMALE)/2-(AD.MALE+CONTROL.MALE)/2",
-                                     "(AD.FEMALE-CONTROL.FEMALE)/2-(AD.MALE-CONTROL.MALE)/2",
-                                     "AD.FEMALE-CONTROL.FEMALE",
-                                     "AD.MALE-CONTROL.MALE"),
+contrast = makeContrasts(contrasts=c("apoe_genotype2-apoe_genotype0",
+                                     "apoe_genotype2-apoe_genotype1",
+                                     "apoe_genotype1-apoe_genotype0"),
                          levels = colnames(FIT$coefficients))
@@
 }, FIT.CONTR)
-names(DE) = c('AD-CONTROL','FEMALE-MALE','AD-CONTROL.IN.FEMALE-MALE','AD-CONTROL.IN.FEMALE', 'AD-CONTROL.IN.MALE')
+names(DE) = colnames(contrast)
 
@@
 
-all.diff.exp = c(all.diff.exp, list(Diagnosis.Gender = DE))
-all.fit = c(all.fit, list(Diagnosis.Gender = FIT))
-```
-
-### Differential expression analysis (with Diagnosis and AOD)
-Differential expression is performed on the Diagnosis x AOD variable by controlling for covariates identified above
+all.diff.exp = c(all.diff.exp, list(Apoe = DE))
+all.fit = c(all.fit, list(Apoe = FIT))
+```
+
+### Differential expression analysis (with Diagnosis and Gender)
+Differential expression is performed on the Diagnosis x Gender variable by controlling for covariates identified above
 
@@
 ```{r diff.exp3, fig.height=6, fig.width=18, cache=FALSE}
+## Modeling Diagnosis, msex and age_death conjointly
+COVARIATES$Diagnosis.msex = paste(COVARIATES$Diagnosis, 
+                                  factor(COVARIATES$msex, labels = c('0' = 'FEMALE', '1' = 'MALE')), sep = '.') %>%
+  factor
+
 # Get design matrix
-DESIGN = getDesignMatrix(COVARIATES[, c('Diagnosis', setdiff(postAdjustCovars, 'age_death')), drop = F], Intercept = F)
-ind = grep('Diagnosis', colnames(DESIGN$design))
-DESIGN$design[,ind] = DESIGN$design[,ind] * COVARIATES[rownames(DESIGN$design),'age_death']
+DESIGN = getDesignMatrix(COVARIATES[, c('Diagnosis.msex', setdiff(postAdjustCovars, 'msex')), drop = F], Intercept = F)
 DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
-colnames(DESIGN$design) = gsub('Diagnosis','AOD.',colnames(DESIGN$design))
+colnames(DESIGN$design) = gsub('Diagnosis.msex','',colnames(DESIGN$design))
 
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(counts, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
 
 # Fit linear model using new weights and new design
-FIT = lmFit(CQN.GENE_EXPRESSION$E.NA,
+FIT = lmFit(CQN.GENE_EXPRESSION$E,
             design = DESIGN$design,
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("AOD.AD-AOD.CONTROL",
-                                     "(AOD.AD+AOD.CONTROL)/2"),
+contrast = makeContrasts(contrasts=c("(AD.FEMALE+AD.MALE)/2-(CONTROL.FEMALE+CONTROL.MALE)/2",
+                                     "(AD.FEMALE+CONTROL.FEMALE)/2-(AD.MALE+CONTROL.MALE)/2",
+                                     "(AD.FEMALE-CONTROL.FEMALE)/2-(AD.MALE-CONTROL.MALE)/2",
+                                     "AD.FEMALE-CONTROL.FEMALE",
+                                     "AD.MALE-CONTROL.MALE"),
                          levels = colnames(FIT$coefficients))
@@
 }, FIT.CONTR)
-names(DE) = c('AOD.AD-AOD.CONTROL','AOD')
+names(DE) = c('AD-CONTROL','FEMALE-MALE','AD-CONTROL.IN.FEMALE-MALE','AD-CONTROL.IN.FEMALE', 'AD-CONTROL.IN.MALE')
 
@@
 
+all.diff.exp = c(all.diff.exp, list(Diagnosis.Gender = DE))
+all.fit = c(all.fit, list(Diagnosis.Gender = FIT))
+```
+
+### Differential expression analysis (with Diagnosis and AOD)
+Differential expression is performed on the Diagnosis x AOD variable by controlling for covariates identified above
+
+Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
+```{r diff.exp4, fig.height=6, fig.width=18, cache=FALSE}
+# Get design matrix
+DESIGN = getDesignMatrix(COVARIATES[, c('Diagnosis', setdiff(postAdjustCovars, 'age_death')), drop = F], Intercept = F)
+ind = grep('Diagnosis', colnames(DESIGN$design))
+DESIGN$design[,ind] = DESIGN$design[,ind] * COVARIATES[rownames(DESIGN$design),'age_death']
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+colnames(DESIGN$design) = gsub('Diagnosis','AOD.',colnames(DESIGN$design))
+
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
+
+# Fit linear model using new weights and new design
+FIT = lmFit(CQN.GENE_EXPRESSION$E,
+            design = DESIGN$design,
+            weights = VOOM.WEIGHTS$weights)
+
+# Fit contrast
+contrast = makeContrasts(contrasts=c("AOD.AD-AOD.CONTROL",
+                                     "(AOD.AD+AOD.CONTROL)/2"),
+                         levels = colnames(FIT$coefficients))
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differential expression
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = T) %>%
+    rownameToFirstColumn('ensembl_gene_id')
+}, FIT.CONTR)
+names(DE) = c('AOD.AD-AOD.CONTROL','AOD')
+
+DE = DE %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  left_join(GENE.PARAM %>%
+              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
+              unique()) %>%
+  dplyr::mutate(Study = 'ROSMAP', Region = 'DLPFC',
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction))
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
+
+writeLines('Differentially expressed genes at an FDR 0.05 and logFC 1.2')
+tmp = DE %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
+  group_by(Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05_FC_1.2) 
+kable(tmp)
+
+p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point()
+p = p + scale_color_manual(values = c('green','grey','red'))
+p = p + facet_grid(.~Comparison, scales = 'free_x')
+p
+
 all.diff.exp = c(all.diff.exp, list(Diagnosis.AOD = DE))
@@
 
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
+
+# Fit linear model using new weights and new design
+FIT = lmFit(CQN.GENE_EXPRESSION$E,
+            design = DESIGN$design,
+            weights = VOOM.WEIGHTS$weights)
+
 # Get (null) design matrix
@@
 # Estimate surrogate variables
-SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$E.zero, 
+SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$E, 
                     MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv
@@
 # Re-estimate voom weights
-VOOM.SVA.WEIGHTS = voom(counts, 
+VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
                         design = cbind(MODEL1, SURR.VAR), plot=F)
@@
 # Fit linear model using new weights and new design
-VOOM.SVA.FIT = lmFit(CQN.GENE_EXPRESSION$E.NA, 
+VOOM.SVA.FIT = lmFit(CQN.GENE_EXPRESSION$E, 
                      design = cbind(MODEL1, SURR.VAR),
@@
   VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
+
+# Assign NA values to outlier expressions
+RESIDUAL.SVA.GENE_EXPRESSION = apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, function(x){
+  mn = mean(x, na.rm = T)
+  std.dev = sd(x, na.rm = T)
+
+  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
+  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
+
+  x[x < (mn-3*std.dev)] = NA
+  x[x > (mn+3*std.dev)] = NA
+  return(x)
+}) %>% t
 ```
@@
 # Estimate voom weights for dispersion control
-VOOM.NET.WEIGHTS = voom(counts, 
+VOOM.NET.WEIGHTS = voom(NEW.COUNTS, 
                         design=DESIGN.NET, plot=F)
@@
 # Fit linear model using new weights and new design
-VOOM.NET.FIT = lmFit(CQN.GENE_EXPRESSION$E.NA,
+VOOM.NET.FIT = lmFit(CQN.GENE_EXPRESSION$E,
                      design = DESIGN.NET,
@@
 RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
-                                                  CQN.GENE_EXPRESSION$E.NA)
+                                                  CQN.GENE_EXPRESSION$E)
+
+# Winsorise outlier expressions
+RESIDUAL.NET.GENE_EXPRESSION = apply(RESIDUAL.NET.GENE_EXPRESSION, 1, function(x){
+  mn = mean(x, na.rm = T)
+  std.dev = sd(x, na.rm = T)
+
+  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
+  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
+
+  x[x < (mn-3*std.dev)] = mn-3*std.dev
+  x[x > (mn+3*std.dev)] = mn+3*std.dev
+  return(x)
+}) %>% t
 ```
@@
 
+# Store filtered counts
+NEW.COUNTS %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  write.table(file = 'ROSMAP_DLPFC_eCounts.tsv', sep = '\t', row.names=F, quote=F)
+ECOUNT_OBJ = File('ROSMAP_DLPFC_eCounts.tsv', name = 'Counts (estimated)', parentId = CODE$properties$id)
+annotations(ECOUNT_OBJ) = annotations(CODE)
+annotations(ECOUNT_OBJ)$dataSubType = 'filteredCounts'
+ECOUNT_OBJ = synStore(ECOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
 # Store logCPM
@@
 |  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
+|  Counts (estimated)                         |  `r ECOUNT_OBJ$properties$id`    |
 |  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |

### 2017-08-10 code/Rmd/ROSMAP_REPROCESSED.Rmd
Minor changes

@@
 ```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
+# Function to change entity ACL (also creates new acls if it doesn't exist)
+synChangeEntityACL <- function(id, new.racl){
+  tryCatch({ 
+    # If old acl exist update
+    old.acl = synGetEntityACL(id)
+    old.acl@resourceAccess = new.racl
+    synUpdateEntityACL(old.acl)
+  }, error = function(e){ 
+    # If not create a new acl
+    new.acl = AccessControlList(id = id, resourceAccess = new.racl)
+    synCreateEntityACL(new.acl)
+  })
+}
+
 # Set annotations

### 2017-08-10 code/Rmd/MAYO_REPROCESSED.Rmd
Added ApoE specific differential expression in Mayo data

--- 2017-04-23 code/Rmd/MAYO_REPROCESSED.Rmd
@@
   dplyr::mutate(AgeAtDeath = gsub("_or_above", "", AgeAtDeath),
+                Diagnosis = gsub("Pathologic Aging", "OTHER", Diagnosis),
+                Diagnosis = gsub("PSP", "OTHER", Diagnosis),
                 Diagnosis = factor(Diagnosis, 
-                                   levels = c("AD", "Control", "Pathologic Aging", "PSP"), 
-                                   labels = c("AD" = "AD", "Control" = "CONTROL", "Pathologic Aging" = "OTHER", "PSP" = "OTHER")),
+                                   levels = c("AD", "Control", "OTHER")),
                 BrainRegion.Diagnosis = paste(BrainRegion, Diagnosis, sep = '.'),
-                RIN2 = RIN^2) %>%
+                RIN2 = RIN^2,
+                ApoE = factor(ApoE)) %>%
   filter(!is.na(Source)) %>%
   dplyr::filter(!(ID %in% SAMPLES.EXCLUDE))
+levels(METADATA$ApoE) = c('0','0','1','0','1','2')
+METADATA$BrainRegion.ApoE = paste(METADATA$BrainRegion, METADATA$ApoE, sep = '.')
 ```
@@
 ```{r covariates.clustering}
-primaryVariable <- "BrainRegion.Diagnosis"
-FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
+primaryVariable <- c("BrainRegion.Diagnosis", "BrainRegion.ApoE")
+FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis", "BrainRegion.ApoE")
 ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
@@
 
-### Conditional Quantile Normalisation (CQN)
-Initial normalisation usign cqn and dispersion control with voom (with NULL design)
-```{r initial.voom.normalisation, fig.height=5, fig.width=5}
+### Winsorise counts (Gene outliers in samples)
+```{r winsorise.data}
+LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)
+
+# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
+LOG.CPM = apply(LOG.CPM, 1, function(x){
+  mn = mean(x, na.rm = T)
+  std.dev = sd(x, na.rm = T)
+
+  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
+  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
+
+  x[x < (mn-3*std.dev)] = x[ind.low]
+  x[x > (mn+3*std.dev)] = x[ind.high]
+  return(x)
+}) %>% t
+
+# Back calculate counts for further analysis
+LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
+NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
+```
+
+### Library Normalisation
+Initial library normalisation usign cqn
+```{r cqn}
 # Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
-                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
-                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
+CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
+                          x = GENE.GC[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
+                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'gene.length'],
                           lengthMethod = "smooth", 
                           verbose = FALSE)
-CQN.GENE_EXPRESSION$E.NA = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-
-# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
-log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
+CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+
+# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
+LOG.CPM = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
   mn = mean(x, na.rm = T)
   std.dev = sd(x, na.rm = T)
-  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
+
+  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
+  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
+
+  x[x < (mn-3*std.dev)] = x[ind.low]
+  x[x > (mn+3*std.dev)] = x[ind.high]
+  return(x)
 }) %>% t
-
-PROCESSED_COUNTS$filteredExprMatrix$counts.NA = PROCESSED_COUNTS$filteredExprMatrix$counts 
-PROCESSED_COUNTS$filteredExprMatrix$counts.NA[log.mat] = NA
-PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0
-
-CQN.GENE_EXPRESSION$E.NA[log.mat] = NA
-CQN.GENE_EXPRESSION$E.zero = CQN.GENE_EXPRESSION$E.NA
-CQN.GENE_EXPRESSION$E.zero[log.mat] = 0
-```
-
-```{r explore.data, cache=FALSE, fig.height=5, fig.width=10}
-# Plot abberent distribution of logcpm counts
-tmp1 = (CQN.GENE_EXPRESSION$E.NA) %>%
-  rownameToFirstColumn('Gene.ID') %>%
-  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
-  left_join(COVARIATES %>%
-              rownameToFirstColumn('SampleID')) %>%
-  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
-
-p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
-p = p + theme(legend.position = 'NONE') + facet_grid(BrainRegion+.~Diagnosis, scale = 'free')
-p
-```
-Coexpression of genes 
-```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(CQN.GENE_EXPRESSION$E.zero))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
-### Decomposition of expression
-Clustering of library normalised unadjusted data (with NULL design)
-```{r decompse.normalise.data1.1, fig.height=6, fig.width=10, results='asis'}
+CQN.GENE_EXPRESSION$E = LOG.CPM
+NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
+```
+
+### Sample clustering
+```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
 # Find principal components of expression to plot
-PC <- prcomp(CQN.GENE_EXPRESSION$E.zero, scale.=T, center = T)
+PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)
 
@@
 plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion','Diagnosis'), sep = '\\.') %>% 
-  dplyr::mutate(BrainRegion = factor(BrainRegion),
-                Source = factor(Source))
+  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
 
@@
 ```
-Tree based classification of samples
-```{r decompse.normalise.data1.2, fig.height=6, fig.width=10, results='asis'}
-# Eucledian tree based analysis
+Tree based clustering of samples
+```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
 COVARIATES.tmp = (COVARIATES[,c(FactorCovariates), drop = F])
@@
   # Estimate voom weights for dispersion control
-  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)
+  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F)
 
   # Calculate correlation values of random effects
-  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E.NA
+  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
   correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
@@
   # Re-estimate voom weights
-  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, 
                               design=DM1$design, plot=F, 
@@
   # Residuals after normalisation
-  RESIDUAL.GENE_EXPRESSION = list()
-  RESIDUAL.GENE_EXPRESSION$expr.NA = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
-                                                CQN.GENE_EXPRESSION$E.NA)
-  RESIDUAL.GENE_EXPRESSION$expr.zero = RESIDUAL.GENE_EXPRESSION$expr.NA
-  RESIDUAL.GENE_EXPRESSION$expr.zero[is.na(RESIDUAL.GENE_EXPRESSION$expr.zero)] = 0
+  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
+                                                CQN.GENE_EXPRESSION$E)
 
@@
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
 
 # Calculate correlation values of random effects
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
 correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Donor_ID)
@@
 # Re-estimate voom weights
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
+VOOM.WEIGHTS = voom(NEW.COUNTS,
                     design=DESIGN$design, plot=F,
@@
 contrast = makeContrasts(contrasts=c(
-  "BrainRegion.DiagnosisCER.CONTROL-BrainRegion.DiagnosisCER.AD",
-  "BrainRegion.DiagnosisCER.CONTROL-BrainRegion.DiagnosisCER.OTHER",
+  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.AD",
+  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.OTHER",
   "BrainRegion.DiagnosisCER.AD-BrainRegion.DiagnosisCER.OTHER",
-  "BrainRegion.DiagnosisTCX.CONTROL-BrainRegion.DiagnosisTCX.AD",
-  "BrainRegion.DiagnosisTCX.CONTROL-BrainRegion.DiagnosisTCX.OTHER",
+  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.AD",
+  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.OTHER",
   "BrainRegion.DiagnosisTCX.AD-BrainRegion.DiagnosisTCX.OTHER"),
@@
 # Get differnetial expression
-DE = lapply(1:6, function(i, FIT){
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
   topTable(FIT, coef=i, number = 50000, confint = T) %>%
@@
   tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
-  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
-  tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
+  tidyr::separate(from.state, c('BrainRegion.ref','reference'), sep = '\\.') %>%
+  tidyr::separate(to.state, c('BrainRegion.ag','against'), sep = '\\.') %>%
   tidyr::unite(Comparison, reference, against, sep = '-') %>%
@@
 tmp = DE %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion) %>%
-  group_by(BrainRegion, Comparison, Direction) %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion.ref) %>%
+  group_by(BrainRegion.ref, Comparison, Direction) %>%
   dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
@@
 p = p + scale_color_manual(values = c('green','grey','red'))
-p = p + facet_grid(BrainRegion+.~Comparison, scales = 'fixed')
+p = p + facet_grid(BrainRegion.ref+.~Comparison, scales = 'fixed')
 p
+
+all.diff.exp = list(Diagnosis = DE)
+all.fit = list(Diagnosis = FIT)
 ```
@@
 pl[[1]] = ggplot(DE %>%
-                   dplyr::filter(Comparison == 'CONTROL-AD'), 
+                   dplyr::filter(Comparison == 'Control-AD'), 
                  aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
 pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(BrainRegion+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
+pl[[1]] = pl[[1]] + facet_grid(BrainRegion.ref+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
 pl[[1]] = pl[[1]] + theme(legend.position = 'top')
@@
 pl[[2]] = ggplot(DE %>%
-                   dplyr::filter(Comparison == 'CONTROL-AD'), 
+                   dplyr::filter(Comparison == 'Control-AD'), 
                  aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
 pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(BrainRegion+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
+pl[[2]] = pl[[2]] + facet_grid(BrainRegion.ref+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
 pl[[2]] = pl[[2]] + theme(legend.position = 'top')
@@
 pl[[3]] = ggplot(DE %>%
-                   dplyr::filter(Comparison == 'CONTROL-AD'), 
+                   dplyr::filter(Comparison == 'Control-AD'), 
                  aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
 pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(BrainRegion+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
+pl[[3]] = pl[[3]] + facet_grid(BrainRegion.ref+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
 pl[[3]] = pl[[3]] + theme(legend.position = 'top')
@@
 multiplot(plotlist = pl, cols = 3)
+```
+
+### Differential expression analysis (with BrainRegion.ApoE as primary variable)
+Differential expression is performed on the primary variable by controlling for covariates identified above
+
+Interpretation of BrainRegion.Diagnosis
+1. CER.0: No ApoE4 allele in cerebellum
+2. CER.1: 1 ApoE4 allele in cerebellum
+3. CER.2: 2 ApoE4 allele in cerebellum
+4. TCX.0: No ApoE4 allele in temporal cortex
+5. TCX.1: 1 ApoE4 allele in temporal cortex
+6. TCX.2: 2 ApoE4 allele in temporal cortex
+
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diff.exp1, fig.height=10, fig.width=18}
+# Get design matrix
+DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[2], postAdjustCovars), drop = F], Intercept = F)
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
+
+# Calculate correlation values of random effects
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Donor_ID)
+
+# Re-estimate voom weights
+VOOM.WEIGHTS = voom(NEW.COUNTS,
+                    design=DESIGN$design, plot=F,
+                    block = COVARIATES$Donor_ID, 
+                    correlation = correlation$cor)
+
+# Fit linear model using new weights and new design
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+FIT = lmFit(VOOM.WEIGHTS)
+
+# Fit contrast
+contrast = makeContrasts(contrasts=c(
+  "BrainRegion.ApoECER.2-BrainRegion.ApoECER.0",
+  "BrainRegion.ApoECER.2-BrainRegion.ApoECER.1",
+  "BrainRegion.ApoECER.1-BrainRegion.ApoECER.0",
+  "BrainRegion.ApoETCX.2-BrainRegion.ApoETCX.0",
+  "BrainRegion.ApoETCX.2-BrainRegion.ApoETCX.1",
+  "BrainRegion.ApoETCX.1-BrainRegion.ApoETCX.0"),
+  levels = colnames(FIT$coefficients))
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = T) %>%
+    rownameToFirstColumn('ensembl_gene_id')
+}, FIT.CONTR)
+names(DE) = colnames(contrast)
+
+DE = DE %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Comparison = gsub('BrainRegion.ApoE','',Comparison),
+                Study = 'MAYO',
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
+  tidyr::separate(from.state, c('BrainRegion.ref','reference'), sep = '\\.') %>%
+  tidyr::separate(to.state, c('BrainRegion.ag','against'), sep = '\\.') %>%
+  dplyr::mutate(reference = paste0('ApoE',reference),
+                against = paste0('ApoE',against)) %>%
+  tidyr::unite(Comparison, reference, against, sep = '-') %>%
+  left_join(GENE.PARAM %>%
+              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
+              unique())
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion.ref) %>%
+  group_by(BrainRegion.ref, Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05_FC_1.2) 
+kable(tmp)
+
+p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red'))
+p = p + facet_grid(BrainRegion.ref+.~Comparison, scales = 'fixed')
+p
+
+all.diff.exp = c(all.diff.exp, list(ApoE = DE))
+all.fit = c(all.fit, list(ApoE = FIT))
+```
+
+### Differential expression analysis (with BrainRegion.Diagnosis x Gender as primary variable)
+Differential expression is performed on the primary variable by controlling for covariates identified above
+
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diff.exp2, fig.height=5, fig.width=18}
+# Get design matrix
+COVARIATES$BrainRegion.Diagnosis.Gender = paste(COVARIATES$BrainRegion.Diagnosis, COVARIATES$Gender, sep = '.') %>%
+  factor()
+DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis.Gender', setdiff(postAdjustCovars, 'Gender')), drop = F], 
+                         Intercept = F)
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
+
+# Calculate correlation values of random effects
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Donor_ID)
+
+# Re-estimate voom weights
+VOOM.WEIGHTS = voom(NEW.COUNTS,
+                    design=DESIGN$design, plot=F,
+                    block = COVARIATES$Donor_ID, 
+                    correlation = correlation$cor)
+
+# Fit linear model using new weights and new design
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+FIT = lmFit(VOOM.WEIGHTS)
+
+# Fit contrast
+contrast = makeContrasts(contrasts=c(
+  "BrainRegion.Diagnosis.GenderCER.AD.F-BrainRegion.Diagnosis.GenderCER.Control.F",
+  "BrainRegion.Diagnosis.GenderCER.AD.M-BrainRegion.Diagnosis.GenderCER.Control.M",
+
+  "BrainRegion.Diagnosis.GenderTCX.AD.F-BrainRegion.Diagnosis.GenderTCX.Control.F",
+  "BrainRegion.Diagnosis.GenderTCX.AD.M-BrainRegion.Diagnosis.GenderTCX.Control.M"),
+  levels = colnames(FIT$coefficients))
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = T) %>%
+    rownameToFirstColumn('ensembl_gene_id')
+}, FIT.CONTR)
+names(DE) = colnames(contrast)
+
+DE = DE %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis.Gender','',Comparison),
+                Study = 'MAYO',
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
+  tidyr::separate(from.state, c('BrainRegion.ref','Dx.ref','Gender.ref'), sep = '\\.') %>%
+  tidyr::separate(to.state, c('BrainRegion.ag','Dx.ag','Gender.ag'), sep = '\\.') %>%
+  tidyr::unite(Comparison, Dx.ref, Dx.ag, sep = '-') %>%
+  left_join(GENE.PARAM %>%
+              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
+              unique())
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion.ref, Gender.ref) %>%
+  group_by(BrainRegion.ref, Gender.ref, Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05_FC_1.2) 
+kable(tmp)
+
+p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red'))
+p = p + facet_grid(Comparison+.~BrainRegion.ref+Gender.ref, scales = 'fixed')
+p
+
+all.diff.exp = c(all.diff.exp, list(Diagnosis.Gender = DE))
+all.fit = c(all.fit, list(Diagnosis.Gender = FIT))
 ```
@@
 ```{r sva.adjust}
+# Get design matrix
+DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[1], postAdjustCovars), drop = F], Intercept = F)
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
+
+# Calculate correlation values of random effects
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Donor_ID)
+
+# Re-estimate voom weights
+VOOM.WEIGHTS = voom(NEW.COUNTS,
+                    design=DESIGN$design, plot=F,
+                    block = COVARIATES$Donor_ID, 
+                    correlation = correlation$cor)
+
+# Fit linear model using new weights and new design
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+FIT = lmFit(VOOM.WEIGHTS)
+
 # Get (null) design matrix
@@
 # Estimate surrogate variables
-SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E.zero,
+SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E,
                         MODEL1, MODEL0, n.sv = NUM.SV, 
@@
 # Estimate voom weights for dispersion control
-VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
                         design=cbind(MODEL1, SURR.VAR), plot=F)
@@
 # Calculate correlation values of random effects
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
+VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
 correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$Donor_ID)
@@
 # Re-estimate voom weights
-VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
                         design = cbind(MODEL1, SURR.VAR), plot=F,
@@
 # Estimate voom weights for dispersion control
-VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+VOOM.NET.WEIGHTS = voom(NEW.COUNTS, 
                         design=DESIGN.NET, plot=F)
@@
 # Calculate correlation values of random effects
-VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
+VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E
 correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$Donor_ID)
@@
 # Re-estimate voom weights
-VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+VOOM.NET.WEIGHTS = voom(NEW.COUNTS, 
                         design = DESIGN.NET, plot=F,
@@
 COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
+# Store filtered counts
+NEW.COUNTS %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  write.table(file = 'MAYO_CBE_TCX_eCounts.tsv', sep = '\t', row.names=F, quote=F)
+ECOUNT_OBJ = File('MAYO_CBE_TCX_eCounts.tsv', name = 'Counts (estimated)', parentId = CODE$properties$id)
+annotations(ECOUNT_OBJ) = annotations(CODE)
+annotations(ECOUNT_OBJ)$dataSubType = 'estimatedCounts'
+ECOUNT_OBJ = synStore(ECOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 

### 2017-08-10 code/Rmd/ROSMAP_REPROCESSED.Rmd
Added ApoE specific differential expression in Mayo data

@@
 annotations(ECOUNT_OBJ) = annotations(CODE)
-annotations(ECOUNT_OBJ)$dataSubType = 'filteredCounts'
+annotations(ECOUNT_OBJ)$dataSubType = 'estimatedCounts'
 ECOUNT_OBJ = synStore(ECOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

### 2017-08-11 code/Rmd/MSSM_REPROCESSED.Rmd
Added ApoE differential expression to MSSM processing markdown

@@
 METADATA = METADATA %>%
+  dplyr::mutate(BrainRegion1 = BrainRegion) %>%
   tidyr::unite(BrainRegion.Diagnosis, BrainRegion, Diagnosis, sep = '.')
@@
   dplyr::filter(RIN >= 5)
+
+METADATA$ApoE[METADATA$Apo1 != 4 & METADATA$Apo2 != 4] = 0
+METADATA$ApoE[METADATA$Apo1 == 4 | METADATA$Apo2 == 4] = 1
+METADATA$ApoE[METADATA$Apo1 == 4 & METADATA$Apo2 == 4] = 2
+METADATA$BrainRegion.ApoE = paste(METADATA$BrainRegion1, METADATA$ApoE, sep = '.') %>%
+  factor()
 ```
@@
 ```{r covariates.clustering}
-primaryVariable <- c("CDR", "BrainRegion.Diagnosis")
-FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrainRegion.Diagnosis", "CDR")
+primaryVariable <- c("CDR", "BrainRegion.Diagnosis", "BrainRegion.ApoE")
+FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrainRegion.Diagnosis", "CDR", "BrainRegion.ApoE")
 ContCovariates <- c("RIN", "RIN2", "AOD","PMI",  "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
@@
 Distribution of samples are: `r kable(tmp)`
+
 ### Winsorise counts (Gene outliers in samples)
@@
 CQN.GENE_EXPRESSION$E = LOG.CPM
-
-PROCESSED_COUNTS$filteredExprMatrix$counts = NEW.COUNTS
+NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
 ```
@@
   # Estimate voom weights for dispersion control
-  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)
-
-  # Calculate correlation values of random effects
+  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F)
+
+  # # Calculate correlation values of random effects
   correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
                                              block = COVARIATES$individualIdentifier)
+  # correlation = list(); correlation$cor = 0.4
 
   # Re-estimate voom weights
-  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, 
                               design=DM1$design, plot=F, 
@@
   loopCount = loopCount + 1
-  print(loopCount)
+  # print(loopCount)
 }
-RESIDUAL.GENE_EXPRESSION
 modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
@@
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
 
@@
 
+### Differential expression analysis (with BrainRegion.ApoE as primary variable)
+Differential expression is performed on the primary variable by controlling for covariates identified above
+
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diff.exp2, fig.height=15, fig.width=13}
+# Get design matrix
+DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[3], postAdjustCovars), drop = F], Intercept = F)
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
+
+# Calculate correlation values of random effects
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)
+
+# Re-estimate voom weights
+VOOM.WEIGHTS = voom(NEW.COUNTS,
+                    design=DESIGN$design, plot=F,
+                    block = COVARIATES$individualIdentifier, 
+                    correlation = correlation$cor)
+
+# Fit linear model using new weights and new design
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+FIT = lmFit(VOOM.WEIGHTS)
+
+# Fit contrast
+cntr = c()
+for (region in c('FP', 'STG', 'PHG', 'IFG')){
+  tmp = combn(paste0('BrainRegion.ApoE', region, '.', c('0', '1', '2')),2)
+  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
+}
+
+contrast = makeContrasts(contrasts=cntr,
+                         levels = colnames(FIT$coefficients))
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = T) %>%
+    rownameToFirstColumn('ensembl_gene_id')
+}, FIT.CONTR)
+names(DE) = colnames(contrast)
+
+DE = DE %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Comparison = gsub('BrainRegion.ApoE','',Comparison),
+                Study = 'MSSM',
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
+  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
+  tidyr::separate(to.state, c('BrainRegion1','against'), sep = '\\.') %>%
+  dplyr::mutate(Comparison = paste(reference, against, sep = '-')) %>%
+  dplyr::select(-BrainRegion1) %>%
+  left_join(GENE.PARAM %>%
+              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
+              unique())
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion) %>%
+  group_by(BrainRegion, Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05_FC_1.2) 
+kable(tmp)
+
+p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position='top')
+p = p + facet_grid(BrainRegion+.~Comparison, scales = 'fixed')
+p
+
+all.diff.exp = c(all.diff.exp, list(ApoE = DE))
+all.fit = c(all.fit, list(ApoE = FIT))
+```
+
 ### Differential expression analysis (with BrainRegion.Diagnosis and Gender)
@@
 Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
-```{r diff.exp2, fig.height=15, fig.width=13}
+```{r diff.exp3, fig.height=15, fig.width=13}
 ## Modeling Diagnosis, msex and age_death conjointly
@@
 Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
-```{r diff.exp3, fig.height=15, fig.width=13}
+```{r diff.exp4, fig.height=15, fig.width=13}
 # Get design matrix
@@
 ```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
-# Set annotations
-all.annotations = list(
-  dataType = 'mRNA',
-  dataSubType = 'geneExp',
-  summaryLevel = 'gene',
-  assay	 = 'RNAseq',
-  assaySubType = 'Reprocessed',
-
-  tissueTypeAbrv	= 'FP, STG, PHG, IFG', 
-  study = 'MSSMRNASeq', 
-  center = 'MSBB',
-
-  organism = 'HomoSapiens',
-  consortium	= 'AMP-AD',
-
-  normalizationStatus	= TRUE,
-  modelSystem	= FALSE,
-
-  normalizationType	= 'CQN'
-)
-
 # Function to change entity ACL (also creates new acls if it doesn't exist)
@@
 
+# Set annotations
+all.annotations = list(
+  dataType = 'mRNA',
+  dataSubType = 'geneExp',
+  summaryLevel = 'gene',
+  assay	 = 'RNAseq',
+  assaySubType = 'Reprocessed',
+
+  tissueTypeAbrv	= 'FP, STG, PHG, IFG', 
+  study = 'MSSMRNASeq', 
+  center = 'MSBB',
+
+  organism = 'HomoSapiens',
+  consortium	= 'AMP-AD',
+
+  normalizationStatus	= TRUE,
+  modelSystem	= FALSE,
+
+  normalizationType	= 'CQN'
+)
+
 # Create new resource acls
@@
 
+# Store filtered counts
+NEW.COUNTS %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  write.table(file = 'MSSM_FP_STG_PHG_IFG_eCounts.tsv', sep = '\t', row.names=F, quote=F)
+ECOUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_eCounts.tsv', name = 'Counts (estimated)', parentId = CODE$properties$id)
+annotations(ECOUNT_OBJ) = annotations(CODE)
+annotations(ECOUNT_OBJ)$dataSubType = 'estimatedCounts'
+ECOUNT_OBJ = synStore(ECOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
 # Store logCPM
-CQN.GENE_EXPRESSION$E %>%
+CQN.GENE_EXPRESSION$y %>%
   rownameToFirstColumn('ensembl_gene_id') %>%

### 2017-08-12 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_qSVA.Rmd
Added markdown to process HBCC data with qSVA variables

--- 2017-08-08 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
+knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_qSVA.Rmd",
                                  parentId ="syn7501880",
-                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
+                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - qSVA')
 ```
@@
 library(biomaRt)
+library(sva)
 
@@
 levels(METADATA$RibozeroBatch)[1] = 'NotAvailable'
+
+# Get quality surrogate variables
+qSV_DEG_MAT_ID = 'syn10284941'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = qSV_DEG_MAT_ID
+qSV_DEG_MAT = downloadFile(qSV_DEG_MAT_ID)
+tmp = t(qSV_DEG_MAT[,-(1)])
+colnames(tmp) = qSV_DEG_MAT$V1
+qSV_DEG_MAT = tmp[,colnames(tmp) %in% colnames(COUNT)]
+rownames(qSV_DEG_MAT) = rownames(tmp)
+qSV = sva::qsva(qSV_DEG_MAT)
+colnames(qSV) = paste0('qSV', 1:dim(qSV)[2])
+
+METADATA = METADATA %>%
+  left_join(qSV %>% rownameToFirstColumn('SampleID'))
 ```
@@
 METADATA <- METADATA  %>% dplyr::filter(Age_of_Death >= 18)
+
+writeLines('Remove samples missing qSVs')
+writeLines(paste('Following',length(METADATA$SampleID[is.na(METADATA$qSV1)]), 'samples are removed:'))
+writeLines(paste(METADATA$SampleID[is.na(METADATA$qSV1)], collapse = ', '))
+METADATA <- METADATA  %>% dplyr::filter(!is.na(qSV1))
 ```
@@
                     "IntronicRate", "rRNARate", "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency",   
-                    "PercentMapped")
+                    "PercentMapped", "qSV1", "qSV2", "qSV3", "qSV4")
 
@@
 ```{r custom.normalisation, results='asis'}
-adjust.covars = c('Dx.Tissue', 'Gender', 'RIN', 'IntronicRate')
+adjust.covars = c('Dx.Tissue', 'Gender', 'RIN', 'IntronicRate', 'qSV1', 'qSV2', 'qSV3', 'qSV4')
 writeLines(paste('Using following covariates in the model:',
@@
 COVARIATES$Dx.Tissue.Gender = factor(paste(COVARIATES$Dx.Tissue, COVARIATES$Gender, sep = '_'))
-postAdjustCovars = c('Dx.Tissue.Gender', 'RIN', 'IntronicRate')
+postAdjustCovars = c('Dx.Tissue.Gender', 'RIN', 'IntronicRate', 'qSV1', 'qSV2', 'qSV3', 'qSV4')
 
@@
 ```{r diffExp2, fig.height=10, fig.width=6}
-postAdjustCovars = c('Dx.Tissue', 'Gender', 'RIN', 'IntronicRate')
+postAdjustCovars = c('Dx.Tissue', 'Gender', 'RIN', 'IntronicRate', 'qSV1', 'qSV2', 'qSV3', 'qSV4')
 
@@
 
-  normalizationType	= 'CQN'
+  normalizationType	= 'CQNandQSVA'
 )
@@
 # Code
-CODE <- Folder(name = "HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
+CODE <- Folder(name = "HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - qSVA", parentId = parentId)
 annotations(CODE) = all.annotations

### 2017-08-12 code/Rmd/ALL_REPROCESSED.Rmd
Minor changes

@@
                 msex = factor(msex, levels = c('1', '0'), labels = c('1'='MALE','0'='FEMALE')),
-                Batch = factor(Batch)) %>%
+                Batch = factor(Batch),
+                BrainRegion.ApoE = paste('DLPFC', apoe_genotype, sep = '.')) %>%
   dplyr::rename(Gender = msex, RIN = RINcontinuous, AOD = age_death, PMI = pmi) %>%
   dplyr::select(SampleID, BrainRegion.Diagnosis, Batch, Gender, AOD, PMI, RIN, PCT_PF_READS_ALIGNED, 
-                PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES)
+                PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES, BrainRegion.ApoE)
 covariates$ROSMAP$individualIdentifier = factor(paste0('ROSMAPpseudoid',1:dim(covariates$ROSMAP)[1]))
@@
   dplyr::select(SampleID, BrainRegion.Diagnosis, individualIdentifier, Batch, Gender, AOD, PMI, RIN, 
-                PCT_PF_READS_ALIGNED, PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES)
+                PCT_PF_READS_ALIGNED, PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, 
+                PCT_RIBOSOMAL_BASES, BrainRegion.ApoE)
 
@@
   dplyr::select(SampleID, BrainRegion.Diagnosis, individualIdentifier, Batch, Gender, AOD, PMI, RIN, 
-                PCT_PF_READS_ALIGNED, PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES)
+                PCT_PF_READS_ALIGNED, PCT_CODING_BASES, PCT_INTERGENIC_BASES, PCT_INTRONIC_BASES, 
+                PCT_RIBOSOMAL_BASES, BrainRegion.ApoE)
+levels(covariates$MAYO$BrainRegion.Diagnosis) = c("CER.AD", "CER.CONTROL", "CER.OTHER", "TCX.AD", "TCX.CONTROL", "TCX.OTHER")
 
@@
   dplyr::mutate(BrainRegion = factor(BrainRegion, levels = c('DLPFC', 'FP', 'IFG', 'PHG', 'STG', 'CER', 'TCX')),
-                BrainRegion.Diagnosis = factor(paste(BrainRegion, Diagnosis, sep = '.')))
+                BrainRegion.Diagnosis = factor(paste(BrainRegion, Diagnosis, sep = '.')),
+                BrainRegion.ApoE = factor(as.character(BrainRegion.ApoE)))
 
@@
 FactorCovariates <- c("individualIdentifier", "Study", "BrainRegion.Diagnosis", "BrainRegion", "Diagnosis",
-                      "Gender", "Batch")
+                      "Gender", "Batch", "BrainRegion.ApoE")
 ContCovariates <- c("AOD", "PMI", "RIN", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES",
@@
 
+### Differential expression analysis (with BrainRegion.ApoE as primary variable)
+Differential expression is performed on the primary variable by controlling for covariates identified above
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diff.exp2, fig.height=7, fig.width=24}
+# Get design matrix
+DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.ApoE', setdiff(adjust.covars, 'BrainRegion.Diagnosis')), drop = F], 
+                         Intercept = F)
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
+
+# Calculate correlation values of random effects
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)
+
+# Re-estimate voom weights
+VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
+                    design=DESIGN$design, plot=F,
+                    block = COVARIATES$individualIdentifier, 
+                    correlation = correlation$cor)
+
+# Fit linear model using new weights and new design
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+FIT = lmFit(VOOM.WEIGHTS)
+
+# Fit contrast
+cntr = c()
+for (region in c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
+  tmp = combn(paste0('BrainRegion.ApoE', region, '_', c('0', '1', '2')),2)
+  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
+}
+
+contrast = makeContrasts(contrasts=cntr,
+                         levels = colnames(FIT$coefficients))
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = lapply(1:dim(contrast)[2], function(i, FIT.CONTR){
+  topTable(FIT.CONTR, coef=i, number = 50000, confint = T) %>%
+    rownameToFirstColumn('Gene.ID')
+}, FIT.CONTR)
+names(DE) = colnames(contrast)
+
+DE = DE %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Comparison = gsub('BrainRegion.ApoE','',Comparison),
+                Study = 'ALL',
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  left_join(GENE.PARAM %>%
+              dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
+              unique())
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
+  group_by(Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05_FC_1.2) 
+kable(tmp)
+
+p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position='top')
+p = p + facet_grid(.~Comparison, scales = 'fixed')
+p
+
+all.diff.exp = c(all.diff.exp, list(ApoE = DE))
+all.fit = c(all.fit, list(ApoE = FIT))
+```
+
 ### Differential expression analysis (with BrainRegion.Diagnosis and Gender)
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diff.exp2, fig.height=7, fig.width=24}
+```{r diff.exp3, fig.height=7, fig.width=24}
 ## Modeling Diagnosis, msex and age_death conjointly

### 2017-08-16 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_qSVA.Rmd
Changed qSVA degration matrix id for HBCC

@@
 # Get quality surrogate variables
-qSV_DEG_MAT_ID = 'syn10284941'
+qSV_DEG_MAT_ID = 'syn10315615'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = qSV_DEG_MAT_ID

### 2017-08-16 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_qSVA.Rmd
Added markdown to process CMC data with qSVA

--- 2017-07-28 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)"
+title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN and qSVA normlaisation)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
+knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_qSVA.Rmd",
                                  parentId = "syn7501880",
-                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
+                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - qSVA')
 ```
@@
 
-thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd'
+thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_qSVA.Rmd'
 
@@
 levels(METADATA$Dx) = c("Other", "Other", "Control", "SCZ")
+
+# Get quality surrogate variables
+qSV_DEG_MAT_DLPFC_ID = 'syn10284943'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = qSV_DEG_MAT_DLPFC_ID
+qSV_DEG_MAT_DLPFC = downloadFile(qSV_DEG_MAT_DLPFC_ID)
+
+qSV_DEG_MAT_ACC_ID = 'syn10284942'
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = qSV_DEG_MAT_ACC_ID
+qSV_DEG_MAT_ACC = downloadFile(qSV_DEG_MAT_ACC_ID)
+
+qSV_DEG_MAT = rbindlist(list(qSV_DEG_MAT_DLPFC, qSV_DEG_MAT_ACC), use.names = T, fill = T)
+tmp = t(qSV_DEG_MAT[,-(1)])
+colnames(tmp) = qSV_DEG_MAT$V1
+qSV_DEG_MAT = tmp[,colnames(tmp) %in% colnames(COUNT)]
+rownames(qSV_DEG_MAT) = rownames(tmp)
+
+qSV = sva::qsva(qSV_DEG_MAT)
+colnames(qSV) = paste0('qSV', 1:dim(qSV)[2])
+
+METADATA = METADATA %>%
+  left_join(qSV %>% rownameToFirstColumn('SampleID'))
 ```
@@
 METADATA <- METADATA  %>% filter(!is.na(METADATA$EV.1))
+
+writeLines('Remove samples missing qSVs')
+writeLines(paste('Following',length(METADATA$SampleID[is.na(METADATA$qSV1)]), 'samples are removed:'))
+writeLines(paste(METADATA$SampleID[is.na(METADATA$qSV1)], collapse = ', '))
+METADATA <- METADATA  %>% dplyr::filter(!is.na(qSV1))
 ```
@@
                     "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
-                    "PercentAligned", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")
+                    "PercentAligned", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5", "qSV1", "qSV2", "qSV3")
 
@@
 ```{r cpmnormalisation}
-genesToAnalyze = dlply(METADATA, .(Dx.Tissue), .fun = function(mtd, count){
-  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID %>% droplevels()],
+genesToAnalyze = COVARIATES %>% 
+  rownameToFirstColumn('SampleID') %>%
+  dlply(.(Dx.Tissue), .fun = function(mtd, count){
+  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                    MIN_GENE_CPM=1, 
@@
 ### Normalisation
-Based on different model selection methodology (as seen here syn8468116, syn9772932, and syn9774491), following variables are used as covariates: Institution, Gender, RIN, and IntronicRate
+Based on different model selection methodology (as seen here syn8468116, syn9772932, and syn9774491), following variables are used as covariates: Institution, Gender, RIN, and IntronicRate, with qSV 1 to 3
 
@@
 # Primary variable of interest
-postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'RIN', 'IntronicRate')
+postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'RIN', 'IntronicRate', 'qSV1', 'qSV2', 'qSV3')
 
@@
 COVARIATES$Dx.Tissue.Gender = factor(paste(COVARIATES$Dx.Tissue, COVARIATES$Gender, sep = '_'))
-postAdjustCovars = c('Dx.Tissue.Gender', 'Institution', 'RIN', 'IntronicRate')
+postAdjustCovars = c('Dx.Tissue.Gender', 'Institution', 'RIN', 'IntronicRate', 'qSV1', 'qSV2', 'qSV3')
 
@@
                 Tissue = factor(Tissue))
-postAdjustCovars = c('Tissue', 'Dx', 'Institution', 'Gender', 'RIN', 'IntronicRate')
+postAdjustCovars = c('Tissue', 'Dx', 'Institution', 'Gender', 'RIN', 'IntronicRate', 'qSV1', 'qSV2', 'qSV3')
 
@@
 
-  normalizationType	= 'CQN'
+  normalizationType	= 'qSVA'
 )
@@
 # Code
-CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
+CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - qSVA", parentId = parentId)
 annotations(CODE) = all.annotations

### 2017-08-26 code/Rmd/CMC_HBCC_COMPARISON.Rmd
Added markdown to comapre cmc and hbcc dlpfc differential expression

---
title: "Compare CMC and HBCC differential expression"
author: "Thanneer M Perumal"
date: "`r base::date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE, include=FALSE, echo=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./CMC_HBCC_COMPARISON.Rmd",
                                 parentId = "syn7501880",
                                 entityName = 'CMC - HBCC - Reproducibility Analysis')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(limma)
library(psych)
library(edgeR)
library(cqn)
library(biomaRt)

library(synapseClient)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

synapseLogin()

library(doParallel)
library(foreach)

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE, echo=FALSE}
parentId = 'syn8672415';
activityName = 'Reproducibility Analysis b/w cmc and hbcc samples';
activityDescription = 'Reproducibility Analysis b/w cmc and hbcc samples';

thisFileName <- 'CMC_HBCC_COMPARISON.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```

### Data download
```{r download.data, cache=TRUE}
# Get differential expression results
diff.exp.ids = c(CMC = 'syn9874406',
                 CMC_qSVA = 'syn10341319',
                 HBCC = 'syn10263511',
                 HBCC_qSVA = 'syn10294725')

diff.exp = lapply(diff.exp.ids, downloadFile) 
diff.exp$HBCC = diff.exp$HBCC %>%
  tidyr::unite(a, to.Dx, to.Tissue, sep = '_') %>%
  tidyr::unite(b, ref.Dx, ref.Tissue, sep = '_') %>%
  tidyr::unite(Comparison, a, b, sep = '-')
diff.exp$HBCC_qSVA = diff.exp$HBCC_qSVA %>%
  tidyr::unite(a, to.Dx, to.Tissue, sep = '_') %>%
  tidyr::unite(b, ref.Dx, ref.Tissue, sep = '_') %>%
  tidyr::unite(Comparison, a, b, sep = '-')

diff.exp = data.table::rbindlist(diff.exp, use.names = T, fill = T, idcol = 'Model')
```

### Analyse results
```{r anlz.results}
rep.rate = list()
for (pval in seq(0,-6,-1)){
  de.genes = diff.exp %>%
    dplyr::filter(Comparison == 'SCZ_DLPFC-Control_DLPFC') %>%
    dplyr::mutate(Direction = factor(sign(logFC), levels = c('-1','1'), labels = c('-1' = 'DOWN','1' = 'UP'))) %>%
    plyr::dlply(.(Model, Direction), .fun = function(x){
      x$ensembl_gene_id[x$adj.P.Val <= 10^pval] %>% unique()
    })

  l1 = length(intersect(de.genes$CMC.DOWN, de.genes$HBCC.DOWN))
  l2 = length(intersect(de.genes$CMC.UP, de.genes$HBCC.UP))
  l3 = length(unique(c(de.genes$CMC.UP, de.genes$HBCC.UP, de.genes$CMC.DOWN, de.genes$HBCC.DOWN)))
  wo.qsva = (l1+l2)/l3
  n.wo.qsva = l1+l2

  l1 = length(intersect(de.genes$CMC_qSVA.DOWN, de.genes$HBCC_qSVA.DOWN))
  l2 = length(intersect(de.genes$CMC_qSVA.UP, de.genes$HBCC_qSVA.UP))
  l3 = length(unique(c(de.genes$CMC_qSVA.UP, de.genes$HBCC_qSVA.UP, de.genes$CMC_qSVA.DOWN, de.genes$HBCC_qSVA.DOWN)))
  w.qsva = (l1+l2)/l3
  n.w.qsva = l1+l2

  rep.rate = c(rep.rate, list(data.frame(fdr = 10^pval, wo.qsva = wo.qsva, w.qsva = w.qsva,
                                         n.wo.qsva = n.wo.qsva, n.w.qsva = n.w.qsva)))
}
rep.rate = rbindlist(rep.rate)
rep.rate %>%
  dplyr::select(fdr, wo.qsva, w.qsva) %>%
  tidyr::gather(method, rep.rate, -fdr) %>%
  dplyr::mutate(rep.rate = rep.rate * 100) %>%
  dplyr::left_join(rep.rate %>%
                     dplyr::select(fdr, n.wo.qsva, n.w.qsva) %>%
                     tidyr::gather(method, ncount, -fdr) %>%
                     dplyr::mutate(method = gsub('n.','',method))) %>%
  dplyr::filter(fdr <= 1e-1) %>%
  ggplot(aes(x=fdr, y=rep.rate, size = ncount, color = method)) + geom_point() + geom_line() + scale_x_log10() + xlab('FDR') + ylab('Replication Rate (%)')
```

```{r anlz.results1}
rep.rate = list()
for (pval in c(-1, -1.30103, -2)){
  de.genes = diff.exp %>%
    dplyr::filter(Comparison == 'SCZ_DLPFC-Control_DLPFC') %>%
    plyr::dlply(.(Model), .fun = function(x){
      x$ensembl_gene_id[x$adj.P.Val <= 10^pval] %>% unique()
    })

  tmp = diff.exp %>%
    dplyr::filter(Comparison == 'SCZ_DLPFC-Control_DLPFC', Model == 'CMC', ensembl_gene_id %in% de.genes$HBCC) %>%
    dplyr::select(P.Value) %>%
    qvalue::qvalue(fdr.level = 10^pval)
  rs = data.frame(method = 'wo.qsva', ref.data = 'HBCC', rep.data = 'CMC',
                  rep.rate = sum(tmp$significant)/length(tmp$significant), 
                  ncount = sum(tmp$significant))

  tmp = diff.exp %>%
    dplyr::filter(Comparison == 'SCZ_DLPFC-Control_DLPFC', Model == 'HBCC', ensembl_gene_id %in% de.genes$CMC) %>%
    dplyr::select(P.Value) %>%
    qvalue::qvalue(fdr.level = 10^pval)
  rs = rbind(rs,
             data.frame(method = 'wo.qsva', ref.data = 'CMC', rep.data = 'HBCC',
                        rep.rate = sum(tmp$significant)/length(tmp$significant), 
                        ncount = sum(tmp$significant)))

  tmp = diff.exp %>%
    dplyr::filter(Comparison == 'SCZ_DLPFC-Control_DLPFC', Model == 'CMC_qSVA', ensembl_gene_id %in% de.genes$HBCC_qSVA) %>%
    dplyr::select(P.Value) %>%
    qvalue::qvalue(fdr.level = 10^pval)
  rs = rbind(rs,
             data.frame(method = 'w.qsva', ref.data = 'HBCC', rep.data = 'CMC',
                        rep.rate = sum(tmp$significant)/length(tmp$significant), 
                        ncount = sum(tmp$significant)))

  tmp = diff.exp %>%
    dplyr::filter(Comparison == 'SCZ_DLPFC-Control_DLPFC', Model == 'HBCC_qSVA', ensembl_gene_id %in% de.genes$CMC_qSVA) %>%
    dplyr::select(P.Value) %>%
    qvalue::qvalue(fdr.level = 10^pval)
  rs = rbind(rs,
             data.frame(method = 'w.qsva', ref.data = 'CMC', rep.data = 'HBCC',
                        rep.rate = sum(tmp$significant)/length(tmp$significant), 
                        ncount = sum(tmp$significant)))

  rs$fdr = 10^pval
  rep.rate = c(rep.rate, list(rs))
}
rep.rate = rbindlist(rep.rate)
rep.rate %>%
  ggplot(aes(x=fdr, y=rep.rate*100, color = method)) + geom_point(aes(size = ncount)) + geom_line() + scale_x_log10() + xlab('FDR') + ylab('Replication Rate (%)') + facet_grid(rep.data~., scales = 'free_y')
```

### 2017-08-26 code/Rmd/CMC_HBCC_COMPARISON.Rmd
Minor modifications

@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_HBCC_COMPARISON.Rmd",
+knit2synapse::storeAndKnitToFileEntity(file = "./CMC_HBCC_COMPARISON.Rmd",
                                        parentId = "syn7501880",
@@
 ### Analyse results
+Replication rate based on number of gene differentiallty expressed in the same directiona at an FDR of 0.05
 ```{r anlz.results}
@@
   dplyr::filter(fdr <= 1e-1) %>%
-  ggplot(aes(x=fdr, y=rep.rate, size = ncount, color = method)) + geom_point() + geom_line() + scale_x_log10() + xlab('FDR') + ylab('Replication Rate (%)')
+  ggplot(aes(x=fdr, y=rep.rate, color = method)) + geom_point(aes(size = ncount)) + geom_line() + scale_x_log10() + xlab('FDR') + ylab('Replication Rate (%)')
 ```
-
+Replication rate based on qvalue
 ```{r anlz.results1}

### 2017-08-28 code/Rmd/CMC_HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Added CMC HBCC age matched analysis

---
title: "Covariate analysis of CMC and HBCC matched RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./CMC_HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd",
                                 parentId = "syn7501880",
                                 entityName = 'CMC and HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(ggpubr)

library(reshape2)
library(limma)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)
library(cqn)

library(synapseClient)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

synapseLogin()

library(doParallel)
library(foreach)

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = "syn7501880";
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (CMC and HBCC DLPFC)';

thisFileName <- 'CMC_HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download reprocessed counts (CMC and HBCC)
COUNT_CMC_ID = 'syn9874379'
ALL_USED_IDs = COUNT_CMC_ID
COUNT_CMC = downloadFile(COUNT_CMC_ID)

COUNT_HBCC_ID = 'syn10263483';
ALL_USED_IDs[length(ALL_USED_IDs)+1] = COUNT_HBCC_ID
COUNT_HBCC = downloadFile(COUNT_HBCC_ID)

# Join HBCC and CMC together
COUNT = join_all(list(COUNT_CMC, COUNT_HBCC), type = 'full') %>%
  as.data.frame()

rownames(COUNT) = COUNT$ensembl_gene_id
COUNT$ensembl_gene_id = NULL

# Get covariates (CMC and HBCC)
COVARIATES_CMC_ID = 'syn9874376'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = COVARIATES_CMC_ID
COVARIATES_CMC = downloadFile(COVARIATES_CMC_ID) %>%
  dplyr::mutate(SampleID = colnames(COUNT_CMC)[-(1)],
                LibraryBatch = paste0('CMC',LibraryBatch),
                RibozeroBatch = paste0('CMC',RibozeroBatch),
                FlowcellBatch = paste0('CMC',FlowcellBatch),
                Dx.Tissue = paste(Dx, Tissue, sep=  '_')) %>%
  dplyr::select(-TotalReads, -PercentAligned, -EV.1, -EV.2, -EV.3, -EV.4, -EV.5, -RIN2, -Dx.Tissue.Gender,
                -Dx, -Tissue)

COVARIATES_HBCC_ID = 'syn10263480'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = COVARIATES_HBCC_ID
COVARIATES_HBCC = downloadFile(COVARIATES_HBCC_ID) %>%
  dplyr::mutate(LibraryBatch = paste0('HBCC',LibraryBatch),
                RibozeroBatch = paste0('HBCC',RibozeroBatch),
                FlowcellBatch = paste0('HBCC',FlowcellBatch),
                Institution = 'HBCC',
                Individual_ID = seq(1,length(Institution),1),
                IntronicRate = IntronicRate * 100,
                IntergenicRate = IntergenicRate * 100,
                rRNARate = rRNARate * 100) %>%
  dplyr::rename(MappedReads = Mapped) %>%
  dplyr::select(-TotalReads, -X5prime_Norm, -Mean_Per_Base_Cov., -Mean_CV, -Fragment_Length_StdDev,
                -Mapped_Pairs, -End_1_Mismatch_Rate, -End_2_Mismatch_Rate, -ExonicRate,
                -PercentMapped, -RIN2, -Dx.Tissue.Gender)

COVARIATES = rbindlist(list(CMC = COVARIATES_CMC, 
                            HBCC = COVARIATES_HBCC), 
                       use.names = T, fill = T, idcol = 'Study')
```

### Data preprocessing
Filter DLPFC SCZ and Control samples alone
```{r preprocess.data}
# Filter all ACC samples and Other samples
COVARIATES = COVARIATES %>%
  dplyr::filter(Dx.Tissue %in% c("SCZ_DLPFC", "Control_DLPFC"))

COVARIATES %>%
  dplyr::group_by(Dx.Tissue, Study) %>%
  dplyr::summarise(count = n()) %>%
  tidyr::spread(Study, count) %>% 
  kable()

my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1),
                  plot.title=element_text(hjust=0.5))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN, color = Study)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = Age_of_Death, color = Study)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI, color = Study)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme

# Intronic Rate
p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate, color = Study)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme

# IntergenicRate
p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate, color = Study)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme

# Transcripts Detected
p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TranscriptsDetected, color = Study)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Transcripts Detected') + my.theme

# Mapped Reads
p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads, color = Study)) + geom_boxplot()
p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme

# rRNARate
p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate, color = Study)) + geom_boxplot()
p[[8]] = p[[8]] + ggtitle('rRNARate') + my.theme

ggpubr::ggarrange(plotlist = p, ncol = 4, nrow = 2)
```

Age match CMC and HBCC (per Dx.Tissue category)
```{r preprocess.data1}
# Filter all ACC samples and Other samples
tmp = COVARIATES %>%
  dplyr::group_by(Dx.Tissue, Study) %>%
  dplyr::summarise(mn = quantile(Age_of_Death, 0.25), mx = quantile(Age_of_Death, 0.75)) %>%
  dplyr::group_by(Dx.Tissue) %>%
  dplyr::summarise(mn = max(mn), mx = min(mx))
COVARIATES = COVARIATES %>%
  dplyr::filter((Dx.Tissue == tmp$Dx.Tissue[1] & Age_of_Death >= tmp$mn[1] & Age_of_Death <= tmp$mx[1]) |
                  (Dx.Tissue == tmp$Dx.Tissue[2] & Age_of_Death >= tmp$mn[2] & Age_of_Death <= tmp$mx[2]))

COVARIATES %>%
  dplyr::group_by(Dx.Tissue, Study) %>%
  dplyr::summarise(count = n()) %>%
  tidyr::spread(Study, count) %>% 
  kable()
```

```{r gene.param}
## Get GC content from biomart
backgroundGenes = data.frame(ensembl_gene_id = rownames(COUNT))

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "dec2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content", "gene_biotype"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::select(ensembl_gene_id, percentage_gc_content) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$ensembl_gene_id

# Get gene lengths from synapse
GENE.LEN = downloadFile('syn8414424') %>%
  dplyr::select(Geneid, Length) %>%
  dplyr::filter(Geneid %in% rownames(COUNT)) %>%
  unique() %>%
  dplyr::rename(ensembl_gene_id = Geneid)
rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id
```

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Study', 'Individual_ID', 'Institution', 'Gender', 'LibraryBatch', 'Dx.Tissue', 'RibozeroBatch',
                      'FlowcellBatch')
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate")

# Find inter relation between factor covariates
rownames(COVARIATES) = COVARIATES$SampleID
COVARIATES = COVARIATES[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){
  x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], function(x){
  x = as.numeric(as.character(gsub('[\\,\\%]','',x)))
})

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=8}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
tmp = COVARIATES.CORRELATION$ESTIMATE
tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
draw(h, heatmap_legend_side = 'left')
```

### Explore metatdata
```{r data.explore, fig.width = 12, fig.height = 12}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN, color = Study)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = Age_of_Death, color = Study)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI, color = Study)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme

# Intronic Rate
p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate, color = Study)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme

# IntergenicRate
p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate, color = Study)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme

# Transcripts Detected
p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TranscriptsDetected, color = Study)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Transcripts Detected') + my.theme

# Mapped Reads
p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads, color = Study)) + geom_boxplot()
p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme

# rRNARate
p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate, color = Study)) + geom_boxplot()
p[[8]] = p[[8]] + ggtitle('rRNARate') + my.theme

ggpubr::ggarrange(plotlist = p, ncol = 4, nrow = 2)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
COUNT[is.na(COUNT)] = 0
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('SampleID') %>%
  dlply(.(Dx.Tissue), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                     MIN_GENE_CPM=1, 
                                                     MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
    processed.counts$filteredExprMatrix$genes
  }, COUNT)

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC.CONT$ensembl_gene_id[!is.na(GENE.GC.CONT$percentage_gc_content)]) %>%
  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$Length)])

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, intersect(colnames(COUNT), rownames(COVARIATES))],
                                                 MIN_GENE_CPM=0,
                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
pct.pc = PROCESSED_COUNTS$filteredExprMatrix$genes %>%
  dplyr::rename(ensembl_gene_id = genes) %>%
  left_join(Ensemble2HGNC) %>%
  group_by(gene_biotype) %>%
  summarise(fraction  = n()) %>%
  filter(fraction > 100) %>%
  dplyr::mutate(fraction = fraction/length(PROCESSED_COUNTS$filteredExprMatrix$genes[,1]))
```
`r dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]` genes are used for the analysis.

Following is the distribution of biotypes
`r kable(pct.pc)`

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
# Find principal components of expression to plot
PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
  dplyr::mutate(label = SampleID)
plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p

tmp = tidyr::separate(COVARIATES, Dx.Tissue, c('Dx', 'Tissue'), sep = '_') %>%
  group_by(Study, Dx, Tissue) %>%
  summarise(count = n()) %>%
  spread(Dx, count)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals

Distribution of samples are:
  `r kable(tmp)`

### Winsorise counts (Gene outliers in samples)
```{r winsorise.data}
LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
LOG.CPM = apply(LOG.CPM, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))

  x[x < (mn-3*std.dev)] = x[ind.low]
  x[x > (mn+3*std.dev)] = x[ind.high]
  return(x)
}) %>% t

# Back calculate counts for further analysis
LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
```

### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
CQN.GENE_EXPRESSION$E = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))

  x[x < (mn-3*std.dev)] = x[ind.low]
  x[x > (mn+3*std.dev)] = x[ind.high]
  return(x)
}) %>% t

# Back calculate counts for further analysis
LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
NEW.COUNTS = (2^CQN.GENE_EXPRESSION$E) * t(replicate(dim(CQN.GENE_EXPRESSION$E)[1], LIB.SIZE))/1e6
```

### Co-expression distribution
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$E))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Sample clustering
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```
```{r temp, include = F}
dev.off()
gc()
```

### Distribution of samples (log cpm)
```{r lcpm.dist, cache=FALSE}
# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$E) %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID') %>%
              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'NONE') + facet_grid(Dx+.~Tissue, scale = 'free')
p
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation
Based on different model selection methodology (as seen here syn8468116, syn9772932, and syn9774491), following variables are used as covariates: Institution, Gender, RIN, and IntronicRate.

NOTE: Dx.Tissue.Study is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue)

```{r custom.normalisation, results='asis'}
# Primary variable of interest
COVARIATES$Dx.Tissue.Study = paste(COVARIATES$Dx.Tissue, COVARIATES$Study, sep='_') %>%
  factor
postAdjustCovars = c('Dx.Tissue.Study', 'Institution', 'Gender', 'RIN', 'IntronicRate')

writeLines(paste('Using following covariates in the model:',
                 paste(postAdjustCovars, collapse=', ')))

# Post adjusted design matrix
DESIGN = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
DESIGN = DESIGN[,linColumnFinder(DESIGN)$indepCols]

# Estimate voom weights
VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design = DESIGN)
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E

# Fit linear model using new weights and new design
ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
```

### Residual calculation
Calculate weighted residuals and add back "Dx.Tissue.Study" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue.Study", colnames(DESIGN), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DESIGN[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
  dplyr::mutate(label = SampleID)
# plotdata$label[!(plotdata$SampleID %in% top.nsamples)] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= label), size=4, hjust=0)
p
```

```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue','Institution')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue','Institution')]))
```

```{r temp1, include=F}
dev.off()
gc()
```

### Differential expression analysis (SCZ - Control)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=15, fig.width=20}
# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.Tissue.StudySCZ_DLPFC_CMC-Dx.Tissue.StudyControl_DLPFC_CMC",
                                     "Dx.Tissue.StudySCZ_DLPFC_HBCC-Dx.Tissue.StudyControl_DLPFC_HBCC"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR) 
names(DE) = gsub('Dx.Tissue.Study', '', colnames(contrast))

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
                Direction = sign(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(Ensemble2HGNC) %>%
  left_join(GENE.LEN)
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, logFC) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))

kable(full_join(tmp,tmp1))

p = DE %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(Comparison~., scales = 'free_y')
p = p + ggtitle('Differential Expression between Schizophrenia and Control')
p
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Set annotations
all.annotations = list(
dataType = 'mRNA',
dataSubType = 'geneExp',
summaryLevel = 'gene',
assay	 = 'RNAseq',

tissueTypeAbrv	= 'DLPFC', 
study = 'CMC and HBCC',

organism = 'HomoSapiens',
consortium	= 'CMC',

normalizationStatus	= TRUE,

normalizationType	= 'CQN'
)

# Code
CODE <- Folder(name = "CMC and HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'CMC_HBCC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('CMC_HBCC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
annotations(COV_OBJ) = annotations(CODE)
annotations(COV_OBJ)$dataSubType = 'covariates'
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
executed = thisFile, activityDescription = activityDescription)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_HBCC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_HBCC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store logCPM
CQN.GENE_EXPRESSION$y[is.na(CQN.GENE_EXPRESSION$E)] = NA
CQN.GENE_EXPRESSION$y %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_HBCC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('CMC_HBCC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store cqn offsets
CQN.GENE_EXPRESSION$offset[is.na(CQN.GENE_EXPRESSION$E)] = NA
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_HBCC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('CMC_HBCC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
annotations(OFFSET_OBJ) = annotations(CODE)
annotations(OFFSET_OBJ)$dataSubType = 'offset'
OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store design matrix
DESIGN %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'CMC_HBCC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('CMC_HBCC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
annotations(DM_OBJ) = annotations(CODE)
annotations(DM_OBJ)$dataSubType = 'designMatrix'
DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store differential expression results
write.table(all.diff.exp, file = 'CMC_HBCC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_HBCC_DLPFC_DiffExpression.tsv', 
name = 'Differential Expression Results (Dx.Tissue)', 
parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |

### Source Code
[R markdown](`r thisFile`)

### 2017-08-28 code/Rmd/CMC_HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Minor bug fix

@@
 Filter DLPFC SCZ and Control samples alone
-```{r preprocess.data}
+```{r preprocess.data, fig.height=10, fig.width=15}
 # Filter all ACC samples and Other samples
@@
 
-ggpubr::ggarrange(plotlist = p, ncol = 4, nrow = 2)
+p = ggpubr::ggarrange(plotlist = p, ncol = 4, nrow = 2) 
+p
 ```
@@
 ### Explore metatdata
-```{r data.explore, fig.width = 12, fig.height = 12}
+```{r data.explore, fig.width = 15, fig.height = 10}
 my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))
@@
 
-ggpubr::ggarrange(plotlist = p, ncol = 4, nrow = 2)
+p = ggpubr::ggarrange(plotlist = p, ncol = 4, nrow = 2)
+p
 ```

### 2017-08-29 code/Rmd/MAYO_REPROCESSED.Rmd
Minor modifications

@@
 ```{r outlier.analysis, cache = FALSE, fig.width = 10}
-indToRemove = c("11311_CER", "1923_CER", "1923_TCX", "11396_TCX", "11294_TCX", "11408_TCX")
+indToRemove = c("11311_CER", "1923_CER", "1923_TCX", "11396_TCX", "11294_TCX", "11408_TCX",
+                '1950_TCX', '1950_CER', '1925_TCX', '1957_CER')
 
@@
 # Store differential expression results
-write.table(DE, file = 'MAYO_CBE_TCX_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+rbindlist(all.diff.exp, use.names = T, fill = T, idcol = 'Model') %>%
+  write.table(file = 'MAYO_CBE_TCX_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
 DEXP_OBJ = File('MAYO_CBE_TCX_DiffExpression.tsv', 

### 2017-08-29 code/Rmd/MSSM_REPROCESSED.Rmd
Minor modifications

@@
   write.table(file = 'MSSM_FP_STG_PHG_IFG_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('MSSM_FP_STG_PHG_IFG_DiffExpression.tsv', name = 'Differential Expression Results (cogdx)', 
+DEXP_OBJ = File('MSSM_FP_STG_PHG_IFG_DiffExpression.tsv', name = 'Differential Expression Results (BrainRegion.Diagnosis)', 
                 parentId = CODE$properties$id)

### 2017-08-30 code/Rmd/CMC_HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Changed the logic of age and PMI matching criteria

@@
 
-p = ggpubr::ggarrange(plotlist = p, ncol = 4, nrow = 2) 
-p
+p1 = ggpubr::ggarrange(plotlist = p, ncol = 4, nrow = 2) 
+p1
 ```
@@
 # Filter all ACC samples and Other samples
-tmp = COVARIATES %>%
+max.age = COVARIATES %>%
   dplyr::group_by(Dx.Tissue, Study) %>%
-  dplyr::summarise(mn = quantile(Age_of_Death, 0.25), mx = quantile(Age_of_Death, 0.75)) %>%
+  dplyr::filter(Study == 'CMC') %>%
+  dplyr::summarise(mn = quantile(Age_of_Death, 0.5))
+
+COVARIATES = COVARIATES %>%
+  dplyr::filter((Dx.Tissue == max.age$Dx.Tissue[1] & Age_of_Death <= max.age$mn[1]) |
+                  (Dx.Tissue == max.age$Dx.Tissue[2] & Age_of_Death <= max.age$mn[2]))
+
+COVARIATES %>%
+  dplyr::group_by(Dx.Tissue, Study) %>%
+  dplyr::summarise(count = n()) %>%
+  tidyr::spread(Study, count) %>% 
+  kable()
+```
+
+PMI match CMC and HBCC (per Dx.Tissue category)
+```{r preprocess.data2}
+max.pmi = COVARIATES %>%
   dplyr::group_by(Dx.Tissue) %>%
-  dplyr::summarise(mn = max(mn), mx = min(mx))
+  dplyr::filter(Study == 'HBCC') %>%
+  dplyr::summarise(mn = quantile(PMI, 0.5))
+
 COVARIATES = COVARIATES %>%
-  dplyr::filter((Dx.Tissue == tmp$Dx.Tissue[1] & Age_of_Death >= tmp$mn[1] & Age_of_Death <= tmp$mx[1]) |
-                  (Dx.Tissue == tmp$Dx.Tissue[2] & Age_of_Death >= tmp$mn[2] & Age_of_Death <= tmp$mx[2]))
+  dplyr::filter((Dx.Tissue == max.age$Dx.Tissue[1] & PMI <= max.pmi$mn[1]) |
+                  (Dx.Tissue == max.age$Dx.Tissue[2] & PMI <= max.pmi$mn[2]))
 
@@
 
-p = ggpubr::ggarrange(plotlist = p, ncol = 4, nrow = 2)
-p
+p1 = ggpubr::ggarrange(plotlist = p, ncol = 4, nrow = 2)
+p1
 ```
@@
 ```
+
+### Iterative Normalisation
+Dx.Tissue.Study is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue)
+```{r iterative.normalisation}
+# Primary variable of interest
+COVARIATES$Dx.Tissue.Study = paste(COVARIATES$Dx.Tissue, COVARIATES$Study, sep = '_') %>%
+  factor()
+postAdjustCovars = c('Dx.Tissue.Study');
+avoid.covariates = c('Dx.Tissue', 'Study', 'Individual_ID')
+
+# Assign residual covariates
+residualCovars = setdiff(preAdjustedSigCovars$significantCovars, 
+                         c(postAdjustCovars, avoid.covariates))
+residualSigCovars = preAdjustedSigCovars
+covariatesEffects = preAdjustedSigCovars$Effects.significantCovars[residualCovars]
+postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
+
+loopCount = 0 
+while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 20){
+  writeLines(paste('Using following covariates in the model:',
+                   paste(postAdjustCovars, collapse=', ')))
+
+  # Post adjusted design matrix
+  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
+  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
+
+  # Estimate voom weights for dispersion control
+  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F)
+  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
+
+  # Fit linear model using new weights and new design
+  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$E,
+                            design = DM1$design,
+                            weights = VOOM.GENE_EXPRESSION$weights)
+
+  # Residuals after normalisation
+  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
+                                                CQN.GENE_EXPRESSION$E)
+
+  # Residual covariates to choose from
+  residCovars <- setdiff(c(FactorCovariates,ContCovariates),
+                         c(postAdjustCovars, avoid.covariates))
+
+  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
+  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+                                                COVARIATES[, residCovars, drop=F], 
+                                                'adjusted design(voom-normalized)',
+                                                isKeyPlot=TRUE)
+
+  # Add postadjusted covariates (if any)
+  residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, avoid.covariates))
+  covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]
+
+  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
+  loopCount = loopCount + 1
+}
+
+modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
+                  'as fixed effects')
+
+tmp <- paste('Using following covariates in the final model:', modelStr)
+```
+
 ### Normalisation
-Based on different model selection methodology (as seen here syn8468116, syn9772932, and syn9774491), following variables are used as covariates: Institution, Gender, RIN, and IntronicRate.
-
-NOTE: Dx.Tissue.Study is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue)
-
 ```{r custom.normalisation, results='asis'}
-# Primary variable of interest
-COVARIATES$Dx.Tissue.Study = paste(COVARIATES$Dx.Tissue, COVARIATES$Study, sep='_') %>%
-  factor
-postAdjustCovars = c('Dx.Tissue.Study', 'Institution', 'Gender', 'RIN', 'IntronicRate')
-
 writeLines(paste('Using following covariates in the model:',
@@
 p
+
+all.diff.exp = list(Diagnosis.Study = DE)
+all.fit = list(Diagnosis.Study = ADJUSTED.FIT)
+```
+
+### Associate differential expression results with gc content, gene length and average expression
+```{r associate.de, fig.height=10, fig.width=15}
+pl = list()
+pl[[1]] = ggplot(DE, aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
+pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
+pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[1]] = pl[[1]] + theme(legend.position = 'top')
+
+pl[[2]] = ggplot(DE, aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
+pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
+pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[2]] = pl[[2]] + theme(legend.position = 'top')
+
+pl[[3]] = ggplot(DE, aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
+pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
+pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[3]] = pl[[3]] + theme(legend.position = 'top')
+
+multiplot(plotlist = pl, cols = 3)
+```
+
+### Differential expression analysis (SCZ - Control gender specific)
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diffExp1, fig.height=15, fig.width=20}
+COVARIATES$Dx.Tissue.Study.Gender = paste(COVARIATES$Dx.Tissue.Study,
+                                          COVARIATES$Gender,
+                                          sep = '_') %>%
+  factor()
+
+postAdjustCovars = c('Dx.Tissue.Study.Gender', setdiff(postAdjustCovars, c('Dx.Tissue.Study', 'Gender')))
+writeLines(paste('Using following covariates in the model:', paste(postAdjustCovars, collapse=', ')))
+
+# Post adjusted design matrix
+DESIGN = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
+DESIGN = DESIGN[,linColumnFinder(DESIGN)$indepCols]
+
+# Estimate voom weights
+VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design = DESIGN)
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
+
+# Fit linear model using new weights and new design
+ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
+
+# Fit contrast
+contrast = makeContrasts(contrasts=
+                           c("Dx.Tissue.Study.GenderSCZ_DLPFC_CMC_Female-Dx.Tissue.Study.GenderControl_DLPFC_CMC_Female",
+                             "Dx.Tissue.Study.GenderSCZ_DLPFC_CMC_Male-Dx.Tissue.Study.GenderControl_DLPFC_CMC_Male",
+                             "Dx.Tissue.Study.GenderSCZ_DLPFC_HBCC_Female-Dx.Tissue.Study.GenderControl_DLPFC_HBCC_Female",
+                             "Dx.Tissue.Study.GenderSCZ_DLPFC_HBCC_Male-Dx.Tissue.Study.GenderControl_DLPFC_HBCC_Male"),
+                         levels = colnames(ADJUSTED.FIT$coefficients))
+FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
+    rownameToFirstColumn('ensembl_gene_id') 
+}, FIT.CONTR) 
+names(DE) = gsub('Dx.Tissue.Study.Gender', '', colnames(contrast))
+
+DE = DE %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Direction = sign(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  left_join(Ensemble2HGNC) %>%
+  left_join(GENE.LEN)
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE %>%
+  dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(ensembl_gene_id, Comparison, logFC) %>%
+  group_by(Comparison) %>%
+  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))
+
+tmp1 = DE %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
+  dplyr::group_by(Comparison) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
+                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))
+
+kable(full_join(tmp,tmp1))
+
+p = DE %>%
+  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
+p = p + facet_grid(Comparison~., scales = 'free_y')
+p = p + ggtitle('Differential Expression between Schizophrenia and Control')
+p
+
+all.diff.exp = c(all.diff.exp, list(Diagnosis.Study.Gender = DE))
+all.fit = c(all.fit, list(Diagnosis.Study.Gender = ADJUSTED.FIT))
 ```
@@
 # Store differential expression results
-write.table(all.diff.exp, file = 'CMC_HBCC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+all.diff.exp %>%
+  rbindlist(use.names = T, fill = T) %>%
+  write.table(file = 'CMC_HBCC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
 DEXP_OBJ = File('CMC_HBCC_DLPFC_DiffExpression.tsv', 

### 2017-08-30 code/Rmd/CMC_HBCC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Added markdown to process CMC and HBCC samples together

--- 2017-08-30 code/Rmd/CMC_HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd",
+knit2synapse::knitToFolderEntity(file = "./CMC_HBCC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd",
                                  parentId = "syn7501880",
-                                 entityName = 'CMC and HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
+                                 entityName = 'CMC and HBCC - ACC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
 ```
@@
 
-thisFileName <- 'CMC_HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd'
+thisFileName <- 'CMC_HBCC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd'
 
@@
                        use.names = T, fill = T, idcol = 'Study')
+COVARIATES$Dx.Tissue.Study = paste(COVARIATES$Dx.Tissue, COVARIATES$Study, sep = '_')
 ```
@@
 ### Data preprocessing
-Filter DLPFC SCZ and Control samples alone
 ```{r preprocess.data, fig.height=10, fig.width=15}
-# Filter all ACC samples and Other samples
 COVARIATES = COVARIATES %>%
-  dplyr::filter(Dx.Tissue %in% c("SCZ_DLPFC", "Control_DLPFC"))
+  dplyr::filter(!(Dx.Tissue %in% c('BP_DLPFC', 'Other_ACC', 'Other_DLPFC')))
 
 COVARIATES %>%
-  dplyr::group_by(Dx.Tissue, Study) %>%
+  dplyr::group_by(Dx.Tissue.Study) %>%
   dplyr::summarise(count = n()) %>% 
-  tidyr::spread(Study, count) %>% 
   kable()
@@
 
-Age match CMC and HBCC (per Dx.Tissue category)
-```{r preprocess.data1}
-# Filter all ACC samples and Other samples
-max.age = COVARIATES %>%
-  dplyr::group_by(Dx.Tissue, Study) %>%
-  dplyr::filter(Study == 'CMC') %>%
-  dplyr::summarise(mn = quantile(Age_of_Death, 0.5))
-
-COVARIATES = COVARIATES %>%
-  dplyr::filter((Dx.Tissue == max.age$Dx.Tissue[1] & Age_of_Death <= max.age$mn[1]) |
-                  (Dx.Tissue == max.age$Dx.Tissue[2] & Age_of_Death <= max.age$mn[2]))
-
-COVARIATES %>%
-  dplyr::group_by(Dx.Tissue, Study) %>%
-  dplyr::summarise(count = n()) %>%
-  tidyr::spread(Study, count) %>% 
-  kable()
-```
-
-PMI match CMC and HBCC (per Dx.Tissue category)
-```{r preprocess.data2}
-max.pmi = COVARIATES %>%
-  dplyr::group_by(Dx.Tissue) %>%
-  dplyr::filter(Study == 'HBCC') %>%
-  dplyr::summarise(mn = quantile(PMI, 0.5))
-
-COVARIATES = COVARIATES %>%
-  dplyr::filter((Dx.Tissue == max.age$Dx.Tissue[1] & PMI <= max.pmi$mn[1]) |
-                  (Dx.Tissue == max.age$Dx.Tissue[2] & PMI <= max.pmi$mn[2]))
-
-COVARIATES %>%
-  dplyr::group_by(Dx.Tissue, Study) %>%
-  dplyr::summarise(count = n()) %>%
-  tidyr::spread(Study, count) %>% 
-  kable()
-```
-
 ```{r gene.param}
@@
 ```{r covariates.clustering}
-FactorCovariates <- c('Study', 'Individual_ID', 'Institution', 'Gender', 'LibraryBatch', 'Dx.Tissue', 'RibozeroBatch',
-                      'FlowcellBatch')
-ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
-                    "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate")
+FactorCovariates <- c('Individual_ID', 'Institution', 'Gender', 'LibraryBatch', 'Dx.Tissue.Study',
+                      'RibozeroBatch', 'FlowcellBatch')
+ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate",
+                    "IntergenicRate", "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate")
 
@@
 
-### Explore metatdata
-```{r data.explore, fig.width = 15, fig.height = 10}
-my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))
-
-# RIN
-p = list()
-p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN, color = Study)) + geom_boxplot()
-p[[1]] = p[[1]] + ggtitle('RIN') + my.theme
-
-# Age of Death
-p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = Age_of_Death, color = Study)) + geom_boxplot()
-p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme
-
-# PMI
-p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI, color = Study)) + geom_boxplot()
-p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme
-
-# Intronic Rate
-p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate, color = Study)) + geom_boxplot()
-p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme
-
-# IntergenicRate
-p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate, color = Study)) + geom_boxplot()
-p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme
-
-# Transcripts Detected
-p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TranscriptsDetected, color = Study)) + geom_boxplot()
-p[[6]] = p[[6]] + ggtitle('Transcripts Detected') + my.theme
-
-# Mapped Reads
-p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads, color = Study)) + geom_boxplot()
-p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme
-
-# rRNARate
-p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate, color = Study)) + geom_boxplot()
-p[[8]] = p[[8]] + ggtitle('rRNARate') + my.theme
-
-p1 = ggpubr::ggarrange(plotlist = p, ncol = 4, nrow = 2)
-p1
-```
 ### Filter genes
@@
   rownameToFirstColumn('SampleID') %>%
-  dlply(.(Dx.Tissue), .fun = function(mtd, count){
+  dlply(.(Dx.Tissue.Study), .fun = function(mtd, count){
     processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
@@
 plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
+  tidyr::separate(Dx.Tissue.Study, c('Dx','Tissue', 'Study'), sep = '_') %>%
   dplyr::mutate(label = SampleID)
-plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''
+# plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
+p <- p + geom_point(aes(color=Study, shape=Tissue, size=Age_of_Death))
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
@@
 
-tmp = tidyr::separate(COVARIATES, Dx.Tissue, c('Dx', 'Tissue'), sep = '_') %>%
-  group_by(Study, Dx, Tissue) %>%
+tmp = tidyr::separate(COVARIATES, Dx.Tissue.Study, c('Dx', 'Tissue', 'Study'), sep = '_') %>%
+  group_by(Study, Dx, Tissue, Study) %>%
   summarise(count = n()) %>%
@@
 plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
+  tidyr::separate(Dx.Tissue.Study, c('Dx','Tissue','Study'), sep = '_')
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
+p <- p + geom_point(aes(color=Study, shape=Tissue, size=Age_of_Death))
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
@@
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
+COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue.Study")])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
@@
 gc()
-```
-
-### Distribution of samples (log cpm)
-```{r lcpm.dist, cache=FALSE}
-# Plot abberent distribution of logcpm counts
-tmp1 = (CQN.GENE_EXPRESSION$E) %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
-  left_join(COVARIATES %>%
-              rownameToFirstColumn('SampleID') %>%
-              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))
-
-p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
-p = p + theme(legend.position = 'NONE') + facet_grid(Dx+.~Tissue, scale = 'free')
-p
 ```
@@
 plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
+  tidyr::separate(Dx.Tissue.Study, c('Dx','Tissue','Study'), sep = '_') %>%
   dplyr::mutate(label = SampleID)
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
+p <- p + geom_point(aes(color=Study, shape=Tissue, size=Age_of_Death))
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
@@
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue','Institution')])
+COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue.Study','Institution')])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
@@
                            main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue','Institution')]))
+                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue.Study','Institution')]))
 ```
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.Tissue.StudySCZ_DLPFC_CMC-Dx.Tissue.StudyControl_DLPFC_CMC",
-                                     "Dx.Tissue.StudySCZ_DLPFC_HBCC-Dx.Tissue.StudyControl_DLPFC_HBCC"),
+contrast = makeContrasts(contrasts=c("Dx.Tissue.StudySCZ_DLPFC_CMC_-Dx.Tissue.StudyControl_DLPFC_CMC_",
+                                     "Dx.Tissue.StudySCZ_DLPFC_HBCC_-Dx.Tissue.StudyControl_DLPFC_HBCC_",
+                                     "Dx.Tissue.StudySCZ_ACC_CMC_-Dx.Tissue.StudyControl_ACC_CMC_"),
                          levels = colnames(ADJUSTED.FIT$coefficients))
@@
 ```{r diffExp1, fig.height=15, fig.width=20}
-COVARIATES$Dx.Tissue.Study.Gender = paste(COVARIATES$Dx.Tissue.Study,
-                                          COVARIATES$Gender,
-                                          sep = '_') %>%
+COVARIATES$Dx.Tissue.Study.Gender = paste0(COVARIATES$Dx.Tissue.Study,
+                                           COVARIATES$Gender) %>%
   factor()
@@
                              "Dx.Tissue.Study.GenderSCZ_DLPFC_HBCC_Female-Dx.Tissue.Study.GenderControl_DLPFC_HBCC_Female",
-                             "Dx.Tissue.Study.GenderSCZ_DLPFC_HBCC_Male-Dx.Tissue.Study.GenderControl_DLPFC_HBCC_Male"),
+                             "Dx.Tissue.Study.GenderSCZ_DLPFC_HBCC_Male-Dx.Tissue.Study.GenderControl_DLPFC_HBCC_Male",
+                             "Dx.Tissue.Study.GenderSCZ_ACC_CMC_Female-Dx.Tissue.Study.GenderControl_ACC_CMC_Female",
+                             "Dx.Tissue.Study.GenderSCZ_ACC_CMC_Male-Dx.Tissue.Study.GenderControl_ACC_CMC_Male"),
                          levels = colnames(ADJUSTED.FIT$coefficients))
@@
 # Code
-CODE <- Folder(name = "CMC and HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
+CODE <- Folder(name = "CMC and HBCC - ACC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
 annotations(CODE) = all.annotations
@@
 |  Design Matrix                              |  `r DM_OBJ$properties$id`        |
-|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
 |  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
-|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |
 

### 2017-08-30 code/Rmd/CMC_HBCC_COMPARISON.Rmd
Added markdown to process CMC and HBCC samples together

@@
                  HBCC = 'syn10263511',
-                 HBCC_qSVA = 'syn10294725')
+                 HBCC_qSVA = 'syn10294725',
+                 MATCHED = 'syn10490519')
 
@@
 rep.rate = list()
-for (pval in seq(0,-6,-1)){
+for (pval in seq(0,-4,-1)){
   de.genes = diff.exp %>%
@@
 
-  rep.rate = c(rep.rate, list(data.frame(fdr = 10^pval, wo.qsva = wo.qsva, w.qsva = w.qsva,
-                                         n.wo.qsva = n.wo.qsva, n.w.qsva = n.w.qsva)))
+  de.genes = diff.exp %>%
+    dplyr::filter(Comparison %in% c('SCZ_DLPFC_CMC-Control_DLPFC_CMC', 'SCZ_DLPFC_HBCC-Control_DLPFC_HBCC')) %>%
+    dplyr::mutate(Direction = factor(sign(logFC), levels = c('-1','1'), labels = c('-1' = 'DOWN','1' = 'UP'))) %>%
+    plyr::dlply(.(Comparison, Model, Direction), .fun = function(x){
+      x$ensembl_gene_id[x$adj.P.Val <= 10^pval] %>% unique()
+    })
+
+  l1 = length(intersect(de.genes$`SCZ_DLPFC_CMC-Control_DLPFC_CMC.MATCHED.DOWN`,
+                        de.genes$`SCZ_DLPFC_HBCC-Control_DLPFC_HBCC.MATCHED.DOWN`))
+  l2 = length(intersect(de.genes$`SCZ_DLPFC_CMC-Control_DLPFC_CMC.MATCHED.UP`, 
+                        de.genes$`SCZ_DLPFC_HBCC-Control_DLPFC_HBCC.MATCHED.UP`))
+  l3 = length(unique(Reduce(c, de.genes)))
+  matched = (l1+l2)/l3
+  n.matched = l1+l2
+
+  rep.rate = c(rep.rate, list(data.frame(fdr = 10^pval, 
+                                         wo.qsva = wo.qsva, w.qsva = w.qsva, matched = matched,
+                                         n.wo.qsva = n.wo.qsva, n.w.qsva = n.w.qsva, n.matched = n.matched)))
 }
@@
 rep.rate %>%
-  dplyr::select(fdr, wo.qsva, w.qsva) %>%
+  dplyr::select(fdr, wo.qsva, w.qsva, matched) %>%
   tidyr::gather(method, rep.rate, -fdr) %>%
@@
   dplyr::left_join(rep.rate %>%
-                     dplyr::select(fdr, n.wo.qsva, n.w.qsva) %>%
-                     tidyr::gather(method, ncount, -fdr) %>%
+                     dplyr::select(fdr, n.wo.qsva, n.w.qsva, n.matched) %>%
+                     tidyr::gather(method, nOverlap, -fdr) %>%
                      dplyr::mutate(method = gsub('n.','',method))) %>%
   dplyr::filter(fdr <= 1e-1) %>%
-  ggplot(aes(x=fdr, y=rep.rate, color = method)) + geom_point(aes(size = ncount)) + geom_line() + scale_x_log10() + xlab('FDR') + ylab('Replication Rate (%)')
+  ggplot(aes(x=fdr, y=rep.rate, color = method)) + geom_point(aes(size = nOverlap)) + geom_line() + scale_x_log10() + xlab('FDR') + ylab('Replication Rate (%)') + ggtitle('CMC - HBCC - DLPFC')
 ```
@@
   rs = data.frame(method = 'wo.qsva', ref.data = 'HBCC', rep.data = 'CMC',
-                  rep.rate = sum(tmp$significant)/length(tmp$significant), 
+                  rep.rate = 1-tmp$pi0, 
                   ncount = sum(tmp$significant))
@@
              data.frame(method = 'wo.qsva', ref.data = 'CMC', rep.data = 'HBCC',
-                        rep.rate = sum(tmp$significant)/length(tmp$significant), 
+                        rep.rate = 1-tmp$pi0, 
                         ncount = sum(tmp$significant)))
@@
              data.frame(method = 'w.qsva', ref.data = 'HBCC', rep.data = 'CMC',
-                        rep.rate = sum(tmp$significant)/length(tmp$significant), 
+                        rep.rate = 1-tmp$pi0, 
                         ncount = sum(tmp$significant)))
@@
              data.frame(method = 'w.qsva', ref.data = 'CMC', rep.data = 'HBCC',
-                        rep.rate = sum(tmp$significant)/length(tmp$significant), 
+                        rep.rate = 1-tmp$pi0, 
+                        ncount = sum(tmp$significant)))
+
+  de.genes = diff.exp %>%
+    dplyr::filter(Comparison %in% c('SCZ_DLPFC_CMC-Control_DLPFC_CMC', 'SCZ_DLPFC_HBCC-Control_DLPFC_HBCC')) %>%
+    plyr::dlply(.(Comparison), .fun = function(x){
+      x$ensembl_gene_id[x$adj.P.Val <= 10^pval] %>% unique()
+    })
+
+  tmp = diff.exp %>%
+    dplyr::filter(Comparison == 'SCZ_DLPFC_CMC-Control_DLPFC_CMC', 
+                  ensembl_gene_id %in% de.genes$`SCZ_DLPFC_HBCC-Control_DLPFC_HBCC`) %>%
+    dplyr::select(P.Value)
+
+  if (dim(tmp)[1] != 0){
+    tmp %>%
+      qvalue::qvalue(fdr.level = 10^pval)
+    rs = rbind(rs,
+               data.frame(method = 'matched', ref.data = 'HBCC', rep.data = 'CMC',
+                          rep.rate = 1-tmp$pi0, 
+                          ncount = sum(tmp$significant)))
+  } else {
+    rs = rbind(rs,
+               data.frame(method = 'matched', ref.data = 'HBCC', rep.data = 'CMC',
+                          rep.rate = 0, 
+                          ncount = 0))
+  }
+
+  tmp = diff.exp %>%
+    dplyr::filter(Comparison == 'SCZ_DLPFC_HBCC-Control_DLPFC_HBCC', 
+                  ensembl_gene_id %in% de.genes$`SCZ_DLPFC_CMC-Control_DLPFC_CMC`) %>%
+    dplyr::select(P.Value) %>%
+    qvalue::qvalue(fdr.level = 10^pval)
+  rs = rbind(rs,
+             data.frame(method = 'matched', ref.data = 'CMC', rep.data = 'HBCC',
+                        rep.rate = 1-tmp$pi0, 
                         ncount = sum(tmp$significant)))

### 2017-09-05 code/Rmd/MAYO_REPROCESSED.Rmd
Minor modifications and added parallelDuplicateCorrelation.R with lmer

@@
 options(xtable.type="html")
+source('../R/parallelDuplicateCorrelation.R')
 
@@
 postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
+notEstimable = c()
 
@@
   correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
-                                             block = COVARIATES$Donor_ID)
-
+                                             block = COVARIATES$Donor_ID,
+                                             method = 'lmer')
+  # correlation = list(); correlation$cor = 0.457197
+
+  if (!is.nan(correlation$cor)){
     # Re-estimate voom weights
@@
                                                   CQN.GENE_EXPRESSION$E)
+  } else {
+    notEstimable = c(notEstimable, postAdjustCovars[length(postAdjustCovars)])
+    postAdjustCovars = postAdjustCovars[1:(length(postAdjustCovars)-1)]
+  }
 
@@
   residCovars <- setdiff(c(FactorCovariates,ContCovariates),
-                         c(postAdjustCovars, primaryVariable, 'Donor_ID'))
+                         c(postAdjustCovars, primaryVariable, 'Donor_ID', notEstimable))
 
@@
   # Add postadjusted covariates (if any)
-  residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, primaryVariable))
+  residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, primaryVariable, notEstimable))
   covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]
@@
 VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Donor_ID)
+correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS,
+                                           block = COVARIATES$Donor_ID,
+                                           method = 'lmer')
 
@@
 VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-FIT = lmFit(VOOM.WEIGHTS)
+FIT = lmFit(VOOM.WEIGHTS$E,
+            weights = VOOM.WEIGHTS$weights,
+            design = VOOM.WEIGHTS$design,
+            block = COVARIATES$Donor_ID, 
+            correlation = correlation$cor)
 
@@
 VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
+correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS,
+                                           block = COVARIATES$Donor_ID,
+                                           method = 'lmer')
 
@@
 VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Donor_ID)
 
@@
 VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$Donor_ID)
+correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS,
+                                           block = COVARIATES$Donor_ID,
+                                           method = 'lmer')
 
@@
 VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)
+VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS$E,
+                     weights = VOOM.SVA.WEIGHTS$weights,
+                     design = VOOM.SVA.WEIGHTS$design,
+                     block = COVARIATES$Donor_ID, 
+                     correlation = correlation$cor)
 
@@
 }) %>% t
+
+all.fit = c(all.fit, list(SVA = VOOM.SVA.FIT))
 ```
@@
 VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$Donor_ID)
+correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS,
+                                           block = COVARIATES$Donor_ID,
+                                           method = 'lmer')
 
@@
 VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)
+VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS$E,
+                     weights = VOOM.NET.WEIGHTS$weights,
+                     design = VOOM.NET.WEIGHTS$design,
+                     block = COVARIATES$Donor_ID, 
+                     correlation = correlation$cor)
 
@@
 }) %>% t
+
+all.fit = c(all.fit, list(NET = VOOM.NET.FIT))
 ```
@@
 DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
+# Store all model fits
+save(list = 'all.fit', file = 'MAYO_CBE_TCX_allfits.RData')
+AFIT_OBJ = File('MAYO_CBE_TCX_allfits.RData', 
+                name = 'All model fits (BrainRegion.Diagnosis)', 
+                parentId = CODE$properties$id)
+annotations(AFIT_OBJ) = annotations(CODE)
+annotations(AFIT_OBJ)$dataSubType = 'modelFits'
+AFIT_OBJ = synStore(AFIT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 # stopCluster(cl)

### 2017-09-05 code/Rmd/MSSM_REPROCESSED.Rmd
Minor modifications and added parallelDuplicateCorrelation.R with lmer

@@
   # # Calculate correlation values of random effects
-  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
-                                             block = COVARIATES$individualIdentifier)
-  # correlation = list(); correlation$cor = 0.4
+  # correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
+  #                                            block = COVARIATES$individualIdentifier)
+  correlation = list(); correlation$cor = 0.457197
 

### 2017-09-05 code/R/parallelDuplicateCorrelation.R

@@ -1,0 +1,209 @@
+parallelDuplicateCorrelation <- function (object, design = NULL, ndups = 2, spacing = 1, block = NULL, 
+                                          trim = 0.15, weights = NULL, method = 'lmer'){
+  if (method == 'lmer'){
+    parallelDuplicateCorrelation.lmer(object, design, ndups, spacing, block, trim, weights)
+  } else if (method == 'mm2'){
+    parallelDuplicateCorrelation.mm2(object, design, ndups, spacing, block, trim, weights)
+  } else {
+    error('Unknown Method')
+  }
+}
+  
+parallelDuplicateCorrelation.lmer <- function (object, design = NULL, ndups = 2, spacing = 1, block = NULL, 
+                                              trim = 0.15, weights = NULL) 
+{
+  y <- getEAWP(object)
+  M <- y$exprs
+  ngenes <- nrow(M)
+  narrays <- ncol(M)
+  if (is.null(design)) 
+    design <- y$design
+  if (is.null(design)) 
+    design <- matrix(1, ncol(y$exprs), 1)
+  else {
+    design <- as.matrix(design)
+    if (mode(design) != "numeric") 
+      stop("design must be a numeric matrix")
+  }
+  if (nrow(design) != narrays) 
+    stop("Number of rows of design matrix does not match number of arrays")
+  ne <- nonEstimable(design)
+  if (!is.null(ne)) 
+    cat("Coefficients not estimable:", paste(ne, collapse = " "), 
+        "\n")
+  nbeta <- ncol(design)
+  if (missing(ndups) && !is.null(y$printer$ndups)) 
+    ndups <- y$printer$ndups
+  if (missing(spacing) && !is.null(y$printer$spacing)) 
+    spacing <- y$printer$spacing
+  if (missing(weights) && !is.null(y$weights)) 
+    weights <- y$weights
+  if (!is.null(weights)) {
+    weights <- as.matrix(weights)
+    if (any(dim(weights) != dim(M))) 
+      weights <- array(weights, dim(M))
+    M[weights < 1e-15] <- NA
+    weights[weights < 1e-15] <- NA
+  }
+  if (is.null(block)) {
+    if (ndups < 2) {
+      warning("No duplicates: correlation between duplicates not estimable")
+      return(list(cor = NA, cor.genes = rep(NA, nrow(M))))
+    }
+    if (is.character(spacing)) {
+      if (spacing == "columns") 
[…]

### 2017-09-05 code/R/collateDiffExp.R
Added R script to collate all differential expression results to genesets

@@ -1,0 +1,310 @@
+library(synapseClient)
+library(knitr)
+library(githubr)
+
+library(CovariateAnalysis)
+library(data.table)
+library(tidyr)
+library(plyr)
+library(dplyr)
+library(stringr)
+
+synapseLogin()
+
+diff.exp = lapply(c(MAYO = 'syn8468023',
+                    MSSM = 'syn10157628',
+                    ROSMAP = 'syn8456721'),
+                  downloadFile)
+
+############################################
+# AD-CONTROL
+gs = list()
+gs[['DLPFC.AD-CONTROL.UP']] = diff.exp$ROSMAP %>%
+  dplyr::filter(Model == 'Diagnosis', Comparison == 'AD-CONTROL', adj.P.Val <= 0.05, logFC >= 0) %>%
+  dplyr::select(hgnc_symbol) %>%
+  unlist %>% unique()
+
+gs[['DLPFC.AD-CONTROL.DOWN']] = diff.exp$ROSMAP %>%
+  dplyr::filter(Model == 'Diagnosis', Comparison == 'AD-CONTROL', adj.P.Val <= 0.05, logFC <= 0) %>%
+  dplyr::select(hgnc_symbol) %>%
+  unlist %>% unique()
+
+tmp = diff.exp$MSSM %>%
+  dplyr::filter(Model == 'Diagnosis', 
+                BrainRegion %in% c('IFG', 'PHG' , 'FP', 'STG'), 
+                Comparison == 'CONTROL-AD', adj.P.Val <= 0.05, logFC <= 0) %>%
+  plyr::dlply(.(BrainRegion), .fun = function(x){
+    dplyr::select(x, hgnc_symbol) %>%
+      unlist %>% unique()
+  })
+names(tmp) = paste(names(tmp), 'AD-CONTROL.UP', sep = '.')
+gs = c(gs, tmp)
+
+tmp = diff.exp$MSSM %>%
+  dplyr::filter(Model == 'Diagnosis', 
+                BrainRegion %in% c('IFG', 'PHG' , 'FP', 'STG'), 
+                Comparison == 'CONTROL-AD', adj.P.Val <= 0.05, logFC >= 0) %>%
+  plyr::dlply(.(BrainRegion), .fun = function(x){
+    dplyr::select(x, hgnc_symbol) %>%
+      unlist %>% unique()
+  })
[…]

### 2017-09-05 code/Rmd/MAYO_REPROCESSED.Rmd
Minor modifications, added AOD specific differential expression to MAYO markdown

@@
 
+### Differential expression analysis (with BrainRegion.Diagnosis x AgeAtDeath as primary variable)
+Differential expression is performed on the primary variable by controlling for covariates identified above
+
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diff.exp3, fig.height=5, fig.width=18}
+# Get design matrix
+DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis', setdiff(postAdjustCovars, 'AgeAtDeath')), drop = F], 
+                         Intercept = F)
+DESIGN$design[,grep('BrainRegion.Diagnosis',colnames(DESIGN$design))] = DESIGN$design[,grep('BrainRegion.Diagnosis',colnames(DESIGN$design))] * COVARIATES[,'AgeAtDeath']
+colnames(DESIGN$design)[grep('BrainRegion.Diagnosis',colnames(DESIGN$design))] = 
+  paste(colnames(DESIGN$design)[grep('BrainRegion.Diagnosis',colnames(DESIGN$design))],'AOD', sep = '.')
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
+
+# Calculate correlation values of random effects
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS,
+                                           block = COVARIATES$Donor_ID,
+                                           method = 'lmer')
+
+# Re-estimate voom weights
+VOOM.WEIGHTS = voom(NEW.COUNTS,
+                    design=DESIGN$design, plot=F,
+                    block = COVARIATES$Donor_ID, 
+                    correlation = correlation$cor)
+
+# Fit linear model using new weights and new design
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+FIT = lmFit(VOOM.WEIGHTS$E,
+            weights = VOOM.WEIGHTS$weights,
+            design = VOOM.WEIGHTS$design,
+            block = COVARIATES$Donor_ID, 
+            correlation = correlation$cor)
+
+# Fit contrast
+contrast = makeContrasts(contrasts=c(
+  "BrainRegion.DiagnosisCER.AD.AOD-BrainRegion.DiagnosisCER.Control.AOD",
+  "BrainRegion.DiagnosisCER.AD.AOD-BrainRegion.DiagnosisCER.OTHER.AOD",
+  "BrainRegion.DiagnosisCER.OTHER.AOD-BrainRegion.DiagnosisCER.Control.AOD",
+  "BrainRegion.DiagnosisTCX.AD.AOD-BrainRegion.DiagnosisTCX.Control.AOD",
+  "BrainRegion.DiagnosisTCX.AD.AOD-BrainRegion.DiagnosisTCX.OTHER.AOD",
+  "BrainRegion.DiagnosisTCX.OTHER.AOD-BrainRegion.DiagnosisTCX.Control.AOD"),
+  levels = colnames(FIT$coefficients))
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = T) %>%
+    rownameToFirstColumn('ensembl_gene_id')
+}, FIT.CONTR)
+names(DE) = colnames(contrast)
+
+DE = DE %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
+                Study = 'MAYO',
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
+  tidyr::separate(from.state, c('BrainRegion.ref','Dx.ref','AOD.ref'), sep = '\\.') %>%
+  tidyr::separate(to.state, c('BrainRegion.ag','Dx.ag','AOD.ag'), sep = '\\.') %>%
+  tidyr::unite(Comparison, Dx.ref, Dx.ag, sep = '-') %>%
+  left_join(GENE.PARAM %>%
+              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
+              unique())
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion.ref, AOD.ref) %>%
+  group_by(BrainRegion.ref, AOD.ref, Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05_FC_1.2) 
+kable(tmp)
+
+p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red'))
+p = p + facet_grid(Comparison+.~BrainRegion.ref+AOD.ref, scales = 'fixed')
+p
+
+all.diff.exp = c(all.diff.exp, list(Diagnosis.AOD = DE))
+all.fit = c(all.fit, list(Diagnosis.AOD = FIT))
+```
+
 ### SVA Adjustments for eQTL analysis

### 2017-09-05 code/R/collateDiffExp.R

@@ -307,4 +307,4 @@ thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/R/',t
 amp.ad.de.geneSets = c(gs, gs1, gs2, gs3)
 save(list='amp.ad.de.geneSets', file = 'all.diff.exp.gs.RData')
 obj = File('all.diff.exp.gs.RData', name = 'All Differential Expression GeneSets (RData format)', parentId = 'syn8672415')
-obj = synStore(obj, used = c('syn8468023', 'syn10157628', 'syn8456721'), executed = thisFile, activity = 'Collate differential expression results')
+obj = synStore(obj, used = c('syn8468023', 'syn10157628', 'syn8456721'), executed = thisFile, activityName = 'Collate differential expression results')

### 2017-09-05 code/Rmd/MAYO_REPROCESSED.Rmd
Minor modifications, added AOD specific differential expression to MAYO markdown

@@
   rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
+  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis.Gender','',Comparison),
                 Study = 'MAYO',
@@
   tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
-  tidyr::separate(from.state, c('BrainRegion.ref','Dx.ref','AOD.ref'), sep = '\\.') %>%
-  tidyr::separate(to.state, c('BrainRegion.ag','Dx.ag','AOD.ag'), sep = '\\.') %>%
+  tidyr::separate(from.state, c('BrainRegion.ref','Dx.ref','Gender.ref'), sep = '\\.') %>%
+  tidyr::separate(to.state, c('BrainRegion.ag','Dx.ag','Gender.ag'), sep = '\\.') %>%
   tidyr::unite(Comparison, Dx.ref, Dx.ag, sep = '-') %>%
@@
 tmp = DE %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion.ref, AOD.ref) %>%
-  group_by(BrainRegion.ref, AOD.ref, Comparison, Direction) %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion.ref, Gender.ref) %>%
+  group_by(BrainRegion.ref, Gender.ref, Comparison, Direction) %>%
   dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
@@
 p = p + scale_color_manual(values = c('green','grey','red'))
-p = p + facet_grid(Comparison+.~BrainRegion.ref+AOD.ref, scales = 'fixed')
+p = p + facet_grid(Comparison+.~BrainRegion.ref+Gender.ref, scales = 'fixed')
 p
 
-all.diff.exp = c(all.diff.exp, list(Diagnosis.AOD = DE))
-all.fit = c(all.fit, list(Diagnosis.AOD = FIT))
+all.diff.exp = c(all.diff.exp, list(Diagnosis.Gender = DE))
+all.fit = c(all.fit, list(Diagnosis.Gender = FIT))
 ```

### 2017-09-07 code/Rmd/ROSMAP_REPROCESSED_ISOFORM.Rmd
Added markdown to process isoform results from ROSMAP

--- 2017-08-10 code/Rmd/ROSMAP_REPROCESSED.Rmd
@@
 ---
-title: "Covariate analysis of ROSMAP reprocessed counts (with CQN normalisation and covariates adjusted)"
+title: "Covariate analysis of ROSMAP reprocessed isfoform counts (with CQN normalisation and covariates adjusted)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./ROSMAP_REPROCESSED.Rmd",
+knit2synapse::knitToFolderEntity(file = "./ROSMAP_REPROCESSED_ISOFORM.Rmd",
                                  parentId = "syn8672910",
-                                 entityName = 'ROSMAP Reprocessed RNASeq (Normalised)')
+                                 entityName = 'ROSMAP Reprocessed Isoform (Normalised)')
 ```
@@
 activityName = 'Covariate analysis';
-activityDescription = 'Covariate analysis of ROSMAP reprocessed RNASeq data';
-
-thisFileName <- 'ROSMAP_REPROCESSED.Rmd'
+activityDescription = 'Covariate analysis of ROSMAP reprocessed isoform data';
+
+thisFileName <- 'ROSMAP_REPROCESSED_ISOFORM.Rmd'
 
 # Github link
-thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
+thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_isoform')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
@@
 # Download expression data
-COUNT_ID <- 'syn8691134';#'syn7750270';
+COUNT_ID <- 'syn10507749';
 ALL_USED_IDs <- COUNT_ID
@@
 COUNT[,grep('150_120419', colnames(COUNT))[2]] = NULL
-
-# Convert rownames of counts from tracking id to ensemble gene id
-tmp = data.frame(Gene.ID = rownames(COUNT)) %>%
-  dplyr::mutate(ID = Gene.ID) %>%
-  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')
-rownames(tmp) = tmp$Gene.ID
-rownames(COUNT) = tmp[rownames(COUNT), 'ensembl_gene_id']
 
@@
 METADATA.TECH <- read.table(METADATA.TECH_OBJ@filePath,sep='\t',header=T)
-
-# Get picard metrics from synapse
-METADATA.PICARD_ID <- 'syn8698240';#'syn8512191'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
-# METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
-# METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
-# METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
-# METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)
-# METADATA.PICARD$Sampleid = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
-#   gsub('Aligned.out/picard.CombinedMetrics.csv','',.)
-
-METADATA.PICARD <- CovariateAnalysis::downloadFile(METADATA.PICARD_ID) %>%
-  dplyr::rename(Sampleid = sample)
-colnames(METADATA.PICARD) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD))
-colnames(METADATA.PICARD) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD))
 
@@
 METADATA <- METADATA.TECH %>%
-  left_join(METADATA.PICARD) %>%
   dplyr::select(-projid) %>%
@@
   top_n(1, RINcontinuous)
-
-# Get gene specific parameters from synapse
-GENE.PARAM = downloadFile('syn8449369')
-ALL_USED_IDs = c(ALL_USED_IDs, 'syn8449369')
-
-GENE.LEN = dplyr::select(GENE.PARAM, ensembl_gene_id, gene.length) %>% 
-  unique() 
-rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id
-
-GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% 
-  unique() 
-rownames(GENE.GC) = GENE.GC$ensembl_gene_id 
 ```
@@
   dplyr::filter(!is.na(pmi)) %>%
-  dplyr::filter(!is.na(PCT_INTRONIC_BASES)) %>%
   dplyr::filter(!is.na(age_death)) %>%
@@
 ```{r covariates.clustering, cache=TRUE}
-#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv", "PF_INDEL_RATE", "PF_MISMATCH_RATE", "age_first_ad_dx","age_at_visit_max",
 primaryVariable <- c("cogdx", "Diagnosis", "apoe_genotype")
-FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx", "Diagnosis", "apoe_genotype")
-ContCovariates <- c("RINcontinuous", "age_death", "pmi", "educ", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
-                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
+FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx", 
+                      "Diagnosis", "apoe_genotype")
+ContCovariates <- c("RINcontinuous", "age_death", "pmi", "educ")
 
@@
 
-levels(COVARIATES$apoe_genotype) = c('0','0','1','0','1','2')
+levels(COVARIATES$apoe_genotype) = c('ApoE0','ApoE0','ApoE1',
+                                     'ApoE0','ApoE1','ApoE2')
 ```
@@
 COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
-draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left', padding  = unit(c(18,2,2,18), 'mm'))
+draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left', 
+     padding  = unit(c(18,2,2,18), 'mm'))
 ```
@@
 
-# Intronic bases
-p[[5]] = ggplot(COVARIATES, aes(x = Diagnosis, y = PCT_INTRONIC_BASES)) + geom_boxplot()
-p[[5]] = p[[5]] + ggtitle('Fraction Intronic Bases') + theme(legend.position = 'top')
-
-# Ribosomal bases
-p[[6]] = ggplot(COVARIATES, aes(x = Diagnosis, y = PCT_RIBOSOMAL_BASES)) + geom_boxplot()
-p[[6]] = p[[6]] + ggtitle('Fraction Ribosomal Bases') + theme(legend.position = 'top')
-
 multiplot(plotlist = p, cols = 2)
@@
 ```
-### Filter genes
-Remove genes that have less than 1 cpm counts in at least 50% of samples per Diagnosis. Also remove genes with missing gene length and percentage GC content
+### Filter Transcripts
+Remove transcripts that have less than 5 tpm in at least 50% of samples per Diagnosis
 ```{r cpmnormalisation}
@@
   processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$Sampleid],
-                                                   MIN_GENE_CPM=1, 
+                                                   MIN_GENE_CPM=5, 
                                                    MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
@@
   unique() %>%
-  intersect(GENE.GC$ensembl_gene_id[!is.na(GENE.GC$percentage_gc_content)]) %>%
-  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$gene.length)]) %>%
   setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous"))
@@
                                                  MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
-
-# Check gene biotype
-## Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "dec2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
-
-## Query biomart
-Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "gene_biotype"),
-                       filters = "ensembl_gene_id", 
-                       values = PROCESSED_COUNTS$filteredExprMatrix$genes,
-                       mart = mart)
-
-summary(factor(Ensemble2HGNC$gene_biotype)) %>%
-  rownameToFirstColumn('Biotype') %>%
-  dplyr::rename(fraction = DF) %>%
-  dplyr::mutate(fraction = fraction/dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]) %>%
-  dplyr::filter(fraction >= 0.01) %>%
-  kable
 ```
@@
 tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
-  rownameToFirstColumn('Gene.ID') %>%
-  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
+  rownameToFirstColumn('Tx.ID') %>%
+  tidyr::gather(SampleID, logCPM, -Tx.ID) %>%
   left_join(COVARIATES %>%
@@
 
-### Library Normalisation
-Initial library normalisation usign cqn
-```{r cqn}
-# Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
-                          x = GENE.GC[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
-                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'gene.length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
-CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+### Library Normalisation (TMM)
+```{r tmm}
+# Compute calc norm factors
+TMM.GENE_EXPRESSION = calcNormFactors(NEW.COUNTS, lib.size = LIB.SIZE, method = 'TMM')
+VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design = NULL, 
+                            lib.size = TMM.GENE_EXPRESSION*LIB.SIZE)
 
 # Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
-LOG.CPM = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
+LOG.CPM = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
   mn = mean(x, na.rm = T)
@@
 }) %>% t
-CQN.GENE_EXPRESSION$E = LOG.CPM
-NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
+VOOM.GENE_EXPRESSION$E = LOG.CPM
+NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], TMM.GENE_EXPRESSION * LIB.SIZE))/1e6
 ```
@@
 # Find principal components of expression to plot
-PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)
+PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E)))
+tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 # Plot abberent distribution of logcpm counts
-tmp1 = (CQN.GENE_EXPRESSION$E) %>%
-  rownameToFirstColumn('Gene.ID') %>%
-  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
+tmp1 = (VOOM.GENE_EXPRESSION$E) %>%
+  rownameToFirstColumn('Tx.ID') %>%
+  tidyr::gather(SampleID, logCPM, -Tx.ID) %>%
   left_join(COVARIATES %>%
@@
 ```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(CQN.GENE_EXPRESSION$E))
+cr = cor(t(VOOM.GENE_EXPRESSION$E))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find correlation between PC's of gene expression with covariates
-preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E, 
+preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
                                                  COVARIATES,
@@
                               plot=F,
-                              na.rm = T)
+                              na.rm = T,
+                              lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
 
   # Fit linear model using new weights and new design
-  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$E,
+  VOOM.ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION$E,
                             design = DM1$design,
@@
   RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
-                                                CQN.GENE_EXPRESSION$E)
+                                                VOOM.GENE_EXPRESSION$E)
 
@@
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F, lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
 
 # Fit linear model using new weights and new design
-FIT = lmFit(CQN.GENE_EXPRESSION$E,
+FIT = lmFit(VOOM.WEIGHTS$E,
             design = DESIGN$design,
@@
   topTable(FIT, coef=i, number = 50000, confint = T) %>%
-    rownameToFirstColumn('ensembl_gene_id')
+    rownameToFirstColumn('Tx.ID')
 }, FIT.CONTR)
@@
   rbindlist(idcol = 'Comparison') %>%
-  left_join(GENE.PARAM %>%
-              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
-              unique()) %>%
   dplyr::mutate(Study = 'ROSMAP', Region = 'DLPFC',
@@
 tmp = DE %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
+  dplyr::select(Tx.ID, Comparison, Direction) %>%
   group_by(Comparison, Direction) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(Tx.ID))) %>%
   spread(Direction, FDR_0_05_FC_1.2) 
@@
 all.fit = c(all.fit, list(Diagnosis = FIT))
-```
-
-### Associate differential expression results with gc content, gene length and average expression
-```{r associate.de, fig.height=10, fig.width=15}
-pl = list()
-pl[[1]] = ggplot(DE, aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
-pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
-pl[[1]] = pl[[1]] + theme(legend.position = 'top')
-
-pl[[2]] = ggplot(DE, aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
-pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
-pl[[2]] = pl[[2]] + theme(legend.position = 'top')
-
-pl[[3]] = ggplot(DE, aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
-pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
-pl[[3]] = pl[[3]] + theme(legend.position = 'top')
-
-multiplot(plotlist = pl, cols = 3)
 ```
@@
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(NEW.COUNTS[,rownames(DESIGN$design)], design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(NEW.COUNTS[,rownames(DESIGN$design)], design=DESIGN$design, plot=F, 
+                    lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
 
 # Fit linear model using new weights and new design
-FIT = lmFit(CQN.GENE_EXPRESSION$E[,rownames(DESIGN$design)],
+FIT = lmFit(VOOM.WEIGHTS$E[,rownames(DESIGN$design)],
             design = DESIGN$design,
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("apoe_genotype2-apoe_genotype0",
-                                     "apoe_genotype2-apoe_genotype1",
-                                     "apoe_genotype1-apoe_genotype0"),
+contrast = makeContrasts(contrasts=c("apoe_genotypeApoE2-apoe_genotypeApoE0",
+                                     "apoe_genotypeApoE2-apoe_genotypeApoE1",
+                                     "apoe_genotypeApoE1-apoe_genotypeApoE0"),
                          levels = colnames(FIT$coefficients))
@@
 all.fit = c(all.fit, list(Diagnosis.AOD = FIT))
-```
-
-### SVA Adjustments for eQTL analysis
-Conditioned on primary variable (Diagnosis) and identified covariates estimate surrogate variables using SVA package
-```{r sva.adjust}
-# Get design matrix
-DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[2], postAdjustCovars), drop = F], Intercept = F)
-DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
-
-# Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
-
-# Fit linear model using new weights and new design
-FIT = lmFit(CQN.GENE_EXPRESSION$E,
-            design = DESIGN$design,
-            weights = VOOM.WEIGHTS$weights)
-
-# Get (null) design matrix
-MODEL0 = DESIGN$design[,-grep(primaryVariable[2],colnames(DESIGN$design))] # Get null model by removing variable of interest
-MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
-MODEL1 = DESIGN$design
-MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
-
-# Calculate the number of surrogate variables (be method)
-# NUM.SV = sva::num.sv(CQN.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456, B = 30)
-
-# Get residuals from differential expression model
-RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, CQN.GENE_EXPRESSION$E)
-
-# Compute actual variance of all principal components
-tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
-actual.var = tmp$d^2/sum(tmp$d^2)
-
-# Compute permuted variance of all principal components
-permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
-  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
-  tmp = svd(tmp.residual)
-  permuted.var = tmp$d^2/sum(tmp$d^2)
-}
-permuted.var = apply(permuted.var, 2, mean)
-NUM.SV = sum(actual.var >= permuted.var[1])
-
-# Plot variance components
-var.comp = data.frame(component = 1:length(actual.var), 
-                      residual = actual.var, 
-                      permuted.residual = permuted.var) %>%
-  tidyr::gather(data, value, -component)
-p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.3)), aes(x = component, y = value, color = data)) + geom_point()
-p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
-p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
-p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
-p
-
-# Estimate surrogate variables
-SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$E, 
-                    MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv
-SURR.VAR = data.frame(SURR.VAR)
-
-colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
-rownames(SURR.VAR) = rownames(MODEL1)
-
-# Re-estimate voom weights
-VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
-                        design = cbind(MODEL1, SURR.VAR), plot=F)
-
-# Fit linear model using new weights and new design
-VOOM.SVA.FIT = lmFit(CQN.GENE_EXPRESSION$E, 
-                     design = cbind(MODEL1, SURR.VAR),
-                     weights = VOOM.SVA.WEIGHTS$weights)
-
-# Residuals after normalisation
-RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
-                                                  CQN.GENE_EXPRESSION$E)
-
-# Add variable of interest back to the residuals
-varsToAddIn = grep("Diagnosis", colnames(MODEL1), value = T)
-RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
-  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
-
-# Assign NA values to outlier expressions
-RESIDUAL.SVA.GENE_EXPRESSION = apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, function(x){
-  mn = mean(x, na.rm = T)
-  std.dev = sd(x, na.rm = T)
-
-  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
-  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
-
-  x[x < (mn-3*std.dev)] = NA
-  x[x > (mn+3*std.dev)] = NA
-  return(x)
-}) %>% t
-```
-
-### Covariates Adjustments for Network Analysis
-Identified covariates are regressed out from the expression matrix for network analysis
-```{r network.adjust}
-# Get design matrix
-DESIGN.NET = getDesignMatrix(COVARIATES[, postAdjustCovars, drop = F], Intercept = F)
-DESIGN.NET = DESIGN.NET$design[,linColumnFinder(DESIGN.NET$design)$indepCols]
-
-# Estimate voom weights for dispersion control
-VOOM.NET.WEIGHTS = voom(NEW.COUNTS, 
-                        design=DESIGN.NET, plot=F)
-
-# Fit linear model using new weights and new design
-VOOM.NET.FIT = lmFit(CQN.GENE_EXPRESSION$E,
-                     design = DESIGN.NET,
-                     weights = VOOM.NET.WEIGHTS$weights)
-
-# Residuals after normalisation
-RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
-                                                  CQN.GENE_EXPRESSION$E)
-
-# Winsorise outlier expressions
-RESIDUAL.NET.GENE_EXPRESSION = apply(RESIDUAL.NET.GENE_EXPRESSION, 1, function(x){
-  mn = mean(x, na.rm = T)
-  std.dev = sd(x, na.rm = T)
-
-  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
-  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
-
-  x[x < (mn-3*std.dev)] = mn-3*std.dev
-  x[x > (mn+3*std.dev)] = mn+3*std.dev
-  return(x)
-}) %>% t
 ```
@@
   dataType = 'mRNA',
-  dataSubType = 'geneExp',
-  summaryLevel = 'gene',
+  dataSubType = 'isoformExp',
+  summaryLevel = 'transcript',
   assay	 = 'RNAseq',
@@
 # Code
-CODE <- Folder(name = "ROSMAP Reprocessed RNASeq (Normalised)", parentId = parentId)
+CODE <- Folder(name = "ROSMAP Reprocessed Isoform (Normalised)", parentId = parentId)
 annotations(CODE) = all.annotations
@@
   write.table(file = 'ROSMAP_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('ROSMAP_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
+COUNT_OBJ = File('ROSMAP_DLPFC_Counts.tsv', name = 'TPM (raw)', parentId = CODE$properties$id)
 annotations(COUNT_OBJ) = annotations(CODE)
@@
   write.table(file = 'ROSMAP_DLPFC_eCounts.tsv', sep = '\t', row.names=F, quote=F)
-ECOUNT_OBJ = File('ROSMAP_DLPFC_eCounts.tsv', name = 'Counts (estimated)', parentId = CODE$properties$id)
+ECOUNT_OBJ = File('ROSMAP_DLPFC_eCounts.tsv', name = 'TPM (estimated)', parentId = CODE$properties$id)
 annotations(ECOUNT_OBJ) = annotations(CODE)
-annotations(ECOUNT_OBJ)$dataSubType = 'filteredCounts'
+annotations(ECOUNT_OBJ)$dataSubType = 'estimatedCounts'
 ECOUNT_OBJ = synStore(ECOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
@@
 # Store logCPM
-CQN.GENE_EXPRESSION$y %>%
+VOOM.GENE_EXPRESSION$E %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
   write.table(file = 'ROSMAP_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('ROSMAP_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
+LCOUNT_OBJ = File('ROSMAP_DLPFC_logCPM.tsv', name = 'TPM (filtered)', parentId = CODE$properties$id)
 annotations(LCOUNT_OBJ) = annotations(CODE)
-annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
+annotations(LCOUNT_OBJ)$dataSubType = 'filteredLTPM'
 LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store cqn offsets
-CQN.GENE_EXPRESSION$offset %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'ROSMAP_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('ROSMAP_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
-annotations(OFFSET_OBJ) = annotations(CODE)
-annotations(OFFSET_OBJ)$dataSubType = 'offset'
-OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 
@@
 
-# Store surrogate variables
-SURR.VAR %>%
-  rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'ROSMAP_DLPFC_SurVar.tsv', sep = '\t', row.names=F, quote=F)
-SURR_OBJ = File('ROSMAP_DLPFC_SurVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
-annotations(SURR_OBJ) = annotations(CODE)
-annotations(SURR_OBJ)$dataSubType = 'surrogateVariables'
-SURR_OBJ = synStore(SURR_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store residual gene expression for eQTL analysis
-RESIDUAL.SVA.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'ROSMAP_DLPFC_eQTLResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-eEXP_OBJ = File('ROSMAP_DLPFC_eQTLResidualExpression.tsv', 
-                name = 'Normalised, covariates and surrogate variable adjusted residual expression (for eQTL)',
-                parentId = CODE$properties$id)
-annotations(eEXP_OBJ) = annotations(CODE)
-annotations(eEXP_OBJ)$dataSubType = 'residualGeneExpForeQTL'
-eEXP_OBJ = synStore(eEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store residual gene expression for network analysis
-RESIDUAL.NET.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'ROSMAP_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('ROSMAP_DLPFC_netResidualExpression.tsv', 
-                name = 'Normalised, covariates removed residual expression (for network analysis)', 
-                parentId = CODE$properties$id)
-annotations(nEXP_OBJ) = annotations(CODE)
-annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
-nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
 # Store differential expression results
@@
 |  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
-|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
 |  Design Matrix                              |  `r DM_OBJ$properties$id`        |
-|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
-|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
-|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
 |  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

### 2017-09-08 code/Rmd/MAYO_REPROCESSED_ISOFORM.Rmd
Added markdown to process MAYO isoform

--- 2017-09-05 code/Rmd/MAYO_REPROCESSED.Rmd
@@
 ---
-title: "Covariate analysis of MAYO reprocessed counts (with CQN normalisation and covariates adjusted)"
+title: "Covariate analysis of MAYO reprocessed transcripts (with TMM normalisation and covariates adjusted)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED_ISOFORM.Rmd",
                                  parentId = "syn8672415",
-                                 entityName = 'MAYO Reprocessed RNASeq (Normalised)')
+                                 entityName = "MAYO Reprocessed Isoform (Normalised)")
 ```
@@
 library(RColorBrewer)
-library(cqn)
-library(glmnet)
 
@@
 activityName = 'Covariate analysis';
-activityDescription = 'Covariate analysis of MAYO reprocessed RNASeq data';
-
-thisFileName <- 'MAYO_REPROCESSED.Rmd'
+activityDescription = 'Covariate analysis of MAYO reprocessed transcript data';
+
+thisFileName <- 'MAYO_REPROCESSED_ISOFORM.Rmd'
 
 # Github link
-thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
+thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_isoform')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
@@
 # Get reprocessed sample counts for temporal cortex
-COUNT_TC_ID <- 'syn8690799';#'syn7987765';
+COUNT_TC_ID <- 'syn10507725';
 ALL_USED_IDs <- COUNT_TC_ID
@@
 # Get reprocessed sample counts for cerebellum
-COUNT_CE_ID <- 'syn8690904';#'syn7513807';
+COUNT_CE_ID <- 'syn10499229';
 ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_CE_ID
@@
 COUNT = cbind(COUNT_TC, COUNT_CE[rownames(COUNT_TC),])
-
-# Convert rownames of counts from tracking id to ensemble gene id
-tmp = data.frame(Gene.ID = rownames(COUNT)) %>%
-  dplyr::mutate(ID = Gene.ID) %>%
-  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')
-rownames(tmp) = tmp$Gene.ID
-rownames(COUNT) = tmp[rownames(COUNT), 'ensembl_gene_id']
 
@@
 
-# Get picard metrics from synapse
-METADATA.PICARD_CE_ID <- 'syn8698214';#'syn8380392'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_CE_ID
-# METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = FALSE, exdir = getwd())
-# METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = TRUE, exdir = getwd())
-# METADATA.PICARD_CE_FPATHS = METADATA.PICARD_CE_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_CE_FPATHS)]
-# METADATA.PICARD_CE <- plyr::ldply(METADATA.PICARD_CE_FPATHS, read.csv)
-# 
-# METADATA.PICARD_CE$ID = gsub('picard/', '', METADATA.PICARD_CE_FPATHS) %>%
-#   gsub('/picard.CombinedMetrics.csv','',.) %>%
-#   sapply(function(x){str_split(x, '\\.')[[1]][1]})
-METADATA.PICARD_CE <- CovariateAnalysis::downloadFile(METADATA.PICARD_CE_ID) %>%
-  dplyr::rename(ID = sample)
-colnames(METADATA.PICARD_CE) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_CE))
-colnames(METADATA.PICARD_CE) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD_CE))
-
-METADATA.PICARD_TC_ID <- 'syn8698211';#'syn8383233'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_TC_ID
-# METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = FALSE, exdir = getwd())
-# METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = TRUE, exdir = getwd())
-# METADATA.PICARD_TC_FPATHS = METADATA.PICARD_TC_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_TC_FPATHS)]
-# METADATA.PICARD_TC <- plyr::ldply(METADATA.PICARD_TC_FPATHS, read.csv)
-# METADATA.PICARD_TC$ID = gsub('picard/', '', METADATA.PICARD_TC_FPATHS) %>%
-#   gsub('/picard.CombinedMetrics.csv','',.) %>%
-#   sapply(function(x){str_split(x, '\\.')[[1]][1]}) %>%
-#   gsub('Aligned','',.)
-
-METADATA.PICARD_TC <- CovariateAnalysis::downloadFile(METADATA.PICARD_TC_ID) %>%
-  dplyr::rename(ID = sample)
-colnames(METADATA.PICARD_TC) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_TC))
-colnames(METADATA.PICARD_TC) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD_TC))
+# Get PMI estimates from gene expression analysis
+METADATA_PMI = downloadFile('syn8466814') %>%
+  dplyr::select(SampleID, PMI) %>%
+  dplyr::rename(ID = SampleID)
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = 'syn8466814'
 
@@
 METADATA = rbindlist(list(METADATA_TC, METADATA_CE), use.names = T, fill = T) %>%
-  inner_join(rbindlist(list(METADATA.PICARD_TC, METADATA.PICARD_CE), use.names = T, fill = T)) %>%
+  left_join(METADATA_PMI) %>%
   as.data.frame
-
-# Get gene specific parameters from synapse
-GENE.PARAM = downloadFile('syn8449369')
-ALL_USED_IDs = c(ALL_USED_IDs, 'syn8449369')
-
-GENE.LEN = dplyr::select(GENE.PARAM, ensembl_gene_id, gene.length) %>% unique()
-rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id
-
-GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% unique()
-rownames(GENE.GC) = GENE.GC$ensembl_gene_id
 
@@
 
-### Predict missing PMI
-```{r predict.pmi}
-# Get PMI
-PMI = METADATA[,c('PMI', 'ID', 'Donor_ID'), drop = F]
-PMI$PMI = PMI$PMI
-
-# Get expression (convert counts to cpm)
-expr = voom(COUNT, design = NULL)$E
-ind = which(rowSums(expr>=1)/dim(expr)[2] >= 0.5)
-expr = expr[ind,]
-
-# Combine expression from different regions of brain of an individual into one
-expr = plyr::dlply(PMI, .(Donor_ID), .fun = function(x, expr){
-  data.frame(value = rowMeans(expr[,x$ID,drop = F], na.rm = T)) %>%
-    plyr::rename(c('value' = unique(x$Donor_ID)))
-}, expr, .parallel = F) %>%
-  do.call(cbind,.)
-
-# Split data into trainining and prediction set
-PMI = dplyr::select(PMI, Donor_ID, PMI) %>% unique()
-rownames(PMI) = PMI$Donor_ID
-PMI = PMI[colnames(expr),]
-
-# Fit a logistic regression model
-x.train = expr[,!is.na(PMI$PMI)]
-x.predict = expr[,is.na(PMI$PMI)]
-
-y.train = PMI[!is.na(PMI$PMI),]
-y.train = y.train[colnames(x.train), 'PMI']
-
-plr.model = cv.glmnet(t(x.train), y.train, family = 'gaussian')
-coeff = as.matrix(coef.glmnet(plr.model, s = 'lambda.min'))
-ssres = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train)^2)
-sstot = sum((y.train-mean(y.train, na.rm = T))^2, na.rm = T)
-R2.allgenes = 1-(ssres/sstot)
-
-y.predict.all = predict(plr.model, t(x.predict), s = 'lambda.min', type = 'response') %>%
-  rownameToFirstColumn('Donor_ID') %>%
-  plyr::rename(c('1' = 'PMI'))
-
-# Get genes that are predictors of PMI and use them as control probes
-CONTROL.PROBES_ID = 'syn6145639'
-ALL_USED_IDs = c(ALL_USED_IDs, CONTROL.PROBES_ID)
-CONTROL.PROBES = fread(synGet(CONTROL.PROBES_ID)@filePath, data.table = F) %>%
-  dplyr::mutate(ID = x) %>%
-  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')
-
-# Initial normalisation of gene expression
-ind = rownames(expr) %in% CONTROL.PROBES$ensembl_gene_id 
-expr = expr[ind,]
-
-# Fit a logistic regression model
-x.train = expr[,!is.na(PMI$PMI)]
-x.predict = expr[,is.na(PMI$PMI)]
-
-y.train = PMI[!is.na(PMI$PMI),]
-y.train = y.train[colnames(x.train), 'PMI']
-
-plr.model = cv.glmnet(t(x.train), y.train, family = 'gaussian')
-coeff = as.matrix(coef.glmnet(plr.model, s = 'lambda.min'))
-ssres = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train)^2)
-sstot = sum((y.train-mean(y.train, na.rm = T))^2, na.rm = T)
-R2.smallgenes = 1-(ssres/sstot)
-
-y.predict = predict(plr.model, t(x.predict), s = 'lambda.min', type = 'response') %>%
-  rownameToFirstColumn('Donor_ID') %>%
-  plyr::rename(c('1' = 'PMI'))
-
-PMI = PMI %>% 
-  dplyr::select(Donor_ID, PMI) %>%
-  dplyr::filter(!is.na(PMI)) %>%
-  list(y.predict) %>%
-  rbindlist(use.names = T, fill = T)
-PMI$PMI[PMI$PMI < 0] = 0
-
-METADATA = METADATA %>% 
-  dplyr::select(-PMI) %>%
-  left_join(PMI)
-```
-Fitted R2 values are, (i) with all gx features `r round(R2.allgenes, 3)`, (ii) with small set of inferred gx features `r round(R2.smallgenes,3)`
-
 ### Covariate clustering
@@
 FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis", "BrainRegion.ApoE")
-ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
-                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
+ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI")
 
@@
 
-```{r data.explore, fig.width = 12, fig.height = 8}
+```{r data.explore, fig.width = 12, fig.height = 5}
 my.theme = theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
@@
 
-# Ribosomal bases
-p[[4]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_RIBOSOMAL_BASES))) + geom_boxplot()
-p[[4]] = p[[4]] + ggtitle('Fraction of Ribosomal Bases') + my.theme
-
-# Intronic bases
-p[[5]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTRONIC_BASES))) + geom_boxplot()
-p[[5]] = p[[5]] + ggtitle('Fraction Intronic Bases') + my.theme
-
-# Intergenic bases
-p[[6]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTERGENIC_BASES))) + geom_boxplot()
-p[[6]] = p[[6]] + ggtitle('Fraction Intergenic Bases') + my.theme
-
 multiplot(plotlist = p, cols = 3)
 ```
-### Filter genes
-Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion x Diagnosis. Also remove genes with missing gene length and percentage GC content
+### Filter transcripts
+Remove genes that have less than 5 cpm counts in at least 50% of samples per BrainRegion x Diagnosis
 ```{r cpmnormalisation}
@@
     processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
-                                                   MIN_GENE_CPM=1, 
+                                                   MIN_GENE_CPM=5, 
                                                    MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
@@
 ### Library Normalisation
-Initial library normalisation usign cqn
-```{r cqn}
+Initial library normalisation usign tmm
+```{r tmm}
 # Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
-                          x = GENE.GC[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
-                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'gene.length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
-CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+TMM.GENE_EXPRESSION = calcNormFactors(NEW.COUNTS, method = 'TMM')
+VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design = NULL, lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
 
 # Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
-LOG.CPM = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
+LOG.CPM = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
   mn = mean(x, na.rm = T)
@@
 }) %>% t
-CQN.GENE_EXPRESSION$E = LOG.CPM
+VOOM.GENE_EXPRESSION$E = LOG.CPM
 NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
@@
   # Estimate voom weights for dispersion control
-  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F)
+  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F, lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
 
   # Calculate correlation values of random effects
-  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
+  VOOM.GENE_EXPRESSION$E = VOOM.GENE_EXPRESSION$E
   correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
@@
     # Fit linear model using new weights and new design
-    VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$E,
+    VOOM.ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION$E,
                               design = DM1$design,
@@
     RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
-                                                  CQN.GENE_EXPRESSION$E)
+                                                  VOOM.GENE_EXPRESSION$E)
   } else {
@@
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F, lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
 
 # Calculate correlation values of random effects
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+VOOM.WEIGHTS$E = VOOM.GENE_EXPRESSION$E
 correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS,
@@
                     block = COVARIATES$Donor_ID, 
-                    correlation = correlation$cor)
+                    correlation = correlation$cor,
+                    lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
 
 # Fit linear model using new weights and new design
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+VOOM.WEIGHTS$E = VOOM.GENE_EXPRESSION$E
 FIT = lmFit(VOOM.WEIGHTS$E,
@@
   tidyr::separate(to.state, c('BrainRegion.ag','against'), sep = '\\.') %>%
-  tidyr::unite(Comparison, reference, against, sep = '-') %>%
-  left_join(GENE.PARAM %>%
-              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
-              unique())
+  tidyr::unite(Comparison, reference, against, sep = '-')
 DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
@@
 tmp = DE %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion.ref) %>%
+  dplyr::select(Tx.ID, Comparison, Direction, BrainRegion.ref) %>%
   group_by(BrainRegion.ref, Comparison, Direction) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(Tx.ID))) %>%
   spread(Direction, FDR_0_05_FC_1.2) 
@@
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F, lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
 correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS,
@@
 # Calculate correlation values of random effects
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+VOOM.WEIGHTS$E = VOOM.GENE_EXPRESSION$E
 
@@
                 against = paste0('ApoE',against)) %>%
-  tidyr::unite(Comparison, reference, against, sep = '-') %>%
-  left_join(GENE.PARAM %>%
-              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
-              unique())
+  tidyr::unite(Comparison, reference, against, sep = '-')
 DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
@@
   tidyr::separate(to.state, c('BrainRegion.ag','Dx.ag','Gender.ag'), sep = '\\.') %>%
-  tidyr::unite(Comparison, Dx.ref, Dx.ag, sep = '-') %>%
-  left_join(GENE.PARAM %>%
-              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
-              unique())
+  tidyr::unite(Comparison, Dx.ref, Dx.ag, sep = '-')
 DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
@@
 tmp = DE %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion.ref, Gender.ref) %>%
+  dplyr::select(Tx.ID, Comparison, Direction, BrainRegion.ref, Gender.ref) %>%
   group_by(BrainRegion.ref, Gender.ref, Comparison, Direction) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(Tx.ID))) %>%
   spread(Direction, FDR_0_05_FC_1.2) 
@@
   rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis.Gender','',Comparison),
+  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
                 Study = 'MAYO',
@@
   tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
-  tidyr::separate(from.state, c('BrainRegion.ref','Dx.ref','Gender.ref'), sep = '\\.') %>%
-  tidyr::separate(to.state, c('BrainRegion.ag','Dx.ag','Gender.ag'), sep = '\\.') %>%
-  tidyr::unite(Comparison, Dx.ref, Dx.ag, sep = '-') %>%
-  left_join(GENE.PARAM %>%
-              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
-              unique())
+  tidyr::separate(from.state, c('BrainRegion.ref','Dx.ref','AOD.ref'), sep = '\\.') %>%
+  tidyr::separate(to.state, c('BrainRegion.ag','Dx.ag','AOD.ag'), sep = '\\.') %>%
+  tidyr::unite(Comparison, Dx.ref, Dx.ag, sep = '-')
 DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
@@
 tmp = DE %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion.ref, Gender.ref) %>%
-  group_by(BrainRegion.ref, Gender.ref, Comparison, Direction) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  dplyr::select(Tx.ID, Comparison, Direction, BrainRegion.ref, AOD.ref) %>%
+  group_by(BrainRegion.ref, AOD.ref, Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(Tx.ID))) %>%
   spread(Direction, FDR_0_05_FC_1.2) 
@@
 p = p + scale_color_manual(values = c('green','grey','red'))
-p = p + facet_grid(Comparison+.~BrainRegion.ref+Gender.ref, scales = 'fixed')
+p = p + facet_grid(Comparison+.~BrainRegion.ref+AOD.ref, scales = 'fixed')
 p
 
-all.diff.exp = c(all.diff.exp, list(Diagnosis.Gender = DE))
-all.fit = c(all.fit, list(Diagnosis.Gender = FIT))
-```
-
-### SVA Adjustments for eQTL analysis
-Conditioned on primary variable (BrainRegion.Diagnosis) and identified covariates estimate surrogate variables using SVA package
-```{r sva.adjust}
-# Get design matrix
-DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[1], postAdjustCovars), drop = F], Intercept = F)
-DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
-
-# Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
-
-# Calculate correlation values of random effects
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS,
-                                           block = COVARIATES$Donor_ID,
-                                           method = 'lmer')
-
-# Re-estimate voom weights
-VOOM.WEIGHTS = voom(NEW.COUNTS,
-                    design=DESIGN$design, plot=F,
-                    block = COVARIATES$Donor_ID, 
-                    correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-FIT = lmFit(VOOM.WEIGHTS$E,
-            weights = VOOM.WEIGHTS$weights,
-            design = VOOM.WEIGHTS$design,
-            block = COVARIATES$Donor_ID, 
-            correlation = correlation$cor)
-
-# Get (null) design matrix
-MODEL0 = DESIGN$design[,-grep(primaryVariable,colnames(DESIGN$design))] # Get null model by removing variable of interest
-MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
-MODEL1 = DESIGN$design
-MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
-
-# Source modified functions from sva package
-source('../R/irwsva.build.R')
-source('../R/f.pvalue.R')
-
-# Get residuals from differential expression model
-RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, CQN.GENE_EXPRESSION$E)
-
-# Compute actual variance of all principal components
-tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
-actual.var = tmp$d^2/sum(tmp$d^2)
-
-# Compute permuted variance of all principal components
-permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
-  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
-  tmp = svd(tmp.residual)
-  permuted.var = tmp$d^2/sum(tmp$d^2)
-}
-permuted.var = apply(permuted.var, 2, mean)
-NUM.SV = sum(actual.var >= permuted.var[1])
-
-# Plot variance components
-var.comp = data.frame(component = 1:length(actual.var), 
-                      residual = actual.var, 
-                      permuted.residual = permuted.var) %>%
-  tidyr::gather(data, value, -component)
-p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.2)), aes(x = component, y = value, color = data)) + geom_point()
-p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
-p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
-p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
-p
-
-# Estimate surrogate variables
-SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E,
-                        MODEL1, MODEL0, n.sv = NUM.SV, 
-                        B = 30, tol = 1e-20)$sv
-
-colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
-rownames(SURR.VAR) = rownames(MODEL1)
-
-# Estimate voom weights for dispersion control
-VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
-                        design=cbind(MODEL1, SURR.VAR), plot=F)
-
-# Calculate correlation values of random effects
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS,
-                                           block = COVARIATES$Donor_ID,
-                                           method = 'lmer')
-
-# Re-estimate voom weights
-VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
-                        design = cbind(MODEL1, SURR.VAR), plot=F,
-                        block = COVARIATES$Donor_ID, 
-                        correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS$E,
-                     weights = VOOM.SVA.WEIGHTS$weights,
-                     design = VOOM.SVA.WEIGHTS$design,
-                     block = COVARIATES$Donor_ID, 
-                     correlation = correlation$cor)
-
-# Residuals after normalisation
-RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
-                                                  CQN.GENE_EXPRESSION$E)
-
-# Add variable of interest back to the residuals
-varsToAddIn = grep("BrainRegion.Diagnosis", colnames(MODEL1), value = T)
-RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
-  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
-
-# Assign NA values to outlier expressions
-RESIDUAL.SVA.GENE_EXPRESSION = apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, function(x){
-  mn = mean(x, na.rm = T)
-  std.dev = sd(x, na.rm = T)
-
-  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
-  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
-
-  x[x < (mn-3*std.dev)] = NA
-  x[x > (mn+3*std.dev)] = NA
-  return(x)
-}) %>% t
-
-all.fit = c(all.fit, list(SVA = VOOM.SVA.FIT))
-```
-
-### Covariates Adjustments for network analysis
-Identified covariates are regressed out from the expression matrix for network analysis
-```{r network.adjust}
-# Get design matrix
-DESIGN.NET = getDesignMatrix(COVARIATES[, postAdjustCovars, drop = F], Intercept = F)
-DESIGN.NET = DESIGN.NET$design[,linColumnFinder(DESIGN.NET$design)$indepCols]
-
-# Estimate voom weights for dispersion control
-VOOM.NET.WEIGHTS = voom(NEW.COUNTS, 
-                        design=DESIGN.NET, plot=F)
-
-# Calculate correlation values of random effects
-VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS,
-                                           block = COVARIATES$Donor_ID,
-                                           method = 'lmer')
-
-# Re-estimate voom weights
-VOOM.NET.WEIGHTS = voom(NEW.COUNTS, 
-                        design = DESIGN.NET, plot=F,
-                        block = COVARIATES$Donor_ID, 
-                        correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS$E,
-                     weights = VOOM.NET.WEIGHTS$weights,
-                     design = VOOM.NET.WEIGHTS$design,
-                     block = COVARIATES$Donor_ID, 
-                     correlation = correlation$cor)
-
-# Residuals after normalisation
-RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
-                                                  CQN.GENE_EXPRESSION$E)
-
-# Winsorise outlier expressions
-RESIDUAL.NET.GENE_EXPRESSION = apply(RESIDUAL.NET.GENE_EXPRESSION, 1, function(x){
-  mn = mean(x, na.rm = T)
-  std.dev = sd(x, na.rm = T)
-
-  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
-  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
-
-  x[x < (mn-3*std.dev)] = mn-3*std.dev
-  x[x > (mn+3*std.dev)] = mn+3*std.dev
-  return(x)
-}) %>% t
-
-all.fit = c(all.fit, list(NET = VOOM.NET.FIT))
-```
-
+all.diff.exp = c(all.diff.exp, list(Diagnosis.AOD = DE))
+all.fit = c(all.fit, list(Diagnosis.AOD = FIT))
+```
 ### Store files in synapse
@@
 # Code
-CODE <- Folder(name = "MAYO Reprocessed RNASeq (Normalised)", parentId = parentId)
+CODE <- Folder(name = "MAYO Reprocessed Isoform (Normalised)", parentId = parentId)
 annotations(CODE) = all.annotations
@@
 PROCESSED_COUNTS$filteredExprMatrix$counts %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
+  rownameToFirstColumn('Tx.ID') %>%
   write.table(file = 'MAYO_CBE_TCX_Counts.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('MAYO_CBE_TCX_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
+COUNT_OBJ = File('MAYO_CBE_TCX_Counts.tsv', name = 'TPM (raw)', parentId = CODE$properties$id)
 annotations(COUNT_OBJ) = annotations(CODE)
@@
 NEW.COUNTS %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
+  rownameToFirstColumn('Tx.ID') %>%
   write.table(file = 'MAYO_CBE_TCX_eCounts.tsv', sep = '\t', row.names=F, quote=F)
-ECOUNT_OBJ = File('MAYO_CBE_TCX_eCounts.tsv', name = 'Counts (estimated)', parentId = CODE$properties$id)
+ECOUNT_OBJ = File('MAYO_CBE_TCX_eCounts.tsv', name = 'TPM (estimated)', parentId = CODE$properties$id)
 annotations(ECOUNT_OBJ) = annotations(CODE)
@@
 # Store logCPM
-CQN.GENE_EXPRESSION$y %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
+VOOM.GENE_EXPRESSION$E %>%
+  rownameToFirstColumn('Tx.ID') %>%
   write.table(file = 'MAYO_CBE_TCX_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('MAYO_CBE_TCX_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
+LCOUNT_OBJ = File('MAYO_CBE_TCX_logCPM.tsv', name = 'TPM (filtered)', parentId = CODE$properties$id)
 annotations(LCOUNT_OBJ) = annotations(CODE)
@@
 
-# Store cqn offsets
-CQN.GENE_EXPRESSION$offset %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MAYO_CBE_TCX_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('MAYO_CBE_TCX_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
-annotations(OFFSET_OBJ) = annotations(CODE)
-annotations(OFFSET_OBJ)$dataSubType = 'offset'
-OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
 # Store design matrix
-MODEL1 %>%
+DM1$design %>%
   rownameToFirstColumn('SampleID') %>%
@@
 DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store surrogate variables
-SURR.VAR %>%
-  rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'MAYO_CBE_TCX_SurVar.tsv', sep = '\t', row.names=F, quote=F)
-SURR_OBJ = File('MAYO_CBE_TCX_SurVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
-annotations(SURR_OBJ) = annotations(CODE)
-annotations(SURR_OBJ)$dataSubType = 'surrogateVariables'
-SURR_OBJ = synStore(SURR_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store residual gene expression for eQTL analysis
-RESIDUAL.SVA.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MAYO_CBE_TCX_eQTLResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-eEXP_OBJ = File('MAYO_CBE_TCX_eQTLResidualExpression.tsv', 
-                name = 'Normalised, covariates and surrogate variable adjusted residual expression (for eQTL)',
-                parentId = CODE$properties$id)
-annotations(eEXP_OBJ) = annotations(CODE)
-annotations(eEXP_OBJ)$dataSubType = 'residualGeneExpForeQTL'
-eEXP_OBJ = synStore(eEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store residual gene expression for network analysis
-RESIDUAL.NET.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MAYO_CBE_TCX_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('MAYO_CBE_TCX_netResidualExpression.tsv', 
-                name = 'Normalised, covariates removed residual expression (for network analysis)', 
-                parentId = CODE$properties$id)
-annotations(nEXP_OBJ) = annotations(CODE)
-annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
-nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 
@@
 |  Design Matrix                              |  `r DM_OBJ$properties$id`        |
-|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
-|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
-|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
 |  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

### 2017-09-08 code/Rmd/MAYO_REPROCESSED_ISOFORM.Rmd
Modified markdown to process MAYO isoform

@@
 
-### Associate differential expression results with gc content, gene length and average expression
-```{r associate.de, fig.height=10, fig.width=15}
-pl = list()
-pl[[1]] = ggplot(DE %>%
-                   dplyr::filter(Comparison == 'Control-AD'), 
-                 aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
-pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(BrainRegion.ref+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
-pl[[1]] = pl[[1]] + theme(legend.position = 'top')
-
-pl[[2]] = ggplot(DE %>%
-                   dplyr::filter(Comparison == 'Control-AD'), 
-                 aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
-pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(BrainRegion.ref+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
-pl[[2]] = pl[[2]] + theme(legend.position = 'top')
-
-pl[[3]] = ggplot(DE %>%
-                   dplyr::filter(Comparison == 'Control-AD'), 
-                 aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
-pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(BrainRegion.ref+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
-pl[[3]] = pl[[3]] + theme(legend.position = 'top')
-
-multiplot(plotlist = pl, cols = 3)
-```
-
 ### Differential expression analysis (with BrainRegion.ApoE as primary variable)
@@
   rownameToFirstColumn('Tx.ID') %>%
-  write.table(file = 'MAYO_CBE_TCX_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('MAYO_CBE_TCX_logCPM.tsv', name = 'TPM (filtered)', parentId = CODE$properties$id)
+  write.table(file = 'MAYO_CBE_TCX_tpm.tsv', sep = '\t', row.names=F, quote=F)
+LCOUNT_OBJ = File('MAYO_CBE_TCX_TPM.tsv', name = 'TPM (filtered)', parentId = CODE$properties$id)
 annotations(LCOUNT_OBJ) = annotations(CODE)

### 2017-09-13 code/Rmd/MAYO_REPROCESSED_ISOFORM.Rmd
Added markdown to process MSSM isoforms

@@
   rownameToFirstColumn('Tx.ID') %>%
-  write.table(file = 'MAYO_CBE_TCX_tpm.tsv', sep = '\t', row.names=F, quote=F)
+  write.table(file = 'MAYO_CBE_TCX_TPM.tsv', sep = '\t', row.names=F, quote=F)
 LCOUNT_OBJ = File('MAYO_CBE_TCX_TPM.tsv', name = 'TPM (filtered)', parentId = CODE$properties$id)

### 2017-09-13 code/Rmd/MSSM_REPROCESSED_ISOFORM.Rmd
Added markdown to process MSSM isoforms

--- 2017-08-29 code/Rmd/MSSM_REPROCESSED.Rmd
@@
 ---
-title: "Covariate analysis of MSSM reprocessed counts (with CQN normalisation and covariates adjusted)"
+title: "Covariate analysis of MSSM reprocessed transcripts (with TMM normalisation and covariates adjusted)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MSSM_REPROCESSED.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MSSM_REPROCESSED_ISOFORM.Rmd",
                                  parentId = "syn8672415",
-                                 entityName = 'MSSM Reprocessed RNASeq (Normalised)')
+                                 entityName = 'MSSM Reprocessed Isoform (Normalised)')
 ```
@@
 synapseLogin()
+source('../R/parallelDuplicateCorrelation.R')
 
@@
 activityName = 'Covariate analysis';
-activityDescription = 'Covariate analysis of MSSM reprocessed RNASeq data';
-
-thisFileName <- 'MSSM_REPROCESSED.Rmd'
+activityDescription = 'Covariate analysis of MSSM reprocessed transcript data';
+
+thisFileName <- 'MSSM_REPROCESSED_ISOFORM.Rmd'
 
 # Github link
-thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
+thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_isoform')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
@@
 # Get reprocessed sample counts for temporal cortex
-COUNT_ID <- 'syn8691099';#'syn7994853';
+COUNT_ID <- 'syn10507727';
 ALL_USED_IDs <- COUNT_ID
 COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F, row.names = 1)
-
-# Convert rownames of counts from tracking id to ensemble gene id
-tmp = data.frame(Gene.ID = rownames(COUNT)) %>%
-  dplyr::mutate(ID = Gene.ID) %>%
-  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')
-rownames(tmp) = tmp$Gene.ID
-rownames(COUNT) = tmp[rownames(COUNT), 'ensembl_gene_id']
 
@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
-# METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
-# METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
-# METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
-# METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)
-# METADATA.PICARD$sampleIdentifier = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
-#   gsub('.accepted_hits.sort.coord/picard.CombinedMetrics.csv','',.)
-
 METADATA.PICARD <- CovariateAnalysis::downloadFile(METADATA.PICARD_ID) %>%
@@
   dplyr::left_join(METADATA_AGE)
-
-# Get gene specific parameters from synapse
-GENE.PARAM = downloadFile('syn8449369')
-ALL_USED_IDs = c(ALL_USED_IDs, 'syn8449369')
-
-GENE.LEN = dplyr::select(GENE.PARAM, ensembl_gene_id, gene.length) %>% unique()
-rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id
-
-GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% unique()
-rownames(GENE.GC) = GENE.GC$ensembl_gene_id
 ```
@@
 FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrainRegion.Diagnosis", "CDR", "BrainRegion.ApoE")
-ContCovariates <- c("RIN", "RIN2", "AOD","PMI",  "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
-                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
+ContCovariates <- c("RIN", "RIN2", "AOD","PMI")
 
@@
 
-# Intronic bases
-p[[5]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTRONIC_BASES))) + geom_boxplot()
-p[[5]] = p[[5]] + ggtitle('Fraction Intronic Bases') + my.theme
-
-# Intergenic bases
-p[[6]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTERGENIC_BASES))) + geom_boxplot()
-p[[6]] = p[[6]] + ggtitle('Fraction Intergenic Bases') + my.theme
-
-multiplot(plotlist = p, cols = 3)
+multiplot(plotlist = p, cols = 2)
 ```
 ### Filter genes
-Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion x Diagnosis. Also remove genes with missing gene length and percentage GC content
+Remove genes that have less than 5 tpm counts in at least 50% of samples per BrainRegion x Diagnosis
 ```{r cpmnormalisation}
@@
     processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
-                                                     MIN_GENE_CPM=1, 
+                                                     MIN_GENE_CPM=5, 
                                                      MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
@@
 
-### Conditional Quantile Normalisation (CQN)
-Initial normalisation usign cqn and dispersion control with voom (with NULL design)
-```{r initial.voom.normalisation, fig.height=5, fig.width=5}
+### Library Normalisation
+Initial library normalisation usign tmm
+```{r tmm}
 # Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
-                          x = GENE.GC[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
-                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'gene.length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
-CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+TMM.GENE_EXPRESSION = calcNormFactors(NEW.COUNTS, method = 'TMM')
+VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design = NULL, lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
 
 # Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
-LOG.CPM = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
+LOG.CPM = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
   mn = mean(x, na.rm = T)
@@
 # Plot abberent distribution of logcpm counts
-tmp1 = (CQN.GENE_EXPRESSION$E) %>%
+tmp1 = (VOOM.GENE_EXPRESSION$E) %>%
   rownameToFirstColumn('Gene.ID') %>%
@@
 ### Normalisation (iterative design)
-Since many covariates are correlated, re-normalising and re-adjusting COUNTS with an iterative design matrix.
+Since many covariates are correlated, re-normalising and re-adjusting tpm counts with an iterative design matrix.
 
@@
   # Estimate voom weights for dispersion control
-  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F)
+  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F, lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
 
@@
                                              block = COVARIATES$individualIdentifier)
-  # correlation = list(); correlation$cor = 0.4
 
@@
                               block = COVARIATES$individualIdentifier, 
-                              correlation = correlation$cor)
+                              correlation = correlation$cor,
+                              lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
 
   # Fit linear model using new weights and new design
-  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$E,
+  VOOM.ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION$E,
                             design = DM1$design,
@@
   loopCount = loopCount + 1
-  # print(loopCount)
 }
@@
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F, lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
 
 # Calculate correlation values of random effects
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
 correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)
@@
                     block = COVARIATES$individualIdentifier, 
+                    correlation = correlation$cor,
+                    lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
+
+# Fit linear model using new weights and new design
+FIT = lmFit(VOOM.WEIGHTS$E,
+            weights = VOOM.WEIGHTS$weights,
+            design=DESIGN$design,
+            block = COVARIATES$individualIdentifier, 
             correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-FIT = lmFit(VOOM.WEIGHTS)
 
@@
 # Get differnetial expression
-DE = lapply(1:4, function(i, FIT){
+DE = lapply(1:dim(contrast)[2], function(i, FIT){
   topTable(FIT, coef=i, number = 50000, confint = T) %>%
-    rownameToFirstColumn('ensembl_gene_id')
+    rownameToFirstColumn('Tx.ID')
 }, FIT.CONTR)
@@
   dplyr::rename(BrainRegion = Comparison) %>%
-  dplyr::mutate(Comparison = 'CDR') %>%
-  left_join(GENE.PARAM %>%
-              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
-              unique())
+  dplyr::mutate(Comparison = 'CDR')
 DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
@@
 tmp = DE %>%
-  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion) %>%
+  dplyr::select(Tx.ID, Comparison, Direction, BrainRegion) %>%
   group_by(BrainRegion, Comparison, Direction) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(Tx.ID))) %>%
   spread(Direction, FDR_0_05_FC_1.2) 
@@
                     block = COVARIATES$individualIdentifier, 
+                    correlation = correlation$cor, 
+                    lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
+
+# Fit linear model using new weights and new design
+FIT = lmFit(VOOM.WEIGHTS$E,
+            weights = VOOM.WEIGHTS$weights,
+            design=DESIGN$design,
+            block = COVARIATES$individualIdentifier, 
             correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-FIT = lmFit(VOOM.WEIGHTS)
 
@@
 for (region in c('FP', 'STG', 'PHG', 'IFG')){
-  tmp = combn(paste0('BrainRegion.Diagnosis', region, '.', c('CONTROL', 'AD', 'OTHER')),2)
+  tmp = combn(paste0('BrainRegion.Diagnosis', region, '.', c('AD', 'OTHER', 'CONTROL')),2)
   cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
@@
   dplyr::mutate(Comparison = paste(reference, against, sep = '-')) %>%
-  dplyr::select(-BrainRegion1) %>%
-  left_join(GENE.PARAM %>%
-              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
-              unique())
+  dplyr::select(-BrainRegion1)
 DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
@@
 all.fit = c(all.fit, list(Diagnosis = FIT))
-```
-
-### Associate differential expression results with gc content, gene length and average expression
-```{r associate.de, fig.height=15, fig.width=15}
-pl = list()
-pl[[1]] = ggplot(DE %>% dplyr::filter(Comparison == 'CONTROL-AD'), 
-                 aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
-pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(BrainRegion~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
-pl[[1]] = pl[[1]] + theme(legend.position = 'top')
-
-pl[[2]] = ggplot(DE %>% dplyr::filter(Comparison == 'CONTROL-AD'), 
-                 aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
-pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(BrainRegion~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
-pl[[2]] = pl[[2]] + theme(legend.position = 'top')
-
-pl[[3]] = ggplot(DE %>% dplyr::filter(Comparison == 'CONTROL-AD'), 
-                 aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
-pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(BrainRegion~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
-pl[[3]] = pl[[3]] + theme(legend.position = 'top')
-
-multiplot(plotlist = pl, cols = 3)
 ```
@@
                     block = COVARIATES$individualIdentifier, 
+                    correlation = correlation$cor,
+                    lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
+
+# Fit linear model using new weights and new design
+FIT = lmFit(VOOM.WEIGHTS$E,
+            weights = VOOM.WEIGHTS$weights,
+            design=DESIGN$design, 
+            block = COVARIATES$individualIdentifier, 
             correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-FIT = lmFit(VOOM.WEIGHTS)
 
@@
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F, lib.size = TMM.GENE_EXPRESSION * LIB.SIZE)
 
 # Calculate correlation values of random effects
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
 correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)
@@
          -0.5*BrainRegion.Diagnosis.SEXSTG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.MALE')
-contrast = makeContrasts(contrasts=cntr,
+contrast = makeContrasts(contrasts = cntr,
                          levels = colnames(FIT$coefficients))
@@
                 Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
-                Direction = as.character(Direction)) %>%
-  left_join(GENE.PARAM %>%
-              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
-              unique())
+                Direction = as.character(Direction))
 DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
@@
 # Estimate voom weights for dispersion control
-VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F, lib.size = LIB.SIZE * TMM.GENE_EXPRESSION)
 
 # Calculate correlation values of random effects
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
 correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)
@@
                     block = COVARIATES$individualIdentifier, 
+                    correlation = correlation$cor,
+                    lib.size = LIB.SIZE * TMM.GENE_EXPRESSION)
+
+# Fit linear model using new weights and new design
+FIT = lmFit(VOOM.WEIGHTS$E,
+            weights = VOOM.WEIGHTS$weights,
+            design=DESIGN$design,
+            block = COVARIATES$individualIdentifier, 
             correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-FIT = lmFit(VOOM.WEIGHTS)
 
@@
 all.fit = c(all.fit, list(Diagnosis.AOD = FIT))
-```
-
-### SVA Adjustments for eQTL analysis
-Conditioned on primary variable (BrainRegion.Diagnosis) and identified covariates estimate surrogate variables using SVA package
-```{r sva.adjust}
-# Get design matrix
-DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[2], postAdjustCovars), drop = F], Intercept = F)
-DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
-
-# Get (null) design matrix
-MODEL0 = DESIGN$design[,-grep(primaryVariable[2],colnames(DESIGN$design))] # Get null model by removing variable of interest
-MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
-MODEL1 = DESIGN$design
-MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
-
-# Source modified functions from sva package
-source('../R/irwsva.build.R')
-source('../R/f.pvalue.R')
-
-# Get residuals from differential expression model
-RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, CQN.GENE_EXPRESSION$E)
-
-# Compute actual variance of all principal components
-tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
-actual.var = tmp$d^2/sum(tmp$d^2)
-
-# Compute permuted variance of all principal components
-permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
-  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
-  tmp = svd(tmp.residual)
-  permuted.var = tmp$d^2/sum(tmp$d^2)
-}
-permuted.var = apply(permuted.var, 2, mean)
-NUM.SV = sum(actual.var >= permuted.var[1])
-
-# Plot variance components
-var.comp = data.frame(component = 1:length(actual.var), 
-                      residual = actual.var, 
-                      permuted.residual = permuted.var) %>%
-  tidyr::gather(data, value, -component)
-p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.2)), aes(x = component, y = value, color = data)) + geom_point()
-p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
-p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
-p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
-p
-
-# Estimate surrogate variables
-SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$E,
-                    MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv
-
-colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
-rownames(SURR.VAR) = rownames(MODEL1)
-
-# Estimate voom weights for dispersion control
-VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
-                        design=cbind(MODEL1, SURR.VAR), plot=F)
-
-# Calculate correlation values of random effects
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$individualIdentifier)
-
-# Re-estimate voom weights
-VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
-                        design = cbind(MODEL1, SURR.VAR), plot=F,
-                        block = COVARIATES$individualIdentifier, 
-                        correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)
-
-# Residuals after normalisation
-RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
-                                                  CQN.GENE_EXPRESSION$E)
-
-# Add variable of interest back to the residuals
-varsToAddIn = grep("BrainRegion.Diagnosis", colnames(MODEL1), value = T)
-RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
-  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
-```
-
-### Covariates Adjustments for network analysis
-Identified covariates are regressed out from the expression matrix for network analysis
-```{r network.adjust}
-# Get design matrix
-DESIGN.NET = getDesignMatrix(COVARIATES[, postAdjustCovars, drop = F], Intercept = F)
-DESIGN.NET = DESIGN.NET$design[,linColumnFinder(DESIGN.NET$design)$indepCols]
-
-# Estimate voom weights for dispersion control
-VOOM.NET.WEIGHTS = voom(NEW.COUNTS, 
-                        design=DESIGN.NET, plot=F)
-
-# Calculate correlation values of random effects
-VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$individualIdentifier)
-
-# Re-estimate voom weights
-VOOM.NET.WEIGHTS = voom(NEW.COUNTS, 
-                        design = DESIGN.NET, plot=F,
-                        block = COVARIATES$individualIdentifier, 
-                        correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E
-VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)
-
-# Residuals after normalisation
-RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
-                                                  CQN.GENE_EXPRESSION$E)
 ```
@@
 # Code
-CODE <- Folder(name = "MSSM Reprocessed RNASeq (Normalised)", parentId = parentId)
+CODE <- Folder(name = "MSSM Reprocessed Isoform (Normalised)", parentId = parentId)
 annotations(CODE) = all.annotations
@@
 
-# Store filtered counts
+# Store filtered TPM
 PROCESSED_COUNTS$filteredExprMatrix$counts %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MSSM_FP_STG_PHG_IFG_Counts.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
+  rownameToFirstColumn('Tx.ID') %>%
+  write.table(file = 'MSSM_FP_STG_PHG_IFG_TPM.tsv', sep = '\t', row.names=F, quote=F)
+COUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_TPM.tsv', name = 'TPM (raw)', parentId = CODE$properties$id)
 annotations(COUNT_OBJ) = annotations(CODE)
-annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
+annotations(COUNT_OBJ)$dataSubType = 'filteredTPM'
 COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 
-# Store filtered counts
+# Store filtered TPM
 NEW.COUNTS %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MSSM_FP_STG_PHG_IFG_eCounts.tsv', sep = '\t', row.names=F, quote=F)
-ECOUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_eCounts.tsv', name = 'Counts (estimated)', parentId = CODE$properties$id)
+  rownameToFirstColumn('Tx.ID') %>%
+  write.table(file = 'MSSM_FP_STG_PHG_IFG_eTPM.tsv', sep = '\t', row.names=F, quote=F)
+ECOUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_eTPM.tsv', name = 'TPM (estimated)', parentId = CODE$properties$id)
 annotations(ECOUNT_OBJ) = annotations(CODE)
-annotations(ECOUNT_OBJ)$dataSubType = 'estimatedCounts'
+annotations(ECOUNT_OBJ)$dataSubType = 'estimatedTPM'
 ECOUNT_OBJ = synStore(ECOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
@@
 # Store logCPM
-CQN.GENE_EXPRESSION$y %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
+VOOM.GENE_EXPRESSION$E %>%
+  rownameToFirstColumn('Tx.ID') %>%
   write.table(file = 'MSSM_FP_STG_PHG_IFG_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
+LCOUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_logCPM.tsv', name = 'Log TPM (filtered)', parentId = CODE$properties$id)
 annotations(LCOUNT_OBJ) = annotations(CODE)
@@
 
-# Store cqn offsets
-CQN.GENE_EXPRESSION$offset %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MSSM_FP_STG_PHG_IFG_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('MSSM_FP_STG_PHG_IFG_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
-annotations(OFFSET_OBJ) = annotations(CODE)
-annotations(OFFSET_OBJ)$dataSubType = 'offset'
-OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
 # Store design matrix
-MODEL1 %>%
+DM1$design %>%
   rownameToFirstColumn('SampleID') %>%
@@
 synChangeEntityACL(DM_OBJ@properties$id, new.racl)
-
-# Store surrogate variables
-SURR.VAR %>%
-  rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'MSSM_FP_STG_PHG_IFG_SurVar.tsv', sep = '\t', row.names=F, quote=F)
-SURR_OBJ = File('MSSM_FP_STG_PHG_IFG_SurVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
-annotations(SURR_OBJ) = annotations(CODE)
-annotations(SURR_OBJ)$dataSubType = 'surrogateVariables'
-SURR_OBJ = synStore(SURR_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store residual gene expression for eQTL analysis
-RESIDUAL.SVA.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MSSM_FP_STG_PHG_IFG_eQTLResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-eEXP_OBJ = File('MSSM_FP_STG_PHG_IFG_eQTLResidualExpression.tsv', 
-                name = 'Normalised, covariates and surrogate variable adjusted residual expression (for eQTL)',
-                parentId = CODE$properties$id)
-annotations(eEXP_OBJ) = annotations(CODE)
-annotations(eEXP_OBJ)$dataSubType = 'residualGeneExpForeQTL'
-eEXP_OBJ = synStore(eEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
-
-# Store residual gene expression for network analysis
-RESIDUAL.NET.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'MSSM_FP_STG_PHG_IFG_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('MSSM_FP_STG_PHG_IFG_netResidualExpression.tsv', 
-                name = 'Normalised, covariates removed residual expression (for network analysis)', 
-                parentId = CODE$properties$id)
-annotations(nEXP_OBJ) = annotations(CODE)
-annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
-nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 
@@
 |  Covariates                                 |  `r COV_OBJ$properties$id`       |
-|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
-|  Counts (estimated)                         |  `r ECOUNT_OBJ$properties$id`    |
-|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
-|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
+|  TPM (raw)                                  |  `r COUNT_OBJ$properties$id`     |
+|  TPM (estimated)                            |  `r ECOUNT_OBJ$properties$id`    |
+|  TPM (lcpm)                                 |  `r LCOUNT_OBJ$properties$id`    |
 |  Design Matrix                              |  `r DM_OBJ$properties$id`        |
-|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
-|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
-|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
 |  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

### 2017-09-15 code/Rmd/MSSM_REPROCESSED_ISOFORM.Rmd
Minor modifications

@@
 
-# Ribosomal bases
-p[[4]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_RIBOSOMAL_BASES))) + geom_boxplot()
-p[[4]] = p[[4]] + ggtitle('Fraction of Ribosomal Bases') + my.theme
-
-multiplot(plotlist = p, cols = 2)
+multiplot(plotlist = p, cols = 3)
 ```

### 2017-09-15 code/Rmd/ROSMAP_REPROCESSED_ISOFORM.Rmd
Minor modifications

@@
 ---
-title: "Covariate analysis of ROSMAP reprocessed isfoform counts (with CQN normalisation and covariates adjusted)"
+title: "Covariate analysis of ROSMAP reprocessed transcript counts (with TMM normalisation and covariates adjusted)"
 author: "Thanneer Perumal"
@@
 CODE <- synStore(CODE)
-old.acl = synGetEntityACL(CODE@properties$id)
+# old.acl = synGetEntityACL(CODE@properties$id)
 synChangeEntityACL(CODE@properties$id, new.racl)
@@
 PROCESSED_COUNTS$filteredExprMatrix$counts %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
+  rownameToFirstColumn('Tx.ID') %>%
   write.table(file = 'ROSMAP_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
@@
 NEW.COUNTS %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
+  rownameToFirstColumn('Tx.ID') %>%
   write.table(file = 'ROSMAP_DLPFC_eCounts.tsv', sep = '\t', row.names=F, quote=F)
@@
 VOOM.GENE_EXPRESSION$E %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
+  rownameToFirstColumn('Tx.ID') %>%
   write.table(file = 'ROSMAP_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
@@
 # Store design matrix
-MODEL1 %>%
+DM1$design %>%
   rownameToFirstColumn('SampleID') %>%
@@
 # Delete folder acl
-synChangeEntityACL(CODE@properties$id, old.acl@resourceAccess)
+synDeleteEntityACL(CODE@properties$id)
 
-stopCluster(cl)
+# stopCluster(cl)
 ```

### 2017-09-16 code/Rmd/CMC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Added markdown to process DLPFC samples in CMC

--- 2017-07-29 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)"
+title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN normlaisation)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
-                                 parentId = "syn9872607",
-                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
+knit2synapse::knitToFolderEntity(file = "./CMC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
+                                 parentId = "syn7501880",
+                                 entityName = 'DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
 ```
@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = "syn9872607";
+parentId = "syn7501880";
 activityName = 'Covariate adjustments';
-activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (ACC and DLPFC)';
-
-thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd'
+activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (DLPFC)';
+
+thisFileName <- 'CMC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd'
 
@@
 ```{r download.data, cache=TRUE}
-# Download reprocessed counts (ACC)
-COUNT_ACC_ID = 'syn8414424'
-ALL_USED_IDs = COUNT_ACC_ID
-COUNT_ACC = downloadFile(COUNT_ACC_ID)
-
 # Download reprocessed counts (DLPFC)
 COUNT_DLPFC_ID = 'syn8413221';
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = COUNT_DLPFC_ID
-COUNT_DLPFC = downloadFile(COUNT_DLPFC_ID)
-
-# Join ACC and DLPFC together
-COUNT = join_all(list(COUNT_ACC, COUNT_DLPFC), type = 'full') %>%
+ALL_USED_IDs = COUNT_DLPFC_ID
+COUNT = downloadFile(COUNT_DLPFC_ID) %>%
   as.data.frame()
 
-GENE.LEN = dplyr::select(COUNT_ACC, Geneid, Length) %>%
+GENE.LEN = dplyr::select(COUNT, Geneid, Length) %>%
   dplyr::rename(ensembl_gene_id = Geneid) %>%
@@
 # Get RNA qc metrics from synapse
-METADATA_QC_ACC_ID = 'syn2929053'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_ACC_ID
-METADATA_QC_ACC = read.table(synGet(METADATA_QC_ACC_ID)@filePath, fill = T, header = T, sep = ',') %>%
-  lapply(as.character)
-
-# Get RNA qc metrics from synapse
 METADATA_QC_DLPFC_ID = 'syn2279446'
@@
 METADATA_QC_DLPFC = read.table(synGet(METADATA_QC_DLPFC_ID)@filePath, fill = T, header = T, sep = ',') %>%
-  lapply(as.character)
-
-METADATA_QC = rbindlist(list(METADATA_QC_ACC, METADATA_QC_DLPFC), use.names = T, fill = T) %>%
   dplyr::filter(is.na(Exclude.)) %>%
@@
 
-                          "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
-                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?",
-
                           "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
@@
 colnames(METADATA) = gsub('[^[:alnum:]]','_',colnames(METADATA))
-
-METADATA_ACC = METADATA %>%
-  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
-                DNA_genotyping_report__Genotyping_Sample_ID, ACC_RNA_isolation__Exclude_, 
-                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_) %>%
-  dplyr::rename(PMI = PMI__in_hours_, 
-                IsolationExclude = ACC_RNA_isolation__Exclude_, 
-                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
-                RIN = ACC_RNA_isolation__RIN, 
-                ReportExclude = ACC_RNA_report__Exclude_,
-                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
-  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
-  dplyr::mutate(Tissue = 'ACC')
 
@@
 
-METADATA = rbindlist(list(METADATA_ACC, METADATA_DLPFC), use.names = T, fill = T) %>%
+METADATA = METADATA_DLPFC %>%
   droplevels() %>%
-  list(.,METADATA_QC) %>%
+  list(.,METADATA_QC_DLPFC) %>%
   join_all %>%
@@
 # Based on PC
-indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250',
-                'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220',
-                'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
+indToRemove = c('MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
                 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
-                'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_ACC_310",
-                "MSSM_RNA_ACC_51",  "MSSM_RNA_ACC_254", "MSSM_RNA_ACC_28", "MSSM_RNA_ACC_293", "MSSM_RNA_ACC_233",
-                "MSSM_RNA_ACC_167", "MSSM_RNA_ACC_353", "MSSM_RNA_ACC_355", "MSSM_RNA_ACC_211", "PENN_RNA_ACC_122",
-                "PENN_RNA_ACC_049", "PENN_RNA_ACC_052", "PENN_RNA_ACC_068", "PENN_RNA_ACC_070", "PITT_RNA_ACC_10020",
-                "PITT_RNA_ACC_1240", "PITT_RNA_ACC_1284", "PITT_RNA_ACC_700", "PITT_RNA_ACC_857", "PITT_RNA_ACC_BP_1589",
-                "MSSM_RNA_PFC_230", "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
+                'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_PFC_230",
+                "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
                 "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")
@@
 VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
-correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
-
-# Re-calculate voom weights with correlation of random effects
-VOOM.GENE_EXPRESSION = voom(NEW.COUNTS,   
-                            design = DESIGN, plot=F,
-                            block = COVARIATES$Individual_ID, 
-                            correlation = correlation$cor)
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
 
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
-                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
-                                     "Dx.TissueOther_ACC-Dx.TissueControl_ACC",
-
-                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
+contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                      "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC",
-                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
-
-                                     "Dx.TissueSCZ_DLPFC-Dx.TissueSCZ_ACC",
-                                     "Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",
-                                     "Dx.TissueOther_DLPFC-Dx.TissueOther_ACC",
-
-                                     "Dx.TissueSCZ_DLPFC+Dx.TissueSCZ_ACC-Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",         
-                                     "Dx.TissueSCZ_DLPFC+Dx.TissueControl_DLPFC-Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
-                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC-Dx.TissueSCZ_ACC+Dx.TissueControl_ACC"),
+                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC"),
                          levels = colnames(ADJUSTED.FIT$coefficients))
@@
 names(DE) = gsub('Dx.Tissue', '', colnames(contrast))
-names(DE)[10:12] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC', 'ALL_ALL-ALL_ALL')
 
@@
 p
-
-p = DE %>%
-  dplyr::filter(ref.Dx == 'ALL', to.Dx == 'ALL') %>%
-  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
-p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
-p = p + facet_grid(ref.Tissue+.~to.Tissue, scales = 'fixed')
-p = p + ggtitle('Differential Expression between DLPFC and ACC')
-p
 ```
@@
 VOOM.GENE_EXPRESSION1$E = CQN.GENE_EXPRESSION$E
-correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION1, block = COVARIATES$Individual_ID)
-
-# Re-calculate voom weights with correlation of random effects
-VOOM.GENE_EXPRESSION1 = voom(NEW.COUNTS,   
-                             design = DESIGN1, plot=F,
-                             block = COVARIATES$Individual_ID, 
-                             correlation = correlation$cor)
-VOOM.GENE_EXPRESSION1$E = CQN.GENE_EXPRESSION$E
 
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.Tissue.GenderSCZ_ACC_Male-Dx.Tissue.GenderControl_ACC_Male",
-                                     "Dx.Tissue.GenderSCZ_ACC_Female-Dx.Tissue.GenderControl_ACC_Female",
-
-                                     "Dx.Tissue.GenderSCZ_DLPFC_Male-Dx.Tissue.GenderControl_DLPFC_Male",
+contrast = makeContrasts(contrasts=c("Dx.Tissue.GenderSCZ_DLPFC_Male-Dx.Tissue.GenderControl_DLPFC_Male",
                                      "Dx.Tissue.GenderSCZ_DLPFC_Female-Dx.Tissue.GenderControl_DLPFC_Female"),
@@
 
-### Differential expression analysis (ACC - DLPFC, SCZ - Control, Female - Male)
-Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp2, fig.height=16, fig.width=8}
-COVARIATES = tidyr::separate(COVARIATES, Dx.Tissue, c('Dx', 'Tissue')) %>%
-  dplyr::mutate(Dx = factor(Dx, levels = c('SCZ', 'Control', 'Other')),
-                Tissue = factor(Tissue))
-postAdjustCovars = c('Tissue', 'Dx', 'Institution', 'Gender', 'RIN', 'IntronicRate')
-
-# Post adjusted design matrix
-DESIGN2 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
-DESIGN2 = DESIGN2[,linColumnFinder(DESIGN2)$indepCols]
-
-# Estimate voom weights
-VOOM.GENE_EXPRESSION2 = voom(NEW.COUNTS, design = DESIGN2)
-VOOM.GENE_EXPRESSION2$E = CQN.GENE_EXPRESSION$E
-correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION2, block = COVARIATES$Individual_ID)
-
-# Re-calculate voom weights with correlation of random effects
-VOOM.GENE_EXPRESSION2 = voom(NEW.COUNTS,   
-                             design = DESIGN2, plot=F,
-                             block = COVARIATES$Individual_ID, 
-                             correlation = correlation$cor)
-VOOM.GENE_EXPRESSION2$E = CQN.GENE_EXPRESSION$E
-
-# Fit linear model using new weights and new design
-ADJUSTED.FIT2 = lmFit(VOOM.GENE_EXPRESSION2)
-
-# Fit contrast
-contrast = makeContrasts(contrasts=c("TissueACC-TissueDLPFC",
-                                     "DxSCZ-DxControl",
-                                     "GenderFemale"),
-                         levels = colnames(ADJUSTED.FIT2$coefficients))
-FIT.CONTR2 = contrasts.fit(ADJUSTED.FIT2, contrasts=contrast)
-FIT.CONTR2 = eBayes(FIT.CONTR2)
-
-# Get differnetial expression
-DE2 = lapply(1:dim(contrast)[2], function(i, FIT){
-  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
-    rownameToFirstColumn('ensembl_gene_id') 
-}, FIT.CONTR2) 
-names(DE2) = c('ACC-DLPFC', 'SCZ-CONTROL', 'FEMALE-MALE')
-
-DE2 = DE2 %>% 
-  rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Direction = sign(logFC),
-                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
-                Direction = as.character(Direction)) %>%
-  left_join(Ensemble2HGNC) 
-DE2$Direction[DE2$adj.P.Val > 0.05 | abs(DE2$logFC) < log2(1.2)] = 'NONE'
-
-tmp = DE2 %>%
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
-  group_by(Comparison) %>%
-  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
-                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))
-
-tmp1 = DE2 %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
-  group_by(Comparison) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
-                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))
-
-kable(full_join(tmp,tmp1))
-
-p = DE2 %>%
-  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
-p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
-p = p + facet_grid(Comparison~., scales = 'free_y')
-p = p + ggtitle('Differential Expression')
-p
-```
-
 ### Gene set enrichment analysis
@@
                       tidyr::unite(Comparison2, ref.Dx, ref.Tissue, ref.Gender) %>%
-                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
-                    DE2) %>%
+                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-')) %>%
   data.table::rbindlist(fill = TRUE, use.names = TRUE)
@@
 
-  tissueTypeAbrv	= 'DLPFC, ACC', 
+  tissueTypeAbrv	= 'DLPFC', 
   study = 'CMC',
@@
 # Code
-CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
+CODE <- Folder(name = "DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
 annotations(CODE) = all.annotations
@@
 COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
-write.table(COVARIATES, file = 'CMC_ACC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
-COV_OBJ = File('CMC_ACC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+write.table(COVARIATES, file = 'CMC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
+COV_OBJ = File('CMC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
 annotations(COV_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
+  write.table(file = 'CMC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
+COUNT_OBJ = File('CMC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
 annotations(COUNT_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
+  write.table(file = 'CMC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
+LCOUNT_OBJ = File('CMC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
 annotations(LCOUNT_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('CMC_ACC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
+  write.table(file = 'CMC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
+OFFSET_OBJ = File('CMC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
 annotations(OFFSET_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'CMC_ACC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('CMC_ACC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+  write.table(file = 'CMC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
+DM_OBJ = File('CMC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
 annotations(DM_OBJ) = annotations(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_ACC_DLPFC_netResidualExpression.tsv', 
+  write.table(file = 'CMC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+nEXP_OBJ = File('CMC_DLPFC_netResidualExpression.tsv', 
                 name = 'Normalised, covariates removed residual expression (for network analysis)', 
@@
 # Store differential expression results
-write.table(all.diff.exp, file = 'CMC_ACC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_ACC_DLPFC_DiffExpression.tsv', 
+write.table(all.diff.exp, file = 'CMC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('CMC_DLPFC_DiffExpression.tsv', 
                 name = 'Differential Expression Results (Dx.Tissue)', 
@@
 # Store enrichment results
-write.table(enrichment.results, file = 'CMC_ACC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
-ENRICH_OBJ = File('CMC_ACC_DLPFC_EnrichResults.tsv', 
+write.table(enrichment.results, file = 'CMC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
+ENRICH_OBJ = File('CMC_DLPFC_EnrichResults.tsv', 
                   name = 'Enrichment Analysis Results (Dx.Tissue)', 

### 2017-09-18 code/Rmd/networkAssociationTest.Rmd
Added markdown to perform assocition test b/w modules and Dx effects

---
title: "Function to find association b/w n significant eigene genes with Diagnosis"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./networkAssociationTest.Rmd",
                                 parentId = "syn10306449",
                                 entityName = 'Network Association Tests')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

# Load Libraries
library(CovariateAnalysis)
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(synapseClient)
library(knitr)
library(githubr)
library(ggplot2)
library(ComplexHeatmap)
library(circlize)

library(doParallel)
library(foreach)

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

synapseLogin()

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn10306449';
activityName = 'Network associaiton statistics';

thisFileName <- 'networkAssociationTest.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_networks')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```

### Data download
```{r download.data, cache=TRUE}
# Download data
MODULES = synTableQuery('SELECT * FROM syn10338156')@values

MOD.MEMBERS = plyr::dlply(MODULES, .(ModuleNameFull), .fun = function(x) unique(x$GeneID))

COVARIATES = downloadFile('syn10160582')
COVARIATES$BrainRegion[COVARIATES$BrainRegion == 'CER'] = 'CBE'

FactorCovariates = c("individualIdentifier", "Study", "BrainRegion.Diagnosis", "BrainRegion", 
                     "Diagnosis", "Gender", "Batch", "BrainRegion.ApoE", "BrainRegion.Diagnosis.Gender")
ContCovariates = c("AOD", "PMI", "RIN", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES", "PCT_INTERGENIC_BASES",
                   "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES", "RIN2") 
COVARIATES[,FactorCovariates] = lapply(COVARIATES[, FactorCovariates], as.factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[, ContCovariates], as.numeric)

EXPR = downloadFile('syn10160584')
gene.id.map = data.frame(ID = EXPR$Gene.ID, Gene.ID = EXPR$Gene.ID) %>%
  tidyr::separate(ID, c('GeneID','position'), sep = '\\.')
EXPR = left_join(EXPR, gene.id.map)
rownames(EXPR) = EXPR$GeneID
EXPR$Gene.ID = NULL
EXPR$GeneID = NULL

MODULE.OVRLP = synTableQuery('SELECT * FROM syn10339153')@values

MODULE.ENRICH = synTableQuery('SELECT * FROM syn10492048')@values

MAGMA.ENRICH = synTableQuery('SELECT * FROM syn10380432')@values
```

```{r fxn}
# Function to find association statistics
findAssociationStatistics <- function(x, COVARIATES, EXPR, FORMULA){

  covar = dplyr::filter(COVARIATES, BrainRegion == unique(x$brainRegion))
  expr = EXPR[x$GeneID,covar$SampleID]

  result = list( assoc.stat = data.frame(total.eff.sz = 0,
                                         ngenes = length(unique(x$GeneID)),
                                         npc = 0,
                                         npc.sig = 0,
                                         pc1.eff.sz = 0,
                                         pc1.fdr = 1),
                 anova.stat = data.frame() )

  if (dim(expr)[1] <= 1){
    return(result)
  }

  pc = prcomp(expr, center = T, scale. = T)
  pc.sdev = pc$sdev^2/sum(pc$sdev^2)
  ind = which(pc.sdev >= 0.01)
  pc.sdev = pc.sdev[ind]
  pc.rot = pc$rotation[,ind, drop = F]

  getAnovaAssociationStats <- function(dat, formula){
    mdl = as.data.frame(anova(lm(formula = formula, data = dat)))
    mdl$fdr = p.adjust(mdl[,'Pr(>F)'])
    eff.sz = mdl['Diagnosis','Sum Sq']/sum(mdl[,'Sum Sq'], na.rm = T)
    fdr = mdl['Diagnosis','fdr']
    return(data.frame(eff.sz = eff.sz, fdr = fdr))
  }

  anova.stat = apply(pc.rot, 2, function(x, covar, formula){ 
    dat = cbind(data.frame(PC = x), covar)
    getAnovaAssociationStats(dat, formula)
  }, covar, FORMULA) %>%
    rbindlist(use.names = T, fill = T)
  anova.stat$fve = pc.sdev

  total.eff.sz = dplyr::filter(anova.stat, fdr <= 0.05) %>%
    dplyr::mutate(ws = eff.sz * fve) %>%
    dplyr::select(ws)

  result = data.frame(total.eff.sz = sum(total.eff.sz$ws, na.rm = T),
                      ngenes = length(unique(x$GeneID)),
                      npc = dim(anova.stat)[1],
                      npc.sig = sum(anova.stat$fdr <= 0.05),
                      pc1.eff.sz = anova.stat$eff.sz[1],
                      pc1.fdr = anova.stat$fdr[1])

  return(list(assoc.stat = result, anova.stat = anova.stat))
}

# Function to get module overlap plots
getOverlapPlots <- function(imp.mods, MOD.MEMBERS, MODULE.OVRLP){
    mdl.ovrlp = matrix(0, length(unique(imp.mods$ModuleNameFull)),length(unique(imp.mods$ModuleNameFull)))
    rownames(mdl.ovrlp) = unique(imp.mods$ModuleNameFull)
    colnames(mdl.ovrlp) = unique(imp.mods$ModuleNameFull)
    for( i in rownames(mdl.ovrlp)){
      for( j in colnames(mdl.ovrlp)){
        mdl.ovrlp[i,j] = length(intersect(MOD.MEMBERS[[i]],MOD.MEMBERS[[j]]))/length(unique(c(MOD.MEMBERS[[i]],MOD.MEMBERS[[j]])))
      }
    }

    mdl.ovrlp.pval = MODULE.OVRLP %>%
      dplyr::filter(ModuleNameFull %in% unique(imp.mods$ModuleNameFull), category %in% unique(imp.mods$ModuleNameFull)) %>%
      dplyr::select(ModuleNameFull, category, fisherPval) %>%
      igraph::graph_from_data_frame(directed=F) %>%
      igraph::as_adj(attr = 'fisherPval')
    mdl.ovrlp.pval = mdl.ovrlp.pval/2
    mdl.ovrlp.log = matrix('*', dim(mdl.ovrlp.pval)[1], dim(mdl.ovrlp.pval)[2])
    mdl.ovrlp.log[as.matrix(mdl.ovrlp.pval) > 0.05] = ''
    rownames(mdl.ovrlp.log) = rownames(mdl.ovrlp.pval)
    colnames(mdl.ovrlp.log) = colnames(mdl.ovrlp.pval)
    mdl.ovrlp.log = mdl.ovrlp.log[rownames(mdl.ovrlp), colnames(mdl.ovrlp)]

    Heatmap(mdl.ovrlp, col = colorRamp2(seq(0,1,length.out = 5), viridis::viridis(5)),
            name = 'JacDist', cell_fun = function(j,i,x,y,wifht,height,fill){
              grid.text(mdl.ovrlp.log[i,j] %>% format(digits = 1), x, y)
            })
  }
```

### Associaiton with Diagnosis
```{r Dx, fig.height=7, fig.width=15}
# Find assocition b/w Diagnosis category with n eigen genes
FORMULA = as.formula('PC~Diagnosis+Batch+RIN+PCT_INTRONIC_BASES+PMI+PCT_INTERGENIC_BASES+AOD+PCT_PF_READS_ALIGNED+Gender')
assoc.stat.diagnosis = MODULES %>%
  dplyr::filter(GeneID %in% rownames(EXPR)) %>%
  plyr::dlply(.(ModuleName, brainRegion), .fun = findAssociationStatistics, 
              COVARIATES, EXPR, FORMULA,
              .parallel = T, .paropts = list(.packages = c('dplyr', 'data.table')))

Dx = lapply(assoc.stat.diagnosis, function(x) x$assoc.stat) %>% 
  data.table::rbindlist(use.names = T, fill = T, idcol = 'ModuleName.BrainRegion') %>%
  tidyr::separate(ModuleName.BrainRegion, c('ModuleName', 'brainRegion'), sep = '\\.') %>%
  dplyr::left_join(MODULES %>%
                     dplyr::select(Module, method, ModuleName, brainRegion, ModuleNameFull) %>%
                     unique())
Dx$pc1.fdr.l = Dx$pc1.fdr <= 0.05

p = ggplot(Dx, aes(x = ngenes, y = total.eff.sz, color = method)) + geom_point() + xlim(0,150)
p = p + facet_grid(.~brainRegion, scales = 'free_x') + xlab('No. of genes') + ylab('Total Effect Size')
p + theme(legend.position = 'top')

p = ggplot(Dx, aes(x = as.factor(npc.sig), y = total.eff.sz)) + geom_boxplot()
p = p + facet_grid(.~brainRegion, scales = 'free_x') + xlab('No. of genes') + ylab('Total Effect Size')
p + theme(legend.position = 'top')

p = ggplot(Dx, aes(x = method, y = total.eff.sz, color = method)) + geom_boxplot() + geom_jitter()
p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free_x')  + xlab('Method') + ylab('Total Effect Size')
p + theme(legend.position = 'top')

p = ggplot(Dx, aes(x = method, y = pc1.eff.sz, color = -log10(pc1.fdr))) + geom_boxplot() + geom_jitter()
p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free_x')  + xlab('Method') + ylab('PC1 Effect Size')
p = p + scale_color_continuous(low = 'white', high = 'red')
p + theme(legend.position = 'top')
```
Top 5 modules per method
```{r Dx.1, fig.height=15, fig.width=15}
# Top 5 in each method
h = lapply(unique(Dx$brainRegion), function(br, Dx, MOD.MEMBERS, MODULE.OVRLP){
  imp.mods = Dx %>%
    dplyr::group_by(brainRegion, method) %>%
    dplyr::top_n(5, total.eff.sz) %>%
    dplyr::top_n(5, pc1.eff.sz) %>%
    dplyr::filter(brainRegion == br)

  getOverlapPlots(imp.mods, MOD.MEMBERS, MODULE.OVRLP)
}, Dx, MOD.MEMBERS, MODULE.OVRLP)
draw(h[[1]], heatmap_legend_side='left')
draw(h[[2]], heatmap_legend_side='left')
draw(h[[3]], heatmap_legend_side='left')
draw(h[[4]], heatmap_legend_side='left')
draw(h[[5]], heatmap_legend_side='left')
draw(h[[6]], heatmap_legend_side='left')
draw(h[[7]], heatmap_legend_side='left')
```
Top 5 modules any method
```{r Dx.2, fig.height=15, fig.width=15}
# Top 5 any method
h = lapply(unique(Dx$brainRegion), function(br, Dx, MOD.MEMBERS, MODULE.OVRLP){
  imp.mods = Dx %>%
    dplyr::group_by(brainRegion) %>%
    dplyr::top_n(25, total.eff.sz) %>%
    dplyr::top_n(25, pc1.eff.sz) %>%
    dplyr::filter(brainRegion == br)
  print(summary(factor(imp.mods$method)))

  getOverlapPlots(imp.mods, MOD.MEMBERS, MODULE.OVRLP)
}, Dx, MOD.MEMBERS, MODULE.OVRLP)
draw(h[[1]], heatmap_legend_side='left')
draw(h[[2]], heatmap_legend_side='left')
draw(h[[3]], heatmap_legend_side='left')
draw(h[[4]], heatmap_legend_side='left')
draw(h[[5]], heatmap_legend_side='left')
draw(h[[6]], heatmap_legend_side='left')
draw(h[[7]], heatmap_legend_side='left')
```

### Associaiton with Diagnosis.Gender
```{r Dx.Gender, fig.height=7, fig.width=15}
# Find assocition b/w Diagnosis.Gender category with n eigen genes
COVARIATES$Diagnosis.Gender = paste(COVARIATES$Diagnosis, COVARIATES$Gender, sep= '.') %>%
  as.factor
FORMULA = as.formula('PC~Diagnosis.Gender+Batch+RIN+PCT_INTRONIC_BASES+PMI+PCT_INTERGENIC_BASES+AOD+PCT_PF_READS_ALIGNED')
assoc.stat.diagnosis.gender = MODULES %>%
  dplyr::filter(GeneID %in% rownames(EXPR)) %>%
  plyr::dlply(.(ModuleName, brainRegion), .fun = findAssociationStatistics, 
              COVARIATES, EXPR, FORMULA,
              .parallel = T, .paropts = list(.packages = c('dplyr', 'data.table')))

Dx.gender = lapply(assoc.stat.diagnosis.gender, function(x) x$assoc.stat) %>% 
  data.table::rbindlist(use.names = T, fill = T, idcol = 'ModuleName.BrainRegion') %>%
  tidyr::separate(ModuleName.BrainRegion, c('ModuleName', 'brainRegion'), sep = '\\.') %>%
  dplyr::left_join(MODULES %>%
                     dplyr::select(Module, method, ModuleName, brainRegion, ModuleNameFull) %>%
                     unique())
Dx.gender$pc1.fdr.l = Dx.gender$pc1.fdr <= 0.05

p = ggplot(Dx.gender, aes(x = ngenes, y = total.eff.sz, color = method)) + geom_point() + xlim(0,150)
p = p + facet_grid(.~brainRegion, scales = 'free_x') + xlab('No. of genes') + ylab('Total Effect Size')
p + theme(legend.position = 'top')

p = ggplot(Dx.gender, aes(x = as.factor(npc.sig), y = total.eff.sz)) + geom_boxplot()
p = p + facet_grid(.~brainRegion, scales = 'free_x') + xlab('No. of genes') + ylab('Total Effect Size')
p + theme(legend.position = 'top')

p = ggplot(Dx.gender, aes(x = method, y = total.eff.sz, color = method)) + geom_boxplot() + geom_jitter()
p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free_x')  + xlab('Method') + ylab('Total Effect Size')
p + theme(legend.position = 'top')

p = ggplot(Dx.gender, aes(x = method, y = pc1.eff.sz, color = -log10(pc1.fdr))) + geom_boxplot() + geom_jitter()
p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free_x')  + xlab('Method') + ylab('PC1 Effect Size')
p = p + scale_color_continuous(low = 'white', high = 'red')
p + theme(legend.position = 'top')
```
Top 5 modules per method
```{r Dx.gender.1, fig.height=15, fig.width=15}
# Top 5 in each method
h = lapply(unique(Dx.gender$brainRegion), function(br, Dx, MOD.MEMBERS, MODULE.OVRLP){
  imp.mods = Dx %>%
    dplyr::group_by(brainRegion, method) %>%
    dplyr::top_n(5, total.eff.sz) %>%
    dplyr::top_n(5, pc1.eff.sz) %>%
    dplyr::filter(brainRegion == br)

  getOverlapPlots(imp.mods, MOD.MEMBERS, MODULE.OVRLP)
}, Dx.gender, MOD.MEMBERS, MODULE.OVRLP)
draw(h[[1]], heatmap_legend_side='left')
draw(h[[2]], heatmap_legend_side='left')
draw(h[[3]], heatmap_legend_side='left')
draw(h[[4]], heatmap_legend_side='left')
draw(h[[5]], heatmap_legend_side='left')
draw(h[[6]], heatmap_legend_side='left')
draw(h[[7]], heatmap_legend_side='left')
```
Top 5 modules any method
```{r Dx.gender.2, fig.height=15, fig.width=15}
# Top 5 any method
h = lapply(unique(Dx.gender$brainRegion), function(br, Dx, MOD.MEMBERS, MODULE.OVRLP){
  imp.mods = Dx %>%
    dplyr::group_by(brainRegion) %>%
    dplyr::top_n(25, total.eff.sz) %>%
    dplyr::top_n(25, pc1.eff.sz) %>%
    dplyr::filter(brainRegion == br)
  print(summary(factor(imp.mods$method)))

  getOverlapPlots(imp.mods, MOD.MEMBERS, MODULE.OVRLP)
}, Dx.gender, MOD.MEMBERS, MODULE.OVRLP)
draw(h[[1]], heatmap_legend_side='left')
draw(h[[2]], heatmap_legend_side='left')
draw(h[[3]], heatmap_legend_side='left')
draw(h[[4]], heatmap_legend_side='left')
draw(h[[5]], heatmap_legend_side='left')
draw(h[[6]], heatmap_legend_side='left')
draw(h[[7]], heatmap_legend_side='left')
```

### Associaiton with ApoE
```{r ApoE, fig.height=7, fig.width=15}
# Find assocition b/w ApoE category with n eigen genes
FORMULA = as.formula('PC~BrainRegion.ApoE+Batch+RIN+PCT_INTRONIC_BASES+PMI+PCT_INTERGENIC_BASES+AOD+PCT_PF_READS_ALIGNED+Gender')
assoc.stat.apoe = MODULES %>%
  dplyr::filter(GeneID %in% rownames(EXPR)) %>%
  plyr::dlply(.(ModuleName, brainRegion), .fun = findAssociationStatistics, 
              COVARIATES, EXPR, FORMULA,
              .parallel = T, .paropts = list(.packages = c('dplyr', 'data.table')))

ApoE = lapply(assoc.stat.apoe, function(x) x$assoc.stat) %>% 
  data.table::rbindlist(use.names = T, fill = T, idcol = 'ModuleName.BrainRegion') %>%
  tidyr::separate(ModuleName.BrainRegion, c('ModuleName', 'brainRegion'), sep = '\\.') %>%
  dplyr::left_join(MODULES %>%
                     dplyr::select(Module, method, ModuleName, brainRegion, ModuleNameFull) %>%
                     unique())
ApoE$pc1.fdr.l = ApoE$pc1.fdr <= 0.05

p = ggplot(ApoE, aes(x = ngenes, y = total.eff.sz, color = method)) + geom_point() + xlim(0,150)
p = p + facet_grid(.~brainRegion, scales = 'free_x') + xlab('No. of genes') + ylab('Total Effect Size')
p + theme(legend.position = 'top')

p = ggplot(ApoE, aes(x = as.factor(npc.sig), y = total.eff.sz)) + geom_boxplot()
p = p + facet_grid(.~brainRegion, scales = 'free_x') + xlab('No. of genes') + ylab('Total Effect Size')
p + theme(legend.position = 'top')

p = ggplot(ApoE, aes(x = method, y = total.eff.sz, color = method)) + geom_boxplot() + geom_jitter()
p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free_x')  + xlab('Method') + ylab('Total Effect Size')
p + theme(legend.position = 'top')

p = ggplot(ApoE, aes(x = method, y = pc1.eff.sz, color = -log10(pc1.fdr))) + geom_boxplot() + geom_jitter()
p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free_x')  + xlab('Method') + ylab('PC1 Effect Size')
p = p + scale_color_continuous(low = 'white', high = 'red')
p + theme(legend.position = 'top')
```

### Select information rich module (based on Dx association, MAGMA enrichment, dbgap enrichment and AD related pathway enrichment)
```{r imp.mdl, fig.height=7, fig.width=12}
ad.magma = MAGMA.ENRICH %>%
  plyr::ddply(.(brainRegion), .fun = function(x){
    x %>%
      dplyr::mutate(fdr = p.adjust(P),
                    rnk = rank(rank(-log10(fdr))+rank(BETA)),
                    rnk = rnk/max(rnk)) %>%
      dplyr::arrange(desc(rnk))%>%
      dplyr::group_by(SET) %>%
      dplyr::top_n(1, rnk) %>%
      dplyr::top_n(1, -log10(fdr))
  })

ad.dbgap = MODULE.ENRICH %>%
  dplyr::filter(geneSet == 'dbgap', category == 'Alzheimer.Disease') %>%
  unique() %>%
  dplyr::left_join(Dx %>% dplyr::select(ModuleName, brainRegion, ModuleNameFull, method)) %>%
  plyr::ddply(.(brainRegion), .fun = function(x){
    x %>%
      dplyr::mutate(fdr = p.adjust(fisherPval),
                    rnk = rank(rank(-log10(fdr))+rank(fisherOR)),
                    rnk = rnk/max(rnk)) %>%
      dplyr::arrange(desc(rnk)) %>%
      dplyr::group_by(ModuleNameFull) %>%
      dplyr::top_n(1, rnk) %>%
      dplyr::top_n(1, -log10(fdr))
  })

other.ad.pathways = c("alzheimer_disease", "Alzheimer.s.disease_Homo.sapiens_hsa05010", 
                      "Alzheimers.Disease_Homo.sapiens_WP2059", "Alzheimer.disease.amyloid.secretase.pathway_Homo.sapiens_P00003",
                      "Alzheimer.s_disease", "Alzheimer.Disease")
ad.others = MODULE.ENRICH %>%
  dplyr::filter(geneSet != 'dbgap', category %in% other.ad.pathways) %>%
  unique() %>%
  dplyr::left_join(Dx %>% dplyr::select(ModuleName, brainRegion, ModuleNameFull, method)) %>%
  plyr::ddply(.(brainRegion), .fun = function(x){
    x %>%
      dplyr::mutate(fdr = p.adjust(fisherPval),
                    rnk = rank(rank(-log10(fdr))+rank(fisherOR)),
                    rnk = rnk/max(rnk)) %>%
      dplyr::arrange(desc(rnk))%>%
      dplyr::group_by(ModuleNameFull) %>%
      dplyr::top_n(1, rnk) %>%
      dplyr::top_n(1, -log10(fdr))
  })

# Combine and rank all the information
scr1 = Dx %>%
  plyr::ddply(.(brainRegion), .fun = function(x){
    x %>%
      dplyr::mutate(rnk = rank(total.eff.sz),
                    rnk = rnk/max(rnk)) %>%
      dplyr::arrange(desc(rnk))%>%
      dplyr::group_by(ModuleNameFull) %>%
      dplyr::top_n(1, rnk) %>%
      dplyr::top_n(1, -log10(pc1.fdr))
  }) %>%
  dplyr::mutate(Dx = rank(rnk),
                Dx = Dx/max(Dx)) %>%
  dplyr::select(ModuleNameFull, brainRegion, method, ModuleName, Dx)

scr2 = ad.magma %>%
  dplyr::mutate(magma = rank(rnk),
                magma = magma/max(magma)) %>%
  dplyr::rename(ModuleNameFull = SET) %>%
  dplyr::select(ModuleNameFull, brainRegion, method, magma)

scr3 = ad.dbgap %>%
  dplyr::mutate(dbgap = rank(rnk),
                dbgap = dbgap/max(dbgap)) %>%
  dplyr::select(ModuleNameFull, ModuleName, brainRegion, method, dbgap)

scr4 = ad.others %>%
  dplyr::mutate(other = rank(rnk),
                other = other/max(other)) %>%
  dplyr::select(ModuleNameFull, ModuleName, brainRegion, method, other)

all.rank = plyr::join_all(list(scr1, scr2, scr3, scr4), type = 'full')
all.rank[is.na(all.rank)] = 0
all.rank = all.rank %>%
  plyr::ddply(.(brainRegion), .fun = function(x){
    x = x %>%
      dplyr::mutate(rnk = rank(Dx+magma+dbgap+other),
                    rnk = rnk/max(rnk))
    print(paste('BrainRegion:', unique(x$brainRegion)))
    print(paste('Correlation with Dx', round(cor(x$rnk, x$Dx),2)))
    print(paste('Correlation with magma', round(cor(x$rnk, x$magma),2)))
    print(paste('Correlation with dbgap', round(cor(x$rnk, x$dbgap),2)))
    print(paste('Correlation with others', round(cor(x$rnk, x$other),2)))
    return(x)
  })

ggplot(all.rank, aes(x = Dx, y = magma))+geom_point()+facet_grid(.~brainRegion)

ggplot(all.rank, aes(x = dbgap, y = other))+geom_point()+facet_grid(.~brainRegion)+geom_smooth(method = 'lm')
ggplot(all.rank, aes(x = dbgap, y = magma))+geom_point()+facet_grid(.~brainRegion)+geom_smooth(method = 'lm')

ggplot(all.rank, aes(x = other, y = magma))+geom_point()+facet_grid(.~brainRegion)+geom_smooth(method = 'lm')
```
Top 5 modules per method
```{r imp.mdl.1, fig.height=15, fig.width=15}
# Top 5 in each method
h = lapply(unique(all.rank$brainRegion), function(br, Dx, MOD.MEMBERS, MODULE.OVRLP){
  imp.mods = Dx %>%
    dplyr::group_by(brainRegion, method) %>%
    dplyr::top_n(5, rnk) %>%
    dplyr::top_n(5, Dx) %>%
    dplyr::filter(brainRegion == br)

  getOverlapPlots(imp.mods, MOD.MEMBERS, MODULE.OVRLP)
}, all.rank, MOD.MEMBERS, MODULE.OVRLP)
draw(h[[1]], heatmap_legend_side='left')
draw(h[[2]], heatmap_legend_side='left')
draw(h[[3]], heatmap_legend_side='left')
draw(h[[4]], heatmap_legend_side='left')
draw(h[[5]], heatmap_legend_side='left')
draw(h[[6]], heatmap_legend_side='left')
draw(h[[7]], heatmap_legend_side='left')
```
Top 5 modules any method
```{r imp.mdl.2, fig.height=15, fig.width=15}
# Top 5 any method
h = lapply(unique(all.rank$brainRegion), function(br, Dx, MOD.MEMBERS, MODULE.OVRLP){
  imp.mods = Dx %>%
    dplyr::group_by(brainRegion) %>%
    dplyr::top_n(25, rnk) %>%
    dplyr::top_n(25, Dx) %>%
    dplyr::filter(brainRegion == br)
  print(summary(factor(imp.mods$method)))

  getOverlapPlots(imp.mods, MOD.MEMBERS, MODULE.OVRLP)
}, all.rank, MOD.MEMBERS, MODULE.OVRLP)
draw(h[[1]], heatmap_legend_side='left')
draw(h[[2]], heatmap_legend_side='left')
draw(h[[3]], heatmap_legend_side='left')
draw(h[[4]], heatmap_legend_side='left')
draw(h[[5]], heatmap_legend_side='left')
draw(h[[6]], heatmap_legend_side='left')
draw(h[[7]], heatmap_legend_side='left')
```

### Modules missing enrichment scores
```{r missing}
writeLines('Modules missing enrich enrichment scores')
filter(Dx, ModuleNameFull %in% unique(setdiff(Dx$ModuleNameFull, MODULE.ENRICH$ModuleNameFull))) %>%
  kable

writeLines('Modules missing MAGMA enrichment scores')
filter(Dx, ModuleNameFull %in% unique(setdiff(Dx$ModuleNameFull, MAGMA.ENRICH$Set))) %>%
  kable
```

### 2017-09-20 code/Rmd/networkAssociationTest.Rmd
Added total effect size randomisation and sanity modifications

@@
 ---
-title: "Function to find association b/w n significant eigene genes with Diagnosis"
+title: "Function to find association b/w n significant eigen genes with Diagnosis, Diagnosis.Gender, ApoE"
 author: "Thanneer Perumal"
@@
 # Load Libraries
-library(CovariateAnalysis)
+library(CovariateAnalysis) #devtools::install_github('th1vairam/CovariateAnalysis@dev')
 library(data.table)
@@
 ### Data download
-```{r download.data, cache=TRUE}
+```{r download.data, cache=FALSE, include=FALSE}
 # Download data
 MODULES = synTableQuery('SELECT * FROM syn10338156')@values
+all.used.ids = c('syn10338156')
 
@@
 COVARIATES = downloadFile('syn10160582')
+all.used.ids = c(all.used.ids, 'syn10160582')
 COVARIATES$BrainRegion[COVARIATES$BrainRegion == 'CER'] = 'CBE'
@@
 EXPR = downloadFile('syn10160584')
+all.used.ids = c(all.used.ids, 'syn10160584')
 gene.id.map = data.frame(ID = EXPR$Gene.ID, Gene.ID = EXPR$Gene.ID) %>%
@@
 MODULE.OVRLP = synTableQuery('SELECT * FROM syn10339153')@values
-
+all.used.ids = c(all.used.ids, 'syn10339153')
 MODULE.ENRICH = synTableQuery('SELECT * FROM syn10492048')@values
+all.used.ids = c(all.used.ids, 'syn10492048')
 
 MAGMA.ENRICH = synTableQuery('SELECT * FROM syn10380432')@values
+all.used.ids = c(all.used.ids, 'syn10380432')
 ```
@@
 # Function to find association statistics
-findAssociationStatistics <- function(x, COVARIATES, EXPR, FORMULA){
+findAssociationStatistics <- function(x, covars, exprs, frmla){
 
-  covar = dplyr::filter(COVARIATES, BrainRegion == unique(x$brainRegion))
-  expr = EXPR[x$GeneID,covar$SampleID]
+  covar = dplyr::filter(covars, BrainRegion == unique(x$brainRegion))
+  expr = exprs[x$GeneID,covar$SampleID]
 
   result = list( assoc.stat = data.frame(total.eff.sz = 0,
+                                         total.pval = 1,
                                          ngenes = length(unique(x$GeneID)),
@@
 
+  # Function to get anova association stats
+  getAnovaAssociationStats <- function(dat, frmla){
+    mdl = as.data.frame(anova(lm(formula = frmla, data = dat)))
+    mdl$fdr = p.adjust(mdl[,'Pr(>F)'])
+    eff.sz = mdl['Diagnosis','Sum Sq']/sum(mdl[,'Sum Sq'], na.rm = T)
+    fdr = mdl['Diagnosis','fdr']
+    return(data.frame(eff.sz = eff.sz, fdr = fdr))
+  }
+
+  # Function to get PCs
+  getPCs <- function(expr){
     pc = prcomp(expr, center = T, scale. = T)
@@
 
-  getAnovaAssociationStats <- function(dat, formula){
-    mdl = as.data.frame(anova(lm(formula = formula, data = dat)))
-    mdl$fdr = p.adjust(mdl[,'Pr(>F)'])
-    eff.sz = mdl['Diagnosis','Sum Sq']/sum(mdl[,'Sum Sq'], na.rm = T)
-    fdr = mdl['Diagnosis','fdr']
-    return(data.frame(eff.sz = eff.sz, fdr = fdr))
+    return(list(sdev = pc.sdev, rot = pc.rot))
   }
 
-  anova.stat = apply(pc.rot, 2, function(x, covar, formula){ 
+  # Function to get PCs
+  bootstrapAssociationStats <- function(expr, covar, frmla){
+    pc = getPCs(expr)
+
+    anova.stat = apply(pc$rot, 2, function(x, covar, frmla){ 
       dat = cbind(data.frame(PC = x), covar)
-    getAnovaAssociationStats(dat, formula)
-  }, covar, FORMULA) %>%
+      getAnovaAssociationStats(dat, frmla)
+    }, covar, frmla) %>%
       rbindlist(use.names = T, fill = T)
-  anova.stat$fve = pc.sdev
+    anova.stat$fve = pc$sdev
 
@@
 
-  result = data.frame(total.eff.sz = sum(total.eff.sz$ws, na.rm = T),
+    sum(total.eff.sz$ws, na.rm = T)
+  }
+
+  pc = getPCs(expr)
+
+  anova.stat = apply(pc$rot, 2, function(x, covar, frmla){ 
+    dat = cbind(data.frame(PC = x), covar)
+    getAnovaAssociationStats(dat, frmla)
+  }, covar, frmla) %>%
+    rbindlist(use.names = T, fill = T)
+  anova.stat$fve = pc$sdev
+
+  total.eff.sz = dplyr::filter(anova.stat, fdr <= 0.05) %>%
+    dplyr::mutate(ws = eff.sz * fve) %>%
+    dplyr::select(ws) 
+  total.eff.sz = sum(total.eff.sz$ws, na.rm = T)
+
+  rn.eff.sz = plyr::laply(1:100, .fun = function(i, expr, l, covar, frmla){
+    tryCatch({ 
+      bootstrapAssociationStats(expr[sample(1:dim(expr)[1], l), ], covar, frmla) 
+    }, error = function(e){ 0 })
+  }, exprs[,covar$SampleID], length(x$GeneID), covar, frmla, .parallel = T)
+
+  result = data.frame(total.eff.sz = total.eff.sz,
+                      total.pval = sum(rn.eff.sz >= total.eff.sz, na.rm = T)/length(rn.eff.sz),
                       ngenes = length(unique(x$GeneID)),
@@
 ### Associaiton with Diagnosis
-```{r Dx, fig.height=7, fig.width=15}
+```{r Dx, fig.height=5, fig.width=10}
 # Find assocition b/w Diagnosis category with n eigen genes
@@
                      dplyr::select(Module, method, ModuleName, brainRegion, ModuleNameFull) %>%
-                     unique())
-Dx$pc1.fdr.l = Dx$pc1.fdr <= 0.05
+                     unique()) %>%
+  plyr::ddply(.(brainRegion), .fun = function(x){
+    x %>%
+      dplyr::mutate(total.fdr = p.adjust(total.pval))
+  })
 
@@
 
-p = ggplot(Dx, aes(x = method, y = total.eff.sz, color = method)) + geom_boxplot() + geom_jitter()
+p = ggplot(Dx, aes(x = method, y = total.eff.sz, color = total.fdr)) + geom_boxplot() + geom_jitter()
 p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free_x')  + xlab('Method') + ylab('Total Effect Size')
+p = p + scale_color_continuous(low = 'red', high = 'white')
 p + theme(legend.position = 'top')
@@
 Top 5 modules per method
-```{r Dx.1, fig.height=15, fig.width=15}
+```{r Dx.1, fig.height=12, fig.width=12}
 # Top 5 in each method
@@
     dplyr::group_by(brainRegion, method) %>%
+    dplyr::filter(total.fdr <= 0.05) %>%
     dplyr::top_n(5, total.eff.sz) %>%
@@
 Top 5 modules any method
-```{r Dx.2, fig.height=15, fig.width=15}
+```{r Dx.2, fig.height=12, fig.width=12}
 # Top 5 any method
@@
     dplyr::group_by(brainRegion) %>%
+    dplyr::filter(total.fdr <= 0.05) %>%
     dplyr::top_n(25, total.eff.sz) %>%
@@
 ### Associaiton with Diagnosis.Gender
-```{r Dx.Gender, fig.height=7, fig.width=15}
+```{r Dx.Gender, fig.height=5, fig.width=10}
 # Find assocition b/w Diagnosis.Gender category with n eigen genes
@@
                      dplyr::select(Module, method, ModuleName, brainRegion, ModuleNameFull) %>%
-                     unique())
-Dx.gender$pc1.fdr.l = Dx.gender$pc1.fdr <= 0.05
+                     unique()) %>%
+  plyr::ddply(.(brainRegion), .fun = function(x){
+    x %>%
+      dplyr::mutate(total.fdr = p.adjust(total.pval))
+  })
 
@@
 
-p = ggplot(Dx.gender, aes(x = method, y = total.eff.sz, color = method)) + geom_boxplot() + geom_jitter()
+p = ggplot(Dx.gender, aes(x = method, y = total.eff.sz, color = total.fdr)) + geom_boxplot() + geom_jitter()
 p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free_x')  + xlab('Method') + ylab('Total Effect Size')
+p = p + scale_color_continuous(low = 'red', high = 'white')
 p + theme(legend.position = 'top')
@@
 Top 5 modules per method
-```{r Dx.gender.1, fig.height=15, fig.width=15}
+```{r Dx.gender.1, fig.height=12, fig.width=12}
 # Top 5 in each method
@@
 Top 5 modules any method
-```{r Dx.gender.2, fig.height=15, fig.width=15}
+```{r Dx.gender.2, fig.height=12, fig.width=12}
 # Top 5 any method
@@
 ### Associaiton with ApoE
-```{r ApoE, fig.height=7, fig.width=15}
+```{r ApoE, fig.height=5, fig.width=10}
 # Find assocition b/w ApoE category with n eigen genes
@@
                      dplyr::select(Module, method, ModuleName, brainRegion, ModuleNameFull) %>%
-                     unique())
-ApoE$pc1.fdr.l = ApoE$pc1.fdr <= 0.05
+                     unique()) %>%
+  plyr::ddply(.(brainRegion), .fun = function(x){
+    x %>%
+      dplyr::mutate(total.fdr = p.adjust(total.pval))
+  })
 
@@
 
-p = ggplot(ApoE, aes(x = method, y = total.eff.sz, color = method)) + geom_boxplot() + geom_jitter()
+p = ggplot(ApoE, aes(x = method, y = total.eff.sz, color = total.fdr)) + geom_boxplot() + geom_jitter()
 p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free_x')  + xlab('Method') + ylab('Total Effect Size')
+p = p + scale_color_continuous(low = 'red', high = 'white')
 p + theme(legend.position = 'top')
@@
     x %>%
-      dplyr::mutate(rnk = rank(total.eff.sz),
+      dplyr::mutate(rnk = rank(rank(total.eff.sz) + rank(-log10(total.fdr))),
                     rnk = rnk/max(rnk)) %>%
@@
 Top 5 modules per method
-```{r imp.mdl.1, fig.height=15, fig.width=15}
+```{r imp.mdl.1, fig.height=12, fig.width=12}
 # Top 5 in each method
@@
 Top 5 modules any method
-```{r imp.mdl.2, fig.height=15, fig.width=15}
+```{r imp.mdl.2, fig.height=12, fig.width=12}
 # Top 5 any method
@@
 ### Modules missing enrichment scores
-```{r missing}
+```{r missing, include=FALSE}
 writeLines('Modules missing enrich enrichment scores')
 filter(Dx, ModuleNameFull %in% unique(setdiff(Dx$ModuleNameFull, MODULE.ENRICH$ModuleNameFull))) %>%
-  kable
+  dplyr::select(ModuleNameFull) %>% unique() %>% unlist() %>% paste(collapse = ', ') %>% print()
 
@@
 filter(Dx, ModuleNameFull %in% unique(setdiff(Dx$ModuleNameFull, MAGMA.ENRICH$Set))) %>%
-  kable
+  dplyr::select(ModuleNameFull) %>% unique() %>% unlist() %>% paste(collapse = ', ') %>% print()
 ```
+
+### Store results in synapse
+```{r syn.store, cache = FALSE, include=FALSE}
+CODE = Folder('Network Association Tests', parentId = "syn10306449")
+CODE = synStore(CODE)
+
+write.table(Dx, file = 'AssociaitonStatWithDx.tsv', sep = '\t', quote = F, row.names = F)
+OBJ = File('AssociaitonStatWithDx.tsv', name = 'Association with Dx', parentId = CODE@properties$id)
+OBJ = synStore(OBJ, used = all.used.ids, executed = thisFile, activityName = activityName)
+
+write.table(Dx.gender, file = 'AssociaitonStatWithDxGender.tsv', sep = '\t', quote = F, row.names = F)
+OBJ1 = File('AssociaitonStatWithDxGender.tsv', name = 'Association with Dx.Gender', parentId = CODE@properties$id)
+OBJ1 = synStore(OBJ1, used = all.used.ids, executed = thisFile, activityName = activityName)
+
+write.table(ApoE, file = 'AssociaitonStatWithApoE4.tsv', sep = '\t', quote = F, row.names = F)
+OBJ2 = File('AssociaitonStatWithApoE4.tsv', name = 'Association with ApoE4', parentId = CODE@properties$id)
+OBJ2 = synStore(OBJ2, used = all.used.ids, executed = thisFile, activityName = activityName)
+
+save(list = c('assoc.stat.diagnosis', 'assoc.stat.diagnosis.gender', 'assoc.stat.apoe'), file = 'AllAssociationResults.RData')
+OBJ3 = File('AllAssociationResults.RData', name = 'All Assocition Statistics (RData)', parentId = CODE@properties$id)
+OBJ3 = synStore(OBJ3, used = all.used.ids, executed = thisFile, activityName = activityName)
+```

### 2017-09-20 code/Rmd/networkAssociationTest.Rmd
Minor bug fix

@@
   tidy = FALSE,
-  cache = TRUE)
+  cache = FALSE)
 ```
 
-```{r synapse.parameters, include=FALSE, cache=TRUE}
+```{r synapse.parameters, include=FALSE}
 parentId = 'syn10306449';
@@
 ### Data download
-```{r download.data, cache=FALSE, include=FALSE}
+```{r download.data, include=FALSE}
 # Download data
@@
 
+  print(dim(imp.mods)[1])
+  print(br)
+
+  if(dim(imp.mods)[1] != 0 ){
     getOverlapPlots(imp.mods, MOD.MEMBERS, MODULE.OVRLP)
+  }
 }, Dx.gender, MOD.MEMBERS, MODULE.OVRLP)
 draw(h[[1]], heatmap_legend_side='left')
-draw(h[[2]], heatmap_legend_side='left')
+# draw(h[[2]], heatmap_legend_side='left')
 draw(h[[3]], heatmap_legend_side='left')
@@
     dplyr::filter(brainRegion == br)
+  print(dim(imp.mods)[1])
+  print(br)
+
+  if(dim(imp.mods)[1] != 0 ){
     print(summary(factor(imp.mods$method)))
-
     getOverlapPlots(imp.mods, MOD.MEMBERS, MODULE.OVRLP)
+  }
 }, Dx.gender, MOD.MEMBERS, MODULE.OVRLP)
 draw(h[[1]], heatmap_legend_side='left')
-draw(h[[2]], heatmap_legend_side='left')
+# draw(h[[2]], heatmap_legend_side='left')
 draw(h[[3]], heatmap_legend_side='left')
@@
 ### Store results in synapse
-```{r syn.store, cache = FALSE, include=FALSE}
+```{r syn.store, include=FALSE}
 CODE = Folder('Network Association Tests', parentId = "syn10306449")

### 2017-09-21 code/R/collateDiffExp.R
Minor bug fix

@@ -83,7 +83,7 @@ tmp = diff.exp$ROSMAP %>%
   dplyr::filter(Model == 'Diagnosis.Gender', Comparison %in% c("AD-CONTROL.IN.FEMALE", "AD-CONTROL.IN.MALE"), 
                 adj.P.Val <= 0.05, logFC >= 0) %>%
   plyr::dlply(.(Comparison), .fun = function(x){
-    select(x, hgnc_symbol) %>%
+    dplyr::select(x, hgnc_symbol) %>%
       unlist %>% unique()
   })
 names(tmp) = c('DLPFC.AD-CONTROL.FEMALE.UP', 'DLPFC.AD-CONTROL.MALE.UP')
@@ -93,7 +93,7 @@ tmp = diff.exp$ROSMAP %>%
   dplyr::filter(Model == 'Diagnosis.Gender', Comparison %in% c("AD-CONTROL.IN.FEMALE", "AD-CONTROL.IN.MALE"), 
                 adj.P.Val <= 0.05, logFC <= 0) %>%
   plyr::dlply(.(Comparison), .fun = function(x){
-    select(x, hgnc_symbol) %>%
+    dplyr::select(x, hgnc_symbol) %>%
       unlist %>% unique()
   })
 names(tmp) = c('DLPFC.AD-CONTROL.FEMALE.DOWN', 'DLPFC.AD-CONTROL.MALE.DOWN')
@@ -185,7 +185,7 @@ tmp = diff.exp$ROSMAP %>%
                                   "apoe_genotype1-apoe_genotype0"), 
                 adj.P.Val <= 0.05, logFC <= 0) %>%
   plyr::dlply(.(Comparison), .fun = function(x){
-    select(x, hgnc_symbol) %>%
+    dplyr::select(x, hgnc_symbol) %>%
       unlist %>% unique()
   })
 
@@ -250,7 +250,7 @@ tmp = diff.exp$ROSMAP %>%
                 Comparison %in% c("AOD.AD-AOD.CONTROL"), 
                 adj.P.Val <= 0.05, logFC >= 0) %>%
   plyr::dlply(.(Comparison), .fun = function(x){
-    select(x, hgnc_symbol) %>%
+    dplyr::select(x, hgnc_symbol) %>%
       unlist %>% unique()
   })
 names(tmp) = 'DLPFC.AD-CONTROL.AOD.UP'
@@ -261,7 +261,7 @@ tmp = diff.exp$ROSMAP %>%
[…]

### 2017-09-21 code/Rmd/ALL_REPROCESSED.Rmd

@@
 ```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
-indToRemove = c('367_120502')
+indToRemove = c('367_120502', '1950_TCX', '1950_CER', '1925_TCX', '1957_CER')
 
@@
 
-### Model Identification
-```{r fxn}
-# Function to fit a mixed effect model and return BIC
-fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){
-
-  # Get design matrix
-  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
-  designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]
-
-  # Estimate voom weights
-  voomWeights = limma::voom(cnt[rownames(lcpm),colnames(lcpm)], design = designMat, plot=F)$weights
-
-  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
-                              .combine = rbind,
-                              .packages = c('lme4', 'dplyr'),
-                              .export = c('voomWeights')) %dopar% {
-
-                                # Perform weighted mixed linear models for every gene
-                                expr = data.frame(Gx = lcpm[i,])
-                                data = cbind(expr,
-                                             designMat[rownames(expr),],
-                                             covars[rownames(designMat),groupingVar,drop=F])
-                                formula = paste(setdiff(colnames(data), c('Gx', groupingVar)), collapse = '+') %>%
-                                  paste('Gx ~',.) %>%
-                                  c(.,paste0('(1|',groupingVar,')')) %>%
-                                  paste(collapse = '+')
-
-                                weights = voomWeights[i,]
-                                BIC = NA
-                                tryCatch({
-                                  mdl = lmer(formula = formula, 
-                                             data = data, 
-                                             weights = weights)
-                                  BIC = BIC(mdl)
-                                }, error = function(e){
-                                  BIC = NA
-                                })
-                              return(data.frame(Gene.ID = rownames(lcpm)[i],
-                                                BIC = BIC))
-                            }
-  return(modelBIC)
-  }
-
-# Function to select covariates in a step-wise regression fashion
-selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){
-
-  varsSelected = c()
-  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
-           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
-  while(length(covarsToSelect) != 0){
-    allModelBIC = foreach(i = 1:length(covarsToSelect),
-                          .combine = c,
-                          .packages = pkgs,
-                          .export = c('fitMixedEffectModel')) %dopar% {
-
-                            # Fit a mixed effect linear model and get all BIC
-                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
-                                                           c('individualIdentifier'), 
-                                                           lcpm, 
-                                                           cnt, 
-                                                           covars) %>%
-                            plyr::rename(c('BIC' = covarsToSelect[i]))
-
-                            return(list(modelBIC))
-                          }
-    allModelBIC = plyr::join_all(allModelBIC)
-
-    # Find number of genes that are changed based on BIC
-    nChange = allModelBIC %>%
-      tidyr::gather(ModelName, Value, -Gene.ID) %>%
-      dplyr::left_join(baseModelBIC) %>%
-      dplyr::mutate(diff = baseModel - Value) %>%
-      dplyr::group_by(ModelName) %>%
-      dplyr::summarise(better = sum(diff >= 5), 
-                       worse = sum(diff <= -5)) %>%
-      dplyr::mutate(nchange = (better/(better+worse)) * 100)
-
-    print(nChange)
-
-    # Select variable
-    tmp = nChange %>%
-      dplyr::filter(nchange > 1) %>%
-      dplyr::top_n(1, nchange)
-
-    if (dim(tmp)[1] == 0){
-      covarsToSelect = c()
-    } else {
-      baseModelBIC = allModelBIC %>%
-        dplyr::select(Gene.ID, one_of(tmp$ModelName))
-      colnames(baseModelBIC)[2] = 'baseModel'
-      varsSelected = c(varsSelected, tmp$ModelName)
-      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
-      print(paste('Variable chosen to be added:',tmp$ModelName))
-    }
-  }
-
-  return(varsSelected)
-}
-```
-#### Phase I (clinical, ancestry and sample specific technical variables)
-```{r phase1}
-# Fit a base model with the following covariates
-adjust.covars = c('BrainRegion.Diagnosis')
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   c('individualIdentifier'), 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
-                                   COVARIATES) %>%
-  dplyr::rename(baseModel = BIC)
-
-# Iteratively identify covariates in phase I
-phase1.covars = c('Study', 'Gender', 'PMI', 'AOD', 'RIN', 'RIN2')
-vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
-
-adjust.covars = c(adjust.covars, vars.selected)
-```
-Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`
-
-#### Phase II (batch effects)
-```{r phase2}
-# Fit a base model with the following covariates
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   c('individualIdentifier'), 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
-                                   COVARIATES) %>%
-  dplyr::rename(baseModel = BIC)
-
-# Iteratively identify covariates in phase I
-phase2.covars = c('Batch')
-vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
-
-adjust.covars = c(adjust.covars, vars.selected)
-```
-Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`
-
-#### Phase III (RNASeq alignment specific covariates)
-```{r phase3}
-# Fit a base model with the following covariates
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   c('individualIdentifier'), 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
-                                   COVARIATES) %>%
-  dplyr::rename(baseModel = BIC)
-
-# Iteratively identify covariates in phase I
-phase3.covars = c("PCT_PF_READS_ALIGNED", "PCT_CODING_BASES", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
-vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
-
-adjust.covars = c(adjust.covars, vars.selected)
-```
-Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`
-
-### Normalisation
+### Normalisation (iterative design)
 1. BrainRegion.Diagnosis is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
 2. Individual_ID is chosen as random effect
-```{r iterative.normalisation, results='asis'}
+```{r iterative.norm, results='asis'}
+# Primary variable of interest
+primaryVariable = c('BrainRegion.Diagnosis');
+notEstimable = c('BrainRegion','Diagnosis','BrainRegion.ApoE')
+postAdjustCovars = c()
+
+# Assign residual covariates
+residualCovars = setdiff(preAdjustedSigCovars$significantCovars, 
+                         c(postAdjustCovars, primaryVariable, notEstimable, 'individualIdentifier'))
+residualSigCovars = preAdjustedSigCovars
+covariatesEffects = preAdjustedSigCovars$Effects.significantCovars[residualCovars]
+postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()
+
+loopCount = 0 
+while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 20){
   writeLines(paste('Using following covariates in the model:',
-                 paste(adjust.covars, collapse=', '),
-                 'as fixed effects and Individual_ID is chosen as random effect'))
+                   paste(postAdjustCovars, collapse=', '),
+                   'as fixed effects and individualIdentifier as random effect'))
 
   # Post adjusted design matrix
-DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
+  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
   DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
 
-# Estimate voom weights
+  # Estimate voom weights for dispersion control
   VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F)
 
-# Estimate consensus correlation coefficient
-correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
-
-# Re-calculate voom weights with correlation of random effects
+  # Calculate correlation values of random effects
+  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
+  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
+                                             block = COVARIATES$individualIdentifier,
+                                             method = 'lmer')
+
+  if (!is.nan(correlation$cor)){
+    # Re-estimate voom weights
     VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, 
                                 design=DM1$design, plot=F, 
-                            block = COVARIATES$Individual_ID, 
+                                block = COVARIATES$individualIdentifier, 
                                 correlation = correlation$cor)
@@
     # Fit linear model using new weights and new design
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
-ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
+    VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$E,
+                              design = DM1$design,
+                              weights = VOOM.GENE_EXPRESSION$weights,
+                              block = COVARIATES$individualIdentifier, 
+                              correlation = correlation$cor)
 
     # Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
+    RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
+                                                  CQN.GENE_EXPRESSION$E)
+  } else {
+    notEstimable = c(notEstimable, postAdjustCovars[length(postAdjustCovars)])
+    postAdjustCovars = postAdjustCovars[1:(length(postAdjustCovars)-1)]
+  }
+
+  # Residual covariates to choose from
+  residCovars <- setdiff(c(FactorCovariates,ContCovariates),
+                         c(postAdjustCovars, primaryVariable, 'COVARIATES$individualIdentifier', notEstimable))
 
   # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, COVARIATES,
+  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
+                                                COVARIATES[, residCovars, drop=F], 
                                                 'adjusted design(voom-normalized)',
                                                 isKeyPlot=TRUE)
+
+  # Add postadjusted covariates (if any)
+  residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, primaryVariable,
+                                                               'individualIdentifier', notEstimable))
+  covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]
+
+  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
+  loopCount = loopCount + 1
+}
+
+modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
+                  'as fixed effects')
+
+tmp <- paste('Using following covariates in the final model:', modelStr)
 ```
@@
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
-  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(VOOM.ADJUSTED.FIT$design[,varsToAddIn])
 ```
@@
 # Get design matrix
-DESIGN = getDesignMatrix(COVARIATES[, adjust.covars, drop = F], Intercept = F)
+DESIGN = getDesignMatrix(COVARIATES[, c("BrainRegion.Diagnosis", postAdjustCovars), drop = F], Intercept = F)
 DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
@@
 # Get design matrix
-DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.ApoE', setdiff(adjust.covars, 'BrainRegion.Diagnosis')), drop = F], 
+DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.ApoE', setdiff(postAdjustCovars, 'BrainRegion.Diagnosis')), drop = F],
                          Intercept = F)
@@
 # Get design matrix
-DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis.Gender', setdiff(adjust.covars, c('BrainRegion.Diagnosis', 'Gender'))), drop = F], Intercept = F)
+DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis.Gender', setdiff(postAdjustCovars, c('BrainRegion.Diagnosis', 'Gender'))), drop = F], Intercept = F)
 DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
@@
 Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
-```{r diff.exp3, fig.height=7, fig.width=24}
+```{r diff.exp4, fig.height=7, fig.width=24}
 # Get design matrix
-DESIGN = getDesignMatrix(COVARIATES[, c(setdiff(adjust.covars, 'AOD')), drop = F], Intercept = F)
+DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis', setdiff(postAdjustCovars, 'AOD')), drop = F], Intercept = F)
 ind = grep('BrainRegion.Diagnosis', colnames(DESIGN$design))
@@
 Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
-```{r diff.exp4.1, fig.height=10, fig.width=15}
+```{r diff.exp5, fig.height=10, fig.width=15, eval = F}
 # Cell type markers
@@
 geneSet.ref = GeneSets[c('KEGG_2016', 'GO_Biological_Process_2015', 'Reactome_2016', 'WikiPathways_2016')] %>%
+  lapply(function(x){
+    lapply(x, function(y){
+      sapply(y, function(y) str_split(y, '\\,')[[1]][1])
+    })
+  }) %>%
   CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
@@
 
+# Store all model fits
+save(list = 'all.fit', file = 'AMP_allModelFits.RData')
+AFIT_OBJ = File('AMP_allModelFits.RData',
+                name = 'All Model Fit', 
+                parentId = CODE$properties$id)
+annotations(AFIT_OBJ) = annotations(CODE)
+annotations(AFIT_OBJ)$dataSubType = 'allModelFit'
+AFIT_OBJ = synStore(AFIT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+synChangeEntityACL(AFIT_OBJ@properties$id, new.racl)
+
 synDeleteEntityACL(CODE@properties$id)

### 2017-10-09 code/Rmd/networkAssociationTest.Rmd
Added markdown

@@
   getPCs <- function(expr){
-    pc = prcomp(expr, center = T, scale. = T)
+    expr = scale(t(scale(t(expr))))
+    pc = prcomp(expr, center = F, scale. = F)
     pc.sdev = pc$sdev^2/sum(pc$sdev^2)
@@
 p + theme(legend.position = 'top')
+
+writeLines('Number of significantly associated modules')
+Dx %>%
+  dplyr::group_by(brainRegion, method) %>%
+  dplyr::summarise(nsig = sum(total.fdr <= 0.05)) %>%
+  tidyr::spread(method, nsig) %>%
+  kable()
+
+writeLines('Percent of significantly associated modules')
+Dx %>%
+  dplyr::group_by(brainRegion, method) %>%
+  dplyr::summarise(nsig = round(sum(total.fdr <= 0.05)/sum(total.fdr <= 1) * 100, 2)) %>%
+  tidyr::spread(method, nsig) %>%
+  kable()
+
+writeLines('Percent of modules whose PC1 is significantly associated')
+Dx %>%
+  dplyr::group_by(brainRegion, method) %>%
+  dplyr::summarise(nsig = round(sum(pc1.fdr <= 0.05)/sum(pc1.fdr <= 1) * 100, 2)) %>%
+  tidyr::spread(method, nsig) %>%
+  kable()
 ```
@@
 p + theme(legend.position = 'top')
+
+writeLines('Number of significantly associated modules')
+Dx.gender %>%
+  dplyr::group_by(brainRegion, method) %>%
+  dplyr::summarise(nsig = sum(total.fdr <= 0.05)) %>%
+  tidyr::spread(method, nsig) %>%
+  kable()
+
+writeLines('Percent of significantly associated modules')
+Dx.gender %>%
+  dplyr::group_by(brainRegion, method) %>%
+  dplyr::summarise(nsig = round(sum(total.fdr <= 0.05)/sum(total.fdr <= 1) * 100, 2)) %>%
+  tidyr::spread(method, nsig) %>%
+  kable()
+
+writeLines('Percent of modules whose PC1 is significantly associated')
+Dx.gender %>%
+  dplyr::group_by(brainRegion, method) %>%
+  dplyr::summarise(nsig = round(sum(pc1.fdr <= 0.05)/sum(pc1.fdr <= 1) * 100, 2)) %>%
+  tidyr::spread(method, nsig) %>%
+  kable()
 ```
@@
 
-  print(dim(imp.mods)[1])
-  print(br)
-
   if(dim(imp.mods)[1] != 0 ){
@@
 draw(h[[1]], heatmap_legend_side='left')
-# draw(h[[2]], heatmap_legend_side='left')
+draw(h[[2]], heatmap_legend_side='left')
 draw(h[[3]], heatmap_legend_side='left')
@@
     dplyr::filter(brainRegion == br)
-  print(dim(imp.mods)[1])
-  print(br)
 
@@
 other.ad.pathways = c("alzheimer_disease", "Alzheimer.s.disease_Homo.sapiens_hsa05010", 
-                      "Alzheimers.Disease_Homo.sapiens_WP2059", "Alzheimer.disease.amyloid.secretase.pathway_Homo.sapiens_P00003",
+                      "Alzheimers.Disease_Homo.sapiens_WP2059",
+                      "Alzheimer.disease.amyloid.secretase.pathway_Homo.sapiens_P00003",
                       "Alzheimer.s_disease", "Alzheimer.Disease")
@@
 ggplot(all.rank, aes(x = other, y = magma))+geom_point()+facet_grid(.~brainRegion)+geom_smooth(method = 'lm')
+
+ggplot(tmp, aes(x = method, y = rnk)) + geom_boxplot() + facet_grid(.~brainRegion) + theme(axis.text.x = element_text(angle=45, hjust = 1)) + coord_flip() + geom_jitter()
+
 ```
@@
 ```{r syn.store, include=FALSE}
-CODE = Folder('Network Association Tests', parentId = "syn10306449")
-CODE = synStore(CODE)
-
 write.table(Dx, file = 'AssociaitonStatWithDx.tsv', sep = '\t', quote = F, row.names = F)
-OBJ = File('AssociaitonStatWithDx.tsv', name = 'Association with Dx', parentId = CODE@properties$id)
+OBJ = File('AssociaitonStatWithDx.tsv', name = 'Association with Dx', parentId = 'syn10831307')
 OBJ = synStore(OBJ, used = all.used.ids, executed = thisFile, activityName = activityName)
@@
 write.table(Dx.gender, file = 'AssociaitonStatWithDxGender.tsv', sep = '\t', quote = F, row.names = F)
-OBJ1 = File('AssociaitonStatWithDxGender.tsv', name = 'Association with Dx.Gender', parentId = CODE@properties$id)
+OBJ1 = File('AssociaitonStatWithDxGender.tsv', name = 'Association with Dx.Gender', parentId = 'syn10831307')
 OBJ1 = synStore(OBJ1, used = all.used.ids, executed = thisFile, activityName = activityName)
@@
 write.table(ApoE, file = 'AssociaitonStatWithApoE4.tsv', sep = '\t', quote = F, row.names = F)
-OBJ2 = File('AssociaitonStatWithApoE4.tsv', name = 'Association with ApoE4', parentId = CODE@properties$id)
+OBJ2 = File('AssociaitonStatWithApoE4.tsv', name = 'Association with ApoE4', parentId = 'syn10831307')
 OBJ2 = synStore(OBJ2, used = all.used.ids, executed = thisFile, activityName = activityName)
@@
 save(list = c('assoc.stat.diagnosis', 'assoc.stat.diagnosis.gender', 'assoc.stat.apoe'), file = 'AllAssociationResults.RData')
-OBJ3 = File('AllAssociationResults.RData', name = 'All Assocition Statistics (RData)', parentId = CODE@properties$id)
+OBJ3 = File('AllAssociationResults.RData', name = 'All Assocition Statistics (RData)', parentId = 'syn10831307')
 OBJ3 = synStore(OBJ3, used = all.used.ids, executed = thisFile, activityName = activityName)

### 2017-10-09 code/Rmd/networkAssociationTest_metaModules.Rmd
Added markdown

--- 2017-09-20 code/Rmd/networkAssociationTest.Rmd
@@
 ---
-title: "Function to find association b/w n significant eigen genes with Diagnosis, Diagnosis.Gender, ApoE"
+title: "Function to find association b/w n significant eigen genes of meta modules with Diagnosis, Diagnosis.Gender, ApoE"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./networkAssociationTest.Rmd",
+knit2synapse::knitToFolderEntity(file = "./networkAssociationTest_metaModules.Rmd",
                                  parentId = "syn10306449",
-                                 entityName = 'Network Association Tests')
+                                 entityName = 'Network Association Tests (aggregated modules)')
 ```
@@
 
-thisFileName <- 'networkAssociationTest.Rmd'
+thisFileName <- 'networkAssociationTest_metaModules.Rmd'
 
@@
 # Download data
-MODULES = synTableQuery('SELECT * FROM syn10338156')@values
-all.used.ids = c('syn10338156')
+MODULES = synTableQuery('SELECT * FROM syn10915669')@values
+all.used.ids = c('syn10915669')
 
@@
 
-MODULE.OVRLP = synTableQuery('SELECT * FROM syn10339153')@values
-all.used.ids = c(all.used.ids, 'syn10339153')
-MODULE.ENRICH = synTableQuery('SELECT * FROM syn10492048')@values
-all.used.ids = c(all.used.ids, 'syn10492048')
-
-MAGMA.ENRICH = synTableQuery('SELECT * FROM syn10380432')@values
-all.used.ids = c(all.used.ids, 'syn10380432')
+# MODULE.OVRLP = synTableQuery('SELECT * FROM syn10339153')@values
+# all.used.ids = c(all.used.ids, 'syn10339153')
+# MODULE.ENRICH = synTableQuery('SELECT * FROM syn10492048')@values
+# all.used.ids = c(all.used.ids, 'syn10492048')
+# 
+# MAGMA.ENRICH = synTableQuery('SELECT * FROM syn10380432')@values
+# all.used.ids = c(all.used.ids, 'syn10380432')
 ```
@@
 # Function to get module overlap plots
-getOverlapPlots <- function(imp.mods, MOD.MEMBERS, MODULE.OVRLP){
+getOverlapPlots <- function(imp.mods, MOD.MEMBERS){
     mdl.ovrlp = matrix(0, length(unique(imp.mods$ModuleNameFull)),length(unique(imp.mods$ModuleNameFull)))
@@
         mdl.ovrlp[i,j] = length(intersect(MOD.MEMBERS[[i]],MOD.MEMBERS[[j]]))/length(unique(c(MOD.MEMBERS[[i]],MOD.MEMBERS[[j]])))
+        fisherEnrichment()
       }
@@
       dplyr::mutate(total.fdr = p.adjust(total.pval))
-  })
+  }) %>%
+  dplyr::mutate(total.sig = total.fdr <= 0.05)
 
-p = ggplot(Dx, aes(x = ngenes, y = total.eff.sz, color = method)) + geom_point() + xlim(0,150)
-p = p + facet_grid(.~brainRegion, scales = 'free_x') + xlab('No. of genes') + ylab('Total Effect Size')
+p = ggplot(Dx, aes(x = ngenes, y = total.eff.sz, color = brainRegion)) + geom_point()
+p = p + xlab('No. of genes') + ylab('Total Effect Size')
 p + theme(legend.position = 'top')
 
-p = ggplot(Dx, aes(x = as.factor(npc.sig), y = total.eff.sz)) + geom_boxplot()
-p = p + facet_grid(.~brainRegion, scales = 'free_x') + xlab('No. of genes') + ylab('Total Effect Size')
+p = ggplot(Dx, aes(x = as.factor(npc.sig), y = total.eff.sz, color = brainRegion)) + geom_boxplot()
+p = p + xlab('No. of genes') + ylab('Total Effect Size')
 p + theme(legend.position = 'top')
 
-p = ggplot(Dx, aes(x = method, y = total.eff.sz, color = total.fdr)) + geom_boxplot() + geom_jitter()
-p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free_x')  + xlab('Method') + ylab('Total Effect Size')
-p = p + scale_color_continuous(low = 'red', high = 'white')
+p = ggplot(Dx, aes(x = brainRegion, y = total.eff.sz)) + geom_boxplot() + geom_jitter(aes(color = total.sig))
+p = p + coord_flip() + xlab('Brain Region') + ylab('Total Effect Size')
+# p = p + scale_color_continuous(low = 'white', high = 'red')
 p + theme(legend.position = 'top')
 
-p = ggplot(Dx, aes(x = method, y = pc1.eff.sz, color = -log10(pc1.fdr))) + geom_boxplot() + geom_jitter()
-p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free_x')  + xlab('Method') + ylab('PC1 Effect Size')
+p = ggplot(Dx, aes(x = brainRegion, y = pc1.eff.sz, color = -log10(pc1.fdr))) + geom_boxplot() + geom_jitter()
+p = p + coord_flip() + xlab('Brain Region') + ylab('PC1 Effect Size')
 p = p + scale_color_continuous(low = 'white', high = 'red')
 p + theme(legend.position = 'top')
-```
-Top 5 modules per method
-```{r Dx.1, fig.height=12, fig.width=12}
-# Top 5 in each method
-h = lapply(unique(Dx$brainRegion), function(br, Dx, MOD.MEMBERS, MODULE.OVRLP){
-  imp.mods = Dx %>%
-    dplyr::group_by(brainRegion, method) %>%
-    dplyr::filter(total.fdr <= 0.05) %>%
-    dplyr::top_n(5, total.eff.sz) %>%
-    dplyr::top_n(5, pc1.eff.sz) %>%
-    dplyr::filter(brainRegion == br)
 
-  getOverlapPlots(imp.mods, MOD.MEMBERS, MODULE.OVRLP)
-}, Dx, MOD.MEMBERS, MODULE.OVRLP)
-draw(h[[1]], heatmap_legend_side='left')
-draw(h[[2]], heatmap_legend_side='left')
-draw(h[[3]], heatmap_legend_side='left')
-draw(h[[4]], heatmap_legend_side='left')
-draw(h[[5]], heatmap_legend_side='left')
-draw(h[[6]], heatmap_legend_side='left')
-draw(h[[7]], heatmap_legend_side='left')
-```
-Top 5 modules any method
-```{r Dx.2, fig.height=12, fig.width=12}
-# Top 5 any method
-h = lapply(unique(Dx$brainRegion), function(br, Dx, MOD.MEMBERS, MODULE.OVRLP){
-  imp.mods = Dx %>%
+writeLines('Number of significantly associated modules')
+Dx %>%
   dplyr::group_by(brainRegion) %>%
-    dplyr::filter(total.fdr <= 0.05) %>%
-    dplyr::top_n(25, total.eff.sz) %>%
-    dplyr::top_n(25, pc1.eff.sz) %>%
-    dplyr::filter(brainRegion == br)
-  print(summary(factor(imp.mods$method)))
+  dplyr::summarise(nsig = sum(total.fdr <= 0.05),
+                   ntotal = sum(total.fdr <= 1)) %>%
+  kable()
 
-  getOverlapPlots(imp.mods, MOD.MEMBERS, MODULE.OVRLP)
-}, Dx, MOD.MEMBERS, MODULE.OVRLP)
-draw(h[[1]], heatmap_legend_side='left')
-draw(h[[2]], heatmap_legend_side='left')
-draw(h[[3]], heatmap_legend_side='left')
-draw(h[[4]], heatmap_legend_side='left')
-draw(h[[5]], heatmap_legend_side='left')
-draw(h[[6]], heatmap_legend_side='left')
-draw(h[[7]], heatmap_legend_side='left')
+writeLines('Percent of significantly associated modules')
+Dx %>%
+  dplyr::group_by(brainRegion) %>%
+  dplyr::summarise(nsig = round(sum(total.fdr <= 0.05)/sum(total.fdr <= 1) * 100, 2)) %>%
+  kable()
+
+writeLines('Percent of modules whose PC1 is significantly associated')
+Dx %>%
+  dplyr::group_by(brainRegion) %>%
+  dplyr::summarise(nsig = round(sum(pc1.fdr <= 0.05)/sum(pc1.fdr <= 1) * 100, 2)) %>%
+  kable()
 ```
@@
       dplyr::mutate(total.fdr = p.adjust(total.pval))
-  })
+  }) %>%
+  dplyr::mutate(total.sig = total.fdr <= 0.05)
 
-p = ggplot(Dx.gender, aes(x = ngenes, y = total.eff.sz, color = method)) + geom_point() + xlim(0,150)
-p = p + facet_grid(.~brainRegion, scales = 'free_x') + xlab('No. of genes') + ylab('Total Effect Size')
+p = ggplot(Dx.gender, aes(x = ngenes, y = total.eff.sz, color = brainRegion)) + geom_point()
+p = p + xlab('No. of genes') + ylab('Total Effect Size')
 p + theme(legend.position = 'top')
 
-p = ggplot(Dx.gender, aes(x = as.factor(npc.sig), y = total.eff.sz)) + geom_boxplot()
-p = p + facet_grid(.~brainRegion, scales = 'free_x') + xlab('No. of genes') + ylab('Total Effect Size')
+p = ggplot(Dx.gender, aes(x = as.factor(npc.sig), y = total.eff.sz, color = brainRegion)) + geom_boxplot()
+p = p + xlab('No. of genes') + ylab('Total Effect Size')
 p + theme(legend.position = 'top')
 
-p = ggplot(Dx.gender, aes(x = method, y = total.eff.sz, color = total.fdr)) + geom_boxplot() + geom_jitter()
-p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free_x')  + xlab('Method') + ylab('Total Effect Size')
-p = p + scale_color_continuous(low = 'red', high = 'white')
+p = ggplot(Dx.gender, aes(x = brainRegion, y = total.eff.sz)) + geom_boxplot() + geom_jitter(aes(color = total.sig))
+p = p + coord_flip() + xlab('Brain Region') + ylab('Total Effect Size')
+# p = p + scale_color_continuous(low = 'white', high = 'red')
 p + theme(legend.position = 'top')
 
-p = ggplot(Dx.gender, aes(x = method, y = pc1.eff.sz, color = -log10(pc1.fdr))) + geom_boxplot() + geom_jitter()
-p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free_x')  + xlab('Method') + ylab('PC1 Effect Size')
+p = ggplot(Dx.gender, aes(x = brainRegion, y = pc1.eff.sz, color = -log10(pc1.fdr))) + geom_boxplot() + geom_jitter()
+p = p + coord_flip() + xlab('Brain Region') + ylab('PC1 Effect Size')
 p = p + scale_color_continuous(low = 'white', high = 'red')
 p + theme(legend.position = 'top')
-```
-Top 5 modules per method
-```{r Dx.gender.1, fig.height=12, fig.width=12}
-# Top 5 in each method
-h = lapply(unique(Dx.gender$brainRegion), function(br, Dx, MOD.MEMBERS, MODULE.OVRLP){
-  imp.mods = Dx %>%
-    dplyr::group_by(brainRegion, method) %>%
-    dplyr::filter(total.fdr <= 0.05) %>%
-    dplyr::top_n(5, total.eff.sz) %>%
-    dplyr::top_n(5, pc1.eff.sz) %>%
-    dplyr::filter(brainRegion == br)
 
-  print(dim(imp.mods)[1])
-  print(br)
+writeLines('Number of significantly associated modules')
+Dx.gender %>%
+  dplyr::group_by(brainRegion) %>%
+  dplyr::summarise(nsig = sum(total.fdr <= 0.05),
+                   ntotal = sum(total.fdr <= 1)) %>%
+  kable()
 
-  if(dim(imp.mods)[1] != 0 ){
-    getOverlapPlots(imp.mods, MOD.MEMBERS, MODULE.OVRLP)
-  }
-}, Dx.gender, MOD.MEMBERS, MODULE.OVRLP)
-draw(h[[1]], heatmap_legend_side='left')
-# draw(h[[2]], heatmap_legend_side='left')
-draw(h[[3]], heatmap_legend_side='left')
-draw(h[[4]], heatmap_legend_side='left')
-draw(h[[5]], heatmap_legend_side='left')
-draw(h[[6]], heatmap_legend_side='left')
-draw(h[[7]], heatmap_legend_side='left')
-```
-Top 5 modules any method
-```{r Dx.gender.2, fig.height=12, fig.width=12}
-# Top 5 any method
-h = lapply(unique(Dx.gender$brainRegion), function(br, Dx, MOD.MEMBERS, MODULE.OVRLP){
-  imp.mods = Dx %>%
+writeLines('Percent of significantly associated modules')
+Dx.gender %>%
   dplyr::group_by(brainRegion) %>%
-    dplyr::filter(total.fdr <= 0.05) %>%
-    dplyr::top_n(25, total.eff.sz) %>%
-    dplyr::top_n(25, pc1.eff.sz) %>%
-    dplyr::filter(brainRegion == br)
-  print(dim(imp.mods)[1])
-  print(br)
+  dplyr::summarise(nsig = round(sum(total.fdr <= 0.05)/sum(total.fdr <= 1) * 100, 2)) %>%
+  kable()
 
-  if(dim(imp.mods)[1] != 0 ){
-    print(summary(factor(imp.mods$method)))
-    getOverlapPlots(imp.mods, MOD.MEMBERS, MODULE.OVRLP)
-  }
-}, Dx.gender, MOD.MEMBERS, MODULE.OVRLP)
-draw(h[[1]], heatmap_legend_side='left')
-# draw(h[[2]], heatmap_legend_side='left')
-draw(h[[3]], heatmap_legend_side='left')
-draw(h[[4]], heatmap_legend_side='left')
-draw(h[[5]], heatmap_legend_side='left')
-draw(h[[6]], heatmap_legend_side='left')
-draw(h[[7]], heatmap_legend_side='left')
+writeLines('Percent of modules whose PC1 is significantly associated')
+Dx.gender %>%
+  dplyr::group_by(brainRegion) %>%
+  dplyr::summarise(nsig = round(sum(pc1.fdr <= 0.05)/sum(pc1.fdr <= 1) * 100, 2)) %>%
+  kable()
 ```
 
-### Associaiton with ApoE
-```{r ApoE, fig.height=5, fig.width=10}
-# Find assocition b/w ApoE category with n eigen genes
-FORMULA = as.formula('PC~BrainRegion.ApoE+Batch+RIN+PCT_INTRONIC_BASES+PMI+PCT_INTERGENIC_BASES+AOD+PCT_PF_READS_ALIGNED+Gender')
-assoc.stat.apoe = MODULES %>%
-  dplyr::filter(GeneID %in% rownames(EXPR)) %>%
-  plyr::dlply(.(ModuleName, brainRegion), .fun = findAssociationStatistics, 
-              COVARIATES, EXPR, FORMULA,
-              .parallel = T, .paropts = list(.packages = c('dplyr', 'data.table')))
+### Enrichment for differential expression
+```{r Dx.enrich}
+# Get differential expression gene list from synapse
+load(synGet('syn10496554')@filePath)
+all.used.ids = c(all.used.ids, 'syn10496554')
+names(amp.ad.de.geneSets) = gsub('CER','CBE', names(amp.ad.de.geneSets))
 
-ApoE = lapply(assoc.stat.apoe, function(x) x$assoc.stat) %>% 
-  data.table::rbindlist(use.names = T, fill = T, idcol = 'ModuleName.BrainRegion') %>%
-  tidyr::separate(ModuleName.BrainRegion, c('ModuleName', 'brainRegion'), sep = '\\.') %>%
-  dplyr::left_join(MODULES %>%
-                     dplyr::select(Module, method, ModuleName, brainRegion, ModuleNameFull) %>%
-                     unique()) %>%
-  plyr::ddply(.(brainRegion), .fun = function(x){
-    x %>%
-      dplyr::mutate(total.fdr = p.adjust(total.pval))
+# Perform enrichment analysis
+MOD.MEMBERS = plyr::dlply(MODULES, .(ModuleNameFull), .fun = function(x) unique(x$external_gene_name))
+Dx.enrich = plyr::ldply(unique(MODULES$brainRegion), .fun = function(x){
+  modMembers = MOD.MEMBERS[grep(x, names(MOD.MEMBERS))]
+  deGenes = amp.ad.de.geneSets[grep(x, names(amp.ad.de.geneSets))]
+  backgroundGenes = unique(MODULES$external_gene_name)
+
+  enrichResults = plyr::ldply(modMembers, .fun = function(y, deGenes, backgroundGenes){
+    plyr::ldply(deGenes, .fun = function(z, y, backgroundGenes){
+      fisherEnrichment(y,z,backgroundGenes)
+    }, y, backgroundGenes, .id = 'Dx.GeneSet')
+  }, deGenes, backgroundGenes, .id = 'ModuleNameFull') %>%
+    dplyr::mutate(fdr = p.adjust(pval))
+}, .id = 'brainRegion')
+```
+### Enrichment for gene sets
+```{r pathway.enrich}
+# Download all related genesets from synapse
+load(synGet('syn4867851')@filePath)
+all.used.ids = c(all.used.ids, 'syn4867851')
+
+# Download cell type specific genesets
+load(synGet('syn9818080')@filePath)
+all.used.ids = c(all.used.ids, 'syn9818080')
+
+backgroundGenes = unique(MODULES$external_gene_name)
+geneSet.ref = c(GeneSets[c('KEGG_2016', 'GO_Biological_Process_2015', 'Reactome_2016', 'WikiPathways_2016')],
+                list(CellType = CellTypeList) )%>%
+  lapply(function(x){
+    lapply(x, function(y){
+      sapply(y, function(y) str_split(y, '\\,')[[1]][1])
     })
+  }) %>%
+  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
+  do.call(c,.)
 
-p = ggplot(ApoE, aes(x = ngenes, y = total.eff.sz, color = method)) + geom_point() + xlim(0,150)
-p = p + facet_grid(.~brainRegion, scales = 'free_x') + xlab('No. of genes') + ylab('Total Effect Size')
-p + theme(legend.position = 'top')
+# Perform enrichment analysis
+Pathway.enrich = plyr::ddply(MODULES, .(ModuleNameFull), .fun = function(mod, geneSetRef, backgroundGenes){
+  modMembers = unique(mod$external_gene_name)
 
-p = ggplot(ApoE, aes(x = as.factor(npc.sig), y = total.eff.sz)) + geom_boxplot()
-p = p + facet_grid(.~brainRegion, scales = 'free_x') + xlab('No. of genes') + ylab('Total Effect Size')
-p + theme(legend.position = 'top')
-
-p = ggplot(ApoE, aes(x = method, y = total.eff.sz, color = total.fdr)) + geom_boxplot() + geom_jitter()
-p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free_x')  + xlab('Method') + ylab('Total Effect Size')
-p = p + scale_color_continuous(low = 'red', high = 'white')
-p + theme(legend.position = 'top')
-
-p = ggplot(ApoE, aes(x = method, y = pc1.eff.sz, color = -log10(pc1.fdr))) + geom_boxplot() + geom_jitter()
-p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free_x')  + xlab('Method') + ylab('PC1 Effect Size')
-p = p + scale_color_continuous(low = 'white', high = 'red')
-p + theme(legend.position = 'top')
-```
-
-### Select information rich module (based on Dx association, MAGMA enrichment, dbgap enrichment and AD related pathway enrichment)
-```{r imp.mdl, fig.height=7, fig.width=12}
-ad.magma = MAGMA.ENRICH %>%
-  plyr::ddply(.(brainRegion), .fun = function(x){
-    x %>%
-      dplyr::mutate(fdr = p.adjust(P),
-                    rnk = rank(rank(-log10(fdr))+rank(BETA)),
-                    rnk = rnk/max(rnk)) %>%
-      dplyr::arrange(desc(rnk))%>%
-      dplyr::group_by(SET) %>%
-      dplyr::top_n(1, rnk) %>%
-      dplyr::top_n(1, -log10(fdr))
-  })
-
-ad.dbgap = MODULE.ENRICH %>%
-  dplyr::filter(geneSet == 'dbgap', category == 'Alzheimer.Disease') %>%
-  unique() %>%
-  dplyr::left_join(Dx %>% dplyr::select(ModuleName, brainRegion, ModuleNameFull, method)) %>%
-  plyr::ddply(.(brainRegion), .fun = function(x){
-    x %>%
-      dplyr::mutate(fdr = p.adjust(fisherPval),
-                    rnk = rank(rank(-log10(fdr))+rank(fisherOR)),
-                    rnk = rnk/max(rnk)) %>%
-      dplyr::arrange(desc(rnk)) %>%
-      dplyr::group_by(ModuleNameFull) %>%
-      dplyr::top_n(1, rnk) %>%
-      dplyr::top_n(1, -log10(fdr))
-  })
-
-other.ad.pathways = c("alzheimer_disease", "Alzheimer.s.disease_Homo.sapiens_hsa05010", 
-                      "Alzheimers.Disease_Homo.sapiens_WP2059", "Alzheimer.disease.amyloid.secretase.pathway_Homo.sapiens_P00003",
-                      "Alzheimer.s_disease", "Alzheimer.Disease")
-ad.others = MODULE.ENRICH %>%
-  dplyr::filter(geneSet != 'dbgap', category %in% other.ad.pathways) %>%
-  unique() %>%
-  dplyr::left_join(Dx %>% dplyr::select(ModuleName, brainRegion, ModuleNameFull, method)) %>%
-  plyr::ddply(.(brainRegion), .fun = function(x){
-    x %>%
-      dplyr::mutate(fdr = p.adjust(fisherPval),
-                    rnk = rank(rank(-log10(fdr))+rank(fisherOR)),
-                    rnk = rnk/max(rnk)) %>%
-      dplyr::arrange(desc(rnk))%>%
-      dplyr::group_by(ModuleNameFull) %>%
-      dplyr::top_n(1, rnk) %>%
-      dplyr::top_n(1, -log10(fdr))
-  })
-
-# Combine and rank all the information
-scr1 = Dx %>%
-  plyr::ddply(.(brainRegion), .fun = function(x){
-    x %>%
-      dplyr::mutate(rnk = rank(rank(total.eff.sz) + rank(-log10(total.fdr))),
-                    rnk = rnk/max(rnk)) %>%
-      dplyr::arrange(desc(rnk))%>%
-      dplyr::group_by(ModuleNameFull) %>%
-      dplyr::top_n(1, rnk) %>%
-      dplyr::top_n(1, -log10(pc1.fdr))
-  }) %>%
-  dplyr::mutate(Dx = rank(rnk),
-                Dx = Dx/max(Dx)) %>%
-  dplyr::select(ModuleNameFull, brainRegion, method, ModuleName, Dx)
-
-scr2 = ad.magma %>%
-  dplyr::mutate(magma = rank(rnk),
-                magma = magma/max(magma)) %>%
-  dplyr::rename(ModuleNameFull = SET) %>%
-  dplyr::select(ModuleNameFull, brainRegion, method, magma)
-
-scr3 = ad.dbgap %>%
-  dplyr::mutate(dbgap = rank(rnk),
-                dbgap = dbgap/max(dbgap)) %>%
-  dplyr::select(ModuleNameFull, ModuleName, brainRegion, method, dbgap)
-
-scr4 = ad.others %>%
-  dplyr::mutate(other = rank(rnk),
-                other = other/max(other)) %>%
-  dplyr::select(ModuleNameFull, ModuleName, brainRegion, method, other)
-
-all.rank = plyr::join_all(list(scr1, scr2, scr3, scr4), type = 'full')
-all.rank[is.na(all.rank)] = 0
-all.rank = all.rank %>%
-  plyr::ddply(.(brainRegion), .fun = function(x){
-    x = x %>%
-      dplyr::mutate(rnk = rank(Dx+magma+dbgap+other),
-                    rnk = rnk/max(rnk))
-    print(paste('BrainRegion:', unique(x$brainRegion)))
-    print(paste('Correlation with Dx', round(cor(x$rnk, x$Dx),2)))
-    print(paste('Correlation with magma', round(cor(x$rnk, x$magma),2)))
-    print(paste('Correlation with dbgap', round(cor(x$rnk, x$dbgap),2)))
-    print(paste('Correlation with others', round(cor(x$rnk, x$other),2)))
-    return(x)
-  })
-
-ggplot(all.rank, aes(x = Dx, y = magma))+geom_point()+facet_grid(.~brainRegion)
-
-ggplot(all.rank, aes(x = dbgap, y = other))+geom_point()+facet_grid(.~brainRegion)+geom_smooth(method = 'lm')
-ggplot(all.rank, aes(x = dbgap, y = magma))+geom_point()+facet_grid(.~brainRegion)+geom_smooth(method = 'lm')
-
-ggplot(all.rank, aes(x = other, y = magma))+geom_point()+facet_grid(.~brainRegion)+geom_smooth(method = 'lm')
-```
-Top 5 modules per method
-```{r imp.mdl.1, fig.height=12, fig.width=12}
-# Top 5 in each method
-h = lapply(unique(all.rank$brainRegion), function(br, Dx, MOD.MEMBERS, MODULE.OVRLP){
-  imp.mods = Dx %>%
-    dplyr::group_by(brainRegion, method) %>%
-    dplyr::top_n(5, rnk) %>%
-    dplyr::top_n(5, Dx) %>%
-    dplyr::filter(brainRegion == br)
-
-  getOverlapPlots(imp.mods, MOD.MEMBERS, MODULE.OVRLP)
-}, all.rank, MOD.MEMBERS, MODULE.OVRLP)
-draw(h[[1]], heatmap_legend_side='left')
-draw(h[[2]], heatmap_legend_side='left')
-draw(h[[3]], heatmap_legend_side='left')
-draw(h[[4]], heatmap_legend_side='left')
-draw(h[[5]], heatmap_legend_side='left')
-draw(h[[6]], heatmap_legend_side='left')
-draw(h[[7]], heatmap_legend_side='left')
-```
-Top 5 modules any method
-```{r imp.mdl.2, fig.height=12, fig.width=12}
-# Top 5 any method
-h = lapply(unique(all.rank$brainRegion), function(br, Dx, MOD.MEMBERS, MODULE.OVRLP){
-  imp.mods = Dx %>%
-    dplyr::group_by(brainRegion) %>%
-    dplyr::top_n(25, rnk) %>%
-    dplyr::top_n(25, Dx) %>%
-    dplyr::filter(brainRegion == br)
-  print(summary(factor(imp.mods$method)))
-
-  getOverlapPlots(imp.mods, MOD.MEMBERS, MODULE.OVRLP)
-}, all.rank, MOD.MEMBERS, MODULE.OVRLP)
-draw(h[[1]], heatmap_legend_side='left')
-draw(h[[2]], heatmap_legend_side='left')
-draw(h[[3]], heatmap_legend_side='left')
-draw(h[[4]], heatmap_legend_side='left')
-draw(h[[5]], heatmap_legend_side='left')
-draw(h[[6]], heatmap_legend_side='left')
-draw(h[[7]], heatmap_legend_side='left')
-```
-
-### Modules missing enrichment scores
-```{r missing, include=FALSE}
-writeLines('Modules missing enrich enrichment scores')
-filter(Dx, ModuleNameFull %in% unique(setdiff(Dx$ModuleNameFull, MODULE.ENRICH$ModuleNameFull))) %>%
-  dplyr::select(ModuleNameFull) %>% unique() %>% unlist() %>% paste(collapse = ', ') %>% print()
-
-writeLines('Modules missing MAGMA enrichment scores')
-filter(Dx, ModuleNameFull %in% unique(setdiff(Dx$ModuleNameFull, MAGMA.ENRICH$Set))) %>%
-  dplyr::select(ModuleNameFull) %>% unique() %>% unlist() %>% paste(collapse = ', ') %>% print()
+  enrichResults = plyr::ldply(geneSetRef, .fun = function(y, modMembers, backgroundGenes){
+    fisherEnrichment(y, modMembers, backgroundGenes)
+  }, modMembers, backgroundGenes, .id = 'GeneSetName', .parallel = T, 
+  .paropts = list(.packages = c('CovariateAnalysis'))) %>%
+    dplyr::mutate(fdr = p.adjust(pval))
+}, 
+.id = 'ModuleNameFull', geneSet.ref, backgroundGenes)
 ```
@@
 ### Store results in synapse
-```{r syn.store, include=FALSE}
-CODE = Folder('Network Association Tests', parentId = "syn10306449")
-CODE = synStore(CODE)
-
+```{r syn.store, include=FALSE, eval = F}
 write.table(Dx, file = 'AssociaitonStatWithDx.tsv', sep = '\t', quote = F, row.names = F)
-OBJ = File('AssociaitonStatWithDx.tsv', name = 'Association with Dx', parentId = CODE@properties$id)
+OBJ = File('AssociaitonStatWithDx.tsv', name = 'Association with Dx', parentId = 'syn10831307')
 OBJ = synStore(OBJ, used = all.used.ids, executed = thisFile, activityName = activityName)
@@
 write.table(Dx.gender, file = 'AssociaitonStatWithDxGender.tsv', sep = '\t', quote = F, row.names = F)
-OBJ1 = File('AssociaitonStatWithDxGender.tsv', name = 'Association with Dx.Gender', parentId = CODE@properties$id)
+OBJ1 = File('AssociaitonStatWithDxGender.tsv', name = 'Association with Dx.Gender', parentId = 'syn10831307')
 OBJ1 = synStore(OBJ1, used = all.used.ids, executed = thisFile, activityName = activityName)
 
-write.table(ApoE, file = 'AssociaitonStatWithApoE4.tsv', sep = '\t', quote = F, row.names = F)
-OBJ2 = File('AssociaitonStatWithApoE4.tsv', name = 'Association with ApoE4', parentId = CODE@properties$id)
+write.table(Pathway.enrich, file = 'PathwayEnrichment.tsv', sep = '\t', quote = F, row.names = F)
+OBJ2 = File('PathwayEnrichment.tsv', name = 'Pathway Enrichment', parentId = 'syn10831307')
 OBJ2 = synStore(OBJ2, used = all.used.ids, executed = thisFile, activityName = activityName)
 
-save(list = c('assoc.stat.diagnosis', 'assoc.stat.diagnosis.gender', 'assoc.stat.apoe'), file = 'AllAssociationResults.RData')
-OBJ3 = File('AllAssociationResults.RData', name = 'All Assocition Statistics (RData)', parentId = CODE@properties$id)
+save(list = c('assoc.stat.diagnosis', 'assoc.stat.diagnosis.gender'), file = 'AllAssociationResults.RData')
+OBJ3 = File('AllAssociationResults.RData', name = 'All Assocition Statistics (RData)', parentId = 'syn10831307')
 OBJ3 = synStore(OBJ3, used = all.used.ids, executed = thisFile, activityName = activityName)
+
+write.table(Dx.enrich, file = 'DiagnosisEnrichment.tsv', sep = '\t', quote = F, row.names = F)
+OBJ4 = File('DiagnosisEnrichment.tsv', name = 'Diagnosis Enrichment', parentId = 'syn10831307')
+OBJ4 = synStore(OBJ4, used = all.used.ids, executed = thisFile, activityName = activityName)
 ```

### 2017-10-09 code/Rmd/networkAssociationTest_metaModules.Rmd
Added markdown to perform association test between Dx and aggregated modules

@@
 
+### Find overlap between modules in different brainregion
+```{r module.overlap}
+
+```
+
+### Intereating modules
+```{r int.modules}
+
+```
+
 ### Store results in synapse
@@
 write.table(Dx, file = 'AssociaitonStatWithDx.tsv', sep = '\t', quote = F, row.names = F)
-OBJ = File('AssociaitonStatWithDx.tsv', name = 'Association with Dx', parentId = 'syn10831307')
+OBJ = File('AssociaitonStatWithDx.tsv', name = 'Association with Dx', parentId = 'syn11140497')
 OBJ = synStore(OBJ, used = all.used.ids, executed = thisFile, activityName = activityName)
@@
 write.table(Dx.gender, file = 'AssociaitonStatWithDxGender.tsv', sep = '\t', quote = F, row.names = F)
-OBJ1 = File('AssociaitonStatWithDxGender.tsv', name = 'Association with Dx.Gender', parentId = 'syn10831307')
+OBJ1 = File('AssociaitonStatWithDxGender.tsv', name = 'Association with Dx.Gender', parentId = 'syn11140497')
 OBJ1 = synStore(OBJ1, used = all.used.ids, executed = thisFile, activityName = activityName)
@@
 write.table(Pathway.enrich, file = 'PathwayEnrichment.tsv', sep = '\t', quote = F, row.names = F)
-OBJ2 = File('PathwayEnrichment.tsv', name = 'Pathway Enrichment', parentId = 'syn10831307')
+OBJ2 = File('PathwayEnrichment.tsv', name = 'Pathway Enrichment', parentId = 'syn11140497')
 OBJ2 = synStore(OBJ2, used = all.used.ids, executed = thisFile, activityName = activityName)
@@
 save(list = c('assoc.stat.diagnosis', 'assoc.stat.diagnosis.gender'), file = 'AllAssociationResults.RData')
-OBJ3 = File('AllAssociationResults.RData', name = 'All Assocition Statistics (RData)', parentId = 'syn10831307')
+OBJ3 = File('AllAssociationResults.RData', name = 'All Assocition Statistics (RData)', parentId = 'syn11140497')
 OBJ3 = synStore(OBJ3, used = all.used.ids, executed = thisFile, activityName = activityName)
@@
 write.table(Dx.enrich, file = 'DiagnosisEnrichment.tsv', sep = '\t', quote = F, row.names = F)
-OBJ4 = File('DiagnosisEnrichment.tsv', name = 'Diagnosis Enrichment', parentId = 'syn10831307')
+OBJ4 = File('DiagnosisEnrichment.tsv', name = 'Diagnosis Enrichment', parentId = 'syn11140497')
 OBJ4 = synStore(OBJ4, used = all.used.ids, executed = thisFile, activityName = activityName)

### 2017-10-10 code/Rmd/networkAssociationTest_metaModules.Rmd
Temp commit

@@
 EXPR$GeneID = NULL
-
-# MODULE.OVRLP = synTableQuery('SELECT * FROM syn10339153')@values
-# all.used.ids = c(all.used.ids, 'syn10339153')
-# MODULE.ENRICH = synTableQuery('SELECT * FROM syn10492048')@values
-# all.used.ids = c(all.used.ids, 'syn10492048')
-# 
-# MAGMA.ENRICH = synTableQuery('SELECT * FROM syn10380432')@values
-# all.used.ids = c(all.used.ids, 'syn10380432')
 ```
@@
 ```
-
 ### Associaiton with Diagnosis
@@
 ```
-
 ### Associaiton with Diagnosis.Gender
-```{r Dx.Gender, fig.height=5, fig.width=10}
+```{r Dx.Gender, fig.height=10, fig.width=10}
 # Find assocition b/w Diagnosis.Gender category with n eigen genes
@@
   kable()
+
+writeLines('Comparison b/w Dx and Dx.gender')
+tmp = dplyr::left_join(Dx %>%
+                         dplyr::select(ModuleName, brainRegion, total.eff.sz, total.fdr) %>%
+                         dplyr::rename(Dx = total.eff.sz, Dx.fdr = total.fdr),
+                       Dx.gender %>%
+                         dplyr::select(ModuleName, brainRegion, total.eff.sz, total.fdr) %>%
+                         dplyr::rename(Dx.gender = total.eff.sz, Dx.gender.fdr = total.fdr))
+p = ggplot(tmp, aes(x = Dx, y = Dx.gender, color = -log10(Dx.fdr+1e-100))) + geom_point() + facet_grid(.~brainRegion) + geom_smooth(method = 'lm')
+p = p +scale_color_continuous(low = 'white', high = 'red')
+p
 ```
-
 ### Enrichment for differential expression
-```{r Dx.enrich}
+```{r Dx.enrich, fig.height=10, fig.width=10}
 # Get differential expression gene list from synapse
@@
 }, .id = 'brainRegion')
+
+writeLines('Comparison b/w Dx and Dx.gender')
+Dx.enrich1 = Dx.enrich %>%
+  dplyr::select(-Genes) %>%
+  tidyr::separate(Dx.GeneSet, c('brainRegion', 'Comparison', 'Gender', 'Direction'), sep = '\\.')
+Dx.enrich1$Direction[is.na(Dx.enrich1$Direction)] = Dx.enrich1$Gender[is.na(Dx.enrich1$Direction)]
+Dx.enrich1$Gender = gsub('UP', 'ALL', Dx.enrich1$Gender)
+Dx.enrich1$Gender = gsub('DOWN', 'ALL', Dx.enrich1$Gender)
+Dx.enrich1 = dplyr::filter(Dx.enrich1, Comparison == 'AD-CONTROL', fdr <= 0.05)
+p = ggplot(Dx.enrich1, aes(x = Gender, y = -log10(fdr))) + geom_boxplot() 
+p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free') + geom_jitter(aes(color = Direction))
+pl = list();pl[[1]] = p
+
+p = ggplot(Dx.enrich1, aes(x = Gender, y = Odds.Ratio)) + geom_boxplot() 
+p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free') + geom_jitter(aes(color = Direction))
+pl[[2]] = p
+multiplot(plotlist = pl, cols = 1)
 ```
 ### Enrichment for gene sets
-```{r pathway.enrich}
+```{r pathway.enrich, results='asis'}
 # Download all related genesets from synapse
@@
 .id = 'ModuleNameFull', geneSet.ref, backgroundGenes)
+
+# Get AD specific pathways
+other.ad.pathways = c("KEGG_2016.Alzheimer's disease_Homo sapiens_hsa05010",
+                      "WikiPathways_2016.Alzheimers Disease_Homo sapiens_WP2059")
+
+writeLines('AD related pathways enriched an fdr of 0.05')
+ad.enrich = Pathway.enrich %>%
+  dplyr::select(ModuleNameFull, GeneSetName, Odds.Ratio, fdr) %>%
+  dplyr::filter(GeneSetName %in% other.ad.pathways, fdr <= 0.05) %>%
+  dplyr::left_join(MODULES %>%
+                     dplyr::select(ModuleNameFull, brainRegion)) %>%
+  unique()
+kable(ad.enrich)
 ```
+### Find overlap between modules in different brainregion
+```{r module.overlap, fig.height=5, fig.width=10}
+# FisherEnrichment b/w modules
+MOD.MEMBERS = plyr::dlply(MODULES, .(ModuleNameFull), .fun = function(x) unique(x$external_gene_name))
+MOD.OVRLP = plyr::ldply(MOD.MEMBERS, .fun = function(x, modMembers, backgroundGenes){
+  plyr::ldply(modMembers, .fun = function(y, x, backgroundGenes){
+    fisherEnrichment(y, x, backgroundGenes)
+  }, x, backgroundGenes, .parallel = T, .paropts = list(.packages = c('CovariateAnalysis')), .id = 'MOD1') %>%
+    dplyr::rename(MOD1 = ModuleNameFull)
+}, MOD.MEMBERS, backgroundGenes, .id = 'MOD2') %>%
+  dplyr::select(-Genes) %>%
+  dplyr::rename(MOD2 = ModuleNameFull)
+```
+### Interesting modules
+```{r int.modules, fig.height=7, fig.width=7, results='asis'}
+# Wrt case control
+tmp = inner_join(Dx %>% 
+                   dplyr::filter(total.fdr <= 0.05) %>%
+                   dplyr::select(ModuleNameFull, brainRegion, total.eff.sz, total.fdr) %>%
+                   dplyr::rename(assoc.eff.sz = total.eff.sz, assoc.fdr = total.fdr),
+                 Dx.enrich1 %>%
+                   dplyr::filter(fdr <= 0.05, Gender == 'ALL') %>%
+                   dplyr::select(ModuleNameFull, brainRegion, Direction, Odds.Ratio, fdr) %>%
+                   dplyr::rename(diffexp.fdr = fdr, diffexp.or = Odds.Ratio)) %>%
+  left_join(ad.enrich %>%
+              dplyr::select(ModuleNameFull, fdr, brainRegion) %>%
+              dplyr::rename(ad.pathway.fdr = fdr))
+writeLines('Imporatant Modules for case control comparison')
+kable(tmp)
 
-### Find overlap between modules in different brainregion
-```{r module.overlap}
+writeLines('Up regulated modules')
+tmp1 = filter(tmp, Direction == 'UP')
+tmp1 = filter(MOD.OVRLP, MOD1 %in% tmp1$ModuleNameFull, MOD2 %in% tmp1$ModuleNameFull)
+g = igraph::graph_from_data_frame(tmp1)
+tmp1.or = igraph::as_adj(g, attr = 'Odds.Ratio') %>% data.matrix
+tmp1.or[is.infinite(tmp1.or)] = 0
 
-```
+tmp1.fdr = igraph::as_adj(g, attr = 'pval') %>% data.matrix()
+tmp1.fdr = tmp1.fdr <= 0.05
+tmp1.fdr[tmp1.fdr] = '*'
+tmp1.fdr[tmp1.fdr == 'FALSE'] = ''
 
-### Intereating modules
-```{r int.modules}
+h = ComplexHeatmap::Heatmap(log2(tmp1.or+1e-2), col = viridis::viridis(5), 
+                        cell_fun = function(j,i,x,y,wifht,height,fill){
+                          grid.text(tmp1.fdr[i,j] %>% format(digits = 1), x, y)
+                        }, name = 'log2(OR)')
+draw(h, heatmap_legend_side = 'left')
+
+dplyr::filter(Pathway.enrich, 
+              GeneSetName %in% grep('CellType', unique(Pathway.enrich$GeneSetName), value = T),
+              fdr <= 0.05, 
+              ModuleNameFull %in% rownames(tmp1.or)) %>%
+  dplyr::select(-Genes)
+
+writeLines('Down regulated modules')
+tmp1 = filter(tmp, Direction == 'DOWN')
+tmp1 = filter(MOD.OVRLP, MOD1 %in% tmp1$ModuleNameFull, MOD2 %in% tmp1$ModuleNameFull)
+g = igraph::graph_from_data_frame(tmp1)
+tmp1.or = igraph::as_adj(g, attr = 'Odds.Ratio') %>% data.matrix
+tmp1.or[is.infinite(tmp1.or)] = 0
+
+tmp1.fdr = igraph::as_adj(g, attr = 'pval') %>% data.matrix()
+tmp1.fdr = tmp1.fdr <= 0.05
+tmp1.fdr[tmp1.fdr] = '*'
+tmp1.fdr[tmp1.fdr == 'FALSE'] = ''
+
+h = ComplexHeatmap::Heatmap(log2(tmp1.or+1e-2), col = viridis::viridis(5), 
+                        cell_fun = function(j,i,x,y,wifht,height,fill){
+                          grid.text(tmp1.fdr[i,j] %>% format(digits = 1), x, y)
+                        }, name = 'log2(OR)')
+draw(h, heatmap_legend_side = 'left')
+
+dplyr::filter(Pathway.enrich, 
+              GeneSetName %in% grep('CellType', unique(Pathway.enrich$GeneSetName), value = T),
+              fdr <= 0.05, 
+              ModuleNameFull %in% rownames(tmp1.or)) %>%
+  dplyr::select(-Genes)
 
@@
 ### Store results in synapse
-```{r syn.store, include=FALSE, eval = F}
+```{r syn.store, include=FALSE, eval = F, cache=FALSE}
 write.table(Dx, file = 'AssociaitonStatWithDx.tsv', sep = '\t', quote = F, row.names = F)

### 2017-10-10 code/R/collateDifferentialExpressionResults.R
Added R function to merge all differential expression results

@@ -1,0 +1,142 @@
+library(synapseClient)
+library(knitr)
+library(githubr)
+
+library(CovariateAnalysis)
+library(data.table)
+library(tidyr)
+library(plyr)
+library(dplyr)
+library(stringr)
+
+synapseLogin()
+
+diff.exp = lapply(c(MAYO = 'syn8468023',
+                    MSSM = 'syn10157628',
+                    ROSMAP = 'syn8456721'),
+                  downloadFile)
+all.used.ids = c('syn8468023', 'syn10157628', 'syn8456721')
+############################################
+# AD-CONTROL
+all.de = rbindlist(list(
+  diff.exp$ROSMAP %>%
+    dplyr::filter(Model == 'Diagnosis', Comparison == 'AD-CONTROL') %>%
+    dplyr::mutate(Study = 'ROSMAP') %>%
+    dplyr::select(-Model, -Comparison) %>%
+    dplyr::rename(Tissue = Region),
+  diff.exp$MSSM %>%
+    dplyr::filter(Model == 'Diagnosis', BrainRegion %in% c('IFG', 'PHG' , 'FP', 'STG'), Comparison == 'CONTROL-AD') %>%
+    dplyr::mutate(Study = 'MSSM', logFC = -logFC, CI.L = -CI.L, CI.R = -CI.R) %>%
+    dplyr::select(-reference, -against, -Model, -Comparison) %>%
+    dplyr::rename(Tissue = BrainRegion, CI.L = CI.R, CI.R = CI.L),
+  diff.exp$MAYO %>%
+    dplyr::filter(BrainRegion.ref %in% c('CER', 'TCX'), Model == 'Diagnosis', Comparison == 'Control-AD') %>%
+    dplyr::mutate(Study = 'MAYO', logFC = -logFC, CI.L = -CI.L, CI.R = -CI.R) %>%
+    dplyr::select(-Model, -Comparison, -BrainRegion.ag, -Gender.ref, -Gender.ag, -AOD.ref, -AOD.ag) %>%
+    dplyr::rename(Tissue = BrainRegion.ref, CI.L = CI.R, CI.R = CI.L)),
+  use.names = T, fill = T) %>%
+  dplyr::mutate(Model = 'Diagnosis', Gender = 'ALL', Comparison = 'AD-CONTROL')
+all.de$Direction[all.de$logFC > 0] = 'UP'
+all.de$Direction[all.de$logFC < 0] = 'DOWN'
+############################################
+
[…]

### 2017-10-16 code/R/collateDifferentialExpressionResults.R
Minor bug fix

@@ -138,5 +138,5 @@ thisFile <- getPermlink(repository = thisRepo, repositoryPath=thisFileName)
 tmp = rbindlist(list(all.de, gender.de, aod.de, apoe.de), use.names = T, fill = T) %>%
   write.table('differentialExpressionSummary.tsv', sep = '\t', row.names = F, quote = F)
 obj = File('differentialExpressionSummary.tsv', name = 'All Differential Expression (Merged)', parentId = 'syn8672415')
-obj = synStore(obj, used = all.used.ids, executed = thisFile, acitivityName = 'Merge all differential expression')
+obj = synStore(obj, used = all.used.ids, executed = thisFile, activityName = 'Merge all differential expression')
 ############################################

### 2017-10-16 code/Rmd/networkAssociationTest_metaModules.Rmd
Major modifications

--- 2017-10-09 code/Rmd/networkAssociationTest_metaModules.Rmd
@@
 }
+
+# Pull gene sets code from Ben 
+pullReferenceGeneSets <- function(url){
+  #pull gene set from Enrichr
+  foo <- RCurl::getURL(url)
+
+  #split by carriage returns
+  fooSplit <- strsplit(foo,"\n")
+
+  #unlist
+  fooSplit <- unlist(fooSplit)
+
+  #split by tab
+  fooSplit <- strsplit(fooSplit,"\t")
+
+  #name them by the first element
+  catNames <- lapply(fooSplit,function(x) x[1])
+  names(fooSplit) <- catNames
+
+  #drop the first two elements
+  fooSplit <- lapply(fooSplit,function(x) x[-c(1,2)])
+
+  fxn1 <- function(x){
+    library(dplyr)
+    strsplit(x,',') %>%
+      sapply(function(y) y[1]) %>%
+      return
+  }
+  #remove 1.0 from elements
+  fooSplit2 <- lapply(fooSplit,fxn1)
+  names(fooSplit2) <- names(fooSplit)
+  return(fooSplit2)
+}
+
+pullADgenes <- function(){
+  genesets1 <- synapseClient::synGet('syn5923958')
+  load(synapseClient::getFileLocation(genesets1))
+  dbgap <- pullReferenceGeneSets("http://amp.pharm.mssm.edu/Enrichr/geneSetLibrary?mode=text&libraryName=dbGaP")
+  #kegg, wikipathways, biocarta, panther, jensen_diseases, omim_disease, omim_expanded
+  kegg <- pullReferenceGeneSets("http://amp.pharm.mssm.edu/Enrichr/geneSetLibrary?mode=text&libraryName=KEGG_2016")
+  wikipathways <- pullReferenceGeneSets("http://amp.pharm.mssm.edu/Enrichr/geneSetLibrary?mode=text&libraryName=WikiPathways_2016")
+  jensen_diseases <- pullReferenceGeneSets("http://amp.pharm.mssm.edu/Enrichr/geneSetLibrary?mode=text&libraryName=Jensen_DISEASES")
+  biocarta <- pullReferenceGeneSets("http://amp.pharm.mssm.edu/Enrichr/geneSetLibrary?mode=text&libraryName=BioCarta_2016")
+  panther <- pullReferenceGeneSets("http://amp.pharm.mssm.edu/Enrichr/geneSetLibrary?mode=text&libraryName=Panther_2016")
+  omim_disease <- pullReferenceGeneSets("http://amp.pharm.mssm.edu/Enrichr/geneSetLibrary?mode=text&libraryName=OMIM_Disease")
+  omim_expanded <- pullReferenceGeneSets("http://amp.pharm.mssm.edu/Enrichr/geneSetLibrary?mode=text&libraryName=OMIM_Expanded")
+
+  adList <- GeneSets$Alzheimers$`AD:GeneticLoci`
+  adList <- c(adList,'HLA-DRB5','HLA-DRB1')
+  adList <- adList[-which(adList=='HLA-DRB5-DRB1')]
+  adList <- list(igap = adList)
+  adList$dbgap <- dbgap$`Alzheimer Disease`
+  adList$kegg <- kegg$`Alzheimer's disease_Homo sapiens_hsa05010`
+  adList$wikipathwaysMouse <- wikipathways$`Alzheimers Disease_Mus musculus_WP2075`
+  adList$wikipathwaysHuman <- wikipathways$`Alzheimers Disease_Homo sapiens_WP2059`
+  adList$jensenDisease <- jensen_diseases$`Alzheimer's_disease`
+  adList$biocarta <- biocarta$`Deregulation of CDK5 in Alzheimers Disease_Homo sapiens_h_p35alzheimersPathway`
+  adList$pantherAmyloid <- panther$`Alzheimer disease-amyloid secretase pathway_Homo sapiens_P00003`
+  adList$pantherPresenilin <- panther$`Alzheimer disease-presenilin pathway_Homo sapiens_P00004`
+  adList$omim <- omim_disease$`alzheimer_disease`
+  adList$omimExpanded <- omim_expanded$`alzheimer_disease`
+
+  genecardsObj <- synapseClient::synGet('syn10507702')
+  genecards <- data.table::fread(synapseClient::getFileLocation(genecardsObj),data.table=F)
+
+  adList$genecards <- genecards$`Gene Symbol`
+  return(adList)
+}
 ```
 
+## Association test
 ### Associaiton with Diagnosis
@@
 ```
-
 ### Associaiton with Diagnosis.Gender
@@
 
-p = ggplot(Dx.gender, aes(x = ngenes, y = total.eff.sz, color = brainRegion)) + geom_point()
-p = p + xlab('No. of genes') + ylab('Total Effect Size')
-p + theme(legend.position = 'top')
-
-p = ggplot(Dx.gender, aes(x = as.factor(npc.sig), y = total.eff.sz, color = brainRegion)) + geom_boxplot()
-p = p + xlab('No. of genes') + ylab('Total Effect Size')
-p + theme(legend.position = 'top')
-
 p = ggplot(Dx.gender, aes(x = brainRegion, y = total.eff.sz)) + geom_boxplot() + geom_jitter(aes(color = total.sig))
@@
 
-p = ggplot(Dx.gender, aes(x = brainRegion, y = pc1.eff.sz, color = -log10(pc1.fdr))) + geom_boxplot() + geom_jitter()
-p = p + coord_flip() + xlab('Brain Region') + ylab('PC1 Effect Size')
-p = p + scale_color_continuous(low = 'white', high = 'red')
+writeLines('Significantly associated modules')
+Dx.gender %>%
+  dplyr::filter(total.fdr <= 0.05) %>%
+  dplyr::group_by(brainRegion) %>%
+  dplyr::arrange(desc(total.eff.sz)) %>%
+  kable()
+```
+Comparison b/w Dx and Dx.gender
+```{r Dx.Gender.1, fig.height=5, fig.width=10}
+tmp = dplyr::left_join(Dx %>%
+                         dplyr::select(ModuleName, brainRegion, total.eff.sz, total.fdr) %>%
+                         dplyr::rename(Dx = total.eff.sz, Dx.fdr = total.fdr),
+                       Dx.gender %>%
+                         dplyr::select(ModuleName, brainRegion, total.eff.sz, total.fdr) %>%
+                         dplyr::rename(Dx.gender = total.eff.sz, Dx.gender.fdr = total.fdr))
+p = ggplot(tmp, aes(x = Dx, y = Dx.gender, color = -log10(Dx.fdr+1e-100))) + geom_point() + facet_grid(.~brainRegion) + geom_smooth(method = 'lm')
+p = p +scale_color_continuous(low = 'white', high = 'red')
+p
+```
+### Association with Diagnosis in female samples
+```{r Dx.female, fig.height=5, fig.width=10}
+# Find assocition b/w Diagnosis category with n eigen genes in female samples
+FORMULA = as.formula('PC~Diagnosis+Batch+RIN+PCT_INTRONIC_BASES+PMI+PCT_INTERGENIC_BASES+AOD+PCT_PF_READS_ALIGNED')
+ind = COVARIATES$SampleID[COVARIATES$Gender == 'FEMALE']
+assoc.stat.diagnosis.female = MODULES %>%
+  dplyr::filter(GeneID %in% rownames(EXPR)) %>%
+  plyr::dlply(.(ModuleName, brainRegion), .fun = findAssociationStatistics, 
+              COVARIATES[COVARIATES$SampleID %in% ind,], EXPR[,ind], FORMULA,
+              .parallel = T, .paropts = list(.packages = c('dplyr', 'data.table')))
+
+Dx.female = lapply(assoc.stat.diagnosis.female, function(x) x$assoc.stat) %>%
+  data.table::rbindlist(use.names = T, fill = T, idcol = 'ModuleName.BrainRegion') %>%
+  tidyr::separate(ModuleName.BrainRegion, c('ModuleName', 'brainRegion'), sep = '\\.') %>%
+  dplyr::left_join(MODULES %>%
+                     dplyr::select(Module, method, ModuleName, brainRegion, ModuleNameFull) %>%
+                     unique()) %>%
+  plyr::ddply(.(brainRegion), .fun = function(x){
+    x %>%
+      dplyr::mutate(total.fdr = p.adjust(total.pval))
+  }) %>%
+  dplyr::mutate(total.sig = total.fdr <= 0.05)
+
+p = ggplot(Dx.female, aes(x = brainRegion, y = total.eff.sz)) + geom_boxplot() + geom_jitter(aes(color = total.sig))
+p = p + coord_flip() + xlab('Brain Region') + ylab('Total Effect Size')
+# p = p + scale_color_continuous(low = 'white', high = 'red')
 p + theme(legend.position = 'top')
 
-writeLines('Number of significantly associated modules')
-Dx.gender %>%
+writeLines('Significantly associated modules')
+Dx.female %>%
+  dplyr::filter(total.fdr <= 0.05) %>%
   dplyr::group_by(brainRegion) %>%
-  dplyr::summarise(nsig = sum(total.fdr <= 0.05),
-                   ntotal = sum(total.fdr <= 1)) %>%
+  dplyr::arrange(desc(total.eff.sz)) %>%
   kable()
+```
+### Association with Diagnosis in male samples
+```{r Dx.male, fig.height=5, fig.width=10}
+# Find assocition b/w Diagnosis category with n eigen genes in female samples
+FORMULA = as.formula('PC~Diagnosis+Batch+RIN+PCT_INTRONIC_BASES+PMI+PCT_INTERGENIC_BASES+AOD+PCT_PF_READS_ALIGNED')
+ind = COVARIATES$SampleID[COVARIATES$Gender == 'MALE']
+assoc.stat.diagnosis.male = MODULES %>%
+  dplyr::filter(GeneID %in% rownames(EXPR)) %>%
+  plyr::dlply(.(ModuleName, brainRegion), .fun = findAssociationStatistics, 
+              COVARIATES[COVARIATES$SampleID %in% ind,], EXPR[,ind], FORMULA,
+              .parallel = T, .paropts = list(.packages = c('dplyr', 'data.table')))
 
-writeLines('Percent of significantly associated modules')
-Dx.gender %>%
+Dx.male = lapply(assoc.stat.diagnosis.female, function(x) x$assoc.stat) %>%
+  data.table::rbindlist(use.names = T, fill = T, idcol = 'ModuleName.BrainRegion') %>%
+  tidyr::separate(ModuleName.BrainRegion, c('ModuleName', 'brainRegion'), sep = '\\.') %>%
+  dplyr::left_join(MODULES %>%
+                     dplyr::select(Module, method, ModuleName, brainRegion, ModuleNameFull) %>%
+                     unique()) %>%
+  plyr::ddply(.(brainRegion), .fun = function(x){
+    x %>%
+      dplyr::mutate(total.fdr = p.adjust(total.pval))
+  }) %>%
+  dplyr::mutate(total.sig = total.fdr <= 0.05)
+
+p = ggplot(Dx.male, aes(x = brainRegion, y = total.eff.sz)) + geom_boxplot() + geom_jitter(aes(color = total.sig))
+p = p + coord_flip() + xlab('Brain Region') + ylab('Total Effect Size')
+# p = p + scale_color_continuous(low = 'white', high = 'red')
+p + theme(legend.position = 'top')
+
+writeLines('Significantly associated modules')
+Dx.male %>%
+  dplyr::filter(total.fdr <= 0.05) %>%
   dplyr::group_by(brainRegion) %>%
-  dplyr::summarise(nsig = round(sum(total.fdr <= 0.05)/sum(total.fdr <= 1) * 100, 2)) %>%
-  kable()
-
-writeLines('Percent of modules whose PC1 is significantly associated')
-Dx.gender %>%
-  dplyr::group_by(brainRegion) %>%
-  dplyr::summarise(nsig = round(sum(pc1.fdr <= 0.05)/sum(pc1.fdr <= 1) * 100, 2)) %>%
+  dplyr::arrange(desc(total.eff.sz)) %>%
   kable()
@@
 
+## Enrichment analysis
 ### Enrichment for differential expression
-```{r Dx.enrich}
+```{r Dx.enrich, fig.height=10, fig.width=12}
 # Get differential expression gene list from synapse
@@
 }, .id = 'brainRegion')
+
+writeLines('Comparison b/w Dx and Dx.gender')
+Dx.enrich1 = Dx.enrich %>%
+  dplyr::select(-Genes) %>%
+  tidyr::separate(Dx.GeneSet, c('brainRegion', 'Comparison', 'Gender', 'Direction'), sep = '\\.')
+Dx.enrich1$Direction[is.na(Dx.enrich1$Direction)] = Dx.enrich1$Gender[is.na(Dx.enrich1$Direction)]
+Dx.enrich1$Gender = gsub('UP', 'ALL', Dx.enrich1$Gender)
+Dx.enrich1$Gender = gsub('DOWN', 'ALL', Dx.enrich1$Gender)
+Dx.enrich1 = dplyr::filter(Dx.enrich1, Comparison == 'AD-CONTROL', fdr <= 0.05)
+p = ggplot(Dx.enrich1, aes(x = Gender, y = -log10(fdr))) + geom_boxplot() 
+p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free') + geom_jitter(aes(color = Direction))
+pl = list();pl[[1]] = p + ylab('') + theme(legend.position = 'top')
+
+p = ggplot(Dx.enrich1, aes(x = Gender, y = Odds.Ratio)) + geom_boxplot() 
+p = p + coord_flip() + facet_grid(.~brainRegion, scales = 'free') + geom_jitter(aes(color = Direction))
+pl[[2]] = p + ylab('') + theme(legend.position = 'top')
+multiplot(plotlist = pl, cols = 1)
 ```
-### Enrichment for gene sets
-```{r pathway.enrich}
-# Download all related genesets from synapse
+
+### Enrichment for AD related gene sets
+```{r AD.enrich, results='asis'}
+# Get all AD related genesets
+GeneSets = pullADgenes()
+
+backgroundGenes = unique(MODULES$external_gene_name)
+geneSet.ref = list(AD = GeneSets) %>%
+  CovariateAnalysis::filterGeneSets(backgroundGenes)
+geneSet.ref = geneSet.ref$AD
+
+# Perform enrichment analysis
+AD.enrich = plyr::ddply(MODULES, .(ModuleNameFull), .fun = function(mod, geneSetRef, backgroundGenes){
+  modMembers = unique(mod$external_gene_name)
+
+  enrichResults = plyr::ldply(geneSetRef, .fun = function(y, modMembers, backgroundGenes){
+    fisherEnrichment(y, modMembers, backgroundGenes)
+  }, modMembers, backgroundGenes, .id = 'GeneSetName', .parallel = T, 
+  .paropts = list(.packages = c('CovariateAnalysis'))) %>%
+    dplyr::mutate(fdr = p.adjust(pval))
+}, 
+.id = 'ModuleNameFull', geneSet.ref, backgroundGenes)
+
+writeLines('AD related pathways enriched an fdr of 0.05')
+AD.enrich %>%
+  dplyr::select(ModuleNameFull, GeneSetName, Odds.Ratio, fdr) %>%
+  dplyr::filter(fdr <= 0.05) %>%
+  dplyr::left_join(MODULES %>%
+                     dplyr::select(ModuleNameFull, brainRegion)) %>%
+  unique() %>%
+  dplyr::arrange(desc(brainRegion), desc(Odds.Ratio)) %>%
+  kable()
+```
+
+### Enrichment for other gene sets
+```{r pathway.enrich, results='asis'}
+# Get all enrichr genesets
 load(synGet('syn4867851')@filePath)
@@
 
-### Find overlap between modules in different brainregion
-```{r module.overlap}
-
+## Module Selection
+### Interesting modules for diagnosis
+```{r int.modules.1, fig.height=7, fig.width=7, results='asis'}
+# Wrt case control
+Dx.modules = dplyr::inner_join(Dx %>% 
+                                 dplyr::filter(total.fdr <= 0.05) %>%
+                                 dplyr::select(ModuleNameFull, brainRegion, total.eff.sz, total.fdr) %>%
+                                 dplyr::rename(assoc.eff.sz = total.eff.sz, assoc.fdr = total.fdr),
+                               Dx.enrich1 %>%
+                                 dplyr::filter(fdr <= 0.05, Gender == 'ALL') %>%
+                                 dplyr::select(ModuleNameFull, brainRegion, Direction, Odds.Ratio, fdr) %>%
+                                 dplyr::rename(diffexp.fdr = fdr, diffexp.or = Odds.Ratio)) %>%
+  dplyr::left_join(AD.enrich %>%
+                     dplyr::filter(fdr <= 0.05) %>%
+                     dplyr::select(ModuleNameFull, GeneSetName, Odds.Ratio, fdr) %>%
+                     dplyr::rename(ad.fdr = fdr, ad.or = Odds.Ratio)) %>%
+  dplyr::arrange((brainRegion), (Direction), desc(assoc.eff.sz))
+writeLines('Imporatant Modules for case control comparison')
+kable(Dx.modules)
 ```
 
-### Intereating modules
-```{r int.modules}
+### Interesting modules for diagnosis.gender
+```{r int.modules.2, fig.height=7, fig.width=7, results='asis'}
+# Wrt gender specific case control
+Dx.gender.modules = dplyr::inner_join(Dx.gender %>% 
+                                        dplyr::filter(total.fdr <= 0.05) %>%
+                                        dplyr::select(ModuleNameFull, brainRegion, total.eff.sz, total.fdr) %>%
+                                        dplyr::rename(assoc.eff.sz = total.eff.sz, assoc.fdr = total.fdr),
+                                      Dx.enrich1 %>%
+                                        dplyr::filter(fdr <= 0.05, Gender %in% c('MALE', 'FEMALE')) %>%
+                                        dplyr::select(ModuleNameFull, brainRegion, Direction, Odds.Ratio, fdr) %>%
+                                        dplyr::rename(diffexp.fdr = fdr, diffexp.or = Odds.Ratio)) %>%
+  dplyr::left_join(AD.enrich %>%
+                     dplyr::filter(fdr <= 0.05) %>%
+                     dplyr::select(ModuleNameFull, GeneSetName, Odds.Ratio, fdr) %>%
+                     dplyr::rename(ad.fdr = fdr, ad.or = Odds.Ratio)) %>%
+  dplyr::arrange((brainRegion), (Direction), desc(assoc.eff.sz))
+writeLines('Imporatant Modules for gender specific case control comparison')
+kable(Dx.gender.modules)
 
+# Wrt case control in females
+Dx.female.modules = dplyr::inner_join(Dx.female %>% 
+                                        dplyr::filter(total.fdr <= 0.05) %>%
+                                        dplyr::select(ModuleNameFull, brainRegion, total.eff.sz, total.fdr) %>%
+                                        dplyr::rename(assoc.eff.sz = total.eff.sz, assoc.fdr = total.fdr),
+                                      Dx.enrich1 %>%
+                                        dplyr::filter(fdr <= 0.05, Gender %in% c('FEMALE')) %>%
+                                        dplyr::select(ModuleNameFull, brainRegion, Direction, Odds.Ratio, fdr) %>%
+                                        dplyr::rename(diffexp.fdr = fdr, diffexp.or = Odds.Ratio)) %>%
+  dplyr::left_join(AD.enrich %>%
+                     dplyr::filter(fdr <= 0.05) %>%
+                     dplyr::select(ModuleNameFull, GeneSetName, Odds.Ratio, fdr) %>%
+                     dplyr::rename(ad.fdr = fdr, ad.or = Odds.Ratio)) %>%
+  dplyr::arrange((brainRegion), (Direction), desc(assoc.eff.sz))
+writeLines('Imporatant Modules for case control comparison in females')
+kable(Dx.female.modules)
+
+# Wrt case control in males
+Dx.male.modules = dplyr::inner_join(Dx.male %>% 
+                                      dplyr::filter(total.fdr <= 0.05) %>%
+                                      dplyr::select(ModuleNameFull, brainRegion, total.eff.sz, total.fdr) %>%
+                                      dplyr::rename(assoc.eff.sz = total.eff.sz, assoc.fdr = total.fdr),
+                                    Dx.enrich1 %>%
+                                      dplyr::filter(fdr <= 0.05, Gender %in% c('MALE')) %>%
+                                      dplyr::select(ModuleNameFull, brainRegion, Direction, Odds.Ratio, fdr) %>%
+                                      dplyr::rename(diffexp.fdr = fdr, diffexp.or = Odds.Ratio)) %>%
+  dplyr::left_join(AD.enrich %>%
+                     dplyr::filter(fdr <= 0.05) %>%
+                     dplyr::select(ModuleNameFull, GeneSetName, Odds.Ratio, fdr) %>%
+                     dplyr::rename(ad.fdr = fdr, ad.or = Odds.Ratio)) %>%
+  dplyr::arrange((brainRegion), (Direction), desc(assoc.eff.sz))
+writeLines('Imporatant Modules for case control comparison in males')
+kable(Dx.male.modules)
+```
+### Find overlap between modules in different brainregion
+```{r module.overlap, fig.height=5, fig.width=10, eval = F}
+# FisherEnrichment b/w modules
+MOD.MEMBERS = plyr::dlply(MODULES, .(ModuleNameFull), .fun = function(x) unique(x$external_gene_name))
+MOD.OVRLP = plyr::ldply(MOD.MEMBERS, .fun = function(x, modMembers, backgroundGenes){
+  plyr::ldply(modMembers, .fun = function(y, x, backgroundGenes){
+    fisherEnrichment(y, x, backgroundGenes)
+  }, x, backgroundGenes, .parallel = T, .paropts = list(.packages = c('CovariateAnalysis')), .id = 'MOD1') %>%
+    dplyr::rename(MOD1 = ModuleNameFull)
+}, MOD.MEMBERS, backgroundGenes, .id = 'MOD2') %>%
+  dplyr::select(-Genes) %>%
+  dplyr::rename(MOD2 = ModuleNameFull)
 ```
@@
 
-write.table(Pathway.enrich, file = 'PathwayEnrichment.tsv', sep = '\t', quote = F, row.names = F)
-OBJ2 = File('PathwayEnrichment.tsv', name = 'Pathway Enrichment', parentId = 'syn11140497')
+write.table(Dx.feamle, file = 'AssociaitonStatWithDxFemale.tsv', sep = '\t', quote = F, row.names = F)
+OBJ2 = File('AssociaitonStatWithDxFemale.tsv', name = 'Association with Dx.Female', parentId = 'syn11140497')
 OBJ2 = synStore(OBJ2, used = all.used.ids, executed = thisFile, activityName = activityName)
 
-save(list = c('assoc.stat.diagnosis', 'assoc.stat.diagnosis.gender'), file = 'AllAssociationResults.RData')
-OBJ3 = File('AllAssociationResults.RData', name = 'All Assocition Statistics (RData)', parentId = 'syn11140497')
+write.table(Dx.male, file = 'AssociaitonStatWithDxMale.tsv', sep = '\t', quote = F, row.names = F)
+OBJ3 = File('AssociaitonStatWithDxMale.tsv', name = 'Association with Dx.Male', parentId = 'syn11140497')
 OBJ3 = synStore(OBJ3, used = all.used.ids, executed = thisFile, activityName = activityName)
 
-write.table(Dx.enrich, file = 'DiagnosisEnrichment.tsv', sep = '\t', quote = F, row.names = F)
-OBJ4 = File('DiagnosisEnrichment.tsv', name = 'Diagnosis Enrichment', parentId = 'syn11140497')
+write.table(Dx.enrich, file = 'DiffEnrichment.tsv', sep = '\t', quote = F, row.names = F)
+OBJ4 = File('DiffEnrichment.tsv', name = 'Differential Expression Enrichment', parentId = 'syn11140497')
 OBJ4 = synStore(OBJ4, used = all.used.ids, executed = thisFile, activityName = activityName)
+
+write.table(AD.enrich, file = 'ADEnrichment.tsv', sep = '\t', quote = F, row.names = F)
+OBJ5 = File('ADEnrichment.tsv', name = 'AD Enrichment', parentId = 'syn11140497')
+OBJ5 = synStore(OBJ5, used = all.used.ids, executed = thisFile, activityName = activityName)
+
+write.table(Pathway.enrich, file = 'PathwayEnrichment.tsv', sep = '\t', quote = F, row.names = F)
+OBJ6 = File('PathwayEnrichment.tsv', name = 'Pathway Enrichment', parentId = 'syn11140497')
+OBJ6 = synStore(OBJ6, used = all.used.ids, executed = thisFile, activityName = activityName)
+
+save(list = c('assoc.stat.diagnosis', 'assoc.stat.diagnosis.gender', 
+              'assoc.stat.diagnosis.female', 'assoc.stat.diagnosis.male'), 
+     file = 'AllAssociationResults.RData')
+OBJ7 = File('AllAssociationResults.RData', name = 'All Assocition Statistics (RData)', parentId = 'syn11140497')
+OBJ7 = synStore(OBJ7, used = all.used.ids, executed = thisFile, activityName = activityName)
 ```

### 2017-10-19 code/R/myGetGeneLengthAndGCContent.R
Added function to get gene length and gc content

@@ -1,0 +1,41 @@
+myGetGeneLengthAndGCContent <- function(id){
+  
+  org = 'hsa'
+  id.type = 'ensembl_gene_id'
+  host = 'dec2016.archive.ensembl.org'
+  
+  message("Connecting to BioMart ...")
+  ensembl <- useMart("ENSEMBL_MART_ENSEMBL", host = host)
+  ds <- listDatasets(ensembl)[, "dataset"]
+  ds <- grep(paste0("^", org), ds, value = TRUE)
+  
+  if (length(ds) == 0){
+    stop(paste("Mart not found for:", org))
+  } else if (length(ds) > 1) {
+    message("Found several marts")
+    sapply(ds, function(d) message(paste(which(ds == d), d, sep = ": ")))
+    n <- readline(paste0("Choose mart (1-", length(ds), ") : "))
+    ds <- ds[as.integer(n)]
+  }
+  
+  ensembl <- useDataset(ds, mart = ensembl)
+  message(paste0("Downloading sequence", ifelse(length(id) > 1, "s", ""), " ..."))
+  if (length(id) > 100)
+    message("This may take a few minutes ...")
+  attrs <- c(id.type, "ensembl_exon_id", "chromosome_name", "exon_chrom_start", "exon_chrom_end")
+  coords <- getBM(filters = id.type, attributes = attrs, values = id, mart = ensembl)
+  id <- unique(coords[, id.type])
+  coords <- sapply(id, function(i) {
+    i.coords <- coords[coords[, 1] == i, 3:5]
+    g <- GRanges(i.coords[, 1], IRanges(i.coords[, 2], i.coords[, 3]))
+    return(g)
+  })
+  coords <- lapply(coords[id], reduce)
+  len <- plyr::ldply(coords, function(x) sum(IRanges::width(x)), .id = 'ensembl_gene_id') %>%
+    dplyr::rename(gne.length = V1)
+  
+  gc.cont <- getBM(filters = id.type, attributes = c(id.type, 'hgnc_symbol', 'percentage_gc_content'), values = id, mart = ensembl)
+  
+  res <- dplyr::full_join(gc.cont, len)
+  return(res)
+}

### 2017-12-21 code/Rmd/MSSM_REPROCESSED.Rmd
Minor fix in MSSM code to accommodate recent changes in individual ID issue.

@@
 # Get reprocessed sample counts for temporal cortex
-COUNT_ID <- 'syn8691099';#'syn7994853';
+COUNT_ID <- 'syn8691099';
 ALL_USED_IDs <- COUNT_ID
@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_AGE_ID
-METADATA_AGE <- read.table(synGet(METADATA_AGE_ID)@filePath, sep='\t', header=T) %>%
-  dplyr::rename(individualIdentifier = BBid)
+METADATA_AGE <- read.table(synGet(METADATA_AGE_ID)@filePath, sep='\t', header=T)
 

### 2017-12-22 code/Rmd/ALL_REPROCESSED.Rmd
temp commit

@@
 ---
-title: "Differential expression analysis of reprocessed AMPAD RNASeq data (All 7 brain regions with CQN normlaisation and BIC based model selection)"
+title: "Differential expression analysis of reprocessed AMPAD RNASeq data (All 7 brain regions with CQN normlaisation and PCA
+based model selection)"
 author: "Thanneer Perumal"

### 2017-12-22 code/Rmd/ALL_REPROCESSED_v2.Rmd
temp commit

--- 2017-07-07 code/Rmd/ALL_REPROCESSED.Rmd
@@
 
-knit2synapse::knitToFolderEntity(file = "./ALL_REPROCESSED.Rmd",
+knit2synapse::knitToFolderEntity(file = "./ALL_REPROCESSED_v2.Rmd",
                                  parentId = "syn8672415",
-                                 entityName = 'ALL Reprocessed RNASeq (Normalised)')
+                                 entityName = 'ALL Reprocessed RNASeq (Normalised - sparse model)')
 ```
@@
 activityName = 'Covariate adjustments';
-activityDescription = 'Covariate analysis of reprocessed rnaseq from all 7 brain regions';
-
-thisFileName <- 'ALL_REPROCESSED.Rmd'
+activityDescription = 'Covariate analysis of reprocessed RNASeq from all 7 brain regions';
+
+thisFileName <- 'ALL_REPROCESSED_v2.Rmd'
 
@@
 FactorCovariates <- c("individualIdentifier", "Study", "BrainRegion.Diagnosis", "BrainRegion", "Diagnosis",
-                      "Gender", "Batch")
-ContCovariates <- c("AOD", "PMI", "RIN", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES",
-                    "PCT_RIBOSOMAL_BASES")
+                      "Gender", "Batch", "BrainRegion.ApoE")
+ContCovariates <- c("AOD", "PMI", "RIN", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES", 
+                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
 
@@
 
-### Model Identification
+### Normalisation (iterative design)
+1. BrainRegion.Diagnosis is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
+2. Individual_ID is chosen as random effect
 ```{r fxn}
@@
   # Estimate voom weights
-  voomWeights = limma::voom(cnt[rownames(lcpm),colnames(lcpm)], design = designMat, plot=F)$weights
+  cnt[is.na(cnt)] = 0
+  voomWeights = limma::voom(cnt, design = designMat, plot=F)$weights
 
@@
                                              designMat[rownames(expr),],
-                                             covars[rownames(designMat),groupingVar,drop=F])
-                                formula = paste(setdiff(colnames(data), c('Gx', groupingVar)), collapse = '+') %>%
-                                  paste('Gx ~',.) %>%
-                                  c(.,paste0('(1|',groupingVar,')')) %>%
-                                  paste(collapse = '+')
+                                             Individual_ID = groupingVar)
+                                formula = paste(setdiff(colnames(data), c('Gx', 'Individual_ID')), collapse = '+') %>%
+                                  paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')
 
@@
                                 })
-                              return(data.frame(Gene.ID = rownames(lcpm)[i],
+                              return(data.frame(ensembl_gene_id = rownames(lcpm)[i],
                                                 BIC = BIC))
@@
                             modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
-                                                           c('individualIdentifier'), 
+                                                           covars[colnames(cnt),'individualIdentifier'], 
                                                            lcpm, 
@@
     nChange = allModelBIC %>%
-      tidyr::gather(ModelName, Value, -Gene.ID) %>%
+      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
       dplyr::left_join(baseModelBIC) %>%
@@
                        worse = sum(diff <= -5)) %>%
-      dplyr::mutate(nchange = (better/(better+worse)) * 100)
+      dplyr::mutate(nchange = better)
 
@@
     tmp = nChange %>%
-      dplyr::filter(nchange > 1) %>%
+      dplyr::filter(nchange > round(dim(cnt)[1]*0.01)) %>%
       dplyr::top_n(1, nchange)
@@
       baseModelBIC = allModelBIC %>%
-        dplyr::select(Gene.ID, one_of(tmp$ModelName))
+        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
       colnames(baseModelBIC)[2] = 'baseModel'
@@
 baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   c('individualIdentifier'), 
+                                   COVARIATES[colnames(NEW.COUNTS),'individualIdentifier'], 
                                    CQN.GENE_EXPRESSION$E, 
@@
 # Iteratively identify covariates in phase I
-phase3.covars = c("PCT_PF_READS_ALIGNED", "PCT_CODING_BASES", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
+phase3.covars = c("PCT_PF_READS_ALIGNED", "PCT_CODING_BASES", "PCT_INTERGENIC_BASES", 
+                  "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
 vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
@@
 Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`
-
-### Normalisation
-1. BrainRegion.Diagnosis is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
-2. Individual_ID is chosen as random effect
-```{r iterative.normalisation, results='asis'}
-writeLines(paste('Using following covariates in the model:',
-                 paste(adjust.covars, collapse=', '),
-                 'as fixed effects and Individual_ID is chosen as random effect'))
-
-# Post adjusted design matrix
-DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
-DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
-
-# Estimate voom weights
-VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F)
-
-# Estimate consensus correlation coefficient
-correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
-
-# Re-calculate voom weights with correlation of random effects
-VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, 
-                            design=DM1$design, plot=F,
-                            block = COVARIATES$Individual_ID, 
-                            correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
-ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
-
-# Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
-
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, COVARIATES,
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
-```
 
@@
 
+### Differential expression analysis (with BrainRegion.ApoE as primary variable)
+Differential expression is performed on the primary variable by controlling for covariates identified above
+Genes that are differentially expressed at an FDR <= 0.05 are
+```{r diff.exp2, fig.height=7, fig.width=24}
+# Get design matrix
+DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.ApoE', setdiff(postAdjustCovars, 'BrainRegion.Diagnosis')), drop = F],
+                         Intercept = F)
+DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
+
+# Estimate voom weights for dispersion control
+VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)
+
+# Calculate correlation values of random effects
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)
+
+# Re-estimate voom weights
+VOOM.WEIGHTS = voom(NEW.COUNTS,
+                    design=DESIGN$design, plot=F,
+                    block = COVARIATES$individualIdentifier, 
+                    correlation = correlation$cor)
+
+# Fit linear model using new weights and new design
+VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
+FIT = lmFit(VOOM.WEIGHTS)
+
+# Fit contrast
+cntr = c()
+for (region in c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
+  tmp = combn(paste0('BrainRegion.ApoE', region, '_', c('0', '1', '2')),2)
+  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
+}
+
+contrast = makeContrasts(contrasts=cntr,
+                         levels = colnames(FIT$coefficients))
+FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = lapply(1:dim(contrast)[2], function(i, FIT.CONTR){
+  topTable(FIT.CONTR, coef=i, number = 50000, confint = T) %>%
+    rownameToFirstColumn('Gene.ID')
+}, FIT.CONTR)
+names(DE) = colnames(contrast)
+
+DE = DE %>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Comparison = gsub('BrainRegion.ApoE','',Comparison),
+                Study = 'ALL',
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>%
+  left_join(GENE.PARAM %>%
+              dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
+              unique())
+DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE %>%
+  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
+  group_by(Comparison, Direction) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
+  spread(Direction, FDR_0_05_FC_1.2) 
+kable(tmp)
+
+p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position='top')
+p = p + facet_grid(.~Comparison, scales = 'fixed')
+p
+
+all.diff.exp = c(all.diff.exp, list(ApoE = DE))
+all.fit = c(all.fit, list(ApoE = FIT))
+```
+
 ### Differential expression analysis (with BrainRegion.Diagnosis and Gender)
@@
 geneSet.ref = GeneSets[c('KEGG_2016', 'GO_Biological_Process_2015', 'Reactome_2016', 'WikiPathways_2016')] %>%
+lapply(function(x){
+lapply(x, function(y){
+sapply(y, function(y) str_split(y, '\\,')[[1]][1])
+})
+}) %>%
 CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
@@
 DEG_OBJ = synStore(DEG_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+
+# Store all model fits
+save(list = 'all.fit', file = 'AMP_allModelFits.RData')
+AFIT_OBJ = File('AMP_allModelFits.RData',
+                name = 'All Model Fit', 
+                parentId = CODE$properties$id)
+annotations(AFIT_OBJ) = annotations(CODE)
+annotations(AFIT_OBJ)$dataSubType = 'allModelFit'
+AFIT_OBJ = synStore(AFIT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+synChangeEntityACL(AFIT_OBJ@properties$id, new.racl)
 

### 2017-12-22 code/Rmd/MAYO_REPROCESSED.Rmd
temp commit

--- 2017-09-05 code/Rmd/MAYO_REPROCESSED.Rmd
@@
 Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diff.exp3, fig.height=5, fig.width=18}
+```{r diff.exp3, fig.height=5, fig.width=18, cache = FALSE}
 # Get design matrix

### 2017-12-22 code/Rmd/MAYO_REPROCESSED_DS.Rmd
Added reprocessing markdown for downsampled Mayo rnaseq data

--- 2017-12-22 code/Rmd/MAYO_REPROCESSED.Rmd
@@
 ---
-title: "Covariate analysis of MAYO reprocessed counts (with CQN normalisation and covariates adjusted)"
+title: "Covariate analysis of MAYO 50% downsampled reprocessed counts (with CQN normalisation and covariates adjusted)"
 author: "Thanneer Perumal"
@@
 
-knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED.Rmd",
+knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED_DS.Rmd",
                                  parentId = "syn8672415",
-                                 entityName = 'MAYO Reprocessed RNASeq (Normalised)')
+                                 entityName = 'MAYO 50percent Down Sampled Reprocessed RNASeq')
 ```
@@
 
-thisFileName <- 'MAYO_REPROCESSED.Rmd'
+thisFileName <- 'MAYO_REPROCESSED_DS.Rmd'
 
@@
 ```{r download.data}
-# Get reprocessed sample counts for temporal cortex
-COUNT_TC_ID <- 'syn8690799';#'syn7987765';
-ALL_USED_IDs <- COUNT_TC_ID
-COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F, row.names = 1)
-
-# Get reprocessed sample counts for cerebellum
-COUNT_CE_ID <- 'syn8690904';#'syn7513807';
-ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_CE_ID
-COUNT_CE <- read.table(synGet(COUNT_CE_ID)@filePath, header=T, sep='\t', check.names = F, row.names = 1)
-
-COUNT = cbind(COUNT_TC, COUNT_CE[rownames(COUNT_TC),])
+# Get reprocessed sample counts for cerebellum and temporal cortex
+COUNT_ID = 'syn11635112'
+ALL_USED_IDs = COUNT_ID
+COUNT = read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F, row.names = 1)
 
@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TC_ID
-METADATA_TC <- read.table(synGet(METADATA_TC_ID)@filePath,sep='\t',header=T, stringsAsFactors=F) %>%
+METADATA_TC <- fread(synGet(METADATA_TC_ID)@filePath,fill=T) %>%
   tidyr::separate(ID, c('Donor_ID', 'BrainRegion'), sep = '_') %>%
@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CE_ID
-METADATA_CE <- read.table(synGet(METADATA_CE_ID)@filePath,sep='\t',header=T, stringsAsFactors=F) %>%
+METADATA_CE <- fread(synGet(METADATA_CE_ID)@filePath, fill = T) %>%
   tidyr::separate(SampleID, c('Donor_ID', 'BrainRegion')) %>%
@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_CE_ID
-# METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = FALSE, exdir = getwd())
-# METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = TRUE, exdir = getwd())
-# METADATA.PICARD_CE_FPATHS = METADATA.PICARD_CE_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_CE_FPATHS)]
-# METADATA.PICARD_CE <- plyr::ldply(METADATA.PICARD_CE_FPATHS, read.csv)
-# 
-# METADATA.PICARD_CE$ID = gsub('picard/', '', METADATA.PICARD_CE_FPATHS) %>%
-#   gsub('/picard.CombinedMetrics.csv','',.) %>%
-#   sapply(function(x){str_split(x, '\\.')[[1]][1]})
 METADATA.PICARD_CE <- CovariateAnalysis::downloadFile(METADATA.PICARD_CE_ID) %>%
@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_TC_ID
-# METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = FALSE, exdir = getwd())
-# METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = TRUE, exdir = getwd())
-# METADATA.PICARD_TC_FPATHS = METADATA.PICARD_TC_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_TC_FPATHS)]
-# METADATA.PICARD_TC <- plyr::ldply(METADATA.PICARD_TC_FPATHS, read.csv)
-# METADATA.PICARD_TC$ID = gsub('picard/', '', METADATA.PICARD_TC_FPATHS) %>%
-#   gsub('/picard.CombinedMetrics.csv','',.) %>%
-#   sapply(function(x){str_split(x, '\\.')[[1]][1]}) %>%
-#   gsub('Aligned','',.)
-
 METADATA.PICARD_TC <- CovariateAnalysis::downloadFile(METADATA.PICARD_TC_ID) %>%
@@
   tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
-  group_by(BrainRegion, Diagnosis) %>%
-  summarise(count = n()) %>%
-  spread(Diagnosis, count)
+  dplyr::group_by(BrainRegion, Diagnosis) %>%
+  dplyr::count() %>%
+  tidyr::spread(Diagnosis, n)
 ```
@@
 
-# Fit linear model using new weights and new design
+# Fit linear model using new weights dplyr::count()and new design
 VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
@@
   assay	 = 'RNAseq',
-  assaySubType = 'Reprocessed',
+  assaySubType = 'Reprocessed down sampled',
 
@@
 # Code
-CODE <- Folder(name = "MAYO Reprocessed RNASeq (Normalised)", parentId = parentId)
+CODE <- Folder(name = "MAYO 50percent Down Sampled Reprocessed RNASeq", parentId = parentId)
 annotations(CODE) = all.annotations

### 2018-01-23 manuscriptFigures/diffExpFigure.Rmd
Added notebook to plot differential expression results

---
title: "Differential expression figures"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_notebook
---
```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
# Load required libraries
library(CovariateAnalysis)
library(data.table)
library(plyr)
library(tidyverse)

library(synapseClient)
library(githubr)
library(knitr)

synapseLogin()

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

### Covariates
```{r technical.cov}
cov = downloadFile('syn10160582')
all.used.ids = c('syn10160582')
```
Total number of samples
```{r total.sample}
cov %>%
  dplyr::select(Study, SampleID, BrainRegion, Diagnosis) %>%
  dplyr::group_by(BrainRegion, Diagnosis) %>%
  dplyr::count() %>%
  tidyr::spread(Diagnosis, n) %>%
  kable()
```
Total number of samples per sex
```{r total.sample.sex}
cov %>%
  dplyr::select(Study, SampleID, BrainRegion, Diagnosis, Gender) %>%
  dplyr::group_by(BrainRegion, Gender) %>%
  dplyr::count() %>%
  tidyr::spread(Gender, n) %>%
  kable()
```
Different types of RNAs
```{r diff.rnas}
tmp = cov %>%
  dplyr::select(Study, PCT_CODING_BASES, PCT_INTERGENIC_BASES, 
                PCT_INTRONIC_BASES, PCT_RIBOSOMAL_BASES) %>%
  tidyr::gather(Feature, Value, -Study) %>%
  dplyr::group_by(Study, Feature) %>%
  dplyr::summarise(Value = mean(Value, na.rm = T)) %>%
  tidyr::spread(Feature, Value) %>%
  dplyr::group_by(Study) %>%
  dplyr::mutate(PCT_UTR_BASES = 1 - PCT_CODING_BASES - PCT_INTERGENIC_BASES - PCT_INTRONIC_BASES - PCT_RIBOSOMAL_BASES) %>%
  tidyr::gather(TypeOfRNA, Value, -Study)

p = ggplot(tmp, aes(x = Study, y= Value, fill = TypeOfRNA)) + geom_bar(stat = 'identity')
p = p + ylab('Fraction of aligned reads') + xlab('Study') + theme_bw()
p
```
### Differential Expression
```{r get.results}
dexp = downloadFile('syn11180450')
all.used.ids = c(all.used.ids, 'syn11180450')

ndexp = dplyr::filter(dexp, Model == 'Diagnosis') %>% 
  dplyr::group_by(Study, Tissue, Comparison, Gender) %>% 
  dplyr::summarise(UP = sum(adj.P.Val <= 0.05 & logFC >= 0), 
                   DOWN = sum(adj.P.Val <= 0.05 & logFC <= 0)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-Comparison, -Gender)
kable(ndexp)
```
pca plot
```{r pca}
expr = downloadFile('syn10160584') %>%
  tidyr::separate(Gene.ID, c('ensembl_gene_id','position'),sep = '\\.') %>%
  dplyr::filter(ensembl_gene_id %in% unique(dexp$ensembl_gene_id[dexp$adj.P.Val <= 0.05 & dexp$Model == 'Diagnosis']))

pc = prcomp(expr[,-(1:2)], scale. = T, center = T)

cov = data.frame(SampleID = colnames(expr)[-(1:2)],
           PC1 = pc$rotation[,1],
           PC2 = pc$rotation[,2]) %>%
  dplyr::left_join(cov)
p = ggplot(cov, aes(x = PC1, y = PC2, color = Study, shape = BrainRegion))
p = p + geom_point() + scale_shape_manual(values = 1:7)
p = p + theme(legend.position = 'top')
p
```

Volcano plots
```{r vp, fig.height=5, fig.width=10}
dexp$Direction = 'None'
dexp$Direction[dexp$logFC >= 0 & dexp$adj.P.Val <= 0.05] = 'Up'
dexp$Direction[dexp$logFC <= 0 & dexp$adj.P.Val <= 0.05] = 'Down'

p = ggplot(dexp %>%
             dplyr::filter(Model == 'Diagnosis'), 
           aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point(alpha = 0.1)
p = p + facet_grid(.~Study+Tissue, scales = 'free') + theme(legend.position = 'top') 
p = p + scale_color_manual(values = viridis::viridis(3))
# dev.off()
# svg(filename = 'VolcanoPlots.svg', height = 6, width = 12)
# print(p)
# dev.off()
p
```
Average expression
```{r aexp, fig.height=5, fig.width=10}
col = viridis::viridis(3)
p = ggplot(dexp%>%
             dplyr::filter(Model == 'Diagnosis'), 
           aes(x = AveExpr, y = logFC, color = Direction)) + geom_point(alpha = 0.1)
p = p + facet_grid(.~Study+Tissue, scales = 'free') + theme(legend.position = 'top') 
p = p + scale_color_manual(values = col)
p
```
Number of differentially expressed genes in each foldchange range (histogram)
```{r ndiff.range, fig.height=5, fig.width=10}
dexp$lfc1 = ifelse(dexp$logFC >= 0, dexp$logFC, NA)
dexp$lfc2 = ifelse(dexp$logFC <= 0, dexp$logFC, NA)
p = ggplot(dexp %>% dplyr::filter(adj.P.Val <= 0.05 & Model == 'Diagnosis'))
p = p + geom_histogram( aes(x = lfc1, y = ..count..), fill = "red")
p = p + geom_histogram( aes(x = lfc2, y = -..count..), fill = "green")
p = p + facet_grid(.~Study+Tissue, scales = 'free') + ylab('Gene count') + xlab('Log Fold Change')
p
```
### Overlap statistics
overlap between different brain regions
```{r ovlp.stats.up}
dexp.enrich.plots = dexp %>%
  dplyr::filter(Model %in% c('Diagnosis'), 
                adj.P.Val <= 0.05) %>%
  plyr::dlply(.(Direction), .fun = function(x){
    gr = x %>%
      plyr::dlply(.(Tissue), .fun = function(x){unique(x$ensembl_gene_id)}) %>%
      plyr::ldply(.,function(x, gs, backgroundGenes){
        plyr::ldply(gs, .fun = function(y,x,backgroundGenes){
          CovariateAnalysis::fisherEnrichment(x,y,backgroundGenes)
        }, x, backgroundGenes) %>%
          dplyr::rename(To = Tissue)
      }, ., background.genes) %>%
      dplyr::rename(From = Tissue) %>%
      dplyr::mutate(FDR = p.adjust(pval, method = 'BH')) %>%
      dplyr::select(From, To, Odds.Ratio, FDR, noverlap) %>%
      igraph::graph_from_data_frame()

    or.mat = igraph::as_adj(gr, type = 'both', attr = 'Odds.Ratio') %>% as.matrix()
    or.mat[is.infinite(or.mat)] = 0
    no.mat = igraph::as_adj(gr, type = 'both', attr = 'noverlap')

    if (unique(x$Direction) == 'Up'){
      col = c('white','red')
    } else {
      col = c('white','darkgreen')
    }
    ComplexHeatmap::Heatmap(as.matrix(or.mat), 
                            col= circlize::colorRamp2(c(0,40),col),
                            cell_fun = function(j, i, x, y, widht, height, fill){
                              no.mat[i,j] %>%
                                format(digit = 1) %>%
                                as.character() %>%
                                grid.text(x, y)
                            },
                            name = 'Odds Ratio',
                            column_title = paste(unique(x$Direction), 
                                                 'regulated'))
  })
purrr::walk(dexp.enrich.plots, function(x){ print(x) })
```

### Sex Specific Differential Expression
```{r ndiff.sex}
ndexp = dplyr::filter(dexp, Model == 'Diagnosis.Gender') %>% 
  dplyr::group_by(Study, Tissue, Comparison, Gender) %>% 
  dplyr::summarise(UP = sum(adj.P.Val <= 0.05 & logFC >= 0), 
                   DOWN = sum(adj.P.Val <= 0.05 & logFC <= 0)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-Comparison)
ndexp.female = dplyr::filter(ndexp, Gender == 'FEMALE') %>%
  dplyr::rename(FEMALE.UP = UP, FEMALE.DOWN = DOWN) %>%
  dplyr::select(-Gender)
ndexp.male = dplyr::filter(ndexp, Gender == 'MALE') %>%
  dplyr::rename(MALE.UP = UP, MALE.DOWN = DOWN) %>%
  dplyr::select(-Gender)
kable(dplyr::full_join(ndexp.female, ndexp.male, by = c('Tissue', 'Study')))
```
Volcano plots
```{r sex.vp, fig.height=10, fig.width=10}
p = ggplot(dexp %>%
             dplyr::filter(Model == 'Diagnosis.Gender'), 
           aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point(alpha = 0.1)
p = p + facet_grid(Gender+.~Study+Tissue, scales = 'free') + theme(legend.position = 'top') 
p = p + scale_color_manual(values = viridis::viridis(3))
# dev.off()
# svg(filename = 'VolcanoPlots.svg', height = 6, width = 12)
# print(p)
# dev.off()
p
```
Average expression
```{r sex.aexp, fig.height=10, fig.width=10}
col = viridis::viridis(3)
p = ggplot(dexp%>%
             dplyr::filter(Model == 'Diagnosis.Gender'), 
           aes(x = AveExpr, y = logFC, color = Direction)) + geom_point(alpha = 0.1)
p = p + facet_grid(Gender+.~Study+Tissue, scales = 'free') + theme(legend.position = 'top') 
p = p + scale_color_manual(values = col)
p
```
Number of differentially expressed genes in each foldchange range (histogram)
```{r sex.ndiff.range, fig.height=10, fig.width=10}
dexp$lfc1 = ifelse(dexp$logFC >= 0, dexp$logFC, NA)
dexp$lfc2 = ifelse(dexp$logFC <= 0, dexp$logFC, NA)
p = ggplot(dexp %>% dplyr::filter(adj.P.Val <= 0.05 & Model == 'Diagnosis.Gender'))
p = p + geom_histogram( aes(x = lfc1, y = ..count..), fill = "red")
p = p + geom_histogram( aes(x = lfc2, y = -..count..), fill = "green")
p = p + facet_grid(Gender+.~Study+Tissue, scales = 'free') + ylab('Gene count') + xlab('Log Fold Change')
p
```
### Overlap statistics
overlap between different brain regions in female
```{r ovlp.stats.up.female}
dexp.enrich.plots = dexp %>%
  dplyr::filter(Model %in% c('Diagnosis.Gender'), 
                adj.P.Val <= 0.05) %>%
  plyr::dlply(.(Gender, Direction), .fun = function(x){
    gr = x %>%
      plyr::dlply(.(Tissue), .fun = function(x){unique(x$ensembl_gene_id)}) %>%
      plyr::ldply(.,function(x, gs, backgroundGenes){
        plyr::ldply(gs, .fun = function(y,x,backgroundGenes){
          CovariateAnalysis::fisherEnrichment(x,y,backgroundGenes)
        }, x, backgroundGenes) %>%
          dplyr::rename(To = Tissue)
      }, ., background.genes) %>%
      dplyr::rename(From = Tissue) %>%
      dplyr::mutate(FDR = p.adjust(pval, method = 'BH')) %>%
      dplyr::select(From, To, Odds.Ratio, FDR, noverlap) %>%
      igraph::graph_from_data_frame()

    or.mat = igraph::as_adj(gr, type = 'both', attr = 'Odds.Ratio') %>% as.matrix()
    or.mat[is.infinite(or.mat)] = 0
    no.mat = igraph::as_adj(gr, type = 'both', attr = 'noverlap')

    if (unique(x$Direction) == 'Up'){
      col = c('white','red')
    } else {
      col = c('white','darkgreen')
    }
    ComplexHeatmap::Heatmap(as.matrix(or.mat), 
                            col= circlize::colorRamp2(c(0,40),col),
                            cell_fun = function(j, i, x, y, widht, height, fill){
                              no.mat[i,j] %>%
                                format(digit = 1) %>%
                                as.character() %>%
                                grid.text(x, y)
                            },
                            name = 'Odds Ratio',
                            column_title = paste(unique(x$Direction), 
                                                 'regulated in', 
                                                 unique(x$Gender)))
  })
purrr::walk(dexp.enrich.plots, function(x){ print(x) })
```
### Store in synapse
```{r syn.store}
# Get github commit
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Network", ref="branch", refName='AMPAD_manuscript')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('manuscriptFigures/',thisFileName))

obj = rmarkdown::render('diffExpFigure.Rmd', output_format = 'html_document')
obj = File(obj, name = 'Differential expression summary', parentId = 'syn11703045')
obj = synStore(obj, executed = thisFile)
# wikiPg = WikiPage(owner = obj, title = 'Figure1')
wikiPg = synGetWiki(obj)
wikiPg@properties$markdown = paste0("${preview?entityId=",obj$properties$id,"}")
wikiPg = synStore(wikiPg)
```

### 2018-01-26 manuscriptFigures/diffExpFigure.Rmd
Minor changes

@@
 library(tidyverse)
-
+library(ComplexHeatmap)
+library(circlize)
+library(gridExtra)
 library(synapseClient)
@@
 ### Store in synapse
-```{r syn.store}
+```{r syn.store, eval = FALSE, include = FALSE}
 # Get github commit
-thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Network", ref="branch", refName='AMPAD_manuscript')
-thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('manuscriptFigures/',thisFileName))
+thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_manuscript')
+thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('manuscriptFigures/diffExpFigure.Rmd'))
 
@@
 obj = synStore(obj, executed = thisFile)
-# wikiPg = WikiPage(owner = obj, title = 'Figure1')
+wikiPg = WikiPage(owner = obj, title = 'Differential expression summary')
 wikiPg = synGetWiki(obj)

### 2018-01-27 code/Rmd/DEXP_ENRICH.Rmd
Added markdown to perform enrichment analysis for differentially expressed genes

---
title: "Enrichment analysis for differential expression results"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./DEXP_ENRICH.Rmd",
                                 parentId = "syn8672415",
                                 entityName = 'Enrichment analysis of all differentialy expressed genes')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(plyr)
library(tidyverse)
library(biomaRt)

library(synapseClient)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

synapseLogin()

library(doParallel)
library(foreach)

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn8672415';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of reprocessed rnaseq from all 7 brain regions';

thisFileName <- 'DEXP_ENRICH.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```

## Data download
### Get summary differential expression results from synapse
```{r get.diffExp}
dexp = downloadFile('syn11180450')
all.used.ids = c(all.used.ids, 'syn11180450')

dexp$Direction = 'NONE'
dexp$Direction[dexp$adj.P.Val <= 0.05 & dexp$logFC >= log2(1.2)] = 'UP'
dexp$Direction[dexp$adj.P.Val <= 0.05 & dexp$logFC <= -log2(1.2)] = 'DOWN'
dexp$Tissue[dexp$Tissue == 'CER'] = 'CBE'

# Get hgnc symbol from biomart
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
                host = "www.ensembl.org", 
                dataset = "hsapiens_gene_ensembl")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", 
                       values = unique(dexp$ensembl_gene_id),
                       mart = mart)
dexp = dexp %>%
  dplyr::select(-hgnc_symbol) %>%
  dplyr::left_join(Ensemble2HGNC)

bck.genes = unique(dexp$hgnc_symbol)
```

### Get all gene sets
```{r get.gs}
# Get enrichr gene sets from synapse
load(synGet('syn4867851')@filePath)
all.used.ids = c('syn4867851')
GeneSets = GeneSets[c('BioCarta', 'Chromosome_Location', 'Cross_Species_Phenotype', 
                      'Disease_Signatures_from_GEO_down', 'Disease_Signatures_from_GEO_up',
                      'ENCODE_TF_ChIP-seq', 'GO_Biological_Process', 
                      'KEGG_2015', 'Mouse_Gene_Atlas', 'OMIM_Disease', 'OMIM_Expanded', 'Panther', 
                      'Reactome', 'TargetScan_microRNA', 'TF-LOF_Expression_from_GEO', 'TRANSFAC_and_JASPAR_PWMs',
                      'WikiPathways', 'Single_Gene_Perturbations_from_GEO_up',
                      'Single_Gene_Perturbations_from_GEO_down', 'Aging_Perturbations_from_GEO_down',
                      'Aging_Perturbations_from_GEO_up', 'dbGaP', 'Disease_Perturbations_from_GEO_down',
                      'Disease_Perturbations_from_GEO_up', 'DrugMatrix', 
                      'Drug_Perturbations_from_GEO_down', 'Drug_Perturbations_from_GEO_up', 
                      'GTEx_Tissue_Sample_Gene_Expression_Profiles_down',
                      'GTEx_Tissue_Sample_Gene_Expression_Profiles_up', 'Jensen_DISEASES', 
                      'RNA-Seq_Disease_Gene_and_Drug_Signatures_from_GEO')]
gs = list()
gs$HypothesisFree = do.call(c, GeneSets) %>%
  lapply(function(y){
    str_split(y, '\\,') %>% 
      sapply(function(x){ x[[1]] })
  })

tmp = synTableQuery('select * from syn11707052')@values
all.used.ids = c(all.used.ids, 'syn11707052')

gs$CellType = tmp %>%
  dplyr::filter(GeneSet %in% grep('Zhang:', unique(tmp$GeneSet), value = T)) %>%
  plyr::dlply(.(GeneSet), .fun = function(x){
    unique(x$external_gene)
  })

gs$HypothesisDriven = tmp %>%
  dplyr::filter(!(GeneSet %in% grep('Zhang:', unique(tmp$GeneSet), value = T))) %>%
  plyr::dlply(.(GeneSet), .fun = function(x){
    unique(x$external_gene)
  })

gs = purrr::map(gs, function(geneSet, bckGenes){
  geneSet = purrr::map(geneSet, function(x, bckGenes){
    intersect(x, bckGenes) %>% unique()
  }, bckGenes)

  n = sapply(geneSet, length)
  geneSet = geneSet[n >= 10 & n <= 3000]
}, bck.genes)
```

### Perform enrichment analysis
```{r enrich.de}
enrich.results = dexp %>%
  dplyr::filter(Direction != 'NONE') %>%
  plyr::ddply(.(Model, Tissue, Comparison, Gender, Direction), .fun = function(x, gs, bckGenes){
    gsToTest = setdiff(unique(x$hgnc_symbol),'')
    enrich = plyr::ldply(gs, .fun = function(y, gsToTest, bckGenes){
      plyr::ldply(y, .fun = function(gsToTest1, gsToTest, bckGenes){
        CovariateAnalysis::fisherEnrichment(gsToTest1, gsToTest, bckGenes)
      }, gsToTest, bckGenes, .id = 'SetName', .parallel = T, .paropts = list(.packages = c('CovariateAnalysis')))
    }, gsToTest, bckGenes, .id = 'Category')
  }, gs, bck.genes)

fwrite(enrich.results, file = 'enrichmentResults.tsv', sep = '\t')
obj = File('enrichmentResults.tsv', name = 'Enrichment Results (for all differential expression)', parentId = parentId)
obj = synStore(obj, executed = thisFile, used = all.used.ids, 
               activityName = activityName, activityDescription = activityDescription)
```

### 2018-02-18 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Minor changes

--- 2017-07-28 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
@@
 # Get differnetial expression
-DE = lapply(1:dim(contrast)[2], function(i, FIT){
-  limma::topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
+DE = lapply(1:6, function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
     rownameToFirstColumn('ensembl_gene_id') 
@@
 DE = DE %>% 
-  data.table::rbindlist(idcol = 'Comparison') %>%
+  rbindlist(idcol = 'Comparison') %>%
   dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
@@
   tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
-  tidyr::separate(ref.state, into = c('Dx1','Tissue1'), sep = '_') %>%
-  tidyr::separate(to.state, into = c('Dx2','Tissue2'), sep = '_') %>%
+  tidyr::separate(ref.state, into = c('Dx1','Tissue'), sep = '_') %>%
+  tidyr::separate(to.state, into = c('Dx2','Tissue'), sep = '_') %>%
   tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_') %>%
@@
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue1) %>%
-  group_by(Comparison, Tissue1) %>%
+  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
+  group_by(Comparison, Tissue) %>%
   dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
@@
   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue1) %>%
-  group_by(Comparison, Tissue1) %>%
+  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
+  group_by(Comparison, Tissue) %>%
   dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
@@
 p = p + scale_color_manual(values = c('green','grey','red'))
-p = p + facet_grid(Tissue1+.~Comparison, scales = 'fixed')
+p = p + facet_grid(Tissue+.~Comparison, scales = 'fixed')
 p
@@
 diffexp.genes = DE %>%
-  dlply(.(Comparison, Tissue1), .fun = function(x){
+  dlply(.(Comparison, Tissue), .fun = function(x){
     tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
@@
 
-### Source Code
-[R markdown](`r thisFile`)
+### Source Rmd
+[Source markdown](`r thisFile`)

### 2018-02-18 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor changes

--- 2017-07-28 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)"
-author: "Thanneer Perumal"
-date: "`r date()`"
-output: html_document
+title: Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)
+author: Thanneer Perumal
+editor_options:
+  chunk_output_type: console
 ---
+Date of analysis update: "`r date()`"
 ```{r knit2synapse, eval=FALSE}
@@
 registerDoParallel(cl)
+
+# Source modified utility 
+source('../R/parallelDuplicateCorrelation.R')
 
@@
 ```
+
 ### Filter genes
-Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
-```{r cpmnormalisation}
+* Remove genes that have less than 1 cpm counts in at least 50% of samples per Tissue x Diagnosis
+* Remove genes with missing gene length and percentage GC content
+```{r filter.genes}
 genesToAnalyze = dlply(METADATA, .(Dx.Tissue), .fun = function(mtd, count){
@@
 
+### Library Normalisation
+Initial library normalisation usign cqn
+```{r cqn}
+# Compute offset for gene length and gc content
+CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
+                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
+                          lengthMethod = "smooth", 
+                          verbose = FALSE)
+CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+```
+
 ### Outlier analysis
+#### Sample outliers
 Detect outlier samples based on their expression (logCPM) pattern
@@
 # Find principal components of expression to plot
-PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)
+PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)
 
@@
 # Plot abberent distribution of logcpm counts
-tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
+tmp1 = CQN.GENE_EXPRESSION$E %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
@@
 PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
+CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E[,indToRetain]
 COVARIATES = COVARIATES[indToRetain,]
@@
 `r kable(tmp)`
-
-### Winsorise counts (Gene outliers in samples)
+#### Gene outliers
+Assign NA values to genes that are above and below 3 std deviation of its distribution
 ```{r winsorise.data}
-LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)
-
 # Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
-LOG.CPM = apply(LOG.CPM, 1, function(x){
+LOG.CPM = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
   mn = mean(x, na.rm = T)
@@
 
-  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
-  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
-
-  x[x < (mn-3*std.dev)] = x[ind.low]
-  x[x > (mn+3*std.dev)] = x[ind.high]
+  x[x < (mn-3*std.dev)] = NA
+  x[x > (mn+3*std.dev)] = NA
   return(x)
 }) %>% t
-
-# Back calculate counts for further analysis
+CQN.GENE_EXPRESSION$E = LOG.CPM
+CQN.GENE_EXPRESSION$E.no.na = CQN.GENE_EXPRESSION$E
+CQN.GENE_EXPRESSION$E.no.na[is.na(CQN.GENE_EXPRESSION$E.no.na)] = 0
 LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
@@
 
-### Library Normalisation
-Initial library normalisation usign cqn
-```{r cqn}
-# Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
-                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
-                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
-CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-
-# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
-CQN.GENE_EXPRESSION$E = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
-  mn = mean(x, na.rm = T)
-  std.dev = sd(x, na.rm = T)
-
-  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
-  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
-
-  x[x < (mn-3*std.dev)] = x[ind.low]
-  x[x > (mn+3*std.dev)] = x[ind.high]
-  return(x)
-}) %>% t
-
-# Back calculate counts for further analysis
-LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
-NEW.COUNTS = (2^CQN.GENE_EXPRESSION$E) * t(replicate(dim(CQN.GENE_EXPRESSION$E)[1], LIB.SIZE))/1e6
-```
-
-### Co-expression distribution
-Coexpression of genes 
-```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(CQN.GENE_EXPRESSION$E))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
-
 ### Sample clustering
+PCA based clustering of samples
 ```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
 # Find principal components of expression to plot
-PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)
+PC <- prcomp(CQN.GENE_EXPRESSION$E.no.na, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E)))
+tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E.no.na)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 ```
-```{r temp, include = F}
-dev.off()
-gc()
-```
-
 ### Distribution of samples (log cpm)
@@
 ```
+### Co-expression distribution
+Coexpression of genes 
+```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
+cr = cor(t(CQN.GENE_EXPRESSION$E.no.na))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
+```
 
@@
 # Find correlation between PC's of gene expression with covariates
-preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E, 
+preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E.no.na, 
                                                  COVARIATES,
@@
 ```
-### Normalisation
+
+### Normalisation (custom design)
 Based on different model selection methodology (as seen here syn8468116, syn9772932, and syn9774491), following variables are used as covariates: Institution, Gender, RIN, and IntronicRate
 
-NOTE: Dx.Tissue is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue), and Individual\_ID is chosen as the random effect
-
-```{r custom.normalisation, results='asis'}
+NOTE: Dx.Tissue is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue), and Individual_ID is chosen as the random effect
+```{r custom.design}
 # Primary variable of interest
@@
 # Estimate voom weights
-VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design = DESIGN)
+cnts = NEW.COUNTS
+cnts[is.na(cnts)] = 0
+VOOM.GENE_EXPRESSION = voom(cnts, design = DESIGN)
 VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
-correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
+correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, 
+                                            block = COVARIATES$Individual_ID,
+                                            method = 'lmer')
 
 # Re-calculate voom weights with correlation of random effects
-VOOM.GENE_EXPRESSION = voom(NEW.COUNTS,   
+VOOM.GENE_EXPRESSION = voom(cnts,   
                             design = DESIGN, plot=F,
@@
 
-### Residual calculation
-Calculate weighted residuals and add back "Dx.Tissue" to the residuals
-```{r varsToAddBack}
-# Add variable of interest back to the residuals
-varsToAddIn = grep("Dx.Tissue", colnames(DESIGN), value = T)
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
-  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DESIGN[,varsToAddIn])
-```
-Coexpression of genes 
-```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.GENE_EXPRESSION))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
-
 ### Clustering residual data
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
+expr = RESIDUAL.GENE_EXPRESSION;expr[is.na(expr)] = 0
+PC <- prcomp(expr, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
+tree = hclust(as.dist(t(expr)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
                            groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue','Institution')]))
-```
-
-```{r temp1, include=F}
-dev.off()
-gc()
 ```
@@
 # Estimate voom weights
-VOOM.GENE_EXPRESSION1 = voom(NEW.COUNTS, design = DESIGN1)
+VOOM.GENE_EXPRESSION1 = voom(cnts, design = DESIGN1)
 VOOM.GENE_EXPRESSION1$E = CQN.GENE_EXPRESSION$E
-correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION1, block = COVARIATES$Individual_ID)
+correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION1, 
+                                            block = COVARIATES$Individual_ID,
+                                            method = 'lmer')
 
 # Re-calculate voom weights with correlation of random effects
-VOOM.GENE_EXPRESSION1 = voom(NEW.COUNTS,   
+VOOM.GENE_EXPRESSION1 = voom(cnts,   
                              design = DESIGN1, plot=F,
@@
 # Estimate voom weights
-VOOM.GENE_EXPRESSION2 = voom(NEW.COUNTS, design = DESIGN2)
+VOOM.GENE_EXPRESSION2 = voom(cnts, design = DESIGN2)
 VOOM.GENE_EXPRESSION2$E = CQN.GENE_EXPRESSION$E
-correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION2, block = COVARIATES$Individual_ID)
+correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION2, 
+                                            block = COVARIATES$Individual_ID,
+                                            method = 'lmer')
 
 # Re-calculate voom weights with correlation of random effects
-VOOM.GENE_EXPRESSION2 = voom(NEW.COUNTS,   
+VOOM.GENE_EXPRESSION2 = voom(cnts,   
                              design = DESIGN2, plot=F,

### 2018-02-19 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Minor modifications to ACC and DLPFC processing markdown with SVA

@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)"
-author: "Thanneer Perumal"
-date: "`r date()`"
-output: html_document
+title: Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN + SVA normlaisation)
+author: Thanneer Perumal
+editor_options:
+  chunk_output_type: console
 ---
@@
 library(cqn)
-library(sva)
 
@@
 registerDoParallel(cl)
+
+# Source modified utility 
+source('../R/parallelDuplicateCorrelation.R')
+source('../R/irwsva.build.R')
+source('../R/f.pvalue.R')
 
@@
 
-# Source modified functions from sva package
-source('../R/irwsva.build.R')
-source('../R/f.pvalue.R')
-
 # Compute actual variance of all principal components
-tmp = svd(RESIDUAL.GENE_EXPRESSION)
+expr = RESIDUAL.GENE_EXPRESSION; expr[is.na(expr)] = 0
+tmp = svd(expr)
 actual.var = tmp$d^2/sum(tmp$d^2)
@@
 permuted.var = foreach(i = 1:20, .combine = rbind, .packages = c('dplyr')) %dopar% {
-  tmp.residual = t(apply(RESIDUAL.GENE_EXPRESSION, 1, sample, replace = FALSE))
+  tmp.residual = t(apply(expr, 1, sample, replace = FALSE))
   tmp = svd(tmp.residual)
@@
 # Estimate surrogate variables
-SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E, MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
+SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E.no.na, MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
 colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
@@
 # Estimate voom weights for dispersion control
-VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
+VOOM.SVA.WEIGHTS = voom(cnts, 
                         design=cbind(MODEL1, SURR.VAR), plot=F)
@@
 # Calculate correlation values of random effects
-correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$Individual_ID)
+correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, 
+                                           block = COVARIATES$Individual_ID,
+                                           method = 'lmer')
 
 # Re-estimate voom weights
-VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
+VOOM.SVA.WEIGHTS = voom(cnts, 
                         design = cbind(MODEL1, SURR.VAR), plot=F,
@@
 ```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.SVA.GENE_EXPRESSION))
+expr = RESIDUAL.SVA.GENE_EXPRESSION;
+expr[is.na(expr)] = 0
+cr = cor(t(expr))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.SVA.GENE_EXPRESSION, scale.=T, center = T)
+PC <- prcomp(expr, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(RESIDUAL.SVA.GENE_EXPRESSION)))
+tree = hclust(as.dist(t(expr)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);

### 2018-02-19 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor modifications in HBCC Rmd to change winsorising from 3SD to NA values

@@
 ---
-title: "Covariate analysis of HBCC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN normlaisation)"
-author: "Thanneer Perumal"
-date: "`r date()`"
-output: html_document
+title: Covariate analysis of HBCC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)
+author: Thanneer Perumal
+editor_options:
+  chunk_output_type: console
 ---
+Date of analysis update: "`r date()`"
+
 ```{r knit2synapse, eval=FALSE}
@@
 
+### Library Normalisation
+Initial library normalisation usign cqn
+```{r cqn}
+# Compute offset for gene length and gc content
+CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
+                          lengths = as.numeric(GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length']),
+                          lengthMethod = "smooth", 
+                          verbose = FALSE)
+CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+```
+
 ### Outlier analysis
+#### Sample Outliers
 Detect outlier samples based on their expression (logCPM) pattern
@@
 
-### Winsorise counts (Gene outliers in samples)
+#### Gene outliers
+Assign NA values to genes that are above and below 3 std deviation of its distribution
 ```{r winsorise.data}
-LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)
-
 # Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
-LOG.CPM = apply(LOG.CPM, 1, function(x){
+LOG.CPM = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
   mn = mean(x, na.rm = T)
@@
 
-  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
-  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
-
-  x[x < (mn-3*std.dev)] = x[ind.low]
-  x[x > (mn+3*std.dev)] = x[ind.high]
+  x[x < (mn-3*std.dev)] = NA
+  x[x > (mn+3*std.dev)] = NA
   return(x)
 }) %>% t
-
-# Back calculate counts for further analysis
+CQN.GENE_EXPRESSION$E = LOG.CPM
+CQN.GENE_EXPRESSION$E.no.na = CQN.GENE_EXPRESSION$E
+CQN.GENE_EXPRESSION$E.no.na[is.na(CQN.GENE_EXPRESSION$E.no.na)] = 0
 LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
 NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
-```
-
-### Library Normalisation
-Initial library normalisation usign cqn
-```{r cqn}
-# Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
-                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
-                          lengths = as.numeric(GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length']),
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
-CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-
-# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
-CQN.GENE_EXPRESSION$E = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
-  mn = mean(x, na.rm = T)
-  std.dev = sd(x, na.rm = T)
-
-  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
-  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
-
-  x[x < (mn-3*std.dev)] = x[ind.low]
-  x[x > (mn+3*std.dev)] = x[ind.high]
-  return(x)
-}) %>% t
-
-# Back calculate counts for further analysis
-LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
-NEW.COUNTS = (2^CQN.GENE_EXPRESSION$E) * t(replicate(dim(CQN.GENE_EXPRESSION$E)[1], LIB.SIZE))/1e6
-```
-
-### Co-expression distribution
-Coexpression of genes 
-```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-cr = cor(t(tmp))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
 ```
@@
 # Find principal components of expression to plot
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-PC <- prcomp(tmp, scale.=T, center = T)
+PC <- prcomp(CQN.GENE_EXPRESSION$E.no.na, scale.=T, center = T)
 
@@
 
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-tree = hclust(as.dist(t(tmp)))
+tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E.no.na)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 ```
-```{r temp, include = F}
-dev.off()
-gc()
-```
 
@@
 # Plot abberent distribution of logcpm counts
-tmp1 = (CQN.GENE_EXPRESSION$E) %>%
+tmp1 = (CQN.GENE_EXPRESSION$E.no.na) %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
@@
 
+### Co-expression distribution
+Coexpression of genes 
+```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
+cr = cor(t(CQN.GENE_EXPRESSION$E.no.na))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
+```
+
 ### Significant Covariates
@@
 # Find correlation between PC's of gene expression with covariates
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-preAdjustedSigCovars = runPCAandPlotCorrelations(tmp, 
+preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E.no.na, 
                                                  COVARIATES,
@@
 # Estimate voom weights
-tmp = NEW.COUNTS
-tmp[is.na(tmp)] = 0
-VOOM.GENE_EXPRESSION = voom(tmp, design=DM1$design, plot=F)
+cnts = NEW.COUNTS; cnts[is.na(cnts)] = 0
+VOOM.GENE_EXPRESSION = voom(cnts, design=DM1$design, plot=F)
 VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
@@
 RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
-
+```
+
+### Sanity check
+```{r residualSigCovars.manual, fig.width=12, fig.height=8}
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-tmp = RESIDUAL.GENE_EXPRESSION
-tmp[is.na(tmp)] = 0
-residualSigCovars = runPCAandPlotCorrelations(tmp, COVARIATES,
+expr = RESIDUAL.GENE_EXPRESSION
+expr[is.na(expr)] = 0
+residualSigCovars = runPCAandPlotCorrelations(expr, 
+                                              COVARIATES,
                                               'adjusted design(voom-normalized)',
                                               isKeyPlot=TRUE)
-```
-
-### Sanity check
-```{r residualSigCovars.manual, fig.width=12, fig.height=8}
+
 residualSigCovars[["PC_res"]][[2]]$plotData
@@
 ```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-tmp = RESIDUAL.GENE_EXPRESSION
-tmp[is.na(tmp)] = 0
-cr = cor(t(tmp))
+cr = cor(t(expr))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(tmp, scale.=T, center = T)
+PC <- prcomp(expr, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(tmp)))
+tree = hclust(as.dist(t(expr)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
                            groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue','Gender')]))
-```
-
-```{r temp1, include=F}
-dev.off()
-gc()
 ```
@@
 # Estimate voom weights
-VOOM.GENE_EXPRESSION1 = limma::voom(NEW.COUNTS, design = DESIGN1)
+VOOM.GENE_EXPRESSION1 = limma::voom(cnts, design = DESIGN1)
 VOOM.GENE_EXPRESSION1$E = CQN.GENE_EXPRESSION$E
@@
 # Estimate voom weights
-VOOM.GENE_EXPRESSION2 = voom(NEW.COUNTS, design = DESIGN2)
+VOOM.GENE_EXPRESSION2 = voom(cnts, design = DESIGN2)
 VOOM.GENE_EXPRESSION2$E = CQN.GENE_EXPRESSION$E

### 2018-02-19 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor modifications in HBCC SVA Rmd to change winsorising from 3SD to NA values

@@
 knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
-                                 parentId ="syn7501880",
+                                 parentId ="syn9872607",
                                  entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')

### 2018-02-19 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Minor modifications in HBCC SVA Rmd to change winsorising from 3SD to NA values

--- 2017-08-08 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
@@
 ---
-title: "Covariate analysis of HBCC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN and SVA normlaisation)"
-author: "Thanneer Perumal"
-date: "`r date()`"
-output: html_document
+title: Covariate analysis of HBCC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN + SVA normlaisation)
+author: Thanneer Perumal
+editor_options:
+  chunk_output_type: console
 ---
+Date of analysis update: "`r date()`"
+
 ```{r knit2synapse, eval=FALSE}
@@
 registerDoParallel(cl)
+
+# Source modified functions from sva package
+source('../R/irwsva.build.R')
+source('../R/f.pvalue.R')
 
@@
 
+### Library Normalisation
+Initial library normalisation usign cqn
+```{r cqn}
+# Compute offset for gene length and gc content
+CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
+                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
+                          lengths = as.numeric(GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length']),
+                          lengthMethod = "smooth", 
+                          verbose = FALSE)
+CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
+```
+
 ### Outlier analysis
+#### Sample Outliers
 Detect outlier samples based on their expression (logCPM) pattern
-```{r detect.outliers, fig.height=5, fig.width=5, results='asis', cache = FALSE}
+```{r detect.outliers, fig.height=5, fig.width=6, results='asis', cache = FALSE}
 indToRemove = c("CMC_HBCC_RNA_PFC_3402")
@@
 
-### Library Normalisation
-Initial library normalisation usign cqn
-```{r cqn}
-# Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
-                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
-                          lengths = as.numeric(GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length']),
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
-CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-
-# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
-CQN.GENE_EXPRESSION$E = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
-  mn = mean(x, na.rm = T)
-  std.dev = sd(x, na.rm = T)
-
-  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
-  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
-
-  x[x < (mn-3*std.dev)] = x[ind.low]
-  x[x > (mn+3*std.dev)] = x[ind.high]
-  return(x)
-}) %>% t
-
-# Back calculate counts for further analysis
-LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
-NEW.COUNTS = (2^CQN.GENE_EXPRESSION$E) * t(replicate(dim(CQN.GENE_EXPRESSION$E)[1], LIB.SIZE))/1e6
-```
-
-### Co-expression distribution
-Coexpression of genes 
-```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-cr = cor(t(tmp))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
-
 ### Sample clustering
-```{r decompse.normalise.data, fig.height=6, fig.width=8, results='asis', cache=FALSE}
+```{r decompse.normalise.data, fig.height=5, fig.width=5, results='asis', cache=FALSE}
 # Find principal components of expression to plot
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-PC <- prcomp(tmp, scale.=T, center = T)
+PC <- prcomp(CQN.GENE_EXPRESSION$E.no.na, scale.=T, center = T)
 
@@
   tidyr::gather(data, value, -component)
-p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.2)), aes(x = component, y = value, color = data)) + geom_point()
+p = ggplot(var.comp %>% 
+             dplyr::filter(component <= round(NUM.SV*1.2)), 
+           aes(x = component, y = value, color = data)) + geom_point()
 p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
@@
 # Estimate surrogate variables
-SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E, MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
+SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E.no.na, MODEL1, MODEL0, 
+                        n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
 colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
@@
 VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)
+
+# Residuals after normalisation
+RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
+                                                  CQN.GENE_EXPRESSION$E)
 ```
@@
 ```{r varsToAddBack}
-# Residuals after normalisation
-RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
-                                                  CQN.GENE_EXPRESSION$E)
-
 # Add variable of interest back to the residuals
@@
 ```
-Coexpression of genes 
-```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.SVA.GENE_EXPRESSION))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
 Clustering residual data
@@
 # Find principal components of expression to plot
-PC <- prcomp(RESIDUAL.SVA.GENE_EXPRESSION, scale.=T, center = T)
+expr = RESIDUAL.SVA.GENE_EXPRESSION; expr[is.na(expr)] = 0
+PC <- prcomp(expr, scale.=T, center = T)
 
@@
                            groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue'),drop=FALSE]))
-```
-
-```{r temp1, include=F}
-dev.off()
-gc()
 ```

### 2018-07-25 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Modified model selection markdown for HBCC DLPFC analysis to account for metadata (individual id) changes

@@
                                  parentId ="syn9846623",
-                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)')
+                                 folderName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)')
 ```
@@
 library(data.table)
-library(tidyr)
 library(plyr)
-library(dplyr)
-library(stringr)
-
-library(ggplot2)
+library(tidyverse)
+
 library(limma)
@@
 ```{r download.data, cache=TRUE}
+downloadFile <- function(id, sep){
+  tmp = data.table::fread(synGet(id)@filePath,  data.table = F, header = T, sep = sep)
+}
+
 # Download reprocessed counts (DLPFC)
-COUNT_DLPFC_ID = 'syn8465067';
-ALL_USED_IDs = COUNT_DLPFC_ID
-COUNT = downloadFile(COUNT_DLPFC_ID) %>%
+COUNT = downloadFile('syn8465067', sep = '\t') %>%
   as.data.frame()
+ALL_USED_IDs = 'syn8465067'
 
@@
 
-# Get ancestry vector calculated using gemtools
-# ANCESTRY_ID = 'syn2511399'
-# ALL_USED_IDs[length(ALL_USED_IDs)+1] = ANCESTRY_ID
-# ANCESTRY = read.table(synGet(ANCESTRY_ID)@filePath, fill = T, header = T, sep = '\t') %>%
-#   plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'GenotypingSampleID'))
+# Get clinical metadata
+METADATA_CLINICAL = downloadFile('syn2279441', sep = ',') %>%
+  dplyr::rename(Individual_ID = `Individual ID`)
+ALL_USED_IDs = c(ALL_USED_IDs, 'syn2279441')
+
+# Get RNA alignment report from synapse
+METADATA_ALG_REP_DLPFC = downloadFile('syn2279446', sep = ',') %>%
+  dplyr::select(`Exclude?`, `Sample RNA ID`, `Ribozero Batch`, `Library Batch`, `Flowcell Batch`) %>%
+  dplyr::rename(Exclude = `Exclude?`,
+                SampleID = `Sample RNA ID`,
+                LibraryBatch = `Library Batch`, 
+                RibozeroBatch = `Ribozero Batch`, 
+                FlowcellBatch = `Flowcell Batch`) %>%
+  distinct()
+ALL_USED_IDs = c(ALL_USED_IDs, 'syn2279446')
 
 # Get RNA qc metrics from synapse
-METADATA_ALG_REP_DLPFC_ID = 'syn2279446'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ALG_REP_DLPFC_ID
-METADATA_ALG_REP_DLPFC = read.table(synGet(METADATA_ALG_REP_DLPFC_ID)@filePath, 
-                                    fill = T, header = T, sep = ',') %>%
-  dplyr::select(Sample.RNA.ID, Ribozero.Batch, Library.Batch, Flowcell.Batch) %>%
-  dplyr::rename(SampleID = Sample.RNA.ID,
-                LibraryBatch = Library.Batch, 
-                RibozeroBatch = Ribozero.Batch, 
-                FlowcellBatch = Flowcell.Batch) %>%
-  unique()
-
-# Get RNA qc metrics from synapse
-METADATA_QC_DLPFC_ID = 'syn8539839'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
-METADATA_QC = read.table(synGet(METADATA_QC_DLPFC_ID)@filePath, fill = T, header = T, sep = ',', row.names = 1) %>%
-  dplyr::select(Sample, Total_Purity_Filtered_Reads_Sequenced, X5prime_Norm, Mean_Per_Base_Cov.,  
+METADATA_RNASEQ_QC = downloadFile('syn8539839', sep = ',') %>%
+  dplyr::select(V1, Sample, Total_Purity_Filtered_Reads_Sequenced, X5prime_Norm, Mean_Per_Base_Cov.,  
                 Mean_CV, Fragment_Length_StdDev,  Mapped_Pairs, End_1_Mismatch_Rate, End_2_Mismatch_Rate,
@@
                 PercentMapped = UNIQUE_READS._Uniquely_mapped_reads_percent)
-
-# Get merged metadata
-METADATA_ID = 'syn2299154'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
-METADATA = read.csv(synGet(METADATA_ID)@filePath, fill = T, header = T, check.names = F)
-
-## Metadata with specific information for RNASeq processing
-metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", "Dx",
-
-                          "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
-                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?")
-
-# Filter metadata 
-ind = colnames(METADATA) %in% metadataVarOfInterest
-METADATA = METADATA[,ind]
-colnames(METADATA) = gsub('[^[:alnum:]]','_',colnames(METADATA))
-
-METADATA = METADATA %>%
-  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
-                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, 
-                DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_) %>%
-  dplyr::rename(PMI = PMI__in_hours_, 
-                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
-                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
-                RIN = DLPFC_RNA_isolation__RIN, 
-                ReportExclude = DLPFC_RNA_report__Exclude_) %>%
-  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
-  dplyr::mutate(Tissue = 'DLPFC') %>%
-  list(.,METADATA_QC, METADATA_ALG_REP_DLPFC) %>%
-  join_all %>% droplevels() %>%
-  unique %>% as.data.frame()
-METADATA$SampleID = as.character(METADATA$SampleID)
-
-# Fix missing values
-levels(METADATA$LibraryBatch)[1] = 'NotAvailable'
-levels(METADATA$FlowcellBatch)[1] = 'NotAvailable'
-levels(METADATA$RibozeroBatch)[1] = 'NotAvailable'
+ALL_USED_IDs = c(ALL_USED_IDs, 'syn8539839')
+
+# Get RNASeq isolation report
+METADATA_RNA_ISOLATION = downloadFile('syn2279445', sep = ',') %>%
+  dplyr::select(`Sample RNA ID`, RIN) %>%
+  dplyr::rename(SampleID = `Sample RNA ID`)
+ALL_USED_IDs = c(ALL_USED_IDs, 'syn2279445')
+
+# Get individual id mapping
+METADATA_IID = downloadFile('syn12979139', sep = '\t') %>%
+  dplyr::select(Data_ID, IID_SAMPLE, label) %>%
+  dplyr::rename(Individual_ID = IID_SAMPLE,
+                SampleID = Data_ID) %>%
+  distinct()
+ALL_USED_IDs = c(ALL_USED_IDs, 'syn12979139')
+
+# Merge metadata
+METADATA = dplyr::inner_join(METADATA_ALG_REP_DLPFC, METADATA_RNASEQ_QC) %>%
+  dplyr::inner_join(METADATA_IID) %>%
+  dplyr::inner_join(METADATA_CLINICAL) %>%
+  dplyr::inner_join(METADATA_RNA_ISOLATION) %>%
+  dplyr::select(SampleID, Individual_ID, label, 
+                LibraryBatch, RibozeroBatch, FlowcellBatch,
+                Dx, Gender, `PMI (in hours)`, `Age of Death`, RIN,
+                TotalReads, X5prime_Norm, Mean_Per_Base_Cov.,
+                Mean_CV, Fragment_Length_StdDev, Mapped_Pairs, End_1_Mismatch_Rate, End_2_Mismatch_Rate, 
+                Mapped, IntragenicRate, ExonicRate, IntergenicRate, IntronicRate, rRNARate, GenesDetected,
+                TranscriptsDetected, ExpProfEfficiency, PercentMapped) %>%
+  dplyr::rename(PMI = `PMI (in hours)`,
+                Age_of_Death = `Age of Death`) %>%
+  dplyr::mutate_at(.vars = c('LibraryBatch', 'RibozeroBatch', 'FlowcellBatch', 'Dx', 'Gender'), 
+                   .funs = function(x){
+                     x = forcats::fct_explicit_na(x, na_level = 'NotAvailable')
+                     levels(x)[levels(x) == ''] = 'NotAvailable'
+                     return(x)
+                   }) %>%
+  dplyr::mutate(Tissue = 'DLPFC')
 ```
@@
 writeLines('Following samples are marked exclude')
-writeLines(paste(METADATA$SampleID[METADATA$IsolationExclude == 1 | METADATA$ReportExclude == 1], collapse = ', '))
-METADATA <- METADATA  %>% dplyr::filter(IsolationExclude == 0, ReportExclude == 0)
+writeLines(paste(METADATA$SampleID[METADATA$label == 'exclude'], collapse = ', '))
+METADATA <- METADATA  %>% dplyr::filter(label == 'accept')
 
@@
 
-```{r gene.param}
+```{r gene.param, include = FALSE}
 ## Get GC content from biomart
@@
 
-### Winsorise counts (Gene outliers in samples)
-```{r winsorise.data}
-LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)
-
-# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
-LOG.CPM = apply(LOG.CPM, 1, function(x){
-  mn = mean(x, na.rm = T)
-  std.dev = sd(x, na.rm = T)
-
-  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
-  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
-
-  x[x < (mn-3*std.dev)] = x[ind.low]
-  x[x > (mn+3*std.dev)] = x[ind.high]
-  return(x)
-}) %>% t
-
-# Back calculate counts for further analysis
-LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
-NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
-```
-
-### Library Normalisation
-Initial library normalisation usign cqn
+### Library Normalisation and Winsorise data
+Initial library normalisation usign cqn and setting gene outliers to NA
 ```{r cqn}
 # Compute offset for gene length and gc content
-CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
+CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                           x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
@@
 
-  x[x < (mn-3*std.dev)] = x[ind.low]
-  x[x > (mn+3*std.dev)] = x[ind.high]
+  x[x < (mn-3*std.dev)] = NA
+  x[x > (mn+3*std.dev)] = NA
   return(x)
@@
 # Estimate voom weights
-VOOM.GENE_EXPRESSION = limma::voom(NEW.COUNTS, design = DESIGN)
+tmp = NEW.COUNTS
+tmp[is.na(tmp)] = 0
+VOOM.GENE_EXPRESSION = limma::voom(tmp, design = DESIGN)
 VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
@@
 # Estimate voom weights
-VOOM.GENE_EXPRESSION1 = limma::voom(NEW.COUNTS, design = DESIGN1)
+tmp = NEW.COUNTS
+tmp[is.na(tmp)] = 0
+VOOM.GENE_EXPRESSION1 = limma::voom(tmp, design = DESIGN1)
 VOOM.GENE_EXPRESSION1$E = CQN.GENE_EXPRESSION$E
@@
 # Estimate voom weights
-VOOM.GENE_EXPRESSION2 = voom(NEW.COUNTS, design = DESIGN2)
+tmp = NEW.COUNTS
+tmp[is.na(tmp)] = 0
+VOOM.GENE_EXPRESSION2 = voom(tmp, design = DESIGN2)
 VOOM.GENE_EXPRESSION2$E = CQN.GENE_EXPRESSION$E
@@
 # Download all related genesets from synapse
-geneset.files = synQuery('select * from file where parentId == "syn3240583"')
+synapseClient::synapseLogin()
+geneset.files = synapseClient::synQuery('select * from file where parentId == "syn3240583"')
 ALL_USED_IDs = c(ALL_USED_IDs, geneset.files$file.id)
 genesets.to.test = sapply(geneset.files$file.name, function(dataSource, geneset.files){
-  tmp = downloadFile(geneset.files$file.id[geneset.files$file.name == dataSource]) %>%
+  tmp = downloadFile(geneset.files$file.id[geneset.files$file.name == dataSource], sep = '\t') %>%
     dlply(.(Name), .fun = function(x){
@@
 # Code
-CODE <- Folder(name = "HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)", parentId = parentId)
-annotations(CODE) = all.annotations
-CODE <- synStore(CODE)
+CODE <- synapseClient::Folder("HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)", 
+                              parentId = parentId)
+synapseClient::synSetAnnotations(CODE, annotations = all.annotations)
+CODE <- synapseClient::synStore(CODE)
 
@@
 write.table(COVARIATES, file = 'CMC_ACC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
-COV_OBJ = File('CMC_ACC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
-annotations(COV_OBJ) = annotations(CODE)
-annotations(COV_OBJ)$dataSubType = 'covariates'
-COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+COV_OBJ = synapseClient::File('CMC_ACC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+synapseClient::annotations(COV_OBJ) = synapseClient::annotations(CODE)
+synapseClient::annotations(COV_OBJ)$dataSubType = 'covariates'
+COV_OBJ = synapseClient::synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                                   executed = thisFile, activityDescription = activityDescription)
@@
   write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
-annotations(COUNT_OBJ) = annotations(CODE)
-annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
-COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+COUNT_OBJ = synapseClient::File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
+synapseClient::annotations(COUNT_OBJ) = synapseClient::annotations(CODE)
+synapseClient::annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
+COUNT_OBJ = synapseClient::synStore(COUNT_OBJ, activity = synapseClient::synGetActivity(COV_OBJ$properties$id))
 
@@
   write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
-annotations(LCOUNT_OBJ) = annotations(CODE)
-annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
-LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+LCOUNT_OBJ = synapseClient::File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
+synapseClient::annotations(LCOUNT_OBJ) = synapseClient::annotations(CODE)
+synapseClient::annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
+LCOUNT_OBJ = synapseClient::synStore(LCOUNT_OBJ, activity = synapseClient::synGetActivity(COV_OBJ$properties$id))
 
@@
   write.table(file = 'CMC_ACC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('CMC_ACC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
-annotations(OFFSET_OBJ) = annotations(CODE)
-annotations(OFFSET_OBJ)$dataSubType = 'offset'
-OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+OFFSET_OBJ = synapseClient::File('CMC_ACC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
+synapseClient::annotations(OFFSET_OBJ) = synapseClient::annotations(CODE)
+synapseClient::annotations(OFFSET_OBJ)$dataSubType = 'offset'
+OFFSET_OBJ = synapseClient::synStore(OFFSET_OBJ, activity = synapseClient::synGetActivity(COV_OBJ$properties$id))
 
@@
   write.table(file = 'CMC_ACC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('CMC_ACC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
-annotations(DM_OBJ) = annotations(CODE)
-annotations(DM_OBJ)$dataSubType = 'designMatrix'
-DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+DM_OBJ = synapseClient::File('CMC_ACC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+synapseClient::annotations(DM_OBJ) = synapseClient::annotations(CODE)
+synapseClient::annotations(DM_OBJ)$dataSubType = 'designMatrix'
+DM_OBJ = synapseClient::synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 
@@
   write.table(file = 'CMC_ACC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_ACC_DLPFC_netResidualExpression.tsv', 
+nEXP_OBJ = synapseClient::File('CMC_ACC_DLPFC_netResidualExpression.tsv', 
                                name = 'Normalised, covariates removed residual expression (for network analysis)', 
                                parentId = CODE$properties$id)
-annotations(nEXP_OBJ) = annotations(CODE)
-annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
-nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+synapseClient::annotations(nEXP_OBJ) = synapseClient::annotations(CODE)
+synapseClient::annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
+nEXP_OBJ = synapseClient::synStore(nEXP_OBJ, activity = synapseClient::synGetActivity(COV_OBJ$properties$id))
 
@@
 write.table(DE, file = 'CMC_ACC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_ACC_DLPFC_DiffExpression.tsv', 
+DEXP_OBJ = synapseClient::File('CMC_ACC_DLPFC_DiffExpression.tsv', 
                                name = 'Differential Expression Results (Dx.Tissue)', 
                                parentId = CODE$properties$id)
-annotations(DEXP_OBJ) = annotations(CODE)
-annotations(DEXP_OBJ)$dataSubType = 'diffExp'
-DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+synapseClient::annotations(DEXP_OBJ) = synapseClient::annotations(CODE)
+synapseClient::annotations(DEXP_OBJ)$dataSubType = 'diffExp'
+DEXP_OBJ = synapseClient::synStore(DEXP_OBJ, activity = synapseClient::synGetActivity(COV_OBJ$properties$id))
 
@@
 write.table(enrichment.results, file = 'CMC_ACC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
-ENRICH_OBJ = File('CMC_ACC_DLPFC_EnrichResults.tsv', 
+ENRICH_OBJ = synapseClient::File('CMC_ACC_DLPFC_EnrichResults.tsv', 
                                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
                                  parentId = CODE$properties$id)
-annotations(ENRICH_OBJ) = annotations(CODE)
-annotations(ENRICH_OBJ)$dataSubType = 'enrichResults'
-ENRICH_OBJ = synStore(ENRICH_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+synapseClient::annotations(ENRICH_OBJ) = synapseClient::annotations(CODE)
+synapseClient::annotations(ENRICH_OBJ)$dataSubType = 'enrichResults'
+ENRICH_OBJ = synapseClient::synStore(ENRICH_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
 # stopCluster(cl)

### 2018-07-25 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Minor bug fix

@@
                                  parentId ="syn9846623",
-                                 folderName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)')
-```
-
-```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
+                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)')
+```
+
+```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
 ## It is assumed your working directory is where this file
@@
 
-```{r synapse.parameters, include=FALSE, cache=TRUE}
+```{r synapse.parameters, include=FALSE, cache=FALSE}
 parentId = 'syn9846623';
@@
 # Code
-CODE <- synapseClient::Folder("HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)", 
+CODE <- synapseClient::Folder(name = "HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)", 
                               parentId = parentId)
-synapseClient::synSetAnnotations(CODE, annotations = all.annotations)
+synapseClient::annotations(CODE) = all.annotations
 CODE <- synapseClient::synStore(CODE)

### 2018-07-25 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Modifications in HBCC DLPFC markdown to account for individual ID changes in metadata

--- 2017-08-08 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
@@
 knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
-                                 parentId ="syn7501880",
+                                 parentId = "syn7501880", #syn9872607",
                                  entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = 'syn7501880';
+parentId = 'syn7501880'; #'syn9872607';
 activityName = 'Covariate adjustments';
@@
 Correlation/association between covariates at an FDR <= 0.1
-```{r covariates.correlation, fig.width=20, fig.height=10}
+```{r covariates.correlation, fig.width=15, fig.height=15}
 COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
@@
 Detect outlier samples based on their expression (logCPM) pattern
-```{r detect.outliers, fig.height=5, fig.width=6, results='asis', cache = FALSE}
+```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
 indToRemove = c("CMC_HBCC_RNA_PFC_3402")
@@
 ### Sample clustering
-```{r decompse.normalise.data, fig.height=5, fig.width=5, results='asis', cache=FALSE}
+```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
 # Find principal components of expression to plot
@@
 genesets.to.test = sapply(geneset.files$file.name, function(dataSource, geneset.files){
-  tmp = downloadFile(geneset.files$file.id[geneset.files$file.name == dataSource]) %>%
+  tmp = downloadFile(geneset.files$file.id[geneset.files$file.name == dataSource], sep = '\t') %>%
     dlply(.(Name), .fun = function(x){

### 2018-07-25 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Modifications in HBCC DLPFC SVA markdown to account for individual ID changes in metadata

@@
 knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
-                                 parentId = "syn7501880", #syn9872607",
+                                 parentId = "syn9872607", #"syn7501880", 
                                  entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = 'syn7501880'; #'syn9872607';
+parentId = 'syn9872607'; #'syn7501880'; #
 activityName = 'Covariate adjustments';

### 2018-07-25 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Modifications in HBCC DLPFC SVA markdown to account for individual ID changes in metadata

--- 2017-08-08 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
@@
 knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd",
-                                 parentId ="syn7501880",
+                                 parentId = "syn7501880", #"syn9872607",
                                  entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - SVA')
@@
 
+# Source modified functions from sva package
+source('../R/irwsva.build.R')
+source('../R/f.pvalue.R')
+
 options(xtable.type="html")
@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = 'syn7501880';
+parentId = "syn7501880", #'syn9872607';
 activityName = 'Covariate adjustments';
@@
 ```
-### Data download
-Obtain count matrix and metadata from synapse.
-```{r download.data, cache=TRUE}
-# Download reprocessed counts (DLPFC)
-COUNT_DLPFC_ID = 'syn8465067';
-ALL_USED_IDs = COUNT_DLPFC_ID
-COUNT = downloadFile(COUNT_DLPFC_ID) %>%
-  as.data.frame()
-
-GENE.LEN = dplyr::select(COUNT, GeneID, Length) %>%
-  dplyr::rename(ensembl_gene_id = GeneID) %>%
-  unique()
-rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id
-
-rownames(COUNT) = COUNT$GeneID
-ind = setdiff(colnames(COUNT), c('GeneID', 'Chr', 'Start', 'End', 'Strand', 'Length'))
-COUNT = data.matrix(COUNT[,ind])
-
-# Get ancestry vector calculated using gemtools
-# ANCESTRY_ID = 'syn2511399'
-# ALL_USED_IDs[length(ALL_USED_IDs)+1] = ANCESTRY_ID
-# ANCESTRY = read.table(synGet(ANCESTRY_ID)@filePath, fill = T, header = T, sep = '\t') %>%
-#   plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'GenotypingSampleID'))
-
-# Get RNA qc metrics from synapse
-METADATA_ALG_REP_DLPFC_ID = 'syn2279446'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ALG_REP_DLPFC_ID
-METADATA_ALG_REP_DLPFC = read.table(synGet(METADATA_ALG_REP_DLPFC_ID)@filePath, 
-                                    fill = T, header = T, sep = ',') %>%
-  dplyr::select(Sample.RNA.ID, Ribozero.Batch, Library.Batch, Flowcell.Batch) %>%
-  dplyr::rename(SampleID = Sample.RNA.ID,
-                LibraryBatch = Library.Batch, 
-                RibozeroBatch = Ribozero.Batch, 
-                FlowcellBatch = Flowcell.Batch) %>%
-  unique()
-
-# Get RNA qc metrics from synapse
-METADATA_QC_DLPFC_ID = 'syn8539839'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
-METADATA_QC = read.table(synGet(METADATA_QC_DLPFC_ID)@filePath, fill = T, header = T, sep = ',', row.names = 1) %>%
-  dplyr::select(Sample, Total_Purity_Filtered_Reads_Sequenced, X5prime_Norm, Mean_Per_Base_Cov.,  
-                Mean_CV, Fragment_Length_StdDev,  Mapped_Pairs, End_1_Mismatch_Rate, End_2_Mismatch_Rate,
-                End_1_Mapping_Rate, End_2_Mapping_Rate, Mapped, Intragenic_Rate, Exonic_Rate, Intergenic_Rate,
-                Intronic_Rate, rRNA_rate, Genes_Detected, Transcripts_Detected, Expression_Profiling_Efficiency,
-                UNIQUE_READS._Uniquely_mapped_reads_percent) %>%
-  dplyr::rename(SampleID = Sample, 
-                IntragenicRate = Intragenic_Rate, 
-                ExonicRate = Exonic_Rate, 
-                IntronicRate = Intronic_Rate, 
-                IntergenicRate = Intergenic_Rate, 
-                GenesDetected = Genes_Detected, 
-                TranscriptsDetected = Transcripts_Detected, 
-                ExpProfEfficiency = Expression_Profiling_Efficiency, 
-                rRNARate = rRNA_rate, 
-                TotalReads = Total_Purity_Filtered_Reads_Sequenced, 
-                PercentMapped = UNIQUE_READS._Uniquely_mapped_reads_percent)
-
-# Get merged metadata
-METADATA_ID = 'syn2299154'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
-METADATA = read.csv(synGet(METADATA_ID)@filePath, fill = T, header = T, check.names = F)
-
-## Metadata with specific information for RNASeq processing
-metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", "Dx",
-
-                          "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
-                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?")
-
-# Filter metadata 
-ind = colnames(METADATA) %in% metadataVarOfInterest
-METADATA = METADATA[,ind]
-colnames(METADATA) = gsub('[^[:alnum:]]','_',colnames(METADATA))
-
-METADATA = METADATA %>%
-  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
-                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, 
-                DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_) %>%
-  dplyr::rename(PMI = PMI__in_hours_, 
-                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
-                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
-                RIN = DLPFC_RNA_isolation__RIN, 
-                ReportExclude = DLPFC_RNA_report__Exclude_) %>%
-  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
-  dplyr::mutate(Tissue = 'DLPFC') %>%
-  list(.,METADATA_QC, METADATA_ALG_REP_DLPFC) %>%
-  join_all %>% droplevels() %>%
-  unique %>% as.data.frame()
-METADATA$SampleID = as.character(METADATA$SampleID)
-
-# Fix missing values
-levels(METADATA$LibraryBatch)[1] = 'NotAvailable'
-levels(METADATA$FlowcellBatch)[1] = 'NotAvailable'
-levels(METADATA$RibozeroBatch)[1] = 'NotAvailable'
-```
-
 ### Data preprocessing
@@
 Detect outlier samples based on their expression (logCPM) pattern
-```{r detect.outliers, fig.height=5, fig.width=5, results='asis', cache = FALSE}
+```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
 indToRemove = c("CMC_HBCC_RNA_PFC_3402")
@@
 ### Sample clustering
-```{r decompse.normalise.data, fig.height=6, fig.width=8, results='asis', cache=FALSE}
+```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
 # Find principal components of expression to plot
@@
 # Estimate surrogate variables
-SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E, MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
+expr = CQN.GENE_EXPRESSION$E
+expr[is.na(expr)] = 0
+SURR.VAR = irwsva.build(expr, MODEL1, MODEL0, 
+                        n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
 colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])

### 2018-09-26 code/Rmd/CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd
Added Model selection markdown for RSEM aligned PFC from CMC

---
title: "Covariate analysis of CMC RNASeq data reprocessed with RSEM-feature counts workflow with GRCh38 (DLPFC with CQN normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapser)
library(knit2synapse)

synLogin()

knit2synapse::knitToFolderEntity(file = "./CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd",
                                 parentId ="syn9846623",
                                 entityName = 'DLPFC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(limma)
library(psych)
library(edgeR)
library(cqn)
library(lme4)
library(biomaRt)

library(synapser)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

synapseLogin()

library(doParallel)
library(foreach)

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn9846623';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of RSEM aligned effective counts with GRCh38 with CQN normalisation (DLPFC)';

thisFileName <- 'CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain effective count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
downloadFile <- function(id){
  fread(synGet(id)$path, data.table = F)
}

# Download reprocessed counts (DLPFC)
COUNT_ID = 'syn16783526'
ALL_USED_IDs = COUNT_ID
COUNT = downloadFile(COUNT_ID) %>% data.frame()
rownames(COUNT) = COUNT$gene_id
COUNT$gene_id = NULL
COUNT$transcript_id.s. = NULL

# Download gene lengths (DLPFC)
GENE.LEN = downloadFile('syn16783530') %>%
  tidyr::gather(sampleID, Length, -gene_id, -`transcript_id(s)`) %>%
  group_by(gene_id) %>%
  summarise(Length = median(Length, na.rm = T)) %>%
  ungroup() %>% data.frame()
ALL_USED_IDs = c(ALL_USED_IDs, 'syn16783530')

# Get ancestry vector calculated using gemtools
ANCESTRY_ID = 'syn2511399'
ALL_USED_IDs = c(ALL_USED_IDs, ANCESTRY_ID)
ANCESTRY = downloadFile(ANCESTRY_ID) %>%
  plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'SNP_report:Genotyping_Sample_ID'))

# Get genotype ids from synapse
GENOTYPE_ID = 'syn16816490'
ALL_USED_IDs = c(ALL_USED_IDs, GENOTYPE_ID)
GENOTYPE = downloadFile(GENOTYPE_ID) %>%
  dplyr::select(Individual_ID, `SNP_report:Genotyping_Sample_ID`) %>%
  dplyr::inner_join(ANCESTRY)

# Get RNASeq alignment qc metrics from synapse
METADATA_ALGMETRIC_DLPFC_ID = 'syn16783867'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ALGMETRIC_DLPFC_ID
METADATA_ALGMETRIC_DLPFC = downloadFile(METADATA_ALGMETRIC_DLPFC_ID) %>%
  dplyr::rename(Sample_RNA_ID = ID)

# Get RNASeq QCmetadata
METADATA_QC_DLPFC_ID = 'syn16816488'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
METADATA_QC_DLPFC = downloadFile(METADATA_QC_DLPFC_ID) %>%
  dplyr::select(Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
                Sample_RNA_ID, one_of('rnaSeq_isolation:Exclude?', 'rnaSeq_isolation:RIN',
                                      'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?'))
```

### Data preprocessing
```{r preprocess.data}
# Merge metadata
METADATA = METADATA_QC_DLPFC %>%
  dplyr::left_join(GENOTYPE) %>%
  dplyr::left_join(METADATA_ALGMETRIC_DLPFC) %>%
  dplyr::rename(PMI = `PMI_(in_hours)`,
                IsolationExclude = `rnaSeq_isolation:Exclude?`, 
                RIN = `rnaSeq_isolation:RIN`, 
                ReportExclude = `rnaSeq_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                MappedReads = Mapped,
                IntragenicRate = `Intragenic Rate`, IntronicRate = `Intronic Rate`, 
                IntergenicRate = `Intergenic Rate`, GenesDetected = `Genes Detected`,
                ExpProfEfficiency = `Expression Profiling Efficiency`, rRNARate = `rRNA rate`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, PMI, Dx, 
                RIN, IsolationExclude, ReportExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, RibozeroBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate) %>%
  dplyr::filter(SampleID %in% colnames(COUNT), !is.na(SampleID)) %>%
  dplyr::mutate(Dx = forcats::fct_recode(Dx,  Other = "AFF", Other = "BP", Control = "Control", SCZ = "SCZ"))

ind = METADATA$SampleID[which(METADATA$IsolationExclude == 1 | METADATA$ReportExclude == 1)]
writeLines(paste('Following',length(ind),'samples are marked exclude'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$Ethnicity) | is.na(METADATA$Institution) | is.na(METADATA$Dx)]
writeLines(paste('Following', length(ind), 'counts are missing any metadata'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$PMI)]
writeLines(paste('Following', length(ind), 'counts are missing PMI'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$Gender)]
writeLines(paste('Following', length(ind), 'counts are missing gender'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$Age_of_Death)]
writeLines(paste('Following', length(ind), 'counts are missing age of death'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$EV.1)]
writeLines(paste('Following', length(ind), 'counts are missing ancestry information'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))
```

```{r preprocess.data1, results='asis'}
# Match covariates to expression data
indToRetain = intersect(METADATA$SampleID, colnames(COUNT))
removedIDs = setdiff(colnames(COUNT), METADATA$SampleID)

COUNT = COUNT[,indToRetain]

rownames(METADATA) = METADATA$SampleID
METADATA = METADATA[indToRetain,]

METADATA %>% 
  group_by(Dx) %>% 
  summarise(count = n()) %>% 
  kable()
```
Following sample are removed `r paste(removedIDs, collapse = ',')`

```{r gene.param, include = F}
## Get GC content from biomart
backgroundGenes = data.frame(gene_id = rownames(COUNT)) %>%
  dplyr::mutate(id = gene_id) %>%
  tidyr::separate(id, c('ensembl_gene_id','position'), sep = '\\.')

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "ensembl.org", dataset = "hsapiens_gene_ensembl")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::left_join(backgroundGenes) %>%
  dplyr::select(gene_id, percentage_gene_gc_content) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
```

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx", "FlowcellBatch")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
                    "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){
  x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
rownames(COVARIATES) <- METADATA$SampleID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], function(x){
  x = as.numeric(as.character(gsub('[\\,\\%]','',x)))
})

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=8}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
tmp = COVARIATES.CORRELATION$ESTIMATE
tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
draw(h, heatmap_legend_side = 'left')
```

## Explore metatdata
```{r data.explore, fig.width = 12, fig.height = 12}
my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = Dx, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = Dx, y = Age_of_Death)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = Dx, y = PMI)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme

# Intronic Rate
p[[4]] = ggplot(COVARIATES, aes(x = Dx, y = IntronicRate)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme

# IntergenicRate
p[[5]] = ggplot(COVARIATES, aes(x = Dx, y = IntergenicRate)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme

# Transcripts Detected
p[[6]] = ggplot(COVARIATES, aes(x = Dx, y = IntragenicRate)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Intragenic Rate') + my.theme

# Mapped Reads
p[[7]] = ggplot(COVARIATES, aes(x = Dx, y = MappedReads)) + geom_boxplot()
p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme

# PercentAligned
p[[8]] = ggplot(COVARIATES, aes(x = Dx, y = TotalReads)) + geom_boxplot()
p[[8]] = p[[8]] + ggtitle('Total Reads') + my.theme

# rRNARate
p[[9]] = ggplot(COVARIATES, aes(x = Dx, y = rRNARate)) + geom_boxplot()
p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme

multiplot(plotlist = p, cols = 3)

# Institution
# vcd::mosaic(~ Institution + Dx, data = COVARIATES)

# Gender
# vcd::mosaic(~ Gender + Dx, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = dlply(METADATA, .(Dx), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC.CONT$gene_id[!is.na(GENE.GC.CONT$percentage_gene_gc_content)]) %>%
  intersect(GENE.LEN$gene_id[!is.na(GENE.LEN$Length)])

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
pct.pc = PROCESSED_COUNTS$filteredExprMatrix$genes %>%
  dplyr::rename(gene_id = genes) %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  group_by(gene_biotype) %>%
  summarise(fraction  = n()) %>%
  filter(fraction > 100) %>%
  dplyr::mutate(fraction = fraction/length(PROCESSED_COUNTS$filteredExprMatrix$genes[,1]))
```
`r dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]` genes are used for the analysis.

Following is the distribution of biotypes
`r kable(pct.pc)`

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250',
                'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220',
                'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
                'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
                'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_ACC_310",
                "MSSM_RNA_ACC_51",  "MSSM_RNA_ACC_254", "MSSM_RNA_ACC_28", "MSSM_RNA_ACC_293", "MSSM_RNA_ACC_233",
                "MSSM_RNA_ACC_167", "MSSM_RNA_ACC_353", "MSSM_RNA_ACC_355", "MSSM_RNA_ACC_211", "PENN_RNA_ACC_122",
                "PENN_RNA_ACC_049", "PENN_RNA_ACC_052", "PENN_RNA_ACC_068", "PENN_RNA_ACC_070", "PITT_RNA_ACC_10020",
                "PITT_RNA_ACC_1240", "PITT_RNA_ACC_1284", "PITT_RNA_ACC_700", "PITT_RNA_ACC_857", "PITT_RNA_ACC_BP_1589",
                "MSSM_RNA_PFC_230", "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
                "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")

# Find principal components of expression to plot
PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  dplyr::mutate(label = SampleID)
plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
p <- p + theme_bw() %+replace% theme(legend.position="right")
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p

# Plot abberent distribution of logcpm counts
tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID'))

p = ggplot(tmp1 %>%
             dplyr::filter(SampleID %in% indToRemove),
           aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme_bw() %+replace% theme(legend.position = 'none') + facet_grid(Dx~., scale = 'free')
p

indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
COVARIATES = COVARIATES[indToRetain,]

tmp = COVARIATES %>%
  group_by(Dx) %>%
  summarise(count = n())
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals

Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`

Distribution of samples are:
`r kable(tmp)`

### Winsorise counts (Gene outliers in samples)
```{r winsorise.data}
LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
LOG.CPM = apply(LOG.CPM, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))

  x[x < (mn-3*std.dev)] = NA
  x[x > (mn+3*std.dev)] = NA
  return(x)
}) %>% t

# Back calculate counts for further analysis
NEW.COUNTS = PROCESSED_COUNTS$filteredExprMatrix$counts
NEW.COUNTS[is.na(LOG.CPM)] = 0
```

### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Compute offset for gene length and gc content
GENE.LEN = tibble::column_to_rownames(GENE.LEN, var = 'gene_id') %>% as.data.frame()
CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gene_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
NEW.COUNTS[log.mat] = NA
CQN.GENE_EXPRESSION$E[log.mat] = NA
```

### Co-expression distribution
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
cr = cor(t(tmp))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Sample clustering
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
PC <- prcomp(tmp, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
p <- p + theme_bw() %+replace% theme(legend.position="top")
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
tree = hclust(as.dist(t(tmp)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```
```{r temp, include = F}
dev.off()
gc()
```

### Distribution of samples (log cpm)
```{r lcpm.dist, cache=FALSE}
# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$E) %>%
  rownameToFirstColumn('gene_id') %>%
  tidyr::gather(SampleID, logCPM, -gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID'))

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme_bw() %+replace% theme(legend.position = 'NONE') + facet_grid(.~Dx, scale = 'free')
p
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
preAdjustedSigCovars = runPCAandPlotCorrelations(tmp, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Model Identification
```{r fxn}
# Function to fit a mixed effect model and return BIC
fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){

  # Get design matrix
  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
  designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]

  # Estimate voom weights
  cnt[is.na(cnt)] = 0
  voomWeights = limma::voom(cnt, design = designMat, plot=F)$weights

  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
                              .combine = rbind,
                              .packages = c('lme4', 'dplyr'),
                              .export = c('voomWeights')) %dopar% {

                                # Perform weighted mixed linear models for every gene
                                expr = data.frame(Gx = lcpm[i,])
                                data = cbind(expr,
                                             designMat[rownames(expr),],
                                             Individual_ID = groupingVar)
                                formula = paste(setdiff(colnames(data), c('Gx', 'Individual_ID')), collapse = '+') %>%
                                  paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')

                                weights = voomWeights[i,]
                                BIC = NA
                                tryCatch({
                                  mdl = lmer(formula = formula, 
                                             data = data, 
                                             weights = weights)
                                  BIC = BIC(mdl)
                                }, error = function(e){
                                  BIC = NA
                                })
                              return(data.frame(ensembl_gene_id = rownames(lcpm)[i],
                                                BIC = BIC))
                            }
  return(modelBIC)
  }

# Function to select covariates in a step-wise regression fashion
selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){

  varsSelected = c()
  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
  while(length(covarsToSelect) != 0){
    allModelBIC = foreach(i = 1:length(covarsToSelect),
                          .combine = c,
                          .packages = pkgs,
                          .export = c('fitMixedEffectModel')) %dopar% {

                            # Fit a mixed effect linear model and get all BIC
                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
                                                           covars[colnames(cnt),'Individual_ID'], 
                                                           lcpm, 
                                                           cnt, 
                                                           covars) %>%
                            plyr::rename(c('BIC' = covarsToSelect[i]))

                            return(list(modelBIC))
                          }
    allModelBIC = plyr::join_all(allModelBIC)

    # Find number of genes that are changed based on BIC
    nChange = allModelBIC %>%
      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
      dplyr::left_join(baseModelBIC) %>%
      dplyr::mutate(diff = baseModel - Value) %>%
      dplyr::group_by(ModelName) %>%
      dplyr::summarise(better = sum(diff >= 5), 
                       worse = sum(diff <= -5)) %>%
      dplyr::mutate(nchange = better - worse)

    print(nChange)

    # Select variable
    tmp = nChange %>%
      dplyr::filter(nchange > round(dim(cnt)[1]*0.01)) %>%
      dplyr::top_n(1, nchange)

    if (dim(tmp)[1] == 0){
      covarsToSelect = c()
    } else {
      baseModelBIC = allModelBIC %>%
        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
      colnames(baseModelBIC)[2] = 'baseModel'
      varsSelected = c(varsSelected, tmp$ModelName)
      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
      print(paste('Variable chosen to be added:',tmp$ModelName))
    }
  }

  return(varsSelected)
}
```
#### Phase I (clinical, ancestry and sample specific technical variables)
```{r phase1}
# Fit a base model with the following covariates
adjust.covars = c('Dx')
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase1.covars = c('Institution', 'Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`

#### Phase II (batch effects)
```{r phase2}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase2.covars = c('LibraryBatch', 'FlowcellBatch')
vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`

#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase3.covars = c("MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate", "GenesDetected",
                  "ExpProfEfficiency", "rRNARate", "TotalReads")
vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`

### Normalisation
1. Dx is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
2. Individual_ID is chosen as random effect
```{r iterative.normalisation, results='asis'}
writeLines(paste('Using following covariates in the model:',
                 paste(adjust.covars, collapse=', '),
                 'as fixed effects and Individual_ID is chosen as random effect'))

# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

# Estimate voom weights
tmp = NEW.COUNTS
tmp[is.na(tmp)] = 0
VOOM.GENE_EXPRESSION = voom(tmp, design=DM1$design, plot=F)

# Estimate consensus correlation coefficient
correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.GENE_EXPRESSION = voom(tmp, 
                            design=DM1$design, plot=F,
                            block = COVARIATES$Individual_ID, 
                            correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
tmp = RESIDUAL.GENE_EXPRESSION
tmp[is.na(tmp)] = 0
residualSigCovars = runPCAandPlotCorrelations(tmp, COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```

### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

### Coexpression of residual expression 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
tmp = RESIDUAL.GENE_EXPRESSION
tmp[is.na(tmp)] = 0
cr = cor(t(tmp))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(tmp, scale.=T, center = T)

# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
p <- p + theme_bw() %+replace% theme(legend.position="top") + facet_grid(.~Dx, scales = 'free_y')
p
```

```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx','Institution')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(tmp)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Dx','Institution')]))
```

```{r temp1, include=F}
dev.off()
gc()
```

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=15}
# Fit contrast
contrast = makeContrasts(contrasts=c("DxOther-DxControl",
                                     "DxSCZ-DxControl",
                                     "DxSCZ-DxOther"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:3, function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('gene_id') %>%
    left_join(backgroundGenes)
}, FIT.CONTR) 
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx','',Comparison),
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(Ensemble2HGNC) %>%
  left_join(GENE.LEN %>%
              rownameToFirstColumn('gene_id'))
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))

kable(full_join(tmp,tmp1))

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
p = p + facet_grid(.~Comparison, scales = 'fixed')
p
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=10, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE, aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme_bw() %+replace% theme(legend.position = 'top')

pl[[2]] = ggplot(DE, aes(x = percentage_gene_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gene_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme_bw() %+replace% theme(legend.position = 'top')

pl[[3]] = ggplot(DE, aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme_bw() %+replace% theme(legend.position = 'top')

multiplot(plotlist = pl, cols = 3)
```

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16}
diffexp.genes = DE %>%
  dlply(.(Comparison), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
backgroundGenes = unique(DE$hgnc_symbol)

# Download all related genesets from synapse
geneset.files = synQuery('select * from file where parentId == "syn3240583"')$results
genesets.to.test = purrr::map(geneset.files, .f = function(fileList){
  tmp = downloadFile(fileList$file.id) %>%
    dlply(.(Name), .fun = function(x){
      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
    })
  }) %>%
  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
  do.call(c,.)

# Perform enrichment analysis
enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
    }, x, backgroundGenes, .parallel = TRUE) %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
}, genesets.to.test, backgroundGenes) %>%
  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
  tidyr::separate(Comparison, c('Name', 'Region'), sep = '\\.') %>%
  dplyr::rename(Comparison = Name, Category = .id)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Set annotations
all.annotations = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay	 = 'RNAseq',

  tissueTypeAbrv	= 'DLPFC, ACC', 
  study = 'CMC',

  organism = 'HomoSapiens',
  consortium	= 'CMC',

  normalizationStatus	= TRUE,

  normalizationType	= 'CQNandMS',
  rnaquantification = 'RSEM',
  genomeAssemblyID = 'GRCh38'
)

# Code
CODE <- Folder(name = "DLPFC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'CMC_ACC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = synapser::File('CMC_ACC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'covariates'
synSetAnnotations(COV_OBJ, annotations = all.annotations)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                     executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'filteredCounts'
synSetAnnotations(COUNT_OBJ, annotations = all.annotations)

# Store logCPM
CQN.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'filteredLCPM'
synSetAnnotations(LCOUNT_OBJ, annotations = all.annotations)

# Store cqn offsets
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('CMC_ACC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
OFFSET_OBJ = synStore(OFFSET_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'offset'
synSetAnnotations(OFFSET_OBJ, annotations = all.annotations)

# Store design matrix
DM1$design %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'CMC_ACC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('CMC_ACC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'designMatrix'
synSetAnnotations(DM_OBJ, annotations = all.annotations)

# Store residual gene expression for network analysis
RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('CMC_ACC_DLPFC_netResidualExpression.tsv', 
                name = 'Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'residualGeneExpForNetAnlz'
synSetAnnotations(nEXP_OBJ, annotations = all.annotations)

# Store differential expression results
write.table(DE, file = 'CMC_ACC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_ACC_DLPFC_DiffExpression.tsv', 
                name = 'Differential Expression Results (Dx.Tissue)', 
                parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'diffExp'
synSetAnnotations(DEXP_OBJ, annotations = all.annotations)

# Store enrichment results
write.table(enrichment.results, file = 'CMC_ACC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
ENRICH_OBJ = File('CMC_ACC_DLPFC_EnrichResults.tsv', 
                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
                  parentId = CODE$properties$id)
ENRICH_OBJ = synStore(ENRICH_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'enrichResults'
synSetAnnotations(ENRICH_OBJ, annotations = all.annotations)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |

### Source Rmd
[Source markdown](`r thisFile`)

### 2018-09-26 code/Rmd/CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd
Minor sanity changes

@@
 
-knit2synapse::knitToFolderEntity(file = "./CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd",
+knit2synapse::createAndKnitToFolderEntity(file = "./CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd",
                                           parentId ="syn9846623",
-                                 entityName = 'DLPFC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model')
+                                          folderName = 'DLPFC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model')
 ```
@@
 
-synapseLogin()
+synLogin()
 

### 2018-10-04 code/Rmd/CMC_ACC_RSEM_GRCh38_CQN_ModSel.Rmd
modified first code block

--- 2018-09-26 code/Rmd/CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd
@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with RSEM-feature counts workflow with GRCh38 (DLPFC with CQN normlaisation)"
-author: "Thanneer Perumal"
+title: "Covariate analysis of CMC RNASeq data reprocessed with RSEM-feature counts workflow with GRCh38 (ACC with CQN normalisation)"
+author: "Kelsey Montgomery"
 date: "`r date()`"
@@
 
-knit2synapse::createAndKnitToFolderEntity(file = "./CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd",
+knit2synapse::createAndKnitToFolderEntity(file = "./CMC_ACC_RSEM_GRCh38_CQN_ModSel.Rmd",
                                           parentId ="syn9846623",
-                                          folderName = 'DLPFC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model')
+                                          folderName = 'ACC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model')
 ```

### 2018-10-17 code/Rmd/CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd
notes where to make new modifications

@@
 
-cl = makeCluster(detectCores()-2)
+#cl = makeCluster(detectCores()-2)
+cl = makeCluster(detectCores())
 registerDoParallel(cl)
@@
 # Github link
-thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
+thisRepo <- getRepo(repository = "kelshmo/Brain_Reg_Net", ref="branch", refName='CMC')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
@@
 # Query biomart
-Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype"),
+Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
                        filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
@@
 ### Filter genes
-Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx. Also remove genes with missing gene length and percentage GC content
+Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx and per Dx.sex. Also remove genes with missing gene length and percentage GC content
+
 ```{r cpmnormalisation}
@@
 `r kable(pct.pc)`
+
+```{r sex-specific_expr, fig.height=8, fig.width=8, results='asis', cache = FALSE}
+#filter for sex specific chromosomes here. Plot x and y expression separately for males and females 
+```
 
@@
 ```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
+#plotMDS of CQN normalized expression data plotMDS(x) where x is any data object which can be coerced to a matrix
+
 # Find principal components of expression to plot

### 2018-10-23 code/Rmd/CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd
Fix metadata

@@
 
+Notes: 
+- 3 Kleinfelter individuals
+
 ```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
@@
 library(biomaRt)
+library(ComplexHeatmap)
 
@@
 library(githubr) # get the package from devtools::install_github('brian-bot/githubr')
-
-synLogin()
 
@@
 #### Obtain effective count matrix and metadata from synapse.
+
+#readRDS(file = "logcpm.rds")
 ```{r download.data, cache=TRUE}
@@
 }
-
+downloadFile_version <- function(id , version){
+  fread(synGet(id, version = version)$path, data.table = F)
+}
 # Download reprocessed counts (DLPFC)
@@
 
-# Get RNASeq alignment qc metrics from synapse
-METADATA_ALGMETRIC_DLPFC_ID = 'syn16783867'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ALGMETRIC_DLPFC_ID
-METADATA_ALGMETRIC_DLPFC = downloadFile(METADATA_ALGMETRIC_DLPFC_ID) %>%
-  dplyr::rename(Sample_RNA_ID = ID)
-
 # Get RNASeq QCmetadata
@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
-METADATA_QC_DLPFC = downloadFile(METADATA_QC_DLPFC_ID) %>%
-  dplyr::select(Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
+METADATA_QC_DLPFC = downloadFile_version(METADATA_QC_DLPFC_ID, version = 7) %>%
+  dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
                 Sample_RNA_ID, one_of('rnaSeq_isolation:Exclude?', 'rnaSeq_isolation:RIN',
                                       'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
-                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?'))
+                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?', 'rnaSeq_report:Mapped_Reads',
+                                      'rnaSeq_report:Intragenic_Rate','rnaSeq_report:Intronic_Rate',
+                                      'rnaSeq_report:Intergenic_Rate','rnaSeq_report:Genes_Detected',
+                                      'rnaSeq_report:Expression_Profiling_Efficiency', 'rnaSeq_report:rRNA_Rate',
+                                      'rnaSeq_report:Total_Reads', 'rnaSeq_report:Alignment_Rate'))
 ```
@@
   dplyr::left_join(GENOTYPE) %>%
-  dplyr::left_join(METADATA_ALGMETRIC_DLPFC) %>%
   dplyr::rename(PMI = `PMI_(in_hours)`,
@@
                 RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
-                MappedReads = Mapped,
-                IntragenicRate = `Intragenic Rate`, IntronicRate = `Intronic Rate`, 
-                IntergenicRate = `Intergenic Rate`, GenesDetected = `Genes Detected`,
-                ExpProfEfficiency = `Expression Profiling Efficiency`, rRNARate = `rRNA rate`) %>%
-  dplyr::select(SampleID, Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, PMI, Dx, 
+                MappedReads = `rnaSeq_report:Mapped_Reads`,
+                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, IntronicRate = `rnaSeq_report:Intronic_Rate`, 
+                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, GenesDetected = `rnaSeq_report:Genes_Detected`,
+                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, rRNARate = `rnaSeq_report:rRNA_Rate`,
+                TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Alignment_Rate`) %>%
+  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, Age_of_Death, PMI, Dx, 
                 RIN, IsolationExclude, ReportExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
@@
 
-ind = METADATA$SampleID [is.na(METADATA$Gender)]
+ind = METADATA$SampleID [is.na(METADATA$Reported_Gender)]
 writeLines(paste('Following', length(ind), 'counts are missing gender'))
@@
   dplyr::left_join(backgroundGenes) %>%
-  dplyr::select(gene_id, percentage_gene_gc_content) %>%
+  dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
   unique
@@
 ```{r covariates.clustering}
-FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx", "FlowcellBatch")
+FactorCovariates <- c('Individual_ID', "Institution", "Reported_Gender", "LibraryBatch", "Dx", "FlowcellBatch")
 ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
@@
 # Gender
-# vcd::mosaic(~ Gender + Dx, data = COVARIATES)
+# vcd::mosaic(~ Reported_Gender + Dx, data = COVARIATES)
 ```
@@
 # Iteratively identify covariates in phase I
-phase1.covars = c('Institution', 'Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
+phase1.covars = c('Institution', 'Reported_Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
 vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 

### 2018-10-25 code/Rmd/CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd
sex-specific chromosome gene expression plots, MDS plot, Reported_Gender as a variable in gene filtering

@@
                                           parentId ="syn9846623",
-                                          folderName = 'DLPFC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model')
-```
-
-Notes: 
-- 3 Kleinfelter individuals
+                                          folderName = 'DLPFC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model - Kelsey')
+```
 
@@
 library(ComplexHeatmap)
+library(statmod)
 
@@
 
+synLogin()
+
 library(doParallel)
@@
 
-#cl = makeCluster(detectCores()-2)
-cl = makeCluster(detectCores())
+cl = makeCluster(detectCores()-2)
+
 registerDoParallel(cl)
@@
 #### Obtain effective count matrix and metadata from synapse.
-
-#readRDS(file = "logcpm.rds")
 ```{r download.data, cache=TRUE}
@@
 ```
+### Sex Chromosome-specific Gene Expression Patterns
+
+```{r Reported_Gender_plots}
+COUNT$gene_id = rownames(COUNT) 
+REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
+  left_join(COUNT) %>%
+  dplyr::select(-one_of("percentage_gene_gc_content")) %>%
+  filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
+  tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
+  mutate(value = log(value)) %>%
+  rename(`counts(log)`= value) %>% 
+  rename(SampleID = item) %>%
+  left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
+  rename(`Reported Gender` = Reported_Gender)
+
+my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))
+
+p = list()
+p[[1]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "X"), aes(x = `Reported Gender`, y = `counts(log)`)) + geom_boxplot()
+p[[1]] = p[[1]] + ggtitle('X') + my.theme
+p[[2]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "Y"), aes(x = `Reported Gender`, y = `counts(log)`)) + geom_boxplot()
+p[[2]] = p[[2]] + ggtitle('Y') + my.theme
+multiplot(plotlist = p, cols = 2)
+```
 
@@
 ```{r cpmnormalisation}
-genesToAnalyze = dlply(METADATA, .(Dx), .fun = function(mtd, count){
+rownames(COUNT) = COUNT$gene_id
+COUNT$gene_id = NULL
+
+genesToAnalyze = dlply(METADATA, .(Dx, Reported_Gender), .fun = function(mtd, count){
   processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
@@
 `r kable(pct.pc)`
-
-```{r sex-specific_expr, fig.height=8, fig.width=8, results='asis', cache = FALSE}
-#filter for sex specific chromosomes here. Plot x and y expression separately for males and females 
-```
 
@@
 
+```{r testCPM}
+# TEMP <- LOG.CPM
+# TEMP$gene_id = rownames(TEMP) 
+# t <- data.frame(TEMP) %>% 
+#   mutate(gene_id = rownames(TEMP))
+# REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
+#   left_join(t) %>%
+#   dplyr::select(-one_of("percentage_gene_gc_content")) %>%
+#   filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
+#   tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
+#   rename(`CPM(log)`= value) %>% 
+#   rename(SampleID = item) %>%
+#   left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
+#   rename(`Reported Gender` = Reported_Gender)
+
+# my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))
+# 
+# p = list()
+# p[[1]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "X"), aes(x = `Reported Gender`, y = `CPM(log)`)) + geom_boxplot()
+# p[[1]] = p[[1]] + ggtitle('X') + my.theme
+# p[[2]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "Y"), aes(x = `Reported Gender`, y = `CPM(log)`)) + geom_boxplot()
+# p[[2]] = p[[2]] + ggtitle('Y') + my.theme
+#multiplot(plotlist = p, cols = 2)
+```
 ### Library Normalisation
@@
 ```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
-#plotMDS of CQN normalized expression data plotMDS(x) where x is any data object which can be coerced to a matrix
-
 # Find principal components of expression to plot
@@
 
+p = list()
 # Plot first 2 PCs
@@
 
-p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
-p <- p + theme_bw() %+replace% theme(legend.position="top")
+p[[1]] <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p[[1]] <- p[[1]] + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
+p[[1]] <- p[[1]] + theme_bw() %+replace% theme(legend.position="top") + guides(colour = guide_legend(nrow = 2),
+                                                                               shape = guide_legend(nrow = 2),
+                                                                               size = guide_legend(nrow = 2))
 # p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
-p
+
+#MDS
+tmp.mds <- dist(t(tmp))
+mds <- cmdscale(tmp.mds)
+#Plot MDS
+plotdata <- data.frame(SampleID = rownames(mds),
+                       MDS1= mds[,1],
+                       MDS2= mds[,2])
+
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
+
+p[[2]] <- ggplot(plotdata, aes(MDS1,MDS2))
+p[[2]] <- p[[2]] + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
+p[[2]] <- p[[2]] + theme_bw() %+replace% theme(legend.position="top") + guides(colour = guide_legend(nrow = 2),
+                                                                               shape = guide_legend(nrow = 2),
+                                                                               size = guide_legend(nrow = 2))
+multiplot(plotlist = p, cols = 1)
 ```
@@
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx")])
+COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Reported_Gender", "LibraryBatch", "Dx")])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
@@
 # Download all related genesets from synapse
-geneset.files = synQuery('select * from file where parentId == "syn3240583"')$results
+geneset.files = synGetChildren("syn3240583", includeTypes=list("file"))$asList()
 genesets.to.test = purrr::map(geneset.files, .f = function(fileList){
-  tmp = downloadFile(fileList$file.id) %>%
+  tmp = downloadFile(fileList$id) %>%
     dlply(.(Name), .fun = function(x){
@@
 # Code
-CODE <- Folder(name = "DLPFC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
+CODE <- Folder(name = "DLPFC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model - Kelsey", parentId = parentId)
 CODE <- synStore(CODE)

### 2018-10-25 code/Rmd/CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd
possible code block for plotting XX - XY ratio, XIST/UTY difference

@@
 multiplot(plotlist = p, cols = 2)
+
+##XIST and UTY expression 
+#ENSG00000229807.10 and ENSG00000183878.15 
+FILT <- REPORTED.GENDER.COUNTS %>% 
+  filter(gene_id == "ENSG00000229807.10" | gene_id == "ENSG00000183878.15") %>% 
+  select(-one_of("chromosome_name")) %>% 
+  tidyr::spread(FILT, key = gene_id, value = `counts(log)`) %>% 
+  mutate(XIST = as.numeric(`ENSG00000229807.10`)) %>% 
+  mutate(UTY = as.numeric(`ENSG00000183878.15`))
+
+p = ggplot(FILT, aes (x= XIST, y = UTY)) 
+p = p + geom_point(aes(color=`Reported Gender`))
+
+female.Y <- filter(REPORTED.GENDER.COUNTS, chromosome_name == "Y" & `Reported Gender` == "Female") %>% 
+  filter(`counts(log)` > 1 | `counts(log)` != "-Inf")
+
+COUNTS.sum <- REPORTED.GENDER.COUNTS %>%
+  filter(`counts(log)` != "-Inf") %>% 
+  group_by(SampleID, chromosome_name, `Reported Gender`) %>% 
+  summarize(`counts(log)sum` = sum(`counts(log)`)) %>% 
+  ungroup() %>% 
+  group_by(SampleID) %>% 
+  mutate(`ratio` = `counts(log)sum`/sum(`counts(log)sum`))
 ```
@@
   dplyr::rename(baseModel = BIC)
-
+baseModelBIC
 # Iteratively identify covariates in phase I

### 2018-10-29 code/Rmd/CMC_ACC_RSEM_GRCh38_CQN_ModSel.Rmd
ready to run in ec2

@@
 ---
-title: "Covariate analysis of CMC RNASeq data reprocessed with RSEM-feature counts workflow with GRCh38 (ACC with CQN normalisation)"
-author: "Kelsey Montgomery"
+title: "Covariate analysis of CMC RNASeq data reprocessed with RSEM-feature counts workflow with GRCh38 (ACC with CQN normlaisation)"
+author: "Thanneer Perumal"
 date: "`r date()`"
@@
                                           parentId ="syn9846623",
-                                          folderName = 'ACC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model')
+                                          folderName = 'ACC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model - Kelsey')
 ```
@@
 library(biomaRt)
+library(ComplexHeatmap)
+library(statmod)
 
@@
 cl = makeCluster(detectCores()-2)
+
 registerDoParallel(cl)
@@
 activityName = 'Covariate adjustments';
-activityDescription = 'Covariate analysis of RSEM aligned effective counts with GRCh38 with CQN normalisation (DLPFC)';
-
-thisFileName <- 'CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd'
+activityDescription = 'Covariate analysis of RSEM aligned effective counts with GRCh38 with CQN normalisation (ACC)';
+
+thisFileName <- 'CMC_ACC_RSEM_GRCh38_CQN_ModSel.Rmd'
 
 # Github link
-thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
+thisRepo <- getRepo(repository = "kelshmo/Brain_Reg_Net", ref="branch", refName='CMC')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
@@
 }
-
-# Download reprocessed counts (DLPFC)
-COUNT_ID = 'syn16783526'
+downloadFile_version <- function(id , version){
+  fread(synGet(id, version = version)$path, data.table = F)
+}
+# Download reprocessed counts (ACC)
+COUNT_ID = 'syn16783564'
 ALL_USED_IDs = COUNT_ID
@@
 
-# Download gene lengths (DLPFC)
-GENE.LEN = downloadFile('syn16783530') %>%
+# Download gene lengths (ACC)
+GENE.LEN = downloadFile('syn16783573') %>%
   tidyr::gather(sampleID, Length, -gene_id, -`transcript_id(s)`) %>%
@@
 
-# Get RNASeq alignment qc metrics from synapse
-METADATA_ALGMETRIC_DLPFC_ID = 'syn16783867'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ALGMETRIC_DLPFC_ID
-METADATA_ALGMETRIC_DLPFC = downloadFile(METADATA_ALGMETRIC_DLPFC_ID) %>%
-  dplyr::rename(Sample_RNA_ID = ID)
-
 # Get RNASeq QCmetadata
-METADATA_QC_DLPFC_ID = 'syn16816488'
-ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
-METADATA_QC_DLPFC = downloadFile(METADATA_QC_DLPFC_ID) %>%
-  dplyr::select(Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
+METADATA_QC_ACC_ID = 'syn16816488' 
+ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_ACC_ID
+METADATA_QC_ACC = downloadFile_version(METADATA_QC_ACC_ID, version = 7) %>%
+  dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
                 Sample_RNA_ID, one_of('rnaSeq_isolation:Exclude?', 'rnaSeq_isolation:RIN',
                                       'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
-                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?'))
+                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?', 'rnaSeq_report:Mapped_Reads',
+                                      'rnaSeq_report:Intragenic_Rate','rnaSeq_report:Intronic_Rate',
+                                      'rnaSeq_report:Intergenic_Rate','rnaSeq_report:Genes_Detected',
+                                      'rnaSeq_report:Expression_Profiling_Efficiency', 'rnaSeq_report:rRNA_Rate',
+                                      'rnaSeq_report:Total_Reads', 'rnaSeq_report:Alignment_Rate'))
 ```
@@
 # Merge metadata
-METADATA = METADATA_QC_DLPFC %>%
+METADATA = METADATA_QC_ACC %>%
   dplyr::left_join(GENOTYPE) %>%
-  dplyr::left_join(METADATA_ALGMETRIC_DLPFC) %>%
   dplyr::rename(PMI = `PMI_(in_hours)`,
@@
   dplyr::left_join(backgroundGenes) %>% 
-  dplyr::select(gene_id, percentage_gene_gc_content) %>%
+  dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
   unique
 rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
+```
+### Sex Chromosome-specific Gene Expression Patterns
+
+```{r Reported_Gender_plots}
+COUNT$gene_id = rownames(COUNT) 
+REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
+  left_join(COUNT) %>%
+  dplyr::select(-one_of("percentage_gene_gc_content")) %>%
+  filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
+  tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
+  mutate(value = log(value)) %>%
+  rename(`counts(log)`= value) %>% 
+  rename(SampleID = item) %>%
+  left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
+  rename(`Reported Gender` = Reported_Gender)
+
+my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))
+
+p = list()
+p[[1]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "X"), aes(x = `Reported Gender`, y = `counts(log)`)) + geom_boxplot()
+p[[1]] = p[[1]] + ggtitle('X') + my.theme
+p[[2]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "Y"), aes(x = `Reported Gender`, y = `counts(log)`)) + geom_boxplot()
+p[[2]] = p[[2]] + ggtitle('Y') + my.theme
+multiplot(plotlist = p, cols = 2)
+
+##XIST and UTY expression 
+#ENSG00000229807.10 and ENSG00000183878.15 
+FILT <- REPORTED.GENDER.COUNTS %>% 
+  filter(gene_id == "ENSG00000229807.10" | gene_id == "ENSG00000183878.15") %>% 
+  select(-one_of("chromosome_name")) %>% 
+  tidyr::spread(FILT, key = gene_id, value = `counts(log)`) %>% 
+  mutate(XIST = as.numeric(`ENSG00000229807.10`)) %>% 
+  mutate(UTY = as.numeric(`ENSG00000183878.15`))
+
+p = ggplot(FILT, aes (x= XIST, y = UTY)) 
+p = p + geom_point(aes(color=`Reported Gender`))
+
+female.Y <- filter(REPORTED.GENDER.COUNTS, chromosome_name == "Y" & `Reported Gender` == "Female") %>% 
+  filter(`counts(log)` > 1 | `counts(log)` != "-Inf")
+
+COUNTS.sum <- REPORTED.GENDER.COUNTS %>%
+  filter(`counts(log)` != "-Inf") %>% 
+  group_by(SampleID, chromosome_name, `Reported Gender`) %>% 
+  summarize(`counts(log)sum` = sum(`counts(log)`)) %>% 
+  ungroup() %>% 
+  group_by(SampleID) %>% 
+  mutate(`ratio` = `counts(log)sum`/sum(`counts(log)sum`))
 ```
@@
 # Gender
-# vcd::mosaic(~ Gender + Dx, data = COVARIATES)
+# vcd::mosaic(~ Reported_Gender + Dx, data = COVARIATES)
 ```
 ### Filter genes
-Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx. Also remove genes with missing gene length and percentage GC content
+Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx and per Dx.sex. Also remove genes with missing gene length and percentage GC content
+
 ```{r cpmnormalisation}
-genesToAnalyze = dlply(METADATA, .(Dx), .fun = function(mtd, count){
+rownames(COUNT) = COUNT$gene_id
+COUNT$gene_id = NULL
+
+genesToAnalyze = dlply(METADATA, .(Dx, Reported_Gender), .fun = function(mtd, count){
   processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
@@
   dplyr::rename(baseModel = BIC)
-
+baseModelBIC
 # Iteratively identify covariates in phase I
-phase1.covars = c('Institution', 'Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
+phase1.covars = c('Institution', 'Reported_Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
 vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
@@
 # Code
-CODE <- Folder(name = "DLPFC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
+CODE <- Folder(name = "ACC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
 CODE <- synStore(CODE)

### 2018-11-02 code/Rmd/CMC_ACC_RSEM_GRCh38_CQN_ModSel.Rmd
CQN not running, NAN values produced in covariate correlation

@@
                                           parentId ="syn9846623",
-                                          folderName = 'ACC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model - Kelsey')
+                                          folderName = 'ACC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model')
 ```
@@
 ```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
-## It is assumed your working directory is where this file
+# It is assumed your working directory is where this file
 
@@
   ungroup() %>% data.frame()
-ALL_USED_IDs = c(ALL_USED_IDs, 'syn16783530')
+ALL_USED_IDs = c(ALL_USED_IDs, 'syn16783573')
 
@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_ACC_ID
-METADATA_QC_ACC = downloadFile_version(METADATA_QC_ACC_ID, version = 7) %>%
+METADATA_QC_ACC = downloadFile_version(METADATA_QC_ACC_ID, version = 8) %>% 
   dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
@@
                                       'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
-                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?', 'rnaSeq_report:Mapped_Reads',
+                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?',
+                                      'rnaSeq_report:Mapped_Reads',
                                       'rnaSeq_report:Intragenic_Rate','rnaSeq_report:Intronic_Rate',
                                       'rnaSeq_report:Intergenic_Rate','rnaSeq_report:Genes_Detected',
-                                      'rnaSeq_report:Expression_Profiling_Efficiency', 'rnaSeq_report:rRNA_Rate',
-                                      'rnaSeq_report:Total_Reads', 'rnaSeq_report:Alignment_Rate'))
+                                      'rnaSeq_report:Expression_Profiling_Efficiency',
+                                      'rnaSeq_report:rRNA_Rate',
+                                      'rnaSeq_report:Total_Reads', 'rnaSeq_report:Percent_Aligned'))
 ```
@@
                 MappedReads = `rnaSeq_report:Mapped_Reads`,
-                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, IntronicRate = `rnaSeq_report:Intronic_Rate`, 
-                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, GenesDetected = `rnaSeq_report:Genes_Detected`,
-                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, rRNARate = `rnaSeq_report:rRNA_Rate`,
-                TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Alignment_Rate`) %>%
+                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, 
+                IntronicRate = `rnaSeq_report:Intronic_Rate`, 
+                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, 
+                GenesDetected = `rnaSeq_report:Genes_Detected`,
+                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, 
+                rRNARate = `rnaSeq_report:rRNA_Rate`,
+                TotalReads = `rnaSeq_report:Total_Reads`, 
+                AlignmentRate = `rnaSeq_report:Percent_Aligned`) %>%
   dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, Age_of_Death, PMI, Dx, 
@@
 ind = METADATA$SampleID [is.na(METADATA$Ethnicity) | is.na(METADATA$Institution) | is.na(METADATA$Dx)]
-writeLines(paste('Following', length(ind), 'counts are missing any metadata'))
+writeLines(paste('Following', length(ind), 'counts are missing clinical metadata'))
 writeLines(paste(ind, collapse = ', '))
@@
 # Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "ensembl.org", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "ensembl.org", dataset = "hsapiens_gene_ensembl")#change to specific ensembl archive....host
 
@@
   filter(gene_id == "ENSG00000229807.10" | gene_id == "ENSG00000183878.15") %>% 
-  select(-one_of("chromosome_name")) %>% 
-  tidyr::spread(FILT, key = gene_id, value = `counts(log)`) %>% 
+  dplyr::select(-one_of("chromosome_name")) %>% 
+  tidyr::spread(key = gene_id, value = `counts(log)`) %>% 
   mutate(XIST = as.numeric(`ENSG00000229807.10`)) %>% 
@@
 p = p + geom_point(aes(color=`Reported Gender`)) 
-
-female.Y <- filter(REPORTED.GENDER.COUNTS, chromosome_name == "Y" & `Reported Gender` == "Female") %>% 
-  filter(`counts(log)` > 1 | `counts(log)` != "-Inf")
-
-COUNTS.sum <- REPORTED.GENDER.COUNTS %>%
-  filter(`counts(log)` != "-Inf") %>% 
-  group_by(SampleID, chromosome_name, `Reported Gender`) %>% 
-  summarize(`counts(log)sum` = sum(`counts(log)`)) %>% 
-  ungroup() %>% 
-  group_by(SampleID) %>% 
-  mutate(`ratio` = `counts(log)sum`/sum(`counts(log)sum`))
+p
 ```
@@
 Determine relationship between covariates.
+
+To summarize NAN values:
+sapply(METADATA, function(x) sum(is.nan(x)))
 ```{r covariates.clustering}
@@
 
-p = list()
 # Plot first 2 PCs
@@
 
-p[[1]] <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p[[1]] <- p[[1]] + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
-p[[1]] <- p[[1]] + theme_bw() %+replace% theme(legend.position="top") + guides(colour = guide_legend(nrow = 2),
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
+p <- p + theme_bw() %+replace% theme(legend.position="top") + guides(colour = guide_legend(nrow = 2),
                                                                      shape = guide_legend(nrow = 2),
                                                                      size = guide_legend(nrow = 2))
+p
 # p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
@@
 
-p[[2]] <- ggplot(plotdata, aes(MDS1,MDS2))
-p[[2]] <- p[[2]] + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
-p[[2]] <- p[[2]] + theme_bw() %+replace% theme(legend.position="top") + guides(colour = guide_legend(nrow = 2),
+p <- ggplot(plotdata, aes(MDS1,MDS2))
+p <- p + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
+p <- p + theme_bw() %+replace% theme(legend.position="top") + guides(colour = guide_legend(nrow = 2),
                                                                      shape = guide_legend(nrow = 2),
                                                                      size = guide_legend(nrow = 2))
-multiplot(plotlist = p, cols = 1)
+p
 ```
@@
   dplyr::rename(baseModel = BIC)
-baseModelBIC
+
 # Iteratively identify covariates in phase I

### 2018-11-06 compareCor.R
in-progress: 1. Sample to sample, and gene to gene correlations and plot the distribution of correlation coefficients in both the cases 2. L1 or L2 Norm of the difference between your residual matrix and mine 3. Compare the resulting covariates matrix between your and mine for quick metadata discrepancy.

@@ -1,0 +1,23 @@
+#sanity check
+library(synapser)
+synLogin()
+##1 compare correlation between my counts and Thanneer's counts (logCPM)
+#current counts, kelsey
+cc <- read.table(synGet("syn17015639")$path, sep = '\t', header=TRUE)
+rownames(cc) <- cc[,1]
+cc <- cc[,-1]
+#former counts, thanneer
+fc <- read.table(synGet("syn16847979")$path, sep = '\t', header=TRUE)
+rownames(fc) <- fc[,1]
+fc <- fc[,-1]
+
+intsct <- intersect(colnames(cc), colnames(fc))
+genes <- intersect(rownames(cc), rownames(fc))
+
+c <- cc[genes,intsct]
+f <- fc[genes,intsct]
+
+compare <- cor(c,f, use = "complete.obs")
+viz <- image(compare)
+
+viz

### 2018-11-07 code/Rmd/CMC_ACC_RSEM_GRCh38_CQN_ModSel.Rmd
Minor changes

@@
 output: html_document
+editor_options: 
+  chunk_output_type: console
 ---
@@
 registerDoParallel(cl)
+
+library(future)
+library(furrr)
+
+plan(multiprocess)
 
@@
 # Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "ensembl.org", dataset = "hsapiens_gene_ensembl")#change to specific ensembl archive....host
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "oct2018.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")#change to specific ensembl archive....host
 
@@
 ```{r winsorise.data}
-LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)
-
-# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
-LOG.CPM = apply(LOG.CPM, 1, function(x){
-  mn = mean(x, na.rm = T)
-  std.dev = sd(x, na.rm = T)
-
-  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
-  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
-
-  x[x < (mn-3*std.dev)] = NA
-  x[x > (mn+3*std.dev)] = NA
+LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T) %>%
+  rownameToFirstColumn('gene_id') %>%
+  tidyr::gather(sample_id, value, -gene_id) %>%
+  group_by(gene_id) %>%
+  nest() %>%
+  tibble::deframe() %>%
+  future_map(function(x){
+    mn = mean(x$value, na.rm = T)
+    std.dev = sd(x$value, na.rm = T)
+
+    x[which(x$value <= mn-3*std.dev), 'value'] = NA 
+    x[which(x$value >= mn+3*std.dev), 'value'] = NA
+
     return(x)
-}) %>% t
-
-# Back calculate counts for further analysis
+  }, .progress = T) %>%
+  bind_rows(.id = 'gene_id') %>%
+  tidyr::spread(sample_id, value) %>%
+  tibble::column_to_rownames(var = 'gene_id') %>%
+  as.data.frame()
+
+# Assign zero values to NAs
+LOG.CPM = LOG.CPM[rownames(PROCESSED_COUNTS$filteredExprMatrix$counts), colnames(PROCESSED_COUNTS$filteredExprMatrix$counts)]
 NEW.COUNTS = PROCESSED_COUNTS$filteredExprMatrix$counts
@@
 GENE.LEN = tibble::column_to_rownames(GENE.LEN, var = 'gene_id') %>% as.data.frame()
-CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
+CQN.GENE_EXPRESSION = cqn(round(NEW.COUNTS[,c(1:190, 192:476)]), 
                           x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gene_gc_content'],
@@
                           lengthMethod = "smooth", 
-                          verbose = FALSE)
+                          verbose = TRUE)
 CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

### 2018-11-07 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
minor changes

@@
 knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd",
-                                 parentId = "syn7501880", #"syn9872607",
+                                 parentId = 'syn9872607', #"syn7501880", #"syn9872607",
                                  entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - SVA')
@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = "syn7501880", #'syn9872607';
+parentId = 'syn9872607'; #"syn7501880"; #
 activityName = 'Covariate adjustments';
